### 2015-05-16 Ceres/Old_versions/ArtigoV7/CeresV7.tex
Adicionado Artigos dos irregulares e de Ceres

% mn2esample.tex
%
% v2.1 released 22nd May 2002 (G. Hutton)
%
% The mnsample.tex file has been amended to highlight
% the proper use of LaTeX2e code with the class file
% and using natbib cross-referencing. These changes
% do not reflect the original paper by A. V. Raveendran.
%
% Previous versions of this sample document were
% compatible with the LaTeX 2.09 style file mn.sty
% v1.2 released 5th September 1994 (M. Reed)
% v1.1 released 18th July 1994
% v1.0 released 28th January 1994

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage[T1]{fontenc}
\usepackage{prettyref}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage[breaklinks=true]{hyperref} %% to avoid \citeads line

% If your system does not have the AMS fonts version 2.0 installed, then
% remove the useAMS option.
%
% useAMS allows you to obtain upright Greek characters.
% e.g. \umu, \upi etc.  See the section on "Upright Greek characters" in
% this guide for further information.
%
% If you are using AMS 2.0 fonts, bold math letters/symbols are available
% at a larger range of sizes for NFSS release 1 and 2 (using \boldmath or
% preferably \bmath).
%
% The usenatbib command allows the use of Patrick Daly's natbib.sty for
% cross-referencing.
%
% If you wish to typeset the paper in Times font (if you do not have the
% PostScript Type 1 Computer Modern fonts you will need to do this to get
% smoother fonts in a PDF file) then uncomment the next line
% \usepackage{Times}

%%%%% AUTHORS - PLACE YOUR OWN MACROS HERE %%%%%

\title[Results of two stellar occultations by Ceres]{Results of two multi-chords stellar occultations by dwarf planet (1) Ceres}
\author[A. R. Gomes-J\'unior, B. L. Giacchini, F. Braga-Ribas et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail:
altair08@astro.ufrj.br},
B. L. Giacchini$^{2,3,4}$\thanks{E-mail: breno@cbpf.br},
F. Braga-Ribas$^{5, 6}$,
M. Assafin$^{1}$,\newauthor
R. Vieira-Martins$^{1,5}$
T. Blank$^{4}$, 
J. Brooks$^{4}$, 
J. Broughton$^{7}$, 
R. F. Dantowitz$^{8}$, \newauthor
D. W. Dunham$^{4}$, 
J. B. Dunham$^{4}$, 
C. K. Ellington$^{4}$, 
T. George$^{4}$, 
P. D. Maley$^{4,9}$, \newauthor
B. Timerson$^{4}$, 
N. S. Weber$^{8}$\\
$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
$^{2}$Centro Brasileiro de Pesquisas F\'isicas, Rua Dr. Xavier Sigaud 150, Rio de Janeiro  22290-180, Brazil\\
$^{3}$Se\c{c}\~ao de Oculta\c{c}\~oes/REA-Brasil, Av. Prof. Magalh\~aes Penido 403/302, Belo Horizonte 31270-700, Brazil\\
$^{4}$International Occultation Timing Association, P.O. Box 131034, Houston, TX 77219-1034, USA\\
$^{5}$Observat\'orio Nacional/MCT, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
$^{6}$Federal University of Technology - Paraná (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil\\
$^{7}$RASNZ Occultation Section, 18 Branch Cr. Reedy Creek, QLD 4227, Australia\\
$^{8}$Clay Center Observatory at Dexter Southfield, 20 Newton Street, Brookline, MA 02445, USA\\
$^{9}$NASA Johnson Space Center Astronomical Society, 4535 Cedar Ridge Trail, Houston, TX 77059 USA}

\begin{document}

\date{Accepted . Received ; in original form }

\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2015}

\maketitle

\label{firstpage}

\begin{abstract}
We report the results from two multi-chord stellar occultations by (1) Ceres that occurred in 2010 August 17, observed at Brazil, and 2013 October 25, observed at USA. Four positive chords were obtained for the 2010 occultation and nine for the 2013 occultation. An elliptic model was used to fit the apparent limb of Ceres. Two limb fitting solution was studied. The first one is a nominal solution with an indeterminate polar aspect angle. The second one, we constrain the pole coordinates as given by \cite{Drummond2014}. Assuming a Maclaurin spheroid, we determine an equatorial radius of $972 \pm 6~km$ and an apparent oblateness of $0.08 \pm 0.03$.
\end{abstract}

\begin{keywords}
minor planets, asteroids: individual (1, Ceres) - occultations - planets and satellites: fundamental parameters
\end{keywords}

\section{Introduction}

Ceres is the sole exemplar of a dwarf planet in the inner Solar System. Far from being a mere taxonomic information, this suggests the great impact its study can have on the understanding of planetary formation and evolution of the Solar System. Indeed, it was proposed that Ceres could have its origin as a transneptunian object, later scattered to the Main Belt due to the giant planets' migration predicted by the `Nice Model'. It may well had been formed close to its current location, however, even on this scenario, the dynamical history of the Solar System must have let its signatures on Ceres. Not only on the late heavy bombardment features that might exist on its surface, but also on its volatiles, which could have been transported from the outer region.

Since the 1970's it has been speculated that Ceres could contain water ice, what was recently verified \citep{Kuppers2014}. Although the water regime on this object is still unknown, some internal structure models suggest the existence of a water ice -- or even a liquid water -- layer. Yet the very question of whether Ceres suffered differentiation is open and, on the assumption of an affirmative answer, it is natural to ask if it ever had tectonics, how its geological evolution was, and if it is still active. Inarguably NASA's \textit{Dawn} mission will shed light on several open issues concerning Ceres.

Owning approximately one fifth of the whole Main Belt's mass, Ceres is expected to have an equilibrium figure, i.e. a Maclaurin or a Jacobi ellipsoid. In fact, direct observations of Ceres by means of adaptive optics confirmed it to be an oblate spheroid \citep{Drummond2014}. The precise knowledge of its shape and dimensions is important, for the models of density, internal structure and differentiation can depend on the size and oblateness of the body.

The best ground-based technique for determining shape and size of a faraway object is the study of its shadow, cast by a star during an occultation. Since the 1960's occultations have provided measurements of hundreds of asteroids, thanks partially to the proficuous professional-amateur collaboration on the field. More recently, this technique has been applied to objects of the outer Solar System and has unveiled outstanding features of distant bodies, e.g. the ring system around the Centaur (10199) Chariklo \citep{BragaRibas2014}.

The first stellar occultation by Ceres was observed in 1984 \citep{Millis1987} and led to the determination of its size to the precision of the kilometre, in a time when the uncertainties were often ten times larger. The low apparent magnitude of Ceres, if compared to most asteroids, imposes a somewhat strong constraint on the stars capable of causing a detectable magnitude drop when occulted. For instance, after the 1984 event, to our knowledge, only four stellar occultations have been observed. Two of them had only two chords each, not sufficient, thus, for providing accurate results\footnote{These events took place on 22 August 1994 and 30 October 2010.}. The two remaining events, which occurred on 17 August 2010 and 25 October 2013, are reported on the present work. The former was observed in Brazil from 4 stations, while the latter was from the United States and led to 9 chords. Throughout the paper we shall refer to these events as the `2010' and the `2013 occultation'.

Both events were foreseen by Steve Preston\footnote{Predictions are published at http://asteroidoccultation.com.} on behalf of the International Occultation Timing Association, during routine prediction of asteroidal occultations of bright stars.

This work is organized as follows. In Sections 2 and 3 we analyse, the 2010 and the 2013 events, respectively. The comparison of both results to those on the literature is carried out in Section 4.

\section{The 2010 occultation}

\subsection{Observations}

%Asteroidal occultations of bright stars are routinely predicted by Steve Preston\footnote{Predictions are published at http://asteroidoccultation.com.}, on behalf of the International Occultation Timing Association. Within this context it was identified that on 17 August 2010 Ceres would occult the star TYC 6833-163-1 (UCAC2 20678706), which has magnitude $\text{K}=9.18$ and whose ICRF/J200 position is \citep{Zacharias2004}:
%
On 17 August 2010 Ceres was predicted to occult the star TYC 6833-163-1 (UCAC2 20678706), which has magnitude $K=9.18$ and whose ICRS position is \citep{Zacharias2004}:
% UCAC4 313-111823
\begin{equation}
\left\{ 
  \begin{array}{l l}
    \alpha = 17^{\text{h}}18^{\text{m}}29^{\text{s}}.008\\
    \delta = -27\degr 26\arcmin 38\arcsec.890.
  \end{array}
\right.
\label{Eq: 2010-coord}
\end{equation}
%
% Abaixo a coordenada UCAC4 aplicada de Movimento Próprio
%
%\begin{equation}
%\left\{ 
%  \begin{array}{l l}
%    \alpha = 17^{\text{h}}18^{\text{m}}29^{\text{s}}.009\\
%    \delta = -27\degr 26\arcmin 38\arcsec.866.
%  \end{array}
%\right.
%\label{Eq: 2010-coord}
%\end{equation}
%
The shadow's path would cross South America, the Atlantic and reach the south part of Africa.

\begin{figure}
\includegraphics[scale=0.42]{figures/Ceres_2010.png} 
\caption{Post-occultation reconstruction of Ceres' shadow path on Earth for the 2010 August 17 event. The big red dot is the geocentric closest approach at 22:40:25 UT. The small red ones represent the centre of the shadow separated by one minute, shadow moves from the left to the right. Blue dots are the sites that have observed the event. As described in text, UFSC had a negative chord.
\label{Fig: Ceres-2010-map}}
\end{figure}

Observations were carried out in Brazil at five stations as displayed in Table \ref{Tab: obs-2010} and Fig. \ref{Fig: Ceres-2010-map}. The occultation was detected on four of them. UFSC had a negative chord. From the positive ones, INPE started the shots with the event already in progress and only provided star's reappearance time; the other three recorded the whole phenomenon.

A remarkable circumstance of this event was the low velocity of Ceres: only $3.9~km.s^{-1}$ in the plane of the sky. Therefore, even exposures of a few seconds would translate in a relevant spatial resolution.

[Falar sobre a quest\~ao do tempo em cada corda. A leitura do header era no segundo inteiro? Foi analisada a cad\^encia das imagens? Qual a precis\~ao?]

[Durante o per\'iodo de observac\~ao Ceres rodou apenas 3.1$\degr$. Mas isso n\~ao \'e relevante no caso de Maclaurin...]

\begin{table*}
 \centering
 \begin{minipage}{140mm}
  \caption{Circumstances of observation for all observing stations of the 2010 event. \label{Tab: obs-2010}}
  \begin{tabular}{@{}lccccc}
  \hline
          & Longitude & Telescope & Exposure & Result &   \\
     Site & Latitude  & Aperture  & Cycle time & Ingress & Observer\\          
          & Height    & Camera    &        & Egress    & \\          
\hline
 Belo Horizonte & 43$\degr$59'51.1'' W & LX200 & 5 s & Positive & C. Jacques  \\
 CEAMIG-REA &19$\degr$49'49.0'' S & 31 cm & 12 s & 22:39:03.9 $\pm$ 0.6 &  E. Pimentel \\
            & 825 m                &       &     & 22:40:20 $\pm$ 5 &   \\
 & & & & & \\
 Pico dos Dias    & 45$\degr$34'45.1'' W &       & 1 s & Positive & J. I. B. Camargo \\
 OPD/LNA    &22$\degr$32'03.7'' S & 60 cm & 2 s & 22:37:30.3 $\pm$ 0.6 &  G. B. Rossi \\
            & 1864 m               &       &     & 22:41:55.3 $\pm$ 0.7 &              \\
 & & & & & \\
 S\~ao Jos\'e dos       & 45$\degr$51'44.0'' W &  C11  & 2 s & Egress only & A. C. Milone\\
 Campos       &23$\degr$12'33.0'' S & 28 cm & 5 s & Start: 22:39:44 & T. Maldonado\\
 INPE      & 975 m                & SBIG ST7 &     & 22:42:03.0 $\pm$ 0.2 & M. Okada    \\
 & & & & & \\
 Ponta Grossa       & 50$\degr$05'56.0'' W & RCX 400 &30 s & Positive & M. Emilio   \\
 UEPG       &25$\degr$05'22.2'' S & 40 cm & 52 s & 22:37:17 $\pm$ 13 & L. Mehret   \\
            & 910 m                & SBIG STL6E &     & 22:39:56 $\pm$ 13 & \\
 & & & & & \\
 Florian\'opolis       & 48$\degr$31'20.5'' W &       & 3 s & No occultation  & W. Schoenell\\
 UFSC       &27$\degr$36'12.3'' S & 28 cm & 6 s&  & A. J. T. Mello\\
            & 20 m                 &       &  &         & F. R. Herpich \\
\hline
\end{tabular}
\end{minipage}
\end{table*}

\subsection{Light curves and timing \label{Sec: light-curve-2010}}

The flux of the star in the five occultation chords was obtained from the FITS images with the Platform for Reduction of Astronomical Images Automatically (PRAIA) \citep{2011gfun.conf...85A}. The light curves were normalized to an unocculted star flux. Additionally, with exception of INPE, they were normalized by fitting a polynomial curve (of first or second order) outside the flux drop, so that the flux ratio was set to 1 outside the occultation.

The start and end times of the occultation were obtained for each light curve by fitting a sharp edge occultation model. This model is convolved by Fresnel diffraction, the CCD bandwidth, the stellar aparent diameter, and the finite integration time; see \cite{Widemann2009}.

The Fresnel scale ($F = \sqrt{\lambda D/2}$) for the present Ceres' geocentric distance $D = 2.29 \text{ au} = 3.42 \times 10^{8}$~km is 0.3~km for a typical wavelength of $\lambda = 0.65$~$\micron$. The star diameter is estimated using the formulae of \cite{vanBelle1999}. Its $B$, $V$, and $K$ apparent magnitudes are 12.00, 11.78 and 9.18, respectively, in the Tycho catalog \citep{Hog2000}. This yields a diameter of about 0.11~km projected at Ceres' distance. The smallest integration time used in the positive observations was 1.0~s, which translates to almost 3.9~km in the celestial plane. Therefore, the occultation light curves are largely dominated by the integration times, not by Fresnel diffraction or star diameter.

The occultation fits consist in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011}. The free parameter to adjust is the ingress (disappearance) or egress (reappearance) time $t_{occ}$, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$. The best fittings to the 2010 occultation light curves are shown in Fig. \ref{Fig: Ceres-2010-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2010}.

\begin{figure}
\includegraphics[scale=0.58]{figures/Ceres_2010_fluxratio.png} 
\caption{The four occultation light curves normalized and vertically shifted by a factor of 0.2 for better viewing. The solid black lines are the best fit of the square-well model to the data. Red lines are the square-well model convoluted with the Fresnel diffraction, the star diameter, and the applied exposure time. The mid-times of the occultations do not coincide due to the propagation delays of the shadow due to the distinct longitude of the sites. Exposures at INPE started after the immersion, as explained in the text. \label{Fig: Ceres-2010-curves}}
\end{figure}

\subsection{Limb fitting methodology} \label{Sec: limbfittingmethod}

The methodology used to analyse Ceres' profile from the observations is the same described by \cite{Sicardy2011} and \cite{BragaRibas2013}. To each combination of site position and recorded ingress/egress time, together with star coordinates and Ceres' ephemeris, corresponds a point $(f_{i,obs},g_{i,obs})$ on the plane of the sky. The collection of those points ideally determines the apparent limb of Ceres.

We adopt an elliptic model for the limb profile, resulting from the projection of an oblate spheroid onto the sky plane. This model is supported by the work of \cite{Drummond2014}, by means of direct imaging of Ceres. Hence, we have $N=7$ chord extremities to adjust the $M=5$ parameters which define an ellipse: apparent semimajor and semiminor axis ($a^\prime$ and $b^\prime$, respectively), position angle $P$ of its semiminor axis and the position $(f_c,g_c)$ of its centre with respect to the occulted star. The coordinates $f_{c}$ and $g_{c}$, in kilometres, are calculated using the JPL\#26 Ceres' ephemeris and the occulted star's position. They are positive toward the local celestial east and north, respectively. The position angle $P$ is counted positively from the direction of local celestial north to celestial east. The apparent oblateness can be defined by $\epsilon^\prime = 1 - (b^\prime/a^\prime)$. We define the number of degrees of freedom of the problem as $\mathcal{N} \equiv N - M$.

%We adopt an elliptic model for the limb profile, resulting from the projection of an oblate spheroid onto the sky plane. This model is supported by the work of \cite{Drummond2014}, by means of direct imaging of Ceres. Hence, we have $N=7$ chord extremities to adjust the $M=5$ parameters which define an ellipse: apparent semimajor and semiminor axis -- $a^\prime$ and $b^\prime$, respectively --, position angle $P$ of its semiminor axis and the position $(f_c,g_c)$ of its centre with respect to the occulted star\footnote{The coordinates $f_{c}$ and $g_{c}$, in kilometres, are calculated using the JPL\#33 Ceres' ephemeris and the occulted star's position. They are positive toward the local celestial east and north, respectively. The position angle $P$ is counted positively from the direction of local celestial north to celestial east.}. The first two parameters can be interchanged with the oblateness and the equivalent radius, defined, on this order, by $\epsilon^\prime = 1 - (b^\prime/a^\prime)$ and $R_{equiv} \equiv \sqrt{a^\prime b^\prime} = a^\prime\sqrt{1-\epsilon^\prime}$. We define the number of degrees of freedom of the problem as $\mathcal{N} \equiv N - M$.

The apparent oblateness is related to the true one $\epsilon$ through
%
\begin{equation}
\epsilon^\prime = 1 - \sqrt{\cos^2(\zeta) + (1-\epsilon)^2\sin^2(\zeta)},
\label{Eq: TrueObla}
\end{equation}
%
where $\zeta$ is the polar aspect angle, encompassed by the (true) polar semiminor axis and the line of sight. Of course, the apparent semimajor axis $a^\prime$ equals the equatorial radius $R_{equa}$ of the ellipsoid.

%Correspondence between true and apparent figures depends on the polar aspect angle $\zeta$, encompassed by the (true) polar $b$-axis and the line of sight. The apparent oblateness is related to the true one, $\epsilon$, through

The search for the best-fitting ellipse is carried out by minimizing the reduced $\chi^2$ function:
%
\begin{equation}
\chi^2_r = \frac{1}{\mathcal{N}} \sum_{i=1}^N \frac{1}{\sigma_i^2}  \left[ (f_{i,obs}-f_{i,cal})^2 + (g_{i,obs}-g_{i,cal})^2\right].
\end{equation}
%
Here $\sigma_i$ is the radial uncertainty of the $i$th chord extremity, obtained by multiplying the time uncertainty reported in Table~1 by the normal velocity of the star (with respect to the limb model). $f_{i,cal}$ and $g_{i,cal}$ are the coordinates of the point on the model ellipse, where a line that goes from the centre of the model to the data point $(f_{i,obs},g_{i,obs})$ intersects the ellipse.%:
%
%\begin{equation}
%f_{i,cal} \equiv f_{i,obs} + \rho \cos(\theta), \quad g_{i,cal} \equiv %g_{i,obs} + \rho \sin(\theta),
%\end{equation}
%
%with:
%\begin{equation}
%\theta \equiv \arctan\left( \frac{g_{i,obs}-g_{c}}{f_{i,obs}-f_{c}} %\right) ,
%\end{equation}
%
%\begin{equation}
%\rho \equiv a^\prime b^\prime / \sqrt{[a^\prime\sin(\theta+P)]^2 +[b^\prime\cos(\theta+P)]^2 }.
%\end{equation}

Evaluation of the reduced $\chi^2$ function over the $\lbrace R_{equa}, \epsilon^\prime, P, f_c, g_c \rbrace$--space produces a map related to the likelihood of a solution. For instance, the 1--sigma reported time uncertainties correspond to the subset of the parameter space for which $\chi^2_r$ lies between ${\chi^2_r}_{min}$ and ${\chi^2_r}_{min} + 1$. This allows the determination of the error bars of the physical parameters. It is worthwhile to mention that we consider all the uncertainty as due to timing issues.

\subsection{Limb fitting solutions}\label{Sec: limbfitting-2010}

Two possible solutions were considered for the limb fitting. The first is the nominal one, which consists of determining the five parameters that characterize an ellipse from the seven observed contacts. As we shortly show, it led to a rather large error bar on the position angle. Furthermore, the nominal solution solely is not capable of returning the true oblateness, which can be evaluated through equation \ref{Eq: TrueObla}, provided that the angle $\zeta$ is known.% or constrained by some physical hypothesis.

From observations by adaptive optics spanning a 10-year period, \cite{Drummond2014} determined the position of Ceres' polar axis within a range of only 3$\degr$:
\begin{equation}
%\left\{ 
%\begin{array}{l}
\alpha = (287 \pm 3) \degr, \quad %\\
\delta = (+64 \pm 3) \degr,
%\end{array} \right .
\label{Eq: Pole}
\end{equation}
%
in equatorial J2000 coordinates. Together with Ceres' ephemeris at the moment of the occultation, this corresponds to the polar aspect angle $\zeta = 86.1\degr$, which is very close to an equator-on geometry. Hence, we expect true figures to be similar to apparent ones.

The knowledge of Ceres' pole not only allows the determination of its polar aspect, it suffices to set its position angle. Therefore Eq. \ref{Eq: Pole} may act as a constraint for $P$, and a second solution can be obtained by probing the parameter space with the restriction that the position angle is confined to the range that follows from Eq. \ref{Eq: Pole}. We call this the ``pole-constrained solution''.

\subsubsection{Nominal solution}

With the seven observed contacts it is possible to adjust the five parameters which define an ellipse. For the best-fitting solution we find $\chi^2_{r,min} = 0.24$, which could be interpreted as a slightly overestimation of the error bars with regard to the good quality of the fit. However, inasmuch as the problem has only two degrees of freedom, it is far from the statistical realm and low $\chi^2_{r,min}$ may eventually happen.

The resulting values of equatorial diameter, oblateness, position angle and centre coordinates are presented in the second column of Table~2.%, and the corresponding ellipse is plotted with the observed chords in Fig.~2.
% The equivalent radius of the best-fitting solution is $R_{equiv} = 469 \pm 10$ km.

Already mentioned, the parameter with the largest uncertainty is the position angle: spanning on a 20$\degr$ interval, its determination has a relative precision worse than 10$\%$. Clearly, the coordinates of Ceres' pole (Eq. \ref{Eq: Pole}) can impose a strong constraint on the position angle, as the next solution shows.

Finally, the correction to the oblateness due to Ceres' polar aspect angle lies within the 1-sigma error bar and has no statistical relevance; hence $\epsilon = 0.08 \pm 0.03$.

\subsubsection{Pole-constrained solution}

At the moment of the occultation, the coordinates (Eq. \ref{Eq: Pole}) of Ceres' rotational pole correspond to the position angle $P = (12 \pm 3)\degr$. Exploration of the parameter space, restricted to ellipses whose position angle lies on this range, results in the pole-constrained solution. The related physical parameters are displayed on the third column of Table \ref{Tab: results}, while the best-fitting solution is depicted in Fig.\ref{Fig:Ceres-2010-body}.

We notice that the constraint corresponds to the upper limit of nominal solution's 1-sigma error bar for $P$. On the other hand, it selects the smallest values of equatorial radius, improving its determination by a factor of about 2. Notwithstanding, oblateness' figures remain the same.

\begin{table*}
 \centering
 \begin{minipage}{140mm}
  \caption{Results of limb fitting to the data of the 2010 and 2013 events.\label{Tab: results}}
  \begin{tabular}{@{}lcccc}
  \hline
     Solution & 2010/Nominal & 2010/Pole-constrained & 2013/Nominal & 2013/Pole-constrained \\
\hline
Equatorial diameter (km) & 982 $\pm$ 14 & 972 $\pm$ 6  & 971 $\pm$ 7  & 971 $\pm$ 7\\
%Equatorial diameter (km) & 491 $\pm$ 7  & 486 $\pm$ 3  & 485.5 $\pm$ 3.5 & 485.5 $\pm$ 3.5\\
True oblateness        & 0.08 $\pm$ 0.03 & 0.08 $\pm$ 0.03 & 0.08 $\pm$ 0.04 & 0.08 $\pm$ 0.04\\
Position angle (deg)   & 5 $\pm$ 10    & 12 $\pm$ 3 (*)& 22 $\pm$ 5    & 25 $\pm$ 3 (*)\\
$f_c$ (km)             & 133 $\pm$ 9   & 138 $\pm$ 5   & 77 $\pm$ 6    & 78 $\pm$ 6\\
$g_c$ (km)             & -17 $\pm$ 15  & -11 $\pm$ 11  & 13 $\pm$ 16   & 13 $\pm$ 16\\
$\chi^2_{r,min}$       & 0.24          &  0.42         & 1.27          & 1.27\\
\hline
\end{tabular}
\textbf{Notes.} Error bars are at 1-sigma level.
(*) Position angle derived from Ceres' rotational pole coordinates determined by \cite{Drummond2014}.
\end{minipage}
\end{table*}

\begin{figure}
\includegraphics[scale=0.36]{figures/Ceres_2010_body.pdf} 
\caption{The best elliptical fit for the occultation chords for the event of 2010 using the times from Table~\ref{Tab: obs-2010} and the pole-constrained solution. \label{Fig:Ceres-2010-body}}
\end{figure}

\section{The 2013 occultation}

\subsection{Observations}\label{Sec: observation-2013}

The event which took place on 25 October 2013 involved the star TYC 865-911-1 (UCAC4 496-058191), of magnitude $K = 8.76$. Based on UCAC4 \citep{Zacharias2013}, the ICRS position to the date of occultation corrected of proper motion is :
%
\begin{equation}
\left\{ 
  \begin{array}{l l}
    \alpha = 11^{h}57^{m}52^{s}.7641\\
    \delta = +09\degr 07\arcmin 49\arcsec.835
  \end{array}
\right.
\end{equation}
%
The occultation could only be visible in the United States, before dawn. The shadow would then cross the Atlantic and Africa, but during daylight time.

\begin{figure*}
\begin{centering}
\subfigure{\includegraphics[scale=0.42]{figures/Ceres_2013.png}}
\hspace{5mm}
\subfigure{\includegraphics[scale=0.42]{figures/Ceres_2013-zoom.png}}
\caption{Post-occultation reconstruction of Ceres' shadow path on Earth for the 2013 October 25 event at the east coast of USA. Left panel: Overview of the occultation. The big red dot is the geocentric closest approach at 09:43:03 UT and the small ones represent the centre of the shadow separated by 30 seconds where the shadow moves from the left to the right. The blue dots are the sites that have observed the event. Right Panel: Upper view of the occultation over the sites that observed the event. Red points are the centre of the shadow separated by 5 seconds. \label{Fig: Ceres-2013-map}}
\end{centering}
\end{figure*}

Nine positive chords were obtained by the variety of equipment listed in Table \ref{Tab: obs-2013}. Each station was equipped with a video camera (whose readout time may be considered negligible). This is of particular importance in this event, since Ceres' shadow had the much faster speed of 42.6~km~s$^{-1}$, if compared to the 2010 event. All the observing sites are in the United States, as depicted in Fig.~\ref{Fig: Ceres-2013-map}.

Three different timing synchronization procedures were adopted among the set of observing stations. At Greenbelt and Owings the 1PPS signal of a GPS unit was used to calibrate time stamps which were inserted at each frame of the video. Time extraction is thus straightforward, after taking camera delays into account. On the other hand, at Brookline the clock would be synchronized by an internet server. A lack of connection, however, resulted in spurious times. In fact, comparison between the times obtained at this station and the others suggests that the former have a delay of about 64 s. Therefore, we do not use Brookline's absolute times in the analysis that follows. Finally, at the remaining six stations the videos were recorded by camcorders on digital tapes. The timing method consisted in the comparison of the camcorder internal clock to a 1PPS GPS signal, before and after the recording of the occultation. Absolute timing errors of this procedure are expected to be inferior to 0.1 s.

A remarkable observing circumstance of this event is the low altitude of Ceres with respect to the horizon. Precisely, its altitude at moment of the occultation ranged between $20\degr$ (Hampton) and $15\degr$ (Winchester). Strong scintillation is expected in such a scenario which, combined to short integration times and the low magnitude drop of the event, resulted in rather noisy light curves and thus larger uncertainty in the time of the contacts, as it is shown in the next section.

\begin{table*}
 \centering
 \begin{minipage}{140mm}
  \caption{Circumstances of observation for the observing stations of the 2013 event.\label{Tab: obs-2013}}
  \begin{tabular}{@{}lccccc}
  \hline
          & Longitude & Telescope: & Camera  & Result &   \\
     Site & Latitude  & Aperture  & Cycle time & Ingress & Observer\\          
          & Height    & f-ratio    &           & Egress    & \\          
\hline
 Hampton & 70$\degr$48'59.7'' W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
  &42$\degr$53'52.8'' N & f/5 & 0.033 s     & 09:40:46.9 $\pm$ 0.1 &   \\
            & 7 m       &  &     & 09:40:57.26  $\pm$ 0.08 &   \\
 & & & & & \\
 Topsfield & 70$\degr$55'16.6'' W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
  &42$\degr$37'55.9'' N & f/5      & 0.033 s   & 09:40:45.4 $\pm$ 0.1 &   \\
            & 45 m      &  &     & 09:40:58.0  $\pm$ 0.1 &   \\
 & & & & & \\
 Brookline & 71$\degr$08'14.5'' W & 64 cm & Infinity2-1R & Positive & N. Weber \\
  &42$\degr$18'27.4'' N & f/9.6 & 0.015 s     & Duration only &  R. Dantowitz  \\
            & 109 m     &       &     & (14.93 $\pm$ 0.01) s &   \\
 & & & & & \\
 Winchester & 78$\degr$14'39.6'' W & 36 cm & PC164C & Positive & J. Brooks \\
  &39$\degr$16'21.5'' N & f/5 & 0.033 s    & 09:40:33.26 $\pm$ 0.08 &   \\
            & 211 m     &  &     & 09:40:55.86  $\pm$ 0.09 &   \\
 & & & & & \\
 Greenbelt & 76$\degr$52'09.4'' W & 12 cm & PC164C-EX2 & Positive & J. Dunham \\
  &38$\degr$59'12.1'' N & f/2.5 & 0.033 s    & 09:40:33.6 $\pm$ 0.1 & D. Dunham  \\
            & 52 m      &   &     & 09:40:56.4  $\pm$ 0.1 &   \\
 & & & & & \\
 Alexandria & 77$\degr$02'28.3'' W & 7 cm & Watec120N & Positive & P. Maley \\
  &38$\degr$49'19.1'' N & f/10 &  0.067 s   & 09:40:33.3 $\pm$ 0.1 &   \\
            & 8 m       &   &     & 09:40:56.1  $\pm$ 0.1 &   \\
 & & & & & \\
 Owings & 76$\degr$38'06.3'' W & 25 cm & PC164C  & Positive & C. Ellington \\
  &38$\degr$41'26.5'' N & f/3.3 &  0.033 s & 09:40:34.27 $\pm$ 0.05 &   \\
            & 38 m                &       &     & 09:40:56.0  $\pm$ 0.2 &   \\
 & & & & & \\
 Mechanicsville & 77$\degr$23'06.7'' W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
  &37$\degr$41'26.1'' N & f/2.5 & 0.033 s    & 09:40:33.0 $\pm$ 0.1 &   \\
            & 60 m      &   &     & 09:40:54.8  $\pm$ 0.1 &   \\
 & & & & & \\
 Varina & 77$\degr$19'49.3'' W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
  &37$\degr$25'58.6'' N & f/2.5 &  0.033 s   & 09:40:32.4 $\pm$ 0.3 &   \\
            & 19 m      &   &     & 09:40:53.1  $\pm$ 0.2 &   \\

\hline
\end{tabular}
\end{minipage}
\end{table*}

\subsection{Light curves and timing}

All videos were converted to FITS images and the photometry of the target was obtained via PRAIA \citep{2011gfun.conf...85A}. The light curves were normalized to the flux of a reference star when such was available on the field.

To reduce the noise, the data were binned by groups of five images -- with the exception of  Greenbelt, which was averaged in sets of ten. This procedure enlarges the effective integration times reported in Table~3 by a factor of 5 (or 10). As in the 2010 event, an additional normalization by a polynomial curve was applied.

The start and end times of the occultation were obtained by the same procedure explained in Section~2.2. The Fresnel scale of 0.4~km for this event was calculated using Ceres' geocentric distance $D = 3.28 \text{ au} = 4.91 \times 10^{8}$~km, for a typical wavelength of $\lambda = 0.65$~$\micron$. Considering the apparent magnitudes of the star, $B = 10.47$, $V=9.97$, and $K=8.76$ \citep{Hog2000}, we estimate its diameter projected at Ceres' distance as 0.16~km. Thence, since the typical effective integration time used (0.17~s) translates to about 7 km in the celestial plane, theoretical occultation light curves are largely dominated by the integration times, as for the 2010 event.

The best-fittings to the occultation light curves are shown in Fig.~\ref{Fig: Ceres-2013-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2013}.

\begin{figure}
\includegraphics[scale=0.58]{figures/Ceres_2013_fluxratio} 
\caption{The nine occultation light curves normalized and vertically shifted by a factor of 1.0 for better viewing. The solid black lines are the best fit of the square-well model to the data. The red lines are the square-well model convoluted with the Fresnel diffraction, the star diameter, and the applied exposure time. The mid-times of the occultations do not coincide due to the propagation delays of the shadow due to the distinct longitude of the sites. The light curve of Brookline is shifted by -64 s as explained in the text. \label{Fig: Ceres-2013-curves}}
\end{figure}

\subsection{Limb fitting solutions}

Elliptic limb profiles were adjusted to the eight\footnote{Brookline's chord was not used for limb fitting since it actually measured duration instead of absolute times.} chords by the same procedure described in Section \ref{Sec: limbfittingmethod}\footnote{For this event, the JPL\#33 Ceres' ephemeris was used to obtain the $f_{c}$ and $g_{c}$ coordinates.}. This yielded $\chi^2_{r,min} = 13$, suggesting that an elliptic model is not satisfactory to the data. Indeed, a quick glance at a plot of the observed chords, such as in Fig. \ref{Fig: Ceres-2013-body}, shows that the Varina chord seems to be somewhat advanced with respect to the others. Taking into account that in this station time stamps were not inserted on the video frames, it is always possible to attribute this advance to an eventual problem on the correspondence between camcorder's and GPS' times. This could be caused, for example, if the camcorder delayed to start the recording and, since this was an unattended pre-pointed station, this fact would not be noticed by the analysis of the video itself.

The immersion recorded at Owings also seems to be shifted (delayed) with respect to the nearby chords (see Fig. \ref{Fig: Ceres-2013-body}). This chord, actually, has roughly the same length of Mechanicsville's, despite the fact that they are separated by circa $100~km$. Differently from Varina, though, this station had time stamps inserted in each frame of the video, which makes it more unlike to justify an eventual timing issue. Another possible explanation for the times of this chord is the determination of the events in the light curve analysis, which could have been affected by noise. Finally, the delay during the ingress could be caused by a relief feature in Ceres; we shall soon return to this hypothesis.

%In addition to a time basis problem, other possible explanations to the times of this chord are the determination of the events in the light curve analysis, due to noise; and finally that the delay during the ingress could be caused by a relief feature in Ceres.

In a second limb fitting, thus, we did not consider Brookline, Varina and Owings chords. The adjust of the five parameters which define an ellipse to the 12 contacts then resulted in $\chi^2_{r,min} = 1.27$, indicating that the fitting is in good agreement with the observed data within the error bars. This is the solution depicted in Fig.~\ref{Fig: Ceres-2013-body}, where we see that the chord length measured in Brookline is compatible to the model. The associated physical parameters are presented in Table~2 as the nominal solution.

For this event the polar aspect angle is $\zeta = 90.7\degr$, which makes true oblateness be equal to the apparent one, within the error bars.

It is worth noting that the position angle of this nominal solution is better constrained than the respective of the 2010 occultation. Actually, the uncertainty of the former is of $5\degr$ in contrast to $10\degr$ of the latter. This suggests that the pole-constrained solution obtained via the position of Ceres' pole (Eq. \ref{Eq: Pole}) would not be significantly different of the nominal one.

This assumption is confirmed when we carry out the limb fitting with the constraint $P = (25 \pm 3)\degr$, the position angle at the moment of the occultation which follows from (Eq. \ref{Eq: Pole}). The physical parameters related to this pole-constrained solution are presented in the last column of Table~2, and are essentially the same of the nominal solution.

We close this section returning to the hypothesis that the delay observed in the immersion at Owings could be associated to a limb topography feature. The recorded contact would then correspond to a negative elevation of $31 \pm 4$ km with respect to the best-fitting ellipse. Theoretical models, however, predict that relief in Ceres should be no higher than about 10--20 km \citep{Johnson1973}, while published observational data sets the bound of 18 km \citep{Carry2008}. More recently, images by the probe \textit{Dawn} also reveal an even smoother surface. The association of Owings first contact to a relief must, therefore, be ruled out.

\begin{figure}
\includegraphics[scale=0.36]{figures/Ceres_2013_body.pdf}
\caption{The best elliptical fit for the occultation chords for the event of 2013 using timing from Table \ref{Tab: obs-2013} and the nominal solution. The chord of Brookline if shifted by -64s as explained in the text\label{Fig: Ceres-2013-body}}
\end{figure}

\section{Discussion}

A quick glance at Table~2 shows there is an overall agreement between the physical parameters derived from both occultations, specially in the equatorial diameter. The differences of the solutions occur basically on the size of their error bars, and can be justified by the particularities of each set of data.

The 2010 event, for example, had only seven contacts; none the less, they were well distributed over Ceres' disc (see Fig. \ref{Fig:Ceres-2010-body}) acting as a constraint to is shape. On the other hand, the 2013 event had five more contacts, but they were concentrated in certain regions of the body. In particular, the absence of chords closer to Ceres' southern limb made its oblateness be worse determined here than in the 2010 event.

However, even our best measurement for the oblateness, $\epsilon=0.08 \pm 0.03$, has still a big uncertainty with regard to other figures published in the literature, as Table~4 shows. A larger number of uniformly spaced chords would be necessary to offer a best constraint to the oblateness.

The few chords of the 2010 occultation could themselves only constraint the position angle of the object to a $20\degr$ interval. This range was reduced by a factor of two in the 2013 event, approaching --and verifying-- the result predicted by the work of \cite{Drummond2014}. As it was shown, using the coordinates of Ceres' polar axis to limit the position angle was not an efficient procedure in the case of this occultation, in the sense that the aforementioned limitation had not correspond to a stronger constraint on the other parameters.

On the other hand, constraining the position angle on the 2010 occultation was proved to reduce the error bars of the other parameters (disregarding oblateness). Moreover, this procedure resulted in an excellent agreement between the equatorial radius figures of both events.

The 2013 occultation, therefore, offers not only an independent verification of the figures resulted from the 2010 event, but also validates the procedure carried out there --and which led to the best-constrained parameters in this work.

Comparison of Ceres' equatorial diameter as measured by different techniques is carried out in Table \ref{Tab: Ceres-final}. We note an overall agreement between our result to those obtained via direct imaging by the Hubble Space Telescope (HST) \citep{Thomas2005}, the Keck Observatory and the ESO VLT \citep{Drummond2014}. The smaller figure reported by \cite{Carry2008} may be justified by the fact that in this study the effects of limb-darking were not taken into account, as pointed out by \cite{Drummond2014}.

As mentioned in the Introduction, the 1984 event \citep{Millis1987} is the only occultation data to which we can compare our result; and it is interesting to notice that the diameters measured in each case do not agree, being our result larger than the other. It is difficult to state with sure the reasons for this divergence since we had no access to the original 1987 data or reduction methodology. We note, however, that in this occultation a variation of 0.5 s in the contact times correspond to almost 7~km on Ceres' limb. Therefore, in order to properly compare their result to ours it would be necessary to reduce their light curves with the same methodology used in the present work.

The geocentric positions for Ceres at the time of the occultations could also be obtained with great accuracy. These are given in Eqs. \ref{Eq: Ceres-2010_pos} and \ref{Eq: Ceres-2013_pos}.

\begin{equation}
2010 ~ Aug ~ 17
\left\{ 
  \begin{array}{l l}
    Time = 22:40:25\\
    \alpha = 17^{\text{h}}18^{\text{m}}29^{\text{s}}.008\\
    \delta = -27\degr 26\arcmin 38\arcsec.890.
  \end{array}
\right.
\label{Eq: Ceres-2010_pos}
\end{equation}

\begin{equation}
2013 ~ Oct ~ 25
\left\{ 
  \begin{array}{l l}
    Time = 09:43:03\\
    \alpha = 11^{h}57^{m}52^{s}.7641\\
    \delta = +09\degr 07\arcmin 49\arcsec.835
  \end{array}
\right.
\label{Eq: Ceres-2013_pos}
\end{equation}

\begin{table}
  \caption{Ceres' equatorial diameter and oblateness \label{Tab: Ceres-final}}
  \begin{centering}
  \begin{tabular}{@{}llcc}
  \hline
     Eq. diameter (km) & Oblateness & Method & Ref. \\
\hline
$972 \pm 6$  & $0.08  \pm 0.03$  & Occultation & 1\\
$967 \pm 10$ & $0.078 \pm 0.015$ & Keck+VTL    & 2 \\
$959 \pm 5$  & $0.074 \pm 0.007$ & Keck        & 3\\
$975 \pm 4$  & $0.068 \pm 0.005$ & HST         & 4\\
$959 \pm 5$  & $0.05  \pm 0.01$  & Occultation & 5\\
\hline
\end{tabular}
\par\end{centering}
\textbf{References.} 1: Present work. 2: \cite{Drummond2014}. 3: \cite{Carry2008}. 4: \cite{Thomas2005}. 5: \cite{Millis1987}.
\end{table}

\section*{Acknowledgements}

ARGJ thanks the financial support of CAPES. BLG thanks CNPq. F.B.R. acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014. We also acknowledge Steve Preston for the predictions of the occultations.

\begin{thebibliography}{99}

\bibitem[\protect\citeauthoryear{Assafin et al}{2011}] {2011gfun.conf...85A} Assafin M. et al., 2011, Gaia follow-up network for the solar system objects : Gaia FUN-SSO workshop proceedings, held at IMCCE -Paris Observatory, France, November 29 - December 1, 2010. ISBN 2-910015-63-7

%@INPROCEEDINGS{2011gfun.conf...85A,
%   author = {{Assafin}, M. and {Vieira Martins}, R. and {Camargo}, J.~I.~B. and 
%	{Andrei}, A.~H. and {Da Silva Neto}, D.~N. and {Braga-Ribas}, F.
%	},
%    title = "{PRAIA - Platform for Reduction of Astronomical Images Automatically}",
%booktitle = {Gaia follow-up network for the solar system objects : Gaia FUN-SSO workshop proceedings, held at IMCCE -Paris Observatory, France, November 29 - December 1, 2010. ISBN 2-910015-63-7},
%     year = 2011,
%   editor = {{Tanga}, P. and {Thuillot}, W.},
%    month = jun,
%    pages = {85-88},
%   adsurl = {http://adsabs.harvard.edu/abs/2011gfun.conf...85A},
%  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
%}

\bibitem[\protect\citeauthoryear{Braga-Ribas et al.}{2013}]{BragaRibas2013} Braga-Ribas F. et al., 2013,
ApJ, 773, 26

\bibitem[\protect\citeauthoryear{Braga-Ribas et al.}{2014}]{BragaRibas2014} Braga-Ribas F. et al., 2014,
Nature, 508, 72

\bibitem[\protect\citeauthoryear{Drummond et al.}{2014}]{Drummond2014} Drummond J.D. et al., 2014,
Icarus, 236, 28

\bibitem[\protect\citeauthoryear{Carry et al.}{2008}] {Carry2008} Carry B. et al., 2008,
A\&A, 478, 235

\bibitem[\protect\citeauthoryear{Hog et al.}{2000}] {Hog2000} Hog E. et al., 2000,
A\&A, 355, L27

\bibitem[\protect\citeauthoryear{Johnson \& McGetchin}{1973}] {Johnson1973} Johnson T.V., McGetchin T.R., 1973,
Icarus, 18, 612

\bibitem[\protect\citeauthoryear{K\"{u}ppers et al.}{2014}]{Kuppers2014} K\"{u}ppers M. et al., 2014,
Nature, 505, 525

\bibitem[\protect\citeauthoryear{Millis et al.}{1987}]{Millis1987} Millis R.L. et al., 1987,
Icarus, 72, 507

%\bibitem[\protect\citeauthoryear{Ortiz et al.}{2012}]{Ortiz2012} Ortiz J.L. et al., 2012,
%Nature, 491, 566

\bibitem[\protect\citeauthoryear{Sicardy et al.}{2011}]{Sicardy2011} Sicardy B. et al., 2011,
Nature, 478, 493

\bibitem[\protect\citeauthoryear{Thomas et al.}{2011}]{Thomas2005} Thomas P.C. et al., 2005,
Nature, 437, 224

\bibitem[\protect\citeauthoryear{van Belle}{1999}] {vanBelle1999} van Belle G.T., 1999,
PASP, 111, 1515

\bibitem[\protect\citeauthoryear{Widemann et al.}{2009}] {Widemann2009} Widemann T. et al., 2009,
Icarus, 199, 458

\bibitem[\protect\citeauthoryear{Zacharias et al.}{2004}]{Zacharias2004} Zacharias N. et al., 2004,
AJ, 127, 3043

\bibitem[\protect\citeauthoryear{Zacharias et al.}{2013}]{Zacharias2013} Zacharias N. et al., 2013,
AJ, 145, 44

%Artigos do exemplo

%\bibitem[\protect\citeauthoryear{Dawson}{1979}]{b4} Dawson D.W., 1979,
%ApJS, 41, 97
%\bibitem[\protect\citeauthoryear{Gerhz}{1972}]{b5} Gerhz R.D., 1972, ApJ,
%178, 715
%\bibitem[\protect\citeauthoryear{Gerhz \& Ney}{1972}]{b6} Gerhz R.D., Ney
%E.P., 1972, PASP, 84, 768
%\bibitem[\protect\citeauthoryear{Gerhz \& Woolf}{1970}]{b7} Gerhz R.D., Woolf N.J.,
%1970, ApJ, 161, L213
%\bibitem[\protect\citeauthoryear{Gilman}{1972}]{b8} Gilman R.C., 1972, ApJ, 178, 423
%\bibitem[\protect\citeauthoryear{Goldsmith et al.}{1987}]{b9} Goldsmith M.J., Evans A.,
%Albinson J.S., Bode M.F., 1987, MNRAS, 227, 143
%\bibitem[\protect\citeauthoryear{Hacking et al.}{1985}]{b10} Hacking P. et al., 1985,
%PASP, 97, 616
%\bibitem[\protect\citeauthoryear{Harvey, Thronson \& Gatley}{Harvey et al.}{1979}]{b11}
%Harvey P.M., Thronson H.A., Gatley I., 1979, ApJ, 231, 115
%\bibitem[\protect\citeauthoryear{Jura}{1986}]{b12} Jura M., 1986, ApJ, 309, 732
%\bibitem[\protect\citeauthoryear{Kukarkin et al.}{1969}]{b13} Kukarkin B.V. et al.,
%1969, General Catalogue of Variable Stars. Moscow
%\bibitem[\protect\citeauthoryear{Lloyd Evans}{1974}]{b14} Lloyd Evans T., 1974, MNRAS,
%167, 17{\sc p}
%\bibitem[\protect\citeauthoryear{Lloyd Evans}{1985}]{b15} Lloyd Evans T., 1985, MNRAS,
%217, 493
%\bibitem[\protect\citeauthoryear{Low et al.}{1984}]{b16} Low F.J. et al., 1984, ApJ,
%278, L19
%\bibitem[\protect\citeauthoryear{McLaughlin}{1932}]{b17} McLaughlin D.B., 1932, Publ. Univ.
%Obs. Mich., 4, 135
%\bibitem[\protect\citeauthoryear{O'Connell}{1961}]{b18} O'Connell J.K., 1961, Specola
%Vaticana Ric. Astron., 6, 341
%\bibitem[\protect\citeauthoryear{Olnon \& Raimond}{1986}]{b19} Olnon F.M., Raimond E., 1986,
%A\&AS, 65, 607
%\bibitem[\protect\citeauthoryear{Preston et al.}{1963}]{b20} Preston G.W., Krzeminski W., Smak J.,
%Williams J.A., 1963, ApJ, 137, 401
%\bibitem[\protect\citeauthoryear{Rowan-Robinson \& Harris}{1983a}]{b21} Rowan-Robinson M., Harris
%S., 1983a, MNRAS, 202, 767
%\bibitem[\protect\citeauthoryear{Rowan-Robinson \& Harris}{1983b}]{b22} Rowan-Robinson M., Harris
%S., 1983b, MNRAS, 202, 797
%\bibitem[\protect\citeauthoryear{van der Veen \& Habing}{1988}]{b23} van der Veen W.E.C.J., Habing
%H.J., 1988, A\&A, 194, 125
%\bibitem[\protect\citeauthoryear{Willems \& de Jong}{1988}]{b24} Willems F.J., de Jong T., 1988,
%A\&A, 196, 173
%\bibitem[\protect\citeauthoryear{Zuckerman \& Dyck}{1986}]{b25} Zuckerman B., Dyck H.M., 1986, ApJ,
%311, 345
\end{thebibliography}

%\bsp

\label{lastpage}

\end{document}

### 2015-05-16 Ceres/Old_versions/CeresV2.tex
Adicionado Artigos dos irregulares e de Ceres

--- 2015-05-16 Ceres/Old_versions/ArtigoV7/CeresV7.tex
@@
 \documentclass[useAMS,usenatbib]{mn2e}
-\usepackage[T1]{fontenc}
-\usepackage{prettyref}
+\usepackage{graphicx}
 \usepackage{amsmath}
-\usepackage{graphicx}
-\usepackage{subfigure}
-\usepackage[breaklinks=true]{hyperref} %% to avoid \citeads line
 
@@
 
-\title[Results of two stellar occultations by Ceres]{Results of two multi-chords stellar occultations by dwarf planet (1) Ceres}
-\author[A. R. Gomes-J\'unior, B. L. Giacchini, F. Braga-Ribas et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail:
-altair08@astro.ufrj.br},
-B. L. Giacchini$^{2,3,4}$\thanks{E-mail: breno@cbpf.br},
-F. Braga-Ribas$^{5, 6}$,
-M. Assafin$^{1}$,\newauthor
-R. Vieira-Martins$^{1,5}$
-T. Blank$^{4}$, 
-J. Brooks$^{4}$, 
-J. Broughton$^{7}$, 
-R. F. Dantowitz$^{8}$, \newauthor
-D. W. Dunham$^{4}$, 
-J. B. Dunham$^{4}$, 
-C. K. Ellington$^{4}$, 
-T. George$^{4}$, 
-P. D. Maley$^{4,9}$, \newauthor
-B. Timerson$^{4}$, 
-N. S. Weber$^{8}$\\
-$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
-CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
-$^{2}$Centro Brasileiro de Pesquisas F\'isicas, Rua Dr. Xavier Sigaud 150, Rio de Janeiro  22290-180, Brazil\\
-$^{3}$Se\c{c}\~ao de Oculta\c{c}\~oes/REA-Brasil, Av. Prof. Magalh\~aes Penido 403/302, Belo Horizonte 31270-700, Brazil\\
-$^{4}$International Occultation Timing Association, P.O. Box 131034, Houston, TX 77219-1034, USA\\
-$^{5}$Observat\'orio Nacional/MCT, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
-$^{6}$Federal University of Technology - Paraná (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil\\
-$^{7}$RASNZ Occultation Section, 18 Branch Cr. Reedy Creek, QLD 4227, Australia\\
-$^{8}$Clay Center Observatory at Dexter Southfield, 20 Newton Street, Brookline, MA 02445, USA\\
-$^{9}$NASA Johnson Space Center Astronomical Society, 4535 Cedar Ridge Trail, Houston, TX 77059 USA}
-
+\title[Results of two stellar occultations by Ceres]{Results of two stellar occultations by dwarf planet (1) Ceres}
+\author[A. V. Raveendran and A. N. Other]{A. V. Raveendran$^{1}$\thanks{E-mail:
+email@address (AVR); otheremail@otheraddress (ANO)} and A. N.
+Other$^{2}$\footnotemark[1]\thanks{This file has been amended to
+highlight the proper use of \LaTeXe\ code with the class file.
+These changes are for illustrative purposes and do not reflect the
+original paper by A. V. Raveendran.}\\
+$^{1}$Indian Institute of Astrophysics, Bangalore 560034, India\\
+$^{2}$Building, Institute, Street Address, City, Code, Country}
 \begin{document}
 
-\date{Accepted . Received ; in original form }
+\date{Accepted 1988 December 15. Received 1988 December 14; in original form 1988 October 11}
 
-\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2015}
+\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2002}
 
@@
 \begin{abstract}
-We report the results from two multi-chord stellar occultations by (1) Ceres that occurred in 2010 August 17, observed at Brazil, and 2013 October 25, observed at USA.
-Four positive chords were obtained for the 2010 occultation and nine for the 2013 occultation.
-An elliptic model was used to fit the apparent limb of Ceres.
-Two limb fitting solution was studied.
-The first one is a nominal solution with an indeterminate polar aspect angle.
-The second one, we constrain the pole coordinates as given by \cite{Drummond2014}.
-Assuming a Maclaurin spheroid, we determine an equatorial radius of $972 \pm 6~km$ and an apparent oblateness of $0.08 \pm 0.03$.
+Abstract
 \end{abstract}
@@
 \begin{keywords}
-minor planets, asteroids: individual (1, Ceres) - occultations - planets and satellites: fundamental parameters
+keywords.
 \end{keywords}
@@
 Far from being a mere taxonomic information, this suggests the great impact its study can have on the understanding of planetary formation and evolution of the Solar System.
-Indeed, it was proposed that Ceres could have its origin as a transneptunian object, later scattered to the Main Belt due to the giant planets' migration predicted by the `Nice Model'.
+Indeed, it was proposed that Ceres could have its origin as a transneptunian object, latter scattered to the Main Belt due to the giant planets' migration predicted by the ``Nice Model''.
 It may well had been formed close to its current location, however, even on this scenario, the dynamical history of the Solar System must have let its signatures on Ceres.
@@
 Yet the very question of whether Ceres suffered differentiation is open and, on the assumption of an affirmative answer, it is natural to ask if it ever had tectonics, how its geological evolution was, and if it is still active.
-Inarguably NASA's \textit{Dawn} mission will shed light on several open issues concerning Ceres.
+Inarguably, \textit{NASA}'s \textit{Dawn} mission will shed light on several open issues concerning Ceres.
 
-Owning approximately one fifth of the whole Main Belt's mass, Ceres is expected to have an equilibrium figure, i.e.
-a Maclaurin or a Jacobi ellipsoid.
+Owning approximately one fifth of the whole Main Belt's mass, Ceres is expected to have an equilibrium figure, \textit{i.e.}, a Maclaurin or a Jacobi ellipsoid.
 In fact, direct observations of Ceres by means of adaptive optics confirmed it to be an oblate spheroid \citep{Drummond2014}.
@@
 Since the 1960's occultations have provided measurements of hundreds of asteroids, thanks partially to the proficuous professional-amateur collaboration on the field.
-More recently, this technique has been applied to objects of the outer Solar System and has unveiled outstanding features of distant bodies, e.g.
-the ring system around the Centaur (10199) Chariklo \citep{BragaRibas2014}.
+More recently, this technique has been applied to objects of the outer Solar System and has unveiled outstanding features of distant bodies, \textit{e.g.}, the ring system around the Centaur (10199) Chariklo \citep{BragaRibas2014}.
 
-The first stellar occultation by Ceres was observed in 1984 \citep{Millis1987} and led to the determination of its size to the precision of the kilometre, in a time when the uncertainties were often ten times larger.
+The first stellar occultation by Ceres was observed in 1984 \citep{Millis1987} and lead to the determination of its size to the precision of the kilometre, at a time when the uncertainties were often ten times larger.
 The low apparent magnitude of Ceres, if compared to most asteroids, imposes a somewhat strong constraint on the stars capable of causing a detectable magnitude drop when occulted.
-For instance, after the 1984 event, to our knowledge, only four stellar occultations have been observed.
-Two of them had only two chords each, not sufficient, thus, for providing accurate results\footnote{These events took place on 22 August 1994 and 30 October 2010.}.
+For instance, after the 1984 event, to our knowledge, only four like this have been observed.
+Two of them had only two chords each, not sufficient, thus, for providing accurate measurements\footnote{These events took place on 22 August 1994 and 30 October 2010.}.
 The two remaining events, which occurred on 17 August 2010 and 25 October 2013, are reported on the present work.
-The former was observed in Brazil from 4 stations, while the latter was from the United States and led to 9 chords.
-Throughout the paper we shall refer to these events as the `2010' and the `2013 occultation'.
+The former was observed in Brazil from 5 stations, while the later was from the United States and led to 9 chords.
+Throughout the paper we shall refer to these events as the ``2010'' and the ``2013 occultation''.
 
-Both events were foreseen by Steve Preston\footnote{Predictions are published at http://asteroidoccultation.com.} on behalf of the International Occultation Timing Association, during routine prediction of asteroidal occultations of bright stars.
+This work is organized as follows.
+In Sections 2 and 3 we analyse, on this order, the 2010 and the 2013 events.
+The comparison of both results to those on the literature is carried out in Section 4.
+We summarise our conclusions in Section 5.
 
-This work is organized as follows. In Sections 2 and 3 we analyse, the 2010 and the 2013 events, respectively. The comparison of both results to those on the literature is carried out in Section 4.
+\section[]{The 2010 occultation}
 
-\section{The 2010 occultation}
-
-\subsection{Observations}
-
-%Asteroidal occultations of bright stars are routinely predicted by Steve Preston\footnote{Predictions are published at http://asteroidoccultation.com.}, on behalf of the International Occultation Timing Association.
-Within this context it was identified that on 17 August 2010 Ceres would occult the star TYC 6833-163-1 (UCAC2 20678706), which has magnitude $\text{K}=9.18$ and whose ICRF/J200 position is \citep{Zacharias2004}:
-%
-On 17 August 2010 Ceres was predicted to occult the star TYC 6833-163-1 (UCAC2 20678706), which has magnitude $K=9.18$ and whose ICRS position is \citep{Zacharias2004}:
-% UCAC4 313-111823
 \begin{equation}
@@
   \begin{array}{l l}
-    \alpha = 17^{\text{h}}18^{\text{m}}29^{\text{s}}.008\\
-    \delta = -27\degr 26\arcmin 38\arcsec.890.
+    \alpha = 17^{h}18^{m}29^{s}.0080\\
+    \delta = -27\degr 26\arcmin 38\arcsec.890
   \end{array}
 \right.
-\label{Eq: 2010-coord}
 \end{equation}
-%
-% Abaixo a coordenada UCAC4 aplicada de Movimento Próprio
-%
-%\begin{equation}
-%\left\{ 
-%  \begin{array}{l l}
-%    \alpha = 17^{\text{h}}18^{\text{m}}29^{\text{s}}.009\\
-%    \delta = -27\degr 26\arcmin 38\arcsec.866.
-%  \end{array}
-%\right.
-%\label{Eq: 2010-coord}
-%\end{equation}
-%
-The shadow's path would cross South America, the Atlantic and reach the south part of Africa.
 
@@
 \caption{Post-occultation reconstruction of Ceres' shadow path on Earth for the 2010 August 17 event.
-The big red dot is the geocentric closest approach at 22:40:25 UT.
-The small red ones represent the centre of the shadow separated by one minute, shadow moves from the left to the right.
-Blue dots are the sites that have observed the event.
+The big red dot is the geocentric closest approach at 22:41:28 UT.
+The small red ones represents the center of the shadow separated by one minute where the shadow moves from the left to the right.
+The blue dots are the sites that have observed the event.
 As described in text, UFSC had a negative chord.
@@
 
-Observations were carried out in Brazil at five stations as displayed in Table \ref{Tab: obs-2010} and Fig.
-\ref{Fig: Ceres-2010-map}.
+\subsection{Observations}
+
+Observations were carried out in Brazil at five stations as displayed in Table 1.
 The occultation was detected on four of them.
-UFSC had a negative chord.
-From the positive ones, INPE started the shots with the event already in progress and only provided star's reappearance time; the other three recorded the whole phenomenon.
-
-A remarkable circumstance of this event was the low velocity of Ceres: only $3.9~km.s^{-1}$ in the plane of the sky.
-Therefore, even exposures of a few seconds would translate in a relevant spatial resolution.
+Among these, one started the shots with the event already in progress and only provided star's reappearance time; the other three recorded the whole phenomenon.
+Since this event had a very low velocity -- only 3.9 km/s -- the exposure times (which for faster occultations would result in a big time uncertainty) led to quite small error bars, with the exception of the UEPG chord.
+Moreover, as the larger timing errors occurred closer to the edges of the shadow's path, their radial-projected uncertainty came to be still smaller.
 
@@
  \begin{minipage}{140mm}
-  \caption{Circumstances of observation for all observing stations of the 2010 event. \label{Tab: obs-2010}}
+  \caption{Circumstances of observation for all observing stations of the 2010 event.}
   \begin{tabular}{@{}lccccc}
   \hline
-          & Longitude & Telescope & Exposure & Result &   \\
-     Site & Latitude  & Aperture  & Cycle time & Ingress & Observer\\          
-          & Height    & Camera    &        & Egress    & \\          
+     Site & Longitude & Telescope & Exposure & Result & Observer  \\
+          & Latitude  & Aperture  & Readout overhead  & Ingress & \\          
+          & Height    & Camera    &    S/N    & Egress    & \\          
 \hline
  Belo Horizonte & 43$\degr$59'51.1'' W & LX200 & 5 s & Positive & C. Jacques  \\
- CEAMIG-REA &19$\degr$49'49.0'' S & 31 cm & 12 s & 22:39:03.9 $\pm$ 0.6 &  E. Pimentel \\
+ CEAMIG-REA &19$\degr$49'49.0'' S & 31 cm &     & 22:39:03.9 $\pm$ 0.6 &  E. Pimentel \\
             & 825 m                &       &     & 22:40:20 $\pm$ 5 &   \\
@@
  Pico dos Dias    & 45$\degr$34'45.1'' W &       & 1 s & Positive & J. I. B. Camargo \\
- OPD/LNA    &22$\degr$32'03.7'' S & 60 cm & 2 s & 22:37:30.3 $\pm$ 0.6 &  G. B. Rossi \\
+ OPD/LNA    &22$\degr$32'03.7'' S & 60 cm &     & 22:37:30.3 $\pm$ 0.6 &  G. B. Rossi \\
             & 1864 m               &       &     & 22:41:55.3 $\pm$ 0.7 &              \\
@@
  S\~ao Jos\'e dos       & 45$\degr$51'44.0'' W &  C11  & 2 s & Egress only & A. C. Milone\\
- Campos       &23$\degr$12'33.0'' S & 28 cm & 5 s & Start: 22:39:44 & T. Maldonado\\
+ Campos       &23$\degr$12'33.0'' S & 28 cm &     & Start: 22:39:44 & T. Maldonado\\
  INPE      & 975 m                & SBIG ST7 &     & 22:42:03.0 $\pm$ 0.2 & M. Okada    \\
@@
  Ponta Grossa       & 50$\degr$05'56.0'' W & RCX 400 &30 s & Positive & M. Emilio   \\
- UEPG       &25$\degr$05'22.2'' S & 40 cm & 52 s & 22:37:17 $\pm$ 13 & L. Mehret   \\
+ UEPG       &25$\degr$05'22.2'' S & 40 cm &     & 22:37:17 $\pm$ 13 & L. Mehret   \\
             & 910 m                & SBIG STL6E &     & 22:39:56 $\pm$ 13 & \\
@@
  Florian\'opolis       & 48$\degr$31'20.5'' W &       & 3 s & No occultation  & W. Schoenell\\
- UFSC       &27$\degr$36'12.3'' S & 28 cm & 6 s&  & A. J. T. Mello\\
+ UFSC       &27$\degr$36'12.3'' S & 28 cm &     &  & A. J. T. Mello\\
             & 20 m                 &       &     &         & F. R. Herpich \\
@@
 
-\subsection{Light curves and timing \label{Sec: light-curve-2010}}
+\subsection{Light curves and timing}
 
-The flux of the star in the five occultation chords was obtained from the FITS images with the Platform for Reduction of Astronomical Images Automatically (PRAIA) \citep{2011gfun.conf...85A}.
-The light curves were normalized to an unocculted star flux.
-Additionally, with exception of INPE, they were normalized by fitting a polynomial curve (of first or second order) outside the flux drop, so that the flux ratio was set to 1 outside the occultation.
-
-The start and end times of the occultation were obtained for each light curve by fitting a sharp edge occultation model.
-This model is convolved by Fresnel diffraction, the CCD bandwidth, the stellar aparent diameter, and the finite integration time; see \cite{Widemann2009}.
-
-The Fresnel scale ($F = \sqrt{\lambda D/2}$) for the present Ceres' geocentric distance $D = 2.29 \text{ au} = 3.42 \times 10^{8}$~km is 0.3~km for a typical wavelength of $\lambda = 0.65$~$\micron$.
-The star diameter is estimated using the formulae of \cite{vanBelle1999}.
-Its $B$, $V$, and $K$ apparent magnitudes are 12.00, 11.78 and 9.18, respectively, in the Tycho catalog \citep{Hog2000}.
-This yields a diameter of about 0.11~km projected at Ceres' distance.
-The smallest integration time used in the positive observations was 1.0~s, which translates to almost 3.9~km in the celestial plane.
-Therefore, the occultation light curves are largely dominated by the integration times, not by Fresnel diffraction or star diameter.
-
-The occultation fits consist in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011}.
-The free parameter to adjust is the ingress (disappearance) or egress (reappearance) time $t_{occ}$, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$.
-The best fittings to the 2010 occultation light curves are shown in Fig.
-\ref{Fig: Ceres-2010-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2010}.
+[Curvas de luz, modelagem, difraç\~ao, di\^ametro estelar.]
 
@@
 The solid black lines are the best fit of the square-well model to the data.
-Red lines are the square-well model convoluted with the Fresnel diffraction, the star diameter, and the applied exposure time.
+The red lines are the square-well model convoluted with the Fresnel diffraction, the star diameter, and the applied exposure time.
 The mid-times of the occultations do not coincide due to the propagation delays of the shadow due to the distinct longitude of the sites.
-Exposures at INPE started after the immersion, as explained in the text.
+The exposures at INPE were started after the immersion, as explained in the text.
 \label{Fig: Ceres-2010-curves}}
@@
 
-\subsection{Limb fitting methodology} \label{Sec: limbfittingmethod}
+\subsection{Limb fitting methodology}
 
-The methodology used to analyse Ceres' profile from the observations is the same described by \cite{Sicardy2011} and \cite{BragaRibas2013}.
+The methodology used to analyse Ceres' profile from the observations is the standard one, applied, for example, by \cite{Ortiz2012} and \cite{BragaRibas2013}.
 To each combination of site position and recorded ingress/egress time, together with star coordinates and Ceres' ephemeris, corresponds a point $(f_{i,obs},g_{i,obs})$ on the plane of the sky.
@@
 This model is supported by the work of \cite{Drummond2014}, by means of direct imaging of Ceres.
-Hence, we have $N=7$ chord extremities to adjust the $M=5$ parameters which define an ellipse: apparent semimajor and semiminor axis ($a^\prime$ and $b^\prime$, respectively), position angle $P$ of its semiminor axis and the position $(f_c,g_c)$ of its centre with respect to the occulted star.
-The coordinates $f_{c}$ and $g_{c}$, in kilometres, are calculated using the JPL\#26 Ceres' ephemeris and the occulted star's position.
-They are positive toward the local celestial east and north, respectively.
-The position angle $P$ is counted positively from the direction of local celestial north to celestial east.
-The apparent oblateness can be defined by $\epsilon^\prime = 1 - (b^\prime/a^\prime)$.
-We define the number of degrees of freedom of the problem as $\mathcal{N} \equiv N - M$.
-
-%We adopt an elliptic model for the limb profile, resulting from the projection of an oblate spheroid onto the sky plane.
-This model is supported by the work of \cite{Drummond2014}, by means of direct imaging of Ceres.
-Hence, we have $N=7$ chord extremities to adjust the $M=5$ parameters which define an ellipse: apparent semimajor and semiminor axis -- $a^\prime$ and $b^\prime$, respectively --, position angle $P$ of its semiminor axis and the position $(f_c,g_c)$ of its centre with respect to the occulted star\footnote{The coordinates $f_{c}$ and $g_{c}$, in kilometres, are calculated using the JPL\#33 Ceres' ephemeris and the occulted star's position.
-They are positive toward the local celestial east and north, respectively.
-The position angle $P$ is counted positively from the direction of local celestial north to celestial east.}.
+Hence, we have $N=7$ chord extremities to adjust the $M=5$ parameters which define an ellipse: apparent semimajor and semiminor axis -- $a^\prime$ and $b^\prime$, respectively --, position angle $P$ and the position $(f_c,g_c)$ of its centre with respect to the occulted star.
 The first two parameters can be interchanged with the oblateness and the equivalent radius, defined, on this order, by $\epsilon^\prime = 1 - (b^\prime/a^\prime)$ and $R_{equiv} \equiv \sqrt{a^\prime b^\prime} = a^\prime\sqrt{1-\epsilon^\prime}$.
@@
 
-The apparent oblateness is related to the true one $\epsilon$ through
+Correspondence between true and apparent figures depends on the polar aspect angle $\zeta$, encompassed by the (true) polar $b$-axis and the line of sight.
+The apparent oblateness is related to the true one, $\epsilon$, through
 %
 \begin{equation}
-\epsilon^\prime = 1 - \sqrt{\cos^2(\zeta) + (1-\epsilon)^2\sin^2(\zeta)},
-\label{Eq: TrueObla}
+\epsilon^\prime = 1 - \sqrt{\cos^2(\zeta) + (1-\epsilon)^2\sin^2(\zeta)}.
 \end{equation}
-%
-where $\zeta$ is the polar aspect angle, encompassed by the (true) polar semiminor axis and the line of sight.
-Of course, the apparent semimajor axis $a^\prime$ equals the equatorial radius $R_{equa}$ of the ellipsoid.
 
-%Correspondence between true and apparent figures depends on the polar aspect angle $\zeta$, encompassed by the (true) polar $b$-axis and the line of sight.
-The apparent oblateness is related to the true one, $\epsilon$, through
+Of course, the apparent semimajor axis $a^\prime$ equals the equatorial radius $a$ of the ellipsoid. %The polar radius $b$ can then be determined from $\epsilon = 1 - (b/a)$.
 
-The search for the best-fitting ellipse is carried out by minimizing the reduced $\chi^2$ function:
+The search for the best fit ellipse is carried out by minimizing the reduced $\chi^2$ function:
 %
@@
 
-Evaluation of the reduced $\chi^2$ function over the $\lbrace R_{equa}, \epsilon^\prime, P, f_c, g_c \rbrace$--space produces a map related to the likelihood of a solution.
+Evaluation of the reduced $\chi^2$ function over the $\lbrace a, \epsilon^\prime, P, f_c, g_c \rbrace$--space produces a map related to the likelihood of a solution.
 For instance, the 1--sigma reported time uncertainties correspond to the subset of the parameter space for which $\chi^2_r$ lies between ${\chi^2_r}_{min}$ and ${\chi^2_r}_{min} + 1$.
@@
 
-\subsection{Limb fitting solutions}\label{Sec: limbfitting-2010}
+\subsection{Limb fitting solutions}
 
-Two possible solutions were considered for the limb fitting.
-The first is the nominal one, which consists of determining the five parameters that characterize an ellipse from the seven observed contacts.
+We considered two possible solutions for the limb fitting.
+The first is the nominal one, which is the determination the five parameters which characterize an ellipse from the seven observed contacts.
 As we shortly show, it led to a rather large error bar on the position angle.
-Furthermore, the nominal solution solely is not capable of returning the true oblateness, which can be evaluated through equation \ref{Eq: TrueObla}, provided that the angle $\zeta$ is known.% or constrained by some physical hypothesis.
+Furthermore, the nominal solution solely is not capable of returning the true oblateness, which can be evaluated through equation (1), provided that the angle $\zeta$ is known.% or constrained by some physical hypothesis.
 
-From observations by adaptive optics spanning a 10-year period, \cite{Drummond2014} determined the position of Ceres' polar axis within a range of only 3$\degr$:
+From observations by adaptive optics spanning a 9-year period, \cite{Drummond2014} determined the position of Ceres' polar axis within a range of only 3$\degr$:
 \begin{equation}
@@
 %\end{array} \right .
-\label{Eq: Pole}
 \end{equation}
 %
-in equatorial J2000 coordinates.
-Together with Ceres' ephemeris at the moment of the occultation, this corresponds to the polar aspect angle $\zeta = 86.1\degr$, which is very close to an equator-on geometry.
-Hence, we expect true figures to be similar to apparent ones.
+in equatorial ICRF/J2000 coordinates.
+Together with Ceres' ephemeris at the moment of the occultation, this corresponds to a polar aspect angle $\zeta = 93.9\degr$, which is very close to an equator-on geometry.
+Hence, we expect true figures to be close to apparent ones.
 
 The knowledge of Ceres' pole not only allows the determination of its polar aspect, it suffices to set its position angle.
-Therefore Eq.
-\ref{Eq: Pole} may act as a constraint for $P$, and a second solution can be obtained by probing the parameter space with the restriction that the position angle is confined to the range that follows from Eq.
-\ref{Eq: Pole}.
+Therefore, (3) may act as a constraint for $P$, and a second solution can be obtained by probing the parameter space with the restriction that the position angle is confined to the range that follows from (3).
 We call this the ``pole-constrained solution''.
@@
 With the seven observed contacts it is possible to adjust the five parameters which define an ellipse.
-For the best-fitting solution we find $\chi^2_{r,min} = 0.24$, which could be interpreted as a slightly overestimation of the error bars with regard to the good quality of the fit.
+For the best fit solution we find $\chi^2_{r,min} = 0.24$, which could be interpreted as a slightly overestimation of the error bars with regard to the good quality of the fit.
 However, inasmuch as the problem has only two degrees of freedom, it is far from the statistical realm and low $\chi^2_{r,min}$ may eventually happen.
 
-The resulting values of equatorial diameter, oblateness, position angle and centre coordinates are presented in the second column of Table~2.%, and the corresponding ellipse is plotted with the observed chords in Fig.~2.
-% The equivalent radius of the best-fitting solution is $R_{equiv} = 469 \pm 10$ km.
+The resulting values of semimajor axis, oblateness, position angle and centre coordinates are presented in the second column of Table~2.%, and the corresponding ellipse is plotted with the observed chords in Fig.~2.
+% The equivalent radius of the best fit solution is $R_{equiv} = 469 \pm 10$ km.
 
 Already mentioned, the parameter with the largest uncertainty is the position angle: spanning on a 20$\degr$ interval, its determination has a relative precision worse than 10$\%$.
-Clearly, the coordinates of Ceres' pole (Eq.
-\ref{Eq: Pole}) can impose a strong constraint on the position angle, as the next solution shows.
+Clearly, the coordinates of Ceres' pole (3) can impose a strong constraint on the position angle, as the next solution shows.
 
@@
 
-At the moment of the occultation, the coordinates (Eq.
-\ref{Eq: Pole}) of Ceres' rotational pole correspond to the position angle $P = (12 \pm 3)\degr$.
-Exploration of the parameter space, restricted to ellipses whose position angle lies on this range, results in the pole-constrained solution.
-The related physical parameters are displayed on the third column of Table \ref{Tab: results}, while the best-fitting solution is depicted in Fig.\ref{Fig:Ceres-2010-body}.
+At the moment of the occultation, the coordinates (3) of Ceres' rotational pole correspond to the position angle $P = (12 \pm 3)\degr$.
+Exploration of the parameter space, restricted to ellipses whose position angle lie on this range, results in the pole-constrained solution.
+The related physical parameters are displayed on the third column of Table~2.
 
-We notice that the constraint corresponds to the upper limit of nominal solution's 1-sigma error bar for $P$.
-On the other hand, it selects the smallest values of equatorial radius, improving its determination by a factor of about 2.
+We notice that the constraint corresponds to the upper limit of nominal solution's 1-sigma error bar.
+On the other hand, it selects the smallest values of semimajor axis, improving its determination by a factor of about 2.
 Notwithstanding, oblateness' figures remain the same.
@@
  \begin{minipage}{140mm}
-  \caption{Results of limb fitting to the data of the 2010 and 2013 events.\label{Tab: results}}
+  \caption{Results of limb fitting to the data of the 2010 and 2013 events.}
   \begin{tabular}{@{}lcccc}
@@
 \hline
-Equatorial diameter (km) & 982 $\pm$ 14 & 972 $\pm$ 6  & 971 $\pm$ 7  & 971 $\pm$ 7\\
-%Equatorial diameter (km) & 491 $\pm$ 7  & 486 $\pm$ 3  & 485.5 $\pm$ 3.5 & 485.5 $\pm$ 3.5\\
-True oblateness        & 0.08 $\pm$ 0.03 & 0.08 $\pm$ 0.03 & 0.08 $\pm$ 0.04 & 0.08 $\pm$ 0.04\\
-Position angle (deg)   & 5 $\pm$ 10    & 12 $\pm$ 3 (*)& 22 $\pm$ 5    & 25 $\pm$ 3 (*)\\
-$f_c$ (km)             & 133 $\pm$ 9   & 138 $\pm$ 5   & 77 $\pm$ 6    & 78 $\pm$ 6\\
-$g_c$ (km)             & -17 $\pm$ 15  & -11 $\pm$ 11  & 13 $\pm$ 16   & 13 $\pm$ 16\\
-$\chi^2_{r,min}$       & 0.24          &  0.42         & 1.27          & 1.27\\
+Equatorial Radius (km) & 491 $\pm$ 7   & 486 $\pm$ 3   & 486 $\pm$ 2   & \\
+Oblateness          & 0.08$\pm$0.03 & 0.08$\pm$0.03 & 0.09$\pm$0.05 & \\
+Position angle (deg)& 5 $\pm$ 10    & 12 $\pm$ 3 (*)& 22 $\pm$ 5    &  \\
+$f_c$ (km)          & 133 $\pm$ 9   & 138 $\pm$ 5   & 78 $\pm$ 7    & \\
+$g_c$ (km)          & -17 $\pm$ 15  & -11 $\pm$ 11  & 14 $\pm$ 16   & \\
+$\chi^2_{r,min}$    & 0.24          &  0.42         & 1.27          & \\
 \hline
 \end{tabular}
-\textbf{Notes.} Error bars are at 1-sigma level.
+\textbf{Notes.} Error bars are at 1-sigma level. The reported oblateness derived from the 2010 event are the true ones, see text for details. \\
 (*) Position angle derived from Ceres' rotational pole coordinates determined by \cite{Drummond2014}.
@@
 \includegraphics[scale=0.36]{figures/Ceres_2010_body.pdf} 
-\caption{The best elliptical fit for the occultation chords for the event of 2010 using the times from Table~\ref{Tab: obs-2010} and the pole-constrained solution. \label{Fig:Ceres-2010-body}}
+\caption{\label{Fig:Ceres-2010-body}}
 \end{figure}
 
-\section{The 2013 occultation}
+\section[]{The 2013 occultation}
 
-\subsection{Observations}\label{Sec: observation-2013}
-
-The event which took place on 25 October 2013 involved the star TYC 865-911-1 (UCAC4 496-058191), of magnitude $K = 8.76$.
-Based on UCAC4 \citep{Zacharias2013}, the ICRS position to the date of occultation corrected of proper motion is :
-%
 \begin{equation}
@@
 \end{equation}
-%
-The occultation could only be visible in the United States, before dawn. The shadow would then cross the Atlantic and Africa, but during daylight time.
 
-\begin{figure*}
-\begin{centering}
-\subfigure{\includegraphics[scale=0.42]{figures/Ceres_2013.png}}
-\hspace{5mm}
-\subfigure{\includegraphics[scale=0.42]{figures/Ceres_2013-zoom.png}}
+\begin{figure}
+\includegraphics[scale=0.42]{figures/Ceres_2013.png} 
 \caption{Post-occultation reconstruction of Ceres' shadow path on Earth for the 2013 October 25 event at the east coast of USA.
-Left panel: Overview of the occultation.
-The big red dot is the geocentric closest approach at 09:43:03 UT and the small ones represent the centre of the shadow separated by 30 seconds where the shadow moves from the left to the right.
-The blue dots are the sites that have observed the event.
-Right Panel: Upper view of the occultation over the sites that observed the event.
-Red points are the centre of the shadow separated by 5 seconds.
-\label{Fig: Ceres-2013-map}}
-\end{centering}
-\end{figure*}
+The red dots represents the center of the shadow separated by one minute where the shadow moves from the left to the right.
+The geocentric closest approach at 09:43:11 UT.
+The blue dots are the sites that have observed the event.\label{Fig: Ceres-2013-map}}
+\end{figure}
 
-Nine positive chords were obtained by the variety of equipment listed in Table \ref{Tab: obs-2013}.
+\subsection{Observations}\label{Sec: observation-2013}
+
+Nine positive chords were obtained for the 2013 occutation by a variety of equipment as listed in Table \ref{Tab: obs-2013}.
 Each station was equipped with a video camera (whose readout time may be considered negligible).
-This is of particular importance in this event, since Ceres' shadow had the much faster speed of 42.6~km~s$^{-1}$, if compared to the 2010 event.
-All the observing sites are in the United States, as depicted in Fig.~\ref{Fig: Ceres-2013-map}.
+This is of particular importance in this event, since Ceres' shadow had the much faster speed of 42.6 $km/s$, if compared to the 2010 event.
+All the observing sites are in the United States, as depicted in Figure \ref{Fig: Ceres-2013-map}.
 
@@
 In fact, comparison between the times obtained at this station and the others suggests that the former have a delay of about 64 s.
-Therefore, we do not use Brookline's absolute times in the analysis that follows.
-Finally, at the remaining six stations the videos were recorded by camcorders on digital tapes.
+Therefore, for Brookline we do not use its absolute times in the analysis that follows.
+Finally, at the six stations left the videos were recorded by camcorders on digital tapes.
 The timing method consisted in the comparison of the camcorder internal clock to a 1PPS GPS signal, before and after the recording of the occultation.
@@
 Precisely, its altitude at moment of the occultation ranged between $20\degr$ (Hampton) and $15\degr$ (Winchester).
-Strong scintillation is expected in such a scenario which, combined to short integration times and the low magnitude drop of the event, resulted in rather noisy light curves and thus larger uncertainty in the time of the contacts, as it is shown in the next section.
+Strong scintillation is expected in such a scenario which, combined to the low magnitude drop of the event, resulted in a larger uncertainty in the time of the contacts, as it is shown in the next section.
 
@@
   \hline
-          & Longitude & Telescope: & Camera  & Result &   \\
-     Site & Latitude  & Aperture  & Cycle time & Ingress & Observer\\          
-          & Height    & f-ratio    &           & Egress    & \\          
+     Site & Longitude & Telescope: & Camera  & Result & Observer  \\
+          & Latitude  & Aperture  & Cycle time & Ingress & \\          
+          & Height    & f-ratio    &    S/N    & Egress    & \\          
 \hline
@@
 
-All videos were converted to FITS images and the photometry of the target was obtained via PRAIA \citep{2011gfun.conf...85A}.
-The light curves were normalized to the flux of a reference star when such was available on the field.
+All videos were converted to FITS images and the flux of the star was obtained with the Platform for Reduction of Astronomical Images Automatically (PRAIA) \citep{2011gfun.conf...85A}.
+The light curves were normalized to an unocculted star flux when a star was available.
 
-To reduce the noise, the data were binned by groups of five images -- with the exception of  Greenbelt, which was averaged in sets of ten.
-This procedure enlarges the effective integration times reported in Table~3 by a factor of 5 (or 10).
-As in the 2010 event, an additional normalization by a polynomial curve was applied.
+The data were binned by 5 or 10 points to reduce the noise. A linear or second-degree polynomial fit was applied to the observed flux before and after the occultation.
 
-The start and end times of the occultation were obtained by the same procedure explained in Section~2.2.
-The Fresnel scale of 0.4~km for this event was calculated using Ceres' geocentric distance $D = 3.28 \text{ au} = 4.91 \times 10^{8}$~km, for a typical wavelength of $\lambda = 0.65$~$\micron$.
-Considering the apparent magnitudes of the star, $B = 10.47$, $V=9.97$, and $K=8.76$ \citep{Hog2000}, we estimate its diameter projected at Ceres' distance as 0.16~km.
-Thence, since the typical effective integration time used (0.17~s) translates to about 7 km in the celestial plane, theoretical occultation light curves are largely dominated by the integration times, as for the 2010 event.
+For the observations at Owings and Greenbelt, the time was inserted on each frame.
+For Brookline, the time was extracted from the image header.
+For the others, the time was obtained from the reports of IOTA.
+The cycle time of the images is shown in Table \ref{Tab: obs-2013}.
 
-The best-fittings to the occultation light curves are shown in Fig.~\ref{Fig: Ceres-2013-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2013}.
+The start and end times of the occultation were obtained for each light curve by fitting a sharp edge occultation model.
+This model is convolved by Fresnel diffraction, the CCD bandwidth, the stellar diameter in kilometers, and the finite integration time; see \cite{Widemann2009}.
+
+The Fresnel scale ($F = \sqrt{\lambda D/2}$) for the present Ceres’ geocentric distance $D = 3.28 AU = 4.91 \times 10^{8} km$ is 0.4 km for a typical wavelength of $\lambda = 0.65 \micron m$.The star diameter is estimated using the formulae of \cite{vanBelle1999}.
+Its B, V , and K apparent magnitudes are 10.47, 9.97 and 8.76, respectively, in the Tycho catalog \citep{Hog2000}.
+This yields a diameter of about 0.16 km projected at Ceres’s distance.
+The smallest integration time used in the positive observations was 0.033 s, which translates to almost 1.4 km in the celestial plane.
+Therefore, the occultation light curves are largely dominated by the integration times, not by Fresnel diffraction or star diameter.
+
+The occultation fits consist in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011}.
+The free parameter to adjust is the ingress (disappearance) or egress (re-appearance) time $t_{occ}$, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$.
+
+The best fits to the occultation light curves are shown in Figure \ref{Fig: Ceres-2013-curves}, and the derived occultation times are listed in Table \ref{Tab: obs-2013}.
 
@@
 \includegraphics[scale=0.58]{figures/Ceres_2013_fluxratio} 
-\caption{The nine occultation light curves normalized and vertically shifted by a factor of 1.0 for better viewing.
-The solid black lines are the best fit of the square-well model to the data.
-The red lines are the square-well model convoluted with the Fresnel diffraction, the star diameter, and the applied exposure time.
-The mid-times of the occultations do not coincide due to the propagation delays of the shadow due to the distinct longitude of the sites.
-The light curve of Brookline is shifted by -64 s as explained in the text.
-\label{Fig: Ceres-2013-curves}}
+\caption{Same as in Fig. \ref{Fig: Ceres-2010-curves} for the nini light curves of the 2013 event. \label{Fig: Ceres-2013-curves}}
 \end{figure}
@@
 
-Elliptic limb profiles were adjusted to the eight\footnote{Brookline's chord was not used for limb fitting since it actually measured duration instead of absolute times.} chords by the same procedure described in Section \ref{Sec: limbfittingmethod}\footnote{For this event, the JPL\#33 Ceres' ephemeris was used to obtain the $f_{c}$ and $g_{c}$ coordinates.}.
-This yielded $\chi^2_{r,min} = 13$, suggesting that an elliptic model is not satisfactory to the data.
-Indeed, a quick glance at a plot of the observed chords, such as in Fig.
-\ref{Fig: Ceres-2013-body}, shows that the Varina chord seems to be somewhat advanced with respect to the others.
-Taking into account that in this station time stamps were not inserted on the video frames, it is always possible to attribute this advance to an eventual problem on the correspondence between camcorder's and GPS' times.
-This could be caused, for example, if the camcorder delayed to start the recording and, since this was an unattended pre-pointed station, this fact would not be noticed by the analysis of the video itself.
+From the occultation times obtained, the shape for Ceres' limb was adjusted by an ellipse characterized by five parameters: the coordinates of the body center, relative to the star in the plane of the sky ($f_{c}, g_{c}$); the apparent semimajor axis a, equivalent to the equatorial radius $R_{Equa}$ ; the apparent oblateness = $(a'-b')/a'$ (where $b'$ is the apparent semiminor axis); and the position angle P of the semiminor axis $b'$ .
+The coordinates $f_{c}$ and $g_{c}$, in kilometers, are calculated using the JPL\#33  Ceres' ephemeris and the star position given in Section \ref{Sec: observation-2013}.
+They are positive toward the local celestial east and north, respectively.
+The position angle P is counted positively from the direction of local celestial north to celestial east.
 
-The immersion recorded at Owings also seems to be shifted (delayed) with respect to the nearby chords (see Fig.
-\ref{Fig: Ceres-2013-body}).
-This chord, actually, has roughly the same length of Mechanicsville's, despite the fact that they are separated by circa $100~km$.
-Differently from Varina, though, this station had time stamps inserted in each frame of the video, which makes it more unlike to justify an eventual timing issue.
-Another possible explanation for the times of this chord is the determination of the events in the light curve analysis, which could have been affected by noise.
-Finally, the delay during the ingress could be caused by a relief feature in Ceres; we shall soon return to this hypothesis.
-
-%In addition to a time basis problem, other possible explanations to the times of this chord are the determination of the events in the light curve analysis, due to noise; and finally that the delay during the ingress could be caused by a relief feature in Ceres.
-
-In a second limb fitting, thus, we did not consider Brookline, Varina and Owings chords.
-The adjust of the five parameters which define an ellipse to the 12 contacts then resulted in $\chi^2_{r,min} = 1.27$, indicating that the fitting is in good agreement with the observed data within the error bars.
-This is the solution depicted in Fig.~\ref{Fig: Ceres-2013-body}, where we see that the chord length measured in Brookline is compatible to the model.
-The associated physical parameters are presented in Table~2 as the nominal solution.
-
-For this event the polar aspect angle is $\zeta = 90.7\degr$, which makes true oblateness be equal to the apparent one, within the error bars.
-
-It is worth noting that the position angle of this nominal solution is better constrained than the respective of the 2010 occultation.
-Actually, the uncertainty of the former is of $5\degr$ in contrast to $10\degr$ of the latter.
-This suggests that the pole-constrained solution obtained via the position of Ceres' pole (Eq.
-\ref{Eq: Pole}) would not be significantly different of the nominal one.
-
-This assumption is confirmed when we carry out the limb fitting with the constraint $P = (25 \pm 3)\degr$, the position angle at the moment of the occultation which follows from (Eq.
-\ref{Eq: Pole}).
-The physical parameters related to this pole-constrained solution are presented in the last column of Table~2, and are essentially the same of the nominal solution.
-
-We close this section returning to the hypothesis that the delay observed in the immersion at Owings could be associated to a limb topography feature.
-The recorded contact would then correspond to a negative elevation of $31 \pm 4$ km with respect to the best-fitting ellipse.
-Theoretical models, however, predict that relief in Ceres should be no higher than about 10--20 km \citep{Johnson1973}, while published observational data sets the bound of 18 km \citep{Carry2008}.
-More recently, images by the probe \textit{Dawn} also reveal an even smoother surface.
-The association of Owings first contact to a relief must, therefore, be ruled out.
+From the nine positive chords, six were able to give a well-constrained limb fit solution. The elliptical fit to the N = 12
+chord extremities is obtained by minimizing the relevant $\chi^{2}$ function (see, for more details, \cite{Sicardy2011} and \cite{BragaRibas2013}).
 
@@
 \includegraphics[scale=0.36]{figures/Ceres_2013_body.pdf}
-\caption{The best elliptical fit for the occultation chords for the event of 2013 using timing from Table \ref{Tab: obs-2013} and the nominal solution.
-The chord of Brookline if shifted by -64s as explained in the text\label{Fig: Ceres-2013-body}}
+\caption{\label{Fig: Ceres-2013-body}}
 \end{figure}
 
-\section{Discussion}
+\section[]{Discussion}
 
-A quick glance at Table~2 shows there is an overall agreement between the physical parameters derived from both occultations, specially in the equatorial diameter.
-The differences of the solutions occur basically on the size of their error bars, and can be justified by the particularities of each set of data.
+\section{Conclusions}
 
-The 2010 event, for example, had only seven contacts; none the less, they were well distributed over Ceres' disc (see Fig.
-\ref{Fig:Ceres-2010-body}) acting as a constraint to is shape.
-On the other hand, the 2013 event had five more contacts, but they were concentrated in certain regions of the body.
-In particular, the absence of chords closer to Ceres' southern limb made its oblateness be worse determined here than in the 2010 event.
+\section*{Acknowledgments}
 
-However, even our best measurement for the oblateness, $\epsilon=0.08 \pm 0.03$, has still a big uncertainty with regard to other figures published in the literature, as Table~4 shows.
-A larger number of uniformly spaced chords would be necessary to offer a best constraint to the oblateness.
-
-The few chords of the 2010 occultation could themselves only constraint the position angle of the object to a $20\degr$ interval.
-This range was reduced by a factor of two in the 2013 event, approaching --and verifying-- the result predicted by the work of \cite{Drummond2014}.
-As it was shown, using the coordinates of Ceres' polar axis to limit the position angle was not an efficient procedure in the case of this occultation, in the sense that the aforementioned limitation had not correspond to a stronger constraint on the other parameters.
-
-On the other hand, constraining the position angle on the 2010 occultation was proved to reduce the error bars of the other parameters (disregarding oblateness).
-Moreover, this procedure resulted in an excellent agreement between the equatorial radius figures of both events.
-
-The 2013 occultation, therefore, offers not only an independent verification of the figures resulted from the 2010 event, but also validates the procedure carried out there --and which led to the best-constrained parameters in this work.
-
-Comparison of Ceres' equatorial diameter as measured by different techniques is carried out in Table \ref{Tab: Ceres-final}.
-We note an overall agreement between our result to those obtained via direct imaging by the Hubble Space Telescope (HST) \citep{Thomas2005}, the Keck Observatory and the ESO VLT \citep{Drummond2014}.
-The smaller figure reported by \cite{Carry2008} may be justified by the fact that in this study the effects of limb-darking were not taken into account, as pointed out by \cite{Drummond2014}.
-
-As mentioned in the Introduction, the 1984 event \citep{Millis1987} is the only occultation data to which we can compare our result; and it is interesting to notice that the diameters measured in each case do not agree, being our result larger than the other.
-It is difficult to state with sure the reasons for this divergence since we had no access to the original 1987 data or reduction methodology.
-We note, however, that in this occultation a variation of 0.5 s in the contact times correspond to almost 7~km on Ceres' limb.
-Therefore, in order to properly compare their result to ours it would be necessary to reduce their light curves with the same methodology used in the present work.
-
-The geocentric positions for Ceres at the time of the occultations could also be obtained with great accuracy. These are given in Eqs. \ref{Eq: Ceres-2010_pos} and \ref{Eq: Ceres-2013_pos}.
-
-\begin{equation}
-2010 ~ Aug ~ 17
-\left\{ 
-  \begin{array}{l l}
-    Time = 22:40:25\\
-    \alpha = 17^{\text{h}}18^{\text{m}}29^{\text{s}}.008\\
-    \delta = -27\degr 26\arcmin 38\arcsec.890.
-  \end{array}
-\right.
-\label{Eq: Ceres-2010_pos}
-\end{equation}
-
-\begin{equation}
-2013 ~ Oct ~ 25
-\left\{ 
-  \begin{array}{l l}
-    Time = 09:43:03\\
-    \alpha = 11^{h}57^{m}52^{s}.7641\\
-    \delta = +09\degr 07\arcmin 49\arcsec.835
-  \end{array}
-\right.
-\label{Eq: Ceres-2013_pos}
-\end{equation}
-
-\begin{table}
-  \caption{Ceres' equatorial diameter and oblateness \label{Tab: Ceres-final}}
-  \begin{centering}
-  \begin{tabular}{@{}llcc}
-  \hline
-     Eq. diameter (km) & Oblateness & Method & Ref. \\
-\hline
-$972 \pm 6$  & $0.08  \pm 0.03$  & Occultation & 1\\
-$967 \pm 10$ & $0.078 \pm 0.015$ & Keck+VTL    & 2 \\
-$959 \pm 5$  & $0.074 \pm 0.007$ & Keck        & 3\\
-$975 \pm 4$  & $0.068 \pm 0.005$ & HST         & 4\\
-$959 \pm 5$  & $0.05  \pm 0.01$  & Occultation & 5\\
-\hline
-\end{tabular}
-\par\end{centering}
-\textbf{References.} 1: Present work. 2: \cite{Drummond2014}. 3: \cite{Carry2008}. 4: \cite{Thomas2005}. 5: \cite{Millis1987}.
-\end{table}
-
-\section*{Acknowledgements}
-
-ARGJ thanks the financial support of CAPES.
-BLG thanks CNPq.
-F.B.R.
-acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014.
-We also acknowledge Steve Preston for the predictions of the occultations.
+Agradecimentos
 
@@
 
-\bibitem[\protect\citeauthoryear{Carry et al.}{2008}] {Carry2008} Carry B. et al., 2008,
-A\&A, 478, 235
-
-\bibitem[\protect\citeauthoryear{Hog et al.}{2000}] {Hog2000} Hog E. et al., 2000,
-A\&A, 355, L27
-
-\bibitem[\protect\citeauthoryear{Johnson \& McGetchin}{1973}] {Johnson1973} Johnson T.V., McGetchin T.R., 1973,
-Icarus, 18, 612
+\bibitem[\protect\citeauthoryear{Hog E. et al}{2000}] {Hog2000} Hog E. et al. 2000, A\&A, 355, L27-30 (2000)
 
@@
 
-%\bibitem[\protect\citeauthoryear{Ortiz et al.}{2012}]{Ortiz2012} Ortiz J.L. et al., 2012,
-%Nature, 491, 566
+\bibitem[\protect\citeauthoryear{Ortiz et al.}{2012}]{Ortiz2012} Ortiz J. L. et al., 2012,
+Nature, 491, 566
 
@@
 
-\bibitem[\protect\citeauthoryear{Thomas et al.}{2011}]{Thomas2005} Thomas P.C. et al., 2005,
-Nature, 437, 224
+\bibitem[\protect\citeauthoryear{Widemann et al.}{2009}] {Widemann2009} Widemann, T., Sicardy, B., Dusser, R., et al. 2009, Icarus, 199, 458
 
-\bibitem[\protect\citeauthoryear{van Belle}{1999}] {vanBelle1999} van Belle G.T., 1999,
-PASP, 111, 1515
-
-\bibitem[\protect\citeauthoryear{Widemann et al.}{2009}] {Widemann2009} Widemann T. et al., 2009,
-Icarus, 199, 458
-
-\bibitem[\protect\citeauthoryear{Zacharias et al.}{2004}]{Zacharias2004} Zacharias N. et al., 2004,
-AJ, 127, 3043
-
-\bibitem[\protect\citeauthoryear{Zacharias et al.}{2013}]{Zacharias2013} Zacharias N. et al., 2013,
-AJ, 145, 44
+\bibitem[\protect\citeauthoryear{van Belle}{1999}] {vanBelle1999} van Belle, G. T. 1999, PASP, 111, 1515
 

### 2015-05-16 Ceres/paper-Ceres-source/paper-Ceres.tex
Adicionado Artigos dos irregulares e de Ceres

% mn2esample.tex
%
% v2.1 released 22nd May 2002 (G. Hutton)
%
% The mnsample.tex file has been amended to highlight
% the proper use of LaTeX2e code with the class file
% and using natbib cross-referencing. These changes
% do not reflect the original paper by A. V. Raveendran.
%
% Previous versions of this sample document were
% compatible with the LaTeX 2.09 style file mn.sty
% v1.2 released 5th September 1994 (M. Reed)
% v1.1 released 18th July 1994
% v1.0 released 28th January 1994

\documentclass[useAMS,usenatbib]{mn2e}
\usepackage[T1]{fontenc}
\usepackage{prettyref}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{epstopdf}
\usepackage{color}
\usepackage[breaklinks=true]{hyperref}

% If your system does not have the AMS fonts version 2.0 installed, then
% remove the useAMS option.
%
% useAMS allows you to obtain upright Greek characters.
% e.g. \umu, \upi etc.  See the section on "Upright Greek characters" in
% this guide for further information.
%
% If you are using AMS 2.0 fonts, bold math letters/symbols are available
% at a larger range of sizes for NFSS release 1 and 2 (using \boldmath or
% preferably \bmath).
%
% The usenatbib command allows the use of Patrick Daly's natbib.sty for
% cross-referencing.
%
% If you wish to typeset the paper in Times font (if you do not have the
% PostScript Type 1 Computer Modern fonts you will need to do this to get
% smoother fonts in a PDF file) then uncomment the next line
% \usepackage{Times}

%%%%% AUTHORS - PLACE YOUR OWN MACROS HERE %%%%%

\title[Results of two stellar occultations by Ceres]{Results of two multi-chord stellar occultations by dwarf planet (1) Ceres}
\author[A. R. Gomes-J\'unior, B. L. Giacchini, F. Braga-Ribas et al.]{A. R. Gomes-J\'unior$^{1,}$\thanks{E-mail: altair08@astro.ufrj.br},
B. L. Giacchini$^{2,3,4}$,%\thanks{E-mail: breno@cbpf.br},
F. Braga-Ribas$^{5, 6, \dag}$,
M. Assafin$^{1, \dag, \ddag}$,\newauthor
R. Vieira-Martins$^{1,5, \dag, \ddag}$,
J.I.B. Camargo$^{5, \dag}$,
B. Sicardy$^{7}$,
B. Timerson$^{4}$, 
T. George$^{4}$, \newauthor
J. Broughton$^{8}$, 
T. Blank$^{4}$,
G. Benedetti-Rossi$^{5}$, 
J. Brooks$^{4}$,  
R. F. Dantowitz$^{9}$,  \newauthor
D. W. Dunham$^{4}$, 
J. B. Dunham$^{4}$, 
C. K. Ellington$^{4}$,
M. Emilio$^{10}$,
F.R. Herpich$^{11}$,  \newauthor
C. Jacques$^{3, 12}$,
P. D. Maley$^{4,13}$,
L. Mehret$^{10}$,
A.J.T. Mello$^{14}$,
A.C. Milone$^{15}$,\newauthor
E. Pimentel$^{3, 12}$,  
W. Schoenell$^{11}$,
N. S. Weber$^{9}$
\\
$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
$^{2}$Centro Brasileiro de Pesquisas F\'isicas, Rua Dr. Xavier Sigaud 150, Rio de Janeiro  22290-180, Brazil\\
$^{3}$Se\c{c}\~ao de Oculta\c{c}\~oes/REA-Brasil, Belo Horizonte, Brazil\\
$^{4}$International Occultation Timing Association, P.O. Box 131034, Houston, TX 77219-1034, USA\\
$^{5}$Observat\'orio Nacional/MCTI, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil\\
$^{7}$LESIA, Observatoire de Paris, CNRS UMR 8109, Universit\'{e} Pierre et Marie Curie, Universit\'{e} Paris-Diderot, 5 place Jules Janssen,\\ F-92195 Meudon Cedex, France\\
$^{8}$RASNZ Occultation Section, P.O. Box 3181, Wellington, New Zealand\\
$^{9}$Clay Center Observatory at Dexter Southfield, 20 Newton Street, Brookline, MA 02445, USA\\
$^{10}$Universidade Estadual de Ponta Grossa, Ponta Grossa, Brazil\\
$^{11}$Universidade Federal de Santa Catarina, Florian\'opolis, Brazil\\
$^{12}$Centro de Estudos Astron\^omicos de Minas Gerais, Belo Horizonte, Brazil\\
$^{13}$NASA Johnson Space Center Astronomical Society, Houston, Texas, USA\\
$^{14}$Federal University of Technology - Paran\'a (UTFPR / DAELT), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil\\
$^{15}$Instituto Nacional de Pesquisas Espaciais, S\~ao Jos\'e dos Campos, Brazil\\
$^\dag$ Associated to Laborat\'{o}rio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, CEP 20921-400,\\ Rio de Janeiro, Brazil\\
$^\ddag$Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France
}
\begin{document}

\date{Accepted . Received ; in original form }

\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2015}

\maketitle

\label{firstpage}

\begin{abstract}
We report the results of two multi-chord stellar occultations by the dwarf planet (1) Ceres that were observed  from Brazil on 2010 August 17, and from the USA on 2013 October 25. Four positive detections were obtained for the 2010 occultation, and nine for the 2013 occultation. Elliptical models were adjusted to the observed chords to obtain Ceres' size and shape. Two limb fitting solutions were studied for each event. The first one is a nominal solution with an indeterminate polar aspect angle. The second one was constrained by the pole coordinates as given by Drummond et al. Assuming a Maclaurin spheroid, we determine an equatorial diameter of 972 $\pm$ 6 km and an apparent oblateness of $0.08 \pm 0.03$ as our best solution. These results are compared to all available size and shape determinations for Ceres made so far, and shall be confirmed by the NASA's \textit{Dawn} space mission.
\end{abstract}

\begin{keywords}
minor planets, asteroids: individual (1, Ceres) - occultations - planets and satellites: fundamental parameters
\end{keywords}

\section{Introduction}

Ceres is the sole example of a dwarf planet in the inner Solar System. Far from being mere taxonomic information, this suggests the great impact its study can have on the understanding of planetary formation and evolution of the Solar System. Indeed, it was proposed that Ceres' origin could have been as a transneptunian object \citep{McKinnon12}, later scattered to the Main Belt due to the giant planets' migration predicted by the `Nice Model' \citep{Gomes05}. Even if it was formed close to its current location, the dynamical history of the Solar System must have let its signatures on Ceres. These could include not only the late heavy bombardment features that might exist on its surface, but also the makeup of its volatiles, which could have been transported from the outer region.

Since the 1970's it has been speculated that Ceres could contain water ice, which was recently verified \citep{Kuppers2014}. Although the water regime on this object is still unknown, some internal structure models suggest the existence of a water ice -- or even a liquid water -- layer \citep{CastilloR2011}. Yet the very question of whether Ceres underwent differentiation is open and, on the assumption of an affirmative answer, it is natural to ask if it ever had tectonics, what its geological evolution was, and if it is still active. Inarguably NASA's \textit{Dawn} mission \citep{Russell2004}, which is currently orbiting the dwarf planet, will shed light on several open issues concerning Ceres.

Containing approximately one fifth of the whole Main Belt's mass, Ceres is expected to have an equilibrium figure, i.e. a Maclaurin or a Jacobi ellipsoid. In fact, direct observations of Ceres by means of adaptive optics indicate that it is an oblate spheroid \citep{Drummond2014}. The precise knowledge of its size and shape is of utmost importance, for the models of density, internal structure and differentiation.% can depend on the size and oblateness of the body.

The best ground-based technique for determining shape and size of a faraway object is the study of its shadow, cast by a star during an occultation. Since the 1960's occultations have provided measurements of hundreds of asteroids, thanks partially to the fruitful professional-amateur collaboration on the field. More recently, this technique has been applied to objects of the outer Solar System and has unveiled outstanding features of distant bodies, e.g. the ring system around the Centaur (10199) Chariklo \citep{BragaRibas2014}.

The first stellar occultation by Ceres was observed in 1984 \citep{Millis1987} and led to the determination of its size to the precision of some kilometres, at a time when the uncertainties were often ten times larger. The high apparent brightness of Ceres, as compared to most asteroids, imposes a somewhat strong constraint on the stars capable of causing a detectable magnitude drop when occulted. For instance, after the 1984 event, to our knowledge, only four stellar occultations have been observed \citep{Dunham2014}. Two of them had only two chords each, thus not sufficient for providing accurate results\footnote{These events took place on 1994 August 22 and 2010 October 30.}. The two remaining events, which occurred on 2010 August 17 and 2013 October 25, are reported in the present work and are the first ones that used charged-couple devices (CCD) as recording systems. % The former was observed in Brazil from 4 stations, while the latter was from the United States and led to 9 chords.
Throughout the paper we shall refer to these events as the `2010 occultation' and the `2013 occultation'.

Both events were predicted by Steve Preston\footnote{Predictions are published at http://asteroidoccultation.com.} on behalf of the International Occultation Timing Association, during routine prediction of asteroidal occultations of bright stars.

This work is organized as follows. In Sections 2 and 3 we analyse the 2010 and the 2013 occultations, respectively. In Section 4 we give the geocentric positions of Ceres derived from the occultations. The comparison of our results to those in the literature is carried out in Section 5.

\section{The 2010 occultation}

\subsection{Observations}

As predicted, on 2010 August 17 Ceres occulted the star TYC 6833-163-1 (UCAC4 313-111823), which has magnitude $V = 11.55$ and has the ICRS position, based on UCAC4 catalogue \citep{Zacharias2013}, to the date of occultation:

%abaixo (UCAC2 20678706),
\begin{equation}
\left\{ 
  \begin{array}{l l}
    \alpha = 17^{\rmn{h}}18^{\rmn{m}}29\fs0085\\
    \delta = -27\degr 26\arcmin 38\farcs 867
  \end{array}
\right.
\label{Eq: 2010-coord}
\end{equation}

% Abaixo a coordenada UCAC4 aplicada de Movimento Próprio
% UCAC4 313-111823
%
%\begin{equation}
%\left\{ 
%  \begin{array}{l l}
%    \alpha = 17^{\text{h}}18^{\text{m}}29^{\text{s}}.009\\
%    \delta = -27\degr 26\arcmin 38\arcsec.866.
%  \end{array}
%\right.
%\label{Eq: 2010-coord}
%\end{equation}
%
%The shadow's path would cross South America, the Atlantic and reach the south part of Africa.

\begin{figure}
\includegraphics[scale=0.42]{figures/Ceres_2010.eps} 
\caption{Post-occultation reconstruction of Ceres' shadow path on Earth for the 2010 August 17 event. The big red dot is the geocentric closest approach at 22:40:25 UT. The small red ones represent the centre of the shadow separated by one minute, shadow moves from the left to the right. Blue dots are the sites that have observed the event. As described in text, UFSC had a negative chord.
\label{Fig: Ceres-2010-map}}
\end{figure}

Observations were carried out in Brazil from five different sites as displayed in Table \ref{Tab: obs-2010} and Fig. \ref{Fig: Ceres-2010-map}. The occultation was detected from four of them. The southern-most one (UFSC) had a negative chord. From the positive sites, the one named INPE started observing after the event was already in progress, due to technical difficulties, thus providing only the star's reappearance time; the other three recorded the whole phenomenon.

A remarkable circumstance of this event was the low velocity of Ceres: only 3.9 km s$^{-1}$ in the plane of the sky. Therefore, even exposures of a few seconds would translate in a relevant spatial resolution.

All the observations were made through a sequence of images obtained with CCDs. The times of each exposure were available on the header of each image with an internal accuracy of a few hundredths of a second. CEAMIG had only the integer part of the second available, due to the acquisition software used. The fraction of a second could be retrieved as described on the Section~3.1 of \cite{BragaRibas2013}  (which shall be referred as BR13 hereafter). Cycle times (exposure plus read-out) varied from 2 to 52 seconds, as can be verified on Table~\ref{Tab: obs-2010}, making it a heterogeneous set of observations and imposing an error of a few seconds for the ingress / egress times of some sites.

%\textcolor{blue}{[Falar sobre a quest\~ao do tempo em cada corda. A leitura do header era no segundo inteiro? Foi analisada a cad\^encia das imagens? Qual a precis\~ao?] [Durante o per\'iodo de observac\~ao Ceres rodou apenas 3.1$\degr$. Mas isso n\~ao \'e relevante no caso de Maclaurin...]}

\begin{table*}
 \centering
 \begin{minipage}{140mm}
  \caption{Circumstances of observation for all observing stations of the 2010 event. \label{Tab: obs-2010}}
  \begin{tabular}{@{}lccccc}
  \hline
          & Longitude & Telescope & Exposure & Result &   \\
     Site & Latitude  & Aperture  & Cycle time & Ingress & Observer\\          
          & Height    & f-ratio & Camera & Egress    & \\          
\hline
 Belo Horizonte & 43$\degr$59\arcmin 51\farcs1~W & LX200 & 5 s & Positive & C. Jacques  \\
 CEAMIG-REA &19$\degr$49\arcmin 49\farcs0~S & 31 cm & 12 s & 22:39:03.9 $\pm$ 0.6 s &  E. Pimentel \\
            & 825 m       &  f/10         &   SBIG ST10        & 22:40:20 $\pm$ 5 s &   \\
 & & & & & \\
 Pico dos Dias    & 45$\degr$34\arcmin45\farcs1~W &  Zeiss     & 1 s & Positive & J. I. B. Camargo \\
 LNA    &22$\degr$32\arcmin03\farcs7~S & 60 cm & 2 s & 22:37:30.3 $\pm$ 0.6 s &  G. B. Rossi \\
            & 1864 m     & f/12.5          &    Andor Ikon       & 22:41:55.3 $\pm$ 0.7 s &              \\
 & & & & & \\
 S\~ao Jos\'e dos       & 45$\degr$51\arcmin44\farcs0~W &  C11  & 2 s & Egress only & A. C. Milone\\
 Campos       &23$\degr$12\arcmin33\farcs0~S & 28 cm & 5 s & Start obs.: 22:39:44 & T. Maldonado\\
 INPE      & 620 m      & f/6.3          & SBIG ST7     & 22:42:03.0 $\pm$ 0.2 s & M. Okada    \\
 & & & & & \\
 Ponta Grossa       & 50$\degr$05\arcmin56\farcs0~W & RCX 400 &30 s & Positive & M. Emilio   \\
 UEPG       &25$\degr$05\arcmin22\farcs2~S & 40 cm & 52 s & 22:37:17 $\pm$ 13 s & L. Mehret   \\
            & 910 m      & f/8          & SBIG STL6E     & 22:39:56 $\pm$ 13 s & \\
 & & & & & \\
 Florian\'opolis       & 48$\degr$31\arcmin20\farcs5~W &   C11      & 3 s & No occultation  & W. Schoenell\\
 UFSC       &27$\degr$36\arcmin12\farcs3~S & 28 cm & 6 s&  Start obs.: 21:49:27 & A. J. T. Mello\\
            & 20 m       & f/10          &  SBIG ST7      &   End obs.: 22:51:21      & F. R. Herpich \\
\hline
\end{tabular}
\end{minipage}
\end{table*}

\subsection{Light curves \label{Sec: light-curve-2010}}

The flux of the star in the five occultation chords was obtained from the FITS images with the \textsc{platform for reduction of astronomical images automatically (praia)} \citep{2011gfun.conf...85A}. The light curves were normalized to the flux of the star plus Ceres, as they were merged right before and after the occultation. Additionally,
%, with exception of INPE
 they were normalized by fitting a polynomial curve (of first or second order) outside the flux drop, so that the flux ratio was set to 1 outside the occultation.

The ingress (disappearance) and egress (reappearance) instants of the occultation were obtained for each light curve by fitting a square-well model convoluted with the Fresnel diffraction, the CCD bandwidth, the stellar apparent diameter, and the applied finite exposure time; see \cite{Widemann2009} and BR13. %\cite{BragaRibas2013}.
%The Fresnel scale ($F = \sqrt{\lambda D/2}$) for the present Ceres' geocentric distance $D = 2.29 \text{ au} = 3.42 \times 10^{8}$~km is 0.3~km for a typical wavelength of $\lambda = 0.65$~$\micron$. The star diameter is estimated using the formulae of \cite{vanBelle1999}. Its $B$, $V$, and $K$ apparent magnitudes are 12.00, 11.78 and 9.18, respectively, in the Tycho catalog \citep{Hog2000}. This yields a diameter of about 0.11~km projected at Ceres' distance. 
The smallest integration time used in the positive observations was 1.0~s, which translates to almost 3.9 km in the celestial plane. Therefore, the error on the time determination of the ingress and egress is largely dominated by the integration times, not by Fresnel diffraction or star diameter, which are both on the order of a few hundred meters for this event.
The occultation data fit consists in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011} and BR13.\ The free parameter to adjust is the ingress or egress instant, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$. The best fittings to the 2010 occultation light curves are shown in Fig. \ref{Fig: Ceres-2010-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2010}.
%The occultation fits consist in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011} and BR13.\ The free parameter to adjust is the ingress (disappearance) or egress (reappearance) time $t_{occ}$, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$. The best fittings to the 2010 occultation light curves are shown in Fig. \ref{Fig: Ceres-2010-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2010}.

\begin{figure}
\includegraphics[scale=0.58]{figures/Ceres_2010_fluxratio.eps} 
\caption{The four occultation light curves normalized and vertically shifted by a factor of 0.2 for better visualisation. The solid black lines are the best fit of the square-well model to the data. Red lines are the square-well model convoluted with the Fresnel diffraction, the star diameter, and the applied exposure time. The mid-times of the occultations do not coincide due to the propagation delays of the shadow due to the distinct longitude of the sites. Exposures at INPE started after the immersion, as explained in the text. \label{Fig: Ceres-2010-curves}}
\end{figure}

\subsection{Limb fitting methodology} \label{Sec: limbfittingmethod}

The methodology used to analyse Ceres' profile from the observations is the same described by \cite{Sicardy2011} and BR13.%\cite{BragaRibas2013}. 
 ~Each combination of site position and recorded ingress/egress instant, together with star coordinates and Ceres' ephemeris, corresponds to a point in the plane of the sky. The collection of all these points ideally determines the apparent limb of Ceres.
%~To each combination of site position and recorded ingress/egress time, together with star coordinates and Ceres' ephemeris, corresponds a point $(f_{i,obs},g_{i,obs})$ on the plane of the sky. The collection of those points ideally determines the apparent limb of Ceres.

We adopt an elliptic model for the limb profile, resulting from the projection of an oblate spheroid onto the sky plane. This choice is supported by the work of \cite{Drummond2014}, by means of direct imaging of Ceres. Hence, we have $N=7$ chord extremities to adjust the $M=5$ parameters which define an ellipse: apparent semimajor and semiminor axis ($a^\prime$ and $b^\prime$, respectively), position angle $P$ of its semiminor axis and the position $(f_c,g_c)$ of its centre with respect to the occulted star. 
Of course, the apparent semimajor axis {$a^\prime$} is equivalent to the equatorial radius $R_{equa}$ of the ellipsoid.
The coordinates $f_{c}$ and $g_{c}$, in kilometres, are calculated using the JPL\#33 Ceres' ephemeris \citep{Giorgini1996} and the occulted star's position. They are positive toward the local celestial east and north, respectively. The position angle $P$ is counted positively from the direction of local celestial north to celestial east. The apparent oblateness can be defined by $\epsilon^\prime = 1 - (b^\prime/a^\prime)$. The best-fitting solution is obtained minimizing a reduced $\chi^{2}_{r}$ function, where we define the number of degrees of freedom of the problem as $\mathcal{N} \equiv N - M$.  All the procedures that allow the determination of the error bars of the physical parameters can be found on BR13.

%We adopt an elliptic model for the limb profile, resulting from the projection of an oblate spheroid onto the sky plane. This model is supported by the work of \cite{Drummond2014}, by means of direct imaging of Ceres. Hence, we have $N=7$ chord extremities to adjust the $M=5$ parameters which define an ellipse: apparent semimajor and semiminor axis -- $a^\prime$ and $b^\prime$, respectively --, position angle $P$ of its semiminor axis and the position $(f_c,g_c)$ of its centre with respect to the occulted star\footnote{The coordinates $f_{c}$ and $g_{c}$, in kilometres, are calculated using the JPL\#33 Ceres' ephemeris and the occulted star's position. They are positive toward the local celestial east and north, respectively. The position angle $P$ is counted positively from the direction of local celestial north to celestial east.}. The first two parameters can be interchanged with the oblateness and the equivalent radius, defined, on this order, by $\epsilon^\prime = 1 - (b^\prime/a^\prime)$ and $R_{equiv} \equiv \sqrt{a^\prime b^\prime} = a^\prime\sqrt{1-\epsilon^\prime}$. We define the number of degrees of freedom of the problem as $\mathcal{N} \equiv N - M$.

%%The apparent oblateness is related to the true one $\epsilon$ through
%%%
%%\begin{equation}
%%\epsilon^\prime = 1 - \sqrt{\cos^2(\zeta) + (1-\epsilon)^2\sin^2(\zeta)},
%%\label{Eq: TrueObla}
%%\end{equation}
%%%
%%where $\zeta$ is the polar aspect angle, encompassed by the (true) polar semiminor axis and the line of sight. Of course, the apparent semimajor axis $a^\prime$ equals the equatorial radius $R_{equa}$ of the ellipsoid.
%%
%%%Correspondence between true and apparent figures depends on the polar aspect angle $\zeta$, encompassed by the (true) polar $b$-axis and the line of sight. The apparent oblateness is related to the true one, $\epsilon$, through
%%
%%The search for the best-fitting ellipse is carried out by minimizing the reduced $\chi^2$ function:
%%%
%%\begin{equation}
%%\chi^2_r = \frac{1}{\mathcal{N}} \sum_{i=1}^N \frac{1}{\sigma_i^2}  \left[ (f_{i,obs}-f_{i,cal})^2 + (g_{i,obs}-g_{i,cal})^2\right].
%%\end{equation}
%%%
%%Here $\sigma_i$ is the radial uncertainty of the $i$th chord extremity, obtained by multiplying the time uncertainty reported in Table~1 by the normal velocity of the star (with respect to the limb model). $f_{i,cal}$ and $g_{i,cal}$ are the coordinates of the point on the model ellipse, where a line that goes from the centre of the model to the data point $(f_{i,obs},g_{i,obs})$ intersects the ellipse.%:
%%%
%%%\begin{equation}
%%%f_{i,cal} \equiv f_{i,obs} + \rho \cos(\theta), \quad g_{i,cal} \equiv %g_{i,obs} + \rho \sin(\theta),
%%%\end{equation}
%%%
%%%with:
%%%\begin{equation}
%%%\theta \equiv \arctan\left( \frac{g_{i,obs}-g_{c}}{f_{i,obs}-f_{c}} %\right) ,
%%%\end{equation}
%%%
%%%\begin{equation}
%%%\rho \equiv a^\prime b^\prime / \sqrt{[a^\prime\sin(\theta+P)]^2 +[b^\prime\cos(\theta+P)]^2 }.
%%%\end{equation}
%%
%%Evaluation of the reduced $\chi^2$ function over the $\lbrace R_{equa}, \epsilon^\prime, P, f_c, g_c \rbrace$--space produces a map related to the likelihood of a solution. For instance, the 1--sigma reported time uncertainties correspond to the subset of the parameter space for which $\chi^2_r$ lies between ${\chi^2_r}_{min}$ and ${\chi^2_r}_{min} + 1$. This allows the determination of the error bars of the physical parameters. It is worthwhile to mention that we consider all the uncertainty as due to timing issues.

\subsection{Limb fitting solutions}\label{Sec: limbfitting-2010}

Two possible solutions were considered for the limb fitting. The first, which we call nominal solution, consists of determining the five parameters that characterize an ellipse from the seven observed contacts. As we shortly show, it led to a rather large uncertainty on the position angle. Furthermore, the nominal solution alone is not capable of returning the true oblateness, which can be evaluated through equation 2 of BR13  provided that the aspect angle $\zeta$ is known.% or constrained by some physical hypothesis.

From observations with adaptive optics spanning a 10-year period, \cite{Drummond2014} determined the position of Ceres' polar axis within an error of only 3$\degr$:
\begin{equation}
%\left\{ 
%\begin{array}{l}
\alpha = (287 \pm 3) \degr, \quad %\\
\delta = (+64 \pm 3) \degr,
%\end{array} \right .
\label{Eq: Pole}
\end{equation}
%
in equatorial J2000 coordinates. Together with Ceres' ephemeris at the moment of the occultation, this corresponds to the polar aspect angle $\zeta = 86.1\degr$, which is very close to an equator-on geometry. Hence, we expect true figures to be similar to apparent ones.

The knowledge of Ceres' pole not only allows the determination of its polar aspect, it suffices to set its position angle. Therefore Eq.~\ref{Eq: Pole} may act as a constraint for $P$, and a second solution can be obtained by probing the parameter space with the restriction that the position angle is confined to the range that follows from Eq.~\ref{Eq: Pole}. We call this the `pole-constrained solution'.

\subsubsection{Nominal solution}

With the seven observed contacts it is possible to adjust the five parameters which define an ellipse. For the best-fitting solution we find $\chi^2_{r,min} = 0.24$, which could be interpreted as a slightly overestimation of the error bars with regard to the good quality of the fit. However, inasmuch as the problem has only two degrees of freedom, it is far from the statistical realm and relatively small $\chi^2_{r,min}$ are acceptable.

The resulting values of equatorial diameter, oblateness, position angle and centre coordinates are presented in the second column of Table~\ref{Tab: results}.%, and the corresponding ellipse is plotted with the observed chords in Fig.~2.
% The equivalent radius of the best-fitting solution is $R_{equiv} = 469 \pm 10$ km.

Already mentioned, the parameter with the largest uncertainty is the position angle: spanning on a 20$\degr$ interval, its determination has a relative precision worse than 10 per cent. Clearly, the coordinates of Ceres' pole (Eq. \ref{Eq: Pole}) can impose a strong constraint on the position angle, as the next solution shows.

Finally, the correction to the oblateness due to Ceres' polar aspect angle lies within the 1-sigma error bar and has no statistical relevance; hence $\epsilon = 0.08 \pm 0.03$.

\subsubsection{Pole-constrained solution}

At the occultation, the coordinates (Eq. \ref{Eq: Pole}) of Ceres' rotational pole correspond to the position angle $P = (12 \pm 3)\degr$. Exploration of the parameter space, restricted to ellipses with position angles laying within this range, results in the pole-constrained solution. The related physical parameters are displayed on the third column of Table \ref{Tab: results} in boldface, while the best-fitting solution is depicted in Fig.\ref{Fig:Ceres-2010-body}.

We notice that the constraint corresponds to the upper limit of nominal solution's 1-sigma error bar for $P$. On the other hand, it selects the smallest values of semimajor axis, improving its determination by a factor of about 2. Notwithstanding, oblateness' figures remain the same.

\begin{table*}
 \centering
 \begin{minipage}{140mm}
  \caption{Results of limb fitting to the data of the 2010 and 2013 events.\label{Tab: results}}
  \begin{tabular}{@{}lcccc}
  \hline
     Solution & 2010/Nominal & \textbf{2010/Pole-constrained} & 2013/Nominal & 2013/Pole-constrained \\
\hline
Equatorial diameter (km) & 982 $\pm$ 14 & \textbf{972 $\pm$ 6}  & 971 $\pm$ 7  & 971 $\pm$ 7\\
%Equatorial radius (km) & 491 $\pm$ 7  & \textbf{486 $\pm$ 3}  & 485.5 $\pm$ 3.5 & 485.5 $\pm$ 3.5\\
True oblateness        & 0.08 $\pm$ 0.03 & \textbf{0.08 $\pm$ 0.03} & 0.08 $\pm$ 0.04 & 0.08 $\pm$ 0.04\\
Position angle (deg)   & 5 $\pm$ 10    & \textbf{12 $\pm$ 3} (*)& 22 $\pm$ 5    & 25 $\pm$ 3 (*)\\
$f_c$ (km)             & 97 $\pm$ 9   & \textbf{102 $\pm$ 5}   & 77 $\pm$ 6    & 78 $\pm$ 6\\
$g_c$ (km)             & 16 $\pm$ 15  & \textbf{21 $\pm$ 11}  & 13 $\pm$ 16   & 13 $\pm$ 16\\
$\chi^2_{r,min}$       & 0.24          &  \textbf{0.42}         & 1.27          & 1.27\\
\hline
\end{tabular}
\textbf{Notes.} In bold we highlight our best solution. Error bars are at 1-sigma level. The polar diameter ($D_{pol}$) can be easily calculated from $D_{pol}=D_{equa}(1 - \epsilon)$.
(*) Position angle derived from Ceres' rotational pole coordinates determined by \cite{Drummond2014}.
\end{minipage}
\end{table*}

\begin{figure}
\includegraphics[scale=0.37, angle=-90]{figures/Ceres_2010_sphere.eps} 
\caption{The best elliptical fit for the occultation chords for the event of 2010 using the times from Table~\ref{Tab: obs-2010} and the pole-constrained solution. The arrow indicates the direction of motion, blue lines are the observed chords, the red segments are the ingress, egress and mid-occultation error bars at 1-$\sigma$ level. \label{Fig:Ceres-2010-body}}
\end{figure}

\section{The 2013 occultation}

\subsection{Observations}\label{Sec: observation-2013}

The event which took place on 2013 October 25 involved the star TYC 865-911-1 (UCAC4 496-058191), of magnitude $V = 10.05$. Based on UCAC4 \citep{Zacharias2013}, the ICRS position to the date of occultation is:
%
\begin{equation}
\left\{ 
  \begin{array}{l l}
    \alpha = 11^{\rmn{h}}57^{\rmn{m}}52\fs7641\\
    \delta = +09\degr 07\arcmin 49\farcs835
  \end{array}
\right.
\end{equation}
%
The occultation could only be visible in the United States, before dawn, as depicted in Fig. \ref{Fig: Ceres-2013-map}.% The shadow would then cross the Atlantic and Africa, but during daylight time.

\begin{figure}
%\begin{centering}
%\subfigure{\includegraphics[scale=0.42]{figures/Ceres_2013_2.png}}
%\hspace{5mm}
%\subfigure
{\includegraphics[scale=0.42]{figures/Ceres_2013-zoom.eps}}
\caption{Post-occultation reconstruction of Ceres' shadow path on Earth for the 2013 October 25 event at the east coast of USA. 
%Left panel: Overview of the occultation. The big red dot is the geocentric closest approach at 09:43:03 UT and the small ones represent the centre of the shadow separated by 30 seconds where the shadow moves from the left to the right. The blue dots are the sites that have observed the event. 
Upper view of the occultation over the sites that observed the event (blue dots). Red points are the centre of the shadow separated by 15 seconds.
\label{Fig: Ceres-2013-map}}
%\end{centering}
\end{figure}

Nine positive chords were obtained by the variety of instruments listed in Table \ref{Tab: obs-2013}. Each station was equipped with a video camera with negligible readout times. This was of particular importance, since in this event Ceres' shadow speed was 42.6 km s$^{-1}$, much faster than in the 2010 event.

Three different timing synchronization procedures were adopted among the set of observing stations. At Greenbelt and Owings the 1PPS signal of a GPS unit was used to calibrate time stamps which were inserted at each frame of the video. Time extraction is thus straightforward, after taking camera delays into account. On the other hand, at Brookline the clock would be synchronized by an internet server. A lack of connection, however, resulted in spurious times. In fact, comparison between the times obtained at this station and the others suggests that the former have a delay of about 64 s. Therefore, we do not use Brookline's %absolute 
times in the analysis that follows. Finally, at the remaining six stations the videos were recorded by camcorders on digital tapes. The timing method consisted in the comparison of the camcorder internal clock to a 1PPS GPS signal, before and after the recording of the occultation. Absolute timing errors of this procedure are expected to be less than 0.1 s.

During the event Ceres was low in the horizon, with altitudes between $15\degr$ (Winchester) and $20\degr$ (Hampton). Strong scintillation is expected in such a scenario which, combined to short integration times and the low magnitude drop of the event, resulted in rather noisy light curves and thus larger uncertainties in the time of the contacts, as is shown in the next section.

\begin{table*}
 \centering
 \begin{minipage}{140mm}
  \caption{Circumstances of observation for the observing stations of the 2013 event.\label{Tab: obs-2013}}
  \begin{tabular}{@{}lccccc}
  \hline
          & Longitude & Telescope: & Camera  & Result &   \\
     Site & Latitude  & Aperture  & Cycle time & Ingress & Observer\\          
          & Height    & f-ratio    &           & Egress    & \\          
\hline
 Hampton & 70$\degr$48\arcmin59\farcs7~W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
  &42$\degr$53\arcmin52\farcs8~N & f/5 & 0.033 s     & 09:40:46.9 $\pm$ 0.1 s &   \\
            & 7 m       &  &     & 09:40:57.26  $\pm$ 0.08 s &   \\
 & & & & & \\
 Topsfield & 70$\degr$55\arcmin16\farcs6~W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
  &42$\degr$37\arcmin55\farcs9~N & f/5      & 0.033 s   & 09:40:45.4 $\pm$ 0.1 s &   \\
            & 45 m      &  &     & 09:40:58.0  $\pm$ 0.1 s &   \\
 & & & & & \\
 Brookline & 71$\degr$08\arcmin14\farcs5~W & 64 cm & Infinity2-1R & Positive & N. Weber \\
  &42$\degr$18\arcmin27\farcs4~N & f/9.6 & 0.015 s     & 09:41:48.00 $\pm$ 0.01 s &  R. Dantowitz  \\
            & 109 m     &       &     & 09:42:02.93 $\pm$ 0.01 s   \\
 & & & & & \\
 Winchester & 78$\degr$14\arcmin39\farcs6~W & 36 cm & PC164C & Positive & J. Brooks \\
  &39$\degr$16\arcmin21\farcs5~N & f/5 & 0.033 s    & 09:40:33.26 $\pm$ 0.08 s &   \\
            & 211 m     &  &     & 09:40:55.86  $\pm$ 0.09 s &   \\
 & & & & & \\
 Greenbelt & 76$\degr$52\arcmin09\farcs4~W & 12 cm & PC164C-EX2 & Positive & J. Dunham \\
  &38$\degr$59\arcmin12\farcs1~N & f/2.5 & 0.033 s    & 09:40:33.6 $\pm$ 0.1 s & D. Dunham  \\
            & 52 m      &   &     & 09:40:56.4  $\pm$ 0.1 s &   \\
 & & & & & \\
 Alexandria & 77$\degr$02\arcmin28\farcs3~W & 7 cm & Watec120N & Positive & P. Maley \\
  &38$\degr$49\arcmin19\farcs1~N & f/10 &  0.067 s   & 09:40:33.3 $\pm$ 0.1 s &   \\
            & 8 m       &   &     & 09:40:56.1  $\pm$ 0.1 s &   \\
 & & & & & \\
 Owings & 76$\degr$38\arcmin06\farcs3~W & 25 cm & PC164C  & Positive & C. Ellington \\
  &38$\degr$41\arcmin26\farcs5~N & f/3.3 &  0.033 s & 09:40:34.27 $\pm$ 0.05 s &   \\
            & 38 m                &       &     & 09:40:56.0  $\pm$ 0.2 s &   \\
 & & & & & \\
 Mechanicsville & 77$\degr$23\arcmin06\farcs7~W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
  &37$\degr$41\arcmin26\farcs1~N & f/2.5 & 0.033 s    & 09:40:33.0 $\pm$ 0.1 s &   \\
            & 60 m      &   &     & 09:40:54.8  $\pm$ 0.1 s &   \\
 & & & & & \\
 Varina & 77$\degr$19\arcmin49\farcs3~W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
  &37$\degr$25\arcmin58\farcs6~N & f/2.5 &  0.033 s   & 09:40:32.4 $\pm$ 0.3 s &   \\
            & 19 m      &   &     & 09:40:53.1  $\pm$ 0.2 s &   \\

\hline
\end{tabular}
\end{minipage}
\end{table*}

\subsection{Light curves}

All videos were converted to FITS images and the photometry of the target was obtained via \textsc{praia} \citep{2011gfun.conf...85A}. The light curves were normalized to the flux of a reference star when available on the field.

To reduce the noise, the data were binned by groups of five images -- with the exception of  Greenbelt, which was averaged in sets of ten. This procedure enlarges the effective integration times reported in Table~\ref{Tab: obs-2013} by a factor of 5 (or 10). As in the 2010 event, an additional normalization by a polynomial curve was applied.

The ingress and egress instants of the occultation were obtained by the same procedure explained in Section \ref{Sec: light-curve-2010}. %The Fresnel scale of 0.4~km for this event was calculated using Ceres' geocentric distance $D = 3.28 \text{ au} = 4.91 \times 10^{8}$~km, for a typical wavelength of $\lambda = 0.65$~$\micron$. Considering the apparent magnitudes of the star, $B = 10.47$, $V=9.97$, and $K=8.76$ \citep{Hog2000}, we estimate its diameter projected at Ceres' distance as 0.16~km. 
Since the typical effective integration time used (0.17~s) translates to about 7 km in the celestial plane, and the Fresnel scale and star diameter are again of the order of a few hundreds of meters, the theoretical occultation light curves are largely dominated by the integration times, as for the 2010 event.

The best-fittings to the occultation light curves are shown in Fig.~\ref{Fig: Ceres-2013-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2013}.

\begin{figure}
\includegraphics[scale=0.58]{figures/Ceres_2013_fluxratio.eps} 
\caption{The nine occultation light curves normalized and vertically shifted by a factor of 1.0 for better viewing, see Fig.~\ref{Fig: Ceres-2010-curves} for the explanation of the graph.% The solid black lines are the best fit of the square-well model to the data. The red lines are the square-well model convoluted with the Fresnel diffraction, the star diameter, and the applied exposure time. The mid-times of the occultations do not coincide due to the propagation delays of the shadow due to the distinct longitude of the sites. 
The light curve of Brookline is shifted by -64 s as explained in the text. \label{Fig: Ceres-2013-curves}}
\end{figure}

\subsection{Limb fitting solutions}

Elliptic limb profiles were adjusted to all the available\footnote{Brookline's chord was not used for limb fitting since it had an inaccurate absolute time.} chords by the same procedure described in Section \ref{Sec: limbfittingmethod}.%\footnote{For this event, the JPL\#33 Ceres' ephemeris was used to obtain the $f_{c}$ and $g_{c}$ coordinates.}
This yielded $\chi^2_{r,min} = 13$, suggesting that an elliptic model is not satisfactory to the data. Indeed, a quick glance at a plot of the observed chords, such as in Fig. \ref{Fig: Ceres-2013-body}, shows that the Varina chord seems to be somewhat advanced with respect to the others. Taking into account that in this station time stamps were not inserted on the video frames, it is fairly possible to attribute this advance to an eventual problem on the correspondence between camcorder's and GPS' times. This could be caused, for example, if the camcorder delayed to start the recording and, since this was an unattended pre-pointed station, this fact would not be noticed by the analysis of the video itself.

The immersion recorded at Owings also seems to be shifted (delayed) with respect to the nearby chords (see Fig. \ref{Fig: Ceres-2013-body}). This chord, actually, has roughly the same length of Mechanicsville's, despite the fact that they are separated by about 100 km. Differently from Varina, though, this station had time stamps inserted in each frame of the video, which makes it more unlikely to justify an eventual timing issue. Another possible explanation for the apparent problem of the times of this chord is the determination of the ingress and egress instants in the light curve analysis, which could have been affected by noise. Finally, the delay during the ingress could be caused by a relief feature in Ceres; we shall soon return to this hypothesis.

In a second limb fitting, thus, we did not consider Brookline, Varina and Owings chords. The adjust of the five parameters which define an ellipse to the 12 contacts then resulted in $\chi^2_{r,min} = 1.27$, indicating that the fitting is in good agreement with the observed data within the error bars. This is the solution depicted in Fig.~\ref{Fig: Ceres-2013-body}, where we see that the chord length measured in Brookline is compatible to the model. The associated physical parameters are presented in Table~\ref{Tab: results} as the nominal solution.

For this event the polar aspect angle is $\zeta = 90.7\degr$, which makes the true oblateness equal to the apparent one, within the error bars.

The position angle of this nominal solution is better constrained than the 2010 occultation. Actually, the uncertainty in the former is of $5\degr$ in contrast to $10\degr$ of the latter. This suggests that the pole-constrained solution obtained via the position of Ceres' pole (Eq. \ref{Eq: Pole}) would not be significantly different from the nominal one.

This assumption was confirmed when we carried out the limb fitting with the constraint $P = (25 \pm 3)\degr$, the position angle at the occultation which follows from Eq. \ref{Eq: Pole}. The physical parameters related to this pole-constrained solution are presented in the last column of Table~\ref{Tab: results}, and are essentially the same of the nominal solution.

In the hypothesis that the delay observed in the immersion at Owings could be associated to a limb topography feature, the recorded contact would then correspond to a negative elevation of 31 $\pm$ 4 km with respect to the best-fitting ellipse. Theoretical models, however, predict that reliefs in Ceres should not be higher than about 10--20 km \citep{Johnson1973}, while published observational data sets the bound of 18 km \citep{Carry2008}. More recently, images by the probe \textit{Dawn} also reveal an even smoother surface. Therefore, the association of Owings first contact to a relief is improbable.

\begin{figure}
\includegraphics[scale=0.36]{figures/Ceres_2013_sphere.pdf}
\caption{The best elliptical fit for the occultation chords for the event of 2013 using timing from Table \ref{Tab: obs-2013} and the pole-constrained solution.  The arrow indicates the direction of motion, blue lines are the observed chords, the red and green segments are the ingress, egress and mid-occultation error bars at 1-$\sigma$ level. The chords with green error bars were not used during the limb fit process. The chord of Brookline if shifted by -64s as explained in the text.\label{Fig: Ceres-2013-body}}
\end{figure}

\section{Astrometry from Occultations}

The objects' geocentric positions derived from stellar occultations are most valuable for improving their orbits \citep{Desmars2015}. Usually the precision of the astrometric positions are limited by the accuracy of the occulted star position, not by the limb fit. Ceres' geocentric J2000 positions at the time of each occultation are displayed in Eqs.~\ref{Eq: Ceres-2010_pos} and \ref{Eq: Ceres-2013_pos}. The errors of the positions come from the errors of the star positions, taken from the catalogues and from the errors of the relative apparent distances between star and Ceres, derived from the limb fit (which are displayed as $\Delta\alpha \cos\delta$, $\Delta\delta$).

%\caption{Ceres' geocentric positions from the reported stellar occultations \label{Tab: Ceres-astrometry}}
\begin{equation}
2010 ~ \text{Aug} ~ 17
\left\{
 \begin{array}{l l}
    \text{Time} = 22:40:00\\
    \alpha = 17^{\rmn{h}}18^{\rmn{m}}29\fs0122\pm0\farcs027\\
    \delta = -27\degr 26\arcmin 38\farcs617\pm0\farcs028\\
    \Delta\alpha\cos\delta = 0\farcs003; ~~\Delta\delta = 0\farcs007
  \end{array}
\right.
\label{Eq: Ceres-2010_pos}
\end{equation}
\begin{equation}
2013 ~ \text{Oct} ~ 25
\left\{
  \begin{array}{l l}
    \text{Time} = 09:45:00\\
    \alpha = 11^{\rmn{h}}57^{\rmn{m}}52\fs9154\pm0\farcs019\\
    \delta = +09\degr 07\arcmin 49\farcs865\pm0\farcs021\\
    \Delta\alpha\cos\delta = 0\farcs002; ~~\Delta\delta = 0\farcs007
  \end{array}
\right.
\label{Eq: Ceres-2013_pos}
\end{equation}

\section{Discussion}

A quick glance at Table~\ref{Tab: results} shows an overall agreement between the physical parameters derived from both occultations, specially in the equatorial diameter. The differences of the solutions occur basically on the size of their error bars, and can be justified by the particularities of each set of data, as discussed below.

The 2010 event, for example, had only seven contacts; none the less, they were well distributed over Ceres' disc (see Fig. \ref{Fig:Ceres-2010-body}) acting as a constraint to its shape. On the other hand, the 2013 event had five more exploitable contacts, but they were concentrated in certain regions of the body. In particular, the absence of chords close to Ceres' south pole made its oblateness worse determined here than in the 2010 event.

However, even our best measurement for the oblateness, $\epsilon=0.08 \pm 0.03$, has still a higher uncertainty with regard to other figures published in the literature, as Table~\ref{Tab: Ceres-final} shows. A larger number of uniformly spaced chords would be necessary to offer a best constraint to the oblateness.

The few chords of the 2010 occultation could themselves only constraint the position angle of the object to a uncertainty of $10\degr$. This uncertainty was reduced by a factor of two in the 2013 event, approaching --and verifying-- the result predicted by the work of \cite{Drummond2014}. As was shown, using the coordinates of Ceres' polar axis to limit the position angle was not an efficient procedure in the 2013 event, in the sense that it did not result in significant changes in the parameters obtained in the nominal solution (see Table \ref{Tab: results}).

On the other hand, constrain the position angle on the 2010 occultation was proved to reduce the error bars of the other parameters (disregarding oblateness). Moreover, this procedure resulted in an excellent agreement between the equatorial radius figures of both events.

The 2013 occultation, therefore, offers not only an independent verification of the figures resulted from the 2010 event, but also validates the procedure carried out there which led to the best-constrained parameters in this work.

Comparison of Ceres' equatorial diameter as measured by different techniques is carried out in Table \ref{Tab: Ceres-final}. We note an overall agreement between our result to those obtained via direct imaging by the Hubble Space Telescope (HST) \citep{Thomas2005}, the Keck Observatory and the ESO VLT \citep{Drummond2014}. The smaller figure reported by \cite{Carry2008} may be justified by the fact that in this study the effects of limb-darking were not taken into account, as pointed out by \cite{Drummond2014}.

As mentioned in the Introduction, the 1984 event \citep{Millis1987} is the only occultation data to which we can compare our result. The measured diameters do not agree, ours being larger by 2-sigma. It is difficult to state for sure the reasons for this divergence since we had no access to the original 1984 data or reduction methodology. We note, however, that in this occultation a variation of 0.5 s in the contact times correspond to almost 7 km on Ceres' limb, which is on the same order of the residuals of their elliptical fit. Therefore, in order to properly compare their result to ours it would be necessary to reduce their light curves with the same methodology used in the present work.

NASA's \textit{Dawn} mission shall bring light to these questions, which will be important not only for the knowledge of Ceres itself, but also for all the techniques used so far to study the physical properties of small Solar System objects, such as the stellar occultations.

\begin{table}
  \caption{Ceres' equatorial diameter and oblateness \label{Tab: Ceres-final}}
  \begin{centering}
  \begin{tabular}{@{}lccc}
  \hline
     Eq. diameter (km) & Oblateness & Method & Ref. \\
\hline
$972 \pm 6$  & $0.08  \pm 0.03$  & Occultation & 1\\
$967 \pm 10$ & $0.078 \pm 0.015$ & Keck+VTL    & 2 \\
$959 \pm 5$  & $0.074 \pm 0.007$ & Keck        & 3\\
$975 \pm 4$  & $0.067 \pm 0.005$ & HST         & 4\\
$959 \pm 5$  & $0.05  \pm 0.01$  & Occultation & 5\\
\hline
\end{tabular}
\par\end{centering}
\textbf{References.} 1: Present work. 2: \cite{Drummond2014}. 3: \cite{Carry2008}. 4: \cite{Thomas2005}. 5: \cite{Millis1987}.
\end{table}

\section*{Acknowledgements}

ARGJ thanks the financial support of CAPES. BLG thanks CNPq. F.B.R. acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014 and CDFB-CAPES/Brazil. M.A. acknowledges CNPq grants 473002/2013-2, 482080/2009-4 and 312394/2014-4, and FAPERJ grant 111.488/2013. RVM thanks grants: CNPq-306885/2013, Capes/Cofecub-2506/2015, Faperj/PAPDRJ-45/2013. J.I.B. Camargo acknowledges CNPq for a PQ2 fellowship (process number 308489/2013-6). We also acknowledge Steve Preston for the predictions of the occultations.

\begin{thebibliography}{99}

\bibitem[\protect\citeauthoryear{Assafin et al.}{2011}] {2011gfun.conf...85A} Assafin M., et al., 2011, in Tanga P., Thuillot W., eds, Gaia follow-up network for the solar system objects: Gaia FUN-SSO workshop proceedings. IMCCE-Paris Observatory, Paris, p. 85

\bibitem[\protect\citeauthoryear{Braga-Ribas et al.}{2013}]{BragaRibas2013} Braga-Ribas F. et al., 2013,
ApJ, 773, 26

\bibitem[\protect\citeauthoryear{Braga-Ribas et al.}{2014}]{BragaRibas2014} Braga-Ribas F. et al., 2014,
Nature, 508, 72

\bibitem[\protect\citeauthoryear{Carry et al.}{2008}] {Carry2008} Carry B. et al., 2008,
A\&A, 478, 235

\bibitem[\protect\citeauthoryear{Castillo-Rogez}{2011}]{CastilloR2011} Castillo-Rogez J.~C.,
2011, Icarus, 215, 599

\bibitem[\protect\citeauthoryear{Desmars et al.}{2015}]{Desmars2015} Desmars J. et al., 2015, A\&A, submitted

\bibitem[\protect\citeauthoryear{Drummond et al.}{2014}]{Drummond2014} Drummond J.D. et al., 2014,
Icarus, 236, 28

\bibitem[\protect\citeauthoryear{Dunham et al.}{2014}]{Dunham2014}
Dunham D.W. et al., 2014, Asteroid Occultations V12.0. EAR-A-3-RDR-OCCULTATIONS-V12.0. NASA Planetary Data System

\bibitem[\protect\citeauthoryear{Gomes et al.}{2005}]{Gomes05} Gomes R., Levison H.~F., Tsiganis K., Morbidelli A.,\ 2005, Nature, 435, 466

\bibitem[\protect\citeauthoryear{Giorgini et al.}{1996}]{Giorgini1996} Giorgini, J. D. et al., 1996, BAAS, 28, 1158

%\bibitem[\protect\citeauthoryear{Hog et al.}{2000}] {Hog2000} Hog E. et al., 2000,
%A\&A, 355, L27

\bibitem[\protect\citeauthoryear{Johnson \& McGetchin}{1973}] {Johnson1973} Johnson T.V., McGetchin T.R., 1973,
Icarus, 18, 612

\bibitem[\protect\citeauthoryear{K\"{u}ppers et al.}{2014}]{Kuppers2014} K\"{u}ppers M. et al., 2014,
Nature, 505, 525

\bibitem[\protect\citeauthoryear{McKinnon}{2012}]{McKinnon12} McKinnon W.~B.,\ 2012, LPI 
Contributions, 1667, 6475

\bibitem[\protect\citeauthoryear{Millis et al.}{1987}]{Millis1987} Millis R.L. et al., 1987,
Icarus, 72, 507

%\bibitem[\protect\citeauthoryear{Ortiz et al.}{2012}]{Ortiz2012} Ortiz J.L. et al., 2012,
%Nature, 491, 566

\bibitem[\protect\citeauthoryear{Russell et al.}{2004}]{Russell2004} Russell C.~T. et al.,\ 2004, P\&SS, 52, 465

\bibitem[\protect\citeauthoryear{Sicardy et al.}{2011}]{Sicardy2011} Sicardy B. et al., 2011,
Nature, 478, 493

\bibitem[\protect\citeauthoryear{Thomas et al.}{2005}]{Thomas2005} Thomas P.C. et al., 2005,
Nature, 437, 224

%\bibitem[\protect\citeauthoryear{van Belle}{1999}] {vanBelle1999} van Belle G.T., 1999,
%PASP, 111, 1515

\bibitem[\protect\citeauthoryear{Widemann et al.}{2009}] {Widemann2009} Widemann T. et al., 2009,
Icarus, 199, 458

%\bibitem[\protect\citeauthoryear{Zacharias et al.}{2004}]{Zacharias2004} Zacharias N. et al., 2004,
%AJ, 127, 3043

\bibitem[\protect\citeauthoryear{Zacharias et al.}{2013}]{Zacharias2013} Zacharias N. et al., 2013,
AJ, 145, 44

\end{thebibliography}

%\bsp

\label{lastpage}

\end{document}

### 2015-05-16 Irregulares/AA-2015-26273/Irregulars.tex
Adicionado Artigos dos irregulares e de Ceres

%% LyX 2.0.0 created this file.  For more info, see http://www.lyx.org/.
%% Do not edit unless you really know what you are doing.
\documentclass[english,traditabstract, referee]{aa}
\usepackage[varg]{txfonts}
%\usepackage[english,brazilian]{babel}
\usepackage[T1]{fontenc}
\setcounter{tocdepth}{3}
\usepackage{prettyref}
\usepackage{array}
\usepackage{rotating}
\usepackage{float}
\usepackage{multirow}
\usepackage{amstext}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{subfigure}
\usepackage{gensymb}
\usepackage{longtable}
%\usepackage{setspace}
%
%\usepackage{etoolbox}
%\AtBeginEnvironment{longtable}{\singlespacing}
%\setlength{\LTcapwidth}{\linewidth}

\makeatletter

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
%% Because html converters don't know tabularnewline
\providecommand{\tabularnewline}{\\}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
\usepackage{natbib,twoopt}
\usepackage[breaklinks=true]{hyperref} %% to avoid \citeads line fills
\usepackage{arydshln}
\bibpunct{(}{)}{;}{a}{}{,} %% natbib format for A&A and ApJ
\makeatletter
\newcommandtwoopt{\citeads}[3][][]{\href{http://adsabs.harvard.edu/abs/#3}%
{\def\hyper@linkstart##1##2{}%
\let\hyper@linkend\@empty\citealp[#1][#2]{#3}}}
\newcommandtwoopt{\citepads}[3][][]{\href{http://adsabs.harvard.edu/abs/#3}%
{\def\hyper@linkstart##1##2{}%
\let\hyper@linkend\@empty\citep[#1][#2]{#3}}}
\newcommandtwoopt{\citetads}[3][][]{\href{http://adsabs.harvard.edu/abs/#3}%
{\def\hyper@linkstart##1##2{}%
\let\hyper@linkend\@empty\citet[#1][#2]{#3}}}
\newcommandtwoopt{\citeyearads}[3][][]%
{\href{http://adsabs.harvard.edu/abs/#3}
{\def\hyper@linkstart##1##2{}%
\let\hyper@linkend\@empty\citeyear[#1][#2]{#3}}}
\makeatother
\titlerunning{Astrometric Positions of the Irregular Satellites of Giant Planets}
\authorrunning{Gomes-Júniot et al.}

\makeatother

\begin{document}

\title{Astrometric positions for 18 irregular satellites of giant planets from 23 years of observations
\fnmsep\thanks{The complete version of Table 8 is available through CDS and IAU NSDC data base at www.imcce.fr/nsdc.}
\fnmsep\thanks{Partially based on observations made at Laborat\'orio Nacional de Astrof\'{\i}sica (LNA), Itajub\'a-MG, Brazil.}
\fnmsep\thanks{Partially based on observations through the ESO runs 079.A-9202(A), 075.C-0154, 077.C-0283 and 079.C-0345.}
\fnmsep\thanks{\textbf{Partially based on observations made at
Observatoire de Haute Provence (OHP), F-04870 Saint-Michel l'observatoire, France}}
}

\author{ A. R. Gomes-Júnior
          \inst{1},
          M. Assafin 
          \inst{1} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France},
          R. Vieira-Martins
          \inst{1,2,3} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France},
          J.-E. Arlot
          \inst{4},      
          J. I. B. Camargo
          \inst{2,3},     
          F. Braga-Ribas
          \inst{2,5}, 
          D. N. da Silva Neto
          \inst{6},
          A. H. Andrei
          \inst{1,2} \fnmsep\thanks{Affiliated researcher at the Observatoire de Paris/SYRTE, 77 av. Denfert Rochereau, 75014 Paris, France},
          A. Dias-Oliveira
          \inst{2},
          B. E. Morgado
          \inst{1},
          G. Benedetti-Rossi
          \inst{2},
          Y. Duchemin
          \inst{4,7},
          J. Desmars
          \inst{4},
          V. Lainey
          \inst{4},
          W. Thuillot
          \inst{4}
          }

\offprints{A. R. Gomes-Júnior}

   \institute{Observatório do Valongo/UFRJ, Ladeira Pedro Antônio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
              \email{altair08@astro.ufrj.br}
              \and
Observatório Nacional/MCT, R. General José Cristino 77,
CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
              \email{rvm@on.br}
              \and
Laboratório Interinstitucional de e-Astronomia - LIneA, Rua Gal. José Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil
              \and
Institut de mécanique céleste et de calcul des éphémérides - Observatoire de Paris, UMR 8028 du CNRS,
77 Av. Denfert-Rochereau, 75014 Paris, France\\
              \email{arlot@imcce.fr}
              \and
Federal University of Technology - Paraná (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
              \and
Centro Universitário Estadual da Zona Oeste, Av. Manual Caldeira de Alvarenga 1203, CEP 23.070-200 Rio de Janeiro RJ, Brazil
              \and
              ESIGELEC-IRSEEM, Technopôle du Madrillet, Avenue Galilée, 76801 Saint-Etienne du Rouvray, France
              }

\date{Received ; accepted }

\abstract
{The irregular satellites of the giant planets are believed to have been captured during the evolution of the solar system. Knowing their physical parameters, such as size, density and albedo is important to constrain where they came from and how they were captured. The best way to obtain these parameters are observations in situ by spacecrafts or from stellar occultations by the objects. Both techniques demand that the orbits are well known.}
{We aimed to obtain good astrometric positions of irregular satellites in order to improve their orbits and ephemeris.}
{We identified and reduced observations of several irregular satellites from three databases containing more than 8000 images obtained between 1992 and 2014 at three sites (Observatório do Pico dos Dias, Observatoire de Haute-Provence and European Southern Observatory - La Silla).  We used the software PRAIA (Platform for Reduction of Astronomical Images Automatically) to make the astrometric reduction of the CCD frames. The UCAC4 catalogue represented the International Celestial Reference System in the reductions. The identification of the satellites in the frames was done through their ephemerides as determined from the SPICE/NAIF kernels. Some procedures were taken to overcome missing or incomplete information (coordinates, date), mostly for the older images.}
{We managed to obtain more than 6000 positions for 18 irregular satellites, being 12 of Jupiter, 4 of Saturn, 1 of Uranus (Sycorax) and 1 of Neptune (Nereid). For some satellites the number of obtained positions is more than 50\% of that used in earlier orbital numerical integrations.}
{Comparison of our positions with recent JPL ephemeris suggests the presence of systematic errors in the orbits for some of the irregular satellites. The most evident case was an error in the inclination of Carme.}

\keywords{Astrometry - Planets and satellites: general - Planets and satellites: individual: Jovian and Saturnian irregular satellites}

\maketitle

\section{Introduction} \label{Sec: introducao}

The irregular satellites of the giant planets are smaller than the regular moons, having more eccentric, inclined, distant and, in most cases, retrograde orbits. Due to their orbital configurations, it is largely accepted that these objects were captured in the early solar system \citep{Sheppard2003}.

Because they are faint, the majority of these objects was discovered only in the last century\footnote{Website: http://ssd.jpl.nasa.gov/?sat\_discovery} . They were never visited by a spacecraft, with the exception of Himalia, Phoebe and Nereid, in a flyby by the Cassini space probe in 2000 for Himalia \citep{Porco2003} and in 2004 for Phoebe \citep{Desmars2013} and in a flyby by the Voyager 2 space probe in 1989 for Nereid \citep{Smith1989}. Even in situ, they were still opportunity target observations resulting in not optimal measurements, with size errors of $10 km$ for Himalia and $25 km$ for Nereid \citep{Thomas1991}. The exception is Phoebe with a very accurate measurement of size with a mean radius error of $0.7 km$ \citep{Thomas2010}.

%There is a number of capture mechanisms of objects by giant planets proposed in the literature. There is the Gas Drag in the primordial circumplanetary nebulae \citep{Sheppard2006} where the object would be affected by the gas drag and its velocity slowed down until it be captured by the planet. Another mechanism is called pull-down capture \citep{Sheppard2006}, where the mass of the planet would increase while the object was temporarily captured.

%A mechanism based in the Nice model \citep{Morbidelli2005, Tsiganis2005, Gomes2005} was proposed by \citealp{Nesvorny2007} and, in the specific case of Jupiter with the modern Nice model, by \citealp{Nesvorny2014}. During the early solar system instability, encounters between the outer planets occurred. These planetary encounters could exchange energy and angular momentum between planets and the objects nearby making it possible for the capture of irregular bodies by the giant planets. In this scenario, the survival rate of prior-LHB (Late Heavy Bombardment) satellites is very small.

%Another important mechanism is the capture through collisional interactions \citep{Sheppard2006}. A collision between two small bodies in the Hill's sphere of the planet could generate fragmented objects and the dissipated energy could be such that some of these objects could be captured.

%Some of these objects are in dynamical groups with similar orbital elements, called families, similar to families found in the Main Asteroid Belt. These families may have been created by a parent body disrupted by collisions with comets or other satellites \citep{Nesvorny2004}. Collisions with comets are more likely to have occurred during the Late Heavy Bombardment (LHB) \citep{Gomes2005}.

%\citealp{Nesvorny2003} studied the collision rates between irregular satellites and concluded that some satellites could have been removed by collision with a bigger satellite. The rate collision between satellites of the Himalia Group (Himalia, Elara, Lysithea and Leda, mainly), for instance, was found to be more than one during the solar system age suggesting that their current structure was originated by satellite-satellite collision.

%For Phoebe, ejected material from its surface caused by impacts could evolve due to Poynting-Robertson drag and collide with Iapetus causing the large variation in albedo observed on it \citep{Nesvorny2003}. Indeed, Cassini was able to detected in Phoebe an absorption feature at 2.42 $\mu m$ (probably CN combinations) that was also detected in the dark side of Iapetus \citep{Clark2005}.

If these objects were captured, there remains the question of where they came from. \citealp{Clark2005} showed from imaging spectroscopy from Cassini that Phoebe has a surface probably covered by material from the outer solar system and \citealp{Grav2003} showed that the satellites of the Jovian Prograde Group Himalia have grey colors implying that their surfaces are similar to that of C-type asteroids. In that same work, the Jovian Retrograde Group Carme was found to have surface colors similar to the D-type asteroids like Hilda or Trojan families while JXIII Kalyke has a redder color like Centaurs or trans-neptunian objects (TNOs).

For Saturnian satellites, \citealp{Grav2007} showed by their colors and spectral slopes that these satellites contain a more or less equal fraction of C-, P- and D-like objects but SXXII Ijiraq is marginally redder than D-type objects. These works may suggest different origins for the irregular satellites.

In this context, we used 3 databases for deriving precise positions for the irregular satellites observed at Observatório do Pico dos Dias (1.6 m and 0.6 m telescopes, IAU code 874), Observatoire Haute-Provence (1.2m telescope, IAU code 511) and ESO (2.2 m telescope, IAU code 809). Many irregular satellites were observed between 1992 and 2014 covering a few orbital periods of these objects (12 satellites of Jupiter, 4 of Saturn, Sycorax of Uranus and Nereid of Neptune).

\textbf{Since their ephemerides are not very precise, predict and observe stellar occultations are very difficult and no observation of such an event for an irregular satellite is found in the literature. The precise star positions to be derived by the ESA astrometry satellite Gaia \citep{deBruijne2012} will render better predictions with the only source of error being the ephemeris. The positions derived from \textbf{our} observations can be used in new orbital numerical integrations, generating more precise ephemerides.}

\textbf{The power of stellar occultations for observing relatively small diameter solar system objects is supported by recent works such as the discovery of a ring system around the Centaur (10199) Chariklo \citep{Braga-Ribas2014}. Once irregular satellites start to be observed by this technique, it will be possible to obtain their physical parameters (shape, size, albedo, density) with unprecedented precision. For instance, in this case, sizes could be obtained with kilometer accuracy. The knowledge of these parameters would in turn bring valuable information for the study of the capture mechanisms and origin of the irregular satellites.}

The databases are described in Sect. \ref{Sec: observations}. The astrometric procedures in Sect. \ref{Sec: reduction}. The obtained positions are presented in Sect \ref{Sec: positions} and analysed in Sect. \ref{Sec: comparison}. Conclusions are given in Sect. \ref{Sec: conclusions}.

\section{Databases} \label{Sec: observations}

Our three databases consist in optical CCD images from many observational programs performed with different telescopes/detectors targeting a variety of objects, among which irregular satellites. The observations were made at 3 sites: Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO). Altogether there are more than 8000 FITS images obtained in a large time span (1992-2014) for the irregular satellites. Since the OHP and mostly the OPD database registers were not well organized, we had to start from scratch and develop an automatic procedure to identify and filter only the images of interest, that is, of the irregular satellites. The instruments and images characteristics are described in the following subsections.

\subsection{OPD} \label{Subsec: observations-opd}

The OPD database was produced at Observatório do Pico dos Dias (OPD, IAU code 874, $45^{\circ} ~34\arcmin ~57\arcsec$ W, $22^{\circ} ~32\arcmin ~04\arcsec$ S, 1864 m)\footnote{Website: http://www.lna.br/opd/opd.html - in Portuguese}, located at geographical longitude , in Brazil. The observations were made between 1992 and 2014 by our group in a variety of observational programs. %In Fig \ref{Fig: obsxyear-opd} we plot the number of frames obtained per satellite over time and in Fig \ref{Fig: obsxsat-opd} the number of frames per satellite for each telescope. 
Two telescopes of 0.6 m diameter (Zeiss and Boller \& Chivens) and one 1.6 m diameter (Perkin-Elmer) were used for the observations. It was identified 5248 observations containing irregular satellites, being 3168 from the Boller \& Chivens, 1967 from the Perkin-Elmer and 113 from the Zeiss.

%\begin{figure*}
%\begin{centering}
%\includegraphics[width=17cm]{figures/framexobj_ano_opd}
%\par\end{centering}
%\caption{Distribution of the observations of the satellites over time at OPD. The graphic is limited in number of frames to 300 for better viewing. The number of frames of Phoebe in 2010 is 533.}
%\label{Fig: obsxyear-opd}
%\end{figure*}

This is an inhomogeneous database with observations made with 9 different detectors (see Table \ref{Tab: OPD-CCDs}) and 6 different filters. The headers of most of the older FITS images had missing, incomplete or incorrect coordinates or date. In some cases, we could not identify the detector origin. The procedures used to overcome these problems are described in Sect. \ref{Sec: reduction}.

\begin{table}[h!]
\caption{\label{Tab: OPD-CCDs} Characteristics of OPD detectors used in this work.}
\begin{centering}
\begin{tabular}{ccc}
\hline
\hline
\multicolumn{3}{c}{Perkin-Elmer}\tabularnewline
% & \multicolumn{2}{c}{Exposure Times}\tabularnewline
Detector & Image Size (pixel) & Pixel Scale ($\mu m$/px)\tabularnewline
\hline
CCD048 & 770 x 1152 & 22.5\tabularnewline
CCD098 & 2048 x 2048 & 13.5\tabularnewline
CCD101 & 1024 x 1024 & 24.0\tabularnewline
CCD105 & 2048 x 2048 & 13.5\tabularnewline
CCD106 & 1024 x 1024 & 24.0\tabularnewline
CCD301 & 385 x 578 & 22.0\tabularnewline
CCD523 & 455 x 512 & 19.0\tabularnewline
IKON & 2048 x 2048 & 13.5\tabularnewline
IXON & 1024 x 1024 & 13.5\tabularnewline
\hline 
\end{tabular} 
\par\end{centering}
The plate scale of the telescopes are 13.09"/mm for Perkin-Elmer, 25.09"/mm for Boller \& Chivens and 27.5"/mm for Zeiss.
\end{table}

%\begin{figure}
%\begin{centering}
%\includegraphics[scale=0.45]{figures/framexobj_opd}
%\par\end{centering}
%
%\caption{Number of frames observed per satellite by OPD telescope.}
%
%
%\label{Fig: obsxsat-opd}
%\end{figure}

\subsection{OHP} \label{Subsec: observations-ohp}

The instrument used at the Observatoire de Haute Provence (OHP, IAU code 511, $5^{\circ} ~42\arcmin ~56.5\arcsec$ E, $43^{\circ} ~55\arcmin ~54.7\arcsec $N, 633.9 m)\footnote{Website: www.obs-hp.fr/guide/t120.shtml - in French} was the 1.2m-telescope in a Newton configuration. The focal length is 7.2 m. The observations were made between 1997 and 2008. During this time only one CCD detector $1024 \times 1024$ was used. The size of field is $12\arcmin \times 12\arcmin$ with a pixel scale of $0.69\arcsec$.  %Fig. \ref{Fig: obsxyear-ohp} shows the distribution of the observation of the satellites over time and Fig. \ref{Fig: obsxsat-ohp} the number of frames observed for each satellite. 
From these observations, 2408 were identified containing irregular satellites.

%\begin{figure*}
%\begin{centering}
%\includegraphics[scale=0.48]{figures/framexobj_ano_ohp}
%\par\end{centering}
%
%\caption{Distribution of the observations of the satellites over time from observations at OHP.}
%\label{Fig: obsxyear-ohp}
%\end{figure*}
%
%\begin{figure}
%\begin{centering}
%\includegraphics[scale=0.45]{figures/framexobj_ohp}
%\par\end{centering}
%
%\caption{Number of frames per satellite observed at OHP.}
%
%
%\label{Fig: obsxsat-ohp}
%\end{figure}

\subsection{ESO} \label{Subsec: observations-eso}

Observations were made at the 2.2 m Max-Planck ESO (ESO2p2) telescope (IAU code 809, $70\degr44\arcmin1.5\arcsec$ W, $29\degr15\arcmin31.8\arcsec$ S, 2345.4 m)\footnote{Website: www.eso.org/sci/facilities/lasilla/telescopes/\\national/2p2.html} with the Wide Field Imager (WFI) CCD mosaic detector. Each mosaic is composed by eight CCDs of $7.5\arcmin \times 15\arcmin$ ($\alpha$, $\delta$) sizes, resulting in a total coverage of $30\arcmin \times 30\arcmin$ per mosaic. Each CCD has $4 k \times 2 k$ pixels with a pixel scale of $0.238\arcsec$. The filter used was a broad-band R filter (ESO\#844) with $\lambda _{c}  = 651.725$ nm and $\Delta \lambda = 162.184$ nm. The telescope was shifted between exposures in such a way that each satellite was observed at least twice in different CCDs.

The satellites were observed in 24 nights, divided in 5 runs, between April 2007 and May 2009 in paralel with, and using the same observational and astrometric procedures of the program that observed stars along the sky path of trans-neptunian objects (TNOs) to identify candidates to stellar occultation \citep[see][]{2010A&A...515A..32A, 2012A&A...541A.142A}. A total of 810 observations for irregular satellites were obtained. %Fig \ref{Fig: obsxsat-eso} shows the number of frames per satellite.

%\begin{figure}
%\begin{centering}
%\includegraphics[scale=0.45]{figures/framexobj_eso}
%\par\end{centering}
%
%\caption{Number of frames per satellite observed at ESO.}
%
%
%\label{Fig: obsxsat-eso}
%\end{figure}

\section{Astrometry} \label{Sec: reduction}

Almost all the frames were photometrically calibrated with auxiliary bias and flat-field frames by means of standard procedures using IRAF\footnote{Website: http://iraf.noao.edu/} and, for the mosaics, using the esowfi \citep{Jones2000} and mscred \citep{Valdes1998} packages. Some of the nights at OPD didn't have bias and flat-field images so the correction was not possible.

The astrometric treatment was made with the Platform for Reduction of Astronomical Images Automatically (PRAIA) \citep{2011gfun.conf...85A}. The (x, y) measurements were performed with 2-dimensional circular symmetric Gaussian fits within one Full Width Half Maximum (FWHM = seeing). Within one FWHM, the image profile is well described by a Gaussian profile, free from the wing distortions, which may jeopardize the center determination. PRAIA automatically recognizes catalog stars and determines ($\alpha$, $\delta$) with a user-defined model relating the (x, y) measured and (X, Y) standard coordinates projected in the sky tangent plane.

We used the UCAC4 \citep{2013AJ....145...44Z} as the practical representative of the International Celestial Reference System (ICRS). For each frame, we used the six constants polynomial model to relate the (x, y) measurements with the (X, Y) tangent plane coordinates. For ESO, we followed the same astrometric procedures described in detail in \cite{2012A&A...541A.142A}; the (x, y) measurements of the individual CCDs were pre-corrected by a field distortion pattern, and all positions coming from different CCDs and mosaics were then combined using a 3rd degree polynomial model to produce a global solution for each night and field observed, and final ($\alpha$, $\delta$) object positions were obtained in the UCAC4 system.

In Table \ref{Tab: stars-errors} we list the average mean error in $\alpha$ and $\delta$ for the reference stars obtained by telescope, the average (x, y) measurement errors of the Gaussian fits described above and the mean number of UCAC4 stars used by frame. For all databases, about 20\% of outlier reference stars were eliminated for presenting (O-C) position residuals higher than 120 mas in the ($\alpha$, $\delta$) reductions. %##### OHP = 21\% de estrelas eliminadas, erro gaussiano: média de 26mas e 90\% entre 5 e 60mas para x e y; ESO = erro gaussiano: média de 14mas e 90\% entre 2 e 41mas; 160 = 13\% de estrelas eliminadas,  média de 15mas e 90\% entre 3 e 38mas ; IAG = 24\% d estrelas eliminadas, média de 29mas e 90\% entre 3 e 75mas ; Zeiss = 37\% de estrelas eliminadas, média de 26mas e 90\% entre 5 e 56mas#####

\begin{table}
\caption{\label{Tab: stars-errors} Astrometric ($\alpha$, $\delta$) reduction by telescope.}
\begin{centering}
\begin{tabular}{lccccc}
\hline %\multicolumn{6}{c}{Perkin-Elmer} \tabularnewline
\hline
 & \multicolumn{2}{c}{Mean errors} &  UCAC4 & \multicolumn{2}{c}{Gaus. errors}       \tabularnewline
Telescope  & $\sigma_{\alpha}$  & $\sigma_{\delta}$  & stars & x & y  \tabularnewline
  &   mas   &   mas & &   mas   &   mas  \tabularnewline
\hline
PE(OPD) & 51 & 48 & 24 & 15 & 15 \tabularnewline
B\&C (OPD) & 56 & 55 & 36 & 29 & 29 \tabularnewline
Zeiss (OPD) & 58 & 57 & 95 & 26 & 26 \tabularnewline
OHP & 50 & 49 & 46 & 26 & 26 \tabularnewline
ESO & 26 & 25 & 632 & 15 & 15 \tabularnewline
\hline
\end{tabular}
\par\end{centering}
Mean errors are the standard deviations in the (O$-$C) residuals from ($\alpha$, $\delta$) reductions with the UCAC4 catalog. Gaussian errors are the errors in the Gaussian fit used to perform the (x, y) measurements.
\end{table}

To help identifying the satellites in the frames, and derive the ephemeris for the instants of the observations for comparisons (see Sect \ref{Sec: comparison}), we used the kernels from SPICE/JPL\footnote{Website: http://naif.jpl.nasa.gov/naif/toolkit.html}. \cite{Emelyanov2008} and references therein also provided ephemeris of similar quality for the irregular satellites. For instance, for Himalia, which has relatively good orbit solutions, the ephemerides differ by less than $20 mas$, and in the case of less-known orbits, like Ananke, the differences are less than $90 mas$. Here, we choose to use the JPL ephemeris because they used more recent observations \cite[see][]{Jacobson2012}. The JPL ephemeris that represented the Jovian satellites in this work was the DE421 $+$ JUP300. For the Saturnian satellites the ephemeris was DE421 $+$ SAT359 to Hyperion, Iapetus and Phoebe and DE421 $+$ SAT361 to Albiorix, Siarnaq and Paaliaq. The DE421 $+$ URA095 was used for Sycorax and DE421 $+$ NEP081 for Nereid. More recent JPL ephemeris versions became available after completion of this work, but this did not affect the results.

In the OPD database, there were some images (mostly the older ones) with missing coordinates or wrong date in their headers. In the case of missing or wrong coordinates, we adopted the ephemeris as the central coordinates of the frames. When the time was not correct, the FOV identification failed. In this case, a search for wrong date (year) displaying was performed. Problems like registering local time instead of UTC were also identified and corrected.

In all databases, for each night a sigma-clipping procedure was performed to eliminate discrepant positions (outliers). A threshold of 120 mas and a deviation of more than 2.5 sigmas from the nightly average ephemeris offsets were adopted.

From Table \ref{Tab: Reductions-160} to \ref{Tab: Reductions-eso} we list the average dispersion (standard deviation) of the position offsets with regard to the ephemeris for $\alpha$ and $\delta$ obtained by telescope for each satellite. The final number of frames, number of nights (in parenthesis), the mean number of UCAC4 stars used in the reduction and the approximate V magnitude are also given. The dashed lines separate the satellites from different families with similar orbital parameters: Himalia Group (Himalia, Elara, Lysithea and Leda), Pasiphae Group (Pasiphae, Callirrhoe and Megaclite) and Ananke Group (Ananke and Praxidike). Carme and Sinope are the only samples of their groups. From Saturn, Siarnaq and Paaliaq are from the Inuit Group while Phoebe and Albiorix are the only samples of their groups.

The differences in the dispersion of the ephemeris offsets of the same satellite for distinct telescopes seen in Tables \ref{Tab: Reductions-160} to \ref{Tab: Reductions-eso} are caused by the different distribution of observations along the orbit for each telescope. This can be seen in Fig \ref{Fig: carme_anom} for Carme, \ref{Fig: pasiphae_anom} for Pasiphae and for all objects in the online material. Since the observations cover different segments of the orbit, the dispersion of the offsets may vary for different telescopes for a single satellite, with larger covered segments usually implying in larger dispersions and vice-versa. For Nereid, due to its high eccentric orbit, the observations are located between 90$\degree$ and 270$\degree$ of True Anomaly where Nereid remains most of the time.

\begin{table}
\caption{\label{Tab: Reductions-160} Astrometric ($\alpha$, $\delta$) reduction for each satellite observed with the Perkin-Elmer telescope.}
\begin{centering}
\begin{tabular}{lcccccc}
\hline 
%\multicolumn{6}{c}{Perkin-Elmer} \tabularnewline
\hline
 & \multicolumn{2}{c}{Offsets (sigma)} & Nr &  UCAC4 & \tabularnewline
Satellite  & $\sigma_{\alpha}$  & $\sigma_{\delta}$ & frames  & stars & Mag \tabularnewline
  &   mas   &   mas  & (nights)  & &\tabularnewline
\hline
Himalia & 290 & 45 & 238 (18) & 37 & 14 \tabularnewline
Elara & 230 & 118 & 99 (12) & 32 & 16\tabularnewline
Lysithea & 107 & 79 & 53 (8) & 41 & 18\tabularnewline
Leda & 207 & 79 & 6 (2) & 46 & 19\tabularnewline
\hdashline
Pasiphae & 157 & 92 & 144 (13) & 22 & 17 \tabularnewline
Callirrhoe & 66 & 35  &   9 (1) & 3 & 21\tabularnewline
\hdashline
Carme & 97 & 94 & 68 (7) & 49 & 18\tabularnewline
Sinope & 155 & 77 & 37 (8) & 42 & 18\tabularnewline
Ananke & 93 & 185 & 52 (7) & 40 & 19\tabularnewline
\hline
Phoebe & 73 & 95 & 410 (22) & 6 & 16\tabularnewline
\hline
Nereid & 200 & 142 & 289 (29) & 8 & 19\tabularnewline
\hline
\end{tabular}
\par\end{centering}
The offsets (sigma) are the average standard deviations of the ephemeris offsets from the ($\alpha$, $\delta$) positions of the satellites. Also given are the approximate satellite V magnitude and the average number of UCAC4 reference stars per frame.
\end{table}

\begin{table}
\caption{\label{Tab: Reductions-iag} Astrometric ($\alpha$, $\delta$) reduction for each satellite observed with the Boller \& Chivens telescope.}
\begin{centering}
\begin{tabular}{lccccc}
\hline 
%\multicolumn{6}{c}{Boller \& Chivens} \tabularnewline
\hline
 & \multicolumn{2}{c}{Offsets (sigma)} & Nr &  UCAC4 & \tabularnewline
Satellite  & $\sigma_{\alpha}$  & $\sigma_{\delta}$ & frames  & stars & Mag \tabularnewline
  &   mas   &   mas  & (nights)  & &\tabularnewline
\hline
Himalia & 83 & 43 & 560 (31) & 57 & 14 \tabularnewline
Elara & 55 & 43 & 294 (23) & 53 & 16 \tabularnewline
Lysithea & 23 & 42 & 7 (2) & 60 & 18 \tabularnewline
\hdashline
Pasiphae & 128 & 71 & 140 (14) & 57 & 17 \tabularnewline
Carme & 68 & 111 & 22 (4) & 45 & 18\tabularnewline
Sinope & 59 & 17  &   4 (1) & 22 & 18\tabularnewline
\hline
Phoebe & 43 & 48 & 810 (42) & 17 & 16 \tabularnewline
\hline
Nereid & 61 & 45 & 514 (38) & 20 & 19 \tabularnewline
\hline
\end{tabular}
\par\end{centering}
Same as in Table \ref{Tab: Reductions-160}.
%Mean errors are the standard deviations in the (O$-$C) residuals from ($\alpha$, $\delta$) reductions with the UCAC4 catalog.
\end{table}

\begin{table}
\caption{\label{Tab: Reductions-zei} Astrometric ($\alpha$, $\delta$) reduction for each satellite observed with the Zeiss telescope.}
\begin{centering}
\begin{tabular}{lccccc}
\hline 
%\multicolumn{6}{c}{Zeiss} \tabularnewline
\hline
 & \multicolumn{2}{c}{Offsets (sigma)} & Nr &  UCAC4 & \tabularnewline
Satellite  & $\sigma_{\alpha}$  & $\sigma_{\delta}$ & frames  & stars & Mag \tabularnewline
  &   mas   &   mas  & (nights)  & &\tabularnewline
\hline
Himalia & 112 & 72 & 56 (4) & 91 & 14\tabularnewline
Elara & 17 & 21  &  10 (1) & 146 & 16 \tabularnewline
\hdashline
Pasiphae & 24 & 25  &  11 (1) & 140 & 17\tabularnewline
\hline
Phoebe & 37 & 30  &  19 (1) & 16 & 16\tabularnewline
\hline
\end{tabular}
\par\end{centering}
Same as in Table \ref{Tab: Reductions-160}.
%Mean errors are the standard deviations in the (O$-$C) residuals from ($\alpha$, $\delta$) reductions with the UCAC4 catalog.
\end{table}

\begin{table}
\caption{\label{Tab: Reductions-ohp} Astrometric ($\alpha$, $\delta$) reduction for each satellite observed with the OHP telescope.}
\begin{centering}
\begin{tabular}{lccccc}
\hline
%\multicolumn{6}{c}{OHP} \tabularnewline
\hline
 & \multicolumn{2}{c}{Offsets (sigma)} & Nr &  UCAC4 & \tabularnewline
Satellite  & $\sigma_{\alpha}$  & $\sigma_{\delta}$ & frames  & stars & Mag \tabularnewline
  &   mas   &   mas  & (nights)  & &\tabularnewline
\hline
Himalia & 49 & 66 & 357 (43) & 49 & 14\tabularnewline
Elara & 52 & 61 & 187 (25) & 37 & 16\tabularnewline
Lysithea & 63 & 50 & 84 (13) & 56 & 18\tabularnewline
Leda & 118 & 33 & 48 (7) & 14 & 19\tabularnewline
\hdashline
Pasiphae & 101 & 75 & 248 (32) & 39 & 17\tabularnewline
Carme & 114 & 96 & 204 (29) & 39 & 18\tabularnewline
Sinope & 196 & 73 & 169 (25) & 43 & 18\tabularnewline
Ananke & 100 & 89 & 141 (20) & 62 & 19\tabularnewline
\hline
Phoebe & 30 & 31 & 516 (63) & 51 & 16\tabularnewline
Siarnaq & 46 & 98 & 20 (6) & 32 & 20\tabularnewline
\hline
\end{tabular}
\par\end{centering}
Same as in Table \ref{Tab: Reductions-160}.
%Mean errors are the standard deviations in the (O$-$C) residuals from ($\alpha$, $\delta$) reductions with the UCAC4 catalog.
\end{table}

\begin{table}
\caption{\label{Tab: Reductions-eso} Astrometric ($\alpha$, $\delta$) reduction for each satellite observed with the ESO telescope.}
\begin{centering}
\begin{tabular}{lccccc}
\hline 
%\multicolumn{6}{c}{ESO} \tabularnewline
\hline
 & \multicolumn{2}{c}{Offsets (sigma)} & Nr &  UCAC4 & \tabularnewline
Satellite  & $\sigma_{\alpha}$  & $\sigma_{\delta}$ & frames  & stars & Mag \tabularnewline
  &   mas   &   mas  & (nights)  & &\tabularnewline
\hline
Himalia & 76 & 74 & 23 (2) & 1153 & 14\tabularnewline
Elara & 112 & 87 & 46 (4) & 1492 & 16\tabularnewline
Lysithea & 76 & 88 & 90 (6) & 695 & 18\tabularnewline
Leda & 60 & 125 & 44 (3) & 632 & 19\tabularnewline
\hdashline
Pasiphae & 70 & 114 & 66 (5) & 836 & 17\tabularnewline
Callirrhoe & 29 & 33  &  16 (1) & 493 & 21\tabularnewline
Megaclite & 52 & 34  &  10 (1) & 445  & 22\tabularnewline
\hdashline
Ananke & 225 & 19 & 57 (3) & 761 & 18\tabularnewline
Praxidike & 7 & 38  &   2 (1) & 1934 & 21\tabularnewline
\hdashline
Carme & 140 & 110 & 37 (4) & 1074 & 18\tabularnewline
Sinope & 339 & 70 & 11 (2) & 1542 & 18\tabularnewline
Themisto & 894 & 28 & 16 (2) & 1232 & 21\tabularnewline
\hline
Phoebe & 102 & 57 & 32 (5) & 312 & 16\tabularnewline
\hdashline
Siarnaq & 86 & 66 & 56 (6) & 283 & 20\tabularnewline
Paaliaq & 301 & 59 & 11 (4) & 382 & 21\tabularnewline
\hdashline
Albiorix & 76 & 50 & 46 (6) & 330 & 20\tabularnewline
\hline
Sycorax & 150 & 82 & 35 (9) & 375 & 21\tabularnewline
\hline
Nereid & 115 & 78 & 99 (12) & 362 & 19\tabularnewline
\hline
\end{tabular}
\par\end{centering}
Same as in Table \ref{Tab: Reductions-160}.
%Mean errors are the standard deviations in the (O$-$C) residuals from ($\alpha$, $\delta$) reductions with the UCAC4 catalog.
\end{table}

%\begin{table*}
%\caption{\label{Tab: offsets} Error by satellite by telescope}
%\begin{centering}
%\begin{tabular}{lccccccccccccccc}
%\hline\hline 
%Satellite & \multicolumn{3}{c}{Perkin-Elmer} & \multicolumn{3}{c}{Boller \& Chivens} & \multicolumn{3}{c}{Zeiss} & \multicolumn{3}{c}{OHP} & \multicolumn{3}{c}{ESO}\tabularnewline
% & $\sigma_\alpha$ & $\sigma_\delta$ & N & $\sigma_\alpha$ & $\sigma_\delta$ & N & $\sigma_\alpha$ & $\sigma_\delta$ & N & $\sigma_\alpha$ & $\sigma_\delta$ & N & $\sigma_\alpha$ & $\sigma_\delta$ & N\tabularnewline
%  & mas & mas & stars & mas & mas & stars & mas & mas & stars & mas & mas & stars & mas & mas & stars\tabularnewline
%\hline
%Ananke & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 18.8 & 70 & 179 & 249 & 1995& 600 & 1951 & 0 \tabularnewline
%Callirrhoe & 20.6 & 9 & 5 & 14 & 2000 & 95 & 1999 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
%Carme & 17.8 & 62 & 245 & 307 & 1995 & 973 & 1938 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
%Elara & 16.6 & 349 & 223 & 572 & 1995 & 1115 & 1905 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
%Himalia & 14.6 & 645 & 444 & 1089 & 1995 & 1757 & 1894 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
%Leda & 20.1 & 12 & 58 & 70 & 1996 & 178 & 1974 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
%Lysithea & 18.3 & 80 & 103 & 183 & 1995 & 431 & 1938 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
%Pasiphae & 16.8 & 216 & 283 & 499 & 1995 & 1629 & 1908 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
%Sinope & 18.2 & 52 & 210 & 262 & 1996 & 854 & 1914 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
%\hline
%Albiorix & 20.5 & - & 2 & 2 & 2001 & 137 & 2000 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
%Phoebe & 16.4 & 937 & 582 & 1519 & 1995 & 3479 & 1898 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
%Siarnaq & 19.9 & - & 25 & 25 & 2001 & 239 & 2000 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
%\hline
%Sycorax & 19.9 & - & 25 & 25 & 2001 & 239 & 2000 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
%\hline
%Nereid & 19.9 & - & 25 & 25 & 2001 & 239 & 2000 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
%\hline 
%\end{tabular} 
%\par\end{centering}
%\end{table*}

No solar phase correction was applied to the positions. For the biggest irregular satellite of Jupiter, Himalia, it was verified that the maximum deviation in the position due to phase angle is 1.94 \textit{mas} using the phase correction described in \cite{Lindegren1977}. For the other satellites, which are smaller objects, this deviation is even smaller. Since our position error is one order of magnitude higher, this effect was neglected.
%Phoebe = 1.32 e Nereid = 1.36

\section{Satellite positions} \label{Sec: positions}

The final set of positions of the satellites consists in 6523 catalogued positions observed between 1992 and 2014 for 12 satellites of Jupiter, 4 of Saturn, 1 of Uranus and 1 of Neptune. The topocentric positions are in the ICRS. The catalogues (one for each satellite) contain epoch of observations, the position error, filter used, estimated magnitude (from PSF fitting) and telescope origin. The magnitude errors can be as high as 1 mag; they are not photometrically calibrated and should be used with care. The position errors were estimated from the dispersion of the ephemeris offsets of the night of observation of each position. Thus, these position errors are probably overestimated, as there must be ephemeris errors present in the dispersion of the offsets. These position catalogues are freely available in electronic form at the CDS (see a sample in Table \ref{Tab: sample-cds}) and at the IAU NSDC data base at www.imcce.fr/nsdc.

The number of positions acquired is significant compared to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012} as shown in Table \ref{Tab: comparison-horizons}.

\begin{table*}
\caption{\label{Tab: sample-cds} CDS data table sample for Himalia.}
\begin{centering}
\begin{tabular}{ccccccccc}
\hline 
%\multicolumn{9}{c}{Himalia} \tabularnewline
\hline
% & \multicolumn{2}{c}{Mean errors} & Nr & Nr &  UCAC4       \tabularnewline
\multicolumn{2}{c}{RA  (ICRS)  Dec} & RA error & Dec error & Epoch & Mag & Filter & Telescope & \textbf{IAU code} \tabularnewline
\multicolumn{1}{l}{\ h \ m \ \ \ s} & \multicolumn{1}{l}{\ \ \ $\degree$ \ \ \ $\arcmin$ \ \ \ $\arcsec$} &   (mas)   &   (mas)  & (jd) & & & & \tabularnewline
\hline
 16 59 11.6508 & -22 00 44.855 &  17 &  12 & 2454147.78241319 &   16.0 &    C &      BC & \textbf{874} \tabularnewline
 16 59 11.6845 & -22 00 44.932 &  17 &  12 & 2454147.78332384 &   15.8 &    C &      BC & \textbf{874} \tabularnewline
 16 59 11.7181 & -22 00 44.978 &  17 &  12 & 2454147.78422477 &   16.0 &    C &      BC & \textbf{874} \tabularnewline
 16 59 11.7818 & -22 00 45.143 &  17 &  12 & 2454147.78602662 &   15.9 &    C &      BC & \textbf{874} \tabularnewline
 16 59 11.8188 & -22 00 45.232 &  17 &  12 & 2454147.78693750 &   16.0 &    C &      BC & \textbf{874} \tabularnewline
 17 17 11.0344 & -22 47 19.415 &  30 &  24 & 2454205.63885463 &   16.1 &        U &      BC & \textbf{874} \tabularnewline
 17 17 11.0270 & -22 47 19.381 &  30 &  24 & 2454205.63959167 &   16.1 &        U &      BC & \textbf{874} \tabularnewline
 17 17 11.0258 & -22 47 19.366 &  30 &  24 & 2454205.64031875 &   16.1 &        U &      BC & \textbf{874} \tabularnewline
 17 17 11.0192 & -22 47 19.417 &  30 &  24 & 2454205.64104583 &   16.1 &        U &      BC & \textbf{874} \tabularnewline
\hline
\end{tabular}
\par\end{centering}
This sample corresponds to 9 observations of Himalia from February 16, 2007 and April 15, 2007. Tables contain the topocentric ICRS coordinates of the irregular satellites, the position error estimated from the dispersion of the ephemeris offsets of the night of observation, \textbf{the UTC time of the frame's mid-exposure in julian date}, the estimated magnitude, the filter used, the telescope origin \textbf{and correspondent IAU code}. The filters may be U, B, V, R or I following the Johnson system; C stands for clear (no filter used), resulting in a broader R band magnitude, RE for the broad-band R filter ESO\#844 with $\lambda_{c} = 651.725$ nm and $\Delta\lambda = 162.184 $ nm (full width at half maximum) and "un" for unknown filter. E, OH, PE, BC and Z stand respectively for the ESO, OHP, Perkin-Elmer, Bollen \& Chivens and Zeiss telescopes.
\end{table*}

\begin{table}
\caption{\label{Tab: comparison-horizons} Comparison of positions obtained with \citealp{Jacobson2012}.}
\begin{centering}
\begin{tabular}{lccccc}
\hline  \hline
 & \multicolumn{4}{c}{Number of Positions}   &    \tabularnewline
Satellite  & OPD  & OHP & ESO & Total  & Jacobson \tabularnewline
\hline
Himalia & 854 & 357 & 23 & 1234 & 1757 \tabularnewline
Elara & 403 & 187 & 46 & 636 & 1115 \tabularnewline
Lysithea & 60 & 84 & 90 & 234 & 431 \tabularnewline
Leda & 6 & 48 & 44 & 98 & 178 \tabularnewline
\hdashline
Pasiphae & 295 & 248 & 66 & 609 & 1629 \tabularnewline
Callirrhoe & 9 & -  &  16 & 25 & 95 \tabularnewline
Megaclite & - & -  &  10 & 10 & 50  \tabularnewline
\hdashline
Ananke & 52 & 141 & 57 & 250 & 600 \tabularnewline
Praxidike & - & -  &   2 & 2 & 59 \tabularnewline
\hdashline
Carme & 90 & 204 & 37 & 331 & 973 \tabularnewline
Sinope & 41 & 169 & 11 & 221 & 854 \tabularnewline
Themisto & - & - & 16 & 16 & 55 \tabularnewline
\hline
Phoebe & 1239 & 516 & 32 & 1787 & 3479 \tabularnewline
\hdashline
Siarnaq & - & 20 & 56 & 76 & 239 \tabularnewline
Paaliaq & - & - & 11 & 11 & 82 \tabularnewline
\hdashline
Albiorix & - & - & 46 & 46 & 137 \tabularnewline
\hline
Sycorax & - & - & 35 & 35 & 237 \tabularnewline
\hline
Nereid & 803 & - & 99 & 902 & 716 \tabularnewline
\hline
\end{tabular}
\par\end{centering}
Comparison between the number of positions obtained in our work with the number used in the numerical integration of orbits by the JPL as published by \citealp{Jacobson2012}.
\end{table}

\section{Comparison with ephemeris} \label{Sec: comparison}

Intending to see the potential of our results to improve the orbit of the irregular satellites observed, we analysed the offsets of our positions with regard to the ephemeris mentioned in Sect. \ref{Sec: reduction}. Taking Carme as example, we plot in Fig. \ref{Fig: carme_anom} the mean ephemeris offsets for each night and their dispersions  (one sigma error bars) as a function of the true anomaly in right ascension (\ref{Fig: carme_alfa}) and declination (\ref{Fig: carme_delta}). Fig. \ref{Fig: carme_delta} clearly shows a systematic error in declination. When Carme is close to its apojove (true anomaly = 180\degree) its offsets are more likely to be more negative than those close to its perijove (true anomaly = 0\degree). The offsets obtained from observations by four telescopes using different cameras and filters are in good agreement, meaning that there is an error in the ephemeris of Carme, most probably due to an error in its orbital inclination.

\begin{figure*}
\begin{centering}
\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Carme_RA}\label{Fig: carme_alfa}}
\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Carme_DEC}\label{Fig: carme_delta}}
\caption{Mean ephemeris offsets and dispersions (1 sigma error bars) in the coordinates of Carme taken night by night by true anomaly for each telescope. \textbf{The red square is for the observations with the Perkin-Elmer telescope from OPD, the blue circle for Boller \& Chivens, the magenta triangle down for Zeiss, the black triangle up for OHP and the green star for ESO.}}
\label{Fig: carme_anom}
\end{centering}
\end{figure*}

\begin{figure*}
\begin{centering}
\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Pasiphae_RA}\label{Fig: pasiphae_alfa}}
\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Pasiphae_DEC}\label{Fig: pasiphae_delta}}
\caption{Same as in Fig \ref{Fig: carme_anom} for Pasiphae.}
\label{Fig: pasiphae_anom}
\end{centering}
\end{figure*}

This pattern in declination was also seen for other satellites like Pasiphae (Fig: \ref{Fig: pasiphae_anom}) and Ananke (plots for other satellites with significant number of observations can be seen in the online material). For some satellites, the orbital coverage is not enough to clearly indicate the presence of systematic errors in specific orbital elements. However, comparing the internal position mean errors of the reductions (Table \ref{Tab: stars-errors}) with the external position errors estimated from the dispersion of the ephemeris offsets (Tables \ref{Tab: Reductions-160} to \ref{Tab: Reductions-eso}), we see position error values much larger than expected from the mean errors. This means that besides some expected astrometric errors, significant ephemeris errors must also be present.

\section{Conclusions} \label{Sec: conclusions}

\textbf{We managed a large database with FITS images acquired by 5 telescopes in 3 sites between 1992 and 2014. From that, we identified 8466 observations of irregular satellites, from which we managed to obtain 6523 suitable astrometric positions, giving a total of 3666 positions for 12 satellites of Jupiter, 1920 positions for 4 satellites of Saturn, 35 positions for Sycorax (Uranus) and 902 positions for Nereid (Neptune).}

\textbf{The positions of all the objects were determined using the PRAIA package. The package was suited to cope with the huge amount of observations and the task of identifying the satellites within the database. PRAIA tasks were also useful to deal with the missing or incorrect coordinate and time stamps present mostly in the old observations.}

\textbf{The UCAC4 was used as the reference frame. Based in the comparisons with ephemeris, we estimate that the position errors are about 60 mas to 80 mas depending on the satellite brightness. }

For some satellites the number \textbf{of positions obtained in this work} is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012} (see Table \ref{Tab: comparison-horizons}). \textbf{For instance, the amount of new positions for Himalia is about 70\% of the number used in the numerical integation of orbits by JPL}. Systematic errors in the ephemeris were found for at least some satellites (Ananke, Carme, Elara and Pasiphae). In the case of Carme, we evidenced an error in the orbital inclination \textbf{(see Fig. \ref{Fig: carme_anom})}.

\textbf{The positions derived in this work can be used in new orbital numerical integrations, generating more precise ephemerides. Stellar occultations by irregular satellites could then be better predicted. Based in this work, our group has already computed occultation predictions for the 8 major irregular satellites of Jupiter. These predictions will be published in a forthcoming paper.}

\begin{acknowledgements}

ARGJ thanks the financial support of CAPES. M. A. thanks the CNPq (Grants 473002/2013-2 and 308721/2011-0) and FAPERJ (Grant E-26/111.488/2013). RVM thanks grants: CNPq-306885/2013, Capes/Cofecub-2506/2015, Faperj/PAPDRJ-45/2013. J-E.A. thanks the "Programme National de Planétologie" of INSU-CNRS-CNES for its financial support. J.I.B. Camargo acknowledges CNPq for a PQ2 fellowship (process number 308489/2013-6). F.B.R. acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014. B.E.M. thanks the financial support of CAPES.

\end{acknowledgements}

\bibliographystyle{aa}
\nocite{*}
\bibliography{references.bib}

\Online

\begin{appendix}

\onecolumn

\section{Ephemeris offsets as a function of true anomaly for all observed irregular satellites} \label{anexo: offsets}

The distribution of ephemeris offsets along the orbit of the satellites are shown below. The red square is for the observations with the Perkin-Elmer telescope from OPD, the blue circle for Boller \& Chivens, the magenta triangle down for Zeiss, the black triangle up for OHP and the green star for ESO. For Carme and Pasiphae see Figs. \ref{Fig: carme_anom} and \ref{Fig: pasiphae_anom} in Section \ref{Sec: comparison}.

\begin{figure}[h!]
\begin{centering}
\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Himalia_RA}\label{Fig: an_himalia_alfa}}
\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Himalia_DEC}\label{Fig: an_himalia_delta}}
\caption{Mean ephemeris offset and dispersion (1 sigma error bars) in the coordinates of Himalia taken night by night as a function of true anomaly.}
\label{Fig: an_himalia_anom}
\end{centering}
\end{figure}

\begin{figure}[h!]
\begin{centering}
\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Elara_RA}\label{Fig: an_elara_alfa}}
\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Elara_DEC}\label{Fig: an_elara_delta}}
\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Elara.}
\label{Fig: an_elara_anom}
\end{centering}
\end{figure}

\begin{figure}[h!]
\begin{centering}
\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Lysithea_RA}\label{Fig: an_lysithea_alfa}}
\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Lysithea_DEC}\label{Fig: an_lysithea_delta}}
\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Lysithea.}
\label{Fig: an_lysithea_anom}
\end{centering}
\end{figure}

\begin{figure}[h!]
\begin{centering}
\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Leda_RA}\label{Fig: an_leda_alfa}}
\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Leda_DEC}\label{Fig: an_leda_delta}}
\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Leda.}
\label{Fig: an_leda_anom}
\end{centering}
\end{figure}

\begin{figure}[h!]
\begin{centering}
\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Carme_RA}\label{Fig: an_carme_alfa}}
\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Carme_DEC}\label{Fig: an_carme_delta}}
\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Carme.}
\label{Fig: an_carme_anom}
\end{centering}
\end{figure}

\begin{figure}[h!]
\begin{centering}
\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Ananke_RA}\label{Fig: an_ananke_alfa}}
\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Ananke_DEC}\label{Fig: an_ananke_delta}}
\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Ananke.}
\label{Fig: an_ananke_anom}
\end{centering}
\end{figure}

\begin{figure}[h!]
\begin{centering}
\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Sinope_RA}\label{Fig: an_sinope_alfa}}
\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Sinope_DEC}\label{Fig: an_sinope_delta}}
\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Sinope.}
\label{Fig: an_sinope_anom}
\end{centering}
\end{figure}

\begin{figure}[h!]
\begin{centering}
\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Phoebe_RA}\label{Fig: an_phoebe_alfa}}
\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Phoebe_DEC}\label{Fig: an_phoebe_delta}}
\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Phoebe.}
\label{Fig: an_phoebe_anom}
\end{centering}
\end{figure}

\begin{figure}[h!]
\begin{centering}
\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Nereid_RA}\label{Fig: an_nereid_alfa}}
\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Nereid_DEC}\label{Fig: an_nereid_delta}}
\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Nereid.}
\label{Fig: an_nereid_anom}
\end{centering}
\end{figure}

\end{appendix}

\end{document}

### 2015-05-16 Irregulares/Irregulares.tex
Adicionado Artigos dos irregulares e de Ceres

--- 2015-05-16 Irregulares/AA-2015-26273/Irregulars.tex
@@
 %% Do not edit unless you really know what you are doing.
-\documentclass[english,traditabstract, referee]{aa}
+\documentclass[english,traditabstract]{aa}
 \usepackage[varg]{txfonts}
@@
 \title{Astrometric positions for 18 irregular satellites of giant planets from 23 years of observations
-\fnmsep\thanks{The complete version of Table 8 is available through CDS and IAU NSDC data base at www.imcce.fr/nsdc.}
+\fnmsep\thanks{Table 8 are only available in electronic form
+at the CDS via anonymous ftp to cdsarc.u-strasbg.fr (130.79.128.5)
+or via http://cdsweb.u-strasbg.fr/cgi-bin/qcat?J/A+A/ and IAU NSDC data base at www.imcce.fr/nsdc.}
+%\fnmsep\thanks{The complete version of Table 8 is available through CDS and IAU NSDC data base at www.imcce.fr/nsdc.}
 \fnmsep\thanks{Partially based on observations made at Laborat\'orio Nacional de Astrof\'{\i}sica (LNA), Itajub\'a-MG, Brazil.}
@@
           A. H. Andrei
-          \inst{1,2} \fnmsep\thanks{Affiliated researcher at the Observatoire de Paris/SYRTE, 77 av. Denfert Rochereau, 75014 Paris, France},
+          \inst{1,2} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau, 75014 Paris, France},
           A. Dias-Oliveira

### 2015-05-16 Ceres/dados/DadosAltair/DadosAltair/grafico.py

@@ -1,0 +1,80 @@
+import numpy as np
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+from astropy.time import Time
+import os
+
+################## Definicao de Parametros  #################################################
+
+arquivo = 'data_ratio_poly_outCUT'			### arquivo de dados de entrada
+arqsaida = 'saida.dat'				### arquivo de dados de saida
+datainicio = '2010-08-17 00:00:00'		### data de referencia de contagem para o plot
+
+#############################################################################################
+
+Ceres = [
+('BH', 'CEAMIG'),
+('OPD', 'LNA'),
+('INPE', 'INPE'),
+('UEPG', 'UEPG')
+]
+
+Ceres = Ceres[::-1]
+
+axes=plt.gca()
+plt.xlabel('Time (seconds after {inicio}, UTC)'.format(inicio=datainicio))
+plt.ylabel('Flux Ratio')
+
+t = Time(datainicio, format='iso', scale='utc')
+fator = 0.2
+for i in np.arange(len(Ceres)):
+    
+    pasta = Ceres[i][0]
+    x = np.loadtxt('{}/data_ratio_poly_out'.format(pasta), unpack=True)
+    y, z, t, u = [], [], [], []
+    if os.path.isdir('{}/im'.format(pasta)):
+        y = np.loadtxt('{}/im/fort.20'.format(pasta), unpack=True)
+        z = np.loadtxt('{}/im/fort.21'.format(pasta), unpack=True)
+        t = np.loadtxt('{}/im/fort.22'.format(pasta), unpack=True)
+        u = np.loadtxt('{}/im/fort.23'.format(pasta), unpack=True)
+    if os.path.isdir('{}/em'.format(pasta)):
+        ye = np.loadtxt('{}/em/fort.20'.format(pasta), unpack=True)
+        ze = np.loadtxt('{}/em/fort.21'.format(pasta), unpack=True)
+        te = np.loadtxt('{}/em/fort.22'.format(pasta), unpack=True)
+        ue = np.loadtxt('{}/em/fort.23'.format(pasta), unpack=True)
+        if y == []:
+            y = ye
+        else:
+            y = np.append(y, ye, axis=1)
+        if z == []:
+            z = ze
+        else:
+            z = np.append(z, ze, axis=1)
+        if t == []:
+            t = te
[…]

### 2015-05-16 Ceres/dados/Mapa_Drummond_PA/2010_Drummond_PA/plot_pla.f

@@ -1,0 +1,1298 @@
+c
+c programme de trace de fonctions au choix.
+c
+	INTEGER UNIT,JUST,AXIS
+	CHARACTER*150 DEVICE,XLBL,YLBL,TOPLBL
+ 1000	FORMAT(A)
+c
+c
+	write(*,*) 'tt/gf=console micro-VAX, tt/tek= console VAX, /dw= Xwindow'
+	write(*,*) 'toto/ps'
+	write(*,*) '/xw: fenetre temporaire'
+	write(*,*) '/xserve: fenetre permanente'
+	read 1000, device
+	write(*,*) 'Nombre de dessins en x et y ?'
+	read(*,*)  nx,ny
+	unit= 20
+	CALL PGBEGIN (UNIT,DEVICE,NX,NY)
+	write(*,*)  'Epaisseur du trait ?'
+	read(*,*)  iepai
+	call pgslw(iepai)
+c
+	write(*,*)  '1= echelle automatique'
+	write(*,*)  '0= pleine page'
+c	just=0
+	read(*,*)  just
+
+  2	write(*,*)  'axis= -2, pas d''annotation'
+	write(*,*)  'axis= -1, cadre seulement'
+	write(*,*)  'axis=  0, cadre avec valeurs sur les axes'
+	write(*,*)  'axis=  1, tirets sur les axes'
+	write(*,*)  'axis=  2, trace une grille'
+	write(*,*)  'axis= 10, trace axe des x en logarithmique'
+	write(*,*)  'axis= 20, trace axe des y en logarithmique'
+	write(*,*)  'axis= 30, trace axe des x et y en logarithmique'	
+	read(*,*)  axis
+
+	write(*,*) 'coin inf. gauche et coin sup. droit ? Taille char.?'
+	write(*,*) 'Xmin, Ymin'
+	write(*,*) 'Xmax, Ymax, taille'
+	read(*,*) xmin,ymin
+	read(*,*) xmax,ymax, char_size
+	call pgsch (char_size)
+	call pgenv(xmin,xmax,ymin,ymax,just,axis)
+C
+	write(*,*) 'LEGENDE EN X,Y ET SOMMET ?'
+	READ 1000 ,XLBL
+	READ 1000 ,YLBL
+	READ 1000 ,TOPLBL
+	CALL PGLABEL (XLBL,YLBL,TOPLBL)
+	call pgsch (1.)
+
+	write(*,*) 'afficher un tableau en grise?'
+	read(*,*)  iop
[…]

### 2015-05-16 Ceres/dados/Mapa_Drummond_PA/bash_positionv_ellipse.sh

@@ -1,0 +1,20 @@
+#nedit ellipse_fit.i positionv.i plot_* fort.53 &
+rm -f fort.66
+./positionv < positionv_err.i > out_poserr
+cp fort.21 fort.21err
+./positionv < positionv_mid.i > out_posmid
+cp fort.21 fort.21mid
+#./positionv < positionv_mid_err.i > out_posmid_err
+#cp fort.21 fort.21mid_err
+#./positionv < positionv_neg.i > out_posneg
+#cp fort.21 fort.21neg
+./positionv < positionv_all.i > out_posall
+cp fort.21 fort.21all
+./positionv < positionv.i > out_positionv
+cp fort.21 fort.21nom
+./ellipse_fit < ellipse_fit.i > out_ellipse2
+cp fort.51 fort.51nom
+#read ok
+#plot_data < plot_data_lat.i
+#plot_pla < plot_pla.i
+

### 2015-05-16 Ceres/dados/Predicao2010/mapa.py

@@ -1,0 +1,218 @@
+from mpl_toolkits.basemap import Basemap
+import matplotlib.pyplot as plt
+import numpy as np
+import os
+import astropy.units as u
+from astropy.time import Time, TimeDelta
+from astropy.coordinates import SkyCoord, EarthLocation
+
+######################################### lendo arquivo de dados da ocultacao e de observatorios ##############################
+
+def lerdados():
+    dados = np.loadtxt(arquivo, skiprows=41, usecols=(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 18, 19, 20, 21, 22, 25, 26, 28, 29), \
+        dtype={'names': ('dia', 'mes', 'ano', 'hor', 'min', 'sec', 'afh', 'afm', 'afs', 'ded', 'dem', 'des', 'ca', 'pa', 'vel', 'delta', 'mR', 'mK', 'long', 'ora', 'ode'), 
+        'formats': ('S30', 'S30', 'S30','S30', 'S30', 'f8','i4', 'i4', 'f8','i4', 'i4', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8')})
+
+################## lendo coordenadas #################
+
+    alfa = np.core.defchararray.add(np.array(map(str, dados['afh'])), [' '])
+    alfa = np.core.defchararray.add(alfa, np.array(map(str, dados['afm'])))
+    alfa = np.core.defchararray.add(alfa, [' '])
+    alfa = np.core.defchararray.add(alfa, np.array(map(str, dados['afs'])))
+    delta = np.core.defchararray.add(np.array(map(str, dados['ded'])), [' '])
+    delta = np.core.defchararray.add(delta, np.array(map(str, dados['dem'])))
+    delta = np.core.defchararray.add(delta, [' '])
+    delta = np.core.defchararray.add(delta, np.array(map(str, dados['des'])))
+    coor = np.core.defchararray.add(alfa, [' '])
+    coor = np.core.defchararray.add(coor, delta)
+    stars = SkyCoord(coor, frame='icrs', unit=(u.hourangle, u.degree))
+
+################### lendo tempo ########################
+
+    tempo = np.core.defchararray.add(np.array(dados['ano']), ['-'])
[…]

### 2015-05-16 Ceres/dados/Predicao2013/PRAIA_occ_star_search_10.f

@@ -1,0 +1,2010 @@
+c
+c     Program PRAIA_occ_star_search
+c
+c     
+c
+c     Identify occultation candidate stars from body ephemeris and "xy"
+c     output file from PRAIA astrometric reductions.
+c
+c
+c     Valid xy files: PRAIA_astrometry version 12 (or earlier) and
+c     PRAIA_global_reduction version 14.  
+c
+C
+c     This version.
+c
+c     Quadrant of chord checked. Normalized R magnitude to 20km/s computed. 
+c
+c     A mini version of the full (RA,DEC) catalog of positions is also
+c     output for plot purposes.
+c
+c     A linear drift of ephemeris in (RA,DEC) can be set.
+c
+c     Optional Solar Local Time range of exclusion of events
+c
+c     (RA,DEC) errors of star are also listed in prediction table.
+c
+c     (RA,DEC) proper motions  are also listed in prediction table.
+c
+c     Option for choosing what type of ephemeris format (Horizons, NAIF)
+c     and TNO/Pluto ephemeris format (Horizons option only)
+c
+c
+c     Ephemeris data points around close approach are spaced by 1 hour
+c     in accord to B. Sicardy procedures for computing C/A and P/A.
+c
+c
+c
+c     A bug on proper motion computations in version 7 was corrected. Do
+c     not use version 7 unless the code was explicitly corrected. For that,
+c     use version 8 or later instead.
+c
+c
+c     In this version, the geocentric astrometric TNO (RA,DEC) position,
+c     corrected by the ephemeris offsets, is furnished for the instant of
+c     occultation. 
+c
+c
+c
+c     Last update: M. Assafin - 20/Mar/2012
+c
+c
+
+      IMPLICIT REAL *8 (A-H,O-Z)
+
+      parameter(stdin=5,stdout=6)
+
+      dimension x(2600000),y(2600000),cseng(2600000),altu(2600000),
[…]

### 2015-05-16 Ceres/dados/Predicao2013/mapa.py

@@ -1,0 +1,260 @@
+from mpl_toolkits.basemap import Basemap
+import matplotlib.pyplot as plt
+import numpy as np
+import os
+import astropy.units as u
+from astropy.time import Time, TimeDelta
+from astropy.coordinates import SkyCoord, EarthLocation
+
+######################################### lendo arquivo de dados da ocultacao e de observatorios ##############################
+
+def lerdados():
+    dados = np.loadtxt(arquivo, skiprows=41, usecols=(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 18, 19, 20, 21, 22, 25, 26, 28, 29), \
+        dtype={'names': ('dia', 'mes', 'ano', 'hor', 'min', 'sec', 'afh', 'afm', 'afs', 'ded', 'dem', 'des', 'ca', 'pa', 'vel', 'delta', 'mR', 'mK', 'long', 'ora', 'ode'), 
+        'formats': ('S30', 'S30', 'S30','S30', 'S30', 'f8','i4', 'i4', 'f8','i4', 'i4', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8')}, ndmin=1)
+
+################## lendo coordenadas #################
+
+    alfa = np.core.defchararray.add(np.array(map(str, dados['afh'])), [' '])
+    alfa = np.core.defchararray.add(alfa, np.array(map(str, dados['afm'])))
+    alfa = np.core.defchararray.add(alfa, [' '])
+    alfa = np.core.defchararray.add(alfa, np.array(map(str, dados['afs'])))
+    delta = np.core.defchararray.add(np.array(map(str, dados['ded'])), [' '])
+    delta = np.core.defchararray.add(delta, np.array(map(str, dados['dem'])))
+    delta = np.core.defchararray.add(delta, [' '])
+    delta = np.core.defchararray.add(delta, np.array(map(str, dados['des'])))
+    coor = np.core.defchararray.add(alfa, [' '])
+    coor = np.core.defchararray.add(coor, delta)
+    stars = SkyCoord(coor, frame='icrs', unit=(u.hourangle, u.degree))
+
+################### lendo tempo ########################
+
+    tempo = np.core.defchararray.add(np.array(dados['ano']), ['-'])
[…]

### 2015-05-16 Ceres/dados/Predicao2013/mapa2.py

@@ -1,0 +1,218 @@
+from mpl_toolkits.basemap import Basemap
+import matplotlib.pyplot as plt
+import numpy as np
+import os
+import astropy.units as u
+from astropy.time import Time, TimeDelta
+from astropy.coordinates import SkyCoord, EarthLocation
+
+######################################### lendo arquivo de dados da ocultacao e de observatorios ##############################
+
+def lerdados():
+    dados = np.loadtxt(arquivo, skiprows=41, usecols=(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 18, 19, 20, 21, 22, 25, 26, 28, 29), \
+        dtype={'names': ('dia', 'mes', 'ano', 'hor', 'min', 'sec', 'afh', 'afm', 'afs', 'ded', 'dem', 'des', 'ca', 'pa', 'vel', 'delta', 'mR', 'mK', 'long', 'ora', 'ode'), 
+        'formats': ('S30', 'S30', 'S30','S30', 'S30', 'f8','i4', 'i4', 'f8','i4', 'i4', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8')})
+
+################## lendo coordenadas #################
+
+    alfa = np.core.defchararray.add(np.array(map(str, dados['afh'])), [' '])
+    alfa = np.core.defchararray.add(alfa, np.array(map(str, dados['afm'])))
+    alfa = np.core.defchararray.add(alfa, [' '])
+    alfa = np.core.defchararray.add(alfa, np.array(map(str, dados['afs'])))
+    delta = np.core.defchararray.add(np.array(map(str, dados['ded'])), [' '])
+    delta = np.core.defchararray.add(delta, np.array(map(str, dados['dem'])))
+    delta = np.core.defchararray.add(delta, [' '])
+    delta = np.core.defchararray.add(delta, np.array(map(str, dados['des'])))
+    coor = np.core.defchararray.add(alfa, [' '])
+    coor = np.core.defchararray.add(coor, delta)
+    stars = SkyCoord(coor, frame='icrs', unit=(u.hourangle, u.degree))
+
+################### lendo tempo ########################
+
+    tempo = np.core.defchararray.add(np.array(dados['ano']), ['-'])
[…]

### 2015-05-16 Ceres/dados/Predicao2013/observation.py

@@ -1,0 +1,289 @@
+# coding=UTF-8
+
+import numpy as np
+import itertools
+from astropy.time import Time, TimeDelta
+import astropy.units as u
+from astropy.coordinates import SkyCoord, Latitude, Longitude, FK5, EarthLocation, Angle
+
+######################################################################
+
+class Observation(object):
+    """
+    """
+
+    def __init__(self, fuse = 0, latitude = 0.0, longitude = 0.0, height = 0.0):
+        self.__set_fuse(fuse)
+        self.__set_site(longitude, latitude, height)
+        
+    def __set_site(self,longitude, latitude, height=0.0*u.m):
+        """
+        """
+        self.site = EarthLocation(longitude, latitude, height)
+
+    def __set_fuse(self, fuse):
+        """
+        """
+        self.fuse = TimeDelta(fuse*3600, format='sec', scale='tai')
+
+    def read(self, datafile, name_col=None, coord_col=None, comment_col=None):
+        """
+        """
+        nome, coord, comment, retornar = None, None, None, []
+        if name_col != None:
+            if type(name_col) != list and type(name_col) != tuple and type(name_col) != numpy.ndarray:
+                name_col = [name_col]
+            nomes = np.loadtxt(datafile, usecols=(name_col), unpack=True, dtype ='S30', ndmin=1)
+            if len(name_col) > 1:
+                nome = nomes[0]
+                for i in np.arange(len(name_col))[1:]:
+                    nome = np.core.defchararray.add(nome, ' ')
+                    nome = np.core.defchararray.add(nome, nomes[i])
+            else:
+                nome = nomes
+        if coord_col != None:
+            if type(coord_col) != list and type(coord_col) != tuple and type(coord_col) != numpy.ndarray:
+                coord_col = [coord_col]
+            coords = np.loadtxt(datafile, usecols=(coord_col), unpack=True, dtype ='S20', ndmin=1)
+            coor = coords[0]
+            for i in np.arange(len(coord_col))[1:]:
+                coor = np.core.defchararray.add(coor, ' ')
+                coor = np.core.defchararray.add(coor, coords[i])
+            coord = SkyCoord(coor, frame='fk5', unit=(u.hourangle, u.degree))
+        if comment_col != None:
[…]

### 2015-05-16 Ceres/dados/grafico.py

@@ -1,0 +1,64 @@
+import numpy as np
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+from astropy.time import Time
+
+################## Definicao de Parametros  #################################################
+
+arquivo = 'data_ratio_poly_outCUT'			### arquivo de dados de entrada
+arqsaida = 'saida.dat'				### arquivo de dados de saida
+datainicio = '2013-10-25 00:00:00'		### data de referencia de contagem para o plot
+
+#############################################################################################
+
+Ceres = [
+('Ted_Hampton', 'Hampton'),
+('Ted_Topsfield', 'Topsfield'),
+('ClayCenter', 'Brookline'),
+('JBrooks', 'Winchester'),
+('Greenbelt_Dunham', 'Greenbelt'),
+('Maley', 'Alexandria'),
+('CHAD', 'Owings'),
+('Mechanicsville_Dunham', 'Mechanicville'),
+('Varina_Dunham', 'Varina')
+]
+
+Ceres = Ceres[::-1]
+
+axes=plt.gca()
+plt.xlabel('Time (seconds after {inicio})'.format(inicio=datainicio))
+plt.ylabel('Flux Ratio')
+
+t = Time(datainicio, format='iso', scale='utc')
+fator = 1.0
+for i in np.arange(len(Ceres)):
+    
+    pasta = Ceres[i][0]
+    x = np.loadtxt('{}/immersion/data_ratio_poly_out'.format(pasta), unpack=True)
+    yi = np.loadtxt('{}/immersion/fort.20'.format(pasta), unpack=True)
+    zi = np.loadtxt('{}/immersion/fort.21'.format(pasta), unpack=True)
+    ti = np.loadtxt('{}/immersion/fort.22'.format(pasta), unpack=True)
+    ui = np.loadtxt('{}/immersion/fort.23'.format(pasta), unpack=True)
+    ye = np.loadtxt('{}/emersion/fort.20'.format(pasta), unpack=True)
+    ze = np.loadtxt('{}/emersion/fort.21'.format(pasta), unpack=True)
+    te = np.loadtxt('{}/emersion/fort.22'.format(pasta), unpack=True)
+    ue = np.loadtxt('{}/emersion/fort.23'.format(pasta), unpack=True)
+    y = np.append(yi, ye, axis=1)
+    z = np.append(zi, ze, axis=1)
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/ESO/Nereid/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="$1"
+nummin="2"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
+  tiramedia .aux3
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/ESO/gera_graph.py

@@ -1,0 +1,16 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+a = np.loadtxt('arquivo', dtype={'names':('number','name'), 'formats':('i8','S20')})
+
+b = np.sort(a)[::-1]
+
+fig, ax = plt.subplots()
+ind = np.arange(len(b))
+rects = ax.bar(ind, b['number'], 0.9, color='grey')
+
+ax.set_ylabel('Number of Frames', fontsize=15)
+ax.set_xticks(ind+1)
+ax.set_xticklabels(b['name'], fontweight='bold', rotation=-45)
+fig.set_size_inches(8.0,4.5)
+fig.savefig('framexobj_eso.eps',dpi=300, format='eps', bbox_inches='tight')

### 2015-05-16 Irregulares/dados/Irregulares_results/OHP/Ananke/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Ananke_ucac4_OHP.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OHP/Carme/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Carme_ucac4_OHP.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OHP/Elara/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Elara_ucac4_OHP.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OHP/Iapetus/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Iapetus_ucac4_OHP.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OHP/Leda/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Leda_ucac4_OHP.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OHP/Lysithea/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Lysithea_ucac4_OHP.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OHP/Pasiphae/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Pasiphae_ucac4_OHP.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OHP/Phoebe/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Phoebe_ucac4_OHP.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OHP/Siarnaq/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Siarnaq_ucac4_OHP.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OHP/Sinope/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Sinope_ucac4_OHP.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OHP/gera_graph.py

@@ -1,0 +1,16 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+a = np.loadtxt('arquivo', dtype={'names':('number','name'), 'formats':('i8','S20')})
+
+b = np.sort(a)[::-1]
+
+fig, ax = plt.subplots()
+ind = np.arange(len(b))
+rects = ax.bar(ind, b['number'], 0.9, color='grey')
+
+ax.set_ylabel('Number of Frames', fontsize=15)
+ax.set_xticks(ind+1)
+ax.set_xticklabels(b['name'], fontweight='bold', rotation=-45)
+fig.set_size_inches(8.0,4.5)
+fig.savefig('framexobj_ohp.eps',dpi=300, format='eps', bbox_inches='tight')

### 2015-05-16 Irregulares/dados/Irregulares_results/OPD-BC/Carme/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Carme_IAG1.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
+  tiramedia .aux3
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OPD-BC/Elara/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Elara_IAG_total.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OPD-BC/Himalia/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Himalia_IAG_total.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OPD-BC/Lysithea/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Lysithea_IAG1.table"
+nummin="3"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OPD-BC/Nereida/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Nereida_IAG_total.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OPD-BC/Phoebe/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Phoebe_IAG_total.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OPD-BC/Sinope/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Sinope_IAG1.table"
+nummin="3"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OPD-PE/Ananke/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Ananke_160-1.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OPD-PE/Carme/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Carme_160-total.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OPD-PE/Himalia/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Himalia_160-total.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OPD-PE/Leda/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Leda_160-total.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OPD-ZEI/Himalia/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Himalia_ZEI_total.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/OPD-ZEI/Phoebe/statistics.sh

@@ -1,0 +1,72 @@
+#
+### 
+#
+arquivo="Phoebe_ZEI1.table"
+nummin="4"
+sigma="2.5"
+#
+### Código
+#
+tiramedia(){
+  totalalfa="0"
+  totaldelta="0"
+  totalalfa2="0"
+  totaldelta2="0"
+  while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+    totalalfa=$(calc "$totalalfa + $ac"  < /dev/null | sed 's/\~//')
+    totalalfa2=$(calc "$totalalfa2 + $ac*$ac"  < /dev/null | sed 's/\~//')
+    totaldelta=$(calc "$totaldelta + $ad"  < /dev/null | sed 's/\~//')
+    totaldelta2=$(calc "$totaldelta2 + $ad*$ad"  < /dev/null | sed 's/\~//')
+  done < $1
+  numarq=$(cat $1 | wc -l)
+  medalfa=$(calc "$totalalfa / $numarq"  < /dev/null | sed 's/\~//'); echo "medalfa = $medalfa"
+  meddelta=$(calc "$totaldelta / $numarq"  < /dev/null | sed 's/\~//'); echo "meddelta = $meddelta"
+  sigmaalfa=$(calc "sqrt(($totalalfa2 - $numarq*$medalfa*$medalfa)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmaalfa = $sigmaalfa"
+  sigmadelta=$(calc "sqrt(($totaldelta2 - $numarq*$meddelta*$meddelta)/($numarq -1))" < /dev/null | sed 's/\~//'); echo "sigmadelta = $sigmadelta"
+}
+#
+#
+#
+nlines=$(cat $arquivo | wc -l)
+depois=$(echo "$nlines - 2" | bc)
+head -$depois $arquivo > .aux1
+antes=$(echo "$depois - 43" | bc)
+tail -$antes .aux1 > .aux2
+while read aa ab ac ad ae af ag ah ai aj ak al am an ao ap aq aar as at au av aw ax ay az ba;do
+  if [ "$ae" -lt "$nummin" ];then
+    cat $arquivo | grep "$az" >> "$arquivo"_eliminated
+  else
+    cat $arquivo | grep "$az" >> .aux3
+  fi
+done < .aux2
+refazer="1"
+until [ "$refazer" -eq "0" ];do
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/final_CDS/CDS.py

@@ -1,0 +1,351 @@
+import numpy as np
+from astropy.coordinates import SkyCoord
+from astropy.time import Time, TimeDelta
+import astropy.units as u
+import os
+
+###############################################################
+
+arquivo = ['search_results_ananke_OHP_final',
+ 'search_results_carme_OHP_final',
+ 'search_results_elara_OHP_final',
+ 'search_results_himalia_OHP_final',
+ 'search_results_leda_OHP_final',
+ 'search_results_lysithea_OHP_final',
+ 'search_results_pasiphae_OHP_final',
+ 'search_results_phoebe_OHP_final',
+ 'search_results_siarnaq_OHP_final',
+ 'search_results_sinope_OHP_final',
+ 'ucac4_Albiorix_eso_global_filtered_final',
+ 'ucac4_Ananke_160-1_filtered_final',
+ 'ucac4_Ananke_eso_global_filtered_final',
+ 'ucac4_Callirrhoe_160-1_filtered_final',
+ 'ucac4_Callirrhoe_eso_global_filtered_final',
+ 'ucac4_Carme_160_filtered_final',
+ 'ucac4_Carme_eso_global_filtered_final',
+ 'ucac4_Carme_IAG1_filtered_final',
+ 'ucac4_Elara_160_filtered_final',
+ 'ucac4_Elara_eso_global_filtered_final',
+ 'ucac4_Elara_IAG_filtered_final',
+ 'ucac4_Elara_zei1_filtered_final',
+ 'ucac4_Himalia_160_filtered_final',
+ 'ucac4_Himalia_eso_global_filtered_final',
+ 'ucac4_Himalia_IAG_filtered_final',
+ 'ucac4_Himalia_zei_filtered_final',
+ 'ucac4_Leda_160_filtered_final',
+ 'ucac4_Leda_eso_global_filtered_final',
+ 'ucac4_Lysithea_160-1_filtered_final',
+ 'ucac4_Lysithea_eso_global_filtered_final',
+ 'ucac4_Lysithea_IAG1_filtered_final',
+ 'ucac4_Megaclite_eso_global_filtered_final',
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/gera_graph.py

@@ -1,0 +1,22 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+a = np.loadtxt('arquivo2', dtype={'names':('name','IAG', '160','ZEI'), 'formats':('S20','f16', 'f16', 'f16')})
+
+c = a['IAG']+ a['160'] + a['ZEI']
+
+b = np.argsort(c)[::-1]
+
+fig, ax = plt.subplots()
+ind = np.arange(len(b))
+
+p1 = plt.bar(ind, a['160'][b], 0.9, color='black')
+p2 = plt.bar(ind, a['IAG'][b], 0.9, color='grey', bottom=a['160'][b])
+p3 = plt.bar(ind, a['ZEI'][b], 0.9, color='white', bottom=a['160'][b]+ a['IAG'][b])
+
+ax.set_ylabel('Number of Frames', fontsize=15)
+ax.set_xticks(ind+1)
+ax.set_xticklabels(a['name'][b], fontweight='bold', rotation=-45)
+fig.set_size_inches(8.0,4.5)
+plt.legend( (p1[0], p2[0], p3[0]), ('Perkin-Elmer', 'Boller & Chivens', 'Zeiss') )
+fig.savefig('framexobj_opd.eps',dpi=300, format='eps', bbox_inches='tight')

### 2015-05-16 Irregulares/dados/Irregulares_results/gera_graph_ano.py

@@ -1,0 +1,54 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+f = open('arquivo', 'r')
+lines = f.readlines()
+f.close()
+
+arr = np.array([], dtype={'names': ('ano', 'Ananke', 'Callirrhoe', 'Carme', 'Elara', 'Himalia', 'Leda', 'Lysithea', 'Pasiphae', 'Sinope', 'Phoebe', 'Nereid'),
+'formats':('i8','i8','i8','i8','i8','i8','i8','i8','i8','i8','i8','i8')})
+
+for i in np.arange(1992,2015):
+    a = [(i, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0)]
+    b = np.asarray(a, dtype={'names': ('ano', 'Ananke', 'Callirrhoe', 'Carme', 'Elara', 'Himalia', 'Leda', 'Lysithea', 'Pasiphae', 'Sinope', 'Phoebe', 'Nereid'),
+'formats':('i8','i8','i8','i8','i8','i8','i8','i8','i8','i8','i8','i8')})
+    arr = np.append(arr,b)
+
+for i in lines:
+    g = i.strip().split('/')[1]
+    a= np.loadtxt(i.strip(), usecols=[40], dtype={'names':['ano'],'formats':['i8']}, unpack=True, ndmin=1)
+    for k in arr['ano']:
+        l = np.where(a['ano'] == k)[0]
+        if len(l) > 0:
+            tam = len(l)
+            arr[g][np.where(arr['ano']==k)] = arr[g][np.where(arr['ano']==k)] + tam
+
+fig, ax = plt.subplots(2, sharey=True)
+ind = np.arange(len(arr))
+space = np.linspace(-0.5,0.5-1.0/len(arr.dtype.names[1:]),len(arr.dtype.names[1:]))
+width = space[1]-space[0]
+colormap = plt.cm.gist_ncar
+cores = [colormap(i) for i in np.linspace(0, 0.95, len(arr.dtype.names[1:]))]
+k = 0
+a = np.where(arr['ano'] < 2004)
+b = np.where(arr['ano'] >= 2004)
+for i in arr.dtype.names[1:]:
+    ax[0].bar(arr['ano'][a]+space[k], arr[i][a], width, color=cores[-k-1], label=i)
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/gera_graph_ano_ohp.py

@@ -1,0 +1,47 @@
+import numpy as np
+import matplotlib.pyplot as plt
+
+f = open('arquivo3', 'r')
+lines = f.readlines()
+f.close()
+
+arr = np.array([], dtype={'names': ('ano', 'Ananke', 'Callirrhoe', 'Carme', 'Elara', 'Himalia', 'Leda', 'Lysithea', 'Pasiphae', 'Sinope', 'Phoebe', 'Siarnaq'),
+'formats':('i8','i8','i8','i8','i8','i8','i8','i8','i8','i8','i8','i8')})
+
+for i in np.arange(1998,2009):
+    a = [(i, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0, 0.0)]
+    b = np.asarray(a, dtype={'names': ('ano', 'Ananke', 'Callirrhoe', 'Carme', 'Elara', 'Himalia', 'Leda', 'Lysithea', 'Pasiphae', 'Sinope', 'Phoebe', 'Siarnaq'),
+'formats':('i8','i8','i8','i8','i8','i8','i8','i8','i8','i8','i8','i8')})
+    arr = np.append(arr,b)
+
+for i in lines:
+    g = i.strip().split('/')[1]
+    a= np.loadtxt(i.strip(), usecols=[40], dtype={'names':['ano'],'formats':['i8']}, unpack=True, ndmin=1)
+    for k in arr['ano']:
+        l = np.where(a['ano'] == k)[0]
+        if len(l) > 0:
+            tam = len(l)
+            arr[g][np.where(arr['ano']==k)] = arr[g][np.where(arr['ano']==k)] + tam
+
+fig, ax = plt.subplots()
+ind = np.arange(len(arr))
+space = np.linspace(-0.5,0.5-1.0/len(arr.dtype.names[1:]),len(arr.dtype.names[1:]))
+width = space[1]-space[0]
+colormap = plt.cm.gist_ncar
+cores = [colormap(i) for i in np.linspace(0, 0.95, len(arr.dtype.names[1:]))]
+k = 0
+for i in arr.dtype.names[1:]:
+    ax.bar(arr['ano']+space[k], arr[i], width, color=cores[-k-1], label=i)
+    k = k +1
+#p2 = plt.bar(ind, a['IAG'][b], 0.9, color='grey', bottom=a['160'][b])
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/grafico_anom.py

@@ -1,0 +1,190 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from astropy.time import Time, TimeDelta
+import os
+
+######################################################################3
+
+plt.rcParams['text.latex.preamble']=[r"\usepackage{txfonts}"]
+
+params = {'text.usetex' : True,
+          'font.size' : 22,
+          'font.family' : 'txfonts',
+          'text.latex.unicode': True,
+          }
+          
+plt.rcParams.update(params)
+
+obj = ['Ananke','Carme','Elara','Himalia','Leda','Lysithea','Nereid','Pasiphae','Phoebe','Sinope']
+nobj= [512, 511, 507, 506, 513, 510, 802, 508, 609, 509]
+pee = ['160-1', '160-total', '160-total', '160-total', '160-total', '160-1', '160-total', '160-total', '160-total', '160-total']
+iage = ['IAG_total', 'IAG1', 'IAG_total', 'IAG_total', 'IAG_total', 'IAG1', 'IAG_total', 'IAG1', 'IAG_total', 'IAG1']
+zeie = ['ZEI1', 'ZEI1', 'ZEI1', 'ZEI_total', 'ZEI1', 'ZEI1', 'ZEI1', 'ZEI1', 'ZEI1', 'ZEI1']
+locald = [2, 1, 1, 1, 1, 1, 1, 1, 1, 1]
+localr = [2, 1, 1, 1, 1, 1, 1, 1, 1, 1]
+
+sizel = 18
+
+def julian(idx):
+    f = open('julian_date_{}'.format(obj[idx]), 'w')
+    datas = []
+    if os.path.isfile(arq160) == True:
+        datas = np.hstack((datas, dado160[4]))
+    if os.path.isfile(arqiag) == True:
+        datas = np.hstack((datas, dadoiag[4]))
+    if os.path.isfile(arqzei) == True:
+        datas = np.hstack((datas, dadozei[4]))
+    if os.path.isfile(arqohp) == True:
+        datas = np.hstack((datas, dadoohp[4]))
+    if os.path.isfile(arqeso) == True:
+        datas = np.hstack((datas, dadoeso[4]))
+    for i in datas:
+        f.write('{:.10f} JD\n'.format(i))
+    f.close()
+    g = open('entry_naif_{}'.format(obj[idx]), 'w')
+    g.write('874\n')
+    g.write('{}\n'.format(nobj[idx]))
+    g.write('de421\n')
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/rodanaif.sh

@@ -1,0 +1,3 @@
+cd /homeA/NAIF64/toolkit/work/
+./ephem_hv5 < /home/altair/Documentos/Irregulares/dados/Irregulares_results/entry_naif_Sinope > /home/altair/Documentos/Irregulares/dados/Irregulares_results/Sinope_ephem.dat
+cd /home/altair/Documentos/Irregulares/dados/Irregulares_results/

### 2015-05-16 Irregulares/dados/Irregulares_results/teste.py

@@ -1,0 +1,43 @@
+import numpy as np
+from astropy.time import Time
+from astropy import units
+from astropy.coordinates import SkyCoord
+
+arquivo1 = 'ucac4_Himalia_IAG_filtered_final'
+arquivo2 = 'Himalia_IAG_total.table_filtered'
+local = 'IAG'
+
+x = np.loadtxt(arquivo1, usecols=(10, 35, 36, 43), unpack=True)
+filt = np.genfromtxt(arquivo1, usecols=45, dtype='str')
+y = np.loadtxt(arquivo2, usecols=(2, 3, 25), unpack=True)
+
+t = Time(x[3], format='jd', scale='utc')
+u = Time(y[2], format='jd', scale='utc')
+
+k = t
+k.precision = 0
+l = u
+l.precision = 0
+
+b = [local for k in range(len(x[0]))]
+
+n = []
+o = []
+for i in t:
+    m = u - i
+    for index, item in enumerate(m.jd):
+        if -1.0 < item < 1.0:
+            valor = index
+    n.append(y[0][valor])
+    o.append(y[1][valor])
+
+r = SkyCoord(ra=[x[1]]*units.hours, dec=[x[2]]*units.degree, frame='icrs')
+print r.ra
+
+#print len(x[1]), len(x[2]), len(n), len(o), len(t.jd), len(x[0]), filt, local
+
+z = np.array(zip(x[1], x[2], n, o, t.jd, x[0], filt, b), dtype=[('x[1]', float), ('x[2]', float), ('n', float), ('o', float), ('t.jd', float), ('x[0]', float), ('filt', 'S8'), ('b', 'S8')])
+
+#np.savetxt('output.txt', data, fmt=["%.3f",]*3 + ["%s"])
+#np.savetxt('saida.dat', z, fmt=["%10.6f",]*2 + ["%5.1f",]*2 + ["%15.8f",] + ["%6.1f",] + ["%8s",]*2)
+#np.savetxt('saida2.dat', z, delimiter=' & ', fmt=["%10.6f",]*2 + ["%5.1f",]*2 + ["%15.8f",] + ["%6.1f",] + ["%8s",]*2)

### 2015-05-16 Irregulares/dados/Irregulares_results/teste/binagem.py

@@ -1,0 +1,137 @@
+import numpy as np
+import matplotlib
+matplotlib.use('Agg')
+import matplotlib.pyplot as plt
+from astropy.time import Time
+from multiprocessing import Pool
+
+################## Definicao de Parametros  #################################################
+
+
+
+arquivo = 'photometry.dat'			### arquivo de dados de entrada
+binagem = [1, 3, 5, 15]				### lista de binagens a plotar
+coltime = [2]                                   ### coluna do tempo
+coltarg = [4, 7, 10, 13, 16]                    ### colunas das estrelas alvo
+targcom = ['6px', '8px', '10px', '12px', '14px']			### nomes das colunas da estrela alvo - plot issue
+calib = 'No'					### caso haja estrela de calibracao => calib = 'Yes'
+colcalb = [19]					### coluna da estrela de calibracao
+calbcom = ['Calib-10px']			### abertura da estrela de calibracao - plot issue
+proc = 4					### numero maximo de processos paralelos
+#  #  #  ### plot config ###  #  #  #
+nomelocal = 'JBrooks'				### Nome do sitio de observacao
+timelimuser = 'No'				### usar limites do eixo x fornecido pelo usuario => Yes ou No
+timeminlim = -5					### tempo minimo do plot; segundos a partir do tempo inicial
+timemaxlim = 100				### tempo maximo do plot; segundos a partir do tempo inicial
+datainicio = '2013-10-25 09:40:00'		### data de referencia de contagem para o plot
+fluxlimuser = 'Yes'				### usar limites de fluxo do usuario => Yes ou No
+fluxminlim = 100000				### fluxo minimo do plot
+fluxmaxlim = 900000				### fluxo maximo do plot
+fluxlimcalib = 'No'				### usar limites de fluxo do usuario => Yes ou No
+fluxmincalib = 0				### fluxo minimo do plot
+fluxmaxcalib = 10000				### fluxo maximo do plot
+fluxlimrelat = 'No'				### usar limites de fluxo do usuario => Yes ou No
+fluxminrelat = 0.5				### fluxo minimo do plot
+fluxmaxrelat = 1.5				### fluxo maximo do plot
+
+
+
+#############################################################################################
+
+colunas = np.hstack((coltime, coltarg))
+coments = ['Time (jd)']
+coments = np.hstack((coments, targcom))
+if calib == 'Yes':
+    colunas = np.hstack((colunas, colcalb))
[…]

### 2015-05-16 Irregulares/dados/Irregulares_results/teste/teste.py

@@ -1,0 +1,49 @@
+import numpy as np
+from astropy.time import Time
+from astropy import units
+from astropy.coordinates import SkyCoord
+
+arquivo1 = 'ucac4_Himalia_IAG_filtered_final'
+arquivo2 = 'Himalia_IAG_total.table_filtered'
+arquivo3 = 'Filtros_Flags'
+local = 'IAG'
+
+x = np.loadtxt(arquivo1, usecols=(10, 35, 36, 43), unpack=True)
+filt = np.genfromtxt(arquivo1, usecols=45, dtype='str')
+y = np.loadtxt(arquivo2, usecols=(2, 3, 25), unpack=True)
+
+t = Time(x[3], format='jd', scale='utc')
+u = Time(y[2], format='jd', scale='utc')
+
+k = t
+k.precision = 0
+l = u
+l.precision = 0
+
+b = [local for k in range(len(x[0]))]
+
+n = []
+o = []
+for i in t:
+    m = u - i
+    for index, item in enumerate(m.jd):
+        if -0.5 < item < 0.5:
+            valor = index
+    n.append(y[0][valor])
+    o.append(y[1][valor])
+
+r = SkyCoord(ra=[x[1]]*units.degree, dec=[x[2]]*units.degree, frame='icrs')
+rrad = r.ra.dms.d
+rram = r.ra.dms.m
+rras = r.ra.dms.s
+rdecd = r.dec.dms.d
+rdecm = np.absolute(r.dec.dms.m)
+rdecs = np.absolute(r.dec.dms.s)
+
+#print len(rrad.T), len(rram), len(rras), len(rdecd), len(rdecm), len(rdecs), len(n), len(o), len(t.jd), len(x[0]), len(filt), len(b)
+
+z = np.array(zip(rrad.T, rram.T, rras.T, rdecd.T, rdecm.T, rdecs.T, n, o, t.jd, x[0], filt, b), dtype=[('rrad', int), ('rram', int), ('rras', float), ('rdecd', int), ('rdecm', int), ('rdecs', float), ('n', float), ('o', float), ('t.jd', float), ('x[0]', float), ('filt', 'S8'), ('b', 'S8')])
+
+#np.savetxt('output.txt', data, fmt=["%.3f",]*3 + ["%s"])
[…]

### 2015-05-18 Irregular-occ/Irregular-occ.tex
Adicionado pasta Irregular-occ

--- 2015-05-16 Irregulares/Irregulares.tex
@@
 
-\begin{figure*}
-\begin{centering}
-\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Carme_RA}\label{Fig: carme_alfa}}
-\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Carme_DEC}\label{Fig: carme_delta}}
-\caption{Mean ephemeris offsets and dispersions (1 sigma error bars) in the coordinates of Carme taken night by night by true anomaly for each telescope.
-\textbf{The red square is for the observations with the Perkin-Elmer telescope from OPD, the blue circle for Boller \& Chivens, the magenta triangle down for Zeiss, the black triangle up for OHP and the green star for ESO.}}
-\label{Fig: carme_anom}
-\end{centering}
-\end{figure*}
-
-\begin{figure*}
-\begin{centering}
-\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Pasiphae_RA}\label{Fig: pasiphae_alfa}}
-\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Pasiphae_DEC}\label{Fig: pasiphae_delta}}
-\caption{Same as in Fig \ref{Fig: carme_anom} for Pasiphae.}
-\label{Fig: pasiphae_anom}
-\end{centering}
-\end{figure*}
-
 This pattern in declination was also seen for other satellites like Pasiphae (Fig: \ref{Fig: pasiphae_anom}) and Ananke (plots for other satellites with significant number of observations can be seen in the online material).
@@
 
-\Online
-
-\begin{appendix}
-
-\onecolumn
-
-\section{Ephemeris offsets as a function of true anomaly for all observed irregular satellites} \label{anexo: offsets}
-
-The distribution of ephemeris offsets along the orbit of the satellites are shown below.
-The red square is for the observations with the Perkin-Elmer telescope from OPD, the blue circle for Boller \& Chivens, the magenta triangle down for Zeiss, the black triangle up for OHP and the green star for ESO.
-For Carme and Pasiphae see Figs.
-\ref{Fig: carme_anom} and \ref{Fig: pasiphae_anom} in Section \ref{Sec: comparison}.
-
-\begin{figure}[h!]
-\begin{centering}
-\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Himalia_RA}\label{Fig: an_himalia_alfa}}
-\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Himalia_DEC}\label{Fig: an_himalia_delta}}
-\caption{Mean ephemeris offset and dispersion (1 sigma error bars) in the coordinates of Himalia taken night by night as a function of true anomaly.}
-\label{Fig: an_himalia_anom}
-\end{centering}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{centering}
-\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Elara_RA}\label{Fig: an_elara_alfa}}
-\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Elara_DEC}\label{Fig: an_elara_delta}}
-\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Elara.}
-\label{Fig: an_elara_anom}
-\end{centering}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{centering}
-\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Lysithea_RA}\label{Fig: an_lysithea_alfa}}
-\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Lysithea_DEC}\label{Fig: an_lysithea_delta}}
-\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Lysithea.}
-\label{Fig: an_lysithea_anom}
-\end{centering}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{centering}
-\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Leda_RA}\label{Fig: an_leda_alfa}}
-\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Leda_DEC}\label{Fig: an_leda_delta}}
-\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Leda.}
-\label{Fig: an_leda_anom}
-\end{centering}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{centering}
-\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Carme_RA}\label{Fig: an_carme_alfa}}
-\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Carme_DEC}\label{Fig: an_carme_delta}}
-\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Carme.}
-\label{Fig: an_carme_anom}
-\end{centering}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{centering}
-\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Ananke_RA}\label{Fig: an_ananke_alfa}}
-\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Ananke_DEC}\label{Fig: an_ananke_delta}}
-\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Ananke.}
-\label{Fig: an_ananke_anom}
-\end{centering}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{centering}
-\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Sinope_RA}\label{Fig: an_sinope_alfa}}
-\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Sinope_DEC}\label{Fig: an_sinope_delta}}
-\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Sinope.}
-\label{Fig: an_sinope_anom}
-\end{centering}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{centering}
-\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Phoebe_RA}\label{Fig: an_phoebe_alfa}}
-\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Phoebe_DEC}\label{Fig: an_phoebe_delta}}
-\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Phoebe.}
-\label{Fig: an_phoebe_anom}
-\end{centering}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{centering}
-\subfigure[Right Ascension]{\includegraphics[scale=0.5]{figures/Nereid_RA}\label{Fig: an_nereid_alfa}}
-\subfigure[Declination]{\includegraphics[scale=0.5]{figures/Nereid_DEC}\label{Fig: an_nereid_delta}}
-\caption{Same as in Fig \ref{Fig: an_himalia_anom} for Nereid.}
-\label{Fig: an_nereid_anom}
-\end{centering}
-\end{figure}
-
-\end{appendix}
-
 \end{document}

### 2015-05-20 qualify/Qualify-altair.tex
Adicionado tex da qualificacao

\documentclass[12pt,a4paper]{monografia}
\usepackage[english,brazilian]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amsthm,amsfonts,amssymb,textcomp}
%\usepackage{latexsym}
\usepackage[]{graphicx}
\graphicspath{{figuras}}
\usepackage{subfigure}
\usepackage{float}
\usepackage{longtable}
\usepackage{color}
\usepackage{pdflscape}
%\usepackage[comma,authoryear]{natbib}
\usepackage{hyperref}
\usepackage{pdfpages}
\usepackage[nonumberlist]{glossaries}
%\usepackage[alf]{abntcite}
\bibliographystyle{apalike}

\newcommand{\PE}{Perkin-Elmer}
\newcommand{\BC}{Boller \& Chivens}

\makeglossaries

\input{glossario}

\begin{document}
\doutorado

\titulo{Aplicações Astrométricas e Fotométricas para o Estudo do Sistema Solar Exterior}
\autor{Altair Ramos Gomes Júnior}
\ultimonome{Gomes Júnior}
\nome{Altair Ramos}
\orientador{Marcelo Assafin}
\ttorientador{Professor Doutor}
\curso{Astronomia}
\instituicao{Universidade Federal do Rio de Janeiro}
\sigla{UFRJ}
\unidadeacademica{Centro de Ciências Matemáticas e da Natureza\\Observatório do Valongo}
\ano{2015}
\data{Junho de 2015}
\cidade{Rio de Janeiro}
\examinadorum{Primeira Pessoa}
\examinadordois{Segunda Pessoa}
\examinadortres{Terceira Pessoa}
\examinadorquatro{Quarta Pessoa}
\ttexaminadorum{Doutor}
\ttexaminadordois{Doutor}
\ttexaminadortres{bacharel}
\ttexaminadorquatro{licenciado}
\CDU{521.9}
\areas{Astrometria}
\npaginas{160}

\maketitle

%\begin{agradecimento}{Dedicatória}
%\indent \indent Agradeço à minha mãe Adelma e meu pai Altair por todo o apoio emocional e financeiro durante a minha trajetória acadêmica. Ao meu irmão, que sempre me incentivou, aos meus amigos e colegas pela troca de experiências.\\
%\indent À minha tia Maria do Carmo pelo auxílio dado para minha permanência no Rio de Janeiro, sem a qual não teria sido possível ficar no curso.\\
%\indent Ao meu orientador Marcelo Assafin por ter me aceito como aluno e me dado a oportunidade de trabalhar com as empolgantes áreas de Astrometria e Sistema Solar e por sua paciência ao me ensinar e me atualizar sobre o campo de trabalho.
%\end{agradecimento}

%\newpage
%\begin{epigrafe}
%"Queremos ter certezas e não dúvidas, resultados e não experiências, mas nem mesmo percebemos que as certezas só podem surgir através das dúvidas e os resultados somente através das experiências."\\
%\hfill Carl Gustav Jung
%\end{epigrafe}

\resumo{Resumo}
\indent \indent O estudo da estrutura e evolução do Sistema Solar tem muita importância atualmente, por exemplo, na compreensão dos mecanismos de formação dos planetas de outras estrelas (exoplanetas), trazendo em seus desdobramentos valiosas informações quanto a viabilidade de formação de ambientes que comportem vida.\\
\indent Praticamente cada satélite é um problema particular de Mecânica Celeste. Dessa forma, a observação de seus movimentos é muito útil por razões teóricas. A preparação e realização de missões espaciais para alguns planetas, e mesmo alguns satélites, como Titã de Saturno ou os Galileanos de Júpiter, requer observações frequentes e posições muito acuradas de todos os satélites do sistema planetário visitado.\\
\indent Os satélites irregulares de planetas gigantes, principalmente de Júpiter e de Saturno, são substancialmente menores do que os satélites regulares e possuem órbitas mais distantes, excêntricas, inclinadas e, na maioria dos casos, retrógradas. Explicar a existência dos satélites irregulares dos planetas gigantes é um estudo interessante em dinâmica orbital.\\
\indent É amplamente aceito que eles, devido à configuração orbital, foram capturados por seus planetas. Porém, ainda não são certos a origem e o método de captura desses objetos embora as teorias mais prováveis sejam a de arrasto gasoso durante a formação do Sistema Solar e a captura de três corpos.\\
\indent A maioria dos satélites formam famílias com características orbitais semelhantes sugerindo uma evolução dada por colisões. A compreensão desses mecanismos nos fará entender melhor o nosso próprio sistema solar e, possivelmente, de sistemas extra-solares.\\
\indent Com o pacote astrométrico PRAIA o tempo empregado na redução de grandes quantidades de imagens foi diminuído significativamente. Os objetivos científicos dos nossos programas observacionais agora tem sido atingidos em curtíssimo prazo, em consonância com a atual demanda astronômica e astrofísica de nossa área.\\
\indent Em uma base de dados com mais de 100 mil imagens obtidas entre 1992 e 2012 no telescópios de 1.2m do Observatoire de Haute Province, França, e nos telescópios \PE, \BC{} e Zeiss no Observatório do Pico dos Dias, mais de 4 mil contem satélites irregulares de Júpiter e Saturno. Reduzir essa grande quantidade de observações com precisão só foi possível com a utilização do PRAIA. Como é possível ver nos resultados obtidos, os offsets de posição refletem o erro natural da astrometria e o erro das efemérides dos satélites mais fracos, e sugerem que podemos efetivamente contribuir para que novas integrações numéricas forneçam posições mais precisas. Essas novas efemérides irão permitir que ocultações estelares por esses objetos possam ser preditas com maior precisão. A observações dessas ocultações permitirão um maior conhecimento das propriedades físicas dos satélites.

\par
\vspace{1em}
\noindent\textbf{Palavras-chave:} Astrometria, Satélites Irregulares de Júpiter, Satélites Irregulares de Saturno

\tableofcontents % para gerar o sumário
\thispagestyle{empty} % para que a pagina não seja enumerada
%\listoffigures % cria a lista de figuras
%\thispagestyle{empty}
%\listoftables % cria a lista de tabela
%\thispagestyle{empty}

\pagestyle{ruledheader}

\chapter{Introdução}
\label{Cap:intro}

\chapter{Observações}
\label{Cap:observacoes}

\chapter{Tratamento de Imagens. Obtenção de Posições ($\alpha$,$ \delta$)}
\label{Cap:reducao}

\chapter{Resultados}
\label{Cap:resultados}

\chapter{Discussão}%graf anom. verdad. e ceu(ind,media por noite)
%discussão sobre provável contribuição para melhorar elementos orbitais
\label{Cap:Analise}

\chapter{Conclusões}
\label{Cap:Conclusao}

\glsaddall
\printglossary

\bibliography{Referencias}

\end{document}

### 2015-05-20 qualify/glossario.tex
Adicionado tex da qualificacao

\newglossaryentry{Offset}{name={Offset}, description={Diferença entre a posição obtida pela redução da observação e a posição dada pela efeméride}}
\newglossaryentry{OPD}{name={OPD}, description={Observatório do Pico dos Dias - Brasópolis, MG}}
\newglossaryentry{LNA}{name={LNA}, description={Laboratório Nacional de Astrofísica - Itajubá, MG}}
\newglossaryentry{OHP}{name={OHP}, description={Observatoire Haute Provence - Saint-Michel-l'Observatoire, França}}
\newglossaryentry{RA}{name={RA}, description={Sigla para Ascensão Reta ($ \alpha $)}}
\newglossaryentry{DEC}{name={DEC}, description={Sigla para Declinação ($ \delta $)}}
\newglossaryentry{Anomalia Verdadeira}{name={Anomalia Verdadeira}, description={Ângulo formado entre o Periastro e a posição instantânea do objeto na órbita centrado no planeta e contada na direção do movimento orbital}}

### 2015-05-20 qualify/comando-pdf.sh

@@ -1,0 +1,1 @@
+pdftk Qualify-altair.pdf cat 1 5-end output Qualify-altair_output.pdf

### 2015-05-25 qualify/Qualify-altair.tex
Iniciado introdução

@@
 \usepackage{hyperref}
-\usepackage{pdfpages}
 \usepackage[nonumberlist]{glossaries}
@@
 
-\chapter{Observações}
-\label{Cap:observacoes}
+\indent \indent O estudo de objetos como TNOs, Centauros e Satélites Irregulares (remanescentes relativamente inalterados da formação do sistema solar) nos ajudam a compreender a formação e evolução do Sistema Solar.
+Atualmente, é aceito que TNOs e Centauros tenham sido formados nas partes mais internas do sistema solar.
+Eles teriam então sido colocados em suas posições atuais devido a troca de momento angular entre os planetas e planetésimos quando da migração dos planetas gigantes.
+A evolução se deu de tal forma que a passagem dos planetesimais e planetas por zonas de ressonância de movimento médio redefiniu as órbitas desses corpos \cite{Tsiganis2005}.\\
+\indent Sabe-se que poucas sondas espaciais foram enviados para estudar o Sistema Solar Externo e que a quantidade de objetos estudados é muito pequena.
+Por isso, ainda hoje, as observações de solo tem se mostrado de grande importância.\\
+\indent Os sistemas de Júpiter e Saturno já foram pelas Voyager I e II, Galileu (Júpiter) e Cassini (Saturno), porém apenas Saturno continua sendo investigado por uma sonda.
+Porém as sondas observaram apenas os planetas, os anéis e satélites mais internos.
+Os satélites externos, que acredita-se ser oriundo de capturas ou foram pouco observados (como Phoebe) ou simplesmente não foram observados.\\
+\indent Já no caso de Urano e Netuno, nenhuma sonda exclusiva foi enviada, apenas as Voyagers I e II os visitaram, mas não permaneceram nos sistemas.
+A sonda New Horizons visitará Plutão em 2015 e obterá parâmetros físicos para Plutão e seus satélites (primeira visita por sonda a um satélite do cinturão de Kuiper), porém será uma passagem rápida e o acompanhamento da evolução do sistema, incluindo a evolução da atmosfera de Plutão se dará por observações de solo.\\
+\indent A quantidade de objetos descobertos além da órbita de Saturno tem aumentado muito desde o fim do século passado.
+Como são raras as oportunidades em que uma sonda se aproxima desses objetos, a obtenção de suas características físicas ficam a cargo de observações de solo ou de
+telescópios espaciais.\\
+\indent Um método que tem se mostrado eficiente para a obtenção desses parâmetros é o métodode ocultações estelares, que proporciona medidas tão precisas que são apenas superadas por
+medidas oriundas de sondas.
 
-\chapter{Tratamento de Imagens. Obtenção de Posições ($\alpha$,$ \delta$)}
-\label{Cap:reducao}
+\chapter{Ocultações}
+\label{Cap: observacoes}
 
-\chapter{Resultados}
-\label{Cap:resultados}
+\section{TNOs}
+\label{Sec: TNOs}
 
-\chapter{Discussão}%graf anom. verdad. e ceu(ind,media por noite)
-%discussão sobre provável contribuição para melhorar elementos orbitais
-\label{Cap:Analise}
+\section{Ceres}
+\label{Sec: Ceres}
 
-\chapter{Conclusões}
-\label{Cap:Conclusao}
+\section{Satélites Irregulares}
+\label{Sec: Irregulares}
+
+\chapter{Astrometria de Netuno e Tritão}
+\label{Cap: Netuno}
+
+%\chapter{Resultados}
+%\label{Cap:resultados}
+%
+%
+%
+%\chapter{Discussão}%graf anom. verdad. e ceu(ind,media por noite)
+%%discussão sobre provável contribuição para melhorar elementos orbitais
+%\label{Cap:Analise}
+%
+%
+%\chapter{Conclusões}
+%\label{Cap:Conclusao}
 

### 2015-05-25 qualify/Qualify-altair.tex
Adicionado figuras de Ceres

@@
 \usepackage{color}
+\usepackage{epstopdf}
 \usepackage{pdflscape}
-%\usepackage[comma,authoryear]{natbib}
+\usepackage[comma,authoryear]{natbib}
 \usepackage{hyperref}
@@
 Eles teriam então sido colocados em suas posições atuais devido a troca de momento angular entre os planetas e planetésimos quando da migração dos planetas gigantes.
-A evolução se deu de tal forma que a passagem dos planetesimais e planetas por zonas de ressonância de movimento médio redefiniu as órbitas desses corpos \cite{Tsiganis2005}.\\
+A evolução se deu de tal forma que a passagem dos planetesimais e planetas por zonas de ressonância de movimento médio redefiniu as órbitas desses corpos \citep{Tsiganis2005}.\\
 \indent Sabe-se que poucas sondas espaciais foram enviados para estudar o Sistema Solar Externo e que a quantidade de objetos estudados é muito pequena.
@@
 telescópios espaciais.\\
-\indent Um método que tem se mostrado eficiente para a obtenção desses parâmetros é o métodode ocultações estelares, que proporciona medidas tão precisas que são apenas superadas por
-medidas oriundas de sondas.
+\indent Um método que tem se mostrado eficiente para a obtenção desses parâmetros é o métodode ocultações estelares, que proporciona medidas tão precisas que são apenas superadas por medidas oriundas de sondas.
 
@@
 \label{Sec: Ceres}
+
+\begin{figure}
+\begin{centering}
+\subfigure[Ocultação de 2010]{\includegraphics[scale=0.39]{figuras/Ceres_2010.eps}}
+\subfigure[Ocultação de 2013]{\includegraphics[scale=0.39]{figuras/Ceres_2013-zoom.eps}}
+\caption{Reconstrução pós-ocultação do caminho da sombra de Ceres na Terra para os eventos de 17 de Agosto de 2010 (a) e 25 de Outubro de 2013 (b).
+Os pontos em azul são os sítios que observaram os eventos.
+a) O ponto vermelho grande é a máxima aproximação geocêntrica às 22:40:25 UT.
+Os pequenos representam o centro da sombra separados por um minuto.
+b) Visão superior da ocultação sobre os sítios que observaram o evento.
+Os pontos vermelhos são os centros da sombra separados por 15 segundos.
+Nos dois eventos a sombra se move da esquerda pra direita.
+\label{Fig: Ceres-map}}
+\end{centering}
+\end{figure}
+
+\begin{figure}
+\begin{centering}
+\subfigure[Ocultação de 2010]{\includegraphics[angle=-90, scale=0.31]{figuras/Ceres_2010_sphere.eps} }
+\subfigure[Ocultação de 2013]{\includegraphics[scale=0.31]{figuras/Ceres_2013_sphere.pdf} }
+\caption{Melhores ajustes elípticos para as cordas das ocultações de 2010 e 2013.
+As setas indicam a direção de movimento, as linhas azuis são as cordas observadas, os segmentos vermelhos são as barras de erro dos ingressos, egressos e centro da ocultação em 1 $\sigma$.
+\label{Fig: Ceres-limb}}
+\end{centering}
+\end{figure}
 

### 2015-05-26 qualify/Qualify-altair.tex
Comecado a escrever Ceres

@@
 \chapter{Introdução}
-\label{Cap:intro}
+\label{Cap: intro}
 
@@
 
-\section{TNOs}
-\label{Sec: TNOs}
-
 \section{Ceres}
@@
 
+\indent \indent Apesar de Ceres não ser um objeto do sistema solar exterior, ele é o único planeta-anão no sistema solar interno e, por isso, é um objeto de grande importância e seu estudo pode ter grande impacto na formação e evolução do sistema solar.
+Na verdade, foi proposto que a origem de Ceres pode ser como um objeto transnetuniano \citep{McKinnon2012}, espalhado posteriormente para o cinturão principal de asteroides devido à migração dos planetas gigantes predito pelo Modelo de Nice \citep{Gomes2005}.
+Mesmo que ele tenha sido formado próximo à sua localização atual, a história dinâmica do sistema solar deve ter deixado sua assinatura em Ceres.\\
+\indent Contendo aproximadamente um quinto de toda a massa do cinturão de asteroides, espera-se que Ceres esteja em equilíbrio gravitacional e seja, portanto, um elipsóide Maclaurin ou Jacobi.
+De fato, observações diretas de Ceres com a utilização de ótica adaptativa indica que ele é um esferóide achatado nos pólos \citep{Drummond2014}.
+O conhecimento preciso de seu tamanho e forma é extrema importância para modelos de densidade, estrutura interna e diferenciação.
+
 \begin{figure}
 \begin{centering}
-\subfigure[Ocultação de 2010]{\includegraphics[scale=0.39]{figuras/Ceres_2010.eps}}
-\subfigure[Ocultação de 2013]{\includegraphics[scale=0.39]{figuras/Ceres_2013-zoom.eps}}
+\subfigure[Ocultação de 2010]{\includegraphics[scale=0.39]{figuras/Ceres_2010.eps}\label{Fig: Ceres-map-2010}}
+\subfigure[Ocultação de 2013]{\includegraphics[scale=0.39]{figuras/Ceres_2013-zoom.eps}\label{Fig: Ceres-map-2013}}
 \caption{Reconstrução pós-ocultação do caminho da sombra de Ceres na Terra para os eventos de 17 de Agosto de 2010 (a) e 25 de Outubro de 2013 (b).
 Os pontos em azul são os sítios que observaram os eventos.
-a) O ponto vermelho grande é a máxima aproximação geocêntrica às 22:40:25 UT.
+a) O ponto grande vermelho é a máxima aproximação geocêntrica às 22:40:25 UT.
 Os pequenos representam o centro da sombra separados por um minuto.
-b) Visão superior da ocultação sobre os sítios que observaram o evento.
+b) Visão superior da ocultação sobre os sítios que observaram o evento de 25 de Outubro de 2013.
 Os pontos vermelhos são os centros da sombra separados por 15 segundos.
@@
 \begin{centering}
-\subfigure[Ocultação de 2010]{\includegraphics[angle=-90, scale=0.31]{figuras/Ceres_2010_sphere.eps} }
-\subfigure[Ocultação de 2013]{\includegraphics[scale=0.31]{figuras/Ceres_2013_sphere.pdf} }
+\subfigure[Ocultação de 2010]{\includegraphics[scale=0.35]{figuras/Ceres_2010_sphere.eps}\label{Fig: Ceres-limb-2010} }
+\subfigure[Ocultação de 2013]{\includegraphics[scale=0.35]{figuras/Ceres_2013_sphere.pdf}\label{Fig: Ceres-limb-2013} }
 \caption{Melhores ajustes elípticos para as cordas das ocultações de 2010 e 2013.
-As setas indicam a direção de movimento, as linhas azuis são as cordas observadas, os segmentos vermelhos são as barras de erro dos ingressos, egressos e centro da ocultação em 1 $\sigma$.
+As setas indicam a direção de movimento, as linhas azuis são as cordas observadas, os segmentos vermelhos são as barras de erro dos ingressos, egressos e centro da ocultação em $1\sigma$.
+A linha verde em (a) é uma corda negativa.
+Os instantes marcados em verde em (b) não foram utilizados para o ajuste como descrito no texto.
 \label{Fig: Ceres-limb}}
@@
 
+\section{TNOs}
+\label{Sec: TNOs}
+
 \chapter{Astrometria de Netuno e Tritão}
 \label{Cap: Netuno}
-
-%\chapter{Resultados}
-%\label{Cap:resultados}
-%
-%
-%
-%\chapter{Discussão}%graf anom. verdad. e ceu(ind,media por noite)
-%%discussão sobre provável contribuição para melhorar elementos orbitais
-%\label{Cap:Analise}
-%
-%
-%\chapter{Conclusões}
-%\label{Cap:Conclusao}
 

### 2015-05-27 qualify/Qualify-altair.tex
continuado a escrever a parte de Ceres

@@
 Na verdade, foi proposto que a origem de Ceres pode ser como um objeto transnetuniano \citep{McKinnon2012}, espalhado posteriormente para o cinturão principal de asteroides devido à migração dos planetas gigantes predito pelo Modelo de Nice \citep{Gomes2005}.
-Mesmo que ele tenha sido formado próximo à sua localização atual, a história dinâmica do sistema solar deve ter deixado sua assinatura em Ceres.\\
-\indent Contendo aproximadamente um quinto de toda a massa do cinturão de asteroides, espera-se que Ceres esteja em equilíbrio gravitacional e seja, portanto, um elipsóide Maclaurin ou Jacobi.
+Mesmo que ele tenha sido formado próximo à sua localização atual, a história dinâmica do sistema solar deve ter deixado sua assinatura em Ceres.
+
+Contendo aproximadamente um quinto de toda a massa do cinturão de asteroides, espera-se que Ceres esteja em equilíbrio gravitacional e seja, portanto, um elipsóide Maclaurin ou Jacobi.
 De fato, observações diretas de Ceres com a utilização de ótica adaptativa indica que ele é um esferóide achatado nos pólos \citep{Drummond2014}.
@@
 
-\begin{figure}
+A primeira ocultação estelar por Ceres foi observada em 1984 \citep{Millis1987} e determinou seu tamanho com precisão de alguns quilômetros em uma época que as incertezas eram, normalmente, dez vezes maiores.
+Devido ao brilho aparente de Ceres ser alto, comparado à maioria dos asteroides, estrelas capazes de causar uma queda de magnitude detectável quando ocultadas são limitadas às mais brilhantes.
+Por exemplo, depois do evento de 1984, apenas 4 ocultações estelares por Ceres foram observadas \citep{Dunham2014}.
+Duas delas tiveram apenas duas cordas cada que não foram suficientes para prover resultados acurados\footnote{Esses eventos ocorreram em 22 de Agosto de 1994 e 30 de Outubro de 2010.}.
+Os dois restantes, que ocorreram em 17 de Agosto de 2010 e 25 de Outubro de 2013, foram trabalhos por mim em colaboração com o grupo do Rio \citep[aceito]{GomesJunior2015-Ceres}.
+
+Os dois eventos foram preditos por Steve Preston\footnote{Predições publicadas em http://asteroidoccultation.com.} para a IOTA (International Occultation Timing Association), durante predições de rotina de ocultações de asteroides de estrelas brilhantes.
+Os caminhos das sombras podem ser visualizados na Fig.
+\ref{Fig: Ceres-map}.
+
+\begin{figure}[h]
 \begin{centering}
@@
 Os pontos vermelhos são os centros da sombra separados por 15 segundos.
-Nos dois eventos a sombra se move da esquerda pra direita.
+Nos dois eventos a sombra se move da esquerda para a direita.
 \label{Fig: Ceres-map}}

### 2015-05-27 Irregulares/corecoes/26273_corr/26273_ja.tex
Continuado a escrever Ceres no qualify

--- 2015-05-16 Irregulares/AA-2015-26273/Irregulars.tex
@@
 \title{Astrometric positions for 18 irregular satellites of giant planets from 23 years of observations
-\fnmsep\thanks{The complete version of Table 8 is available through CDS and IAU NSDC data base at www.imcce.fr/nsdc.}
-\fnmsep\thanks{Partially based on observations made at Laborat\'orio Nacional de Astrof\'{\i}sica (LNA), Itajub\'a-MG, Brazil.}
+\fnmsep\thanks{The complete version of Table 8 is available through the CDS and IAU NSDC data base at www.imcce.fr/nsdc.}
+\fnmsep\thanks{Partially based on observations made at the Laborat\'orio Nacional de Astrof\'{\i}sica (LNA), Itajub\'a-MG, Brazil.}
 \fnmsep\thanks{Partially based on observations through the ESO runs 079.A-9202(A), 075.C-0154, 077.C-0283 and 079.C-0345.}
 \fnmsep\thanks{\textbf{Partially based on observations made at
-Observatoire de Haute Provence (OHP), F-04870 Saint-Michel l'observatoire, France}}
+the Observatoire de Haute Provence (OHP), F-04870 Saint-Michel l'observatoire, France}}
 }
@@
 {The irregular satellites of the giant planets are believed to have been captured during the evolution of the solar system.
-Knowing their physical parameters, such as size, density and albedo is important to constrain where they came from and how they were captured.
+Knowing their physical parameters, such as size, density, and albedo is important for constraining where they came from and how they were captured.
 The best way to obtain these parameters are observations in situ by spacecrafts or from stellar occultations by the objects.
 Both techniques demand that the orbits are well known.}
-{We aimed to obtain good astrometric positions of irregular satellites in order to improve their orbits and ephemeris.}
-{We identified and reduced observations of several irregular satellites from three databases containing more than 8000 images obtained between 1992 and 2014 at three sites (Observatório do Pico dos Dias, Observatoire de Haute-Provence and European Southern Observatory - La Silla).
+{We aimed to obtain good astrometric positions of irregular satellites to improve their orbits and ephemeris.}
+{We identified and reduced observations of several irregular satellites from three databases containing more than 8000 images obtained between 1992 and 2014 at three sites (Observatório do Pico dos Dias, Observatoire de Haute-Provence, and European Southern Observatory - La Silla).
 We used the software PRAIA (Platform for Reduction of Astronomical Images Automatically) to make the astrometric reduction of the CCD frames.
-The UCAC4 catalogue represented the International Celestial Reference System in the reductions.
-The identification of the satellites in the frames was done through their ephemerides as determined from the SPICE/NAIF kernels.
-Some procedures were taken to overcome missing or incomplete information (coordinates, date), mostly for the older images.}
-{We managed to obtain more than 6000 positions for 18 irregular satellites, being 12 of Jupiter, 4 of Saturn, 1 of Uranus (Sycorax) and 1 of Neptune (Nereid).
-For some satellites the number of obtained positions is more than 50\% of that used in earlier orbital numerical integrations.}
-{Comparison of our positions with recent JPL ephemeris suggests the presence of systematic errors in the orbits for some of the irregular satellites.
+The UCAC4 catalog represented the International Celestial Reference System in the reductions.
+Identification of the satellites in the frames was done through their ephemerides as determined from the SPICE/NAIF kernels.
+Some procedures were followed\LEt{ Or "used".} to overcome missing or incomplete information (coordinates, date), mostly for the older images.}
+{We managed to obtain more than 6000 positions for 18 irregular satellites: 12 of Jupiter, 4 of Saturn, 1 of Uranus (Sycorax), and 1 of Neptune (Nereid).
+For some satellites, the number of obtained positions is more than 50\% of what was used in earlier orbital numerical integrations.}
+{Comparison of our positions with recent JPL ephemeris suggests there are systematic errors in the orbits for some of the irregular satellites.
 The most evident case was an error in the inclination of Carme.}
@@
 
-The irregular satellites of the giant planets are smaller than the regular moons, having more eccentric, inclined, distant and, in most cases, retrograde orbits.
-Due to their orbital configurations, it is largely accepted that these objects were captured in the early solar system \citep{Sheppard2003}.
+The irregular satellites of the giant planets are smaller than the regular moons, having more eccentric, inclined, distant, and in most cases, retrograde orbits.
+Owing to their orbital configurations, it is largely accepted that these objects were captured in the early solar system \citep{Sheppard2003}.
 
-Because they are faint, the majority of these objects was discovered only in the last century\footnote{Website: http://ssd.jpl.nasa.gov/?sat\_discovery} .
-They were never visited by a spacecraft, with the exception of Himalia, Phoebe and Nereid, in a flyby by the Cassini space probe in 2000 for Himalia \citep{Porco2003} and in 2004 for Phoebe \citep{Desmars2013} and in a flyby by the Voyager 2 space probe in 1989 for Nereid \citep{Smith1989}.
+Because they are faint, the majority of these objects were only discovered in the last century\footnote{Website: http://ssd.jpl.nasa.gov/?sat\_discovery} .
+They were never visited by a spacecraft, with the exception of Himalia, Phoebe, and Nereid, in a flyby by the Cassini space probe in 2000 for Himalia \citep{Porco2003} and in 2004 for Phoebe \citep{Desmars2013} and in a flyby by the Voyager 2 space probe in 1989 for Nereid \citep{Smith1989}.
 Even in situ, they were still opportunity target observations resulting in not optimal measurements, with size errors of $10 km$ for Himalia and $25 km$ for Nereid \citep{Thomas1991}.
@@
 If these objects were captured, there remains the question of where they came from.
-\citealp{Clark2005} showed from imaging spectroscopy from Cassini that Phoebe has a surface probably covered by material from the outer solar system and \citealp{Grav2003} showed that the satellites of the Jovian Prograde Group Himalia have grey colors implying that their surfaces are similar to that of C-type asteroids.
-In that same work, the Jovian Retrograde Group Carme was found to have surface colors similar to the D-type asteroids like Hilda or Trojan families while JXIII Kalyke has a redder color like Centaurs or trans-neptunian objects (TNOs).
+\citealp{Clark2005} show from imaging spectroscopy from Cassini that Phoebe has a surface probably covered by material from the outer solar system and \citealp{Grav2003} show that the satellites of the Jovian Prograde Group Himalia have gray colors, implying that their surfaces are similar to that of C-type asteroids.
+In that same work, the Jovian Retrograde Group Carme was found to have surface colors similar to the D-type asteroids as for the Hilda or Trojan families, while JXIII Kalyke has a redder color like Centaurs or trans-Neptunian objects (TNOs).
 
-For Saturnian satellites, \citealp{Grav2007} showed by their colors and spectral slopes that these satellites contain a more or less equal fraction of C-, P- and D-like objects but SXXII Ijiraq is marginally redder than D-type objects.
+For Saturnian satellites, \citealp{Grav2007} show by their colors and spectral slopes that these satellites contain a more or less equal fraction of C-, P-, and D-like objects, but SXXII Ijiraq is marginally redder than D-type objects.
 These works may suggest different origins for the irregular satellites.
 
-In this context, we used 3 databases for deriving precise positions for the irregular satellites observed at Observatório do Pico dos Dias (1.6 m and 0.6 m telescopes, IAU code 874), Observatoire Haute-Provence (1.2m telescope, IAU code 511) and ESO (2.2 m telescope, IAU code 809).
-Many irregular satellites were observed between 1992 and 2014 covering a few orbital periods of these objects (12 satellites of Jupiter, 4 of Saturn, Sycorax of Uranus and Nereid of Neptune).
+In this context, we used three databases for deriving precise positions for the irregular satellites observed at the Observatório do Pico dos Dias (1.6 m and 0.6 m telescopes, IAU code 874), the Observatoire Haute-Provence (1.2m telescope, IAU code 511), and ESO (2.2 m telescope, IAU code 809).
+Many irregular satellites were observed between 1992 and 2014, covering a few orbital periods of these objects (12 satellites of Jupiter, 4 of Saturn, Sycorax of Uranu,s and Nereid of Neptune).
 
-\textbf{Since their ephemerides are not very precise, predict and observe stellar occultations are very difficult and no observation of such an event for an irregular satellite is found in the literature.
+\textbf{Since their ephemerides are not very precise, predicting and observing stellar occultations are very difficult, and no observation of such an event for an irregular satellite is found in the literature.
 The precise star positions to be derived by the ESA astrometry satellite Gaia \citep{deBruijne2012} will render better predictions with the only source of error being the ephemeris.
@@
 
-\textbf{The power of stellar occultations for observing relatively small diameter solar system objects is supported by recent works such as the discovery of a ring system around the Centaur (10199) Chariklo \citep{Braga-Ribas2014}.
+\textbf{The power of stellar occultations for observing relatively small diameter solar system objects is supported by recent works, such as the discovery of a ring system around the Centaur (10199) Chariklo \citep{Braga-Ribas2014}.
 Once irregular satellites start to be observed by this technique, it will be possible to obtain their physical parameters (shape, size, albedo, density) with unprecedented precision.
 For instance, in this case, sizes could be obtained with kilometer accuracy.
-The knowledge of these parameters would in turn bring valuable information for the study of the capture mechanisms and origin of the irregular satellites.}
+The knowledge of these parameters would in turn bring valuable information for studying the capture mechanisms and origin of the irregular satellites.}
 
@@
 \ref{Sec: reduction}.
-The obtained positions are presented in Sect \ref{Sec: positions} and analysed in Sect.
+The obtained positions are presented in Sect \ref{Sec: positions} and analyzed in Sect.
 \ref{Sec: comparison}.
@@
 
-Our three databases consist in optical CCD images from many observational programs performed with different telescopes/detectors targeting a variety of objects, among which irregular satellites.
-The observations were made at 3 sites: Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO).
-Altogether there are more than 8000 FITS images obtained in a large time span (1992-2014) for the irregular satellites.
-Since the OHP and mostly the OPD database registers were not well organized, we had to start from scratch and develop an automatic procedure to identify and filter only the images of interest, that is, of the irregular satellites.
-The instruments and images characteristics are described in the following subsections.
+Our three databases consist of optical CCD images from many observational programs performed with different telescopes and detectors that target a variety of objects, among which are irregular satellites.
+The observations were made at three sites: Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP), and the European Southern Observatory (ESO).
+All together there are more than 8000 FITS images obtained in a large time span (1992-2014) for the irregular satellites.
+Since the OHP and mostly the OPD database registers were not well organized, we had to start from scratch and develop an automatic procedure to identify and filter only the images of interest, that is, for the irregular satellites.
+The instrument and image characteristics are described in the following sections.
 
@@
 
-The OPD database was produced at Observatório do Pico dos Dias (OPD, IAU code 874, $45^{\circ} ~34\arcmin ~57\arcsec$ W, $22^{\circ} ~32\arcmin ~04\arcsec$ S, 1864 m)\footnote{Website: http://www.lna.br/opd/opd.html - in Portuguese}, located at geographical longitude , in Brazil.
+The OPD database was produced at the Observatório do Pico dos Dias (OPD, IAU code 874, $45^{\circ} ~34\arcmin ~57\arcsec$ W, $22^{\circ} ~32\arcmin ~04\arcsec$ S, 1864 m)\footnote{Website: http://www.lna.br/opd/opd.html - in Portuguese}, located at geographical longitude, in Brazil.
 The observations were made between 1992 and 2014 by our group in a variety of observational programs.
@@
 Two telescopes of 0.6 m diameter (Zeiss and Boller \& Chivens) and one 1.6 m diameter (Perkin-Elmer) were used for the observations.
-It was identified 5248 observations containing irregular satellites, being 3168 from the Boller \& Chivens, 1967 from the Perkin-Elmer and 113 from the Zeiss.
+Identified were 5248 observations containing irregular satellites, with 3168 from the Boller \& Chivens, 1967 from the Perkin-Elmer, and 113 from the Zeiss.
 
@@
 
-This is an inhomogeneous database with observations made with 9 different detectors (see Table \ref{Tab: OPD-CCDs}) and 6 different filters.
-The headers of most of the older FITS images had missing, incomplete or incorrect coordinates or date.
-In some cases, we could not identify the detector origin.
+This is an inhomogeneous database with observations made with nine different detectors (see Table \ref{Tab: OPD-CCDs}) and six different filters.
+The headers of most of the older FITS images had missing, incomplete, or incorrect coordinates or dates.
+In some cases, we could not identify the detector's origin.
 The procedures used to overcome these problems are described in Sect.
@@
 % & \multicolumn{2}{c}{Exposure Times}\tabularnewline
-Detector & Image Size (pixel) & Pixel Scale ($\mu m$/px)\tabularnewline
+Detector & Image size (pixel) & Pixel Scale ($\mu m$/px)\tabularnewline
 \hline
@@
 \par\end{centering}
-The plate scale of the telescopes are 13.09"/mm for Perkin-Elmer, 25.09"/mm for Boller \& Chivens and 27.5"/mm for Zeiss.
+The plate scale of the telescopes are 13.09"/mm for Perkin-Elmer, 25.09"/mm for Boller \& Chivens, and 27.5"/mm for Zeiss.
 \end{table}
@@
 Observations were made at the 2.2 m Max-Planck ESO (ESO2p2) telescope (IAU code 809, $70\degr44\arcmin1.5\arcsec$ W, $29\degr15\arcmin31.8\arcsec$ S, 2345.4 m)\footnote{Website: www.eso.org/sci/facilities/lasilla/telescopes/\\national/2p2.html} with the Wide Field Imager (WFI) CCD mosaic detector.
-Each mosaic is composed by eight CCDs of $7.5\arcmin \times 15\arcmin$ ($\alpha$, $\delta$) sizes, resulting in a total coverage of $30\arcmin \times 30\arcmin$ per mosaic.
+Each mosaic is composed of eight CCDs of $7.5\arcmin \times 15\arcmin$ ($\alpha$, $\delta$) sizes, resulting in a total coverage of $30\arcmin \times 30\arcmin$ per mosaic.
 Each CCD has $4 k \times 2 k$ pixels with a pixel scale of $0.238\arcsec$.
@@
 
-The satellites were observed in 24 nights, divided in 5 runs, between April 2007 and May 2009 in paralel with, and using the same observational and astrometric procedures of the program that observed stars along the sky path of trans-neptunian objects (TNOs) to identify candidates to stellar occultation \citep[see][]{2010A&A...515A..32A, 2012A&A...541A.142A}.
-A total of 810 observations for irregular satellites were obtained.
+The satellites were observed in 24 nights, divided in five runs between April 2007 and May 2009 in parallel with, and using the same observational and astrometric procedures of the program that observed stars along the sky path of trans-Neptunian objects (TNOs) to identify candidates for stellar occultation \citep[see][]{2010A&A...515A..32A, 2012A&A...541A.142A}.
+A total of 810 observations were obtained for irregular satellites.
 %Fig \ref{Fig: obsxsat-eso} shows the number of frames per satellite.
@@
 Almost all the frames were photometrically calibrated with auxiliary bias and flat-field frames by means of standard procedures using IRAF\footnote{Website: http://iraf.noao.edu/} and, for the mosaics, using the esowfi \citep{Jones2000} and mscred \citep{Valdes1998} packages.
-Some of the nights at OPD didn't have bias and flat-field images so the correction was not possible.
+Some of the nights at OPD did not have bias and flat-field images so the correction was not possible.
 
 The astrometric treatment was made with the Platform for Reduction of Astronomical Images Automatically (PRAIA) \citep{2011gfun.conf...85A}.
-The (x, y) measurements were performed with 2-dimensional circular symmetric Gaussian fits within one Full Width Half Maximum (FWHM = seeing).
-Within one FWHM, the image profile is well described by a Gaussian profile, free from the wing distortions, which may jeopardize the center determination.
+The (x,y) measurements were performed with two-dimensional circular symmetric Gaussian fits within one full width half maximum (FWHM = seeing).
+Within one FWHM, the image profile is described well by a Gaussian profile, which is free of the wing distortions and may jeopardize the center determination\LEt{ Do you mean "the determination of the center" here?}.
 PRAIA automatically recognizes catalog stars and determines ($\alpha$, $\delta$) with a user-defined model relating the (x, y) measured and (X, Y) standard coordinates projected in the sky tangent plane.
@@
 For each frame, we used the six constants polynomial model to relate the (x, y) measurements with the (X, Y) tangent plane coordinates.
-For ESO, we followed the same astrometric procedures described in detail in \cite{2012A&A...541A.142A}; the (x, y) measurements of the individual CCDs were pre-corrected by a field distortion pattern, and all positions coming from different CCDs and mosaics were then combined using a 3rd degree polynomial model to produce a global solution for each night and field observed, and final ($\alpha$, $\delta$) object positions were obtained in the UCAC4 system.
+For ESO, we followed the same astrometric procedures as described in detail in \cite{2012A&A...541A.142A}; the (x, y) measurements of the individual CCDs were precorrected by a field distortion pattern, and all positions coming from different CCDs and mosaics were then combined using a third-degree polynomial model to produce a global solution\LEt{ A complete solution?\ Or do you mean the faux ami sense of "general" or "overall"?\ If so, use one of those words instead.} for each night and field observed, and final ($\alpha$, $\delta$) object positions were obtained in the UCAC4 system.
 
-In Table \ref{Tab: stars-errors} we list the average mean error in $\alpha$ and $\delta$ for the reference stars obtained by telescope, the average (x, y) measurement errors of the Gaussian fits described above and the mean number of UCAC4 stars used by frame.
+In Table \ref{Tab: stars-errors} we list the average mean error in $\alpha$ and $\delta$ for the reference stars obtained by telescope, the average (x, y) measurement errors of the Gaussian fits described above, and the mean number of UCAC4 stars used by frame.
 For all databases, about 20\% of outlier reference stars were eliminated for presenting (O-C) position residuals higher than 120 mas in the ($\alpha$, $\delta$) reductions.
@@
 
-To help identifying the satellites in the frames, and derive the ephemeris for the instants of the observations for comparisons (see Sect \ref{Sec: comparison}), we used the kernels from SPICE/JPL\footnote{Website: http://naif.jpl.nasa.gov/naif/toolkit.html}.
+To help identify the satellites in the frames and derive the ephemeris for the instants of the observations for comparisons (see Sect \ref{Sec: comparison}), we used the kernels from SPICE/JPL\footnote{Website: http://naif.jpl.nasa.gov/naif/toolkit.html}.
 \cite{Emelyanov2008} and references therein also provided ephemeris of similar quality for the irregular satellites.
-For instance, for Himalia, which has relatively good orbit solutions, the ephemerides differ by less than $20 mas$, and in the case of less-known orbits, like Ananke, the differences are less than $90 mas$.
-Here, we choose to use the JPL ephemeris because they used more recent observations \cite[see][]{Jacobson2012}.
+For instance, for Himalia, which has relatively good orbit solutions, the ephemerides differ by less than $20~ mas$, and in the case of less-known orbits, like Ananke, the differences are less than $90 ~mas$.
+We chose to use the JPL ephemeris because they used more recent observations \cite[see][]{Jacobson2012}.
 The JPL ephemeris that represented the Jovian satellites in this work was the DE421 $+$ JUP300.
-For the Saturnian satellites the ephemeris was DE421 $+$ SAT359 to Hyperion, Iapetus and Phoebe and DE421 $+$ SAT361 to Albiorix, Siarnaq and Paaliaq.
+For the Saturnian satellites, the ephemeris was DE421 $+$ SAT359 to Hyperion, Iapetus, and Phoebe and DE421 $+$ SAT361 to Albiorix, Siarnaq, and Paaliaq.
 The DE421 $+$ URA095 was used for Sycorax and DE421 $+$ NEP081 for Nereid.
@@
 
-In the OPD database, there were some images (mostly the older ones) with missing coordinates or wrong date in their headers.
-In the case of missing or wrong coordinates, we adopted the ephemeris as the central coordinates of the frames.
+In the OPD database, there were some images (mostly the older ones) with missing coordinates or the wrong date in their headers.
+In the case of missing or incorrect coordinates, we adopted the ephemeris as the central coordinates of the frames.
 When the time was not correct, the FOV identification failed.
-In this case, a search for wrong date (year) displaying was performed.
+In this case, a search for displays of the wrong date (year) was performed.
 Problems like registering local time instead of UTC were also identified and corrected.
@@
 In all databases, for each night a sigma-clipping procedure was performed to eliminate discrepant positions (outliers).
-A threshold of 120 mas and a deviation of more than 2.5 sigmas from the nightly average ephemeris offsets were adopted.
+A threshold of 120 mas and a deviation of more than 2.5 sigma from the nightly average ephemeris offsets were adopted.
 
-From Table \ref{Tab: Reductions-160} to \ref{Tab: Reductions-eso} we list the average dispersion (standard deviation) of the position offsets with regard to the ephemeris for $\alpha$ and $\delta$ obtained by telescope for each satellite.
-The final number of frames, number of nights (in parenthesis), the mean number of UCAC4 stars used in the reduction and the approximate V magnitude are also given.
-The dashed lines separate the satellites from different families with similar orbital parameters: Himalia Group (Himalia, Elara, Lysithea and Leda), Pasiphae Group (Pasiphae, Callirrhoe and Megaclite) and Ananke Group (Ananke and Praxidike).
+From Tables \ref{Tab: Reductions-160} to \ref{Tab: Reductions-eso}, we list the average dispersion (standard deviation) of the position offsets with regard to the ephemeris for $\alpha$ and $\delta$ obtained by telescope for each satellite.
+The final number of frames, number of nights (in parenthesis), the mean number of UCAC4 stars used in the reduction, and the approximate V magnitude are also given.
+The dashed lines separate the satellites from different families with similar orbital parameters: Himalia Group (Himalia, Elara, Lysithea and Leda), Pasiphae Group (Pasiphae, Callirrhoe and Megaclite), and Ananke Group (Ananke and Praxidike).
 Carme and Sinope are the only samples of their groups.
-From Saturn, Siarnaq and Paaliaq are from the Inuit Group while Phoebe and Albiorix are the only samples of their groups.
+From Saturn, Siarnaq, and Paaliaq are from the Inuit Group, while Phoebe and Albiorix are the only samples in their groups.
 
 The differences in the dispersion of the ephemeris offsets of the same satellite for distinct telescopes seen in Tables \ref{Tab: Reductions-160} to \ref{Tab: Reductions-eso} are caused by the different distribution of observations along the orbit for each telescope.
-This can be seen in Fig \ref{Fig: carme_anom} for Carme, \ref{Fig: pasiphae_anom} for Pasiphae and for all objects in the online material.
-Since the observations cover different segments of the orbit, the dispersion of the offsets may vary for different telescopes for a single satellite, with larger covered segments usually implying in larger dispersions and vice-versa.
-For Nereid, due to its high eccentric orbit, the observations are located between 90$\degree$ and 270$\degree$ of True Anomaly where Nereid remains most of the time.
+This can be seen in Fig.
+\ref{Fig: carme_anom} for Carme and Fig, \ref{Fig: pasiphae_anom} for Pasiphae and for all objects in the online material.
+Since the observations cover different segments of the orbit, the dispersion of the offsets may vary for different telescopes for a single satellite, with larger covered segments usually implying larger dispersions and vice versa.
+For Nereid, owing to its high eccentric orbit, the observations are located between 90$\degree$ and 270$\degree$ of true anomaly where Nereid remains most of the time.
 
@@
 
-The final set of positions of the satellites consists in 6523 catalogued positions observed between 1992 and 2014 for 12 satellites of Jupiter, 4 of Saturn, 1 of Uranus and 1 of Neptune.
+The final set of positions of the satellites consists in 6523 cataloged positions observed between 1992 and 2014 for 12 satellites of Jupiter, 4 of Saturn, 1 of Uranus, and 1 of Neptune.
 The topocentric positions are in the ICRS.
-The catalogues (one for each satellite) contain epoch of observations, the position error, filter used, estimated magnitude (from PSF fitting) and telescope origin.
+The catalogs (one for each satellite) contain epoch of observations, the position error, filter used, estimated magnitude (from PSF fitting) and telescope origin.
 The magnitude errors can be as high as 1 mag; they are not photometrically calibrated and should be used with care.
 The position errors were estimated from the dispersion of the ephemeris offsets of the night of observation of each position.
-Thus, these position errors are probably overestimated, as there must be ephemeris errors present in the dispersion of the offsets.
-These position catalogues are freely available in electronic form at the CDS (see a sample in Table \ref{Tab: sample-cds}) and at the IAU NSDC data base at www.imcce.fr/nsdc.
+Thus, these position errors are probably overestimated because there must be ephemeris errors present in the dispersion of the offsets.
+These position catalogs are freely available in electronic form at the CDS (see a sample in Table \ref{Tab: sample-cds}) and at the IAU NSDC data base at www.imcce.fr/nsdc.
 
@@
 This sample corresponds to 9 observations of Himalia from February 16, 2007 and April 15, 2007.
-Tables contain the topocentric ICRS coordinates of the irregular satellites, the position error estimated from the dispersion of the ephemeris offsets of the night of observation, \textbf{the UTC time of the frame's mid-exposure in julian date}, the estimated magnitude, the filter used, the telescope origin \textbf{and correspondent IAU code}.
-The filters may be U, B, V, R or I following the Johnson system; C stands for clear (no filter used), resulting in a broader R band magnitude, RE for the broad-band R filter ESO\#844 with $\lambda_{c} = 651.725$ nm and $\Delta\lambda = 162.184 $ nm (full width at half maximum) and "un" for unknown filter.
-E, OH, PE, BC and Z stand respectively for the ESO, OHP, Perkin-Elmer, Bollen \& Chivens and Zeiss telescopes.
+Tables contain the topocentric ICRS coordinates of the irregular satellites, the position error estimated from the dispersion of the ephemeris offsets of the night of observation, \textbf{the UTC time of the frame's mid-exposure in Julian date}, the estimated magnitude, the filter used, the telescope origin \textbf{and correspondent IAU code}.
+The filters may be U, B, V, R, or I following the Johnson system; C stands for clear (no filter used), resulting in a broader R band magnitude, RE for the broad-band R filter ESO\#844 with $\lambda_{c} = 651.725$ nm, and $\Delta\lambda = 162.184 $ nm (full width at half maximum) and "un" for unknown filter.
+E, OH, PE, BC, and Z stand for the ESO, OHP, Perkin-Elmer, Bollen \& Chivens and Zeiss telescopes, respectively.
 \end{table*}
@@
 \hline  \hline
- & \multicolumn{4}{c}{Number of Positions}   &    \tabularnewline
+ & \multicolumn{4}{c}{Number of positions}   &    \tabularnewline
 Satellite  & OPD  & OHP & ESO & Total  & Jacobson \tabularnewline
@@
 \par\end{centering}
-Comparison between the number of positions obtained in our work with the number used in the numerical integration of orbits by the JPL as published by \citealp{Jacobson2012}.
+Comparison between the number of positions obtained in our work with the number used in the numerical integration of orbits by the JPL as published by \citealp{Jacobson2012}.\LEt{ This just repeats the title in detail, which is not the point of a table note.
+Move it to the running text, if that helps after removing here.}
 \end{table}
@@
 
-Intending to see the potential of our results to improve the orbit of the irregular satellites observed, we analysed the offsets of our positions with regard to the ephemeris mentioned in Sect.
+Intending to see the potential of our results to improve the orbit of the irregular satellites observed, we analyzed the offsets of our positions with regard to the ephemeris mentioned in Sect.
 \ref{Sec: reduction}.
-Taking Carme as example, we plot in Fig.
-\ref{Fig: carme_anom} the mean ephemeris offsets for each night and their dispersions  (one sigma error bars) as a function of the true anomaly in right ascension (\ref{Fig: carme_alfa}) and declination (\ref{Fig: carme_delta}).
-Fig.
-\ref{Fig: carme_delta} clearly shows a systematic error in declination.
-When Carme is close to its apojove (true anomaly = 180\degree) its offsets are more likely to be more negative than those close to its perijove (true anomaly = 0\degree).
+Taking Carme as example, we plot the mean ephemeris offsets for each night in Fig.
+\ref{Fig: carme_anom} and their dispersions  (one sigma error bars) as a function of the true anomaly in right ascension (\ref{Fig: carme_alfa}) and declination (\ref{Fig: carme_delta}).
+Figure \ref{Fig: carme_delta} clearly shows a systematic error in declination.
+When Carme is close to its apojove (true anomaly = 180\degree), its offsets are more likely to be more negative than those close to its perijove (true anomaly = 0\degree).
 The offsets obtained from observations by four telescopes using different cameras and filters are in good agreement, meaning that there is an error in the ephemeris of Carme, most probably due to an error in its orbital inclination.
@@
 For some satellites, the orbital coverage is not enough to clearly indicate the presence of systematic errors in specific orbital elements.
-However, comparing the internal position mean errors of the reductions (Table \ref{Tab: stars-errors}) with the external position errors estimated from the dispersion of the ephemeris offsets (Tables \ref{Tab: Reductions-160} to \ref{Tab: Reductions-eso}), we see position error values much larger than expected from the mean errors.
+However, after comparing the internal position mean errors of the reductions (Table \ref{Tab: stars-errors}) with the external position errors estimated from the dispersion of the ephemeris offsets (Tables \ref{Tab: Reductions-160} to \ref{Tab: Reductions-eso}), we see position error values that are much higher than expected from the mean errors.
 This means that besides some expected astrometric errors, significant ephemeris errors must also be present.
@@
 
-\textbf{We managed a large database with FITS images acquired by 5 telescopes in 3 sites between 1992 and 2014.
-From that, we identified 8466 observations of irregular satellites, from which we managed to obtain 6523 suitable astrometric positions, giving a total of 3666 positions for 12 satellites of Jupiter, 1920 positions for 4 satellites of Saturn, 35 positions for Sycorax (Uranus) and 902 positions for Nereid (Neptune).}
+\textbf{We managed a large database with FITS images acquired by five telescopes in three sites between 1992 and 2014.
+From that, we identified 8466 observations of irregular satellites, from which we managed to obtain 6523 suitable astrometric positions, giving a total of 3666 positions for 12 satellites of Jupiter, 1920 positions for 4 satellites of Saturn, 35 positions for Sycorax (Uranus), and 902 positions for Nereid (Neptune).}
 
 \textbf{The positions of all the objects were determined using the PRAIA package.
-The package was suited to cope with the huge amount of observations and the task of identifying the satellites within the database.
-PRAIA tasks were also useful to deal with the missing or incorrect coordinate and time stamps present mostly in the old observations.}
+The package was suited to coping with the huge number of observations and with the task of identifying the satellites within the database.
+PRAIA tasks were also useful for dealing with the missing or incorrect coordinate and time stamps present mostly in the old observations.
+}
 
-\textbf{The UCAC4 was used as the reference frame. Based in the comparisons with ephemeris, we estimate that the position errors are about 60 mas to 80 mas depending on the satellite brightness. }
+\textbf{The UCAC4 was used as the reference frame.
+Based on the comparisons with ephemeris, we estimate that the position errors are about 60 mas to 80 mas depending on the satellite brightness.
+\LEt{ A single sentence may not be a paragraph in an expository paper like this one.}}
 
 For some satellites the number \textbf{of positions obtained in this work} is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012} (see Table \ref{Tab: comparison-horizons}).
-\textbf{For instance, the amount of new positions for Himalia is about 70\% of the number used in the numerical integation of orbits by JPL}.
+\textbf{For instance, the number of new positions for Himalia is about 70\% of the number used in the numerical integation of orbits by JPL}.
 Systematic errors in the ephemeris were found for at least some satellites (Ananke, Carme, Elara and Pasiphae).
-In the case of Carme, we evidenced an error in the orbital inclination \textbf{(see Fig.
+In the case of Carme, we showed an error in the orbital inclination \textbf{(see Fig.
 \ref{Fig: carme_anom})}.
@@
 \textbf{The positions derived in this work can be used in new orbital numerical integrations, generating more precise ephemerides.
-Stellar occultations by irregular satellites could then be better predicted.
-Based in this work, our group has already computed occultation predictions for the 8 major irregular satellites of Jupiter.
+Stellar occultations by irregular satellites could then be predicted better.
+Based on this work, our group has already computed occultation predictions for the eight major irregular satellites of Jupiter.
 These predictions will be published in a forthcoming paper.}
@@
 
-ARGJ thanks the financial support of CAPES.
+ARGJ acknowledges the financial support of CAPES.
 M.
 A.
-thanks the CNPq (Grants 473002/2013-2 and 308721/2011-0) and FAPERJ (Grant E-26/111.488/2013).
-RVM thanks grants: CNPq-306885/2013, Capes/Cofecub-2506/2015, Faperj/PAPDRJ-45/2013.
+is grateful to the CNPq (Grants 473002/2013-2 and 308721/2011-0) and FAPERJ (Grant E-26/111.488/2013).
+RVM acknowledges the following grants: CNPq-306885/2013, Capes/Cofecub-2506/2015, Faperj/PAPDRJ-45/2013.
 J-E.A.
-thanks the "Programme National de Planétologie" of INSU-CNRS-CNES for its financial support.
+is grateful to the "Programme National de Planétologie" of INSU-CNRS-CNES for its financial support.
 J.I.B.
@@
 B.E.M.
-thanks the financial support of CAPES.
+is grateful for the financial support of CAPES.
 
@@
 The distribution of ephemeris offsets along the orbit of the satellites are shown below.
-The red square is for the observations with the Perkin-Elmer telescope from OPD, the blue circle for Boller \& Chivens, the magenta triangle down for Zeiss, the black triangle up for OHP and the green star for ESO.
+The red square is for the observations with the Perkin-Elmer telescope from OPD, the blue circle for Boller \& Chivens, the magenta triangle down for Zeiss, the black triangle up for OHP, and the green star for ESO.
 For Carme and Pasiphae see Figs.

### 2015-05-27 qualify/Qualify-altair.tex
Continuado a escrever Ceres no qualify

@@
 %\usepackage{latexsym}
-\usepackage[]{graphicx}
+\usepackage{graphicx}
 \graphicspath{{figuras}}
@@
 
+%%% newcommand %%%%%%%%%%%%%%%%%
 \newcommand{\PE}{Perkin-Elmer}
 \newcommand{\BC}{Boller \& Chivens}
+
+\newcommand{\degr}{\ensuremath{^{\circ}}}%                    % degree symbol:  °
+\newcommand{\arcmin}{\ensuremath{^{\prime}}}%                    % degree symbol:  °
+\newcommand{\fs}{\mbox{\ensuremath{.\!\!^s}}}
+\newcommand{\farcm}{\mbox{\ensuremath{.\mkern-4mu^\prime}}}%    % fractional arcminute symbol: 0.'0
+\newcommand{\farcs}{\mbox{\ensuremath{.\!\!^{\prime\prime}}}}%  % fractional arcsecond symbol: 0.''0
+\newcommand{\fdg}{\mbox{\ensuremath{.\!\!^\circ}}}%             % fractional degree symbol:     0.°0
 
@@
 \resumo{Resumo}
-\indent \indent O estudo da estrutura e evolução do Sistema Solar tem muita importância atualmente, por exemplo, na compreensão dos mecanismos de formação dos planetas de outras estrelas (exoplanetas), trazendo em seus desdobramentos valiosas informações quanto a viabilidade de formação de ambientes que comportem vida.\\
-\indent Praticamente cada satélite é um problema particular de Mecânica Celeste.
+%\indent \indent O estudo da estrutura e evolução do Sistema Solar tem muita importância atualmente, por exemplo, na compreensão dos mecanismos de formação dos planetas de outras estrelas (exoplanetas), trazendo em seus desdobramentos valiosas informações quanto a viabilidade de formação de ambientes que comportem vida.\\
+%\indent Praticamente cada satélite é um problema particular de Mecânica Celeste.
 Dessa forma, a observação de seus movimentos é muito útil por razões teóricas.
 A preparação e realização de missões espaciais para alguns planetas, e mesmo alguns satélites, como Titã de Saturno ou os Galileanos de Júpiter, requer observações frequentes e posições muito acuradas de todos os satélites do sistema planetário visitado.\\
-\indent Os satélites irregulares de planetas gigantes, principalmente de Júpiter e de Saturno, são substancialmente menores do que os satélites regulares e possuem órbitas mais distantes, excêntricas, inclinadas e, na maioria dos casos, retrógradas.
+%\indent Os satélites irregulares de planetas gigantes, principalmente de Júpiter e de Saturno, são substancialmente menores do que os satélites regulares e possuem órbitas mais distantes, excêntricas, inclinadas e, na maioria dos casos, retrógradas.
 Explicar a existência dos satélites irregulares dos planetas gigantes é um estudo interessante em dinâmica orbital.\\
-\indent É amplamente aceito que eles, devido à configuração orbital, foram capturados por seus planetas.
+%\indent É amplamente aceito que eles, devido à configuração orbital, foram capturados por seus planetas.
 Porém, ainda não são certos a origem e o método de captura desses objetos embora as teorias mais prováveis sejam a de arrasto gasoso durante a formação do Sistema Solar e a captura de três corpos.\\
-\indent A maioria dos satélites formam famílias com características orbitais semelhantes sugerindo uma evolução dada por colisões.
+%\indent A maioria dos satélites formam famílias com características orbitais semelhantes sugerindo uma evolução dada por colisões.
 A compreensão desses mecanismos nos fará entender melhor o nosso próprio sistema solar e, possivelmente, de sistemas extra-solares.\\
-\indent Com o pacote astrométrico PRAIA o tempo empregado na redução de grandes quantidades de imagens foi diminuído significativamente.
+%\indent Com o pacote astrométrico PRAIA o tempo empregado na redução de grandes quantidades de imagens foi diminuído significativamente.
 Os objetivos científicos dos nossos programas observacionais agora tem sido atingidos em curtíssimo prazo, em consonância com a atual demanda astronômica e astrofísica de nossa área.\\
-\indent Em uma base de dados com mais de 100 mil imagens obtidas entre 1992 e 2012 no telescópios de 1.2m do Observatoire de Haute Province, França, e nos telescópios \PE, \BC{} e Zeiss no Observatório do Pico dos Dias, mais de 4 mil contem satélites irregulares de Júpiter e Saturno.
+%\indent Em uma base de dados com mais de 100 mil imagens obtidas entre 1992 e 2012 no telescópios de 1.2m do Observatoire de Haute Province, França, e nos telescópios \PE, \BC{} e Zeiss no Observatório do Pico dos Dias, mais de 4 mil contem satélites irregulares de Júpiter e Saturno.
 Reduzir essa grande quantidade de observações com precisão só foi possível com a utilização do PRAIA.
@@
 \vspace{1em}
-\noindent\textbf{Palavras-chave:} Astrometria, Satélites Irregulares de Júpiter, Satélites Irregulares de Saturno
+\noindent\textbf{Palavras-chave:} Astrometria, Ocultações
 
@@
 
-\begin{figure}
+\subsection{Ocultação de 2010}
+\label{Subsec: 2010-occ}
+
+\indent \indent Em 17 de Agosto de 2010 Ceres ocultou a estrela TYC 6833-163-1 (UCAC4 313-111823), cuja magnitude é $V = 11.55$ e tem posição no ICRS para a data do evento baseada no catálogo UCAC4 \citep{Zacharias2013}:
+
+\begin{equation}
+\left\{ 
+  \begin{array}{l l}
+    \alpha = 17^{h}18^{m}29\fs 0085\\
+    \delta = -27\degr 26\arcmin 38\farcs 867
+  \end{array}
+\right.
+\label{Eq: 2010-coord}
+\end{equation}
+
+\indent O evento foi observado no Brasil a partir de cinco diferentes sítios (ver Fig.
+\ref{Fig: Ceres-map-2010}).
+Destes, 4 obtiveram cordas positivas enquanto UFSC teve uma corda negativa.
+Das positivas, a observação proveniente do INPE iniciciou-se após o início do evento devido a dificuldades técnicas e, portanto, apenas a emersão da curva de luz foi detectada.
+
+Uma das características mais importantes desse evento foi a velocidade com que ocorreu (apenas 3.9 km s$^{-1}$) acarretando que mesmo exposições de poucos segundos representariam resoluções espaciais significantes.
+
+Todas as observações foram feitas com a utilização de CCDs.
+As curvas de luz de cada observação foram obtidas das imagens FITS com a utilização do pacote PRAIA \citep[Plataforma de Redução Astrométrica de Imagens Astronômicas,][]{Assafin2011}.
+As curvas foram normalizadas para o fluxo da estrela mais Ceres, uma vez que eles estavam indistinguíveis logo antes e depois da ocultação.
+Por fim, elas foram normalizadas pelo ajuste de uma curva polinomial (de primeira ou segunda ordem) fora da queda de fluxo assim fixando em 1 a razão de fluxo fora da ocultação.
+
+Os instantes de ingresso e egresso foram obtidas de cada curva de luz ajustando-se um modelo de poço quadrado levando em consideração a difração de Fresnel, a banda do CCD, o diâmetro aparente da estrela e o tempo de exposição utilizado \citep[ver][]{Widemann2009, Braga-Ribas2013}.
+
+O menor tempo de integração usado nas observações positivas foi de $1.0s$, que corresponde a aproximadamente $3.9km$ no plano do céu.
+Portanto, o erro na determinação dos instantes de ingresso e egresso é dominado principalmente pelo tempo de integração, não pela difração de Fresnel ou diâmetro da estrela, ambos da ordem de algumas centenas de metros para esse evento
+
+O ajuste dos dados da ocultação consiste em minimizar uma função de $\chi^{2}$ clássica para cada curva de luz, como descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}.
+
+The occultation data fit consists in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011} and BR13.\ The free parameter to adjust is the ingress or egress instant, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$.
+The best fittings to the 2010 occultation light curves are shown in Fig.
+\ref{Fig: Ceres-2010-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2010}.
+\begin{figure}[h]
 \begin{centering}

### 2015-05-28 qualify/Qualify-altair.tex
Continuado a escrever a parte de Ceres

@@
 \usepackage{pdflscape}
+\usepackage[breaklinks]{hyperref}
 \usepackage[comma,authoryear]{natbib}
-\usepackage{hyperref}
+\usepackage{url}
 \usepackage[nonumberlist]{glossaries}
 %\usepackage[alf]{abntcite}
-\bibliographystyle{apalike}
 
@@
 
-Os dois eventos foram preditos por Steve Preston\footnote{Predições publicadas em http://asteroidoccultation.com.} para a IOTA (International Occultation Timing Association), durante predições de rotina de ocultações de asteroides de estrelas brilhantes.
+Os dois eventos foram preditos por Steve Preston\footnote{Predições publicadas em \url{http://asteroidoccultation.com}.} para a IOTA (International Occultation Timing Association), durante predições de rotina de ocultações de asteroides de estrelas brilhantes.
 Os caminhos das sombras podem ser visualizados na Fig.
@@
 O menor tempo de integração usado nas observações positivas foi de $1.0s$, que corresponde a aproximadamente $3.9km$ no plano do céu.
-Portanto, o erro na determinação dos instantes de ingresso e egresso é dominado principalmente pelo tempo de integração, não pela difração de Fresnel ou diâmetro da estrela, ambos da ordem de algumas centenas de metros para esse evento
+Portanto, o erro na determinação dos instantes de ingresso e egresso é dominado principalmente pelo tempo de integração, não pela difração de Fresnel ou diâmetro da estrela, ambos da ordem de algumas centenas de metros para esse evento.
 
 O ajuste dos dados da ocultação consiste em minimizar uma função de $\chi^{2}$ clássica para cada curva de luz, como descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}.
+Os parâmetros livres para ajustar são os instantes de ingresso e egresso que fornece o valor mínimo de $\chi^{2}$ ($\chi^{2}_{min}$).
+O melhor ajuste das curvas de luz para a ocultação de 2010 está mostrado na Fig.
+\ref{Fig: Ceres-2010-curves}.
 
-The occultation data fit consists in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011} and BR13.\ The free parameter to adjust is the ingress or egress instant, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$.
-The best fittings to the 2010 occultation light curves are shown in Fig.
-\ref{Fig: Ceres-2010-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2010}.
+\begin{figure}[h]
+\begin{centering}
+\subfigure[Ocultação de 2010]{\includegraphics[scale=0.51]{figuras/Ceres_2010_fluxratio.eps}\label{Fig: Ceres-2010-curves} }
+\subfigure[Ocultação de 2013]{\includegraphics[scale=0.51]{figuras/Ceres_2013_fluxratio.eps}\label{Fig: Ceres-2013-curves} }
+\caption{Curvas de luz normalizadas das cordas positivas dos eventos.
+As curvas estão desviadas por um fator de 0.2 (a) e 1.0 (b) para melhor visualização.
+As linhas pretas são os melhores ajustes com o modelo de poço quadrado.
+As linhas vermelhas são os melhores ajustes com o modelo de poço quadrado, porém levando em conta a difração de Fresnel, o diâmetro da estrela e o tempo de exposição.
+Os instantes médios de cada curva não coincidem devido às diferentes longitudes dos sítios.
+A curva de luz de Brookline (b) está desviada por um fator de -64 s como explicado no texto.
+\label{Fig: Ceres-curves}}
+\end{centering}
+\end{figure}
+
+A metodologia usada para analizar o perfil de Ceres a partir das observações é o mesmo descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}.
+Cada combinação de posição do sítio, instantes de ingresso e egresso, junto com as coordenadas da estrela e as efemérides de Ceres, correspondem a um ponto no plano do céu.
+A coleção de todos esses pontos determina o limbo aparente de Ceres.
+
+Adotamos um modelo elíptico para o perfil do limbo, resultante da projeção de um esferóide com achatamento nos pólos no plano do céu.
+Essa escolha é suportada pelo trabalho de \cite{Drummond2014}, por meio de imagem direta de Ceres.
+Dessa forma, nós temos $N=7$ extremidades das cordas para ajustar $M=5$ parâmetros que definem uma elipse: semi-eixo maior e semi-eixo menor aparentes ($a^\prime$ and $b^\prime$, respectivamente), ângulo de posição $P$ do seu semi-eixo maior e as posição $(f_c,g_c)$ do seu centro com respeito à estrela ocultada.
+O semi-eixo maior {$a^\prime$} é equivalente ao raio equatorial $R_{equa}$ do elipsoide.
+
+As coordenadas $f_{c}$ e $g_{c}$, em quilômetros, foram calculadas usando a efeméride de Ceres JPL\#33 \citep{Giorgini1996} e a posição da estrela ocultada.
+Elas são positivas na direção Leste e Norte celestes, respectivamente.
+O ângulo de posição $P$ é contado positivamente a partir do norte celeste local em direção ao leste celeste.
+O achatamento aparente pode ser definido por $\epsilon^\prime = 1 - (b^\prime/a^\prime)$.
+O melhor ajuste é obtido minimizando uma função de $\chi^{2}_{r}$ reduzido, onde definimos o número de graus de liberdade do problema como $\mathcal{N} \equiv N - M$.
+Todos os procedimentos que permitem a determinação das barras de erro dos paramêtros físicos podem ser encontradas em \cite{Braga-Ribas2013}.
+
 \begin{figure}[h]
@@
 \end{figure}
+
+\subsection{Ocultação de 2013}
+\label{Subsec: 2013-occ}
 
@@
 \bibliography{Referencias}
+\bibliographystyle{apalike}
 

### 2015-05-28 Irregulares/Irregulares.tex
modificado author running
Feitos as correcoes do editor no meu texto

@@
 \titlerunning{Astrometric Positions of the Irregular Satellites of Giant Planets}
-\authorrunning{Gomes-Júniot et al.}
+\authorrunning{Gomes-Júnior et al.}
 
@@
 \fnmsep\thanks{Partially based on observations through the ESO runs 079.A-9202(A), 075.C-0154, 077.C-0283 and 079.C-0345.}
-\fnmsep\thanks{\textbf{Partially based on observations made at
-Observatoire de Haute Provence (OHP), F-04870 Saint-Michel l'observatoire, France}}
+\fnmsep\thanks{Partially based on observations made at
+Observatoire de Haute Provence (OHP), F-04870 Saint-Michel l'observatoire, France}
 }
@@
 {The irregular satellites of the giant planets are believed to have been captured during the evolution of the solar system.
-Knowing their physical parameters, such as size, density and albedo is important to constrain where they came from and how they were captured.
+Knowing their physical parameters, such as size, density, and albedo is important for constraining where they came from and how they were captured.
 The best way to obtain these parameters are observations in situ by spacecrafts or from stellar occultations by the objects.
 Both techniques demand that the orbits are well known.}
-{We aimed to obtain good astrometric positions of irregular satellites in order to improve their orbits and ephemeris.}
-{We identified and reduced observations of several irregular satellites from three databases containing more than 8000 images obtained between 1992 and 2014 at three sites (Observatório do Pico dos Dias, Observatoire de Haute-Provence and European Southern Observatory - La Silla).
+{We aimed to obtain good astrometric positions of irregular satellites to improve their orbits and ephemeris.}
+{We identified and reduced observations of several irregular satellites from three databases containing more than 8000 images obtained between 1992 and 2014 at three sites (Observatório do Pico dos Dias, Observatoire de Haute-Provence, and European Southern Observatory - La Silla).
 We used the software PRAIA (Platform for Reduction of Astronomical Images Automatically) to make the astrometric reduction of the CCD frames.
-The UCAC4 catalogue represented the International Celestial Reference System in the reductions.
-The identification of the satellites in the frames was done through their ephemerides as determined from the SPICE/NAIF kernels.
-Some procedures were taken to overcome missing or incomplete information (coordinates, date), mostly for the older images.}
-{We managed to obtain more than 6000 positions for 18 irregular satellites, being 12 of Jupiter, 4 of Saturn, 1 of Uranus (Sycorax) and 1 of Neptune (Nereid).
-For some satellites the number of obtained positions is more than 50\% of that used in earlier orbital numerical integrations.}
-{Comparison of our positions with recent JPL ephemeris suggests the presence of systematic errors in the orbits for some of the irregular satellites.
+The UCAC4 catalog represented the International Celestial Reference System in the reductions.
+Identification of the satellites in the frames was done through their ephemerides as determined from the SPICE/NAIF kernels.
+Some procedures were followed to overcome missing or incomplete information (coordinates, date), mostly for the older images.}
+{We managed to obtain more than 6000 positions for 18 irregular satellites: 12 of Jupiter, 4 of Saturn, 1 of Uranus (Sycorax), and 1 of Neptune (Nereid).
+For some satellites the number of obtained positions is more than 50\% of what was used in earlier orbital numerical integrations.}
+{Comparison of our positions with recent JPL ephemeris suggests there are systematic errors in the orbits for some of the irregular satellites.
 The most evident case was an error in the inclination of Carme.}
@@
 
-The irregular satellites of the giant planets are smaller than the regular moons, having more eccentric, inclined, distant and, in most cases, retrograde orbits.
-Due to their orbital configurations, it is largely accepted that these objects were captured in the early solar system \citep{Sheppard2003}.
+The irregular satellites of the giant planets are smaller than the regular moons, having more eccentric, inclined, distant, and in most cases, retrograde orbits.
+Owing to their orbital configurations, it is largely accepted that these objects were captured in the early solar system \citep{Sheppard2003}.
 
-Because they are faint, the majority of these objects was discovered only in the last century\footnote{Website: http://ssd.jpl.nasa.gov/?sat\_discovery} .
+Because they are faint, the majority of these objects were only discovered in the last century\footnote{Website: http://ssd.jpl.nasa.gov/?sat\_discovery} .
 They were never visited by a spacecraft, with the exception of Himalia, Phoebe and Nereid, in a flyby by the Cassini space probe in 2000 for Himalia \citep{Porco2003} and in 2004 for Phoebe \citep{Desmars2013} and in a flyby by the Voyager 2 space probe in 1989 for Nereid \citep{Smith1989}.
@@
 If these objects were captured, there remains the question of where they came from.
-\citealp{Clark2005} showed from imaging spectroscopy from Cassini that Phoebe has a surface probably covered by material from the outer solar system and \citealp{Grav2003} showed that the satellites of the Jovian Prograde Group Himalia have grey colors implying that their surfaces are similar to that of C-type asteroids.
-In that same work, the Jovian Retrograde Group Carme was found to have surface colors similar to the D-type asteroids like Hilda or Trojan families while JXIII Kalyke has a redder color like Centaurs or trans-neptunian objects (TNOs).
+\citealp{Clark2005} show from imaging spectroscopy from Cassini that Phoebe has a surface probably covered by material from the outer solar system and \citealp{Grav2003} show that the satellites of the Jovian Prograde Group Himalia have gray colors implying that their surfaces are similar to that of C-type asteroids.
+In that same work, the Jovian Retrograde Group Carme was found to have surface colors similar to the D-type asteroids as for the Hilda or Trojan families, while JXIII Kalyke has a redder color like Centaurs or trans-Neptunian objects (TNOs).
 
-For Saturnian satellites, \citealp{Grav2007} showed by their colors and spectral slopes that these satellites contain a more or less equal fraction of C-, P- and D-like objects but SXXII Ijiraq is marginally redder than D-type objects.
+For Saturnian satellites, \citealp{Grav2007} show by their colors and spectral slopes that these satellites contain a more or less equal fraction of C-, P-, and D-like objects, but SXXII Ijiraq is marginally redder than D-type objects.
 These works may suggest different origins for the irregular satellites.
 
-In this context, we used 3 databases for deriving precise positions for the irregular satellites observed at Observatório do Pico dos Dias (1.6 m and 0.6 m telescopes, IAU code 874), Observatoire Haute-Provence (1.2m telescope, IAU code 511) and ESO (2.2 m telescope, IAU code 809).
-Many irregular satellites were observed between 1992 and 2014 covering a few orbital periods of these objects (12 satellites of Jupiter, 4 of Saturn, Sycorax of Uranus and Nereid of Neptune).
+In this context, we used three databases for deriving precise positions for the irregular satellites observed at the Observatório do Pico dos Dias (1.6 m and 0.6 m telescopes, IAU code 874), the Observatoire Haute-Provence (1.2m telescope, IAU code 511), and ESO (2.2 m telescope, IAU code 809).
+Many irregular satellites were observed between 1992 and 2014, covering a few orbital periods of these objects (12 satellites of Jupiter, 4 of Saturn, Sycorax of Uranus, and Nereid of Neptune).
 
-\textbf{Since their ephemerides are not very precise, predict and observe stellar occultations are very difficult and no observation of such an event for an irregular satellite is found in the literature.
+Since their ephemerides are not very precise, predicting and observing stellar occultations are very difficult, and no observation of such an event for an irregular satellite is found in the literature.
 The precise star positions to be derived by the ESA astrometry satellite Gaia \citep{deBruijne2012} will render better predictions with the only source of error being the ephemeris.
-The positions derived from \textbf{our} observations can be used in new orbital numerical integrations, generating more precise ephemerides.}
+The positions derived from our observations can be used in new orbital numerical integrations, generating more precise ephemerides.
 
-\textbf{The power of stellar occultations for observing relatively small diameter solar system objects is supported by recent works such as the discovery of a ring system around the Centaur (10199) Chariklo \citep{Braga-Ribas2014}.
+The power of stellar occultations for observing relatively small diameter solar system objects is supported by recent works, such as the discovery of a ring system around the Centaur (10199) Chariklo \citep{Braga-Ribas2014}.
 Once irregular satellites start to be observed by this technique, it will be possible to obtain their physical parameters (shape, size, albedo, density) with unprecedented precision.
 For instance, in this case, sizes could be obtained with kilometer accuracy.
-The knowledge of these parameters would in turn bring valuable information for the study of the capture mechanisms and origin of the irregular satellites.}
+The knowledge of these parameters would in turn bring valuable information for studying the capture mechanisms and origin of the irregular satellites.
 
@@
 
-The OPD database was produced at Observatório do Pico dos Dias (OPD, IAU code 874, $45^{\circ} ~34\arcmin ~57\arcsec$ W, $22^{\circ} ~32\arcmin ~04\arcsec$ S, 1864 m)\footnote{Website: http://www.lna.br/opd/opd.html - in Portuguese}, located at geographical longitude , in Brazil.
+The OPD database was produced at the Observatório do Pico dos Dias (OPD, IAU code 874, $45^{\circ} ~34\arcmin ~57\arcsec$ W, $22^{\circ} ~32\arcmin ~04\arcsec$ S, 1864 m)\footnote{Website: http://www.lna.br/opd/opd.html - in Portuguese}, located at geographical longitude , in Brazil.
 The observations were made between 1992 and 2014 by our group in a variety of observational programs.
@@
 Almost all the frames were photometrically calibrated with auxiliary bias and flat-field frames by means of standard procedures using IRAF\footnote{Website: http://iraf.noao.edu/} and, for the mosaics, using the esowfi \citep{Jones2000} and mscred \citep{Valdes1998} packages.
-Some of the nights at OPD didn't have bias and flat-field images so the correction was not possible.
+Some of the nights at OPD did not have bias and flat-field images so the correction was not possible.
 
 The astrometric treatment was made with the Platform for Reduction of Astronomical Images Automatically (PRAIA) \citep{2011gfun.conf...85A}.
-The (x, y) measurements were performed with 2-dimensional circular symmetric Gaussian fits within one Full Width Half Maximum (FWHM = seeing).
-Within one FWHM, the image profile is well described by a Gaussian profile, free from the wing distortions, which may jeopardize the center determination.
+The (x, y) measurements were performed with two-dimensional circular symmetric Gaussian fits within one full width half maximum (FWHM = seeing).
+Within one FWHM, the image profile is described well by a Gaussian profile, which is free of the wing distortions, which may jeopardize the determination of the center.
 PRAIA automatically recognizes catalog stars and determines ($\alpha$, $\delta$) with a user-defined model relating the (x, y) measured and (X, Y) standard coordinates projected in the sky tangent plane.
@@
 For each frame, we used the six constants polynomial model to relate the (x, y) measurements with the (X, Y) tangent plane coordinates.
-For ESO, we followed the same astrometric procedures described in detail in \cite{2012A&A...541A.142A}; the (x, y) measurements of the individual CCDs were pre-corrected by a field distortion pattern, and all positions coming from different CCDs and mosaics were then combined using a 3rd degree polynomial model to produce a global solution for each night and field observed, and final ($\alpha$, $\delta$) object positions were obtained in the UCAC4 system.
+For ESO, we followed the same astrometric procedures as described in detail in \cite{2012A&A...541A.142A}; the (x, y) measurements of the individual CCDs were precorrected by a field distortion pattern, and all positions coming from different CCDs and mosaics were then combined using a third degree polynomial model to produce a global solution for each night and field observed, and final ($\alpha$, $\delta$) object positions were obtained in the UCAC4 system.
 
-In Table \ref{Tab: stars-errors} we list the average mean error in $\alpha$ and $\delta$ for the reference stars obtained by telescope, the average (x, y) measurement errors of the Gaussian fits described above and the mean number of UCAC4 stars used by frame.
+In Table \ref{Tab: stars-errors} we list the average mean error in $\alpha$ and $\delta$ for the reference stars obtained by telescope, the average (x, y) measurement errors of the Gaussian fits described above, and the mean number of UCAC4 stars used by frame.
 For all databases, about 20\% of outlier reference stars were eliminated for presenting (O-C) position residuals higher than 120 mas in the ($\alpha$, $\delta$) reductions.
@@
 
-To help identifying the satellites in the frames, and derive the ephemeris for the instants of the observations for comparisons (see Sect \ref{Sec: comparison}), we used the kernels from SPICE/JPL\footnote{Website: http://naif.jpl.nasa.gov/naif/toolkit.html}.
+To help identify the satellites in the frames and derive the ephemeris for the instants of the observations for comparisons (see Sect \ref{Sec: comparison}), we used the kernels from SPICE/JPL\footnote{Website: http://naif.jpl.nasa.gov/naif/toolkit.html}.
 \cite{Emelyanov2008} and references therein also provided ephemeris of similar quality for the irregular satellites.
-For instance, for Himalia, which has relatively good orbit solutions, the ephemerides differ by less than $20 mas$, and in the case of less-known orbits, like Ananke, the differences are less than $90 mas$.
-Here, we choose to use the JPL ephemeris because they used more recent observations \cite[see][]{Jacobson2012}.
+For instance, for Himalia, which has relatively good orbit solutions, the ephemerides differ by less than $20~mas$, and in the case of less-known orbits, like Ananke, the differences are less than $90~mas$.
+We chose to use the JPL ephemeris because they used more recent observations \cite[see][]{Jacobson2012}.
 The JPL ephemeris that represented the Jovian satellites in this work was the DE421 $+$ JUP300.
-For the Saturnian satellites the ephemeris was DE421 $+$ SAT359 to Hyperion, Iapetus and Phoebe and DE421 $+$ SAT361 to Albiorix, Siarnaq and Paaliaq.
+For the Saturnian satellites, the ephemeris was DE421 $+$ SAT359 to Hyperion, Iapetus, and Phoebe and DE421 $+$ SAT361 to Albiorix, Siarnaq, and Paaliaq.
 The DE421 $+$ URA095 was used for Sycorax and DE421 $+$ NEP081 for Nereid.
@@
 
-In the OPD database, there were some images (mostly the older ones) with missing coordinates or wrong date in their headers.
-In the case of missing or wrong coordinates, we adopted the ephemeris as the central coordinates of the frames.
+In the OPD database, there were some images (mostly the older ones) with missing coordinates or the wrong date in their headers.
+In the case of missing or incorrect coordinates, we adopted the ephemeris as the central coordinates of the frames.
 When the time was not correct, the FOV identification failed.
-In this case, a search for wrong date (year) displaying was performed.
+In this case, a search for displays of wrong date (year) was performed.
 Problems like registering local time instead of UTC were also identified and corrected.
@@
 In all databases, for each night a sigma-clipping procedure was performed to eliminate discrepant positions (outliers).
-A threshold of 120 mas and a deviation of more than 2.5 sigmas from the nightly average ephemeris offsets were adopted.
+A threshold of 120 mas and a deviation of more than 2.5 sigma from the nightly average ephemeris offsets were adopted.
 
-From Table \ref{Tab: Reductions-160} to \ref{Tab: Reductions-eso} we list the average dispersion (standard deviation) of the position offsets with regard to the ephemeris for $\alpha$ and $\delta$ obtained by telescope for each satellite.
+From Tables \ref{Tab: Reductions-160} to \ref{Tab: Reductions-eso} we list the average dispersion (standard deviation) of the position offsets with regard to the ephemeris for $\alpha$ and $\delta$ obtained by telescope for each satellite.
 The final number of frames, number of nights (in parenthesis), the mean number of UCAC4 stars used in the reduction and the approximate V magnitude are also given.
-The dashed lines separate the satellites from different families with similar orbital parameters: Himalia Group (Himalia, Elara, Lysithea and Leda), Pasiphae Group (Pasiphae, Callirrhoe and Megaclite) and Ananke Group (Ananke and Praxidike).
+The dashed lines separate the satellites from different families with similar orbital parameters: Himalia Group (Himalia, Elara, Lysithea and Leda), Pasiphae Group (Pasiphae, Callirrhoe and Megaclite), and Ananke Group (Ananke and Praxidike).
 Carme and Sinope are the only samples of their groups.
-From Saturn, Siarnaq and Paaliaq are from the Inuit Group while Phoebe and Albiorix are the only samples of their groups.
+From Saturn, Siarnaq and Paaliaq are from the Inuit Group while Phoebe and Albiorix are the only samples in their groups.
 
 The differences in the dispersion of the ephemeris offsets of the same satellite for distinct telescopes seen in Tables \ref{Tab: Reductions-160} to \ref{Tab: Reductions-eso} are caused by the different distribution of observations along the orbit for each telescope.
-This can be seen in Fig \ref{Fig: carme_anom} for Carme, \ref{Fig: pasiphae_anom} for Pasiphae and for all objects in the online material.
-Since the observations cover different segments of the orbit, the dispersion of the offsets may vary for different telescopes for a single satellite, with larger covered segments usually implying in larger dispersions and vice-versa.
-For Nereid, due to its high eccentric orbit, the observations are located between 90$\degree$ and 270$\degree$ of True Anomaly where Nereid remains most of the time.
+This can be seen in Fig.
+\ref{Fig: carme_anom} for Carme and Fig.
+\ref{Fig: pasiphae_anom} for Pasiphae and for all objects in the online material.
+Since the observations cover different segments of the orbit, the dispersion of the offsets may vary for different telescopes for a single satellite, with larger covered segments usually implying larger dispersions and vice versa.
+For Nereid, due to its high eccentric orbit, the observations are located between 90$\degree$ and 270$\degree$ of true anomaly where Nereid remains most of the time.
 
@@
 This sample corresponds to 9 observations of Himalia from February 16, 2007 and April 15, 2007.
-Tables contain the topocentric ICRS coordinates of the irregular satellites, the position error estimated from the dispersion of the ephemeris offsets of the night of observation, \textbf{the UTC time of the frame's mid-exposure in julian date}, the estimated magnitude, the filter used, the telescope origin \textbf{and correspondent IAU code}.
+Tables contain the topocentric ICRS coordinates of the irregular satellites, the position error estimated from the dispersion of the ephemeris offsets of the night of observation, the UTC time of the frame's mid-exposure in Julian date, the estimated magnitude, the filter used, the telescope origin and correspondent IAU code.
 The filters may be U, B, V, R or I following the Johnson system; C stands for clear (no filter used), resulting in a broader R band magnitude, RE for the broad-band R filter ESO\#844 with $\lambda_{c} = 651.725$ nm and $\Delta\lambda = 162.184 $ nm (full width at half maximum) and "un" for unknown filter.
-E, OH, PE, BC and Z stand respectively for the ESO, OHP, Perkin-Elmer, Bollen \& Chivens and Zeiss telescopes.
+E, OH, PE, BC and Z stand for the ESO, OHP, Perkin-Elmer, Bollen \& Chivens and Zeiss telescopes, respectively.
 \end{table*}
@@
 \caption{Mean ephemeris offsets and dispersions (1 sigma error bars) in the coordinates of Carme taken night by night by true anomaly for each telescope.
-\textbf{The red square is for the observations with the Perkin-Elmer telescope from OPD, the blue circle for Boller \& Chivens, the magenta triangle down for Zeiss, the black triangle up for OHP and the green star for ESO.}}
+The red square is for the observations with the Perkin-Elmer telescope from OPD, the blue circle for Boller \& Chivens, the magenta triangle down for Zeiss, the black triangle up for OHP and the green star for ESO.}
 \label{Fig: carme_anom}
@@
 
-\textbf{We managed a large database with FITS images acquired by 5 telescopes in 3 sites between 1992 and 2014.
-From that, we identified 8466 observations of irregular satellites, from which we managed to obtain 6523 suitable astrometric positions, giving a total of 3666 positions for 12 satellites of Jupiter, 1920 positions for 4 satellites of Saturn, 35 positions for Sycorax (Uranus) and 902 positions for Nereid (Neptune).}
+We managed a large database with FITS images acquired by five telescopes in three sites between 1992 and 2014.
+From that, we identified 8466 observations of irregular satellites, from which we managed to obtain 6523 suitable astrometric positions, giving a total of 3666 positions for 12 satellites of Jupiter, 1920 positions for 4 satellites of Saturn, 35 positions for Sycorax (Uranus) and 902 positions for Nereid (Neptune).
 
-\textbf{The positions of all the objects were determined using the PRAIA package.
-The package was suited to cope with the huge amount of observations and the task of identifying the satellites within the database.
-PRAIA tasks were also useful to deal with the missing or incorrect coordinate and time stamps present mostly in the old observations.}
+The positions of all the objects were determined using the PRAIA package.
+The package was suited to coping with the huge number of observations and with the task of identifying the satellites within the database.
+PRAIA tasks were also useful for dealing with the missing or incorrect coordinate and time stamps present mostly in the old observations.
 
-\textbf{The UCAC4 was used as the reference frame. Based in the comparisons with ephemeris, we estimate that the position errors are about 60 mas to 80 mas depending on the satellite brightness. }
+The UCAC4 was used as the reference frame.
+Based on the comparisons with ephemeris, we estimate that the position errors are about 60 mas to 80 mas depending on the satellite brightness.
+For some satellites the number of positions obtained in this work is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012} (see Table \ref{Tab: comparison-horizons}).
+For instance, the number of new positions for Himalia is about 70\% of the number used in the numerical integation of orbits by JPL.
+Systematic errors in the ephemeris were found for at least some satellites (Ananke, Carme, Elara and Pasiphae).
+In the case of Carme, we showed an error in the orbital inclination (see Fig.
+\ref{Fig: carme_anom}).
 
-For some satellites the number \textbf{of positions obtained in this work} is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012} (see Table \ref{Tab: comparison-horizons}).
-\textbf{For instance, the amount of new positions for Himalia is about 70\% of the number used in the numerical integation of orbits by JPL}.
-Systematic errors in the ephemeris were found for at least some satellites (Ananke, Carme, Elara and Pasiphae).
-In the case of Carme, we evidenced an error in the orbital inclination \textbf{(see Fig.
-\ref{Fig: carme_anom})}.
-
-\textbf{The positions derived in this work can be used in new orbital numerical integrations, generating more precise ephemerides.
-Stellar occultations by irregular satellites could then be better predicted.
-Based in this work, our group has already computed occultation predictions for the 8 major irregular satellites of Jupiter.
-These predictions will be published in a forthcoming paper.}
+The positions derived in this work can be used in new orbital numerical integrations, generating more precise ephemerides.
+Stellar occultations by irregular satellites could then be predicted better.
+Based on this work, our group has already computed occultation predictions for the eight major irregular satellites of Jupiter.
+These predictions will be published in a forthcoming paper.
 

### 2015-05-28 qualify/Qualify-altair.tex
Terminado de escrever ocultacao de 2010 do qualify

@@
 
+Duas possíveis soluções foram consideradas para o ajuste do limbo.
+O primeiro, que chamamos de solução nominal, consiste em determinar os cinco parâmetros que caracterizam uma elipse a partir dos sete contatos observados.
+A segunda solução consiste em calcular o ângulo de posição $P$ a partir das coordenadas do pólo de Ceres obtidas por \cite{Drummond2014} ($\alpha_{p} = (287 \pm 3) \degr$, $\delta_{p} = (+64 \pm 3) \degr$ no ICRS) e da efeméride de Ceres no instante da ocultação.
+Chamamos de solução de pólo fixo.
+
+Para o evento de 2010, a solução nominal teve como melhor ajuste $\chi^2_{r,min} = 0.24$, que podem ser interpretadas como as barras de erro estarem superestimadas com respeito à boa qualidade do ajuste.
+Porém, como o problema tem somente dois graus de liberdade, $\chi^2_{r,min}$ relativamente pequenos são aceitáveis.
+Os resultados obtidos para o diâmetro equatorial, achatamento, ângulo de posição e coordenadas do centro esao apresentadas na segunda coluna da tabela \ref{Tab: resultados}.
+
+\begin{table}[h]
+ \begin{centering}
+  \caption{Resultados do ajuste de limbo de Ceres com os dados dos eventos de 2010 e 2013.\label{Tab: resultados}}
+  \begin{tabular}{@{}lcccc}
+  \hline
+     Solution & 2010/Nominal & \textbf{2010/Pólo fixo} & 2013/Nominal & 2013/Pólo fixo \\
+\hline
+Diam. equat.  (km) & 982 $\pm$ 14 & \textbf{972 $\pm$ 6}  & 971 $\pm$ 7  & 971 $\pm$ 7\\
+%Equatorial radius (km) & 491 $\pm$ 7  & \textbf{486 $\pm$ 3}  & 485.5 $\pm$ 3.5 & 485.5 $\pm$ 3.5\\
+Achatamento        & 0.08 $\pm$ 0.03 & \textbf{0.08 $\pm$ 0.03} & 0.08 $\pm$ 0.04 & 0.08 $\pm$ 0.04\\
+Âng. de pos. (deg)   & 5 $\pm$ 10    & \textbf{12 $\pm$ 3} (*)& 22 $\pm$ 5    & 25 $\pm$ 3 (*)\\
+$f_c$ (km)             & 97 $\pm$ 9   & \textbf{102 $\pm$ 5}   & 77 $\pm$ 6    & 78 $\pm$ 6\\
+$g_c$ (km)             & 16 $\pm$ 15  & \textbf{21 $\pm$ 11}  & 13 $\pm$ 16   & 13 $\pm$ 16\\
+$\chi^2_{r,min}$       & 0.24          &  \textbf{0.42}         & 1.27          & 1.27\\
+\hline
+\end{tabular}
+\textbf{Notas:} Em negrito destacamos nossa melhor solução.
+As barras de erro estão no nível de 1$\sigma$.
+O diâmetro polar ($D_{pol}$) pode ser facilmente calculado a partir de $D_{pol}=D_{equa}(1 - \epsilon)$.
+(*) Ângulos de posição derivados a partir das coordenadas do pólo de Ceres determinapas por \cite{Drummond2014}.
+\end{centering}
+\end{table}
+
+Como pode ser visto, o parâmetro com a maior incerteza é o ângulo de posição cobrindo um intervalo de 20$\degr$.
+Claramente, fixar as coordenadas do pólo pode melhorar a solução.
+Por fim, a correção do achatamento devido ao ângulo do aspecto polar está dentro da barra de erro 1$\sigma$ e não tem relevância estatística, dessa forma $\epsilon = 0.08 \pm 0.03$.
+
+No momento da ocultação, as coordenadas do pólo de Ceres correspondiam a um ângulo de posição $P = (12 \pm 3)\degr$.
+Explorar o espaço de parâmetros restringindo a elipses com ângulo de posição dentro deste intervalo resulta na solução de pólo fixo.
+O parâmetros físicos do melhor ajuste estão mostrados na Tab.
+\ref{Tab: resultados} enquanto a solução está esquematizada na Fig.
+\ref{Fig: Ceres-limb-2010}.
+
+Essa solução corresponde ao limite superior da barra de erro 1$\sigma$ da solução nominal para $P$.
+Por outro lado, ela obtem os menores valores para o diâmetro equatorial, melhorando sua determinação por um fator de 2.
+
 \begin{figure}[h]
@@
 A linha verde em (a) é uma corda negativa.
-Os instantes marcados em verde em (b) não foram utilizados para o ajuste como descrito no texto.
+Os instantes marcados em verde em (b) não foram utilizados para o ajuste, como descrito no texto.
 \label{Fig: Ceres-limb}}

### 2015-05-29 qualify/Qualify-altair.tex
Terminado de digitar Ceres no qualify

@@
 
-\chapter{Ocultações}
+\chapter{Ocultações Estelares}
 \label{Cap: observacoes}
@@
 \end{tabular}
-\textbf{Notas:} Em negrito destacamos nossa melhor solução.
+\textbf{Notas:} Em negrito destaca-se a melhor solução obtida.
 As barras de erro estão no nível de 1$\sigma$.
@@
 
+\indent \indent Em 25 de Outubro de 2010, Ceres ocultou a estrela TYC 865-911-1 (UCAC4 496-058191), de magnitude $V = 10.05$.
+Baseada no UCAC4 \citep{Zacharias2013}, sua posição ICRS para a data da ocultação é:
+
+%
+\begin{equation}
+\left\{ 
+  \begin{array}{l l}
+    \alpha = 11^{h}57^{m}52\fs7641\\
+    \delta = +09\degr 07\arcmin 49\farcs835
+  \end{array}
+\right.
+\end{equation}
+%
+Esse evento foi observado na costa Leste dos Estados Unidos logo antes do amanhecer, como mostrado na Fig. \ref{Fig: Ceres-map-2013}.
+
+Nove cordas positivas foram obtidas em vários sítios (ver Fig.
+\ref{Fig: Ceres-map-2013}).
+Cada estação foi equipada com uma câmera de vídeo com tempo de leitura desprezível.
+Isso é particularmente importante, já que a velocidade da sombra de Ceres para esse evento foi de 42.6 km s$^{-1}$, muito mais rápido que o evento de 2010.
+
+Durante o evento Ceres estava muito baixo no céu com alturas entre $15\degr$ (Winchester) e $20\degr$ (Hampton).
+Forte cintilação era esperada e, combinada com o curto tempo de integração e baixa diminuição de brilho, resultou em curvas de luz ruidosas e assim grandes incertezas nos instantes de imersão e emersão.
+
+Todos os vídeos foram convertidos para imagens FITS e a fotometria foi obtidia via \textsc{praia} \cite{Assafin2011}.
+As curvas de luz foram normalizadas por uma estrela de referência quando havia uma estrela no campo.
+
+Para reduzir o ruído, os dados foram binados por grupos de cinco imagens -- com exceção de Greenbelt, onde gupos de dez imagens foram utilizadas.
+Esse procedimento o tempo de integração efetivo por um fator de 5 (ou 10).
+Da mesma forma que para o evento de 2010, uma normalização adicional por um polinômio foi aplicada.
+
+Os instantes de ingresso e egresso da ocultação foram obtidos pelo mesmo procedimento descrito na seção \ref{Subsec: 2010-occ}.
+Uma vez que o tempo de integração efetivo usado (0.17 s) representa aproximadamente 7 km no plano do céu, e a escala de Fresnel e  o diâmetro da estrela estão novamente na ordem de centenas de metros, o erro da determinação dos instantes de imersão e emersão são dominados principalmente pelo tempo de integração, da mesma forma que para o evento de 2010.
+Os melhores ajustes para as curvas de luz da ocultação estão mostrados na Fig.
+\ref{Fig: Ceres-2013-curves}.
+
+Uma comparação entre os tempos obtidos pela estação de Brookline e o restante mostrou que esses têm um atraso de aproximadamente 64 s. Dessa forma, não utilizamos os tempos de Brookline na análise.
+
+Perfis elípticos foram ajustados para todas as cordas restantes.
+pelo mesmo procedimento descrito na seção \ref{Subsec: 2010-occ}.
+O resultado foi $\chi^2_{r,min} = 13$ , sugerindo que um modelo elíptico não é satisfatório para os dados.
+De fato, ao olharmos para a Fig.
+\ref{Fig: Ceres-limb-2013} vemos que a corda de Varina adiantada com respeito às outras.
+Como nessa estação o tempo não foi inserido diretamente nos frames do vídeo é possível que essa diferença seja oriunda de um eventual problema da correspondência entre os tempos do camcorder e do GPS.
+
+A imersão gravada em Owings também parece atrasada com respeito às cordas próximas (ver Fig.
+\ref{Fig: Ceres-limb-2013}).
+Essa corda tem aproximadamente o mesmo tamanho da corda de Mechanicsville, apesar de estarem separadas por cerca de 100 km.
+Diferentemente de Varina, essa estação teve os tempos inseridos em cada frame do vídeo o que torna mais difícil justificar um problema de tempo.
+Outras possibilidades seriam uma má determinação dos instantes de ingresso e egresso dessa curva ou uma característica do relevo de Ceres.
+
+Em um segundo ajuste, não consideramos as cordas de Brookline, Varina e Owings.
+O ajuste dos cinco parâmetros que definem uma elipse para os doze contatos resultou em $\chi^2_{r,min} = 1.27$, indicando  que está em bom acordo com os dados observados dentro das barras de erro.
+Essa é a solução mostrada na Fig.
+\ref{Fig: Ceres-limb-2013}, onde podemos ver que o tamanho da corda de Brookline é compatível com o modelo.
+
+Os resultados das soluções nominais e de pólo fixo não são significantemente diferentes.
+Isso se deve ao fato da barra de erro do ângulo de posição da solução nominal (que é muito menor que o da solução nominal de 2010) ser muito similar à barra de erro do ângulo de posição da solução de pólo fixo ($P = (25 \pm 3)\degr$).
+Os dois resultados estão apresentados nas colunas 4 e 5 da tabela \ref{Tab: resultados}.
+
+Sobre a hipótese do atraso observado na imersão da curva de luz de Owings ser associado a uma característica topográfica, o contato gravado corresponderia a uma elevação negativa de 31 $\pm$ 4 km com respeito à elipse de melhor ajuste.
+Porém modelos teóricos prevêem que relevos em Ceres não devem ser maiores que 10--20 km \citep{Johnson1973}, enquanto dados observacionais limitam o contorno em 18 km \citep{Carry2008}.
+Imagens mais recentes da sonda \textit{Dawn} também revelam uma superfície mais suave.
+Portanto, a associação da imersão em Owings com um relevo é improvável.
+
+\subsection{Discussão}
+
+\indent \indent Os resultados apresentados na tabela \ref{Tab: resultados} mostram um acordo entre os parâmetros físicos obtidos nas duas ocultações, especialmente no diâmetro equatorial.
+As diferenças ocorrem basicamente nos tamanhos das barras de erro e podem ser justificadas pelas particularidades de cada conjunto de dados.
+
+O evento de 2010, por exemplo, teve somente sete contatos, porém bem distribuídos sobre o disco de Ceres.
+Por outro lado, o evento de 2013 teve cinco contatos a mais, todavia concentrados em certas regiões do corpo.
+Em particular, a ausência de cordas prómimas ao pólo sul fez seu achatamente ser pior determinado para o evento de 2013 que para o evento de 2010.
+Mesmo nossa melhor medida de achatamento, $\epsilon=0.08 \pm 0.03$, tem alta incerteza se comparado com outros valores publicados na literatura, como mostra a tabela \ref{Tab: Ceres-final}.
+
+Como foi mostrado, usar as coordenadas do pólo de Ceres determinadas por \cite{Drummond2014} para limitar o ângulo de posição não foi um procedimento eficiente para o evento de 2013.
+Por outro lado, fixar o ângulo de posição para a ocultação de 2010 reduziu as barras de erro dos outros parâmetros (com exceção do achatamento).
+Por fim, Esse procedimento resultou em um excelente acordo entre os raios equatoriais obtidos para ambos os eventos.
+
+Uma comparação do diâmetro equatorial de Ceres medido por diferentes técnicas está mostrado na tabela \ref{Tab: Ceres-final}.
+Vemos um acordo entre os nossos resultados e aqueles obtidos por imageamente direto do Hubble Space Telescope (HST) \citep{Thomas2005}, do Keck Observatory e do ESO VLT \citep{Drummond2014}.
+O menor valor, reportado por \cite{Carry2008}, pode ser justificado pelo fato desse estudo não levar em conta o efeito de escurecimento de bordo.
+
+O evento de 1984 \citep{Millis1987} é a única outra ocultação que podemos comparar nosssos resultados.
+As medidas do diâmetro não se acordam dentro de 2$\sigma$.
+É difícil dizer com certeza as razões dessa divergência.
+Uma forma de clarificar o problema seria redeterminar os instantes de imersão e emersão das curvas de luz originais usando a mesma metodologia aplicada nesse trabalho.
+Infelizmente, não temos acesso aos dados da curva de luz original do evento de 1984.
+
+A sonda da NASA \textit{Dawn} poderá responder esses questões que são importantes não só no conhecimento do próprio Ceres, mas também para todas as técnicas usadas até agora para o estudo das propriedades físicas dos pequenos objetos do Sistema Solar, como as ocultações estelares.
+
+\begin{table}[h]
+  \caption{Diâmetro equatorial e achatamento de Ceres \label{Tab: Ceres-final}}
+  \begin{centering}
+  \begin{tabular}{@{}lccc}
+  \hline
+     Diâmetro Equatorial (km) & Achatamento & Método & Ref. \\
+\hline
+$972 \pm 6$  & $0.08  \pm 0.03$  & Occultação & 1\\
+$967 \pm 10$ & $0.078 \pm 0.015$ & Keck+VTL    & 2\\
+$959 \pm 5$  & $0.074 \pm 0.007$ & Keck        & 3\\
+$975 \pm 4$  & $0.067 \pm 0.005$ & HST         & 4\\
+$959 \pm 5$  & $0.05  \pm 0.01$  & Occultação & 5\\
+\hline
+\end{tabular}
+\par\end{centering}
+\textbf{Referências.} 1: \cite{GomesJunior2015-Ceres}. 2: \cite{Drummond2014}. 3: \cite{Carry2008}. 4: \cite{Thomas2005}. 5: \cite{Millis1987}.
+\end{table}
+
 \section{Satélites Irregulares}

### 2015-05-29 Irregulares/AA-2015-26273/Irregulares.tex
...

--- 2015-05-28 Irregulares/Irregulares.tex
@@
 %\usepackage[english,brazilian]{babel}
+\usepackage[utf8]{inputenc}
 \usepackage[T1]{fontenc}
@@
 
-\date{Received ; accepted }
+\date{Received: Abr 08, 2015; accepted: May 06, 2015}
 
@@
 % & \multicolumn{2}{c}{Mean errors} & Nr & Nr &  UCAC4       \tabularnewline
-\multicolumn{2}{c}{RA  (ICRS)  Dec} & RA error & Dec error & Epoch & Mag & Filter & Telescope & \textbf{IAU code} \tabularnewline
+\multicolumn{2}{c}{RA  (ICRS)  Dec} & RA error & Dec error & Epoch & Mag & Filter & Telescope & IAU code \tabularnewline
 \multicolumn{1}{l}{\ h \ m \ \ \ s} & \multicolumn{1}{l}{\ \ \ $\degree$ \ \ \ $\arcmin$ \ \ \ $\arcsec$} &   (mas)   &   (mas)  & (jd) & & & & \tabularnewline
 \hline
- 16 59 11.6508 & -22 00 44.855 &  17 &  12 & 2454147.78241319 &   16.0 &    C &      BC & \textbf{874} \tabularnewline
- 16 59 11.6845 & -22 00 44.932 &  17 &  12 & 2454147.78332384 &   15.8 &    C &      BC & \textbf{874} \tabularnewline
- 16 59 11.7181 & -22 00 44.978 &  17 &  12 & 2454147.78422477 &   16.0 &    C &      BC & \textbf{874} \tabularnewline
- 16 59 11.7818 & -22 00 45.143 &  17 &  12 & 2454147.78602662 &   15.9 &    C &      BC & \textbf{874} \tabularnewline
- 16 59 11.8188 & -22 00 45.232 &  17 &  12 & 2454147.78693750 &   16.0 &    C &      BC & \textbf{874} \tabularnewline
- 17 17 11.0344 & -22 47 19.415 &  30 &  24 & 2454205.63885463 &   16.1 &        U &      BC & \textbf{874} \tabularnewline
- 17 17 11.0270 & -22 47 19.381 &  30 &  24 & 2454205.63959167 &   16.1 &        U &      BC & \textbf{874} \tabularnewline
- 17 17 11.0258 & -22 47 19.366 &  30 &  24 & 2454205.64031875 &   16.1 &        U &      BC & \textbf{874} \tabularnewline
- 17 17 11.0192 & -22 47 19.417 &  30 &  24 & 2454205.64104583 &   16.1 &        U &      BC & \textbf{874} \tabularnewline
+ 16 59 11.6508 & -22 00 44.855 &  17 &  12 & 2454147.78241319 &   16.0 &    C &      BC & 874 \tabularnewline
+ 16 59 11.6845 & -22 00 44.932 &  17 &  12 & 2454147.78332384 &   15.8 &    C &      BC & 874 \tabularnewline
+ 16 59 11.7181 & -22 00 44.978 &  17 &  12 & 2454147.78422477 &   16.0 &    C &      BC & 874 \tabularnewline
+ 16 59 11.7818 & -22 00 45.143 &  17 &  12 & 2454147.78602662 &   15.9 &    C &      BC & 874 \tabularnewline
+ 16 59 11.8188 & -22 00 45.232 &  17 &  12 & 2454147.78693750 &   16.0 &    C &      BC & 874 \tabularnewline
+ 17 17 11.0344 & -22 47 19.415 &  30 &  24 & 2454205.63885463 &   16.1 &        U &      BC & 874 \tabularnewline
+ 17 17 11.0270 & -22 47 19.381 &  30 &  24 & 2454205.63959167 &   16.1 &        U &      BC & 874 \tabularnewline
+ 17 17 11.0258 & -22 47 19.366 &  30 &  24 & 2454205.64031875 &   16.1 &        U &      BC & 874 \tabularnewline
+ 17 17 11.0192 & -22 47 19.417 &  30 &  24 & 2454205.64104583 &   16.1 &        U &      BC & 874 \tabularnewline
 \hline

### 2015-06-01 Irregulares/Irregulares.tex
Escrito astrometria dos irregulares para o qualify

@@
 
-\date{Received ; accepted }
+\date{Received: ; accepted: }
 
@@
 \begin{table}
-\caption{\label{Tab: comparison-horizons} Comparison of positions obtained with \citealp{Jacobson2012}.}
+\caption{\label{Tab: comparison-horizons} Comparison between the number of positions obtained in our work with the number used in the numerical integration of orbits by the JPL as published by \citealp{Jacobson2012}.}
 \begin{centering}
@@
 \par\end{centering}
-Comparison between the number of positions obtained in our work with the number used in the numerical integration of orbits by the JPL as published by \citealp{Jacobson2012}.
 \end{table}

### 2015-06-01 qualify/Qualify-altair.tex
...
Escrito astrometria dos irregulares para o qualify

@@
 \usepackage{pdflscape}
-\usepackage[breaklinks]{hyperref}
+\usepackage[breaklinks=true]{hyperref}
 \usepackage[comma,authoryear]{natbib}
-\usepackage{url}
 \usepackage[nonumberlist]{glossaries}
+\usepackage{arydshln}
 %\usepackage[alf]{abntcite}
@@
 
+\subsection{Astrometria dos sat\'elites irregulares}
+
+\indent \indent Durante o mestrado, foi realizado junto ao grupo um trabalho de caráter astrométrico para a melhoria das efemérides dos satélites irregulares de Júpiter Saturno.
+Em colaboração com o Dr.
+Jean-Eudes Arlot do IMCCE, reduzi um banco de dados observado no Observatoire Haute-Provence (OHP) entre 1998 e 2008, contendo mais de 28 mil posições para 10 satélites.
+Reduzi também um banco de dados com mais de 100 mil imagens obtidas no Observatório do Pido dos Dias (OPD) entre 1992 e 2014.
+Já no doutorado, neste mesmo âmbito, reduzi 810 observações feitas no European Southern Observatory (ESO) em 24 noites utilizando o detector mosaico CCD Wide Field Imager (WFI).
+
+Mais de 8000 posições foram identificadas como pertencentes a satélites irregulares.
+Porém, devido à grande gama de configurações (3 sítios, 5 telescópios, mais de 10 câmeras e mais de 10 filtros) e às condições observacionais de algumas noites, 6523 posições foram selecionadas como boas posições, ou seja, possuem 3 ou mais estrelas do catálogo de referência (UCAC4), estão dentro de 2$\sigma$ da dispersão das posições da noite e a dispersão da noite à qual pertence não é maior que 2$\sigma$ da média das dispersões das noites.
+Essa estatística é feita satélite por satélite.
+O trabalho está aceito para publicação \citep{GomesJunior2015-Irregular} e o catálogo de posições está disponível no CDS.
+
+Um dos principais resultados desse trabalho foi a grande quantidade de posições obtidas para os satélites irregulares em comparação com a quantidade utilizada nas integrações numéricas atuais (ver tabela \ref{Tab: comparison-horizons}).
+
+\begin{table}
+\caption{\label{Tab: comparison-horizons} Comparação entre o número de posições obtidas em \cite{GomesJunior2015-Irregular} e o número utilizado nas integrações numéricas das órbitas pelo JPL como publicado por  \cite{Jacobson2012}.}
+\begin{centering}
+\begin{tabular}{lccccc}
+\hline  \hline
+ & \multicolumn{4}{c}{Número de Posições}   &    \tabularnewline
+Satélite  & OPD  & OHP & ESO & Total  & Jacobson \tabularnewline
+\hline
+Himalia & 854 & 357 & 23 & 1234 & 1757 \tabularnewline
+Elara & 403 & 187 & 46 & 636 & 1115 \tabularnewline
+Lysithea & 60 & 84 & 90 & 234 & 431 \tabularnewline
+Leda & 6 & 48 & 44 & 98 & 178 \tabularnewline
+\hdashline
+Pasiphae & 295 & 248 & 66 & 609 & 1629 \tabularnewline
+Callirrhoe & 9 & -  &  16 & 25 & 95 \tabularnewline
+Megaclite & - & -  &  10 & 10 & 50  \tabularnewline
+\hdashline
+Ananke & 52 & 141 & 57 & 250 & 600 \tabularnewline
+Praxidike & - & -  &   2 & 2 & 59 \tabularnewline
+\hdashline
+Carme & 90 & 204 & 37 & 331 & 973 \tabularnewline
+Sinope & 41 & 169 & 11 & 221 & 854 \tabularnewline
+Themisto & - & - & 16 & 16 & 55 \tabularnewline
+\hline
+Phoebe & 1239 & 516 & 32 & 1787 & 3479 \tabularnewline
+\hdashline
+Siarnaq & - & 20 & 56 & 76 & 239 \tabularnewline
+Paaliaq & - & - & 11 & 11 & 82 \tabularnewline
+\hdashline
+Albiorix & - & - & 46 & 46 & 137 \tabularnewline
+\hline
+Sycorax & - & - & 35 & 35 & 237 \tabularnewline
+\hline
+Nereid & 803 & - & 99 & 902 & 716 \tabularnewline
+\hline
+\end{tabular}
+\par\end{centering}
+\end{table}
+
+Com o objetivo ver o potencial das nossas posições em melhorar as órbitas dos satélites irregulares, analizamos os offsets das nossas posições com respeito às efemérides do JPL.
+Tomando Carme como exemplo, plotamos os offsets médios das efemérides para cada noite na Fig.
+\ref{Fig: carme_anom} e suas respectivas dispersões (erro de barra 1$\sigma$) em função da anomalia verdadeira em ascenção reta e declinação.
+A figura mostra claramente um erro sistemático em declinação.
+Quando Carme está próximo a seu apojove (anomalia verdadeira = 180$\degr$), seus offsets tem maior probabilidade de serem mais negativos do que aqueles próximos ao seu perijove (anomalia verdadeira = 0$\degr$).
+Todos os offsets obtidos de observações feitas em quatro telescópios diferentes usando câmeras e filtros diferentes estão em bom acordo, o que significa que há um erro nas efemérides de Carme, muito provavelmente devido a um erro em sua inclinação orbital.
+
+\begin{figure}[h]
+\begin{centering}
+\subfigure[Right Ascension]{\includegraphics[scale=0.45]{figuras/Carme_RA}\label{Fig: carme_alfa}}
+\subfigure[Declination]{\includegraphics[scale=0.45]{figuras/Carme_DEC}\label{Fig: carme_delta}}
+\caption{Offsets médios das efemérides e dispersões (barra de erro 1$\sigma$) nas coordenadas de Carme tomadas noite a noite por anomalia verdadeira para cada telescópio.
+O quadrado vermelho são as observações com o telescópio do OPD Perkin-Elmer, o círculo azul para o Boller \& Chivens também do OPD, o triângulo pra cima preto para o OHP e a estrela verde para o ESO.}
+\label{Fig: carme_anom}
+\end{centering}
+\end{figure}
+
+Esse padrão em declinação também foi identificado para outros satélites como Pasiphae e Ananke.
+Para alguns satélites a cobertura orbital não foi suficiente para indicar claramente a presença de erros sistemáticos em elementos orbitais específicos.
+Os gráficos para todos os satélites estão disponíveis em \cite{GomesJunior2015-Irregular}.
+
+\subsection{Previs\~ao das ocultaç\~oes}
+
+\indent \indent
+
 \section{TNOs}

### 2015-06-02 EKOS/Ceres/Ceres_EKO.tex
consertado referencia como pedido pela revista

%========================================================================
%=================LaTeX TEMPLATE FOR ARTICLE SUBMISSION==================
%========================================================================
%
% When submitting an abstract to the ``Distant EKOs'' Newsletter, please
% use the following LaTeX template and submit it to:
%     ekonews@boulder.swri.edu
% Search for the "%%%" strings for places to put your text.
%
%====================================== version 1.12 (2015 April 17) ====

\documentstyle[11pt]{article}
\textheight 9.25in
\parindent 0pt
\topmargin -0.75in
\textwidth 6.58in
\oddsidemargin 0in

\newcommand{\AbsYes}[4]
  {\begin{center}{\Large\bf #1}\end{center} \begin{center}{\bf #2}\end{center}
   \vspace*{-2ex} {\footnotesize #3}  \par \vspace*{2ex} {\parindent=3ex #4}
   \vspace*{2ex} \par }
\newcommand{\AbsNo}[3]
  {{\Large \bf #1} \par {\bf #2} \par {\footnotesize #3} \par }

\newcommand{\Published}[5] {\AbsYes{#1}{#2}{#3}{#4} \mbox{\bf Published in: #5}}
\newcommand{\Accepted}[5] {\AbsYes{#1}{#2}{#3}{#4} \mbox{\bf To appear in: #5}}
\newcommand{\Submitted}[5] {\AbsYes{#1}{#2}{#3}{#4} \mbox{\bf Submitted to: #5}}
\newcommand{\Conference}[5] {\AbsNo{#1}{#2}{#3} \smallskip \mbox{To
appear in: #5}}
\newcommand{\Other}[5] {\AbsYes{#1}{#2}{#3}{#4} {\em #5} }
\newcommand{\OtherN}[5] {\AbsNo{#1}{#2}{#3} \smallskip \mbox{#5} }

\newcommand {\gtrsim}
  {{\hbox{\rlap{\hbox{\lower4pt\hbox{$\sim$}}}\hbox{$>$}}}}
\newcommand {\lesssim}
  {{\hbox{\rlap{\hbox{\lower4pt\hbox{$\sim$}}}\hbox{$<$}}}}

%%% IF YOU USE ANY SPECIAL MACROS, PLEASE INCLUDE THEM BELOW THIS LINE:

\begin{document}

%%% SELECT THE APPROPRIATE DESCRIPTION FOR YOUR SUBMISSION BY UNCOMMENTING
%%% THE CORRESPONDING LINE (I.E., REMOVE THE "%" AT THE BEGINNING OF THE LINE):
%%% "Published = A PAPER THAT HAS BEEN PUBLISHED IN A REFEREED JOURNAL
%%% "Accepted" = A PAPER ACCEPTED FOR PUBLICATION IN A REFEREED JOURNAL
%%% "Submitted" = A PAPER SUBMITTED (BUT NOT YET ACCEPTED) TO A JOURNAL
%%% "Conference" =  A PAPER THAT HAS BEEN PRESENTED AT A CONFERENCE
%%% "Other" =  THESIS OR NEWS/ANNOUNCEMENT/EDITORIAL

%\Published{
\Accepted{
%\Submitted{
%\Conference{
%\Other{

%%% ENTER THE TITLE OF YOUR ARTICLE

Results of two multi-chord stellar occultations by dwarf planet (1) Ceres

}{
%%% ENTER LIST OF AUTHORS BELOW.  PLEASE INDICATE WITHIN $^...$
%%% THE NUMBER WHICH CORRESPONDS TO THE INSTITUTE OF EACH AUTHOR.

A. R. Gomes-J\'unior$^{1},$%\thanks{E-mail: altair08@astro.ufrj.br},
B. L. Giacchini$^{2,3,4}$,%\thanks{E-mail: breno@cbpf.br},
F. Braga-Ribas$^{5, 6, \dag}$,
M. Assafin$^{1, \dag, \ddag}$,
R. Vieira-Martins$^{1,5, \dag, \ddag}$,
J.I.B. Camargo$^{5, \dag}$,
B. Sicardy$^{7}$,
B. Timerson$^{4}$, 
T. George$^{4}$,
J. Broughton$^{8}$, 
T. Blank$^{4}$,
G. Benedetti-Rossi$^{5}$, 
J. Brooks$^{4}$,  
R. F. Dantowitz$^{9}$,
D. W. Dunham$^{4}$, 
J. B. Dunham$^{4}$, 
C. K. Ellington$^{4}$,
M. Emilio$^{10}$,
F.R. Herpich$^{11}$,
C. Jacques$^{3, 12}$,
P. D. Maley$^{4,13}$,
L. Mehret$^{10}$,
A.J.T. Mello$^{14}$,
A.C. Milone$^{15}$,
E. Pimentel$^{3, 12}$,  
W. Schoenell$^{11}$,
N. S. Weber$^{9}$

}{
%%% ENTER LIST OF AFFILIATIONS/ADDRESS BELOW.  ADD/DELETE LINES AS NEEDED.
%%% THE NUMBER IN $^..$ INDICATES AUTHOR NUMBER AS GIVEN ABOVE

$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
$^{2}$Centro Brasileiro de Pesquisas F\'isicas, Rua Dr. Xavier Sigaud 150, Rio de Janeiro  22290-180, Brazil\\
$^{3}$Se\c{c}\~ao de Oculta\c{c}\~oes/REA-Brasil, Belo Horizonte, Brazil\\
$^{4}$International Occultation Timing Association, P.O. Box 131034, Houston, TX 77219-1034, USA\\
$^{5}$Observat\'orio Nacional/MCTI, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil\\
$^{7}$LESIA, Observatoire de Paris, CNRS UMR 8109, Universit\'{e} Pierre et Marie Curie, Universit\'{e} Paris-Diderot, 5 place Jules Janssen,\\ F-92195 Meudon Cedex, France\\
$^{8}$RASNZ Occultation Section, P.O. Box 3181, Wellington, New Zealand\\
$^{9}$Clay Center Observatory at Dexter Southfield, 20 Newton Street, Brookline, MA 02445, USA\\
$^{10}$Universidade Estadual de Ponta Grossa, Ponta Grossa, Brazil\\
$^{11}$Universidade Federal de Santa Catarina, Florian\'opolis, Brazil\\
$^{12}$Centro de Estudos Astron\^omicos de Minas Gerais, Belo Horizonte, Brazil\\
$^{13}$NASA Johnson Space Center Astronomical Society, Houston, Texas, USA\\
$^{14}$Federal University of Technology - Paran\'a (UTFPR / DAELT), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil\\
$^{15}$Instituto Nacional de Pesquisas Espaciais, S\~ao Jos\'e dos Campos, Brazil\\
$^\dag$ Associated to Laborat\'{o}rio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, CEP 20921-400,\\ Rio de Janeiro, Brazil\\
$^\ddag$Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France

}{
%%% ENTER TEXT OF ABSTRACT ONLY FOR *ACCEPTED* PAPERS.  CONFERENCE PROCEEDINGS
%%% OR PAPERS THAT HAVE BEEN SUBMITTED BUT NOT YET ACCEPTED WILL HAVE ONLY THE
%%% TITLE, AUTHOR LIST, AND CONTACT INFORMATION PUBLISHED IN THE NEWSLETTER.

We report the results of two multi-chord stellar occultations by the dwarf planet (1) Ceres that were observed  from Brazil on 2010 August 17, and from the USA on 2013 October 25. Four positive detections were obtained for the 2010 occultation, and nine for the 2013 occultation. Elliptical models were adjusted to the observed chords to obtain Ceres' size and shape. Two limb fitting solutions were studied for each event. The first one is a nominal solution with an indeterminate polar aspect angle. The second one was constrained by the pole coordinates as given by Drummond et al. Assuming a Maclaurin spheroid, we determine an equatorial diameter of 972 $\pm$ 6 km and an apparent oblateness of $0.08 \pm 0.03$ as our best solution. These results are compared to all available size and shape determinations for Ceres made so far, and shall be confirmed by the NASA's \textit{Dawn} space mission.

}{
%%% ENTER NAME OF JOURNAL OR CONFERENCE PROCEEDINGS BELOW.  YOU MAY INCLUDE
%%% VOLUME AND PAGE NUMBER IF KNOWN.
%%% IF SUBMITTING A THESIS ABSTRACT, PLEASE STATE WHEN, WHERE, AND UNDER WHOSE
%%% DIRECTION YOUR DEGREE WAS COMPLETED, AND ANY RELEVANT CONTACT INFORMATION.
%%% e.g., Dissertation directed by G. Kuiper \\
%%%       Ph.D. awarded Month, Year from University X \\
%%%       Address as of Month, Year: Institute X, postal address
%%%

Montly Notices of the Royal Astronomical Society

}
%%% ENTER PREPRINT AVAILABILITY INFORMATION

{\it For preprints, contact \ }  {\tt altair08@astro.ufrj.br, breno@cbpf.br, ribas@on.br} \\
%{\it or by anonymous ftp to \ } {\tt FTP ADDRESS} \\
{\it or on the web at \ }       {\verb| http://arxiv.org/abs/1504.04902 |}

\end{document}

### 2015-06-07 qualify/Qualify-altair.tex
Escrito parte de previsão das ocultações de irregulares no qualify

@@
 \resumo{Resumo}
-%\indent \indent O estudo da estrutura e evolução do Sistema Solar tem muita importância atualmente, por exemplo, na compreensão dos mecanismos de formação dos planetas de outras estrelas (exoplanetas), trazendo em seus desdobramentos valiosas informações quanto a viabilidade de formação de ambientes que comportem vida.\\
-%\indent Praticamente cada satélite é um problema particular de Mecânica Celeste.
+%\indent \indent O estudo da estrutura e evolução do Sistema Solar tem muita importância atualmente, por exemplo, na compreensão dos mecanismos de formação dos planetas de outras estrelas (exoplanetas), trazendo em seus desdobramentos valiosas informações quanto a viabilidade de formação de ambientes que comportem vida.
+
+%Praticamente cada satélite é um problema particular de Mecânica Celeste.
 Dessa forma, a observação de seus movimentos é muito útil por razões teóricas.
@@
 
-O ajuste dos dados da ocultação consiste em minimizar uma função de $\chi^{2}$ clássica para cada curva de luz, como descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}.
+O ajuste dos dados da ocultação consiste em minimizar uma função dep $\chi^{2}$ clássica para cada curva de luz, como descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}.
 Os parâmetros livres para ajustar são os instantes de ingresso e egresso que fornece o valor mínimo de $\chi^{2}$ ($\chi^{2}_{min}$).
@@
 
-\indent \indent
+\indent \indent Dentre os satélite irregulares dos planetas gigantes, poucos são aqueles que possuem medidas de seus parâmetros físicos.
+Apenas Himalia, Phoebe e Nereida foram observados por sondas, apesar de não serem medidas ideais, já que foram alvos de oportunidade.
+A sonda Cassini observou Himalia em 2000 ao passar próximo a Júpiter e obteve o tamanho de Himalia com um erro médio de 10 km \citep{Porco2003}.
+Em 2004, a Cassini, aproximando-se de Saturno, observou Phoebe em alta resolução obtendo um erro médio na medida de seu tamanho de 0.7 km \citep{Thomas2010}.
+Por fim, Nereida foi observado em 1989 pela sonda Voyager 2 e seu tamanho foi obtido com um erro de 25 km \citep{Smith1989}.
+Para outros satélites irregulares de Júpiter, seus tamanhos foram estimados por \cite{Rettig2001} impondo um albedo de 4\% justificando que esse valor é um albedo nominalmente utilizado para objetos do Sistema Solar Externo.
 
-\section{TNOs}
+Com o objetivo de obter parâmetros físicos para os satélites irregulares com maior precisão foram feitas predições de ocultações estelares para os 7 maiores satélites de Júpiter, o satélite Phoebe de Saturno e Nereida de Netuno.
+Esses objetos são pequenos (ver Tabela \ref{Tab: comparison-horizons}), se comparados aos TNOs e Centauros estudados na seção \ref{Sec: TNOs}, sendo que o menor dos satélites da amostra, Ananke, possui aproximadamente 28 km.
+
+As predições foram feitas utilizando o catálogo UCAC4 e efemérides do JPL.
+Um total de 588 eventos foram obtidos entre 01 de Janeiro de 2015 e 31 de Dezembro de 2017 para os 9 objetos sendo que a maioria será descartada por passar em regiões de oceano ou onde não há observadores disponíveis.
+Para Nereida, apenas uma estrela UCAC4 foi identificada para ser ocultada.
+Nesse caso, utilizamos um catálogo de estrelas feito a partir do caminho aparente do satélite de Netuno Tritão no céu para predições de ocultações estelares pelo mesmo (ver seção \ref{Sec: TNOs}).
+
+Como esses objetos são muito pequenos, em sua maioria as ocultações durarão poucos segundos, portanto apenas eventos com estrelas brilhantes serão selecionados e se houver câmeras de integração rápida disponíveis.
+Por outro lado, os satélites de Júpiter estão muito mais pertos que os TNOs e como o erro astrométrico é uma medida angular, consequentemente, o erro da sombra da ocultação projetada na Terra será muito menor que aquelas medidas para TNOs.
+Assim, a área a ser coberta para que uma ocultação por satélite irregular de Júpiter seja detectada será da ordem de poucas centenas de quilômetros.
+
+\subsection{Testes}
+
+\section{Centauros e TNOs}
 \label{Sec: TNOs}

### 2015-06-10 qualify/Qualify-altair.tex
Criado Resumo
Começado teste de ocultações de irregulares

@@
 \usepackage{arydshln}
+\usepackage{tablefootnote}
 %\usepackage[alf]{abntcite}
@@
 
-\titulo{Aplicações Astrométricas e Fotométricas para o Estudo do Sistema Solar Exterior}
+\titulo{Estudo Astrométrico e Fotométrico do Sistema Solar Exterior}
 \autor{Altair Ramos Gomes Júnior}
@@
 \subsection{Astrometria dos sat\'elites irregulares}
+\label{Subsec: Irr-astrometria}
 
@@
 \begin{table}
-\caption{\label{Tab: comparison-horizons} Comparação entre o número de posições obtidas em \cite{GomesJunior2015-Irregular} e o número utilizado nas integrações numéricas das órbitas pelo JPL como publicado por  \cite{Jacobson2012}.}
+\caption{\label{Tab: comparison-horizons} Comparação entre o número de posições obtidas em \cite{GomesJunior2015-Irregular} e o número utilizado nas integrações numéricas das órbitas pelo JPL como publicado por  \cite{Jacobson2012}.
+Os satélites estão separados por planeta (linha reta) e por família orbital (linha tracejada)}
 \begin{centering}
-\begin{tabular}{lccccc}
+\begin{tabular}{lccccccc}
 \hline  \hline
- & \multicolumn{4}{c}{Número de Posições}   &    \tabularnewline
-Satélite  & OPD  & OHP & ESO & Total  & Jacobson \tabularnewline
+Satélite & Diâm. (km)\tablefootnote{Planetary Satellite Physical Parameters - JPL: \url{http://ssd.jpl.nasa.gov/?sat_phys_par}} & Mag V  & OPD  & OHP & ESO & Total  & Jacobson \tabularnewline
 \hline
-Himalia & 854 & 357 & 23 & 1234 & 1757 \tabularnewline
-Elara & 403 & 187 & 46 & 636 & 1115 \tabularnewline
-Lysithea & 60 & 84 & 90 & 234 & 431 \tabularnewline
-Leda & 6 & 48 & 44 & 98 & 178 \tabularnewline
+Himalia & 170 & 14 & 854 & 357 & 23 & 1234 & 1757 \tabularnewline
+Elara & 86 & 16 & 403 & 187 & 46 & 636 & 1115 \tabularnewline
+Lysithea & 36 & 18 & 60 & 84 & 90 & 234 & 431 \tabularnewline
+Leda & 20 & 19 & 6 & 48 & 44 & 98 & 178 \tabularnewline
 \hdashline
-Pasiphae & 295 & 248 & 66 & 609 & 1629 \tabularnewline
-Callirrhoe & 9 & -  &  16 & 25 & 95 \tabularnewline
-Megaclite & - & -  &  10 & 10 & 50  \tabularnewline
+Pasiphae & 60 & 17 & 295 & 248 & 66 & 609 & 1629 \tabularnewline
+Callirrhoe & 9 & 21 & 9 & -  &  16 & 25 & 95 \tabularnewline
+Megaclite & 5 & 22 & - & -  &  10 & 10 & 50  \tabularnewline
 \hdashline
-Ananke & 52 & 141 & 57 & 250 & 600 \tabularnewline
-Praxidike & - & -  &   2 & 2 & 59 \tabularnewline
+Ananke & 28 & 18 & 52 & 141 & 57 & 250 & 600 \tabularnewline
+Praxidike & 7 & 21 & - & -  &   2 & 2 & 59 \tabularnewline
 \hdashline
-Carme & 90 & 204 & 37 & 331 & 973 \tabularnewline
-Sinope & 41 & 169 & 11 & 221 & 854 \tabularnewline
-Themisto & - & - & 16 & 16 & 55 \tabularnewline
+Carme & 46 & 18 & 90 & 204 & 37 & 331 & 973 \tabularnewline
+Sinope & 38 & 18 & 41 & 169 & 11 & 221 & 854 \tabularnewline
+Themisto & 8 & 21 & - & - & 16 & 16 & 55 \tabularnewline
 \hline
-Phoebe & 1239 & 516 & 32 & 1787 & 3479 \tabularnewline
+Phoebe & 213 & 16 & 1239 & 516 & 32 & 1787 & 3479 \tabularnewline
 \hdashline
-Siarnaq & - & 20 & 56 & 76 & 239 \tabularnewline
-Paaliaq & - & - & 11 & 11 & 82 \tabularnewline
+Siarnaq & 40 & 20 & - & 20 & 56 & 76 & 239 \tabularnewline
+Paaliaq & 22 & 21 & - & - & 11 & 11 & 82 \tabularnewline
 \hdashline
-Albiorix & - & - & 46 & 46 & 137 \tabularnewline
+Albiorix & 32 & 20 & - & - & 46 & 46 & 137 \tabularnewline
 \hline
-Sycorax & - & - & 35 & 35 & 237 \tabularnewline
+Sycorax & 150 & 21 & - & - & 35 & 35 & 237 \tabularnewline
 \hline
-Nereid & 803 & - & 99 & 902 & 716 \tabularnewline
+Nereid & 340 & 19 & 803 & - & 99 & 902 & 716 \tabularnewline
 \hline
@@
 \subsection{Previs\~ao das ocultaç\~oes}
+\label{Subsec: Irr-predic}
 
@@
 \subsection{Testes}
+\label{Subsec: Irr-testes}
+
+\indent \indent Até que o primeiro catálogo GAIA seja lançado, as posições das estrelas usadas serão as do catálogo UCAC4.
+Porém esse catálogo possui imprecisões relevantes principalmente para as estrelas mais fracas.
+Uma das estrelas testadas apresentou um offset da posição do catálogo para a data da observação de mais de 600 mas.
+Para a estrela em questão, esse valor representa uma variação de aproximadamente 1900 km da sombra projetada na Terra.
+
+Para corrigir esses erros, será necessário observar as estrelas a serem ocultadas dias antes do evento obtendo os offsets correspondentes e calibrando a predição.
+Erros em ascensão reta interferem majoritariamente no instante em que a ocultação ocorrerá.
+Já erros em declinação principalmente na latitude geográfica em que a sombra passará.
+
+Outra fonte de erro na predição é a posição do satélite.
+Como mostradas na seção \ref{Subsec: Irr-astrometria}, as órbitas dos satélites irregulares não são muito precisas havendo erros sistemáticos em seus elementos orbitais.
+Infelizmente o JPL não provê suas efemérides com os erros estimados das posições.
+Dessa forma, há três alternativas para a correção das posições dos satélites.
+
+A primeira delas é observar o satélite dias antes, quanto mais próximo do evento melhor, e determinar um offset.
+O offset, portanto, seria considerado estável dentro desses poucos dias uma vez que ele não se move muito em sua órbita.
+Porém, nem sempre é possível observar o alvo com antecedêndia.
+
+A segunda alternativa é recalcular as órbitas dos satélites, gerar novas efemérides e estimar seus erros.
+Espera-se que com o grande número de posições publicadas em \cite{GomesJunior2015-Irregular} as órbitas sejam calculadas com maior precisão.
+Esse integração numérica será trabalhada durante o doutorado sanduíche em Paris em colaboração com o Dr.
+Jean-Eudes Arlot e o Dr.
+Josselin Desmars.
+Um trabalho semelhante foi abordado para ocultação de Centauros e TNOs (ver seção \ref{Sec: TNOs}).
+
+Por fim, a terceira alternativa consiste em se fazer um ajuste dos offset obtidos em \cite{GomesJunior2015-Irregular} em função da anomalia verdadeira, tempo e/ou anomalia média.
+Essa escolha é suportada pelas dispersões dos offsets em função da anomalia verdadeira (por exemplo, Fig.
+\ref{Fig: carme_anom}, para Carme) que apresentaram características notáveis.
+Os parâmetros a serem utilizados para cada satélite serão estudados e aplicados.
+Esse é o método que será utilizado para o artigo de predições de satélites irregulares que será submetido logo.
 

### 2015-06-10 qualify/Qualify-altair.tex
Feito predição e testes de ocultações, falta terminar testes

@@
 
-\subsection{Previs\~ao das ocultaç\~oes}
+\subsection{Prediç\~ao das ocultaç\~oes}
 \label{Subsec: Irr-predic}
@@
 
+Os primeiros ajustes foram feitos apenas considerando os offsets em declinação, pois são os que apresentaram características mais evidentes e, por seus erros representarem principalmente erros em latitudes, são os principais responsáveis pela não detecção da ocultação.
+Nesse caso, utilizamos um ajuste em função de três parâmetros: sen${f}, \cos{f}$ e uma constante, onde $f$ é a anomalia verdadeira dada pela efeméride.
+Esse ajuste é mostrado na Fig.
+\ref{Fig: Carme-dec-fit}.
+
+\begin{figure}
+\begin{centering}
+\includegraphics[scale=0.45]{figuras/Carme-DEC_fit.png} 
+\caption{Ajuste dos offsets em declinação de Carme publicados em \cite{GomesJunior2015-Irregular} em função da anomalia verdadeira}
+\label{Fig: Carme-dec-fit}
+\end{centering}
+\end{figure}
+
+É possível ver para o caso de Carme que o ajuste representa razoavelmente bem a dispersão dos offsets em função da anomalia verdadeira.
+Para outros satélites, ajuste em primeira ordem do seno e do cosseno da anomalia verdadeira também representam razoavelmente bem os offsets em declinação.
+Para os offsets em ascenção reta, o ajuste com três parâmetros não é suficiente.
+
+Observar uma ocultação estelar de um satélite irregular exige um esforço muito grande por parte dos observadores.
+A sombra cobrirá uma região muito restrita na Terra devido aos satélites irregulares serem pequenos.
+Portanto, antes de iniciar uma grande campanha observacional, testaremos algumas ocultações, principalmente dos objetos maiores para confirmar a predição.
+
+Esse teste consiste em se observar o objeto ocultante e a estrela a ser ocultada próximo ao evento, quando os dois objetos estiverem presentes no mesmo campo, preferencialmente próximo ao instante do evento quando os objetos estarão o mais próximo possível.
+Assim as posições relativas entre os dois objetos terá mínima influência dos erros das estrelas de catálogo utilizadas na redução e das possíveis distorções de campo.
+
+Até o momento, apenas dois testes de ocultações foram feitas, uma de Himalia que ocorreu dia 03 de Março de 2015 e a segunda de Elara que ocorreu dia 30 de Março de 2015.
+Para cada evento quatro mapas foram gerados: o primeiro com as posições nominais da estrela e do satélite para o instante previsto; a segunda com o offset calculado como descrito anteriomente; o terceiro com os offsets da estrela e do satélite a partir de observações feitas alguns dias antes da ocultação quando os dois objetos estavam separados (campos diferentes); e o quarto a partir de observações feitas com menos de 24 horas de diferença da data prevista com a estrela e o satélite próximos no mesmo campo.
+
+\begin{figure}
+\begin{centering}
+\subfigure[]{\includegraphics[scale=0.16]{figuras/Elara_1_2015-03-30T01:45:15.png} \label{Fig: carme_alfa}}
+\subfigure[]{\includegraphics[scale=0.16]{figuras/Elara_2_2015-03-30T01:45:03.png} \label{Fig: carme_delta}}
+\subfigure[]{\includegraphics[scale=0.16]{figuras/Elara_3_2015-03-30T01:45:41.png}  \label{Fig: carme_delta}}
+\subfigure[]{\includegraphics[scale=0.16]{figuras/Elara_4_2015-03-30T01:44:13.png}  \label{Fig: carme_delta}}
+\caption{Predições para Elara}
+\label{Fig: occ-Elara}
+\end{centering}
+\end{figure}
+
+\begin{figure}
+\begin{centering}
+\subfigure[]{\includegraphics[scale=0.16]{figuras/Himalia_1_2015-03-03T00:39:25.png}  \label{Fig: carme_alfa}}
+\subfigure[]{\includegraphics[scale=0.16]{figuras/Himalia_2_2015-03-03T00:39:34.png}  \label{Fig: carme_delta}}
+\subfigure[]{\includegraphics[scale=0.16]{figuras/Himalia_3_2015-03-03T00:39:14.png}   \label{Fig: carme_delta}}
+\subfigure[]{\includegraphics[scale=0.16]{figuras/Himalia_4_2015-03-03T00:39:31.png}   \label{Fig: carme_delta}}
+\caption{Predições para Himalia}
+\label{Fig: occ-Himalia}
+\end{centering}
+\end{figure}
+
 \section{Centauros e TNOs}

### 2015-06-11 qualify/Qualify-altair.tex
Terminei a parte de ocultações de irregulares

@@
 
-\titulo{Estudo Astrométrico e Fotométrico do Sistema Solar Exterior}
+\titulo{Um Estudo Astrométrico e Fotométrico do Sistema Solar Exterior}
 \autor{Altair Ramos Gomes Júnior}
@@
 
+\chapter{Astrometria dos sat\'elites irregulares dos planetas gigantes}
+\label{Cap: Irr-astrometria}
+
+\indent \indent Os satélites irregulares dos planetas gigantes são menores que as regulares, possuindo órbitas mais excêntricas, inclinadas e distantes.
+Na maioria dos casos, essas órbitas são retrógradas.
+Devido suas configurações orbitais, é altamente aceito que estes objetos foram capturados nos estágios iniciais da formação do sistema solar \citep{Sheppard2003}.
+
+Por serem pequenos, eles são pouco brilhantes e só foram descobertos no último século\footnote{Website: http://ssd.jpl.nasa.gov/?sat\_discovery}.
+Dentre os satélite irregulares dos planetas gigantes, poucos são aqueles que possuem medidas de seus parâmetros físicos.
+Apenas Himalia, Phoebe e Nereida foram observados por sondas, apesar de não serem medidas ideais, já que foram alvos de oportunidade.
+A sonda Cassini observou Himalia em 2000 ao passar próximo a Júpiter e obteve o tamanho de Himalia com um erro médio de 10 km \citep{Porco2003}.
+Em 2004, a Cassini, aproximando-se de Saturno, observou Phoebe em alta resolução obtendo um erro médio na medida de seu tamanho de 0.7 km \citep{Thomas2010}.
+Por fim, Nereida foi observado em 1989 pela sonda Voyager 2 e seu tamanho foi obtido com um erro de 25 km \citep{Smith1989}.
+Para outros satélites irregulares de Júpiter, seus tamanhos foram estimados por \cite{Rettig2001} impondo um albedo de 4\% cam a justifica de que esse valor é um albedo nominalmente utilizado para objetos do Sistema Solar Externo.
+
+Existem alguns mecanismos de captura de objetos por planetas gigantes propostos na literatura.
+Há o Arrasto Gasoso na nebulosa circumplanetária primordial \citep{Sheppard2005} onde o movimento objeto seria afetado gás e sua velocidade diminuiria até que fosse capturado pelo planeta.
+Outro mecanismo se chama captura pull-down \citep{Sheppard2005}, onde a massa do planeta aumentaria enquanto o objeto estivesse temporariamente capturado.
+
+Um mecanismo baseado no modelo de Nice \citep{Morbidelli2005, Tsiganis2005, Gomes2005} foi proposto por \cite{Nesvorny2007} e, especificamente para Júpiter, por \citealp{Nesvorny2014}.
+Durante a instabilidade do sistema solar primordial, ocorreram encontros entre os planetas externos.
+Esses encontros planetários poderiam trocar energia e momento angular entre os planetas e os objetos próximos tornando possível a captura de um corpo irregular por um planeta gigante.
+Nesse cenário, a taxa de sobrevivência de satélites anteriores ao LHB (Bombardeamento Pesado Tardio\footnote{"Late Heavy Bombardment", em inglês}) é muito pequena.
+
+Um outro mecanismo importante é a captura através de interações colisionais \citep{Sheppard2005}.
+Uma colisão entre dois corpos pequenos dentro da esfera de Hill do planeta poderia gerar objetos fragmentados e a energia dissipativa poderia ser tal que alguns desses objetos seriam capturados.
+
+Alguns desses objetos formam grupos dinâmicos com elementos orbitais semelhantes, chamados famílias, similar às famílias encontradas no Cinturão Principal de Asteróides.
+Essas famífias podem ter sido criadas por um corpo pai que se fragmentou por colisões com cometas ou outros asteróides \citep{Nesvorny2004}.
+Colisões com cometas tem uma probabilidade maior de ocorrer durante o Bombardeamento Pesado Tardio (LHB) \citep{Gomes2005}.
+
+\cite{Nesvorny2003} estudou as taxas de colisões entre satélites irregulares e concluiu que alguns satélites podem ter sido removidos por colisão com um satélite maior.
+A taxa de colisão entre os satélites do grupo de Himalia (Himalia, Elara, Lysithea and Leda, principalmente), por exemplo, foi encontrado como sendo maior que "um" durante a idade do sistema solar sugerindo que sua estrutura atual foi originada por colisão satélite-satélite.
+
+Para Phoebe, materiais ejetados de sua superfície causadas por impactos poderiam evoluir devido ao arrasto de Poynting-Robertson e colidir com Iapetus causando causando a grande variação de albedo observada no satélite \citep{Nesvorny2003}.
+De fato, A Cassini detectou em Phoebe uma característica de absorção em 2.42 $\mu m$, provavelmente combinações de CN, que foi também detectada no lado escuro de Iapetus \citep{Clark2005}
+
+Se esses objetos foram capturados, permanece a pergunta de onde eles vieram.
+\citealp{Clark2005} mostraram a partir de espectroscopia da Cassini que Phoebe tem uma superfície provavelmente coberta por materiam do sistema solar externo e \citealp{Grav2003} mostraram que os satélites de grupo prógrado de Júpiter Himalia tem cores cinzas implicando que suas superfícies são similares a asteróides tipo C.
+Nesse mesmo trabalho, o grupo retrógrado de Júpiter Carme foi encontrado como tendo cores superficiais semelhantes à asteróides tipo D como a família de Hilda ou famílias de troianos, enquanto JXIII Kalyke tem uma cor mais avermelhada como centauros ou objetos trans-Netunianos (TNOs).
+
+Para os satélites de saturno, \citealp{Grav2007} mostraram através de suas cores e inclinações espectrais que esses satélites contêm uma fração mais ou menos igual de objetos semelhantes a asteróides tipo c, P e D, mas SXXII Ijiraq é um pouco mais avermelhado que objetos tipo D.
+Esses trabalhos sugerem origins diferentes para os satélites irregulares.
+
+Nesse contexto, durante o mestrado, foi realizado junto ao grupo um trabalho de caráter astrométrico para a melhoria das efemérides dos satélites irregulares de Júpiter Saturno.
+Em colaboração com o Dr.
+Jean-Eudes Arlot do IMCCE, reduzi um banco de dados observado no Observatoire Haute-Provence (OHP) entre 1998 e 2008, contendo mais de 28 mil posições para 10 satélites.
+Reduzi também um banco de dados com mais de 100 mil imagens obtidas no Observatório do Pido dos Dias (OPD) entre 1992 e 2014.
+Já no doutorado, neste mesmo âmbito, reduzi 810 observações feitas no European Southern Observatory (ESO) em 24 noites utilizando o detector mosaico CCD Wide Field Imager (WFI).
+
+Mais de 8000 posições foram identificadas como pertencentes a satélites irregulares.
+Porém, devido à grande gama de configurações (3 sítios, 5 telescópios, mais de 10 câmeras e mais de 10 filtros) e às condições observacionais de algumas noites, 6523 posições foram selecionadas como boas posições, ou seja, possuem 3 ou mais estrelas do catálogo de referência (UCAC4), estão dentro de 2$\sigma$ da dispersão das posições da noite e a dispersão da noite à qual pertence não é maior que 2$\sigma$ da média das dispersões das noites.
+Essa estatística é feita satélite por satélite.
+O trabalho está aceito para publicação \citep{GomesJunior2015-Irregular} e o catálogo de posições está disponível no CDS.
+
+Um dos principais resultados desse trabalho foi a grande quantidade de posições obtidas para os satélites irregulares em comparação com a quantidade utilizada nas integrações numéricas atuais (ver tabela \ref{Tab: comparison-horizons}).
+
+\begin{table}[h]
+\caption{\label{Tab: comparison-horizons} Comparação entre o número de posições obtidas em \cite{GomesJunior2015-Irregular} e o número utilizado nas integrações numéricas das órbitas pelo JPL como publicado por  \cite{Jacobson2012}.
+Os satélites estão separados por planeta (linha reta) e por família orbital (linha tracejada)}
+\begin{centering}
+\begin{tabular}{lccccccc}
+\hline  \hline
+Satélite & Diâm. (km)\tablefootnote{Planetary Satellite Physical Parameters - JPL: \url{http://ssd.jpl.nasa.gov/?sat_phys_par}} & Mag V  & OPD  & OHP & ESO & Total  & Jacobson \tabularnewline
+\hline
+Himalia & 170 & 14 & 854 & 357 & 23 & 1234 & 1757 \tabularnewline
+Elara & 86 & 16 & 403 & 187 & 46 & 636 & 1115 \tabularnewline
+Lysithea & 36 & 18 & 60 & 84 & 90 & 234 & 431 \tabularnewline
+Leda & 20 & 19 & 6 & 48 & 44 & 98 & 178 \tabularnewline
+\hdashline
+Pasiphae & 60 & 17 & 295 & 248 & 66 & 609 & 1629 \tabularnewline
+Callirrhoe & 9 & 21 & 9 & -  &  16 & 25 & 95 \tabularnewline
+Megaclite & 5 & 22 & - & -  &  10 & 10 & 50  \tabularnewline
+\hdashline
+Ananke & 28 & 18 & 52 & 141 & 57 & 250 & 600 \tabularnewline
+Praxidike & 7 & 21 & - & -  &   2 & 2 & 59 \tabularnewline
+\hdashline
+Carme & 46 & 18 & 90 & 204 & 37 & 331 & 973 \tabularnewline
+Sinope & 38 & 18 & 41 & 169 & 11 & 221 & 854 \tabularnewline
+Themisto & 8 & 21 & - & - & 16 & 16 & 55 \tabularnewline
+\hline
+Phoebe & 213 & 16 & 1239 & 516 & 32 & 1787 & 3479 \tabularnewline
+\hdashline
+Siarnaq & 40 & 20 & - & 20 & 56 & 76 & 239 \tabularnewline
+Paaliaq & 22 & 21 & - & - & 11 & 11 & 82 \tabularnewline
+\hdashline
+Albiorix & 32 & 20 & - & - & 46 & 46 & 137 \tabularnewline
+\hline
+Sycorax & 150 & 21 & - & - & 35 & 35 & 237 \tabularnewline
+\hline
+Nereid & 340 & 19 & 803 & - & 99 & 902 & 716 \tabularnewline
+\hline
+\end{tabular}
+\par\end{centering}
+\end{table}
+
+Com o objetivo ver o potencial das nossas posições em melhorar as órbitas dos satélites irregulares, analizamos os offsets das nossas posições com respeito às efemérides do JPL.
+Tomando Carme como exemplo, plotamos os offsets médios das efemérides para cada noite na Fig.
+\ref{Fig: carme_anom} e suas respectivas dispersões (erro de barra 1$\sigma$) em função da anomalia verdadeira em ascenção reta e declinação.
+A figura mostra claramente um erro sistemático em declinação.
+Quando Carme está próximo a seu apojove (anomalia verdadeira = 180$\degr$), seus offsets tem maior probabilidade de serem mais negativos do que aqueles próximos ao seu perijove (anomalia verdadeira = 0$\degr$).
+Todos os offsets obtidos de observações feitas em quatro telescópios diferentes usando câmeras e filtros diferentes estão em bom acordo, o que significa que há um erro nas efemérides de Carme, muito provavelmente devido a um erro em sua inclinação orbital.
+
+\begin{figure}[h]
+\begin{centering}
+\subfigure[Right Ascension]{\includegraphics[scale=0.45]{figuras/Carme_RA}\label{Fig: carme_alfa}}
+\subfigure[Declination]{\includegraphics[scale=0.45]{figuras/Carme_DEC}\label{Fig: carme_delta}}
+\caption{Offsets médios das efemérides e dispersões (barra de erro 1$\sigma$) nas coordenadas de Carme tomadas noite a noite por anomalia verdadeira para cada telescópio.
+O quadrado vermelho são as observações com o telescópio do OPD Perkin-Elmer, o círculo azul para o Boller \& Chivens também do OPD, o triângulo pra cima preto para o OHP e a estrela verde para o ESO.}
+\label{Fig: carme_anom}
+\end{centering}
+\end{figure}
+
+Esse padrão em declinação também foi identificado para outros satélites como Pasiphae e Ananke.
+Para alguns satélites a cobertura orbital não foi suficiente para indicar claramente a presença de erros sistemáticos em elementos orbitais específicos.
+Os gráficos para todos os satélites estão disponíveis em \cite{GomesJunior2015-Irregular}.
+
+Um método que tem se mostrado eficiente para a obtenção desses parâmetros é o métodode ocultações estelares, que proporciona medidas tão precisas que são apenas superadas por medidas oriundas de sondas.
+Uma vez que suas efemérides não são muito precisas, predizer e observar ocultações estelares é muito difícil e nenhuma observação de um evento desses para um satélite irregular é encontrado na literatura.
+Um projeto nesse sentido está sendo encaminhado e será discutido na seção \ref{Sec: Irregulares}.
+
+\chapter{Astrometria de Netuno e Tritão}
+\label{Cap: Netuno}
+
 \chapter{Ocultações Estelares}
 \label{Cap: observacoes}
+
+%Since their ephemerides are not very precise, predicting and observing stellar occultations are very difficult, and no observation of such an event for an irregular satellite is found in the literature.
+The precise star positions to be derived by the ESA astrometry satellite Gaia \citep{deBruijne2012} will render better predictions with the only source of error being the ephemeris.
+The positions derived from our observations can be used in new orbital numerical integrations, generating more precise ephemerides.
+
+%The power of stellar occultations for observing relatively small diameter solar system objects is supported by recent works, such as the discovery of a ring system around the Centaur (10199) Chariklo \citep{Braga-Ribas2014}.
+Once irregular satellites start to be observed by this technique, it will be possible to obtain their physical parameters (shape, size, albedo, density) with unprecedented precision.
+For instance, in this case, sizes could be obtained with kilometer accuracy.
+The knowledge of these parameters would in turn bring valuable information for studying the capture mechanisms and origin of the irregular satellites.
 
@@
 
-A sonda da NASA \textit{Dawn} poderá responder esses questões que são importantes não só no conhecimento do próprio Ceres, mas também para todas as técnicas usadas até agora para o estudo das propriedades físicas dos pequenos objetos do Sistema Solar, como as ocultações estelares.
+A sonda da NASA \textit{Dawn} poderá responder esses questões que são importantes não só no conhecimento do próprio Ceres, mas também para calibrar todas as técnicas usadas até agora para o estudo das propriedades físicas dos pequenos objetos do Sistema Solar, como as ocultações estelares.
 
@@
 
-\subsection{Astrometria dos sat\'elites irregulares}
-\label{Subsec: Irr-astrometria}
-
-\indent \indent Durante o mestrado, foi realizado junto ao grupo um trabalho de caráter astrométrico para a melhoria das efemérides dos satélites irregulares de Júpiter Saturno.
-Em colaboração com o Dr.
-Jean-Eudes Arlot do IMCCE, reduzi um banco de dados observado no Observatoire Haute-Provence (OHP) entre 1998 e 2008, contendo mais de 28 mil posições para 10 satélites.
-Reduzi também um banco de dados com mais de 100 mil imagens obtidas no Observatório do Pido dos Dias (OPD) entre 1992 e 2014.
-Já no doutorado, neste mesmo âmbito, reduzi 810 observações feitas no European Southern Observatory (ESO) em 24 noites utilizando o detector mosaico CCD Wide Field Imager (WFI).
-
-Mais de 8000 posições foram identificadas como pertencentes a satélites irregulares.
-Porém, devido à grande gama de configurações (3 sítios, 5 telescópios, mais de 10 câmeras e mais de 10 filtros) e às condições observacionais de algumas noites, 6523 posições foram selecionadas como boas posições, ou seja, possuem 3 ou mais estrelas do catálogo de referência (UCAC4), estão dentro de 2$\sigma$ da dispersão das posições da noite e a dispersão da noite à qual pertence não é maior que 2$\sigma$ da média das dispersões das noites.
-Essa estatística é feita satélite por satélite.
-O trabalho está aceito para publicação \citep{GomesJunior2015-Irregular} e o catálogo de posições está disponível no CDS.
-
-Um dos principais resultados desse trabalho foi a grande quantidade de posições obtidas para os satélites irregulares em comparação com a quantidade utilizada nas integrações numéricas atuais (ver tabela \ref{Tab: comparison-horizons}).
-
-\begin{table}
-\caption{\label{Tab: comparison-horizons} Comparação entre o número de posições obtidas em \cite{GomesJunior2015-Irregular} e o número utilizado nas integrações numéricas das órbitas pelo JPL como publicado por  \cite{Jacobson2012}.
-Os satélites estão separados por planeta (linha reta) e por família orbital (linha tracejada)}
-\begin{centering}
-\begin{tabular}{lccccccc}
-\hline  \hline
-Satélite & Diâm. (km)\tablefootnote{Planetary Satellite Physical Parameters - JPL: \url{http://ssd.jpl.nasa.gov/?sat_phys_par}} & Mag V  & OPD  & OHP & ESO & Total  & Jacobson \tabularnewline
-\hline
-Himalia & 170 & 14 & 854 & 357 & 23 & 1234 & 1757 \tabularnewline
-Elara & 86 & 16 & 403 & 187 & 46 & 636 & 1115 \tabularnewline
-Lysithea & 36 & 18 & 60 & 84 & 90 & 234 & 431 \tabularnewline
-Leda & 20 & 19 & 6 & 48 & 44 & 98 & 178 \tabularnewline
-\hdashline
-Pasiphae & 60 & 17 & 295 & 248 & 66 & 609 & 1629 \tabularnewline
-Callirrhoe & 9 & 21 & 9 & -  &  16 & 25 & 95 \tabularnewline
-Megaclite & 5 & 22 & - & -  &  10 & 10 & 50  \tabularnewline
-\hdashline
-Ananke & 28 & 18 & 52 & 141 & 57 & 250 & 600 \tabularnewline
-Praxidike & 7 & 21 & - & -  &   2 & 2 & 59 \tabularnewline
-\hdashline
-Carme & 46 & 18 & 90 & 204 & 37 & 331 & 973 \tabularnewline
-Sinope & 38 & 18 & 41 & 169 & 11 & 221 & 854 \tabularnewline
-Themisto & 8 & 21 & - & - & 16 & 16 & 55 \tabularnewline
-\hline
-Phoebe & 213 & 16 & 1239 & 516 & 32 & 1787 & 3479 \tabularnewline
-\hdashline
-Siarnaq & 40 & 20 & - & 20 & 56 & 76 & 239 \tabularnewline
-Paaliaq & 22 & 21 & - & - & 11 & 11 & 82 \tabularnewline
-\hdashline
-Albiorix & 32 & 20 & - & - & 46 & 46 & 137 \tabularnewline
-\hline
-Sycorax & 150 & 21 & - & - & 35 & 35 & 237 \tabularnewline
-\hline
-Nereid & 340 & 19 & 803 & - & 99 & 902 & 716 \tabularnewline
-\hline
-\end{tabular}
-\par\end{centering}
-\end{table}
-
-Com o objetivo ver o potencial das nossas posições em melhorar as órbitas dos satélites irregulares, analizamos os offsets das nossas posições com respeito às efemérides do JPL.
-Tomando Carme como exemplo, plotamos os offsets médios das efemérides para cada noite na Fig.
-\ref{Fig: carme_anom} e suas respectivas dispersões (erro de barra 1$\sigma$) em função da anomalia verdadeira em ascenção reta e declinação.
-A figura mostra claramente um erro sistemático em declinação.
-Quando Carme está próximo a seu apojove (anomalia verdadeira = 180$\degr$), seus offsets tem maior probabilidade de serem mais negativos do que aqueles próximos ao seu perijove (anomalia verdadeira = 0$\degr$).
-Todos os offsets obtidos de observações feitas em quatro telescópios diferentes usando câmeras e filtros diferentes estão em bom acordo, o que significa que há um erro nas efemérides de Carme, muito provavelmente devido a um erro em sua inclinação orbital.
-
-\begin{figure}[h]
-\begin{centering}
-\subfigure[Right Ascension]{\includegraphics[scale=0.45]{figuras/Carme_RA}\label{Fig: carme_alfa}}
-\subfigure[Declination]{\includegraphics[scale=0.45]{figuras/Carme_DEC}\label{Fig: carme_delta}}
-\caption{Offsets médios das efemérides e dispersões (barra de erro 1$\sigma$) nas coordenadas de Carme tomadas noite a noite por anomalia verdadeira para cada telescópio.
-O quadrado vermelho são as observações com o telescópio do OPD Perkin-Elmer, o círculo azul para o Boller \& Chivens também do OPD, o triângulo pra cima preto para o OHP e a estrela verde para o ESO.}
-\label{Fig: carme_anom}
-\end{centering}
-\end{figure}
-
-Esse padrão em declinação também foi identificado para outros satélites como Pasiphae e Ananke.
-Para alguns satélites a cobertura orbital não foi suficiente para indicar claramente a presença de erros sistemáticos em elementos orbitais específicos.
-Os gráficos para todos os satélites estão disponíveis em \cite{GomesJunior2015-Irregular}.
-
 \subsection{Prediç\~ao das ocultaç\~oes}
@@
 
-\indent \indent Dentre os satélite irregulares dos planetas gigantes, poucos são aqueles que possuem medidas de seus parâmetros físicos.
-Apenas Himalia, Phoebe e Nereida foram observados por sondas, apesar de não serem medidas ideais, já que foram alvos de oportunidade.
-A sonda Cassini observou Himalia em 2000 ao passar próximo a Júpiter e obteve o tamanho de Himalia com um erro médio de 10 km \citep{Porco2003}.
-Em 2004, a Cassini, aproximando-se de Saturno, observou Phoebe em alta resolução obtendo um erro médio na medida de seu tamanho de 0.7 km \citep{Thomas2010}.
-Por fim, Nereida foi observado em 1989 pela sonda Voyager 2 e seu tamanho foi obtido com um erro de 25 km \citep{Smith1989}.
-Para outros satélites irregulares de Júpiter, seus tamanhos foram estimados por \cite{Rettig2001} impondo um albedo de 4\% justificando que esse valor é um albedo nominalmente utilizado para objetos do Sistema Solar Externo.
-
-Com o objetivo de obter parâmetros físicos para os satélites irregulares com maior precisão foram feitas predições de ocultações estelares para os 7 maiores satélites de Júpiter, o satélite Phoebe de Saturno e Nereida de Netuno.
+\indent \indent Com o objetivo de obter parâmetros físicos para os satélites irregulares com maior precisão foram feitas predições de ocultações estelares para os 7 maiores satélites de Júpiter, o satélite Phoebe de Saturno e Nereida de Netuno.
 Esses objetos são pequenos (ver Tabela \ref{Tab: comparison-horizons}), se comparados aos TNOs e Centauros estudados na seção \ref{Sec: TNOs}, sendo que o menor dos satélites da amostra, Ananke, possui aproximadamente 28 km.
@@
 Outra fonte de erro na predição é a posição do satélite.
-Como mostradas na seção \ref{Subsec: Irr-astrometria}, as órbitas dos satélites irregulares não são muito precisas havendo erros sistemáticos em seus elementos orbitais.
+Como mostradas no capítulo \ref{Cap: Irr-astrometria}, as órbitas dos satélites irregulares não são muito precisas havendo erros sistemáticos em seus elementos orbitais.
 Infelizmente o JPL não provê suas efemérides com os erros estimados das posições.
@@
 
+A Fig.
+\ref{Fig: occ-Himalia} mostra os quatro mapas para o teste da ocultação de Himalia de 03 de Março de 2015.
+O mapa \ref{Fig: occ-Himalia-off22fev} foi feito a partir de posições obtidas em 22 de fevereiro observadas no telescópio Zeiss do Observatório do Pico dos Dias (OPD).
+Nesse dia, Himalia e a estrela a ser ocultada foram observados em campos separados por ainda estarem distantes.
+Na noite do evento, 03 de março, os objetos foram observados com o telescópio Perkin-Elmer do OPD pouco mais de uma hora após os instante previsto para o evento.
+Satélite e estrela estavam separados por aproximadamente 16 arcsec, portanto muito próximos no campo.
+A partir dos offsets calculados, o mapa \ref{Fig: occ-Himalia-off03mar} foi gerado.
+
 \begin{figure}
 \begin{centering}
-\subfigure[]{\includegraphics[scale=0.16]{figuras/Elara_1_2015-03-30T01:45:15.png} \label{Fig: carme_alfa}}
-\subfigure[]{\includegraphics[scale=0.16]{figuras/Elara_2_2015-03-30T01:45:03.png} \label{Fig: carme_delta}}
-\subfigure[]{\includegraphics[scale=0.16]{figuras/Elara_3_2015-03-30T01:45:41.png}  \label{Fig: carme_delta}}
-\subfigure[]{\includegraphics[scale=0.16]{figuras/Elara_4_2015-03-30T01:44:13.png}  \label{Fig: carme_delta}}
-\caption{Predições para Elara}
+\subfigure[Sem offsets]{\includegraphics[scale=0.16]{figuras/Himalia1_2015-03-03T00:39:25.png}  \label{Fig: occ-Himalia-nooff}}
+\subfigure[Offset de Himalia a partir do ajuste]{\includegraphics[scale=0.16]{figuras/Himalia2_2015-03-03T00:39:18.png}  \label{Fig: occ-Himalia-ajuste}}
+\subfigure[Obs. dia 22 de Fevereiro (campos separados)]{\includegraphics[scale=0.16]{figuras/Himalia3_2015-03-03T00:39:40.png}   \label{Fig: occ-Himalia-off22fev}}
+\subfigure[Obs. dia 03 de Março (mesmo campo)]{\includegraphics[scale=0.16]{figuras/Himalia4_2015-03-03T00:39:15.png}   \label{Fig: occ-Himalia-off03mar}}
+\caption{Predições para Himalia.
+O ponto grande em vermelho mostra o local de máxima aproximação da sombra na Terra, os pontos vermelhos menores são os centros da ocultação separados por um minuto, as linhas retas são os limites das sombras dado o tamanho do objeto e as linhas tracejadas demonstram uma diferença de 40 mas do centro da sombra.
+(a) é o mapa utilizando apenas as posições nominais da estrela e do satélite.
+(b) mostra a sombra dado um offset estimado pra posição de Himalia dado sua anomalia verdadeira segundo um ajuste feito em cima dos offsets publicados em \cite{GomesJunior2015-Irregular}.
+Em (c) são aplicados offsets às posições da estrela e do satélite dado observações feitas em 22 de fevereiro no telescópio Zeiss.
+Em (d), temos o mesmo que para (c) porém com offsets obtidos de observações feitas em 03 de Março no telescópio Perkin-Elmer quando os objetos estavam próximos no campo}
+\label{Fig: occ-Himalia}
+\end{centering}
+\end{figure}
+
+Para esse evento é possível ver que a sombra não varia muito dentre os quatro mapas sugerindo que para Himalia uma probabilidade maior de observar um evento.
+De fato, a maior diferença entre as sombras dos quatros mapas é de $25 s$ e $130 km$ na direção perpendicular às sombras.
+
+Na Fig.
+\ref{Fig: occ-Elara} temos os mapas para a ocultação de Elara do dia 30 de Março.
+As observações de teste foram tomadas dias 25 de março e 02 de abril no Zeiss.
+Na noite do dia 02 eles ainda puderam ser observadas no mesmo campo.
+Devido à Elara ser bem mais fraco, as dispersões das posições do satélite nas duas noites acabaram sendo maiores que para Himalia.
+Ainda assim, as diferenças entre os mapas obtidos foram pequenos.
+A maior diferença entre eles é de $48 s$ e $302 km$.
+
+\begin{figure}
+\begin{centering}
+\subfigure[Sem offsets]{\includegraphics[scale=0.16]{figuras/Elara1_2015-03-30T01:45:15.png} \label{Fig: occ-Elara-nooff}}
+\subfigure[Offset de Elara a partir do ajuste]{\includegraphics[scale=0.16]{figuras/Elara2_2015-03-30T01:44:59.png} \label{Fig: occ-Elara-ajuste}}
+\subfigure[Obs. dia 25 de Março (campos separados)]{\includegraphics[scale=0.16]{figuras/Elara3_2015-03-30T01:44:17.png}  \label{Fig: occ-Elara-off25mar}}
+\subfigure[Obs. dia 02 de Abril (mesmo campo)]{\includegraphics[scale=0.16]{figuras/Elara4_2015-03-30T01:45:30.png}  \label{Fig: occ-Elara-off02abr}}
+\caption{Predições para Elara. Mesmo que a Fig. \ref{Fig: occ-Himalia}, mas para Elara. O mapa (c) feito a partir de observações tomadas dia 25 de março e (d) a partir de 02 de abril.}
 \label{Fig: occ-Elara}
@@
 
+Um pedido de tempo foi feito para o telescópio Zeiss para se tentar observar uma ocultação de Himalia  dia 23 de Outubro de 2015.
+A sombra dessa ocultação passará pela região sudeste do Brasil (ver Fig.
+\ref{Fig: occ-Himalia-itajuba}).
+A ocultação passará muito baixo do céu (~22º de altura) e estará próximo ao amanhecer, porém se observar será a primera ocultação de um satélite irregular.
+
 \begin{figure}
-\begin{centering}
-\subfigure[]{\includegraphics[scale=0.16]{figuras/Himalia_1_2015-03-03T00:39:25.png}  \label{Fig: carme_alfa}}
-\subfigure[]{\includegraphics[scale=0.16]{figuras/Himalia_2_2015-03-03T00:39:34.png}  \label{Fig: carme_delta}}
-\subfigure[]{\includegraphics[scale=0.16]{figuras/Himalia_3_2015-03-03T00:39:14.png}   \label{Fig: carme_delta}}
-\subfigure[]{\includegraphics[scale=0.16]{figuras/Himalia_4_2015-03-03T00:39:31.png}   \label{Fig: carme_delta}}
-\caption{Predições para Himalia}
-\label{Fig: occ-Himalia}
-\end{centering}
+\includegraphics[scale=0.35]{figuras/Himalia_2015-10-23T07:54:47.png}
+\caption{Mapa da ocultação de Himalia que ocorrerá em 23 de Outubro de 2015}
+\label{Fig: occ-Himalia-itajuba}
 \end{figure}
@@
 \label{Sec: TNOs}
-
-\chapter{Astrometria de Netuno e Tritão}
-\label{Cap: Netuno}
 

### 2015-06-12 qualify/Qualify-altair.tex
Começado TNOs e Centauros

@@
 \indent \indent Com o objetivo de obter parâmetros físicos para os satélites irregulares com maior precisão foram feitas predições de ocultações estelares para os 7 maiores satélites de Júpiter, o satélite Phoebe de Saturno e Nereida de Netuno.
-Esses objetos são pequenos (ver Tabela \ref{Tab: comparison-horizons}), se comparados aos TNOs e Centauros estudados na seção \ref{Sec: TNOs}, sendo que o menor dos satélites da amostra, Ananke, possui aproximadamente 28 km.
+Esses objetos são pequenos (ver Tabela \ref{Tab: comparison-horizons}), se comparados aos TNOs e Centauros estudados na seção \ref{Sec: TNO-occ}, sendo que o menor dos satélites da amostra, Ananke, possui aproximadamente 28 km.
 
@@
 Para Nereida, apenas uma estrela UCAC4 foi identificada para ser ocultada.
-Nesse caso, utilizamos um catálogo de estrelas feito a partir do caminho aparente do satélite de Netuno Tritão no céu para predições de ocultações estelares pelo mesmo (ver seção \ref{Sec: TNOs}).
+Nesse caso, utilizamos um catálogo de estrelas feito a partir do caminho aparente do satélite de Netuno Tritão no céu para predições de ocultações estelares pelo mesmo (ver seção \ref{Sec: TNO-occ}).
 
@@
 Josselin Desmars.
-Um trabalho semelhante foi abordado para ocultação de Centauros e TNOs (ver seção \ref{Sec: TNOs}).
+Um trabalho semelhante foi abordado para ocultação de Centauros e TNOs (ver seção \ref{Sec: TNO-occ}).
 
@@
 Ainda assim, as diferenças entre os mapas obtidos foram pequenos.
-A maior diferença entre eles é de $48 s$ e $302 km$.
+A maior diferença entre elas é de $48 s$ e $302 km$.
 
@@
 \subfigure[Obs. dia 02 de Abril (mesmo campo)]{\includegraphics[scale=0.16]{figuras/Elara4_2015-03-30T01:45:30.png}  \label{Fig: occ-Elara-off02abr}}
-\caption{Predições para Elara. Mesmo que a Fig. \ref{Fig: occ-Himalia}, mas para Elara. O mapa (c) feito a partir de observações tomadas dia 25 de março e (d) a partir de 02 de abril.}
+\caption{Predições para Elara. Mesmo que a Fig. \ref{Fig: occ-Himalia}, mas para Elara. O mapa (c) feito a partir de observações tomadas dia 25 de março e (d) a partir de observações de 02 de abril.}
 \label{Fig: occ-Elara}
@@
 \ref{Fig: occ-Himalia-itajuba}).
-A ocultação passará muito baixo do céu (~22º de altura) e estará próximo ao amanhecer, porém se observar será a primera ocultação de um satélite irregular.
+O evento ocorrerá muito baixo do céu (~22º de altura) e estará próximo ao amanhecer, porém se observar será a primera ocultação de um satélite irregular.
 
@@
 \section{Centauros e TNOs}
-\label{Sec: TNOs}
+\label{Sec: TNO-occ}
+
+\indent \indent Como dito no Cap.
+\ref{Cap: intro}, os objetos do sistema solar exterior podem ser considerados fósseis do sistema solar.
+Dentre esses, os TNOs, por estarem mais distantes, são os que guardam as informações mais antigas.
+Ainda é incerta a distribuição de tamanhos dos TNOs e pouco se conhece sobre suas composições, estruturas internas, presença de atmosfera, densidade, etc.
+
+Com o objetivo de obter os parâmetro físicos desses objetos uma grande campanha observacional é realizada para se observar ocultações estelares por TNOs e Centauros.
+Essa campanha é liderada por pesquisadores brasileiros, franceses e espanhóis e tem colaboração com observadores espalhados por todo o mundo.
+
+\chapter{Cronograma das Atividades}
+\label{Cap: cronograma}
 

### 2015-06-12 qualify/Qualify-altair.tex
continuando texto dos tnos

@@
 
-Com o objetivo de obter os parâmetro físicos desses objetos uma grande campanha observacional é realizada para se observar ocultações estelares por TNOs e Centauros.
+Com o objetivo de obter os parâmetro físicos desses objetos uma grande campanha observacional é realizada para se observar ocultações estelares por TNOs e Centauros (a partir daqui, "TNOs" se referem a TNOs e Centauros).
 Essa campanha é liderada por pesquisadores brasileiros, franceses e espanhóis e tem colaboração com observadores espalhados por todo o mundo.
+
+Assim como para os satélites irregulares, observar uma ocultação exige um grande trabalho de predição e sua melhoria.
+Os TNOs estão muito distantes significando que erros pequenos em sua posição ou na posição da estrela podem significar um erro do local onde a sombra irá passar de muitos quilômetros.
+Preferencialmente, esse erro tem que ser menor que o tamanho angular do objeto no plano do céu.
+Por exemplo, Plutão tem um raio de $1190 \pm 5$ km \cite[em preparação]{DiasOliveira2015} que a uma distância de aproximadamente 32 UA representam um diâmetro angular de 102 mas.
+Para a maioria dos objetos selecionados, menores que Plutão e/ou mais distantes, seus tamanhos angulares são manores que 50 mas.
+
+Da mesma forma, estando muito distantes, eles se movem pouco no céu ocultando poucas esrelas, porém sendo menos brilhantes (Mag V > 18) é possível observar ocultações de estrelas mais fracas, onde a queda de brilho da estrela durante a ocultação poderá ser detectada com baixo sinal-ruído.
+
+Para identificar candidatas a ocultações estelares foi realizado uma campanha observacional no ESO para observar os caminhos aparentes dos TNOs entre 2009 e 2015.
+As observações foram feitas com o telescópio de 2.2 m e o detector mosaico WFI.
+Posições para estrelas até magnitude R=21, com nível de completeza em R=19, foram obtidas com precisões menores que 20 mas até R=20 \citep{Assafin2012}.
+Assim, estrelas que não estão no UCAC4, catálogo comumente utilizado como referência astrométrica, foram identificados bem como estrelas que estão mas não tem posições ou movimentos próprios com precisão suficiente.
+Tantos os catálogos de estrelas quanto as cadidatas a ocultação foram publicados em periódicos \citep{Assafin2010, Assafin2012, Camargo2013}.
+
+Não somente as posições das estrelas precisam ser atualizadas, mas também as posições dos TNOs.
+Portanto os TNOs também foram observados tanto no ESO quanto No Observatório do Pico dos Dias.
+À princípio as predições das ocultações eram atualizadas utilizando o método de offsets, porém, em colaboração, integrações numéricas das órbitas dos TNOs foram realizadas desenvolvendo-se uma ferramenta denominada NIMA\footnote{Numerical Integration of the Motion of an Asteroid} \citep[submetido]{Desmars2015}.
+
+Nesse trabalho, posições obtidas de ocultações já realizadas tem um peso muito maior que as restantes, diferentemente do que é feito pelo JPL.
+Da mesma forma, noites em que houveram um número de observações maior que as restantes terá um peso moderado já que se fosse utilizado um peso individual, o peso final de uma noite poderia estar superestimado levando a um erro na integração.
+Neste trabalho contribuí com observações de TNOs feitas no OPD.
+
+Vários objetos já foram observados pelo grupo, sendo que o Sistema Plutão-Caronte conta com várias ocultações observadas.
+Essa trabalho já gerou diversas publicações, incluindo na revista Nature, por exemplo \cite{Sicardy2011} e \cite{BragaRibas2014} nas quais contribuí em caráter observacional.
+A ocultação de Plutão de 04 de Maio de 2013, a qual observei, gerou duas publicações \citep{Olkin2015, DiasOliveira2015}.
+
+\cite{Sicardy2011} identificaram, a partir de uma ocultação de 06 de Novembro de 2010, que Eris é um planeta anão menor do que se esperava.
+Devido a seu brilho e distância acreditava-se que ele fosse muito maior, mas na verdade ele tem um albedo geométrico no visível de $p_v=0.96^{+0.09}_{-0.04}$ e que seu tamanho é de $1163 \pm 6$ km, muito próximo ao tamanho de Plutão.
+O alto albedo de Eris pode ser relacionada à presença de uma atmosfera colapsada devido ao fato dele estar se aproximando de seu afélio.
+
+\cite{BragaRibas2014} observaram uma ocultação do Centauro Chariklo que ocorreu dia 03 de Junho de 2013.
+O ocultação revelou que o objeto possui um sistema de anéis nunca visto antes em um objeto tão pequeno.
+Até o momento, conhecia-se apenas anéis em torno dos planetas gigantes.
+Essa descoberta levanta muitas questão sobre formação dos anéis, estabilididade ou tempo de vida e nos leva a crer que pode haver mais objetos com anéis no sistema solar.
+
+\cite{Olkin2015} determinaram a partir de observações de ocultações por Plutão que a atmosfera do objeto não colapsa durante os 248 anos de sua órbita, como acontece com Eris.
+Modelos anteriores previam um colapso de sua atmosfera devido a Plutão receber três vezes menos lus do Sol em afélio que em seu periélio.
+
+\cite{DiasOliveira2015} também a partir de ocultações de Plutão utilizaram modelos para obter os perfis de temperatura, pressão e densidade da atmosfera do objeto.
+Os modelos ajustam perfeitamente os perfis de temperatura para as duas ocultações utilizadas assumindo uma atmosfera esfericamente simétrica.
+Foi obtido que a pressão da atmosfera aumentou cerca de 6\% entre 2012 e 2013 significando que a atmosfera de Plutão ainda está se expandindo e confirmando o trabalho de \cite{Olkin2015}.
 

### 2015-06-12 qualify/Qualify-altair.tex
Terminado parte  ocultações de TNOs

@@
 \indent Já no caso de Urano e Netuno, nenhuma sonda exclusiva foi enviada, apenas as Voyagers I e II os visitaram, mas não permaneceram nos sistemas.
-A sonda New Horizons visitará Plutão em 2015 e obterá parâmetros físicos para Plutão e seus satélites (primeira visita por sonda a um satélite do cinturão de Kuiper), porém será uma passagem rápida e o acompanhamento da evolução do sistema, incluindo a evolução da atmosfera de Plutão se dará por observações de solo.\\
+A sonda New Horizons já está próxima a Plutão e obterá parâmetros físicos para Plutão e seus satélites (primeira visita por sonda a um objeto do cinturão de Kuiper), porém será uma passagem rápida e o acompanhamento da evolução do sistema, incluindo a evolução da atmosfera de Plutão se dará por observações de solo.\\
 \indent A quantidade de objetos descobertos além da órbita de Saturno tem aumentado muito desde o fim do século passado.
@@
 
-\chapter{Astrometria dos sat\'elites irregulares dos planetas gigantes}
+\chapter{Astrometria dos satélites irregulares dos planetas gigantes}
 \label{Cap: Irr-astrometria}
@@
 \label{Cap: observacoes}
+
+\indent \indent Ocultação estelar é um evento que ocorre quando um objeto do sistema solar passa na frente de uma estrela de forma que o brilho da estrela é completamente ou parcialmente apagado durante um intervalo de tempo.
+A variação de brilho da estrela pode ser medido fazendo-se observações com exposições curtas.
+O produto gerado é uma curva de luz luminosidade x tempo.
+Essa técnica pode ser utilizada para se obter parâmetros físicos dos objetos ocultantes como tamanho, densidade, detecção da presença de atmosfera, satélites ou anéis, etc \citep{Elliot1979}.
+
+A Fig.
+\ref{Fig: occ-scheme}.1 mostra uma curva de luz esquemática de uma ocultação estelar central por um planeta com atmosfera e anéis.
+Ao ocultar a estrela, os anéis passarão primeiro, a densidade do anel, tamanho e tempo de exposição determinará a queda de brilho da estrela.
+Ao chegar no corpo principal, caso não possua atmosfera, a queda de brilho será abrupta, caso possua, a queda de brilho será mais suave devido à refração da luz da estrela na atmosfera dificultando a determinação da superfície do objeto.
+
+\begin{figure}[h]
+\begin{centering}
+\includegraphics[scale=0.75]{figuras/stellarocc.png} 
+\label{Fig: occ-scheme}
+\caption{Curva de luz esquemática de uma ocultação estelar por um planeta com atmosfera e anéis. Fonte: \cite{Elliot1979}}
+\end{centering} 
+\end{figure}
+
+A suavidade da curva portanto será determinada pelo gradiente de temperatura, pressão e densidade da atmosfera, presença de nuvens, etc.
+A luz da estrela nunca será completamente apagada quando ocultada por um objeto com atmosfera devido à refração da luz na atmosfera onde parte dela convergirá na direção do observador.
+Caso a ocultação seja central para o observador haverá um aumento de brilho no meio do evento pois a convergência da luz será máxima.
+A atmosfera, no caso, agiria como uma lente como mostrado na Fig.
+\ref{Fig: occ-scheme}.1.
+
+Quando o evento é observado por diversos telescópios espalhados, cada observador verá a ocultação em latitudes diferentes.
+Essas diversas curvas de luz permitirá reconstruir a forma do objeto possivelmente identificando estruturas superficiais como montanhas e crateras.
+Como a estrela está muito distante, podendo ser considerada quase pontual, a sombra do objeto durante a ocultação é muito próximo ao seu tamanho real.
+
+Para se observar esses eventos, portanto, é necessário uma campanha grande com observadores espalhados por várias regiões para cobrir a sombra do objeto.
+A precisão nas posições da estrela e do objeto são cruciais pois o evento ocorre em instantes específicos e passará por regiões específicas da Terra alheias à escolha do observador.
 
@@
 
-\subsection{Prediç\~ao das ocultaç\~oes}
+\subsection{Predição das ocultações}
 \label{Subsec: Irr-predic}
@@
 \cite{Olkin2015} determinaram a partir de observações de ocultações por Plutão que a atmosfera do objeto não colapsa durante os 248 anos de sua órbita, como acontece com Eris.
-Modelos anteriores previam um colapso de sua atmosfera devido a Plutão receber três vezes menos lus do Sol em afélio que em seu periélio.
+Modelos anteriores previam um colapso de sua atmosfera devido a Plutão receber três vezes menos luz do Sol em afélio que em seu periélio.
 
-\cite{DiasOliveira2015} também a partir de ocultações de Plutão utilizaram modelos para obter os perfis de temperatura, pressão e densidade da atmosfera do objeto.
+\cite{DiasOliveira2015}, também a partir de ocultações de Plutão, utilizaram modelos para obter os perfis de temperatura, pressão e densidade da atmosfera do objeto.
 Os modelos ajustam perfeitamente os perfis de temperatura para as duas ocultações utilizadas assumindo uma atmosfera esfericamente simétrica.
 Foi obtido que a pressão da atmosfera aumentou cerca de 6\% entre 2012 e 2013 significando que a atmosfera de Plutão ainda está se expandindo e confirmando o trabalho de \cite{Olkin2015}.
+
+No dia 29 de Junho de 2015 haverá uma ocultação de Plutão de máxima importância.
+Será a última antes da chegada da sonda \textit{New Horizons} no sistema e poderá ser observada na Austrália.
+Observada a ocultação, os modelos de perfil de atmosfera para Plutão poderão ser calibrados com os dados da \textit{New Horizons} para serem utilizadas em futuras ocultações e estudar a evolução da atmosfera do planeta anão.
+
+Em 29 de Abril de 2014, uma ocultação de Chariklo foi observada na África.
+Eu trabalhei com a redução de três curvas de luz (Gifberb, SAAO e Springbok).
+Um fato interessante dessa ocultação foi que a estrela era uma binária não catalogada.
+Como a segunda estrela era bem mais fraca, acreditava-se que poderia ser algum material próximo à Chariklo.
+
+Essa ocultação foi importante para obter com total certeza a direção do pólo de Chariklo.
+Além disso, as câmeras utilizadas eram de alta qualidade possibilitando obter que pelo menos um dos anéis é opaco.
+Uma das cordas, fortuitamente, passou contígua ao anel observando apenas o anel externo.
+Os resultados ainda não foram publicados.
+
+Outro evento que observei foi uma ocultação de 2003VS2 dia 07 de Novembro de 2014 no OPD.
+As variações do céu não permitiram uma boa observação de forma que a curva de luz estava muito variada com instantes de tempo bom próximos ao instante do evento.
+Um trabalho foi feito para identificar alguma ocultação mas nada pôde ser extraído.
+O trabalho ainda está em análise.
 

### 2015-06-14 qualify/Qualify-altair.tex
Aumentei a introdução e fiz o começo da astrometria

@@
 \newcommand{\arcmin}{\ensuremath{^{\prime}}}%                    % degree symbol:  °
+\newcommand{\arcsec}{\ensuremath{^{\prime\prime}}}%                    % degree symbol:  °
 \newcommand{\fs}{\mbox{\ensuremath{.\!\!^s}}}
@@
 \resumo{Resumo}
-%\indent \indent O estudo da estrutura e evolução do Sistema Solar tem muita importância atualmente, por exemplo, na compreensão dos mecanismos de formação dos planetas de outras estrelas (exoplanetas), trazendo em seus desdobramentos valiosas informações quanto a viabilidade de formação de ambientes que comportem vida.
+\indent \indent O estudo de objetos como TNOs, Centauros e Satélites Irregulares (remanescentes relativamente inalterados da formação do sistema solar) nos ajudam a compreender a formação e evolução do Sistema Solar.
+Poucas foram as sondas enviadas para explorar essas regiões e muito do conhecimento sobre seus constituintes foram adquiridos a partir de observações de solo.
+Atualmente, é aceito que TNOs e Centauros tenham sido formados nas partes mais internas do sistema solar, que satélites irregulares de planetas gigantes foram capturados pelos planetas.
+Resta saber quando e como.
+Para ajudar a responder essas perguntas é necessário caracterizar seus parâmetros físicos.
 
-%Praticamente cada satélite é um problema particular de Mecânica Celeste.
-Dessa forma, a observação de seus movimentos é muito útil por razões teóricas.
-A preparação e realização de missões espaciais para alguns planetas, e mesmo alguns satélites, como Titã de Saturno ou os Galileanos de Júpiter, requer observações frequentes e posições muito acuradas de todos os satélites do sistema planetário visitado.\\
-%\indent Os satélites irregulares de planetas gigantes, principalmente de Júpiter e de Saturno, são substancialmente menores do que os satélites regulares e possuem órbitas mais distantes, excêntricas, inclinadas e, na maioria dos casos, retrógradas.
-Explicar a existência dos satélites irregulares dos planetas gigantes é um estudo interessante em dinâmica orbital.\\
-%\indent É amplamente aceito que eles, devido à configuração orbital, foram capturados por seus planetas.
-Porém, ainda não são certos a origem e o método de captura desses objetos embora as teorias mais prováveis sejam a de arrasto gasoso durante a formação do Sistema Solar e a captura de três corpos.\\
-%\indent A maioria dos satélites formam famílias com características orbitais semelhantes sugerindo uma evolução dada por colisões.
-A compreensão desses mecanismos nos fará entender melhor o nosso próprio sistema solar e, possivelmente, de sistemas extra-solares.\\
-%\indent Com o pacote astrométrico PRAIA o tempo empregado na redução de grandes quantidades de imagens foi diminuído significativamente.
-Os objetivos científicos dos nossos programas observacionais agora tem sido atingidos em curtíssimo prazo, em consonância com a atual demanda astronômica e astrofísica de nossa área.\\
-%\indent Em uma base de dados com mais de 100 mil imagens obtidas entre 1992 e 2012 no telescópios de 1.2m do Observatoire de Haute Province, França, e nos telescópios \PE, \BC{} e Zeiss no Observatório do Pico dos Dias, mais de 4 mil contem satélites irregulares de Júpiter e Saturno.
-Reduzir essa grande quantidade de observações com precisão só foi possível com a utilização do PRAIA.
-Como é possível ver nos resultados obtidos, os offsets de posição refletem o erro natural da astrometria e o erro das efemérides dos satélites mais fracos, e sugerem que podemos efetivamente contribuir para que novas integrações numéricas forneçam posições mais precisas.
-Essas novas efemérides irão permitir que ocultações estelares por esses objetos possam ser preditas com maior precisão.
-A observações dessas ocultações permitirão um maior conhecimento das propriedades físicas dos satélites.
+Um dos métodos mais robustos para caracterizar tamanho, forma, densidade, atmosfera, anéis, etc é a ocultação estelar por objetos do Sistema Solar Externo.
+Este trabalho apresenta a análise de duas ocultações por Ceres, a predição de ocultações por satélites irregulares e o trabalho de ocultações por TNOs e Centauros desenvolvido pelo grupo em colaboração com pesquisadores internacionais.
+
+Parâmetros internos de Netuno e Tritão serão trabalhados em um doutorado sanduíche com o Dr.
+Valéry Lainey estudando os efeitos de maré do sistema.
+Uma vez que esses objetos estão muito próximos no céu, uma análise será feita para obter as posições astrométricas relativas Netuno-Tritão utilizando a técnica de coronografia digital para eliminar a contribuição da luz de Netuno na posição de Tritão.
+Usaremos nosso banco de dados de 23 anos de observações CCD do sistema no OPD/LNA, bem como dados inéditos de telemetria de sondas do espaço (Voyager) de posse do Dr.
+Lainey, dentro do projeto ESPaCE.
 
@@
 \vspace{1em}
-\noindent\textbf{Palavras-chave:} Astrometria, Ocultações
+\noindent\textbf{Palavras-chave:} astrometria, fotometria, ocultações estelares, marés, Ceres, satélites irregulares, objetos trans-Neptunianos, Netuno, Tritao
 
@@
 
-\indent \indent O estudo de objetos como TNOs, Centauros e Satélites Irregulares (remanescentes relativamente inalterados da formação do sistema solar) nos ajudam a compreender a formação e evolução do Sistema Solar.
-Atualmente, é aceito que TNOs e Centauros tenham sido formados nas partes mais internas do sistema solar.
+\indent \indent O estudo de objetos como TNOs, Centauros e Satélites Irregulares nos ajudam a compreender a formação e evolução do Sistema Solar.
+Nesta região distante do Sol, de baixas temperaturas, objetos principalmente de tamanho relativamente pequeno, e mais dispersos no espaço, provavelmente sofreram muito pouca diferenciação, seja por mecanismos internos, seja por choques com outros corpos, comparados a objetos formados mais próximos do Sol.
+
+Por serem corpos asteroidais localizados de forma dispersa além da órbita de Netuno, considera-se que os objetos trans-Neptunianos (TNOs) guardam estruturas e composições relativamente inalteradas em relação a sua época de formação, constituindo-se assim em corpos de prova de grande valor para o estudo da origem do sistema solar, ao menos para essa região exterior.
+
+Sendo que o primeiro TNO, além de Plutão, foi descoberto a menos de 20 anos \citep{Jewitt1993}, as propriedades básicas desta população ainda não estão inteiramente estabelecidas, como a distribuição de tamanhos, composição, estruturas internas.
+
+Centauros são uma população transiente entre TNOs e cometas da família de Júpiter, orbitando em uma região entre Júpiter e Netuno.
+Atualmente, cerca de 400 deles são conhecidos e é geralmente aceito que eles compartilham uma origem comum com os objetos do cinturão de Kuiper.
+Uma vez que Centauros são tipicamente mais brilhantes que TNOs por estarem mais próximos, eles servem como representantes nos quais são possíveis de inferir propriedades mais gerais sobre os objetos mais distantes \citep{Fernandez2002}.
+Dessa forma, a partir daqui, TNOs e Centauros serão tratados indistintamente como uma única população e referida unicamente como "TNOs".
+
+Atualmente, é aceito que TNOs tenham sido formados nas partes mais internas do sistema solar.
 Eles teriam então sido colocados em suas posições atuais devido a troca de momento angular entre os planetas e planetésimos quando da migração dos planetas gigantes.
-A evolução se deu de tal forma que a passagem dos planetesimais e planetas por zonas de ressonância de movimento médio redefiniu as órbitas desses corpos \citep{Tsiganis2005}.\\
-\indent Sabe-se que poucas sondas espaciais foram enviados para estudar o Sistema Solar Externo e que a quantidade de objetos estudados é muito pequena.
-Por isso, ainda hoje, as observações de solo tem se mostrado de grande importância.\\
-\indent Os sistemas de Júpiter e Saturno já foram pelas Voyager I e II, Galileu (Júpiter) e Cassini (Saturno), porém apenas Saturno continua sendo investigado por uma sonda.
-Porém as sondas observaram apenas os planetas, os anéis e satélites mais internos.
-Os satélites externos, que acredita-se ser oriundo de capturas ou foram pouco observados (como Phoebe) ou simplesmente não foram observados.\\
-\indent Já no caso de Urano e Netuno, nenhuma sonda exclusiva foi enviada, apenas as Voyagers I e II os visitaram, mas não permaneceram nos sistemas.
-A sonda New Horizons já está próxima a Plutão e obterá parâmetros físicos para Plutão e seus satélites (primeira visita por sonda a um objeto do cinturão de Kuiper), porém será uma passagem rápida e o acompanhamento da evolução do sistema, incluindo a evolução da atmosfera de Plutão se dará por observações de solo.\\
-\indent A quantidade de objetos descobertos além da órbita de Saturno tem aumentado muito desde o fim do século passado.
+A evolução se deu de tal forma que a passagem dos planetesimais e planetas por zonas de ressonância de movimento médio redefiniu as órbitas desses corpos \citep{Tsiganis2005}.
+
+Muitos dos objetos que pertenciam ao cinturão de Kuiper primordial acabaram sendo enviados pela interação com os planetas gigantes para as partes mais internas do sistema solar.
+Alguns podem ter sido capturados pelos planetas criando a população de satélites irregulares ou satélites externos \citep{Nesvorny2007}, troianos \citep{Morbidelli2005} ou até mesmo para o cinturão principal de asteroides \citep{McKinnon2012}.
+Estudar esses objetos é de grande importância para entender a evolução do sistema solar.
+
+Desses objetos, Tritão possui um interesse particular.
+Ele foi capturado por Netuno \citep{McKinnon2007} em uma órbita retrógrada e próxima ao planeta.
+Sua superfície é deformada com características tectônicas e possivelmente criovulcânicas \citep{Nimmo2015} e as propriedades físicas de Tritão já conhecidas mostram uma similaridade com as de Plutão.
+Além do mais, é um dos poucos objetos do sistema solar que se sabe possuir atmosfera.
+
+Sabe-se que poucas sondas espaciais foram enviados para estudar o sistema solar externo e que a quantidade de objetos estudados é muito pequena.
+Por isso, ainda hoje, as observações de solo tem se mostrado de grande importância.
+
+Os sistemas de Júpiter e Saturno já foram pelas Voyager I e II, Galileu (Júpiter) e Cassini (Saturno), porém apenas Saturno continua sendo investigado por uma sonda.
+Todavia, as sondas observaram apenas os planetas, os anéis e satélites mais internos.
+Os satélites externos ou foram pouco observados (como Phoebe) ou simplesmente não foram observados.
+
+Já no caso de Urano e Netuno, nenhuma sonda exclusiva foi enviada, apenas as Voyagers I e II os visitaram, mas não permaneceram nos sistemas.
+A sonda New Horizons já está próxima a Plutão e obterá parâmetros físicos para Plutão e seus satélites (primeira visita por sonda a um objeto do cinturão de Kuiper), porém será uma passagem rápida e o acompanhamento da evolução do sistema, incluindo a evolução da atmosfera de Plutão se dará por observações de solo.
+
+A quantidade de objetos descobertos além da órbita de Saturno tem aumentado muito desde o fim do século passado.
 Como são raras as oportunidades em que uma sonda se aproxima desses objetos, a obtenção de suas características físicas ficam a cargo de observações de solo ou de
-telescópios espaciais.\\
-\indent Um método que tem se mostrado eficiente para a obtenção desses parâmetros é o métodode ocultações estelares, que proporciona medidas tão precisas que são apenas superadas por medidas oriundas de sondas.
+telescópios espaciais.
 
-\chapter{Astrometria dos satélites irregulares dos planetas gigantes}
-\label{Cap: Irr-astrometria}
+Um método que tem se mostrado eficiente para a obtenção desses parâmetros é o método de ocultações estelares, que proporciona medidas tão precisas que são apenas superadas por medidas oriundas de sondas.
+De fato, a sonda \textit{Dawn} já está orbitando Ceres, um dos objetos deste estudo, e a sonda \textit{New Horizons} está prestes a chegar em Plutão.
+As observações feitas por essas sondas trazem a oportunidade perfeita para comparar os resultados com os obtidos pelas nossas ocultações e calibrar a técnica, de forma que ocultações feitas para outros objetos possam ser mais precisas.
 
-\indent \indent Os satélites irregulares dos planetas gigantes são menores que as regulares, possuindo órbitas mais excêntricas, inclinadas e distantes.
+O objetivo desse Doutorado é observar objetos do sistema solar exterior, fazer astrometria para melhorar suas efemérides de forma que ocultações estelares por esses objetos possam ser preditas com precisão, observar as ocultações e obter os parâmetros físicos.
+Além disso, será feito um estudo astrométrico do sistema Netuno-Tritão para obter parâmetros relacionados a efeitos de maré.
+
+\chapter{Astrometria}
+\label{Cap: Astrometria}
+
+\section{Procedimento de Redução}
+
+O tratamento astrométrico das imagens é feito utilizando a Plataforma para Redução Automática de Imagens Astronômias \citep[PRAIA][]{Assafin2011}.
+As medidas (x, y) são obtidas através de uma Gaussiana bidimensional simétrica ajustada até uma largura à altura (FWHM = seeing).
+Em uma FWHM, o perfil da imagem é bem descrito por um perfil gaussiano, livre das distorções de borda que prejudicam a determinação do centro.
+O PRAIA identifica automaticamente as estrelas de catálogo e determina ($\alpha$, $\delta$) com um modelo definido pelo usuário que relaciona as medidas (x, y) da imagem e as coordenadas projetadas no plano do céu (X, Y).
+
+Majoritariamente, o catálogo UCAC4 \citep{Zacharias2013} é utilizado como representante prático do ICRS.
+Como será exposto na seção \ref{Sec: TNOs-astrometria}, o catálogo WFI é utilizado para determinados TNOs até 2015.
+Esses catálogos serão substituídos pelo catálogo GAIA uma vez que seja lançado em meados de 2016.
+
+Para cada frame, usamos um modelo polinomial de seis constantes para relacionar as medidas (x, y) com as coordenadas (X, Y) do plano tangente.
+Para observações feitas no ESO, o procedimento astrométrico seguido é o mesmo descrito em \citep{Assafin2012}; as medidas (x, y) dos CCDs individuais são pré-corrigidos por um padrão de distorção de campo e todas as posições vindas de diferentes CCDs e mosaicos são combinadas usando o polinômio de grau três para produzir uma solução global para cada noite e campo observado, a as posições ($\alpha$, $\delta$) são então obtidas no sistema do catálogo utilizado.
+
+\section{Centauros e TNOs}
+\label{Sec: TNOs-astrometria}
+
+\indent \indent Observar uma ocultação exige um grande trabalho de predição e sua melhoria.
+Os TNOs estão muito distantes significando que mesmo sem grandes erros nas suas efemérides ou na posição da estrela o erro do local por onde a sombra irá passar pode ser de milhares de quilômetros.
+Preferencialmente, esse erro tem que ser menor que o tamanho angular do objeto no plano do céu.
+Por exemplo, Plutão tem um raio de $1190 \pm 5$ km \cite[submetido]{DiasOliveira2015} que a uma distância de aproximadamente 32 UA representam um diâmetro angular de 102 mas.
+Para a maioria dos objetos selecionados, menores que Plutão e/ou mais distantes, seus tamanhos angulares são menores que 50 mas.
+
+Da mesma forma, estando muito distantes, eles se movem relativamente devagar no céu ocultando relativamente poucas estrelas por ano.
+Porém, sendo menos brilhantes (Mag $V > 18$) é possível observar ocultações de estrelas mais fracas (estrelas até $R = 17$), onde a queda de brilho da estrela durante a ocultação poderá ser detectada e, mesmo com baixo sinal-ruído (ex: 5 a 10), o tamanho do corpo pode ser calculado com precisão de pelo menos 10 a 20 km, melhorando a medida em que a velocidade da sombra seja menor que 20 km/s (velocidade típica para um corpo em oposição a 30 UA).
+
+Para identificar candidatas a ocultações estelares foi realizada uma campanha observacional no ESO para observar os caminhos aparentes dos TNOs entre 2009 e 2015.
+As observações foram feitas com o telescópio de 2.2 m Max-Planck (código IAU 809, $70\degr44\arcmin1\farcs5$ L, $29\degr15\arcmin31\farcs8$ S, 2345.4 m)\footnote{Website: \url{www.eso.org/sci/facilities/lasilla/telescopes/national/2p2.html}} e o detector mosaico Wide Field Imager (WFI).
+Posições para estrelas até magnitude R=21, com nível de completeza em R=19, foram obtidas com precisões menores que 20 mas até R=20 \citep{Assafin2012}.
+Assim, estrelas que não estão no UCAC4, catálogo comumente utilizado como referência astrométrica, foram identificados bem como estrelas que estão mas não tem posições ou movimentos próprios com precisão suficiente.
+Tantos os catálogos de estrelas quanto as cadidatas a ocultação foram publicados \citep{Assafin2010, Assafin2012, Camargo2013}.
+Esses catálogos são utilizados como representante prático do ICRS no tratamento astrométrico e o conjunto é denominado catálogo WFI.
+
+Não somente as posições das estrelas precisam ser atualizadas, mas também as posições dos TNOs.
+Portanto os TNOs também foram observados tanto no ESO quanto no Observatório do Pico dos Dias (OPD, LNA/MCTI, Itajuba, MG., código IAU 874, $45^{\circ} ~34\arcmin ~57\arcsec$ W, $22^{\circ} ~32\arcmin ~04\arcsec$ S, 1864 m).
+À princípio as predições das ocultações eram atualizadas utilizando o método de offsets, porém, em colaboração com o Dr.
+Josselin Desmars, integrações numéricas das órbitas dos TNOs foram realizadas desenvolvendo-se uma ferramenta denominada NIMA\footnote{Numerical Integration of the Motion of an Asteroid} \citep[submetido]{Desmars2015}.
+
+Nesse trabalho, posições obtidas de ocultações já realizadas tem um peso muito maior que as restantes, diferentemente do que é feito pelo JPL.
+Da mesma forma, noites em que houveram um número de observações maior que as restantes terá um peso moderado já que se fosse utilizado um peso individual, o peso final de uma noite poderia estar superestimado levando a um erro na integração.
+Neste trabalho contribuí com observações de TNOs feitas no OPD.
+
+\section{Satélites Irregulares dos planetas gigantes}
+\label{Sec: Irr-astrometria}
+
+\indent \indent Os satélites irregulares dos planetas gigantes são menores que os regulares, possuindo órbitas mais excêntricas, inclinadas e distantes.
 Na maioria dos casos, essas órbitas são retrógradas.
-Devido suas configurações orbitais, é altamente aceito que estes objetos foram capturados nos estágios iniciais da formação do sistema solar \citep{Sheppard2003}.
+Devido a suas configurações orbitais, é amplamente aceito que estes objetos foram capturados nos estágios iniciais da formação do sistema solar \citep{Sheppard2003}.
 
-Por serem pequenos, eles são pouco brilhantes e só foram descobertos no último século\footnote{Website: http://ssd.jpl.nasa.gov/?sat\_discovery}.
+Por serem pequenos, eles são pouco brilhantes e só foram descobertos no último século\footnote{Website: \url{http://ssd.jpl.nasa.gov/?sat\_discovery}}.
 Dentre os satélite irregulares dos planetas gigantes, poucos são aqueles que possuem medidas de seus parâmetros físicos.
@@
 Por fim, Nereida foi observado em 1989 pela sonda Voyager 2 e seu tamanho foi obtido com um erro de 25 km \citep{Smith1989}.
-Para outros satélites irregulares de Júpiter, seus tamanhos foram estimados por \cite{Rettig2001} impondo um albedo de 4\% cam a justifica de que esse valor é um albedo nominalmente utilizado para objetos do Sistema Solar Externo.
+Para outros satélites irregulares de Júpiter, seus tamanhos foram estimados por \cite{Rettig2001} impondo um albedo de 4\% com a justifica de que esse valor é um albedo nominalmente utilizado para objetos do Sistema Solar Externo.
 
 Existem alguns mecanismos de captura de objetos por planetas gigantes propostos na literatura.
-Há o Arrasto Gasoso na nebulosa circumplanetária primordial \citep{Sheppard2005} onde o movimento objeto seria afetado gás e sua velocidade diminuiria até que fosse capturado pelo planeta.
+Há o Arrasto Gasoso na nebulosa circumplanetária primordial \citep{Sheppard2005} onde o movimento do objeto seria afetado pelo gás e sua velocidade diminuiria até que fosse capturado pelo planeta.
 Outro mecanismo se chama captura pull-down \citep{Sheppard2005}, onde a massa do planeta aumentaria enquanto o objeto estivesse temporariamente capturado.
@@
 Durante a instabilidade do sistema solar primordial, ocorreram encontros entre os planetas externos.
-Esses encontros planetários poderiam trocar energia e momento angular entre os planetas e os objetos próximos tornando possível a captura de um corpo irregular por um planeta gigante.
+Esses encontros planetários poderia haver troca de energia e momento angular entre os planetas e os objetos próximos tornando possível a captura de um corpo irregular por um planeta gigante.
 Nesse cenário, a taxa de sobrevivência de satélites anteriores ao LHB (Bombardeamento Pesado Tardio\footnote{"Late Heavy Bombardment", em inglês}) é muito pequena.
@@
 
-Nesse contexto, durante o mestrado, foi realizado junto ao grupo um trabalho de caráter astrométrico para a melhoria das efemérides dos satélites irregulares de Júpiter Saturno.
+Nesse contexto, durante o Mestrado, foi realizado junto ao grupo um trabalho de caráter astrométrico para a melhoria das efemérides dos satélites irregulares de Júpiter Saturno.
 Em colaboração com o Dr.
@@
 Reduzi também um banco de dados com mais de 100 mil imagens obtidas no Observatório do Pido dos Dias (OPD) entre 1992 e 2014.
-Já no doutorado, neste mesmo âmbito, reduzi 810 observações feitas no European Southern Observatory (ESO) em 24 noites utilizando o detector mosaico CCD Wide Field Imager (WFI).
+Já no Doutorado, neste mesmo âmbito, reduzi 810 observações feitas no European Southern Observatory (ESO) em 24 noites utilizando o detector mosaico CCD Wide Field Imager (WFI).
 
 Mais de 8000 posições foram identificadas como pertencentes a satélites irregulares.
-Porém, devido à grande gama de configurações (3 sítios, 5 telescópios, mais de 10 câmeras e mais de 10 filtros) e às condições observacionais de algumas noites, 6523 posições foram selecionadas como boas posições, ou seja, possuem 3 ou mais estrelas do catálogo de referência (UCAC4), estão dentro de 2$\sigma$ da dispersão das posições da noite e a dispersão da noite à qual pertence não é maior que 2$\sigma$ da média das dispersões das noites.
+Porém, devido à grande gama de configurações (3 sítios, 5 telescópios, mais de 10 câmeras e mais de 10 filtros) e às condições observacionais de algumas noites, 6523 posições foram selecionadas como boas posições, ou seja, possuem cinco ou mais estrelas do catálogo de referência (UCAC4), estão dentro de 2$\sigma$ da dispersão das posições da noite e a dispersão da noite à qual pertence não é maior que 2$\sigma$ da média das dispersões do conjunto total de noites.
 Essa estatística é feita satélite por satélite.
@@
 \caption{\label{Tab: comparison-horizons} Comparação entre o número de posições obtidas em \cite{GomesJunior2015-Irregular} e o número utilizado nas integrações numéricas das órbitas pelo JPL como publicado por  \cite{Jacobson2012}.
-Os satélites estão separados por planeta (linha reta) e por família orbital (linha tracejada)}
+Os satélites estão separados por planeta (linha reta) e por família orbital (linha tracejada).}
 \begin{centering}
@@
 \caption{Offsets médios das efemérides e dispersões (barra de erro 1$\sigma$) nas coordenadas de Carme tomadas noite a noite por anomalia verdadeira para cada telescópio.
-O quadrado vermelho são as observações com o telescópio do OPD Perkin-Elmer, o círculo azul para o Boller \& Chivens também do OPD, o triângulo pra cima preto para o OHP e a estrela verde para o ESO.}
-\label{Fig: carme_anom}
+O quadrado vermelho são as observações com o telescópio do OPD Perkin-Elmer, o círculo azul para o Boller \& Chivens também do OPD, o triângulo pra cima preto para o OHP e a estrela verde para o ESO.
+\label{Fig: carme_anom}}
 \end{centering}
@@
 
-Um método que tem se mostrado eficiente para a obtenção desses parâmetros é o métodode ocultações estelares, que proporciona medidas tão precisas que são apenas superadas por medidas oriundas de sondas.
-Uma vez que suas efemérides não são muito precisas, predizer e observar ocultações estelares é muito difícil e nenhuma observação de um evento desses para um satélite irregular é encontrado na literatura.
+Um método que tem se mostrado eficiente para a obtenção dos parâmetros físicos (tamanho, forma, albedo, etc) desses objetos é o método de ocultações estelares, que proporciona medidas tão precisas que são apenas superadas por medidas oriundas de sondas.
+Uma vez que suas efemérides não são muito precisas, predizer e observar ocultações estelares é muito difícil e nenhuma observação de um evento desses para um satélite irregular ainda foi relatado na literatura.
 Um projeto nesse sentido está sendo encaminhado e será discutido na seção \ref{Sec: Irregulares}.
 
-\chapter{Astrometria de Netuno e Tritão}
-\label{Cap: Netuno}
+\section{Netuno e Tritão}
+\label{Sec: Netuno-astrometria}
 
 \chapter{Ocultações Estelares}
-\label{Cap: observacoes}
+\label{Cap: ocultacoes}
+
+\section{Noções Gerais}
 
@@
 A Fig.
-\ref{Fig: occ-scheme}.1 mostra uma curva de luz esquemática de uma ocultação estelar central por um planeta com atmosfera e anéis.
+\ref{Fig: occ-scheme} mostra uma curva de luz esquemática de uma ocultação estelar central por um planeta com atmosfera e anéis.
 Ao ocultar a estrela, os anéis passarão primeiro, a densidade do anel, tamanho e tempo de exposição determinará a queda de brilho da estrela.
@@
 A atmosfera, no caso, agiria como uma lente como mostrado na Fig.
-\ref{Fig: occ-scheme}.1.
+\ref{Fig: occ-scheme}.
 
 Quando o evento é observado por diversos telescópios espalhados, cada observador verá a ocultação em latitudes diferentes.
-Essas diversas curvas de luz permitirá reconstruir a forma do objeto possivelmente identificando estruturas superficiais como montanhas e crateras.
+Essas diversas curvas de luz permitirá reconstruir a forma do objeto, possivelmente identificando estruturas superficiais como montanhas e crateras.
 Como a estrela está muito distante, podendo ser considerada quase pontual, a sombra do objeto durante a ocultação é muito próximo ao seu tamanho real.
@@
 The knowledge of these parameters would in turn bring valuable information for studying the capture mechanisms and origin of the irregular satellites.
+
+\section{Centauros e TNOs}
+\label{Sec: TNO-occ}
+
+\indent \indent Como dito no Cap.
+\ref{Cap: intro}, certos objetos do sistema solar exterior podem ser considerados fósseis do sistema solar.
+Dentre esses, os TNOs, por estarem mais distantes, são os que guardam as informações mais antigas.
+Com o objetivo de obter os parâmetro físicos desses objetos, para melhor caracterizar as propriedades básicas dessas populações, uma grande campanha observacional é realizada para se observar ocultações estelares por TNOs e Centauros.
+Essa campanha é liderada por pesquisadores brasileiros, franceses e espanhóis e tem colaboração com observadores espalhados por todo o mundo.
+
+Vários objetos já foram observados pelo grupo, sendo que o Sistema Plutão-Caronte conta com várias ocultações observadas além de um trabalho astrométrico minucioso baseado em 19 anos de observações no OPD, mais observações no ESO \citep{Rossi2014}.
+As observações bem sucedidas de ocultações estelares resultantes desse trabalho de predição já geraram diversas publicações, inclusive na revista Nature, por exemplo \cite{Sicardy2011} e \cite{BragaRibas2014} nas quais contribuí em caráter observacional.
+A ocultação de Plutão de 04 de Maio de 2013, a qual observei, gerou duas publicações \citep{Olkin2015, DiasOliveira2015}.
+
+\cite{Sicardy2011} identificaram, a partir de uma ocultação de 06 de Novembro de 2010, que Eris é um planeta anão menor do que se esperava.
+Devido a seu brilho e distância acreditava-se que ele fosse muito maior, mas na verdade ele tem um albedo geométrico no visível de $p_v=0.96^{+0.09}_{-0.04}$ e seu tamanho é de $1163 \pm 6$ km, muito próximo ao tamanho de Plutão.
+O alto albedo de Eris pode ser relacionado à presença de uma atmosfera colapsada devido ao fato dele estar se aproximando de seu afélio.
+
+\cite{BragaRibas2014} observaram uma ocultação do Centauro Chariklo que ocorreu dia 03 de Junho de 2013.
+O ocultação revelou que o objeto possui um sistema de anéis nunca visto antes em um objeto tão pequeno.
+Até o momento, conhecia-se apenas anéis em torno dos planetas gigantes.
+Essa descoberta levanta muitas questões sobre formação dos anéis, estabilidade e tempo de vida e nos leva a crer que pode haver mais objetos com anéis no sistema solar, ao menos nessa região.
+
+\cite{Olkin2015} determinaram a partir de observações de ocultações por Plutão que a atmosfera do objeto não colapsa durante os 248 anos de sua órbita, como acontece com Eris.
+Modelos anteriores previam um colapso de sua atmosfera devido a Plutão receber três vezes menos luz do Sol em afélio que em seu periélio.
+
+\cite{DiasOliveira2015}, também a partir de ocultações de Plutão, utilizaram modelos para obter os perfis de temperatura, pressão e densidade da atmosfera do objeto.
+Os modelos ajustam perfeitamente os perfis de temperatura para as duas ocultações utilizadas assumindo uma atmosfera esfericamente simétrica.
+Foram usadas as curvas de luz de melhor resolução até hoje obtidas para Plutão, sendo a melhor obtida em  Paranal em 18 de Julho de 2012, usando a câmera NAOS-CONICA (NACO) ligada ao 8.2-m “Yepun” Very Large Telescope (VLT) do ESO a uma taxa de 5 frames por segundo na banda H.
+Por fim, foi obtido que a pressão da atmosfera aumentou cerca de 6\% entre 2012 e 2013 significando que a atmosfera de Plutão ainda está se expandindo e confirmando o trabalho de \cite{Olkin2015}.
+
+No dia 29 de Junho de 2015 haverá uma ocultação de Plutão de máxima importância.
+Será a última antes da chegada da sonda \textit{New Horizons} no sistema e poderá ser observada na Austrália.
+Observada a ocultação, os modelos de perfil de atmosfera para Plutão poderão ser calibrados com os dados da \textit{New Horizons} para serem utilizados em futuras ocultações e estudar a evolução da atmosfera do planeta anão.
+
+Em 29 de Abril de 2014, uma ocultação de Chariklo foi observada na África.
+Eu trabalhei com a redução de três curvas de luz (Gifberb, SAAO e Springbok).
+Um fato interessante dessa ocultação foi que a estrela era uma binária não catalogada.
+Como a segunda estrela era bem mais fraca, acreditava-se que poderia ser algum material próximo à Chariklo.
+
+Essa ocultação foi importante para obter com total certeza a direção do pólo de Chariklo.
+Além disso, as câmeras utilizadas eram de alta qualidade possibilitando obter que pelo menos um dos anéis é opaco.
+Uma das cordas, fortuitamente, passou contígua ao anel observando apenas o anel externo.
+Os resultados ainda não foram publicados.
+
+Outro evento que observei foi uma ocultação de 2003VS2 dia 07 de Novembro de 2014 no OPD.
+As variações do céu não permitiram uma boa observação de forma que a curva de luz estava muito ruidosa, mas boa o suficiente próximo ao instante do evento.
+Um trabalho foi feito para identificar alguma ocultação mas nada pôde ser extraído.
+Nesse evento foram obtidas 3 cordas positivas e 7 negativas.
+A redução preliminar revelou que 2003VS2 é possui um corpo levemente alongado com aproximadamente 600 km de diâmetro.
+O trabalho ainda está em análise.
 
@@
   \hline
-     Solution & 2010/Nominal & \textbf{2010/Pólo fixo} & 2013/Nominal & 2013/Pólo fixo \\
+     Solução & 2010/Nominal & \textbf{2010/Pólo fixo} & 2013/Nominal & 2013/Pólo fixo \\
 \hline
@@
 
-\indent \indent Com o objetivo de obter parâmetros físicos para os satélites irregulares com maior precisão foram feitas predições de ocultações estelares para os 7 maiores satélites de Júpiter, o satélite Phoebe de Saturno e Nereida de Netuno.
-Esses objetos são pequenos (ver Tabela \ref{Tab: comparison-horizons}), se comparados aos TNOs e Centauros estudados na seção \ref{Sec: TNO-occ}, sendo que o menor dos satélites da amostra, Ananke, possui aproximadamente 28 km.
+\indent \indent Com o objetivo de obter parâmetros físicos para os satélites irregulares com maior precisão foram feitas predições de ocultações estelares para os 7 maiores satélites de Júpiter (Himalia, Elara, Pasiphae, Sinope, Lysithea, Carme, Ananke), o satélite Phoebe de Saturno e Nereida de Netuno.
+Esses objetos são pequenos (ver Tabela \ref{Tab: comparison-horizons}), se comparados aos TNOs estudados na seção \ref{Sec: TNO-occ}, sendo que o menor dos satélites da amostra, Ananke, possui um diâmetro estimado de 28 km.
 
-As predições foram feitas utilizando o catálogo UCAC4 e efemérides do JPL.
+As predições foram feitas utilizando as posições de estrelas dadas no cataálogo UCAC4.
+Tomou-se como referência as efemérides do JPL, corrigidas com as posições de satélites obtidas, de acordo com o descrito a seguir.
 Um total de 588 eventos foram obtidos entre 01 de Janeiro de 2015 e 31 de Dezembro de 2017 para os 9 objetos sendo que a maioria será descartada por passar em regiões de oceano ou onde não há observadores disponíveis.
@@
 Como esses objetos são muito pequenos, em sua maioria as ocultações durarão poucos segundos, portanto apenas eventos com estrelas brilhantes serão selecionados e se houver câmeras de integração rápida disponíveis.
-Por outro lado, os satélites de Júpiter estão muito mais pertos que os TNOs e como o erro astrométrico é uma medida angular, consequentemente, o erro da sombra da ocultação projetada na Terra será muito menor que aquelas medidas para TNOs.
-Assim, a área a ser coberta para que uma ocultação por satélite irregular de Júpiter seja detectada será da ordem de poucas centenas de quilômetros.
+Por outro lado, os satélites de Júpiter estão muito mais perto que os TNOs e como o erro astrométrico é uma medida angular, consequentemente, o erro da sombra da ocultação projetada na Terra será muito menor que para TNOs.
+Assim, as latituds a serem cobertas para que uma ocultação por satélite irregular de Júpiter seja detectada corresponde a poucas centenas de quilômetros.
 
@@
 Erros em ascensão reta interferem majoritariamente no instante em que a ocultação ocorrerá.
-Já erros em declinação principalmente na latitude geográfica em que a sombra passará.
+Já erros em declinação, principalmente na latitude geográfica em que a sombra passará.
 
 Outra fonte de erro na predição é a posição do satélite.
-Como mostradas no capítulo \ref{Cap: Irr-astrometria}, as órbitas dos satélites irregulares não são muito precisas havendo erros sistemáticos em seus elementos orbitais.
+Como mostradas no capítulo \ref{Sec: Irr-astrometria}, as órbitas dos satélites irregulares não são muito precisas havendo erros sistemáticos em seus elementos orbitais.
 Infelizmente o JPL não provê suas efemérides com os erros estimados das posições.
@@
 Josselin Desmars.
-Um trabalho semelhante foi abordado para ocultação de Centauros e TNOs (ver seção \ref{Sec: TNO-occ}).
+Um trabalho semelhante foi abordado para ocultação de TNOs (ver seção \ref{Sec: TNO-occ}).
 
-Por fim, a terceira alternativa consiste em se fazer um ajuste dos offset obtidos em \cite{GomesJunior2015-Irregular} em função da anomalia verdadeira, tempo e/ou anomalia média.
+Por fim, a terceira alternativa consiste em se fazer um ajuste dos offsets de efeméride obtidos em \cite{GomesJunior2015-Irregular} em função da anomalia verdadeira, tempo e/ou anomalia média.
 Essa escolha é suportada pelas dispersões dos offsets em função da anomalia verdadeira (por exemplo, Fig.
@@
 A sombra cobrirá uma região muito restrita na Terra devido aos satélites irregulares serem pequenos.
-Portanto, antes de iniciar uma grande campanha observacional, testaremos algumas ocultações, principalmente dos objetos maiores para confirmar a predição.
+Portanto, antes de iniciar uma grande campanha observacional, testamos algumas predições de ocultações para objetos maiores, para avaliar a qualidade da predição.
 
-Esse teste consiste em se observar o objeto ocultante e a estrela a ser ocultada próximo ao evento, quando os dois objetos estiverem presentes no mesmo campo, preferencialmente próximo ao instante do evento quando os objetos estarão o mais próximo possível.
+Esse teste consiste em se observar o objeto ocultante e a estrela a ser ocultada próximo ao evento, quando os dois objetos estiverem presentes no mesmo campo, preferencialmente próximo ao instante do evento quando os objetos estarão próximos pum do outro.
 Assim as posições relativas entre os dois objetos terá mínima influência dos erros das estrelas de catálogo utilizadas na redução e das possíveis distorções de campo.
 
-Até o momento, apenas dois testes de ocultações foram feitas, uma de Himalia que ocorreu dia 03 de Março de 2015 e a segunda de Elara que ocorreu dia 30 de Março de 2015.
-Para cada evento quatro mapas foram gerados: o primeiro com as posições nominais da estrela e do satélite para o instante previsto; a segunda com o offset calculado como descrito anteriomente; o terceiro com os offsets da estrela e do satélite a partir de observações feitas alguns dias antes da ocultação quando os dois objetos estavam separados (campos diferentes); e o quarto a partir de observações feitas com menos de 24 horas de diferença da data prevista com a estrela e o satélite próximos no mesmo campo.
+Até o momento, dois testes de ocultações foram feitos, um de Himalia que ocorreu dia 03 de Março de 2015 e o segundo de Elara que ocorreu dia 30 de Março de 2015.
+Para cada evento quatro mapas foram gerados: o primeiro com as posições nominais da estrela e do satélite para o instante previsto; o segundo com o offset calculado como descrito anteriomente; o terceiro com os offsets da estrela e do satélite a partir de observações feitas alguns dias antes da ocultação quando os dois objetos estavam separados (campos diferentes); e o quarto a partir de observações feitas com menos de 24 horas de diferença da data prevista com a estrela e o satélite próximos no mesmo campo.
 
@@
 \ref{Fig: occ-Himalia} mostra os quatro mapas para o teste da ocultação de Himalia de 03 de Março de 2015.
-O mapa \ref{Fig: occ-Himalia-off22fev} foi feito a partir de posições obtidas em 22 de fevereiro observadas no telescópio Zeiss do Observatório do Pico dos Dias (OPD).
+O mapa \ref{Fig: occ-Himalia-off22fev} foi feito a partir de posições obtidas em 22 de Fevereiro observadas no telescópio Zeiss do Observatório do Pico dos Dias (OPD).
 Nesse dia, Himalia e a estrela a ser ocultada foram observados em campos separados por ainda estarem distantes.
-Na noite do evento, 03 de março, os objetos foram observados com o telescópio Perkin-Elmer do OPD pouco mais de uma hora após os instante previsto para o evento.
+Na noite do evento, 03 de Março, os objetos foram observados com o telescópio Perkin-Elmer do OPD pouco mais de uma hora após os instante previsto para o evento.
 Satélite e estrela estavam separados por aproximadamente 16 arcsec, portanto muito próximos no campo.
@@
 
-\begin{figure}
+\begin{figure}[h!]
 \begin{centering}
@@
 \caption{Predições para Himalia.
-O ponto grande em vermelho mostra o local de máxima aproximação da sombra na Terra, os pontos vermelhos menores são os centros da ocultação separados por um minuto, as linhas retas são os limites das sombras dado o tamanho do objeto e as linhas tracejadas demonstram uma diferença de 40 mas do centro da sombra.
+O ponto grande em vermelho mostra o ponto de máxima aproximação geocêntrica da sombra na Terra, os pontos vermelhos menores são os centros da ocultação separados por um minuto, as linhas retas são os limites das sombras dado o tamanho estimado do objeto e as linhas tracejadas equivalem a uma diferença de 40 mas do centro da sombra representando o erro estimado da predição.
 (a) é o mapa utilizando apenas as posições nominais da estrela e do satélite.
-(b) mostra a sombra dado um offset estimado pra posição de Himalia dado sua anomalia verdadeira segundo um ajuste feito em cima dos offsets publicados em \cite{GomesJunior2015-Irregular}.
-Em (c) são aplicados offsets às posições da estrela e do satélite dado observações feitas em 22 de fevereiro no telescópio Zeiss.
-Em (d), temos o mesmo que para (c) porém com offsets obtidos de observações feitas em 03 de Março no telescópio Perkin-Elmer quando os objetos estavam próximos no campo}
+(b) mostra a sombra dado um offset estimado para a posição de Himalia dada sua anomalia verdadeira segundo um ajuste feito em cima dos offsets de efeméride encntrados em \cite{GomesJunior2015-Irregular}.
+Em (c) são aplicados offsets às posições da estrela e do satélite a partir de observações feitas em 22 de Fevereiro no telescópio Zeiss.
+Em (d), temos o mesmo que para (c) porém com offsets obtidos de observações feitas em 03 de Março no telescópio Perkin-Elmer quando os objetos estavam próximos no campo.}
 \label{Fig: occ-Himalia}
@@
 
-Para esse evento é possível ver que a sombra não varia muito dentre os quatro mapas sugerindo que para Himalia uma probabilidade maior de observar um evento.
+Para esse evento é possível ver que a sombra não varia muito dentre os quatro mapas sugerindo que para Himalia há uma probabilidade maior de observar um evento.
 De fato, a maior diferença entre as sombras dos quatros mapas é de $25 s$ e $130 km$ na direção perpendicular às sombras.
@@
 Na Fig.
-\ref{Fig: occ-Elara} temos os mapas para a ocultação de Elara do dia 30 de Março.
-As observações de teste foram tomadas dias 25 de março e 02 de abril no Zeiss.
+\ref{Fig: occ-Elara} temos os mapas para a ocultação de Elara do dia 30 de Março de 2015.
+As observações de teste foram tomadas nos dias 25 de Março e 02 de Abril de 2015 no Zeiss.
 Na noite do dia 02 eles ainda puderam ser observadas no mesmo campo.
 Devido à Elara ser bem mais fraco, as dispersões das posições do satélite nas duas noites acabaram sendo maiores que para Himalia.
-Ainda assim, as diferenças entre os mapas obtidos foram pequenos.
+Ainda assim, as diferenças entre os mapas obtidos foram relativamente pequenas.
 A maior diferença entre elas é de $48 s$ e $302 km$.
 
-\begin{figure}
+\begin{figure}[h]
 \begin{centering}
@@
 \subfigure[Obs. dia 02 de Abril (mesmo campo)]{\includegraphics[scale=0.16]{figuras/Elara4_2015-03-30T01:45:30.png}  \label{Fig: occ-Elara-off02abr}}
-\caption{Predições para Elara. Mesmo que a Fig. \ref{Fig: occ-Himalia}, mas para Elara. O mapa (c) feito a partir de observações tomadas dia 25 de março e (d) a partir de observações de 02 de abril.}
+\caption{Predições para Elara.
+Mesmo que a Fig.
+\ref{Fig: occ-Himalia}, mas para Elara.
+O mapa (c) feito a partir de observações tomadas no dia 25 de Março e (d) a partir de observações de 02 de Abril de 2015.}
 \label{Fig: occ-Elara}
@@
 \ref{Fig: occ-Himalia-itajuba}).
-O evento ocorrerá muito baixo do céu (~22º de altura) e estará próximo ao amanhecer, porém se observar será a primera ocultação de um satélite irregular.
+O evento ocorrerá muito baixo do céu (~22º de altura) e estará próximo ao amanhecer, porém se formos bem sucedidos será a primeira ocultação estelar por um satélite irregular a ser observada.
 
-\begin{figure}
+\begin{figure}[h]
 \includegraphics[scale=0.35]{figuras/Himalia_2015-10-23T07:54:47.png}
@@
 
-\section{Centauros e TNOs}
-\label{Sec: TNO-occ}
+\section{Tritão}
+\label{Sec: occ-Tritao}
 
-\indent \indent Como dito no Cap.
-\ref{Cap: intro}, os objetos do sistema solar exterior podem ser considerados fósseis do sistema solar.
-Dentre esses, os TNOs, por estarem mais distantes, são os que guardam as informações mais antigas.
-Ainda é incerta a distribuição de tamanhos dos TNOs e pouco se conhece sobre suas composições, estruturas internas, presença de atmosfera, densidade, etc.
-
-Com o objetivo de obter os parâmetro físicos desses objetos uma grande campanha observacional é realizada para se observar ocultações estelares por TNOs e Centauros (a partir daqui, "TNOs" se referem a TNOs e Centauros).
-Essa campanha é liderada por pesquisadores brasileiros, franceses e espanhóis e tem colaboração com observadores espalhados por todo o mundo.
-
-Assim como para os satélites irregulares, observar uma ocultação exige um grande trabalho de predição e sua melhoria.
-Os TNOs estão muito distantes significando que erros pequenos em sua posição ou na posição da estrela podem significar um erro do local onde a sombra irá passar de muitos quilômetros.
-Preferencialmente, esse erro tem que ser menor que o tamanho angular do objeto no plano do céu.
-Por exemplo, Plutão tem um raio de $1190 \pm 5$ km \cite[em preparação]{DiasOliveira2015} que a uma distância de aproximadamente 32 UA representam um diâmetro angular de 102 mas.
-Para a maioria dos objetos selecionados, menores que Plutão e/ou mais distantes, seus tamanhos angulares são manores que 50 mas.
-
-Da mesma forma, estando muito distantes, eles se movem pouco no céu ocultando poucas esrelas, porém sendo menos brilhantes (Mag V > 18) é possível observar ocultações de estrelas mais fracas, onde a queda de brilho da estrela durante a ocultação poderá ser detectada com baixo sinal-ruído.
-
-Para identificar candidatas a ocultações estelares foi realizado uma campanha observacional no ESO para observar os caminhos aparentes dos TNOs entre 2009 e 2015.
-As observações foram feitas com o telescópio de 2.2 m e o detector mosaico WFI.
-Posições para estrelas até magnitude R=21, com nível de completeza em R=19, foram obtidas com precisões menores que 20 mas até R=20 \citep{Assafin2012}.
-Assim, estrelas que não estão no UCAC4, catálogo comumente utilizado como referência astrométrica, foram identificados bem como estrelas que estão mas não tem posições ou movimentos próprios com precisão suficiente.
-Tantos os catálogos de estrelas quanto as cadidatas a ocultação foram publicados em periódicos \citep{Assafin2010, Assafin2012, Camargo2013}.
-
-Não somente as posições das estrelas precisam ser atualizadas, mas também as posições dos TNOs.
-Portanto os TNOs também foram observados tanto no ESO quanto No Observatório do Pico dos Dias.
-À princípio as predições das ocultações eram atualizadas utilizando o método de offsets, porém, em colaboração, integrações numéricas das órbitas dos TNOs foram realizadas desenvolvendo-se uma ferramenta denominada NIMA\footnote{Numerical Integration of the Motion of an Asteroid} \citep[submetido]{Desmars2015}.
-
-Nesse trabalho, posições obtidas de ocultações já realizadas tem um peso muito maior que as restantes, diferentemente do que é feito pelo JPL.
-Da mesma forma, noites em que houveram um número de observações maior que as restantes terá um peso moderado já que se fosse utilizado um peso individual, o peso final de uma noite poderia estar superestimado levando a um erro na integração.
-Neste trabalho contribuí com observações de TNOs feitas no OPD.
-
-Vários objetos já foram observados pelo grupo, sendo que o Sistema Plutão-Caronte conta com várias ocultações observadas.
-Essa trabalho já gerou diversas publicações, incluindo na revista Nature, por exemplo \cite{Sicardy2011} e \cite{BragaRibas2014} nas quais contribuí em caráter observacional.
-A ocultação de Plutão de 04 de Maio de 2013, a qual observei, gerou duas publicações \citep{Olkin2015, DiasOliveira2015}.
-
-\cite{Sicardy2011} identificaram, a partir de uma ocultação de 06 de Novembro de 2010, que Eris é um planeta anão menor do que se esperava.
-Devido a seu brilho e distância acreditava-se que ele fosse muito maior, mas na verdade ele tem um albedo geométrico no visível de $p_v=0.96^{+0.09}_{-0.04}$ e que seu tamanho é de $1163 \pm 6$ km, muito próximo ao tamanho de Plutão.
-O alto albedo de Eris pode ser relacionada à presença de uma atmosfera colapsada devido ao fato dele estar se aproximando de seu afélio.
-
-\cite{BragaRibas2014} observaram uma ocultação do Centauro Chariklo que ocorreu dia 03 de Junho de 2013.
-O ocultação revelou que o objeto possui um sistema de anéis nunca visto antes em um objeto tão pequeno.
-Até o momento, conhecia-se apenas anéis em torno dos planetas gigantes.
-Essa descoberta levanta muitas questão sobre formação dos anéis, estabilididade ou tempo de vida e nos leva a crer que pode haver mais objetos com anéis no sistema solar.
-
-\cite{Olkin2015} determinaram a partir de observações de ocultações por Plutão que a atmosfera do objeto não colapsa durante os 248 anos de sua órbita, como acontece com Eris.
-Modelos anteriores previam um colapso de sua atmosfera devido a Plutão receber três vezes menos luz do Sol em afélio que em seu periélio.
-
-\cite{DiasOliveira2015}, também a partir de ocultações de Plutão, utilizaram modelos para obter os perfis de temperatura, pressão e densidade da atmosfera do objeto.
-Os modelos ajustam perfeitamente os perfis de temperatura para as duas ocultações utilizadas assumindo uma atmosfera esfericamente simétrica.
-Foi obtido que a pressão da atmosfera aumentou cerca de 6\% entre 2012 e 2013 significando que a atmosfera de Plutão ainda está se expandindo e confirmando o trabalho de \cite{Olkin2015}.
-
-No dia 29 de Junho de 2015 haverá uma ocultação de Plutão de máxima importância.
-Será a última antes da chegada da sonda \textit{New Horizons} no sistema e poderá ser observada na Austrália.
-Observada a ocultação, os modelos de perfil de atmosfera para Plutão poderão ser calibrados com os dados da \textit{New Horizons} para serem utilizadas em futuras ocultações e estudar a evolução da atmosfera do planeta anão.
-
-Em 29 de Abril de 2014, uma ocultação de Chariklo foi observada na África.
-Eu trabalhei com a redução de três curvas de luz (Gifberb, SAAO e Springbok).
-Um fato interessante dessa ocultação foi que a estrela era uma binária não catalogada.
-Como a segunda estrela era bem mais fraca, acreditava-se que poderia ser algum material próximo à Chariklo.
-
-Essa ocultação foi importante para obter com total certeza a direção do pólo de Chariklo.
-Além disso, as câmeras utilizadas eram de alta qualidade possibilitando obter que pelo menos um dos anéis é opaco.
-Uma das cordas, fortuitamente, passou contígua ao anel observando apenas o anel externo.
-Os resultados ainda não foram publicados.
-
-Outro evento que observei foi uma ocultação de 2003VS2 dia 07 de Novembro de 2014 no OPD.
-As variações do céu não permitiram uma boa observação de forma que a curva de luz estava muito variada com instantes de tempo bom próximos ao instante do evento.
-Um trabalho foi feito para identificar alguma ocultação mas nada pôde ser extraído.
-O trabalho ainda está em análise.
-
-\chapter{Cronograma das Atividades}
+\chapter{Perspectivas}
 \label{Cap: cronograma}

### 2015-06-15 qualify/Qualify-altair.tex
Falta tritao

@@
 \resumo{Resumo}
-\indent \indent O estudo de objetos como TNOs, Centauros e Satélites Irregulares (remanescentes relativamente inalterados da formação do sistema solar) nos ajudam a compreender a formação e evolução do Sistema Solar.
+\indent \indent O estudo de objetos como TNOs, Centauros e Satélites Irregulares (remanescentes relativamente inalterados da formação do Sistema Solar) nos ajudam a compreender a formação e evolução do Sistema Solar.
 Poucas foram as sondas enviadas para explorar essas regiões e muito do conhecimento sobre seus constituintes foram adquiridos a partir de observações de solo.
-Atualmente, é aceito que TNOs e Centauros tenham sido formados nas partes mais internas do sistema solar, que satélites irregulares de planetas gigantes foram capturados pelos planetas.
+Atualmente, é aceito que TNOs e Centauros tenham sido formados nas partes mais internas do Sistema Solar, que satélites irregulares de planetas gigantes foram capturados pelos planetas.
 Resta saber quando e como.
@@
 
-\indent \indent O estudo de objetos como TNOs, Centauros e Satélites Irregulares nos ajudam a compreender a formação e evolução do Sistema Solar.
+\indent \indent O estudo de objetos como TNOs, Centauros e Satélites Irregulares nos ajudam a compreender a formação e evolução do Sistema Solar Exterior.
 Nesta região distante do Sol, de baixas temperaturas, objetos principalmente de tamanho relativamente pequeno, e mais dispersos no espaço, provavelmente sofreram muito pouca diferenciação, seja por mecanismos internos, seja por choques com outros corpos, comparados a objetos formados mais próximos do Sol.
 
-Por serem corpos asteroidais localizados de forma dispersa além da órbita de Netuno, considera-se que os objetos trans-Neptunianos (TNOs) guardam estruturas e composições relativamente inalteradas em relação a sua época de formação, constituindo-se assim em corpos de prova de grande valor para o estudo da origem do sistema solar, ao menos para essa região exterior.
+Por serem corpos asteroidais localizados de forma dispersa além da órbita de Netuno, considera-se que os objetos trans-Neptunianos (TNOs) guardam estruturas e composições relativamente inalteradas em relação a sua época de formação, constituindo-se assim em corpos de prova de grande valor para o estudo da origem do Sistema Solar, ao menos para essa região exterior.
 
-Sendo que o primeiro TNO, além de Plutão, foi descoberto a menos de 20 anos \citep{Jewitt1993}, as propriedades básicas desta população ainda não estão inteiramente estabelecidas, como a distribuição de tamanhos, composição, estruturas internas.
+Tirando Plutão e Tritão, o primeiro TNO foi descoberto a pouco mais de 20 anos \citep{Jewitt1993}.
+Por isso, as propriedades básicas desta população ainda não estão inteiramente estabelecidas, como a distribuição de tamanhos, composição, estruturas internas.
 
@@
 Atualmente, cerca de 400 deles são conhecidos e é geralmente aceito que eles compartilham uma origem comum com os objetos do cinturão de Kuiper.
-Uma vez que Centauros são tipicamente mais brilhantes que TNOs por estarem mais próximos, eles servem como representantes nos quais são possíveis de inferir propriedades mais gerais sobre os objetos mais distantes \citep{Fernandez2002}.
+Uma vez que Centauros são tipicamente mais brilhantes que TNOs por estarem mais próximos, eles servem como representantes a partir dos quais é possível inferir propriedades mais gerais sobre os objetos mais distantes \citep{Fernandez2002}.
 Dessa forma, a partir daqui, TNOs e Centauros serão tratados indistintamente como uma única população e referida unicamente como "TNOs".
 
-Atualmente, é aceito que TNOs tenham sido formados nas partes mais internas do sistema solar.
+Atualmente, é aceito que TNOs tenham sido formados nas partes mais internas do Sistema Solar.
 Eles teriam então sido colocados em suas posições atuais devido a troca de momento angular entre os planetas e planetésimos quando da migração dos planetas gigantes.
@@
 
-Muitos dos objetos que pertenciam ao cinturão de Kuiper primordial acabaram sendo enviados pela interação com os planetas gigantes para as partes mais internas do sistema solar.
-Alguns podem ter sido capturados pelos planetas criando a população de satélites irregulares ou satélites externos \citep{Nesvorny2007}, troianos \citep{Morbidelli2005} ou até mesmo para o cinturão principal de asteroides \citep{McKinnon2012}.
-Estudar esses objetos é de grande importância para entender a evolução do sistema solar.
+Muitos dos objetos que pertenciam ao cinturão de Kuiper primordial acabaram sendo enviados pela interação com os planetas gigantes para as partes mais internas do Sistema Solar.
+Alguns podem ter sido capturados pelos planetas criando a população de satélites irregulares ou satélites externos \citep{Nesvorny2007}, troianos \citep{Morbidelli2005} ou até mesmo para o cinturão principal de asteroides como proposto para Ceres por \cite{McKinnon2012}.
+Estudar esses objetos é de grande importância para entender a evolução do Sistema Solar.
 
@@
 Ele foi capturado por Netuno \citep{McKinnon2007} em uma órbita retrógrada e próxima ao planeta.
-Sua superfície é deformada com características tectônicas e possivelmente criovulcânicas \citep{Nimmo2015} e as propriedades físicas de Tritão já conhecidas mostram uma similaridade com as de Plutão.
-Além do mais, é um dos poucos objetos do sistema solar que se sabe possuir atmosfera.
+Sua superfície é deformada com características tectônicas e possivelmente criovulcânicas \citep{Nimmo2015} e as propriedades físicas já conhecidas de Tritão mostram uma similaridade com as de Plutão.
+Além do mais, é um dos poucos objetos do Sistema Solar que se sabe possuir atmosfera.
 
-Sabe-se que poucas sondas espaciais foram enviados para estudar o sistema solar externo e que a quantidade de objetos estudados é muito pequena.
+Sabe-se que poucas sondas espaciais foram enviados para estudar o Sistema Solar externo e que a quantidade de objetos estudados é muito pequena.
 Por isso, ainda hoje, as observações de solo tem se mostrado de grande importância.
 
-Os sistemas de Júpiter e Saturno já foram pelas Voyager I e II, Galileu (Júpiter) e Cassini (Saturno), porém apenas Saturno continua sendo investigado por uma sonda.
+Os sistemas de Júpiter e Saturno já foram visitados pelas Voyager I e II, Galileu (Júpiter) e Cassini (Saturno), porém apenas Saturno continua sendo investigado por uma sonda.
 Todavia, as sondas observaram apenas os planetas, os anéis e satélites mais internos.
@@
 
-Já no caso de Urano e Netuno, nenhuma sonda exclusiva foi enviada, apenas as Voyagers I e II os visitaram, mas não permaneceram nos sistemas.
-A sonda New Horizons já está próxima a Plutão e obterá parâmetros físicos para Plutão e seus satélites (primeira visita por sonda a um objeto do cinturão de Kuiper), porém será uma passagem rápida e o acompanhamento da evolução do sistema, incluindo a evolução da atmosfera de Plutão se dará por observações de solo.
+Já no caso de Urano e Netuno, nenhuma sonda exclusiva foi enviada, apenas as \textit{Voyagers I} e \textit{II} os visitaram, mas não permaneceram nos sistemas.
+A sonda \textit{New Horizons} já está próxima a Plutão e obterá parâmetros físicos para Plutão e seus satélites (primeira visita por sonda a um objeto do cinturão de Kuiper), porém será uma passagem rápida e o acompanhamento da evolução do sistema, incluindo a evolução da atmosfera de Plutão se dará por observações de solo.
 
@@
 
-O objetivo desse Doutorado é observar objetos do sistema solar exterior, fazer astrometria para melhorar suas efemérides de forma que ocultações estelares por esses objetos possam ser preditas com precisão, observar as ocultações e obter os parâmetros físicos.
-Além disso, será feito um estudo astrométrico do sistema Netuno-Tritão para obter parâmetros relacionados a efeitos de maré.
+O objetivo desse Doutorado é observar objetos do Sistema Solar exterior, fazer astrometria para melhorar suas efemérides de forma que ocultações estelares por esses objetos possam ser preditas com precisão, observar as ocultações e obter os parâmetros físicos.
+Além disso, será feito um estudo astrométrico do sistema Netuno-Tritão para obter parâmetros relacionados a efeitos de maré, que permitem em princípio inferir a estrutura interna desses corpos.
+
+Assim, além do trabalho rotineiro de estudo de TNOs e Centauros feito pelo grupo, do qual eu participo ativamente, nos propomos a estudar aqui, no contexto observacional acima, alguns objetos que, por hipótese, também podem ser representativos da população original de TNOs, isto é, podem ter uma origem comum aos TNOs que hoje habitam o Cinturão de Kuiper.
+São eles alguns dos satélites irregulares de planetas gigantes, Ceres e Tritão.
+
+No capítulo \ref{Cap: Astrometria} descrevemos a astrometria feita para ajudar a corrigir as efemérides dos corpos aqui estudados, etapa essencial para melhorar a predição e viabilizar a efetiva observação de ocultações estelares.
+No capítulo \ref{Cap: ocultacoes} destacamos a importância da técnica de ocultações estelares e sua capacidade de obter parâmetros físicos dos objetos do Sistema Solar a partir de observações de solo.
+
+Nos capítulos \ref{Cap: TNOs}, \ref{Cap: Ceres}, \ref{Cap: Irregulares} e \ref{Cap: Netuno-Tritao} relatamos os trabalhos desenvolvidos envolvendo predições, observações e testes de ocultações estelares com Centauros e TNOs, Ceres, satélites irregulares dos planetas gigantes e com o sistema Netuno-Tritão, respectivamente.
+Por fim, no capítulo \ref{Cap: perspectivas} resumimos o que já foi desenvolvido e descrevemos o que ainda está para ser desenvolvido.
 
@@
 
-\section{Procedimento de Redução}
-
-O tratamento astrométrico das imagens é feito utilizando a Plataforma para Redução Automática de Imagens Astronômias \citep[PRAIA][]{Assafin2011}.
+\indent \indent O tratamento astrométrico das imagens é feito utilizando a Plataforma para Redução Automática de Imagens Astronômicas \citep[PRAIA][]{Assafin2011}.
 As medidas (x, y) são obtidas através de uma Gaussiana bidimensional simétrica ajustada até uma largura à altura (FWHM = seeing).
@@
 O PRAIA identifica automaticamente as estrelas de catálogo e determina ($\alpha$, $\delta$) com um modelo definido pelo usuário que relaciona as medidas (x, y) da imagem e as coordenadas projetadas no plano do céu (X, Y).
+Os alvos são automaticamente identificados nas reduções, a partir das suas efemérides (JPL, NAIF, etc) e suas posições, offsets de efeméride e estimativa de erros de medida e de posição guardados para análise posterior.
 
@@
 
-Para cada frame, usamos um modelo polinomial de seis constantes para relacionar as medidas (x, y) com as coordenadas (X, Y) do plano tangente.
+Para cada frame, usamos um modelo polinomial que pode ser de grau 1, 2 ou 3 para relacionar as medidas (x, y) com as coordenadas (X, Y) do plano tangente.
 Para observações feitas no ESO, o procedimento astrométrico seguido é o mesmo descrito em \citep{Assafin2012}; as medidas (x, y) dos CCDs individuais são pré-corrigidos por um padrão de distorção de campo e todas as posições vindas de diferentes CCDs e mosaicos são combinadas usando o polinômio de grau três para produzir uma solução global para cada noite e campo observado, a as posições ($\alpha$, $\delta$) são então obtidas no sistema do catálogo utilizado.
+Estrelas de referência com resíduos de posição (O-C) que desviam mais de 120 mas (ou valor similar adotado pelo usuario, usualmente 3 sigma do erro do catálogo) são automaticamente eliminadas uma a uma.
 
-\section{Centauros e TNOs}
+Um fator importante do trabalho está na utilidade estatística de se usar uma base de dados com um número extremamente elevado de observações (imagens).
+Esta é a maneira pela qual os efeitos atmosféricos de turbulência e extinção são efetivamente alisados e as informações de posição e de brilho são corretamente extraídas das imagens CCD.
+
+O tratamento astrométrico e fotométrico do PRAIA é capaz de lidar com uma imensa quantidade de dados de forma rápida, sendo capaz de obter a qualidade necessária para atingir os objetivos científicos dos programas observacionais em tempo recorde de processamento.
+As implementações incorporadas no software refletem experiência de mais de 20 anos de atuação na área da redução astrométrica e fotométrica de imagens astronômicas digitalizadas.
+
+Há ainda o cálculo de magnitudes instrumentais via fotometria de abertura e PSF, calculando-se o ponto zero em relação ao sistema fotométrico do catálogo de referência, como o 2MASS, UCAC2 ou o UCAC4 (o ponto zero pode ser re-definido).
+A magnitude do fundo de céu também é calculada.
+A FWHM das estrelas (seeing) é calculada automaticamente, sendo dados valores individuais por estrela, e valores médios, com as respectivas incertezas.
+
+Todas as informações fotométricas e astrométricas da redução, para todos os catálogos, e para cada campo, são guardadas em arquivos de dados.
+Também são gerados arquivos listando as estatísticas gerais de todos os campos.
+Os arquivos são de fácil manuseio nas tarefas de análise numérica e gráfica dos resultados.
+Além das posições obtidas, listam-se os parâmetros Gaussianos dos ajustes de cada estrela, os erros de medida em (x,y) e em posição, número de estrelas de referência, movimentos próprios, época da observação, magnitudes UCAC2, UCAC4, 2MASS, magnitudes instrumentais, informações da observação (filtros usados, tempo de exposição, etc).
+Em particular, são listados em arquivos separados os dados pertinentes aos alvos identificados nos campos, incluindo-se as diferenças de posição entre os ($ \alpha $,$ \delta $) observados reduzidos com UCAC2, UCAC4 e 2MASS e as coordenadas de referência fornecidas previamente pelo usuário em arquivo (efeméride, catálogo, etc.), chamados de offsets.
+Um esquema ilustrativo do PRAIA é mostrado na Fig \ref{Fig:PRAIA}.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.4]{figuras/PRAIA.png}
+\caption[Esquema ilustrativo do pacote PRAIA]{Esquema ilustrativo do pacote PRAIA: a sequência a ser seguida pelo PRAIA está no sentido horário, desde a extração dos dados das imagens, até a identificação dos alvos nos campos CCD e, consequentemente, a obtenção de resultados.}
+\label{Fig:PRAIA}
+\end{figure}
+
+\chapter{Ocultações Estelares}
+\label{Cap: ocultacoes}
+
+\indent \indent Ocultação estelar é um evento que ocorre quando um objeto do Sistema Solar passa na frente de uma estrela de forma que o brilho da estrela é completamente ou parcialmente bloqueado durante um intervalo de tempo.
+A variação de brilho da estrela pode ser medido fazendo-se observações com exposições curtas.
+O produto gerado é uma curva de fluxo de luz x tempo.
+Essa técnica pode ser utilizada para se obter parâmetros físicos dos objetos ocultantes como tamanho, forma, albedo, densidade (se a massa é conhecida), detecção da presença de atmosfera, satélites, anéis, etc \citep{Elliot1979}.
+
+A Fig.
+\ref{Fig: occ-scheme} mostra uma curva de luz esquemática de uma ocultação estelar central por um planeta com atmosfera e anéis.
+Ao ocultar a estrela, os anéis passarão primeiro, a densidade do anel, tamanho e tempo de exposição determinará a queda de brilho da estrela.
+Ao chegar no corpo principal, caso não possua atmosfera, a queda de brilho será abrupta, caso possua, a queda de brilho será mais suave devido à refração da luz da estrela na atmosfera dificultando a determinação da superfície do objeto.
+
+\begin{figure}[h]
+\begin{centering}
+\includegraphics[scale=0.75]{figuras/stellarocc.png} 
+\label{Fig: occ-scheme}
+\caption{Curva de luz esquemática de uma ocultação estelar por um planeta com atmosfera e anéis. Fonte: \cite{Elliot1979}}
+\end{centering} 
+\end{figure}
+
+A suavidade da curva portanto será determinada pelo gradiente de temperatura, pressão e densidade da atmosfera, presença de nuvens, etc.
+A luz da estrela nunca será completamente apagada quando ocultada por um objeto com atmosfera devido à refração da luz na atmosfera onde parte dela convergirá na direção do observador.
+Caso a ocultação seja central para o observador haverá um aumento de brilho no meio do evento ("flash") pois a convergência da luz será máxima.
+A atmosfera, no caso, agiria como uma lente como mostrado na Fig.
+\ref{Fig: occ-scheme}.
+
+Quando o evento é observado por diversos telescópios espalhados, cada observador verá a ocultação em latitudes diferentes.
+Essas diversas curvas de luz permitirão reconstruir a forma do objeto, possivelmente identificando estruturas superficiais como montanhas e crateras.
+Como a estrela está muito distante, podendo ser considerada quase pontual, a sombra do objeto durante a ocultação é muito próximo ao seu tamanho real.
+
+Para se observar esses eventos, portanto, é necessário uma campanha grande com observadores espalhados por várias regiões para cobrir a sombra do objeto.
+A precisão nas posições da estrela e do objeto são cruciais pois o evento ocorre em instantes específicos e passará por regiões específicas da Terra alheias à escolha do observador.
+
+%Since their ephemerides are not very precise, predicting and observing stellar occultations are very difficult, and no observation of such an event for an irregular satellite is found in the literature.
+The precise star positions to be derived by the ESA astrometry satellite Gaia \citep{deBruijne2012} will render better predictions with the only source of error being the ephemeris.
+The positions derived from our observations can be used in new orbital numerical integrations, generating more precise ephemerides.
+
+%The power of stellar occultations for observing relatively small diameter solar system objects is supported by recent works, such as the discovery of a ring system around the Centaur (10199) Chariklo \citep{Braga-Ribas2014}.
+Once irregular satellites start to be observed by this technique, it will be possible to obtain their physical parameters (shape, size, albedo, density) with unprecedented precision.
+For instance, in this case, sizes could be obtained with kilometer accuracy.
+The knowledge of these parameters would in turn bring valuable information for studying the capture mechanisms and origin of the irregular satellites.
+
+\chapter{Centauros e TNOs}
+\label{Cap: TNOs}
+
+\section{Astrometria}
 \label{Sec: TNOs-astrometria}
 
-\indent \indent Observar uma ocultação exige um grande trabalho de predição e sua melhoria.
+\indent \indent Observar uma ocultação exige um grande trabalho de predição e de sua melhoria.
 Os TNOs estão muito distantes significando que mesmo sem grandes erros nas suas efemérides ou na posição da estrela o erro do local por onde a sombra irá passar pode ser de milhares de quilômetros.
@@
 Da mesma forma, noites em que houveram um número de observações maior que as restantes terá um peso moderado já que se fosse utilizado um peso individual, o peso final de uma noite poderia estar superestimado levando a um erro na integração.
-Neste trabalho contribuí com observações de TNOs feitas no OPD.
+Neste trabalho contribuí com observações e reduções astrométricas de imagens de TNOs feitas no OPD.
 
-\section{Satélites Irregulares dos planetas gigantes}
-\label{Sec: Irr-astrometria}
+\section{Ocultações Estelares}
+\label{Sec: TNO-occ}
+
+\indent \indent Como dito no Cap.
+\ref{Cap: intro}, certos objetos do Sistema Solar exterior podem ser considerados fósseis do Sistema Solar.
+Dentre esses, os TNOs, por estarem mais distantes, são os que guardam as informações mais antigas.
+Com o objetivo de obter os parâmetro físicos desses objetos, para melhor caracterizar as propriedades básicas dessas populações, uma grande campanha observacional é realizada para se observar ocultações estelares por TNOs e Centauros.
+Essa campanha é liderada por pesquisadores brasileiros, franceses e espanhóis e tem colaboração com observadores espalhados por todo o mundo.
+
+Os objetos foram basicamente selecionados de acordo o grau de interesse científico (representar uma classe de TNO, ser um  objeto peculiar, etc), tamanho (quanto maior, mais chances de observar a ocultação) e região do céu (quanto mais estrelas, maior  a probabilidade de ocultar estrelas brilhantes, permitindo curvas de luz de melhor resolução espacial).
+
+Vários objetos já foram observados pelo grupo, sendo que o Sistema Plutão-Caronte conta com várias ocultações observadas além de um trabalho astrométrico minucioso baseado em 19 anos de observações no OPD, mais observações no ESO \citep{Rossi2014}.
+As observações bem sucedidas de ocultações estelares resultantes desse trabalho de predição já geraram diversas publicações, inclusive na revista Nature, por exemplo \cite{Sicardy2011} e \cite{BragaRibas2014} nas quais contribuí em caráter observacional.
+A ocultação de Plutão de 04 de Maio de 2013, a qual observei, gerou duas publicações \citep{Olkin2015, DiasOliveira2015}.
+
+\cite{Sicardy2011} identificaram, a partir de uma ocultação de 06 de Novembro de 2010, que Eris é um planeta anão menor do que se esperava.
+Devido a seu brilho e distância acreditava-se que ele fosse muito maior, mas na verdade ele tem um albedo geométrico no visível de $p_v=0.96^{+0.09}_{-0.04}$ e seu tamanho é de $1163 \pm 6$ km, muito próximo ao tamanho de Plutão.
+O alto albedo de Eris pode ser relacionado à presença de uma atmosfera colapsada devido ao fato dele estar se aproximando de seu afélio.
+
+\cite{BragaRibas2014} observaram uma ocultação do Centauro Chariklo que ocorreu dia 03 de Junho de 2013.
+O ocultação revelou que o objeto possui um sistema de anéis nunca visto antes em um objeto tão pequeno.
+Até o momento, conhecia-se apenas anéis em torno dos planetas gigantes.
+Essa descoberta levanta muitas questões sobre formação dos anéis, estabilidade e tempo de vida e nos leva a crer que pode haver mais objetos com anéis no Sistema Solar, ao menos nessa região.
+
+\cite{Olkin2015} determinaram a partir de observações de ocultações por Plutão que a atmosfera do objeto não colapsa durante os 248 anos de sua órbita, como acontece com Eris.
+Modelos anteriores previam um colapso de sua atmosfera devido a Plutão receber três vezes menos luz do Sol em afélio que em seu periélio.
+
+\cite{DiasOliveira2015}, também a partir de ocultações de Plutão, utilizaram modelos para obter os perfis de temperatura, pressão e densidade da atmosfera do objeto.
+Os modelos ajustam perfeitamente os perfis de temperatura para as duas ocultações utilizadas assumindo uma atmosfera esfericamente simétrica.
+Foram usadas as curvas de luz de melhor resolução até hoje obtidas para Plutão, sendo a melhor obtida em  Paranal em 18 de Julho de 2012, usando a câmera NAOS-CONICA (NACO) ligada ao 8.2-m “Yepun” Very Large Telescope (VLT) do ESO a uma taxa de 5 frames por segundo na banda H.
+Por fim, foi obtido que a pressão da atmosfera aumentou cerca de 6\% entre 2012 e 2013 significando que a atmosfera de Plutão ainda está se expandindo e confirmando o trabalho de \cite{Olkin2015}.
+
+No dia 29 de Junho de 2015 haverá uma ocultação de Plutão de máxima importância.
+Será a última antes da chegada da sonda \textit{New Horizons} no sistema e poderá ser observada na Austrália.
+Observada a ocultação, os modelos de perfil de atmosfera para Plutão poderão ser calibrados com os dados da \textit{New Horizons} para serem utilizados em futuras ocultações e estudar a evolução da atmosfera do planeta anão.
+
+Em 29 de Abril de 2014, uma ocultação de Chariklo foi observada na África.
+Eu trabalhei com a redução de três curvas de luz (Gifberb, SAAO e Springbok).
+Um fato interessante dessa ocultação foi que a estrela era uma binária não catalogada.
+Como a segunda estrela era bem mais fraca, acreditava-se que a queda de magnitude causada por ela poderia ser algum material próximo à Chariklo.
+
+Essa ocultação foi importante para obter com total certeza a direção do polo de Chariklo.
+Além disso, as câmeras utilizadas eram de alta qualidade possibilitando obter que pelo menos um dos anéis é opaco.
+Uma das cordas, fortuitamente, passou contígua ao anel observando apenas o anel externo.
+Os resultados ainda não foram publicados.
+
+Outro evento que observei foi uma ocultação de 2003VS2 dia 07 de Novembro de 2014 no OPD.
+As variações do céu não permitiram uma boa observação de forma que a curva de luz estava muito ruidosa, mas boa o suficiente próximo ao instante do evento.
+Um trabalho foi feito para identificar alguma ocultação mas nada pôde ser extraído.
+Nesse evento foram obtidas 3 cordas positivas e 7 negativas.
+A redução preliminar revelou que 2003VS2 é possui um corpo levemente alongado com aproximadamente 600 km de diâmetro.
+O trabalho ainda está em análise.
+
+\chapter{Ceres}
+\label{Cap: Ceres}
+
+\section{Introdução}
+\label{Sec: Ceres-intro}
+
+\indent \indent Apesar de Ceres não ser um objeto do Sistema Solar exterior, ele é o único planeta-anão no Sistema Solar interno e, por isso, é um objeto de grande importância e seu estudo pode ter grande impacto na formação e evolução do Sistema Solar.
+Na verdade, foi proposto que a origem de Ceres pode ser como um objeto trans-Netuniano \citep{McKinnon2012}, espalhado posteriormente para o cinturão principal de asteroides devido à migração dos planetas gigantes predito pelo Modelo de Nice \citep{Gomes2005}.
+Mesmo que ele tenha sido formado próximo à sua localização atual, a história dinâmica do Sistema Solar deve ter deixado sua assinatura em Ceres.
+
+Contendo aproximadamente um quinto de toda a massa do cinturão de asteroides, espera-se que Ceres esteja em equilíbrio gravitacional e seja, portanto, um elipsoide Maclaurin ou Jacobi.
+De fato, observações diretas de Ceres com a utilização de ótica adaptativa indica que ele é um esferoide achatado nos polos \citep{Drummond2014}.
+O conhecimento preciso de seu tamanho e forma é extrema importância para modelos de densidade, estrutura interna e diferenciação.
+
+A primeira ocultação estelar por Ceres foi observada em 1984 \citep{Millis1987} e determinou seu tamanho com precisão de alguns quilômetros em uma época que as incertezas eram, normalmente, dez vezes maiores.
+Devido ao brilho aparente de Ceres ser alto, comparado à maioria dos asteroides, estrelas capazes de causar uma queda de magnitude detectável quando ocultadas são limitadas às mais brilhantes.
+Por exemplo, depois do evento de 1984, apenas 4 ocultações estelares por Ceres foram observadas \citep{Dunham2014}.
+Duas delas tiveram apenas duas cordas cada que não foram suficientes para prover resultados acurados\footnote{Esses eventos ocorreram em 22 de Agosto de 1994 e 30 de Outubro de 2010.}.
+Os dois restantes, que ocorreram em 17 de Agosto de 2010 e 25 de Outubro de 2013, foram trabalhos por mim em colaboração com o grupo do Rio \citep[aceito]{GomesJunior2015-Ceres}.
+
+Os dois eventos foram preditos por Steve Preston\footnote{Predições publicadas em \url{http://asteroidoccultation.com}.} para a IOTA (International Occultation Timing Association), durante predições de rotina de ocultações de asteroides de estrelas brilhantes.
+Os caminhos das sombras podem ser visualizados na Fig.
+\ref{Fig: Ceres-map}.
+
+\begin{figure}[h]
+\begin{centering}
+\subfigure[Ocultação de 2010]{\includegraphics[scale=0.39]{figuras/Ceres_2010.eps}\label{Fig: Ceres-map-2010}}
+\subfigure[Ocultação de 2013]{\includegraphics[scale=0.39]{figuras/Ceres_2013-zoom.eps}\label{Fig: Ceres-map-2013}}
+\caption{Reconstrução pós-ocultação do caminho da sombra de Ceres na Terra para os eventos de 17 de Agosto de 2010 (a) e 25 de Outubro de 2013 (b).
+Os pontos em azul são os sítios que observaram os eventos.
+a) O ponto grande vermelho é a máxima aproximação geocêntrica às 22:40:25 UT.
+Os pequenos representam o centro da sombra separados por um minuto.
+b) Visão superior da ocultação sobre os sítios que observaram o evento de 25 de Outubro de 2013.
+Os pontos vermelhos são os centros da sombra separados por 15 segundos.
+Nos dois eventos a sombra se move da esquerda para a direita.
+\label{Fig: Ceres-map}}
+\end{centering}
+\end{figure}
+
+\section{Ocultação de 2010}
+\label{Subsec: 2010-occ}
+
+\indent \indent Em 17 de Agosto de 2010 Ceres ocultou a estrela TYC 6833-163-1 (UCAC4 313-111823), cuja magnitude é $V = 11.55$ e tem posição no ICRS para a data do evento baseada no catálogo UCAC4 \citep{Zacharias2013}:
+
+\begin{equation}
+\left\{ 
+  \begin{array}{l l}
+    \alpha = 17^{h}18^{m}29\fs 0085\\
+    \delta = -27\degr 26\arcmin 38\farcs 867
+  \end{array}
+\right.
+\label{Eq: 2010-coord}
+\end{equation}
+
+\indent O evento foi observado no Brasil a partir de cinco diferentes sítios (ver Fig.
+\ref{Fig: Ceres-map-2010}).
+Destes, 4 obtiveram cordas positivas enquanto UFSC teve uma corda negativa.
+Das positivas, a observação proveniente do INPE iniciou-se após o início do evento devido a dificuldades técnicas e, portanto, apenas a emersão da curva de luz foi detectada.
+
+Uma das características mais importantes desse evento foi a velocidade com que ocorreu (apenas 3.9 km s$^{-1}$) acarretando que mesmo exposições de poucos segundos representariam resoluções espaciais significantes.
+
+Todas as observações foram feitas com a utilização de CCDs.
+As curvas de luz de cada observação foram obtidas das imagens FITS com a utilização do pacote PRAIA \citep[Plataforma de Redução Astrométrica de Imagens Astronômicas,][]{Assafin2011}.
+As curvas foram normalizadas para o fluxo da estrela mais Ceres, uma vez que eles estavam indistinguíveis logo antes e depois da ocultação.
+Por fim, elas foram normalizadas pelo ajuste de uma curva polinomial (de primeira ou segunda ordem) fora da queda de fluxo assim fixando em 1 a razão de fluxo fora da ocultação.
+
+Os instantes de ingresso e egresso foram obtidas de cada curva de luz ajustando-se um modelo de poço quadrado levando em consideração a difração de Fresnel, a banda do CCD, o diâmetro aparente da estrela e o tempo de exposição utilizado \citep[ver][]{Widemann2009, Braga-Ribas2013}.
+
+O menor tempo de integração usado nas observações positivas foi de $1.0s$, que corresponde a aproximadamente $3.9km$ no plano do céu.
+Portanto, o erro na determinação dos instantes de ingresso e egresso é dominado principalmente pelo tempo de integração, não pela difração de Fresnel ou diâmetro da estrela, ambos da ordem de algumas centenas de metros para esse evento.
+
+O ajuste dos dados da ocultação consiste em minimizar uma função dep $\chi^{2}$ clássica para cada curva de luz, como descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}.
+Os parâmetros livres para ajustar são os instantes de ingresso e egresso que fornece o valor mínimo de $\chi^{2}$ ($\chi^{2}_{min}$).
+O melhor ajuste das curvas de luz para a ocultação de 2010 está mostrado na Fig.
+\ref{Fig: Ceres-2010-curves}.
+
+\begin{figure}[h]
+\begin{centering}
+\subfigure[Ocultação de 2010]{\includegraphics[scale=0.51]{figuras/Ceres_2010_fluxratio.eps}\label{Fig: Ceres-2010-curves} }
+\subfigure[Ocultação de 2013]{\includegraphics[scale=0.51]{figuras/Ceres_2013_fluxratio.eps}\label{Fig: Ceres-2013-curves} }
+\caption{Curvas de luz normalizadas das cordas positivas dos eventos.
+As curvas estão desviadas por um fator de 0.2 (a) e 1.0 (b) para melhor visualização.
+As linhas pretas são os melhores ajustes com o modelo de poço quadrado.
+As linhas vermelhas são os melhores ajustes com o modelo de poço quadrado, porém levando em conta a difração de Fresnel, o diâmetro da estrela e o tempo de exposição.
+Os instantes médios de cada curva não coincidem devido às diferentes longitudes dos sítios.
+A curva de luz de Brookline (b) está desviada por um fator de -64 s como explicado no texto.
+\label{Fig: Ceres-curves}}
+\end{centering}
+\end{figure}
+
+A metodologia usada para analisar o perfil de Ceres a partir das observações é o mesmo descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}.
+Cada combinação de posição do sítio, instantes de ingresso e egresso, junto com as coordenadas da estrela e as efemérides de Ceres, correspondem a um ponto no plano do céu.
+A coleção de todos esses pontos determina o limbo aparente de Ceres.
+
+Adotamos um modelo elíptico para o perfil do limbo, resultante da projeção de um esferoide com achatamento nos polos no plano do céu.
+Essa escolha é suportada pelo trabalho de \cite{Drummond2014}, por meio de imagem direta de Ceres.
+Dessa forma, nós temos $N=7$ extremidades das cordas para ajustar $M=5$ parâmetros que definem uma elipse: semi-eixo maior e semi-eixo menor aparentes ($a^\prime$ and $b^\prime$, respectivamente), ângulo de posição $P$ do seu semi-eixo maior e as posição $(f_c,g_c)$ do seu centro com respeito à estrela ocultada.
+O semi-eixo maior {$a^\prime$} é equivalente ao raio equatorial $R_{equa}$ do elipsoide.
+
+As coordenadas $f_{c}$ e $g_{c}$, em quilômetros, foram calculadas usando a efeméride de Ceres JPL\#33 \citep{Giorgini1996} e a posição da estrela ocultada.
+Elas são positivas na direção Leste e Norte celestes, respectivamente.
+O ângulo de posição $P$ é contado positivamente a partir do norte celeste local em direção ao leste celeste.
+O achatamento aparente pode ser definido por $\epsilon^\prime = 1 - (b^\prime/a^\prime)$.
+O melhor ajuste é obtido minimizando uma função de $\chi^{2}_{r}$ reduzido, onde definimos o número de graus de liberdade do problema como $\mathcal{N} \equiv N - M$.
+Todos os procedimentos que permitem a determinação das barras de erro dos parâmetros físicos podem ser encontradas em \cite{Braga-Ribas2013}.
+
+Duas possíveis soluções foram consideradas para o ajuste do limbo.
+O primeiro, que chamamos de solução nominal, consiste em determinar os cinco parâmetros que caracterizam uma elipse a partir dos sete contatos observados.
+A segunda solução consiste em calcular o ângulo de posição $P$ a partir das coordenadas do polo de Ceres obtidas por \cite{Drummond2014} ($\alpha_{p} = (287 \pm 3) \degr$, $\delta_{p} = (+64 \pm 3) \degr$ no ICRS) e da efeméride de Ceres no instante da ocultação.
+Chamamos de solução de polo fixo.
+
+Para o evento de 2010, a solução nominal teve como melhor ajuste $\chi^2_{r,min} = 0.24$, que podem ser interpretadas como as barras de erro estarem superestimadas com respeito à boa qualidade do ajuste.
+Porém, como o problema tem somente dois graus de liberdade, $\chi^2_{r,min}$ relativamente pequenos são aceitáveis.
+Os resultados obtidos para o diâmetro equatorial, achatamento, ângulo de posição e coordenadas do centro são apresentadas na segunda coluna da tabela \ref{Tab: resultados}.
+
+\begin{table}[h]
+ \begin{centering}
+  \caption{Resultados do ajuste de limbo de Ceres com os dados dos eventos de 2010 e 2013.\label{Tab: resultados}}
+  \begin{tabular}{@{}lcccc}
+  \hline
+     Solução & 2010/Nominal & \textbf{2010/Polo fixo} & 2013/Nominal & 2013/Polo fixo \\
+\hline
+Diam. equat.  (km) & 982 $\pm$ 14 & \textbf{972 $\pm$ 6}  & 971 $\pm$ 7  & 971 $\pm$ 7\\
+%Equatorial radius (km) & 491 $\pm$ 7  & \textbf{486 $\pm$ 3}  & 485.5 $\pm$ 3.5 & 485.5 $\pm$ 3.5\\
+Achatamento        & 0.08 $\pm$ 0.03 & \textbf{0.08 $\pm$ 0.03} & 0.08 $\pm$ 0.04 & 0.08 $\pm$ 0.04\\
+Âng. de pos. (deg)   & 5 $\pm$ 10    & \textbf{12 $\pm$ 3} (*)& 22 $\pm$ 5    & 25 $\pm$ 3 (*)\\
+$f_c$ (km)             & 97 $\pm$ 9   & \textbf{102 $\pm$ 5}   & 77 $\pm$ 6    & 78 $\pm$ 6\\
+$g_c$ (km)             & 16 $\pm$ 15  & \textbf{21 $\pm$ 11}  & 13 $\pm$ 16   & 13 $\pm$ 16\\
+$\chi^2_{r,min}$       & 0.24          &  \textbf{0.42}         & 1.27          & 1.27\\
+\hline
+\end{tabular}
+\textbf{Notas:} Em negrito destaca-se a melhor solução obtida.
+As barras de erro estão no nível de 1$\sigma$.
+O diâmetro polar ($D_{pol}$) pode ser facilmente calculado a partir de $D_{pol}=D_{equa}(1 - \epsilon)$.
+(*) Ângulos de posição derivados a partir das coordenadas do polo de Ceres determinadas por \cite{Drummond2014}.
+\end{centering}
+\end{table}
+
+Como pode ser visto, o parâmetro com a maior incerteza é o ângulo de posição cobrindo um intervalo de 20$\degr$.
+Claramente, fixar as coordenadas do pólo pode melhorar a solução.
+Por fim, a correção do achatamento devido ao ângulo do aspecto polar está dentro da barra de erro 1$\sigma$ e não tem relevância estatística, dessa forma $\epsilon = 0.08 \pm 0.03$.
+
+No momento da ocultação, as coordenadas do polo de Ceres correspondiam a um ângulo de posição $P = (12 \pm 3)\degr$.
+Explorar o espaço de parâmetros restringindo a elipses com ângulo de posição dentro deste intervalo resulta na solução de pólo fixo.
+O parâmetros físicos do melhor ajuste estão mostrados na Tab.
+\ref{Tab: resultados} enquanto a solução está esquematizada na Fig.
+\ref{Fig: Ceres-limb-2010}.
+
+Essa solução corresponde ao limite superior da barra de erro 1$\sigma$ da solução nominal para $P$.
+Por outro lado, ela obtém os menores valores para o diâmetro equatorial, melhorando sua determinação por um fator de 2.
+
+\begin{figure}[h]
+\begin{centering}
+\subfigure[Ocultação de 2010]{\includegraphics[scale=0.35]{figuras/Ceres_2010_sphere.eps}\label{Fig: Ceres-limb-2010} }
+\subfigure[Ocultação de 2013]{\includegraphics[scale=0.35]{figuras/Ceres_2013_sphere.pdf}\label{Fig: Ceres-limb-2013} }
+\caption{Melhores ajustes elípticos para as cordas das ocultações de 2010 e 2013.
+As setas indicam a direção de movimento, as linhas azuis são as cordas observadas, os segmentos vermelhos são as barras de erro dos ingressos, egressos e centro da ocultação em $1\sigma$.
+A linha verde em (a) é uma corda negativa.
+Os instantes marcados em verde em (b) não foram utilizados para o ajuste, como descrito no texto.
+\label{Fig: Ceres-limb}}
+\end{centering}
+\end{figure}
+
+\section{Ocultação de 2013}
+\label{Subsec: 2013-occ}
+
+\indent \indent Em 25 de Outubro de 2010, Ceres ocultou a estrela TYC 865-911-1 (UCAC4 496-058191), de magnitude $V = 10.05$.
+Baseada no UCAC4 \citep{Zacharias2013}, sua posição ICRS para a data da ocultação é:
+
+%
+\begin{equation}
+\left\{ 
+  \begin{array}{l l}
+    \alpha = 11^{h}57^{m}52\fs7641\\
+    \delta = +09\degr 07\arcmin 49\farcs835
+  \end{array}
+\right.
+\end{equation}
+%
+Esse evento foi observado na costa Leste dos Estados Unidos logo antes do amanhecer, como mostrado na Fig. \ref{Fig: Ceres-map-2013}.
+
+Nove cordas positivas foram obtidas em vários sítios (ver Fig.
+\ref{Fig: Ceres-map-2013}).
+Cada estação foi equipada com uma câmera de vídeo com tempo de leitura desprezível.
+Isso é particularmente importante, já que a velocidade da sombra de Ceres para esse evento foi de 42.6 km s$^{-1}$, muito mais rápido que o evento de 2010.
+
+Durante o evento Ceres estava muito baixo no céu com alturas entre $15\degr$ (Winchester) e $20\degr$ (Hampton).
+Forte cintilação era esperada e, combinada com o curto tempo de integração e baixa diminuição de brilho, resultou em curvas de luz ruidosas e assim grandes incertezas nos instantes de imersão e emersão.
+
+Todos os vídeos foram convertidos para imagens FITS e a fotometria foi obtida via \textsc{praia} \cite{Assafin2011}.
+As curvas de luz foram normalizadas por uma estrela de referência quando havia uma estrela no campo.
+
+Para reduzir o ruído, os dados foram binados por grupos de cinco imagens -- com exceção de Greenbelt, onde grupos de dez imagens foram utilizadas.
+Esse procedimento o tempo de integração efetivo por um fator de 5 (ou 10).
+Da mesma forma que para o evento de 2010, uma normalização adicional por um polinômio foi aplicada.
+
+Os instantes de ingresso e egresso da ocultação foram obtidos pelo mesmo procedimento descrito na seção \ref{Subsec: 2010-occ}.
+Uma vez que o tempo de integração efetivo usado (0.17 s) representa aproximadamente 7 km no plano do céu, e a escala de Fresnel e  o diâmetro da estrela estão novamente na ordem de centenas de metros, o erro da determinação dos instantes de imersão e emersão são dominados principalmente pelo tempo de integração, da mesma forma que para o evento de 2010.
+Os melhores ajustes para as curvas de luz da ocultação estão mostrados na Fig.
+\ref{Fig: Ceres-2013-curves}.
+
+Uma comparação entre os tempos obtidos pela estação de Brookline e o restante mostrou que esses têm um atraso de aproximadamente 64 s. Dessa forma, não utilizamos os tempos de Brookline na análise.
+
+Perfis elípticos foram ajustados para todas as cordas restantes.
+pelo mesmo procedimento descrito na seção \ref{Subsec: 2010-occ}.
+O resultado foi $\chi^2_{r,min} = 13$ , sugerindo que um modelo elíptico não é satisfatório para os dados.
+De fato, ao olharmos para a Fig.
+\ref{Fig: Ceres-limb-2013} vemos que a corda de Varina adiantada com respeito às outras.
+Como nessa estação o tempo não foi inserido diretamente nos frames do vídeo é possível que essa diferença seja oriunda de um eventual problema da correspondência entre os tempos do camcorder e do GPS.
+
+A imersão gravada em Owings também parece atrasada com respeito às cordas próximas (ver Fig.
+\ref{Fig: Ceres-limb-2013}).
+Essa corda tem aproximadamente o mesmo tamanho da corda de Mechanicsville, apesar de estarem separadas por cerca de 100 km.
+Diferentemente de Varina, essa estação teve os tempos inseridos em cada frame do vídeo o que torna mais difícil justificar um problema de tempo.
+Outras possibilidades seriam uma má determinação dos instantes de ingresso e egresso dessa curva ou uma característica do relevo de Ceres.
+
+Em um segundo ajuste, não consideramos as cordas de Brookline, Varina e Owings.
+O ajuste dos cinco parâmetros que definem uma elipse para os doze contatos resultou em $\chi^2_{r,min} = 1.27$, indicando  que está em bom acordo com os dados observados dentro das barras de erro.
+Essa é a solução mostrada na Fig.
+\ref{Fig: Ceres-limb-2013}, onde podemos ver que o tamanho da corda de Brookline é compatível com o modelo.
+
+Os resultados das soluções nominais e de polo fixo não são significantemente diferentes.
+Isso se deve ao fato da barra de erro do ângulo de posição da solução nominal (que é muito menor que o da solução nominal de 2010) ser muito similar à barra de erro do ângulo de posição da solução de polo fixo ($P = (25 \pm 3)\degr$).
+Os dois resultados estão apresentados nas colunas 4 e 5 da tabela \ref{Tab: resultados}.
+
+Sobre a hipótese do atraso observado na imersão da curva de luz de Owings ser associado a uma característica topográfica, o contato gravado corresponderia a uma elevação negativa de 31 $\pm$ 4 km com respeito à elipse de melhor ajuste.
+Porém modelos teóricos preveem que relevos em Ceres não devem ser maiores que 10--20 km \citep{Johnson1973}, enquanto dados observacionais limitam o contorno em 18 km \citep{Carry2008}.
+Imagens mais recentes da sonda \textit{Dawn} também revelam uma superfície mais suave.
+Portanto, a associação da imersão em Owings com um relevo é improvável.
+
+\section{Discussão}
+
+\indent \indent Os resultados apresentados na tabela \ref{Tab: resultados} mostram um acordo entre os parâmetros físicos obtidos nas duas ocultações, especialmente no diâmetro equatorial.
+As diferenças ocorrem basicamente nos tamanhos das barras de erro e podem ser justificadas pelas particularidades de cada conjunto de dados.
+
+O evento de 2010, por exemplo, teve somente sete contatos, porém bem distribuídos sobre o disco de Ceres.
+Por outro lado, o evento de 2013 teve cinco contatos a mais, todavia concentrados em certas regiões do corpo.
+Em particular, a ausência de cordas próximas ao polo sul fez seu achatamento ser pior determinado para o evento de 2013 que para o evento de 2010.
+Mesmo nossa melhor medida de achatamento, $\epsilon=0.08 \pm 0.03$, tem alta incerteza se comparado com outros valores publicados na literatura, como mostra a tabela \ref{Tab: Ceres-final}.
+
+Como foi mostrado, usar as coordenadas do polo de Ceres determinadas por \cite{Drummond2014} para limitar o ângulo de posição não foi um procedimento eficiente para o evento de 2013.
+Por outro lado, fixar o ângulo de posição para a ocultação de 2010 reduziu as barras de erro dos outros parâmetros (com exceção do achatamento).
+Por fim, Esse procedimento resultou em um excelente acordo entre os raios equatoriais obtidos para ambos os eventos.
+
+Uma comparação do diâmetro equatorial de Ceres medido por diferentes técnicas está mostrado na tabela \ref{Tab: Ceres-final}.
+Vemos um acordo entre os nossos resultados e aqueles obtidos por imageamente direto do Hubble Space Telescope (HST) \citep{Thomas2005}, do Keck Observatory e do ESO VLT \citep{Drummond2014}.
+O menor valor, reportado por \cite{Carry2008}, pode ser justificado pelo fato desse estudo não levar em conta o efeito de escurecimento de bordo.
+
+O evento de 1984 \citep{Millis1987} é a única outra ocultação que podemos comparar nosssos resultados.
+As medidas do diâmetro não se acordam dentro de 2$\sigma$.
+É difícil dizer com certeza as razões dessa divergência.
+Uma forma de clarificar o problema seria redeterminar os instantes de imersão e emersão das curvas de luz originais usando a mesma metodologia aplicada nesse trabalho.
+Infelizmente, não temos acesso aos dados da curva de luz original do evento de 1984.
+
+A sonda da NASA \textit{Dawn} poderá responder esses questões que são importantes não só no conhecimento do próprio Ceres, mas também para calibrar todas as técnicas usadas até agora para o estudo das propriedades físicas dos pequenos objetos do Sistema Solar, como as ocultações estelares.
+
+\begin{table}[h]
+  \caption{Diâmetro equatorial e achatamento de Ceres \label{Tab: Ceres-final}}
+  \begin{centering}
+  \begin{tabular}{@{}lccc}
+  \hline
+     Diâmetro Equatorial (km) & Achatamento & Método & Ref. \\
+\hline
+$972 \pm 6$  & $0.08  \pm 0.03$  & Occultação & 1\\
+$967 \pm 10$ & $0.078 \pm 0.015$ & Keck+VTL    & 2\\
+$959 \pm 5$  & $0.074 \pm 0.007$ & Keck        & 3\\
+$975 \pm 4$  & $0.067 \pm 0.005$ & HST         & 4\\
+$959 \pm 5$  & $0.05  \pm 0.01$  & Occultação & 5\\
+\hline
+\end{tabular}
+\par\end{centering}
+\textbf{Referências.} 1: \cite{GomesJunior2015-Ceres}. 2: \cite{Drummond2014}. 3: \cite{Carry2008}. 4: \cite{Thomas2005}. 5: \cite{Millis1987}.
+\end{table}
+
+\chapter{Satélites Irregulares dos planetas gigantes}
+\label{Cap: Irregulares}
+
+\section{Introdução}
+\label{Sec: Irr-intro}
 
@@
 Na maioria dos casos, essas órbitas são retrógradas.
-Devido a suas configurações orbitais, é amplamente aceito que estes objetos foram capturados nos estágios iniciais da formação do sistema solar \citep{Sheppard2003}.
+Devido a suas configurações orbitais, é amplamente aceito que estes objetos foram capturados nos estágios iniciais da formação do Sistema Solar \citep{Sheppard2003}.
 
@@
 Apenas Himalia, Phoebe e Nereida foram observados por sondas, apesar de não serem medidas ideais, já que foram alvos de oportunidade.
-A sonda Cassini observou Himalia em 2000 ao passar próximo a Júpiter e obteve o tamanho de Himalia com um erro médio de 10 km \citep{Porco2003}.
-Em 2004, a Cassini, aproximando-se de Saturno, observou Phoebe em alta resolução obtendo um erro médio na medida de seu tamanho de 0.7 km \citep{Thomas2010}.
-Por fim, Nereida foi observado em 1989 pela sonda Voyager 2 e seu tamanho foi obtido com um erro de 25 km \citep{Smith1989}.
+A sonda \textit{Cassini} observou Himalia em 2000 ao passar próximo a Júpiter e obteve o tamanho de Himalia com um erro médio de 10 km \citep{Porco2003}.
+Em 2004, a \textit{Cassini}, aproximando-se de Saturno, observou Phoebe em alta resolução obtendo um erro médio na medida de seu tamanho de 0.7 km \citep{Thomas2010}.
+Por fim, Nereida foi observado em 1989 pela sonda \textit{Voyager II} e seu tamanho foi obtido com um erro de 25 km \citep{Smith1989}.
 Para outros satélites irregulares de Júpiter, seus tamanhos foram estimados por \cite{Rettig2001} impondo um albedo de 4\% com a justifica de que esse valor é um albedo nominalmente utilizado para objetos do Sistema Solar Externo.
@@
 
-Um mecanismo baseado no modelo de Nice \citep{Morbidelli2005, Tsiganis2005, Gomes2005} foi proposto por \cite{Nesvorny2007} e, especificamente para Júpiter, por \citealp{Nesvorny2014}.
-Durante a instabilidade do sistema solar primordial, ocorreram encontros entre os planetas externos.
+Um mecanismo baseado no modelo de Nice \citep{Morbidelli2005, Tsiganis2005, Gomes2005} foi proposto por \cite{Nesvorny2007} e, especificamente para Júpiter, por \cite{Nesvorny2014}.
+Durante a instabilidade do Sistema Solar primordial, ocorreram encontros entre os planetas externos.
 Esses encontros planetários poderia haver troca de energia e momento angular entre os planetas e os objetos próximos tornando possível a captura de um corpo irregular por um planeta gigante.
@@
 
-Alguns desses objetos formam grupos dinâmicos com elementos orbitais semelhantes, chamados famílias, similar às famílias encontradas no Cinturão Principal de Asteróides.
-Essas famífias podem ter sido criadas por um corpo pai que se fragmentou por colisões com cometas ou outros asteróides \citep{Nesvorny2004}.
+Alguns desses objetos formam grupos dinâmicos com elementos orbitais semelhantes, chamados famílias, similar às famílias encontradas no Cinturão Principal de Asteroides.
+Essas famílias podem ter sido criadas por um corpo pai que se fragmentou por colisões com cometas ou outros asteroides \citep{Nesvorny2004}.
 Colisões com cometas tem uma probabilidade maior de ocorrer durante o Bombardeamento Pesado Tardio (LHB) \citep{Gomes2005}.
@@
 \cite{Nesvorny2003} estudou as taxas de colisões entre satélites irregulares e concluiu que alguns satélites podem ter sido removidos por colisão com um satélite maior.
-A taxa de colisão entre os satélites do grupo de Himalia (Himalia, Elara, Lysithea and Leda, principalmente), por exemplo, foi encontrado como sendo maior que "um" durante a idade do sistema solar sugerindo que sua estrutura atual foi originada por colisão satélite-satélite.
+A taxa de colisão entre os satélites do grupo de Himalia (Himalia, Elara, Lysithea and Leda, principalmente), por exemplo, foi encontrado como sendo maior que "um" durante a idade do Sistema Solar sugerindo que sua estrutura atual foi originada por colisão satélite-satélite.
 
 Para Phoebe, materiais ejetados de sua superfície causadas por impactos poderiam evoluir devido ao arrasto de Poynting-Robertson e colidir com Iapetus causando causando a grande variação de albedo observada no satélite \citep{Nesvorny2003}.
-De fato, A Cassini detectou em Phoebe uma característica de absorção em 2.42 $\mu m$, provavelmente combinações de CN, que foi também detectada no lado escuro de Iapetus \citep{Clark2005}
+De fato, a sonda \textit{Cassini} detectou em Phoebe uma característica de absorção em 2.42 $\mu m$, provavelmente combinações de CN, que foi também detectada no lado escuro de Iapetus \citep{Clark2005}
 
 Se esses objetos foram capturados, permanece a pergunta de onde eles vieram.
-\citealp{Clark2005} mostraram a partir de espectroscopia da Cassini que Phoebe tem uma superfície provavelmente coberta por materiam do sistema solar externo e \citealp{Grav2003} mostraram que os satélites de grupo prógrado de Júpiter Himalia tem cores cinzas implicando que suas superfícies são similares a asteróides tipo C.
-Nesse mesmo trabalho, o grupo retrógrado de Júpiter Carme foi encontrado como tendo cores superficiais semelhantes à asteróides tipo D como a família de Hilda ou famílias de troianos, enquanto JXIII Kalyke tem uma cor mais avermelhada como centauros ou objetos trans-Netunianos (TNOs).
+\citealp{Clark2005} mostraram a partir de espectroscopia da Cassini que Phoebe tem uma superfície provavelmente coberta por material do Sistema Solar externo e \citealp{Grav2003} mostraram que os satélites de grupo prógrado de Júpiter Himalia tem cores cinzas implicando que suas superfícies são similares a asteróides tipo C.
+Nesse mesmo trabalho, o grupo retrógrado de Júpiter Carme foi encontrado como tendo cores superficiais semelhantes à asteroides tipo D como a família de Hilda ou famílias de troianos, enquanto JXIII Kalyke tem uma cor mais avermelhada como centauros ou objetos trans-Netunianos (TNOs).
 
-Para os satélites de saturno, \citealp{Grav2007} mostraram através de suas cores e inclinações espectrais que esses satélites contêm uma fração mais ou menos igual de objetos semelhantes a asteróides tipo c, P e D, mas SXXII Ijiraq é um pouco mais avermelhado que objetos tipo D.
-Esses trabalhos sugerem origins diferentes para os satélites irregulares.
+Para os satélites de saturno, \citealp{Grav2007} mostraram através de suas cores e inclinações espectrais que esses satélites contêm uma fração mais ou menos igual de objetos semelhantes a asteroides tipo c, P e D, mas SXXII Ijiraq é um pouco mais avermelhado que objetos tipo D.
+Esses trabalhos sugerem origens diferentes para os satélites irregulares.
+
+\section{Astrometria}
+\label{Sec: Irr-astrometria}
 
@@
 Jean-Eudes Arlot do IMCCE, reduzi um banco de dados observado no Observatoire Haute-Provence (OHP) entre 1998 e 2008, contendo mais de 28 mil posições para 10 satélites.
-Reduzi também um banco de dados com mais de 100 mil imagens obtidas no Observatório do Pido dos Dias (OPD) entre 1992 e 2014.
+Reduzi também um banco de dados com mais de 100 mil imagens obtidas no Observatório do Pico dos Dias (OPD) entre 1992 e 2014.
 Já no Doutorado, neste mesmo âmbito, reduzi 810 observações feitas no European Southern Observatory (ESO) em 24 noites utilizando o detector mosaico CCD Wide Field Imager (WFI).
@@
 
-Com o objetivo ver o potencial das nossas posições em melhorar as órbitas dos satélites irregulares, analizamos os offsets das nossas posições com respeito às efemérides do JPL.
+Com o objetivo ver o potencial das nossas posições em melhorar as órbitas dos satélites irregulares, analisamos os offsets das nossas posições com respeito às efemérides do JPL.
 Tomando Carme como exemplo, plotamos os offsets médios das efemérides para cada noite na Fig.
-\ref{Fig: carme_anom} e suas respectivas dispersões (erro de barra 1$\sigma$) em função da anomalia verdadeira em ascenção reta e declinação.
+\ref{Fig: carme_anom} e suas respectivas dispersões (erro de barra 1$\sigma$) em função da anomalia verdadeira em ascensão reta e declinação.
 A figura mostra claramente um erro sistemático em declinação.
@@
 Uma vez que suas efemérides não são muito precisas, predizer e observar ocultações estelares é muito difícil e nenhuma observação de um evento desses para um satélite irregular ainda foi relatado na literatura.
-Um projeto nesse sentido está sendo encaminhado e será discutido na seção \ref{Sec: Irregulares}.
 
-\section{Netuno e Tritão}
-\label{Sec: Netuno-astrometria}
-
-\chapter{Ocultações Estelares}
-\label{Cap: ocultacoes}
-
-\section{Noções Gerais}
-
-\indent \indent Ocultação estelar é um evento que ocorre quando um objeto do sistema solar passa na frente de uma estrela de forma que o brilho da estrela é completamente ou parcialmente apagado durante um intervalo de tempo.
-A variação de brilho da estrela pode ser medido fazendo-se observações com exposições curtas.
-O produto gerado é uma curva de luz luminosidade x tempo.
-Essa técnica pode ser utilizada para se obter parâmetros físicos dos objetos ocultantes como tamanho, densidade, detecção da presença de atmosfera, satélites ou anéis, etc \citep{Elliot1979}.
-
-A Fig.
-\ref{Fig: occ-scheme} mostra uma curva de luz esquemática de uma ocultação estelar central por um planeta com atmosfera e anéis.
-Ao ocultar a estrela, os anéis passarão primeiro, a densidade do anel, tamanho e tempo de exposição determinará a queda de brilho da estrela.
-Ao chegar no corpo principal, caso não possua atmosfera, a queda de brilho será abrupta, caso possua, a queda de brilho será mais suave devido à refração da luz da estrela na atmosfera dificultando a determinação da superfície do objeto.
-
-\begin{figure}[h]
-\begin{centering}
-\includegraphics[scale=0.75]{figuras/stellarocc.png} 
-\label{Fig: occ-scheme}
-\caption{Curva de luz esquemática de uma ocultação estelar por um planeta com atmosfera e anéis. Fonte: \cite{Elliot1979}}
-\end{centering} 
-\end{figure}
-
-A suavidade da curva portanto será determinada pelo gradiente de temperatura, pressão e densidade da atmosfera, presença de nuvens, etc.
-A luz da estrela nunca será completamente apagada quando ocultada por um objeto com atmosfera devido à refração da luz na atmosfera onde parte dela convergirá na direção do observador.
-Caso a ocultação seja central para o observador haverá um aumento de brilho no meio do evento pois a convergência da luz será máxima.
-A atmosfera, no caso, agiria como uma lente como mostrado na Fig.
-\ref{Fig: occ-scheme}.
-
-Quando o evento é observado por diversos telescópios espalhados, cada observador verá a ocultação em latitudes diferentes.
-Essas diversas curvas de luz permitirá reconstruir a forma do objeto, possivelmente identificando estruturas superficiais como montanhas e crateras.
-Como a estrela está muito distante, podendo ser considerada quase pontual, a sombra do objeto durante a ocultação é muito próximo ao seu tamanho real.
-
-Para se observar esses eventos, portanto, é necessário uma campanha grande com observadores espalhados por várias regiões para cobrir a sombra do objeto.
-A precisão nas posições da estrela e do objeto são cruciais pois o evento ocorre em instantes específicos e passará por regiões específicas da Terra alheias à escolha do observador.
-
-%Since their ephemerides are not very precise, predicting and observing stellar occultations are very difficult, and no observation of such an event for an irregular satellite is found in the literature.
-The precise star positions to be derived by the ESA astrometry satellite Gaia \citep{deBruijne2012} will render better predictions with the only source of error being the ephemeris.
-The positions derived from our observations can be used in new orbital numerical integrations, generating more precise ephemerides.
-
-%The power of stellar occultations for observing relatively small diameter solar system objects is supported by recent works, such as the discovery of a ring system around the Centaur (10199) Chariklo \citep{Braga-Ribas2014}.
-Once irregular satellites start to be observed by this technique, it will be possible to obtain their physical parameters (shape, size, albedo, density) with unprecedented precision.
-For instance, in this case, sizes could be obtained with kilometer accuracy.
-The knowledge of these parameters would in turn bring valuable information for studying the capture mechanisms and origin of the irregular satellites.
-
-\section{Centauros e TNOs}
-\label{Sec: TNO-occ}
-
-\indent \indent Como dito no Cap.
-\ref{Cap: intro}, certos objetos do sistema solar exterior podem ser considerados fósseis do sistema solar.
-Dentre esses, os TNOs, por estarem mais distantes, são os que guardam as informações mais antigas.
-Com o objetivo de obter os parâmetro físicos desses objetos, para melhor caracterizar as propriedades básicas dessas populações, uma grande campanha observacional é realizada para se observar ocultações estelares por TNOs e Centauros.
-Essa campanha é liderada por pesquisadores brasileiros, franceses e espanhóis e tem colaboração com observadores espalhados por todo o mundo.
-
-Vários objetos já foram observados pelo grupo, sendo que o Sistema Plutão-Caronte conta com várias ocultações observadas além de um trabalho astrométrico minucioso baseado em 19 anos de observações no OPD, mais observações no ESO \citep{Rossi2014}.
-As observações bem sucedidas de ocultações estelares resultantes desse trabalho de predição já geraram diversas publicações, inclusive na revista Nature, por exemplo \cite{Sicardy2011} e \cite{BragaRibas2014} nas quais contribuí em caráter observacional.
-A ocultação de Plutão de 04 de Maio de 2013, a qual observei, gerou duas publicações \citep{Olkin2015, DiasOliveira2015}.
-
-\cite{Sicardy2011} identificaram, a partir de uma ocultação de 06 de Novembro de 2010, que Eris é um planeta anão menor do que se esperava.
-Devido a seu brilho e distância acreditava-se que ele fosse muito maior, mas na verdade ele tem um albedo geométrico no visível de $p_v=0.96^{+0.09}_{-0.04}$ e seu tamanho é de $1163 \pm 6$ km, muito próximo ao tamanho de Plutão.
-O alto albedo de Eris pode ser relacionado à presença de uma atmosfera colapsada devido ao fato dele estar se aproximando de seu afélio.
-
-\cite{BragaRibas2014} observaram uma ocultação do Centauro Chariklo que ocorreu dia 03 de Junho de 2013.
-O ocultação revelou que o objeto possui um sistema de anéis nunca visto antes em um objeto tão pequeno.
-Até o momento, conhecia-se apenas anéis em torno dos planetas gigantes.
-Essa descoberta levanta muitas questões sobre formação dos anéis, estabilidade e tempo de vida e nos leva a crer que pode haver mais objetos com anéis no sistema solar, ao menos nessa região.
-
-\cite{Olkin2015} determinaram a partir de observações de ocultações por Plutão que a atmosfera do objeto não colapsa durante os 248 anos de sua órbita, como acontece com Eris.
-Modelos anteriores previam um colapso de sua atmosfera devido a Plutão receber três vezes menos luz do Sol em afélio que em seu periélio.
-
-\cite{DiasOliveira2015}, também a partir de ocultações de Plutão, utilizaram modelos para obter os perfis de temperatura, pressão e densidade da atmosfera do objeto.
-Os modelos ajustam perfeitamente os perfis de temperatura para as duas ocultações utilizadas assumindo uma atmosfera esfericamente simétrica.
-Foram usadas as curvas de luz de melhor resolução até hoje obtidas para Plutão, sendo a melhor obtida em  Paranal em 18 de Julho de 2012, usando a câmera NAOS-CONICA (NACO) ligada ao 8.2-m “Yepun” Very Large Telescope (VLT) do ESO a uma taxa de 5 frames por segundo na banda H.
-Por fim, foi obtido que a pressão da atmosfera aumentou cerca de 6\% entre 2012 e 2013 significando que a atmosfera de Plutão ainda está se expandindo e confirmando o trabalho de \cite{Olkin2015}.
-
-No dia 29 de Junho de 2015 haverá uma ocultação de Plutão de máxima importância.
-Será a última antes da chegada da sonda \textit{New Horizons} no sistema e poderá ser observada na Austrália.
-Observada a ocultação, os modelos de perfil de atmosfera para Plutão poderão ser calibrados com os dados da \textit{New Horizons} para serem utilizados em futuras ocultações e estudar a evolução da atmosfera do planeta anão.
-
-Em 29 de Abril de 2014, uma ocultação de Chariklo foi observada na África.
-Eu trabalhei com a redução de três curvas de luz (Gifberb, SAAO e Springbok).
-Um fato interessante dessa ocultação foi que a estrela era uma binária não catalogada.
-Como a segunda estrela era bem mais fraca, acreditava-se que poderia ser algum material próximo à Chariklo.
-
-Essa ocultação foi importante para obter com total certeza a direção do pólo de Chariklo.
-Além disso, as câmeras utilizadas eram de alta qualidade possibilitando obter que pelo menos um dos anéis é opaco.
-Uma das cordas, fortuitamente, passou contígua ao anel observando apenas o anel externo.
-Os resultados ainda não foram publicados.
-
-Outro evento que observei foi uma ocultação de 2003VS2 dia 07 de Novembro de 2014 no OPD.
-As variações do céu não permitiram uma boa observação de forma que a curva de luz estava muito ruidosa, mas boa o suficiente próximo ao instante do evento.
-Um trabalho foi feito para identificar alguma ocultação mas nada pôde ser extraído.
-Nesse evento foram obtidas 3 cordas positivas e 7 negativas.
-A redução preliminar revelou que 2003VS2 é possui um corpo levemente alongado com aproximadamente 600 km de diâmetro.
-O trabalho ainda está em análise.
-
-\section{Ceres}
-\label{Sec: Ceres}
-
-\indent \indent Apesar de Ceres não ser um objeto do sistema solar exterior, ele é o único planeta-anão no sistema solar interno e, por isso, é um objeto de grande importância e seu estudo pode ter grande impacto na formação e evolução do sistema solar.
-Na verdade, foi proposto que a origem de Ceres pode ser como um objeto transnetuniano \citep{McKinnon2012}, espalhado posteriormente para o cinturão principal de asteroides devido à migração dos planetas gigantes predito pelo Modelo de Nice \citep{Gomes2005}.
-Mesmo que ele tenha sido formado próximo à sua localização atual, a história dinâmica do sistema solar deve ter deixado sua assinatura em Ceres.
-
-Contendo aproximadamente um quinto de toda a massa do cinturão de asteroides, espera-se que Ceres esteja em equilíbrio gravitacional e seja, portanto, um elipsóide Maclaurin ou Jacobi.
-De fato, observações diretas de Ceres com a utilização de ótica adaptativa indica que ele é um esferóide achatado nos pólos \citep{Drummond2014}.
-O conhecimento preciso de seu tamanho e forma é extrema importância para modelos de densidade, estrutura interna e diferenciação.
-
-A primeira ocultação estelar por Ceres foi observada em 1984 \citep{Millis1987} e determinou seu tamanho com precisão de alguns quilômetros em uma época que as incertezas eram, normalmente, dez vezes maiores.
-Devido ao brilho aparente de Ceres ser alto, comparado à maioria dos asteroides, estrelas capazes de causar uma queda de magnitude detectável quando ocultadas são limitadas às mais brilhantes.
-Por exemplo, depois do evento de 1984, apenas 4 ocultações estelares por Ceres foram observadas \citep{Dunham2014}.
-Duas delas tiveram apenas duas cordas cada que não foram suficientes para prover resultados acurados\footnote{Esses eventos ocorreram em 22 de Agosto de 1994 e 30 de Outubro de 2010.}.
-Os dois restantes, que ocorreram em 17 de Agosto de 2010 e 25 de Outubro de 2013, foram trabalhos por mim em colaboração com o grupo do Rio \citep[aceito]{GomesJunior2015-Ceres}.
-
-Os dois eventos foram preditos por Steve Preston\footnote{Predições publicadas em \url{http://asteroidoccultation.com}.} para a IOTA (International Occultation Timing Association), durante predições de rotina de ocultações de asteroides de estrelas brilhantes.
-Os caminhos das sombras podem ser visualizados na Fig.
-\ref{Fig: Ceres-map}.
-
-\begin{figure}[h]
-\begin{centering}
-\subfigure[Ocultação de 2010]{\includegraphics[scale=0.39]{figuras/Ceres_2010.eps}\label{Fig: Ceres-map-2010}}
-\subfigure[Ocultação de 2013]{\includegraphics[scale=0.39]{figuras/Ceres_2013-zoom.eps}\label{Fig: Ceres-map-2013}}
-\caption{Reconstrução pós-ocultação do caminho da sombra de Ceres na Terra para os eventos de 17 de Agosto de 2010 (a) e 25 de Outubro de 2013 (b).
-Os pontos em azul são os sítios que observaram os eventos.
-a) O ponto grande vermelho é a máxima aproximação geocêntrica às 22:40:25 UT.
-Os pequenos representam o centro da sombra separados por um minuto.
-b) Visão superior da ocultação sobre os sítios que observaram o evento de 25 de Outubro de 2013.
-Os pontos vermelhos são os centros da sombra separados por 15 segundos.
-Nos dois eventos a sombra se move da esquerda para a direita.
-\label{Fig: Ceres-map}}
-\end{centering}
-\end{figure}
-
-\subsection{Ocultação de 2010}
-\label{Subsec: 2010-occ}
-
-\indent \indent Em 17 de Agosto de 2010 Ceres ocultou a estrela TYC 6833-163-1 (UCAC4 313-111823), cuja magnitude é $V = 11.55$ e tem posição no ICRS para a data do evento baseada no catálogo UCAC4 \citep{Zacharias2013}:
-
-\begin{equation}
-\left\{ 
-  \begin{array}{l l}
-    \alpha = 17^{h}18^{m}29\fs 0085\\
-    \delta = -27\degr 26\arcmin 38\farcs 867
-  \end{array}
-\right.
-\label{Eq: 2010-coord}
-\end{equation}
-
-\indent O evento foi observado no Brasil a partir de cinco diferentes sítios (ver Fig.
-\ref{Fig: Ceres-map-2010}).
-Destes, 4 obtiveram cordas positivas enquanto UFSC teve uma corda negativa.
-Das positivas, a observação proveniente do INPE iniciciou-se após o início do evento devido a dificuldades técnicas e, portanto, apenas a emersão da curva de luz foi detectada.
-
-Uma das características mais importantes desse evento foi a velocidade com que ocorreu (apenas 3.9 km s$^{-1}$) acarretando que mesmo exposições de poucos segundos representariam resoluções espaciais significantes.
-
-Todas as observações foram feitas com a utilização de CCDs.
-As curvas de luz de cada observação foram obtidas das imagens FITS com a utilização do pacote PRAIA \citep[Plataforma de Redução Astrométrica de Imagens Astronômicas,][]{Assafin2011}.
-As curvas foram normalizadas para o fluxo da estrela mais Ceres, uma vez que eles estavam indistinguíveis logo antes e depois da ocultação.
-Por fim, elas foram normalizadas pelo ajuste de uma curva polinomial (de primeira ou segunda ordem) fora da queda de fluxo assim fixando em 1 a razão de fluxo fora da ocultação.
-
-Os instantes de ingresso e egresso foram obtidas de cada curva de luz ajustando-se um modelo de poço quadrado levando em consideração a difração de Fresnel, a banda do CCD, o diâmetro aparente da estrela e o tempo de exposição utilizado \citep[ver][]{Widemann2009, Braga-Ribas2013}.
-
-O menor tempo de integração usado nas observações positivas foi de $1.0s$, que corresponde a aproximadamente $3.9km$ no plano do céu.
-Portanto, o erro na determinação dos instantes de ingresso e egresso é dominado principalmente pelo tempo de integração, não pela difração de Fresnel ou diâmetro da estrela, ambos da ordem de algumas centenas de metros para esse evento.
-
-O ajuste dos dados da ocultação consiste em minimizar uma função dep $\chi^{2}$ clássica para cada curva de luz, como descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}.
-Os parâmetros livres para ajustar são os instantes de ingresso e egresso que fornece o valor mínimo de $\chi^{2}$ ($\chi^{2}_{min}$).
-O melhor ajuste das curvas de luz para a ocultação de 2010 está mostrado na Fig.
-\ref{Fig: Ceres-2010-curves}.
-
-\begin{figure}[h]
-\begin{centering}
-\subfigure[Ocultação de 2010]{\includegraphics[scale=0.51]{figuras/Ceres_2010_fluxratio.eps}\label{Fig: Ceres-2010-curves} }
-\subfigure[Ocultação de 2013]{\includegraphics[scale=0.51]{figuras/Ceres_2013_fluxratio.eps}\label{Fig: Ceres-2013-curves} }
-\caption{Curvas de luz normalizadas das cordas positivas dos eventos.
-As curvas estão desviadas por um fator de 0.2 (a) e 1.0 (b) para melhor visualização.
-As linhas pretas são os melhores ajustes com o modelo de poço quadrado.
-As linhas vermelhas são os melhores ajustes com o modelo de poço quadrado, porém levando em conta a difração de Fresnel, o diâmetro da estrela e o tempo de exposição.
-Os instantes médios de cada curva não coincidem devido às diferentes longitudes dos sítios.
-A curva de luz de Brookline (b) está desviada por um fator de -64 s como explicado no texto.
-\label{Fig: Ceres-curves}}
-\end{centering}
-\end{figure}
-
-A metodologia usada para analizar o perfil de Ceres a partir das observações é o mesmo descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}.
-Cada combinação de posição do sítio, instantes de ingresso e egresso, junto com as coordenadas da estrela e as efemérides de Ceres, correspondem a um ponto no plano do céu.
-A coleção de todos esses pontos determina o limbo aparente de Ceres.
-
-Adotamos um modelo elíptico para o perfil do limbo, resultante da projeção de um esferóide com achatamento nos pólos no plano do céu.
-Essa escolha é suportada pelo trabalho de \cite{Drummond2014}, por meio de imagem direta de Ceres.
-Dessa forma, nós temos $N=7$ extremidades das cordas para ajustar $M=5$ parâmetros que definem uma elipse: semi-eixo maior e semi-eixo menor aparentes ($a^\prime$ and $b^\prime$, respectivamente), ângulo de posição $P$ do seu semi-eixo maior e as posição $(f_c,g_c)$ do seu centro com respeito à estrela ocultada.
-O semi-eixo maior {$a^\prime$} é equivalente ao raio equatorial $R_{equa}$ do elipsoide.
-
-As coordenadas $f_{c}$ e $g_{c}$, em quilômetros, foram calculadas usando a efeméride de Ceres JPL\#33 \citep{Giorgini1996} e a posição da estrela ocultada.
-Elas são positivas na direção Leste e Norte celestes, respectivamente.
-O ângulo de posição $P$ é contado positivamente a partir do norte celeste local em direção ao leste celeste.
-O achatamento aparente pode ser definido por $\epsilon^\prime = 1 - (b^\prime/a^\prime)$.
-O melhor ajuste é obtido minimizando uma função de $\chi^{2}_{r}$ reduzido, onde definimos o número de graus de liberdade do problema como $\mathcal{N} \equiv N - M$.
-Todos os procedimentos que permitem a determinação das barras de erro dos paramêtros físicos podem ser encontradas em \cite{Braga-Ribas2013}.
-
-Duas possíveis soluções foram consideradas para o ajuste do limbo.
-O primeiro, que chamamos de solução nominal, consiste em determinar os cinco parâmetros que caracterizam uma elipse a partir dos sete contatos observados.
-A segunda solução consiste em calcular o ângulo de posição $P$ a partir das coordenadas do pólo de Ceres obtidas por \cite{Drummond2014} ($\alpha_{p} = (287 \pm 3) \degr$, $\delta_{p} = (+64 \pm 3) \degr$ no ICRS) e da efeméride de Ceres no instante da ocultação.
-Chamamos de solução de pólo fixo.
-
-Para o evento de 2010, a solução nominal teve como melhor ajuste $\chi^2_{r,min} = 0.24$, que podem ser interpretadas como as barras de erro estarem superestimadas com respeito à boa qualidade do ajuste.
-Porém, como o problema tem somente dois graus de liberdade, $\chi^2_{r,min}$ relativamente pequenos são aceitáveis.
-Os resultados obtidos para o diâmetro equatorial, achatamento, ângulo de posição e coordenadas do centro esao apresentadas na segunda coluna da tabela \ref{Tab: resultados}.
-
-\begin{table}[h]
- \begin{centering}
-  \caption{Resultados do ajuste de limbo de Ceres com os dados dos eventos de 2010 e 2013.\label{Tab: resultados}}
-  \begin{tabular}{@{}lcccc}
-  \hline
-     Solução & 2010/Nominal & \textbf{2010/Pólo fixo} & 2013/Nominal & 2013/Pólo fixo \\
-\hline
-Diam. equat.  (km) & 982 $\pm$ 14 & \textbf{972 $\pm$ 6}  & 971 $\pm$ 7  & 971 $\pm$ 7\\
-%Equatorial radius (km) & 491 $\pm$ 7  & \textbf{486 $\pm$ 3}  & 485.5 $\pm$ 3.5 & 485.5 $\pm$ 3.5\\
-Achatamento        & 0.08 $\pm$ 0.03 & \textbf{0.08 $\pm$ 0.03} & 0.08 $\pm$ 0.04 & 0.08 $\pm$ 0.04\\
-Âng. de pos. (deg)   & 5 $\pm$ 10    & \textbf{12 $\pm$ 3} (*)& 22 $\pm$ 5    & 25 $\pm$ 3 (*)\\
-$f_c$ (km)             & 97 $\pm$ 9   & \textbf{102 $\pm$ 5}   & 77 $\pm$ 6    & 78 $\pm$ 6\\
-$g_c$ (km)             & 16 $\pm$ 15  & \textbf{21 $\pm$ 11}  & 13 $\pm$ 16   & 13 $\pm$ 16\\
-$\chi^2_{r,min}$       & 0.24          &  \textbf{0.42}         & 1.27          & 1.27\\
-\hline
-\end{tabular}
-\textbf{Notas:} Em negrito destaca-se a melhor solução obtida.
-As barras de erro estão no nível de 1$\sigma$.
-O diâmetro polar ($D_{pol}$) pode ser facilmente calculado a partir de $D_{pol}=D_{equa}(1 - \epsilon)$.
-(*) Ângulos de posição derivados a partir das coordenadas do pólo de Ceres determinapas por \cite{Drummond2014}.
-\end{centering}
-\end{table}
-
-Como pode ser visto, o parâmetro com a maior incerteza é o ângulo de posição cobrindo um intervalo de 20$\degr$.
-Claramente, fixar as coordenadas do pólo pode melhorar a solução.
-Por fim, a correção do achatamento devido ao ângulo do aspecto polar está dentro da barra de erro 1$\sigma$ e não tem relevância estatística, dessa forma $\epsilon = 0.08 \pm 0.03$.
-
-No momento da ocultação, as coordenadas do pólo de Ceres correspondiam a um ângulo de posição $P = (12 \pm 3)\degr$.
-Explorar o espaço de parâmetros restringindo a elipses com ângulo de posição dentro deste intervalo resulta na solução de pólo fixo.
-O parâmetros físicos do melhor ajuste estão mostrados na Tab.
-\ref{Tab: resultados} enquanto a solução está esquematizada na Fig.
-\ref{Fig: Ceres-limb-2010}.
-
-Essa solução corresponde ao limite superior da barra de erro 1$\sigma$ da solução nominal para $P$.
-Por outro lado, ela obtem os menores valores para o diâmetro equatorial, melhorando sua determinação por um fator de 2.
-
-\begin{figure}[h]
-\begin{centering}
-\subfigure[Ocultação de 2010]{\includegraphics[scale=0.35]{figuras/Ceres_2010_sphere.eps}\label{Fig: Ceres-limb-2010} }
-\subfigure[Ocultação de 2013]{\includegraphics[scale=0.35]{figuras/Ceres_2013_sphere.pdf}\label{Fig: Ceres-limb-2013} }
-\caption{Melhores ajustes elípticos para as cordas das ocultações de 2010 e 2013.
-As setas indicam a direção de movimento, as linhas azuis são as cordas observadas, os segmentos vermelhos são as barras de erro dos ingressos, egressos e centro da ocultação em $1\sigma$.
-A linha verde em (a) é uma corda negativa.
-Os instantes marcados em verde em (b) não foram utilizados para o ajuste, como descrito no texto.
-\label{Fig: Ceres-limb}}
-\end{centering}
-\end{figure}
-
-\subsection{Ocultação de 2013}
-\label{Subsec: 2013-occ}
-
-\indent \indent Em 25 de Outubro de 2010, Ceres ocultou a estrela TYC 865-911-1 (UCAC4 496-058191), de magnitude $V = 10.05$.
-Baseada no UCAC4 \citep{Zacharias2013}, sua posição ICRS para a data da ocultação é:
-
-%
-\begin{equation}
-\left\{ 
-  \begin{array}{l l}
-    \alpha = 11^{h}57^{m}52\fs7641\\
-    \delta = +09\degr 07\arcmin 49\farcs835
-  \end{array}
-\right.
-\end{equation}
-%
-Esse evento foi observado na costa Leste dos Estados Unidos logo antes do amanhecer, como mostrado na Fig. \ref{Fig: Ceres-map-2013}.
-
-Nove cordas positivas foram obtidas em vários sítios (ver Fig.
-\ref{Fig: Ceres-map-2013}).
-Cada estação foi equipada com uma câmera de vídeo com tempo de leitura desprezível.
-Isso é particularmente importante, já que a velocidade da sombra de Ceres para esse evento foi de 42.6 km s$^{-1}$, muito mais rápido que o evento de 2010.
-
-Durante o evento Ceres estava muito baixo no céu com alturas entre $15\degr$ (Winchester) e $20\degr$ (Hampton).
-Forte cintilação era esperada e, combinada com o curto tempo de integração e baixa diminuição de brilho, resultou em curvas de luz ruidosas e assim grandes incertezas nos instantes de imersão e emersão.
-
-Todos os vídeos foram convertidos para imagens FITS e a fotometria foi obtidia via \textsc{praia} \cite{Assafin2011}.
-As curvas de luz foram normalizadas por uma estrela de referência quando havia uma estrela no campo.
-
-Para reduzir o ruído, os dados foram binados por grupos de cinco imagens -- com exceção de Greenbelt, onde gupos de dez imagens foram utilizadas.
-Esse procedimento o tempo de integração efetivo por um fator de 5 (ou 10).
-Da mesma forma que para o evento de 2010, uma normalização adicional por um polinômio foi aplicada.
-
-Os instantes de ingresso e egresso da ocultação foram obtidos pelo mesmo procedimento descrito na seção \ref{Subsec: 2010-occ}.
-Uma vez que o tempo de integração efetivo usado (0.17 s) representa aproximadamente 7 km no plano do céu, e a escala de Fresnel e  o diâmetro da estrela estão novamente na ordem de centenas de metros, o erro da determinação dos instantes de imersão e emersão são dominados principalmente pelo tempo de integração, da mesma forma que para o evento de 2010.
-Os melhores ajustes para as curvas de luz da ocultação estão mostrados na Fig.
-\ref{Fig: Ceres-2013-curves}.
-
-Uma comparação entre os tempos obtidos pela estação de Brookline e o restante mostrou que esses têm um atraso de aproximadamente 64 s. Dessa forma, não utilizamos os tempos de Brookline na análise.
-
-Perfis elípticos foram ajustados para todas as cordas restantes.
-pelo mesmo procedimento descrito na seção \ref{Subsec: 2010-occ}.
-O resultado foi $\chi^2_{r,min} = 13$ , sugerindo que um modelo elíptico não é satisfatório para os dados.
-De fato, ao olharmos para a Fig.
-\ref{Fig: Ceres-limb-2013} vemos que a corda de Varina adiantada com respeito às outras.
-Como nessa estação o tempo não foi inserido diretamente nos frames do vídeo é possível que essa diferença seja oriunda de um eventual problema da correspondência entre os tempos do camcorder e do GPS.
-
-A imersão gravada em Owings também parece atrasada com respeito às cordas próximas (ver Fig.
-\ref{Fig: Ceres-limb-2013}).
-Essa corda tem aproximadamente o mesmo tamanho da corda de Mechanicsville, apesar de estarem separadas por cerca de 100 km.
-Diferentemente de Varina, essa estação teve os tempos inseridos em cada frame do vídeo o que torna mais difícil justificar um problema de tempo.
-Outras possibilidades seriam uma má determinação dos instantes de ingresso e egresso dessa curva ou uma característica do relevo de Ceres.
-
-Em um segundo ajuste, não consideramos as cordas de Brookline, Varina e Owings.
-O ajuste dos cinco parâmetros que definem uma elipse para os doze contatos resultou em $\chi^2_{r,min} = 1.27$, indicando  que está em bom acordo com os dados observados dentro das barras de erro.
-Essa é a solução mostrada na Fig.
-\ref{Fig: Ceres-limb-2013}, onde podemos ver que o tamanho da corda de Brookline é compatível com o modelo.
-
-Os resultados das soluções nominais e de pólo fixo não são significantemente diferentes.
-Isso se deve ao fato da barra de erro do ângulo de posição da solução nominal (que é muito menor que o da solução nominal de 2010) ser muito similar à barra de erro do ângulo de posição da solução de pólo fixo ($P = (25 \pm 3)\degr$).
-Os dois resultados estão apresentados nas colunas 4 e 5 da tabela \ref{Tab: resultados}.
-
-Sobre a hipótese do atraso observado na imersão da curva de luz de Owings ser associado a uma característica topográfica, o contato gravado corresponderia a uma elevação negativa de 31 $\pm$ 4 km com respeito à elipse de melhor ajuste.
-Porém modelos teóricos prevêem que relevos em Ceres não devem ser maiores que 10--20 km \citep{Johnson1973}, enquanto dados observacionais limitam o contorno em 18 km \citep{Carry2008}.
-Imagens mais recentes da sonda \textit{Dawn} também revelam uma superfície mais suave.
-Portanto, a associação da imersão em Owings com um relevo é improvável.
-
-\subsection{Discussão}
-
-\indent \indent Os resultados apresentados na tabela \ref{Tab: resultados} mostram um acordo entre os parâmetros físicos obtidos nas duas ocultações, especialmente no diâmetro equatorial.
-As diferenças ocorrem basicamente nos tamanhos das barras de erro e podem ser justificadas pelas particularidades de cada conjunto de dados.
-
-O evento de 2010, por exemplo, teve somente sete contatos, porém bem distribuídos sobre o disco de Ceres.
-Por outro lado, o evento de 2013 teve cinco contatos a mais, todavia concentrados em certas regiões do corpo.
-Em particular, a ausência de cordas prómimas ao pólo sul fez seu achatamente ser pior determinado para o evento de 2013 que para o evento de 2010.
-Mesmo nossa melhor medida de achatamento, $\epsilon=0.08 \pm 0.03$, tem alta incerteza se comparado com outros valores publicados na literatura, como mostra a tabela \ref{Tab: Ceres-final}.
-
-Como foi mostrado, usar as coordenadas do pólo de Ceres determinadas por \cite{Drummond2014} para limitar o ângulo de posição não foi um procedimento eficiente para o evento de 2013.
-Por outro lado, fixar o ângulo de posição para a ocultação de 2010 reduziu as barras de erro dos outros parâmetros (com exceção do achatamento).
-Por fim, Esse procedimento resultou em um excelente acordo entre os raios equatoriais obtidos para ambos os eventos.
-
-Uma comparação do diâmetro equatorial de Ceres medido por diferentes técnicas está mostrado na tabela \ref{Tab: Ceres-final}.
-Vemos um acordo entre os nossos resultados e aqueles obtidos por imageamente direto do Hubble Space Telescope (HST) \citep{Thomas2005}, do Keck Observatory e do ESO VLT \citep{Drummond2014}.
-O menor valor, reportado por \cite{Carry2008}, pode ser justificado pelo fato desse estudo não levar em conta o efeito de escurecimento de bordo.
-
-O evento de 1984 \citep{Millis1987} é a única outra ocultação que podemos comparar nosssos resultados.
-As medidas do diâmetro não se acordam dentro de 2$\sigma$.
-É difícil dizer com certeza as razões dessa divergência.
-Uma forma de clarificar o problema seria redeterminar os instantes de imersão e emersão das curvas de luz originais usando a mesma metodologia aplicada nesse trabalho.
-Infelizmente, não temos acesso aos dados da curva de luz original do evento de 1984.
-
-A sonda da NASA \textit{Dawn} poderá responder esses questões que são importantes não só no conhecimento do próprio Ceres, mas também para calibrar todas as técnicas usadas até agora para o estudo das propriedades físicas dos pequenos objetos do Sistema Solar, como as ocultações estelares.
-
-\begin{table}[h]
-  \caption{Diâmetro equatorial e achatamento de Ceres \label{Tab: Ceres-final}}
-  \begin{centering}
-  \begin{tabular}{@{}lccc}
-  \hline
-     Diâmetro Equatorial (km) & Achatamento & Método & Ref. \\
-\hline
-$972 \pm 6$  & $0.08  \pm 0.03$  & Occultação & 1\\
-$967 \pm 10$ & $0.078 \pm 0.015$ & Keck+VTL    & 2\\
-$959 \pm 5$  & $0.074 \pm 0.007$ & Keck        & 3\\
-$975 \pm 4$  & $0.067 \pm 0.005$ & HST         & 4\\
-$959 \pm 5$  & $0.05  \pm 0.01$  & Occultação & 5\\
-\hline
-\end{tabular}
-\par\end{centering}
-\textbf{Referências.} 1: \cite{GomesJunior2015-Ceres}. 2: \cite{Drummond2014}. 3: \cite{Carry2008}. 4: \cite{Thomas2005}. 5: \cite{Millis1987}.
-\end{table}
-
-\section{Satélites Irregulares}
-\label{Sec: Irregulares}
-
-\subsection{Predição das ocultações}
-\label{Subsec: Irr-predic}
+\section{Predição das ocultações}
+\label{Sec: Irr-predic}
 
@@
 
-As predições foram feitas utilizando as posições de estrelas dadas no cataálogo UCAC4.
+As predições foram feitas utilizando as posições de estrelas dadas no catálogo UCAC4.
 Tomou-se como referência as efemérides do JPL, corrigidas com as posições de satélites obtidas, de acordo com o descrito a seguir.
@@
 Por outro lado, os satélites de Júpiter estão muito mais perto que os TNOs e como o erro astrométrico é uma medida angular, consequentemente, o erro da sombra da ocultação projetada na Terra será muito menor que para TNOs.
-Assim, as latituds a serem cobertas para que uma ocultação por satélite irregular de Júpiter seja detectada corresponde a poucas centenas de quilômetros.
+Assim, as latitudes a serem cobertas para que uma ocultação por satélite irregular de Júpiter seja detectada corresponde a poucas centenas de quilômetros.
 
-\subsection{Testes}
-\label{Subsec: Irr-testes}
+\section{Testes}
+\label{Sec: Irr-testes}
 
@@
 O offset, portanto, seria considerado estável dentro desses poucos dias uma vez que ele não se move muito em sua órbita.
-Porém, nem sempre é possível observar o alvo com antecedêndia.
+Porém, nem sempre é possível observar o alvo com antecedência.
 
@@
 Para outros satélites, ajuste em primeira ordem do seno e do cosseno da anomalia verdadeira também representam razoavelmente bem os offsets em declinação.
-Para os offsets em ascenção reta, o ajuste com três parâmetros não é suficiente.
+Para os offsets em ascensão reta, o ajuste com três parâmetros não é suficiente.
 
@@
 Até o momento, dois testes de ocultações foram feitos, um de Himalia que ocorreu dia 03 de Março de 2015 e o segundo de Elara que ocorreu dia 30 de Março de 2015.
-Para cada evento quatro mapas foram gerados: o primeiro com as posições nominais da estrela e do satélite para o instante previsto; o segundo com o offset calculado como descrito anteriomente; o terceiro com os offsets da estrela e do satélite a partir de observações feitas alguns dias antes da ocultação quando os dois objetos estavam separados (campos diferentes); e o quarto a partir de observações feitas com menos de 24 horas de diferença da data prevista com a estrela e o satélite próximos no mesmo campo.
+Para cada evento quatro mapas foram gerados: o primeiro com as posições nominais da estrela e do satélite para o instante previsto; o segundo com o offset calculado como descrito anteriormente; o terceiro com os offsets da estrela e do satélite a partir de observações feitas alguns dias antes da ocultação quando os dois objetos estavam separados (campos diferentes); e o quarto a partir de observações feitas com menos de 24 horas de diferença da data prevista com a estrela e o satélite próximos no mesmo campo.
 
@@
 \subfigure[Obs. dia 03 de Março (mesmo campo)]{\includegraphics[scale=0.16]{figuras/Himalia4_2015-03-03T00:39:15.png}   \label{Fig: occ-Himalia-off03mar}}
-\caption{Predições para Himalia.
-O ponto grande em vermelho mostra o ponto de máxima aproximação geocêntrica da sombra na Terra, os pontos vermelhos menores são os centros da ocultação separados por um minuto, as linhas retas são os limites das sombras dado o tamanho estimado do objeto e as linhas tracejadas equivalem a uma diferença de 40 mas do centro da sombra representando o erro estimado da predição.
+\caption{Predições para Himalia: O ponto grande em vermelho mostra o ponto de máxima aproximação geocêntrica da sombra na Terra, os pontos vermelhos menores são os centros da ocultação separados por um minuto, as linhas retas são os limites das sombras dado o tamanho estimado do objeto e as linhas tracejadas equivalem a uma diferença de 40 mas do centro da sombra representando o erro estimado da predição.
 (a) é o mapa utilizando apenas as posições nominais da estrela e do satélite.
-(b) mostra a sombra dado um offset estimado para a posição de Himalia dada sua anomalia verdadeira segundo um ajuste feito em cima dos offsets de efeméride encntrados em \cite{GomesJunior2015-Irregular}.
+(b) mostra a sombra dado um offset estimado para a posição de Himalia dada sua anomalia verdadeira segundo um ajuste feito em cima dos offsets de efeméride encontrados em \cite{GomesJunior2015-Irregular}.
 Em (c) são aplicados offsets às posições da estrela e do satélite a partir de observações feitas em 22 de Fevereiro no telescópio Zeiss.
@@
 \subfigure[Obs. dia 02 de Abril (mesmo campo)]{\includegraphics[scale=0.16]{figuras/Elara4_2015-03-30T01:45:30.png}  \label{Fig: occ-Elara-off02abr}}
-\caption{Predições para Elara.
-Mesmo que a Fig.
+\caption{Predições para Elara: Mesmo que a Fig.
 \ref{Fig: occ-Himalia}, mas para Elara.
@@
 
-\section{Tritão}
-\label{Sec: occ-Tritao}
+\chapter{Netuno e Tritão}
+\label{Cap: Netuno-Tritao}
+
+\section{Introdução}
+\label{Sec: Netuno-introducao}
+
+\section{Astrometria}
+\label{Sec: Netuno-astrometria}
+
+\indent \indent Com o objetivo de estudar os efeitos de maré sobre o sistema Netuno-Tritão, serão obtidos as posições desses objetos em mais de 7000 imagens observadas no OPD desde 1992.
+Uma vez que eles estão muito próximos no céu e Netuno é muito mais brilhante do que Tritão, a posição do satélite pode estar afetada pela luz do planeta.
+Para eliminar essa influência será utilizada uma técnica chamada de "coronografia digital".
+
+O procedimento de coronografia digital faz parte do pacote PRAIA e foi desenvolvido para se observar os satélites principais de Urano durante os fenômenos mútuos de 2008 \citep{Assafin2008}.
+A técnica consiste em computar o perfil de luz do objeto mais brilhante dado uma área.
+A imagem original, então, é subtraída do perfil do objeto gerando uma imagem coronografada.
+
+Para se obter o perfil do objeto, a distância $d$ de cada pixel ($j$,$i$) às coordenadas centrais do baricentro ($x_{c}$,$y_{c}$) da fonte brilhante.
+Para cada anel de 1 pixel de largura a uma distância $d$ do baricentro é feito um histograma de contagens.
+Por fim, o perfil é obtido sendo que, para cada pixel, a contagem é estimada como a moda do histograma de contagens de seu respectivo anel.
+
+No caso, a luz de Netuno será removida diminuindo a sua influência sobre a posição de Tritão. Como subproduto, as coordenadas do baricentro calculadas são uma boa estimativa da posição de Netuno.
+
+Como o tamanho aparente de Netuno ($\approx 2\arcsec$) é da ordem do seeing, uma técnica está sendo desenvolvida para determinar o centro de objetos extensos.
+No caso, supõe-se que, para cada ponto da imagem, a luz é espalhada sobre o CCD como uma gaussiana e a imagem correspondente é a soma de todas estas gaussianas.
+
+\section{Predições ESO/WFI}
+\label{Sec: Netuno-predic}
+
+\section{Ocultação Maio/2008}
+\label{Sec: Netuno-maio2008}
+
+\section{Telemetria de sondas, marés}
+\label{Sec: Netuno-mares}
+
+\indent \indent Para o Doutorado sanduíche, irei trabalhar com as posições do OPD para estudar os efeitos de maré no sistema Netuno-Tritão.
+A vantagem desse banco de dados é sua cobertura temporal e grande número de observações que pode compensar falta de precisão ao se focar em efeitos dinâmicos de longo período \citep{Lainey2012}.
 
 \chapter{Perspectivas}
-\label{Cap: cronograma}
+\label{Cap: perspectivas}
 
-\glsaddall
-\printglossary
+\indent \indent
 

### 2015-06-16 qualify/Qualify-altair.tex
Falta ocultacao de maio de 2008 e espace

@@
 \usepackage{arydshln}
-\usepackage{tablefootnote}
+\usepackage{footnote}
 %\usepackage[alf]{abntcite}
@@
 \begin{document}
-\doutorado
+\qualify
 
@@
 A Fig.
-\ref{Fig: occ-scheme} mostra uma curva de luz esquemática de uma ocultação estelar central por um planeta com atmosfera e anéis.
+\ref{Fig: occ-scheme}.1 mostra uma curva de luz esquemática de uma ocultação estelar central por um planeta com atmosfera e anéis.
 Ao ocultar a estrela, os anéis passarão primeiro, a densidade do anel, tamanho e tempo de exposição determinará a queda de brilho da estrela.
@@
 A atmosfera, no caso, agiria como uma lente como mostrado na Fig.
-\ref{Fig: occ-scheme}.
+\ref{Fig: occ-scheme}.1.
 
@@
 Preferencialmente, esse erro tem que ser menor que o tamanho angular do objeto no plano do céu.
-Por exemplo, Plutão tem um raio de $1190 \pm 5$ km \cite[submetido]{DiasOliveira2015} que a uma distância de aproximadamente 32 UA representam um diâmetro angular de 102 mas.
+Por exemplo, Plutão tem um raio de $1190 \pm 5$ km \cite[submetido]{DiasOliveira2015} que a uma distância de aproximadamente 32 UA representa um diâmetro angular de 102 mas.
 Para a maioria dos objetos selecionados, menores que Plutão e/ou mais distantes, seus tamanhos angulares são menores que 50 mas.
@@
 Duas delas tiveram apenas duas cordas cada que não foram suficientes para prover resultados acurados\footnote{Esses eventos ocorreram em 22 de Agosto de 1994 e 30 de Outubro de 2010.}.
-Os dois restantes, que ocorreram em 17 de Agosto de 2010 e 25 de Outubro de 2013, foram trabalhos por mim em colaboração com o grupo do Rio \citep[aceito]{GomesJunior2015-Ceres}.
+Os dois restantes, que ocorreram em 17 de Agosto de 2010 e 25 de Outubro de 2013, foram trabalhos por mim em colaboração com o grupo do Rio \citep{GomesJunior2015-Ceres}.
 
@@
 
-O menor tempo de integração usado nas observações positivas foi de $1.0s$, que corresponde a aproximadamente $3.9km$ no plano do céu.
+O menor tempo de integração usado nas observações positivas foi de 1.0 s, que corresponde a aproximadamente 3.9 km no plano do céu.
 Portanto, o erro na determinação dos instantes de ingresso e egresso é dominado principalmente pelo tempo de integração, não pela difração de Fresnel ou diâmetro da estrela, ambos da ordem de algumas centenas de metros para esse evento.
@@
 Por outro lado, fixar o ângulo de posição para a ocultação de 2010 reduziu as barras de erro dos outros parâmetros (com exceção do achatamento).
-Por fim, Esse procedimento resultou em um excelente acordo entre os raios equatoriais obtidos para ambos os eventos.
+Por fim, esse procedimento resultou em um excelente acordo entre os raios equatoriais obtidos para ambos os eventos.
 
@@
 \hline  \hline
-Satélite & Diâm. (km)\tablefootnote{Planetary Satellite Physical Parameters - JPL: \url{http://ssd.jpl.nasa.gov/?sat_phys_par}} & Mag V  & OPD  & OHP & ESO & Total  & Jacobson \tabularnewline
+Satélite & Diâm. (km)\footnote{Planetary Satellite Physical Parameters - JPL: \url{http://ssd.jpl.nasa.gov/?sat_phys_par}} & Mag V  & OPD  & OHP & ESO & Total  & Jacobson \tabularnewline
 \hline
@@
 \indent \indent Com o objetivo de obter parâmetros físicos para os satélites irregulares com maior precisão foram feitas predições de ocultações estelares para os 7 maiores satélites de Júpiter (Himalia, Elara, Pasiphae, Sinope, Lysithea, Carme, Ananke), o satélite Phoebe de Saturno e Nereida de Netuno.
-Esses objetos são pequenos (ver Tabela \ref{Tab: comparison-horizons}), se comparados aos TNOs estudados na seção \ref{Sec: TNO-occ}, sendo que o menor dos satélites da amostra, Ananke, possui um diâmetro estimado de 28 km.
+Esses objetos são pequenos (ver Tabela \ref{Tab: comparison-horizons}), se comparados aos TNOs estudados no capítulo \ref{Cap: TNOs}, sendo que o menor dos satélites da amostra, Ananke, possui um diâmetro estimado de 28 km.
 
@@
 Para Nereida, apenas uma estrela UCAC4 foi identificada para ser ocultada.
-Nesse caso, utilizamos um catálogo de estrelas feito a partir do caminho aparente do satélite de Netuno Tritão no céu para predições de ocultações estelares pelo mesmo (ver seção \ref{Sec: TNO-occ}).
+Nesse caso, utilizamos um catálogo de estrelas feito a partir do caminho aparente do satélite de Netuno, Tritão, no céu para predições de ocultações estelares pelo mesmo (ver seção \ref{Sec: Netuno-astrometria}).
 
@@
 Outra fonte de erro na predição é a posição do satélite.
-Como mostradas no capítulo \ref{Sec: Irr-astrometria}, as órbitas dos satélites irregulares não são muito precisas havendo erros sistemáticos em seus elementos orbitais.
+Como mostradas na seção \ref{Sec: Irr-astrometria}, as órbitas dos satélites irregulares não são muito precisas havendo erros sistemáticos em seus elementos orbitais.
 Infelizmente o JPL não provê suas efemérides com os erros estimados das posições.
@@
 Jean-Eudes Arlot e o Dr.
-Josselin Desmars.
-Um trabalho semelhante foi abordado para ocultação de TNOs (ver seção \ref{Sec: TNO-occ}).
+Josselin Desmars como foi feito para o trabalho de ocultações de TNOs na seção \ref{Sec: TNO-occ}.
 
@@
 \section{Introdução}
-\label{Sec: Netuno-introducao}
+\label{Sec: Netuno-intro}
+
+\indent \indent Tritão é um satélite de Netuno que, diferentemente dos satélites regulares dos outros planetas, possui uma órbita retrógrada, circular e altamente inclinada.
+Devido à sua configuração orbital é muito provável que Tritão tenha sido capturado por Netuno \citep{McKinnon2007}.
+Ele possui um diâmetro de 1353 km \citep{Thomas2000}, sendo portanto pouco maior que Plutão, e seus parâmetros físicos mostram similaridades com os do planeta anão \citep{Nimmo2015}.
+
+Dentre os diversos mecanismos propostos para a captura do satélite, o mais provável foi proposto por \cite{Agnor2006} onde Tritão seria capturado através da dissociação de um sistema binário de planetesimais ao passar próximo a Netuno.
+Após a captura, a órbita de Tritão seria encolhida e circularizada para sua órbita atual.
+Um dos possíveis mecanismos seria a interação gravitacional com Netuno, i.e., efeito de maré \citep{Nogueira2011}.
+
+A combinação entre a obliquidade do eixo de rotação de Netuno em relação ao seu eixo orbital e a alta inclinação da órbita de Tritão dá a Tritão um complicado e extremo ciclo sazonal \citep{McKinnon2007}.
+As marés associadas à sua captura podem ter fortemente aquecido e transformado seu interior.
+
+Tritão é um dos poucos objetos do sistema solar que possuem atmosfera e é constituída principalmente de nitrogênio.
+A extrema variação da latitude subsolar em Tritão causam bruscas variações na distribuição de nitrogênio congelado na superfície do objeto de forma que a calota polar pode chegar até próximo o equador do satélite \citep{Hansen1992}.
+Modelos preveem que a pressão da atmosfera de Tritão varia ao longo do tempo de duas ordens de grandeza.
 
@@
 
+\indent \indent Nos mesmos moldes em foram feitos observações dos caminhos de TNOs para predição de ocultações estelares (ver seção \ref{Sec: TNOs-astrometria}), o caminho de Tritão também foi observado gerando um catálogo de estrelas inclusos no catálogo WFI.
+Infelizmente, devido a Netuno estar passando por regiões pouco populada por estrelas brilhantes, nenhum evento pôde ser observado oriundo deste trabalho.
+
+Para 2015, antes que o catálogo WFI não cubra mais o caminho orbital de Tritão, ainda haverá uma oportunidade de observar uma ocultação estelar pelo objeto.
+O evento está predito para ocorrer dia 10 de Novembro de 2015 com a sombra passando pela China.
+Dependendo do offset da estrela e de Tritão, há possibilidade da sombra ir mais ao sul passando pelo Japão ou Austrália, onde temos colaboradores que poderão observar o evento.
+
 \section{Ocultação Maio/2008}

### 2015-06-16 qualify/Qualify-altair.tex
Escrevi ocultacao de maio de 2008

@@
 O evento está predito para ocorrer dia 10 de Novembro de 2015 com a sombra passando pela China.
-Dependendo do offset da estrela e de Tritão, há possibilidade da sombra ir mais ao sul passando pelo Japão ou Austrália, onde temos colaboradores que poderão observar o evento.
+Dependendo do offset da estrela e de Tritão, há a possibilidade da sombra ir mais ao sul passando pelo Japão ou Austrália, onde temos colaboradores que poderão observar o evento.
 
-\section{Ocultação Maio/2008}
+\section{Ocultação de Maio/2008}
 \label{Sec: Netuno-maio2008}
+
+\indent \indent Em 21 de Maio de 2008 houve uma ocultação por Tritão cuja sombra cruzou o sul da África.
+Dos sítios que observaram, apenas duas cordas foram positivas.
+As observações foram feitas de La Réunion e da Namíbia com câmeras rápidas.
+Dentre elas, uma observou somente a atmosfera do objeto.
+Foi a segunda ocultação de Tritão, sendo a primeira observada em Julho de 1997 \citep{Elliot2000}.
+
+Os perfis de pressão da atmosfera de Tritão foram calculadas, porém, devido a uma das cordas ter pego somente atmosfera, não foi possível determinar a posição das cordas em relação ao corpo com precisão.
+Isso resultou em valores com barras de erro muito grandes comparadas às encontradas na literatura.
+
+\cite{Hansen1992} obteve que a pressão atmosférica de Tritão em sua superfície estaria em ascensão.
+Neste trabalho, a pressão atmosférica foi obtida com um valor muito diferente do valor esperado mesmo ao nível de $3\sigma$.
+O resultado obtido também é inconsistente com outros valores da literatura que confirmam o esperado por \cite{Hansen1992}.
+
+Para recuperar essa ocultação, um trabalho astrométrico deve ser feito de forma a medir com precisão maior a posição do centro do corpo no instante da ocultação.
+Dessa forma, será possível localizar as cordas observadas às suas respectivas latitudes diminuindo as barras de erro.
 

### 2015-06-16 qualify/comando-pdf.sh

@@ -1,1 +1,1 @@
-pdftk Qualify-altair.pdf cat 1 5-end output Qualify-altair_output.pdf
+pdftk Qualify-altair.pdf cat 1-2 5-end output Qualify-altair_output.pdf

### 2015-06-16 qualify/Qualify-altair.tex
terminei telemetria e perspectivas

@@
 \indent \indent Para o Doutorado sanduíche, irei trabalhar com as posições do OPD para estudar os efeitos de maré no sistema Netuno-Tritão.
-A vantagem desse banco de dados é sua cobertura temporal e grande número de observações que pode compensar falta de precisão ao se focar em efeitos dinâmicos de longo período \citep{Lainey2012}.
+A vantagem desse banco de dados é sua cobertura temporal e grande número de observações que pode compensar falta de precisão, se comparada a posições oriundas de sondas, ao se focar em efeitos dinâmicos de longo período \citep{Lainey2012}.
 
-\chapter{Perspectivas}
+Além das posições de solo, irei trabalhar com posições telemétricas de sondas espaciais dentro do projeto ESPaCE (European Satellite PArtnership for Computing Ephemerides).
+Esse projeto tem por objetivo extrair e analisar dados astrométricos de medidas feitas por sondas espaciais e utilizá-las para cálculo de órbitas e geração de efemérides dos objetos dos Sistema Solar.
+Essas medidas são ultra precisas e poderão prover novos modelos dinâmicos para muitos satélites naturais\footnote{\url{http://espace.oma.be/}}.
+
+A combinação desses dois bancos de dados, junto com as posições publicadas na literatura permitirá estudar efeitos dissipativos de energia no interior de Tritão.
+Como dito na seção \ref{Sec: Netuno-intro}, a circularização da órbita de Tritão, a partir de uma órbita altamente excêntrica, após a captura aconteceu por dissipação oriunda de efeitos de maré \citep{Nimmo2015}.
+
+Durante a circularização da órbita, o aquecimento causado por efeito de maré teria aumentado chegando a um fluxo de calor de $\approx 2 - 4 W m^{-2}$, comparáveis às medidas atualmente para Io.
+Após o aquecimento, o interior de Tritão começou a esfriar, mas mesmo hoje esse esfriamento não deve ter sido completado \citep{McKinnon2007}.
+
+Pela órbita de Tritão ser retrógrada, espera-se que sua órbita diminua devido ao efeito de maré que provoca em Netuno e seja destruído ao se aproximar do planeta devido à força de maré provocada por Netuno \citep{McKinnon2007}.
+Estima-se que a órbita de Tritão encolha em aproximadamente 15\% nos próximos 5 bilhões de anos.
+
+\chapter{Conclus\~ao e Perspectivas}
 \label{Cap: perspectivas}
 
-\indent \indent
+\indent \indent Esse trabalho tem por objeto estudar os objetos do Sistema Solar exterior, de forma a obter características físicas e dinâmicas relevantes para entender a formação e evolução do Sistema Solar.
+As técnicas utilizadas consistem em obter posições astrométricas precisas para integração orbital e geração de efemérides.
+Além disso, prever, observar e reduzir ocultações estelares por esses objetos e obter seus parâmetros físicos (tamanho, forma, albedo, densidade, presença de atmosfera, satélites ou anéis).
+
+Participo ativamente em observações de TNOs, satélites, e estrelas com o objetivo de melhorar as efemérides e/ou predição de ocultações.
+Nesse âmbito já escrevi 1 trabalho que está aceito para publicação: \cite{GomesJunior2015-Irregular}; e contribuo em 2 trabalhos submetidos: \cite{Camargo2015} e \cite{Desmars2015}.
+
+A busca pela predição de uma ocultação e sua melhoria na era pré-GAIA vai demandar ainda diversas missões observacionais.
+Quando o catálogo do GAIA for lançado, as posições das estrelas serão boas o suficientes para não precisar mais de acompanhamento, porém os objetos do Sistema Solar ainda precisaram ser observados e suas efemérides melhoradas pois serão, portanto, a única fonte de erro na predição de uma ocultação.
+
+Observar uma ocultação pode revelar características inesperadas e fornecerá um rico conhecimento dos mistérios do Sistema Solar.
+Exemplos disso são a descoberta de um sistema de anéis, estrutura antes conhecida apenas nos planetas gigantes, ao redor de um Centauro e a identificação de um alto albedo geométrico em Eris, objeto que está muito distante do Sol e possui um tamanho similar ao de Plutão e Tritão.
+Esses dois trabalhos foram publicados em \cite{BragaRibas2014} e \cite{Sicardy2011}, respectivamente, nos quais contribuí em caráter observacional.
+
+Particularmente, Plutão possui um interesse especial.
+Ele é um planeta anão que possui atmosfera, cinco satélites, sendo que o maior deles possui um diâmetro pouco maior que a metade do de Plutão, e está prestes a ser visitado por uma sonda espacial.
+Ocultações estelares por esse objeto permitirá estudar a evolução de sua atmosfera, a qual será calibrada a partir de observações oriundas da sonda \textit{New Horizons}.
+Para isso uma ocultação que ocorrerá próxima à data da chegada da sonda está sendo planejada e muitos observadores mobilizados.
+Das observações às quais participei, dois estudos sobre a atmosfera de Plutão puderam ser feitos, um publicado: \cite{Olkin2015}; e outro está submetido: \cite{DiasOliveira2015}.
+
+Da mesma forma que para Plutão, Ceres é um planeta anão e está sendo visitado por uma sonda.
+A partir de duas ocultações, calculamos o tamanho de Ceres e obtivemos posições muito precisas.
+Os resultados foram publicados em \cite{GomesJunior2015-Ceres} e serviram como aprendizado pessoal quanto à utilização da técnica.
+Quando os resultados da sonda \textit{Dawn} forem publicados, será possível calibrar a técnica de ocultações estelares utilizada nesse trabalho.
+
+Muitos outros objetos estão na lista para se observar ocultações estelares, todos com grandes potenciais científicos.
+Por exemplo, há a possibilidade de que o Centauro (2060) Chiron também possua anéis \citep{Ruprecht2015}.
+Como o número de ocultações detectadas pelo nosso grupo vem aumentando ao longo do tempo e aumentará muito mais com o GAIA, diversas publicações ainda estão por vim.
+
+Dentre os objetos a serem ocultados, inserimos na lista os satélites irregulares dos planetas gigantes, nunca antes observados por ocultações.
+Diversas ocultações já estão preditas e publicação destas predições está em preparação.
+Observações de ocultações por Himalia, satélite de Júpiter, e Tritão para esse ano estão sendo preparadas.
+
+Por fim, estudos sobre efeitos de maré no sistema Netuno-Tritão para obter parâmetros internos do satélites serão feitos durante o Doutorado Sanduíche que ocorrerá na França em colaboração com o Doutor Valéry Lainey.
+Nesse trabalho, focaremos na obtenção de posições astrométricas do sistema a partir de imagens de solo observadas no OPD e de sondas espaciais dentro do projeto ESPaCE.
 

### 2015-06-16 qualify/Qualify-altair.tex
falta terminar introducao dos TNOs

@@
 
-Nos capítulos \ref{Cap: TNOs}, \ref{Cap: Ceres}, \ref{Cap: Irregulares} e \ref{Cap: Netuno-Tritao} relatamos os trabalhos desenvolvidos envolvendo predições, observações e testes de ocultações estelares com Centauros e TNOs, Ceres, satélites irregulares dos planetas gigantes e com o sistema Netuno-Tritão, respectivamente.
+Nos capítulos \ref{Cap: TNOs}, \ref{Cap: Ceres}, \ref{Cap: Irregulares} e \ref{Cap: Netuno-Tritao} relatamos os trabalhos desenvolvidos envolvendo predições e observações de ocultações estelares com Centauros e TNOs, Ceres, satélites irregulares dos planetas gigantes e com o sistema Netuno-Tritão, respectivamente.
 Por fim, no capítulo \ref{Cap: perspectivas} resumimos o que já foi desenvolvido e descrevemos o que ainda está para ser desenvolvido.
@@
 
-%Since their ephemerides are not very precise, predicting and observing stellar occultations are very difficult, and no observation of such an event for an irregular satellite is found in the literature.
-The precise star positions to be derived by the ESA astrometry satellite Gaia \citep{deBruijne2012} will render better predictions with the only source of error being the ephemeris.
-The positions derived from our observations can be used in new orbital numerical integrations, generating more precise ephemerides.
-
-%The power of stellar occultations for observing relatively small diameter solar system objects is supported by recent works, such as the discovery of a ring system around the Centaur (10199) Chariklo \citep{Braga-Ribas2014}.
-Once irregular satellites start to be observed by this technique, it will be possible to obtain their physical parameters (shape, size, albedo, density) with unprecedented precision.
-For instance, in this case, sizes could be obtained with kilometer accuracy.
-The knowledge of these parameters would in turn bring valuable information for studying the capture mechanisms and origin of the irregular satellites.
+As curvas de luz são geradas a partir das observações seguindo a técnica da fotometria diferencial.
+Em nosso trabalho usamos as tasks de fotometria diferencial do pacote PRAIA.
+As imagens são previamente calibradas de bias e "flatfield".
+Os alvos e objetos de calibração são marcados em uma imagem de referência (normalmente a primeira da série) e depois automaticamente identificados no resto da sequência.
+Técnicas de cálculo do fundo de céu são empregadas, envolvendo a análise da moda de histogramas, e o fluxo dos objetos e do alvo são calculados.
+Diversas aberturas são testadas até que a menor dispersão seja obtida para a curva (melhor relação sinal-ruído).
+Ao final, obtêm-se as razões de fluxo alvo/calibrador(es), sinal-ruído, erros da razão de fluxo, entre outras informações.
+O instante de tempo a cada ponto da curva também é registrado.
+
+Nas seções \ref{Sec: Ceres-2010-occ} e \ref{Sec: Ceres-2013-occ} detalhamos os procedimentos da análise da curva para obtenção dos instantes de imersão e emersão, para o cálculo da dimensão e forma do alvo, no caso de ausência de atmosfera.
+Na presença de atmosfera, técnicas de "ray-tracing" são utilizadas para ajustar a altura e perfis de densidade/temperatura a partir da curva de luz obtida.
+Maiores detalhes desta técnica são dadas em \cite{DiasOliveira2015} e referências.
 
@@
 \label{Cap: TNOs}
+
+\section{Introdução}
+\label{Sec: TNOs-intro}
+
+\indent \indent Os objetos trans-Netunianos (TNOs) estão localizados após a órbita de Netuno, como o própio nome diz.
+Devido às suas grandes distâncias, eles podem ser considerados remanescentes quase inalterados da formação do Sistema Solar \cite{Assafin2012}.
+Já os Centauros se localizam entre as órbitas de Júpiter e Netuno.
+Eles são objetos transientes entre TNOs e cometas da família de Júpiter e acredita-se que tenha origem comum aos TNOs \citep{Camargo2013}.
+
+Estudá-los é essencial para entender a formação e evolução do Sistema Solar.
 
@@
 \section{Ocultação de 2010}
-\label{Subsec: 2010-occ}
+\label{Sec: Ceres-2010-occ}
 
@@
 \section{Ocultação de 2013}
-\label{Subsec: 2013-occ}
+\label{Sec: Ceres-2013-occ}
 
@@
 
-Os instantes de ingresso e egresso da ocultação foram obtidos pelo mesmo procedimento descrito na seção \ref{Subsec: 2010-occ}.
+Os instantes de ingresso e egresso da ocultação foram obtidos pelo mesmo procedimento descrito na seção \ref{Sec: Ceres-2010-occ}.
 Uma vez que o tempo de integração efetivo usado (0.17 s) representa aproximadamente 7 km no plano do céu, e a escala de Fresnel e  o diâmetro da estrela estão novamente na ordem de centenas de metros, o erro da determinação dos instantes de imersão e emersão são dominados principalmente pelo tempo de integração, da mesma forma que para o evento de 2010.
@@
 Perfis elípticos foram ajustados para todas as cordas restantes.
-pelo mesmo procedimento descrito na seção \ref{Subsec: 2010-occ}.
+pelo mesmo procedimento descrito na seção \ref{Sec: Ceres-2010-occ}.
 O resultado foi $\chi^2_{r,min} = 13$ , sugerindo que um modelo elíptico não é satisfatório para os dados.
@@
 \section{Discussão}
+\label{Sec: Ceres-discussao}
 
@@
 
-Tritão é um dos poucos objetos do sistema solar que possuem atmosfera e é constituída principalmente de nitrogênio.
-A extrema variação da latitude subsolar em Tritão causam bruscas variações na distribuição de nitrogênio congelado na superfície do objeto de forma que a calota polar pode chegar até próximo o equador do satélite \citep{Hansen1992}.
+Tritão é um dos poucos objetos do sistema solar que possuem atmosfera e é constituída principalmente de $N_{2}$.
+A extrema variação da latitude subsolar em Tritão causa bruscas variações na distribuição de nitrogênio congelado na superfície do objeto de forma que a calota polar pode chegar até próximo o equador do satélite \citep{Hansen1992}.
 Modelos preveem que a pressão da atmosfera de Tritão varia ao longo do tempo de duas ordens de grandeza.
@@
 
-\indent \indent Com o objetivo de estudar os efeitos de maré sobre o sistema Netuno-Tritão, serão obtidos as posições desses objetos em mais de 7000 imagens observadas no OPD desde 1992.
+\indent \indent Com o objetivo de estudar os efeitos de maré sobre o sistema Netuno-Tritão, serão obtidos as posições desses objetos em mais de 7000 imagens observadas no OPD desde 1992 até o presente.
 Uma vez que eles estão muito próximos no céu e Netuno é muito mais brilhante do que Tritão, a posição do satélite pode estar afetada pela luz do planeta.
@@
 
-O procedimento de coronografia digital faz parte do pacote PRAIA e foi desenvolvido para se observar os satélites principais de Urano durante os fenômenos mútuos de 2008 \citep{Assafin2008}.
+O procedimento de coronografia digital faz parte do pacote PRAIA e foi desenvolvido para se observar os satélites principais de Urano durante os fenômenos mútuos de 2007 \citep{Assafin2009}.
 A técnica consiste em computar o perfil de luz do objeto mais brilhante dado uma área.
@@
 
-Para se obter o perfil do objeto, a distância $d$ de cada pixel ($j$,$i$) às coordenadas centrais do baricentro ($x_{c}$,$y_{c}$) da fonte brilhante.
+Para se obter o perfil do objeto, é calculada a distância $d$ de cada pixel ($j$,$i$) às coordenadas centrais do baricentro ($x_{c}$,$y_{c}$) da fonte brilhante.
 Para cada anel de 1 pixel de largura a uma distância $d$ do baricentro é feito um histograma de contagens.
@@
 
-Como o tamanho aparente de Netuno ($\approx 2\arcsec$) é da ordem do seeing, uma técnica está sendo desenvolvida para determinar o centro de objetos extensos.
+Como o tamanho aparente de Netuno ($\approx 2\arcsec$) é da ordem do seeing, uma técnica está sendo desenvolvida para determinar o centroide de objetos extensos.
 No caso, supõe-se que, para cada ponto da imagem, a luz é espalhada sobre o CCD como uma gaussiana e a imagem correspondente é a soma de todas estas gaussianas.
+O perfil resultante não tem forma analítica fechada, devendo ser calculado numericamente.
+Este estudo ainda está em andamento.
 
@@
 
-\indent \indent Nos mesmos moldes em foram feitos observações dos caminhos de TNOs para predição de ocultações estelares (ver seção \ref{Sec: TNOs-astrometria}), o caminho de Tritão também foi observado gerando um catálogo de estrelas inclusos no catálogo WFI.
-Infelizmente, devido a Netuno estar passando por regiões pouco populada por estrelas brilhantes, nenhum evento pôde ser observado oriundo deste trabalho.
+\indent \indent Nos mesmos moldes em que foram feitas observações dos caminhos de TNOs para predição de ocultações estelares (ver seção \ref{Sec: TNOs-astrometria}), o caminho de Tritão também foi observado gerando um catálogo de estrelas inclusos no catálogo WFI.
+Infelizmente, devido a Netuno estar passando por regiões pouco populadas por estrelas brilhantes, nenhum evento pôde ser observado oriundo deste trabalho.
 
@@
 Dentre elas, uma observou somente a atmosfera do objeto.
-Foi a segunda ocultação de Tritão, sendo a primeira observada em Julho de 1997 \citep{Elliot2000}.
-
-Os perfis de pressão da atmosfera de Tritão foram calculadas, porém, devido a uma das cordas ter pego somente atmosfera, não foi possível determinar a posição das cordas em relação ao corpo com precisão.
-Isso resultou em valores com barras de erro muito grandes comparadas às encontradas na literatura.
-
-\cite{Hansen1992} obteve que a pressão atmosférica de Tritão em sua superfície estaria em ascensão.
+Foi a primeira ocultação de Tritão desde 1997 \citep{Elliot2000}.
+
+Os perfis de pressão da atmosfera de Tritão foram calculados (Sicardy et al., comunicação privada), porém, devido a uma das cordas ter pego somente atmosfera, não foi possível determinar a posição das cordas em relação ao corpo com precisão.
+Isso resultou em valores de pressão próximo à superfície com pequenas barras de erro interno, mas que podem estar desviados em mais de 3 sigmas do valor real.
+
+\begin{figure}[h]
+\begin{centering}
+\subfigure[]{\includegraphics[scale=0.29]{figuras/tritao-2008.png} \label{Fig: Tritao-body}}
+\subfigure[]{\includegraphics[scale=0.29]{figuras/pressao-2008.png} \label{Fig: Tritao-pressao}}
+\caption{a) Reconstrução pós-ocultação da forma de Tritão.
+As linhas são as cordas observadas.
+Como é possível ver, a corda da Namíbia observou apenas a atmosfera de Tritão.
+b) Comparação entre a pressão atmosférica na superfície de Tritão obtida por Sicardy et al.
+(comunicação privada) e pressões publicadas na literatura: \cite{Gurola1995, Olkin1997, Sicardy1998} e \cite{Elliot2000}.
+\label{Fig: Ceres-map}}
+\end{centering}
+\end{figure}
+
+\cite{Hansen1992} geraram modelos que preveem que a pressão atmosférica de Tritão em sua superfície estaria em ascensão.
 Neste trabalho, a pressão atmosférica foi obtida com um valor muito diferente do valor esperado mesmo ao nível de $3\sigma$.
@@
 Para recuperar essa ocultação, um trabalho astrométrico deve ser feito de forma a medir com precisão maior a posição do centro do corpo no instante da ocultação.
-Dessa forma, será possível localizar as cordas observadas às suas respectivas latitudes diminuindo as barras de erro.
+Dessa forma, será possível localizar as cordas observadas às suas respectivas latitudes, diminuindo as barras de erro.
 
@@
 
-\chapter{Conclus\~ao e Perspectivas}
+\chapter{Conclusão e Perspectivas}
 \label{Cap: perspectivas}

### 2015-06-17 qualify/Qualify-altair.tex
Qualify terminado

@@
 \includegraphics[scale=0.75]{figuras/stellarocc.png} 
+\caption{Curva de luz esquemática de uma ocultação estelar por um planeta com atmosfera e anéis. Fonte: \cite{Elliot1979}}
 \label{Fig: occ-scheme}
-\caption{Curva de luz esquemática de uma ocultação estelar por um planeta com atmosfera e anéis. Fonte: \cite{Elliot1979}}
 \end{centering} 
@@
 A atmosfera, no caso, agiria como uma lente como mostrado na Fig.
-\ref{Fig: occ-scheme}.1.
+\ref{Fig: occ-scheme}.
 
@@
 
-\indent \indent Os objetos trans-Netunianos (TNOs) estão localizados após a órbita de Netuno, como o própio nome diz.
-Devido às suas grandes distâncias, eles podem ser considerados remanescentes quase inalterados da formação do Sistema Solar \cite{Assafin2012}.
-Já os Centauros se localizam entre as órbitas de Júpiter e Netuno.
-Eles são objetos transientes entre TNOs e cometas da família de Júpiter e acredita-se que tenha origem comum aos TNOs \citep{Camargo2013}.
-
-Estudá-los é essencial para entender a formação e evolução do Sistema Solar.
+\indent \indent Dentro do contexto explicitado no capítulo \ref{Cap: intro}, os TNOs são uns dos objetos mais significantes por possivelmente serem a fonte de outras populações do Sistema Solar, como por exemplo Centauros \citep{Fernandez2002}, Satélites Irregulares \citep{Nesvorny2007} e Troianos \citep{Morbidelli2005}.
+Também existe a possibilidade de que Ceres tenha uma origem como TNO \citep{McKinnon2012}, assim como Tritão \citep{Agnor2006}.
+
+Pouco se conhece sobre Centauros e TNOs, pois nunca foram visitados por sondas e a grande maioria das observações são de origem fotométrica usual (magnitudes, cores) ou, em menor parte, espectroscópica, ou em menor número ainda por telescópios espaciais.
+Plutão será o primeiro objeto da classe a ser visitado por uma sonda, a \textit{New Horizons}.
+
+Estimar parâmetros físicos para esses objetos é um desafio, porém essencial para acessar a massa atual e composição (material) da população e recuperar sua história evolutiva.
+É possível obter seus tamanhos através de modelos, por exemplo combinando o brilho no visível e emissão térmica obtida em infra-vermelho.
+Todavia os erros obtidos são piores que 10 \% \citep{Camargo2013}.
+
+Para obter parâmetros como tamanho, albedo, densidade, etc, de forma muito mais precisa e acurada, sem a necessidade de se adotar modelos, utilizamos a técnica de ocultações estelares.
+Até 2009, quando foi observada uma ocultação de 2002TX300 \citep{Elliot2010}, Plutão e Caronte eram os únicos TNOs com ocultações observadas.
+Até hoje, cerca de uma dezena de TNOs e Centauros tiveram ocultações observadas, alguns com apenas uma corda, como Varuna, o que permite apenas determinar um tamanho mínimo para o objeto.
+
+Com o objetivo de prever e observar ocultações estelares por TNOs, o nosso grupo conta com colaboração de pesquisadores nacionais e internacionais, profissionais e amadores.
+Utilizamos telescópios de pequeno e grande porte e câmera rápida para observar eventos que proporcionaram grandes descobertas, como por exemplo a descoberta do sistema de anéis ao redor do Centauro Chariklo \citep{BragaRibas2014}.
 
@@
 
-\indent \indent Como dito no Cap.
-\ref{Cap: intro}, certos objetos do Sistema Solar exterior podem ser considerados fósseis do Sistema Solar.
-Dentre esses, os TNOs, por estarem mais distantes, são os que guardam as informações mais antigas.
-Com o objetivo de obter os parâmetro físicos desses objetos, para melhor caracterizar as propriedades básicas dessas populações, uma grande campanha observacional é realizada para se observar ocultações estelares por TNOs e Centauros.
+\indent \indent Com o objetivo de obter os parâmetro físicos desses objetos, para melhor caracterizar as propriedades básicas dessas populações, uma grande campanha observacional é realizada para se observar ocultações estelares por TNOs e Centauros.
 Essa campanha é liderada por pesquisadores brasileiros, franceses e espanhóis e tem colaboração com observadores espalhados por todo o mundo.
@@
 \begin{table}[h]
-\caption{\label{Tab: comparison-horizons} Comparação entre o número de posições obtidas em \cite{GomesJunior2015-Irregular} e o número utilizado nas integrações numéricas das órbitas pelo JPL como publicado por  \cite{Jacobson2012}.
+\caption{Comparação entre o número de posições obtidas em \cite{GomesJunior2015-Irregular} e o número utilizado nas integrações numéricas das órbitas pelo JPL como publicado por  \cite{Jacobson2012}.
 Os satélites estão separados por planeta (linha reta) e por família orbital (linha tracejada).}
+\label{Tab: comparison-horizons}
 \begin{centering}
@@
 
-Os perfis de pressão da atmosfera de Tritão foram calculados (Sicardy et al., comunicação privada), porém, devido a uma das cordas ter pego somente atmosfera, não foi possível determinar a posição das cordas em relação ao corpo com precisão.
-Isso resultou em valores de pressão próximo à superfície com pequenas barras de erro interno, mas que podem estar desviados em mais de 3 sigmas do valor real.
+Os perfis de pressão da atmosfera de Tritão foram calculados (Sicardy et al., comunicação privada), porém, devido a uma das cordas ter pego somente atmosfera, não foi possível determinar a posição das cordas em relação ao corpo com precisão (ver figura \ref{Fig: Tritao-body}).
+Isso resultou em valores de pressão próximo à superfície com pequenas barras de erro interno, mas que podem estar desviados em mais de 3 sigmas do valor real (ver figura \ref{Fig: Tritao-pressao}).
 
@@
 \begin{centering}
-\subfigure[]{\includegraphics[scale=0.29]{figuras/tritao-2008.png} \label{Fig: Tritao-body}}
-\subfigure[]{\includegraphics[scale=0.29]{figuras/pressao-2008.png} \label{Fig: Tritao-pressao}}
-\caption{a) Reconstrução pós-ocultação da forma de Tritão.
-As linhas são as cordas observadas.
-Como é possível ver, a corda da Namíbia observou apenas a atmosfera de Tritão.
-b) Comparação entre a pressão atmosférica na superfície de Tritão obtida por Sicardy et al.
-(comunicação privada) e pressões publicadas na literatura: \cite{Gurola1995, Olkin1997, Sicardy1998} e \cite{Elliot2000}.
-\label{Fig: Ceres-map}}
+\includegraphics[scale=0.58]{figuras/tritao-2008.png} 
+\caption{Reconstrução pós-ocultação da forma de Tritão. As linhas são as cordas observadas. Como é possível ver, a corda da Namíbia observou apenas a atmosfera de Tritão.}
+\label{Fig: Tritao-body}
+\end{centering}
+\end{figure}
+
+\begin{figure}[h]
+\begin{centering}
+\includegraphics[scale=0.58]{figuras/pressao-2008.png} 
+\caption{Comparação entre a pressão atmosférica na superfície de Tritão obtida por Sicardy et al.
+(comunicação privada) e pressões publicadas na literatura: \cite{Gurola1995, Olkin1997, Sicardy1998} e \cite{Elliot2000}}.
+\label{Fig: Tritao-pressao}
 \end{centering}
@@
 \indent \indent Para o Doutorado sanduíche, irei trabalhar com as posições do OPD para estudar os efeitos de maré no sistema Netuno-Tritão.
-A vantagem desse banco de dados é sua cobertura temporal e grande número de observações que pode compensar falta de precisão, se comparada a posições oriundas de sondas, ao se focar em efeitos dinâmicos de longo período \citep{Lainey2012}.
+A vantagem desse banco de dados é sua cobertura temporal (1992 - 2015), o grande número de observações e um tratamento astrométrico diferenciado.
+Faremos uso de coronografia digital para eliminar a influência da luz difusa de Netuno sobre Tritão, e faremos uma determinação direta de posições de Netuno (com PSFs numéricas especialmente deduzidas para o caso de objetos extensos), com Tritão no mesmo frame.
+Esses dados tem o mesmo valor que o de posições oriundas de sondas, em que pesem efeitos dinâmicos de longo período \citep{Lainey2012}.
 
@@
 
-Pela órbita de Tritão ser retrógrada, espera-se que sua órbita diminua devido ao efeito de maré que provoca em Netuno e seja destruído ao se aproximar do planeta devido à força de maré provocada por Netuno \citep{McKinnon2007}.
+Pela órbita de Tritão ser retrógrada, espera-se que o semi-eixo de sua órbita diminua devido ao efeito de maré que provoca em Netuno e que seja destruído ao se aproximar do planeta devido à essa mesma força \citep{McKinnon2007}.
 Estima-se que a órbita de Tritão encolha em aproximadamente 15\% nos próximos 5 bilhões de anos.
@@
 
-\indent \indent Esse trabalho tem por objeto estudar os objetos do Sistema Solar exterior, de forma a obter características físicas e dinâmicas relevantes para entender a formação e evolução do Sistema Solar.
-As técnicas utilizadas consistem em obter posições astrométricas precisas para integração orbital e geração de efemérides.
-Além disso, prever, observar e reduzir ocultações estelares por esses objetos e obter seus parâmetros físicos (tamanho, forma, albedo, densidade, presença de atmosfera, satélites ou anéis).
+\indent \indent Esse trabalho tem por meta estudar os objetos do Sistema Solar exterior, de forma a obter características físicas e dinâmicas relevantes para entender a formação e evolução do Sistema Solar.
+As técnicas utilizadas consistem em obter posições astrométricas precisas desses corpos para corrigir erros sistemáticos em suas efemérides, ou mesmo permitir a determinação de novas órbitas com novas integrações numéricas, a fim de melhorar a predição de futuras ocultações estelares.
+A técnica de fotometria diferencial de abertura é empregada para gerar as curvas de luz das ocultações estelares.
+Prever, observar e reduzir ocultações estelares por esses objetos e obter seus parâmetros físicos (tamanho, forma, albedo, densidade, presença de atmosfera, satélites ou anéis) são os objetivos gerais desta Tese..
 
@@
 
-Observar uma ocultação pode revelar características inesperadas e fornecerá um rico conhecimento dos mistérios do Sistema Solar.
-Exemplos disso são a descoberta de um sistema de anéis, estrutura antes conhecida apenas nos planetas gigantes, ao redor de um Centauro e a identificação de um alto albedo geométrico em Eris, objeto que está muito distante do Sol e possui um tamanho similar ao de Plutão e Tritão.
-Esses dois trabalhos foram publicados em \cite{BragaRibas2014} e \cite{Sicardy2011}, respectivamente, nos quais contribuí em caráter observacional.
+Observar uma ocultação pode revelar características inesperadas e até dar origem a novas descobertas no Sistema Solar.
+Exemplos disso são a descoberta de um sistema de anéis, estrutura antes conhecida apenas nos planetas gigantes, ao redor de um Centauro e a identificação de um alto albedo geométrico em Eris, objeto que está muito distante do Sol e possui um tamanho bem menor do que se supunha, similar ao de Plutão e Tritão.
+Esses dois trabalhos foram publicados na Nature em \cite{BragaRibas2014} e \cite{Sicardy2011}, respectivamente, nos quais contribuí em caráter observacional.
 
 Particularmente, Plutão possui um interesse especial.
-Ele é um planeta anão que possui atmosfera, cinco satélites, sendo que o maior deles possui um diâmetro pouco maior que a metade do de Plutão, e está prestes a ser visitado por uma sonda espacial.
-Ocultações estelares por esse objeto permitirá estudar a evolução de sua atmosfera, a qual será calibrada a partir de observações oriundas da sonda \textit{New Horizons}.
+Prestes a ser visitado por uma sonda espacial, ele é um planeta anão que possui atmosfera, cinco satélites, sendo que o maior deles possui um diâmetro pouco maior que a metade do de Plutão.
+Ocultações estelares por esse objeto permitem estudar a evolução de sua atmosfera, a qual será calibrada a partir de observações oriundas da sonda \textit{New Horizons}.
 Para isso uma ocultação que ocorrerá próxima à data da chegada da sonda está sendo planejada e muitos observadores mobilizados.
@@
 Por exemplo, há a possibilidade de que o Centauro (2060) Chiron também possua anéis \citep{Ruprecht2015}.
-Como o número de ocultações detectadas pelo nosso grupo vem aumentando ao longo do tempo e aumentará muito mais com o GAIA, diversas publicações ainda estão por vim.
+Como o número de ocultações detectadas pelo nosso grupo vem aumentando ao longo do tempo e aumentará muito mais com o GAIA, diversas publicações ainda estão por vir.
 
 Dentre os objetos a serem ocultados, inserimos na lista os satélites irregulares dos planetas gigantes, nunca antes observados por ocultações.
-Diversas ocultações já estão preditas e publicação destas predições está em preparação.
+Diversas ocultações já estão preditas e a publicação destas predições está em preparação (Gomes-Júnior et al.
+2015, A\&A, in prep.).
 Observações de ocultações por Himalia, satélite de Júpiter, e Tritão para esse ano estão sendo preparadas.

### 2015-06-18 qualify/Qualify-altair.tex
Feito figura legal

@@
 
-Para os satélites de saturno, \citealp{Grav2007} mostraram através de suas cores e inclinações espectrais que esses satélites contêm uma fração mais ou menos igual de objetos semelhantes a asteroides tipo c, P e D, mas SXXII Ijiraq é um pouco mais avermelhado que objetos tipo D.
+Para os satélites de saturno, \citealp{Grav2007} mostraram através de suas cores e inclinações espectrais que esses satélites contêm uma fração mais ou menos igual de objetos semelhantes a asteroides tipo C, P e D, mas SXXII Ijiraq é um pouco mais avermelhado que objetos tipo D.
 Esses trabalhos sugerem origens diferentes para os satélites irregulares.

### 2015-06-18 qualify/apresentacao/grafico.py

@@ -1,0 +1,37 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import astropy.units as u
+
+def gaussian(x, mu, sig):
+    return np.exp(-np.power(x - mu, 2.) / (2 * np.power(sig, 2.)))/np.sqrt(2*np.pi*np.power(sig, 2.))
+
+
+tam = [28,21]*u.cm
+
+fig = plt.figure(frameon=False)
+ax = fig.add_axes([0, 0, 1, 1])
+ax.axis('off')
+
+plt.plot([0.0, 2.4], [20.0, 20.0], color='black')
+plt.plot([2.4, 2.4], [20.0, 16.65], color='black')
+plt.plot([2.4, 25.6], [16.65, 16.65], color='black')
+plt.plot([25.6, 25.6], [16.65, 20.0], color='black')
+plt.plot([25.6, 28.0], [20.0, 20.0], color='black')
+
+plt.plot([27.0, 27.0], [0.0, 2.0], color='black', zorder=0.9)
+
+x = np.linspace(0, 28.0, 1000)
+y = 3*gaussian(x, 1.4, 0.5) + gaussian(x, 3.4, 0.5) + 1
+plt.plot(x,y, color='black', zorder=0.9)
+
+circle1=plt.Circle((27.0,1.0),0.15, color='black')
+circle2=plt.Circle((27.0,1.0),0.1, color='white', zorder=1.1)
+plt.gcf().gca().add_artist(circle1)
+plt.gcf().gca().add_artist(circle2)
+
+plt.xlim(0,tam[0].value)
+plt.ylim(0,tam[1].value)
+
+#fig = plt.gcf()
+fig.set_size_inches(tam[0].to(u.imperial.inch).value, tam[1].to(u.imperial.inch).value)
+fig.savefig('Fundo.png', format='png', transparent=True)

### 2015-06-23 qualify/apresentacao/grafico.py
quase terminado a apresentacao

@@ -33,4 +33,4 @@ plt.ylim(0,tam[1].value)
 
 #fig = plt.gcf()
 fig.set_size_inches(tam[0].to(u.imperial.inch).value, tam[1].to(u.imperial.inch).value)
-fig.savefig('Fundo.png', format='png', transparent=True)
+fig.savefig('Fundo.png', format='png', dpi=300, transparent=True)

### 2015-06-25 Irregular-occ/Irregular-occ.tex
.

@@
 
-\title{Astrometric positions for 18 irregular satellites of giant planets from 23 years of observations
-\fnmsep\thanks{Table 8 are only available in electronic form
-at the CDS via anonymous ftp to cdsarc.u-strasbg.fr (130.79.128.5)
-or via http://cdsweb.u-strasbg.fr/cgi-bin/qcat?J/A+A/ and IAU NSDC data base at www.imcce.fr/nsdc.}
+\title{Predictions of stellar occultations of 9 irregular satellites of giant planets
+%\fnmsep\thanks{Table 8 are only available in electronic form at the CDS via anonymous ftp to cdsarc.u-strasbg.fr (130.79.128.5) or via http://cdsweb.u-strasbg.fr/cgi-bin/qcat?J/A+A/ and IAU NSDC data base at www.imcce.fr/nsdc.}
 %\fnmsep\thanks{The complete version of Table 8 is available through CDS and IAU NSDC data base at www.imcce.fr/nsdc.}
-\fnmsep\thanks{Partially based on observations made at Laborat\'orio Nacional de Astrof\'{\i}sica (LNA), Itajub\'a-MG, Brazil.}
-\fnmsep\thanks{Partially based on observations through the ESO runs 079.A-9202(A), 075.C-0154, 077.C-0283 and 079.C-0345.}
-\fnmsep\thanks{\textbf{Partially based on observations made at
-Observatoire de Haute Provence (OHP), F-04870 Saint-Michel l'observatoire, France}}
+%\fnmsep\thanks{Partially based on observations made at Laborat\'orio Nacional de Astrof\'{\i}sica (LNA), Itajub\'a-MG, Brazil.}
+%\fnmsep\thanks{Partially based on observations through the ESO runs 079.A-9202(A), 075.C-0154, 077.C-0283 and 079.C-0345.}
+%\fnmsep\thanks{\textbf{Partially based on observations made at Observatoire de Haute Provence (OHP), F-04870 Saint-Michel l'observatoire, France}}
 }
@@
           \inst{1,2,3} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France},
-          J.-E. Arlot
-          \inst{4},      
+%          J.-E. Arlot
+%          \inst{4},      
           J. I. B. Camargo
@@
           F. Braga-Ribas
-          \inst{2,5}, 
-          D. N. da Silva Neto
-          \inst{6},
-          A. H. Andrei
-          \inst{1,2} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau, 75014 Paris, France},
+          \inst{2,4}, 
+%          D. N. da Silva Neto
+%          \inst{6},
+%          A. H. Andrei
+%          \inst{1,2} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau, 75014 Paris, France},
           A. Dias-Oliveira
@@
           \inst{2},
-          Y. Duchemin
-          \inst{4,7},
-          J. Desmars
-          \inst{4},
-          V. Lainey
-          \inst{4},
-          W. Thuillot
-          \inst{4}
+%          Y. Duchemin
+%          \inst{4,7},
+%          J. Desmars
+%          \inst{4},
+%          V. Lainey
+%          \inst{4},
+%          W. Thuillot
+%          \inst{4}
           }
@@
 Laboratório Interinstitucional de e-Astronomia - LIneA, Rua Gal. José Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil
-              \and
-Institut de mécanique céleste et de calcul des éphémérides - Observatoire de Paris, UMR 8028 du CNRS,
-77 Av. Denfert-Rochereau, 75014 Paris, France\\
-              \email{arlot@imcce.fr}
+%              \and
+%Institut de mécanique céleste et de calcul des éphémérides - Observatoire %de Paris, UMR 8028 du CNRS,
+%77 Av. Denfert-Rochereau, 75014 Paris, France\\
+%              \email{arlot@imcce.fr}
               \and
 Federal University of Technology - Paraná (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
-              \and
-Centro Universitário Estadual da Zona Oeste, Av. Manual Caldeira de Alvarenga 1203, CEP 23.070-200 Rio de Janeiro RJ, Brazil
-              \and
-              ESIGELEC-IRSEEM, Technopôle du Madrillet, Avenue Galilée, 76801 Saint-Etienne du Rouvray, France
+%              \and
+%Centro Universitário Estadual da Zona Oeste, Av. Manual Caldeira de Alvarenga 1203, CEP 23.070-200 Rio de Janeiro RJ, Brazil
+%              \and
+%              ESIGELEC-IRSEEM, Technopôle du Madrillet, Avenue Galilée, 76801 Saint-Etienne du Rouvray, France
               }
@@
 \abstract
-{The irregular satellites of the giant planets are believed to have been captured during the evolution of the solar system.
+%{The irregular satellites of the giant planets are believed to have been captured during the evolution of the solar system.
 Knowing their physical parameters, such as size, density and albedo is important to constrain where they came from and how they were captured.
@@
 Both techniques demand that the orbits are well known.}
-{We aimed to obtain good astrometric positions of irregular satellites in order to improve their orbits and ephemeris.}
-{We identified and reduced observations of several irregular satellites from three databases containing more than 8000 images obtained between 1992 and 2014 at three sites (Observatório do Pico dos Dias, Observatoire de Haute-Provence and European Southern Observatory - La Silla).
+%{We aimed to obtain good astrometric positions of irregular satellites in order to improve their orbits and ephemeris.}
+%{We identified and reduced observations of several irregular satellites from three databases containing more than 8000 images obtained between 1992 and 2014 at three sites (Observatório do Pico dos Dias, Observatoire de Haute-Provence and European Southern Observatory - La Silla).
 We used the software PRAIA (Platform for Reduction of Astronomical Images Automatically) to make the astrometric reduction of the CCD frames.
@@
 Some procedures were taken to overcome missing or incomplete information (coordinates, date), mostly for the older images.}
-{We managed to obtain more than 6000 positions for 18 irregular satellites, being 12 of Jupiter, 4 of Saturn, 1 of Uranus (Sycorax) and 1 of Neptune (Nereid).
+%{We managed to obtain more than 6000 positions for 18 irregular satellites, being 12 of Jupiter, 4 of Saturn, 1 of Uranus (Sycorax) and 1 of Neptune (Nereid).
 For some satellites the number of obtained positions is more than 50\% of that used in earlier orbital numerical integrations.}
-{Comparison of our positions with recent JPL ephemeris suggests the presence of systematic errors in the orbits for some of the irregular satellites.
+%{Comparison of our positions with recent JPL ephemeris suggests the presence of systematic errors in the orbits for some of the irregular satellites.
 The most evident case was an error in the inclination of Carme.}
@@
 
-The irregular satellites of the giant planets are smaller than the regular moons, having more eccentric, inclined, distant and, in most cases, retrograde orbits.
+%The irregular satellites of the giant planets are smaller than the regular moons, having more eccentric, inclined, distant and, in most cases, retrograde orbits.
 Due to their orbital configurations, it is largely accepted that these objects were captured in the early solar system \citep{Sheppard2003}.
 
-Because they are faint, the majority of these objects was discovered only in the last century\footnote{Website: http://ssd.jpl.nasa.gov/?sat\_discovery} .
+%Because they are faint, the majority of these objects was discovered only in the last century\footnote{Website: http://ssd.jpl.nasa.gov/?sat\_discovery} .
 They were never visited by a spacecraft, with the exception of Himalia, Phoebe and Nereid, in a flyby by the Cassini space probe in 2000 for Himalia \citep{Porco2003} and in 2004 for Phoebe \citep{Desmars2013} and in a flyby by the Voyager 2 space probe in 1989 for Nereid \citep{Smith1989}.
@@
 
-If these objects were captured, there remains the question of where they came from.
+%If these objects were captured, there remains the question of where they came from.
 \citealp{Clark2005} showed from imaging spectroscopy from Cassini that Phoebe has a surface probably covered by material from the outer solar system and \citealp{Grav2003} showed that the satellites of the Jovian Prograde Group Himalia have grey colors implying that their surfaces are similar to that of C-type asteroids.
@@
 
-For Saturnian satellites, \citealp{Grav2007} showed by their colors and spectral slopes that these satellites contain a more or less equal fraction of C-, P- and D-like objects but SXXII Ijiraq is marginally redder than D-type objects.
+%For Saturnian satellites, \citealp{Grav2007} showed by their colors and spectral slopes that these satellites contain a more or less equal fraction of C-, P- and D-like objects but SXXII Ijiraq is marginally redder than D-type objects.
 These works may suggest different origins for the irregular satellites.
 
-In this context, we used 3 databases for deriving precise positions for the irregular satellites observed at Observatório do Pico dos Dias (1.6 m and 0.6 m telescopes, IAU code 874), Observatoire Haute-Provence (1.2m telescope, IAU code 511) and ESO (2.2 m telescope, IAU code 809).
+%In this context, we used 3 databases for deriving precise positions for the irregular satellites observed at Observatório do Pico dos Dias (1.6 m and 0.6 m telescopes, IAU code 874), Observatoire Haute-Provence (1.2m telescope, IAU code 511) and ESO (2.2 m telescope, IAU code 809).
 Many irregular satellites were observed between 1992 and 2014 covering a few orbital periods of these objects (12 satellites of Jupiter, 4 of Saturn, Sycorax of Uranus and Nereid of Neptune).
 
-\textbf{Since their ephemerides are not very precise, predict and observe stellar occultations are very difficult and no observation of such an event for an irregular satellite is found in the literature.
+%Since their ephemerides are not very precise, predict and observe stellar occultations are very difficult and no observation of such an event for an irregular satellite is found in the literature.
 The precise star positions to be derived by the ESA astrometry satellite Gaia \citep{deBruijne2012} will render better predictions with the only source of error being the ephemeris.
-The positions derived from \textbf{our} observations can be used in new orbital numerical integrations, generating more precise ephemerides.}
+The positions derived from \textbf{our} observations can be used in new orbital numerical integrations, generating more precise ephemerides.
 
-\textbf{The power of stellar occultations for observing relatively small diameter solar system objects is supported by recent works such as the discovery of a ring system around the Centaur (10199) Chariklo \citep{Braga-Ribas2014}.
+%The power of stellar occultations for observing relatively small diameter solar system objects is supported by recent works such as the discovery of a ring system around the Centaur (10199) Chariklo \citep{Braga-Ribas2014}.
 Once irregular satellites start to be observed by this technique, it will be possible to obtain their physical parameters (shape, size, albedo, density) with unprecedented precision.
 For instance, in this case, sizes could be obtained with kilometer accuracy.
-The knowledge of these parameters would in turn bring valuable information for the study of the capture mechanisms and origin of the irregular satellites.}
+The knowledge of these parameters would in turn bring valuable information for the study of the capture mechanisms and origin of the irregular satellites.
 
-The databases are described in Sect.
+%The databases are described in Sect.
 \ref{Sec: observations}.
@@
 
-\section{Databases} \label{Sec: observations}
-
-Our three databases consist in optical CCD images from many observational programs performed with different telescopes/detectors targeting a variety of objects, among which irregular satellites.
-The observations were made at 3 sites: Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO).
-Altogether there are more than 8000 FITS images obtained in a large time span (1992-2014) for the irregular satellites.
-Since the OHP and mostly the OPD database registers were not well organized, we had to start from scratch and develop an automatic procedure to identify and filter only the images of interest, that is, of the irregular satellites.
-The instruments and images characteristics are described in the following subsections.
-
-\subsection{OPD} \label{Subsec: observations-opd}
-
-The OPD database was produced at Observatório do Pico dos Dias (OPD, IAU code 874, $45^{\circ} ~34\arcmin ~57\arcsec$ W, $22^{\circ} ~32\arcmin ~04\arcsec$ S, 1864 m)\footnote{Website: http://www.lna.br/opd/opd.html - in Portuguese}, located at geographical longitude , in Brazil.
-The observations were made between 1992 and 2014 by our group in a variety of observational programs.
-%In Fig \ref{Fig: obsxyear-opd} we plot the number of frames obtained per satellite over time and in Fig \ref{Fig: obsxsat-opd} the number of frames per satellite for each telescope.
-Two telescopes of 0.6 m diameter (Zeiss and Boller \& Chivens) and one 1.6 m diameter (Perkin-Elmer) were used for the observations.
-It was identified 5248 observations containing irregular satellites, being 3168 from the Boller \& Chivens, 1967 from the Perkin-Elmer and 113 from the Zeiss.
-
-%\begin{figure*}
-%\begin{centering}
-%\includegraphics[width=17cm]{figures/framexobj_ano_opd}
-%\par\end{centering}
-%\caption{Distribution of the observations of the satellites over time at OPD. The graphic is limited in number of frames to 300 for better viewing. The number of frames of Phoebe in 2010 is 533.}
-%\label{Fig: obsxyear-opd}
-%\end{figure*}
-
-This is an inhomogeneous database with observations made with 9 different detectors (see Table \ref{Tab: OPD-CCDs}) and 6 different filters.
-The headers of most of the older FITS images had missing, incomplete or incorrect coordinates or date.
-In some cases, we could not identify the detector origin.
-The procedures used to overcome these problems are described in Sect.
-\ref{Sec: reduction}.
-
-\begin{table}[h!]
-\caption{\label{Tab: OPD-CCDs} Characteristics of OPD detectors used in this work.}
-\begin{centering}
-\begin{tabular}{ccc}
-\hline
-\hline
-\multicolumn{3}{c}{Perkin-Elmer}\tabularnewline
-% & \multicolumn{2}{c}{Exposure Times}\tabularnewline
-Detector & Image Size (pixel) & Pixel Scale ($\mu m$/px)\tabularnewline
-\hline
-CCD048 & 770 x 1152 & 22.5\tabularnewline
-CCD098 & 2048 x 2048 & 13.5\tabularnewline
-CCD101 & 1024 x 1024 & 24.0\tabularnewline
-CCD105 & 2048 x 2048 & 13.5\tabularnewline
-CCD106 & 1024 x 1024 & 24.0\tabularnewline
-CCD301 & 385 x 578 & 22.0\tabularnewline
-CCD523 & 455 x 512 & 19.0\tabularnewline
-IKON & 2048 x 2048 & 13.5\tabularnewline
-IXON & 1024 x 1024 & 13.5\tabularnewline
-\hline 
-\end{tabular} 
-\par\end{centering}
-The plate scale of the telescopes are 13.09"/mm for Perkin-Elmer, 25.09"/mm for Boller \& Chivens and 27.5"/mm for Zeiss.
-\end{table}
-
-%\begin{figure}
-%\begin{centering}
-%\includegraphics[scale=0.45]{figures/framexobj_opd}
-%\par\end{centering}
-%
-%\caption{Number of frames observed per satellite by OPD telescope.}
-%
-%
-%\label{Fig: obsxsat-opd}
-%\end{figure}
-
-\subsection{OHP} \label{Subsec: observations-ohp}
-
-The instrument used at the Observatoire de Haute Provence (OHP, IAU code 511, $5^{\circ} ~42\arcmin ~56.5\arcsec$ E, $43^{\circ} ~55\arcmin ~54.7\arcsec $N, 633.9 m)\footnote{Website: www.obs-hp.fr/guide/t120.shtml - in French} was the 1.2m-telescope in a Newton configuration.
-The focal length is 7.2 m.
-The observations were made between 1997 and 2008.
-During this time only one CCD detector $1024 \times 1024$ was used.
-The size of field is $12\arcmin \times 12\arcmin$ with a pixel scale of $0.69\arcsec$.
-%Fig.
-\ref{Fig: obsxyear-ohp} shows the distribution of the observation of the satellites over time and Fig.
-\ref{Fig: obsxsat-ohp} the number of frames observed for each satellite.
-From these observations, 2408 were identified containing irregular satellites.
-
-%\begin{figure*}
-%\begin{centering}
-%\includegraphics[scale=0.48]{figures/framexobj_ano_ohp}
-%\par\end{centering}
-%
-%\caption{Distribution of the observations of the satellites over time from observations at OHP.}
-%\label{Fig: obsxyear-ohp}
-%\end{figure*}
-%
-%\begin{figure}
-%\begin{centering}
-%\includegraphics[scale=0.45]{figures/framexobj_ohp}
-%\par\end{centering}
-%
-%\caption{Number of frames per satellite observed at OHP.}
-%
-%
-%\label{Fig: obsxsat-ohp}
-%\end{figure}
-
-\subsection{ESO} \label{Subsec: observations-eso}
-
-Observations were made at the 2.2 m Max-Planck ESO (ESO2p2) telescope (IAU code 809, $70\degr44\arcmin1.5\arcsec$ W, $29\degr15\arcmin31.8\arcsec$ S, 2345.4 m)\footnote{Website: www.eso.org/sci/facilities/lasilla/telescopes/\\national/2p2.html} with the Wide Field Imager (WFI) CCD mosaic detector.
-Each mosaic is composed by eight CCDs of $7.5\arcmin \times 15\arcmin$ ($\alpha$, $\delta$) sizes, resulting in a total coverage of $30\arcmin \times 30\arcmin$ per mosaic.
-Each CCD has $4 k \times 2 k$ pixels with a pixel scale of $0.238\arcsec$.
-The filter used was a broad-band R filter (ESO\#844) with $\lambda _{c}  = 651.725$ nm and $\Delta \lambda = 162.184$ nm.
-The telescope was shifted between exposures in such a way that each satellite was observed at least twice in different CCDs.
-
-The satellites were observed in 24 nights, divided in 5 runs, between April 2007 and May 2009 in paralel with, and using the same observational and astrometric procedures of the program that observed stars along the sky path of trans-neptunian objects (TNOs) to identify candidates to stellar occultation \citep[see][]{2010A&A...515A..32A, 2012A&A...541A.142A}.
-A total of 810 observations for irregular satellites were obtained.
-%Fig \ref{Fig: obsxsat-eso} shows the number of frames per satellite.
-
-%\begin{figure}
-%\begin{centering}
-%\includegraphics[scale=0.45]{figures/framexobj_eso}
-%\par\end{centering}
-%
-%\caption{Number of frames per satellite observed at ESO.}
-%
-%
-%\label{Fig: obsxsat-eso}
-%\end{figure}
-
-\section{Astrometry} \label{Sec: reduction}
-
-Almost all the frames were photometrically calibrated with auxiliary bias and flat-field frames by means of standard procedures using IRAF\footnote{Website: http://iraf.noao.edu/} and, for the mosaics, using the esowfi \citep{Jones2000} and mscred \citep{Valdes1998} packages.
-Some of the nights at OPD didn't have bias and flat-field images so the correction was not possible.
-
-The astrometric treatment was made with the Platform for Reduction of Astronomical Images Automatically (PRAIA) \citep{2011gfun.conf...85A}.
-The (x, y) measurements were performed with 2-dimensional circular symmetric Gaussian fits within one Full Width Half Maximum (FWHM = seeing).
-Within one FWHM, the image profile is well described by a Gaussian profile, free from the wing distortions, which may jeopardize the center determination.
-PRAIA automatically recognizes catalog stars and determines ($\alpha$, $\delta$) with a user-defined model relating the (x, y) measured and (X, Y) standard coordinates projected in the sky tangent plane.
-
-We used the UCAC4 \citep{2013AJ....145...44Z} as the practical representative of the International Celestial Reference System (ICRS).
-For each frame, we used the six constants polynomial model to relate the (x, y) measurements with the (X, Y) tangent plane coordinates.
-For ESO, we followed the same astrometric procedures described in detail in \cite{2012A&A...541A.142A}; the (x, y) measurements of the individual CCDs were pre-corrected by a field distortion pattern, and all positions coming from different CCDs and mosaics were then combined using a 3rd degree polynomial model to produce a global solution for each night and field observed, and final ($\alpha$, $\delta$) object positions were obtained in the UCAC4 system.
-
-In Table \ref{Tab: stars-errors} we list the average mean error in $\alpha$ and $\delta$ for the reference stars obtained by telescope, the average (x, y) measurement errors of the Gaussian fits described above and the mean number of UCAC4 stars used by frame.
-For all databases, about 20\% of outlier reference stars were eliminated for presenting (O-C) position residuals higher than 120 mas in the ($\alpha$, $\delta$) reductions.
-%##### OHP = 21\% de estrelas eliminadas, erro gaussiano: média de 26mas e 90\% entre 5 e 60mas para x e y; ESO = erro gaussiano: média de 14mas e 90\% entre 2 e 41mas; 160 = 13\% de estrelas eliminadas,  média de 15mas e 90\% entre 3 e 38mas ; IAG = 24\% d estrelas eliminadas, média de 29mas e 90\% entre 3 e 75mas ; Zeiss = 37\% de estrelas eliminadas, média de 26mas e 90\% entre 5 e 56mas#####
-
-\begin{table}
-\caption{\label{Tab: stars-errors} Astrometric ($\alpha$, $\delta$) reduction by telescope.}
-\begin{centering}
-\begin{tabular}{lccccc}
-\hline %\multicolumn{6}{c}{Perkin-Elmer} \tabularnewline
-\hline
- & \multicolumn{2}{c}{Mean errors} &  UCAC4 & \multicolumn{2}{c}{Gaus. errors}       \tabularnewline
-Telescope  & $\sigma_{\alpha}$  & $\sigma_{\delta}$  & stars & x & y  \tabularnewline
-  &   mas   &   mas & &   mas   &   mas  \tabularnewline
-\hline
-PE(OPD) & 51 & 48 & 24 & 15 & 15 \tabularnewline
-B\&C (OPD) & 56 & 55 & 36 & 29 & 29 \tabularnewline
-Zeiss (OPD) & 58 & 57 & 95 & 26 & 26 \tabularnewline
-OHP & 50 & 49 & 46 & 26 & 26 \tabularnewline
-ESO & 26 & 25 & 632 & 15 & 15 \tabularnewline
-\hline
-\end{tabular}
-\par\end{centering}
-Mean errors are the standard deviations in the (O$-$C) residuals from ($\alpha$, $\delta$) reductions with the UCAC4 catalog.
-Gaussian errors are the errors in the Gaussian fit used to perform the (x, y) measurements.
-\end{table}
-
-To help identifying the satellites in the frames, and derive the ephemeris for the instants of the observations for comparisons (see Sect \ref{Sec: comparison}), we used the kernels from SPICE/JPL\footnote{Website: http://naif.jpl.nasa.gov/naif/toolkit.html}.
-\cite{Emelyanov2008} and references therein also provided ephemeris of similar quality for the irregular satellites.
-For instance, for Himalia, which has relatively good orbit solutions, the ephemerides differ by less than $20 mas$, and in the case of less-known orbits, like Ananke, the differences are less than $90 mas$.
-Here, we choose to use the JPL ephemeris because they used more recent observations \cite[see][]{Jacobson2012}.
-The JPL ephemeris that represented the Jovian satellites in this work was the DE421 $+$ JUP300.
-For the Saturnian satellites the ephemeris was DE421 $+$ SAT359 to Hyperion, Iapetus and Phoebe and DE421 $+$ SAT361 to Albiorix, Siarnaq and Paaliaq.
-The DE421 $+$ URA095 was used for Sycorax and DE421 $+$ NEP081 for Nereid.
-More recent JPL ephemeris versions became available after completion of this work, but this did not affect the results.
-
-In the OPD database, there were some images (mostly the older ones) with missing coordinates or wrong date in their headers.
-In the case of missing or wrong coordinates, we adopted the ephemeris as the central coordinates of the frames.
-When the time was not correct, the FOV identification failed.
-In this case, a search for wrong date (year) displaying was performed.
-Problems like registering local time instead of UTC were also identified and corrected.
-
-In all databases, for each night a sigma-clipping procedure was performed to eliminate discrepant positions (outliers).
-A threshold of 120 mas and a deviation of more than 2.5 sigmas from the nightly average ephemeris offsets were adopted.
-
-From Table \ref{Tab: Reductions-160} to \ref{Tab: Reductions-eso} we list the average dispersion (standard deviation) of the position offsets with regard to the ephemeris for $\alpha$ and $\delta$ obtained by telescope for each satellite.
-The final number of frames, number of nights (in parenthesis), the mean number of UCAC4 stars used in the reduction and the approximate V magnitude are also given.
-The dashed lines separate the satellites from different families with similar orbital parameters: Himalia Group (Himalia, Elara, Lysithea and Leda), Pasiphae Group (Pasiphae, Callirrhoe and Megaclite) and Ananke Group (Ananke and Praxidike).
-Carme and Sinope are the only samples of their groups.
-From Saturn, Siarnaq and Paaliaq are from the Inuit Group while Phoebe and Albiorix are the only samples of their groups.
-
-The differences in the dispersion of the ephemeris offsets of the same satellite for distinct telescopes seen in Tables \ref{Tab: Reductions-160} to \ref{Tab: Reductions-eso} are caused by the different distribution of observations along the orbit for each telescope.
-This can be seen in Fig \ref{Fig: carme_anom} for Carme, \ref{Fig: pasiphae_anom} for Pasiphae and for all objects in the online material.
-Since the observations cover different segments of the orbit, the dispersion of the offsets may vary for different telescopes for a single satellite, with larger covered segments usually implying in larger dispersions and vice-versa.
-For Nereid, due to its high eccentric orbit, the observations are located between 90$\degree$ and 270$\degree$ of True Anomaly where Nereid remains most of the time.
-
-\begin{table}
-\caption{\label{Tab: Reductions-160} Astrometric ($\alpha$, $\delta$) reduction for each satellite observed with the Perkin-Elmer telescope.}
-\begin{centering}
-\begin{tabular}{lcccccc}
-\hline 
-%\multicolumn{6}{c}{Perkin-Elmer} \tabularnewline
-\hline
- & \multicolumn{2}{c}{Offsets (sigma)} & Nr &  UCAC4 & \tabularnewline
-Satellite  & $\sigma_{\alpha}$  & $\sigma_{\delta}$ & frames  & stars & Mag \tabularnewline
-  &   mas   &   mas  & (nights)  & &\tabularnewline
-\hline
-Himalia & 290 & 45 & 238 (18) & 37 & 14 \tabularnewline
-Elara & 230 & 118 & 99 (12) & 32 & 16\tabularnewline
-Lysithea & 107 & 79 & 53 (8) & 41 & 18\tabularnewline
-Leda & 207 & 79 & 6 (2) & 46 & 19\tabularnewline
-\hdashline
-Pasiphae & 157 & 92 & 144 (13) & 22 & 17 \tabularnewline
-Callirrhoe & 66 & 35  &   9 (1) & 3 & 21\tabularnewline
-\hdashline
-Carme & 97 & 94 & 68 (7) & 49 & 18\tabularnewline
-Sinope & 155 & 77 & 37 (8) & 42 & 18\tabularnewline
-Ananke & 93 & 185 & 52 (7) & 40 & 19\tabularnewline
-\hline
-Phoebe & 73 & 95 & 410 (22) & 6 & 16\tabularnewline
-\hline
-Nereid & 200 & 142 & 289 (29) & 8 & 19\tabularnewline
-\hline
-\end{tabular}
-\par\end{centering}
-The offsets (sigma) are the average standard deviations of the ephemeris offsets from the ($\alpha$, $\delta$) positions of the satellites.
-Also given are the approximate satellite V magnitude and the average number of UCAC4 reference stars per frame.
-\end{table}
-
-\begin{table}
-\caption{\label{Tab: Reductions-iag} Astrometric ($\alpha$, $\delta$) reduction for each satellite observed with the Boller \& Chivens telescope.}
-\begin{centering}
-\begin{tabular}{lccccc}
-\hline 
-%\multicolumn{6}{c}{Boller \& Chivens} \tabularnewline
-\hline
- & \multicolumn{2}{c}{Offsets (sigma)} & Nr &  UCAC4 & \tabularnewline
-Satellite  & $\sigma_{\alpha}$  & $\sigma_{\delta}$ & frames  & stars & Mag \tabularnewline
-  &   mas   &   mas  & (nights)  & &\tabularnewline
-\hline
-Himalia & 83 & 43 & 560 (31) & 57 & 14 \tabularnewline
-Elara & 55 & 43 & 294 (23) & 53 & 16 \tabularnewline
-Lysithea & 23 & 42 & 7 (2) & 60 & 18 \tabularnewline
-\hdashline
-Pasiphae & 128 & 71 & 140 (14) & 57 & 17 \tabularnewline
-Carme & 68 & 111 & 22 (4) & 45 & 18\tabularnewline
-Sinope & 59 & 17  &   4 (1) & 22 & 18\tabularnewline
-\hline
-Phoebe & 43 & 48 & 810 (42) & 17 & 16 \tabularnewline
-\hline
-Nereid & 61 & 45 & 514 (38) & 20 & 19 \tabularnewline
-\hline
-\end{tabular}
-\par\end{centering}
-Same as in Table \ref{Tab: Reductions-160}.
-%Mean errors are the standard deviations in the (O$-$C) residuals from ($\alpha$, $\delta$) reductions with the UCAC4 catalog.
-\end{table}
-
-\begin{table}
-\caption{\label{Tab: Reductions-zei} Astrometric ($\alpha$, $\delta$) reduction for each satellite observed with the Zeiss telescope.}
-\begin{centering}
-\begin{tabular}{lccccc}
-\hline 
-%\multicolumn{6}{c}{Zeiss} \tabularnewline
-\hline
- & \multicolumn{2}{c}{Offsets (sigma)} & Nr &  UCAC4 & \tabularnewline
-Satellite  & $\sigma_{\alpha}$  & $\sigma_{\delta}$ & frames  & stars & Mag \tabularnewline
-  &   mas   &   mas  & (nights)  & &\tabularnewline
-\hline
-Himalia & 112 & 72 & 56 (4) & 91 & 14\tabularnewline
-Elara & 17 & 21  &  10 (1) & 146 & 16 \tabularnewline
-\hdashline
-Pasiphae & 24 & 25  &  11 (1) & 140 & 17\tabularnewline
-\hline
-Phoebe & 37 & 30  &  19 (1) & 16 & 16\tabularnewline
-\hline
-\end{tabular}
-\par\end{centering}
-Same as in Table \ref{Tab: Reductions-160}.
-%Mean errors are the standard deviations in the (O$-$C) residuals from ($\alpha$, $\delta$) reductions with the UCAC4 catalog.
-\end{table}
-
-\begin{table}
-\caption{\label{Tab: Reductions-ohp} Astrometric ($\alpha$, $\delta$) reduction for each satellite observed with the OHP telescope.}
-\begin{centering}
-\begin{tabular}{lccccc}
-\hline
-%\multicolumn{6}{c}{OHP} \tabularnewline
-\hline
- & \multicolumn{2}{c}{Offsets (sigma)} & Nr &  UCAC4 & \tabularnewline
-Satellite  & $\sigma_{\alpha}$  & $\sigma_{\delta}$ & frames  & stars & Mag \tabularnewline
-  &   mas   &   mas  & (nights)  & &\tabularnewline
-\hline
-Himalia & 49 & 66 & 357 (43) & 49 & 14\tabularnewline
-Elara & 52 & 61 & 187 (25) & 37 & 16\tabularnewline
-Lysithea & 63 & 50 & 84 (13) & 56 & 18\tabularnewline
-Leda & 118 & 33 & 48 (7) & 14 & 19\tabularnewline
-\hdashline
-Pasiphae & 101 & 75 & 248 (32) & 39 & 17\tabularnewline
-Carme & 114 & 96 & 204 (29) & 39 & 18\tabularnewline
-Sinope & 196 & 73 & 169 (25) & 43 & 18\tabularnewline
-Ananke & 100 & 89 & 141 (20) & 62 & 19\tabularnewline
-\hline
-Phoebe & 30 & 31 & 516 (63) & 51 & 16\tabularnewline
-Siarnaq & 46 & 98 & 20 (6) & 32 & 20\tabularnewline
-\hline
-\end{tabular}
-\par\end{centering}
-Same as in Table \ref{Tab: Reductions-160}.
-%Mean errors are the standard deviations in the (O$-$C) residuals from ($\alpha$, $\delta$) reductions with the UCAC4 catalog.
-\end{table}
-
-\begin{table}
-\caption{\label{Tab: Reductions-eso} Astrometric ($\alpha$, $\delta$) reduction for each satellite observed with the ESO telescope.}
-\begin{centering}
-\begin{tabular}{lccccc}
-\hline 
-%\multicolumn{6}{c}{ESO} \tabularnewline
-\hline
- & \multicolumn{2}{c}{Offsets (sigma)} & Nr &  UCAC4 & \tabularnewline
-Satellite  & $\sigma_{\alpha}$  & $\sigma_{\delta}$ & frames  & stars & Mag \tabularnewline
-  &   mas   &   mas  & (nights)  & &\tabularnewline
-\hline
-Himalia & 76 & 74 & 23 (2) & 1153 & 14\tabularnewline
-Elara & 112 & 87 & 46 (4) & 1492 & 16\tabularnewline
-Lysithea & 76 & 88 & 90 (6) & 695 & 18\tabularnewline
-Leda & 60 & 125 & 44 (3) & 632 & 19\tabularnewline
-\hdashline
-Pasiphae & 70 & 114 & 66 (5) & 836 & 17\tabularnewline
-Callirrhoe & 29 & 33  &  16 (1) & 493 & 21\tabularnewline
-Megaclite & 52 & 34  &  10 (1) & 445  & 22\tabularnewline
-\hdashline
-Ananke & 225 & 19 & 57 (3) & 761 & 18\tabularnewline
-Praxidike & 7 & 38  &   2 (1) & 1934 & 21\tabularnewline
-\hdashline
-Carme & 140 & 110 & 37 (4) & 1074 & 18\tabularnewline
-Sinope & 339 & 70 & 11 (2) & 1542 & 18\tabularnewline
-Themisto & 894 & 28 & 16 (2) & 1232 & 21\tabularnewline
-\hline
-Phoebe & 102 & 57 & 32 (5) & 312 & 16\tabularnewline
-\hdashline
-Siarnaq & 86 & 66 & 56 (6) & 283 & 20\tabularnewline
-Paaliaq & 301 & 59 & 11 (4) & 382 & 21\tabularnewline
-\hdashline
-Albiorix & 76 & 50 & 46 (6) & 330 & 20\tabularnewline
-\hline
-Sycorax & 150 & 82 & 35 (9) & 375 & 21\tabularnewline
-\hline
-Nereid & 115 & 78 & 99 (12) & 362 & 19\tabularnewline
-\hline
-\end{tabular}
-\par\end{centering}
-Same as in Table \ref{Tab: Reductions-160}.
-%Mean errors are the standard deviations in the (O$-$C) residuals from ($\alpha$, $\delta$) reductions with the UCAC4 catalog.
-\end{table}
-
-%\begin{table*}
-%\caption{\label{Tab: offsets} Error by satellite by telescope}
-%\begin{centering}
-%\begin{tabular}{lccccccccccccccc}
-%\hline\hline 
-%Satellite & \multicolumn{3}{c}{Perkin-Elmer} & \multicolumn{3}{c}{Boller \& Chivens} & \multicolumn{3}{c}{Zeiss} & \multicolumn{3}{c}{OHP} & \multicolumn{3}{c}{ESO}\tabularnewline
-% & $\sigma_\alpha$ & $\sigma_\delta$ & N & $\sigma_\alpha$ & $\sigma_\delta$ & N & $\sigma_\alpha$ & $\sigma_\delta$ & N & $\sigma_\alpha$ & $\sigma_\delta$ & N & $\sigma_\alpha$ & $\sigma_\delta$ & N\tabularnewline
-%  & mas & mas & stars & mas & mas & stars & mas & mas & stars & mas & mas & stars & mas & mas & stars\tabularnewline
-%\hline
-%Ananke & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 18.8 & 70 & 179 & 249 & 1995& 600 & 1951 & 0 \tabularnewline
-%Callirrhoe & 20.6 & 9 & 5 & 14 & 2000 & 95 & 1999 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
-%Carme & 17.8 & 62 & 245 & 307 & 1995 & 973 & 1938 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
-%Elara & 16.6 & 349 & 223 & 572 & 1995 & 1115 & 1905 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
-%Himalia & 14.6 & 645 & 444 & 1089 & 1995 & 1757 & 1894 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
-%Leda & 20.1 & 12 & 58 & 70 & 1996 & 178 & 1974 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
-%Lysithea & 18.3 & 80 & 103 & 183 & 1995 & 431 & 1938 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
-%Pasiphae & 16.8 & 216 & 283 & 499 & 1995 & 1629 & 1908 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
-%Sinope & 18.2 & 52 & 210 & 262 & 1996 & 854 & 1914 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
-%\hline
-%Albiorix & 20.5 & - & 2 & 2 & 2001 & 137 & 2000 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
-%Phoebe & 16.4 & 937 & 582 & 1519 & 1995 & 3479 & 1898 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
-%Siarnaq & 19.9 & - & 25 & 25 & 2001 & 239 & 2000 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
-%\hline
-%Sycorax & 19.9 & - & 25 & 25 & 2001 & 239 & 2000 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
-%\hline
-%Nereid & 19.9 & - & 25 & 25 & 2001 & 239 & 2000 & 18.8 & 70 & 179 & 249 & 1995 & 600 & 1951 & 0 \tabularnewline
-%\hline 
-%\end{tabular} 
-%\par\end{centering}
-%\end{table*}
-
-No solar phase correction was applied to the positions.
-For the biggest irregular satellite of Jupiter, Himalia, it was verified that the maximum deviation in the position due to phase angle is 1.94 \textit{mas} using the phase correction described in \cite{Lindegren1977}.
-For the other satellites, which are smaller objects, this deviation is even smaller.
-Since our position error is one order of magnitude higher, this effect was neglected.
-%Phoebe = 1.32 e Nereid = 1.36
-
-\section{Satellite positions} \label{Sec: positions}
-
-The final set of positions of the satellites consists in 6523 catalogued positions observed between 1992 and 2014 for 12 satellites of Jupiter, 4 of Saturn, 1 of Uranus and 1 of Neptune.
-The topocentric positions are in the ICRS.
-The catalogues (one for each satellite) contain epoch of observations, the position error, filter used, estimated magnitude (from PSF fitting) and telescope origin.
-The magnitude errors can be as high as 1 mag; they are not photometrically calibrated and should be used with care.
-The position errors were estimated from the dispersion of the ephemeris offsets of the night of observation of each position.
-Thus, these position errors are probably overestimated, as there must be ephemeris errors present in the dispersion of the offsets.
-These position catalogues are freely available in electronic form at the CDS (see a sample in Table \ref{Tab: sample-cds}) and at the IAU NSDC data base at www.imcce.fr/nsdc.
-
-The number of positions acquired is significant compared to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012} as shown in Table \ref{Tab: comparison-horizons}.
-
-\begin{table*}
-\caption{\label{Tab: sample-cds} CDS data table sample for Himalia.}
-\begin{centering}
-\begin{tabular}{ccccccccc}
-\hline 
-%\multicolumn{9}{c}{Himalia} \tabularnewline
-\hline
-% & \multicolumn{2}{c}{Mean errors} & Nr & Nr &  UCAC4       \tabularnewline
-\multicolumn{2}{c}{RA  (ICRS)  Dec} & RA error & Dec error & Epoch & Mag & Filter & Telescope & \textbf{IAU code} \tabularnewline
-\multicolumn{1}{l}{\ h \ m \ \ \ s} & \multicolumn{1}{l}{\ \ \ $\degree$ \ \ \ $\arcmin$ \ \ \ $\arcsec$} &   (mas)   &   (mas)  & (jd) & & & & \tabularnewline
-\hline
- 16 59 11.6508 & -22 00 44.855 &  17 &  12 & 2454147.78241319 &   16.0 &    C &      BC & \textbf{874} \tabularnewline
- 16 59 11.6845 & -22 00 44.932 &  17 &  12 & 2454147.78332384 &   15.8 &    C &      BC & \textbf{874} \tabularnewline
- 16 59 11.7181 & -22 00 44.978 &  17 &  12 & 2454147.78422477 &   16.0 &    C &      BC & \textbf{874} \tabularnewline
- 16 59 11.7818 & -22 00 45.143 &  17 &  12 & 2454147.78602662 &   15.9 &    C &      BC & \textbf{874} \tabularnewline
- 16 59 11.8188 & -22 00 45.232 &  17 &  12 & 2454147.78693750 &   16.0 &    C &      BC & \textbf{874} \tabularnewline
- 17 17 11.0344 & -22 47 19.415 &  30 &  24 & 2454205.63885463 &   16.1 &        U &      BC & \textbf{874} \tabularnewline
- 17 17 11.0270 & -22 47 19.381 &  30 &  24 & 2454205.63959167 &   16.1 &        U &      BC & \textbf{874} \tabularnewline
- 17 17 11.0258 & -22 47 19.366 &  30 &  24 & 2454205.64031875 &   16.1 &        U &      BC & \textbf{874} \tabularnewline
- 17 17 11.0192 & -22 47 19.417 &  30 &  24 & 2454205.64104583 &   16.1 &        U &      BC & \textbf{874} \tabularnewline
-\hline
-\end{tabular}
-\par\end{centering}
-This sample corresponds to 9 observations of Himalia from February 16, 2007 and April 15, 2007.
-Tables contain the topocentric ICRS coordinates of the irregular satellites, the position error estimated from the dispersion of the ephemeris offsets of the night of observation, \textbf{the UTC time of the frame's mid-exposure in julian date}, the estimated magnitude, the filter used, the telescope origin \textbf{and correspondent IAU code}.
-The filters may be U, B, V, R or I following the Johnson system; C stands for clear (no filter used), resulting in a broader R band magnitude, RE for the broad-band R filter ESO\#844 with $\lambda_{c} = 651.725$ nm and $\Delta\lambda = 162.184 $ nm (full width at half maximum) and "un" for unknown filter.
-E, OH, PE, BC and Z stand respectively for the ESO, OHP, Perkin-Elmer, Bollen \& Chivens and Zeiss telescopes.
-\end{table*}
-
-\begin{table}
-\caption{\label{Tab: comparison-horizons} Comparison of positions obtained with \citealp{Jacobson2012}.}
-\begin{centering}
-\begin{tabular}{lccccc}
-\hline  \hline
- & \multicolumn{4}{c}{Number of Positions}   &    \tabularnewline
-Satellite  & OPD  & OHP & ESO & Total  & Jacobson \tabularnewline
-\hline
-Himalia & 854 & 357 & 23 & 1234 & 1757 \tabularnewline
-Elara & 403 & 187 & 46 & 636 & 1115 \tabularnewline
-Lysithea & 60 & 84 & 90 & 234 & 431 \tabularnewline
-Leda & 6 & 48 & 44 & 98 & 178 \tabularnewline
-\hdashline
-Pasiphae & 295 & 248 & 66 & 609 & 1629 \tabularnewline
-Callirrhoe & 9 & -  &  16 & 25 & 95 \tabularnewline
-Megaclite & - & -  &  10 & 10 & 50  \tabularnewline
-\hdashline
-Ananke & 52 & 141 & 57 & 250 & 600 \tabularnewline
-Praxidike & - & -  &   2 & 2 & 59 \tabularnewline
-\hdashline
-Carme & 90 & 204 & 37 & 331 & 973 \tabularnewline
-Sinope & 41 & 169 & 11 & 221 & 854 \tabularnewline
-Themisto & - & - & 16 & 16 & 55 \tabularnewline
-\hline
-Phoebe & 1239 & 516 & 32 & 1787 & 3479 \tabularnewline
-\hdashline
-Siarnaq & - & 20 & 56 & 76 & 239 \tabularnewline
-Paaliaq & - & - & 11 & 11 & 82 \tabularnewline
-\hdashline
-Albiorix & - & - & 46 & 46 & 137 \tabularnewline
-\hline
-Sycorax & - & - & 35 & 35 & 237 \tabularnewline
-\hline
-Nereid & 803 & - & 99 & 902 & 716 \tabularnewline
-\hline
-\end{tabular}
-\par\end{centering}
-Comparison between the number of positions obtained in our work with the number used in the numerical integration of orbits by the JPL as published by \citealp{Jacobson2012}.
-\end{table}
-
-\section{Comparison with ephemeris} \label{Sec: comparison}
-
-Intending to see the potential of our results to improve the orbit of the irregular satellites observed, we analysed the offsets of our positions with regard to the ephemeris mentioned in Sect.
-\ref{Sec: reduction}.
-Taking Carme as example, we plot in Fig.
-\ref{Fig: carme_anom} the mean ephemeris offsets for each night and their dispersions  (one sigma error bars) as a function of the true anomaly in right ascension (\ref{Fig: carme_alfa}) and declination (\ref{Fig: carme_delta}).
-Fig.
-\ref{Fig: carme_delta} clearly shows a systematic error in declination.
-When Carme is close to its apojove (true anomaly = 180\degree) its offsets are more likely to be more negative than those close to its perijove (true anomaly = 0\degree).
-The offsets obtained from observations by four telescopes using different cameras and filters are in good agreement, meaning that there is an error in the ephemeris of Carme, most probably due to an error in its orbital inclination.
-
-This pattern in declination was also seen for other satellites like Pasiphae (Fig: \ref{Fig: pasiphae_anom}) and Ananke (plots for other satellites with significant number of observations can be seen in the online material).
-For some satellites, the orbital coverage is not enough to clearly indicate the presence of systematic errors in specific orbital elements.
-However, comparing the internal position mean errors of the reductions (Table \ref{Tab: stars-errors}) with the external position errors estimated from the dispersion of the ephemeris offsets (Tables \ref{Tab: Reductions-160} to \ref{Tab: Reductions-eso}), we see position error values much larger than expected from the mean errors.
-This means that besides some expected astrometric errors, significant ephemeris errors must also be present.
-
 \section{Conclusions} \label{Sec: conclusions}
 
-\textbf{We managed a large database with FITS images acquired by 5 telescopes in 3 sites between 1992 and 2014.
-From that, we identified 8466 observations of irregular satellites, from which we managed to obtain 6523 suitable astrometric positions, giving a total of 3666 positions for 12 satellites of Jupiter, 1920 positions for 4 satellites of Saturn, 35 positions for Sycorax (Uranus) and 902 positions for Nereid (Neptune).}
+%We managed a large database with FITS images acquired by 5 telescopes in 3 sites between 1992 and 2014.
+From that, we identified 8466 observations of irregular satellites, from which we managed to obtain 6523 suitable astrometric positions, giving a total of 3666 positions for 12 satellites of Jupiter, 1920 positions for 4 satellites of Saturn, 35 positions for Sycorax (Uranus) and 902 positions for Nereid (Neptune).
 
-\textbf{The positions of all the objects were determined using the PRAIA package.
+%The positions of all the objects were determined using the PRAIA package.
 The package was suited to cope with the huge amount of observations and the task of identifying the satellites within the database.
-PRAIA tasks were also useful to deal with the missing or incorrect coordinate and time stamps present mostly in the old observations.}
+PRAIA tasks were also useful to deal with the missing or incorrect coordinate and time stamps present mostly in the old observations.The UCAC4 was used as the reference frame.
+Based in the comparisons with ephemeris, we estimate that the position errors are about 60 mas to 80 mas depending on the satellite brightness.
 
-\textbf{The UCAC4 was used as the reference frame. Based in the comparisons with ephemeris, we estimate that the position errors are about 60 mas to 80 mas depending on the satellite brightness. }
+%For some satellites the number of positions obtained in this work is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012} (see Table \ref{Tab: comparison-horizons}).
+For instance, the amount of new positions for Himalia is about 70\% of the number used in the numerical integation of orbits by JPL.
+Systematic errors in the ephemeris were found for at least some satellites (Ananke, Carme, Elara and Pasiphae).
+In the case of Carme, we evidenced an error in the orbital inclination (see Fig.
+\ref{Fig: carme_anom}).
 
-For some satellites the number \textbf{of positions obtained in this work} is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012} (see Table \ref{Tab: comparison-horizons}).
-\textbf{For instance, the amount of new positions for Himalia is about 70\% of the number used in the numerical integation of orbits by JPL}.
-Systematic errors in the ephemeris were found for at least some satellites (Ananke, Carme, Elara and Pasiphae).
-In the case of Carme, we evidenced an error in the orbital inclination \textbf{(see Fig.
-\ref{Fig: carme_anom})}.
-
-\textbf{The positions derived in this work can be used in new orbital numerical integrations, generating more precise ephemerides.
+%The positions derived in this work can be used in new orbital numerical integrations, generating more precise ephemerides.
 Stellar occultations by irregular satellites could then be better predicted.
 Based in this work, our group has already computed occultation predictions for the 8 major irregular satellites of Jupiter.
-These predictions will be published in a forthcoming paper.}
+These predictions will be published in a forthcoming paper.
 
@@
 
-ARGJ thanks the financial support of CAPES.
-M.
-A.
-thanks the CNPq (Grants 473002/2013-2 and 308721/2011-0) and FAPERJ (Grant E-26/111.488/2013).
-RVM thanks grants: CNPq-306885/2013, Capes/Cofecub-2506/2015, Faperj/PAPDRJ-45/2013.
-J-E.A.
-thanks the "Programme National de Planétologie" of INSU-CNRS-CNES for its financial support.
-J.I.B.
-Camargo acknowledges CNPq for a PQ2 fellowship (process number 308489/2013-6).
-F.B.R.
-acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014.
-B.E.M.
-thanks the financial support of CAPES.
+ARG-J thanks the financial support of CAPES.
+MA thanks the CNPq (Grants 473002/2013-2 and 308721/2011-0) and FAPERJ (Grant E-26/111.488/2013).
+RV-M thanks grants: CNPq-306885/2013, Capes/Cofecub-2506/2015, Faperj/PAPDRJ-45/2013.
+JIBC acknowledges CNPq for a PQ2 fellowship (process number 308489/2013-6).
+FB-R acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014.
+BEM thanks the financial support of CAPES.
 

### 2015-07-03 qualify/Qualify-altair.tex
começado a correção do qualify

@@
 
-\indent \indent O estudo de objetos como TNOs, Centauros e Satélites Irregulares nos ajudam a compreender a formação e evolução do Sistema Solar Exterior.
+\indent \indent O estudo de objetos como objetos trans-Neptunianos (TNOs), Centauros e Satélites Irregulares nos ajudam a compreender a formação e evolução do Sistema Solar Exterior.
 Nesta região distante do Sol, de baixas temperaturas, objetos principalmente de tamanho relativamente pequeno, e mais dispersos no espaço, provavelmente sofreram muito pouca diferenciação, seja por mecanismos internos, seja por choques com outros corpos, comparados a objetos formados mais próximos do Sol.
 
-Por serem corpos asteroidais localizados de forma dispersa além da órbita de Netuno, considera-se que os objetos trans-Neptunianos (TNOs) guardam estruturas e composições relativamente inalteradas em relação a sua época de formação, constituindo-se assim em corpos de prova de grande valor para o estudo da origem do Sistema Solar, ao menos para essa região exterior.
-
-Tirando Plutão e Tritão, o primeiro TNO foi descoberto a pouco mais de 20 anos \citep{Jewitt1993}.
+Por serem corpos asteroidais localizados de forma dispersa além da órbita de Netuno, considera-se que os TNOs guardam estruturas e composições relativamente inalteradas em relação a sua época de formação, constituindo-se assim em corpos de prova de grande valor para o estudo da origem do Sistema Solar, ao menos para essa região exterior.
+
+Tirando Plutão, o primeiro TNO foi descoberto há pouco mais de 20 anos \citep{Jewitt1993}.
 Por isso, as propriedades básicas desta população ainda não estão inteiramente estabelecidas, como a distribuição de tamanhos, composição, estruturas internas.
@@
 \indent \indent Ocultação estelar é um evento que ocorre quando um objeto do Sistema Solar passa na frente de uma estrela de forma que o brilho da estrela é completamente ou parcialmente bloqueado durante um intervalo de tempo.
-A variação de brilho da estrela pode ser medido fazendo-se observações com exposições curtas.
+A variação de brilho da estrela pode ser medida fazendo-se observações com exposições curtas.
 O produto gerado é uma curva de fluxo de luz x tempo.
@@
 Essas diversas curvas de luz permitirão reconstruir a forma do objeto, possivelmente identificando estruturas superficiais como montanhas e crateras.
-Como a estrela está muito distante, podendo ser considerada quase pontual, a sombra do objeto durante a ocultação é muito próximo ao seu tamanho real.
+Como a estrela está muito distante, podendo ser considerada quase pontual, a sombra do objeto durante a ocultação é muito próxima ao seu tamanho real.
 
@@
 Da mesma forma, estando muito distantes, eles se movem relativamente devagar no céu ocultando relativamente poucas estrelas por ano.
-Porém, sendo menos brilhantes (Mag $V > 18$) é possível observar ocultações de estrelas mais fracas (estrelas até $R = 17$), onde a queda de brilho da estrela durante a ocultação poderá ser detectada e, mesmo com baixo sinal-ruído (ex: 5 a 10), o tamanho do corpo pode ser calculado com precisão de pelo menos 10 a 20 km, melhorando a medida em que a velocidade da sombra seja menor que 20 km/s (velocidade típica para um corpo em oposição a 30 UA).
+Porém, sendo menos brilhantes (Mag $V > 18$) é possível observar ocultações de estrelas mais fracas (estrelas até $R = 17$), onde a queda de brilho da estrela durante a ocultação poderá ser detectada e, mesmo com baixo sinal-ruído (ex: 5 a 10), o tamanho do corpo pode ser calculado com precisão de pelo menos 10 a 20 km, melhorando à medida em que a velocidade da sombra seja menor que 20 km/s (velocidade típica para um corpo em oposição a 30 UA).
 
@@
 Posições para estrelas até magnitude R=21, com nível de completeza em R=19, foram obtidas com precisões menores que 20 mas até R=20 \citep{Assafin2012}.
-Assim, estrelas que não estão no UCAC4, catálogo comumente utilizado como referência astrométrica, foram identificados bem como estrelas que estão mas não tem posições ou movimentos próprios com precisão suficiente.
+Assim, estrelas que não estão no UCAC4, catálogo comumente utilizado como referência astrométrica, foram identificadas bem como estrelas que estão mas não tem posições ou movimentos próprios com precisão suficiente.
 Tantos os catálogos de estrelas quanto as cadidatas a ocultação foram publicados \citep{Assafin2010, Assafin2012, Camargo2013}.

### 2015-07-07 Irregular-occ/Irregular-occ.tex
começado a correção do qualify
.

@@
 
-%The irregular satellites of the giant planets are smaller than the regular moons, having more eccentric, inclined, distant and, in most cases, retrograde orbits.
-Due to their orbital configurations, it is largely accepted that these objects were captured in the early solar system \citep{Sheppard2003}.
+Irregular satellites, also known as outer satellites, of the giant planets are objects orbiting the planets from a distant, eccentric and highly inclined orbit, most of them are retrograde.
+Because of these peculiar orbits, it is largely accepted that these objects were captured by their planets in the early solar system \citep{Sheppard2005}.
+
+There is a number of capture mechanisms of objects by giant planets proposed in the literature.
+There is the Gas Drag in the primordial circumplanetary nebulae \citep{Sheppard2005} where the object would be affected by the gas drag and its velocity slowed down until it be captured by the planet.
+Another mechanism is called pull-down capture \citep{Sheppard2005}, where the mass of the planet would increase while the object was temporarily captured.
+
+A mechanism based in the Nice model \citep{Morbidelli2005, Tsiganis2005, Gomes2005} was proposed by \cite{Nesvorny2007} and, in the specific case of Jupiter with the modern Nice model, by \citealp{Nesvorny2014}.
+During the early solar system instability, encounters between the outer planets occurred.
+These planetary encounters could exchange energy and angular momentum between planets and the objects nearby making it possible for the capture of irregular bodies by the giant planets.
+In this scenario, the survival rate of prior-LHB (Late Heavy Bombardment) satellites is very small.
+
+Another important mechanism is the capture through collisional interactions \citep{Sheppard2005}.
+A collision between two small bodies in the Hill's sphere of the planet could generate fragmented objects and the dissipated energy could be such that some of these objects could be captured.
+
+Some of these objects are in dynamical groups with similar orbital elements, called families, similar to families found in the Main Asteroid Belt.
+These families may have been created by a parent body disrupted by collisions with comets or other satellites \citep{Nesvorny2004}.
+Collisions with comets are more likely to have occurred during the Late Heavy Bombardment (LHB) \citep{Gomes2005}.
+
+\cite{Nesvorny2003} studied the collision rates between irregular satellites and concluded that some satellites could have been removed by collision with a bigger satellite.
+The rate collision between satellites of the Himalia Group (Himalia, Elara, Lysithea and Leda, mainly), for instance, was found to be more than one during the solar system age suggesting that their current structure was originated by satellite-satellite collision.
+
+For Phoebe, ejected material from its surface caused by impacts could evolve due to Poynting-Robertson drag and collide with Iapetus causing the large variation in albedo observed on it \citep{Nesvorny2003}.
+Indeed, Cassini was able to detected in Phoebe an absorption feature at 2.42 $\mu m$ (probably CN combinations) that was also detected in the dark side of Iapetus \citep{Clark2005}.
+
+The region of origin of these object is not well known, \cite{Grav2003} and \cite{Grav2007} showed that the irregular satellites from the giant planest have their colors and spectral slopes similar to C-, D- and P-type asteroids, Centaurs and trans-neptunian objects (TNOs) suggesting that they have been originated from different locations in the early solar system.
+
+In this work, we study these objects as possible representatives of the small TNOs population.
+TNOs are objects that due to their distance may be highly preserved having their properties similar to those they had when they were formed then providing history and evolution of the outer solar system \citep{Camargo2013}.
+Due to their distance, the smaller objects from this region are more difficult to observe and study.
+
+In the intent to obtain their physical parameters (size, shape, albedo, density, etc) and help identify their origin locations, we will make use of the stellar occultation technique.
+This technique is the best one to obtain the physical parameters of the solar system objects from ground-based observations providing more accurate results than other techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
+
+Since their sizes are very small (see table \ref{}), predict the exact location and instant where the shadow will cross the Earth demands a good precision.
+For instance, the bigger irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of 40 mas, must have an error smaller than its size for being observable.
+For the other objects, smaller or farther than Himalia, the situation is more difficult.
+
+As pointed out by \cite{GomesJunior2015}, the ephemeris of the irregular satellites have errors that may reach 200 mas for some satellites.
+For an object at the distance of Jupiter, this represents an error bigger than 700 km in the shadow path.
+
+We present in this paper the stellar occultation predictions for the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke and Sinope), Phoebe from Saturn and Nereid from Neptune.
+In the section \ref{Sec: Rationale} we explore the scientific rationale for study the irregular satellites.
+In section \ref{Sec: Offset} we show the correction made to the ephemeris for better predict stellar occultations.
+In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites.
+Some test realized to confirm the predictions are presented in section \ref{Sec: testes} and the conclusion is given in section \ref{Sec: conclusions}.
+
+\section{Scientific Rationale} \label{Sec: Rationale}
 
@@
 The exception is Phoebe with a very accurate measurement of size with a mean radius error of $0.7 km$ \citep{Thomas2010}.
-
-%There is a number of capture mechanisms of objects by giant planets proposed in the literature.
-There is the Gas Drag in the primordial circumplanetary nebulae \citep{Sheppard2006} where the object would be affected by the gas drag and its velocity slowed down until it be captured by the planet.
-Another mechanism is called pull-down capture \citep{Sheppard2006}, where the mass of the planet would increase while the object was temporarily captured.
-
-%A mechanism based in the Nice model \citep{Morbidelli2005, Tsiganis2005, Gomes2005} was proposed by \citealp{Nesvorny2007} and, in the specific case of Jupiter with the modern Nice model, by \citealp{Nesvorny2014}.
-During the early solar system instability, encounters between the outer planets occurred.
-These planetary encounters could exchange energy and angular momentum between planets and the objects nearby making it possible for the capture of irregular bodies by the giant planets.
-In this scenario, the survival rate of prior-LHB (Late Heavy Bombardment) satellites is very small.
-
-%Another important mechanism is the capture through collisional interactions \citep{Sheppard2006}.
-A collision between two small bodies in the Hill's sphere of the planet could generate fragmented objects and the dissipated energy could be such that some of these objects could be captured.
-
-%Some of these objects are in dynamical groups with similar orbital elements, called families, similar to families found in the Main Asteroid Belt.
-These families may have been created by a parent body disrupted by collisions with comets or other satellites \citep{Nesvorny2004}.
-Collisions with comets are more likely to have occurred during the Late Heavy Bombardment (LHB) \citep{Gomes2005}.
-
-%\citealp{Nesvorny2003} studied the collision rates between irregular satellites and concluded that some satellites could have been removed by collision with a bigger satellite.
-The rate collision between satellites of the Himalia Group (Himalia, Elara, Lysithea and Leda, mainly), for instance, was found to be more than one during the solar system age suggesting that their current structure was originated by satellite-satellite collision.
-
-%For Phoebe, ejected material from its surface caused by impacts could evolve due to Poynting-Robertson drag and collide with Iapetus causing the large variation in albedo observed on it \citep{Nesvorny2003}.
-Indeed, Cassini was able to detected in Phoebe an absorption feature at 2.42 $\mu m$ (probably CN combinations) that was also detected in the dark side of Iapetus \citep{Clark2005}.
 
@@
 
-\section{Conclusions} \label{Sec: conclusions}
+\section{Correction of the ephemeris} \label{Sec: Offset}
+
+\section{Prediction of occultations} \label{Sec: predictions}
+
+%The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the corrected JPL ephemeris as presented in the section \ref{Sec: Offset}.
+
+\section{Occultation tests} \label{Sec: testes}
+
+\section{Conclusion} \label{Sec: conclusions}
 

### 2015-07-07 Irregular-occ/Irregular-occ.tex
.

@@
 %          \inst{1,2} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau, 75014 Paris, France},
+          B. E. Morgado
+          \inst{1},
           A. Dias-Oliveira
           \inst{2},
-          B. E. Morgado
-          \inst{1},
           G. Benedetti-Rossi
@@
 In the intent to obtain their physical parameters (size, shape, albedo, density, etc) and help identify their origin locations, we will make use of the stellar occultation technique.
-This technique is the best one to obtain the physical parameters of the solar system objects from ground-based observations providing more accurate results than other techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
+This technique is the best one to obtain the physical parameters of the solar system objects from ground-based observations providing more accurate results than other groud-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
 

### 2015-07-13 Irregular-occ/Irregular-occ.tex
.

--- 2015-07-07 Irregular-occ/Irregular-occ.tex
@@
 \usepackage{longtable}
+\usepackage{color}
 %\usepackage{setspace}
@@
           \inst{2,3},     
-          F. Braga-Ribas
-          \inst{2,4}, 
+%          F. Braga-Ribas
+%          \inst{2,4}, 
 %          D. N. da Silva Neto
@@
 %          \inst{1,2} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau, 75014 Paris, France},
-          A. Dias-Oliveira
-          \inst{2},
           B. E. Morgado
           \inst{1},
-          G. Benedetti-Rossi
-          \inst{2},
+%          A. Dias-Oliveira
+%          \inst{2},
+%          G. Benedetti-Rossi
+%          \inst{2},
 %          Y. Duchemin
@@
 %              \email{arlot@imcce.fr}
-              \and
-Federal University of Technology - Paraná (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
+%              \and
+%Federal University of Technology - Paraná (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
 %              \and
@@
 
-\keywords{Astrometry - Planets and satellites: general - Planets and satellites: individual: Jovian and Saturnian irregular satellites}
+\keywords{Occultations - Planets and satellites: general - Planets and satellites: individual: Jovian and Saturnian irregular satellites}
 
@@
 In this work, we study these objects as possible representatives of the small TNOs population.
-TNOs are objects that due to their distance may be highly preserved having their properties similar to those they had when they were formed then providing history and evolution of the outer solar system \citep{Camargo2013}.
+TNOs are objects that due to their distance may be highly preserved having their properties similar to those they had when they were formed, then providing history and evolution of the outer solar system \citep{Camargo2013}.
 Due to their distance, the smaller objects from this region are more difficult to observe and study.
 
-In the intent to obtain their physical parameters (size, shape, albedo, density, etc) and help identify their origin locations, we will make use of the stellar occultation technique.
-This technique is the best one to obtain the physical parameters of the solar system objects from ground-based observations providing more accurate results than other techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
+In the intent to obtain the physical parameters (size, shape, albedo, density, etc) of the irregular satellites and help identify their origin locations, we will make use of the stellar occultation technique.
+This technique is the best one to obtain these parameters of the solar system objects from ground-based observations providing more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
 
-Since their sizes are very small (see table \ref{}), predict the exact location and instant where the shadow will cross the Earth demands a good precision.
+Since their sizes are very small (see table \ref{} \textcolor{red}{Ainda vou fazer a tabela}), predict the exact location and instant where the shadow will cross the Earth demands a good precision.
 For instance, the bigger irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of 40 mas, must have an error smaller than its size for being observable.
@@
 
-We present in this paper the stellar occultation predictions for the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke and Sinope), Phoebe from Saturn and Nereid from Neptune.
-In the section \ref{Sec: Rationale} we explore the scientific rationale for study the irregular satellites.
+We present in this paper the stellar occultation predictions for the 7 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke and Sinope), Phoebe from Saturn and Nereid from Neptune.
+In the section \ref{Sec: Rationale} we explore the scientific rationale for study the irregular satellites and the possibility of having a common origin with TNOs.
 In section \ref{Sec: Offset} we show the correction made to the ephemeris for better predict stellar occultations.
-In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites.
+In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites and how they were made.
 Some test realized to confirm the predictions are presented in section \ref{Sec: testes} and the conclusion is given in section \ref{Sec: conclusions}.

### 2015-07-14 Irregular-occ/Irregular-occ.tex
.

@@
 
+\cite{GomesJunior2015} showed from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) that the orbits of the irregular satellites from the the giant planets have systematic errors.
+The offsets of the observations relative to the ephemeris could be up to 200 mas for some satellites.
+These differences could be associated with errors in their orbital elements.
+
 \section{Prediction of occultations} \label{Sec: predictions}

### 2015-07-15 Irregular-occ/Irregular-occ.tex
.

--- 2015-07-13 Irregular-occ/Irregular-occ.tex
@@
 
+As explicited in section \ref{Sec: introducao}
+
 %Because they are faint, the majority of these objects was discovered only in the last century\footnote{Website: http://ssd.jpl.nasa.gov/?sat\_discovery} .
@@
 \section{Correction of the ephemeris} \label{Sec: Offset}
+
+\cite{GomesJunior2015} showed from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) that the orbits of the irregular satellites of the giant planets have systematic errors.
+The offsets of the observations relative to the JPL ephemeris could be up to 200 mas for some satellites.
+These differences could be associated with errors in their orbital elements.
+
+We utilize the offsets obtained by \cite{GomesJunior2015} to identify a pattern in the error of the ephemeris.
+This pattern could be used to extrapolate an offset to the satellite by the time of the occultation predicted.
+Plots of the offsets over time and true anomaly (see Fig.
+\ref{} for ??) clearly show that these two parameters are the most important in the differences observed.
 

### 2015-07-28 Irregular-occ/Irregular-occ.tex
.

@@
 
-Since their sizes are very small (see table \ref{} \textcolor{red}{Ainda vou fazer a tabela}), predict the exact location and instant where the shadow will cross the Earth demands a good precision.
+Since their estimated sizes are very small (see table \ref{} \textcolor{red}{Ainda vou fazer a tabela}), predict the exact location and instant where the shadow will cross the Earth demands a good precision.
 For instance, the bigger irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of 40 mas, must have an error smaller than its size for being observable.
@@
 Plots of the offsets over time and true anomaly (see Fig.
-\ref{} for ??) clearly show that these two parameters are the most important in the differences observed.
+\ref{Fig:offxtime} and \ref{Fig:offxtanom} for Carme) clearly show that these two parameters are the most important in the differences observed.
+
+\begin{figure}
+\begin{centering}
+\includegraphics[scale=0.25]{figures/DEC.png}\label{Fig:offxtime}
+\caption{Offsets of the declination of Carme by time. \textcolor{red}{Figura só para visualização, vou colocar alguma melhor depois}}
+\end{centering}
+\end{figure}
+
+\begin{figure}
+\begin{centering}
+\includegraphics[scale=0.25]{figures/DEC_anom.png} \label{Fig:offxtanom}
+\caption{Offsets of the declination of Carme by true anomaly. \textcolor{red}{Figura só para visualização, vou colocar alguma melhor depois}}
+\end{centering}
+\end{figure}
+
+\begin{equation}
+F(t,f) = p[0]\times \frac{t - jd_{0}}{365.65} + p[1]\times \sin(f) + p[2]\times\cos(f) + p[3],
+\label{eq:linear}
+\end{equation}
+
+\begin{equation}
+\begin{split}
+%F = \{F_{x} \in  F_{c} &: (|S| > |C|) \\
+% &\quad \cap (\text{minPixels}  < |S| < \text{maxPixels}) \\
+% &\quad \cap (|S_{\text{conected}}| > |S| - \epsilon) \}
+F(t,f) = p[0]\times\sin\left(\frac{2\pi}{p[1]}\times \frac{t - jd_{0}}{365.25} + p[2]\right) + p[3]\times\sin(f) + \\ +p[4]\times\cos(f) + p[5],
+\end{split}
+\end{equation}
 

### 2015-07-28 Irregular-occ/Irregular-occ.tex
.

@@
 \makeatother
-\titlerunning{Astrometric Positions of the Irregular Satellites of Giant Planets}
-\authorrunning{Gomes-Júniot et al.}
+\titlerunning{Predictions of stellar occultations by irregular satellites}
+\authorrunning{Gomes-Júnior et al.}
 
@@
 
+\begin{figure*}
+\begin{centering}
+\subfigure[Sem offsets]{\includegraphics[scale=0.19]{figures/Himalia1_2015-03-03T00:39:25.png}  \label{Fig: occ-Himalia-nooff}}
+\subfigure[Offset de Himalia a partir do ajuste]{\includegraphics[scale=0.19]{figures/Himalia2_2015-03-03T00:39:18.png}  \label{Fig: occ-Himalia-ajuste}}
+\subfigure[Obs. dia 22 de Fevereiro (campos separados)]{\includegraphics[scale=0.19]{figures/Himalia3_2015-03-03T00:39:40.png}   \label{Fig: occ-Himalia-off22fev}}
+\subfigure[Obs. dia 03 de Março (mesmo campo)]{\includegraphics[scale=0.19]{figures/Himalia4_2015-03-03T00:39:15.png}   \label{Fig: occ-Himalia-off03mar}}
+\caption{Predictions for Himalia: O ponto grande em vermelho mostra o ponto de máxima aproximação geocêntrica da sombra na Terra, os pontos vermelhos menores são os centros da ocultação separados por um minuto, as linhas retas são os limites das sombras dado o tamanho estimado do objeto e as linhas tracejadas equivalem a uma diferença de 40 mas do centro da sombra representando o erro estimado da predição.
+(a) é o mapa utilizando apenas as posições nominais da estrela e do satélite.
+(b) mostra a sombra dado um offset estimado para a posição de Himalia dada sua anomalia verdadeira segundo um ajuste feito em cima dos offsets de efeméride encontrados em \cite{GomesJunior2015-Irregular}.
+Em (c) são aplicados offsets às posições da estrela e do satélite a partir de observações feitas em 22 de Fevereiro no telescópio Zeiss.
+Em (d), temos o mesmo que para (c) porém com offsets obtidos de observações feitas em 03 de Março no telescópio Perkin-Elmer quando os objetos estavam próximos no campo.\textcolor{red}{Figuras e texto ainda serão mudados.
+Colocado para visualização}}
+\label{Fig: occ-Himalia}
+\end{centering}
+\end{figure*}
+
 \section{Conclusion} \label{Sec: conclusions}

### 2015-07-30 Irregular-occ/Irregular-occ.tex
.

@@
 
-We utilize the offsets obtained by \cite{GomesJunior2015} to identify a pattern in the error of the ephemeris.
-This pattern could be used to extrapolate an offset to the satellite by the time of the occultation predicted.
+Making a new model for the orbits of these objects would demand a lot of time and delay the publication of these predictions.
+Then we utilize the offsets obtained by \cite{GomesJunior2015} to identify a pattern in the error of the ephemeris.
+This pattern could be used to extrapolate an offset to the satellite ephemeris by the time of the occultation predicted.
 Plots of the offsets over time and true anomaly (see Fig.
-\ref{Fig:offxtime} and \ref{Fig:offxtanom} for Carme) clearly show that these two parameters are the most important in the differences observed.
+\ref{Fig:offxtime} and \ref{Fig:offxtanom} for Carme) clearly show the systematic error in the JPL ephemeris.
 

### 2015-07-31 Irregulares/dados/Irregulares_results/final_CDS/NSDS.py
adicionado arquivos para NSDS

@@ -1,0 +1,352 @@
+import numpy as np
+from astropy.coordinates import SkyCoord
+from astropy.time import Time, TimeDelta
+import astropy.units as u
+import os
+
+###############################################################
+
+arquivo = ['search_results_ananke_OHP_final',
+ 'search_results_carme_OHP_final',
+ 'search_results_elara_OHP_final',
+ 'search_results_himalia_OHP_final',
+ 'search_results_leda_OHP_final',
+ 'search_results_lysithea_OHP_final',
+ 'search_results_pasiphae_OHP_final',
+ 'search_results_phoebe_OHP_final',
+ 'search_results_siarnaq_OHP_final',
+ 'search_results_sinope_OHP_final',
+ 'ucac4_Albiorix_eso_global_filtered_final',
+ 'ucac4_Ananke_160-1_filtered_final',
+ 'ucac4_Ananke_eso_global_filtered_final',
+ 'ucac4_Callirrhoe_160-1_filtered_final',
+ 'ucac4_Callirrhoe_eso_global_filtered_final',
+ 'ucac4_Carme_160_filtered_final',
+ 'ucac4_Carme_eso_global_filtered_final',
+ 'ucac4_Carme_IAG1_filtered_final',
+ 'ucac4_Elara_160_filtered_final',
+ 'ucac4_Elara_eso_global_filtered_final',
+ 'ucac4_Elara_IAG_filtered_final',
+ 'ucac4_Elara_zei1_filtered_final',
+ 'ucac4_Himalia_160_filtered_final',
+ 'ucac4_Himalia_eso_global_filtered_final',
+ 'ucac4_Himalia_IAG_filtered_final',
+ 'ucac4_Himalia_zei_filtered_final',
+ 'ucac4_Leda_160_filtered_final',
+ 'ucac4_Leda_eso_global_filtered_final',
+ 'ucac4_Lysithea_160-1_filtered_final',
+ 'ucac4_Lysithea_eso_global_filtered_final',
+ 'ucac4_Lysithea_IAG1_filtered_final',
+ 'ucac4_Megaclite_eso_global_filtered_final',
[…]

### 2015-07-31 Irregulares/dados/Irregulares_results/final_CDS/NSDC.py
.

@@ -110,57 +110,57 @@ table = ['Ananke_ucac4_OHP.table_filtered',
  'Sycorax_ucac4_ESO.table_filtered',
  'Themisto_ucac4_ESO.table_filtered']
 
-output = ['NSDS/NSDS_Ananke_OHP.txt',
-'NSDS/NSDS_Carme_OHP.txt',
-'NSDS/NSDS_Elara_OHP.txt',
-'NSDS/NSDS_Himalia_OHP.txt',
-'NSDS/NSDS_Leda_OHP.txt',
-'NSDS/NSDS_Lysithea_OHP.txt',
-'NSDS/NSDS_Pasiphae_OHP.txt',
-'NSDS/NSDS_Phoebe_OHP.txt',
-'NSDS/NSDS_Siarnaq_OHP.txt',
-'NSDS/NSDS_Sinope_OHP.txt',
-'NSDS/NSDS_Albiorix_ESO.txt',
-'NSDS/NSDS_Ananke_160.txt',
-'NSDS/NSDS_Ananke_ESO.txt',
-'NSDS/NSDS_Callirrhoe_160.txt',
-'NSDS/NSDS_Callirrhoe_ESO.txt',
-'NSDS/NSDS_Carme_160.txt',
-'NSDS/NSDS_Carme_ESO.txt',
-'NSDS/NSDS_Carme_IAG.txt',
-'NSDS/NSDS_Elara_160.txt',
-'NSDS/NSDS_Elara_ESO.txt',
-'NSDS/NSDS_Elara_IAG.txt',
-'NSDS/NSDS_Elara_ZEI.txt',
-'NSDS/NSDS_Himalia_160.txt',
-'NSDS/NSDS_Himalia_ESO.txt',
-'NSDS/NSDS_Himalia_IAG.txt',
-'NSDS/NSDS_Himalia_ZEI.txt',
-'NSDS/NSDS_Leda_160.txt',
-'NSDS/NSDS_Leda_ESO.txt',
-'NSDS/NSDS_Lysithea_160.txt',
-'NSDS/NSDS_Lysithea_ESO.txt',
-'NSDS/NSDS_Lysithea_IAG.txt',
-'NSDS/NSDS_Megaclite_ESO.txt',
-'NSDS/NSDS_Nereid_160.txt',
-'NSDS/NSDS_Nereid_ESO.txt',
[…]

### 2015-07-31 Irregular-occ/Irregular-occ.tex
continuado artigo de ocultacoes dos irregulares

@@
 
-As explicited in section \ref{Sec: introducao}
+As explicited in section \ref{Sec: introducao}, there is no consensus on a single model about the region where the irregular satellites were formed.
+In this work we treat them as having a common origin to the Centaurs and TNOs.
+In this case, they would be the representative objects of the small TNOs population.
+Since that small objects at great distances are difficult to study, the irregular satellites can complement this gap.
+
+This idea is supported by the
 
@@
 
+\textcolor{red}{Falta justificar as funções}
+
 \section{Prediction of occultations} \label{Sec: predictions}
 
-%The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the corrected JPL ephemeris as presented in the section \ref{Sec: Offset}.
+The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the corrected JPL ephemeris as presented in the section \ref{Sec: Offset}.
+The procedure to search for stellar candidates follows the same procedure presented by \cite{Assafin2010, Assafin2012} and \cite{Camargo2013}.
+
+For Nereid, the candidates for stellar occultations in 2015 and 2016 was searched using the WFI catalogue in the same way as the predictions for Centaurs and TNOs occultations by \cite{Assafin2010, Assafin2012} and \cite{Camargo2013}.
+This catalogue contains the stars in the path of Neptune in the sky up to mid-2016.
+It was generated by observations made at the ESO 2p2 telescope (IAU code 809) using the Wide Field Imager (WFI) CCD mosaic detector.
+The filter used was the broad-band R filter ESO\#844 with $\lambda_c$ = 651.725 nm and $\Delta\lambda$ = 162.184 nm
 

### 2015-08-03 Irregular-occ/Irregular-occ.tex
Continuado artigo occ irregulares

@@
 
-We present in this paper the stellar occultation predictions for the 7 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke and Sinope), Phoebe from Saturn and Nereid from Neptune.
+We present in this paper the stellar occultation predictions for the 7 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke and Sinope), Phoebe from Saturn and Triton and Nereid from Neptune.
 In the section \ref{Sec: Rationale} we explore the scientific rationale for study the irregular satellites and the possibility of having a common origin with TNOs.
@@
 
-This idea is supported by the
+%This idea is supported by the
 
@@
 \begin{equation}
-F(t,f) = p[0]\times \frac{t - jd_{0}}{365.65} + p[1]\times \sin(f) + p[2]\times\cos(f) + p[3],
+F(t,f) = p[0]\times t + p[1]\times \sin(f) + p[2]\times\cos(f) + p[3],
 \label{eq:linear}
@@
 % &\quad \cap (|S_{\text{conected}}| > |S| - \epsilon) \}
-F(t,f) = p[0]\times\sin\left(\frac{2\pi}{p[1]}\times \frac{t - jd_{0}}{365.25} + p[2]\right) + p[3]\times\sin(f) + \\ +p[4]\times\cos(f) + p[5],
+F(t,f) = p[0]\times\sin\left(\frac{2\pi}{p[1]}\times t + p[2]\right) + p[3]\times\sin(f) + \\ +p[4]\times\cos(f) + p[5],
 \end{split}
 \end{equation}
+where $F(t,f)$ is the offset obtained, $t$ is time in years counting from J2000.0 and $f$ is the true anomaly.
 
@@
 The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the corrected JPL ephemeris as presented in the section \ref{Sec: Offset}.
-The procedure to search for stellar candidates follows the same procedure presented by \cite{Assafin2010, Assafin2012} and \cite{Camargo2013}.
+The search for stellar candidates follows the same procedure as presented by \cite{Assafin2010, Assafin2012} and \cite{Camargo2013}.
 
-For Nereid, the candidates for stellar occultations in 2015 and 2016 was searched using the WFI catalogue in the same way as the predictions for Centaurs and TNOs occultations by \cite{Assafin2010, Assafin2012} and \cite{Camargo2013}.
+It was predicted occultation by the 7 major irregular satellites of Jupiter, Phoebe of Saturn and Triton and Nereid of Neptune.
+
+For Triton and Nereid, the candidates for stellar occultations in 2015 and 2016 was searched using the WFI catalogue in the same way as the predictions for Centaurs and TNOs occultations by \cite{Assafin2010, Assafin2012} and \cite{Camargo2013}.
 This catalogue contains the stars in the path of Neptune in the sky up to mid-2016.
-It was generated by observations made at the ESO 2p2 telescope (IAU code 809) using the Wide Field Imager (WFI) CCD mosaic detector.
-The filter used was the broad-band R filter ESO\#844 with $\lambda_c$ = 651.725 nm and $\Delta\lambda$ = 162.184 nm
+The catalogue was generated by observations made at the ESO 2p2 telescope (IAU code 809) using the Wide Field Imager (WFI) CCD mosaic detector.
+The filter used was the broad-band R filter ESO\#844 with $\lambda_c$ = 651.725 nm and $\Delta\lambda$ = 162.184 nm.
+
+A total of 588 events \textcolor{red}{(Não estou contando ainda os eventos para Tritão)} were identified between January 2015 and December 2017.
+Table \ref{Tab: occ-list} exemplifies the catalogue of occultations generated and their parameters, which is necessary to produce occultation maps (see Fig.
+\ref{Fig: ocultacao} as an example).
+Since these objects are very small, the duration of each event is a few seconds.
+
+\begin{table*}
+\caption{\label{Tab: sample-cds} Extract of prediction table. \textcolor{red}{Ainda tenho que fazer a tabela melhor, encaixar. Abaixo é um exemplo apenas para visualização.}}
+\begin{centering}
+\begin{tabular}{ccccccccccccccccccccccc}
+\hline 
+%\multicolumn{9}{c}{Himalia} \tabularnewline
+\hline
+% & \multicolumn{2}{c}{Mean errors} & Nr & Nr &  UCAC4       \tabularnewline
+d m Year & h m s & \multicolumn{2}{c}{RA (ICRS) Dec} & C/A & P/A & v & D & R* & $\lambda$ & LST & $\Delta e_{\alpha *}$ & $\Delta e_{\delta}$ & pm & ct & $E_{\alpha *}$ & $E_{\delta}$ & $\mu_{\alpha *}$ & $\mu_{\delta}$ \tabularnewline
+\hline
+ 07 10 2015 & 21 52 34. &  10 53 40.1835 & +08 21  00.489 & 0.465 & 200.31 & 39.31 & 6.25 & 16.6 & 179. & 09:48 & -31.0 & -13.0 & ok & uc &   0  &  0 & -13 &  -7 \tabularnewline
+ 23 10 2015 & 07 54 50. &  11 05 26.6037 & +07 14  32.030 & 1.017 & 201.33  & 36.11 & 6.12 & 15.5 & 16. & 08:59 & -31.0 & -13.0 & ok & uc &  0  &  0 &  -7 &  -4 \tabularnewline
+%\texttt{ d  m year  h:m:s UT     ra\_\_\_dec\_\_\_J2000\_candidate     ra\_dec\_J2000\_target\_geocen     C/A    P/A     vel  Delta  R*   J*   H*   K*   long  loc.
+t.
+off\_ra   off\_de pm ct f E\_ra E\_de pmra pmde} \tabularnewline
+%\hline
+% 05 01 2015  01 07  4.
+09 32 20.9233 +15 35  10.584   09 32 20.9885 +15 35  12.220   1.888   29.94 -11.41  4.51 14.8 -0.6 -0.6 -0.6    22.
+02:35     -31.0    -13.0 ok uc 0    0    0   -4    1 \tabularnewline
+% 15 01 2015  01 03 15.
+09 29  0.9265 +16 01  18.932   09 29  0.9610 +16 01  19.900   1.088   27.21 -13.17  4.45 14.8 -0.5 -0.5 -0.5    12.
+01:52     -31.0    -13.0 ok uc 0    0    0   -2  -16 \tabularnewline
+% 02 02 2015  08 03 30.
+09 21 46.5321 +16 49  43.658   09 21 46.4857 +16 49  42.068   1.723  202.72 -14.32  4.40 16.0 -0.4 -0.4 -0.4   247.
+00:32     -31.0    -13.0 ok uc 0    0    0    1   -8 \tabularnewline
+% 14 02 2015  16 11 59.
+09 16 41.4840 +17 18   2.448   09 16 41.4473 +17 18   0.976   1.563  199.68 -13.64  4.42 12.5 -0.4 -0.4 -0.4   112.
+23:39     -31.0    -13.0 ok uc 0    0    0    5    0 \tabularnewline
+% 25 02 2015  19 27 38.
+09 12 25.6457 +17 38   6.563   09 12 25.6804 +17 38   8.217   1.727   16.72 -12.01  4.47 12.2 -0.6 -0.6 -0.6    51.
+22:51     -31.0    -13.0 ok uc 0    0    0   -3    4 \tabularnewline
+% 28 02 2015  09 17 38.
+09 11 31.5455 +17 41  53.701   09 11 31.5720 +17 41  55.023   1.375   15.96 -11.50  4.49 14.1 -0.6 -0.6 -0.6   201.
+22:40     -31.0    -13.0 ok uc 0    0    0  -16    6 \tabularnewline
+% 03 03 2015  00 39 17.
+09 10 38.6352 +17 45  26.467   09 10 38.6218 +17 45  25.758   0.734  195.15 -10.94  4.50 14.1 -0.7 -0.7 -0.7   327.
+22:28     -31.0    -13.0 ok uc 0    0    0   -5    2 \tabularnewline
+% 20 05 2015  08 59 41.
+09 13 37.6521 +16 48  39.598   09 13 37.6915 +16 48  41.011   1.522   21.84  17.60  5.48 15.2 -0.1 -0.1 -0.1   126.
+17:22     -31.0    -13.0 ok uc 0    0    0    0    0 \tabularnewline
+% 22 05 2015  06 52 27.
+09 14 25.8437 +16 44   5.646   09 14 25.8400 +16 44   5.511   0.147  201.62  18.32  5.51 16.4 -0.1 -0.1 -0.1   156.
+17:16     -31.0    -13.0 ok uc 0    0    0  -13    2 \tabularnewline
+% 02 06 2015  16 14 18.
+09 19 46.8676 +16 14  30.501   09 19 46.8693 +16 14  30.569   0.078   20.53  22.38  5.67 14.5  0.1  0.1  0.1     6.
+16:36     -31.0    -13.0 ok uc 0    0    0   -7  -15 \tabularnewline
+% 06 06 2015  02 41  6.
+09 21 34.2809 +16 04  53.249   09 21 34.2951 +16 04  53.800   0.588   20.32  23.54  5.71 16.4  0.2  0.2  0.2   206.
+16:24     -31.0    -13.0 ok uc 0    0    0    2    5 \tabularnewline
+% 08 06 2015  20 25 31.
+09 23  3.2029 +15 57   1.059   09 23  3.2057 +15 57   1.170   0.122   20.13  24.44  5.75 16.7  0.2  0.2  0.2   297.
+16:15     -31.0    -13.0 ok uc 0    0    0   -3   -6 \tabularnewline
+% 09 06 2015  22 29 56.
+09 23 39.2777 +15 53  51.969   09 23 39.2451 +15 53  50.681   1.372  200.09  24.80  5.76 11.6  0.2  0.2  0.2   265.
+16:11     -31.0    -13.0 ok uc 0    0    0    5  -24 \tabularnewline
+% 20 06 2015  16 15 15.
+09 29 57.6850 +15 21   0.007   09 29 57.7127 +15 21   1.135   1.198   19.56  28.11  5.90 16.4  0.4  0.4  0.4   350.
+15:35     -31.0    -13.0 ok uc 0    0    0   -8   -1 \tabularnewline
+% 05 07 2015  17 02 21.
+09 39 45.6144 +14 31  18.362   09 39 45.6473 +14 31  19.747   1.465   19.04  32.07  6.06 16.7  0.5  0.5  0.5   326.
+14:46     -31.0    -13.0 ok uc 0    0    0    0   -9 \tabularnewline
+% 16 07 2015  03 15 38.
+09 47  5.8941 +13 54  48.764   09 47  5.8853 +13 54  48.387   0.400  198.77  34.42  6.15 16.4  0.6  0.6  0.6   164.
+14:12     -31.0    -13.0 ok uc 0    0    0  -19  -10 \tabularnewline
+% 07 10 2015  21 52 34.
+10 53 40.1835 +08 21   0.489   10 53 40.1726 +08 21   0.054   0.465  200.31  39.31  6.25 16.6  0.7  0.7  0.7   179.
+09:48     -31.0    -13.0 ok uc 0    0    0  -13   -7 \tabularnewline
+% 23 10 2015  07 54 50.
+11 05 26.6037 +07 14  32.030   11 05 26.5789 +07 14  31.083   1.017  201.33  36.11  6.12 15.5  0.6  0.6  0.6    16.
+08:59     -31.0    -13.0 ok uc 0    0    0   -7   -4 \tabularnewline
+% 31 10 2015  00 24 19.
+11 10 56.2964 +06 42   7.110   11 10 56.3414 +06 42   8.782   1.801   21.84  33.94  6.04 16.9  0.6  0.6  0.6   123.
+08:34     -31.0    -13.0 ok uc 0    0    0   -4  -13 \tabularnewline
+% 07 11 2015  21 22 53.
+11 16 14.3041 +06 10   5.194   11 16 14.2963 +06 10   4.912   0.306  202.36  31.33  5.94 16.2  0.5  0.5  0.5   162.
+08:09     -31.0    -13.0 ok uc 0    0    0    0    0 \tabularnewline
+% 08 11 2015  04 27 21.
+11 16 25.7881 +06 08  55.566   11 16 25.7606 +06 08  54.571   1.077  202.38  31.23  5.94 16.9  0.5  0.5  0.5    55.
+08:08     -31.0    -13.0 ok uc 0    0    0    9  -10 \tabularnewline
+% 09 11 2015  09 01  4.
+11 17 11.6890 +06 04  11.474   11 17 11.7030 +06 04  11.980   0.549   22.45  30.79  5.92 15.9  0.5  0.5  0.5   346.
+08:04     -31.0    -13.0 ok uc 0    0    0   19  -13 \tabularnewline
+% 11 11 2015  12 58 46.
+11 18 33.8286 +05 55  43.695   11 18 33.8357 +05 55  43.946   0.274   22.59  29.98  5.90 14.6  0.4  0.4  0.4   285.
+07:57     -31.0    -13.0 ok uc 0    0    0  -14    1 \tabularnewline
+% 12 11 2015  20 14 22.
+11 19 22.2740 +05 50  40.732   11 19 22.3183 +05 50  42.312   1.713   22.68  29.48  5.88 15.7  0.4  0.4  0.4   175.
+07:52     -31.0    -13.0 ok uc 0    0    0   -7   -4 \tabularnewline
+% 04 12 2015  03 31 56.
+11 30 41.9967 +04 37  36.039   11 30 42.0020 +04 37  36.217   0.196   24.18  20.14  5.56 15.0  0.0  0.0  0.0    47.
+06:40     -31.0    -13.0 ok uc 0    0    0    5   -9 \tabularnewline
+
+% 31 10 2015  00 24 19.
+11 10 56.2964 +06 42   7.110   11 10 56.3414 +06 42   8.782   1.801   21.84  33.94  6.04 16.9  0.6  0.6  0.6   123.
+08:34     -31.0    -13.0 ok uc 0    0    0   -4  -13 \tabularnewline
+% 07 11 2015  21 22 53.
+11 16 14.3041 +06 10   5.194   11 16 14.2963 +06 10   4.912   0.306  202.36  31.33  5.94 16.2  0.5  0.5  0.5   162.
+08:09     -31.0    -13.0 ok uc 0    0    0    0    0 \tabularnewline
+% 08 11 2015  04 27 21.
+11 16 25.7881 +06 08  55.566   11 16 25.7606 +06 08  54.571   1.077  202.38  31.23  5.94 16.9  0.5  0.5  0.5    55.
+08:08     -31.0    -13.0 ok uc 0    0    0    9  -10 \tabularnewline
+% 09 11 2015  09 01  4.
+11 17 11.6890 +06 04  11.474   11 17 11.7030 +06 04  11.980   0.549   22.45  30.79  5.92 15.9  0.5  0.5  0.5   346.
+08:04     -31.0    -13.0 ok uc 0    0    0   19  -13 \tabularnewline
+% 11 11 2015  12 58 46.
+11 18 33.8286 +05 55  43.695   11 18 33.8357 +05 55  43.946   0.274   22.59  29.98  5.90 14.6  0.4  0.4  0.4   285.
+07:57     -31.0    -13.0 ok uc 0    0    0  -14    1 \tabularnewline
+% 12 11 2015  20 14 22.
+11 19 22.2740 +05 50  40.732   11 19 22.3183 +05 50  42.312   1.713   22.68  29.48  5.88 15.7  0.4  0.4  0.4   175.
+07:52     -31.0    -13.0 ok uc 0    0    0   -7   -4 \tabularnewline
+% 04 12 2015  03 31 56.
+11 30 41.9967 +04 37  36.039   11 30 42.0020 +04 37  36.217   0.196   24.18  20.14  5.56 15.0  0.0  0.0  0.0    47.
+06:40     -31.0    -13.0 ok uc 0    0    0    5   -9 \tabularnewline
+\hline
+\end{tabular}
+\par\end{centering}
+\textbf{Notes}.
+Occultation table: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - in the original table, these coordinates are immediately followed by the geocentric astrometric equatorial coordinates (corrected for the offset or ephemeris method), not presented here, of the occulting body; C/A: the geocentric closest approach, in arcseconds; P/A: the planet position angle with respect to the occulted star at C/A, in degrees; velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; D: planet range to Earth, in AU; R*, J*, H*, K*: normalized magnitudes to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{Mag}* = \textrm{Mag}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$.
+A value of 50.0 means that the star is not in the 2MASS; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\Delta e_{\alpha *}$ and $\Delta e_{\delta}$:offset in mas applied to the ephemeris right ascension and declination, respectively; pm: ok = proper motion applied, no = no proper motion applied; catalogue cross-identification (ct) = uc (UCAC2), 2m (2MASS), fs (field star); f = multiplicity flag (see Table 3); $E_{\alpha *}$ and $E_{\delta}$: uncertainties (mas) in right ascension and declination.
+A value of 9999 means that there was no estimation of the respective uncertainty; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
+\label{Tab: occ-list}
+\end{table*}
+
+\begin{figure}
+\begin{centering}
+\includegraphics[scale=0.2]{figures/Himalia_2015-10-23T07:54:47.png} 
+\caption{Occultation map for Himalia \textcolor{red}{Mapa de Himalia para exemplificar, posso colocar outro depois}.
+Compare the information given in caption with the second entry of Table \ref{Tab: occ-list}}
+\label{Fig: ocultacao}
+\end{centering}
+\end{figure}
 
@@
 
+Observe a stellar occultation by an irregular satellites demands a great effort.
+The shadow covers a very restricted area on Earth because the irregular satellites are small.
+So before we start a large observational campaign, we tested some occultation predictions for larger objects, to assess the quality of the prediction.
+
+This test consists in observing the object and star to be occulted near the event when the two objects are present in the same field of view, preferably when the objects are close to each other.
+Thus the relative positions between the two objects will have minimal influence of the errors of the catalogue of stars used and possible field distortions.
+
+To date, two occultation tests were performed, one by Himalia that occurred on March 3, 2015 and the second by Elara that occurred on March 30, 2015.
+For each event, four maps were generated: the first with the nominal positions of the star and the satellite to the predicted time; the second with the offset calculated as described in section \ref{Sec: Offset}; the third with star and satellite offsets from observations made a few days before the occultation when the two objects were separated (different FOV); and the fourth from observations made when the star and the satellite were close in the same FOV.
+
+Figure \ref{Fig: occ-Himalia}.
+Shows the four maps for Himalia occultation test in March 03, 2015.
+The map \ref{Fig: occ-Himalia-nooff} is the nominal prediction with the coordinate of the star given by the catalogue and that of the satellite from the ephemeris.
+We then corrected the positions of the satellite by an offset calculated by the method showed in section \ref{Sec: Offset} to make the map \ref{Fig: occ-Himalia-ajuste}.
+
+The map \ref{Fig: occ-Himalia-off22fev} was made from obtained positions on February 22 observed with the Zeiss telescope at the Observatório do Pico dos Dias (OPD).
+On that day, Himalia and the star were observed in separate FOVs as they were still far apart.
+On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope at OPD just over an hour after the time scheduled for the event.
+Satellite and star were separated by about 16 arcsec, so very close to each other.
+From the calculated offsets, the map \ref{Fig: occ-Himalia-off03mar} was generated.
+
 \begin{figure*}
 \begin{centering}
-\subfigure[Sem offsets]{\includegraphics[scale=0.19]{figures/Himalia1_2015-03-03T00:39:25.png}  \label{Fig: occ-Himalia-nooff}}
-\subfigure[Offset de Himalia a partir do ajuste]{\includegraphics[scale=0.19]{figures/Himalia2_2015-03-03T00:39:18.png}  \label{Fig: occ-Himalia-ajuste}}
-\subfigure[Obs. dia 22 de Fevereiro (campos separados)]{\includegraphics[scale=0.19]{figures/Himalia3_2015-03-03T00:39:40.png}   \label{Fig: occ-Himalia-off22fev}}
-\subfigure[Obs. dia 03 de Março (mesmo campo)]{\includegraphics[scale=0.19]{figures/Himalia4_2015-03-03T00:39:15.png}   \label{Fig: occ-Himalia-off03mar}}
-\caption{Predictions for Himalia: O ponto grande em vermelho mostra o ponto de máxima aproximação geocêntrica da sombra na Terra, os pontos vermelhos menores são os centros da ocultação separados por um minuto, as linhas retas são os limites das sombras dado o tamanho estimado do objeto e as linhas tracejadas equivalem a uma diferença de 40 mas do centro da sombra representando o erro estimado da predição.
-(a) é o mapa utilizando apenas as posições nominais da estrela e do satélite.
-(b) mostra a sombra dado um offset estimado para a posição de Himalia dada sua anomalia verdadeira segundo um ajuste feito em cima dos offsets de efeméride encontrados em \cite{GomesJunior2015-Irregular}.
-Em (c) são aplicados offsets às posições da estrela e do satélite a partir de observações feitas em 22 de Fevereiro no telescópio Zeiss.
-Em (d), temos o mesmo que para (c) porém com offsets obtidos de observações feitas em 03 de Março no telescópio Perkin-Elmer quando os objetos estavam próximos no campo.\textcolor{red}{Figuras e texto ainda serão mudados.
-Colocado para visualização}}
+\subfigure[Nominal prediction]{\includegraphics[scale=0.19]{figures/Himalia1_2015-03-03T00:39:25.png}  \label{Fig: occ-Himalia-nooff}}
+\subfigure[\textcolor{red}{Offset de Himalia a partir do ajuste}]{\includegraphics[scale=0.19]{figures/Himalia2_2015-03-03T00:39:18.png}  \label{Fig: occ-Himalia-ajuste}}
+\subfigure[February 22 observations (different FOVs)]{\includegraphics[scale=0.19]{figures/Himalia3_2015-03-03T00:39:40.png}   \label{Fig: occ-Himalia-off22fev}}
+\subfigure[March 03 observation (same FOV)]{\includegraphics[scale=0.19]{figures/Himalia4_2015-03-03T00:39:15.png}   \label{Fig: occ-Himalia-off03mar}}
+\caption{Predictions for Himalia: The big red dot show the geocentric closest approach of the shadow.
+The small red ones are the center of the shadow separated by one minute.
+The straight lines show the size of the shadow.
+\ref{Fig: occ-Himalia-nooff} is the map using the nominal positions of the star and satellite.
+\ref{Fig: occ-Himalia-ajuste} shows the shadow given an estimated offset for the position of Himalia related to the JPL ephemeris obtained in section \ref{Sec: Offset}.
+In \ref{Fig: occ-Himalia-off22fev} we apply offsets to the positions of star and satellite accordingly the observations made at February 22.
+In \ref{Fig: occ-Himalia-off03mar} is as in \ref{Fig: occ-Himalia-off22fev} but with observations made at March 03 when the objects were close to each other.
+\textcolor{red}{Figuras ainda serão mudadas.
+Colocadas para visualização}}
 \label{Fig: occ-Himalia}
@@
 \end{figure*}
+
+In this event, it is possible to see that the shade does not vary much among the four maps suggesting that, at least for Himalia, there is a greater probability of observing an event.
+In fact, the biggest difference between the shadows of the four maps are 25s and 130km in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
+
+\begin{table}
+\caption{\label{Tab: comparison-Himalia} Comparison between the predictions of Himalia. \textcolor{red}{Ainda vou terminar de preencher a tabela e tem que ver direito o offset do ajuste}}
+\begin{centering}
+\begin{tabular}{lcc}
+\hline  \hline
+\multicolumn{3}{c}{Difference from nominal prediction} \tabularnewline
+Method  & Instant of C/A  & C/A   \tabularnewline
+\hline
+Nominal & 2015-03-03 00:39:25 UT & $0.714\arcsec$ \tabularnewline
+Offset & -07 s & \tabularnewline
+Feb. 22 Obs. & +15 s &  \tabularnewline
+Mar. 03 Obs. & -10 s & \tabularnewline
+\hline
+\end{tabular}
+\par\end{centering}
+\end{table}
+
+The second test was with the satellite Elara, which is the second biggest irregular satellites of Jupiter.
+The event was predict the occur at March 30, 2015.
+The observations were taken on March 25 and April 2, 2015 with the Zeiss telescope.
+On the night of April 2 they could still be observed in the same FOV.
+Due to Elara be much weaker, dispersions of the satellite positions on both nights ended up being higher than for Himalia.
+Still, the differences between the maps obtained were relatively small.
+The biggest difference between them is 73s and 302km (see table \ref{Tab: comparison-Elara}).
+
+\begin{table}
+\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara. \textcolor{red}{Ainda vou terminar de preencher a tabela e tem que ver direito o offset do ajuste}}
+\begin{centering}
+\begin{tabular}{lcc}
+\hline  \hline
+\multicolumn{3}{c}{Difference from nominal prediction} \tabularnewline
+Method  & Instant of C/A  & C/A  \tabularnewline
+\hline
+Nominal & 2015-03-30 01:45:15 UT & $1.139\arcsec$ \tabularnewline
+Offset & -16 s & \tabularnewline
+Mar. 25 Obs. & -58 s &  \tabularnewline
+Apr. 02 Obs. & +15 s & \tabularnewline
+\hline
+\end{tabular}
+\par\end{centering}
+\end{table}
 

### 2015-08-04 Irregular-occ/Irregular-occ.tex
.

@@
  23 10 2015 & 07 54 50. &  11 05 26.6037 & +07 14  32.030 & 1.017 & 201.33  & 36.11 & 6.12 & 15.5 & 16. & 08:59 & -31.0 & -13.0 & ok & uc &  0  &  0 &  -7 &  -4 \tabularnewline
-%\texttt{ d  m year  h:m:s UT     ra\_\_\_dec\_\_\_J2000\_candidate     ra\_dec\_J2000\_target\_geocen     C/A    P/A     vel  Delta  R*   J*   H*   K*   long  loc.
-t.
-off\_ra   off\_de pm ct f E\_ra E\_de pmra pmde} \tabularnewline
-%\hline
-% 05 01 2015  01 07  4.
-09 32 20.9233 +15 35  10.584   09 32 20.9885 +15 35  12.220   1.888   29.94 -11.41  4.51 14.8 -0.6 -0.6 -0.6    22.
-02:35     -31.0    -13.0 ok uc 0    0    0   -4    1 \tabularnewline
-% 15 01 2015  01 03 15.
-09 29  0.9265 +16 01  18.932   09 29  0.9610 +16 01  19.900   1.088   27.21 -13.17  4.45 14.8 -0.5 -0.5 -0.5    12.
-01:52     -31.0    -13.0 ok uc 0    0    0   -2  -16 \tabularnewline
-% 02 02 2015  08 03 30.
-09 21 46.5321 +16 49  43.658   09 21 46.4857 +16 49  42.068   1.723  202.72 -14.32  4.40 16.0 -0.4 -0.4 -0.4   247.
-00:32     -31.0    -13.0 ok uc 0    0    0    1   -8 \tabularnewline
-% 14 02 2015  16 11 59.
-09 16 41.4840 +17 18   2.448   09 16 41.4473 +17 18   0.976   1.563  199.68 -13.64  4.42 12.5 -0.4 -0.4 -0.4   112.
-23:39     -31.0    -13.0 ok uc 0    0    0    5    0 \tabularnewline
-% 25 02 2015  19 27 38.
-09 12 25.6457 +17 38   6.563   09 12 25.6804 +17 38   8.217   1.727   16.72 -12.01  4.47 12.2 -0.6 -0.6 -0.6    51.
-22:51     -31.0    -13.0 ok uc 0    0    0   -3    4 \tabularnewline
-% 28 02 2015  09 17 38.
-09 11 31.5455 +17 41  53.701   09 11 31.5720 +17 41  55.023   1.375   15.96 -11.50  4.49 14.1 -0.6 -0.6 -0.6   201.
-22:40     -31.0    -13.0 ok uc 0    0    0  -16    6 \tabularnewline
-% 03 03 2015  00 39 17.
-09 10 38.6352 +17 45  26.467   09 10 38.6218 +17 45  25.758   0.734  195.15 -10.94  4.50 14.1 -0.7 -0.7 -0.7   327.
-22:28     -31.0    -13.0 ok uc 0    0    0   -5    2 \tabularnewline
-% 20 05 2015  08 59 41.
-09 13 37.6521 +16 48  39.598   09 13 37.6915 +16 48  41.011   1.522   21.84  17.60  5.48 15.2 -0.1 -0.1 -0.1   126.
-17:22     -31.0    -13.0 ok uc 0    0    0    0    0 \tabularnewline
-% 22 05 2015  06 52 27.
-09 14 25.8437 +16 44   5.646   09 14 25.8400 +16 44   5.511   0.147  201.62  18.32  5.51 16.4 -0.1 -0.1 -0.1   156.
-17:16     -31.0    -13.0 ok uc 0    0    0  -13    2 \tabularnewline
-% 02 06 2015  16 14 18.
-09 19 46.8676 +16 14  30.501   09 19 46.8693 +16 14  30.569   0.078   20.53  22.38  5.67 14.5  0.1  0.1  0.1     6.
-16:36     -31.0    -13.0 ok uc 0    0    0   -7  -15 \tabularnewline
-% 06 06 2015  02 41  6.
-09 21 34.2809 +16 04  53.249   09 21 34.2951 +16 04  53.800   0.588   20.32  23.54  5.71 16.4  0.2  0.2  0.2   206.
-16:24     -31.0    -13.0 ok uc 0    0    0    2    5 \tabularnewline
-% 08 06 2015  20 25 31.
-09 23  3.2029 +15 57   1.059   09 23  3.2057 +15 57   1.170   0.122   20.13  24.44  5.75 16.7  0.2  0.2  0.2   297.
-16:15     -31.0    -13.0 ok uc 0    0    0   -3   -6 \tabularnewline
-% 09 06 2015  22 29 56.
-09 23 39.2777 +15 53  51.969   09 23 39.2451 +15 53  50.681   1.372  200.09  24.80  5.76 11.6  0.2  0.2  0.2   265.
-16:11     -31.0    -13.0 ok uc 0    0    0    5  -24 \tabularnewline
-% 20 06 2015  16 15 15.
-09 29 57.6850 +15 21   0.007   09 29 57.7127 +15 21   1.135   1.198   19.56  28.11  5.90 16.4  0.4  0.4  0.4   350.
-15:35     -31.0    -13.0 ok uc 0    0    0   -8   -1 \tabularnewline
-% 05 07 2015  17 02 21.
-09 39 45.6144 +14 31  18.362   09 39 45.6473 +14 31  19.747   1.465   19.04  32.07  6.06 16.7  0.5  0.5  0.5   326.
-14:46     -31.0    -13.0 ok uc 0    0    0    0   -9 \tabularnewline
-% 16 07 2015  03 15 38.
-09 47  5.8941 +13 54  48.764   09 47  5.8853 +13 54  48.387   0.400  198.77  34.42  6.15 16.4  0.6  0.6  0.6   164.
-14:12     -31.0    -13.0 ok uc 0    0    0  -19  -10 \tabularnewline
-% 07 10 2015  21 52 34.
-10 53 40.1835 +08 21   0.489   10 53 40.1726 +08 21   0.054   0.465  200.31  39.31  6.25 16.6  0.7  0.7  0.7   179.
-09:48     -31.0    -13.0 ok uc 0    0    0  -13   -7 \tabularnewline
-% 23 10 2015  07 54 50.
-11 05 26.6037 +07 14  32.030   11 05 26.5789 +07 14  31.083   1.017  201.33  36.11  6.12 15.5  0.6  0.6  0.6    16.
-08:59     -31.0    -13.0 ok uc 0    0    0   -7   -4 \tabularnewline
-% 31 10 2015  00 24 19.
-11 10 56.2964 +06 42   7.110   11 10 56.3414 +06 42   8.782   1.801   21.84  33.94  6.04 16.9  0.6  0.6  0.6   123.
-08:34     -31.0    -13.0 ok uc 0    0    0   -4  -13 \tabularnewline
-% 07 11 2015  21 22 53.
-11 16 14.3041 +06 10   5.194   11 16 14.2963 +06 10   4.912   0.306  202.36  31.33  5.94 16.2  0.5  0.5  0.5   162.
-08:09     -31.0    -13.0 ok uc 0    0    0    0    0 \tabularnewline
-% 08 11 2015  04 27 21.
-11 16 25.7881 +06 08  55.566   11 16 25.7606 +06 08  54.571   1.077  202.38  31.23  5.94 16.9  0.5  0.5  0.5    55.
-08:08     -31.0    -13.0 ok uc 0    0    0    9  -10 \tabularnewline
-% 09 11 2015  09 01  4.
-11 17 11.6890 +06 04  11.474   11 17 11.7030 +06 04  11.980   0.549   22.45  30.79  5.92 15.9  0.5  0.5  0.5   346.
-08:04     -31.0    -13.0 ok uc 0    0    0   19  -13 \tabularnewline
-% 11 11 2015  12 58 46.
-11 18 33.8286 +05 55  43.695   11 18 33.8357 +05 55  43.946   0.274   22.59  29.98  5.90 14.6  0.4  0.4  0.4   285.
-07:57     -31.0    -13.0 ok uc 0    0    0  -14    1 \tabularnewline
-% 12 11 2015  20 14 22.
-11 19 22.2740 +05 50  40.732   11 19 22.3183 +05 50  42.312   1.713   22.68  29.48  5.88 15.7  0.4  0.4  0.4   175.
-07:52     -31.0    -13.0 ok uc 0    0    0   -7   -4 \tabularnewline
-% 04 12 2015  03 31 56.
-11 30 41.9967 +04 37  36.039   11 30 42.0020 +04 37  36.217   0.196   24.18  20.14  5.56 15.0  0.0  0.0  0.0    47.
-06:40     -31.0    -13.0 ok uc 0    0    0    5   -9 \tabularnewline
-
-% 31 10 2015  00 24 19.
-11 10 56.2964 +06 42   7.110   11 10 56.3414 +06 42   8.782   1.801   21.84  33.94  6.04 16.9  0.6  0.6  0.6   123.
-08:34     -31.0    -13.0 ok uc 0    0    0   -4  -13 \tabularnewline
-% 07 11 2015  21 22 53.
-11 16 14.3041 +06 10   5.194   11 16 14.2963 +06 10   4.912   0.306  202.36  31.33  5.94 16.2  0.5  0.5  0.5   162.
-08:09     -31.0    -13.0 ok uc 0    0    0    0    0 \tabularnewline
-% 08 11 2015  04 27 21.
-11 16 25.7881 +06 08  55.566   11 16 25.7606 +06 08  54.571   1.077  202.38  31.23  5.94 16.9  0.5  0.5  0.5    55.
-08:08     -31.0    -13.0 ok uc 0    0    0    9  -10 \tabularnewline
+ 31 10 2015 & 00 24 19. &  11 10 56.2964 & +06 42  07.110 &  1.801 &  21.84 & 33.94 &  6.04 & 16.9 &  123. & 08:34 & -31.0 & -13.0 & ok & uc & 0  &  0 & -4 & -13 \tabularnewline
+ 07 11 2015 &  21 22 53. &  11 16 14.3041 & +06 10  05.194 &  0.306 &  202.36 & 31.33 & 5.94 & 16.2 & 162. & 08:09 & -31.0 & -13.0 & ok & uc & 0 & 0 & 0 & 0 \tabularnewline
+ 08 11 2015 & 04 27 21. &  11 16 25.7881 & +06 08  55.566 & 1.077 & 202.38 & 31.23 & 5.94 & 16.9 & 55. & 08:08 & -31.0 & -13.0 & ok & uc & 0 & 0 & 9 & -10 \tabularnewline
 % 09 11 2015  09 01  4.
@@
 \textbf{Notes}.
-Occultation table: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - in the original table, these coordinates are immediately followed by the geocentric astrometric equatorial coordinates (corrected for the offset or ephemeris method), not presented here, of the occulting body; C/A: the geocentric closest approach, in arcseconds; P/A: the planet position angle with respect to the occulted star at C/A, in degrees; velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; D: planet range to Earth, in AU; R*, J*, H*, K*: normalized magnitudes to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{Mag}* = \textrm{Mag}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$.
+Occultation table: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - in the original table, these coordinates are immediately followed by the geocentric astrometric equatorial coordinates (corrected for the offset or ephemeris method), not presented here, of the occulting body; C/A: the geocentric closest approach, in arcseconds; P/A: the planet position angle with respect to the occulted star at C/A, in degrees; velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; D: planet range to Earth, in AU; R*: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{Mag}* = \textrm{Mag}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$.
 A value of 50.0 means that the star is not in the 2MASS; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\Delta e_{\alpha *}$ and $\Delta e_{\delta}$:offset in mas applied to the ephemeris right ascension and declination, respectively; pm: ok = proper motion applied, no = no proper motion applied; catalogue cross-identification (ct) = uc (UCAC2), 2m (2MASS), fs (field star); f = multiplicity flag (see Table 3); $E_{\alpha *}$ and $E_{\delta}$: uncertainties (mas) in right ascension and declination.
@@
 This test consists in observing the object and star to be occulted near the event when the two objects are present in the same field of view, preferably when the objects are close to each other.
-Thus the relative positions between the two objects will have minimal influence of the errors of the catalogue of stars used and possible field distortions.
+Thus the relative positions between the two objects will have minimal influence of the errors of the catalogue of stars used and possible field distortions \citep[and references therein]{Peng2008}.
+The offsets of the positions of the star and the satellites will be used to correct the predictions.
 
@@
 \subfigure[\textcolor{red}{Offset de Himalia a partir do ajuste}]{\includegraphics[scale=0.19]{figures/Himalia2_2015-03-03T00:39:18.png}  \label{Fig: occ-Himalia-ajuste}}
-\subfigure[February 22 observations (different FOVs)]{\includegraphics[scale=0.19]{figures/Himalia3_2015-03-03T00:39:40.png}   \label{Fig: occ-Himalia-off22fev}}
-\subfigure[March 03 observation (same FOV)]{\includegraphics[scale=0.19]{figures/Himalia4_2015-03-03T00:39:15.png}   \label{Fig: occ-Himalia-off03mar}}
+\subfigure[Observation at February 22 (different FOVs)]{\includegraphics[scale=0.19]{figures/Himalia3_2015-03-03T00:39:40.png}   \label{Fig: occ-Himalia-off22fev}}
+\subfigure[Observation at March 03 (same FOV)]{\includegraphics[scale=0.19]{figures/Himalia4_2015-03-03T00:39:15.png}   \label{Fig: occ-Himalia-off03mar}}
 \caption{Predictions for Himalia: The big red dot show the geocentric closest approach of the shadow.

### 2015-08-07 Irregular-occ/Irregular-occ.tex
adicionado comentarios Marcelo 04-08-2015, fiz secoes 1 a 3

@@
 
-\title{Predictions of stellar occultations of 9 irregular satellites of giant planets
+\title{Stellar occultation predictions for  9 irregular satellites of giant planets for 2015-2016
 %\fnmsep\thanks{Table 8 are only available in electronic form at the CDS via anonymous ftp to cdsarc.u-strasbg.fr (130.79.128.5) or via http://cdsweb.u-strasbg.fr/cgi-bin/qcat?J/A+A/ and IAU NSDC data base at www.imcce.fr/nsdc.}
@@
 
-Irregular satellites, also known as outer satellites, of the giant planets are objects orbiting the planets from a distant, eccentric and highly inclined orbit, most of them are retrograde.
-Because of these peculiar orbits, it is largely accepted that these objects were captured by their planets in the early solar system \citep{Sheppard2005}.
+Irregular (outer) satellites revolve around giant planets at large distances in eccentric, highly inclined and frequently retrograde orbits.
+Because of these peculiar orbits, it is largely accepted that these objects were not originated with, but were captured by their planets in the early solar system \citep{Sheppard2005}.
 
@@
 These families may have been created by a parent body disrupted by collisions with comets or other satellites \citep{Nesvorny2004}.
-Collisions with comets are more likely to have occurred during the Late Heavy Bombardment (LHB) \citep{Gomes2005}.
+Collisions with comets are more likely to have occurred during the LHB \citep{Gomes2005}.
 
 \cite{Nesvorny2003} studied the collision rates between irregular satellites and concluded that some satellites could have been removed by collision with a bigger satellite.
-The rate collision between satellites of the Himalia Group (Himalia, Elara, Lysithea and Leda, mainly), for instance, was found to be more than one during the solar system age suggesting that their current structure was originated by satellite-satellite collision.
+The collision rate between satellites of the Himalia Group (Himalia, Elara, Lysithea and Leda, mainly), for instance, was found to be more than one during the solar system age suggesting that their current structure was originated by satellite-satellite collision.
 
@@
 
-The region of origin of these object is not well known, \cite{Grav2003} and \cite{Grav2007} showed that the irregular satellites from the giant planest have their colors and spectral slopes similar to C-, D- and P-type asteroids, Centaurs and trans-neptunian objects (TNOs) suggesting that they have been originated from different locations in the early solar system.
+The region of origin of these object is not well known.
+\cite{Grav2003} and \cite{Grav2007} showed that the irregular satellites from the giant planets have their colors and spectral slopes similar to C-, D- and P-type asteroids, Centaurs and trans-neptunian objects (TNOs) suggesting that they have been originated from different locations in the early solar system.
 
-In this work, we study these objects as possible representatives of the small TNOs population.
-TNOs are objects that due to their distance may be highly preserved having their properties similar to those they had when they were formed, then providing history and evolution of the outer solar system \citep{Camargo2013}.
-Due to their distance, the smaller objects from this region are more difficult to observe and study.
+In this work, we propose to study these objects as possible representatives of the small TNOs' population.
+The basis for this hypothesis is discussed in Section \ref{Sec: Rationale}.
+TNOs are highly interesting objects that due to their large heliocentric distances (low temperatures) may be highly preserved having their physical properties similar to those they had when they were formed, then providing history and evolution of the outer solar system \citep{Camargo2013}.
+This is even more true for the smaller objects, since in principle larger sizes favour physical differentiation processes in the body and vice-versa.
+However, due to the distance, the smaller TNOs from this region are more difficult to observe.
+A clever way to overcome this difficulty is to study  the much closer irregular satellites, under the hypothesis that they share a common origin with the small TNOs' population.
 
-In the intent to obtain the physical parameters (size, shape, albedo, density, etc) of the irregular satellites and help identify their origin locations, we will make use of the stellar occultation technique.
-This technique is the best one to obtain these parameters of the solar system objects from ground-based observations providing more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
+In order to obtain precise fundamental physical parameters like size, shape and albedo for the irregular satellites (and thus, in a sense, to small-sized TNOs), we will make use of stellar occultations, which provide  more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
 
-Since their estimated sizes are very small (see table \ref{} \textcolor{red}{Ainda vou fazer a tabela}), predict the exact location and instant where the shadow will cross the Earth demands a good precision.
-For instance, the bigger irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of 40 mas, must have an error smaller than its size for being observable.
-For the other objects, smaller or farther than Himalia, the situation is more difficult.
+Up to date no stellar occultation by irregular satellite is found in the literature.
+Since their estimated sizes are very small (see table \ref{} \textcolor{red}{Ainda vou fazer a tabela}), prediction of the exact location and instant where the shadow will cross the Earth demands a good accurate ephemeris and star positions.
+For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas( milliarcseconds).Thus, even in the best cases, the overall budget of ephemeris and star position errors must be around 40 mas in order that we get good chances of observing a stellar occultation.
+\textcolor{red}{aqui poderia dizer alguma coisa do outro lado da moeda, do lado bom  das condicoes mais favoraveis em relacao a TNOs.
+(ver comentario que fiz da Secao 2)}
 
@@
 For an object at the distance of Jupiter, this represents an error bigger than 700 km in the shadow path.
+\textcolor{red}{Falar mais de Gomes-Junior 2015, dando a base para que o presente trabalho de predicoes pudesse ser feito.
+Falar que foram feitas XXXX posicoes de NN satelites com precisoes medias de YYY, possibiitando corrigir efeitos sistematicos nas efemerides, e com isso tornando possivel fazer predicoes com chances razoaveis de exito.}
 
-We present in this paper the stellar occultation predictions for the 7 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke and Sinope), Phoebe from Saturn and Triton and Nereid from Neptune.
-In the section \ref{Sec: Rationale} we explore the scientific rationale for study the irregular satellites and the possibility of having a common origin with TNOs.
-In section \ref{Sec: Offset} we show the correction made to the ephemeris for better predict stellar occultations.
+We present in this paper stellar occultation predictions for the 7 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke and Sinope), Phoebe from Saturn and Nereid from Neptune.
+In the section \ref{Sec: Rationale} we explore the scientific rationale for the study of the irregular satellites and the possibility of having a common origin with TNOs.
+In section \ref{Sec: Offset} we show the correction made to the ephemeris of these satellites for better predicting the stellar occultations.
 In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites and how they were made.
-Some test realized to confirm the predictions are presented in section \ref{Sec: testes} and the conclusion is given in section \ref{Sec: conclusions}.
+Some tests made to confirm the predictions are presented in section \ref{Sec: testes} and the conclusions are given in section \ref{Sec: conclusions}.
 
@@
 
-As explicited in section \ref{Sec: introducao}, there is no consensus on a single model about the region where the irregular satellites were formed.
+\textcolor{red}{Como pode ver, comparado ao paragrafo na Introducao que falou a respeito, o texto aqui está muito pequeno e chovendo no molhado.}
+
+\textcolor{red}{Tem que falar mais.
+Citar o paper no livro da Barucci que voce me mostrou outro dia.
+Dar mais citacoes sobre essa hipotese.
+Falar dos albedos parecidos, cores.
+Tem que fundamentar a hipotese de irregulares = TNOs pequenos.}
+
+\textcolor{red}{Tambem falar aqui das vantagens de se observar ocultacoes dos irregulares, porque apesar de alguns satelites terem efemerides ruins, outros nem tanto, as efemerides sao melhores que a de TNOs (os irregulares ja completaram voltas em torno dos seus planetas); no caso de Jupiter, ele esta proximo o que ajuda a diminuir o erro em km.}
+
+\textcolor{red}{E dizer que, mesmo que a hipotese de TNOs que nos motiva a fazer as predicoes, venha a ser refutada no futuro, seja por outros trabalhos, seja pelos resultados futuros de observacoes de nossas proprias ocultacoes, o levantamento em si das propriedades fisicas, tamanho, forma e albedo ainda assim seriam muito uteis para testar as hiposteses de captura, ou seja, contribui para o estudo da formacao e evolucao dos sistemas de Jupiter, Saturno e Netuno.}
+%As explicited in section \ref{Sec: introducao}, there is no consensus on a single model about the region where the irregular satellites were formed.
 In this work we treat them as having a common origin to the Centaurs and TNOs.
@@
 
-\cite{GomesJunior2015} showed from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) that the orbits of the irregular satellites of the giant planets have systematic errors.
+\cite{GomesJunior2015} \textcolor{red}{published XXXX precise positions for NNN irregular satellites from 1992-2015} from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO).
+They showed that the orbits of the irregular satellites of the giant planets have systematic errors.
 The offsets of the observations relative to the JPL ephemeris could be up to 200 mas for some satellites.
@@
 
-Making a new model for the orbits of these objects would demand a lot of time and delay the publication of these predictions.
-Then we utilize the offsets obtained by \cite{GomesJunior2015} to identify a pattern in the error of the ephemeris.
-This pattern could be used to extrapolate an offset to the satellite ephemeris by the time of the occultation predicted.
+Making a new model for the orbits of these objects would demand a lot of time and delay the publication of predictions of possible events to occur in the very near future.
+Thus, we utilized the ephemeris offsets obtained by \cite{GomesJunior2015} to identify error patterns in the ephemeris.
+The error patterns in right ascension and in declination could be used to extrapolate the offsets to the satellite ephemeris by the time of the predicted occultation, improving it.
 Plots of the offsets over time and true anomaly (see Fig.
-\ref{Fig:offxtime} and \ref{Fig:offxtanom} for Carme) clearly show the systematic error in the JPL ephemeris.
+\ref{Fig:offxtime} and \ref{Fig:offxtanom} for Carme) clearly show the systematic errors in the JPL ephemeris.
 
@@
 
+The two kinds of error or offset patterns adopted depending on the case are given by Eqs. \ref{eq:linear} and \ref{eq:seno}:
+
 \begin{equation}
@@
 \end{split}
+\label{eq:seno}
 \end{equation}
 where $F(t,f)$ is the offset obtained, $t$ is time in years counting from J2000.0 and $f$ is the true anomaly.
+The patterns could be applied for right ascension as well as for declination ephemeris offsets.
 

### 2015-08-10 Irregular-occ/Irregular-occ.tex
.

@@
 
-\cite{GomesJunior2015} \textcolor{red}{published XXXX precise positions for NNN irregular satellites from 1992-2015} from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO).
+\cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.
 They showed that the orbits of the irregular satellites of the giant planets have systematic errors.
@@
 
-The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the corrected JPL ephemeris as presented in the section \ref{Sec: Offset}.
+The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the corrected JPL ephemeris as presented in section \ref{Sec: Offset}.
 The search for stellar candidates follows the same procedure as presented by \cite{Assafin2010, Assafin2012} and \cite{Camargo2013}.
 
-It was predicted occultation by the 7 major irregular satellites of Jupiter, Phoebe of Saturn and Triton and Nereid of Neptune.
+We predicted occultation for the 7 major irregular satellites of Jupiter, Phoebe of Saturn and Triton and Nereid of Neptune.
 
@@
 A total of 588 events \textcolor{red}{(Não estou contando ainda os eventos para Tritão)} were identified between January 2015 and December 2017.
-Table \ref{Tab: occ-list} exemplifies the catalogue of occultations generated and their parameters, which is necessary to produce occultation maps (see Fig.
+Table \ref{Tab: occ-list} exemplifies the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps (see Fig.
 \ref{Fig: ocultacao} as an example).
@@
 \begin{table*}
-\caption{\label{Tab: sample-cds} Extract of prediction table. \textcolor{red}{Ainda tenho que fazer a tabela melhor, encaixar. Abaixo é um exemplo apenas para visualização.}}
+\caption{\label{Tab: sample-cds} Stellar occultation sampled predictions for Himalia \textcolor{red}{Consertar tabela}}
 \begin{centering}
@@
 \textbf{Notes}.
-Occultation table: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - in the original table, these coordinates are immediately followed by the geocentric astrometric equatorial coordinates (corrected for the offset or ephemeris method), not presented here, of the occulting body; C/A: the geocentric closest approach, in arcseconds; P/A: the planet position angle with respect to the occulted star at C/A, in degrees; velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; D: planet range to Earth, in AU; R*: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{Mag}* = \textrm{Mag}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$.
-A value of 50.0 means that the star is not in the 2MASS; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\Delta e_{\alpha *}$ and $\Delta e_{\delta}$:offset in mas applied to the ephemeris right ascension and declination, respectively; pm: ok = proper motion applied, no = no proper motion applied; catalogue cross-identification (ct) = uc (UCAC2), 2m (2MASS), fs (field star); f = multiplicity flag (see Table 3); $E_{\alpha *}$ and $E_{\delta}$: uncertainties (mas) in right ascension and declination.
+Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - in the original table, these coordinates are immediately followed by the geocentric astrometric equatorial coordinates of the satellite (corrected the computed ephemeris offset using Eqs \ref{eq:linear} or \ref{eq:seno}); C/A: the geocentric closest approach, in arcseconds; P/A: the planet position angle with respect to the occulted star at C/A, in degrees; velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; D: planet range to Earth, in AU; R*: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{Mag}* = \textrm{Mag}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$.
+A value of 50.0 means that the star is not in the 2MASS; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\Delta e_{\alpha *}$ and $\Delta e_{\delta}$: computed offsets in mas applied to the ephemeris right ascension and declination, respectively; pm: ok = proper motion applied, no = no proper motion applied; catalogue cross-identification (ct) = uc (UCAC2), 2m (2MASS), fs (field star); $E_{\alpha *}$ and $E_{\delta}$: uncertainties (mas) in right ascension and declination.
 A value of 9999 means that there was no estimation of the respective uncertainty; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
@@
 \includegraphics[scale=0.2]{figures/Himalia_2015-10-23T07:54:47.png} 
-\caption{Occultation map for Himalia \textcolor{red}{Mapa de Himalia para exemplificar, posso colocar outro depois}.
-Compare the information given in caption with the second entry of Table \ref{Tab: occ-list}}
+\caption{Occultation map for Himalia regarding to the second event sampled in Table \ref{Tab: occ-list}}
 \label{Fig: ocultacao}
@@
 \end{figure}
+
+\textcolor{red}{Acrescentar um paragrafo falando do GAIA, cujo primeiro release esta previsto para fins de 2016.
+Falar que seremos capazes de refazer o trabalho de Gomes-Junior 2015 e com posicoes de estrelas GAIA poderemos predizer melhor os eventos de 2017 em diante.
+Mas por questao de precaucao e de preparo futuro para campanhas, publicamos ate fim de 2017 com o UCAC4.}
 
@@
 The shadow covers a very restricted area on Earth because the irregular satellites are small.
-So before we start a large observational campaign, we tested some occultation predictions for larger objects, to assess the quality of the prediction.
+So, before we start a large observational campaign, we tested some occultation predictions for larger objects, to assess the quality of the prediction.
 
-This test consists in observing the object and star to be occulted near the event when the two objects are present in the same field of view, preferably when the objects are close to each other.
-Thus the relative positions between the two objects will have minimal influence of the errors of the catalogue of stars used and possible field distortions \citep[and references therein]{Peng2008}.
-The offsets of the positions of the star and the satellites will be used to correct the predictions.
+The tests consisted in observing the object and star to be occulted near the event when the two objects were present in the same field of view (FOV), preferably when the objects were close to each other.
+Thus, the relative positions between the two objects had minimal influence of the errors of the reference catalogue of stars used and possible field distortions \citep[and references therein]{Peng2008}.
+The offsets of the positions of the star and of the satellites were used to check the original predictions.
 
-To date, two occultation tests were performed, one by Himalia that occurred on March 3, 2015 and the second by Elara that occurred on March 30, 2015.
-For each event, four maps were generated: the first with the nominal positions of the star and the satellite to the predicted time; the second with the offset calculated as described in section \ref{Sec: Offset}; the third with star and satellite offsets from observations made a few days before the occultation when the two objects were separated (different FOV); and the fourth from observations made when the star and the satellite were close in the same FOV.
+Two occultation tests were performed, one by Himalia that occurred on March 3, 2015 and the second by Elara that occurred on March 30, 2015.
+For each event, four situations were considered: the first with the nominal positions of the star and the satellite to the predicted time; the second with the offset calculated as described in section \ref{Sec: Offset}; the third with star and satellite offsets from observations made a few days before the occultation when the two objects were separated (different FOV); and the fourth from observations made when the star and the satellite were close in the same FOV.
 
-Figure \ref{Fig: occ-Himalia}.
-Shows the four maps for Himalia occultation test in March 03, 2015.
+Figure \ref{Fig: occ-Himalia}shows four zoomed in occultation maps corresponding to each of the four situations for Himalia..
 The map \ref{Fig: occ-Himalia-nooff} is the nominal prediction with the coordinate of the star given by the catalogue and that of the satellite from the ephemeris.
 We then corrected the positions of the satellite by an offset calculated by the method showed in section \ref{Sec: Offset} to make the map \ref{Fig: occ-Himalia-ajuste}.
-
-The map \ref{Fig: occ-Himalia-off22fev} was made from obtained positions on February 22 observed with the Zeiss telescope at the Observatório do Pico dos Dias (OPD).
+The map \ref{Fig: occ-Himalia-off22fev} was made from obtained positions on February 22 observed with the Zeiss telescope \textcolor{red}{dar informacoes do Zeiss, diametro, FOV em arcmin, pixel scale} at the Observatório do Pico dos Dias (OPD).
 On that day, Himalia and the star were observed in separate FOVs as they were still far apart.
-On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope at OPD just over an hour after the time scheduled for the event.
+On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope \textcolor{red}{dar as mesmas infos, agora para o 1.6m} at OPD just over an hour after the time scheduled for the event.
 Satellite and star were separated by about 16 arcsec, so very close to each other.
 From the calculated offsets, the map \ref{Fig: occ-Himalia-off03mar} was generated.
+Notice that the shadow path was not predicted to cross the OPD.
+This was not necessary for testing the prediction.
 
@@
 In \ref{Fig: occ-Himalia-off22fev} we apply offsets to the positions of star and satellite accordingly the observations made at February 22.
-In \ref{Fig: occ-Himalia-off03mar} is as in \ref{Fig: occ-Himalia-off22fev} but with observations made at March 03 when the objects were close to each other.
+\ref{Fig: occ-Himalia-off03mar} is as in \ref{Fig: occ-Himalia-off22fev} but with observations made at March 03 when the objects were close to each other.
 \textcolor{red}{Figuras ainda serão mudadas.
@@
 
-In this event, it is possible to see that the shade does not vary much among the four maps suggesting that, at least for Himalia, there is a greater probability of observing an event.
-In fact, the biggest difference between the shadows of the four maps are 25s and 130km in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
+In this event, it is possible to see that the shade does not vary by much among the four maps suggesting thatthere was a good probability of observing the event.
+In fact, the biggest differences between the shadows of the four maps were 25s and 130km in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
 
@@
 
-The second test was with the satellite Elara, which is the second biggest irregular satellites of Jupiter.
-The event was predict the occur at March 30, 2015.
-The observations were taken on March 25 and April 2, 2015 with the Zeiss telescope.
+The second test was with the satellite Elara, which is the second largest irregular satellite of Jupiter.
+The event was predicted to occur at March 30, 2015.
+The observations were made on March 25 and April 2, 2015 with the Zeiss telescope.
 On the night of April 2 they could still be observed in the same FOV.
-Due to Elara be much weaker, dispersions of the satellite positions on both nights ended up being higher than for Himalia.
+Due to Elara being fainter than Himalia by 2 magnitudes, dispersions of the satellite positions on both nights ended up being higher than for Himalia.
 Still, the differences between the maps obtained were relatively small.
-The biggest difference between them is 73s and 302km (see table \ref{Tab: comparison-Elara}).
+The biggest differences between them were 73s along the shadow path and 302km perpendicular to it (see table \ref{Tab: comparison-Elara}).
 

### 2015-08-11 Irregular-occ/Irregular-occ.tex
.

@@
 \makeatother
-\titlerunning{Predictions of stellar occultations by irregular satellites}
+\titlerunning{Predictions of stellar occultations by irregular satellites up to 2017}
 \authorrunning{Gomes-Júnior et al.}
@@
 
-\title{Stellar occultation predictions for  9 irregular satellites of giant planets for 2015-2016
+\title{Stellar occultation predictions for  9 irregular satellites of giant planets for 2015-2017
 %\fnmsep\thanks{Table 8 are only available in electronic form at the CDS via anonymous ftp to cdsarc.u-strasbg.fr (130.79.128.5) or via http://cdsweb.u-strasbg.fr/cgi-bin/qcat?J/A+A/ and IAU NSDC data base at www.imcce.fr/nsdc.}

### 2015-08-14 Irregular-occ/Irregular-occ.tex
.

@@
 Since their estimated sizes are very small (see table \ref{} \textcolor{red}{Ainda vou fazer a tabela}), prediction of the exact location and instant where the shadow will cross the Earth demands a good accurate ephemeris and star positions.
-For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas( milliarcseconds).Thus, even in the best cases, the overall budget of ephemeris and star position errors must be around 40 mas in order that we get good chances of observing a stellar occultation.
-\textcolor{red}{aqui poderia dizer alguma coisa do outro lado da moeda, do lado bom  das condicoes mais favoraveis em relacao a TNOs.
-(ver comentario que fiz da Secao 2)}
+For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds).Thus, even in the best cases, the overall budget of ephemeris and star position errors must be around 40 mas in order that we get good chances of observing a stellar occultation.
 
-As pointed out by \cite{GomesJunior2015}, the ephemeris of the irregular satellites have errors that may reach 200 mas for some satellites.
+Unlikely stellar occultations by TNOs, which has been proved effective, the observation of an occultation by a irregular satellite is more favorable.
+They have been known for more time and observed in many orbital period having better ephemeris than TNOs.
+Moreover, the irregular satellites are closer to Earth which means a minor error in kilometers.
+
+\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas.
+For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}.
+They pointed out that the ephemeris of the irregular satellites have systematic errors that may reach 200 mas for some satellites.
 For an object at the distance of Jupiter, this represents an error bigger than 700 km in the shadow path.
-\textcolor{red}{Falar mais de Gomes-Junior 2015, dando a base para que o presente trabalho de predicoes pudesse ser feito.
-Falar que foram feitas XXXX posicoes de NN satelites com precisoes medias de YYY, possibiitando corrigir efeitos sistematicos nas efemerides, e com isso tornando possivel fazer predicoes com chances razoaveis de exito.}
+Using the offsets obtained by \cite{GomesJunior2015} it is possible to correct the ephemeris and better predict stellar occultations for these objects.
 
@@
 
-\textcolor{red}{Acrescentar um paragrafo falando do GAIA, cujo primeiro release esta previsto para fins de 2016.
-Falar que seremos capazes de refazer o trabalho de Gomes-Junior 2015 e com posicoes de estrelas GAIA poderemos predizer melhor os eventos de 2017 em diante.
-Mas por questao de precaucao e de preparo futuro para campanhas, publicamos ate fim de 2017 com o UCAC4.}
+The ESA astrometry satellite GAIA \citep{deBruijne2012} catalogue is expected to be released in the end of 2016.
+The precise star positions to be derived by GAIA will render better predictions with the only source of error being the ephemeris.
+The work of \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions from 2017 onwards will be better predicted.
+For a matter of precaution and preparations of campaigns, we publish stellar occultations with UCAC4 until 2017.
 
@@
 We then corrected the positions of the satellite by an offset calculated by the method showed in section \ref{Sec: Offset} to make the map \ref{Fig: occ-Himalia-ajuste}.
-The map \ref{Fig: occ-Himalia-off22fev} was made from obtained positions on February 22 observed with the Zeiss telescope \textcolor{red}{dar informacoes do Zeiss, diametro, FOV em arcmin, pixel scale} at the Observatório do Pico dos Dias (OPD).
+The map \ref{Fig: occ-Himalia-off22fev} was made from obtained positions on February 22 observed with the Zeiss telescope (Diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / px$) at the Observatório do Pico dos Dias (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
 On that day, Himalia and the star were observed in separate FOVs as they were still far apart.
-On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope \textcolor{red}{dar as mesmas infos, agora para o 1.6m} at OPD just over an hour after the time scheduled for the event.
+On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope (Diameter = 1.6m; FOV = $5\farcm8$; pixel scale = $0\farcs17 / px$) at OPD just over an hour after the time scheduled for the event.
 Satellite and star were separated by about 16 arcsec, so very close to each other.

### 2015-08-19 Irregular-occ/Irregular-occ.tex
.
Adicionado figura mapa consertado

@@
 \textcolor{red}{E dizer que, mesmo que a hipotese de TNOs que nos motiva a fazer as predicoes, venha a ser refutada no futuro, seja por outros trabalhos, seja pelos resultados futuros de observacoes de nossas proprias ocultacoes, o levantamento em si das propriedades fisicas, tamanho, forma e albedo ainda assim seriam muito uteis para testar as hiposteses de captura, ou seja, contribui para o estudo da formacao e evolucao dos sistemas de Jupiter, Saturno e Netuno.}
-%As explicited in section \ref{Sec: introducao}, there is no consensus on a single model about the region where the irregular satellites were formed.
-In this work we treat them as having a common origin to the Centaurs and TNOs.
-In this case, they would be the representative objects of the small TNOs population.
-Since that small objects at great distances are difficult to study, the irregular satellites can complement this gap.
 
+As explicited in section \ref{Sec: introducao}, there is no consensus on a single model about the region where the irregular satellites were formed. 
 %This idea is supported by the
@@
 \begin{centering}
-\subfigure[Nominal prediction]{\includegraphics[scale=0.19]{figures/Himalia1_2015-03-03T00:39:25.png}  \label{Fig: occ-Himalia-nooff}}
-\subfigure[\textcolor{red}{Offset de Himalia a partir do ajuste}]{\includegraphics[scale=0.19]{figures/Himalia2_2015-03-03T00:39:18.png}  \label{Fig: occ-Himalia-ajuste}}
-\subfigure[Observation at February 22 (different FOVs)]{\includegraphics[scale=0.19]{figures/Himalia3_2015-03-03T00:39:40.png}   \label{Fig: occ-Himalia-off22fev}}
-\subfigure[Observation at March 03 (same FOV)]{\includegraphics[scale=0.19]{figures/Himalia4_2015-03-03T00:39:15.png}   \label{Fig: occ-Himalia-off03mar}}
+\subfigure[Nominal prediction]{\includegraphics[width=8.8cm]{figures/Himalia0_2015-03-03T00:39:25.eps}  \label{Fig: occ-Himalia-nooff}}
+\subfigure[\textcolor{red}{Offset de Himalia a partir do ajuste}]{\includegraphics[width=8.8cm]{figures/Himalia2_2015-03-03T00:39:18.png}  \label{Fig: occ-Himalia-ajuste}}
+\subfigure[Observation at February 22 (different FOVs)]{\includegraphics[width=8.8cm]{figures/Himalia3_2015-03-03T00:39:40.png}   \label{Fig: occ-Himalia-off22fev}}
+\subfigure[Observation at March 03 (same FOV)]{\includegraphics[width=8.8cm]{figures/Himalia4_2015-03-03T00:39:15.png}   \label{Fig: occ-Himalia-off03mar}}
 \caption{Predictions for Himalia: The big red dot show the geocentric closest approach of the shadow.
@@
 \ref{Fig: occ-Himalia-off03mar} is as in \ref{Fig: occ-Himalia-off22fev} but with observations made at March 03 when the objects were close to each other.
-\textcolor{red}{Figuras ainda serão mudadas.
-Colocadas para visualização}}
+The green dot at the bottom of the maps is the location of Observatório do Pico dos Dias.
+The black "+" marks at maps (b,c,d) are the nominal prediction closest approach for reference.
+\textcolor{red}{Figura b será mudada após confirmamos o offset para Himalia}}
 \label{Fig: occ-Himalia}

### 2015-08-26 Irregular-occ/Irregular-occ.tex
.

@@
 
-\title{Stellar occultation predictions for  9 irregular satellites of giant planets for 2015-2017
+\title{Stellar occultation predictions for  9 irregular satellites of giant planets and Triton for 2015-2017
 %\fnmsep\thanks{Table 8 are only available in electronic form at the CDS via anonymous ftp to cdsarc.u-strasbg.fr (130.79.128.5) or via http://cdsweb.u-strasbg.fr/cgi-bin/qcat?J/A+A/ and IAU NSDC data base at www.imcce.fr/nsdc.}
@@
 \begin{centering}
-\subfigure[Nominal prediction]{\includegraphics[width=8.8cm]{figures/Himalia0_2015-03-03T00:39:25.eps}  \label{Fig: occ-Himalia-nooff}}
+\subfigure[Nominal prediction]{\includegraphics[width=8.8cm]{figures/Himalia_nominal.eps}  \label{Fig: occ-Himalia-nooff}}
 \subfigure[\textcolor{red}{Offset de Himalia a partir do ajuste}]{\includegraphics[width=8.8cm]{figures/Himalia2_2015-03-03T00:39:18.png}  \label{Fig: occ-Himalia-ajuste}}
-\subfigure[Observation at February 22 (different FOVs)]{\includegraphics[width=8.8cm]{figures/Himalia3_2015-03-03T00:39:40.png}   \label{Fig: occ-Himalia-off22fev}}
-\subfigure[Observation at March 03 (same FOV)]{\includegraphics[width=8.8cm]{figures/Himalia4_2015-03-03T00:39:15.png}   \label{Fig: occ-Himalia-off03mar}}
+\subfigure[Observation at February 22 (different FOVs)]{\includegraphics[width=8.8cm]{figures/Himalia_22fev.eps}   \label{Fig: occ-Himalia-off22fev}}
+\subfigure[Observation at March 03 (same FOV)]{\includegraphics[width=8.8cm]{figures/Himalia_03mar.eps}   \label{Fig: occ-Himalia-off03mar}}
 \caption{Predictions for Himalia: The big red dot show the geocentric closest approach of the shadow.
+The black "+" marks at maps (b,c,d) are the nominal prediction closest approach for reference.
 The small red ones are the center of the shadow separated by one minute.
@@
 The green dot at the bottom of the maps is the location of Observatório do Pico dos Dias.
-The black "+" marks at maps (b,c,d) are the nominal prediction closest approach for reference.
-\textcolor{red}{Figura b será mudada após confirmamos o offset para Himalia}}
+\textcolor{red}{Figura b será mudada após confirmarmos o offset para Himalia}}
 \label{Fig: occ-Himalia}
@@
 In this event, it is possible to see that the shade does not vary by much among the four maps suggesting thatthere was a good probability of observing the event.
-In fact, the biggest differences between the shadows of the four maps were 25s and 130km in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
+In fact, the biggest differences between the shadows of the four maps were \textcolor{red}{22s and 109km} in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
 
 \begin{table}
-\caption{\label{Tab: comparison-Himalia} Comparison between the predictions of Himalia. \textcolor{red}{Ainda vou terminar de preencher a tabela e tem que ver direito o offset do ajuste}}
+\caption{\label{Tab: comparison-Himalia} Comparison between the predictions of the Himalia occultation at March 03, 2015.
+\textcolor{red}{Ainda vou terminar de preencher a tabela e tem que ver direito o offset do ajuste}}
 \begin{centering}
@@
 \hline
-Nominal & 2015-03-03 00:39:25 UT & $0.714\arcsec$ \tabularnewline
-Offset & -07 s & \tabularnewline
-Feb. 22 Obs. & +15 s &  \tabularnewline
-Mar. 03 Obs. & -10 s & \tabularnewline
+Nominal & 00:39:25 UTC & $0.714\arcsec$ \tabularnewline
+Offset &   & \tabularnewline
+Feb. 22 Obs. & +12s & -33mas (109km) \tabularnewline
+Mar. 03 Obs. & -10s & -19mas (61km)\tabularnewline
 \hline
@@
 \par\end{centering}
+
 \end{table}
@@
 \hline
-Nominal & 2015-03-30 01:45:15 UT & $1.139\arcsec$ \tabularnewline
+Nominal & 2015-03-30 01:45:15 UTC & $1.139\arcsec$ \tabularnewline
 Offset & -16 s & \tabularnewline

### 2015-08-31 Irregular-occ/Irregular-occ.tex
.

@@
 As explicited in section \ref{Sec: introducao}, there is no consensus on a single model about the region where the irregular satellites were formed.
-%This idea is supported by the
+
+Phoebe, the most studied irregular satellite, was identified by \cite{Clark2005} that it has surface material probably of cometary origin.
+It was also stated by \cite{Johnson2005} that if the porosity of Phoebe would be of about 15\%, Phoebe would have an uncompressed density similar to those of Pluto and Triton.
+
+Triton is a special satellite.
+Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
+Because Triton's orbit is very small and the its precession is not dominated by Solar perturbations, Triton is frequently excluded of the irregular satellites' class, but studied together by many authors \citep{Sheppard2005, Jewitt2007}.
+
+Similar to the irregular satellites, Triton was probably capture in the early Solar System and may have a similar origin as the TNOs \citep{Agnor2006}.
+Differently, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
 

### 2015-09-02 Irregular-occ/Irregular-occ.tex
.

--- 2015-08-26 Irregular-occ/Irregular-occ.tex
@@
 
-\textcolor{red}{Como pode ver, comparado ao paragrafo na Introducao que falou a respeito, o texto aqui está muito pequeno e chovendo no molhado.}
+%\textcolor{red}{Como pode ver, comparado ao paragrafo na Introducao que falou a respeito, o texto aqui está muito pequeno e chovendo no molhado.}
 
-\textcolor{red}{Tem que falar mais.
+%\textcolor{red}{Tem que falar mais.
 Citar o paper no livro da Barucci que voce me mostrou outro dia.
@@
 
-\textcolor{red}{Tambem falar aqui das vantagens de se observar ocultacoes dos irregulares, porque apesar de alguns satelites terem efemerides ruins, outros nem tanto, as efemerides sao melhores que a de TNOs (os irregulares ja completaram voltas em torno dos seus planetas); no caso de Jupiter, ele esta proximo o que ajuda a diminuir o erro em km.}
+%\textcolor{red}{Tambem falar aqui das vantagens de se observar ocultacoes dos irregulares, porque apesar de alguns satelites terem efemerides ruins, outros nem tanto, as efemerides sao melhores que a de TNOs (os irregulares ja completaram voltas em torno dos seus planetas); no caso de Jupiter, ele esta proximo o que ajuda a diminuir o erro em km.}
 
-\textcolor{red}{E dizer que, mesmo que a hipotese de TNOs que nos motiva a fazer as predicoes, venha a ser refutada no futuro, seja por outros trabalhos, seja pelos resultados futuros de observacoes de nossas proprias ocultacoes, o levantamento em si das propriedades fisicas, tamanho, forma e albedo ainda assim seriam muito uteis para testar as hiposteses de captura, ou seja, contribui para o estudo da formacao e evolucao dos sistemas de Jupiter, Saturno e Netuno.}
+%\textcolor{red}{E dizer que, mesmo que a hipotese de TNOs que nos motiva a fazer as predicoes, venha a ser refutada no futuro, seja por outros trabalhos, seja pelos resultados futuros de observacoes de nossas proprias ocultacoes, o levantamento em si das propriedades fisicas, tamanho, forma e albedo ainda assim seriam muito uteis para testar as hiposteses de captura, ou seja, contribui para o estudo da formacao e evolucao dos sistemas de Jupiter, Saturno e Netuno.}
 
 As explicited in section \ref{Sec: introducao}, there is no consensus on a single model about the region where the irregular satellites were formed.
-%This idea is supported by the
+
+Phoebe, the most studied irregular satellite, was identified by \cite{Clark2005} that its surface is probably covered by material of cometary origin.
+It was also stated by \cite{Johnson2005} that if the porosity of Phoebe would be of about 15\%, Phoebe would have an uncompressed density similar to those of Pluto and Triton.
+
+Triton is a special satellite.
+Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
+Because Triton's orbit is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded of the irregular satellites' class, but studied together by many authors \citep{Sheppard2005, Jewitt2007}.
+
+Similarly to the irregular satellites, Triton was probably capture in the early Solar System and may have a similar origin as the TNOs \citep{Agnor2006}.
+Differently, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
+The main objective to study Triton by stellar occultation is to understand the evolution of its atmosphere due to Triton's complicated and extreme seasonal cycle \citep{McKinnon2007}.
 
@@
 \hline
-Nominal & 00:39:25 UTC & $0.714\arcsec$ \tabularnewline
+Nominal & 00:39:25 UTC & $0\farcs714$ \tabularnewline
 Offset &   & \tabularnewline
@@
 Still, the differences between the maps obtained were relatively small.
-The biggest differences between them were 73s along the shadow path and 302km perpendicular to it (see table \ref{Tab: comparison-Elara}).
+The biggest differences between them were 74s along the shadow path and 307km perpendicular to it (see table \ref{Tab: comparison-Elara}).
 
 \begin{table}
-\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara. \textcolor{red}{Ainda vou terminar de preencher a tabela e tem que ver direito o offset do ajuste}}
+\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara occultation at March 30, 2015.
+\textcolor{red}{Ainda vou terminar de preencher a tabela e tem que ver direito o offset do ajuste}}
 \begin{centering}
@@
 \hline
-Nominal & 2015-03-30 01:45:15 UTC & $1.139\arcsec$ \tabularnewline
-Offset & -16 s & \tabularnewline
-Mar. 25 Obs. & -58 s &  \tabularnewline
-Apr. 02 Obs. & +15 s & \tabularnewline
+Nominal & 01:45:15 UTC & $1\farcs139$ \tabularnewline
+Offset &  & \tabularnewline
+Mar. 25 Obs. & -59 s & -17mas (61km) \tabularnewline
+Apr. 02 Obs. & +15 s & -89mas (307km) \tabularnewline
 \hline

### 2015-09-09 Irregular-occ/Irregular-occ.tex
.

@@
 Up to date no stellar occultation by irregular satellite is found in the literature.
-Since their estimated sizes are very small (see table \ref{} \textcolor{red}{Ainda vou fazer a tabela}), prediction of the exact location and instant where the shadow will cross the Earth demands a good accurate ephemeris and star positions.
+Since their estimated sizes are very small (see table \ref{Tab: satellite-diameter}), prediction of the exact location and instant where the shadow will cross the Earth demands a good accurate ephemeris and star positions.
 For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds).Thus, even in the best cases, the overall budget of ephemeris and star position errors must be around 40 mas in order that we get good chances of observing a stellar occultation.
+
+\begin{table}
+\caption{\label{Tab: satellite-diameter} Estimated diameter of the satellites and correspondent apparent diameter}
+\begin{centering}
+\begin{tabular}{lccc}
+\hline  \hline
+\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
+Satellite  & km  & mas\tablefootmark{a} & References \tabularnewline
+\hline
+Ananke & 29 &  & 1 \tabularnewline
+Carme & 46 &  & 1 \tabularnewline
+Elara & 86 &  & 1 \tabularnewline
+Himalia & $(150 \pm 20) \times (120\pm 20)$ &  & 2 \tabularnewline
+Lysithea & 36 &  & 1 \tabularnewline
+Pasiphae & 62 & & 1 \tabularnewline
+Sinope & 37 & & 1 \tabularnewline
+\hdashline
+Phoebe & $212 \pm 1.4$ & & 3 \tabularnewline
+\hdashline
+Nereid & $340 \pm 50$ &  & 4 \tabularnewline
+Triton & $2707 \pm 2.0$ &  & 5 \tabularnewline
+\hline
+\end{tabular}
+\tablebib{
+(1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}; (4) \cite{Thomas1991} (5) \cite{Thomas2000}.}
+\end{centering}
+\tablefoottext{a}{Using a mean distance from Jupiter of XX AU, from Saturn of XX AU and from Neptune of XX AU.}
+\par
+\end{table}
 
@@
 The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the corrected JPL ephemeris as presented in section \ref{Sec: Offset}.
-The search for stellar candidates follows the same procedure as presented by \cite{Assafin2010, Assafin2012} and \cite{Camargo2013}.
+The search for stellar candidates follows the same procedure as presented by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.
 
-We predicted occultation for the 7 major irregular satellites of Jupiter, Phoebe of Saturn and Triton and Nereid of Neptune.
+We predicted occultation for the 7 major irregular satellites of Jupiter:  Ananke, Carme, Elara, Himalia, Lysithea, Pasiphae, and Sinope; Phoebe of Saturn and Triton and Nereid of Neptune.
 
-For Triton and Nereid, the candidates for stellar occultations in 2015 and 2016 was searched using the WFI catalogue in the same way as the predictions for Centaurs and TNOs occultations by \cite{Assafin2010, Assafin2012} and \cite{Camargo2013}.
+For Triton and Nereid, the candidates for stellar occultations in 2015 and 2016 was searched using the WFI catalogue in the same way as the predictions for Centaurs and TNOs occultations by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.
 This catalogue contains the stars in the path of Neptune in the sky up to mid-2016.
@@
 
-\section{Conclusion} \label{Sec: conclusions}
+\section{Discussion} \label{Sec: conclusions}
+
+We presented stellar occultations for the period of 2015-2017 for seven irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid.
+The procedure used was the same that predicted stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.
+
+The candidate stars was searched in the UCAC4 catalogue, except for the candidates in 2015 and 2016 for Triton and Nereid.
+In this case, we used the WFI catalogue that was generated from observations made with ESO2p2/WFI CCD mosaic that covered the path of Neptune in the sky-plane up to 2016.
+
+In table \ref{Tab: satellite-occultation} it's presented the number of stellar occultations predicted by year for each satellite.
+
+\begin{table}
+\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from 2015 up to 2017}
+\begin{centering}
+\begin{tabular}{lcccc}
+\hline  \hline
+%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
+Satellite  & 2015  & 2016 & 2017 & Total \tabularnewline
+\hline
+Ananke &  &  & & \tabularnewline
+Carme &  &  &  & \tabularnewline
+Elara &  &  &  & \tabularnewline
+Himalia &  &  &  & \tabularnewline
+Lysithea &  &  &  & \tabularnewline
+Pasiphae &  & &  & \tabularnewline
+Sinope &  &  &  & \tabularnewline
+\hdashline
+Phoebe & 13 & 22 & 98 & 133\tabularnewline
+\hdashline
+Nereid & 40 & 11 & 1 & 52 \tabularnewline
+Triton &  &  &  & \tabularnewline
+\hline
+\end{tabular}
+\par \end{centering}
+\end{table}
 

### 2015-09-10 Irregular-occ/Irregular-occ.tex
.

@@
 \hline
-Ananke & 29 &  & 1 \tabularnewline
-Carme & 46 &  & 1 \tabularnewline
-Elara & 86 &  & 1 \tabularnewline
-Himalia & $(150 \pm 20) \times (120\pm 20)$ &  & 2 \tabularnewline
-Lysithea & 36 &  & 1 \tabularnewline
-Pasiphae & 62 & & 1 \tabularnewline
-Sinope & 37 & & 1 \tabularnewline
+Ananke & 29 & 8 & 1 \tabularnewline
+Carme & 46 & 13 & 1 \tabularnewline
+Elara & 86 & 24 & 1 \tabularnewline
+Himalia & $(150\times120) \pm 20$\tablefootmark{b} & 41 & 2 \tabularnewline
+Lysithea & 36 & 10 & 1 \tabularnewline
+Pasiphae & 62 & 17 & 1 \tabularnewline
+Sinope & 37 & 10 & 1 \tabularnewline
 \hdashline
-Phoebe & $212 \pm 1.4$ & & 3 \tabularnewline
+Phoebe & $212 \pm 1.4$\tablefootmark{b} & 32 & 3 \tabularnewline
 \hdashline
-Nereid & $340 \pm 50$ &  & 4 \tabularnewline
-Triton & $2707 \pm 2.0$ &  & 5 \tabularnewline
+Nereid & $340 \pm 50$\tablefootmark{c} & 15 & 4 \tabularnewline
+Triton & $2707 \pm 2.0$\tablefootmark{c} & 124 & 5 \tabularnewline
 \hline
@@
 \tablebib{
-(1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}; (4) \cite{Thomas1991} (5) \cite{Thomas2000}.}
+(1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}; (4) \cite{Thomas1991}; (5) \cite{Thomas2000}.}
 \end{centering}
-\tablefoottext{a}{Using a mean distance from Jupiter of XX AU, from Saturn of XX AU and from Neptune of XX AU.}
+\tablefoottext{a}{Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
+\tablefoottext{b}{From Cassini observation.}
+\tablefoottext{c}{From Voyager-2 observation.}
 \par
@@
 \begin{table}
-\caption{\label{Tab: comparison-Himalia} Comparison between the predictions of the Himalia occultation at March 03, 2015.
-\textcolor{red}{Ainda vou terminar de preencher a tabela e tem que ver direito o offset do ajuste}}
+\caption{\label{Tab: comparison-Himalia} Comparison between the predictions of the Himalia occultation at March 03, 2015. \textcolor{red}{Tem que ver direito o offset do ajuste}}
 \begin{centering}
@@
 \begin{table}
-\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara occultation at March 30, 2015.
-\textcolor{red}{Ainda vou terminar de preencher a tabela e tem que ver direito o offset do ajuste}}
+\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara occultation at March 30, 2015. \textcolor{red}{Tem que ver direito o offset do ajuste}}
 \begin{centering}
@@
 The candidate stars was searched in the UCAC4 catalogue, except for the candidates in 2015 and 2016 for Triton and Nereid.
-In this case, we used the WFI catalogue that was generated from observations made with ESO2p2/WFI CCD mosaic that covered the path of Neptune in the sky-plane up to 2016.
+In this case, we used the WFI catalogue that was generated from observations made with ESO2p2/WFI CCD mosaic that covered the path of Neptune in the sky-plane up to 2016 (see Sec.
+\ref{Sec: predictions}).
 
+A total of \textcolor{red}{588} events were detected to happen between and the period presented.
 In table \ref{Tab: satellite-occultation} it's presented the number of stellar occultations predicted by year for each satellite.
@@
 \begin{table}
-\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from 2015 up to 2017}
+\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from 2015 up to 2017.}
 \begin{centering}
@@
 \hline
-Ananke &  &  & & \tabularnewline
-Carme &  &  &  & \tabularnewline
-Elara &  &  &  & \tabularnewline
-Himalia &  &  &  & \tabularnewline
-Lysithea &  &  &  & \tabularnewline
-Pasiphae &  & &  & \tabularnewline
-Sinope &  &  &  & \tabularnewline
+Ananke & 22 & 12 & 16 & 50 \tabularnewline
+Carme & 25 & 20 & 14 & 59 \tabularnewline
+Elara & 32 & 14 & 16 & 62 \tabularnewline
+Himalia & 25 & 15 & 12 & 52 \tabularnewline
+Lysithea & 22 & 16 & 11 & 49 \tabularnewline
+Pasiphae & 26 & 19 & 20 & 65 \tabularnewline
+Sinope & 19 & 15 & 21 & 55 \tabularnewline
 \hdashline
-Phoebe & 13 & 22 & 98 & 133\tabularnewline
+Phoebe & 13 & 32 & 98 & 143 \tabularnewline
 \hdashline
-Nereid & 40 & 11 & 1 & 52 \tabularnewline
+Nereid & 41 & 11 & 1 & 53 \tabularnewline
 Triton &  &  &  &  \tabularnewline
@@
 \par \end{centering}
+Occultations predicted using the UCAC4 catalogue. For Nereid and Triton, the occultations in 2015 and 2016 were predicted using the WFI catalogue as explained in Sec. \ref{Sec: predictions}.
 \end{table}
+
+Concerning the UCAC4 stars, an accuracy of about 20-50 mas in the star position would give a 15-40\% probability of observe such an event by Ananke, the smallest irregular satellite in the sample.
+For Himalia, the occultation would have 80-100\% probability.
+Combined with a mean accuracy in the position of 30 mas for Himalia and 100 mas for Ananke mas published in Table 2 of \cite{Jacobson2012}, the probability of success would be around 50-80\% for Himalia and $\approx$5\% for Ananke.
+Observations a few days before the date of occultation predicted may diminish the combined accuracy to 40-80 mas, depending on the magnitude of the objects.
+
+The tests made with an occultation expected to happen in March 03, 2015 for Himalia showed that this event could probably be observed in case there were observers available in the shadow area.
+Fig.
+\ref{Fig: occ-Himalia} show the maps made for 4 different offset corrections as explained in Sec.
+\ref{Sec: testes}.
+Table \ref{Tab: comparison-Himalia} show numerically the difference between the maps.
+A similar test was made for an occultation of Elara in March 30, 2015.
+The difference between the maps can be seen summarized in Table \ref{Tab: comparison-Elara}.
+
+Continuous observations of the satellites and candidates are recommended and new numerical integrations of the orbits of the satellites are expected to reduce the respective position errors.
+The GAIA catalogue, which are to be released in mid-2016 will improve the position error of the stars and occultations of fainter stars not present in the UCAC4 catalogue may be found as candidates.
+Consequently, the astrometry of the satellites with the GAIA catalogue will improve the position errors of the satellites that will allow for better predictions and a bigger probability of observe stellar occultations by irregular satellites.
 

### 2015-09-11 Irregular-occ/Irregular-occ.tex
.

@@
 \abstract
-%{The irregular satellites of the giant planets are believed to have been captured during the evolution of the solar system.
-Knowing their physical parameters, such as size, density and albedo is important to constrain where they came from and how they were captured.
-The best way to obtain these parameters are observations in situ by spacecrafts or from stellar occultations by the objects.
-Both techniques demand that the orbits are well known.}
-%{We aimed to obtain good astrometric positions of irregular satellites in order to improve their orbits and ephemeris.}
-%{We identified and reduced observations of several irregular satellites from three databases containing more than 8000 images obtained between 1992 and 2014 at three sites (Observatório do Pico dos Dias, Observatoire de Haute-Provence and European Southern Observatory - La Silla).
+{The irregular satellites of the giant planets are believed, due to their orbital configurations, to not have been formed where they are located, but they were probably captured in the early solar system.
+To constrain their origin location, it's important to know their physical parameters, such as size, density and albedo.
+These informations would to understand better their densities and consequently their composition.
+The best ground-based technique to obtain these parameters is the observation of stellar occultations by these objects.}
+{We aimed to predict stellar occultation for the seven biggest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope and Ananke; for the irregular satellite of Saturn: Phoebe; and the satellites of Neptune: Nereid and Triton.}
+{\textcolor{red}{We identified and reduced observations of several irregular satellites from three databases containing more than 8000 images obtained between 1992 and 2014 at three sites (Observatório do Pico dos Dias, Observatoire de Haute-Provence and European Southern Observatory - La Silla).
 We used the software PRAIA (Platform for Reduction of Astronomical Images Automatically) to make the astrometric reduction of the CCD frames.
@@
 The identification of the satellites in the frames was done through their ephemerides as determined from the SPICE/NAIF kernels.
-Some procedures were taken to overcome missing or incomplete information (coordinates, date), mostly for the older images.}
-%{We managed to obtain more than 6000 positions for 18 irregular satellites, being 12 of Jupiter, 4 of Saturn, 1 of Uranus (Sycorax) and 1 of Neptune (Nereid).
-For some satellites the number of obtained positions is more than 50\% of that used in earlier orbital numerical integrations.}
-%{Comparison of our positions with recent JPL ephemeris suggests the presence of systematic errors in the orbits for some of the irregular satellites.
-The most evident case was an error in the inclination of Carme.}
+Some procedures were taken to overcome missing or incomplete information (coordinates, date), mostly for the older images.}}
+{\textcolor{red}{We managed to obtain more than 6000 positions for 18 irregular satellites, being 12 of Jupiter, 4 of Saturn, 1 of Uranus (Sycorax) and 1 of Neptune (Nereid).
+For some satellites the number of obtained positions is more than 50\% of that used in earlier orbital numerical integrations.}}
+{\textcolor{red}{Comparison of our positions with recent JPL ephemeris suggests the presence of systematic errors in the orbits for some of the irregular satellites.
+The most evident case was an error in the inclination of Carme.}}
 
@@
 In this event, it is possible to see that the shade does not vary by much among the four maps suggesting thatthere was a good probability of observing the event.
-In fact, the biggest differences between the shadows of the four maps were \textcolor{red}{22s and 109km} in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
+In fact, the biggest differences between the shadows of the four maps were \textcolor{red}{22s} along the shadow path and \textcolor{red}{109km} in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
 
@@
 Still, the differences between the maps obtained were relatively small.
-The biggest differences between them were 74s along the shadow path and 307km perpendicular to it (see table \ref{Tab: comparison-Elara}).
+The biggest differences between them were \textcolor{red}{74s} along the shadow path and \textcolor{red}{307km} perpendicular to it (see table \ref{Tab: comparison-Elara}).
 

### 2015-09-11 Irregular-occ/Irregular-occ.tex
.

@@
           \inst{2,3},     
-%          F. Braga-Ribas
-%          \inst{2,4}, 
+          B. E. Morgado
+          \inst{1,2},
+          F. Braga-Ribas
+          \inst{2,4}, 
 %          D. N. da Silva Neto
@@
 %          \inst{1,2} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau, 75014 Paris, France},
-          B. E. Morgado
-          \inst{1},
 %          A. Dias-Oliveira
@@
 %              \email{arlot@imcce.fr}
-%              \and
-%Federal University of Technology - Paraná (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
+              \and
+Federal University of Technology - Paraná (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
 %              \and
@@
 {We aimed to predict stellar occultation for the seven biggest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope and Ananke; for the irregular satellite of Saturn: Phoebe; and the satellites of Neptune: Nereid and Triton.}
-{\textcolor{red}{We identified and reduced observations of several irregular satellites from three databases containing more than 8000 images obtained between 1992 and 2014 at three sites (Observatório do Pico dos Dias, Observatoire de Haute-Provence and European Southern Observatory - La Silla).
-We used the software PRAIA (Platform for Reduction of Astronomical Images Automatically) to make the astrometric reduction of the CCD frames.
-The UCAC4 catalogue represented the International Celestial Reference System in the reductions.
-The identification of the satellites in the frames was done through their ephemerides as determined from the SPICE/NAIF kernels.
-Some procedures were taken to overcome missing or incomplete information (coordinates, date), mostly for the older images.}}
-{\textcolor{red}{We managed to obtain more than 6000 positions for 18 irregular satellites, being 12 of Jupiter, 4 of Saturn, 1 of Uranus (Sycorax) and 1 of Neptune (Nereid).
-For some satellites the number of obtained positions is more than 50\% of that used in earlier orbital numerical integrations.}}
-{\textcolor{red}{Comparison of our positions with recent JPL ephemeris suggests the presence of systematic errors in the orbits for some of the irregular satellites.
-The most evident case was an error in the inclination of Carme.}}
+{We identified stars candidates to stellar occultation by the irregular satellites in the UCAC4 catalogue and a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument.
+These catalogue were crossed with the JPL ephemeris of the satellites using the methodology applied to identify stellar occultation by TNOs by \cite{Assafin2012}.
+The ephemeris were corrected of offsets using an numerical integration \textcolor{red}{(Here we may write something about after the work of Laurène)} of the positions obtained by \cite{GomesJunior2015} to improve the predictions.}
+{We managed to obtain \textcolor{red}{588} candidates of stellar occultations between the period of January, 2015 and December, 2017.
+Tests were develop to obtain offsets for the star and satellite for the events of Himalia in March 03, 2015 and Elara in March 30, 2015.
+The observations were made in the same field to minimize errors and the offsets were used to correct the predictions and evaluate them.}
+{The comparisons between the nominal predictions, the corrected predictions using the numerical integration and the observation tests show a good agreement.
+This reveals that observe an stellar occultation by these objects, at least for Himalia, is quite possible.}
 
@@
 
-We present in this paper stellar occultation predictions for the 7 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke and Sinope), Phoebe from Saturn and Nereid from Neptune.
+We present in this paper stellar occultation predictions for the 7 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke and Sinope), Phoebe from Saturn and Nereid and Triton from Neptune.
 In the section \ref{Sec: Rationale} we explore the scientific rationale for the study of the irregular satellites and the possibility of having a common origin with TNOs.
@@
 As explicited in section \ref{Sec: introducao}, there is no consensus on a single model about the region where the irregular satellites were formed.
+\cite{Cuk2004} showed that the progenitor of the Himalia group may have origin in heliocentric orbits similar to the Hilda asteroid group.
 
@@
 Making a new model for the orbits of these objects would demand a lot of time and delay the publication of predictions of possible events to occur in the very near future.
-Thus, we utilized the ephemeris offsets obtained by \cite{GomesJunior2015} to identify error patterns in the ephemeris.
+Thus, we utilized the positions obtained by \cite{GomesJunior2015} to identify error patterns in the ephemeris.
 The error patterns in right ascension and in declination could be used to extrapolate the offsets to the satellite ephemeris by the time of the predicted occultation, improving it.
-Plots of the offsets over time and true anomaly (see Fig.
-\ref{Fig:offxtime} and \ref{Fig:offxtanom} for Carme) clearly show the systematic errors in the JPL ephemeris.
+Plots of the offsets over true anomaly (see Fig.
+\ref{Fig:offxtanom} for declination offsets of Carme, for instance) clearly show the systematic errors in the JPL ephemeris.
 
-\begin{figure}
+%\begin{figure}
+%\begin{centering}
+%\includegraphics[scale=0.25]{figures/DEC.png}\label{Fig:offxtime}
+%\caption{Offsets of the declination of Carme by time. \textcolor{red}{Figura só para visualização, vou colocar alguma melhor depois}}
+%\end{centering}
+%\end{figure}
+
+\begin{figure*}
 \begin{centering}
-\includegraphics[scale=0.25]{figures/DEC.png}\label{Fig:offxtime}
-\caption{Offsets of the declination of Carme by time. \textcolor{red}{Figura só para visualização, vou colocar alguma melhor depois}}
+\subfigure[Offset from the JPL ephemeris]{\includegraphics[scale=0.25]{figures/DEC_anom.png} \label{Fig:offxtanomxjpl}}
+\subfigure[\textcolor{red}{Offset from the numerical integration}]{\includegraphics[scale=0.25]{figures/DEC_anom.png} \label{Fig:offxtanomxni}}
+\caption{Offsets of the declination of Carme by true anomaly from the JPL ephemeris and \textcolor{red}{from the numerical integration from the positions obtained in \cite{GomesJunior2015}}.}
 \end{centering}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
-\begin{centering}
-\includegraphics[scale=0.25]{figures/DEC_anom.png} \label{Fig:offxtanom}
-\caption{Offsets of the declination of Carme by true anomaly. \textcolor{red}{Figura só para visualização, vou colocar alguma melhor depois}}
-\end{centering}
-\end{figure}
-
-The two kinds of error or offset patterns adopted depending on the case are given by Eqs. \ref{eq:linear} and \ref{eq:seno}:
-
-\begin{equation}
-F(t,f) = p[0]\times t + p[1]\times \sin(f) + p[2]\times\cos(f) + p[3],
-\label{eq:linear}
-\end{equation}
-
-\begin{equation}
-\begin{split}
-%F = \{F_{x} \in  F_{c} &: (|S| > |C|) \\
-% &\quad \cap (\text{minPixels}  < |S| < \text{maxPixels}) \\
-% &\quad \cap (|S_{\text{conected}}| > |S| - \epsilon) \}
-F(t,f) = p[0]\times\sin\left(\frac{2\pi}{p[1]}\times t + p[2]\right) + p[3]\times\sin(f) + \\ +p[4]\times\cos(f) + p[5],
-\end{split}
-\label{eq:seno}
-\end{equation}
-where $F(t,f)$ is the offset obtained, $t$ is time in years counting from J2000.0 and $f$ is the true anomaly.
+%The two kinds of error or offset patterns adopted depending on the case are given by Eqs. \ref{eq:linear} and \ref{eq:seno}:
+%
+%\begin{equation}
+%F(t,f) = p[0]\times t + p[1]\times \sin(f) + p[2]\times\cos(f) + p[3],
+%\label{eq:linear}
+%\end{equation}
+%
+%\begin{equation}
+%\begin{split}
+%%F = \{F_{x} \in  F_{c} &: (|S| > |C|) \\
+%% &\quad \cap (\text{minPixels}  < |S| < \text{maxPixels}) \\
+%% &\quad \cap (|S_{\text{conected}}| > |S| - \epsilon) \}
+%F(t,f) = p[0]\times\sin\left(\frac{2\pi}{p[1]}\times t + p[2]\right) + p[3]\times\sin(f) + \\ +p[4]\times\cos(f) + p[5],
+%\end{split}
+%\label{eq:seno}
+%\end{equation}
+%where $F(t,f)$ is the offset obtained, $t$ is time in years counting from J2000.0 and $f$ is the true anomaly.
 The patterns could be applied for right ascension as well as for declination ephemeris offsets.
 
-\textcolor{red}{Falta justificar as funções}
+\textcolor{red}{...}
 
@@
 \textbf{Notes}.
-Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - in the original table, these coordinates are immediately followed by the geocentric astrometric equatorial coordinates of the satellite (corrected the computed ephemeris offset using Eqs \ref{eq:linear} or \ref{eq:seno}); C/A: the geocentric closest approach, in arcseconds; P/A: the planet position angle with respect to the occulted star at C/A, in degrees; velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; D: planet range to Earth, in AU; R*: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{Mag}* = \textrm{Mag}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$.
+Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - in the original table, these coordinates are immediately followed by the geocentric astrometric equatorial coordinates of the satellite (corrected the computed ephemeris offset derived in Sec.
+\ref{Sec: Offset}); C/A: the geocentric closest approach, in arcseconds; P/A: the planet position angle with respect to the occulted star at C/A, in degrees; velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; D: planet range to Earth, in AU; R*: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{Mag}* = \textrm{Mag}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$.
 A value of 50.0 means that the star is not in the 2MASS; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\Delta e_{\alpha *}$ and $\Delta e_{\delta}$: computed offsets in mas applied to the ephemeris right ascension and declination, respectively; pm: ok = proper motion applied, no = no proper motion applied; catalogue cross-identification (ct) = uc (UCAC2), 2m (2MASS), fs (field star); $E_{\alpha *}$ and $E_{\delta}$: uncertainties (mas) in right ascension and declination.
@@
 Concerning the UCAC4 stars, an accuracy of about 20-50 mas in the star position would give a 15-40\% probability of observe such an event by Ananke, the smallest irregular satellite in the sample.
-For Himalia, the occultation would have 80-100\% probability.
+For Himalia, the occultation would have 80-100\% probability of success.
 Combined with a mean accuracy in the position of 30 mas for Himalia and 100 mas for Ananke mas published in Table 2 of \cite{Jacobson2012}, the probability of success would be around 50-80\% for Himalia and $\approx$5\% for Ananke.
@@
 Fig.
-\ref{Fig: occ-Himalia} show the maps made for 4 different offset corrections as explained in Sec.
+\ref{Fig: occ-Himalia} shows the maps made for 4 different offset corrections as explained in Sec.
 \ref{Sec: testes}.
-Table \ref{Tab: comparison-Himalia} show numerically the difference between the maps.
+Table \ref{Tab: comparison-Himalia} shows numerically the difference between the maps.
 A similar test was made for an occultation of Elara in March 30, 2015.
-The difference between the maps can be seen summarized in Table \ref{Tab: comparison-Elara}.
+The difference between the maps in this case can be seen summarized in Table \ref{Tab: comparison-Elara}.
 
 Continuous observations of the satellites and candidates are recommended and new numerical integrations of the orbits of the satellites are expected to reduce the respective position errors.
-The GAIA catalogue, which are to be released in mid-2016 will improve the position error of the stars and occultations of fainter stars not present in the UCAC4 catalogue may be found as candidates.
+The GAIA catalogue, which are to be released in the end of 2016 will improve the position error of the stars and will allow for the detection of occultations candidates of fainter stars, which are not present in the UCAC4 catalogue.
 Consequently, the astrometry of the satellites with the GAIA catalogue will improve the position errors of the satellites that will allow for better predictions and a bigger probability of observe stellar occultations by irregular satellites.

### 2015-09-14 Irregular-occ/Irregular-occ.tex
.

@@
 \cite{Cuk2004} showed that the progenitor of the Himalia group may have origin in heliocentric orbits similar to the Hilda asteroid group.
+\cite{Sheppard2005} stated that the irregular satellites may be some of the objects that were formed within the giant planets region.
+\cite{Sheppard2005} and \cite{Jewitt2007} also expose the possibility for the irregular satellites having their origin as comets or similar to the TNOs.
 
 Phoebe, the most studied irregular satellite, was identified by \cite{Clark2005} that its surface is probably covered by material of cometary origin.
-It was also stated by \cite{Johnson2005} that if the porosity of Phoebe would be of about 15\%, Phoebe would have an uncompressed density similar to those of Pluto and Triton.
+It was also stated by \cite{Johnson2005} that if the porosity of Phoebe would be of about 15\%, which the authors believe it is possible, Phoebe would have an uncompressed density similar to those of Pluto and Triton.
 
@@
 Plots of the offsets over true anomaly (see Fig.
-\ref{Fig:offxtanom} for declination offsets of Carme, for instance) clearly show the systematic errors in the JPL ephemeris.
+\ref{Fig:offxanom} for declination offsets of Carme, for instance) clearly show the systematic errors in the JPL ephemeris.
 
@@
 \begin{centering}
-\subfigure[Offset from the JPL ephemeris]{\includegraphics[scale=0.25]{figures/DEC_anom.png} \label{Fig:offxtanomxjpl}}
-\subfigure[\textcolor{red}{Offset from the numerical integration}]{\includegraphics[scale=0.25]{figures/DEC_anom.png} \label{Fig:offxtanomxni}}
-\caption{Offsets of the declination of Carme by true anomaly from the JPL ephemeris and \textcolor{red}{from the numerical integration from the positions obtained in \cite{GomesJunior2015}}.}
+\subfigure[Offset from the JPL ephemeris]{\includegraphics[width=8.8cm]{figures/DEC_anom.png} \label{Fig:offxtanomxjpl}}
+\subfigure[\textcolor{red}{Offset from the numerical integration}]{\includegraphics[width=8.8cm]{figures/DEC_anom.png} \label{Fig:offxtanomxni}}
+\caption{Offsets of the declination of Carme by true anomaly from the JPL ephemeris and \textcolor{red}{from the numerical integration of the positions obtained in \cite{GomesJunior2015}}.
+\label{Fig:offxanom}}
 \end{centering}
@@
 
-A total of 588 events \textcolor{red}{(Não estou contando ainda os eventos para Tritão)} were identified between January 2015 and December 2017.
+A total of 588 events \textcolor{red}{(Not counting Triton's event yet)} were identified between January 2015 and December 2017.
 Table \ref{Tab: occ-list} exemplifies the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps (see Fig.
@@
 \begin{table*}
-\caption{\label{Tab: sample-cds} Stellar occultation sampled predictions for Himalia \textcolor{red}{Consertar tabela}}
+\caption{\label{Tab: sample-cds} Stellar occultation sampled predictions for Himalia \textcolor{red}{I still have to fix the table}}
 \begin{centering}
@@
 \begin{centering}
-\subfigure[Nominal prediction]{\includegraphics[width=8.8cm]{figures/Himalia_nominal.eps}  \label{Fig: occ-Himalia-nooff}}
-\subfigure[\textcolor{red}{Offset de Himalia a partir do ajuste}]{\includegraphics[width=8.8cm]{figures/Himalia2_2015-03-03T00:39:18.png}  \label{Fig: occ-Himalia-ajuste}}
+\subfigure[Nominal prediction from JPL]{\includegraphics[width=8.8cm]{figures/Himalia_nominal.eps}  \label{Fig: occ-Himalia-nooff}}
+\subfigure[\textcolor{red}{From the numerical integration}]{\includegraphics[width=8.8cm]{figures/Himalia2_2015-03-03T00:39:18.png}  \label{Fig: occ-Himalia-ajuste}}
 \subfigure[Observation at February 22 (different FOVs)]{\includegraphics[width=8.8cm]{figures/Himalia_22fev.eps}   \label{Fig: occ-Himalia-off22fev}}
@@
 The green dot at the bottom of the maps is the location of Observatório do Pico dos Dias.
-\textcolor{red}{Figura b será mudada após confirmarmos o offset para Himalia}}
+\textcolor{red}{Figure b waiting for the offset of Himalia}}
 \label{Fig: occ-Himalia}
@@
 \begin{table}
-\caption{\label{Tab: comparison-Himalia} Comparison between the predictions of the Himalia occultation at March 03, 2015. \textcolor{red}{Tem que ver direito o offset do ajuste}}
+\caption{\label{Tab: comparison-Himalia} Comparison between the predictions of the Himalia occultation at March 03, 2015. \textcolor{red}{Waiting for the right offset}}
 \begin{centering}
@@
 \begin{table}
-\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara occultation at March 30, 2015. \textcolor{red}{Tem que ver direito o offset do ajuste}}
+\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara occultation at March 30, 2015. \textcolor{red}{Waiting for the right offset}}
 \begin{centering}
@@
 \par \end{centering}
-Occultations predicted using the UCAC4 catalogue. For Nereid and Triton, the occultations in 2015 and 2016 were predicted using the WFI catalogue as explained in Sec. \ref{Sec: predictions}.
+Occultations predicted using the UCAC4 catalogue.
+For Nereid and Triton, the occultations in 2015 and 2016 were predicted using the WFI catalogue as explained in Sec.
+\ref{Sec: predictions} \textcolor{red}{Numbers of Triton yet to be written}.
 \end{table}

### 2015-09-15 Irregular-occ/Irregular-occ.tex
.

@@
 \abstract
-{The irregular satellites of the giant planets are believed, due to their orbital configurations, to not have been formed where they are located, but they were probably captured in the early solar system.
-To constrain their origin location, it's important to know their physical parameters, such as size, density and albedo.
-These informations would to understand better their densities and consequently their composition.
-The best ground-based technique to obtain these parameters is the observation of stellar occultations by these objects.}
-{We aimed to predict stellar occultation for the seven biggest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope and Ananke; for the irregular satellite of Saturn: Phoebe; and the satellites of Neptune: Nereid and Triton.}
-{We identified stars candidates to stellar occultation by the irregular satellites in the UCAC4 catalogue and a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument.
-These catalogue were crossed with the JPL ephemeris of the satellites using the methodology applied to identify stellar occultation by TNOs by \cite{Assafin2012}.
-The ephemeris were corrected of offsets using an numerical integration \textcolor{red}{(Here we may write something about after the work of Laurène)} of the positions obtained by \cite{GomesJunior2015} to improve the predictions.}
+{Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system.
+It is important to know their physical parameters, such as size and albedo, to trace back their true origin.
+The best ground-based technique to obtain such parameters is the observation of stellar occultations by these objects.}
+{We aimed to predict stellar occultations for the seven biggest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope and Ananke; for the irregular satellite of Saturn: Phoebe; and the satellites of Neptune: Nereid and Triton.}
+{We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument.
+These catalogues were crossed with the ephemeris of the satellites to identify stellar occultations.
+The ephemeris were generated by a numerical integration \textcolor{red}{(Here we may write something about after the work of Laurène)} that used the positions obtained by \cite{GomesJunior2015} to improve the predictions.}
 {We managed to obtain \textcolor{red}{588} candidates of stellar occultations between the period of January, 2015 and December, 2017.
-Tests were develop to obtain offsets for the star and satellite for the events of Himalia in March 03, 2015 and Elara in March 30, 2015.
-The observations were made in the same field to minimize errors and the offsets were used to correct the predictions and evaluate them.}
-{The comparisons between the nominal predictions, the corrected predictions using the numerical integration and the observation tests show a good agreement.
-This reveals that observe an stellar occultation by these objects, at least for Himalia, is quite possible.}
+We made observational tests for the predictions of the events of Himalia in March 03, 2015 and Elara in March 30, 2015.
+The observations were made in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions.}
+{The comparisons between the predictions and the observation tests show a good agreement. We discuss how the successful observation of a stellar occultation by these objects is quite possible.}
 
@@
 
-In this work, we propose to study these objects as possible representatives of the small TNOs' population.
-The basis for this hypothesis is discussed in Section \ref{Sec: Rationale}.
-TNOs are highly interesting objects that due to their large heliocentric distances (low temperatures) may be highly preserved having their physical properties similar to those they had when they were formed, then providing history and evolution of the outer solar system \citep{Camargo2013}.
-This is even more true for the smaller objects, since in principle larger sizes favour physical differentiation processes in the body and vice-versa.
-However, due to the distance, the smaller TNOs from this region are more difficult to observe.
-A clever way to overcome this difficulty is to study  the much closer irregular satellites, under the hypothesis that they share a common origin with the small TNOs' population.
+In order to obtain precise fundamental physical parameters like size, shape and albedo for the irregular satellites and thus to contribute to the study of their origin, we will make use of stellar occultations, which provide  more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
 
-In order to obtain precise fundamental physical parameters like size, shape and albedo for the irregular satellites (and thus, in a sense, to small-sized TNOs), we will make use of stellar occultations, which provide  more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
-
-Up to date no stellar occultation by irregular satellite is found in the literature.
-Since their estimated sizes are very small (see table \ref{Tab: satellite-diameter}), prediction of the exact location and instant where the shadow will cross the Earth demands a good accurate ephemeris and star positions.
-For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds).Thus, even in the best cases, the overall budget of ephemeris and star position errors must be around 40 mas in order that we get good chances of observing a stellar occultation.
+No observation of a stellar occultation by an irregular satellite was published up to date.
+Since their estimated sizes are very small (see Table \ref{Tab: satellite-diameter}), this may have prevented earlier tries.
+But, in fact, given their distances to us, current ephemeris and star positions are already good enough for the prediction of the exact location and instant where the shadow of the occultation will cross the Earth.
+For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds).Thus, in this case, the overall budget of ephemeris and star position errors must be around 40 mas in order that we get good chances of observing a stellar occultation, which is quite feasible today.
 
@@
 
-Unlikely stellar occultations by TNOs, which has been proved effective, the observation of an occultation by a irregular satellite is more favorable.
-They have been known for more time and observed in many orbital period having better ephemeris than TNOs.
-Moreover, the irregular satellites are closer to Earth which means a minor error in kilometers.
+Unlike stellar occultations by TNOs, which nevertheless have been proved effective, the observation of an occultation by a irregular satellite is, in principle, more favorable.
+The orbits of their host planets are well known and these satellites have been observed completing already many orbital periods around them, thus presenting better ephemeris than TNOs.
+Moreover, the irregular satellites are closer to Earth which means a minor location error in kilometers.
 
@@
 For an object at the distance of Jupiter, this represents an error bigger than 700 km in the shadow path.
-Using the offsets obtained by \cite{GomesJunior2015} it is possible to correct the ephemeris and better predict stellar occultations for these objects.
+Using the positions obtained by \cite{GomesJunior2015} it is possible to improve the ephemeris and better predict stellar occultations for these objects.
 
@@
 In the section \ref{Sec: Rationale} we explore the scientific rationale for the study of the irregular satellites and the possibility of having a common origin with TNOs.
-In section \ref{Sec: Offset} we show the correction made to the ephemeris of these satellites for better predicting the stellar occultations.
+In section \ref{Sec: Offset} we show the \textcolor{red}{numerical integration to improve the ephemeris}.
 In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites and how they were made.
@@
 
-As explicited in section \ref{Sec: introducao}, there is no consensus on a single model about the region where the irregular satellites were formed.
-\cite{Cuk2004} showed that the progenitor of the Himalia group may have origin in heliocentric orbits similar to the Hilda asteroid group.
+There is no consensus for a single model explaining where the irregular satellites were formed.
+\cite{Cuk2004} showed that the progenitor of the Himalia group may have originated in heliocentric orbits similar to the Hilda asteroid group.
 \cite{Sheppard2005} stated that the irregular satellites may be some of the objects that were formed within the giant planets region.
+
+Phoebe is the most studied irregular satellite.
+\cite{Clark2005} suggest that its surface is probably covered by material of cometary origin.
+It was also stated by \cite{Johnson2005} that if the porosity of Phoebe would be of about 15\%, which the authors believe it is possible, Phoebe would have an uncompressed density similar to those of Pluto and Triton.
+
 \cite{Sheppard2005} and \cite{Jewitt2007} also expose the possibility for the irregular satellites having their origin as comets or similar to the TNOs.
-
-Phoebe, the most studied irregular satellite, was identified by \cite{Clark2005} that its surface is probably covered by material of cometary origin.
-It was also stated by \cite{Johnson2005} that if the porosity of Phoebe would be of about 15\%, which the authors believe it is possible, Phoebe would have an uncompressed density similar to those of Pluto and Triton.
+TNOs are highly interesting objects that due to their large heliocentric distances (low temperatures) may be highly preserved having their physical properties similar to those they had when they were formed, then providing history and evolution of the outer solar system \citep{Camargo2014}.
+This is even more true for the smaller objects, since in principle larger sizes favour physical differentiation processes in the body and vice-versa.
+However, due to the distance, the smaller TNOs from this region are more difficult to observe.
+A clever way to overcome this difficulty is to study  the much closer irregular satellites, under the hypothesis that they share a common origin with the small TNOs' population.
 
@@
 Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
-Because Triton's orbit is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded of the irregular satellites' class, but studied together by many authors \citep{Sheppard2005, Jewitt2007}.
+Because Triton's orbit size is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded from the irregular satellites' class, but studied together by many authors \citep{Sheppard2005, Jewitt2007}.
 
-Similarly to the irregular satellites, Triton was probably capture in the early Solar System and may have a similar origin as the TNOs \citep{Agnor2006}.
+Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have a similar origin as the TNOs \citep{Agnor2006}.
 Differently, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
-The main objective to study Triton by stellar occultation is to understand the evolution of its atmosphere due to Triton's complicated and extreme seasonal cycle \citep{McKinnon2007}.
+The main motivation here to study Triton by stellar occultations is to understand the evolution of its atmosphere due to Triton's complicated and extreme seasonal cycle \citep{McKinnon2007}.
 
@@
 
-\begin{figure}
+\begin{figure*}
 \begin{centering}
-\includegraphics[scale=0.2]{figures/Himalia_2015-10-23T07:54:47.png} 
+\includegraphics[scale=0.4]{figures/Himalia_2015-10-23T07:54:47.png} 
 \caption{Occultation map for Himalia regarding to the second event sampled in Table \ref{Tab: occ-list}}
@@
 \end{centering}
-\end{figure}
+\end{figure*}
 

### 2015-09-18 Irregular-occ/Irregular-occ.tex
.

@@
 The precise star positions to be derived by GAIA will render better predictions with the only source of error being the ephemeris.
-The work of \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions from 2017 onwards will be better predicted.
-For a matter of precaution and preparations of campaigns, we publish stellar occultations with UCAC4 until 2017.
+The work of \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions from 2017 onwards will be better.
+For that, we only publish stellar occultations with UCAC4 until 2017.
 
@@
 
-Observe a stellar occultation by an irregular satellites demands a great effort.
-The shadow covers a very restricted area on Earth because the irregular satellites are small.
-So, before we start a large observational campaign, we tested some occultation predictions for larger objects, to assess the quality of the prediction.
+Observe a stellar occultation demands a great effort.
+And, in our case, the shadow covers a very restricted area on Earth because the irregular satellites are small.
+So, before we start a large observational campaign, we tested some occultation predictions for large targets, to assess the quality of the predictions.
 
@@
 Thus, the relative positions between the two objects had minimal influence of the errors of the reference catalogue of stars used and possible field distortions \citep[and references therein]{Peng2008}.
-The offsets of the positions of the star and of the satellites were used to check the original predictions.
+The relative positions of the star and of the satellites were used to check the original predictions.
 
@@
 We then corrected the positions of the satellite by an offset calculated by the method showed in section \ref{Sec: Offset} to make the map \ref{Fig: occ-Himalia-ajuste}.
-The map \ref{Fig: occ-Himalia-off22fev} was made from obtained positions on February 22 observed with the Zeiss telescope (Diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / px$) at the Observatório do Pico dos Dias (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
+The map \ref{Fig: occ-Himalia-off22fev} was made from obtained positions on February 22 observed with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / px$) at the Observatório do Pico dos Dias (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
 On that day, Himalia and the star were observed in separate FOVs as they were still far apart.
-On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope (Diameter = 1.6m; FOV = $5\farcm8$; pixel scale = $0\farcs17 / px$) at OPD just over an hour after the time scheduled for the event.
+On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope (diameter = 1.6m; FOV = $5\farcm8$; pixel scale = $0\farcs17 / px$) at OPD just over an hour after the time scheduled for the event.
 Satellite and star were separated by about 16 arcsec, so very close to each other.
@@
 
-In this event, it is possible to see that the shade does not vary by much among the four maps suggesting thatthere was a good probability of observing the event.
-In fact, the biggest differences between the shadows of the four maps were \textcolor{red}{22s} along the shadow path and \textcolor{red}{109km} in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
+In this event, it is possible to see that the shade does not vary by much among the four maps suggesting that there was a good probability of observing the event.
+In fact, the largest differences between the shadows of the four maps were \textcolor{red}{22s} along the shadow path and \textcolor{red}{109km} in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
 
@@
 Still, the differences between the maps obtained were relatively small.
-The biggest differences between them were \textcolor{red}{74s} along the shadow path and \textcolor{red}{307km} perpendicular to it (see table \ref{Tab: comparison-Elara}).
+The largest differences between them were \textcolor{red}{74s} along the shadow path and \textcolor{red}{307km} perpendicular to it (see table \ref{Tab: comparison-Elara}).
 
@@
 We presented stellar occultations for the period of 2015-2017 for seven irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid.
-The procedure used was the same that predicted stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.
+The procedure used was the same as that for the prediction of stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.
 
@@
 
-A total of \textcolor{red}{588} events were detected to happen between and the period presented.
-In table \ref{Tab: satellite-occultation} it's presented the number of stellar occultations predicted by year for each satellite.
+A total of \textcolor{red}{588} events were foreseen. In table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite.
 
@@
 
-Concerning the UCAC4 stars, an accuracy of about 20-50 mas in the star position would give a 15-40\% probability of observe such an event by Ananke, the smallest irregular satellite in the sample.
-For Himalia, the occultation would have 80-100\% probability of success.
-Combined with a mean accuracy in the position of 30 mas for Himalia and 100 mas for Ananke mas published in Table 2 of \cite{Jacobson2012}, the probability of success would be around 50-80\% for Himalia and $\approx$5\% for Ananke.
-Observations a few days before the date of occultation predicted may diminish the combined accuracy to 40-80 mas, depending on the magnitude of the objects.
+The probability of successfully observing an occultation is roughly the ratio of the satellite's diameter by the budget error of ephemeris and star position.
+Thus, UCAC4 errors ranging between 20 mas - 50 mas combined with a mean accuracy in the position of 30 mas for Himalia and 100 mas for Ananke mas published in Table 2 of \cite{Jacobson2012} would give 80\%-50\% probability of observing such an event by Himalia and $\approx$5\% for Ananke, the smallest irregular satellite in the sample.
+Observations a few days before the date of occultation predicted may improve the combined accuracy to 40-80 mas, depending on the magnitude of the objects.
 
-The tests made with an occultation expected to happen in March 03, 2015 for Himalia showed that this event could probably be observed in case there were observers available in the shadow area.
+The tests made with an occultation expected to happen in March 03, 2015 for Himalia showed that this event could probably have been observed in case there were observers available in the shadow area.
 Fig.
@@
 A similar test was made for an occultation of Elara in March 30, 2015.
-The difference between the maps in this case can be seen summarized in Table \ref{Tab: comparison-Elara}.
+The difference between the maps in this case are listed in Table \ref{Tab: comparison-Elara}.
 
 Continuous observations of the satellites and candidates are recommended and new numerical integrations of the orbits of the satellites are expected to reduce the respective position errors.
-The GAIA catalogue, which are to be released in the end of 2016 will improve the position error of the stars and will allow for the detection of occultations candidates of fainter stars, which are not present in the UCAC4 catalogue.
-Consequently, the astrometry of the satellites with the GAIA catalogue will improve the position errors of the satellites that will allow for better predictions and a bigger probability of observe stellar occultations by irregular satellites.
+The GAIA catalogue, which is to be released at the end of 2016 will improve the position error of the stars and will allow for the detection of occultations candidates of fainter stars, which are not present in the UCAC4 catalogue.
+Consequently, the astrometry of the satellites with the GAIA catalogue will improve the position errors allowing for better predictions and a higher probabilities of observing stellar occultations by irregular satellites.
 

### 2015-11-11 relatorio_2015/Relatorio.tex
Adicionado relatorio da pos 2015

\documentclass[12pt,a4paper]{report}
\usepackage[english,brazilian]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amsthm,amsfonts,amssymb,textcomp}
%\usepackage{latexsym}
\usepackage{graphicx}
\graphicspath{{figuras}}
\usepackage{subfigure}
\usepackage{float}
\usepackage{longtable}
\usepackage{color}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage[breaklinks=true]{hyperref}
\usepackage[comma,authoryear]{natbib}
\usepackage[nonumberlist]{glossaries}
\usepackage{arydshln}
\usepackage{footnote}
\usepackage[left=3cm,right=2cm,top=3cm,bottom=2cm]{geometry}
%\usepackage[alf]{abntcite}

%%% newcommand %%%%%%%%%%%%%%%%%
\newcommand{\PE}{Perkin-Elmer}
\newcommand{\BC}{Boller \& Chivens}

\newcommand{\degr}{\ensuremath{^{\circ}}}%                    % degree symbol:  °
\newcommand{\arcmin}{\ensuremath{^{\prime}}}%                    % degree symbol:  °
\newcommand{\arcsec}{\ensuremath{^{\prime\prime}}}%                    % degree symbol:  °
\newcommand{\fs}{\mbox{\ensuremath{.\!\!^s}}}
\newcommand{\farcm}{\mbox{\ensuremath{.\mkern-4mu^\prime}}}%    % fractional arcminute symbol: 0.'0
\newcommand{\farcs}{\mbox{\ensuremath{.\!\!^{\prime\prime}}}}%  % fractional arcsecond symbol: 0.''0
\newcommand{\fdg}{\mbox{\ensuremath{.\!\!^\circ}}}%             % fractional degree symbol:     0.°0

%\makeglossaries

\makeatletter
\renewcommand\chapter{\thispagestyle{plain}
                \global\@topnum\z@
                \@afterindentfalse
                \secdef\@chapter\@schapter}
\makeatother

\makeatletter
\renewcommand{\@makechapterhead}[1]{%
\vspace*{50 pt}%
{\setlength{\parindent}{0pt} \raggedright \normalfont
\bfseries\Huge\thechapter.\ #1
\par\nobreak\vspace{40 pt}}}
\makeatother

%\input{glossario}
\author{Altair Ramos Gomes Júnior}
\title{Um Estudo Astrométrico e Fotométrico do Sistema Solar Exterior}
\begin{document}

\maketitle

%\pagestyle{headings}

\chapter{Introdução}
\label{Cap: intro}

\indent \indent O estudo de objetos como objetos trans-Neptunianos (TNOs), Centauros e Satélites Irregulares nos ajudam a compreender a formação e evolução do Sistema Solar Exterior. Nesta região distante do Sol, de baixas temperaturas, objetos principalmente de tamanho relativamente pequeno, e mais dispersos no espaço, provavelmente sofreram muito pouca diferenciação, seja por mecanismos internos, seja por choques com outros corpos, comparados a objetos formados mais próximos do Sol.

Por serem corpos asteroidais localizados de forma dispersa além da órbita de Netuno, considera-se que os TNOs guardam estruturas e composições relativamente inalteradas em relação a sua época de formação, constituindo-se assim em corpos de prova de grande valor para o estudo da origem do Sistema Solar, ao menos para essa região exterior.

Tirando Plutão, o primeiro TNO foi descoberto há pouco mais de 20 anos \citep{Jewitt1993}. Por isso, as propriedades básicas desta população ainda não estão inteiramente estabelecidas, como a distribuição de tamanhos, composição, estruturas internas.

Centauros são uma população transiente entre TNOs e cometas da família de Júpiter, orbitando em uma região entre Júpiter e Netuno. Atualmente, cerca de 400 deles são conhecidos e é geralmente aceito que eles compartilham uma origem comum com os objetos do cinturão de Kuiper. Uma vez que Centauros são tipicamente mais brilhantes que TNOs por estarem mais próximos, eles servem como representantes a partir dos quais é possível inferir propriedades mais gerais sobre os objetos mais distantes \citep{Fernandez2002}. Dessa forma, a partir daqui, TNOs e Centauros serão tratados indistintamente como uma única população e referida unicamente como "TNOs".

Atualmente, é aceito que TNOs tenham sido formados nas partes mais internas do Sistema Solar. Eles teriam então sido colocados em suas posições atuais devido a troca de momento angular entre os planetas e planetésimos quando da migração dos planetas gigantes. A evolução se deu de tal forma que a passagem dos planetesimais e planetas por zonas de ressonância de movimento médio redefiniu as órbitas desses corpos \citep{Tsiganis2005}.

Muitos dos objetos que pertenciam ao cinturão de Kuiper primordial acabaram sendo enviados pela interação com os planetas gigantes para as partes mais internas do Sistema Solar. Alguns podem ter sido capturados pelos planetas criando a população de satélites irregulares ou satélites externos \citep{Nesvorny2007}, troianos \citep{Morbidelli2005} ou até mesmo para o cinturão principal de asteroides como proposto para Ceres por \cite{McKinnon2012}. Estudar esses objetos é de grande importância para entender a evolução do Sistema Solar.

Desses objetos, Tritão possui um interesse particular. Ele foi capturado por Netuno \citep{McKinnon2007} em uma órbita retrógrada e próxima ao planeta. Sua superfície é deformada com características tectônicas e possivelmente criovulcânicas \citep{Nimmo2015} e as propriedades físicas já conhecidas de Tritão mostram uma similaridade com as de Plutão. Além do mais, é um dos poucos objetos do Sistema Solar que se sabe possuir atmosfera.

Sabe-se que poucas sondas espaciais foram enviadas para estudar o Sistema Solar externo e que a quantidade de objetos estudados é muito pequena. Por isso, ainda hoje, as observações de solo tem se mostrado de grande importância.

Os sistemas de Júpiter e Saturno já foram visitados pelas Voyager I e II, Galileu (Júpiter) e Cassini (Saturno), porém apenas Saturno continua sendo investigado por uma sonda. Todavia, as sondas observaram apenas os planetas, os anéis e satélites mais internos. Os satélites externos ou foram pouco observados (como Phoebe) ou simplesmente não foram observados.

Já no caso de Urano e Netuno, nenhuma sonda exclusiva foi enviada, apenas as \textit{Voyagers I} e \textit{II} os visitaram, mas não permaneceram nos sistemas. Recentemente, a sonda \textit{New Horizons} visitou Plutão \citep{Stern2015} e obteu parâmetros físicos para o planeta anão e seus satélites (primeira visita por sonda a um objeto do cinturão de Kuiper), porém foi uma passagem rápida e o acompanhamento da evolução do sistema, incluindo a evolução da atmosfera de Plutão se dará por observações de solo.

A quantidade de objetos descobertos além da órbita de Saturno tem aumentado muito desde o fim do século passado. Como são raras as oportunidades em que uma sonda se aproxima desses objetos, a obtenção de suas características físicas ficam a cargo de observações de solo ou de
telescópios espaciais.

Um método que tem se mostrado eficiente para a obtenção desses parâmetros é o método de ocultações estelares, que proporciona medidas tão precisas que são apenas superadas por medidas oriundas de sondas. De fato, a sonda \textit{Dawn} já está orbitando Ceres, um dos objetos deste estudo, e a sonda \textit{New Horizons} visitou Plutão. As observações feitas por essas sondas trazem a oportunidade perfeita para comparar os resultados com os obtidos pelas nossas ocultações e calibrar a técnica, de forma que ocultações feitas para outros objetos possam ser mais precisas.

O objetivo desse Doutorado é observar objetos do Sistema Solar exterior, fazer astrometria para melhorar suas efemérides de forma que ocultações estelares por esses objetos possam ser preditas com precisão, observar as ocultações e obter os parâmetros físicos. Além disso, será feito um estudo astrométrico do sistema Netuno-Tritão para obter parâmetros relacionados a efeitos de maré, que permitem em princípio inferir a estrutura interna desses corpos.

Assim, além do trabalho rotineiro de estudo de TNOs e Centauros feito pelo grupo, do qual eu participo ativamente, nos propomos a estudar aqui, no contexto observacional acima, alguns objetos que, por hipótese, também podem ser representativos da população original de TNOs, isto é, podem ter uma origem comum aos TNOs. São eles alguns dos satélites irregulares de planetas gigantes, Ceres e Tritão.

%Nos capítulos \ref{Cap: TNOs}, \ref{Cap: Ceres}, \ref{Cap: Irregulares} e \ref{Cap: Netuno-Tritao} relatamos os trabalhos desenvolvidos envolvendo predições e observações de ocultações estelares com Centauros e TNOs, Ceres, satélites irregulares dos planetas gigantes e com o sistema Netuno-Tritão, respectivamente. Por fim, no capítulo \ref{Cap: perspectivas} resumimos o que já foi desenvolvido e descrevemos o que ainda está para ser desenvolvido.

%\chapter{Astrometria}
%\label{Cap: Astrometria}
%
%
%\indent \indent O tratamento astrométrico das imagens é feito utilizando a Plataforma para Redução Automática de Imagens Astronômicas \citep[PRAIA][]{Assafin2011}. As medidas (x, y) são obtidas através de uma Gaussiana bidimensional simétrica ajustada até uma largura à altura (FWHM = seeing). Em uma FWHM, o perfil da imagem é bem descrito por um perfil gaussiano, livre das distorções de borda que prejudicam a determinação do centro. O PRAIA identifica automaticamente as estrelas de catálogo e determina ($\alpha$, $\delta$) com um modelo definido pelo usuário que relaciona as medidas (x, y) da imagem e as coordenadas projetadas no plano do céu (X, Y). Os alvos são automaticamente identificados nas reduções, a partir das suas efemérides (JPL, NAIF, etc) e suas posições, offsets de efeméride e estimativa de erros de medida e de posição guardados para análise posterior.
%
%Majoritariamente, o catálogo UCAC4 \citep{Zacharias2013} é utilizado como representante prático do ICRS. Como será exposto na seção \ref{Sec: TNOs-astrometria}, o catálogo WFI é utilizado para determinados TNOs até 2015. Esses catálogos serão substituídos pelo catálogo GAIA uma vez que seja lançado em meados de 2016.
%
%Para cada frame, usamos um modelo polinomial que pode ser de grau 1, 2 ou 3 para relacionar as medidas (x, y) com as coordenadas (X, Y) do plano tangente. Para observações feitas no ESO, o procedimento astrométrico seguido é o mesmo descrito em \citep{Assafin2012}; as medidas (x, y) dos CCDs individuais são pré-corrigidos por um padrão de distorção de campo e todas as posições vindas de diferentes CCDs e mosaicos são combinadas usando o polinômio de grau três para produzir uma solução global para cada noite e campo observado, a as posições ($\alpha$, $\delta$) são então obtidas no sistema do catálogo utilizado. Estrelas de referência com resíduos de posição (O-C) que desviam mais de 120 mas\footnote{mili-segundos de arco (mili-arcsecond, \textit{mas}, em inglês)} (ou valor similar adotado pelo usuário, usualmente 3 sigma do erro do catálogo) são automaticamente eliminadas uma a uma.
%
%Um fator importante do trabalho está na utilidade estatística de se usar uma base de dados com um número extremamente elevado de observações (imagens). Esta é a maneira pela qual os efeitos atmosféricos de turbulência e extinção são efetivamente alisados e as informações de posição e de brilho são corretamente extraídas das imagens CCD.
%
%O tratamento astrométrico e fotométrico do PRAIA é capaz de lidar com uma imensa quantidade de dados de forma rápida, sendo capaz de obter a qualidade necessária para atingir os objetivos científicos dos programas observacionais em tempo recorde de processamento. As implementações incorporadas no software refletem experiência de mais de 20 anos de atuação na área da redução astrométrica e fotométrica de imagens astronômicas digitalizadas.
%
%Há ainda o cálculo de magnitudes instrumentais via fotometria de abertura e PSF, calculando-se o ponto zero em relação ao sistema fotométrico do catálogo de referência, como o 2MASS, UCAC2 ou o UCAC4 (o ponto zero pode ser re-definido). A magnitude do fundo de céu também é calculada. A FWHM das estrelas (seeing) é calculada automaticamente, sendo dados valores individuais por estrela, e valores médios, com as respectivas incertezas.
%
%Todas as informações fotométricas e astrométricas da redução, para todos os catálogos, e para cada campo, são guardadas em arquivos de dados. Também são gerados arquivos listando as estatísticas gerais de todos os campos. Os arquivos são de fácil manuseio nas tarefas de análise numérica e gráfica dos resultados. Além das posições obtidas, listam-se os parâmetros Gaussianos dos ajustes de cada estrela, os erros de medida em (x,y) e em posição, número de estrelas de referência, movimentos próprios, época da observação, magnitudes UCAC2, UCAC4, 2MASS, magnitudes instrumentais, informações da observação (filtros usados, tempo de exposição, etc). Em particular, são listados em arquivos separados os dados pertinentes aos alvos identificados nos campos, incluindo-se as diferenças de posição entre os ($ \alpha $,$ \delta $) observados reduzidos com UCAC2, UCAC4 e 2MASS e as coordenadas de referência fornecidas previamente pelo usuário em arquivo (efeméride, catálogo, etc.), chamados de offsets. Um esquema ilustrativo do PRAIA é mostrado na Fig \ref{Fig:PRAIA}.
%
%\begin{figure}[h]
%\centering
%\includegraphics[scale=0.4]{figuras/PRAIA.png}
%\caption[Esquema ilustrativo do pacote PRAIA]{Esquema ilustrativo do pacote PRAIA: a sequência a ser seguida pelo PRAIA está no sentido horário, desde a extração dos dados das imagens, até a identificação dos alvos nos campos CCD e, consequentemente, a obtenção de resultados.}
%\label{Fig:PRAIA}
%\end{figure}
%
%\chapter{Ocultações Estelares}
%\label{Cap: ocultacoes}
%
%\indent \indent Ocultação estelar é um evento que ocorre quando um objeto do Sistema Solar passa na frente de uma estrela de forma que o brilho da estrela é completamente ou parcialmente bloqueado durante um intervalo de tempo. A variação de brilho da estrela pode ser medida fazendo-se observações com exposições curtas. O produto gerado é uma curva de fluxo de luz x tempo. Essa técnica pode ser utilizada para se obter parâmetros físicos dos objetos ocultantes como tamanho, forma, albedo, densidade (se a massa é conhecida), detecção da presença de atmosfera, satélites, anéis, etc \citep{Elliot1979}.
%
%A Fig. \ref{Fig: occ-scheme} mostra uma curva de luz esquemática de uma ocultação estelar central por um planeta com atmosfera e anéis. Ao ocultar a estrela, os anéis passarão primeiro, a densidade do anel, tamanho e tempo de exposição determinará a queda de brilho da estrela. Ao chegar no corpo principal, caso não possua atmosfera, a queda de brilho será abrupta, caso possua, a queda de brilho será mais suave devido à refração da luz da estrela na atmosfera dificultando a determinação da superfície do objeto.
%
%\begin{figure}[h]
%\begin{centering}
%\includegraphics[scale=0.75]{figuras/stellarocc.png} 
%\caption{Curva de luz esquemática de uma ocultação estelar por um planeta com atmosfera e anéis. Fonte: \cite{Elliot1979}}
%\label{Fig: occ-scheme}
%\end{centering} 
%\end{figure}
%
%A suavidade da curva portanto será determinada pelo gradiente de temperatura, pressão e densidade da atmosfera, presença de nuvens, etc. A luz da estrela nunca será completamente apagada quando ocultada por um objeto com atmosfera devido à refração da luz na atmosfera onde parte dela convergirá na direção do observador. Caso a ocultação seja central para o observador haverá um aumento de brilho no meio do evento ("flash") pois a convergência da luz será máxima. A atmosfera, no caso, agiria como uma lente como mostrado na Fig. \ref{Fig: occ-scheme}.
%
%Quando o evento é observado por diversos telescópios espalhados, cada observador verá a ocultação em latitudes diferentes. Essas diversas curvas de luz permitirão reconstruir a forma do objeto, possivelmente identificando estruturas superficiais como montanhas e crateras. Como a estrela está muito distante, podendo ser considerada quase pontual, a sombra do objeto durante a ocultação é muito próxima ao seu tamanho real.
%
%Para se observar esses eventos, portanto, é necessário uma campanha grande com observadores espalhados por várias regiões para cobrir a sombra do objeto. A precisão nas posições da estrela e do objeto são cruciais pois o evento ocorre em instantes específicos e passará por regiões específicas da Terra alheias à escolha do observador.
%
%As curvas de luz são geradas a partir das observações seguindo a técnica da fotometria diferencial. Em nosso trabalho usamos as tasks de fotometria diferencial do pacote PRAIA. As imagens são previamente calibradas de bias e "flatfield". Os alvos e objetos de calibração são marcados em uma imagem de referência (normalmente a primeira da série) e depois automaticamente identificados no resto da sequência. Técnicas de cálculo do fundo de céu são empregadas, envolvendo a análise da moda de histogramas, e o fluxo dos objetos e do alvo são calculados. Diversas aberturas são testadas até que a menor dispersão seja obtida para a curva (melhor relação sinal-ruído). Ao final, obtêm-se as razões de fluxo alvo/calibrador(es), sinal-ruído, erros da razão de fluxo, entre outras informações. O instante de tempo a cada ponto da curva também é registrado.
%
%Nas seções \ref{Sec: Ceres-2010-occ} e \ref{Sec: Ceres-2013-occ} detalhamos os procedimentos da análise da curva para obtenção dos instantes de imersão e emersão, para o cálculo da dimensão e forma do alvo, no caso de ausência de atmosfera. Na presença de atmosfera, técnicas de "ray-tracing" são utilizadas para ajustar a altura e perfis de densidade/temperatura a partir da curva de luz obtida. Maiores detalhes desta técnica são dadas em \cite{DiasOliveira2015} e referências.

\chapter{Centauros e TNOs}
\label{Cap: TNOs}

\section{Introdução}
\label{Sec: TNOs-intro}

\indent \indent Dentro do contexto explicitado no capítulo \ref{Cap: intro}, os TNOs são uns dos objetos mais significantes por possivelmente serem a fonte de outras populações do Sistema Solar, como por exemplo Centauros \citep{Fernandez2002}, Satélites Irregulares \citep{Nesvorny2007} e Troianos \citep{Morbidelli2005}. Também existe a possibilidade de que Ceres tenha uma origem como TNO \citep{McKinnon2012}, assim como Tritão \citep{Agnor2006}.

Pouco se conhece sobre Centauros e TNOs, pois nunca foram visitados por sondas, com a exceção de Plutão, e a grande maioria das observações são de origem fotométrica usual (magnitudes, cores) ou, em menor parte, espectroscópica, ou em menor número ainda por telescópios espaciais.

Estimar parâmetros físicos para esses objetos é um desafio, porém essencial para acessar a massa atual e composição (material) da população e recuperar sua história evolutiva. É possível obter seus tamanhos através de modelos, por exemplo combinando o brilho no visível e emissão térmica obtida em infra-vermelho. Todavia os erros obtidos são piores que 10 \% \citep{Camargo2013}.

Para obter parâmetros como tamanho, albedo, densidade, etc, de forma muito mais precisa e acurada, sem a necessidade de se adotar modelos, utilizamos a técnica de ocultações estelares. Até 2009, quando foi observada uma ocultação de 2002TX300 \citep{Elliot2010}, Plutão e Caronte eram os únicos TNOs com ocultações observadas. Até hoje, cerca de uma dezena de TNOs e Centauros tiveram ocultações observadas, alguns com apenas uma corda, como Varuna, o que permite apenas determinar um tamanho mínimo para o objeto.

Com o objetivo de prever e observar ocultações estelares por TNOs, o nosso grupo conta com colaboração de pesquisadores nacionais e internacionais, profissionais e amadores. Utilizamos telescópios de pequeno e grande porte e câmera rápida para observar eventos que proporcionaram grandes descobertas, como por exemplo a descoberta do sistema de anéis ao redor do Centauro Chariklo \citep{BragaRibas2014}.

\section{Astrometria}
\label{Sec: TNOs-astrometria}

\indent \indent Observar uma ocultação exige um grande trabalho de predição e de sua melhoria. Os TNOs estão muito distantes significando que mesmo sem grandes erros nas suas efemérides ou na posição da estrela o erro do local por onde a sombra irá passar pode ser de milhares de quilômetros. Preferencialmente, esse erro tem que ser menor que o tamanho angular do objeto no plano do céu. Por exemplo, Plutão tem um raio de $1190 \pm 5$ km \cite{DiasOliveira2015} que a uma distância de aproximadamente 32 UA representa um diâmetro angular de 102 mas. Para a maioria dos objetos selecionados, menores que Plutão e/ou mais distantes, seus tamanhos angulares são menores que 50 mas.

Da mesma forma, estando muito distantes, eles se movem relativamente devagar no céu ocultando relativamente poucas estrelas por ano. Porém, sendo menos brilhantes (Mag $V > 18$) é possível observar ocultações de estrelas mais fracas (estrelas até Mag $R = 17$), onde a queda de brilho da estrela durante a ocultação poderá ser detectada e, mesmo com baixo sinal-ruído (ex: 5 a 10), o tamanho do corpo pode ser calculado com precisão de pelo menos 10 a 20 km, melhorando à medida em que a velocidade da sombra seja menor que 20 km/s (velocidade típica para um corpo em oposição a 30 UA).

Para identificar candidatas a ocultações estelares foi realizada uma campanha observacional no ESO para observar os caminhos aparentes dos TNOs entre 2009 e 2015. Assim, estrelas que não estão no UCAC4, catálogo comumente utilizado como referência astrométrica, foram identificadas bem como estrelas que estão mas não tem posições ou movimentos próprios com precisão suficiente. Tanto os catálogos de estrelas quanto as estrelas candidatas a ocultação foram publicados \citep{Assafin2010, Assafin2012, Camargo2013}. Esses catálogos são utilizados como representante prático do ICRS no tratamento astrométrico e o conjunto é denominado catálogo WFI.

Não somente as posições das estrelas precisam ser atualizadas, mas também as posições dos TNOs. Portanto os TNOs também foram observados tanto no ESO quanto no Observatório do Pico dos Dias (OPD, LNA/MCTI, Itajuba, MG., código IAU 874, $45^{\circ} ~34\arcmin ~57\arcsec$ W, $22^{\circ} ~32\arcmin ~04\arcsec$ S, 1864 m). A princípio as predições das ocultações eram atualizadas utilizando o método de offsets, porém, em colaboração com o Dr. Josselin Desmars, integrações numéricas das órbitas dos TNOs foram realizadas desenvolvendo-se uma ferramenta denominada NIMA\footnote{Numerical Integration of the Motion of an Asteroid} \citep[aceito]{Desmars2015}.

%Nesse trabalho, posições obtidas de ocultações já realizadas tem um peso muito maior que as restantes, diferentemente do que é feito pelo JPL. Da mesma forma, noites em que houveram um número maior de observações que as restantes terão um peso moderado já que se fosse utilizado um peso individual, o peso final de uma noite poderia estar superestimado levando a um erro na integração. Neste trabalho contribuí com observações e reduções astrométricas de imagens de TNOs feitas no OPD.

\section{Ocultações Estelares}
\label{Sec: TNO-occ}

\indent \indent Com o objetivo de obter os parâmetros físicos desses objetos, para melhor caracterizar as propriedades básicas dessas populações, uma grande campanha observacional é realizada para se observar ocultações estelares por TNOs e Centauros. Essa campanha é liderada por pesquisadores brasileiros, franceses e espanhóis e tem colaboração com observadores espalhados por todo o mundo.

Os objetos foram basicamente selecionados de acordo com o grau de interesse científico (representar uma classe de TNO, ser um  objeto peculiar, etc), tamanho (quanto maior, mais chances de observar a ocultação) e região do céu (quanto mais estrelas, maior  a probabilidade de ocultar estrelas brilhantes, permitindo curvas de luz de melhor resolução espacial).

Vários objetos já foram observados pelo grupo, sendo que o Sistema Plutão-Caronte conta com várias ocultações observadas além de um trabalho astrométrico minucioso baseado em 19 anos de observações no OPD, mais observações no ESO \citep{Rossi2014}. As observações bem sucedidas de ocultações estelares resultantes desse trabalho de predição já geraram diversas publicações, inclusive na revista Nature, por exemplo \cite{Sicardy2011} e \cite{BragaRibas2014} nas quais contribuí em caráter observacional. A ocultação de Plutão de 04 de Maio de 2013, a qual observei, gerou duas publicações \citep{Olkin2015, DiasOliveira2015}.

\cite{Sicardy2011} identificaram, a partir de uma ocultação de 06 de Novembro de 2010, que Eris é um planeta anão menor do que se esperava. Devido a seu brilho e distância acreditava-se que ele fosse muito maior, mas na verdade ele tem um albedo geométrico no visível de $p_v=0.96^{+0.09}_{-0.04}$ e seu tamanho é de $1163 \pm 6$ km, muito próximo ao tamanho de Plutão. O alto albedo de Eris pode ser relacionado à presença de uma atmosfera colapsada devido ao fato dele estar se aproximando de seu afélio.

\cite{BragaRibas2014} observaram uma ocultação do Centauro Chariklo que ocorreu dia 03 de Junho de 2013. O ocultação revelou que o objeto possui um sistema de anéis nunca visto antes em um objeto tão pequeno. Até o momento, conhecia-se apenas anéis em torno dos planetas gigantes. Essa descoberta levanta muitas questões sobre formação dos anéis, estabilidade e tempo de vida e nos leva a crer que pode haver mais objetos com anéis no Sistema Solar, ao menos nessa região.

\cite{Olkin2015} determinaram a partir de observações de ocultações por Plutão que a atmosfera do objeto não colapsa durante os 248 anos de sua órbita, como acontece com Eris. Modelos anteriores previam um colapso de sua atmosfera devido a Plutão receber três vezes menos luz do Sol em afélio que em seu periélio.

\cite{DiasOliveira2015}, também a partir de ocultações de Plutão, utilizaram modelos para obter os perfis de temperatura, pressão e densidade da atmosfera do objeto. Os modelos ajustam perfeitamente os perfis de temperatura para as duas ocultações utilizadas assumindo uma atmosfera esfericamente simétrica. Por fim, foi obtido que a pressão da atmosfera aumentou cerca de 6\% entre 2012 e 2013 significando que a atmosfera de Plutão ainda está se expandindo e confirmando o trabalho de \cite{Olkin2015}.

No dia 29 de Junho de 2015 houve uma ocultação de Plutão de máxima importância devido a esta ser a última antes da chegada da sonda \textit{New Horizons} no sistema e foi observada na Austrália e Nova Zelândia. Os modelos de perfil de atmosfera para Plutão poderão ser calibrados com os dados da \textit{New Horizons} para serem utilizados em futuras ocultações e estudar a evolução da atmosfera do planeta anão.

Em 29 de Abril de 2014, uma ocultação de Chariklo foi observada na África. Eu trabalhei com a redução de três curvas de luz (Gifberb, SAAO e Springbok). Um fato interessante dessa ocultação foi que a estrela era uma binária não catalogada. Essa ocultação foi importante para obter com total certeza a direção do polo de Chariklo. Além disso, as câmeras utilizadas eram de alta qualidade possibilitando obter que pelo menos um dos anéis é opaco. Uma das cordas, fortuitamente, passou contígua ao anel observando apenas o anel externo. Os resultados ainda não foram publicados.

Outro evento que observei foi uma ocultação do TNO 2003VS2 dia 07 de Novembro de 2014 no OPD. As variações do céu não permitiram uma boa observação de forma que a curva de luz estava muito ruidosa, mas boa o suficiente próximo ao instante do evento. Um trabalho foi feito para identificar alguma ocultação mas nada pôde ser extraído. Nesse evento foram obtidas 3 cordas positivas e 7 negativas. A redução preliminar revelou que 2003VS2 possui um corpo levemente alongado com aproximadamente 600 km de diâmetro. O trabalho ainda está em análise.

\chapter{Ceres}
\label{Cap: Ceres}

\section{Introdução}
\label{Sec: Ceres-intro}

\indent \indent Apesar de Ceres não ser um objeto do Sistema Solar exterior, ele é o único planeta-anão no Sistema Solar interno e, por isso, é um objeto de grande importância e seu estudo pode ter grande impacto na formação e evolução do Sistema Solar. Na verdade, foi proposto que a origem de Ceres pode ser comum aos objetos trans-Netunianos \citep{McKinnon2012}, espalhado posteriormente para o cinturão principal de asteroides devido à migração dos planetas gigantes predito pelo Modelo de Nice \citep{Gomes2005}.

Contendo aproximadamente um quinto de toda a massa do cinturão de asteroides, espera-se que Ceres esteja em equilíbrio gravitacional e seja, portanto, um elipsoide Maclaurin ou Jacobi. De fato, observações diretas de Ceres com a utilização de ótica adaptativa indica que ele é um esferoide achatado nos polos \citep{Drummond2014}. O conhecimento preciso de seu tamanho e forma é de extrema importância para modelos de densidade, estrutura interna e diferenciação.

%A primeira ocultação estelar por Ceres foi observada em 1984 \citep{Millis1987} e determinou seu tamanho com precisão de alguns quilômetros em uma época que as incertezas eram, normalmente, dez vezes maiores. Devido ao brilho aparente de Ceres ser alto, comparado à maioria dos asteroides, estrelas capazes de causar uma queda de magnitude detectável quando ocultadas são limitadas às mais brilhantes. Por exemplo, depois do evento de 1984, apenas 4 ocultações estelares por Ceres foram observadas \citep{Dunham2014}. Duas delas tiveram apenas duas cordas cada que não foram suficientes para prover resultados acurados\footnote{Esses eventos ocorreram em 22 de Agosto de 1994 e 30 de Outubro de 2010.}. Os dois restantes, que ocorreram em 17 de Agosto de 2010 e 25 de Outubro de 2013, foram trabalhos por mim em colaboração com o grupo do Rio \citep{GomesJunior2015-Ceres}.

Foram preditos dois eventos por Steve Preston\footnote{Predições publicadas em \url{http://asteroidoccultation.com}.} para a IOTA (International Occultation Timing Association), durante predições de rotina de ocultações de estrelas brilhantes por asteroides. Os caminhos das sombras podem ser visualizados na Fig. \ref{Fig: Ceres-map}.

\begin{figure}[h]
\begin{centering}
\subfigure[Ocultação de 2010]{\includegraphics[scale=0.39]{figuras/Ceres_2010.eps}\label{Fig: Ceres-map-2010}}
\subfigure[Ocultação de 2013]{\includegraphics[scale=0.39]{figuras/Ceres_2013-zoom.eps}\label{Fig: Ceres-map-2013}}
\caption{Reconstrução pós-ocultação do caminho da sombra de Ceres na Terra para os eventos de 17 de Agosto de 2010 (a) e 25 de Outubro de 2013 (b). Os pontos em azul são os sítios que observaram os eventos. a) O ponto grande vermelho é a máxima aproximação geocêntrica às 22:40:25 UT. Os pequenos representam o centro da sombra separados por um minuto. b) Visão superior da ocultação sobre os sítios que observaram o evento de 25 de Outubro de 2013. Os pontos vermelhos são os centros da sombra separados por 15 s.
\label{Fig: Ceres-map}}
\end{centering}
\end{figure}

\section{Ocultaç\~oes de 2010 e 2013}
\label{Sec: Ceres-2010-occ}

\indent \indent Em 17 de Agosto de 2010 Ceres ocultou a estrela TYC 6833-163-1 (UCAC4 313-111823), cuja magnitude é $V = 11.55$ e tem posição no ICRS para a data do evento baseada no catálogo UCAC4 \citep{Zacharias2013}: $\alpha = 17^{h}18^{m}29\fs 0085$  e $\delta = -27\degr 26\arcmin 38\farcs 867$. E em 25 de Outubro de 2013, Ceres ocultou a estrela TYC 865-911-1 (UCAC4 496-058191), de magnitude $V = 10.05$. Baseada no UCAC4 \citep{Zacharias2013}, sua posição ICRS para a data da ocultação é: $\alpha = 11^{h}57^{m}52\fs7641$ e $\delta = +09\degr 07\arcmin 49\farcs835$.

%\begin{equation}
%\left\{ 
%  \begin{array}{l l}
%    \alpha = 17^{h}18^{m}29\fs 0085\\
%    \delta = -27\degr 26\arcmin 38\farcs 867
%  \end{array}
%\right.
%\label{Eq: 2010-coord}
%\end{equation}

\indent O primeiro evento foi observado no Brasil a partir de cinco diferentes sítios (ver Fig. \ref{Fig: Ceres-map-2010}). Destes, 4 obtiveram cordas positivas enquanto UFSC teve uma corda negativa. Das positivas, a observação proveniente do INPE iniciou-se após o início do evento devido a dificuldades técnicas e, portanto, apenas a emersão da curva de luz foi detectada. O segundo foi observado na costa Leste dos Estados Unidos (\ref{Fig: Ceres-map-2013}).

Uma das características mais importantes desse evento foi a velocidade com que ocorreu (apenas 3.9 km s$^{-1}$) acarretando que mesmo exposições de poucos segundos representariam resoluções espaciais significantes.

Todas as observações foram feitas com a utilização de CCDs. As curvas de luz de cada observação foram obtidas das imagens FITS com a utilização do pacote PRAIA \citep[Plataforma de Redução Astrométrica de Imagens Astronômicas,][]{Assafin2011}. As curvas foram normalizadas para o fluxo da estrela mais Ceres, uma vez que eles estavam indistinguíveis logo antes e depois da ocultação. Por fim, elas foram normalizadas pelo ajuste de uma curva polinomial (de primeira ou segunda ordem) fora da queda de fluxo assim fixando em 1 a razão de fluxo fora da ocultação.

Os instantes de ingresso e egresso foram obtidas de cada curva de luz ajustando-se um modelo de poço quadrado levando em consideração a difração de Fresnel, a banda do CCD, o diâmetro aparente da estrela e o tempo de exposição utilizado \citep[ver][]{Widemann2009, Braga-Ribas2013}. Algumas cordas do evento de 2013 não foram utilizadas por haver erros nos tempos fornecidos identificados durante o tratamento das imagens.

O ajuste dos dados da ocultação consiste em minimizar uma função de $\chi^{2}$ clássica para cada curva de luz, como descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}. Os parâmetros livres para ajustar são os instantes de ingresso e egresso que fornece o valor mínimo de $\chi^{2}$ ($\chi^{2}_{min}$). %O melhor ajuste das curvas de luz para as ocultações está mostrado na Fig. \ref{Fig: Ceres-curves}.

%\begin{figure}
%\begin{centering}
%\subfigure[Ocultação de 2010]{\includegraphics[scale=0.51]{figuras/Ceres_2010_fluxratio.eps}\label{Fig: Ceres-2010-curves} }
%\subfigure[Ocultação de 2013]{\includegraphics[scale=0.51]{figuras/Ceres_2013_fluxratio.eps}\label{Fig: Ceres-2013-curves} }
%\caption{Curvas de luz normalizadas das cordas positivas dos eventos. As curvas estão desviadas por um fator de 0.2 (a) e 1.0 (b) para melhor visualização. As linhas pretas são os melhores ajustes com o modelo de poço quadrado. As linhas vermelhas são os melhores ajustes com o modelo de poço quadrado, porém levando em conta a difração de Fresnel, o diâmetro da estrela e o tempo de exposição. Os instantes médios de cada curva não coincidem devido às diferentes longitudes dos sítios. A curva de luz de Brookline (b) está desviada por um fator de -64 s como explicado no texto.
%\label{Fig: Ceres-curves}}
%\end{centering}
%\end{figure}

A metodologia usada para analisar o perfil de Ceres a partir das observações é o mesmo descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}. %Cada combinação de posição do sítio, instantes de ingresso e egresso, junto com as coordenadas da estrela e as efemérides de Ceres, correspondem a um ponto no plano do céu. A coleção de todos esses pontos determina o limbo aparente de Ceres.
%Adotamos um modelo elíptico para o perfil do limbo, resultante da projeção de um esferoide com achatamento nos polos no plano do céu. Essa escolha é suportada pelo trabalho de \cite{Drummond2014}, por meio de imagem direta de Ceres. Dessa forma, nós temos $N=7$ extremidades das cordas para ajustar $M=5$ parâmetros que definem uma elipse: semi-eixo maior e semi-eixo menor aparentes ($a^\prime$ and $b^\prime$, respectivamente), ângulo de posição $P$ do seu semi-eixo maior e as posição $(f_c,g_c)$ do seu centro com respeito à estrela ocultada. O semi-eixo maior {$a^\prime$} é equivalente ao raio equatorial $R_{equa}$ do elipsoide.
%As coordenadas $f_{c}$ e $g_{c}$, em quilômetros, foram calculadas usando a efeméride de Ceres JPL\#33 \citep{Giorgini1996} e a posição da estrela ocultada. Elas são positivas na direção Leste e Norte celestes, respectivamente. O ângulo de posição $P$ é contado positivamente a partir do norte celeste local em direção ao leste celeste. O achatamento aparente pode ser definido por $\epsilon^\prime = 1 - (b^\prime/a^\prime)$. O melhor ajuste é obtido minimizando uma função de $\chi^{2}_{r}$ reduzido, onde definimos o número de graus de liberdade do problema como $\mathcal{N} \equiv N - M$. 
Todos os procedimentos que permitem a determinação das barras de erro dos parâmetros físicos podem ser encontradas em \cite{Braga-Ribas2013}.

Duas possíveis soluções foram consideradas para o ajuste do limbo. A primeira, que chamamos de solução nominal, consiste em determinar os cinco parâmetros que caracterizam uma elipse a partir dos contatos observados. A segunda consiste em calcular o ângulo de posição $P$ a partir das coordenadas do polo de Ceres obtidas por \cite{Drummond2014} ($\alpha_{p} = (287 \pm 3) \degr$, $\delta_{p} = (+64 \pm 3) \degr$ no ICRS) e da efeméride de Ceres no instante da ocultação. Chamamos de solução de polo fixo.

%Para o evento de 2010, a solução nominal teve como melhor ajuste $\chi^2_{r,min} = 0.24$, que podem ser interpretadas como as barras de erro estarem superestimadas com respeito à boa qualidade do ajuste. Porém, como o problema tem somente dois graus de liberdade, $\chi^2_{r,min}$ relativamente pequenos são aceitáveis.
Os resultados obtidos para o diâmetro equatorial, achatamento, ângulo de posição e coordenadas do centro são apresentadas na segunda coluna da tabela \ref{Tab: resultados}. As soluções para cada evento estão esquematizadas na Fig. \ref{Fig: Ceres-limb}

\begin{table}[h]
 \begin{centering}
  \caption{Resultados do ajuste de limbo de Ceres com os dados dos eventos de 2010 e 2013.\label{Tab: resultados}}
  \begin{tabular}{@{}lcccc}
  \hline
     Solução & 2010/Nominal & \textbf{2010/Polo fixo} & 2013/Nominal & 2013/Polo fixo \\
\hline
Diam. equat.  (km) & 982 $\pm$ 14 & \textbf{972 $\pm$ 6}  & 971 $\pm$ 7  & 971 $\pm$ 7\\
%Equatorial radius (km) & 491 $\pm$ 7  & \textbf{486 $\pm$ 3}  & 485.5 $\pm$ 3.5 & 485.5 $\pm$ 3.5\\
Achatamento        & 0.08 $\pm$ 0.03 & \textbf{0.08 $\pm$ 0.03} & 0.08 $\pm$ 0.04 & 0.08 $\pm$ 0.04\\
Âng. de pos. (deg)   & 5 $\pm$ 10    & \textbf{12 $\pm$ 3} (*)& 22 $\pm$ 5    & 25 $\pm$ 3 (*)\\
$f_c$ (km)             & 97 $\pm$ 9   & \textbf{102 $\pm$ 5}   & 77 $\pm$ 6    & 78 $\pm$ 6\\
$g_c$ (km)             & 16 $\pm$ 15  & \textbf{21 $\pm$ 11}  & 13 $\pm$ 16   & 13 $\pm$ 16\\
$\chi^2_{r,min}$       & 0.24          &  \textbf{0.42}         & 1.27          & 1.27\\
\hline
\end{tabular}
\textbf{Notas:} Em negrito destaca-se a melhor solução obtida. As barras de erro estão no nível de 1$\sigma$. O diâmetro polar ($D_{pol}$) pode ser facilmente calculado a partir de $D_{pol}=D_{equa}(1 - \epsilon)$. (*) Ângulos de posição derivados a partir das coordenadas do polo de Ceres determinadas por \cite{Drummond2014}.
\end{centering}
\end{table}

%Como pode ser visto, o parâmetro com a maior incerteza é o ângulo de posição cobrindo um intervalo de 20$\degr$. Claramente, fixar as coordenadas do pólo pode melhorar a solução. Por fim, a correção do achatamento devido ao ângulo do aspecto polar está dentro da barra de erro 1$\sigma$ e não tem relevância estatística, dessa forma $\epsilon = 0.08 \pm 0.03$.

%No momento da ocultação, as coordenadas do polo de Ceres correspondiam a um ângulo de posição $P = (12 \pm 3)\degr$. Explorar o espaço de parâmetros restringindo a elipses com ângulo de posição dentro deste intervalo resulta na solução de pólo fixo. O parâmetros físicos do melhor ajuste estão mostrados na Tab. \ref{Tab: resultados} enquanto a solução está esquematizada na Fig. \ref{Fig: Ceres-limb-2010}.

%Essa solução corresponde ao limite superior da barra de erro 1$\sigma$ da solução nominal para $P$. Por outro lado, ela obtém os menores valores para o diâmetro equatorial, melhorando sua determinação por um fator de 2.

\begin{figure}
\begin{centering}
\subfigure[Ocultação de 2010]{\includegraphics[scale=0.35]{figuras/Ceres_2010_sphere.eps}\label{Fig: Ceres-limb-2010} }
\subfigure[Ocultação de 2013]{\includegraphics[scale=0.35]{figuras/Ceres_2013_sphere.pdf}\label{Fig: Ceres-limb-2013} }
\caption{Melhores ajustes elípticos para as cordas das ocultações de 2010 e 2013. As setas indicam a direção de movimento, as linhas azuis são as cordas observadas, os segmentos vermelhos são as barras de erro dos ingressos, egressos e centro da ocultação em $1\sigma$. A linha verde em (a) é uma corda negativa. Os instantes marcados em verde em (b) não foram utilizados para o ajuste, como descrito no texto.
\label{Fig: Ceres-limb}}
\end{centering}
\end{figure}

%\section{Ocultação de 2013}
%\label{Sec: Ceres-2013-occ}

%\indent \indent Em 25 de Outubro de 2013, Ceres ocultou a estrela TYC 865-911-1 (UCAC4 496-058191), de magnitude $V = 10.05$. Baseada no UCAC4 \citep{Zacharias2013}, sua posição ICRS para a data da ocultação é:
%
%%
%\begin{equation}
%\left\{ 
%  \begin{array}{l l}
%    \alpha = 11^{h}57^{m}52\fs7641\\
%    \delta = +09\degr 07\arcmin 49\farcs835
%  \end{array}
%\right.
%\end{equation}
%%
%Esse evento foi observado na costa Leste dos Estados Unidos logo antes do amanhecer, como mostrado na Fig. \ref{Fig: Ceres-map-2013}.
%
%Nove cordas positivas foram obtidas em vários sítios (ver Fig. \ref{Fig: Ceres-map-2013}). Cada estação foi equipada com uma câmera de vídeo com tempo de leitura desprezível. Isso é particularmente importante, já que a velocidade da sombra de Ceres para esse evento foi de 42.6 km s$^{-1}$, muito mais rápido que o evento de 2010.
%
%Durante o evento Ceres estava muito baixo no céu com alturas entre $15\degr$ (Winchester) e $20\degr$ (Hampton). Forte cintilação era esperada e, combinada com o curto tempo de integração e baixa diminuição de brilho, resultou em curvas de luz ruidosas e assim grandes incertezas nos instantes de imersão e emersão.
%
%Todos os vídeos foram convertidos para imagens FITS e a fotometria foi obtida via \textsc{praia} \cite{Assafin2011}. As curvas de luz foram normalizadas por uma estrela de referência quando havia uma estrela no campo.
%
%Para reduzir o ruído, os dados foram binados por grupos de cinco imagens -- com exceção de Greenbelt, onde grupos de dez imagens foram utilizadas. Esse procedimento aumenta o tempo de integração efetivo reportado por um fator de 5 (ou 10). Da mesma forma que para o evento de 2010, uma normalização adicional por um polinômio foi aplicada.
%
%Os instantes de ingresso e egresso da ocultação foram obtidos pelo mesmo procedimento descrito na seção \ref{Sec: Ceres-2010-occ}. Uma vez que o tempo de integração efetivo usado (0.17 s) representa aproximadamente 7 km no plano do céu, e a escala de Fresnel e  o diâmetro da estrela estão novamente na ordem de centenas de metros, o erro da determinação dos instantes de imersão e emersão são dominados principalmente pelo tempo de integração, da mesma forma que para o evento de 2010. Os melhores ajustes para as curvas de luz da ocultação estão mostrados na Fig. \ref{Fig: Ceres-2013-curves}.
%
%Uma comparação entre os tempos obtidos pela estação de Brookline e o restante mostrou que Brookline têm um atraso de aproximadamente 64 s. Dessa forma, não utilizamos os tempos de Brookline na análise.
%
%Perfis elípticos foram ajustados para todas as cordas restantes. pelo mesmo procedimento descrito na seção \ref{Sec: Ceres-2010-occ}. O resultado foi $\chi^2_{r,min} = 13$ , sugerindo que um modelo elíptico não é satisfatório para os dados. De fato, ao olharmos para a Fig. \ref{Fig: Ceres-limb-2013} vemos que a corda de Varina adiantada com respeito às outras. Como nessa estação o tempo não foi inserido diretamente nos frames do vídeo é possível que essa diferença seja oriunda de um eventual problema da correspondência entre os tempos do camcorder e do GPS.
%
%A imersão gravada em Owings também parece atrasada com respeito às cordas próximas (ver Fig. \ref{Fig: Ceres-limb-2013}). Essa corda tem aproximadamente o mesmo tamanho da corda de Mechanicsville, apesar de estarem separadas por cerca de 100 km. Diferentemente de Varina, essa estação teve os tempos inseridos em cada frame do vídeo o que torna mais difícil justificar um problema de tempo. Outras possibilidades seriam uma má determinação dos instantes de ingresso e egresso dessa curva ou uma característica do relevo de Ceres.
%
%Em um segundo ajuste, não consideramos as cordas de Brookline, Varina e Owings. O ajuste dos cinco parâmetros que definem uma elipse para os doze contatos resultou em $\chi^2_{r,min} = 1.27$, indicando  que está em bom acordo com os dados observados dentro das barras de erro. Essa é a solução mostrada na Fig. \ref{Fig: Ceres-limb-2013}, onde podemos ver que o tamanho da corda de Brookline é compatível com o modelo.
%
%Os resultados das soluções nominais e de polo fixo não são significantemente diferentes. Isso se deve ao fato da barra de erro do ângulo de posição da solução nominal (que é muito menor que o da solução nominal de 2010) ser muito similar à barra de erro do ângulo de posição da solução de polo fixo ($P = (25 \pm 3)\degr$). Os dois resultados estão apresentados nas colunas 4 e 5 da tabela \ref{Tab: resultados}.
%
%Sobre a hipótese do atraso observado na imersão da curva de luz de Owings ser associado a uma característica topográfica, o contato gravado corresponderia a uma elevação negativa de 31 $\pm$ 4 km com respeito à elipse de melhor ajuste. Porém modelos teóricos preveem que relevos em Ceres não devem ser maiores que 10--20 km \citep{Johnson1973}, enquanto dados observacionais limitam o contorno em 18 km \citep{Carry2008}. Imagens mais recentes da sonda \textit{Dawn} também revelam uma superfície mais suave. Portanto, a associação da imersão em Owings com um relevo é improvável.

\section{Discussão}
\label{Sec: Ceres-discussao}

\indent \indent Os resultados apresentados na tabela \ref{Tab: resultados} mostram um acordo entre os parâmetros físicos obtidos nas duas ocultações, especialmente no diâmetro equatorial. As diferenças ocorrem basicamente nos tamanhos das barras de erro e podem ser justificadas pelas particularidades de cada conjunto de dados.

%O evento de 2010, por exemplo, teve somente sete contatos, porém bem distribuídos sobre o disco de Ceres. Por outro lado, o evento de 2013 teve cinco contatos a mais, todavia concentrados em certas regiões do corpo. Em particular, a ausência de cordas próximas ao polo sul fez seu achatamento ser pior determinado para o evento de 2013 que para o evento de 2010. Mesmo nossa melhor medida de achatamento, $\epsilon=0.08 \pm 0.03$, tem alta incerteza se comparado com outros valores publicados na literatura, como mostra a tabela \ref{Tab: Ceres-final}.

Como foi mostrado, usar as coordenadas do polo de Ceres determinadas por \cite{Drummond2014} para limitar o ângulo de posição não foi um procedimento eficiente para o evento de 2013. Por outro lado, fixar o ângulo de posição para a ocultação de 2010 reduziu as barras de erro dos outros parâmetros (com exceção do achatamento). Por fim, esse procedimento resultou em um excelente acordo entre os raios equatoriais obtidos para ambos os eventos.

Uma comparação do diâmetro equatorial de Ceres medido por diferentes técnicas mostra um acordo entre os nossos resultados e aqueles obtidos por imageamente direto do Hubble Space Telescope (HST) \citep{Thomas2005}, do Keck Observatory e do ESO VLT \citep{Drummond2014}. O valor diferente e menor reportado por \cite{Carry2008} pode ser justificado pelo fato desse estudo não levar em conta o efeito de escurecimento de bordo.

%O evento de 1984 \citep{Millis1987} é a única outra ocultação com a qual podemos comparar nossos resultados. As medidas do diâmetro não estão de acordo dentro de um fator de 2$\sigma$. É difícil dizer com certeza as razões dessa divergência. Uma forma de clarificar o problema seria redeterminar os instantes de imersão e emersão das curvas de luz originais usando a mesma metodologia aplicada nesse trabalho. Infelizmente, não temos acesso aos dados da curva de luz original do evento de 1984.

A sonda da NASA \textit{Dawn} poderá elucidar muitas perguntas sobre Ceres e seus resultados serão para calibrar todas as técnicas usadas até agora para o estudo das propriedades físicas dos pequenos objetos do Sistema Solar, como as ocultações estelares.

O resultado deste trabalho está foi publicado em \cite{GomesJunior2015-Ceres}

%\begin{table}
%  \caption{Diâmetro equatorial e achatamento de Ceres \label{Tab: Ceres-final}}
%  \begin{centering}
%  \begin{tabular}{@{}lccc}
%  \hline
%     Diâmetro Equatorial (km) & Achatamento & Método & Ref. \\
%\hline
%$972 \pm 6$  & $0.08  \pm 0.03$  & Ocultação & 1\\
%$967 \pm 10$ & $0.078 \pm 0.015$ & Keck+VTL    & 2\\
%$959 \pm 5$  & $0.074 \pm 0.007$ & Keck        & 3\\
%$975 \pm 4$  & $0.067 \pm 0.005$ & HST         & 4\\
%$959 \pm 5$  & $0.05  \pm 0.01$  & Ocultação & 5\\
%\hline
%\end{tabular}
%\par\end{centering}
%\textbf{Referências.} 1: \cite{GomesJunior2015-Ceres}. 2: \cite{Drummond2014}. 3: \cite{Carry2008}. 4: \cite{Thomas2005}. 5: \cite{Millis1987}.
%\end{table}

\chapter{Satélites Irregulares dos planetas gigantes}
\label{Cap: Irregulares}

\section{Introdução}
\label{Sec: Irr-intro}

\indent \indent Os satélites irregulares dos planetas gigantes são menores que os regulares, possuindo órbitas mais excêntricas, inclinadas e distantes. Na maioria dos casos, essas órbitas são retrógradas. Devido a suas configurações orbitais, é amplamente aceito que estes objetos foram capturados nos estágios iniciais da formação do Sistema Solar \citep{Sheppard2003}.

Por serem pequenos, eles são pouco brilhantes e só foram descobertos no último século. Dentre os satélite irregulares dos planetas gigantes, poucos são aqueles que possuem medidas de seus parâmetros físicos. Apenas Himalia, Phoebe e Nereida foram observados por sondas, apesar de não serem medidas ideais, já que foram alvos de oportunidade. A sonda \textit{Cassini} observou Himalia em 2000 ao passar próximo a Júpiter e obteve o tamanho de Himalia com um erro médio de 10 km \citep{Porco2003}. Em 2004, a \textit{Cassini}, aproximando-se de Saturno, observou Phoebe em alta resolução obtendo um erro médio na medida de seu tamanho de 0.7 km \citep{Thomas2010}. Por fim, Nereida foi observado em 1989 pela sonda \textit{Voyager II} e seu tamanho foi obtido com um erro de 25 km \citep{Smith1989}. Para outros satélites irregulares de Júpiter, seus tamanhos foram estimados por \cite{Rettig2001} impondo um albedo de 4\% com a justificativa de que esse valor é um albedo nominalmente utilizado para objetos do Sistema Solar Externo.

%Existem alguns mecanismos de captura de objetos por planetas gigantes propostos na literatura. Há o Arrasto Gasoso na nebulosa circumplanetária primordial \citep{Sheppard2005} onde o movimento do objeto seria afetado pelo gás e sua velocidade diminuiria até que fosse capturado pelo planeta. Outro mecanismo se chama captura pull-down \citep{Sheppard2005}, onde a massa do planeta aumentaria enquanto o objeto estivesse temporariamente capturado.

%Um mecanismo baseado no modelo de Nice \citep{Morbidelli2005, Tsiganis2005, Gomes2005} foi proposto por \cite{Nesvorny2007} e, especificamente para Júpiter, por \cite{Nesvorny2014}. Durante a instabilidade do Sistema Solar primordial, ocorreram encontros entre os planetas externos. Nesses encontros planetários poderia haver troca de energia e momento angular entre os planetas e os objetos próximos tornando possível a captura de um corpo irregular por um planeta gigante. Nesse cenário, a taxa de sobrevivência de satélites anteriores ao LHB (Bombardeamento Pesado Tardio\footnote{"Late Heavy Bombardment", em inglês}) é muito pequena.

%Um outro mecanismo importante é a captura através de interações colisionais \citep{Sheppard2005}. Uma colisão entre dois corpos pequenos dentro da esfera de Hill do planeta poderia gerar objetos fragmentados e a energia dissipativa poderia ser tal que alguns desses objetos seriam capturados.

%Alguns desses objetos formam grupos dinâmicos com elementos orbitais semelhantes, chamados famílias, similar às famílias encontradas no Cinturão Principal de Asteroides. Essas famílias podem ter sido criadas por um corpo pai que se fragmentou por colisões com cometas ou outros asteroides \citep{Nesvorny2004}. Colisões com cometas tem uma probabilidade maior de ocorrer durante o Bombardeamento Pesado Tardio (LHB) \citep{Gomes2005}.

%\cite{Nesvorny2003} estudou as taxas de colisões entre satélites irregulares e concluiu que alguns satélites podem ter sido removidos por colisão com um satélite maior. A taxa de colisão entre os satélites do grupo de Himalia (Himalia, Elara, Lysithea and Leda, principalmente), por exemplo, foi encontrado como sendo maior que 1 durante a idade do Sistema Solar sugerindo que sua estrutura atual foi originada por colisão satélite-satélite.

%Para Phoebe, materiais ejetados de sua superfície causadas por impactos poderiam evoluir devido ao arrasto de Poynting-Robertson e colidir com Iapetus causando a grande variação de albedo observada no satélite \citep{Nesvorny2003}. De fato, a sonda \textit{Cassini} detectou em Phoebe uma característica de absorção em 2.42 $\mu m$, provavelmente combinações de CN, que foi também detectada no lado escuro de Iapetus \citep{Clark2005}

Se esses objetos foram capturados, permanece a pergunta de onde eles vieram. Estudos como os de \citealp{Grav2003, Clark2005, Grav2007} mostraram através de suas cores e inclinações espectrais que esses satélites contêm uma fração mais ou menos igual de objetos semelhantes a asteroides tipo C, P e D ou a Centauros e TNOs. Esses trabalhos sugerem origens diferentes para os satélites irregulares.

Nesse contexto, durante o Mestrado, foi realizado junto ao grupo um trabalho de caráter astrométrico para a melhoria das efemérides dos satélites irregulares de Júpiter e Saturno. Em colaboração com o Dr. Jean-Eudes Arlot do IMCCE, reduzi um banco de dados observado no Observatoire Haute-Provence (OHP) entre 1998 e 2008, contendo mais de 28 mil posições para 10 satélites. Reduzi também um banco de dados com mais de 100 mil imagens obtidas no Observatório do Pico dos Dias (OPD) entre 1992 e 2014. Já no Doutorado, neste mesmo âmbito, reduzi 810 observações feitas no European Southern Observatory (ESO) em 24 noites utilizando o detector mosaico CCD Wide Field Imager (WFI).

Um total de 6523 posições de 18 satélites foram identificadas. Esse trabalho está publicado \citep{GomesJunior2015-Irregular} e o catálogo com as posições está disponível no CDS.

Um dos principais resultados obtidos foi a grande quantidade de posições identificadas para os satélites irregulares em comparação com a quantidade utilizada nas integrações numéricas atuais publicadas por \cite{Jacobson2012}. Dessa forma, é de se esperar que nossas posições contribuam significativamente para a melhoria da órbita desses objetos.

%\begin{table}[h]
%\caption{Comparação entre o número de posições obtidas em \cite{GomesJunior2015-Irregular} e o número utilizado nas integrações numéricas das órbitas pelo JPL como publicado por  \cite{Jacobson2012}. Os satélites estão separados por planeta (linha reta) e por família orbital (linha tracejada).}
%\label{Tab: comparison-horizons}
%\begin{centering}
%\begin{tabular}{lccccccc}
%\hline  \hline
%Satélite & Diâm. (km)\footnote{Planetary Satellite Physical Parameters - JPL: \url{http://ssd.jpl.nasa.gov/?sat_phys_par}} & Mag V  & OPD  & OHP & ESO & Total  & Jacobson \tabularnewline
%\hline
%Himalia & 170 & 14 & 854 & 357 & 23 & 1234 & 1757 \tabularnewline
%Elara & 86 & 16 & 403 & 187 & 46 & 636 & 1115 \tabularnewline
%Lysithea & 36 & 18 & 60 & 84 & 90 & 234 & 431 \tabularnewline
%Leda & 20 & 19 & 6 & 48 & 44 & 98 & 178 \tabularnewline
%\hdashline
%Pasiphae & 60 & 17 & 295 & 248 & 66 & 609 & 1629 \tabularnewline
%Callirrhoe & 9 & 21 & 9 & -  &  16 & 25 & 95 \tabularnewline
%Megaclite & 5 & 22 & - & -  &  10 & 10 & 50  \tabularnewline
%\hdashline
%Ananke & 28 & 18 & 52 & 141 & 57 & 250 & 600 \tabularnewline
%Praxidike & 7 & 21 & - & -  &   2 & 2 & 59 \tabularnewline
%\hdashline
%Carme & 46 & 18 & 90 & 204 & 37 & 331 & 973 \tabularnewline
%Sinope & 38 & 18 & 41 & 169 & 11 & 221 & 854 \tabularnewline
%Themisto & 8 & 21 & - & - & 16 & 16 & 55 \tabularnewline
%\hline
%Phoebe & 213 & 16 & 1239 & 516 & 32 & 1787 & 3479 \tabularnewline
%\hdashline
%Siarnaq & 40 & 20 & - & 20 & 56 & 76 & 239 \tabularnewline
%Paaliaq & 22 & 21 & - & - & 11 & 11 & 82 \tabularnewline
%\hdashline
%Albiorix & 32 & 20 & - & - & 46 & 46 & 137 \tabularnewline
%\hline
%Sycorax & 150 & 21 & - & - & 35 & 35 & 237 \tabularnewline
%\hline
%Nereid & 340 & 19 & 803 & - & 99 & 902 & 716 \tabularnewline
%\hline
%\end{tabular}
%\par\end{centering}
%\end{table}

Observando os offsets das posições com respeito às efemérides ao longo do tempo e anomalia verdadeira foram identificados erros sistemáticos nas efemérides. O erro mais claro é na declinação do satélite Carme o qual foi associado a um erro em sua inclinação orbital.

%\begin{figure}[h]
%\begin{centering}
%\subfigure[Right Ascension]{\includegraphics[scale=0.45]{figuras/Carme_RA}\label{Fig: carme_alfa}}
%\subfigure[Declination]{\includegraphics[scale=0.45]{figuras/Carme_DEC}\label{Fig: carme_delta}}
%\caption{Offsets médios das efemérides e dispersões (barra de erro 1$\sigma$) nas coordenadas de Carme tomadas noite a noite por anomalia verdadeira para cada telescópio. O quadrado vermelho são as observações com o telescópio do OPD Perkin-Elmer, o círculo azul para o Boller \& Chivens também do OPD, o triângulo pra cima preto para o OHP e a estrela verde para o ESO.
%\label{Fig: carme_anom}}
%\end{centering}
%\end{figure}

Esse padrão em declinação também foi identificado para outros satélites como Pasiphae e Ananke. Para alguns satélites a cobertura orbital não foi suficiente para indicar claramente a presença de erros sistemáticos em elementos orbitais específicos. Os gráficos para todos os satélites estão disponíveis em \cite{GomesJunior2015-Irregular}.

%Um método que tem se mostrado eficiente para a obtenção dos parâmetros físicos (tamanho, forma, albedo, etc) desses objetos é o método de ocultações estelares, que proporciona medidas tão precisas que são apenas superadas por medidas oriundas de sondas. Uma vez que suas efemérides não são muito precisas dado os seus tamanhos estimados, predizer e observar ocultações estelares é muito difícil e nenhuma observação de um evento desses para um satélite irregular ainda foi relatado na literatura.

\section{Predição das ocultações}
\label{Sec: Irr-predic}

\indent \indent Com o objetivo de obter parâmetros físicos para os satélites irregulares com maior precisão foram feitas predições de ocultações estelares para os 7 maiores satélites de Júpiter (Himalia, Elara, Pasiphae, Sinope, Lysithea, Carme, Ananke), o satélite Phoebe de Saturno e Nereida de Netuno. Esses objetos são pequenos, se comparados aos TNOs estudados no capítulo \ref{Cap: TNOs}, sendo que o menor dos satélites da amostra, Ananke, possui um diâmetro estimado de 28 km.

As predições foram feitas utilizando as posições de estrelas dadas no catálogo UCAC4. Após o lançamento do catálogo GAIA novas predições serão feitas. Para as posições dos satélites de Júpiter foi feita uma nova integração numérica utilizando as posições publicadas em \cite{GomesJunior2015-Irregular}. Essa integração foi realizada pela pós-doc Laurène Beauvalet. Para Phoebe e Nereida foram utilizadas efemérides do JPL.

Um total de 588 eventos foram obtidos entre 01 de Janeiro de 2015 e 31 de Dezembro de 2017 para os 9 objetos sendo que a maioria será descartada por passar em regiões de oceano ou onde não há observadores disponíveis. Estima-se que restem ao final cerca de 10\% desse número. Para Nereida, apenas uma estrela UCAC4 foi identificada para ser ocultada. Nesse caso, utilizamos um catálogo de estrelas feito a partir do caminho aparente do satélite de Netuno, Tritão, no céu para predições de ocultações estelares pelo mesmo.

%Como esses objetos são muito pequenos, em sua maioria as ocultações durarão poucos segundos, portanto apenas eventos com estrelas brilhantes serão selecionados e se houver câmeras de integração rápida disponíveis. Por outro lado, os satélites de Júpiter estão muito mais perto que os TNOs e como o erro astrométrico é uma medida angular, consequentemente, o erro da sombra da ocultação projetada na Terra será muito menor que para TNOs. Assim, as latitudes a serem cobertas para que uma ocultação por satélite irregular de Júpiter seja detectada corresponde a poucas centenas de quilômetros.

%\section{Testes}
%\label{Sec: Irr-testes}

%\indent \indent Até que o primeiro catálogo GAIA seja lançado, as posições das estrelas usadas serão as do catálogo UCAC4. Porém esse catálogo possui imprecisões relevantes principalmente para as estrelas mais fracas. Uma das estrelas testadas apresentou um offset da posição do catálogo para a data da observação de mais de 600 mas. Para a estrela em questão, esse valor representa uma variação de aproximadamente 1900 km da sombra projetada na Terra.

%Para corrigir esses erros, será necessário observar as estrelas a serem ocultadas dias antes do evento obtendo os offsets correspondentes e calibrando a predição. Erros em ascensão reta interferem majoritariamente no instante em que a ocultação ocorrerá. Já erros em declinação principalmente na latitude geográfica em que a sombra passará.

%Outra fonte de erro na predição é a posição do satélite. Como mostradas na seção \ref{Sec: Irr-astrometria}, as órbitas dos satélites irregulares não são muito precisas havendo erros sistemáticos em seus elementos orbitais. Infelizmente o JPL não provê suas efemérides com os erros estimados das posições. Dessa forma, há três alternativas para a correção das posições dos satélites.

%A primeira delas é observar o satélite dias antes, quanto mais próximo do evento melhor, e determinar um offset. O offset, portanto, seria considerado estável dentro desses poucos dias uma vez que ele não se move muito em sua órbita. Porém, nem sempre é possível observar o alvo com antecedência.

%A segunda alternativa é recalcular as órbitas dos satélites, gerar novas efemérides e estimar seus erros. Espera-se que com o grande número de posições publicadas em \cite{GomesJunior2015-Irregular} as órbitas sejam calculadas com maior precisão. Essa integração numérica será trabalhada durante o doutorado sanduíche em Paris em colaboração com o Dr. Jean-Eudes Arlot e o Dr. Josselin Desmars como foi feito para o trabalho de ocultações de TNOs na seção \ref{Sec: TNO-occ}.

%Por fim, a terceira alternativa consiste em se fazer um ajuste dos offsets de efeméride obtidos em \cite{GomesJunior2015-Irregular} em função da anomalia verdadeira, tempo e/ou anomalia média. Essa escolha é suportada pelas dispersões dos offsets em função da anomalia verdadeira (por exemplo, Fig. \ref{Fig: carme_anom}, para Carme) que apresentaram características notáveis. Os parâmetros a serem utilizados para cada satélite serão estudados e aplicados. Esse é o método que será utilizado para o artigo de predições de satélites irregulares que será submetido logo.

%Os primeiros ajustes foram feitos apenas considerando os offsets em declinação, pois são os que apresentaram características mais evidentes e, por seus erros representarem principalmente erros em latitudes, são os principais responsáveis pela não detecção da ocultação. Nesse caso, utilizamos um ajuste em função de três parâmetros: sen${f}, \cos{f}$ e uma constante, onde $f$ é a anomalia verdadeira dada pela efeméride. Esse ajuste é mostrado na Fig. \ref{Fig: Carme-dec-fit}.

%\begin{figure}
%\begin{centering}
%\includegraphics[scale=0.45]{figuras/Carme-DEC_fit.png} 
%\caption{Ajuste dos offsets em declinação de Carme publicados em \cite{GomesJunior2015-Irregular} em função da anomalia verdadeira}
%\label{Fig: Carme-dec-fit}
%\end{centering}
%\end{figure}

%É possível ver para o caso de Carme que o ajuste representa razoavelmente bem a dispersão dos offsets em função da anomalia verdadeira. Para outros satélites, ajuste em primeira ordem do seno e do cosseno da anomalia verdadeira também representam razoavelmente bem os offsets em declinação. Para os offsets em ascensão reta, o ajuste com três parâmetros não é suficiente.

Observar uma ocultação estelar de um satélite irregular exige um esforço muito grande por parte dos observadores. A sombra cobrirá uma região muito restrita na Terra devido aos satélites irregulares serem pequenos. Portanto, antes de iniciar uma grande campanha observacional, testamos algumas predições de ocultações para objetos maiores, para avaliar a qualidade da predição.

Esse teste consiste em se observar o objeto ocultante e a estrela a ser ocultada próximo ao evento, quando os dois objetos estiverem presentes no mesmo campo, preferencialmente próximo ao instante do evento quando os objetos estarão próximos um do outro. Assim as posições relativas entre os dois objetos terão influências mínimas dos erros das estrelas de catálogo utilizadas na redução e das possíveis distorções de campo.

Até o momento, dois testes de ocultações foram feitos, um de Himalia que ocorreu dia 03 de Março de 2015 e o segundo de Elara que ocorreu dia 30 de Março de 2015.
Foram comparadas predições oriundas da nossa integração numérica, das efemérides do JPL, e de offsets de observações tomadas próximas à data do evento.
%Para cada evento quatro mapas foram gerados: o primeiro com as posições nominais da estrela e do satélite para o instante previsto; o segundo com o offset calculado como descrito anteriormente; o terceiro com os offsets da estrela e do satélite a partir de observações feitas alguns dias antes da ocultação quando os dois objetos estavam separados (campos diferentes); e o quarto a partir de observações feitas com menos de 24 horas de diferença da data prevista com a estrela e o satélite próximos no mesmo campo.

%A Fig. \ref{Fig: occ-Himalia} mostra os quatro mapas para o teste da ocultação de Himalia de 03 de Março de 2015. O mapa \ref{Fig: occ-Himalia-off22fev} foi feito a partir de posições obtidas em 22 de Fevereiro observadas no telescópio Zeiss do Observatório do Pico dos Dias (OPD). Nesse dia, Himalia e a estrela a ser ocultada foram observados em campos separados por ainda estarem distantes. Na noite do evento, 03 de Março, os objetos foram observados com o telescópio Perkin-Elmer do OPD pouco mais de uma hora após os instante previsto para o evento. Satélite e estrela estavam separados por aproximadamente 16 arcsec, portanto muito próximos no campo. A partir dos offsets calculados, o mapa \ref{Fig: occ-Himalia-off03mar} foi gerado.

%\begin{figure}[h!]
%\begin{centering}
%\subfigure[Sem offsets]{\includegraphics[scale=0.16]{figuras/Himalia1_2015-03-03T00:39:25.png}  \label{Fig: occ-Himalia-nooff}}
%\subfigure[Offset de Himalia a partir do ajuste]{\includegraphics[scale=0.16]{figuras/Himalia2_2015-03-03T00:39:18.png}  \label{Fig: occ-Himalia-ajuste}}
%\subfigure[Obs. dia 22 de Fevereiro (campos separados)]{\includegraphics[scale=0.16]{figuras/Himalia3_2015-03-03T00:39:40.png}   \label{Fig: occ-Himalia-off22fev}}
%\subfigure[Obs. dia 03 de Março (mesmo campo)]{\includegraphics[scale=0.16]{figuras/Himalia4_2015-03-03T00:39:15.png}   \label{Fig: occ-Himalia-off03mar}}
%\caption{Predições para Himalia: O ponto grande em vermelho mostra o ponto de máxima aproximação geocêntrica da sombra na Terra, os pontos vermelhos menores são os centros da ocultação separados por um minuto, as linhas retas são os limites das sombras dado o tamanho estimado do objeto e as linhas tracejadas equivalem a uma diferença de 40 mas do centro da sombra representando o erro estimado da predição. (a) é o mapa utilizando apenas as posições nominais da estrela e do satélite. (b) mostra a sombra dado um offset estimado para a posição de Himalia dada sua anomalia verdadeira segundo um ajuste feito em cima dos offsets de efeméride encontrados em \cite{GomesJunior2015-Irregular}. Em (c) são aplicados offsets às posições da estrela e do satélite a partir de observações feitas em 22 de Fevereiro no telescópio Zeiss. Em (d), temos o mesmo que para (c) porém com offsets obtidos de observações feitas em 03 de Março no telescópio Perkin-Elmer quando os objetos estavam próximos no campo.}
%\label{Fig: occ-Himalia}
%\end{centering}
%\end{figure}

Para o evento de Himalia identificamos uma boa concordância entre as predições. De fato, a maior diferença entre as sombras calculadas é de $25 s$ e $130 km$ na direção perpendicular às sombras. Para o evento de Elara as observações tomadas no Zeiss prejudicaram a qualidade das posições uma vez que Elara é um objeto pouco brilhante.
%Na Fig. \ref{Fig: occ-Elara} temos os mapas para a ocultação de Elara do dia 30 de Março de 2015. As observações de teste foram tomadas nos dias 25 de Março e 02 de Abril de 2015 no Zeiss. Na noite do dia 02 eles ainda puderam ser observadas no mesmo campo. Devido à Elara ser bem mais fraco, as dispersões das posições do satélite nas duas noites acabaram sendo maiores que para Himalia.
Ainda assim, as diferenças entre os mapas obtidos foram relativamente pequenas. A maior diferença entre elas é de $48 s$ e $302 km$.

%\begin{figure}[h]
%\begin{centering}
%\subfigure[Sem offsets]{\includegraphics[scale=0.16]{figuras/Elara1_2015-03-30T01:45:15.png} \label{Fig: occ-Elara-nooff}}
%\subfigure[Offset de Elara a partir do ajuste]{\includegraphics[scale=0.16]{figuras/Elara2_2015-03-30T01:44:59.png} \label{Fig: occ-Elara-ajuste}}
%\subfigure[Obs. dia 25 de Março (campos separados)]{\includegraphics[scale=0.16]{figuras/Elara3_2015-03-30T01:44:17.png}  \label{Fig: occ-Elara-off25mar}}
%\subfigure[Obs. dia 02 de Abril (mesmo campo)]{\includegraphics[scale=0.16]{figuras/Elara4_2015-03-30T01:45:30.png}  \label{Fig: occ-Elara-off02abr}}
%\caption{Predições para Elara: Mesmo que a Fig. \ref{Fig: occ-Himalia}, mas para Elara. O mapa (c) feito a partir de observações tomadas no dia 25 de Março e (d) a partir de observações de 02 de Abril de 2015.}
%\label{Fig: occ-Elara}
%\end{centering}
%\end{figure}

Um pedido de tempo foi feito para o telescópio Zeiss para se tentar observar uma ocultação de Himalia  dia 23 de Outubro de 2015. A sombra dessa ocultação estava predita para passar pela região sudeste do Brasil. Infelizmente o tempo estava fechado e nenhum dos sítios disponíveis pôde observar o evento.

Uma publicação das predições das ocultações está sendo preparada. A observação de novos eventos serão tentados.

%\begin{figure}[h]
%\includegraphics[scale=0.35]{figuras/Himalia_2015-10-23T07:54:47.png}
%\caption{Mapa da ocultação de Himalia que ocorrerá em 23 de Outubro de 2015}
%\label{Fig: occ-Himalia-itajuba}
%\end{figure}

\chapter{Netuno e Tritão}
\label{Cap: Netuno-Tritao}

\section{Introdução}
\label{Sec: Netuno-intro}

\indent \indent Tritão é um satélite de Netuno que, diferentemente dos satélites regulares dos outros planetas, possui uma órbita retrógrada, circular e altamente inclinada. Devido à sua configuração orbital é muito provável que Tritão tenha sido capturado por Netuno \citep{McKinnon2007}. Ele possui um diâmetro de 1353 km \citep{Thomas2000}, sendo portanto pouco maior que Plutão, e seus parâmetros físicos mostram similaridades com os do planeta anão \citep{Nimmo2015}.

%Dentre os diversos mecanismos propostos para a captura do satélite, o mais provável foi proposto por \cite{Agnor2006} onde Tritão seria capturado através da dissociação de um sistema binário de planetesimais ao passar próximo a Netuno. Após a captura, a órbita de Tritão seria encolhida e circularizada para sua órbita atual. Um dos possíveis mecanismos seria a interação gravitacional com Netuno, i.e., efeito de maré \citep{Nogueira2011}.

%A combinação entre a obliquidade do eixo de rotação de Netuno em relação ao seu eixo orbital e a alta inclinação da órbita de Tritão dá a Tritão um complicado e extremo ciclo sazonal \citep{McKinnon2007}. As marés associadas à sua captura podem ter fortemente aquecido e transformado seu interior.

Tritão é um dos poucos objetos do sistema solar que possuem atmosfera e é constituída principalmente de $N_{2}$. A extrema variação da latitude subsolar em Tritão causa bruscas variações na distribuição de nitrogênio congelado na superfície do objeto de forma que a calota polar pode chegar até próximo o equador do satélite \citep{Hansen1992}. Modelos preveem que a pressão da atmosfera de Tritão varia ao longo do tempo de duas ordens de grandeza.

%\section{Astrometria}
%\label{Sec: Netuno-astrometria}

%\indent \indent Com o objetivo de estudar os efeitos de maré sobre o sistema Netuno-Tritão, serão obtidas as posições desses objetos em mais de 7000 imagens observadas no OPD desde 1992 até o presente. Uma vez que eles estão muito próximos no céu e Netuno é muito mais brilhante do que Tritão, a posição do satélite pode estar afetada pela luz do planeta. Para eliminar essa influência será utilizada uma técnica chamada de "coronografia digital" \citep{Assafin2009}.

%O procedimento de coronografia digital faz parte do pacote PRAIA e foi desenvolvido para se observar os satélites principais de Urano durante os fenômenos mútuos de 2007 \citep{Assafin2009}. A técnica consiste em computar o perfil de luz do objeto mais brilhante dado uma área. A imagem original, então, é subtraída do perfil do objeto gerando uma imagem coronografada.

%Para se obter o perfil do objeto, é calculada a distância $d$ de cada pixel ($j$,$i$) às coordenadas centrais do baricentro ($x_{c}$,$y_{c}$) da fonte brilhante. Para cada anel de 1 pixel de largura a uma distância $d$ do baricentro é feito um histograma de contagens. Por fim, o perfil é obtido sendo que, para cada pixel, a contagem é estimada como a moda do histograma de contagens de seu respectivo anel.

%No caso, a luz de Netuno será removida diminuindo a sua influência sobre a posição de Tritão. Como subproduto, as coordenadas do baricentro calculadas são uma boa estimativa da posição de Netuno.

%Como o tamanho aparente de Netuno ($\approx 2\arcsec$) é da ordem do seeing, uma técnica está sendo desenvolvida para determinar o centroide de objetos extensos. No caso, supõe-se que, para cada ponto da imagem, a luz é espalhada sobre o CCD como uma gaussiana e a imagem correspondente é a soma de todas estas gaussianas. O perfil resultante não tem forma analítica fechada, devendo ser calculado numericamente. Ainda estamos estudando a possibilidade de se obter funções analíticas para a PSF.

\section{Ocultaç\~oes}
\label{Sec: Netuno-predic}

\indent \indent Nos mesmos moldes em que foram feitas observações dos caminhos de TNOs para predição de ocultações estelares (ver seção \ref{Sec: TNOs-astrometria}), o caminho de Tritão também foi observado gerando um catálogo de estrelas nos moldes do catálogo WFI. Infelizmente, devido a Netuno estar passando por regiões pouco populadas por estrelas brilhantes, nenhum evento pôde ser observado oriundo deste trabalho.

%\section{Ocultação de Maio/2008}
%\label{Sec: Netuno-maio2008}

Em 21 de Maio de 2008 houve uma ocultação por Tritão cuja sombra cruzou o sul da África. Dos sítios que observaram, apenas duas cordas foram positivas. As observações foram feitas de La Réunion e da Namíbia com câmeras rápidas. Dentre elas, uma observou somente a atmosfera do objeto. Foi a primeira ocultação de Tritão desde 1997 \citep{Elliot2000}.

Os perfis de pressão da atmosfera de Tritão foram calculados (Sicardy et al., comunicação privada), porém, devido a uma das cordas ter pego somente atmosfera, não foi possível determinar a posição das cordas em relação ao corpo com precisão. Isso resultou em valores de pressão próximo à superfície com pequenas barras de erro interno, mas que podem estar desviados em mais de 3 sigmas do valor real.

Novas ocultações são necessárias para estudar o perfil da atmosfera de Tritão e sua evolução ao longo do tempo. Para 2015, antes que o catálogo WFI não cubra mais o caminho orbital de Tritão, ainda haverá uma oportunidade de observar uma ocultação estelar pelo objeto. O evento está predito para ocorrer dia 10 de Novembro de 2015 com a sombra passando pela China. Dependendo do offset da estrela e de Tritão, há a possibilidade da sombra ir mais ao sul passando pelo Japão ou Austrália, onde temos colaboradores que poderão observar o evento.

%\begin{figure}[h]
%\begin{centering}
%\includegraphics[scale=0.58]{figuras/tritao-2008.png} 
%\caption{Reconstrução pós-ocultação da forma de Tritão. As linhas são as cordas observadas. Como é possível ver, a corda da Namíbia observou apenas a atmosfera de Tritão.}
%\label{Fig: Tritao-body}
%\end{centering}
%\end{figure}
%
%\begin{figure}[h]
%\begin{centering}
%\includegraphics[scale=0.58]{figuras/pressao-2008.png} 
%\caption{Comparação entre a pressão atmosférica na superfície de Tritão obtida por Sicardy et al. (comunicação privada) e pressões publicadas na literatura: \cite{Gurola1995, Olkin1997, Sicardy1998} e \cite{Elliot2000}}.
%\label{Fig: Tritao-pressao}
%\end{centering}
%\end{figure}

%\cite{Hansen1992} geraram modelos que preveem que a pressão atmosférica de Tritão em sua superfície estaria em ascensão. Neste trabalho, a pressão atmosférica foi obtida com um valor muito diferente do valor esperado mesmo ao nível de $3\sigma$. O resultado obtido também é inconsistente com outros valores da literatura que confirmam o esperado por \cite{Hansen1992}.

%Para recuperar essa ocultação, um trabalho astrométrico deve ser feito de forma a medir com precisão maior a posição do centro do corpo no instante da ocultação. Dessa forma, será possível localizar as cordas observadas às suas respectivas latitudes, diminuindo as barras de erro.

\section{Astrometria}
\label{Sec: Netuno-mares}

\indent \indent Com o objetivo de estudar os efeitos de maré sobre o sistema Netuno-Tritão, serão obtidas as posições desses objetos em mais de 7000 imagens observadas no OPD desde 1992 até o presente. Uma vez que eles estão muito próximos no céu e Netuno é muito mais brilhante do que Tritão, a posição do satélite pode estar afetada pela luz do planeta. Para eliminar essa influência será utilizada uma técnica chamada de "coronografia digital" \citep{Assafin2009}.

A vantagem desse banco de dados é sua cobertura temporal (1992 - 2015), o grande número de observações e um tratamento astrométrico diferenciado. Faremos uso de coronografia digital para eliminar a influência da luz difusa de Netuno sobre Tritão, e faremos uma determinação direta de posições de Netuno (com PSFs numéricas especialmente deduzidas para o caso de objetos extensos), com Tritão no mesmo frame. Esses dados tem o mesmo valor que o de posições oriundas de sondas, em que pesem efeitos dinâmicos de longo período \citep{Lainey2012}.

Além das posições de solo, irei trabalhar com posições telemétricas de sondas espaciais dentro do projeto ESPaCE (European Satellite PArtnership for Computing Ephemerides). Esse projeto tem por objetivo extrair e analisar dados astrométricos de medidas feitas por sondas espaciais e utilizá-las para cálculo de órbitas e geração de efemérides dos objetos dos Sistema Solar. Essas medidas são ultra precisas e poderão prover novos modelos dinâmicos para muitos satélites naturais\footnote{\url{http://espace.oma.be/}}.

A combinação desses dois bancos de dados, junto com as posições publicadas na literatura permitirá estudar efeitos dissipativos de energia no interior de Tritão. Como dito na seção \ref{Sec: Netuno-intro}, a circularização da órbita de Tritão, a partir de uma órbita altamente excêntrica, após a captura aconteceu por dissipação oriunda de efeitos de maré \citep{Nimmo2015}.

%Durante a circularização da órbita, o aquecimento causado por efeito de maré teria aumentado chegando a um fluxo de calor de $\approx 2 - 4 W m^{-2}$, comparáveis às medidas atualmente para Io. Após o aquecimento, o interior de Tritão começou a esfriar, mas mesmo hoje esse esfriamento não deve ter sido completado \citep{McKinnon2007}.

%Pela órbita de Tritão ser retrógrada, espera-se que o semi-eixo de sua órbita diminua devido ao efeito de maré que provoca em Netuno e que seja destruído ao se aproximar do planeta devido à essa mesma força \citep{McKinnon2007}. Estima-se que a órbita de Tritão encolha em aproximadamente 15\% nos próximos 5 bilhões de anos.

\chapter{Conclusão e Perspectivas}
\label{Cap: perspectivas}

\indent \indent Esse trabalho tem por meta estudar os objetos do Sistema Solar exterior, de forma a obter características físicas e dinâmicas relevantes para entender a formação e evolução do Sistema Solar. As técnicas utilizadas consistem em obter posições astrométricas precisas desses corpos para corrigir erros sistemáticos em suas efemérides, ou mesmo permitir a determinação de novas órbitas com novas integrações numéricas, a fim de melhorar a predição de futuras ocultações estelares.% A técnica de fotometria diferencial de abertura é empregada para gerar as curvas de luz das ocultações estelares.
Prever, observar e reduzir ocultações estelares por esses objetos e obter seus parâmetros físicos (tamanho, forma, albedo, densidade, presença de atmosfera, satélites ou anéis) são os objetivos gerais desta Tese.

Participo ativamente em observações de TNOs, satélites, e estrelas com o objetivo de melhorar as efemérides e/ou predição de ocultações. Nesse âmbito já escrevi 1 trabalho: \cite{GomesJunior2015-Irregular}; e contribuo em 2 trabalhos: \cite{Camargo2015} e \cite{Desmars2015}.

%A busca pela predição de uma ocultação e sua melhoria na era pré-GAIA vai demandar ainda diversas missões observacionais. Quando o catálogo do GAIA for lançado, as posições das estrelas serão boas o suficientes para não precisar mais de acompanhamento, porém os objetos do Sistema Solar ainda precisaram ser observados e suas efemérides melhoradas pois serão, portanto, a única fonte de erro na predição de uma ocultação.

Observar uma ocultação pode revelar características inesperadas e até dar origem a novas descobertas no Sistema Solar. Exemplos disso são a descoberta de um sistema de anéis, estrutura antes conhecida apenas nos planetas gigantes, ao redor de um Centauro e a identificação de um alto albedo geométrico em Eris, objeto que está muito distante do Sol e possui um tamanho bem menor do que se supunha, similar ao de Plutão e Tritão. Esses dois trabalhos foram publicados na Nature em \cite{BragaRibas2014} e \cite{Sicardy2011}, respectivamente, nos quais contribuí em caráter observacional.

Particularmente, Plutão possui um interesse especial. Recém visitado por uma sonda espacial, ele é um planeta anão que possui atmosfera, cinco satélites, sendo que o maior deles possui um diâmetro pouco maior que a metade do de Plutão. Ocultações estelares por esse objeto permitem estudar a evolução de sua atmosfera, a qual será calibrada a partir de observações oriundas da sonda \textit{New Horizons}. Para isso uma ocultação que ocorrei próxima à data da chegada da sonda foi observada. Das observações às quais participei, dois estudos sobre a atmosfera de Plutão puderam ser feitos: \cite{Olkin2015} e \cite{DiasOliveira2015}.

Da mesma forma que para Plutão, Ceres é um planeta anão e está sendo visitado por uma sonda. A partir de duas ocultações, calculamos o tamanho de Ceres e obtivemos posições muito precisas. Os resultados foram publicados em \cite{GomesJunior2015-Ceres} e serviram como aprendizado pessoal quanto à utilização da técnica. Quando os resultados da sonda \textit{Dawn} forem publicados, será possível calibrar a técnica de ocultações estelares utilizada nesse trabalho.

%Muitos outros objetos estão na lista para se observar ocultações estelares, todos com grandes potenciais científicos. Por exemplo, há a possibilidade de que o Centauro (2060) Chiron também possua anéis \citep{Ruprecht2015}. Como o número de ocultações detectadas pelo nosso grupo vem aumentando ao longo do tempo e aumentará muito mais com o GAIA, diversas publicações ainda estão por vir.

Dentre os objetos a serem ocultados, inserimos na lista os satélites irregulares dos planetas gigantes, nunca antes observados por ocultações. Diversas ocultações já estão preditas e a publicação destas predições está em preparação (Gomes-Júnior et al. 2016, A\&A, in prep.). Observações de ocultações por Himalia, satélite de Júpiter, e Tritão estão sendo preparadas.

Por fim, estudos sobre efeitos de maré no sistema Netuno-Tritão para obter parâmetros internos dos satélites serão feitos durante o Doutorado Sanduíche que ocorrerá na França em colaboração com o Doutor Valéry Lainey. Nesse trabalho, focaremos na obtenção de posições astrométricas do sistema a partir de imagens de solo observadas no OPD e de sondas espaciais dentro do projeto ESPaCE.

\bibliography{Referencias}
\bibliographystyle{apalike}

\end{document}

### 2015-11-13 Irregular-occ/Laurene/article.tex
comentarios Laurene

\documentclass[traditabstract]{aa}
\usepackage{txfonts}
\usepackage{graphicx}
\usepackage{longtable}
\usepackage{natbib}
\usepackage{color}
\bibpunct{(}{)}{;}{a}{}{,} % to follow the A&A style
\begin{document}

\title{Dynamical parameter determinations in Pluto's system}
\subtitle{Expected constraints from the New Horizons mission to Pluto}

\author{L. Beauvalet\inst{1}
\and V. Lainey \inst{1}
\and J.-E. Arlot\inst{1}
\and R. P. Binzel\inst{2}\fnmsep\inst{1}
}

\institute{Institut de M\'ecanique C\'eleste et de Calcul des \'Eph\'em\'erides-Observatoire de Paris, UMR 8028 du CNRS, Universit\'e Pierre et Marie Curie, Universit\'e Lille 1, 77 avenue Denfert Rochereau, 75014 PARIS, France \\
\and
Department of Earth, Atmospheric, and Planetary Sciences, Massachusetts Institute of Technology, 77 Massachusetts Avenue, Cambridge, MA 02139 USA
}
%\date{Received ; accepted}

\abstract
{
Pluto is the multiple system that has been observed the longest. Yet, the masses of its smallest satellites, Nix and Hydra, which were discovered in 2005, are still imprecisely known, because of the short time span and number of available observations. We present a numerical model that takes into account the second order gravity fields and Pluto's orbital motion in the solar system. We investigated the dynamical parameters that may be reliably determined today. We also assessed the possible improvements on the parameter uncertainties with the future increase of observations, including the New Horizons mission. Fitting our model to simulated data, we show that the precision of observations prevents the quantification of the polar oblateness $J_2$ and equatorial bulge $c_{22}$ of Pluto and Charon. Similarly, we show that the masses are on the detection limit.  In particular, unless 25 observations are made every year, the mass of Nix may be constrained with confidence only with New Horizons data. Hydra's mass will only be constrained by the probe. The recent discovery of P4 might change this situation, but our knowledge of this object is still too vague to draw any conclusion.
}

   \keywords{Celestial mechanics -- Planets and satellites: fundamental parameters -- planets and satellites: individual: Pluto -- Methods: numerical -- Kuiper belt: general
               }

   \maketitle

\section{Special tailored Ephemerides (STE) for Jupiter irregular satellites}
\newpage
\subsection{Observations used}

The last observations used to develop JPL current ephemeris of the irregular satellites of Jupiter were obtained in 2012 \citep{Jacobson2012}. As a result, the errors on the residuals of recent observations are too important for the ephemeris to accurately predict stellar occultations without anny corrections. In order to have efficient ephemerides for predicting the stellar occultations by irregular satellites, we need predictions based on recent observations. This is why we decided to develop our own ephemeris based on the observations published in \cite{GomesJunior2015}.

{\color{red}Here comes your paragraphs from \emph{Correction of the ephemeris} about your observations from Gomes-Junior 2015 and how they present systematic errors. Also, the little blurb I write just before will need to be reread and edited (because editing others is sooooo easier and more efficient than editing oneself)}

\newpage
\subsection{Description of the model}

Our numerical model describes the dynamical evolution of the irregular satellites of Jupiter in a jovicentric reference frame. The satellites are submitted to the influence of the Sun and the rest of the solar system, as well as those of the Galilean satellites and the first harmonics of Jupiter's gravity field. The axis of the reference frame are those of the equatorial reference frame J2000.

We use the following notations: \begin{itemize}
\item $i$ and $l$ one of the irregular satellites of Jupiter
\item $J$ Jupiter 
\item $j$ another body of the Solar System
\item $M_j$ the mass of the $j$ body, not an irregular satellite
\item $m_i$ the mass of the irregular satellite $i$
\item $\vec{r_i}$ the position of the $i$ body with respect to the barycentre of Jupiter System
\item $r_{ij}$ the distance between bodies $i$ and $j$  
\item $R_J$ the radius of Jupiter
\item $J_n$ the dynamic polar oblateness of the nth order for Jupiter's gravity field
\item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
\item $\Phi_i$ is the inclination of the $i$ satellite with respect to Jupiter's equator.
\end{itemize}

For an irregular satellite $i$, under the gravitational influence of Jupiter, the $\mathcal{N'}-1$ other irregular satellites, the regular Jovian satellites and the rest of the Solar System ($\mathcal{N}$ bodies), the equation of motion is:
\begin{equation}\begin{array}{ll}

\ddot{\vec{r_i}}= & \displaystyle -GM_J\frac{\vec{r_J}-\vec{r_i}}{r_{iJ}^3}-\sum_{l=1,l\neq i}^\mathcal{N'}Gm_l\frac{\vec{r_l}-\vec{r_i}}{r_{il}^3}\\
&\displaystyle -\sum_{j=1}^\mathcal{N}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \underbrace{\frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3}}_\mathtt{undirect perturbations} \right)\\
 & \displaystyle +GM_J \nabla U_{\bar{l}\hat{J}} -\underbrace{\sum_{l=1}^\mathcal{N} Gm_l\nabla U_{\bar{l}\hat{J}}}_\mathtt{undirect perturbations}
\end{array}
\end{equation}

where the oblateness potential seen by the body $i$ because of Jupiter is:
\begin{equation}\begin{array}{ll}

U_{\bar{l}\hat{J}}=&\displaystyle -\frac{R_J^2 J_2}{r_{iJ}^3}\left(\frac{3}{2}\sin^2 \Phi_i-\frac{1}{2}\right)\\ &\\ & 
\displaystyle-\frac{R_J^4 J_4}{r_{iJ}^5}\left(\frac{35}{8}\sin^4 \Phi_i-\frac{15}{4}\sin^2 \Phi_i+\frac{3}{8}\right)\\
& \\
&\displaystyle-\frac{R_J^6 J_6}{r_{iJ}^7}\left(\frac{231}{16}\sin^6 \Phi_i-\frac{315}{16}\sin^4 \Phi_i+\frac{105}{16}\sin^2 \Phi_i-\frac{5}{16}\right)

\end{array}
\end{equation}

The expressions of $\nabla_lU_{\bar{l}\, \hat{i}}$ and $\nabla_i U_{\bar{i}\, \hat{l}}.$ have been developed in \citet{Lainey2004}.The equations of motion are integrated with the numerical integrator RADAU \citep{Everhart1985}. 
Our model was fitted to the observations through a least-squares procedure. The satellites were integrated one dynamical family at a time, to gain computing time, while losing minimum precision. Indeed, the interactions between satellites not belonging to the same dynamical family are negligible considering the short timespan of our integration.

The initial osculating elements at the origin of integration are presented in Table \ref{sat_ell}, while the residuals are presented in Fig \ref{omcHim} to \ref{}. %don't remember wich one

All the orbits determined for the satellites show satisfying residuals. Yet, one of the satellites has a peculiar situation. Its small number of observations in our set means that its orbit is definitely loosely constrained. 
The residuals are lower than those obtained with JPL ephemeris, but the accuracy of an ephemeris decreases when we get further from the time of observations. The main risk of divergence over time comes from the possible absence of long-term effects when fitting to a short timespan of observations. If that were the case, our ephemeris would diverge too quickly to be of any use. JPL ephemerides are fitted over all the available observations. As a result, they will diverge less quickly than our own. Though they are no longer precise enough for our use, they remain a recious reference to identify whether our own model presents a quick divergence. We compared our ephemeris to the JPL for all the Jupiter satellites we fitted, until 2017. The divergence between 2015 and 2018 is at most \textit{\color{red}insert value here} for \textit{\color{red}insert right satellite name here}. A example of this difference for satellites Himalia and \textit{\color{red}whichever satellite you want, if you want)} is presented in Figure \ref{JPL-STE}.

The obtained ephemeris is hereafter refered as STE, for special-tailored ephemeris.

\begin{table}
\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0. For Themisto's case, the small number of observations meant that the statistical uncertainty was even greater than the obtained elements, though the fitting proved satisfying. We decided to provide its elements with significant number comparable to the other satellite for information, but left it without uncertainty.  }\label{sat_ell}
\begin{center}
\begin{tabular}{ccccccc}
\hline\hline
Satellite & a (km) & e & incl. (EQJ2000) & $\Omega$ & $\omega$ & $v$ \\ 
\hline
Himalia &   11372100 $\pm$ 500    &    0.166 $\pm$ 0.002      &   45.14 $\pm$ 0.15      &   39.77 $\pm$ 0.19      &   351.48 $\pm$ 0.46      &   97.35 $\pm$ 0.48    \\
Elara &   11741170 $\pm$ 690  &      0.222 $\pm$ 0.002      &   28.64 $\pm$ 0.18      &   68.42 $\pm$ 0.43      &   179.82 $\pm$ 0.56      &   339.08 $\pm$ 0.82  \\
Pasiphae &  23425000  $\pm$ 5000    &     0.379  $\pm$ 0.001       &   152.44 $\pm$ 0.10      &   284.59 $\pm$ 0.21      &   135.96 $\pm$ 0.19      &   236.97 $\pm$ 0.16 \\
Sinope &   22968800 $\pm$ 5200   &     0.316 $\pm$ 0.002      &   157.76 $\pm$ 0.12      &   256.62 $\pm$ 0.55      &   298.38 $\pm$ 0.55      &   167.57 $\pm$ 0.19    \\
Lysithea &   11739900 $\pm$ 1300  &      0.136 $\pm$ 0.004      &    51.12 $\pm$ 0.27     &   5.53 $\pm$ 0.52      &   53.0 $\pm$ 1.5      &   318.9 $\pm$ 2.0   \\
Carme &   24202924 $\pm$ 4800      &  0.242 $\pm$ 0.001      &   147.13 $\pm$ 0.10      &   154.01 $\pm$ 0.25      &   47.90 $\pm$ 0.29      &   234.41 $\pm$ 0.19  \\
Ananke &  21683800  $\pm$ 7200  &     0.380 $\pm$ 0.002      &   172.29 $\pm$ 0.20      &   56.9 $\pm$ 1.2      &   123.3 $\pm$ 1.2      &   231.24 $\pm$ 0.21  \\
Leda &   11140300  $\pm$ 4300  &     0.173  $\pm$ 0.007     &   16.15  $\pm$ 0.75    &   272.6  $\pm$ 1.7    &   212.2  $\pm$ 3.6          &   218.8  $\pm$ 3.2  \\
Themisto  & 7393800    &     0.198       &   25.77  &   220.0     &   216.3    &   262.1     \\ \hline
\end{tabular}

\end{center}
\end{table}
\bibliographystyle{plainnat} 
\bibliography{biblio} % your references Yourfile.bib

\end{document}

### 2015-11-14 Irregular-occ/Irregular-occ.tex
.

@@
           \inst{1} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France},
+          L. Beauvalet
+          \inst{2},
           R. Vieira-Martins
@@
           B. E. Morgado
-          \inst{1,2},
-          F. Braga-Ribas
-          \inst{2,4}, 
+          \inst{1,2}
+%          F. Braga-Ribas
+%          \inst{2,4}, 
 %          D. N. da Silva Neto
@@
 
-\cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.
+\subsection{Description of the model}
+
+Our numerical model describes the dynamical evolution of the irregular satellites of Jupiter in a jovicentric reference frame.
+The satellites are submitted to the influence of the Sun and the rest of the solar system, as well as those of the Galilean satellites and the first harmonics of Jupiter's gravity field.
+The axis of the reference frame are those of the equatorial reference frame J2000.
+
+We use the following notations: \begin{itemize}
+\item $i$ and $l$ one of the irregular satellites of Jupiter
+\item $J$ Jupiter 
+\item $j$ another body of the Solar System
+\item $M_j$ the mass of the $j$ body, not an irregular satellite
+\item $m_i$ the mass of the irregular satellite $i$
+\item $\vec{r_i}$ the position of the $i$ body with respect to the barycentre of Jupiter System
+\item $r_{ij}$ the distance between bodies $i$ and $j$  
+\item $R_J$ the radius of Jupiter
+\item $J_n$ the dynamic polar oblateness of the nth order for Jupiter's gravity field
+\item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
+\item $\Phi_i$ is the inclination of the $i$ satellite with respect to Jupiter's equator.
+\end{itemize}
+
+For an irregular satellite $i$, under the gravitational influence of Jupiter, the $\mathcal{N'}-1$ other irregular satellites, the regular Jovian satellites and the rest of the Solar System ($\mathcal{N}$ bodies), the equation of motion is:
+\begin{equation}\begin{array}{ll}
+
+\ddot{\vec{r_i}}= & \displaystyle -GM_J\frac{\vec{r_J}-\vec{r_i}}{r_{iJ}^3}-\sum_{l=1,l\neq i}^\mathcal{N'}Gm_l\frac{\vec{r_l}-\vec{r_i}}{r_{il}^3}\\
+&\displaystyle -\sum_{j=1}^\mathcal{N}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \underbrace{\frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3}}_\mathtt{undirect perturbations} \right)\\
+ & \displaystyle +GM_J \nabla U_{\bar{l}\hat{J}} -\underbrace{\sum_{l=1}^\mathcal{N} Gm_l\nabla U_{\bar{l}\hat{J}}}_\mathtt{undirect perturbations}
+\end{array}
+\end{equation}
+
+where the oblateness potential seen by the body $i$ because of Jupiter is:
+\begin{equation}\begin{array}{ll}
+
+U_{\bar{l}\hat{J}}=&\displaystyle -\frac{R_J^2 J_2}{r_{iJ}^3}\left(\frac{3}{2}\sin^2 \Phi_i-\frac{1}{2}\right)\\ &\\ & 
+\displaystyle-\frac{R_J^4 J_4}{r_{iJ}^5}\left(\frac{35}{8}\sin^4 \Phi_i-\frac{15}{4}\sin^2 \Phi_i+\frac{3}{8}\right)\\
+& \\
+&\displaystyle-\frac{R_J^6 J_6}{r_{iJ}^7}\left(\frac{231}{16}\sin^6 \Phi_i-\frac{315}{16}\sin^4 \Phi_i+\frac{105}{16}\sin^2 \Phi_i-\frac{5}{16}\right)
+
+\end{array}
+\end{equation}
+
+The expressions of $\nabla_lU_{\bar{l}\, \hat{i}}$ and $\nabla_i U_{\bar{i}\, \hat{l}}.$ have been developed in \citet{Lainey2004}.The equations of motion are integrated with the numerical integrator RADAU \citep{Everhart1985}.
+Our model was fitted to the observations through a least-squares procedure.
+The satellites were integrated one dynamical family at a time, to gain computing time, while losing minimum precision.
+Indeed, the interactions between satellites not belonging to the same dynamical family are negligible considering the short timespan of our integration.
+
+The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}, while the residuals are presented in Fig \ref{omcHim} to \ref{}. %don't remember wich one
+
+All the orbits determined for the satellites show satisfying residuals.
+Yet, one of the satellites has a peculiar situation.
+Its small number of observations in our set means that its orbit is definitely loosely constrained.
+The residuals are lower than those obtained with JPL ephemeris, but the accuracy of an ephemeris decreases when we get further from the time of observations.
+The main risk of divergence over time comes from the possible absence of long-term effects when fitting to a short timespan of observations.
+If that were the case, our ephemeris would diverge too quickly to be of any use.
+JPL ephemerides are fitted over all the available observations.
+As a result, they will diverge less quickly than our own.
+Though they are no longer precise enough for our use, they remain a recious reference to identify whether our own model presents a quick divergence.
+We compared our ephemeris to the JPL for all the Jupiter satellites we fitted, until 2017.
+The divergence between 2015 and 2018 is at most \textit{\color{red}insert value here} for \textit{\color{red}insert right satellite name here}.
+A example of this difference for satellites Himalia and \textit{\color{red}whichever satellite you want, if you want)} is presented in Figure \ref{JPL-STE}.
+
+The obtained ephemeris is hereafter refered as STE, for special-tailored ephemeris.
+
+\begin{table*}
+\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0.
+For Themisto's case, the small number of observations meant that the statistical uncertainty was even greater than the obtained elements, though the fitting proved satisfying.
+We decided to provide its elements with significant number comparable to the other satellite for information, but left it without uncertainty.
+}\label{Tab: sat_ell}
+\begin{center}
+\begin{tabular}{ccccccc}
+\hline\hline
+Satellite & a (km) & e & incl. (EQJ2000) & $\Omega$ & $\omega$ & $v$ \\ 
+\hline
+Himalia &   11372100 $\pm$ 500    &    0.166 $\pm$ 0.002      &   45.14 $\pm$ 0.15      &   39.77 $\pm$ 0.19      &   351.48 $\pm$ 0.46      &   97.35 $\pm$ 0.48    \\
+Elara &   11741170 $\pm$ 690  &      0.222 $\pm$ 0.002      &   28.64 $\pm$ 0.18      &   68.42 $\pm$ 0.43      &   179.82 $\pm$ 0.56      &   339.08 $\pm$ 0.82  \\
+Pasiphae &  23425000  $\pm$ 5000    &     0.379  $\pm$ 0.001       &   152.44 $\pm$ 0.10      &   284.59 $\pm$ 0.21      &   135.96 $\pm$ 0.19      &   236.97 $\pm$ 0.16 \\
+Sinope &   22968800 $\pm$ 5200   &     0.316 $\pm$ 0.002      &   157.76 $\pm$ 0.12      &   256.62 $\pm$ 0.55      &   298.38 $\pm$ 0.55      &   167.57 $\pm$ 0.19    \\
+Lysithea &   11739900 $\pm$ 1300  &      0.136 $\pm$ 0.004      &    51.12 $\pm$ 0.27     &   5.53 $\pm$ 0.52      &   53.0 $\pm$ 1.5      &   318.9 $\pm$ 2.0   \\
+Carme &   24202924 $\pm$ 4800      &  0.242 $\pm$ 0.001      &   147.13 $\pm$ 0.10      &   154.01 $\pm$ 0.25      &   47.90 $\pm$ 0.29      &   234.41 $\pm$ 0.19  \\
+Ananke &  21683800  $\pm$ 7200  &     0.380 $\pm$ 0.002      &   172.29 $\pm$ 0.20      &   56.9 $\pm$ 1.2      &   123.3 $\pm$ 1.2      &   231.24 $\pm$ 0.21  \\
+Leda &   11140300  $\pm$ 4300  &     0.173  $\pm$ 0.007     &   16.15  $\pm$ 0.75    &   272.6  $\pm$ 1.7    &   212.2  $\pm$ 3.6          &   218.8  $\pm$ 3.2  \\
+Themisto  & 7393800    &     0.198       &   25.77  &   220.0     &   216.3    &   262.1     \\ \hline
+\end{tabular}
+
+\end{center}
+\end{table*}
+%\cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.
 They showed that the orbits of the irregular satellites of the giant planets have systematic errors.
@@
 
-Making a new model for the orbits of these objects would demand a lot of time and delay the publication of predictions of possible events to occur in the very near future.
+%Making a new model for the orbits of these objects would demand a lot of time and delay the publication of predictions of possible events to occur in the very near future.
 Thus, we utilized the positions obtained by \cite{GomesJunior2015} to identify error patterns in the ephemeris.
@@
 
-\textcolor{red}{...}
+\textcolor{red}{Here we write something about the numerical integration}
 

### 2015-11-17 Irregular-occ/Irregular-occ.tex
Iniciado mudancas sugeridas pela Laurene

@@
 {Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system.
-It is important to know their physical parameters, such as size and albedo, to trace back their true origin.
-The best ground-based technique to obtain such parameters is the observation of stellar occultations by these objects.}
-{We aimed to predict stellar occultations for the seven biggest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope and Ananke; for the irregular satellite of Saturn: Phoebe; and the satellites of Neptune: Nereid and Triton.}
+It is important to know their physical parameters, such as size, albedo and composition, to trace back their true origin.
+The best ground-based technique to constrain those parameters is the observation of stellar occultations by these objects.}
+{We aim to predict stellar occultations for the seven biggest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda; the irregular satellite of Saturn: Phoebe; and the satellites of Neptune: Nereid and Triton.}
 {We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument.
 These catalogues were crossed with the ephemeris of the satellites to identify stellar occultations.
-The ephemeris were generated by a numerical integration \textcolor{red}{(Here we may write something about after the work of Laurène)} that used the positions obtained by \cite{GomesJunior2015} to improve the predictions.}
+We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate quick predictions for the short-time future for the satellites of Jupiter.
+For Phoebe, Nereid and Triton we used the JPL ephemeris.}
 {We managed to obtain \textcolor{red}{588} candidates of stellar occultations between the period of January, 2015 and December, 2017.
 We made observational tests for the predictions of the events of Himalia in March 03, 2015 and Elara in March 30, 2015.
-The observations were made in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions.}
-{The comparisons between the predictions and the observation tests show a good agreement. We discuss how the successful observation of a stellar occultation by these objects is quite possible.}
+The stars and objects were observed close to the date predicted and and in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions.}
+{The comparisons between the predictions and the observation tests show a good agreement.
+We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.}
 
@@
 
-Irregular (outer) satellites revolve around giant planets at large distances in eccentric, highly inclined and frequently retrograde orbits.
-Because of these peculiar orbits, it is largely accepted that these objects were not originated with, but were captured by their planets in the early solar system \citep{Sheppard2005}.
+Irregular satellites revolve around giant planets at large distances, on eccentric, highly inclined and frequently retrograde orbits.
+Because of these peculiar orbits, it is largely accepted that these objects did not form by accretion around of their planet, but were captured by their planets in the early solar system \citep{Sheppard2005}.
 
-There is a number of capture mechanisms of objects by giant planets proposed in the literature.
-There is the Gas Drag in the primordial circumplanetary nebulae \citep{Sheppard2005} where the object would be affected by the gas drag and its velocity slowed down until it be captured by the planet.
+There are several of capture mechanisms of objects by giant planets suggested in the literature.
+For example the Gas Drag in the primordial circumplanetary nebulae \citep{Sheppard2005} where the object would be affected by the gas drag and its velocity slowed down until it be captured by the planet.
 Another mechanism is called pull-down capture \citep{Sheppard2005}, where the mass of the planet would increase while the object was temporarily captured.
@@
 A mechanism based in the Nice model \citep{Morbidelli2005, Tsiganis2005, Gomes2005} was proposed by \cite{Nesvorny2007} and, in the specific case of Jupiter with the modern Nice model, by \citealp{Nesvorny2014}.
-During the early solar system instability, encounters between the outer planets occurred.
-These planetary encounters could exchange energy and angular momentum between planets and the objects nearby making it possible for the capture of irregular bodies by the giant planets.
+In this scenario, during the early solar system instability, encounters between the outer planets occurred.
+These planetary encounters could exchange energy and angular momentum between planets and the objects nearby allowing the capture of irregular bodies by the giant planets.
 In this scenario, the survival rate of prior-LHB (Late Heavy Bombardment) satellites is very small.
 
-Another important mechanism is the capture through collisional interactions \citep{Sheppard2005}.
+Another possible mechanism is the capture through collisional interactions \citep{Sheppard2005}.
 A collision between two small bodies in the Hill's sphere of the planet could generate fragmented objects and the dissipated energy could be such that some of these objects could be captured.
 
-Some of these objects are in dynamical groups with similar orbital elements, called families, similar to families found in the Main Asteroid Belt.
+Some of these objects are in dynamical groups with similar semi-major axis, eccentricities and inclinations, called families, similar to families found in the Main Asteroid Belt.
 These families may have been created by a parent body disrupted by collisions with comets or other satellites \citep{Nesvorny2004}.
-Collisions with comets are more likely to have occurred during the LHB \citep{Gomes2005}.
+These collisions are more likely to have occurred during the LHB \citep{Gomes2005}.
 
@@
 For Phoebe, ejected material from its surface caused by impacts could evolve due to Poynting-Robertson drag and collide with Iapetus causing the large variation in albedo observed on it \citep{Nesvorny2003}.
-Indeed, Cassini was able to detected in Phoebe an absorption feature at 2.42 $\mu m$ (probably CN combinations) that was also detected in the dark side of Iapetus \citep{Clark2005}.
+Indeed, Cassini detected in Phoebe an absorption feature at 2.42 $\mu m$ (probably CN combinations) that was also detected in the dark side of Iapetus \citep{Clark2005}.
 
 The region of origin of these object is not well known.
-\cite{Grav2003} and \cite{Grav2007} showed that the irregular satellites from the giant planets have their colors and spectral slopes similar to C-, D- and P-type asteroids, Centaurs and trans-neptunian objects (TNOs) suggesting that they have been originated from different locations in the early solar system.
+\cite{Grav2003} and \cite{Grav2007} showed that the irregular satellites from the giant planets have their colors and spectral slopes similar to C-, D- and P-type asteroids, Centaurs and trans-neptunian objects (TNOs) suggesting that they come from different locations in the early solar system.
 
-In order to obtain precise fundamental physical parameters like size, shape and albedo for the irregular satellites and thus to contribute to the study of their origin, we will make use of stellar occultations, which provide  more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
+In order to obtain precise fundamental physical parameters like size, shape and albedo for the irregular satellites and thus to contribute to the study of their origin, we want to observe stellar occultations, which provide  more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
 
@@
 But, in fact, given their distances to us, current ephemeris and star positions are already good enough for the prediction of the exact location and instant where the shadow of the occultation will cross the Earth.
-For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds).Thus, in this case, the overall budget of ephemeris and star position errors must be around 40 mas in order that we get good chances of observing a stellar occultation, which is quite feasible today.
+For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds).
+Thus, in this case, the cumulated uncertainty of ephemeris and star position1 must be around 40 mas to get good chances of observing a stellar occultation, which is quite feasible today.
 
@@
 \multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
-Satellite  & km  & mas\tablefootmark{a} & References \tabularnewline
+Satellite  & mas\tablefootmark{a}  & km & References \tabularnewline
 \hline
-Ananke & 29 & 8 & 1 \tabularnewline
-Carme & 46 & 13 & 1 \tabularnewline
-Elara & 86 & 24 & 1 \tabularnewline
-Himalia & $(150\times120) \pm 20$\tablefootmark{b} & 41 & 2 \tabularnewline
-Lysithea & 36 & 10 & 1 \tabularnewline
-Pasiphae & 62 & 17 & 1 \tabularnewline
-Sinope & 37 & 10 & 1 \tabularnewline
+Ananke & 8 & 29 & 1 \tabularnewline
+Carme & 13 & 46 & 1 \tabularnewline
+Elara & 24 & 86 & 1 \tabularnewline
+Himalia & 41 & $(150\times120) \pm 20$\tablefootmark{b} & 2 \tabularnewline
+Lysithea & 10 & 36 & 1 \tabularnewline
+Pasiphae & 17 & 62 & 1 \tabularnewline
+Sinope & 10 & 37 & 1 \tabularnewline
 \hdashline
-Phoebe & $212 \pm 1.4$\tablefootmark{b} & 32 & 3 \tabularnewline
+Phoebe & 32 & $212 \pm 1.4$\tablefootmark{b} & 3 \tabularnewline
 \hdashline
-Nereid & $340 \pm 50$\tablefootmark{c} & 15 & 4 \tabularnewline
-Triton & $2707 \pm 2.0$\tablefootmark{c} & 124 & 5 \tabularnewline
+Nereid & 15 & $340 \pm 50$\tablefootmark{c} & 4 \tabularnewline
+Triton & 124 & $2707 \pm 2.0$\tablefootmark{c} & 5 \tabularnewline
 \hline
@@
 
-Unlike stellar occultations by TNOs, which nevertheless have been proved effective, the observation of an occultation by a irregular satellite is, in principle, more favorable.
+From what we have just presented, we can think that the obsevation of an stellar occultation by an irregular satellite should be likelier than occultations by TNOs..
+The orbits of their host planets are well known and their observation time-span covers many orbital periods, contrary to TNOs.
+%Unlike stellar occultations by TNOs, which nevertheless have been proved effective, the observation of an occultation by a irregular satellite is, in principle, more favorable.
 The orbits of their host planets are well known and these satellites have been observed completing already many orbital periods around them, thus presenting better ephemeris than TNOs.
-Moreover, the irregular satellites are closer to Earth which means a minor location error in kilometers.
+Moreover, the irregular satellites are closer to Earth which means a minor localization error in kilometers.
 
@@
 For an object at the distance of Jupiter, this represents an error bigger than 700 km in the shadow path.
-Using the positions obtained by \cite{GomesJunior2015} it is possible to improve the ephemeris and better predict stellar occultations for these objects.
+Using the positions obtained by \cite{GomesJunior2015} we produced a specific ephemeris for short-time future for the satellites of Jupiter and better predict stellar occultations for these objects.
 
@@
 In the section \ref{Sec: Rationale} we explore the scientific rationale for the study of the irregular satellites and the possibility of having a common origin with TNOs.
-In section \ref{Sec: Offset} we show the \textcolor{red}{numerical integration to improve the ephemeris}.
+In section \ref{Sec: integration} we show the the process of th productin of the new ephemeris.
 In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites and how they were made.
-Some tests made to confirm the predictions are presented in section \ref{Sec: testes} and the conclusions are given in section \ref{Sec: conclusions}.
+Some tests made to confirm the predictions are presented in section \ref{Sec: testes} and the final discussion are in section \ref{Sec: discussion}.
 
@@
 \cite{Clark2005} suggest that its surface is probably covered by material of cometary origin.
-It was also stated by \cite{Johnson2005} that if the porosity of Phoebe would be of about 15\%, which the authors believe it is possible, Phoebe would have an uncompressed density similar to those of Pluto and Triton.
+It was also stated by \cite{Johnson2005} that if the porosity of Phoebe is 15\%, Phoebe would have an uncompressed density similar to those of Pluto and Triton.
 
-\cite{Sheppard2005} and \cite{Jewitt2007} also expose the possibility for the irregular satellites having their origin as comets or similar to the TNOs.
-TNOs are highly interesting objects that due to their large heliocentric distances (low temperatures) may be highly preserved having their physical properties similar to those they had when they were formed, then providing history and evolution of the outer solar system \citep{Camargo2014}.
+\cite{Sheppard2005} and \cite{Jewitt2007} also expose the possibility for the irregular satellites having their origin as comets or TNOs.
+TNOs are highly interesting objects that, due to their large heliocentric distances, may be highly preserved having their physical properties similar to those they had when they were formed \citep{Camargo2014}.
 This is even more true for the smaller objects, since in principle larger sizes favour physical differentiation processes in the body and vice-versa.
@@
 
-Triton is a special satellite.
+Triton is a uncommon satellite.
 Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
@@
 
-Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have a similar origin as the TNOs \citep{Agnor2006}.
+Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have the same origin as the TNOs \citep{Agnor2006}.
 Differently, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
-The main motivation here to study Triton by stellar occultations is to understand the evolution of its atmosphere due to Triton's complicated and extreme seasonal cycle \citep{McKinnon2007}.
+The main motivation to study Triton by stellar occultations is to understand the evolution of its atmosphere due to Triton's complicated and extreme seasonal cycle \citep{McKinnon2007}.
 
@@
 
-\section{Correction of the ephemeris} \label{Sec: Offset}
+\section{Special tailored Ephemerides (STE) for Jupiter irregular satellites} \label{Sec: integration}
+
+\subsection{Observations used}
+
+The last observations used to develop JPL current ephemeris of the irregular satellites of Jupiter were obtained in 2012 \citep{Jacobson2012}.
+As a result, the errors on the residuals of recent observations are too important for the ephemeris to accurately predict stellar occultations without any corrections.
+In order to have efficient ephemerides for predicting the stellar occultations by irregular satellites, we need predictions based on recent observations.
+This is why we decided to develop our own ephemeris based on the observations published in \cite{GomesJunior2015}.
+
+\cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.
+%They showed that the orbits of the irregular satellites of the giant planets have systematic errors.
+The offsets of the observations relative to the JPL ephemeris could be up to 200 mas for some satellites. %These differences could be associated with errors in their orbital elements.
 
@@
 All the orbits determined for the satellites show satisfying residuals.
-Yet, one of the satellites has a peculiar situation.
+%Yet, one of the satellites has a peculiar situation.
 Its small number of observations in our set means that its orbit is definitely loosely constrained.
@@
 As a result, they will diverge less quickly than our own.
-Though they are no longer precise enough for our use, they remain a recious reference to identify whether our own model presents a quick divergence.
+Though they are no longer precise enough for our use, they remain a precious reference to identify whether our own model presents a quick divergence.
 We compared our ephemeris to the JPL for all the Jupiter satellites we fitted, until 2017.
@@
 Leda &   11140300  $\pm$ 4300  &     0.173  $\pm$ 0.007     &   16.15  $\pm$ 0.75    &   272.6  $\pm$ 1.7    &   212.2  $\pm$ 3.6          &   218.8  $\pm$ 3.2  \\
-Themisto  & 7393800    &     0.198       &   25.77  &   220.0     &   216.3    &   262.1     \\ \hline
+%Themisto  & 7393800    &     0.198       &   25.77  &   220.0     &   216.3    &   262.1     \\
+\hline
 \end{tabular}
@@
 \end{table*}
-%\cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.
-They showed that the orbits of the irregular satellites of the giant planets have systematic errors.
-The offsets of the observations relative to the JPL ephemeris could be up to 200 mas for some satellites.
-These differences could be associated with errors in their orbital elements.
 
@@
 
-\begin{figure*}
-\begin{centering}
-\subfigure[Offset from the JPL ephemeris]{\includegraphics[width=8.8cm]{figures/DEC_anom.png} \label{Fig:offxtanomxjpl}}
-\subfigure[\textcolor{red}{Offset from the numerical integration}]{\includegraphics[width=8.8cm]{figures/DEC_anom.png} \label{Fig:offxtanomxni}}
-\caption{Offsets of the declination of Carme by true anomaly from the JPL ephemeris and \textcolor{red}{from the numerical integration of the positions obtained in \cite{GomesJunior2015}}.
-\label{Fig:offxanom}}
-\end{centering}
-\end{figure*}
+%\begin{figure*}
+%\begin{centering}
+%\subfigure[Offset from the JPL ephemeris]{\includegraphics[width=8.8cm]{figures/DEC_anom.png} \label{Fig:offxtanomxjpl}}
+%\subfigure[\textcolor{red}{Offset from the numerical integration}]{\includegraphics[width=8.8cm]{figures/DEC_anom.png} \label{Fig:offxtanomxni}}
+%\caption{Offsets of the declination of Carme by true anomaly from the JPL ephemeris and \textcolor{red}{from the numerical integration of the positions obtained in \cite{GomesJunior2015}}.
+%\label{Fig:offxanom}}
+%\end{centering}
+%\end{figure*}
 
@@
 
-The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the corrected JPL ephemeris as presented in section \ref{Sec: Offset}.
+The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the special-tailored ephemeris for Jupiter satellites and JPL ephemeris for the others as presented in section \ref{Sec: integration}.
 The search for stellar candidates follows the same procedure as presented by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.
@@
 A total of 588 events \textcolor{red}{(Not counting Triton's event yet)} were identified between January 2015 and December 2017.
-Table \ref{Tab: occ-list} exemplifies the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps (see Fig.
+Table \ref{Tab: occ-list} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps (see Fig.
 \ref{Fig: ocultacao} as an example).
@@
 Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - in the original table, these coordinates are immediately followed by the geocentric astrometric equatorial coordinates of the satellite (corrected the computed ephemeris offset derived in Sec.
-\ref{Sec: Offset}); C/A: the geocentric closest approach, in arcseconds; P/A: the planet position angle with respect to the occulted star at C/A, in degrees; velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; D: planet range to Earth, in AU; R*: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{Mag}* = \textrm{Mag}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$.
+\ref{Sec: integration}); C/A: the geocentric closest approach, in arcseconds; P/A: the planet position angle with respect to the occulted star at C/A, in degrees; velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; D: planet range to Earth, in AU; R*: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{Mag}* = \textrm{Mag}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$.
 A value of 50.0 means that the star is not in the 2MASS; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\Delta e_{\alpha *}$ and $\Delta e_{\delta}$: computed offsets in mas applied to the ephemeris right ascension and declination, respectively; pm: ok = proper motion applied, no = no proper motion applied; catalogue cross-identification (ct) = uc (UCAC2), 2m (2MASS), fs (field star); $E_{\alpha *}$ and $E_{\delta}$: uncertainties (mas) in right ascension and declination.
@@
 \begin{centering}
-\includegraphics[scale=0.4]{figures/Himalia_2015-10-23T07:54:47.png} 
+\includegraphics[scale=0.4]{figures/Pasiphae_2016-04-09T03:58:19.png}   
 \caption{Occultation map for Himalia regarding to the second event sampled in Table \ref{Tab: occ-list}}
@@
 The ESA astrometry satellite GAIA \citep{deBruijne2012} catalogue is expected to be released in the end of 2016.
-The precise star positions to be derived by GAIA will render better predictions with the only source of error being the ephemeris.
-The work of \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions from 2017 onwards will be better.
-For that, we only publish stellar occultations with UCAC4 until 2017.
+The precise star positions to be derived by GAIA will provide better predictions with the main source of error being the ephemeris.
+Astrometric reduction of observations published in \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions from 2017 onwards will be better.
+Because of that, we only publish stellar occultations with UCAC4 until 2017.
 
@@
 
-Observe a stellar occultation demands a great effort.
-And, in our case, the shadow covers a very restricted area on Earth because the irregular satellites are small.
+Observing a stellar occultation demands a great effort.
+And, in our case, the shadow covers a very restricted area on Earth because of the size of the irregular satellites.
 So, before we start a large observational campaign, we tested some occultation predictions for large targets, to assess the quality of the predictions.
 
-The tests consisted in observing the object and star to be occulted near the event when the two objects were present in the same field of view (FOV), preferably when the objects were close to each other.
+The tests consisted in observing the object and star to be occulted near the date of the event predicted when the two objects were present in the same field of view (FOV), preferably when the objects were close to each other.
 Thus, the relative positions between the two objects had minimal influence of the errors of the reference catalogue of stars used and possible field distortions \citep[and references therein]{Peng2008}.
@@
 Two occultation tests were performed, one by Himalia that occurred on March 3, 2015 and the second by Elara that occurred on March 30, 2015.
-For each event, four situations were considered: the first with the nominal positions of the star and the satellite to the predicted time; the second with the offset calculated as described in section \ref{Sec: Offset}; the third with star and satellite offsets from observations made a few days before the occultation when the two objects were separated (different FOV); and the fourth from observations made when the star and the satellite were close in the same FOV.
+For each event, four situations were considered: the first with the nominal positions of the star and the satellite to the predicted time; the second with the offset calculated as described in section \ref{Sec: integration}; the third with star and satellite offsets from observations made a few days before the occultation when the two objects were separated (different FOV); and the fourth from observations made when the star and the satellite were close in the same FOV.
 
-Figure \ref{Fig: occ-Himalia}shows four zoomed in occultation maps corresponding to each of the four situations for Himalia..
+Figure \ref{Fig: occ-Himalia} shows four zoomed in occultation maps corresponding to each of the four situations for Himalia..
 The map \ref{Fig: occ-Himalia-nooff} is the nominal prediction with the coordinate of the star given by the catalogue and that of the satellite from the ephemeris.
-We then corrected the positions of the satellite by an offset calculated by the method showed in section \ref{Sec: Offset} to make the map \ref{Fig: occ-Himalia-ajuste}.
+We then corrected the positions of the satellite by an offset calculated by the method showed in section \ref{Sec: integration} to make the map \ref{Fig: occ-Himalia-ajuste}.
 The map \ref{Fig: occ-Himalia-off22fev} was made from obtained positions on February 22 observed with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / px$) at the Observatório do Pico dos Dias (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
@@
 \begin{centering}
-\subfigure[Nominal prediction from JPL]{\includegraphics[width=8.8cm]{figures/Himalia_nominal.eps}  \label{Fig: occ-Himalia-nooff}}
-\subfigure[\textcolor{red}{From the numerical integration}]{\includegraphics[width=8.8cm]{figures/Himalia2_2015-03-03T00:39:18.png}  \label{Fig: occ-Himalia-ajuste}}
+\subfigure[Prediction with STE ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_STE.eps}  \label{Fig: occ-Himalia-nooff}}
+\subfigure[Prediction with JPL ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_JPL.eps}  \label{Fig: occ-Himalia-ajuste}}
 \subfigure[Observation at February 22 (different FOVs)]{\includegraphics[width=8.8cm]{figures/Himalia_22fev.eps}   \label{Fig: occ-Himalia-off22fev}}
@@
 \ref{Fig: occ-Himalia-nooff} is the map using the nominal positions of the star and satellite.
-\ref{Fig: occ-Himalia-ajuste} shows the shadow given an estimated offset for the position of Himalia related to the JPL ephemeris obtained in section \ref{Sec: Offset}.
+\ref{Fig: occ-Himalia-ajuste} shows the shadow given an estimated offset for the position of Himalia related to the JPL ephemeris obtained in section \ref{Sec: integration}.
 In \ref{Fig: occ-Himalia-off22fev} we apply offsets to the positions of star and satellite accordingly the observations made at February 22.
@@
 \begin{table}
-\caption{\label{Tab: comparison-Himalia} Comparison between the predictions of the Himalia occultation at March 03, 2015. \textcolor{red}{Waiting for the right offset}}
+\caption{\label{Tab: comparison-Himalia} Comparison between the predictions of the Himalia occultation at March 03, 2015.}
 \begin{centering}
@@
 \hline  \hline
-\multicolumn{3}{c}{Difference from nominal prediction} \tabularnewline
+\multicolumn{3}{c}{Difference from STE prediction} \tabularnewline
 Method  & Instant of C/A  & C/A   \tabularnewline
 \hline
-Nominal & 00:39:25 UTC & $0\farcs714$ \tabularnewline
-Offset &   & \tabularnewline
-Feb. 22 Obs. & +12s & -33mas (109km) \tabularnewline
-Mar. 03 Obs. & -10s & -19mas (61km)\tabularnewline
+STE & 00:39:51 UTC & $0\farcs703$ \tabularnewline
+JPL & -26s & +11mas (36km) \tabularnewline % (284km)
+Feb. 22 Obs. & -14s & -20mas (65km) \tabularnewline % (153km)
+Mar. 03 Obs. & -36s & -09mas (29km)\tabularnewline % (393km)
 \hline
@@
 \begin{table}
-\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara occultation at March 30, 2015. \textcolor{red}{Waiting for the right offset}}
+\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara occultation at March 30, 2015.}
 \begin{centering}
@@
 \hline  \hline
-\multicolumn{3}{c}{Difference from nominal prediction} \tabularnewline
+\multicolumn{3}{c}{Difference from STE prediction} \tabularnewline
 Method  & Instant of C/A  & C/A  \tabularnewline
 \hline
-Nominal & 01:45:15 UTC & $1\farcs139$ \tabularnewline
-Offset &  & \tabularnewline
-Mar. 25 Obs. & -59 s & -17mas (61km) \tabularnewline
-Apr. 02 Obs. & +15 s & -89mas (307km) \tabularnewline
+STE & 01:45:13 UTC & $1\farcs082$ \tabularnewline
+JPL & +02s & +57mas(195km) \tabularnewline
+Mar. 25 Obs. & s & mas (km) \tabularnewline
+Apr. 02 Obs. & s & mas (km) \tabularnewline
 \hline
@@
 
-\section{Discussion} \label{Sec: conclusions}
+\section{Discussion} \label{Sec: discussion}
 

### 2015-11-18 Irregular-occ/Irregular-occ.tex
.

@@
 Himalia & 41 & $(150\times120) \pm 20$\tablefootmark{b} & 2 \tabularnewline
+Leda &  &  & 1 \tabularnewline
 Lysithea & 10 & 36 & 1 \tabularnewline
@@
 Elara &   11741170 $\pm$ 690  &      0.222 $\pm$ 0.002      &   28.64 $\pm$ 0.18      &   68.42 $\pm$ 0.43      &   179.82 $\pm$ 0.56      &   339.08 $\pm$ 0.82  \\
+Lysithea &   11739900 $\pm$ 1300  &      0.136 $\pm$ 0.004      &    51.12 $\pm$ 0.27     &   5.53 $\pm$ 0.52      &   53.0 $\pm$ 1.5      &   318.9 $\pm$ 2.0   \\
+Leda &   11140300  $\pm$ 4300  &     0.173  $\pm$ 0.007     &   16.15  $\pm$ 0.75    &   272.6  $\pm$ 1.7    &   212.2  $\pm$ 3.6          &   218.8  $\pm$ 3.2  \\
 Pasiphae &  23425000  $\pm$ 5000    &     0.379  $\pm$ 0.001       &   152.44 $\pm$ 0.10      &   284.59 $\pm$ 0.21      &   135.96 $\pm$ 0.19      &   236.97 $\pm$ 0.16 \\
 Sinope &   22968800 $\pm$ 5200   &     0.316 $\pm$ 0.002      &   157.76 $\pm$ 0.12      &   256.62 $\pm$ 0.55      &   298.38 $\pm$ 0.55      &   167.57 $\pm$ 0.19    \\
-Lysithea &   11739900 $\pm$ 1300  &      0.136 $\pm$ 0.004      &    51.12 $\pm$ 0.27     &   5.53 $\pm$ 0.52      &   53.0 $\pm$ 1.5      &   318.9 $\pm$ 2.0   \\
 Carme &   24202924 $\pm$ 4800      &  0.242 $\pm$ 0.001      &   147.13 $\pm$ 0.10      &   154.01 $\pm$ 0.25      &   47.90 $\pm$ 0.29      &   234.41 $\pm$ 0.19  \\
 Ananke &  21683800  $\pm$ 7200  &     0.380 $\pm$ 0.002      &   172.29 $\pm$ 0.20      &   56.9 $\pm$ 1.2      &   123.3 $\pm$ 1.2      &   231.24 $\pm$ 0.21  \\
-Leda &   11140300  $\pm$ 4300  &     0.173  $\pm$ 0.007     &   16.15  $\pm$ 0.75    &   272.6  $\pm$ 1.7    &   212.2  $\pm$ 3.6          &   218.8  $\pm$ 3.2  \\
 %Themisto  & 7393800    &     0.198       &   25.77  &   220.0     &   216.3    &   262.1     \\
@@
 A total of 588 events \textcolor{red}{(Not counting Triton's event yet)} were identified between January 2015 and December 2017.
+In table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite.
 Table \ref{Tab: occ-list} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps (see Fig.
@@
 Since these objects are very small, the duration of each event is a few seconds.
+
+\begin{table}
+\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from 2015 up to 2017.}
+\begin{centering}
+\begin{tabular}{lcccc}
+\hline  \hline
+%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
+Satellite  & 2015  & 2016 & 2017 & Total \tabularnewline
+\hline
+Ananke & 22 & 12 & 16 & 50 \tabularnewline
+Carme & 25 & 20 & 14 & 59 \tabularnewline
+Elara & 32 & 14 & 16 & 62 \tabularnewline
+Himalia & 25 & 15 & 12 & 52 \tabularnewline
+Lysithea & 22 & 16 & 11 & 49 \tabularnewline
+Pasiphae & 26 & 19 & 20 & 65 \tabularnewline
+Sinope & 19 & 15 & 21 & 55 \tabularnewline
+Leda &  &  &  &  \tabularnewline
+\hdashline
+Phoebe\tablefootmark{a} & 13 & 32 & 98 & 143 \tabularnewline
+\hdashline
+Nereid\tablefootmark{a} & 41\tablefootmark{b} & 11\tablefootmark{b} & 1 & 53 \tabularnewline
+Triton\tablefootmark{a} & \tablefootmark{b} & \tablefootmark{b} &  &  \tabularnewline
+\hline
+\end{tabular}
+\par \end{centering}
+Occultations predicted using the UCAC4 catalogue and STE ephemeris.
+\tablefoottext{a}{Using JPL ephemeris.}
+\tablefoottext{b} {Using the WFI catalogue as explained in Sec. \ref{Sec: predictions}} \textcolor{red}{Numbers of Triton yet to be written}.
+\end{table}
 
@@
 
-Two occultation tests were performed, one by Himalia that occurred on March 3, 2015 and the second by Elara that occurred on March 30, 2015.
-For each event, four situations were considered: the first with the nominal positions of the star and the satellite to the predicted time; the second with the offset calculated as described in section \ref{Sec: integration}; the third with star and satellite offsets from observations made a few days before the occultation when the two objects were separated (different FOV); and the fourth from observations made when the star and the satellite were close in the same FOV.
+Two occultation tests were performed, one by Himalia that occurred on March 3, 2015 and the second by Elara that occurred on March 30, 2015. For each event, four situations were considered:
+\begin{enumerate}
+\item Prediction with the STE ephemeris (see Sec. \ref{Sec: integration}), and the nominal UCAC4 position of the star;
+\item Prediction with the JPL ephemeris and the nominal UCAC4 position of the star;
+\item From star and satellite offsets calculated from observations made a few days before the occultation when the objects were separated (different FOV);
+\item Same as 3 but with the star and the satellite close in the same FOV.
+\end{enumerate}
 
-Figure \ref{Fig: occ-Himalia} shows four zoomed in occultation maps corresponding to each of the four situations for Himalia..
-The map \ref{Fig: occ-Himalia-nooff} is the nominal prediction with the coordinate of the star given by the catalogue and that of the satellite from the ephemeris.
-We then corrected the positions of the satellite by an offset calculated by the method showed in section \ref{Sec: integration} to make the map \ref{Fig: occ-Himalia-ajuste}.
+Figure \ref{Fig: occ-Himalia} shows four zoomed in occultation maps corresponding to each of the four situations for Himalia.
+The map \ref{Fig: occ-Himalia-STE} is for the situation 1 in the list above.
+Map \ref{Fig: occ-Himalia-JPL} using JPL ephemeris (situation 2).
 The map \ref{Fig: occ-Himalia-off22fev} was made from obtained positions on February 22 observed with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / px$) at the Observatório do Pico dos Dias (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
@@
 \begin{centering}
-\subfigure[Prediction with STE ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_STE.eps}  \label{Fig: occ-Himalia-nooff}}
-\subfigure[Prediction with JPL ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_JPL.eps}  \label{Fig: occ-Himalia-ajuste}}
+\subfigure[Prediction with STE ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_STE.eps}  \label{Fig: occ-Himalia-STE}}
+\subfigure[Prediction with JPL ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_JPL.eps}  \label{Fig: occ-Himalia-JPL}}
 \subfigure[Observation at February 22 (different FOVs)]{\includegraphics[width=8.8cm]{figures/Himalia_22fev.eps}   \label{Fig: occ-Himalia-off22fev}}
 \subfigure[Observation at March 03 (same FOV)]{\includegraphics[width=8.8cm]{figures/Himalia_03mar.eps}   \label{Fig: occ-Himalia-off03mar}}
-\caption{Predictions for Himalia: The big red dot show the geocentric closest approach of the shadow.
-The black "+" marks at maps (b,c,d) are the nominal prediction closest approach for reference.
+\caption{Predictions for Himalia: The straight lines show the size of the shadow.
+The big red dot show the geocentric closest approach of the shadow.
+The black "+" marks at maps (b,c,d) are the STE prediction closest approach for reference.
 The small red ones are the center of the shadow separated by one minute.
-The straight lines show the size of the shadow.
-\ref{Fig: occ-Himalia-nooff} is the map using the nominal positions of the star and satellite.
-\ref{Fig: occ-Himalia-ajuste} shows the shadow given an estimated offset for the position of Himalia related to the JPL ephemeris obtained in section \ref{Sec: integration}.
+\ref{Fig: occ-Himalia-STE} is the map using the STE ephemeris showed in Sec.
+\ref{Sec: integration}.
+\ref{Fig: occ-Himalia-JPL} shows the shadow using the JPL ephemeris.
 In \ref{Fig: occ-Himalia-off22fev} we apply offsets to the positions of star and satellite accordingly the observations made at February 22.
 \ref{Fig: occ-Himalia-off03mar} is as in \ref{Fig: occ-Himalia-off22fev} but with observations made at March 03 when the objects were close to each other.
-The green dot at the bottom of the maps is the location of Observatório do Pico dos Dias.
-\textcolor{red}{Figure b waiting for the offset of Himalia}}
+The green dot at the bottom of the maps is the location of Observatório do Pico dos Dias.}
 \label{Fig: occ-Himalia}
@@
 In this event, it is possible to see that the shade does not vary by much among the four maps suggesting that there was a good probability of observing the event.
-In fact, the largest differences between the shadows of the four maps were \textcolor{red}{22s} along the shadow path and \textcolor{red}{109km} in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
+In fact, the largest differences between the shadows of the four maps were 36s in time along the shadow path and 101km in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
 
@@
 \begin{centering}
-\begin{tabular}{lcc}
+\begin{tabular}{lccc}
 \hline  \hline
-\multicolumn{3}{c}{Difference from STE prediction} \tabularnewline
-Method  & Instant of C/A  & C/A   \tabularnewline
+\multicolumn{4}{c}{Difference from STE prediction} \tabularnewline
+Method  & Instant of C/A*  & C/A & Map   \tabularnewline
 \hline
-STE & 00:39:51 UTC & $0\farcs703$ \tabularnewline
-JPL & -26s & +11mas (36km) \tabularnewline % (284km)
-Feb. 22 Obs. & -14s & -20mas (65km) \tabularnewline % (153km)
-Mar. 03 Obs. & -36s & -09mas (29km)\tabularnewline % (393km)
+STE & 00:39:51 UTC & $0\farcs703$ & \ref{Fig: occ-Himalia-STE} \tabularnewline
+JPL & -26s & +11mas (36km) & \ref{Fig: occ-Himalia-JPL} \tabularnewline % (284km)
+Feb. 22 Obs. & -14s & -20mas (65km) & \ref{Fig: occ-Himalia-off22fev} \tabularnewline % (153km)
+Mar. 03 Obs. & -36s & -09mas (29km) & \ref{Fig: occ-Himalia-off03mar} \tabularnewline % (393km)
 \hline
@@
 \par\end{centering}
-
+*Closest Approach
 \end{table}
@@
 The observations were made on March 25 and April 2, 2015 with the Zeiss telescope.
-On the night of April 2 they could still be observed in the same FOV.
+On the night of April 2 the star and satellite could still be observed in the same FOV.
 Due to Elara being fainter than Himalia by 2 magnitudes, dispersions of the satellite positions on both nights ended up being higher than for Himalia.
 Still, the differences between the maps obtained were relatively small.
-The largest differences between them were \textcolor{red}{74s} along the shadow path and \textcolor{red}{307km} perpendicular to it (see table \ref{Tab: comparison-Elara}).
+The largest differences between them were \textcolor{red}{74s} in time along the shadow path and \textcolor{red}{307km} perpendicular to it (see table \ref{Tab: comparison-Elara}).
 
@@
 \ref{Sec: predictions}).
-
-A total of \textcolor{red}{588} events were foreseen. In table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite.
-
-\begin{table}
-\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from 2015 up to 2017.}
-\begin{centering}
-\begin{tabular}{lcccc}
-\hline  \hline
-%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
-Satellite  & 2015  & 2016 & 2017 & Total \tabularnewline
-\hline
-Ananke & 22 & 12 & 16 & 50 \tabularnewline
-Carme & 25 & 20 & 14 & 59 \tabularnewline
-Elara & 32 & 14 & 16 & 62 \tabularnewline
-Himalia & 25 & 15 & 12 & 52 \tabularnewline
-Lysithea & 22 & 16 & 11 & 49 \tabularnewline
-Pasiphae & 26 & 19 & 20 & 65 \tabularnewline
-Sinope & 19 & 15 & 21 & 55 \tabularnewline
-\hdashline
-Phoebe & 13 & 32 & 98 & 143 \tabularnewline
-\hdashline
-Nereid & 41 & 11 & 1 & 53 \tabularnewline
-Triton &  &  &  &  \tabularnewline
-\hline
-\end{tabular}
-\par \end{centering}
-Occultations predicted using the UCAC4 catalogue.
-For Nereid and Triton, the occultations in 2015 and 2016 were predicted using the WFI catalogue as explained in Sec.
-\ref{Sec: predictions} \textcolor{red}{Numbers of Triton yet to be written}.
-\end{table}
+From this, a total of \textcolor{red}{588} events were foreseen.
 
@@
 
-The tests made with an occultation expected to happen in March 03, 2015 for Himalia showed that this event could probably have been observed in case there were observers available in the shadow area.
-Fig.
+The tests made with an occultation expected to happen in March 03, 2015 for Himalia showed that this event would probably have been observed successfully in case there were observers available in the shadow area.
+%Fig.
 \ref{Fig: occ-Himalia} shows the maps made for 4 different offset corrections as explained in Sec.
@@
 Table \ref{Tab: comparison-Himalia} shows numerically the difference between the maps.
-A similar test was made for an occultation of Elara in March 30, 2015.
-The difference between the maps in this case are listed in Table \ref{Tab: comparison-Elara}.
+A similar test was made for an occultation of Elara in March 30, 2015. %The difference between the maps in this case are listed in Table \ref{Tab: comparison-Elara}.
+The results show satisfying small offsets. We can expect similar results for other satellites and our method should allow successful stellar occultation.
 
-Continuous observations of the satellites and candidates are recommended and new numerical integrations of the orbits of the satellites are expected to reduce the respective position errors.
+Continuous observations of the satellites and candidates are recommended and fitting of our dynamical model to those observations are expected to reduce the respective position errors.
 The GAIA catalogue, which is to be released at the end of 2016 will improve the position error of the stars and will allow for the detection of occultations candidates of fainter stars, which are not present in the UCAC4 catalogue.
-Consequently, the astrometry of the satellites with the GAIA catalogue will improve the position errors allowing for better predictions and a higher probabilities of observing stellar occultations by irregular satellites.
+% Consequently, the astrometry of the satellites with the GAIA catalogue will improve the position errors allowing for better predictions and a higher probabilities of observing stellar occultations by irregular satellites.
+The release of the GAIA catalogue should have a positive impact on both the astrometric precision of occulted stars and the reduction of astrometric positions of occulting satellites.
+As a result, stellar occultations by irregular satellites should increase in number as well as success.
 

### 2015-11-18 Irregular-occ/Irregular-occ.tex
.

@@
 
-\title{Stellar occultation predictions for  9 irregular satellites of giant planets and Triton for 2015-2017
+\title{Stellar occultation predictions for  10 irregular satellites of giant planets and Triton for 2015-2017
 %\fnmsep\thanks{Table 8 are only available in electronic form at the CDS via anonymous ftp to cdsarc.u-strasbg.fr (130.79.128.5) or via http://cdsweb.u-strasbg.fr/cgi-bin/qcat?J/A+A/ and IAU NSDC data base at www.imcce.fr/nsdc.}
@@
 For Phoebe, Nereid and Triton we used the JPL ephemeris.}
-{We managed to obtain \textcolor{red}{588} candidates of stellar occultations between the period of January, 2015 and December, 2017.
+{We managed to identify 687 candidates of stellar occultations between the period of January, 2015 and December, 2017.
 We made observational tests for the predictions of the events of Himalia in March 03, 2015 and Elara in March 30, 2015.
-The stars and objects were observed close to the date predicted and and in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions.}
+The stars and objects were observed close to the date predicted and in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions.}
 {The comparisons between the predictions and the observation tests show a good agreement.
@@
 Himalia & 41 & $(150\times120) \pm 20$\tablefootmark{b} & 2 \tabularnewline
-Leda &  &  & 1 \tabularnewline
+Leda & 5 & 20 & 1 \tabularnewline
 Lysithea & 10 & 36 & 1 \tabularnewline
@@
 
-The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}, while the residuals are presented in Fig \ref{omcHim} to \ref{}. %don't remember wich one
+The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}, while the residuals are presented in Fig \ref{Fig: residuals} for Himalia and Carme.
+
+\begin{figure*}
+\subfigure{\includegraphics[width=8.8cm]{Laurene/residuals/Himalia_resid.eps}   \label{Fig: residuals-Himalia}}
+\subfigure{\includegraphics[width=8.8cm]{Laurene/residuals/Carme_resid.eps}  \label{Fig: residuals-Carme}}
+\caption{Residuals of Himalia and Carme \label{Fig: residuals}}
+\end{figure*}
 
@@
 \begin{table*}
-\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0.
-For Themisto's case, the small number of observations meant that the statistical uncertainty was even greater than the obtained elements, though the fitting proved satisfying.
+\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0.% For Themisto's case, the small number of observations meant that the statistical uncertainty was even greater than the obtained elements, though the fitting proved satisfying.
 We decided to provide its elements with significant number comparable to the other satellite for information, but left it without uncertainty.
@@
 
-\textcolor{red}{Here we write something about the numerical integration}
-
 \section{Prediction of occultations} \label{Sec: predictions}
@@
 
-A total of 588 events \textcolor{red}{(Not counting Triton's event yet)} were identified between January 2015 and December 2017.
+A total of 687 events were identified between January 2015 and December 2017.
 In table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite.
@@
 \hline
-Ananke & 22 & 12 & 16 & 50 \tabularnewline
-Carme & 25 & 20 & 14 & 59 \tabularnewline
+Ananke & 21 & 12 & 16 & 49 \tabularnewline
+Carme & 26 & 20 & 14 & 60 \tabularnewline
 Elara & 32 & 14 & 16 & 62 \tabularnewline
-Himalia & 25 & 15 & 12 & 52 \tabularnewline
+Himalia & 26 & 15 & 12 & 53 \tabularnewline
+Leda & 13 & 8 & 24 & 45 \tabularnewline
 Lysithea & 22 & 16 & 11 & 49 \tabularnewline
-Pasiphae & 26 & 19 & 20 & 65 \tabularnewline
+Pasiphae & 25 & 20 & 19 & 64 \tabularnewline
 Sinope & 19 & 15 & 21 & 55 \tabularnewline
-Leda &  &  &  &  \tabularnewline
 \hdashline
@@
 Nereid\tablefootmark{a} & 41\tablefootmark{b} & 11\tablefootmark{b} & 1 & 53 \tabularnewline
-Triton\tablefootmark{a} & \tablefootmark{b} & \tablefootmark{b} &  &  \tabularnewline
+Triton\tablefootmark{a} & 37\tablefootmark{b} & 16\tablefootmark{b} & 1 & 54 \tabularnewline
 \hline
@@
 \tablefoottext{a}{Using JPL ephemeris.}
-\tablefoottext{b} {Using the WFI catalogue as explained in Sec. \ref{Sec: predictions}} \textcolor{red}{Numbers of Triton yet to be written}.
+\tablefoottext{b} {Using the WFI catalogue as explained in Sec. \ref{Sec: predictions}}.
 \end{table}
@@
 \begin{centering}
-\begin{tabular}{ccccccccccccccccccccccc}
+\begin{tabular}{l}
 \hline 
@@
 % & \multicolumn{2}{c}{Mean errors} & Nr & Nr &  UCAC4       \tabularnewline
-d m Year & h m s & \multicolumn{2}{c}{RA (ICRS) Dec} & C/A & P/A & v & D & R* & $\lambda$ & LST & $\Delta e_{\alpha *}$ & $\Delta e_{\delta}$ & pm & ct & $E_{\alpha *}$ & $E_{\delta}$ & $\mu_{\alpha *}$ & $\mu_{\delta}$ \tabularnewline
-\hline
- 07 10 2015 & 21 52 34. &  10 53 40.1835 & +08 21  00.489 & 0.465 & 200.31 & 39.31 & 6.25 & 16.6 & 179. & 09:48 & -31.0 & -13.0 & ok & uc &   0  &  0 & -13 &  -7 \tabularnewline
- 23 10 2015 & 07 54 50. &  11 05 26.6037 & +07 14  32.030 & 1.017 & 201.33  & 36.11 & 6.12 & 15.5 & 16. & 08:59 & -31.0 & -13.0 & ok & uc &  0  &  0 &  -7 &  -4 \tabularnewline
- 31 10 2015 & 00 24 19. &  11 10 56.2964 & +06 42  07.110 &  1.801 &  21.84 & 33.94 &  6.04 & 16.9 &  123. & 08:34 & -31.0 & -13.0 & ok & uc & 0  &  0 & -4 & -13 \tabularnewline
- 07 11 2015 &  21 22 53. &  11 16 14.3041 & +06 10  05.194 &  0.306 &  202.36 & 31.33 & 5.94 & 16.2 & 162. & 08:09 & -31.0 & -13.0 & ok & uc & 0 & 0 & 0 & 0 \tabularnewline
- 08 11 2015 & 04 27 21. &  11 16 25.7881 & +06 08  55.566 & 1.077 & 202.38 & 31.23 & 5.94 & 16.9 & 55. & 08:08 & -31.0 & -13.0 & ok & uc & 0 & 0 & 9 & -10 \tabularnewline
-% 09 11 2015  09 01  4.
-11 17 11.6890 +06 04  11.474   11 17 11.7030 +06 04  11.980   0.549   22.45  30.79  5.92 15.9  0.5  0.5  0.5   346.
-08:04     -31.0    -13.0 ok uc 0    0    0   19  -13 \tabularnewline
-% 11 11 2015  12 58 46.
-11 18 33.8286 +05 55  43.695   11 18 33.8357 +05 55  43.946   0.274   22.59  29.98  5.90 14.6  0.4  0.4  0.4   285.
-07:57     -31.0    -13.0 ok uc 0    0    0  -14    1 \tabularnewline
-% 12 11 2015  20 14 22.
-11 19 22.2740 +05 50  40.732   11 19 22.3183 +05 50  42.312   1.713   22.68  29.48  5.88 15.7  0.4  0.4  0.4   175.
-07:52     -31.0    -13.0 ok uc 0    0    0   -7   -4 \tabularnewline
-% 04 12 2015  03 31 56.
-11 30 41.9967 +04 37  36.039   11 30 42.0020 +04 37  36.217   0.196   24.18  20.14  5.56 15.0  0.0  0.0  0.0    47.
-06:40     -31.0    -13.0 ok uc 0    0    0    5   -9 \tabularnewline
+
 \hline
@@
 \includegraphics[scale=0.4]{figures/Pasiphae_2016-04-09T03:58:19.png}   
-\caption{Occultation map for Himalia regarding to the second event sampled in Table \ref{Tab: occ-list}}
+\caption{Occultation map for Himalia regarding to the fifth event sampled in Table \ref{Tab: occ-list}}
 \label{Fig: ocultacao}

### 2015-11-18 Irregular-occ/Laurene/residuals/grafico_resid.py

@@ -1,0 +1,51 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from astropy.time import Time, TimeDelta
+import astropy.units as u
+import os
+
+######################################################################3
+
+plt.rcParams['text.latex.preamble']=[r"\usepackage{txfonts}"]
+
+params = {'text.usetex' : True,
+          'font.size' : 22,
+          'font.family' : 'txfonts',
+          'text.latex.unicode': True,
+          }
+          
+plt.rcParams.update(params)
+
+obj = ['Ananke','Carme','Elara','Himalia','Leda','Lysithea','Pasiphae','Sinope']
+nobj= ['omc012.dat', 'omc011.dat', 'omc007.dat', 'omc006.dat', 'omc013.dat', 'omc010.dat', 'omc008.dat', 'omc009.dat']
+
+sizel = 17
+
+r = []
+for i in np.arange(1995,2016,2):
+    r.append(Time('{}-01-01 00:00:00'.format(i), format='iso').jd - 2451544.5)
+r = np.array(r)
+
+for t in np.arange(len(obj)):
+    dados = np.loadtxt(nobj[t], usecols=[0,1,2], unpack=True)
+    time = Time(dados[0], format='jd')
+    alfa = dados[1]*u.rad
+    dec = dados[2]*u.rad
+    plt.plot(time.jd - 2451544.5,alfa.to(u.mas), '+', markersize=5, label='RA')
+    plt.plot(time.jd - 2451544.5,dec.to(u.mas), 'x', markersize=5, label='DEC')
+    plt.title('{}'.format(obj[t]), fontsize=sizel)
+    plt.xlim(Time('1995-01-01 00:00:00', format='iso').jd - 2451544.5,Time('2016-01-01 00:00:00', format='iso').jd - 2451544.5)
+    plt.ylim(-400,400)
+    plt.xlabel('Time', fontsize=sizel)
+    plt.xticks(r, ['{}'.format(i) for i in np.arange(1995,2016,2)])
+    plt.ylabel('Residuals (mas)', fontsize=sizel)
+    plt.legend(numpoints=1, labelspacing=0.25, borderpad=0.5, handlelength=0.1, prop={'size':sizel})
+    plt.axhline(0, color='black')
+    fig =plt.gcf()
+    fig.set_size_inches(6.93,3.9)
+    plt.tick_params(axis='both', which='major', labelsize=sizel)
[…]

### 2015-11-18 Irregular-occ/dados/tabela.py

@@ -1,0 +1,65 @@
+import numpy as np
+from astropy.coordinates import SkyCoord
+from astropy.time import Time
+import astropy.units as u
+from astropy.table import Table, Column, vstack
+import matplotlib.pyplot as plt
+import matplotlib.image as mpimg
+
+f = open('lista_tables', 'r')
+lista = f.readlines()
+f.close()
+
+tabela = Table(names=('object', 'instantes', 'RA (ICRS) DEC', 'C/A', 'P/A', 'vel', 'distance', 'magR', 'magK', 'long', 'caminho'),\
+    dtype=(np.str, np.str, np.str, np.float, np.float, np.float, np.float, np.float, np.float, np.float, np.str))
+    
+def texto(i):
+    if i in ['Triton', 'Nereida', 'Phoebe']:
+        return '/JPL/'
+    return '/LAU/'
+    
+def mostrar(i):
+    imagem = tabela[np.where(tabela['numero'] == i)]
+    imagem.pprint(max_width=-1)
+    img = mpimg.imread(imagem['caminho'][0])
+    plt.imshow(img)
+    plt.show()
+
+for i in lista:
+    arquivo = i.strip()
+    dados = np.loadtxt(arquivo, skiprows=41, usecols=(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 18, 19, 20, 21, 22, 25, 26, 28, 29), \
+        dtype={'names': ('dia', 'mes', 'ano', 'hor', 'min', 'sec', 'afh', 'afm', 'afs', 'ded', 'dem', 'des', 'ca', 'pa', 'vel', 'delta', 'mR', 'mK', 'long', 'ora', 'ode'), \
+       'formats': ('S30', 'S30', 'S30','S30', 'S30', 'S30','S20', 'S20', 'S20','S20', 'S20', 'S20', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8')}, ndmin=1)
+################## lendo coordenadas #################
+    coor = dados['afh']
+    for i in ['afm', 'afs', 'ded', 'dem', 'des']:
+        coor = np.core.defchararray.add(coor, ' ')
+        coor = np.core.defchararray.add(coor, dados[i])
+    stars = SkyCoord(coor, frame='fk5', unit=(u.hourangle, u.degree))
+    star_coord = Column(stars.to_string('hmsdms', precision=4, sep=' '), name='RA (ICRS) DEC')
+################### lendo tempo ########################
[…]

### 2015-11-19 Irregular-occ/Irregular-occ.tex
.

@@
 \begin{figure*}
-\subfigure{\includegraphics[width=8.8cm]{Laurene/residuals/Himalia_resid.eps}   \label{Fig: residuals-Himalia}}
-\subfigure{\includegraphics[width=8.8cm]{Laurene/residuals/Carme_resid.eps}  \label{Fig: residuals-Carme}}
+\subfigure{\includegraphics[width=8.8cm]{figures/Himalia_resid.eps}   \label{Fig: residuals-Himalia}}
+\subfigure{\includegraphics[width=8.8cm]{figures/Carme_resid.eps}  \label{Fig: residuals-Carme}}
 \caption{Residuals of Himalia and Carme \label{Fig: residuals}}
@@
 \begin{table*}
-\caption{\label{Tab: sample-cds} Stellar occultation sampled predictions for Himalia \textcolor{red}{I still have to fix the table}}
+\caption{\label{Tab: sample-cds} Stellar occultation sampled predictions for Pasiphae}
 \begin{centering}
-\begin{tabular}{l}
+\begin{tabular}{cccrcccrcrr}
 \hline
-%\multicolumn{9}{c}{Himalia} \tabularnewline
+\hline
+d m Year~~~~h~~m~~~s & RA~~~(ICRS)~~~Dec & C/A & P/A & $\nu$ & {\it D} & R* & $\lambda$ & LST & $\mu_{\alpha*}$ & $\mu_{\delta}$ \tabularnewline
 \hline
 % & \multicolumn{2}{c}{Mean errors} & Nr & Nr &  UCAC4       \tabularnewline
-
+09 04 2016 03:58:19. & 11 14 36.7707 +07 39 20.7610 & 1.003 &  17.9 & -12.88 &  4.54 & 14.9 & 271. & 22:03 &  12. & -33. \tabularnewline 
+13 06 2016 00:16:12. & 11 12 48.5020 +07 06 43.3520 & 0.661 &  30.0 & +14.32 &  5.50 & 13.9 & 262. & 17:45 &  -1. &   1. \tabularnewline 
+27 06 2016 13:56:09. & 11 18 03.4160 +06 23 45.1940 & 1.707 &  28.0 & +20.29 &  5.74 & 11.7 &  44. & 16:53 &   4. & -10. \tabularnewline 
+18 07 2016 15:07:24. & 11 28 15.5076 +05 05 31.8060 & 0.942 &  26.7 & +27.80 &  6.05 & 14.0 &   8. & 15:40 &   4. &   4. \tabularnewline 
+22 07 2016 16:15:07. & 11 30 30.4310 +04 48 43.4340 & 0.644 & 206.5 & +29.04 &  6.11 & 14.6 & 348. & 15:27 &  23. & -24. \tabularnewline 
+24 07 2016 01:37:34. & 11 31 17.8471 +04 42 49.0540 & 0.029 & 206.6 & +29.46 &  6.12 & 15.1 & 206. & 15:22 &   2. &  -8. \tabularnewline 
+24 07 2016 17:37:18. & 11 31 40.7472 +04 39 57.5060 & 0.840 &  26.5 & +29.66 &  6.13 & 14.9 & 326. & 15:20 & -11. &  -1. \tabularnewline 
 \hline
@@
 \textbf{Notes}.
-Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - in the original table, these coordinates are immediately followed by the geocentric astrometric equatorial coordinates of the satellite (corrected the computed ephemeris offset derived in Sec.
-\ref{Sec: integration}); C/A: the geocentric closest approach, in arcseconds; P/A: the planet position angle with respect to the occulted star at C/A, in degrees; velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; D: planet range to Earth, in AU; R*: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{Mag}* = \textrm{Mag}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$.
-A value of 50.0 means that the star is not in the 2MASS; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\Delta e_{\alpha *}$ and $\Delta e_{\delta}$: computed offsets in mas applied to the ephemeris right ascension and declination, respectively; pm: ok = proper motion applied, no = no proper motion applied; catalogue cross-identification (ct) = uc (UCAC2), 2m (2MASS), fs (field star); $E_{\alpha *}$ and $E_{\delta}$: uncertainties (mas) in right ascension and declination.
-A value of 9999 means that there was no estimation of the respective uncertainty; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
+Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - in the original table; C/A: the geocentric closest approach, in arcseconds; P/A: the planet position angle with respect to the occulted star at C/A, in degrees; $\nu$: velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; {\it D}: planet range to Earth, in AU; R*: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
 \label{Tab: occ-list}
@@
 \includegraphics[scale=0.4]{figures/Pasiphae_2016-04-09T03:58:19.png}   
-\caption{Occultation map for Himalia regarding to the fifth event sampled in Table \ref{Tab: occ-list}}
+\caption{Occultation map for Himalia regarding to the first event sampled in Table \ref{Tab: occ-list}}
 \label{Fig: ocultacao}
@@
 \ref{Sec: predictions}).
-From this, a total of \textcolor{red}{588} events were foreseen.
+From this, a total of 687 events were foreseen.
 

### 2015-11-19 Irregular-occ/Laurene/residuals/grafico_resid.py

@@ -32,7 +32,7 @@ for t in np.arange(len(obj)):
     alfa = dados[1]*u.rad
     dec = dados[2]*u.rad
     plt.plot(time.jd - 2451544.5,alfa.to(u.mas), '+', markersize=5, label='RA')
-    plt.plot(time.jd - 2451544.5,dec.to(u.mas), 'x', markersize=5, label='DEC')
+    plt.plot(time.jd - 2451544.5,dec.to(u.mas), 'x', markersize=5, label='Dec')
     plt.title('{}'.format(obj[t]), fontsize=sizel)
     plt.xlim(Time('1995-01-01 00:00:00', format='iso').jd - 2451544.5,Time('2016-01-01 00:00:00', format='iso').jd - 2451544.5)
     plt.ylim(-400,400)

### 2015-11-19 Irregular-occ/Irregular-occ.tex
.

@@
 The best ground-based technique to constrain those parameters is the observation of stellar occultations by these objects.}
-{We aim to predict stellar occultations for the seven biggest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda; the irregular satellite of Saturn: Phoebe; and the satellites of Neptune: Nereid and Triton.}
+{We aim to predict stellar occultations for the eight biggest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda; the irregular satellite of Saturn: Phoebe; and the satellites of Neptune: Nereid and Triton.}
 {We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument.
@@
 For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds).
-Thus, in this case, the cumulated uncertainty of ephemeris and star position1 must be around 40 mas to get good chances of observing a stellar occultation, which is quite feasible today.
+Thus, in this case, the cumulated uncertainty of ephemeris and star position must be around 40 mas to get good chances of observing a stellar occultation, which is quite feasible today.
 
@@
 
-From what we have just presented, we can think that the obsevation of an stellar occultation by an irregular satellite should be likelier than occultations by TNOs..
+From what we have just presented, we can think that the observation of an stellar occultation by an irregular satellite should be likelier than occultations by TNOs.
 The orbits of their host planets are well known and their observation time-span covers many orbital periods, contrary to TNOs.
@@
 The orbits of their host planets are well known and these satellites have been observed completing already many orbital periods around them, thus presenting better ephemeris than TNOs.
-Moreover, the irregular satellites are closer to Earth which means a minor localization error in kilometers.
+Moreover, the irregular satellites are closer to Earth which means a minor localization error in kilometer.
 
@@
 
-We present in this paper stellar occultation predictions for the 7 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke and Sinope), Phoebe from Saturn and Nereid and Triton from Neptune.
+We present in this paper stellar occultation predictions for the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda), Phoebe from Saturn and Nereid and Triton from Neptune.
 In the section \ref{Sec: Rationale} we explore the scientific rationale for the study of the irregular satellites and the possibility of having a common origin with TNOs.
-In section \ref{Sec: integration} we show the the process of th productin of the new ephemeris.
+In section \ref{Sec: integration} we show the the process of the production of the new ephemeris.
 In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites and how they were made.
@@
 
+\section*{Introduction V2}
+
+Irregular satellites revolve around giant planets at large distances, on eccentric, highly inclined and frequently retrograde orbits.
+Because of these peculiar orbits, it is largely accepted that these objects did not form by accretion around of their planet, but were captured by their planets in the early solar system \citep{Sheppard2005}.
+
+There is no consensus for a single model explaining where the irregular satellites were formed.
+\cite{Cuk2004} showed that the progenitor of the Himalia group may have originated in heliocentric orbits similar to the Hilda asteroid group.
+\cite{Sheppard2005} stated that the irregular satellites may be some of the objects that were formed within the giant planets region.
+
+\cite{Grav2003} and \cite{Grav2007} showed that the irregular satellites from the giant planets have their colors and spectral slopes similar to C-, D- and P-type asteroids, Centaurs and trans-neptunian objects (TNOs).
+This suggests that they may have come from different locations in the early solar system.
+
+\cite{Sheppard2005} and \cite{Jewitt2007} also expose the possibility for the irregular satellites having their origin as comets or TNOs.
+TNOs are highly interesting objects that, due to their large heliocentric distances, may be highly preserved having their physical properties similar to those they had when they were formed \citep{Camargo2014}.
+This is even more true for the smaller objects, since in principle larger sizes favour physical differentiation processes in the body and vice-versa.
+However, due to the distance, the smaller TNOs from this region are more difficult to observe.
+A clever way to overcome this difficulty is to study  the much closer irregular satellites, under the hypothesis that they share a common origin with the small TNOs' population.
+
+Phoebe is the most studied irregular satellite.
+\cite{Clark2005} suggest that its surface is probably covered by material of cometary origin.
+It was also stated by \cite{Johnson2005} that if the porosity of Phoebe is 15\%, Phoebe would have an uncompressed density similar to those of Pluto and Triton.
+
+In order to obtain precise fundamental physical parameters like size, shape and albedo for the irregular satellites and thus to contribute to the study of their origin, we want to observe stellar occultations, which provide  more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
+
+We present in this paper stellar occultation predictions for the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda), Phoebe from Saturn and Nereid and Triton from Neptune.
+Phoebe, being the most studied object with a good measured size, can be used to calibrate and evaluate the technique for similar objects.
+
+Triton is a uncommon satellite.
+Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
+Because Triton's orbit size is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded from the irregular satellites' class, but studied together by many authors \citep{Sheppard2005, Jewitt2007}.
+
+Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have the same origin as the TNOs \citep{Agnor2006}.
+Differently, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
+The main motivation to study Triton by stellar occultations is to understand the evolution of its atmosphere due to Triton's complicated and extreme seasonal cycle \citep{McKinnon2007}.
+
+No observation of a stellar occultation by an irregular satellite was published up to date.
+Since their estimated sizes are very small (see Table \ref{Tab: satellite-diameter}), this may have prevented earlier tries.
+But, in fact, given their distances to us, current ephemeris and star positions are already good enough for the prediction of the exact location and instant where the shadow of the occultation will cross the Earth.
+For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds).
+Thus, in this case, the cumulated uncertainty of ephemeris and star position must be around 40 mas to get good chances of observing a stellar occultation, which is quite feasible today.
+
+\begin{table}
+\caption{\label{Tab: satellite-diameter} Estimated diameter of the satellites and correspondent apparent diameter}
+\begin{centering}
+\begin{tabular}{lccc}
+\hline  \hline
+\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
+Satellite  & mas\tablefootmark{a}  & km & References \tabularnewline
+\hline
+Ananke & 8 & 29 & 1 \tabularnewline
+Carme & 13 & 46 & 1 \tabularnewline
+Elara & 24 & 86 & 1 \tabularnewline
+Himalia & 41 & $(150\times120) \pm 20$\tablefootmark{b} & 2 \tabularnewline
+Leda & 5 & 20 & 1 \tabularnewline
+Lysithea & 10 & 36 & 1 \tabularnewline
+Pasiphae & 17 & 62 & 1 \tabularnewline
+Sinope & 10 & 37 & 1 \tabularnewline
+\hdashline
+Phoebe & 32 & $212 \pm 1.4$\tablefootmark{b} & 3 \tabularnewline
+\hdashline
+Nereid & 15 & $340 \pm 50$\tablefootmark{c} & 4 \tabularnewline
+Triton & 124 & $2707 \pm 2.0$\tablefootmark{c} & 5 \tabularnewline
+\hline
+\end{tabular}
+\tablebib{
+(1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}; (4) \cite{Thomas1991}; (5) \cite{Thomas2000}.}
+\end{centering}
+\tablefoottext{a}{Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
+\tablefoottext{b}{From Cassini observation.}
+\tablefoottext{c}{From Voyager-2 observation.}
+\par
+\end{table}
+
+\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas.
+For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}.
+They pointed out that the ephemeris of the irregular satellites have systematic errors that may reach 200 mas for some satellites.
+For an object at the distance of Jupiter, this represents an error bigger than 700 km in the shadow path.
+Using the positions obtained by \cite{GomesJunior2015} we produced a specific ephemeris for short-time future for the satellites of Jupiter and better predict stellar occultations for these objects.
+
+From what we have just presented, we can think that the observation of an stellar occultation by an irregular satellite should be likelier than occultations by TNOs.
+The orbits of their host planets are well known and their observation time-span covers many orbital periods, contrary to TNOs.
+%Unlike stellar occultations by TNOs, which nevertheless have been proved effective, the observation of an occultation by a irregular satellite is, in principle, more favorable.
+The orbits of their host planets are well known and these satellites have been observed completing already many orbital periods around them, thus presenting better ephemeris than TNOs.
+Moreover, the irregular satellites are closer to Earth which means a minor localization error in kilometer.
+
+In section \ref{Sec: integration} we show the the process of the production of the new ephemeris.
+In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites and how they were made.
+Some tests made to confirm the predictions are presented in section \ref{Sec: testes} and the final discussion are in section \ref{Sec: discussion}.
+
 \section{Special tailored Ephemerides (STE) for Jupiter irregular satellites} \label{Sec: integration}
@@
 %They showed that the orbits of the irregular satellites of the giant planets have systematic errors.
-The offsets of the observations relative to the JPL ephemeris could be up to 200 mas for some satellites. %These differences could be associated with errors in their orbital elements.
+Their offsets of the observations relative to the JPL ephemeris could be up to 200 mas for some satellites.
+At the distance of Jupiter, this represents an difference bigger than 700 km in the shadow path of an occultation.%These differences could be associated with errors in their orbital elements.
 
@@
 \end{equation}
-
 where the oblateness potential seen by the body $i$ because of Jupiter is:
@@
 
-The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}, while the residuals are presented in Fig \ref{Fig: residuals} for Himalia and Carme.
+The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}, while the residuals are presented in Fig \ref{Fig: residuals} for Himalia and Carme as representatives.
 
@@
 \subfigure{\includegraphics[width=8.8cm]{figures/Carme_resid.eps}  \label{Fig: residuals-Carme}}
-\caption{Residuals of Himalia and Carme \label{Fig: residuals}}
+\caption{Residuals of the numerical integration for Himalia and Carme. \label{Fig: residuals}}
 \end{figure*}
-
-All the orbits determined for the satellites show satisfying residuals.
-%Yet, one of the satellites has a peculiar situation.
-Its small number of observations in our set means that its orbit is definitely loosely constrained.
-The residuals are lower than those obtained with JPL ephemeris, but the accuracy of an ephemeris decreases when we get further from the time of observations.
-The main risk of divergence over time comes from the possible absence of long-term effects when fitting to a short timespan of observations.
-If that were the case, our ephemeris would diverge too quickly to be of any use.
-JPL ephemerides are fitted over all the available observations.
-As a result, they will diverge less quickly than our own.
-Though they are no longer precise enough for our use, they remain a precious reference to identify whether our own model presents a quick divergence.
-We compared our ephemeris to the JPL for all the Jupiter satellites we fitted, until 2017.
-The divergence between 2015 and 2018 is at most \textit{\color{red}insert value here} for \textit{\color{red}insert right satellite name here}.
-A example of this difference for satellites Himalia and \textit{\color{red}whichever satellite you want, if you want)} is presented in Figure \ref{JPL-STE}.
-
-The obtained ephemeris is hereafter refered as STE, for special-tailored ephemeris.
 
@@
 \end{table*}
+
+All the orbits determined for the satellites show satisfying residuals.
+%Yet, one of the satellites has a peculiar situation.
+Its small number of observations in our set means that its orbit is definitely loosely constrained.
+The residuals are lower than those obtained with JPL ephemeris, but the accuracy of an ephemeris decreases when we get further from the time of observations.
+The main risk of divergence over time comes from the possible absence of long-term effects when fitting to a short timespan of observations.
+If that were the case, our ephemeris would diverge too quickly to be of any use.
+JPL ephemerides are fitted over all the available observations.
+As a result, they will diverge less quickly than our own.
+Though they are no longer precise enough for our use, they remain a precious reference to identify whether our own model presents a quick divergence.
+We compared our ephemeris to the JPL for all the Jupiter satellites we fitted, until 2017.
+The divergence between 2015 and 2018 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.
+A example of this difference for satellites Himalia and Carme over time is presented in Figure \ref{Fig: JPL-STE}.
+
+\begin{figure*}
+\subfigure{\includegraphics[width=8.8cm]{figures/JPL_DEC_Himalia.eps}    \label{Fig: JPL-STE-Himalia}}
+\subfigure{\includegraphics[width=8.8cm]{figures/JPL_DEC_Carme.eps}  \label{Fig: JPL-STE-Carme}}
+\caption{Differences between the JUP300 JPL ephemeris and the special-tailored ephemeris for Himalia and Carme. \label{Fig: JPL-STE}}
+\end{figure*}
+
+The obtained ephemeris is hereafter refered as STE, for special-tailored ephemeris.
 
@@
 
-The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the special-tailored ephemeris for Jupiter satellites and JPL ephemeris for the others as presented in section \ref{Sec: integration}.
+The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the special-tailored ephemeris for Jupiter satellites (see Sec.
+\ref{Sec: integration}) and JPL ephemeris for the others.
 The search for stellar candidates follows the same procedure as presented by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.
 
-We predicted occultation for the 7 major irregular satellites of Jupiter:  Ananke, Carme, Elara, Himalia, Lysithea, Pasiphae, and Sinope; Phoebe of Saturn and Triton and Nereid of Neptune.
+We predicted occultation for the 8 major irregular satellites of Jupiter:  Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope; Phoebe of Saturn and Triton and Nereid of Neptune.
 
@@
 \hline
-d m Year~~~~h~~m~~~s & RA~~~(ICRS)~~~Dec & C/A & P/A & $\nu$ & {\it D} & R* & $\lambda$ & LST & $\mu_{\alpha*}$ & $\mu_{\delta}$ \tabularnewline
+d m Year~~~~h~~m~~~s & RA~~~(ICRS)~~~Dec & C/A & P/A & $\nu$ & {\it D} & $R^*$ & $\lambda$ & LST & $\mu_{\alpha*}$ & $\mu_{\delta}$ \tabularnewline
 \hline
@@
 \textbf{Notes}.
-Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - in the original table; C/A: the geocentric closest approach, in arcseconds; P/A: the planet position angle with respect to the occulted star at C/A, in degrees; $\nu$: velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; {\it D}: planet range to Earth, in AU; R*: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
+Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - in the original table; C/A: the geocentric closest approach, in arcseconds; P/A: the planet position angle with respect to the occulted star at C/A, in degrees; $\nu$: velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; {\it D}: planet range to Earth, in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
 \label{Tab: occ-list}
@@
 \includegraphics[scale=0.4]{figures/Pasiphae_2016-04-09T03:58:19.png}   
-\caption{Occultation map for Himalia regarding to the first event sampled in Table \ref{Tab: occ-list}}
+\caption{Occultation map for Pasiphae regarding to the first event sampled in Table \ref{Tab: occ-list}}
 \label{Fig: ocultacao}
@@
 \item Prediction with the JPL ephemeris and the nominal UCAC4 position of the star;
-\item From star and satellite offsets calculated from observations made a few days before the occultation when the objects were separated (different FOV);
+\item From star and satellite offsets calculated from observations made a few days before the occultation when the objects were separated (different FOVs);
 \item Same as 3 but with the star and the satellite close in the same FOV.
@@
 The map \ref{Fig: occ-Himalia-off22fev} was made from obtained positions on February 22 observed with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / px$) at the Observatório do Pico dos Dias (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
-On that day, Himalia and the star were observed in separate FOVs as they were still far apart.
+On that day, Himalia and the star were observed in separate FOVs as they were still far apart (situation 3).
 On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope (diameter = 1.6m; FOV = $5\farcm8$; pixel scale = $0\farcs17 / px$) at OPD just over an hour after the time scheduled for the event.
-Satellite and star were separated by about 16 arcsec, so very close to each other.
+Satellite and star were separated by about 16 arcsec, so very close to each other (situation 4).
 From the calculated offsets, the map \ref{Fig: occ-Himalia-off03mar} was generated.
-Notice that the shadow path was not predicted to cross the OPD.
+Notice that the shadow path was not predicted to cross the OPD (green point at the bottom of the maps).
 This was not necessary for testing the prediction.
@@
 In this event, it is possible to see that the shade does not vary by much among the four maps suggesting that there was a good probability of observing the event.
-In fact, the largest differences between the shadows of the four maps were 36s in time along the shadow path and 101km in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
+In fact, the largest differences between the shadows of the four maps were 36s, which correspond to 120 mas, in time along the shadow path and 101km (31 mas) in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
 
@@
 \multicolumn{4}{c}{Difference from STE prediction} \tabularnewline
-Method  & Instant of C/A*  & C/A & Map   \tabularnewline
+Method  & Instant of C/A  & C/A & Map   \tabularnewline
 \hline
@@
 \par\end{centering}
-*Closest Approach
+%*Closest Approach
 \end{table}
@@
 Due to Elara being fainter than Himalia by 2 magnitudes, dispersions of the satellite positions on both nights ended up being higher than for Himalia.
-Still, the differences between the maps obtained were relatively small.
-The largest differences between them were \textcolor{red}{74s} in time along the shadow path and \textcolor{red}{307km} perpendicular to it (see table \ref{Tab: comparison-Elara}).
+%Still, the differences between the maps obtained were relatively small.
+The largest differences between them were 144s (186 mas) in time along the shadow path and 525 km (153 mas) perpendicular to it (see table \ref{Tab: comparison-Elara}).
 
@@
 JPL & +02s & +57mas(195km) \tabularnewline
-Mar. 25 Obs. & s & mas (km) \tabularnewline
-Apr. 02 Obs. & s & mas (km) \tabularnewline
+Mar. 25 Obs. & -70s & +20mas (68km) \tabularnewline
+Apr. 02 Obs. & +74s & -96mas (330km) \tabularnewline
 \hline
@@
 
-We presented stellar occultations for the period of 2015-2017 for seven irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid.
+We presented stellar occultations for the period of 2015-2017 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid.
 The procedure used was the same as that for the prediction of stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.
@@
 The probability of successfully observing an occultation is roughly the ratio of the satellite's diameter by the budget error of ephemeris and star position.
-Thus, UCAC4 errors ranging between 20 mas - 50 mas combined with a mean accuracy in the position of 30 mas for Himalia and 100 mas for Ananke mas published in Table 2 of \cite{Jacobson2012} would give 80\%-50\% probability of observing such an event by Himalia and $\approx$5\% for Ananke, the smallest irregular satellite in the sample.
+Thus, UCAC4 errors ranging between 20 mas - 50 mas combined with a mean accuracy in the position of 30 mas for Himalia and 150 mas for Leda published in Table 2 of \cite{Jacobson2012} would give 80\%-50\% probability of observing such an event by Himalia and $\approx$3\% for Leda, the smallest irregular satellite in the sample.
 Observations a few days before the date of occultation predicted may improve the combined accuracy to 40-80 mas, depending on the magnitude of the objects.

### 2015-11-19 Irregular-occ/Laurene/residuals/grafico_resid.py

@@ -31,8 +31,8 @@ for t in np.arange(len(obj)):
     time = Time(dados[0], format='jd')
     alfa = dados[1]*u.rad
     dec = dados[2]*u.rad
-    plt.plot(time.jd - 2451544.5,alfa.to(u.mas), '+', markersize=5, label='RA')
-    plt.plot(time.jd - 2451544.5,dec.to(u.mas), 'x', markersize=5, label='Dec')
+    plt.plot(time.jd - 2451544.5,alfa.to(u.mas), '+', markersize=5, label=r'$\Delta\alpha\cos\delta$')
+    plt.plot(time.jd - 2451544.5,dec.to(u.mas), 'x', markersize=5, label=r'$\Delta\delta$')
     plt.title('{}'.format(obj[t]), fontsize=sizel)
     plt.xlim(Time('1995-01-01 00:00:00', format='iso').jd - 2451544.5,Time('2016-01-01 00:00:00', format='iso').jd - 2451544.5)
     plt.ylim(-400,400)

### 2015-11-23 Irregular-occ/Irregular-occ.tex
.

@@
           L. Beauvalet
-          \inst{2},
+          \inst{2,3},
           R. Vieira-Martins
-          \inst{1,2,3} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France},
+          \inst{1,2,4} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France},
 %          J.-E. Arlot
@@
           J. I. B. Camargo
-          \inst{2,3},     
+          \inst{2,4},     
           B. E. Morgado
@@
               \email{rvm@on.br}
+              \and
+Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France
               \and
@@
 JIBC acknowledges CNPq for a PQ2 fellowship (process number 308489/2013-6).
-FB-R acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014.
+%FB-R acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014.
 BEM thanks the financial support of CAPES.
+The numerical model of the satellites of Jupiter was developped during a post-doctoral contract funded by the Chinese Academy of Sciences (CAS) and supported by the National Scientific Fund of China (NSFC)
 

### 2015-11-26 Irregular-occ/Irregular-occ.tex
.

@@
 
-\title{Stellar occultation predictions for  10 irregular satellites of giant planets and Triton for 2015-2017
+\title{Stellar occultation predictions for  10 irregular satellites of giant planets plus Triton for 2015-2017
 %\fnmsep\thanks{Table 8 are only available in electronic form at the CDS via anonymous ftp to cdsarc.u-strasbg.fr (130.79.128.5) or via http://cdsweb.u-strasbg.fr/cgi-bin/qcat?J/A+A/ and IAU NSDC data base at www.imcce.fr/nsdc.}
@@
           \inst{1,2}
-%          F. Braga-Ribas
-%          \inst{2,4}, 
+          F. Braga-Ribas
+          \inst{2,5}, 
 %          D. N. da Silva Neto
@@
 {Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system.
-It is important to know their physical parameters, such as size, albedo and composition, to trace back their true origin.
-The best ground-based technique to constrain those parameters is the observation of stellar occultations by these objects.}
-{We aim to predict stellar occultations for the eight biggest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda; the irregular satellite of Saturn: Phoebe; and the satellites of Neptune: Nereid and Triton.}
+It is important to know their physical parameters, such as size, shape, albedo and composition, to trace back their true origin.
+The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects.}
+{We aim to predict stellar occultations for the eight biggest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda, and for the irregular satellites Phoebe of Saturn and Nereid of Neptune, and also for Triton.}
 {We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument.
 These catalogues were crossed with the ephemeris of the satellites to identify stellar occultations.
-We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate quick predictions for the short-time future for the satellites of Jupiter.
+We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future for the satellites of Jupiter.
 For Phoebe, Nereid and Triton we used the JPL ephemeris.}
 {We managed to identify 687 candidates of stellar occultations between the period of January, 2015 and December, 2017.
-We made observational tests for the predictions of the events of Himalia in March 03, 2015 and Elara in March 30, 2015.
+We made observational tests for the prediction of the event of Himalia in March 03, 2015.
 The stars and objects were observed close to the date predicted and in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions.}
@@
 
-\section{Introduction} \label{Sec: introducao}
-
-Irregular satellites revolve around giant planets at large distances, on eccentric, highly inclined and frequently retrograde orbits.
+%\section{Introduction} \label{Sec: introducao} 
+%
+%Irregular satellites revolve around giant planets at large distances, on eccentric, highly inclined and frequently retrograde orbits.
 Because of these peculiar orbits, it is largely accepted that these objects did not form by accretion around of their planet, but were captured by their planets in the early solar system \citep{Sheppard2005}.
-
-There are several of capture mechanisms of objects by giant planets suggested in the literature.
+%
+%There are several of capture mechanisms of objects by giant planets suggested in the literature.
 For example the Gas Drag in the primordial circumplanetary nebulae \citep{Sheppard2005} where the object would be affected by the gas drag and its velocity slowed down until it be captured by the planet.
 Another mechanism is called pull-down capture \citep{Sheppard2005}, where the mass of the planet would increase while the object was temporarily captured.
-
-A mechanism based in the Nice model \citep{Morbidelli2005, Tsiganis2005, Gomes2005} was proposed by \cite{Nesvorny2007} and, in the specific case of Jupiter with the modern Nice model, by \citealp{Nesvorny2014}.
+%
+%A mechanism based in the Nice model \citep{Morbidelli2005, Tsiganis2005, Gomes2005} was proposed by \cite{Nesvorny2007} and, in the specific case of Jupiter with the modern Nice model, by \citealp{Nesvorny2014}.
 In this scenario, during the early solar system instability, encounters between the outer planets occurred.
@@
 In this scenario, the survival rate of prior-LHB (Late Heavy Bombardment) satellites is very small.
-
-Another possible mechanism is the capture through collisional interactions \citep{Sheppard2005}.
+%
+%Another possible mechanism is the capture through collisional interactions \citep{Sheppard2005}.
 A collision between two small bodies in the Hill's sphere of the planet could generate fragmented objects and the dissipated energy could be such that some of these objects could be captured.
-
-Some of these objects are in dynamical groups with similar semi-major axis, eccentricities and inclinations, called families, similar to families found in the Main Asteroid Belt.
+%
+%Some of these objects are in dynamical groups with similar semi-major axis, eccentricities and inclinations, called families, similar to families found in the Main Asteroid Belt.
 These families may have been created by a parent body disrupted by collisions with comets or other satellites \citep{Nesvorny2004}.
 These collisions are more likely to have occurred during the LHB \citep{Gomes2005}.
-
-\cite{Nesvorny2003} studied the collision rates between irregular satellites and concluded that some satellites could have been removed by collision with a bigger satellite.
+%
+%\cite{Nesvorny2003} studied the collision rates between irregular satellites and concluded that some satellites could have been removed by collision with a bigger satellite.
 The collision rate between satellites of the Himalia Group (Himalia, Elara, Lysithea and Leda, mainly), for instance, was found to be more than one during the solar system age suggesting that their current structure was originated by satellite-satellite collision.
-
-For Phoebe, ejected material from its surface caused by impacts could evolve due to Poynting-Robertson drag and collide with Iapetus causing the large variation in albedo observed on it \citep{Nesvorny2003}.
+%
+%For Phoebe, ejected material from its surface caused by impacts could evolve due to Poynting-Robertson drag and collide with Iapetus causing the large variation in albedo observed on it \citep{Nesvorny2003}.
 Indeed, Cassini detected in Phoebe an absorption feature at 2.42 $\mu m$ (probably CN combinations) that was also detected in the dark side of Iapetus \citep{Clark2005}.
-
-The region of origin of these object is not well known.
+%
+%The region of origin of these object is not well known.
 \cite{Grav2003} and \cite{Grav2007} showed that the irregular satellites from the giant planets have their colors and spectral slopes similar to C-, D- and P-type asteroids, Centaurs and trans-neptunian objects (TNOs) suggesting that they come from different locations in the early solar system.
-
-In order to obtain precise fundamental physical parameters like size, shape and albedo for the irregular satellites and thus to contribute to the study of their origin, we want to observe stellar occultations, which provide  more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
-
-No observation of a stellar occultation by an irregular satellite was published up to date.
+%
+%In order to obtain precise fundamental physical parameters like size, shape and albedo for the irregular satellites and thus to contribute to the study of their origin, we want to observe stellar occultations, which provide  more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
+%
+%No observation of a stellar occultation by an irregular satellite was published up to date.
 Since their estimated sizes are very small (see Table \ref{Tab: satellite-diameter}), this may have prevented earlier tries.
@@
 Thus, in this case, the cumulated uncertainty of ephemeris and star position must be around 40 mas to get good chances of observing a stellar occultation, which is quite feasible today.
-
-\begin{table}
-\caption{\label{Tab: satellite-diameter} Estimated diameter of the satellites and correspondent apparent diameter}
-\begin{centering}
-\begin{tabular}{lccc}
-\hline  \hline
-\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
-Satellite  & mas\tablefootmark{a}  & km & References \tabularnewline
-\hline
-Ananke & 8 & 29 & 1 \tabularnewline
-Carme & 13 & 46 & 1 \tabularnewline
-Elara & 24 & 86 & 1 \tabularnewline
-Himalia & 41 & $(150\times120) \pm 20$\tablefootmark{b} & 2 \tabularnewline
-Leda & 5 & 20 & 1 \tabularnewline
-Lysithea & 10 & 36 & 1 \tabularnewline
-Pasiphae & 17 & 62 & 1 \tabularnewline
-Sinope & 10 & 37 & 1 \tabularnewline
-\hdashline
-Phoebe & 32 & $212 \pm 1.4$\tablefootmark{b} & 3 \tabularnewline
-\hdashline
-Nereid & 15 & $340 \pm 50$\tablefootmark{c} & 4 \tabularnewline
-Triton & 124 & $2707 \pm 2.0$\tablefootmark{c} & 5 \tabularnewline
-\hline
-\end{tabular}
-\tablebib{
-(1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}; (4) \cite{Thomas1991}; (5) \cite{Thomas2000}.}
-\end{centering}
-\tablefoottext{a}{Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
-\tablefoottext{b}{From Cassini observation.}
-\tablefoottext{c}{From Voyager-2 observation.}
-\par
-\end{table}
-
-From what we have just presented, we can think that the observation of an stellar occultation by an irregular satellite should be likelier than occultations by TNOs.
+%
+%\begin{table}
+%\caption{\label{Tab: satellite-diameter} Estimated diameter of the satellites and correspondent apparent diameter}
+%\begin{centering}
+%\begin{tabular}{lccc}
+%\hline  \hline
+%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
+%Satellite  & mas\tablefootmark{a}  & km & References \tabularnewline
+%\hline
+%Ananke & 8 & 29 & 1 \tabularnewline
+%Carme & 13 & 46 & 1 \tabularnewline
+%Elara & 24 & 86 & 1 \tabularnewline
+%Himalia & 41 & $(150\times120) \pm 20$\tablefootmark{b} & 2 \tabularnewline
+%Leda & 5 & 20 & 1 \tabularnewline
+%Lysithea & 10 & 36 & 1 \tabularnewline
+%Pasiphae & 17 & 62 & 1 \tabularnewline
+%Sinope & 10 & 37 & 1 \tabularnewline
+%\hdashline
+%Phoebe & 32 & $212 \pm 1.4$\tablefootmark{b} & 3 \tabularnewline
+%\hdashline
+%Nereid & 15 & $340 \pm 50$\tablefootmark{c} & 4 \tabularnewline
+%Triton & 124 & $2707 \pm 2.0$\tablefootmark{c} & 5 \tabularnewline
+%\hline
+%\end{tabular}
+%\tablebib{
+%(1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}; (4) \cite{Thomas1991}; (5) \cite{Thomas2000}.}
+%\end{centering}
+%\tablefoottext{a}{Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
+%\tablefoottext{b}{From Cassini observation.}
+%\tablefoottext{c}{From Voyager-2 observation.}
+%\par
+%\end{table}
+%
+%From what we have just presented, we can think that the observation of an stellar occultation by an irregular satellite should be likelier than occultations by TNOs.
 The orbits of their host planets are well known and their observation time-span covers many orbital periods, contrary to TNOs.
-%Unlike stellar occultations by TNOs, which nevertheless have been proved effective, the observation of an occultation by a irregular satellite is, in principle, more favorable.
+%%Unlike stellar occultations by TNOs, which nevertheless have been proved effective, the observation of an occultation by a irregular satellite is, in principle, more favorable.
 The orbits of their host planets are well known and these satellites have been observed completing already many orbital periods around them, thus presenting better ephemeris than TNOs.
-Moreover, the irregular satellites are closer to Earth which means a minor localization error in kilometer.
-
-\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas.
+%Moreover, the irregular satellites are closer to Earth which means a minor localization error in kilometer.
+%
+%\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas.
 For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}.
@@
 Using the positions obtained by \cite{GomesJunior2015} we produced a specific ephemeris for short-time future for the satellites of Jupiter and better predict stellar occultations for these objects.
-
-We present in this paper stellar occultation predictions for the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda), Phoebe from Saturn and Nereid and Triton from Neptune.
+%
+%We present in this paper stellar occultation predictions for the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda), Phoebe from Saturn and Nereid and Triton from Neptune.
 In the section \ref{Sec: Rationale} we explore the scientific rationale for the study of the irregular satellites and the possibility of having a common origin with TNOs.
@@
 Some tests made to confirm the predictions are presented in section \ref{Sec: testes} and the final discussion are in section \ref{Sec: discussion}.
-
-\section{Scientific Rationale} \label{Sec: Rationale}
-
-%\textcolor{red}{Como pode ver, comparado ao paragrafo na Introducao que falou a respeito, o texto aqui está muito pequeno e chovendo no molhado.}
-
-%\textcolor{red}{Tem que falar mais.
+%
+%\section{Scientific Rationale} \label{Sec: Rationale}
+%
+%%\textcolor{red}{Como pode ver, comparado ao paragrafo na Introducao que falou a respeito, o texto aqui está muito pequeno e chovendo no molhado.}
+%
+%%\textcolor{red}{Tem que falar mais.
 Citar o paper no livro da Barucci que voce me mostrou outro dia.
@@
 Tem que fundamentar a hipotese de irregulares = TNOs pequenos.}
-
-%\textcolor{red}{Tambem falar aqui das vantagens de se observar ocultacoes dos irregulares, porque apesar de alguns satelites terem efemerides ruins, outros nem tanto, as efemerides sao melhores que a de TNOs (os irregulares ja completaram voltas em torno dos seus planetas); no caso de Jupiter, ele esta proximo o que ajuda a diminuir o erro em km.}
-
-%\textcolor{red}{E dizer que, mesmo que a hipotese de TNOs que nos motiva a fazer as predicoes, venha a ser refutada no futuro, seja por outros trabalhos, seja pelos resultados futuros de observacoes de nossas proprias ocultacoes, o levantamento em si das propriedades fisicas, tamanho, forma e albedo ainda assim seriam muito uteis para testar as hiposteses de captura, ou seja, contribui para o estudo da formacao e evolucao dos sistemas de Jupiter, Saturno e Netuno.}
-
-There is no consensus for a single model explaining where the irregular satellites were formed.
+%
+%%\textcolor{red}{Tambem falar aqui das vantagens de se observar ocultacoes dos irregulares, porque apesar de alguns satelites terem efemerides ruins, outros nem tanto, as efemerides sao melhores que a de TNOs (os irregulares ja completaram voltas em torno dos seus planetas); no caso de Jupiter, ele esta proximo o que ajuda a diminuir o erro em km.}
+%
+%
+%%\textcolor{red}{E dizer que, mesmo que a hipotese de TNOs que nos motiva a fazer as predicoes, venha a ser refutada no futuro, seja por outros trabalhos, seja pelos resultados futuros de observacoes de nossas proprias ocultacoes, o levantamento em si das propriedades fisicas, tamanho, forma e albedo ainda assim seriam muito uteis para testar as hiposteses de captura, ou seja, contribui para o estudo da formacao e evolucao dos sistemas de Jupiter, Saturno e Netuno.}
+%
+%There is no consensus for a single model explaining where the irregular satellites were formed.
 \cite{Cuk2004} showed that the progenitor of the Himalia group may have originated in heliocentric orbits similar to the Hilda asteroid group.
 \cite{Sheppard2005} stated that the irregular satellites may be some of the objects that were formed within the giant planets region.
-
-Phoebe is the most studied irregular satellite.
+%
+%Phoebe is the most studied irregular satellite.
 \cite{Clark2005} suggest that its surface is probably covered by material of cometary origin.
 It was also stated by \cite{Johnson2005} that if the porosity of Phoebe is 15\%, Phoebe would have an uncompressed density similar to those of Pluto and Triton.
-
-\cite{Sheppard2005} and \cite{Jewitt2007} also expose the possibility for the irregular satellites having their origin as comets or TNOs.
+%
+%\cite{Sheppard2005} and \cite{Jewitt2007} also expose the possibility for the irregular satellites having their origin as comets or TNOs.
 TNOs are highly interesting objects that, due to their large heliocentric distances, may be highly preserved having their physical properties similar to those they had when they were formed \citep{Camargo2014}.
@@
 A clever way to overcome this difficulty is to study  the much closer irregular satellites, under the hypothesis that they share a common origin with the small TNOs' population.
-
-Triton is a uncommon satellite.
+%
+%Triton is a uncommon satellite.
 Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
 Because Triton's orbit size is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded from the irregular satellites' class, but studied together by many authors \citep{Sheppard2005, Jewitt2007}.
-
-Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have the same origin as the TNOs \citep{Agnor2006}.
+%
+%Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have the same origin as the TNOs \citep{Agnor2006}.
 Differently, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
@@
 
-\section*{Introduction V2}
+\section{Introduction}\label{Sec: introducao}
 
-Irregular satellites revolve around giant planets at large distances, on eccentric, highly inclined and frequently retrograde orbits.
-Because of these peculiar orbits, it is largely accepted that these objects did not form by accretion around of their planet, but were captured by their planets in the early solar system \citep{Sheppard2005}.
+Irregular satellites revolve around giant planets at large distances on eccentric, highly inclined and frequently retrograde orbits.
+Because of these peculiar orbits, it is largely accepted that these objects did not form by accretion around their planet, but were captured in the early solar system \citep{Sheppard2005}.
+There is no consensus for a single model explaining where the irregular satellites were formed.
 
-There is no consensus for a single model explaining where the irregular satellites were formed.
 \cite{Cuk2004} showed that the progenitor of the Himalia group may have originated in heliocentric orbits similar to the Hilda asteroid group.
@@
 \cite{Sheppard2005} and \cite{Jewitt2007} also expose the possibility for the irregular satellites having their origin as comets or TNOs.
-TNOs are highly interesting objects that, due to their large heliocentric distances, may be highly preserved having their physical properties similar to those they had when they were formed \citep{Camargo2014}.
+TNOs are highly interesting objects that, due to their large heliocentric distances, may be highly preserved and have physical properties similar to those they had when they were formed \citep{Barucci2008}.
 This is even more true for the smaller objects, since in principle larger sizes favour physical differentiation processes in the body and vice-versa.
 However, due to the distance, the smaller TNOs from this region are more difficult to observe.
-A clever way to overcome this difficulty is to study  the much closer irregular satellites, under the hypothesis that they share a common origin with the small TNOs' population.
+Thus, if irregular satellites - or at least a few of them - do share a common origin with small TNOs, and since these objects are situated at much closer distances, this gives a unique chance of observing and studying representatives of this specific TNO population in much greater detail than could ever be possible by direct observation of this population in the Kuiper Belt.
 
@@
 
-In order to obtain precise fundamental physical parameters like size, shape and albedo for the irregular satellites and thus to contribute to the study of their origin, we want to observe stellar occultations, which provide  more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
+In order to obtain precise fundamental physical parameters like size and shape, thus constraining the albedo and in a broader sense also the composition for the irregular satellites and therefore to contribute to the study of their origin, we aim at observing stellar occultations, which provide  more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
 
@@
 
-Triton is a uncommon satellite.
+Triton is an uncommon satellite.
 Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
@@
 Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have the same origin as the TNOs \citep{Agnor2006}.
-Differently, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
-The main motivation to study Triton by stellar occultations is to understand the evolution of its atmosphere due to Triton's complicated and extreme seasonal cycle \citep{McKinnon2007}.
+However, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
+The main motivation to study Triton by stellar occultations is to understand the evolution of its atmosphere due to Triton's complicated and extreme seasonal cycle \citep{McKinnon2007, Elliot_2000}.
 
 No observation of a stellar occultation by an irregular satellite was published up to date.
-Since their estimated sizes are very small (see Table \ref{Tab: satellite-diameter}), this may have prevented earlier tries.
-But, in fact, given their distances to us, current ephemeris and star positions are already good enough for the prediction of the exact location and instant where the shadow of the occultation will cross the Earth.
+Since their estimated sizes are very small (see Table \ref{Tab: satellite-diameter}), this may have discouraged earlier tries.
+But, in fact, given their distances to us, current ephemeris and star positions are precise enough for the prediction of the exact location and instant where the shadow of the occultation will cross the Earth.
 For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds).
@@
 They pointed out that the ephemeris of the irregular satellites have systematic errors that may reach 200 mas for some satellites.
-For an object at the distance of Jupiter, this represents an error bigger than 700 km in the shadow path.
-Using the positions obtained by \cite{GomesJunior2015} we produced a specific ephemeris for short-time future for the satellites of Jupiter and better predict stellar occultations for these objects.
+For an object at the distance of Jupiter, this represents an error larger than 700 km in the shadow path.
+Using the positions obtained by \cite{GomesJunior2015} we produced a specific short-time ephemeris for the satellites of Jupiter and better predict stellar occultations for these objects.
 
-From what we have just presented, we can think that the observation of an stellar occultation by an irregular satellite should be likelier than occultations by TNOs.
+Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
+However, in the case of irregular satellites, we have much better ephemeris because the orbits of their host planets are well known, their observational time span is much wider and covers many orbital periods in contrast to TNOs.
+Moreover, the irregular satellites are much closer to Earth which implies in a much smaller shadow path error in kilometers.
+These advantages may be somewhat balanced by the smaller sizes estimated for the irregular satellites.Thus, in comparison, the chances for a successful observation of an stellar occultation by an irregular satellite should be considered at least also as good as that of by a TNO.
+
+%From what we have just presented, we can think that the observation of an stellar occultation by an irregular satellite should be likelier than occultations by TNOs.
 The orbits of their host planets are well known and their observation time-span covers many orbital periods, contrary to TNOs.
@@
 The orbits of their host planets are well known and these satellites have been observed completing already many orbital periods around them, thus presenting better ephemeris than TNOs.
-Moreover, the irregular satellites are closer to Earth which means a minor localization error in kilometer.
+%Moreover, the irregular satellites are closer to Earth which means a minor localization error in kilometer.
 
-In section \ref{Sec: integration} we show the the process of the production of the new ephemeris.
+In section \ref{Sec: integration} we show the building of the production of the new ephemeris.
 In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites and how they were made.
-Some tests made to confirm the predictions are presented in section \ref{Sec: testes} and the final discussion are in section \ref{Sec: discussion}.
+Some tests made to check the accuracy of the predictions are presented in section \ref{Sec: testes} and the final discussion is presented in section \ref{Sec: discussion}.
 
@@
 The last observations used to develop JPL current ephemeris of the irregular satellites of Jupiter were obtained in 2012 \citep{Jacobson2012}.
-As a result, the errors on the residuals of recent observations are too important for the ephemeris to accurately predict stellar occultations without any corrections.
-In order to have efficient ephemerides for predicting the stellar occultations by irregular satellites, we need predictions based on recent observations.
+As a result, the errors in the JPL ephemeris for the current epoch are large enough to prevent accurate predictions of stellar occultations without any corrections.
+%In order to have efficient ephemerides for predicting the stellar occultations by irregular satellites, we need predictions based on recent observations.
 This is why we decided to develop our own ephemeris based on the observations published in \cite{GomesJunior2015}.
@@
 %They showed that the orbits of the irregular satellites of the giant planets have systematic errors.
-Their offsets of the observations relative to the JPL ephemeris could be up to 200 mas for some satellites.
-At the distance of Jupiter, this represents an difference bigger than 700 km in the shadow path of an occultation.%These differences could be associated with errors in their orbital elements.
+Their observed position offsets relative to the JPL ephemeris reached up to 200 mas for some satellites.
+At the distance of Jupiter, this represents an difference larger than 700 km in the shadow path of an occultation.%These differences could be associated with errors in their orbital elements.
+
+In order to have efficient ephemerides for predicting the stellar occultations by irregular satellites, we need predictions based on such recent observations. This is why we decided to develop our own
+ephemeris based on the observations published in \cite{GomesJunior2015}.
+First because many of the authors of this paper worked in the production of that set of positions, so that it was very well known.
+Second, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets.
+For these reasons, only this set of recent positions was used for building the new ephemeris for the satellites.
 
@@
 \subfigure{\includegraphics[width=8.8cm]{figures/Carme_resid.eps}  \label{Fig: residuals-Carme}}
-\caption{Residuals of the numerical integration for Himalia and Carme. \label{Fig: residuals}}
+\caption{Residuals in topocentric right ascension and declination of the numerical integration for Himalia and Carme. \label{Fig: residuals}}
 \end{figure*}
@@
 \hline\hline
-Satellite & a (km) & e & incl. (EQJ2000) & $\Omega$ & $\omega$ & $v$ \\ 
+Satellite & a (km) & e & I$\degr$ (EQJ2000) & $\Omega\degr$ & $\omega\degr$ & $v$ \\ 
 \hline
@@
 \subfigure{\includegraphics[width=8.8cm]{figures/JPL_DEC_Carme.eps}  \label{Fig: JPL-STE-Carme}}
-\caption{Differences between the JUP300 JPL ephemeris and the special-tailored ephemeris for Himalia and Carme. \label{Fig: JPL-STE}}
+\caption{Geocentric right ascension and declination differences between the JUP300 JPL ephemeris and the special-tailored ephemeris for Himalia and Carme. \label{Fig: JPL-STE}}
 \end{figure*}
@@
 The obtained ephemeris is hereafter refered as STE, for special-tailored ephemeris.
+
+In this work, the STE was used in all but three satellites, not Jovian ones.
+For these 3 cases, the JPL ephemeris was used instead.
+The JPL ephemeris for Phoebe is of enough high quality for the short term, mostly thanks to the observations by Cassini \textcolor{red}{(é verdade?
+O paper do Tadjedine foi usado pelo JPL?
+dar alguma citacao aqui)}.
+In the case of Triton, the JPL ephemeris is also good enough, as the relatively large diameter of Triton favours the positive detection of occultations.
+A new short term ephemeris for Nereid is however still needed, but its production was beyond the scope of this work, so we used the JPL ephemeris in this case too.
 
@@
 
-The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the special-tailored ephemeris for Jupiter satellites (see Sec.
+The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the STE ephemeris for Jupiter satellites (see Sec.
 \ref{Sec: integration}) and JPL ephemeris for the others.
@@
 Since these objects are very small, the duration of each event is a few seconds.
+All the occultation tables and maps will be publicly available at the CDS.
 
@@
 \textbf{Notes}.
-Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - in the original table; C/A: the geocentric closest approach, in arcseconds; P/A: the planet position angle with respect to the occulted star at C/A, in degrees; $\nu$: velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; {\it D}: planet range to Earth, in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
+Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - in the original table; C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; $\nu$: velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; {\it D}: planet range to Earth, in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
 \label{Tab: occ-list}
@@
 \includegraphics[scale=0.4]{figures/Pasiphae_2016-04-09T03:58:19.png}   
-\caption{Occultation map for Pasiphae regarding to the first event sampled in Table \ref{Tab: occ-list}}
+\caption{Occultation map for Pasiphae regarding to the first event sampled in Table \ref{Tab: occ-list}.
+The central red dot show the geocentric closest approach of the shadow.
+The small ones shows the center of the shadow separated by one minute.
+The lines show the path of the shadow over the Earth.}
 \label{Fig: ocultacao}
@@
 
-The ESA astrometry satellite GAIA \citep{deBruijne2012} catalogue is expected to be released in the end of 2016.
+The first preliminary version of the ESA astrometry satellite GAIA \citep{deBruijne2012} catalogue is expected to be released up to the end of 2016.
 The precise star positions to be derived by GAIA will provide better predictions with the main source of error being the ephemeris.
@@
 
-The candidate stars was searched in the UCAC4 catalogue, except for the candidates in 2015 and 2016 for Triton and Nereid.
+The candidate stars were searched in the UCAC4 catalogue, except for the candidates in 2015 and 2016 for Triton and Nereid.
 In this case, we used the WFI catalogue that was generated from observations made with ESO2p2/WFI CCD mosaic that covered the path of Neptune in the sky-plane up to 2016 (see Sec.
 \ref{Sec: predictions}).
-From this, a total of 687 events were foreseen.
+From this, a total of 687 events are foreseen.
 
 The probability of successfully observing an occultation is roughly the ratio of the satellite's diameter by the budget error of ephemeris and star position.
-Thus, UCAC4 errors ranging between 20 mas - 50 mas combined with a mean accuracy in the position of 30 mas for Himalia and 150 mas for Leda published in Table 2 of \cite{Jacobson2012} would give 80\%-50\% probability of observing such an event by Himalia and $\approx$3\% for Leda, the smallest irregular satellite in the sample.
+Thus, UCAC4 errors ranging between 20 mas - 50 mas combined with a mean accuracy in the JPL ephemeris of 30 mas for Himalia and 150 mas for Leda published in Table 2 of \cite{Jacobson2012} would give 80\%-50\% probability of observing such an event by Himalia and $\approx$3\% for Leda, the smallest irregular satellite in the sample.
 Observations a few days before the date of occultation predicted may improve the combined accuracy to 40-80 mas, depending on the magnitude of the objects.
@@
 A similar test was made for an occultation of Elara in March 30, 2015. %The difference between the maps in this case are listed in Table \ref{Tab: comparison-Elara}.
-The results show satisfying small offsets. We can expect similar results for other satellites and our method should allow successful stellar occultation.
+The results show satisfying small offsets. %We can expect similar results for other satellites and our method should allow successful stellar occultation.
 
-Continuous observations of the satellites and candidates are recommended and fitting of our dynamical model to those observations are expected to reduce the respective position errors.
-The GAIA catalogue, which is to be released at the end of 2016 will improve the position error of the stars and will allow for the detection of occultations candidates of fainter stars, which are not present in the UCAC4 catalogue.
+Continuous observations of the satellites and candidates are recommended and fitting of our dynamical model to those observations are expected to reduce the respective STE ephemeris errors.
+The GAIA catalogue, which first version is to be released up to the end of 2016 will improve the position error of the stars and will allow for the detection of occultations candidates of fainter stars, which are not present in the UCAC4 catalogue.
 % Consequently, the astrometry of the satellites with the GAIA catalogue will improve the position errors allowing for better predictions and a higher probabilities of observing stellar occultations by irregular satellites.
-The release of the GAIA catalogue should have a positive impact on both the astrometric precision of occulted stars and the reduction of astrometric positions of occulting satellites.
-As a result, stellar occultations by irregular satellites should increase in number as well as success.
+The release of the GAIA catalogue should have a positive impact on both the astrometric precision of occulted stars and the reduction of astrometric positions of the satellites.
+As a result, stellar occultations by irregular satellites shall increase in number as well as success.
 
@@
 JIBC acknowledges CNPq for a PQ2 fellowship (process number 308489/2013-6).
-%FB-R acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014.
 BEM thanks the financial support of CAPES.
+FB-R acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014.
 The numerical model of the satellites of Jupiter was developped during a post-doctoral contract funded by the Chinese Academy of Sciences (CAS) and supported by the National Scientific Fund of China (NSFC)

### 2015-11-26 Irregular-occ/Irregular-occ.tex
.

@@
 \hline\hline
-Satellite & a (km) & e & I$\degr$ (EQJ2000) & $\Omega\degr$ & $\omega\degr$ & $v$ \\ 
+Satellite & a (km) & e & I$\degr$ & $\Omega\degr$ & $\omega\degr$ & $v\degr$ \\ 
 \hline
@@
 \end{tabular} 
-
 \end{center}
+\textbf{Notes}: a: Semimajor axis; e : Excentricity; I: Inclination relative to the Equatorial Reference Plane J2000; $\Omega$: Longitude of the ascending node; $\omega$: Argument of periapsis; $v$: True Anomaly.
 \end{table*}
@@
 For these 3 cases, the JPL ephemeris was used instead.
-The JPL ephemeris for Phoebe is of enough high quality for the short term, mostly thanks to the observations by Cassini \textcolor{red}{(é verdade?
-O paper do Tadjedine foi usado pelo JPL?
-dar alguma citacao aqui)}.
+The JPL ephemeris for Phoebe is of enough high quality for the short term, mostly thanks to the observations by Cassini.
 In the case of Triton, the JPL ephemeris is also good enough, as the relatively large diameter of Triton favours the positive detection of occultations.
@@
 
+\begin{figure*}
+\begin{centering}
+\includegraphics[width=17.6cm]{figures/HIMALIA.eps}
+\caption{TESTE DE MAPA.
+STE = AZUL, JPL=VERMELHO, 22FEV=PRETO, 03MAR=VERDE%Predictions for Himalia: The straight lines show the size of the shadow.
+The big red dot show the geocentric closest approach of the shadow.
+The black "+" marks at maps (b,c,d) are the STE prediction closest approach for reference.
+The small red ones are the center of the shadow separated by one minute.
+\ref{Fig: occ-Himalia-STE} is the map using the STE ephemeris showed in Sec.
+\ref{Sec: integration}.
+\ref{Fig: occ-Himalia-JPL} shows the shadow using the JPL ephemeris.
+In \ref{Fig: occ-Himalia-off22fev} we apply offsets to the positions of star and satellite accordingly the observations made at February 22.
+\ref{Fig: occ-Himalia-off03mar} is as in \ref{Fig: occ-Himalia-off22fev} but with observations made at March 03 when the objects were close to each other.
+The green dot at the bottom of the maps is the location of Observatório do Pico dos Dias.
+}
+\label{Fig: occ-Himalia}
+\end{centering}
+\end{figure*}
+
 In this event, it is possible to see that the shade does not vary by much among the four maps suggesting that there was a good probability of observing the event.

### 2015-11-27 Irregular-occ/Irregular-occ.tex
.

@@
 
-\begin{figure*}
-\begin{centering}
-\subfigure[Prediction with STE ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_STE.eps}  \label{Fig: occ-Himalia-STE}}
-\subfigure[Prediction with JPL ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_JPL.eps}  \label{Fig: occ-Himalia-JPL}}
-\subfigure[Observation at February 22 (different FOVs)]{\includegraphics[width=8.8cm]{figures/Himalia_22fev.eps}   \label{Fig: occ-Himalia-off22fev}}
-\subfigure[Observation at March 03 (same FOV)]{\includegraphics[width=8.8cm]{figures/Himalia_03mar.eps}   \label{Fig: occ-Himalia-off03mar}}
-\caption{Predictions for Himalia: The straight lines show the size of the shadow.
+%\begin{figure*}
+%\begin{centering}
+%\subfigure[Prediction with STE ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_STE.eps}  \label{Fig: occ-Himalia-STE}}
+%\subfigure[Prediction with JPL ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_JPL.eps}  \label{Fig: occ-Himalia-JPL}}
+%\subfigure[Observation at February 22 (different FOVs)]{\includegraphics[width=8.8cm]{figures/Himalia_22fev.eps}   \label{Fig: occ-Himalia-off22fev}}
+%\subfigure[Observation at March 03 (same FOV)]{\includegraphics[width=8.8cm]{figures/Himalia_03mar.eps}   \label{Fig: occ-Himalia-off03mar}}
+%\caption{Predictions for Himalia: The straight lines show the size of the shadow.
 The big red dot show the geocentric closest approach of the shadow.
@@
 The green dot at the bottom of the maps is the location of Observatório do Pico dos Dias.}
-\label{Fig: occ-Himalia}
-\end{centering}
-\end{figure*}
+%\label{Fig: occ-Himalia}
+%\end{centering}
+%\end{figure*}
 
@@
 \includegraphics[width=17.6cm]{figures/HIMALIA.eps}
-\caption{TESTE DE MAPA.
-STE = AZUL, JPL=VERMELHO, 22FEV=PRETO, 03MAR=VERDE%Predictions for Himalia: The straight lines show the size of the shadow.
-The big red dot show the geocentric closest approach of the shadow.
-The black "+" marks at maps (b,c,d) are the STE prediction closest approach for reference.
-The small red ones are the center of the shadow separated by one minute.
-\ref{Fig: occ-Himalia-STE} is the map using the STE ephemeris showed in Sec.
+\caption{Predictions for Himalia: The straight lines show the center of the shadow over time.
+In blue is for the prediction using the STE ephemeris showed in Sec.
 \ref{Sec: integration}.
-\ref{Fig: occ-Himalia-JPL} shows the shadow using the JPL ephemeris.
-In \ref{Fig: occ-Himalia-off22fev} we apply offsets to the positions of star and satellite accordingly the observations made at February 22.
-\ref{Fig: occ-Himalia-off03mar} is as in \ref{Fig: occ-Himalia-off22fev} but with observations made at March 03 when the objects were close to each other.
-The green dot at the bottom of the maps is the location of Observatório do Pico dos Dias.
-}
+In red the shadow using the JPL ephemeris.
+In black we apply offsets to the positions of star and satellite accordingly the observations made at February 22.
+In green, the same for the previous one but with observations made at March 03 when the objects were close to each other.
+The dots show the closest approach of the shadow for each prediction.}
 \label{Fig: occ-Himalia}
@@
 \begin{centering}
-\begin{tabular}{lccc}
+\begin{tabular}{lcc}
 \hline  \hline
-\multicolumn{4}{c}{Difference from STE prediction} \tabularnewline
-Method  & Instant of C/A  & C/A & Map   \tabularnewline
+\multicolumn{3}{c}{Difference from STE prediction} \tabularnewline
+Method  & Instant of C/A  & C/A   \tabularnewline
 \hline
-STE & 00:39:51 UTC & $0\farcs703$ & \ref{Fig: occ-Himalia-STE} \tabularnewline
-JPL & -26s & +11mas (36km) & \ref{Fig: occ-Himalia-JPL} \tabularnewline % (284km)
-Feb. 22 Obs. & -14s & -20mas (65km) & \ref{Fig: occ-Himalia-off22fev} \tabularnewline % (153km)
-Mar. 03 Obs. & -36s & -09mas (29km) & \ref{Fig: occ-Himalia-off03mar} \tabularnewline % (393km)
+STE & 00:39:51 UTC & $0\farcs703$ \tabularnewline
+JPL & -26s & +11mas (36km) \tabularnewline % (284km)
+Feb. 22 Obs. & -14s & -20mas (65km) \tabularnewline % (153km)
+Mar. 03 Obs. & -36s & -09mas (29km) \tabularnewline % (393km)
 \hline

### 2015-11-28 Irregular-occ/Irregular-occ.tex
.

@@
 
-Figure \ref{Fig: occ-Himalia} shows four zoomed in occultation maps corresponding to each of the four situations for Himalia.
-The map \ref{Fig: occ-Himalia-STE} is for the situation 1 in the list above.
-Map \ref{Fig: occ-Himalia-JPL} using JPL ephemeris (situation 2).
-The map \ref{Fig: occ-Himalia-off22fev} was made from obtained positions on February 22 observed with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / px$) at the Observatório do Pico dos Dias (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
+Figure \ref{Fig: occ-Himalia} shows a zoomed in occultation map with the four situations for Himalia.
+The line in blue is center of the shadow for the situation 1 in the list above.
+The line in red using JPL ephemeris (situation 2).
+The line in black was made from obtained positions on February 22 observed with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / px$) at the Observatório do Pico dos Dias (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
 On that day, Himalia and the star were observed in separate FOVs as they were still far apart (situation 3).
@@
 Satellite and star were separated by about 16 arcsec, so very close to each other (situation 4).
-From the calculated offsets, the map \ref{Fig: occ-Himalia-off03mar} was generated.
-Notice that the shadow path was not predicted to cross the OPD (green point at the bottom of the maps).
+From the calculated offsets, the center of the shadow was obtained.
+Notice that the shadow path was not predicted to cross the OPD (which was located at almost 2000 km from the shadow path).
 This was not necessary for testing the prediction.
@@
 In green, the same for the previous one but with observations made at March 03 when the objects were close to each other.
-The dots show the closest approach of the shadow for each prediction.}
+The dots show the closest approach of the shadow for each prediction.
+The scales are relative to the closest approach of the prediction using the STE ephemeris.
+The bars show the estimated diameter of the satellite and estimated error of the prediction.}
 \label{Fig: occ-Himalia}
@@
 
-In this event, it is possible to see that the shade does not vary by much among the four maps suggesting that there was a good probability of observing the event.
-In fact, the largest differences between the shadows of the four maps were 36s, which correspond to 120 mas, in time along the shadow path and 101km (31 mas) in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
+In this event, it is possible to see that the shade does not vary by much among the four predictions being smaller than the estimated diameter of the satellite and error.
+This suggests that there was a good probability of observing the event.
+In fact, the largest differences between the shadows of the four situations were 36s, in time along the shadow path and 101km (31 mas) in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
 

### 2015-12-08 Irregular-occ/Irregular-occ.tex
.

@@
 
-\title{Stellar occultation predictions for  10 irregular satellites of giant planets plus Triton for 2015-2017
+\title{Stellar occultation predictions for  10 irregular satellites of giant planets plus Triton up to 2017
 %\fnmsep\thanks{Table 8 are only available in electronic form at the CDS via anonymous ftp to cdsarc.u-strasbg.fr (130.79.128.5) or via http://cdsweb.u-strasbg.fr/cgi-bin/qcat?J/A+A/ and IAU NSDC data base at www.imcce.fr/nsdc.}
@@
           \inst{2,4},     
+          J. Desmars
+          \inst{5},
           B. E. Morgado
-          \inst{1,2}
+          \inst{1,2},
           F. Braga-Ribas
-          \inst{2,5}, 
+          \inst{2,6}, 
 %          D. N. da Silva Neto
@@
 %          \inst{4,7},
-%          J. Desmars
-%          \inst{4},
 %          V. Lainey
@@
 Laboratório Interinstitucional de e-Astronomia - LIneA, Rua Gal. José Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil
-%              \and
-%Institut de mécanique céleste et de calcul des éphémérides - Observatoire %de Paris, UMR 8028 du CNRS,
-%77 Av. Denfert-Rochereau, 75014 Paris, France\\
+              \and
+Institut de mécanique céleste et de calcul des éphémérides - Observatoire de Paris, UMR 8028 du CNRS,
+77 Av. Denfert-Rochereau, 75014 Paris, France
 %              \email{arlot@imcce.fr}
@@
 For Phoebe, Nereid and Triton we used the JPL ephemeris.}
-{We managed to identify 687 candidates of stellar occultations between the period of January, 2015 and December, 2017.
+{We managed to identify 396 candidates of stellar occultations between the period of January, 2016 and December, 2017.
 We made observational tests for the prediction of the event of Himalia in March 03, 2015.
@@
 
-No observation of a stellar occultation by an irregular satellite was published up to date.
+No observation of a stellar occultation by an irregular satellite was published up to date, excluding Triton.
 Since their estimated sizes are very small (see Table \ref{Tab: satellite-diameter}), this may have discouraged earlier tries.
@@
 \multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
-Satellite  & mas\tablefootmark{a}  & km & References \tabularnewline
+Satellite  & mas\tablefootmark{a}  & km & Ref. \tabularnewline
 \hline
@@
 In the case of Triton, the JPL ephemeris is also good enough, as the relatively large diameter of Triton favours the positive detection of occultations.
-A new short term ephemeris for Nereid is however still needed, but its production was beyond the scope of this work, so we used the JPL ephemeris in this case too.
+A new short term ephemeris for Nereid is however still needed, but its production was beyond the scope of this work due to its more complicated dynamic, so we used the JPL ephemeris in this case too.
 
@@
 
-For Triton and Nereid, the candidates for stellar occultations in 2015 and 2016 was searched using the WFI catalogue in the same way as the predictions for Centaurs and TNOs occultations by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.
+For Triton and Nereid, the candidates for stellar occultations in 2016 was searched using the WFI catalogue in the same way as the predictions for Centaurs and TNOs occultations by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.
 This catalogue contains the stars in the path of Neptune in the sky up to mid-2016.
@@
 
-A total of 687 events were identified between January 2015 and December 2017.
+A total of 396 events were identified between January 2016 and December 2017.
 In table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite.
@@
 All the occultation tables and maps will be publicly available at the CDS.
+No star brighter than MagR*=18 was identified to be occulted by Triton in 2016.
+Since the flux drop during the such an occultation is very small, we don't publish this events.
 
 \begin{table}
-\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from 2015 up to 2017.}
+\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from January, 2016 up to December, 2017.}
 \begin{centering}
-\begin{tabular}{lcccc}
+\begin{tabular}{lccc}
 \hline  \hline
 %\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
-Satellite  & 2015  & 2016 & 2017 & Total \tabularnewline
+Satellite  & 2016 & 2017 & Total \tabularnewline
 \hline
-Ananke & 21 & 12 & 16 & 49 \tabularnewline
-Carme & 26 & 20 & 14 & 60 \tabularnewline
-Elara & 32 & 14 & 16 & 62 \tabularnewline
-Himalia & 26 & 15 & 12 & 53 \tabularnewline
-Leda & 13 & 8 & 24 & 45 \tabularnewline
-Lysithea & 22 & 16 & 11 & 49 \tabularnewline
-Pasiphae & 25 & 20 & 19 & 64 \tabularnewline
-Sinope & 19 & 15 & 21 & 55 \tabularnewline
+Ananke & 12 & 16 & 28 \tabularnewline
+Carme & 20 & 14 & 34 \tabularnewline
+Elara & 14 & 16 & 30 \tabularnewline
+Himalia & 15 & 12 & 27 \tabularnewline
+Leda & 8 & 24 & 32 \tabularnewline
+Lysithea & 16 & 11 & 27 \tabularnewline
+Pasiphae & 20 & 19 & 39 \tabularnewline
+Sinope & 15 & 21 & 36 \tabularnewline
 \hdashline
-Phoebe\tablefootmark{a} & 13 & 32 & 98 & 143 \tabularnewline
+Phoebe\tablefootmark{a} & 32 & 98 & 130 \tabularnewline
 \hdashline
-Nereid\tablefootmark{a} & 41\tablefootmark{b} & 11\tablefootmark{b} & 1 & 53 \tabularnewline
-Triton\tablefootmark{a} & 37\tablefootmark{b} & 16\tablefootmark{b} & 1 & 54 \tabularnewline
+Nereid\tablefootmark{a} & 11\tablefootmark{b} & 1 & 12 \tabularnewline
+Triton\tablefootmark{a} & -- & 1 & 1 \tabularnewline
 \hline
@@
 
-\section{Occultation tests} \label{Sec: testes}
+\subsection{Occultation test} \label{Sec: testes}
 
@@
 And, in our case, the shadow covers a very restricted area on Earth because of the size of the irregular satellites.
-So, before we start a large observational campaign, we tested some occultation predictions for large targets, to assess the quality of the predictions.
+Since no stellar occultation by irregular satellite was observed up to date, with the exception of Triton, and before we start a large observational campaign, we tested an occultation prediction for a large target, to assess the quality of the prediction.
 
-The tests consisted in observing the object and star to be occulted near the date of the event predicted when the two objects were present in the same field of view (FOV), preferably when the objects were close to each other.
+The test consisted in observing the object and star to be occulted near the date of the event predicted when the two objects were present in the same field of view (FOV), preferably when the objects were close to each other.
 Thus, the relative positions between the two objects had minimal influence of the errors of the reference catalogue of stars used and possible field distortions \citep[and references therein]{Peng2008}.
@@
 
-Two occultation tests were performed, one by Himalia that occurred on March 3, 2015 and the second by Elara that occurred on March 30, 2015. For each event, four situations were considered:
-\begin{enumerate}
+The occultation tested was performed by Himalia that occurred on March 3, 2015. For the event, four situations were considered:
+\begin{enumerate}[I]
 \item Prediction with the STE ephemeris (see Sec. \ref{Sec: integration}), and the nominal UCAC4 position of the star;
@@
 
-Figure \ref{Fig: occ-Himalia} shows a zoomed in occultation map with the four situations for Himalia.
-The line in blue is center of the shadow for the situation 1 in the list above.
-The line in red using JPL ephemeris (situation 2).
-The line in black was made from obtained positions on February 22 observed with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / px$) at the Observatório do Pico dos Dias (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
-On that day, Himalia and the star were observed in separate FOVs as they were still far apart (situation 3).
+Table \ref{Tab: comparison-Himalia} shows the differences between the predictions in the four situations.
+For situation 3 we observed the objects on February 22 with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / px$) at the Observatório do Pico dos Dias (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
+On that day, Himalia and the star were observed in separate FOVs as they were still far apart.
 On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope (diameter = 1.6m; FOV = $5\farcm8$; pixel scale = $0\farcs17 / px$) at OPD just over an hour after the time scheduled for the event.
@@
 
-\begin{figure*}
-\begin{centering}
-\includegraphics[width=17.6cm]{figures/HIMALIA.eps}
-\caption{Predictions for Himalia: The straight lines show the center of the shadow over time.
+%\begin{figure*}
+%\begin{centering}
+%\includegraphics[width=17.6cm]{figures/HIMALIA.eps}
+%\caption{Predictions for Himalia: The straight lines show the center of the shadow over time.
 In blue is for the prediction using the STE ephemeris showed in Sec.
@@
 The bars show the estimated diameter of the satellite and estimated error of the prediction.}
-\label{Fig: occ-Himalia}
-\end{centering}
-\end{figure*}
+%\label{Fig: occ-Himalia}
+%\end{centering}
+%\end{figure*}
 
@@
 This suggests that there was a good probability of observing the event.
-In fact, the largest differences between the shadows of the four situations were 36s, in time along the shadow path and 101km (31 mas) in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).
+In fact, the largest differences between the shadows of the four situations were 36s in time along the shadow path and 101km (31 mas) in the direction perpendicular to the shadows.
 
@@
 \begin{centering}
-\begin{tabular}{lcc}
+\begin{tabular}{lccc}
 \hline  \hline
-\multicolumn{3}{c}{Difference from STE prediction} \tabularnewline
-Method  & Instant of C/A  & C/A   \tabularnewline
+\multicolumn{4}{c}{Difference from STE prediction} \tabularnewline
+Method  & Instant of C/A  & C/A & Sit.   \tabularnewline
 \hline
-STE & 00:39:51 UTC & $0\farcs703$ \tabularnewline
-JPL & -26s & +11mas (36km) \tabularnewline % (284km)
-Feb. 22 Obs. & -14s & -20mas (65km) \tabularnewline % (153km)
-Mar. 03 Obs. & -36s & -09mas (29km) \tabularnewline % (393km)
+STE & 00:39:51 UTC & $0\farcs703$ & I \tabularnewline
+JPL & -26s & +11mas (36km) & II \tabularnewline % (284km)
+Feb. 22 Obs. & -14s & -20mas (65km) & III \tabularnewline % (153km)
+Mar. 03 Obs. & -36s & -09mas (29km) & IV \tabularnewline % (393km)
 \hline
@@
 \par\end{centering}
-%*Closest Approach
+C/A: geocentric closest approach; Sit: Situation test considered.
 \end{table}
 
-The second test was with the satellite Elara, which is the second largest irregular satellite of Jupiter.
+%The second test was with the satellite Elara, which is the second largest irregular satellite of Jupiter.
 The event was predicted to occur at March 30, 2015.
@@
 %Still, the differences between the maps obtained were relatively small.
-The largest differences between them were 144s (186 mas) in time along the shadow path and 525 km (153 mas) perpendicular to it (see table \ref{Tab: comparison-Elara}).
+%The largest differences between them were 144s (186 mas) in time along the shadow path and 525 km (153 mas) perpendicular to it (see table \ref{Tab: comparison-Elara}).
 
-\begin{table}
-\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara occultation at March 30, 2015.}
-\begin{centering}
-\begin{tabular}{lcc}
-\hline  \hline
-\multicolumn{3}{c}{Difference from STE prediction} \tabularnewline
-Method  & Instant of C/A  & C/A  \tabularnewline
-\hline
-STE & 01:45:13 UTC & $1\farcs082$ \tabularnewline
-JPL & +02s & +57mas(195km) \tabularnewline
-Mar. 25 Obs. & -70s & +20mas (68km) \tabularnewline
-Apr. 02 Obs. & +74s & -96mas (330km) \tabularnewline
-\hline
-\end{tabular}
-\par\end{centering}
-\end{table}
+%\begin{table}
+%\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara occultation at March 30, 2015.}
+%\begin{centering}
+%\begin{tabular}{lcc}
+%\hline  \hline
+%\multicolumn{3}{c}{Difference from STE prediction} \tabularnewline
+%Method  & Instant of C/A  & C/A  \tabularnewline
+%\hline
+%STE & 01:45:13 UTC & $1\farcs082$ \tabularnewline
+%JPL & +02s & +57mas(195km) \tabularnewline
+%Mar. 25 Obs. & -70s & +20mas (68km) \tabularnewline
+%Apr. 02 Obs. & +74s & -96mas (330km) \tabularnewline
+%\hline
+%\end{tabular}
+%\par\end{centering}
+%\end{table}
 
@@
 
-We presented stellar occultations for the period of 2015-2017 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid.
+We presented stellar occultations for the period of 2016-2017 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid.
 The procedure used was the same as that for the prediction of stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.
 
-The candidate stars were searched in the UCAC4 catalogue, except for the candidates in 2015 and 2016 for Triton and Nereid.
+The candidate stars were searched in the UCAC4 catalogue, except for the candidates in 2016 for Triton and Nereid.
 In this case, we used the WFI catalogue that was generated from observations made with ESO2p2/WFI CCD mosaic that covered the path of Neptune in the sky-plane up to 2016 (see Sec.
 \ref{Sec: predictions}).
-From this, a total of 687 events are foreseen.
+From this, a total of 396 events are foreseen.
 
@@
 
-The tests made with an occultation expected to happen in March 03, 2015 for Himalia showed that this event would probably have been observed successfully in case there were observers available in the shadow area.
+The test made with an occultation expected to happen in March 03, 2015 for Himalia showed that this event would probably have been observed successfully in case there were observers available in the shadow area.
 %Fig.
@@
 Table \ref{Tab: comparison-Himalia} shows numerically the difference between the maps.
-A similar test was made for an occultation of Elara in March 30, 2015. %The difference between the maps in this case are listed in Table \ref{Tab: comparison-Elara}.
+%A similar test was made for an occultation of Elara in March 30, 2015. %The difference between the maps in this case are listed in Table \ref{Tab: comparison-Elara}.
 The results show satisfying small offsets. %We can expect similar results for other satellites and our method should allow successful stellar occultation.

### 2016-01-25 Irregular-occ/Irregular-occ.tex
Adicionado texto do Josselin

@@
 
-\section{Special tailored Ephemerides (STE) for Jupiter irregular satellites} \label{Sec: integration}
+\section{Ephemeris} \label{Sec: integration}
 
-\subsection{Observations used}
-
-The last observations used to develop JPL current ephemeris of the irregular satellites of Jupiter were obtained in 2012 \citep{Jacobson2012}.
-As a result, the errors in the JPL ephemeris for the current epoch are large enough to prevent accurate predictions of stellar occultations without any corrections.
-%In order to have efficient ephemerides for predicting the stellar occultations by irregular satellites, we need predictions based on recent observations.
-This is why we decided to develop our own ephemeris based on the observations published in \cite{GomesJunior2015}.
+%\subsection{Observations used}
+In order to have efficient ephemerides for predicting the stellar occultations by irregular satellites, we need predictions based on such recent observations.
+This is why we decided to develop our own
+ephemeris based on the observations published in \cite{GomesJunior2015}.
+First because many of the authors of this paper worked in the production of that set of positions, so that it was very well known.
+Second, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets.
+For these reasons, only this set of recent positions was used for building the new ephemeris for the satellites.
 
@@
 
-In order to have efficient ephemerides for predicting the stellar occultations by irregular satellites, we need predictions based on such recent observations. This is why we decided to develop our own
-ephemeris based on the observations published in \cite{GomesJunior2015}.
+%In order to have efficient ephemerides for predicting the stellar occultations by irregular satellites, we need predictions based on such recent observations.
+This is why we decided to develop our own ephemeris based on the observations published in \cite{GomesJunior2015}.
 First because many of the authors of this paper worked in the production of that set of positions, so that it was very well known.
@@
 
-\subsection{Description of the model}
+\subsection{Special tailored Ephemerides (STE) for Jupiter irregular satellites}
+
+The last observations used to develop JPL current ephemeris of the irregular satellites of Jupiter were obtained in 2012 \citep{Jacobson2012}.
+As a result, the errors in the JPL ephemeris for the current epoch are large enough to prevent accurate predictions of stellar occultations without any corrections.
 
@@
 
-In this work, the STE was used in all but three satellites, not Jovian ones.
-For these 3 cases, the JPL ephemeris was used instead.
-The JPL ephemeris for Phoebe is of enough high quality for the short term, mostly thanks to the observations by Cassini.
-In the case of Triton, the JPL ephemeris is also good enough, as the relatively large diameter of Triton favours the positive detection of occultations.
-A new short term ephemeris for Nereid is however still needed, but its production was beyond the scope of this work due to its more complicated dynamic, so we used the JPL ephemeris in this case too.
+\subsection{Phoebe's ephemeris}
+
+For the specific case of Phoebe, the ninth satellite of Saturn, we have updated the ephemeris published in \cite{Desmars2013}.
+The new ephemeris (PH15) used the same dynamical model, including the perturbations of the Sun and the eight planets, the eight major satellites of Saturn and the $J_2$ parameter.
+The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \cite{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center at \url{http://lnfm1.sai.msu.ru/neb/nss/bsapoouf.htm}), and observations from Flagstaff \citep{NOFS} between 2012 and 2014.
+It represents a total number of 5886 observations from 1898 to 2014.
+
+\subsection{Triton's and Nereid's ephemeris}
 
@@
 
-The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the STE ephemeris for Jupiter satellites (see Sec.
-\ref{Sec: integration}) and JPL ephemeris for the others.
+The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the ephemeris explicited in Sec.
+\ref{Sec: integration}.
 The search for stellar candidates follows the same procedure as presented by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.

### 2016-01-26 Irregular-occ/Irregular-occ.tex
.

@@
 
+\begin{figure*}
+\begin{centering}
+\subfigure[Offset from the JPL ephemeris]{\includegraphics[width=8.8cm]{figures/JPL-EME_Triton.eps} \label{Fig:offxtanomxjpl}}
+\subfigure[\textcolor{red}{Offset from the numerical integration}]{\includegraphics[width=8.8cm]{figures/Zhang-EME.eps} \label{Fig:offxtanomxni}}
+\caption{Offsets of the declination of Carme by true anomaly from the JPL ephemeris and \textcolor{red}{from the numerical integration of the positions obtained in \cite{GomesJunior2015}}.
+\label{Fig:offxanom}}
+\end{centering}
+\end{figure*}
+
 %Making a new model for the orbits of these objects would demand a lot of time and delay the publication of predictions of possible events to occur in the very near future.

### 2016-01-27 Irregular-occ/Irregular-occ.tex
.

--- 2016-01-25 Irregular-occ/Irregular-occ.tex
@@
 \usepackage{natbib,twoopt}
-\usepackage[breaklinks=true]{hyperref} %% to avoid \citeads line fills
+\usepackage[breaklinks=true, draft]{hyperref} %% to avoid \citeads line fills
 \usepackage{arydshln}
@@
 
-The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}, while the residuals are presented in Fig \ref{Fig: residuals} for Himalia and Carme as representatives.
+The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}.%, while the residuals are presented in Fig \ref{Fig: residuals} for Himalia and Carme as representatives.
 
-\begin{figure*}
-\subfigure{\includegraphics[width=8.8cm]{figures/Himalia_resid.eps}   \label{Fig: residuals-Himalia}}
-\subfigure{\includegraphics[width=8.8cm]{figures/Carme_resid.eps}  \label{Fig: residuals-Carme}}
-\caption{Residuals in topocentric right ascension and declination of the numerical integration for Himalia and Carme. \label{Fig: residuals}}
-\end{figure*}
+%\begin{figure*}
+%\subfigure{\includegraphics[width=8.8cm]{figures/Himalia_resid.eps}   \label{Fig: residuals-Himalia}}
+%\subfigure{\includegraphics[width=8.8cm]{figures/Carme_resid.eps}  \label{Fig: residuals-Carme}}
+%\caption{Residuals in topocentric right ascension and declination of the numerical integration for Himalia and Carme. \label{Fig: residuals}}
+%\end{figure*}
 
@@
 Though they are no longer precise enough for our use, they remain a precious reference to identify whether our own model presents a quick divergence.
+
 We compared our ephemeris to the JPL for all the Jupiter satellites we fitted, until 2017.
-The divergence between 2015 and 2018 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.
-A example of this difference for satellites Himalia and Carme over time is presented in Figure \ref{Fig: JPL-STE}.
+For instance, the divergence between 2015 and 2018 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.
 
-\begin{figure*}
-\subfigure{\includegraphics[width=8.8cm]{figures/JPL_DEC_Himalia.eps}    \label{Fig: JPL-STE-Himalia}}
-\subfigure{\includegraphics[width=8.8cm]{figures/JPL_DEC_Carme.eps}  \label{Fig: JPL-STE-Carme}}
-\caption{Geocentric right ascension and declination differences between the JUP300 JPL ephemeris and the special-tailored ephemeris for Himalia and Carme. \label{Fig: JPL-STE}}
-\end{figure*}
+In Fig.
+\ref{Fig: JPL-STE} it is presented the offsets of the positions published by \cite{GomesJunior2015} in declination relative to our ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.
+It is possible to see that the systematic error point out by \cite{GomesJunior2015} is reduced with our ephemeris, as expected.
+
+\begin{figure}
+\includegraphics[width=8.8cm]{figures/Carme_ephemeris.eps} 
+\caption{Offsets in declination of the positions published by \cite{GomesJunior2015} for Carme.
+The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the JUP300 JPL ephemeris and the green dot relative to \cite{Emelyanov2008}.
+\label{Fig: JPL-STE}}
+\end{figure}
+
+%\begin{figure*}
+%\subfigure{\includegraphics[width=8.8cm]{figures/JPL_DEC_Himalia.eps}    \label{Fig: JPL-STE-Himalia}}
+%\subfigure{\includegraphics[width=8.8cm]{figures/JPL_DEC_Carme.eps}  \label{Fig: JPL-STE-Carme}}
+%\caption{Geocentric right ascension and declination differences between the JUP300 JPL ephemeris and the special-tailored ephemeris for Himalia and Carme. \label{Fig: JPL-STE}}
+%\end{figure*}
 
@@
 \subsection{Triton's and Nereid's ephemeris}
+
+\begin{figure*}
+\begin{centering}
+\subfigure[NEP081 JPL and \cite{Emelyanov2015} ephemeris]{\includegraphics[width=8.8cm]{figures/JPL-EME_Triton.eps} \label{Fig:offxtanomxjpl}}
+\subfigure[\cite{Zhang2013} and \cite{Emelyanov2015} ephemeris]{\includegraphics[width=8.8cm]{figures/Zhang-EME.eps} \label{Fig:offxtanomxni}}
+\caption{Comparison between the ephemeris between the \cite{Emelyanov2015}, \cite{Zhang2013} and JPL ephemeris.}
+\label{Fig:offxanom}
+\end{centering}
+\end{figure*}
 

### 2016-01-27 Irregular-occ/Irregular-occ.tex
.

@@
 \usepackage{natbib,twoopt}
-\usepackage[breaklinks=true, draft]{hyperref} %% to avoid \citeads line fills
+\usepackage[breaklinks=true]{hyperref} %% to avoid \citeads line fills
 \usepackage{arydshln}
@@
 
-\subsection{Triton's and Nereid's ephemeris}
+In Fig.
+\ref{Fig:eph-Phoebe} we compare our ephemeris (PH15) with the SAT375 JPL\footnote{Jacobson, R.A.
+2015-Feb-27.
+"Satellite Ephemeris: SAT375", JPL Satellite Ephemeris File Release, \url{ftp://ssd.jpl.nasa.gov/pub/eph/satellites/nio/LINUX_PC/sat375l.txt}} and the \cite{Emelyanov2007} ephemeris.
+We can see that our ephemeris is closer to the SAT375 JPL ephemeris where the difference between them is smaller than 20 mas (< 10 mas in Declination).
+This difference is smaller than the apparent diameter of Phoebe (see Table \ref{Tab: satellite-diameter})
 
@@
 \begin{centering}
-\subfigure[NEP081 JPL and \cite{Emelyanov2015} ephemeris]{\includegraphics[width=8.8cm]{figures/JPL-EME_Triton.eps} \label{Fig:offxtanomxjpl}}
-\subfigure[\cite{Zhang2013} and \cite{Emelyanov2015} ephemeris]{\includegraphics[width=8.8cm]{figures/Zhang-EME.eps} \label{Fig:offxtanomxni}}
-\caption{Comparison between the ephemeris between the \cite{Emelyanov2015}, \cite{Zhang2013} and JPL ephemeris.}
-\label{Fig:offxanom}
+\subfigure[SAT375 JPL and PH15 ephemeris]{\includegraphics[width=8.8cm]{figures/Phoebe_JPL-ph15.eps} \label{Fig:eph-Triton-eme}}
+\subfigure[\cite{Emelyanov2007} and PH15 ephemeris]{\includegraphics[width=8.8cm]{figures/Phoebe_EME-ph15.eps} \label{Fig:eph-Triton-zhang}}
+\caption{Comparison between the PH15, SAT375 JPL and \cite{Emelyanov2007} ephemeris.}
+\label{Fig:eph-Phoebe}
 \end{centering}
 \end{figure*}
+
+\subsection{Triton's and Nereid's ephemeris}
+
+For Triton, we used the most recent ephemeris published by \cite{Emelyanov2015}.
+In Fig.
+\ref{Fig:eph-Triton} we compare the ephemeris with the ephemeris published in \cite{Zhang2014} and the NEP081 JPL ephemeris \citep{Jacobson2009}.
+It is possible to see that the difference between the three ephemeris is smaller than 15 mas for the period 2015-2018.
+This value is smaller than the apparent size of Triton (124 mas, see Table \ref{Tab: satellite-diameter}), which means a good probability of observing an event by this object.
+
+\begin{figure*}
+\begin{centering}
+\subfigure[NEP081 JPL and \cite{Emelyanov2015} ephemeris]{\includegraphics[width=8.8cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
+\subfigure[\cite{Zhang2014} and \cite{Emelyanov2015} ephemeris]{\includegraphics[width=8.8cm]{figures/Zhang-EME.eps} \label{Fig:eph-Triton-zhang}}
+\caption{Comparison between the \cite{Emelyanov2015}, \cite{Zhang2014} and NEP081 JPL \citep{Jacobson2009} ephemeris.}
+\label{Fig:eph-Triton}
+\end{centering}
+\end{figure*}
+
+For Nereid, we used the ephemeris published by \cite{Emelyanov2011}, which is more recent than the JPL ephemeris published by \cite{Jacobson2009}.
+The comparison between the two ephemeris can be seen in Fig.
+\ref{Fig:eph-Nereid}.
+The difference between them is smaller than 60 mas which is bigger than the satellite apparent diameter (see Table \ref{Tab: satellite-diameter}).
+Although this fact, the difference in bigger in Right Ascension than in Declination, which means that, in an event, the error would be smaller in the latitude where the shadow will pass on Earth than the error at the time of the occultation.
+
+\begin{figure}
+\begin{centering}
+\includegraphics[width=8.8cm]{figures/JPL-EME_Nereid.eps}
+\caption{Comparison between the \cite{Emelyanov2011} and NEP081 JPL \citep{Jacobson2009} ephemeris }
+\label{Fig:eph-Nereid}
+\end{centering}
+\end{figure}
 
@@
 \begin{centering}
-\includegraphics[scale=0.4]{figures/Pasiphae_2016-04-09T03:58:19.png}   
-\caption{Occultation map for Pasiphae regarding to the first event sampled in Table \ref{Tab: occ-list}.
+\includegraphics[scale=0.4]{figures/Triton_2017-10-05T23:51:53.png}   
+\caption{Occultation map for Triton.% regarding to the first event sampled in Table \ref{Tab: occ-list}.
 The central red dot show the geocentric closest approach of the shadow.
 The small ones shows the center of the shadow separated by one minute.
-The lines show the path of the shadow over the Earth.}
+The lines show the path of the shadow over the Earth.
+The shadow moves from right to left.}
 \label{Fig: ocultacao}

### 2016-01-28 Irregular-occ/Irregular-occ.tex
feito v.16-01-28

@@
 
-\title{Stellar occultation predictions for  10 irregular satellites of giant planets plus Triton up to 2017
+\title{Stellar occultation predictions for  11 irregular satellites of giant planets up to 2017
 %\fnmsep\thanks{Table 8 are only available in electronic form at the CDS via anonymous ftp to cdsarc.u-strasbg.fr (130.79.128.5) or via http://cdsweb.u-strasbg.fr/cgi-bin/qcat?J/A+A/ and IAU NSDC data base at www.imcce.fr/nsdc.}
@@
 These catalogues were crossed with the ephemeris of the satellites to identify stellar occultations.
-We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future for the satellites of Jupiter.
-For Phoebe, Nereid and Triton we used the JPL ephemeris.}
+We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future for the satellites of Jupiter and Phoebe.
+For Nereid and Triton we used ephemeris from the literature.}
 {We managed to identify 396 candidates of stellar occultations between the period of January, 2016 and December, 2017.
 We made observational tests for the prediction of the event of Himalia in March 03, 2015.
-The stars and objects were observed close to the date predicted and in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions.}
+The stars and objects were observed close to the date predicted and in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions.
+We call attention for an occultation by Triton on a bright star in October 5, 2017.
+The event can be observed from Europe and the east coast of USA and will be a good opportunity to study the atmosphere of the satellite.}
 {The comparisons between the predictions and the observation tests show a good agreement.
@@
 For an object at the distance of Jupiter, this represents an error larger than 700 km in the shadow path.
-Using the positions obtained by \cite{GomesJunior2015} we produced a specific short-time ephemeris for the satellites of Jupiter and better predict stellar occultations for these objects.
+Using the positions obtained by \cite{GomesJunior2015} we produced a specific short-time ephemeris for the satellites of Jupiter and Phoebe to better predict stellar occultations by these objects.
 
@@
 %\subsection{Observations used}
-In order to have efficient ephemerides for predicting the stellar occultations by irregular satellites, we need predictions based on such recent observations.
-This is why we decided to develop our own
+In order to have efficient ephemerides for predicting the stellar occultations by irregular satellites, we need predictions based on recent observations.  This is why we decided to develop our own
 ephemeris based on the observations published in \cite{GomesJunior2015}.
@@
 
-The expressions of $\nabla_lU_{\bar{l}\, \hat{i}}$ and $\nabla_i U_{\bar{i}\, \hat{l}}.$ have been developed in \citet{Lainey2004}.The equations of motion are integrated with the numerical integrator RADAU \citep{Everhart1985}.
+The expressions of $\nabla_lU_{\bar{l}\, \hat{i}}$ and $\nabla_i U_{\bar{i}\, \hat{l}}.$ have been developed in \citet{Lainey2004}.
+The equations of motion are integrated with the numerical integrator RADAU \citep{Everhart1985}.
 Our model was fitted to the observations through a least-squares procedure.
@@
 In Fig.
-\ref{Fig: JPL-STE} it is presented the offsets of the positions published by \cite{GomesJunior2015} in declination relative to our ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.
+\ref{Fig: JPL-STE} it is presented the offsets of the positions published by \cite{GomesJunior2015} for the satellite Carme in declination relative to our ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.
 It is possible to see that the systematic error point out by \cite{GomesJunior2015} is reduced with our ephemeris, as expected.
@@
 The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the JUP300 JPL ephemeris and the green dot relative to \cite{Emelyanov2008}.
+As expected, it is possible to see that the systematic error point out by \cite{GomesJunior2015} is reduced with the STE ephemeris.
 \label{Fig: JPL-STE}}
@@
 
-\begin{figure*}
+\begin{figure}
 \begin{centering}
-\subfigure[SAT375 JPL and PH15 ephemeris]{\includegraphics[width=8.8cm]{figures/Phoebe_JPL-ph15.eps} \label{Fig:eph-Triton-eme}}
-\subfigure[\cite{Emelyanov2007} and PH15 ephemeris]{\includegraphics[width=8.8cm]{figures/Phoebe_EME-ph15.eps} \label{Fig:eph-Triton-zhang}}
-\caption{Comparison between the PH15, SAT375 JPL and \cite{Emelyanov2007} ephemeris.}
+\subfigure[SAT375 JPL and PH15]{\includegraphics[width=8.8cm]{figures/Phoebe_JPL-ph15.eps} \label{Fig:eph-Triton-eme}}
+\subfigure[\cite{Emelyanov2007} and PH15]{\includegraphics[width=8.8cm]{figures/Phoebe_EME-ph15.eps} \label{Fig:eph-Triton-zhang}}
+\caption{Comparison between the PH15, SAT375 JPL and \cite{Emelyanov2007} ephemeris for the satellite Phoebe.}
 \label{Fig:eph-Phoebe}
 \end{centering}
-\end{figure*}
+\end{figure}
 
@@
 \begin{centering}
-\subfigure[NEP081 JPL and \cite{Emelyanov2015} ephemeris]{\includegraphics[width=8.8cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
-\subfigure[\cite{Zhang2014} and \cite{Emelyanov2015} ephemeris]{\includegraphics[width=8.8cm]{figures/Zhang-EME.eps} \label{Fig:eph-Triton-zhang}}
-\caption{Comparison between the \cite{Emelyanov2015}, \cite{Zhang2014} and NEP081 JPL \citep{Jacobson2009} ephemeris.}
+\subfigure[NEP081 JPL and \cite{Emelyanov2015}]{\includegraphics[width=8.8cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
+\subfigure[\cite{Zhang2014} and \cite{Emelyanov2015}]{\includegraphics[width=8.8cm]{figures/Zhang-EME.eps} \label{Fig:eph-Triton-zhang}}
+\caption{Comparison between the \cite{Emelyanov2015}, \cite{Zhang2014} and NEP081 JPL \citep{Jacobson2009} ephemeris for the satellite Triton.}
 \label{Fig:eph-Triton}
@@
 The difference between them is smaller than 60 mas which is bigger than the satellite apparent diameter (see Table \ref{Tab: satellite-diameter}).
-Although this fact, the difference in bigger in Right Ascension than in Declination, which means that, in an event, the error would be smaller in the latitude where the shadow will pass on Earth than the error at the time of the occultation.
+Although this fact, the difference is bigger in Right Ascension than in Declination, which means that, in an event, the error would be smaller in the latitude where the shadow will pass on Earth than the error at the time of the occultation.
 
@@
 \includegraphics[width=8.8cm]{figures/JPL-EME_Nereid.eps}
-\caption{Comparison between the \cite{Emelyanov2011} and NEP081 JPL \citep{Jacobson2009} ephemeris }
+\caption{Comparison between the \cite{Emelyanov2011} and NEP081 JPL \citep{Jacobson2009} ephemeris for the satellite Nereid.}
 \label{Fig:eph-Nereid}
@@
 In table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite.
-Table \ref{Tab: occ-list} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps (see Fig.
-\ref{Fig: ocultacao} as an example).
+Table \ref{Tab: occ-list} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps.
 Since these objects are very small, the duration of each event is a few seconds.
@@
 No star brighter than MagR*=18 was identified to be occulted by Triton in 2016.
-Since the flux drop during the such an occultation is very small, we don't publish this events.
+Since the flux drop during such an occultation is very small, we don't publish this events.
+In Fig.
+\ref{Fig: ocultacao} we show an example of an occultation map.
+In this case is an occultation by Triton that is predicted to happen in October 05, 2017.
+This event can be observed from Europe and the east coast of USA and will be a great opportunity to study the atmosphere of the satellite.
 
@@
 The central red dot show the geocentric closest approach of the shadow.
-The small ones shows the center of the shadow separated by one minute.
+The small ones shows the center of the shadow separated by 60s.
 The lines show the path of the shadow over the Earth.
-The shadow moves from right to left.}
+The shadow moves from right to left.
+\textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for a central observation; C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; vel: velocity of event in km/s; Delta: Geocentric distance to the occulting object; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$; long: east longitude of subplanet point in degrees, positive towards east.}
 \label{Fig: ocultacao}

### 2016-02-01 Irregular-occ/Irregular-occ.tex
.

@@
 
-\title{Stellar occultation predictions for  11 irregular satellites of giant planets up to 2017
+\title{Stellar occultation predictions for  10 irregular satellites of giant planets plus Triton for 2016-2017
 %\fnmsep\thanks{Table 8 are only available in electronic form at the CDS via anonymous ftp to cdsarc.u-strasbg.fr (130.79.128.5) or via http://cdsweb.u-strasbg.fr/cgi-bin/qcat?J/A+A/ and IAU NSDC data base at www.imcce.fr/nsdc.}
@@
 The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects.}
-{We aim to predict stellar occultations for the eight biggest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda, and for the irregular satellites Phoebe of Saturn and Nereid of Neptune, and also for Triton.}
+{We aim to predict stellar occultations for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda, and for the irregular satellites Phoebe of Saturn and Nereid of Neptune, and also for Triton.}
 {We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument.
@@
 We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future for the satellites of Jupiter and Phoebe.
-For Nereid and Triton we used ephemeris from the literature.}
+For Nereid and Triton we used recently published ephemeris.}
 {We managed to identify 396 candidates of stellar occultations between the period of January, 2016 and December, 2017.
@@
 The stars and objects were observed close to the date predicted and in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions.
-We call attention for an occultation by Triton on a bright star in October 5, 2017.
-The event can be observed from Europe and the east coast of USA and will be a good opportunity to study the atmosphere of the satellite.}
+We call attention for an occultation by Triton of a bright star (R=12.5) in October 5, 2017.
+The event can be observed from Europe and the east coast of USA and will be a good opportunity to access the current state of the atmosphere of the satellite.}
 {The comparisons between the predictions and the observation tests show a good agreement.
@@
 
-\keywords{Occultations - Planets and satellites: general - Planets and satellites: individual: Jovian and Saturnian irregular satellites}
+\keywords{Occultations - Planets and satellites: general - Planets and satellites: individual: Jovian and Saturnian irregular satellites - Planets and satellites: individual: Triton}
 
@@
 
-Irregular satellites revolve around giant planets at large distances on eccentric, highly inclined and frequently retrograde orbits.
+Irregular satellites revolve around giant planets at large distances in eccentric, highly inclined and frequently retrograde orbits.
 Because of these peculiar orbits, it is largely accepted that these objects did not form by accretion around their planet, but were captured in the early solar system \citep{Sheppard2005}.
+
 There is no consensus for a single model explaining where the irregular satellites were formed.
-
 \cite{Cuk2004} showed that the progenitor of the Himalia group may have originated in heliocentric orbits similar to the Hilda asteroid group.
@@
 
-\cite{Sheppard2005} and \cite{Jewitt2007} also expose the possibility for the irregular satellites having their origin as comets or TNOs.
-TNOs are highly interesting objects that, due to their large heliocentric distances, may be highly preserved and have physical properties similar to those they had when they were formed \citep{Barucci2008}.
+\cite{Sheppard2005} and \cite{Jewitt2007} also explored the possibility that the irregular satellites originated as comets or TNOs.
+TNOs are highly interesting objects that, due to their large heliocentric distances, may be highly preserved with physical properties similar to those they had when they were formed \citep{Barucci2008}.
 This is even more true for the smaller objects, since in principle larger sizes favour physical differentiation processes in the body and vice-versa.
 However, due to the distance, the smaller TNOs from this region are more difficult to observe.
-Thus, if irregular satellites - or at least a few of them - do share a common origin with small TNOs, and since these objects are situated at much closer distances, this gives a unique chance of observing and studying representatives of this specific TNO population in much greater detail than could ever be possible by direct observation of this population in the Kuiper Belt.
+Thus, if irregular satellites - or at least a few of them - do share a common origin with small TNOs, and since these objects are situated at much closer heliocentric distances now, this gives a unique chance of observing and studying representatives of this specific TNO population in much greater detail than could ever be possible by direct observation of this population in the Kuiper Belt.
 
@@
 In order to obtain precise fundamental physical parameters like size and shape, thus constraining the albedo and in a broader sense also the composition for the irregular satellites and therefore to contribute to the study of their origin, we aim at observing stellar occultations, which provide  more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
+For that, reliable predictions of stellar occultations by these satellites are most needed.
 
@@
 Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
-Because Triton's orbit size is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded from the irregular satellites' class, but studied together by many authors \citep{Sheppard2005, Jewitt2007}.
-
+Because Triton's orbit size is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded from the irregular satellites' class, but still studied together by many authors \citep{Sheppard2005, Jewitt2007}.
 Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have the same origin as the TNOs \citep{Agnor2006}.
@@
 The main motivation to study Triton by stellar occultations is to understand the evolution of its atmosphere due to Triton's complicated and extreme seasonal cycle \citep{McKinnon2007, Elliot_2000}.
+For all these reasons, Triton was also included as a target in this work.
 
-No observation of a stellar occultation by an irregular satellite was published up to date, excluding Triton.
+Excluding Triton \citep{Olkin1997, Elliot_2000}, no observation of a stellar occultation by an irregular satellite was published up to date.
 Since their estimated sizes are very small (see Table \ref{Tab: satellite-diameter}), this may have discouraged earlier tries.
-But, in fact, given their distances to us, current ephemeris and star positions are precise enough for the prediction of the exact location and instant where the shadow of the occultation will cross the Earth.
+But, in fact, given their relatively closer distances as compared to TNOs and Centaurs, and considering the current precision of their ephemeris and of star positions, we can now reliably predict the exact location and instant where the shadow of the occultation will cross the Earth.
 For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds).
-Thus, in this case, the cumulated uncertainty of ephemeris and star position must be around 40 mas to get good chances of observing a stellar occultation, which is quite feasible today.
+Thus, in this case, if the accumulated error (2 sigma or 95\% confidence level) of ephemeris and star position be around 70 mas, we have a probability of about 30\% of observing a stellar occultation, which is quite satisfactory today (see discussion in section \ref{Sec: discussion}).
 
@@
 \tablefoottext{a}{Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
-\tablefoottext{b}{From Cassini observation.}
-\tablefoottext{c}{From Voyager-2 observation.}
+\tablefoottext{b}{From Cassini observations.}
+\tablefoottext{c}{From Voyager-2 observations.}
 \par
@@
 
-Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
+Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
 However, in the case of irregular satellites, we have much better ephemeris because the orbits of their host planets are well known, their observational time span is much wider and covers many orbital periods in contrast to TNOs.
 Moreover, the irregular satellites are much closer to Earth which implies in a much smaller shadow path error in kilometers.
-These advantages may be somewhat balanced by the smaller sizes estimated for the irregular satellites.Thus, in comparison, the chances for a successful observation of an stellar occultation by an irregular satellite should be considered at least also as good as that of by a TNO.
+These advantages may be somewhat balanced by the smaller sizes estimated for the irregular satellites.Thus, in comparison, the chances for a successful observation of an stellar occultation by an irregular satellite should be considered at least also as good as those by TNOs.
 
@@
 %\subsection{Observations used}
-In order to have efficient ephemerides for predicting the stellar occultations by irregular satellites, we need predictions based on recent observations.  This is why we decided to develop our own
-ephemeris based on the observations published in \cite{GomesJunior2015}.
+In order to improve ephemeris in short time spans in the near future for a more efficient prediction of stellar occultations by irregular satellites, we need predictions based on new and recent observations.
+In this context, \cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.
+%They showed that the orbits of the irregular satellites of the giant planets have systematic errors.
+%Their observed position offsets relative to the JPL ephemeris reached up to 200 mas for some satellites.
+At the distance of Jupiter, this represents an difference larger than 700 km in the shadow path of an occultation.%These differences could be associated with errors in their orbital elements.
+
+Here we develop our own ephemeris based on the observations published in \cite{GomesJunior2015}.
 First because many of the authors of this paper worked in the production of that set of positions, so that it was very well known.
-Second, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets.
-For these reasons, only this set of recent positions was used for building the new ephemeris for the satellites.
-
-\cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.
-%They showed that the orbits of the irregular satellites of the giant planets have systematic errors.
-Their observed position offsets relative to the JPL ephemeris reached up to 200 mas for some satellites.
-At the distance of Jupiter, this represents an difference larger than 700 km in the shadow path of an occultation.%These differences could be associated with errors in their orbital elements.
+Second, besides recent observations, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets.
+For these reasons, only this set of positions was used for building the new ephemeris for the satellites of Jupiter.
 
@@
 Our numerical model describes the dynamical evolution of the irregular satellites of Jupiter in a jovicentric reference frame.
-The satellites are submitted to the influence of the Sun and the rest of the solar system, as well as those of the Galilean satellites and the first harmonics of Jupiter's gravity field.
-The axis of the reference frame are those of the equatorial reference frame J2000.
+The satellites are submitted to the influence of the Sun and the rest of the solar system, as well as that of the Galilean satellites and the first harmonics of Jupiter's gravity field.
+The axes of the reference frame are those of the ICRS.
 
@@
 \item $j$ another body of the Solar System
-\item $M_j$ the mass of the $j$ body, not an irregular satellite
+\item $M_j$ the mass of the $j$-th body, not an irregular satellite
 \item $m_i$ the mass of the irregular satellite $i$
-\item $\vec{r_i}$ the position of the $i$ body with respect to the barycentre of Jupiter System
+\item $\vec{r_i}$ the position of the $i$-th body with respect to the barycentre of Jupiter System
 \item $r_{ij}$ the distance between bodies $i$ and $j$  
 \item $R_J$ the radius of Jupiter
-\item $J_n$ the dynamic polar oblateness of the nth order for Jupiter's gravity field
+\item $J_n$ the dynamic polar oblateness of the n-th order for Jupiter's gravity field
 \item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
-\item $\Phi_i$ is the inclination of the $i$ satellite with respect to Jupiter's equator.
+\item $\Phi_i$ is the inclination of the $i$-th satellite with respect to Jupiter's equator.
 \end{itemize}
@@
 \end{center}
-\textbf{Notes}: a: Semimajor axis; e : Excentricity; I: Inclination relative to the Equatorial Reference Plane J2000; $\Omega$: Longitude of the ascending node; $\omega$: Argument of periapsis; $v$: True Anomaly.
+\textbf{Notes}: a: semimajor axis; e : excentricity; I: inclination relative to the equatorial reference plane J2000; $\Omega$: longitude of the ascending node; $\omega$: argument of periapsis; $v$: true anomaly.
 \end{table*}
@@
 Its small number of observations in our set means that its orbit is definitely loosely constrained.
-The residuals are lower than those obtained with JPL ephemeris, but the accuracy of an ephemeris decreases when we get further from the time of observations.
+The residuals are smaller than those obtained with JPL ephemeris, which was expected because the accuracy of an ephemeris decreases when we get further from the time of observations.
 The main risk of divergence over time comes from the possible absence of long-term effects when fitting to a short timespan of observations.
@@
 
-We compared our ephemeris to the JPL for all the Jupiter satellites we fitted, until 2017.
+We compared our ephemeris to the JPL for all the Jupiter satellites we fitted, until 2018.
 For instance, the divergence between 2015 and 2018 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.
 
-In Fig.
-\ref{Fig: JPL-STE} it is presented the offsets of the positions published by \cite{GomesJunior2015} for the satellite Carme in declination relative to our ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.
-It is possible to see that the systematic error point out by \cite{GomesJunior2015} is reduced with our ephemeris, as expected.
+Fig.
+\ref{Fig: JPL-STE} displays the offsets of the positions published by \cite{GomesJunior2015} for the satellite Carme in declination relative to our ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.We see that the systematic JPL ephemeris offsets pointed out by \cite{GomesJunior2015} are reduced with our ephemeris, as expected.
 
@@
 The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the JUP300 JPL ephemeris and the green dot relative to \cite{Emelyanov2008}.
-As expected, it is possible to see that the systematic error point out by \cite{GomesJunior2015} is reduced with the STE ephemeris.
+As expected, the ephemeris systematic errors pointed out by \cite{GomesJunior2015} are reduced with the STE ephemeris.
 \label{Fig: JPL-STE}}
@@
 \begin{centering}
-\subfigure[SAT375 JPL and PH15]{\includegraphics[width=8.8cm]{figures/Phoebe_JPL-ph15.eps} \label{Fig:eph-Triton-eme}}
-\subfigure[\cite{Emelyanov2007} and PH15]{\includegraphics[width=8.8cm]{figures/Phoebe_EME-ph15.eps} \label{Fig:eph-Triton-zhang}}
+\includegraphics[width=8.8cm]{figures/Phoebe.eps} 
 \caption{Comparison between the PH15, SAT375 JPL and \cite{Emelyanov2007} ephemeris for the satellite Phoebe.}
@@
 In Fig.
-\ref{Fig:eph-Triton} we compare the ephemeris with the ephemeris published in \cite{Zhang2014} and the NEP081 JPL ephemeris \citep{Jacobson2009}.
-It is possible to see that the difference between the three ephemeris is smaller than 15 mas for the period 2015-2018.
-This value is smaller than the apparent size of Triton (124 mas, see Table \ref{Tab: satellite-diameter}), which means a good probability of observing an event by this object.
+\ref{Fig:eph-Triton} we compare it with the ephemeris published in \cite{Zhang2014} and the NEP081 JPL ephemeris \citep{Jacobson2009}.
+The offsets between the three ephemeris are smaller than 15 mas for the period 2015-2018.
+This value is much smaller than the apparent size of Triton (124 mas, see Table \ref{Tab: satellite-diameter}), which indicates a good probability of observing an event by this object.
 
@@
 
-For Nereid, we used the ephemeris published by \cite{Emelyanov2011}, which is more recent than the JPL ephemeris published by \cite{Jacobson2009}.
+For Nereid, we used the ephemeris published by \cite{Emelyanov2011}, which uses more recent observations than the JPL ephemeris published by \cite{Jacobson2009}.
 The comparison between the two ephemeris can be seen in Fig.
 \ref{Fig:eph-Nereid}.
-The difference between them is smaller than 60 mas which is bigger than the satellite apparent diameter (see Table \ref{Tab: satellite-diameter}).
+The right ascension offsets between them are smaller than 60 mas and the declination ones are smaller than 15 mas.
+For Nereid, ephemeris errors in right ascension translate to uncertainties in the central instant of a stellar occultation, so offsets of 60 mas are not critical.
+The declination ephemeris offsets are of the order of the apparent diameter (15 mas, see Table \ref{Tab: satellite-diameter}), resulting in fair chances (50\%) that the shadow crosses the expected Earth latitude predicted for an occultation.
+% which is bigger than the satellite apparent diameter (see Table \ref{Tab: satellite-diameter}).
 Although this fact, the difference is bigger in Right Ascension than in Declination, which means that, in an event, the error would be smaller in the latitude where the shadow will pass on Earth than the error at the time of the occultation.
@@
 \end{figure}
+
+\cite{GomesJunior2015} published 902 new positions for Nereid from observations between 1992 and 2014.
+The development of a better ephemeris for Nereid using these new positions in the context discussed here of near future stellar occultations is desirable, but is out of the scope of this paper.
 
@@
 
-The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the ephemeris explicited in Sec.
+The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the ephemeris presented in Sec.
 \ref{Sec: integration}.
@@
 
-We predicted occultation for the 8 major irregular satellites of Jupiter:  Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope; Phoebe of Saturn and Triton and Nereid of Neptune.
+We predicted occultations for the 8 major irregular satellites of Jupiter,  Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope, and for Phoebe Phoebe of Saturn and Triton and Nereid of Neptune.
 
-For Triton and Nereid, the candidates for stellar occultations in 2016 was searched using the WFI catalogue in the same way as the predictions for Centaurs and TNOs occultations by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.
+For Triton and Nereid, the candidates for stellar occultations in 2016 were searched using the WFI catalogue in the same way as the predictions for Centaurs and TNOs occultations by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.
 This catalogue contains the stars in the path of Neptune in the sky up to mid-2016.
@@
 A total of 396 events were identified between January 2016 and December 2017.
-In table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite.
+In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite.
 Table \ref{Tab: occ-list} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps.
@@
 All the occultation tables and maps will be publicly available at the CDS.
-No star brighter than MagR*=18 was identified to be occulted by Triton in 2016.
-Since the flux drop during such an occultation is very small, we don't publish this events.
+No star brighter than MagR*=18 will be occulted by Triton in 2016.
+For this satellite, we cut events with stars fainter than R=18, since for Triton the flux drop during such an occultation would be very small.
 In Fig.
 \ref{Fig: ocultacao} we show an example of an occultation map.
-In this case is an occultation by Triton that is predicted to happen in October 05, 2017.
-This event can be observed from Europe and the east coast of USA and will be a great opportunity to study the atmosphere of the satellite.
+This is the only found occultation by Triton that will happen in October 05, 2017.
+This event can be observed from Europe and the east coast of USA and will be a great opportunity to study in high resolution the atmosphere of the satellite.
 
@@
 \begin{table*}
-\caption{\label{Tab: sample-cds} Stellar occultation sampled predictions for Pasiphae}
+\caption{\label{Tab: sample-cds} A sample of stellar occultation predictions for Pasiphae}
 \begin{centering}
@@
 \textbf{Notes}.
-Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - in the original table; C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; $\nu$: velocity in plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; {\it D}: planet range to Earth, in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
+Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - at the central instant of occultation (corrected by proper motions); C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees (zero at north of the star, increasing clockwise); $\nu$: velocity in the plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; {\it D}: planet range to Earth, in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
 \label{Tab: occ-list}
@@
 The shadow moves from right to left.
-\textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for a central observation; C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; vel: velocity of event in km/s; Delta: Geocentric distance to the occulting object; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$; long: east longitude of subplanet point in degrees, positive towards east.}
+\textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for a central observation; C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; vel: velocity of event in km/s; Delta: Geocentric distance to the occulting object in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$; long: east longitude of subplanet point in degrees, positive towards east.}
 \label{Fig: ocultacao}
@@
 
-The first preliminary version of the ESA astrometry satellite GAIA \citep{deBruijne2012} catalogue is expected to be released up to the end of 2016.
+The first preliminary catalogue version of the ESA astrometry satellite GAIA \citep{deBruijne2012} is expected to be released up to the end of 2016.
 The precise star positions to be derived by GAIA will provide better predictions with the main source of error being the ephemeris.
-Astrometric reduction of observations published in \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions from 2017 onwards will be better.
-Because of that, we only publish stellar occultations with UCAC4 until 2017.
+Astrometric reduction of observations published in \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions from 2017 onwards will be improved.
+Since we cannot foresee exactly when the GAIA catalogue will be released and when new and re-reduced satellite positions in the GAIA frame will become available, we decided to publish only the predictions for occultations restricted to the 2016-2017 time interval.
 
-\subsection{Occultation test} \label{Sec: testes}
+\section{Occultation test} \label{Sec: testes}
 
@@
 And, in our case, the shadow covers a very restricted area on Earth because of the size of the irregular satellites.
-Since no stellar occultation by irregular satellite was observed up to date, with the exception of Triton, and before we start a large observational campaign, we tested an occultation prediction for a large target, to assess the quality of the prediction.
+Since no stellar occultation by an irregular satellite was observed up to date, with the exception of Triton, and since we want to be sure that we can start observational campaigns with reasonable chances of success, we tested an occultation prediction for a large target, to assess the quality of the prediction.
 
-The test consisted in observing the object and star to be occulted near the date of the event predicted when the two objects were present in the same field of view (FOV), preferably when the objects were close to each other.
+The test design consisted in observing the object and star to be occulted near the date of the event predicted when the two objects were present in the same field of view (FOV), close to each other.
 Thus, the relative positions between the two objects had minimal influence of the errors of the reference catalogue of stars used and possible field distortions \citep[and references therein]{Peng2008}.
-The relative positions of the star and of the satellites were used to check the original predictions.
+The relative positions of the star and satellite were used to check the original predictions.
+Notice that in the test we do not attempt to observe any actual occultation.
+The test could be performed at any site, regardless of the Earth location where the occultation would in fact be visible.
 
-The occultation tested was performed by Himalia that occurred on March 3, 2015. For the event, four situations were considered:
+We tested the occultation by Himalia predicted to occur on March 3, 2015. The shadow would cross the northern part of South America. For the event, four situations were considered:
 \begin{enumerate}[I]
-\item Prediction with the STE ephemeris (see Sec. \ref{Sec: integration}), and the nominal UCAC4 position of the star;
+\item Our nominal, published prediction with the STE ephemeris (see Sec. \ref{Sec: integration}), and the nominal UCAC4 position of the star;
 \item Prediction with the JPL ephemeris and the nominal UCAC4 position of the star;
-\item From star and satellite offsets calculated from observations made a few days before the occultation when the objects were separated (different FOVs);
+\item From star and satellite offsets calculated from observations made a few days before the occultation when the objects were very separated (different FOVs);
 \item Same as 3 but with the star and the satellite close in the same FOV.
@@
 Table \ref{Tab: comparison-Himalia} shows the differences between the predictions in the four situations.
-For situation 3 we observed the objects on February 22 with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / px$) at the Observatório do Pico dos Dias (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
+For situation 3 we observed the objects on February 22 with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / pixel$) at the Observatório do Pico dos Dias, Brazil (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
 On that day, Himalia and the star were observed in separate FOVs as they were still far apart.
-On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope (diameter = 1.6m; FOV = $5\farcm8$; pixel scale = $0\farcs17 / px$) at OPD just over an hour after the time scheduled for the event.
+On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope (diameter = 1.6m; FOV = $5\farcm8$; pixel scale = $0\farcs17 / pixel$) at OPD just over an hour after the time scheduled for the event.
 Satellite and star were separated by about 16 arcsec, so very close to each other (situation 4).
 From the calculated offsets, the center of the shadow was obtained.
-Notice that the shadow path was not predicted to cross the OPD (which was located at almost 2000 km from the shadow path).
+Notice that the shadow path was not predicted to cross the OPD (which was located at almost 2000 km south from the shadow path).
 This was not necessary for testing the prediction.
@@
 
-In this event, it is possible to see that the shade does not vary by much among the four predictions being smaller than the estimated diameter of the satellite and error.
-This suggests that there was a good probability of observing the event.
-In fact, the largest differences between the shadows of the four situations were 36s in time along the shadow path and 101km (31 mas) in the direction perpendicular to the shadows.
+The critical parameter in the comparisons is the C/A, which here is related to latitudes.
+The apparent radius of Himalia is about 20 mas (see Table 1).
+In the context of the test, for a 0 mas offset in C/A we would have 100\% probability of observing the occultation, and 0\% in the case of a C/A offset equal to or larger than 20 mas, the radius of Himalia.
+From Table 5, we have nearly 0\% probability of success in situation 3, for which the offset in C/A was -20 mas, but when the relative astrometry was poor, 10 days prior to the event.
+Once at the day of the event in situation 4, the C/A offset dropped to -9 mas only, corresponding to a 55\% probability of success.
+Comparison with the prediction using the JPL ephemeris (situation 2) gives a +11 mas C/A offset, or a compatibility of 45\% between the ephemerides.
+All this suggests that there was a good probability of observing the event.
+The largest differences between the shadows of the four situations were 36s in time along the shadow path and 101km (31 mas) in the direction perpendicular to the shadows, suggesting that observers should be spread in narrow latitude ranges 100 km wide.
 
@@
 \hline  \hline
-\multicolumn{4}{c}{Difference from STE prediction} \tabularnewline
+\multicolumn{4}{c}{Differences with respect to the STE prediction} \tabularnewline
 Method  & Instant of C/A  & C/A & Sit.   \tabularnewline
@@
 
-We presented stellar occultations for the period of 2016-2017 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid.
+We predict stellar occultations for the period of 2016-2017 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid.
 The procedure used was the same as that for the prediction of stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.
@@
 
-The probability of successfully observing an occultation is roughly the ratio of the satellite's diameter by the budget error of ephemeris and star position.
-Thus, UCAC4 errors ranging between 20 mas - 50 mas combined with a mean accuracy in the JPL ephemeris of 30 mas for Himalia and 150 mas for Leda published in Table 2 of \cite{Jacobson2012} would give 80\%-50\% probability of observing such an event by Himalia and $\approx$3\% for Leda, the smallest irregular satellite in the sample.
-Observations a few days before the date of occultation predicted may improve the combined accuracy to 40-80 mas, depending on the magnitude of the objects.
+In a broader, general sense, the probability of successfully observing an occultation is roughly the ratio of the satellite's radius by the budget error (2 sigma for a 95\% confidence level) of ephemeris and star position.
+Thus, UCAC4 errors ranging between 20 mas - 50 mas (1 sigma) combined with a mean error (1 sigma) in the JPL ephemeris of 30 mas for Himalia and 150 mas for Leda published in Table 2 of \cite{Jacobson2012} would give 28\%-17\% probability of observing such an event by Himalia and $\approx2$\% for Leda, the smallest irregular satellite in the sample.
+Observations a few days before the date of occultation predicted may improve the combined errors to 40-80 mas, depending on the magnitude of the objects.
 
@@
 %A similar test was made for an occultation of Elara in March 30, 2015. %The difference between the maps in this case are listed in Table \ref{Tab: comparison-Elara}.
-The results show satisfying small offsets. %We can expect similar results for other satellites and our method should allow successful stellar occultation.
+The results show satisfying small offsets with respect to the local of the prediction. %We can expect similar results for other satellites and our method should allow successful stellar occultation.
 
-Continuous observations of the satellites and candidates are recommended and fitting of our dynamical model to those observations are expected to reduce the respective STE ephemeris errors.
-The GAIA catalogue, which first version is to be released up to the end of 2016 will improve the position error of the stars and will allow for the detection of occultations candidates of fainter stars, which are not present in the UCAC4 catalogue.
+Continuous observations of the satellites are recommended and fitting of our dynamical model to those observations are expected to reduce the respective STE ephemeris errors.
+The first version of the GAIA catalogue is to be released up to the end of 2016 and will improve the position error of the stars to the 1-5 mas level.
+It will allow for the discovery of ocultations by fainter stars not present in the UCAC4 catalogue.
 % Consequently, the astrometry of the satellites with the GAIA catalogue will improve the position errors allowing for better predictions and a higher probabilities of observing stellar occultations by irregular satellites.
 The release of the GAIA catalogue should have a positive impact on both the astrometric precision of occulted stars and the reduction of astrometric positions of the satellites.
-As a result, stellar occultations by irregular satellites shall increase in number as well as success.
+As a result, prediction of stellar occultations by irregular satellites shall increase in number as well as in success.
 

### 2016-02-02 Irregular-occ/Irregular-occ.tex
.

@@
 These catalogues were crossed with the ephemeris of the satellites to identify stellar occultations.
-We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future for the satellites of Jupiter and Phoebe.
-For Nereid and Triton we used recently published ephemeris.}
+We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future or the satellites of Jupiter.
+For Phoebe, we used an updated ephemeris and for Nereid and Triton we used recently published ephemeris.}
 {We managed to identify 396 candidates of stellar occultations between the period of January, 2016 and December, 2017.
@@
 
-We present in this paper stellar occultation predictions for the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda), Phoebe from Saturn and Nereid and Triton from Neptune.
+We present in this paper stellar occultation predictions for the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda), Phoebe of Saturn and, Nereid and Triton of Neptune.
 Phoebe, being the most studied object with a good measured size, can be used to calibrate and evaluate the technique for similar objects.
@@
 Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
-However, in the case of irregular satellites, we have much better ephemeris because the orbits of their host planets are well known, their observational time span is much wider and covers many orbital periods in contrast to TNOs.
+However, in the case of irregular satellites, we have much better ephemeris because the orbits of their host planets are better known, their observational time span is much wider and covers many orbital periods in contrast to TNOs.
 Moreover, the irregular satellites are much closer to Earth which implies in a much smaller shadow path error in kilometers.
@@
 
-For an irregular satellite $i$, under the gravitational influence of Jupiter, the $\mathcal{N'}-1$ other irregular satellites, the regular Jovian satellites and the rest of the Solar System ($\mathcal{N}$ bodies), the equation of motion is:
+For an irregular satellite $i$, under the gravitational influence of Jupiter, the $\mathcal{N}-1$ other irregular satellites, the regular Jovian satellites and the rest of the Solar System ($\mathcal{N}$ bodies), the equation of motion is:
 \begin{equation}\begin{array}{ll}
@@
 \ddot{\vec{r_i}}= & \displaystyle -GM_J\frac{\vec{r_J}-\vec{r_i}}{r_{iJ}^3}-\sum_{l=1,l\neq i}^\mathcal{N'}Gm_l\frac{\vec{r_l}-\vec{r_i}}{r_{il}^3}\\
-&\displaystyle -\sum_{j=1}^\mathcal{N}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \underbrace{\frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3}}_\mathtt{undirect perturbations} \right)\\
- & \displaystyle +GM_J \nabla U_{\bar{l}\hat{J}} -\underbrace{\sum_{l=1}^\mathcal{N} Gm_l\nabla U_{\bar{l}\hat{J}}}_\mathtt{undirect perturbations}
+&\displaystyle -\sum_{j=1}^\mathcal{N}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3} \right)\\
+ & \displaystyle +GM_J \nabla U_{\bar{l}\hat{J}} -\sum_{l=1}^\mathcal{N} Gm_l\nabla U_{\bar{l}\hat{J}}
 \end{array}
+\label{Eq:1}
 \end{equation}
-where the oblateness potential seen by the body $i$ because of Jupiter is:
+where the last term in the brackets and the last term in Eq. \ref{Eq:1} represent undirect perturbations. The oblateness potential seen by the body $i$ because of Jupiter is:
 \begin{equation}\begin{array}{ll}
@@
 The new ephemeris (PH15) used the same dynamical model, including the perturbations of the Sun and the eight planets, the eight major satellites of Saturn and the $J_2$ parameter.
-The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \cite{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center at \url{http://lnfm1.sai.msu.ru/neb/nss/bsapoouf.htm}), and observations from Flagstaff \citep{NOFS} between 2012 and 2014.
+The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \cite{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center\footnote{\url{http://lnfm1.sai.msu.ru/neb/nss/bsapoouf.htm}}), and observations from Flagstaff \citep{NOFS} between 2012 and 2014.
 It represents a total number of 5886 observations from 1898 to 2014.
@@
 
-We predicted occultations for the 8 major irregular satellites of Jupiter,  Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope, and for Phoebe Phoebe of Saturn and Triton and Nereid of Neptune.
+We predicted occultations for the 8 major irregular satellites of Jupiter,  Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope, and for Phoebe of Saturn and Triton and Nereid of Neptune.
 
@@
 \ref{Fig: ocultacao} we show an example of an occultation map.
-This is the only found occultation by Triton that will happen in October 05, 2017.
+This is the only found occultation by Triton and it will happen in October 05, 2017.
 This event can be observed from Europe and the east coast of USA and will be a great opportunity to study in high resolution the atmosphere of the satellite.

### 2016-02-03 Irregular-occ/dados/tabela.py
.

@@ -14,8 +14,10 @@ tabela = Table(names=('object', 'instantes', 'RA (ICRS) DEC', 'C/A', 'P/A', 'vel
     dtype=(np.str, np.str, np.str, np.float, np.float, np.float, np.float, np.float, np.float, np.float, np.str))
     
 def texto(i):
-    if i in ['Triton', 'Nereida', 'Phoebe']:
-        return '/JPL/'
+    if i in ['Triton', 'Nereida']:
+        return '/Emelyanov/'
+    if i == 'Phoebe':
+        return '/Josselin/'
     return '/LAU/'
     
 def mostrar(i):

### 2016-02-08 Irregular-occ/Irregular-occ.tex
.

@@
 Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
-However, in the case of irregular satellites, we have much better ephemeris because the orbits of their host planets are better known, their observational time span is much wider and covers many orbital periods in contrast to TNOs.
+In contrast to TNOs, the irregular satellites have much better ephemeris because the orbits of their host planets are better  known, their observational time span is much wider and covers many orbital periods.
 Moreover, the irregular satellites are much closer to Earth which implies in a much smaller shadow path error in kilometers.
@@
 Here we develop our own ephemeris based on the observations published in \cite{GomesJunior2015}.
-First because many of the authors of this paper worked in the production of that set of positions, so that it was very well known.
+First, because the reduction was made with consistent and precise stellar catalogue and with a robust and reliable software (PRAIA, \cite{Assafin2011}).
 Second, besides recent observations, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets.

### 2016-02-09 Irregular-occ/dados/tabela.py
.

@@ -5,6 +5,7 @@ import astropy.units as u
 from astropy.table import Table, Column, vstack
 import matplotlib.pyplot as plt
 import matplotlib.image as mpimg
+import csv
 
 f = open('lista_tables', 'r')
 lista = f.readlines()
@@ -12,6 +13,9 @@ f.close()
 
 tabela = Table(names=('object', 'instantes', 'RA (ICRS) DEC', 'C/A', 'P/A', 'vel', 'distance', 'magR', 'magK', 'long', 'caminho'),\
     dtype=(np.str, np.str, np.str, np.float, np.float, np.float, np.float, np.float, np.float, np.float, np.str))
+
+c = csv.writer(open("agenda.csv", "wb"))
+c.writerow(["Subject","Start Date","Start Time","End Date","End Time","All Day Event","Description","Private"])
     
 def texto(i):
     if i in ['Triton', 'Nereida']:
@@ -61,6 +65,13 @@ for i in lista:
     caminho = '/home/altair/Documentos/Maps/All/' + np.char.array(datas.iso).rpartition(' ')[:,0].rpartition('-')[:,0].rpartition('-')[:,0] + '/Altair/' + arquivo[15:-11].capitalize() + texto(arquivo[15:-11].capitalize()) + arquivo[15:-11].capitalize() + '_' + np.char.array(datas.isot) + '.png'
     t = Table([name, occ_data, star_coord, ca, pa, vel, dist, magR, magK, longi, caminho], names=('object', 'instantes', 'RA (ICRS) DEC', 'C/A', 'P/A', 'vel', 'distance', 'magR', 'magK', 'long', 'caminho'))
     tabela = vstack([tabela, t])
+
+    for i in np.arange(len(stars)):
+        subj = 'Ocultation by {}'.format(arquivo[15:-11].capitalize())
+        stdate = '{}/{}/{}'.format(datas.iso[i][5:7], datas.iso[i][8:10], datas.iso[i][0:4])
+        sttime = datas.iso[i][11:16]
+        desc = 'Coord Star: {}\nC/A: {} arcsec\nP/A {} deg\nVelocity: {} km/s\nDistance: {} AU\nMagR: {}\n'.format(star_coord[i], ca[i], pa[i], vel[i], dist[i], magR[i])
+        c.writerow([subj,stdate,sttime,stdate,sttime,"False",desc,"False"])
     
 tabela.sort('instantes')
 aa = Column(np.arange(len(tabela)), name='numero')

### 2016-02-29 Irregular-occ/Irregular-occ.tex
.

@@
-%% LyX 2.0.0 created this file.  For more info, see http://www.lyx.org/.
-%% Do not edit unless you really know what you are doing.
-\documentclass[english,traditabstract]{aa}
-\usepackage[varg]{txfonts}
-%\usepackage[english,brazilian]{babel}
+% mn2esample.tex
+%
+% v2.1 released 22nd May 2002 (G. Hutton)
+%
+% The mnsample.tex file has been amended to highlight
+% the proper use of LaTeX2e code with the class file
+% and using natbib cross-referencing. These changes
+% do not reflect the original paper by A. V. Raveendran.
+%
+% Previous versions of this sample document were
+% compatible with the LaTeX 2.09 style file mn.sty
+% v1.2 released 5th September 1994 (M. Reed)
+% v1.1 released 18th July 1994
+% v1.0 released 28th January 1994
+
+\documentclass[useAMS,usenatbib]{mn2e}
 \usepackage[T1]{fontenc}
-\setcounter{tocdepth}{3}
 \usepackage{prettyref}
-\usepackage{array}
-\usepackage{rotating}
-\usepackage{float}
-\usepackage{multirow}
-\usepackage{amstext}
+\usepackage{amsmath}
 \usepackage{graphicx}
+\usepackage{subfigure}
 \usepackage{epstopdf}
-\usepackage{subfigure}
-\usepackage{gensymb}
-\usepackage{longtable}
 \usepackage{color}
-%\usepackage{setspace}
+\usepackage[breaklinks=true]{hyperref}
+
+% If your system does not have the AMS fonts version 2.0 installed, then
+% remove the useAMS option.
 %
-%\usepackage{etoolbox}
-%\AtBeginEnvironment{longtable}{\singlespacing}
-%\setlength{\LTcapwidth}{\linewidth}
+% useAMS allows you to obtain upright Greek characters.
+% e.g. \umu, \upi etc.  See the section on "Upright Greek characters" in
+% this guide for further information.
+%
+% If you are using AMS 2.0 fonts, bold math letters/symbols are available
+% at a larger range of sizes for NFSS release 1 and 2 (using \boldmath or
+% preferably \bmath).
+%
+% The usenatbib command allows the use of Patrick Daly's natbib.sty for
+% cross-referencing.
+%
+% If you wish to typeset the paper in Times font (if you do not have the
+% PostScript Type 1 Computer Modern fonts you will need to do this to get
+% smoother fonts in a PDF file) then uncomment the next line
+% \usepackage{Times}
 
-\makeatletter
+%%%%% AUTHORS - PLACE YOUR OWN MACROS HERE %%%%%
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
-%% Because html converters don't know tabularnewline
-\providecommand{\tabularnewline}{\\}
+\title[Predictions of Stellar Occulations by irregular satellites]{Stellar occultation predictions for 10 irregular satellites of giant planets plus Triton for 2016-2017}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
-\usepackage{natbib,twoopt}
-\usepackage[breaklinks=true, draft]{hyperref} %% to avoid \citeads line fills
-\usepackage{arydshln}
-\bibpunct{(}{)}{;}{a}{}{,} %% natbib format for A&A and ApJ
-\makeatletter
-\newcommandtwoopt{\citeads}[3][][]{\href{http://adsabs.harvard.edu/abs/#3}%
-{\def\hyper@linkstart##1##2{}%
-\let\hyper@linkend\@empty\citealp[#1][#2]{#3}}}
-\newcommandtwoopt{\citepads}[3][][]{\href{http://adsabs.harvard.edu/abs/#3}%
-{\def\hyper@linkstart##1##2{}%
-\let\hyper@linkend\@empty\citep[#1][#2]{#3}}}
-\newcommandtwoopt{\citetads}[3][][]{\href{http://adsabs.harvard.edu/abs/#3}%
-{\def\hyper@linkstart##1##2{}%
-\let\hyper@linkend\@empty\citet[#1][#2]{#3}}}
-\newcommandtwoopt{\citeyearads}[3][][]%
-{\href{http://adsabs.harvard.edu/abs/#3}
-{\def\hyper@linkstart##1##2{}%
-\let\hyper@linkend\@empty\citeyear[#1][#2]{#3}}}
-\makeatother
-\titlerunning{Predictions of stellar occultations by irregular satellites up to 2017}
-\authorrunning{Gomes-Júnior et al.}
-
-\makeatother
+\author[A. R. Gomes-Júnior, M. Assafin, L. Beauvalet et al.]{
+A. R. Gomes-Júnior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
+M. Assafin$^{1,\dag,\ddag}$\thanks{E-mail: massaf@astro.ufrj.br},
+L. Beauvalet$^{2,3}$,
+R. Vieira-Martins$^{1,2,\dag,\ddag}$,
+J.I.B. Camargo$^{2,\dag}$,
+J. Desmars$^{4}$
+B. E. Morgado$^{1,2}$
+F. Braga-Ribas$^{2, 5, \dag}$,
+\\
+$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
+CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
+$^{2}$Observat\'orio Nacional/MCTI, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
+$^{3}$Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France\\
+$^{4}$Institut de mécanique céleste et de calcul des éphémérides - Observatoire de Paris, UMR 8028 du CNRS,
+77 Av. Denfert-Rochereau, 75014 Paris, France\\
+$^{5}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil\\
+$^\dag$ Associated to Laborat\'{o}rio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, CEP 20921-400,\\ Rio de Janeiro, Brazil\\
+$^\ddag$Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France
+}
 
@@
 
-\title{Stellar occultation predictions for  10 irregular satellites of giant planets plus Triton for 2016-2017
-%\fnmsep\thanks{Table 8 are only available in electronic form at the CDS via anonymous ftp to cdsarc.u-strasbg.fr (130.79.128.5) or via http://cdsweb.u-strasbg.fr/cgi-bin/qcat?J/A+A/ and IAU NSDC data base at www.imcce.fr/nsdc.}
-%\fnmsep\thanks{The complete version of Table 8 is available through CDS and IAU NSDC data base at www.imcce.fr/nsdc.}
-%\fnmsep\thanks{Partially based on observations made at Laborat\'orio Nacional de Astrof\'{\i}sica (LNA), Itajub\'a-MG, Brazil.}
-%\fnmsep\thanks{Partially based on observations through the ESO runs 079.A-9202(A), 075.C-0154, 077.C-0283 and 079.C-0345.}
-%\fnmsep\thanks{\textbf{Partially based on observations made at Observatoire de Haute Provence (OHP), F-04870 Saint-Michel l'observatoire, France}}
-}
+\date{Accepted . Received ; in original form }
 
-\author{ A. R. Gomes-Júnior
-          \inst{1},
-          M. Assafin 
-          \inst{1} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France},
-          L. Beauvalet
-          \inst{2,3},
-          R. Vieira-Martins
-          \inst{1,2,4} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France},
-%          J.-E. Arlot
-%          \inst{4},      
-          J. I. B. Camargo
-          \inst{2,4},     
-          J. Desmars
-          \inst{5},
-          B. E. Morgado
-          \inst{1,2},
-          F. Braga-Ribas
-          \inst{2,6}, 
-%          D. N. da Silva Neto
-%          \inst{6},
-%          A. H. Andrei
-%          \inst{1,2} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau, 75014 Paris, France},
-%          A. Dias-Oliveira
-%          \inst{2},
-%          G. Benedetti-Rossi
-%          \inst{2},
-%          Y. Duchemin
-%          \inst{4,7},
-%          V. Lainey
-%          \inst{4},
-%          W. Thuillot
-%          \inst{4}
-          }
+\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2015}
 
-\offprints{A. R. Gomes-Júnior}
+\maketitle
 
-   \institute{Observatório do Valongo/UFRJ, Ladeira Pedro Antônio 43,
-CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
-              \email{altair08@astro.ufrj.br}
-              \and
-Observatório Nacional/MCT, R. General José Cristino 77,
-CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
-              \email{rvm@on.br}
-              \and
-Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France
-              \and
-Laboratório Interinstitucional de e-Astronomia - LIneA, Rua Gal. José Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil
-              \and
-Institut de mécanique céleste et de calcul des éphémérides - Observatoire de Paris, UMR 8028 du CNRS,
-77 Av. Denfert-Rochereau, 75014 Paris, France
-%              \email{arlot@imcce.fr}
-              \and
-Federal University of Technology - Paraná (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
-%              \and
-%Centro Universitário Estadual da Zona Oeste, Av. Manual Caldeira de Alvarenga 1203, CEP 23.070-200 Rio de Janeiro RJ, Brazil
-%              \and
-%              ESIGELEC-IRSEEM, Technopôle du Madrillet, Avenue Galilée, 76801 Saint-Etienne du Rouvray, France
-              }
-
-\date{Received ; accepted }
+\label{firstpage}
 
 \abstract
-{Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system.
+Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system.
 It is important to know their physical parameters, such as size, shape, albedo and composition, to trace back their true origin.
-The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects.}
-{We aim to predict stellar occultations for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda, and for the irregular satellites Phoebe of Saturn and Nereid of Neptune, and also for Triton.}
-{We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument.
+The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects.
+We aim to predict stellar occultations for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda, and for the irregular satellites Phoebe of Saturn and Nereid of Neptune, and also for Triton.
+We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument.
 These catalogues were crossed with the ephemeris of the satellites to identify stellar occultations.
 We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future or the satellites of Jupiter.
-For Phoebe, we used an updated ephemeris and for Nereid and Triton we used recently published ephemeris.}
-{We managed to identify 396 candidates of stellar occultations between the period of January, 2016 and December, 2017.
+For Phoebe, we used an updated ephemeris and for Nereid and Triton we used recently published ephemeris.
+We managed to identify 396 candidates of stellar occultations between the period of January, 2016 and December, 2017.
 We made observational tests for the prediction of the event of Himalia in March 03, 2015.
@@
 We call attention for an occultation by Triton of a bright star (R=12.5) in October 5, 2017.
-The event can be observed from Europe and the east coast of USA and will be a good opportunity to access the current state of the atmosphere of the satellite.}
-{The comparisons between the predictions and the observation tests show a good agreement.
-We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.}
+The event can be observed from Europe and the east coast of USA and will be a good opportunity to access the current state of the atmosphere of the satellite.
+The comparisons between the predictions and the observation tests show a good agreement.
+We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
 
-\keywords{Occultations - Planets and satellites: general - Planets and satellites: individual: Jovian and Saturnian irregular satellites - Planets and satellites: individual: Triton}
-
-\maketitle
+\begin{keywords}
+Occultations - Planets and satellites: general - Planets and satellites: individual: Jovian and Saturnian irregular satellites - Planets and satellites: individual: Triton
+\end{keywords}
 
@@
 
-\section{Introduction}\label{Sec: introducao}
+\include{intro.tex}
 
-Irregular satellites revolve around giant planets at large distances in eccentric, highly inclined and frequently retrograde orbits.
-Because of these peculiar orbits, it is largely accepted that these objects did not form by accretion around their planet, but were captured in the early solar system \citep{Sheppard2005}.
+\include{ephemeris.tex}
 
-There is no consensus for a single model explaining where the irregular satellites were formed.
-\cite{Cuk2004} showed that the progenitor of the Himalia group may have originated in heliocentric orbits similar to the Hilda asteroid group.
-\cite{Sheppard2005} stated that the irregular satellites may be some of the objects that were formed within the giant planets region.
+\include{prediction.tex}
 
-\cite{Grav2003} and \cite{Grav2007} showed that the irregular satellites from the giant planets have their colors and spectral slopes similar to C-, D- and P-type asteroids, Centaurs and trans-neptunian objects (TNOs).
-This suggests that they may have come from different locations in the early solar system.
+\include{discussion.tex}
 
-\cite{Sheppard2005} and \cite{Jewitt2007} also explored the possibility that the irregular satellites originated as comets or TNOs.
-TNOs are highly interesting objects that, due to their large heliocentric distances, may be highly preserved with physical properties similar to those they had when they were formed \citep{Barucci2008}.
-This is even more true for the smaller objects, since in principle larger sizes favour physical differentiation processes in the body and vice-versa.
-However, due to the distance, the smaller TNOs from this region are more difficult to observe.
-Thus, if irregular satellites - or at least a few of them - do share a common origin with small TNOs, and since these objects are situated at much closer heliocentric distances now, this gives a unique chance of observing and studying representatives of this specific TNO population in much greater detail than could ever be possible by direct observation of this population in the Kuiper Belt.
+%\begin{acknowledgements}
+%
+%ARG-J thanks the financial support of CAPES.
+%MA thanks the CNPq (Grants 473002/2013-2 and 308721/2011-0) and FAPERJ (Grant E-26/111.488/2013).
+%RV-M thanks grants: CNPq-306885/2013, Capes/Cofecub-2506/2015, Faperj/PAPDRJ-45/2013.
+%JIBC acknowledges CNPq for a PQ2 fellowship (process number 308489/2013-6).
+%BEM thanks the financial support of CAPES.
+%FB-R acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014.
+%The numerical model of the satellites of Jupiter was developped during a post-doctoral contract funded by the Chinese Academy of Sciences (CAS) and supported by the National Scientific Fund of China (NSFC)
+%
+%\end{acknowledgements}
 
-Phoebe is the most studied irregular satellite.
-\cite{Clark2005} suggest that its surface is probably covered by material of cometary origin.
-It was also stated by \cite{Johnson2005} that if the porosity of Phoebe is 15\%, Phoebe would have an uncompressed density similar to those of Pluto and Triton.
-
-In order to obtain precise fundamental physical parameters like size and shape, thus constraining the albedo and in a broader sense also the composition for the irregular satellites and therefore to contribute to the study of their origin, we aim at observing stellar occultations, which provide  more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
-For that, reliable predictions of stellar occultations by these satellites are most needed.
-
-We present in this paper stellar occultation predictions for the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda), Phoebe of Saturn and, Nereid and Triton of Neptune.
-Phoebe, being the most studied object with a good measured size, can be used to calibrate and evaluate the technique for similar objects.
-
-Triton is an uncommon satellite.
-Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
-Because Triton's orbit size is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded from the irregular satellites' class, but still studied together by many authors \citep{Sheppard2005, Jewitt2007}.
-Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have the same origin as the TNOs \citep{Agnor2006}.
-However, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
-The main motivation to study Triton by stellar occultations is to understand the evolution of its atmosphere due to Triton's complicated and extreme seasonal cycle \citep{McKinnon2007, Elliot_2000}.
-For all these reasons, Triton was also included as a target in this work.
-
-Excluding Triton \citep{Olkin1997, Elliot_2000}, no observation of a stellar occultation by an irregular satellite was published up to date.
-Since their estimated sizes are very small (see Table \ref{Tab: satellite-diameter}), this may have discouraged earlier tries.
-But, in fact, given their relatively closer distances as compared to TNOs and Centaurs, and considering the current precision of their ephemeris and of star positions, we can now reliably predict the exact location and instant where the shadow of the occultation will cross the Earth.
-For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds).
-Thus, in this case, if the accumulated error (2 sigma or 95\% confidence level) of ephemeris and star position be around 70 mas, we have a probability of about 30\% of observing a stellar occultation, which is quite satisfactory today (see discussion in section \ref{Sec: discussion}).
-
-\begin{table}
-\caption{\label{Tab: satellite-diameter} Estimated diameter of the satellites and correspondent apparent diameter}
-\begin{centering}
-\begin{tabular}{lccc}
-\hline  \hline
-\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
-Satellite  & mas\tablefootmark{a}  & km & Ref. \tabularnewline
-\hline
-Ananke & 8 & 29 & 1 \tabularnewline
-Carme & 13 & 46 & 1 \tabularnewline
-Elara & 24 & 86 & 1 \tabularnewline
-Himalia & 41 & $(150\times120) \pm 20$\tablefootmark{b} & 2 \tabularnewline
-Leda & 5 & 20 & 1 \tabularnewline
-Lysithea & 10 & 36 & 1 \tabularnewline
-Pasiphae & 17 & 62 & 1 \tabularnewline
-Sinope & 10 & 37 & 1 \tabularnewline
-\hdashline
-Phoebe & 32 & $212 \pm 1.4$\tablefootmark{b} & 3 \tabularnewline
-\hdashline
-Nereid & 15 & $340 \pm 50$\tablefootmark{c} & 4 \tabularnewline
-Triton & 124 & $2707 \pm 2.0$\tablefootmark{c} & 5 \tabularnewline
-\hline
-\end{tabular}
-\tablebib{
-(1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}; (4) \cite{Thomas1991}; (5) \cite{Thomas2000}.}
-\end{centering}
-\tablefoottext{a}{Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
-\tablefoottext{b}{From Cassini observations.}
-\tablefoottext{c}{From Voyager-2 observations.}
-\par
-\end{table}
-
-\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas.
-For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}.
-They pointed out that the ephemeris of the irregular satellites have systematic errors that may reach 200 mas for some satellites.
-For an object at the distance of Jupiter, this represents an error larger than 700 km in the shadow path.
-Using the positions obtained by \cite{GomesJunior2015} we produced a specific short-time ephemeris for the satellites of Jupiter and Phoebe to better predict stellar occultations by these objects.
-
-Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
-In contrast to TNOs, the irregular satellites have much better ephemeris because the orbits of their host planets are better  known, their observational time span is much wider and covers many orbital periods.
-Moreover, the irregular satellites are much closer to Earth which implies in a much smaller shadow path error in kilometers.
-These advantages may be somewhat balanced by the smaller sizes estimated for the irregular satellites.Thus, in comparison, the chances for a successful observation of an stellar occultation by an irregular satellite should be considered at least also as good as those by TNOs.
-
-%From what we have just presented, we can think that the observation of an stellar occultation by an irregular satellite should be likelier than occultations by TNOs.
-The orbits of their host planets are well known and their observation time-span covers many orbital periods, contrary to TNOs.
-%Unlike stellar occultations by TNOs, which nevertheless have been proved effective, the observation of an occultation by a irregular satellite is, in principle, more favorable.
-The orbits of their host planets are well known and these satellites have been observed completing already many orbital periods around them, thus presenting better ephemeris than TNOs.
-%Moreover, the irregular satellites are closer to Earth which means a minor localization error in kilometer.
-
-In section \ref{Sec: integration} we show the building of the production of the new ephemeris.
-In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites and how they were made.
-Some tests made to check the accuracy of the predictions are presented in section \ref{Sec: testes} and the final discussion is presented in section \ref{Sec: discussion}.
-
-\section{Ephemeris} \label{Sec: integration}
-
-%\subsection{Observations used}
-In order to improve ephemeris in short time spans in the near future for a more efficient prediction of stellar occultations by irregular satellites, we need predictions based on new and recent observations.
-In this context, \cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.
-%They showed that the orbits of the irregular satellites of the giant planets have systematic errors.
-%Their observed position offsets relative to the JPL ephemeris reached up to 200 mas for some satellites.
-At the distance of Jupiter, this represents an difference larger than 700 km in the shadow path of an occultation.%These differences could be associated with errors in their orbital elements.
-
-Here we develop our own ephemeris based on the observations published in \cite{GomesJunior2015}.
-First, because the reduction was made with consistent and precise stellar catalogue and with a robust and reliable software (PRAIA, \cite{Assafin2011}).
-Second, besides recent observations, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets.
-For these reasons, only this set of positions was used for building the new ephemeris for the satellites of Jupiter.
-
-%In order to have efficient ephemerides for predicting the stellar occultations by irregular satellites, we need predictions based on such recent observations.
-This is why we decided to develop our own ephemeris based on the observations published in \cite{GomesJunior2015}.
-First because many of the authors of this paper worked in the production of that set of positions, so that it was very well known.
-Second, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets.
-For these reasons, only this set of recent positions was used for building the new ephemeris for the satellites.
-
-\subsection{Special tailored Ephemerides (STE) for Jupiter irregular satellites}
-
-The last observations used to develop JPL current ephemeris of the irregular satellites of Jupiter were obtained in 2012 \citep{Jacobson2012}.
-As a result, the errors in the JPL ephemeris for the current epoch are large enough to prevent accurate predictions of stellar occultations without any corrections.
-
-Our numerical model describes the dynamical evolution of the irregular satellites of Jupiter in a jovicentric reference frame.
-The satellites are submitted to the influence of the Sun and the rest of the solar system, as well as that of the Galilean satellites and the first harmonics of Jupiter's gravity field.
-The axes of the reference frame are those of the ICRS.
-
-We use the following notations: \begin{itemize}
-\item $i$ and $l$ one of the irregular satellites of Jupiter
-\item $J$ Jupiter 
-\item $j$ another body of the Solar System
-\item $M_j$ the mass of the $j$-th body, not an irregular satellite
-\item $m_i$ the mass of the irregular satellite $i$
-\item $\vec{r_i}$ the position of the $i$-th body with respect to the barycentre of Jupiter System
-\item $r_{ij}$ the distance between bodies $i$ and $j$  
-\item $R_J$ the radius of Jupiter
-\item $J_n$ the dynamic polar oblateness of the n-th order for Jupiter's gravity field
-\item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
-\item $\Phi_i$ is the inclination of the $i$-th satellite with respect to Jupiter's equator.
-\end{itemize}
-
-For an irregular satellite $i$, under the gravitational influence of Jupiter, the $\mathcal{N}-1$ other irregular satellites, the regular Jovian satellites and the rest of the Solar System ($\mathcal{N}$ bodies), the equation of motion is:
-\begin{equation}\begin{array}{ll}
-
-\ddot{\vec{r_i}}= & \displaystyle -GM_J\frac{\vec{r_J}-\vec{r_i}}{r_{iJ}^3}-\sum_{l=1,l\neq i}^\mathcal{N'}Gm_l\frac{\vec{r_l}-\vec{r_i}}{r_{il}^3}\\
-&\displaystyle -\sum_{j=1}^\mathcal{N}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3} \right)\\
- & \displaystyle +GM_J \nabla U_{\bar{l}\hat{J}} -\sum_{l=1}^\mathcal{N} Gm_l\nabla U_{\bar{l}\hat{J}}
-\end{array}
-\label{Eq:1}
-\end{equation}
-where the last term in the brackets and the last term in Eq. \ref{Eq:1} represent undirect perturbations. The oblateness potential seen by the body $i$ because of Jupiter is:
-\begin{equation}\begin{array}{ll}
-
-U_{\bar{l}\hat{J}}=&\displaystyle -\frac{R_J^2 J_2}{r_{iJ}^3}\left(\frac{3}{2}\sin^2 \Phi_i-\frac{1}{2}\right)\\ &\\ & 
-\displaystyle-\frac{R_J^4 J_4}{r_{iJ}^5}\left(\frac{35}{8}\sin^4 \Phi_i-\frac{15}{4}\sin^2 \Phi_i+\frac{3}{8}\right)\\
-& \\
-&\displaystyle-\frac{R_J^6 J_6}{r_{iJ}^7}\left(\frac{231}{16}\sin^6 \Phi_i-\frac{315}{16}\sin^4 \Phi_i+\frac{105}{16}\sin^2 \Phi_i-\frac{5}{16}\right)
-
-\end{array}
-\end{equation}
-
-The expressions of $\nabla_lU_{\bar{l}\, \hat{i}}$ and $\nabla_i U_{\bar{i}\, \hat{l}}.$ have been developed in \citet{Lainey2004}.
-The equations of motion are integrated with the numerical integrator RADAU \citep{Everhart1985}.
-Our model was fitted to the observations through a least-squares procedure.
-The satellites were integrated one dynamical family at a time, to gain computing time, while losing minimum precision.
-Indeed, the interactions between satellites not belonging to the same dynamical family are negligible considering the short timespan of our integration.
-
-The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}.%, while the residuals are presented in Fig \ref{Fig: residuals} for Himalia and Carme as representatives.
-
-%\begin{figure*}
-%\subfigure{\includegraphics[width=8.8cm]{figures/Himalia_resid.eps}   \label{Fig: residuals-Himalia}}
-%\subfigure{\includegraphics[width=8.8cm]{figures/Carme_resid.eps}  \label{Fig: residuals-Carme}}
-%\caption{Residuals in topocentric right ascension and declination of the numerical integration for Himalia and Carme. \label{Fig: residuals}}
-%\end{figure*}
-
-\begin{table*}
-\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0.% For Themisto's case, the small number of observations meant that the statistical uncertainty was even greater than the obtained elements, though the fitting proved satisfying.
-We decided to provide its elements with significant number comparable to the other satellite for information, but left it without uncertainty.
- }\label{Tab: sat_ell}
-\begin{center}
-\begin{tabular}{ccccccc}
-\hline\hline
-Satellite & a (km) & e & I$\degr$ & $\Omega\degr$ & $\omega\degr$ & $v\degr$ \\ 
-\hline
-Himalia &   11372100 $\pm$ 500    &    0.166 $\pm$ 0.002      &   45.14 $\pm$ 0.15      &   39.77 $\pm$ 0.19      &   351.48 $\pm$ 0.46      &   97.35 $\pm$ 0.48    \\
-Elara &   11741170 $\pm$ 690  &      0.222 $\pm$ 0.002      &   28.64 $\pm$ 0.18      &   68.42 $\pm$ 0.43      &   179.82 $\pm$ 0.56      &   339.08 $\pm$ 0.82  \\
-Lysithea &   11739900 $\pm$ 1300  &      0.136 $\pm$ 0.004      &    51.12 $\pm$ 0.27     &   5.53 $\pm$ 0.52      &   53.0 $\pm$ 1.5      &   318.9 $\pm$ 2.0   \\
-Leda &   11140300  $\pm$ 4300  &     0.173  $\pm$ 0.007     &   16.15  $\pm$ 0.75    &   272.6  $\pm$ 1.7    &   212.2  $\pm$ 3.6          &   218.8  $\pm$ 3.2  \\
-Pasiphae &  23425000  $\pm$ 5000    &     0.379  $\pm$ 0.001       &   152.44 $\pm$ 0.10      &   284.59 $\pm$ 0.21      &   135.96 $\pm$ 0.19      &   236.97 $\pm$ 0.16 \\
-Sinope &   22968800 $\pm$ 5200   &     0.316 $\pm$ 0.002      &   157.76 $\pm$ 0.12      &   256.62 $\pm$ 0.55      &   298.38 $\pm$ 0.55      &   167.57 $\pm$ 0.19    \\
-Carme &   24202924 $\pm$ 4800      &  0.242 $\pm$ 0.001      &   147.13 $\pm$ 0.10      &   154.01 $\pm$ 0.25      &   47.90 $\pm$ 0.29      &   234.41 $\pm$ 0.19  \\
-Ananke &  21683800  $\pm$ 7200  &     0.380 $\pm$ 0.002      &   172.29 $\pm$ 0.20      &   56.9 $\pm$ 1.2      &   123.3 $\pm$ 1.2      &   231.24 $\pm$ 0.21  \\
-%Themisto  & 7393800    &     0.198       &   25.77  &   220.0     &   216.3    &   262.1     \\
-\hline
-\end{tabular} 
-\end{center}
-\textbf{Notes}: a: semimajor axis; e : excentricity; I: inclination relative to the equatorial reference plane J2000; $\Omega$: longitude of the ascending node; $\omega$: argument of periapsis; $v$: true anomaly.
-\end{table*}
-
-All the orbits determined for the satellites show satisfying residuals.
-%Yet, one of the satellites has a peculiar situation.
-Its small number of observations in our set means that its orbit is definitely loosely constrained.
-The residuals are smaller than those obtained with JPL ephemeris, which was expected because the accuracy of an ephemeris decreases when we get further from the time of observations.
-The main risk of divergence over time comes from the possible absence of long-term effects when fitting to a short timespan of observations.
-If that were the case, our ephemeris would diverge too quickly to be of any use.
-JPL ephemerides are fitted over all the available observations.
-As a result, they will diverge less quickly than our own.
-Though they are no longer precise enough for our use, they remain a precious reference to identify whether our own model presents a quick divergence.
-
-We compared our ephemeris to the JPL for all the Jupiter satellites we fitted, until 2018.
-For instance, the divergence between 2015 and 2018 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.
-
-Fig.
-\ref{Fig: JPL-STE} displays the offsets of the positions published by \cite{GomesJunior2015} for the satellite Carme in declination relative to our ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.We see that the systematic JPL ephemeris offsets pointed out by \cite{GomesJunior2015} are reduced with our ephemeris, as expected.
-
-\begin{figure}
-\includegraphics[width=8.8cm]{figures/Carme_ephemeris.eps} 
-\caption{Offsets in declination of the positions published by \cite{GomesJunior2015} for Carme.
-The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the JUP300 JPL ephemeris and the green dot relative to \cite{Emelyanov2008}.
-As expected, the ephemeris systematic errors pointed out by \cite{GomesJunior2015} are reduced with the STE ephemeris.
-\label{Fig: JPL-STE}}
-\end{figure}
-
-%\begin{figure*}
-%\subfigure{\includegraphics[width=8.8cm]{figures/JPL_DEC_Himalia.eps}    \label{Fig: JPL-STE-Himalia}}
-%\subfigure{\includegraphics[width=8.8cm]{figures/JPL_DEC_Carme.eps}  \label{Fig: JPL-STE-Carme}}
-%\caption{Geocentric right ascension and declination differences between the JUP300 JPL ephemeris and the special-tailored ephemeris for Himalia and Carme. \label{Fig: JPL-STE}}
-%\end{figure*}
-
-The obtained ephemeris is hereafter refered as STE, for special-tailored ephemeris.
-
-\subsection{Phoebe's ephemeris}
-
-For the specific case of Phoebe, the ninth satellite of Saturn, we have updated the ephemeris published in \cite{Desmars2013}.
-The new ephemeris (PH15) used the same dynamical model, including the perturbations of the Sun and the eight planets, the eight major satellites of Saturn and the $J_2$ parameter.
-The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \cite{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center\footnote{\url{http://lnfm1.sai.msu.ru/neb/nss/bsapoouf.htm}}), and observations from Flagstaff \citep{NOFS} between 2012 and 2014.
-It represents a total number of 5886 observations from 1898 to 2014.
-
-In Fig.
-\ref{Fig:eph-Phoebe} we compare our ephemeris (PH15) with the SAT375 JPL\footnote{Jacobson, R.A.
-2015-Feb-27.
-"Satellite Ephemeris: SAT375", JPL Satellite Ephemeris File Release, \url{ftp://ssd.jpl.nasa.gov/pub/eph/satellites/nio/LINUX_PC/sat375l.txt}} and the \cite{Emelyanov2007} ephemeris.
-We can see that our ephemeris is closer to the SAT375 JPL ephemeris where the difference between them is smaller than 20 mas (< 10 mas in Declination).
-This difference is smaller than the apparent diameter of Phoebe (see Table \ref{Tab: satellite-diameter})
-
-\begin{figure}
-\begin{centering}
-\includegraphics[width=8.8cm]{figures/Phoebe.eps} 
-\caption{Comparison between the PH15, SAT375 JPL and \cite{Emelyanov2007} ephemeris for the satellite Phoebe.}
-\label{Fig:eph-Phoebe}
-\end{centering}
-\end{figure}
-
-\subsection{Triton's and Nereid's ephemeris}
-
-For Triton, we used the most recent ephemeris published by \cite{Emelyanov2015}.
-In Fig.
-\ref{Fig:eph-Triton} we compare it with the ephemeris published in \cite{Zhang2014} and the NEP081 JPL ephemeris \citep{Jacobson2009}.
-The offsets between the three ephemeris are smaller than 15 mas for the period 2015-2018.
-This value is much smaller than the apparent size of Triton (124 mas, see Table \ref{Tab: satellite-diameter}), which indicates a good probability of observing an event by this object.
-
-\begin{figure*}
-\begin{centering}
-\subfigure[NEP081 JPL and \cite{Emelyanov2015}]{\includegraphics[width=8.8cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
-\subfigure[\cite{Zhang2014} and \cite{Emelyanov2015}]{\includegraphics[width=8.8cm]{figures/Zhang-EME.eps} \label{Fig:eph-Triton-zhang}}
-\caption{Comparison between the \cite{Emelyanov2015}, \cite{Zhang2014} and NEP081 JPL \citep{Jacobson2009} ephemeris for the satellite Triton.}
-\label{Fig:eph-Triton}
-\end{centering}
-\end{figure*}
-
-For Nereid, we used the ephemeris published by \cite{Emelyanov2011}, which uses more recent observations than the JPL ephemeris published by \cite{Jacobson2009}.
-The comparison between the two ephemeris can be seen in Fig.
-\ref{Fig:eph-Nereid}.
-The right ascension offsets between them are smaller than 60 mas and the declination ones are smaller than 15 mas.
-For Nereid, ephemeris errors in right ascension translate to uncertainties in the central instant of a stellar occultation, so offsets of 60 mas are not critical.
-The declination ephemeris offsets are of the order of the apparent diameter (15 mas, see Table \ref{Tab: satellite-diameter}), resulting in fair chances (50\%) that the shadow crosses the expected Earth latitude predicted for an occultation.
-% which is bigger than the satellite apparent diameter (see Table \ref{Tab: satellite-diameter}).
-Although this fact, the difference is bigger in Right Ascension than in Declination, which means that, in an event, the error would be smaller in the latitude where the shadow will pass on Earth than the error at the time of the occultation.
-
-\begin{figure}
-\begin{centering}
-\includegraphics[width=8.8cm]{figures/JPL-EME_Nereid.eps}
-\caption{Comparison between the \cite{Emelyanov2011} and NEP081 JPL \citep{Jacobson2009} ephemeris for the satellite Nereid.}
-\label{Fig:eph-Nereid}
-\end{centering}
-\end{figure}
-
-\cite{GomesJunior2015} published 902 new positions for Nereid from observations between 1992 and 2014.
-The development of a better ephemeris for Nereid using these new positions in the context discussed here of near future stellar occultations is desirable, but is out of the scope of this paper.
-
-%Making a new model for the orbits of these objects would demand a lot of time and delay the publication of predictions of possible events to occur in the very near future.
-Thus, we utilized the positions obtained by \cite{GomesJunior2015} to identify error patterns in the ephemeris.
-The error patterns in right ascension and in declination could be used to extrapolate the offsets to the satellite ephemeris by the time of the predicted occultation, improving it.
-Plots of the offsets over true anomaly (see Fig.
-\ref{Fig:offxanom} for declination offsets of Carme, for instance) clearly show the systematic errors in the JPL ephemeris.
-
-%\begin{figure}
-%\begin{centering}
-%\includegraphics[scale=0.25]{figures/DEC.png}\label{Fig:offxtime}
-%\caption{Offsets of the declination of Carme by time. \textcolor{red}{Figura só para visualização, vou colocar alguma melhor depois}}
-%\end{centering}
-%\end{figure}
-
-%\begin{figure*}
-%\begin{centering}
-%\subfigure[Offset from the JPL ephemeris]{\includegraphics[width=8.8cm]{figures/DEC_anom.png} \label{Fig:offxtanomxjpl}}
-%\subfigure[\textcolor{red}{Offset from the numerical integration}]{\includegraphics[width=8.8cm]{figures/DEC_anom.png} \label{Fig:offxtanomxni}}
-%\caption{Offsets of the declination of Carme by true anomaly from the JPL ephemeris and \textcolor{red}{from the numerical integration of the positions obtained in \cite{GomesJunior2015}}.
-%\label{Fig:offxanom}}
-%\end{centering}
-%\end{figure*}
-
-%The two kinds of error or offset patterns adopted depending on the case are given by Eqs. \ref{eq:linear} and \ref{eq:seno}:
+%\begin{thebibliography}{99}
+%\bibliography{references.bib}
+%\end{thebibliography}
 %
-%\begin{equation}
-%F(t,f) = p[0]\times t + p[1]\times \sin(f) + p[2]\times\cos(f) + p[3],
-%\label{eq:linear}
-%\end{equation}
-%
-%\begin{equation}
-%\begin{split}
-%%F = \{F_{x} \in  F_{c} &: (|S| > |C|) \\
-%% &\quad \cap (\text{minPixels}  < |S| < \text{maxPixels}) \\
-%% &\quad \cap (|S_{\text{conected}}| > |S| - \epsilon) \}
-%F(t,f) = p[0]\times\sin\left(\frac{2\pi}{p[1]}\times t + p[2]\right) + p[3]\times\sin(f) + \\ +p[4]\times\cos(f) + p[5],
-%\end{split}
-%\label{eq:seno}
-%\end{equation}
-%where $F(t,f)$ is the offset obtained, $t$ is time in years counting from J2000.0 and $f$ is the true anomaly.
-The patterns could be applied for right ascension as well as for declination ephemeris offsets.
-
-\section{Prediction of occultations} \label{Sec: predictions}
-
-The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the ephemeris presented in Sec.
-\ref{Sec: integration}.
-The search for stellar candidates follows the same procedure as presented by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.
-
-We predicted occultations for the 8 major irregular satellites of Jupiter,  Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope, and for Phoebe of Saturn and Triton and Nereid of Neptune.
-
-For Triton and Nereid, the candidates for stellar occultations in 2016 were searched using the WFI catalogue in the same way as the predictions for Centaurs and TNOs occultations by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.
-This catalogue contains the stars in the path of Neptune in the sky up to mid-2016.
-The catalogue was generated by observations made at the ESO 2p2 telescope (IAU code 809) using the Wide Field Imager (WFI) CCD mosaic detector.
-The filter used was the broad-band R filter ESO\#844 with $\lambda_c$ = 651.725 nm and $\Delta\lambda$ = 162.184 nm.
-
-A total of 396 events were identified between January 2016 and December 2017.
-In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite.
-Table \ref{Tab: occ-list} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps.
-Since these objects are very small, the duration of each event is a few seconds.
-All the occultation tables and maps will be publicly available at the CDS.
-No star brighter than MagR*=18 will be occulted by Triton in 2016.
-For this satellite, we cut events with stars fainter than R=18, since for Triton the flux drop during such an occultation would be very small.
-In Fig.
-\ref{Fig: ocultacao} we show an example of an occultation map.
-This is the only found occultation by Triton and it will happen in October 05, 2017.
-This event can be observed from Europe and the east coast of USA and will be a great opportunity to study in high resolution the atmosphere of the satellite.
-
-\begin{table}
-\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from January, 2016 up to December, 2017.}
-\begin{centering}
-\begin{tabular}{lccc}
-\hline  \hline
-%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
-Satellite  & 2016 & 2017 & Total \tabularnewline
-\hline
-Ananke & 12 & 16 & 28 \tabularnewline
-Carme & 20 & 14 & 34 \tabularnewline
-Elara & 14 & 16 & 30 \tabularnewline
-Himalia & 15 & 12 & 27 \tabularnewline
-Leda & 8 & 24 & 32 \tabularnewline
-Lysithea & 16 & 11 & 27 \tabularnewline
-Pasiphae & 20 & 19 & 39 \tabularnewline
-Sinope & 15 & 21 & 36 \tabularnewline
-\hdashline
-Phoebe\tablefootmark{a} & 32 & 98 & 130 \tabularnewline
-\hdashline
-Nereid\tablefootmark{a} & 11\tablefootmark{b} & 1 & 12 \tabularnewline
-Triton\tablefootmark{a} & -- & 1 & 1 \tabularnewline
-\hline
-\end{tabular}
-\par \end{centering}
-Occultations predicted using the UCAC4 catalogue and STE ephemeris.
-\tablefoottext{a}{Using JPL ephemeris.}
-\tablefoottext{b} {Using the WFI catalogue as explained in Sec. \ref{Sec: predictions}}.
-\end{table}
-
-\begin{table*}
-\caption{\label{Tab: sample-cds} A sample of stellar occultation predictions for Pasiphae}
-\begin{centering}
-\begin{tabular}{cccrcccrcrr}
-\hline
-\hline
-d m Year~~~~h~~m~~~s & RA~~~(ICRS)~~~Dec & C/A & P/A & $\nu$ & {\it D} & $R^*$ & $\lambda$ & LST & $\mu_{\alpha*}$ & $\mu_{\delta}$ \tabularnewline
-\hline
-% & \multicolumn{2}{c}{Mean errors} & Nr & Nr &  UCAC4       \tabularnewline
-09 04 2016 03:58:19. & 11 14 36.7707 +07 39 20.7610 & 1.003 &  17.9 & -12.88 &  4.54 & 14.9 & 271. & 22:03 &  12. & -33. \tabularnewline 
-13 06 2016 00:16:12. & 11 12 48.5020 +07 06 43.3520 & 0.661 &  30.0 & +14.32 &  5.50 & 13.9 & 262. & 17:45 &  -1. &   1. \tabularnewline 
-27 06 2016 13:56:09. & 11 18 03.4160 +06 23 45.1940 & 1.707 &  28.0 & +20.29 &  5.74 & 11.7 &  44. & 16:53 &   4. & -10. \tabularnewline 
-18 07 2016 15:07:24. & 11 28 15.5076 +05 05 31.8060 & 0.942 &  26.7 & +27.80 &  6.05 & 14.0 &   8. & 15:40 &   4. &   4. \tabularnewline 
-22 07 2016 16:15:07. & 11 30 30.4310 +04 48 43.4340 & 0.644 & 206.5 & +29.04 &  6.11 & 14.6 & 348. & 15:27 &  23. & -24. \tabularnewline 
-24 07 2016 01:37:34. & 11 31 17.8471 +04 42 49.0540 & 0.029 & 206.6 & +29.46 &  6.12 & 15.1 & 206. & 15:22 &   2. &  -8. \tabularnewline 
-24 07 2016 17:37:18. & 11 31 40.7472 +04 39 57.5060 & 0.840 &  26.5 & +29.66 &  6.13 & 14.9 & 326. & 15:20 & -11. &  -1. \tabularnewline 
-\hline
-\end{tabular}
-\par\end{centering}
-\textbf{Notes}.
-Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - at the central instant of occultation (corrected by proper motions); C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees (zero at north of the star, increasing clockwise); $\nu$: velocity in the plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; {\it D}: planet range to Earth, in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
-\label{Tab: occ-list}
-\end{table*}
-
-\begin{figure*}
-\begin{centering}
-\includegraphics[scale=0.4]{figures/Triton_2017-10-05T23:51:53.png}   
-\caption{Occultation map for Triton.% regarding to the first event sampled in Table \ref{Tab: occ-list}.
-The central red dot show the geocentric closest approach of the shadow.
-The small ones shows the center of the shadow separated by 60s.
-The lines show the path of the shadow over the Earth.
-The shadow moves from right to left.
-\textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for a central observation; C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; vel: velocity of event in km/s; Delta: Geocentric distance to the occulting object in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$; long: east longitude of subplanet point in degrees, positive towards east.}
-\label{Fig: ocultacao}
-\end{centering}
-\end{figure*}
-
-The first preliminary catalogue version of the ESA astrometry satellite GAIA \citep{deBruijne2012} is expected to be released up to the end of 2016.
-The precise star positions to be derived by GAIA will provide better predictions with the main source of error being the ephemeris.
-Astrometric reduction of observations published in \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions from 2017 onwards will be improved.
-Since we cannot foresee exactly when the GAIA catalogue will be released and when new and re-reduced satellite positions in the GAIA frame will become available, we decided to publish only the predictions for occultations restricted to the 2016-2017 time interval.
-
-\section{Occultation test} \label{Sec: testes}
-
-Observing a stellar occultation demands a great effort.
-And, in our case, the shadow covers a very restricted area on Earth because of the size of the irregular satellites.
-Since no stellar occultation by an irregular satellite was observed up to date, with the exception of Triton, and since we want to be sure that we can start observational campaigns with reasonable chances of success, we tested an occultation prediction for a large target, to assess the quality of the prediction.
-
-The test design consisted in observing the object and star to be occulted near the date of the event predicted when the two objects were present in the same field of view (FOV), close to each other.
-Thus, the relative positions between the two objects had minimal influence of the errors of the reference catalogue of stars used and possible field distortions \citep[and references therein]{Peng2008}.
-The relative positions of the star and satellite were used to check the original predictions.
-Notice that in the test we do not attempt to observe any actual occultation.
-The test could be performed at any site, regardless of the Earth location where the occultation would in fact be visible.
-
-We tested the occultation by Himalia predicted to occur on March 3, 2015. The shadow would cross the northern part of South America. For the event, four situations were considered:
-\begin{enumerate}[I]
-\item Our nominal, published prediction with the STE ephemeris (see Sec. \ref{Sec: integration}), and the nominal UCAC4 position of the star;
-\item Prediction with the JPL ephemeris and the nominal UCAC4 position of the star;
-\item From star and satellite offsets calculated from observations made a few days before the occultation when the objects were very separated (different FOVs);
-\item Same as 3 but with the star and the satellite close in the same FOV.
-\end{enumerate}
-
-Table \ref{Tab: comparison-Himalia} shows the differences between the predictions in the four situations.
-For situation 3 we observed the objects on February 22 with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / pixel$) at the Observatório do Pico dos Dias, Brazil (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
-On that day, Himalia and the star were observed in separate FOVs as they were still far apart.
-On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope (diameter = 1.6m; FOV = $5\farcm8$; pixel scale = $0\farcs17 / pixel$) at OPD just over an hour after the time scheduled for the event.
-Satellite and star were separated by about 16 arcsec, so very close to each other (situation 4).
-From the calculated offsets, the center of the shadow was obtained.
-Notice that the shadow path was not predicted to cross the OPD (which was located at almost 2000 km south from the shadow path).
-This was not necessary for testing the prediction.
-
-%\begin{figure*}
-%\begin{centering}
-%\subfigure[Prediction with STE ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_STE.eps}  \label{Fig: occ-Himalia-STE}}
-%\subfigure[Prediction with JPL ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_JPL.eps}  \label{Fig: occ-Himalia-JPL}}
-%\subfigure[Observation at February 22 (different FOVs)]{\includegraphics[width=8.8cm]{figures/Himalia_22fev.eps}   \label{Fig: occ-Himalia-off22fev}}
-%\subfigure[Observation at March 03 (same FOV)]{\includegraphics[width=8.8cm]{figures/Himalia_03mar.eps}   \label{Fig: occ-Himalia-off03mar}}
-%\caption{Predictions for Himalia: The straight lines show the size of the shadow.
-The big red dot show the geocentric closest approach of the shadow.
-The black "+" marks at maps (b,c,d) are the STE prediction closest approach for reference.
-The small red ones are the center of the shadow separated by one minute.
-\ref{Fig: occ-Himalia-STE} is the map using the STE ephemeris showed in Sec.
-\ref{Sec: integration}.
-\ref{Fig: occ-Himalia-JPL} shows the shadow using the JPL ephemeris.
-In \ref{Fig: occ-Himalia-off22fev} we apply offsets to the positions of star and satellite accordingly the observations made at February 22.
-\ref{Fig: occ-Himalia-off03mar} is as in \ref{Fig: occ-Himalia-off22fev} but with observations made at March 03 when the objects were close to each other.
-The green dot at the bottom of the maps is the location of Observatório do Pico dos Dias.}
-%\label{Fig: occ-Himalia}
-%\end{centering}
-%\end{figure*}
-
-%\begin{figure*}
-%\begin{centering}
-%\includegraphics[width=17.6cm]{figures/HIMALIA.eps}
-%\caption{Predictions for Himalia: The straight lines show the center of the shadow over time.
-In blue is for the prediction using the STE ephemeris showed in Sec.
-\ref{Sec: integration}.
-In red the shadow using the JPL ephemeris.
-In black we apply offsets to the positions of star and satellite accordingly the observations made at February 22.
-In green, the same for the previous one but with observations made at March 03 when the objects were close to each other.
-The dots show the closest approach of the shadow for each prediction.
-The scales are relative to the closest approach of the prediction using the STE ephemeris.
-The bars show the estimated diameter of the satellite and estimated error of the prediction.}
-%\label{Fig: occ-Himalia}
-%\end{centering}
-%\end{figure*}
-
-The critical parameter in the comparisons is the C/A, which here is related to latitudes.
-The apparent radius of Himalia is about 20 mas (see Table 1).
-In the context of the test, for a 0 mas offset in C/A we would have 100\% probability of observing the occultation, and 0\% in the case of a C/A offset equal to or larger than 20 mas, the radius of Himalia.
-From Table 5, we have nearly 0\% probability of success in situation 3, for which the offset in C/A was -20 mas, but when the relative astrometry was poor, 10 days prior to the event.
-Once at the day of the event in situation 4, the C/A offset dropped to -9 mas only, corresponding to a 55\% probability of success.
-Comparison with the prediction using the JPL ephemeris (situation 2) gives a +11 mas C/A offset, or a compatibility of 45\% between the ephemerides.
-All this suggests that there was a good probability of observing the event.
-The largest differences between the shadows of the four situations were 36s in time along the shadow path and 101km (31 mas) in the direction perpendicular to the shadows, suggesting that observers should be spread in narrow latitude ranges 100 km wide.
-
-\begin{table}
-\caption{\label{Tab: comparison-Himalia} Comparison between the predictions of the Himalia occultation at March 03, 2015.}
-\begin{centering}
-\begin{tabular}{lccc}
-\hline  \hline
-\multicolumn{4}{c}{Differences with respect to the STE prediction} \tabularnewline
-Method  & Instant of C/A  & C/A & Sit.   \tabularnewline
-\hline
-STE & 00:39:51 UTC & $0\farcs703$ & I \tabularnewline
-JPL & -26s & +11mas (36km) & II \tabularnewline % (284km)
-Feb. 22 Obs. & -14s & -20mas (65km) & III \tabularnewline % (153km)
-Mar. 03 Obs. & -36s & -09mas (29km) & IV \tabularnewline % (393km)
-\hline
-\end{tabular}
-\par\end{centering}
-C/A: geocentric closest approach; Sit: Situation test considered.
-\end{table}
-
-%The second test was with the satellite Elara, which is the second largest irregular satellite of Jupiter.
-The event was predicted to occur at March 30, 2015.
-The observations were made on March 25 and April 2, 2015 with the Zeiss telescope.
-On the night of April 2 the star and satellite could still be observed in the same FOV.
-Due to Elara being fainter than Himalia by 2 magnitudes, dispersions of the satellite positions on both nights ended up being higher than for Himalia.
-%Still, the differences between the maps obtained were relatively small.
-%The largest differences between them were 144s (186 mas) in time along the shadow path and 525 km (153 mas) perpendicular to it (see table \ref{Tab: comparison-Elara}).
-
-%\begin{table}
-%\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara occultation at March 30, 2015.}
-%\begin{centering}
-%\begin{tabular}{lcc}
-%\hline  \hline
-%\multicolumn{3}{c}{Difference from STE prediction} \tabularnewline
-%Method  & Instant of C/A  & C/A  \tabularnewline
-%\hline
-%STE & 01:45:13 UTC & $1\farcs082$ \tabularnewline
-%JPL & +02s & +57mas(195km) \tabularnewline
-%Mar. 25 Obs. & -70s & +20mas (68km) \tabularnewline
-%Apr. 02 Obs. & +74s & -96mas (330km) \tabularnewline
-%\hline
-%\end{tabular}
-%\par\end{centering}
-%\end{table}
-
-\section{Discussion} \label{Sec: discussion}
-
-We predict stellar occultations for the period of 2016-2017 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid.
-The procedure used was the same as that for the prediction of stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.
-
-The candidate stars were searched in the UCAC4 catalogue, except for the candidates in 2016 for Triton and Nereid.
-In this case, we used the WFI catalogue that was generated from observations made with ESO2p2/WFI CCD mosaic that covered the path of Neptune in the sky-plane up to 2016 (see Sec.
-\ref{Sec: predictions}).
-From this, a total of 396 events are foreseen.
-
-In a broader, general sense, the probability of successfully observing an occultation is roughly the ratio of the satellite's radius by the budget error (2 sigma for a 95\% confidence level) of ephemeris and star position.
-Thus, UCAC4 errors ranging between 20 mas - 50 mas (1 sigma) combined with a mean error (1 sigma) in the JPL ephemeris of 30 mas for Himalia and 150 mas for Leda published in Table 2 of \cite{Jacobson2012} would give 28\%-17\% probability of observing such an event by Himalia and $\approx2$\% for Leda, the smallest irregular satellite in the sample.
-Observations a few days before the date of occultation predicted may improve the combined errors to 40-80 mas, depending on the magnitude of the objects.
-
-The test made with an occultation expected to happen in March 03, 2015 for Himalia showed that this event would probably have been observed successfully in case there were observers available in the shadow area.
-%Fig.
-\ref{Fig: occ-Himalia} shows the maps made for 4 different offset corrections as explained in Sec.
-\ref{Sec: testes}.
-Table \ref{Tab: comparison-Himalia} shows numerically the difference between the maps.
-%A similar test was made for an occultation of Elara in March 30, 2015. %The difference between the maps in this case are listed in Table \ref{Tab: comparison-Elara}.
-The results show satisfying small offsets with respect to the local of the prediction. %We can expect similar results for other satellites and our method should allow successful stellar occultation.
-
-Continuous observations of the satellites are recommended and fitting of our dynamical model to those observations are expected to reduce the respective STE ephemeris errors.
-The first version of the GAIA catalogue is to be released up to the end of 2016 and will improve the position error of the stars to the 1-5 mas level.
-It will allow for the discovery of ocultations by fainter stars not present in the UCAC4 catalogue.
-% Consequently, the astrometry of the satellites with the GAIA catalogue will improve the position errors allowing for better predictions and a higher probabilities of observing stellar occultations by irregular satellites.
-The release of the GAIA catalogue should have a positive impact on both the astrometric precision of occulted stars and the reduction of astrometric positions of the satellites.
-As a result, prediction of stellar occultations by irregular satellites shall increase in number as well as in success.
-
-%We managed a large database with FITS images acquired by 5 telescopes in 3 sites between 1992 and 2014.
-From that, we identified 8466 observations of irregular satellites, from which we managed to obtain 6523 suitable astrometric positions, giving a total of 3666 positions for 12 satellites of Jupiter, 1920 positions for 4 satellites of Saturn, 35 positions for Sycorax (Uranus) and 902 positions for Nereid (Neptune).
-
-%The positions of all the objects were determined using the PRAIA package.
-The package was suited to cope with the huge amount of observations and the task of identifying the satellites within the database.
-PRAIA tasks were also useful to deal with the missing or incorrect coordinate and time stamps present mostly in the old observations.The UCAC4 was used as the reference frame.
-Based in the comparisons with ephemeris, we estimate that the position errors are about 60 mas to 80 mas depending on the satellite brightness.
-
-%For some satellites the number of positions obtained in this work is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012} (see Table \ref{Tab: comparison-horizons}).
-For instance, the amount of new positions for Himalia is about 70\% of the number used in the numerical integation of orbits by JPL.
-Systematic errors in the ephemeris were found for at least some satellites (Ananke, Carme, Elara and Pasiphae).
-In the case of Carme, we evidenced an error in the orbital inclination (see Fig.
-\ref{Fig: carme_anom}).
-
-%The positions derived in this work can be used in new orbital numerical integrations, generating more precise ephemerides.
-Stellar occultations by irregular satellites could then be better predicted.
-Based in this work, our group has already computed occultation predictions for the 8 major irregular satellites of Jupiter.
-These predictions will be published in a forthcoming paper.
-
-\begin{acknowledgements}
-
-ARG-J thanks the financial support of CAPES.
-MA thanks the CNPq (Grants 473002/2013-2 and 308721/2011-0) and FAPERJ (Grant E-26/111.488/2013).
-RV-M thanks grants: CNPq-306885/2013, Capes/Cofecub-2506/2015, Faperj/PAPDRJ-45/2013.
-JIBC acknowledges CNPq for a PQ2 fellowship (process number 308489/2013-6).
-BEM thanks the financial support of CAPES.
-FB-R acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014.
-The numerical model of the satellites of Jupiter was developped during a post-doctoral contract funded by the Chinese Academy of Sciences (CAS) and supported by the National Scientific Fund of China (NSFC)
-
-\end{acknowledgements}
-
-\bibliographystyle{aa}
-\nocite{*}
-\bibliography{references.bib}
+%\label{lastpage}
 

### 2016-02-29 Irregular-occ/discussion.tex
.

\section{Discussion} \label{Sec: discussion}

We predict stellar occultations for the period of 2016-2017 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid. The procedure used was the same as that for the prediction of stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.

The candidate stars were searched in the UCAC4 catalogue, except for the candidates in 2016 for Triton and Nereid. In this case, we used the WFI catalogue that was generated from observations made with ESO2p2/WFI CCD mosaic that covered the path of Neptune in the sky-plane up to 2016 (see Sec. \ref{Sec: predictions}). From this, a total of 396 events are foreseen.

In a broader, general sense, the probability of successfully observing an occultation is roughly the ratio of the satellite's radius by the budget error (2 sigma for a 95\% confidence level) of ephemeris and star position. Thus, UCAC4 errors ranging between 20 mas - 50 mas (1 sigma) combined with a mean error (1 sigma) in the JPL ephemeris of 30 mas for Himalia and 150 mas for Leda published in Table 2 of \cite{Jacobson2012} would give 28\%-17\% probability of observing such an event by Himalia and $\approx2$\% for Leda, the smallest irregular satellite in the sample. Observations a few days before the date of occultation predicted may improve the combined errors to 40-80 mas, depending on the magnitude of the objects.

The test made with an occultation expected to happen in March 03, 2015 for Himalia showed that this event would probably have been observed successfully in case there were observers available in the shadow area. %Fig. \ref{Fig: occ-Himalia} shows the maps made for 4 different offset corrections as explained in Sec. \ref{Sec: testes}. Table \ref{Tab: comparison-Himalia} shows numerically the difference between the maps.
%A similar test was made for an occultation of Elara in March 30, 2015. %The difference between the maps in this case are listed in Table \ref{Tab: comparison-Elara}.
The results show satisfying small offsets with respect to the local of the prediction. %We can expect similar results for other satellites and our method should allow successful stellar occultation.

Continuous observations of the satellites are recommended and fitting of our dynamical model to those observations are expected to reduce the respective STE ephemeris errors. The first version of the GAIA catalogue is to be released up to the end of 2016 and will improve the position error of the stars to the 1-5 mas level. It will allow for the discovery of ocultations by fainter stars not present in the UCAC4 catalogue. % Consequently, the astrometry of the satellites with the GAIA catalogue will improve the position errors allowing for better predictions and a higher probabilities of observing stellar occultations by irregular satellites.
The release of the GAIA catalogue should have a positive impact on both the astrometric precision of occulted stars and the reduction of astrometric positions of the satellites. As a result, prediction of stellar occultations by irregular satellites shall increase in number as well as in success.

%We managed a large database with FITS images acquired by 5 telescopes in 3 sites between 1992 and 2014. From that, we identified 8466 observations of irregular satellites, from which we managed to obtain 6523 suitable astrometric positions, giving a total of 3666 positions for 12 satellites of Jupiter, 1920 positions for 4 satellites of Saturn, 35 positions for Sycorax (Uranus) and 902 positions for Nereid (Neptune).

%The positions of all the objects were determined using the PRAIA package. The package was suited to cope with the huge amount of observations and the task of identifying the satellites within the database. PRAIA tasks were also useful to deal with the missing or incorrect coordinate and time stamps present mostly in the old observations.The UCAC4 was used as the reference frame. Based in the comparisons with ephemeris, we estimate that the position errors are about 60 mas to 80 mas depending on the satellite brightness.

%For some satellites the number of positions obtained in this work is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012} (see Table \ref{Tab: comparison-horizons}). For instance, the amount of new positions for Himalia is about 70\% of the number used in the numerical integation of orbits by JPL. Systematic errors in the ephemeris were found for at least some satellites (Ananke, Carme, Elara and Pasiphae). In the case of Carme, we evidenced an error in the orbital inclination (see Fig. \ref{Fig: carme_anom}).

%The positions derived in this work can be used in new orbital numerical integrations, generating more precise ephemerides. Stellar occultations by irregular satellites could then be better predicted. Based in this work, our group has already computed occultation predictions for the 8 major irregular satellites of Jupiter. These predictions will be published in a forthcoming paper.

### 2016-02-29 Irregular-occ/ephemeris.tex
.

\section{Ephemeris} \label{Sec: integration}

%\subsection{Observations used}
In order to improve ephemeris in short time spans in the near future for a more efficient prediction of stellar occultations by irregular satellites, we need predictions based on new and recent observations. In this context, \cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014. %They showed that the orbits of the irregular satellites of the giant planets have systematic errors.
%Their observed position offsets relative to the JPL ephemeris reached up to 200 mas for some satellites. At the distance of Jupiter, this represents an difference larger than 700 km in the shadow path of an occultation.%These differences could be associated with errors in their orbital elements.

Here we develop our own ephemeris based on the observations published in \cite{GomesJunior2015}. First, because the reduction was made with consistent and precise stellar catalogue and with a robust and reliable software (PRAIA, \cite{Assafin2011}). Second, besides recent observations, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets. For these reasons, only this set of positions was used for building the new ephemeris for the satellites of Jupiter.

%In order to have efficient ephemerides for predicting the stellar occultations by irregular satellites, we need predictions based on such recent observations. This is why we decided to develop our own ephemeris based on the observations published in \cite{GomesJunior2015}. First because many of the authors of this paper worked in the production of that set of positions, so that it was very well known. Second, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets. For these reasons, only this set of recent positions was used for building the new ephemeris for the satellites.

\subsection{Special tailored Ephemerides (STE) for Jupiter irregular satellites}

The last observations used to develop JPL current ephemeris of the irregular satellites of Jupiter were obtained in 2012 \citep{Jacobson2012}. As a result, the errors in the JPL ephemeris for the current epoch are large enough to prevent accurate predictions of stellar occultations without any corrections.

Our numerical model describes the dynamical evolution of the irregular satellites of Jupiter in a jovicentric reference frame. The satellites are submitted to the influence of the Sun and the rest of the solar system, as well as that of the Galilean satellites and the first harmonics of Jupiter's gravity field. The axes of the reference frame are those of the ICRS.

We use the following notations: \begin{itemize}
\item $i$ and $l$ one of the irregular satellites of Jupiter
\item $J$ Jupiter 
\item $j$ another body of the Solar System
\item $M_j$ the mass of the $j$-th body, not an irregular satellite
\item $m_i$ the mass of the irregular satellite $i$
\item $\vec{r_i}$ the position of the $i$-th body with respect to the barycentre of Jupiter System
\item $r_{ij}$ the distance between bodies $i$ and $j$  
\item $R_J$ the radius of Jupiter
\item $J_n$ the dynamic polar oblateness of the n-th order for Jupiter's gravity field
\item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
\item $\Phi_i$ is the inclination of the $i$-th satellite with respect to Jupiter's equator.
\end{itemize}

For an irregular satellite $i$, under the gravitational influence of Jupiter, the $\mathcal{N}-1$ other irregular satellites, the regular Jovian satellites and the rest of the Solar System ($\mathcal{N}$ bodies), the equation of motion is:
\begin{equation}\begin{array}{ll}

\ddot{\vec{r_i}}= & \displaystyle -GM_J\frac{\vec{r_J}-\vec{r_i}}{r_{iJ}^3}-\sum_{l=1,l\neq i}^\mathcal{N'}Gm_l\frac{\vec{r_l}-\vec{r_i}}{r_{il}^3}\\
&\displaystyle -\sum_{j=1}^\mathcal{N}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3} \right)\\
 & \displaystyle +GM_J \nabla U_{\bar{l}\hat{J}} -\sum_{l=1}^\mathcal{N} Gm_l\nabla U_{\bar{l}\hat{J}}
\end{array}
\label{Eq:1}
\end{equation}
where the last term in the brackets and the last term in Eq. \ref{Eq:1} represent undirect perturbations. The oblateness potential seen by the body $i$ because of Jupiter is:
\begin{equation}\begin{array}{ll}

U_{\bar{l}\hat{J}}=&\displaystyle -\frac{R_J^2 J_2}{r_{iJ}^3}\left(\frac{3}{2}\sin^2 \Phi_i-\frac{1}{2}\right)\\ &\\ & 
\displaystyle-\frac{R_J^4 J_4}{r_{iJ}^5}\left(\frac{35}{8}\sin^4 \Phi_i-\frac{15}{4}\sin^2 \Phi_i+\frac{3}{8}\right)\\
& \\
&\displaystyle-\frac{R_J^6 J_6}{r_{iJ}^7}\left(\frac{231}{16}\sin^6 \Phi_i-\frac{315}{16}\sin^4 \Phi_i+\frac{105}{16}\sin^2 \Phi_i-\frac{5}{16}\right)

\end{array}
\end{equation}

The expressions of $\nabla_lU_{\bar{l}\, \hat{i}}$ and $\nabla_i U_{\bar{i}\, \hat{l}}.$ have been developed in \citet{Lainey2004}. The equations of motion are integrated with the numerical integrator RADAU \citep{Everhart1985}. 
Our model was fitted to the observations through a least-squares procedure. The satellites were integrated one dynamical family at a time, to gain computing time, while losing minimum precision. Indeed, the interactions between satellites not belonging to the same dynamical family are negligible considering the short timespan of our integration.

The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}.%, while the residuals are presented in Fig \ref{Fig: residuals} for Himalia and Carme as representatives.

%\begin{figure*}
%\subfigure{\includegraphics[width=8.8cm]{figures/Himalia_resid.eps}   \label{Fig: residuals-Himalia}}
%\subfigure{\includegraphics[width=8.8cm]{figures/Carme_resid.eps}  \label{Fig: residuals-Carme}}
%\caption{Residuals in topocentric right ascension and declination of the numerical integration for Himalia and Carme. \label{Fig: residuals}}
%\end{figure*}

%\begin{table*}
%\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0.% For Themisto's case, the small number of observations meant that the statistical uncertainty was even greater than the obtained elements, though the fitting proved satisfying. We decided to provide its elements with significant number comparable to the other satellite for information, but left it without uncertainty.
% }\label{Tab: sat_ell}
%\begin{center}
%\begin{tabular}{ccccccc}
%\hline\hline
%Satellite & a (km) & e & I$\degr$ & $\Omega\degr$ & $\omega\degr$ & $v\degr$ \\ 
%\hline
%Himalia &   11372100 $\pm$ 500    &    0.166 $\pm$ 0.002      &   45.14 $\pm$ 0.15      &   39.77 $\pm$ 0.19      &   351.48 $\pm$ 0.46      &   97.35 $\pm$ 0.48    \\
%Elara &   11741170 $\pm$ 690  &      0.222 $\pm$ 0.002      &   28.64 $\pm$ 0.18      &   68.42 $\pm$ 0.43      &   179.82 $\pm$ 0.56      &   339.08 $\pm$ 0.82  \\
%Lysithea &   11739900 $\pm$ 1300  &      0.136 $\pm$ 0.004      &    51.12 $\pm$ 0.27     &   5.53 $\pm$ 0.52      &   53.0 $\pm$ 1.5      &   318.9 $\pm$ 2.0   \\
%Leda &   11140300  $\pm$ 4300  &     0.173  $\pm$ 0.007     &   16.15  $\pm$ 0.75    &   272.6  $\pm$ 1.7    &   212.2  $\pm$ 3.6          &   218.8  $\pm$ 3.2  \\
%Pasiphae &  23425000  $\pm$ 5000    &     0.379  $\pm$ 0.001       &   152.44 $\pm$ 0.10      &   284.59 $\pm$ 0.21      &   135.96 $\pm$ 0.19      &   236.97 $\pm$ 0.16 \\
%Sinope &   22968800 $\pm$ 5200   &     0.316 $\pm$ 0.002      &   157.76 $\pm$ 0.12      &   256.62 $\pm$ 0.55      &   298.38 $\pm$ 0.55      &   167.57 $\pm$ 0.19    \\
%Carme &   24202924 $\pm$ 4800      &  0.242 $\pm$ 0.001      &   147.13 $\pm$ 0.10      &   154.01 $\pm$ 0.25      &   47.90 $\pm$ 0.29      &   234.41 $\pm$ 0.19  \\
%Ananke &  21683800  $\pm$ 7200  &     0.380 $\pm$ 0.002      &   172.29 $\pm$ 0.20      &   56.9 $\pm$ 1.2      &   123.3 $\pm$ 1.2      &   231.24 $\pm$ 0.21  \\
%%Themisto  & 7393800    &     0.198       &   25.77  &   220.0     &   216.3    &   262.1     \\
%\hline
%\end{tabular} 
%\end{center}
%\textbf{Notes}: a: semimajor axis; e : excentricity; I: inclination relative to the equatorial reference plane J2000; $\Omega$: longitude of the ascending node; $\omega$: argument of periapsis; $v$: true anomaly.
%\end{table*}

All the orbits determined for the satellites show satisfying residuals. %Yet, one of the satellites has a peculiar situation. Its small number of observations in our set means that its orbit is definitely loosely constrained. 
The residuals are smaller than those obtained with JPL ephemeris, which was expected because the accuracy of an ephemeris decreases when we get further from the time of observations. The main risk of divergence over time comes from the possible absence of long-term effects when fitting to a short timespan of observations. If that were the case, our ephemeris would diverge too quickly to be of any use. JPL ephemerides are fitted over all the available observations. As a result, they will diverge less quickly than our own. Though they are no longer precise enough for our use, they remain a precious reference to identify whether our own model presents a quick divergence.

We compared our ephemeris to the JPL for all the Jupiter satellites we fitted, until 2018. For instance, the divergence between 2015 and 2018 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.

Fig. \ref{Fig: JPL-STE} displays the offsets of the positions published by \cite{GomesJunior2015} for the satellite Carme in declination relative to our ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.We see that the systematic JPL ephemeris offsets pointed out by \cite{GomesJunior2015} are reduced with our ephemeris, as expected.

\begin{figure}
\includegraphics[width=8.8cm]{figures/Carme_ephemeris.eps} 
\caption{Offsets in declination of the positions published by \cite{GomesJunior2015} for Carme. The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the JUP300 JPL ephemeris and the green dot relative to \cite{Emelyanov2008}. As expected, the ephemeris systematic errors pointed out by \cite{GomesJunior2015} are reduced with the STE ephemeris.  \label{Fig:JPL-STE}}
\end{figure}

%\begin{figure*}
%\subfigure{\includegraphics[width=8.8cm]{figures/JPL_DEC_Himalia.eps}    \label{Fig: JPL-STE-Himalia}}
%\subfigure{\includegraphics[width=8.8cm]{figures/JPL_DEC_Carme.eps}  \label{Fig: JPL-STE-Carme}}
%\caption{Geocentric right ascension and declination differences between the JUP300 JPL ephemeris and the special-tailored ephemeris for Himalia and Carme. \label{Fig: JPL-STE}}
%\end{figure*}

The obtained ephemeris is hereafter refered as STE, for special-tailored ephemeris.

\subsection{Phoebe's ephemeris}

For the specific case of Phoebe, the ninth satellite of Saturn, we have updated the ephemeris published in \cite{Desmars2013}. The new ephemeris (PH15) used the same dynamical model, including the perturbations of the Sun and the eight planets, the eight major satellites of Saturn and the $J_2$ parameter. The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \cite{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center\footnote{\url{http://lnfm1.sai.msu.ru/neb/nss/bsapoouf.htm}}), and observations from Flagstaff \citep{NOFS} between 2012 and 2014. It represents a total number of 5886 observations from 1898 to 2014.

In Fig. \ref{Fig:eph-Phoebe} we compare our ephemeris (PH15) with the SAT375 JPL\footnote{Jacobson, R.A. 2015-Feb-27. "Satellite Ephemeris: SAT375", JPL Satellite Ephemeris File Release, \url{ftp://ssd.jpl.nasa.gov/pub/eph/satellites/nio/LINUX_PC/sat375l.txt}} and the \cite{Emelyanov2007} ephemeris. We can see that our ephemeris is closer to the SAT375 JPL ephemeris where the difference between them is smaller than 20 mas (< 10 mas in Declination). This difference is smaller than the apparent diameter of Phoebe (see Table \ref{Tab: satellite-diameter})

\begin{figure}
\begin{centering}
\includegraphics[width=8.8cm]{figures/Phoebe.eps} 
\caption{Comparison between the PH15, SAT375 JPL and \cite{Emelyanov2007} ephemeris for the satellite Phoebe.}
\label{Fig:eph-Phoebe}
\end{centering}
\end{figure}

\subsection{Triton's and Nereid's ephemeris}

For Triton, we used the most recent ephemeris published by \cite{Emelyanov2015}. In Fig. \ref{Fig:eph-Triton} we compare it with the ephemeris published in \cite{Zhang2014} and the NEP081 JPL ephemeris \citep{Jacobson2009}. The offsets between the three ephemeris are smaller than 15 mas for the period 2015-2018. This value is much smaller than the apparent size of Triton (124 mas, see Table \ref{Tab: satellite-diameter}), which indicates a good probability of observing an event by this object.

\begin{figure*}
\begin{centering}
\subfigure[NEP081 JPL and \cite{Emelyanov2015}]{\includegraphics[width=8.8cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
\subfigure[\cite{Zhang2014} and \cite{Emelyanov2015}]{\includegraphics[width=8.8cm]{figures/Zhang-EME.eps} \label{Fig:eph-Triton-zhang}}
\caption{Comparison between the \cite{Emelyanov2015}, \cite{Zhang2014} and NEP081 JPL \citep{Jacobson2009} ephemeris for the satellite Triton.}
\label{Fig:eph-Triton}
\end{centering}
\end{figure*}

For Nereid, we used the ephemeris published by \cite{Emelyanov2011}, which uses more recent observations than the JPL ephemeris published by \cite{Jacobson2009}. The comparison between the two ephemeris can be seen in Fig. \ref{Fig:eph-Nereid}. The right ascension offsets between them are smaller than 60 mas and the declination ones are smaller than 15 mas. For Nereid, ephemeris errors in right ascension translate to uncertainties in the central instant of a stellar occultation, so offsets of 60 mas are not critical. The declination ephemeris offsets are of the order of the apparent diameter (15 mas, see Table \ref{Tab: satellite-diameter}), resulting in fair chances (50\%) that the shadow crosses the expected Earth latitude predicted for an occultation. % which is bigger than the satellite apparent diameter (see Table \ref{Tab: satellite-diameter}). Although this fact, the difference is bigger in Right Ascension than in Declination, which means that, in an event, the error would be smaller in the latitude where the shadow will pass on Earth than the error at the time of the occultation.

\begin{figure}
\begin{centering}
\includegraphics[width=8.8cm]{figures/JPL-EME_Nereid.eps}
\caption{Comparison between the \cite{Emelyanov2011} and NEP081 JPL \citep{Jacobson2009} ephemeris for the satellite Nereid.}
\label{Fig:eph-Nereid}
\end{centering}
\end{figure}

\cite{GomesJunior2015} published 902 new positions for Nereid from observations between 1992 and 2014. The development of a better ephemeris for Nereid using these new positions in the context discussed here of near future stellar occultations is desirable, but is out of the scope of this paper.

%Making a new model for the orbits of these objects would demand a lot of time and delay the publication of predictions of possible events to occur in the very near future. Thus, we utilized the positions obtained by \cite{GomesJunior2015} to identify error patterns in the ephemeris. The error patterns in right ascension and in declination could be used to extrapolate the offsets to the satellite ephemeris by the time of the predicted occultation, improving it. Plots of the offsets over true anomaly (see Fig. \ref{Fig:offxanom} for declination offsets of Carme, for instance) clearly show the systematic errors in the JPL ephemeris.

%\begin{figure}
%\begin{centering}
%\includegraphics[scale=0.25]{figures/DEC.png}\label{Fig:offxtime}
%\caption{Offsets of the declination of Carme by time. \textcolor{red}{Figura só para visualização, vou colocar alguma melhor depois}}
%\end{centering}
%\end{figure}

%\begin{figure*}
%\begin{centering}
%\subfigure[Offset from the JPL ephemeris]{\includegraphics[width=8.8cm]{figures/DEC_anom.png} \label{Fig:offxtanomxjpl}}
%\subfigure[\textcolor{red}{Offset from the numerical integration}]{\includegraphics[width=8.8cm]{figures/DEC_anom.png} \label{Fig:offxtanomxni}}
%\caption{Offsets of the declination of Carme by true anomaly from the JPL ephemeris and \textcolor{red}{from the numerical integration of the positions obtained in \cite{GomesJunior2015}}.
%\label{Fig:offxanom}}
%\end{centering}
%\end{figure*}

%The two kinds of error or offset patterns adopted depending on the case are given by Eqs. \ref{eq:linear} and \ref{eq:seno}:
%
%\begin{equation}
%F(t,f) = p[0]\times t + p[1]\times \sin(f) + p[2]\times\cos(f) + p[3],
%\label{eq:linear}
%\end{equation}
%
%\begin{equation}
%\begin{split}
%%F = \{F_{x} \in  F_{c} &: (|S| > |C|) \\
%% &\quad \cap (\text{minPixels}  < |S| < \text{maxPixels}) \\
%% &\quad \cap (|S_{\text{conected}}| > |S| - \epsilon) \}
%F(t,f) = p[0]\times\sin\left(\frac{2\pi}{p[1]}\times t + p[2]\right) + p[3]\times\sin(f) + \\ +p[4]\times\cos(f) + p[5],
%\end{split}
%\label{eq:seno}
%\end{equation}
%where $F(t,f)$ is the offset obtained, $t$ is time in years counting from J2000.0 and $f$ is the true anomaly. The patterns could be applied for right ascension as well as for declination ephemeris offsets.

### 2016-02-29 Irregular-occ/intro.tex
.

\section{Introduction}\label{Sec: introducao}

Irregular satellites revolve around giant planets at large distances in eccentric, highly inclined and frequently retrograde orbits. Because of these peculiar orbits, it is largely accepted that these objects did not form by accretion around their planet, but were captured in the early solar system \citep{Sheppard2005}.

There is no consensus for a single model explaining where the irregular satellites were formed. \cite{Cuk2004} showed that the progenitor of the Himalia group may have originated in heliocentric orbits similar to the Hilda asteroid group. \cite{Sheppard2005} stated that the irregular satellites may be some of the objects that were formed within the giant planets region.

\cite{Grav2003} and \cite{Grav2007} showed that the irregular satellites from the giant planets have their colors and spectral slopes similar to C-, D- and P-type asteroids, Centaurs and trans-neptunian objects (TNOs). This suggests that they may have come from different locations in the early solar system.

\cite{Sheppard2005} and \cite{Jewitt2007} also explored the possibility that the irregular satellites originated as comets or TNOs. TNOs are highly interesting objects that, due to their large heliocentric distances, may be highly preserved with physical properties similar to those they had when they were formed \citep{Barucci2008}. This is even more true for the smaller objects, since in principle larger sizes favour physical differentiation processes in the body and vice-versa. However, due to the distance, the smaller TNOs from this region are more difficult to observe. Thus, if irregular satellites - or at least a few of them - do share a common origin with small TNOs, and since these objects are situated at much closer heliocentric distances now, this gives a unique chance of observing and studying representatives of this specific TNO population in much greater detail than could ever be possible by direct observation of this population in the Kuiper Belt.

Phoebe is the most studied irregular satellite. \cite{Clark2005} suggest that its surface is probably covered by material of cometary origin. It was also stated by \cite{Johnson2005} that if the porosity of Phoebe is 15\%, Phoebe would have an uncompressed density similar to those of Pluto and Triton.

In order to obtain precise fundamental physical parameters like size and shape, thus constraining the albedo and in a broader sense also the composition for the irregular satellites and therefore to contribute to the study of their origin, we aim at observing stellar occultations, which provide  more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}. For that, reliable predictions of stellar occultations by these satellites are most needed.

We present in this paper stellar occultation predictions for the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda), Phoebe of Saturn and, Nereid and Triton of Neptune. Phoebe, being the most studied object with a good measured size, can be used to calibrate and evaluate the technique for similar objects.

Triton is an uncommon satellite. Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones. Because Triton's orbit size is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded from the irregular satellites' class, but still studied together by many authors \citep{Sheppard2005, Jewitt2007}. Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have the same origin as the TNOs \citep{Agnor2006}. However, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere. The main motivation to study Triton by stellar occultations is to understand the evolution of its atmosphere due to Triton's complicated and extreme seasonal cycle \citep{McKinnon2007, Elliot_2000}. For all these reasons, Triton was also included as a target in this work.

Excluding Triton \citep{Olkin1997, Elliot_2000}, no observation of a stellar occultation by an irregular satellite was published up to date. Since their estimated sizes are very small (see Table \ref{Tab: satellite-diameter}), this may have discouraged earlier tries. But, in fact, given their relatively closer distances as compared to TNOs and Centaurs, and considering the current precision of their ephemeris and of star positions, we can now reliably predict the exact location and instant where the shadow of the occultation will cross the Earth. For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds). Thus, in this case, if the accumulated error (2 sigma or 95\% confidence level) of ephemeris and star position be around 70 mas, we have a probability of about 30\% of observing a stellar occultation, which is quite satisfactory today (see discussion in section \ref{Sec: discussion}).

%\begin{table}
%\caption{\label{Tab: satellite-diameter} Estimated diameter of the satellites and correspondent apparent diameter}
%\begin{centering}
%\begin{tabular}{lccc}
%\hline  \hline
%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
%Satellite  & mas\tablefootmark{a}  & km & Ref. \tabularnewline
%\hline
%Ananke & 8 & 29 & 1 \tabularnewline
%Carme & 13 & 46 & 1 \tabularnewline
%Elara & 24 & 86 & 1 \tabularnewline
%Himalia & 41 & $(150\times120) \pm 20$\tablefootmark{b} & 2 \tabularnewline
%Leda & 5 & 20 & 1 \tabularnewline
%Lysithea & 10 & 36 & 1 \tabularnewline
%Pasiphae & 17 & 62 & 1 \tabularnewline
%Sinope & 10 & 37 & 1 \tabularnewline
%\hdashline
%Phoebe & 32 & $212 \pm 1.4$\tablefootmark{b} & 3 \tabularnewline
%\hdashline
%Nereid & 15 & $340 \pm 50$\tablefootmark{c} & 4 \tabularnewline
%Triton & 124 & $2707 \pm 2.0$\tablefootmark{c} & 5 \tabularnewline
%\hline
%\end{tabular}
%\tablebib{
%(1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}; (4) \cite{Thomas1991}; (5) \cite{Thomas2000}.}
%\end{centering}
%\tablefoottext{a}{Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
%\tablefoottext{b}{From Cassini observations.}
%\tablefoottext{c}{From Voyager-2 observations.}
%\par
%\end{table}

\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas. For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}. They pointed out that the ephemeris of the irregular satellites have systematic errors that may reach 200 mas for some satellites. For an object at the distance of Jupiter, this represents an error larger than 700 km in the shadow path. Using the positions obtained by \cite{GomesJunior2015} we produced a specific short-time ephemeris for the satellites of Jupiter and Phoebe to better predict stellar occultations by these objects.

Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved. In contrast to TNOs, the irregular satellites have much better ephemeris because the orbits of their host planets are better  known, their observational time span is much wider and covers many orbital periods. Moreover, the irregular satellites are much closer to Earth which implies in a much smaller shadow path error in kilometers. These advantages may be somewhat balanced by the smaller sizes estimated for the irregular satellites.Thus, in comparison, the chances for a successful observation of an stellar occultation by an irregular satellite should be considered at least also as good as those by TNOs.

%From what we have just presented, we can think that the observation of an stellar occultation by an irregular satellite should be likelier than occultations by TNOs. The orbits of their host planets are well known and their observation time-span covers many orbital periods, contrary to TNOs.
%Unlike stellar occultations by TNOs, which nevertheless have been proved effective, the observation of an occultation by a irregular satellite is, in principle, more favorable. The orbits of their host planets are well known and these satellites have been observed completing already many orbital periods around them, thus presenting better ephemeris than TNOs. 
%Moreover, the irregular satellites are closer to Earth which means a minor localization error in kilometer.

In section \ref{Sec: integration} we show the building of the production of the new ephemeris. In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites and how they were made. Some tests made to check the accuracy of the predictions are presented in section \ref{Sec: testes} and the final discussion is presented in section \ref{Sec: discussion}.

### 2016-02-29 Irregular-occ/prediction.tex
.

\section{Prediction of occultations} \label{Sec: predictions}

The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the ephemeris presented in Sec. \ref{Sec: integration}. The search for stellar candidates follows the same procedure as presented by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.

We predicted occultations for the 8 major irregular satellites of Jupiter,  Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope, and for Phoebe of Saturn and Triton and Nereid of Neptune.

For Triton and Nereid, the candidates for stellar occultations in 2016 were searched using the WFI catalogue in the same way as the predictions for Centaurs and TNOs occultations by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}. This catalogue contains the stars in the path of Neptune in the sky up to mid-2016. The catalogue was generated by observations made at the ESO 2p2 telescope (IAU code 809) using the Wide Field Imager (WFI) CCD mosaic detector. The filter used was the broad-band R filter ESO\#844 with $\lambda_c$ = 651.725 nm and $\Delta\lambda$ = 162.184 nm.

A total of 396 events were identified between January 2016 and December 2017. In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite. Table \ref{Tab: occ-list} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps. Since these objects are very small, the duration of each event is a few seconds. All the occultation tables and maps will be publicly available at the CDS. No star brighter than MagR*=18 will be occulted by Triton in 2016. For this satellite, we cut events with stars fainter than R=18, since for Triton the flux drop during such an occultation would be very small. In Fig. \ref{Fig: ocultacao} we show an example of an occultation map. This is the only found occultation by Triton and it will happen in October 05, 2017. This event can be observed from Europe and the east coast of USA and will be a great opportunity to study in high resolution the atmosphere of the satellite.

\begin{table}
\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from January, 2016 up to December, 2017.}
\begin{centering}
\begin{tabular}{lccc}
\hline  \hline
%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
Satellite  & 2016 & 2017 & Total \tabularnewline
\hline
Ananke & 12 & 16 & 28 \tabularnewline
Carme & 20 & 14 & 34 \tabularnewline
Elara & 14 & 16 & 30 \tabularnewline
Himalia & 15 & 12 & 27 \tabularnewline
Leda & 8 & 24 & 32 \tabularnewline
Lysithea & 16 & 11 & 27 \tabularnewline
Pasiphae & 20 & 19 & 39 \tabularnewline
Sinope & 15 & 21 & 36 \tabularnewline
\hdashline
Phoebe\tablefootmark{a} & 32 & 98 & 130 \tabularnewline
\hdashline
Nereid\tablefootmark{a} & 11\tablefootmark{b} & 1 & 12 \tabularnewline
Triton\tablefootmark{a} & -- & 1 & 1 \tabularnewline
\hline
\end{tabular}
\par \end{centering}
Occultations predicted using the UCAC4 catalogue and STE ephemeris.
\tablefoottext{a}{Using JPL ephemeris.}
\tablefoottext{b} {Using the WFI catalogue as explained in Sec. \ref{Sec: predictions}}.
\end{table}

\begin{table*}
\caption{\label{Tab: sample-cds} A sample of stellar occultation predictions for Pasiphae}
\begin{centering}
\begin{tabular}{cccrcccrcrr}
\hline
\hline
d m Year~~~~h~~m~~~s & RA~~~(ICRS)~~~Dec & C/A & P/A & $\nu$ & {\it D} & $R^*$ & $\lambda$ & LST & $\mu_{\alpha*}$ & $\mu_{\delta}$ \tabularnewline
\hline
% & \multicolumn{2}{c}{Mean errors} & Nr & Nr &  UCAC4       \tabularnewline
09 04 2016 03:58:19. & 11 14 36.7707 +07 39 20.7610 & 1.003 &  17.9 & -12.88 &  4.54 & 14.9 & 271. & 22:03 &  12. & -33. \tabularnewline 
13 06 2016 00:16:12. & 11 12 48.5020 +07 06 43.3520 & 0.661 &  30.0 & +14.32 &  5.50 & 13.9 & 262. & 17:45 &  -1. &   1. \tabularnewline 
27 06 2016 13:56:09. & 11 18 03.4160 +06 23 45.1940 & 1.707 &  28.0 & +20.29 &  5.74 & 11.7 &  44. & 16:53 &   4. & -10. \tabularnewline 
18 07 2016 15:07:24. & 11 28 15.5076 +05 05 31.8060 & 0.942 &  26.7 & +27.80 &  6.05 & 14.0 &   8. & 15:40 &   4. &   4. \tabularnewline 
22 07 2016 16:15:07. & 11 30 30.4310 +04 48 43.4340 & 0.644 & 206.5 & +29.04 &  6.11 & 14.6 & 348. & 15:27 &  23. & -24. \tabularnewline 
24 07 2016 01:37:34. & 11 31 17.8471 +04 42 49.0540 & 0.029 & 206.6 & +29.46 &  6.12 & 15.1 & 206. & 15:22 &   2. &  -8. \tabularnewline 
24 07 2016 17:37:18. & 11 31 40.7472 +04 39 57.5060 & 0.840 &  26.5 & +29.66 &  6.13 & 14.9 & 326. & 15:20 & -11. &  -1. \tabularnewline 
\hline
\end{tabular}
\par\end{centering}
\textbf{Notes}. Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - at the central instant of occultation (corrected by proper motions); C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees (zero at north of the star, increasing clockwise); $\nu$: velocity in the plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; {\it D}: planet range to Earth, in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
\label{Tab: occ-list}
\end{table*}

\begin{figure*}
\begin{centering}
\includegraphics[scale=0.4]{figures/Triton_2017-10-05T23:51:53.png}   
\caption{Occultation map for Triton.% regarding to the first event sampled in Table \ref{Tab: occ-list}.
The central red dot show the geocentric closest approach of the shadow. The small ones shows the center of the shadow separated by 60s. The lines show the path of the shadow over the Earth. The shadow moves from right to left.
\textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for a central observation; C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; vel: velocity of event in km/s; Delta: Geocentric distance to the occulting object in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$; long: east longitude of subplanet point in degrees, positive towards east.}
\label{Fig: ocultacao}
\end{centering}
\end{figure*}

The first preliminary catalogue version of the ESA astrometry satellite GAIA \citep{deBruijne2012} is expected to be released up to the end of 2016. The precise star positions to be derived by GAIA will provide better predictions with the main source of error being the ephemeris. Astrometric reduction of observations published in \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions from 2017 onwards will be improved. Since we cannot foresee exactly when the GAIA catalogue will be released and when new and re-reduced satellite positions in the GAIA frame will become available, we decided to publish only the predictions for occultations restricted to the 2016-2017 time interval.

\section{Occultation test} \label{Sec: testes}

Observing a stellar occultation demands a great effort. And, in our case, the shadow covers a very restricted area on Earth because of the size of the irregular satellites. Since no stellar occultation by an irregular satellite was observed up to date, with the exception of Triton, and since we want to be sure that we can start observational campaigns with reasonable chances of success, we tested an occultation prediction for a large target, to assess the quality of the prediction.

The test design consisted in observing the object and star to be occulted near the date of the event predicted when the two objects were present in the same field of view (FOV), close to each other. Thus, the relative positions between the two objects had minimal influence of the errors of the reference catalogue of stars used and possible field distortions \citep[and references therein]{Peng2008}. The relative positions of the star and satellite were used to check the original predictions. Notice that in the test we do not attempt to observe any actual occultation. The test could be performed at any site, regardless of the Earth location where the occultation would in fact be visible.

We tested the occultation by Himalia predicted to occur on March 3, 2015. The shadow would cross the northern part of South America. For the event, four situations were considered:
\begin{enumerate}[I]
\item Our nominal, published prediction with the STE ephemeris (see Sec. \ref{Sec: integration}), and the nominal UCAC4 position of the star;
\item Prediction with the JPL ephemeris and the nominal UCAC4 position of the star;
\item From star and satellite offsets calculated from observations made a few days before the occultation when the objects were very separated (different FOVs);
\item Same as 3 but with the star and the satellite close in the same FOV.
\end{enumerate}

Table \ref{Tab: comparison-Himalia} shows the differences between the predictions in the four situations. For situation 3 we observed the objects on February 22 with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / pixel$) at the Observatório do Pico dos Dias, Brazil (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m). On that day, Himalia and the star were observed in separate FOVs as they were still far apart. On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope (diameter = 1.6m; FOV = $5\farcm8$; pixel scale = $0\farcs17 / pixel$) at OPD just over an hour after the time scheduled for the event. Satellite and star were separated by about 16 arcsec, so very close to each other (situation 4). From the calculated offsets, the center of the shadow was obtained. Notice that the shadow path was not predicted to cross the OPD (which was located at almost 2000 km south from the shadow path). This was not necessary for testing the prediction.

%\begin{figure*}
%\begin{centering}
%\subfigure[Prediction with STE ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_STE.eps}  \label{Fig: occ-Himalia-STE}}
%\subfigure[Prediction with JPL ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_JPL.eps}  \label{Fig: occ-Himalia-JPL}}
%\subfigure[Observation at February 22 (different FOVs)]{\includegraphics[width=8.8cm]{figures/Himalia_22fev.eps}   \label{Fig: occ-Himalia-off22fev}}
%\subfigure[Observation at March 03 (same FOV)]{\includegraphics[width=8.8cm]{figures/Himalia_03mar.eps}   \label{Fig: occ-Himalia-off03mar}}
%\caption{Predictions for Himalia: The straight lines show the size of the shadow. The big red dot show the geocentric closest approach of the shadow. The black "+" marks at maps (b,c,d) are the STE prediction closest approach for reference. The small red ones are the center of the shadow separated by one minute. \ref{Fig: occ-Himalia-STE} is the map using the STE ephemeris showed in Sec. \ref{Sec: integration}. \ref{Fig: occ-Himalia-JPL} shows the shadow using the JPL ephemeris. In \ref{Fig: occ-Himalia-off22fev} we apply offsets to the positions of star and satellite accordingly the observations made at February 22. \ref{Fig: occ-Himalia-off03mar} is as in \ref{Fig: occ-Himalia-off22fev} but with observations made at March 03 when the objects were close to each other. The green dot at the bottom of the maps is the location of Observatório do Pico dos Dias.}
%\label{Fig: occ-Himalia}
%\end{centering}
%\end{figure*}

%\begin{figure*}
%\begin{centering}
%\includegraphics[width=17.6cm]{figures/HIMALIA.eps}
%\caption{Predictions for Himalia: The straight lines show the center of the shadow over time. In blue is for the prediction using the STE ephemeris showed in Sec. \ref{Sec: integration}. In red the shadow using the JPL ephemeris. In black we apply offsets to the positions of star and satellite accordingly the observations made at February 22. In green, the same for the previous one but with observations made at March 03 when the objects were close to each other. The dots show the closest approach of the shadow for each prediction. The scales are relative to the closest approach of the prediction using the STE ephemeris. The bars show the estimated diameter of the satellite and estimated error of the prediction.}
%\label{Fig: occ-Himalia}
%\end{centering}
%\end{figure*}

The critical parameter in the comparisons is the C/A, which here is related to latitudes. The apparent radius of Himalia is about 20 mas (see Table 1). In the context of the test, for a 0 mas offset in C/A we would have 100\% probability of observing the occultation, and 0\% in the case of a C/A offset equal to or larger than 20 mas, the radius of Himalia. From Table 5, we have nearly 0\% probability of success in situation 3, for which the offset in C/A was -20 mas, but when the relative astrometry was poor, 10 days prior to the event. Once at the day of the event in situation 4, the C/A offset dropped to -9 mas only, corresponding to a 55\% probability of success. Comparison with the prediction using the JPL ephemeris (situation 2) gives a +11 mas C/A offset, or a compatibility of 45\% between the ephemerides. All this suggests that there was a good probability of observing the event. The largest differences between the shadows of the four situations were 36s in time along the shadow path and 101km (31 mas) in the direction perpendicular to the shadows, suggesting that observers should be spread in narrow latitude ranges 100 km wide.

\begin{table}
\caption{\label{Tab: comparison-Himalia} Comparison between the predictions of the Himalia occultation at March 03, 2015.}
\begin{centering}
\begin{tabular}{lccc}
\hline  \hline
\multicolumn{4}{c}{Differences with respect to the STE prediction} \tabularnewline
Method  & Instant of C/A  & C/A & Sit.   \tabularnewline
\hline
STE & 00:39:51 UTC & $0\farcs703$ & I \tabularnewline
JPL & -26s & +11mas (36km) & II \tabularnewline % (284km)
Feb. 22 Obs. & -14s & -20mas (65km) & III \tabularnewline % (153km)
Mar. 03 Obs. & -36s & -09mas (29km) & IV \tabularnewline % (393km)
\hline
\end{tabular}
\par\end{centering}
C/A: geocentric closest approach; Sit: Situation test considered.
\end{table}

%The second test was with the satellite Elara, which is the second largest irregular satellite of Jupiter. The event was predicted to occur at March 30, 2015. The observations were made on March 25 and April 2, 2015 with the Zeiss telescope. On the night of April 2 the star and satellite could still be observed in the same FOV. Due to Elara being fainter than Himalia by 2 magnitudes, dispersions of the satellite positions on both nights ended up being higher than for Himalia. %Still, the differences between the maps obtained were relatively small. 
%The largest differences between them were 144s (186 mas) in time along the shadow path and 525 km (153 mas) perpendicular to it (see table \ref{Tab: comparison-Elara}).

%\begin{table}
%\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara occultation at March 30, 2015.}
%\begin{centering}
%\begin{tabular}{lcc}
%\hline  \hline
%\multicolumn{3}{c}{Difference from STE prediction} \tabularnewline
%Method  & Instant of C/A  & C/A  \tabularnewline
%\hline
%STE & 01:45:13 UTC & $1\farcs082$ \tabularnewline
%JPL & +02s & +57mas(195km) \tabularnewline
%Mar. 25 Obs. & -70s & +20mas (68km) \tabularnewline
%Apr. 02 Obs. & +74s & -96mas (330km) \tabularnewline
%\hline
%\end{tabular}
%\par\end{centering}
%\end{table}

### 2016-02-29 Irregular-occ/Irregular-occ.tex
modificado para mnras

--- 2016-02-08 Irregular-occ/Irregular-occ.tex
@@
-%% LyX 2.0.0 created this file.  For more info, see http://www.lyx.org/.
-%% Do not edit unless you really know what you are doing.
-\documentclass[english,traditabstract]{aa}
-\usepackage[varg]{txfonts}
-%\usepackage[english,brazilian]{babel}
+% mn2esample.tex
+%
+% v2.1 released 22nd May 2002 (G. Hutton)
+%
+% The mnsample.tex file has been amended to highlight
+% the proper use of LaTeX2e code with the class file
+% and using natbib cross-referencing. These changes
+% do not reflect the original paper by A. V. Raveendran.
+%
+% Previous versions of this sample document were
+% compatible with the LaTeX 2.09 style file mn.sty
+% v1.2 released 5th September 1994 (M. Reed)
+% v1.1 released 18th July 1994
+% v1.0 released 28th January 1994
+
+\documentclass[useAMS,usenatbib]{mn2e}
+\usepackage[brazilian,english]{babel}
+\usepackage[utf8]{inputenc}
 \usepackage[T1]{fontenc}
-\setcounter{tocdepth}{3}
 \usepackage{prettyref}
-\usepackage{array}
-\usepackage{rotating}
-\usepackage{float}
-\usepackage{multirow}
-\usepackage{amstext}
+\usepackage{amsmath}
 \usepackage{graphicx}
+\usepackage{subfigure}
 \usepackage{epstopdf}
-\usepackage{subfigure}
-\usepackage{gensymb}
-\usepackage{longtable}
 \usepackage{color}
-%\usepackage{setspace}
-%
-%\usepackage{etoolbox}
-%\AtBeginEnvironment{longtable}{\singlespacing}
-%\setlength{\LTcapwidth}{\linewidth}
+\usepackage[breaklinks=true]{hyperref}  
+\usepackage{arydshln} % creates horizontal dashed lines in tables
 
-\makeatletter
+%%%%% AUTHORS - PLACE YOUR OWN MACROS HERE %%%%%
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% LyX specific LaTeX commands.
-%% Because html converters don't know tabularnewline
-\providecommand{\tabularnewline}{\\}
+\title[Predictions of stellar occultations by irregular satellites up to 2020]{Stellar occultation predictions for 10 irregular satellites of giant planets plus Triton up to 2020}
+\author[A. R. Gomes-J\'unior, M. Assafin, L. Beauvalet et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
+M. Assafin$^{1,\dag}$,
+L. Beauvalet$^{2,3}$,
+R. Vieira-Martins$^{1,2,4,\dag}$,\newauthor
+J. I. B. Camargo$^{2,4}$,
+J. Desmars$^{5}$,
+B. E. Morgado$^{1,2}$
+F. Braga-Ribas$^{2,6}$
+\\
+$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
+CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
+$^{2}$Observat\'orio Nacional/MCTI, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
+$^{3}$Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France\\
+$^{4}$Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil\\
+$^{5}$Institut de m\'ecanique c\'eleste et de calcul des \'eph\'em\'erides - Observatoire de Paris, UMR 8028 du CNRS, 77 Av. Denfert-Rochereau,\\ 75014 Paris, France\\
+$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
+}
+\begin{document}
+\newcommand{\noccs}{396 } % total number of occultations
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% User specified LaTeX commands.
-\usepackage{natbib,twoopt}
-\usepackage[breaklinks=true, draft]{hyperref} %% to avoid \citeads line fills
-\usepackage{arydshln}
-\bibpunct{(}{)}{;}{a}{}{,} %% natbib format for A&A and ApJ
-\makeatletter
-\newcommandtwoopt{\citeads}[3][][]{\href{http://adsabs.harvard.edu/abs/#3}%
-{\def\hyper@linkstart##1##2{}%
-\let\hyper@linkend\@empty\citealp[#1][#2]{#3}}}
-\newcommandtwoopt{\citepads}[3][][]{\href{http://adsabs.harvard.edu/abs/#3}%
-{\def\hyper@linkstart##1##2{}%
-\let\hyper@linkend\@empty\citep[#1][#2]{#3}}}
-\newcommandtwoopt{\citetads}[3][][]{\href{http://adsabs.harvard.edu/abs/#3}%
-{\def\hyper@linkstart##1##2{}%
-\let\hyper@linkend\@empty\citet[#1][#2]{#3}}}
-\newcommandtwoopt{\citeyearads}[3][][]%
-{\href{http://adsabs.harvard.edu/abs/#3}
-{\def\hyper@linkstart##1##2{}%
-\let\hyper@linkend\@empty\citeyear[#1][#2]{#3}}}
-\makeatother
-\titlerunning{Predictions of stellar occultations by irregular satellites up to 2017}
-\authorrunning{Gomes-Júnior et al.}
+\date{Accepted . Received ; in original form }
 
-\makeatother
+\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2015}
 
-\begin{document}
+\maketitle
 
-\title{Stellar occultation predictions for  10 irregular satellites of giant planets plus Triton for 2016-2017
-%\fnmsep\thanks{Table 8 are only available in electronic form at the CDS via anonymous ftp to cdsarc.u-strasbg.fr (130.79.128.5) or via http://cdsweb.u-strasbg.fr/cgi-bin/qcat?J/A+A/ and IAU NSDC data base at www.imcce.fr/nsdc.}
-%\fnmsep\thanks{The complete version of Table 8 is available through CDS and IAU NSDC data base at www.imcce.fr/nsdc.}
-%\fnmsep\thanks{Partially based on observations made at Laborat\'orio Nacional de Astrof\'{\i}sica (LNA), Itajub\'a-MG, Brazil.}
-%\fnmsep\thanks{Partially based on observations through the ESO runs 079.A-9202(A), 075.C-0154, 077.C-0283 and 079.C-0345.}
-%\fnmsep\thanks{\textbf{Partially based on observations made at Observatoire de Haute Provence (OHP), F-04870 Saint-Michel l'observatoire, France}}
-}
+\label{firstpage}
 
-\author{ A. R. Gomes-Júnior
-          \inst{1},
-          M. Assafin 
-          \inst{1} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France},
-          L. Beauvalet
-          \inst{2,3},
-          R. Vieira-Martins
-          \inst{1,2,4} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France},
-%          J.-E. Arlot
-%          \inst{4},      
-          J. I. B. Camargo
-          \inst{2,4},     
-          J. Desmars
-          \inst{5},
-          B. E. Morgado
-          \inst{1,2},
-          F. Braga-Ribas
-          \inst{2,6}, 
-%          D. N. da Silva Neto
-%          \inst{6},
-%          A. H. Andrei
-%          \inst{1,2} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau, 75014 Paris, France},
-%          A. Dias-Oliveira
-%          \inst{2},
-%          G. Benedetti-Rossi
-%          \inst{2},
-%          Y. Duchemin
-%          \inst{4,7},
-%          V. Lainey
-%          \inst{4},
-%          W. Thuillot
-%          \inst{4}
-          }
-
-\offprints{A. R. Gomes-Júnior}
-
-   \institute{Observatório do Valongo/UFRJ, Ladeira Pedro Antônio 43,
-CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
-              \email{altair08@astro.ufrj.br}
-              \and
-Observatório Nacional/MCT, R. General José Cristino 77,
-CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
-              \email{rvm@on.br}
-              \and
-Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France
-              \and
-Laboratório Interinstitucional de e-Astronomia - LIneA, Rua Gal. José Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil
-              \and
-Institut de mécanique céleste et de calcul des éphémérides - Observatoire de Paris, UMR 8028 du CNRS,
-77 Av. Denfert-Rochereau, 75014 Paris, France
-%              \email{arlot@imcce.fr}
-              \and
-Federal University of Technology - Paraná (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
-%              \and
-%Centro Universitário Estadual da Zona Oeste, Av. Manual Caldeira de Alvarenga 1203, CEP 23.070-200 Rio de Janeiro RJ, Brazil
-%              \and
-%              ESIGELEC-IRSEEM, Technopôle du Madrillet, Avenue Galilée, 76801 Saint-Etienne du Rouvray, France
-              }
-
-\date{Received ; accepted }
-
-\abstract
-{Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system.
+\begin{abstract}
+Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system.
 It is important to know their physical parameters, such as size, shape, albedo and composition, to trace back their true origin.
-The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects.}
-{We aim to predict stellar occultations for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda, and for the irregular satellites Phoebe of Saturn and Nereid of Neptune, and also for Triton.}
-{We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument.
+The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects.
+We aim to predict stellar occultations for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda, and for the irregular satellites Phoebe of Saturn and Nereid of Neptune, and also for Triton.
+We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument.
 These catalogues were crossed with the ephemeris of the satellites to identify stellar occultations.
 We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future or the satellites of Jupiter.
-For Phoebe, we used an updated ephemeris and for Nereid and Triton we used recently published ephemeris.}
-{We managed to identify 396 candidates of stellar occultations between the period of January, 2016 and December, 2017.
+For Phoebe, we used an updated ephemeris and for Nereid and Triton we used recently published ephemeris.
+We managed to identify \noccs candidates of stellar occultations between the period of January, 2016 and December, 2017.
 We made observational tests for the prediction of the event of Himalia in March 03, 2015.
@@
 We call attention for an occultation by Triton of a bright star (R=12.5) in October 5, 2017.
-The event can be observed from Europe and the east coast of USA and will be a good opportunity to access the current state of the atmosphere of the satellite.}
-{The comparisons between the predictions and the observation tests show a good agreement.
-We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.}
+The event can be observed from Europe and the east coast of USA and will be a good opportunity to access the current state of the atmosphere of the satellite.
+The comparisons between the predictions and the observation tests show a good agreement.
+We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
+\end{abstract}
 
-\keywords{Occultations - Planets and satellites: general - Planets and satellites: individual: Jovian and Saturnian irregular satellites - Planets and satellites: individual: Triton}
-
-\maketitle
-
-%\section{Introduction} \label{Sec: introducao} 
-%
-%Irregular satellites revolve around giant planets at large distances, on eccentric, highly inclined and frequently retrograde orbits.
-Because of these peculiar orbits, it is largely accepted that these objects did not form by accretion around of their planet, but were captured by their planets in the early solar system \citep{Sheppard2005}.
-%
-%There are several of capture mechanisms of objects by giant planets suggested in the literature.
-For example the Gas Drag in the primordial circumplanetary nebulae \citep{Sheppard2005} where the object would be affected by the gas drag and its velocity slowed down until it be captured by the planet.
-Another mechanism is called pull-down capture \citep{Sheppard2005}, where the mass of the planet would increase while the object was temporarily captured.
-%
-%A mechanism based in the Nice model \citep{Morbidelli2005, Tsiganis2005, Gomes2005} was proposed by \cite{Nesvorny2007} and, in the specific case of Jupiter with the modern Nice model, by \citealp{Nesvorny2014}.
-In this scenario, during the early solar system instability, encounters between the outer planets occurred.
-These planetary encounters could exchange energy and angular momentum between planets and the objects nearby allowing the capture of irregular bodies by the giant planets.
-In this scenario, the survival rate of prior-LHB (Late Heavy Bombardment) satellites is very small.
-%
-%Another possible mechanism is the capture through collisional interactions \citep{Sheppard2005}.
-A collision between two small bodies in the Hill's sphere of the planet could generate fragmented objects and the dissipated energy could be such that some of these objects could be captured.
-%
-%Some of these objects are in dynamical groups with similar semi-major axis, eccentricities and inclinations, called families, similar to families found in the Main Asteroid Belt.
-These families may have been created by a parent body disrupted by collisions with comets or other satellites \citep{Nesvorny2004}.
-These collisions are more likely to have occurred during the LHB \citep{Gomes2005}.
-%
-%\cite{Nesvorny2003} studied the collision rates between irregular satellites and concluded that some satellites could have been removed by collision with a bigger satellite.
-The collision rate between satellites of the Himalia Group (Himalia, Elara, Lysithea and Leda, mainly), for instance, was found to be more than one during the solar system age suggesting that their current structure was originated by satellite-satellite collision.
-%
-%For Phoebe, ejected material from its surface caused by impacts could evolve due to Poynting-Robertson drag and collide with Iapetus causing the large variation in albedo observed on it \citep{Nesvorny2003}.
-Indeed, Cassini detected in Phoebe an absorption feature at 2.42 $\mu m$ (probably CN combinations) that was also detected in the dark side of Iapetus \citep{Clark2005}.
-%
-%The region of origin of these object is not well known.
-\cite{Grav2003} and \cite{Grav2007} showed that the irregular satellites from the giant planets have their colors and spectral slopes similar to C-, D- and P-type asteroids, Centaurs and trans-neptunian objects (TNOs) suggesting that they come from different locations in the early solar system.
-%
-%In order to obtain precise fundamental physical parameters like size, shape and albedo for the irregular satellites and thus to contribute to the study of their origin, we want to observe stellar occultations, which provide  more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
-%
-%No observation of a stellar occultation by an irregular satellite was published up to date.
-Since their estimated sizes are very small (see Table \ref{Tab: satellite-diameter}), this may have prevented earlier tries.
-But, in fact, given their distances to us, current ephemeris and star positions are already good enough for the prediction of the exact location and instant where the shadow of the occultation will cross the Earth.
-For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds).
-Thus, in this case, the cumulated uncertainty of ephemeris and star position must be around 40 mas to get good chances of observing a stellar occultation, which is quite feasible today.
-%
-%\begin{table}
-%\caption{\label{Tab: satellite-diameter} Estimated diameter of the satellites and correspondent apparent diameter}
-%\begin{centering}
-%\begin{tabular}{lccc}
-%\hline  \hline
-%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
-%Satellite  & mas\tablefootmark{a}  & km & References \tabularnewline
-%\hline
-%Ananke & 8 & 29 & 1 \tabularnewline
-%Carme & 13 & 46 & 1 \tabularnewline
-%Elara & 24 & 86 & 1 \tabularnewline
-%Himalia & 41 & $(150\times120) \pm 20$\tablefootmark{b} & 2 \tabularnewline
-%Leda & 5 & 20 & 1 \tabularnewline
-%Lysithea & 10 & 36 & 1 \tabularnewline
-%Pasiphae & 17 & 62 & 1 \tabularnewline
-%Sinope & 10 & 37 & 1 \tabularnewline
-%\hdashline
-%Phoebe & 32 & $212 \pm 1.4$\tablefootmark{b} & 3 \tabularnewline
-%\hdashline
-%Nereid & 15 & $340 \pm 50$\tablefootmark{c} & 4 \tabularnewline
-%Triton & 124 & $2707 \pm 2.0$\tablefootmark{c} & 5 \tabularnewline
-%\hline
-%\end{tabular}
-%\tablebib{
-%(1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}; (4) \cite{Thomas1991}; (5) \cite{Thomas2000}.}
-%\end{centering}
-%\tablefoottext{a}{Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
-%\tablefoottext{b}{From Cassini observation.}
-%\tablefoottext{c}{From Voyager-2 observation.}
-%\par
-%\end{table}
-%
-%From what we have just presented, we can think that the observation of an stellar occultation by an irregular satellite should be likelier than occultations by TNOs.
-The orbits of their host planets are well known and their observation time-span covers many orbital periods, contrary to TNOs.
-%%Unlike stellar occultations by TNOs, which nevertheless have been proved effective, the observation of an occultation by a irregular satellite is, in principle, more favorable.
-The orbits of their host planets are well known and these satellites have been observed completing already many orbital periods around them, thus presenting better ephemeris than TNOs.
-%Moreover, the irregular satellites are closer to Earth which means a minor localization error in kilometer.
-%
-%\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas.
-For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}.
-They pointed out that the ephemeris of the irregular satellites have systematic errors that may reach 200 mas for some satellites.
-For an object at the distance of Jupiter, this represents an error bigger than 700 km in the shadow path.
-Using the positions obtained by \cite{GomesJunior2015} we produced a specific ephemeris for short-time future for the satellites of Jupiter and better predict stellar occultations for these objects.
-%
-%We present in this paper stellar occultation predictions for the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda), Phoebe from Saturn and Nereid and Triton from Neptune.
-In the section \ref{Sec: Rationale} we explore the scientific rationale for the study of the irregular satellites and the possibility of having a common origin with TNOs.
-In section \ref{Sec: integration} we show the the process of the production of the new ephemeris.
-In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites and how they were made.
-Some tests made to confirm the predictions are presented in section \ref{Sec: testes} and the final discussion are in section \ref{Sec: discussion}.
-%
-%\section{Scientific Rationale} \label{Sec: Rationale}
-%
-%%\textcolor{red}{Como pode ver, comparado ao paragrafo na Introducao que falou a respeito, o texto aqui está muito pequeno e chovendo no molhado.}
-%
-%%\textcolor{red}{Tem que falar mais.
-Citar o paper no livro da Barucci que voce me mostrou outro dia.
-Dar mais citacoes sobre essa hipotese.
-Falar dos albedos parecidos, cores.
-Tem que fundamentar a hipotese de irregulares = TNOs pequenos.}
-%
-%%\textcolor{red}{Tambem falar aqui das vantagens de se observar ocultacoes dos irregulares, porque apesar de alguns satelites terem efemerides ruins, outros nem tanto, as efemerides sao melhores que a de TNOs (os irregulares ja completaram voltas em torno dos seus planetas); no caso de Jupiter, ele esta proximo o que ajuda a diminuir o erro em km.}
-%
-%
-%%\textcolor{red}{E dizer que, mesmo que a hipotese de TNOs que nos motiva a fazer as predicoes, venha a ser refutada no futuro, seja por outros trabalhos, seja pelos resultados futuros de observacoes de nossas proprias ocultacoes, o levantamento em si das propriedades fisicas, tamanho, forma e albedo ainda assim seriam muito uteis para testar as hiposteses de captura, ou seja, contribui para o estudo da formacao e evolucao dos sistemas de Jupiter, Saturno e Netuno.}
-%
-%There is no consensus for a single model explaining where the irregular satellites were formed.
-\cite{Cuk2004} showed that the progenitor of the Himalia group may have originated in heliocentric orbits similar to the Hilda asteroid group.
-\cite{Sheppard2005} stated that the irregular satellites may be some of the objects that were formed within the giant planets region.
-%
-%Phoebe is the most studied irregular satellite.
-\cite{Clark2005} suggest that its surface is probably covered by material of cometary origin.
-It was also stated by \cite{Johnson2005} that if the porosity of Phoebe is 15\%, Phoebe would have an uncompressed density similar to those of Pluto and Triton.
-%
-%\cite{Sheppard2005} and \cite{Jewitt2007} also expose the possibility for the irregular satellites having their origin as comets or TNOs.
-TNOs are highly interesting objects that, due to their large heliocentric distances, may be highly preserved having their physical properties similar to those they had when they were formed \citep{Camargo2014}.
-This is even more true for the smaller objects, since in principle larger sizes favour physical differentiation processes in the body and vice-versa.
-However, due to the distance, the smaller TNOs from this region are more difficult to observe.
-A clever way to overcome this difficulty is to study  the much closer irregular satellites, under the hypothesis that they share a common origin with the small TNOs' population.
-%
-%Triton is a uncommon satellite.
-Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
-Because Triton's orbit size is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded from the irregular satellites' class, but studied together by many authors \citep{Sheppard2005, Jewitt2007}.
-%
-%Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have the same origin as the TNOs \citep{Agnor2006}.
-Differently, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
-The main motivation to study Triton by stellar occultations is to understand the evolution of its atmosphere due to Triton's complicated and extreme seasonal cycle \citep{McKinnon2007}.
-
-%Because they are faint, the majority of these objects was discovered only in the last century\footnote{Website: http://ssd.jpl.nasa.gov/?sat\_discovery} .
-They were never visited by a spacecraft, with the exception of Himalia, Phoebe and Nereid, in a flyby by the Cassini space probe in 2000 for Himalia \citep{Porco2003} and in 2004 for Phoebe \citep{Desmars2013} and in a flyby by the Voyager 2 space probe in 1989 for Nereid \citep{Smith1989}.
-Even in situ, they were still opportunity target observations resulting in not optimal measurements, with size errors of $10 km$ for Himalia and $25 km$ for Nereid \citep{Thomas1991}.
-The exception is Phoebe with a very accurate measurement of size with a mean radius error of $0.7 km$ \citep{Thomas2010}.
-
-%If these objects were captured, there remains the question of where they came from.
-\citealp{Clark2005} showed from imaging spectroscopy from Cassini that Phoebe has a surface probably covered by material from the outer solar system and \citealp{Grav2003} showed that the satellites of the Jovian Prograde Group Himalia have grey colors implying that their surfaces are similar to that of C-type asteroids.
-In that same work, the Jovian Retrograde Group Carme was found to have surface colors similar to the D-type asteroids like Hilda or Trojan families while JXIII Kalyke has a redder color like Centaurs or trans-neptunian objects (TNOs).
-
-%For Saturnian satellites, \citealp{Grav2007} showed by their colors and spectral slopes that these satellites contain a more or less equal fraction of C-, P- and D-like objects but SXXII Ijiraq is marginally redder than D-type objects.
-These works may suggest different origins for the irregular satellites.
-
-%In this context, we used 3 databases for deriving precise positions for the irregular satellites observed at Observatório do Pico dos Dias (1.6 m and 0.6 m telescopes, IAU code 874), Observatoire Haute-Provence (1.2m telescope, IAU code 511) and ESO (2.2 m telescope, IAU code 809).
-Many irregular satellites were observed between 1992 and 2014 covering a few orbital periods of these objects (12 satellites of Jupiter, 4 of Saturn, Sycorax of Uranus and Nereid of Neptune).
-
-%Since their ephemerides are not very precise, predict and observe stellar occultations are very difficult and no observation of such an event for an irregular satellite is found in the literature.
-The precise star positions to be derived by the ESA astrometry satellite Gaia \citep{deBruijne2012} will render better predictions with the only source of error being the ephemeris.
-The positions derived from \textbf{our} observations can be used in new orbital numerical integrations, generating more precise ephemerides.
-
-%The power of stellar occultations for observing relatively small diameter solar system objects is supported by recent works such as the discovery of a ring system around the Centaur (10199) Chariklo \citep{Braga-Ribas2014}.
-Once irregular satellites start to be observed by this technique, it will be possible to obtain their physical parameters (shape, size, albedo, density) with unprecedented precision.
-For instance, in this case, sizes could be obtained with kilometer accuracy.
-The knowledge of these parameters would in turn bring valuable information for the study of the capture mechanisms and origin of the irregular satellites.
-
-%The databases are described in Sect.
-\ref{Sec: observations}.
-The astrometric procedures in Sect.
-\ref{Sec: reduction}.
-The obtained positions are presented in Sect \ref{Sec: positions} and analysed in Sect.
-\ref{Sec: comparison}.
-Conclusions are given in Sect.
-\ref{Sec: conclusions}.
+\begin{keywords}
+occultations - planets and satellites: general - planets and satellites: individual: Jovian and Saturnian irregular satellites - planets and satellites: individual: Triton
+\end{keywords}
 
@@
 \caption{\label{Tab: satellite-diameter} Estimated diameter of the satellites and correspondent apparent diameter}
-\begin{centering}
+\begin{center}
 \begin{tabular}{lccc}
@@
 \multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
-Satellite  & mas\tablefootmark{a}  & km & Ref. \tabularnewline
+Satellite  & mas$^ {a}$  & km & Ref. \tabularnewline
 \hline
@@
 Elara & 24 & 86 & 1 \tabularnewline
-Himalia & 41 & $(150\times120) \pm 20$\tablefootmark{b} & 2 \tabularnewline
+Himalia & 41 & $(150\times120) \pm 20^{b}$ & 2 \tabularnewline
 Leda & 5 & 20 & 1 \tabularnewline
@@
 \hdashline
-Phoebe & 32 & $212 \pm 1.4$\tablefootmark{b} & 3 \tabularnewline
+Phoebe & 32 & $212 \pm 1.4^{b}$ & 3 \tabularnewline
 \hdashline
-Nereid & 15 & $340 \pm 50$\tablefootmark{c} & 4 \tabularnewline
-Triton & 124 & $2707 \pm 2.0$\tablefootmark{c} & 5 \tabularnewline
+Nereid & 15 & $340 \pm 50^{c}$ & 4 \tabularnewline
+Triton & 124 & $2707 \pm 2.0^{c}$ & 5 \tabularnewline
 \hline
 \end{tabular}
-\tablebib{
-(1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}; (4) \cite{Thomas1991}; (5) \cite{Thomas2000}.}
-\end{centering}
-\tablefoottext{a}{Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
-\tablefoottext{b}{From Cassini observations.}
-\tablefoottext{c}{From Voyager-2 observations.}
+\end{center}
+References: (1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}; (4) \cite{Thomas1991}; (5) \cite{Thomas2000}.\\
+$^{a}${Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
+$^{b}${From Cassini observations.}
+$^{c}${From Voyager-2 observations.}
 \par
@@
 
-%From what we have just presented, we can think that the observation of an stellar occultation by an irregular satellite should be likelier than occultations by TNOs.
-The orbits of their host planets are well known and their observation time-span covers many orbital periods, contrary to TNOs.
-%Unlike stellar occultations by TNOs, which nevertheless have been proved effective, the observation of an occultation by a irregular satellite is, in principle, more favorable.
-The orbits of their host planets are well known and these satellites have been observed completing already many orbital periods around them, thus presenting better ephemeris than TNOs.
-%Moreover, the irregular satellites are closer to Earth which means a minor localization error in kilometer.
-
 In section \ref{Sec: integration} we show the building of the production of the new ephemeris.
@@
 
-%\subsection{Observations used}
 In order to improve ephemeris in short time spans in the near future for a more efficient prediction of stellar occultations by irregular satellites, we need predictions based on new and recent observations.
 In this context, \cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.
-%They showed that the orbits of the irregular satellites of the giant planets have systematic errors.
-%Their observed position offsets relative to the JPL ephemeris reached up to 200 mas for some satellites.
-At the distance of Jupiter, this represents an difference larger than 700 km in the shadow path of an occultation.%These differences could be associated with errors in their orbital elements.
 
@@
 For these reasons, only this set of positions was used for building the new ephemeris for the satellites of Jupiter.
-
-%In order to have efficient ephemerides for predicting the stellar occultations by irregular satellites, we need predictions based on such recent observations.
-This is why we decided to develop our own ephemeris based on the observations published in \cite{GomesJunior2015}.
-First because many of the authors of this paper worked in the production of that set of positions, so that it was very well known.
-Second, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets.
-For these reasons, only this set of recent positions was used for building the new ephemeris for the satellites.
 
@@
 
-The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}.%, while the residuals are presented in Fig \ref{Fig: residuals} for Himalia and Carme as representatives.
-
-%\begin{figure*}
-%\subfigure{\includegraphics[width=8.8cm]{figures/Himalia_resid.eps}   \label{Fig: residuals-Himalia}}
-%\subfigure{\includegraphics[width=8.8cm]{figures/Carme_resid.eps}  \label{Fig: residuals-Carme}}
-%\caption{Residuals in topocentric right ascension and declination of the numerical integration for Himalia and Carme. \label{Fig: residuals}}
-%\end{figure*}
+The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}.
 
 \begin{table*}
-\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0.% For Themisto's case, the small number of observations meant that the statistical uncertainty was even greater than the obtained elements, though the fitting proved satisfying.
-We decided to provide its elements with significant number comparable to the other satellite for information, but left it without uncertainty.
+\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0.
  }\label{Tab: sat_ell}
@@
 All the orbits determined for the satellites show satisfying residuals.
-%Yet, one of the satellites has a peculiar situation.
-Its small number of observations in our set means that its orbit is definitely loosely constrained.
 The residuals are smaller than those obtained with JPL ephemeris, which was expected because the accuracy of an ephemeris decreases when we get further from the time of observations.
@@
 \includegraphics[width=8.8cm]{figures/Carme_ephemeris.eps} 
-\caption{Offsets in declination of the positions published by \cite{GomesJunior2015} for Carme.
-The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the JUP300 JPL ephemeris and the green dot relative to \cite{Emelyanov2008}.
-As expected, the ephemeris systematic errors pointed out by \cite{GomesJunior2015} are reduced with the STE ephemeris.
+\caption{Offsets in declination of the positions published by \protect\cite{GomesJunior2015} for Carme.
+The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the JUP300 JPL ephemeris and the green dot relative to \protect\cite{Emelyanov2008}.
+As expected, the ephemeris systematic errors pointed out by \protect\cite{GomesJunior2015} are reduced with the STE ephemeris.
 \label{Fig: JPL-STE}}
 \end{figure}
-
-%\begin{figure*}
-%\subfigure{\includegraphics[width=8.8cm]{figures/JPL_DEC_Himalia.eps}    \label{Fig: JPL-STE-Himalia}}
-%\subfigure{\includegraphics[width=8.8cm]{figures/JPL_DEC_Carme.eps}  \label{Fig: JPL-STE-Carme}}
-%\caption{Geocentric right ascension and declination differences between the JUP300 JPL ephemeris and the special-tailored ephemeris for Himalia and Carme. \label{Fig: JPL-STE}}
-%\end{figure*}
 
@@
 \includegraphics[width=8.8cm]{figures/Phoebe.eps} 
-\caption{Comparison between the PH15, SAT375 JPL and \cite{Emelyanov2007} ephemeris for the satellite Phoebe.}
+\caption{Comparison between the PH15, SAT375 JPL and \protect\cite{Emelyanov2007} ephemeris for the satellite Phoebe.}
 \label{Fig:eph-Phoebe}
@@
 \begin{centering}
-\subfigure[NEP081 JPL and \cite{Emelyanov2015}]{\includegraphics[width=8.8cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
-\subfigure[\cite{Zhang2014} and \cite{Emelyanov2015}]{\includegraphics[width=8.8cm]{figures/Zhang-EME.eps} \label{Fig:eph-Triton-zhang}}
-\caption{Comparison between the \cite{Emelyanov2015}, \cite{Zhang2014} and NEP081 JPL \citep{Jacobson2009} ephemeris for the satellite Triton.}
+\subfigure[NEP081 JPL and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
+\subfigure[\protect\cite{Zhang2014} and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/Zhang-EME.eps} \label{Fig:eph-Triton-zhang}}
+\caption{Comparison between the \protect\cite{Emelyanov2015}, \protect\cite{Zhang2014} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Triton.}
 \label{Fig:eph-Triton}
@@
 The declination ephemeris offsets are of the order of the apparent diameter (15 mas, see Table \ref{Tab: satellite-diameter}), resulting in fair chances (50\%) that the shadow crosses the expected Earth latitude predicted for an occultation.
-% which is bigger than the satellite apparent diameter (see Table \ref{Tab: satellite-diameter}).
-Although this fact, the difference is bigger in Right Ascension than in Declination, which means that, in an event, the error would be smaller in the latitude where the shadow will pass on Earth than the error at the time of the occultation.
 
@@
 \includegraphics[width=8.8cm]{figures/JPL-EME_Nereid.eps}
-\caption{Comparison between the \cite{Emelyanov2011} and NEP081 JPL \citep{Jacobson2009} ephemeris for the satellite Nereid.}
+\caption{Comparison between the \protect\cite{Emelyanov2011} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Nereid.}
 \label{Fig:eph-Nereid}
@@
 The development of a better ephemeris for Nereid using these new positions in the context discussed here of near future stellar occultations is desirable, but is out of the scope of this paper.
-
-%Making a new model for the orbits of these objects would demand a lot of time and delay the publication of predictions of possible events to occur in the very near future.
-Thus, we utilized the positions obtained by \cite{GomesJunior2015} to identify error patterns in the ephemeris.
-The error patterns in right ascension and in declination could be used to extrapolate the offsets to the satellite ephemeris by the time of the predicted occultation, improving it.
-Plots of the offsets over true anomaly (see Fig.
-\ref{Fig:offxanom} for declination offsets of Carme, for instance) clearly show the systematic errors in the JPL ephemeris.
-
-%\begin{figure}
-%\begin{centering}
-%\includegraphics[scale=0.25]{figures/DEC.png}\label{Fig:offxtime}
-%\caption{Offsets of the declination of Carme by time. \textcolor{red}{Figura só para visualização, vou colocar alguma melhor depois}}
-%\end{centering}
-%\end{figure}
-
-%\begin{figure*}
-%\begin{centering}
-%\subfigure[Offset from the JPL ephemeris]{\includegraphics[width=8.8cm]{figures/DEC_anom.png} \label{Fig:offxtanomxjpl}}
-%\subfigure[\textcolor{red}{Offset from the numerical integration}]{\includegraphics[width=8.8cm]{figures/DEC_anom.png} \label{Fig:offxtanomxni}}
-%\caption{Offsets of the declination of Carme by true anomaly from the JPL ephemeris and \textcolor{red}{from the numerical integration of the positions obtained in \cite{GomesJunior2015}}.
-%\label{Fig:offxanom}}
-%\end{centering}
-%\end{figure*}
-
-%The two kinds of error or offset patterns adopted depending on the case are given by Eqs. \ref{eq:linear} and \ref{eq:seno}:
-%
-%\begin{equation}
-%F(t,f) = p[0]\times t + p[1]\times \sin(f) + p[2]\times\cos(f) + p[3],
-%\label{eq:linear}
-%\end{equation}
-%
-%\begin{equation}
-%\begin{split}
-%%F = \{F_{x} \in  F_{c} &: (|S| > |C|) \\
-%% &\quad \cap (\text{minPixels}  < |S| < \text{maxPixels}) \\
-%% &\quad \cap (|S_{\text{conected}}| > |S| - \epsilon) \}
-%F(t,f) = p[0]\times\sin\left(\frac{2\pi}{p[1]}\times t + p[2]\right) + p[3]\times\sin(f) + \\ +p[4]\times\cos(f) + p[5],
-%\end{split}
-%\label{eq:seno}
-%\end{equation}
-%where $F(t,f)$ is the offset obtained, $t$ is time in years counting from J2000.0 and $f$ is the true anomaly.
-The patterns could be applied for right ascension as well as for declination ephemeris offsets.
 
@@
 
-A total of 396 events were identified between January 2016 and December 2017.
+A total of \noccs events were identified between January 2016 and December 2017.
 In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite.
@@
 \begin{table}
-\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from January, 2016 up to December, 2017.}
+\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from January, 2016 up to December, 2020.}
 \begin{centering}
-\begin{tabular}{lccc}
+\begin{tabular}{lcccccc}
 \hline  \hline
 %\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
-Satellite  & 2016 & 2017 & Total \tabularnewline
+Satellite  & 2016 & 2017 & 2018 & 2019 & 2020 & Total \tabularnewline
 \hline
-Ananke & 12 & 16 & 28 \tabularnewline
-Carme & 20 & 14 & 34 \tabularnewline
-Elara & 14 & 16 & 30 \tabularnewline
-Himalia & 15 & 12 & 27 \tabularnewline
-Leda & 8 & 24 & 32 \tabularnewline
-Lysithea & 16 & 11 & 27 \tabularnewline
-Pasiphae & 20 & 19 & 39 \tabularnewline
-Sinope & 15 & 21 & 36 \tabularnewline
+Ananke & 12 & 16 & & & &  \tabularnewline
+Carme & 20 & 14 & & & &  \tabularnewline
+Elara & 14 & 16 & & & &  \tabularnewline
+Himalia & 15 & 12 & & & &  \tabularnewline
+Leda & 8 & 24 & & & &  \tabularnewline
+Lysithea & 16 & 11 & & & &  \tabularnewline
+Pasiphae & 20 & 19 & & & &  \tabularnewline
+Sinope & 15 & 21 & & & &  \tabularnewline
 \hdashline
-Phoebe\tablefootmark{a} & 32 & 98 & 130 \tabularnewline
+Phoebe & 32 & 98 & & & &  \tabularnewline
 \hdashline
-Nereid\tablefootmark{a} & 11\tablefootmark{b} & 1 & 12 \tabularnewline
-Triton\tablefootmark{a} & -- & 1 & 1 \tabularnewline
+Nereid & 11$^{a}$ & 1 & & & &  \tabularnewline
+Triton & -- & 1 & & & &  \tabularnewline
 \hline
@@
 Occultations predicted using the UCAC4 catalogue and STE ephemeris.
-\tablefoottext{a}{Using JPL ephemeris.}
-\tablefoottext{b} {Using the WFI catalogue as explained in Sec. \ref{Sec: predictions}}.
+$^{a}$ {Using the WFI catalogue as explained in Sec. \ref{Sec: predictions}}.
 \end{table}
@@
 \caption{\label{Tab: sample-cds} A sample of stellar occultation predictions for Pasiphae}
-\begin{centering}
+\begin{center}
 \begin{tabular}{cccrcccrcrr}
@@
 \end{tabular}
-\par\end{centering}
+\end{center}
+\begin{flushleft}
 \textbf{Notes}.
@@
 \label{Tab: occ-list}
+\end{flushleft}
 \end{table*}
 
-\begin{figure*}
+\begin{figure}
 \begin{centering}
-\includegraphics[scale=0.4]{figures/Triton_2017-10-05T23:51:53.png}   
+\includegraphics[width = 8.8cm]{figures/Triton_2017-10-05T23:51:53.png}   
 \caption{Occultation map for Triton.% regarding to the first event sampled in Table \ref{Tab: occ-list}.
@@
 \end{centering}
-\end{figure*}
+\end{figure}
 
@@
 
-\section{Occultation test} \label{Sec: testes}
+\subsection{Prediction test} \label{Sec: testes}
 
@@
 
-%\begin{figure*}
-%\begin{centering}
-%\subfigure[Prediction with STE ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_STE.eps}  \label{Fig: occ-Himalia-STE}}
-%\subfigure[Prediction with JPL ephemeris]{\includegraphics[width=8.8cm]{figures/Himalia_JPL.eps}  \label{Fig: occ-Himalia-JPL}}
-%\subfigure[Observation at February 22 (different FOVs)]{\includegraphics[width=8.8cm]{figures/Himalia_22fev.eps}   \label{Fig: occ-Himalia-off22fev}}
-%\subfigure[Observation at March 03 (same FOV)]{\includegraphics[width=8.8cm]{figures/Himalia_03mar.eps}   \label{Fig: occ-Himalia-off03mar}}
-%\caption{Predictions for Himalia: The straight lines show the size of the shadow.
-The big red dot show the geocentric closest approach of the shadow.
-The black "+" marks at maps (b,c,d) are the STE prediction closest approach for reference.
-The small red ones are the center of the shadow separated by one minute.
-\ref{Fig: occ-Himalia-STE} is the map using the STE ephemeris showed in Sec.
-\ref{Sec: integration}.
-\ref{Fig: occ-Himalia-JPL} shows the shadow using the JPL ephemeris.
-In \ref{Fig: occ-Himalia-off22fev} we apply offsets to the positions of star and satellite accordingly the observations made at February 22.
-\ref{Fig: occ-Himalia-off03mar} is as in \ref{Fig: occ-Himalia-off22fev} but with observations made at March 03 when the objects were close to each other.
-The green dot at the bottom of the maps is the location of Observatório do Pico dos Dias.}
-%\label{Fig: occ-Himalia}
-%\end{centering}
-%\end{figure*}
-
-%\begin{figure*}
-%\begin{centering}
-%\includegraphics[width=17.6cm]{figures/HIMALIA.eps}
-%\caption{Predictions for Himalia: The straight lines show the center of the shadow over time.
-In blue is for the prediction using the STE ephemeris showed in Sec.
-\ref{Sec: integration}.
-In red the shadow using the JPL ephemeris.
-In black we apply offsets to the positions of star and satellite accordingly the observations made at February 22.
-In green, the same for the previous one but with observations made at March 03 when the objects were close to each other.
-The dots show the closest approach of the shadow for each prediction.
-The scales are relative to the closest approach of the prediction using the STE ephemeris.
-The bars show the estimated diameter of the satellite and estimated error of the prediction.}
-%\label{Fig: occ-Himalia}
-%\end{centering}
-%\end{figure*}
-
 The critical parameter in the comparisons is the C/A, which here is related to latitudes.
@@
 
-%The second test was with the satellite Elara, which is the second largest irregular satellite of Jupiter.
-The event was predicted to occur at March 30, 2015.
-The observations were made on March 25 and April 2, 2015 with the Zeiss telescope.
-On the night of April 2 the star and satellite could still be observed in the same FOV.
-Due to Elara being fainter than Himalia by 2 magnitudes, dispersions of the satellite positions on both nights ended up being higher than for Himalia.
-%Still, the differences between the maps obtained were relatively small.
-%The largest differences between them were 144s (186 mas) in time along the shadow path and 525 km (153 mas) perpendicular to it (see table \ref{Tab: comparison-Elara}).
-
-%\begin{table}
-%\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara occultation at March 30, 2015.}
-%\begin{centering}
-%\begin{tabular}{lcc}
-%\hline  \hline
-%\multicolumn{3}{c}{Difference from STE prediction} \tabularnewline
-%Method  & Instant of C/A  & C/A  \tabularnewline
-%\hline
-%STE & 01:45:13 UTC & $1\farcs082$ \tabularnewline
-%JPL & +02s & +57mas(195km) \tabularnewline
-%Mar. 25 Obs. & -70s & +20mas (68km) \tabularnewline
-%Apr. 02 Obs. & +74s & -96mas (330km) \tabularnewline
-%\hline
-%\end{tabular}
-%\par\end{centering}
-%\end{table}
-
 \section{Discussion} \label{Sec: discussion}
@@
 \ref{Sec: predictions}).
-From this, a total of 396 events are foreseen.
+From this, a total of \noccs events are foreseen.
 
@@
 The test made with an occultation expected to happen in March 03, 2015 for Himalia showed that this event would probably have been observed successfully in case there were observers available in the shadow area.
-%Fig.
-\ref{Fig: occ-Himalia} shows the maps made for 4 different offset corrections as explained in Sec.
-\ref{Sec: testes}.
-Table \ref{Tab: comparison-Himalia} shows numerically the difference between the maps.
-%A similar test was made for an occultation of Elara in March 30, 2015. %The difference between the maps in this case are listed in Table \ref{Tab: comparison-Elara}.
-The results show satisfying small offsets with respect to the local of the prediction. %We can expect similar results for other satellites and our method should allow successful stellar occultation.
+The results show satisfying small offsets with respect to the local of the prediction.
 
@@
 It will allow for the discovery of ocultations by fainter stars not present in the UCAC4 catalogue.
-% Consequently, the astrometry of the satellites with the GAIA catalogue will improve the position errors allowing for better predictions and a higher probabilities of observing stellar occultations by irregular satellites.
 The release of the GAIA catalogue should have a positive impact on both the astrometric precision of occulted stars and the reduction of astrometric positions of the satellites.
@@
 
-%We managed a large database with FITS images acquired by 5 telescopes in 3 sites between 1992 and 2014.
-From that, we identified 8466 observations of irregular satellites, from which we managed to obtain 6523 suitable astrometric positions, giving a total of 3666 positions for 12 satellites of Jupiter, 1920 positions for 4 satellites of Saturn, 35 positions for Sycorax (Uranus) and 902 positions for Nereid (Neptune).
-
-%The positions of all the objects were determined using the PRAIA package.
-The package was suited to cope with the huge amount of observations and the task of identifying the satellites within the database.
-PRAIA tasks were also useful to deal with the missing or incorrect coordinate and time stamps present mostly in the old observations.The UCAC4 was used as the reference frame.
-Based in the comparisons with ephemeris, we estimate that the position errors are about 60 mas to 80 mas depending on the satellite brightness.
-
-%For some satellites the number of positions obtained in this work is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012} (see Table \ref{Tab: comparison-horizons}).
-For instance, the amount of new positions for Himalia is about 70\% of the number used in the numerical integation of orbits by JPL.
-Systematic errors in the ephemeris were found for at least some satellites (Ananke, Carme, Elara and Pasiphae).
-In the case of Carme, we evidenced an error in the orbital inclination (see Fig.
-\ref{Fig: carme_anom}).
-
-%The positions derived in this work can be used in new orbital numerical integrations, generating more precise ephemerides.
-Stellar occultations by irregular satellites could then be better predicted.
-Based in this work, our group has already computed occultation predictions for the 8 major irregular satellites of Jupiter.
-These predictions will be published in a forthcoming paper.
-
-\begin{acknowledgements}
+\section*{Acknowledgements}
 
@@
 
-\end{acknowledgements}
+\bibliographystyle{mn2e}
+\bibliography{references}
 
-\bibliographystyle{aa}
-\nocite{*}
-\bibliography{references.bib}
+\label{lastpage}
 

### 2016-03-02 Irregular-occ/Irregular-occ.tex
.
Feito ocultacoes 2018-2020

@@
 L. Beauvalet$^{2,3}$,
-R. Vieira-Martins$^{1,2,4,\dag}$,\newauthor
-J. I. B. Camargo$^{2,4}$,
-J. Desmars$^{5}$,
+J. Desmars$^{4}$,
+R. Vieira-Martins$^{1,2,5,\dag}$,\newauthor
+J. I. B. Camargo$^{2,5}$,
 B. E. Morgado$^{1,2}$
@@
 $^{3}$Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France\\
-$^{4}$Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil\\
-$^{5}$Institut de m\'ecanique c\'eleste et de calcul des \'eph\'em\'erides - Observatoire de Paris, UMR 8028 du CNRS, 77 Av. Denfert-Rochereau,\\ 75014 Paris, France\\
+$^{4}$Institut de m\'ecanique c\'eleste et de calcul des \'eph\'em\'erides - Observatoire de Paris, UMR 8028 du CNRS, 77 Av. Denfert-Rochereau,\\ 75014 Paris, France\\
+$^{5}$Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil\\
 $^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
@@
 \begin{document}
-\newcommand{\noccs}{396 } % total number of occultations
+\newcommand{\noccs}{5457 } % total number of occultations
 
@@
 
-\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2015}
+\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2016}
 
@@
 These catalogues were crossed with the ephemeris of the satellites to identify stellar occultations.
-We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future or the satellites of Jupiter.
+We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future of the satellites of Jupiter.
 For Phoebe, we used an updated ephemeris and for Nereid and Triton we used recently published ephemeris.
-We managed to identify \noccs candidates of stellar occultations between the period of January, 2016 and December, 2017.
+We managed to identify \noccs candidates of stellar occultations between the period of January, 2016 and December, 2020.
 We made observational tests for the prediction of the event of Himalia in March 03, 2015.
 The stars and objects were observed close to the date predicted and in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions.
+The comparisons between the predictions and the observation tests show a good agreement.
 We call attention for an occultation by Triton of a bright star (R=12.5) in October 5, 2017.
 The event can be observed from Europe and the east coast of USA and will be a good opportunity to access the current state of the atmosphere of the satellite.
-The comparisons between the predictions and the observation tests show a good agreement.
+We note that Jupiter will cross a populated region of stars in 2019-2020 and Saturn in 2018.
+This results in a large number of occultations predicted for that period.
 We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
@@
 Here we develop our own ephemeris based on the observations published in \cite{GomesJunior2015}.
-First, because the reduction was made with consistent and precise stellar catalogue and with a robust and reliable software (PRAIA, \cite{Assafin2011}).
+First, because the reduction was made with consistent and precise stellar catalogue and with a robust and reliable software \citep[PRAIA,][]{Assafin2011}.
 Second, besides recent observations, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets.
@@
 \end{center}
+\begin{flushleft}
 \textbf{Notes}: a: semimajor axis; e : excentricity; I: inclination relative to the equatorial reference plane J2000; $\Omega$: longitude of the ascending node; $\omega$: argument of periapsis; $v$: true anomaly.
+\end{flushleft}
 \end{table*}
@@
 2015-Feb-27.
-"Satellite Ephemeris: SAT375", JPL Satellite Ephemeris File Release, \url{ftp://ssd.jpl.nasa.gov/pub/eph/satellites/nio/LINUX_PC/sat375l.txt}} and the \cite{Emelyanov2007} ephemeris.
-We can see that our ephemeris is closer to the SAT375 JPL ephemeris where the difference between them is smaller than 20 mas (< 10 mas in Declination).
+"Satellite Ephemeris: SAT375", JPL Satellite Ephemeris File Release, \url{ftp://ssd.jpl.nasa.gov/pub/eph/satellites/nio/LINUX_PC/sat375l.txt}} ephemeris.
+The difference between them is smaller than 30 mas (< 10 mas in Declination).
 This difference is smaller than the apparent diameter of Phoebe (see Table \ref{Tab: satellite-diameter})
@@
 \subfigure[NEP081 JPL and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
-\subfigure[\protect\cite{Zhang2014} and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/Zhang-EME.eps} \label{Fig:eph-Triton-zhang}}
+\subfigure[\protect\cite{Zhang2014} and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/Zhang-EME_Triton.eps} \label{Fig:eph-Triton-zhang}}
 \caption{Comparison between the \protect\cite{Emelyanov2015}, \protect\cite{Zhang2014} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Triton.}
@@
 
-A total of \noccs events were identified between January 2016 and December 2017.
+A total of \noccs events were identified between January 2016 and December 2020.
 In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite.
@@
 \ref{Fig: ocultacao} we show an example of an occultation map.
-This is the only found occultation by Triton and it will happen in October 05, 2017.
+This is one of the only two found occultation by Triton and it will happen in October 05, 2017.
 This event can be observed from Europe and the east coast of USA and will be a great opportunity to study in high resolution the atmosphere of the satellite.
@@
 \hline
-Ananke & 12 & 16 & & & &  \tabularnewline
-Carme & 20 & 14 & & & &  \tabularnewline
-Elara & 14 & 16 & & & &  \tabularnewline
-Himalia & 15 & 12 & & & &  \tabularnewline
-Leda & 8 & 24 & & & &  \tabularnewline
-Lysithea & 16 & 11 & & & &  \tabularnewline
-Pasiphae & 20 & 19 & & & &  \tabularnewline
-Sinope & 15 & 21 & & & &  \tabularnewline
+Ananke & 12 & 16 & 49 & 359 & 187 & 623 \tabularnewline
+Carme & 20 & 14 & 30 & 369 & 220 & 653 \tabularnewline
+Elara & 14 & 16 & 33 & 305 & 193 & 561 \tabularnewline
+Himalia & 15 & 12 & 54 & 257 & 230 & 568 \tabularnewline
+Leda & 8 & 24 & 38 & 362 & 208 & 640 \tabularnewline
+Lysithea & 16 & 11 & 35 & 330 & 212 & 604 \tabularnewline
+Pasiphae & 20 & 19 & 44 & 362 & 206 & 651 \tabularnewline
+Sinope & 15 & 21 & 34 & 356 & 256 & 682 \tabularnewline
 \hdashline
-Phoebe & 32 & 98 & & & &  \tabularnewline
+Phoebe & 32 & 98 & 238 & 79 & 13 & 460 \tabularnewline
 \hdashline
-Nereid & 11$^{a}$ & 1 & & & &  \tabularnewline
-Triton & -- & 1 & & & &  \tabularnewline
+Nereid & 11$^{a}$ & 1 & 1 & 0 & 0 & 13 \tabularnewline
+Triton & -- & 1 & 1 & 0 & 0 & 2 \tabularnewline
 \hline
@@
 \par \end{centering}
-Occultations predicted using the UCAC4 catalogue and STE ephemeris.
 $^{a}$ {Using the WFI catalogue as explained in Sec. \ref{Sec: predictions}}.
@@
 
-The first preliminary catalogue version of the ESA astrometry satellite GAIA \citep{deBruijne2012} is expected to be released up to the end of 2016.
+The first preliminary catalogue version of the ESA astrometry satellite GAIA \citep{deBruijne2012} is expected to be released up to the end of 2016 (The catalogue with five-parameter astrometric solutions is up to the end of 2017).
 The precise star positions to be derived by GAIA will provide better predictions with the main source of error being the ephemeris.
-Astrometric reduction of observations published in \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions from 2017 onwards will be improved.
-Since we cannot foresee exactly when the GAIA catalogue will be released and when new and re-reduced satellite positions in the GAIA frame will become available, we decided to publish only the predictions for occultations restricted to the 2016-2017 time interval.
+Astrometric reduction of observations published in \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions will be improved.
+Since we cannot foresee exactly when the GAIA catalogue will be released and when new and re-reduced satellite positions in the GAIA frame will become available, we decided to publish predictions up to the end of 2020.
+In the GAIA era, the occultation predictions will be updated.
 
@@
 
-We predict stellar occultations for the period of 2016-2017 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid.
+We predict stellar occultations for the period of 2016-2020 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid.
 The procedure used was the same as that for the prediction of stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.

### 2016-03-03 Irregular-occ/dados/tabela.py
.

@@ -17,13 +17,6 @@ tabela = Table(names=('object', 'instantes', 'RA (ICRS) DEC', 'C/A', 'P/A', 'vel
 c = csv.writer(open("agenda.csv", "wb"))
 c.writerow(["Subject","Start Date","Start Time","End Date","End Time","All Day Event","Description","Private"])
     
-def texto(i):
-    if i in ['Triton', 'Nereida']:
-        return '/Emelyanov/'
-    if i == 'Phoebe':
-        return '/Josselin/'
-    return '/LAU/'
-    
 def mostrar(i):
     imagem = tabela[np.where(tabela['numero'] == i)]
     imagem.pprint(max_width=-1)
@@ -62,7 +55,7 @@ for i in lista:
     magK = Column(dados['mK'], name='MagK', format='4.1f')
     longi = Column(dados['long'], name='long', format='+3.0f')
     name = Column([arquivo[15:-11].capitalize()]*len(ca), name='object')
-    caminho = '/home/altair/Documentos/Maps/All/' + np.char.array(datas.iso).rpartition(' ')[:,0].rpartition('-')[:,0].rpartition('-')[:,0] + '/Altair/' + arquivo[15:-11].capitalize() + texto(arquivo[15:-11].capitalize()) + arquivo[15:-11].capitalize() + '_' + np.char.array(datas.isot) + '.png'
+    caminho = '/home/altair/Documentos/Maps/Todos/' + np.char.array(datas.iso).rpartition(' ')[:,0].rpartition('-')[:,0].rpartition('-')[:,0] + arquivo[15:-11].capitalize() + arquivo[15:-11].capitalize() + '_' + np.char.array(datas.isot) + '.png'
     t = Table([name, occ_data, star_coord, ca, pa, vel, dist, magR, magK, longi, caminho], names=('object', 'instantes', 'RA (ICRS) DEC', 'C/A', 'P/A', 'vel', 'distance', 'magR', 'magK', 'long', 'caminho'))
     tabela = vstack([tabela, t])
 

### 2016-03-07 Irregular-occ/Irregular-occ.tex
.

@@
 The event can be observed from Europe and the east coast of USA and will be a good opportunity to access the current state of the atmosphere of the satellite.
-We note that Jupiter will cross a populated region of stars in 2019-2020 and Saturn in 2018.
+We note that Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018.
 This results in a large number of occultations predicted for that period.
@@
 \hline
-STE & 00:39:51 UTC & $0\farcs703$ & I \tabularnewline
-JPL & -26s & +11mas (36km) & II \tabularnewline % (284km)
-Feb. 22 Obs. & -14s & -20mas (65km) & III \tabularnewline % (153km)
-Mar. 03 Obs. & -36s & -09mas (29km) & IV \tabularnewline % (393km)
+STE & 00:39:51 UTC & $0\farcs703$ & i \tabularnewline
+JPL & -26s & +11mas (36km) & ii \tabularnewline % (284km)
+Feb. 22 Obs. & -14s & -20mas (65km) & iii \tabularnewline % (153km)
+Mar. 03 Obs. & -36s & -09mas (29km) & iv \tabularnewline % (393km)
 \hline

### 2016-03-07 Irregular-occ/Irregular-occ.tex
.

@@
 
-\title[Predictions of stellar occultations by irregular satellites up to 2020]{Stellar occultation predictions for 10 irregular satellites of giant planets plus Triton up to 2020}
+\title[New orbits and predictions of stellar occultations by irregular satellites up to 2020]{New numerical integration of orbits of irregular satellites and prediction of stellar occultations up to 2020}
 \author[A. R. Gomes-J\'unior, M. Assafin, L. Beauvalet et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
@@
 The event can be observed from Europe and the east coast of USA and will be a good opportunity to access the current state of the atmosphere of the satellite.
-We note that Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018.
+We also note that Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018.
 This results in a large number of occultations predicted for that period.
@@
 
-We present in this paper stellar occultation predictions for the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda), Phoebe of Saturn and, Nereid and Triton of Neptune.
+We present in this paper new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) for predict stellar occultations.
+This is a specific short-time ephemeris using only the positions obtained by \cite{GomesJunior2015}.
+\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas.
+For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}.
+They pointed out that the ephemeris of the irregular satellites have systematic errors that may reach 200 mas for some satellites.
+For an object at the distance of Jupiter, this represents an error langer than 700 km in the shadow path of an occultation.
+
+For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \cite{GomesJunior2015}, observations from Minor Planet Center and observations from Flagstaff.
+For Nereid and Triton we used the most recent ephemeris available.
+
+Also, with new and precise ephemeris, we present stellar occultation predictions for these 11 objects up to 2020.
 Phoebe, being the most studied object with a good measured size, can be used to calibrate and evaluate the technique for similar objects.
@@
 
-\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas.
-For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}.
-They pointed out that the ephemeris of the irregular satellites have systematic errors that may reach 200 mas for some satellites.
-For an object at the distance of Jupiter, this represents an error larger than 700 km in the shadow path.
-Using the positions obtained by \cite{GomesJunior2015} we produced a specific short-time ephemeris for the satellites of Jupiter and Phoebe to better predict stellar occultations by these objects.
-
 Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
@@
 
-In section \ref{Sec: integration} we show the building of the production of the new ephemeris.
+In section \ref{Sec: integration} we show the building of the new ephemeris.
 In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites and how they were made.
@@
 In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite.
+It is possible to see an increase in the number of events found for Phoebe in 2018 and for the satellites of Jupiter in 2019-2020.
+This is because at that periods these satellites will cross the apparent galactic plane.
+In contrast, for Nereid and Triton, almost no occultation was found because Neptune is in a very low dense region of stars and move more slowly than the other planets.
+We call attention that about 10\% (516) of the events will involve stars brighter than magR=14 (and almost 25\% brighter than magR=15), which helps the attempt of amateur observers.
+
 Table \ref{Tab: occ-list} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps.
@@
 
+For this, we generated new numerical integration of the orbits of the irregular satellites of Jupiter.
+A specific short-time ephemeris were obtained using only the positions of \cite{GomesJunior2015}.
+The systematic errors found in the JPL ephemeris \citep{Jacobson2012} were corrected.
+We also updated the ephemeris of \cite{Desmars2013} using the observations of \cite{GomesJunior2015}, from MPC and from Flagstaff.
+For Nereid and Triton we used the most recent ephemeris available \citep[, respectively]{Emelyanov2011, Emelyanov2015}.
+
 The candidate stars were searched in the UCAC4 catalogue, except for the candidates in 2016 for Triton and Nereid.

### 2016-03-10 Irregular-occ/Irregular-occ.tex
.

@@
 For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \cite{GomesJunior2015}, observations from Minor Planet Center and observations from Flagstaff.
-For Nereid and Triton we used the most recent ephemeris available.
+%For Nereid and Triton we used the most recent ephemeris available.
 
@@
 Phoebe, being the most studied object with a good measured size, can be used to calibrate and evaluate the technique for similar objects.
+Triton, although it is not an actual irregular satellite, was also include in the list because of its atmosphere which has a complicated and extreme seasonal cycle \citep{McKinnon2007, Elliot_2000}.
 
-Triton is an uncommon satellite.
+%Triton is an uncommon satellite.
 Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
@@
 But, in fact, given their relatively closer distances as compared to TNOs and Centaurs, and considering the current precision of their ephemeris and of star positions, we can now reliably predict the exact location and instant where the shadow of the occultation will cross the Earth.
-For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds).
+%For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds).
 Thus, in this case, if the accumulated error (2 sigma or 95\% confidence level) of ephemeris and star position be around 70 mas, we have a probability of about 30\% of observing a stellar occultation, which is quite satisfactory today (see discussion in section \ref{Sec: discussion}).
@@
 
-Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
+%Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
 In contrast to TNOs, the irregular satellites have much better ephemeris because the orbits of their host planets are better  known, their observational time span is much wider and covers many orbital periods.
@@
 
-For Triton, we used the most recent ephemeris published by \cite{Emelyanov2015}.
+Triton is an uncommon satellite.
+Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
+Because Triton's orbit size is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded from the irregular satellites' class, but still studied together by many authors \citep{Sheppard2005, Jewitt2007}.
+Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have the same origin as the TNOs \citep{Agnor2006}.
+However, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
+
+For this satellite, we used the most recent ephemeris published by \cite{Emelyanov2015} which is based on more observations than previous ephemeris and covers 4 years more than the ephemeris of \cite{Jacobson2009}.
 In Fig.
@@
 
+Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
+In contrast to TNOs, the irregular satellites have much better ephemeris because the orbits of their host planets are better  known, their observational time span is much wider and covers many orbital periods.
+Moreover, the irregular satellites are much closer to Earth which implies in a much smaller shadow path error in kilometers.
+These advantages may be somewhat balanced by the smaller sizes estimated for the irregular satellites.Thus, in comparison, the chances for a successful observation of an stellar occultation by an irregular satellite should be considered at least also as good as those by TNOs.
+
 Observing a stellar occultation demands a great effort.

### 2016-03-14 Irregular-occ/Irregular-occ.tex
.

@@
 
-We use the following notations: \begin{itemize}
+We use the following notations:
+\begin{itemize}
 \item $i$ and $l$ one of the irregular satellites of Jupiter
@@
 It represents a total number of 5886 observations from 1898 to 2014.
+In contrast, in \cite{Desmars2013} was used 3367 observations from 1898 to 2012.
+It represents an increase of almost 75\% in the number of observations.
 
@@
 \ref{Fig:eph-Triton} we compare it with the ephemeris published in \cite{Zhang2014} and the NEP081 JPL ephemeris \citep{Jacobson2009}.
-The offsets between the three ephemeris are smaller than 15 mas for the period 2015-2018.
+The offsets between the three ephemeris are smaller than 20 mas for the period 2015-2021.
 This value is much smaller than the apparent size of Triton (124 mas, see Table \ref{Tab: satellite-diameter}), which indicates a good probability of observing an event by this object.

### 2016-04-01 Irregular-occ/Irregular-occ.tex
.

@@
 
-\title[New orbits and predictions of stellar occultations by irregular satellites up to 2020]{New numerical integration of orbits of irregular satellites and prediction of stellar occultations up to 2020}
+\title[New orbits and predictions of stellar occultations by irregular satellites up to 2020]{New orbits for irregular satellites based on thousands of new observations and predictions of stellar occultations up to 2020}
 \author[A. R. Gomes-J\'unior, M. Assafin, L. Beauvalet et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
@@
 \begin{document}
-\newcommand{\noccs}{5457 } % total number of occultations
+\newcommand{\noccs}{5442 } % total number of occultations
 
@@
 \begin{abstract}
-Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system.
+%Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system.
 It is important to know their physical parameters, such as size, shape, albedo and composition, to trace back their true origin.
-The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects.
-We aim to predict stellar occultations for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda, and for the irregular satellites Phoebe of Saturn and Nereid of Neptune, and also for Triton.
-We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument.
-These catalogues were crossed with the ephemeris of the satellites to identify stellar occultations.
-We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future of the satellites of Jupiter.
-For Phoebe, we used an updated ephemeris and for Nereid and Triton we used recently published ephemeris.
-We managed to identify \noccs candidates of stellar occultations between the period of January, 2016 and December, 2020.
-We made observational tests for the prediction of the event of Himalia in March 03, 2015.
-The stars and objects were observed close to the date predicted and in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions.
-The comparisons between the predictions and the observation tests show a good agreement.
-We call attention for an occultation by Triton of a bright star (R=12.5) in October 5, 2017.
-The event can be observed from Europe and the east coast of USA and will be a good opportunity to access the current state of the atmosphere of the satellite.
-We also note that Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018.
-This results in a large number of occultations predicted for that period.
-We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
+The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects but this demands a very precise ephemeris based on new and good positions.
+%\cite{GomesJunior2015} published 6523 observations obtained between 1992 and 2014 for 18 irregular satellites with an estimated error in the positions of about 60 to 80 mas.
+%Based on this observations, we aimed to generate new numerical integration of the orbits of eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda.
+This is a short-time ephemeris that corrected systematic errors from previous ephemeris (JPL JUP300).
+For Phoebe we updated ephemeris from \cite{Desmars2013} using 75\% more positions than the previous one.
+%Also Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018.
+This creates a good opportunity to predict stellar occultations due to the large number of events that can be predicted in these periods.
+Thus, we aimed to predict stellar occultations for the same 9 satellites cited above.
+Using the UCAC4 catalogue we managed to identify \noccs candidates of stellar occultations between the period of January, 2016 and December, 2020.
+%We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
+\cite{GomesJunior2015} published 6523 new positions for 18 irregular satellites from observations made between 1992 and 2014 with an estimated error of about 60 to 80 mas.
+Based on this set of original positons, we performed numerical integrations for deriving new orbits for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda.
+This corrected systematic errors and resulted in an ephemeris more accurate/precise than JPL JUP300 for these bodies at least in the immediate future.
+For Phoebe we updated the ephemeris from \cite{Desmars2013} using 75\% more positions than the previous one.
+Due to their orbital configurations, it is common belief that the irregular satellites were captured by the giant planets in the early Solar System, but there is no consensus for a single model explaining where they were formed.
+Size, shape, albedo and composition would help a lot to trace back their true origin, but these physical parameters are yet poorly known for irregular satellites.
+The observation of stellar occultations would allow for the determination of such parameters.
+Indeed Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018, improving a lot the chances of observing such events in the near future.
+Using the derived ephemerides and the UCAC4 catalogue we managed to identify \noccs candidate stellar occultations between January 2016 and December 2020 for the 9 irregular satellites studied here.
+We discussed how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
 \end{abstract}
@@
 Irregular satellites revolve around giant planets at large distances in eccentric, highly inclined and frequently retrograde orbits.
-Because of these peculiar orbits, it is largely accepted that these objects did not form by accretion around their planet, but were captured in the early solar system \citep{Sheppard2005}.
+Because of these peculiar orbits, it is largely accepted that these objects did not form by accretion around their planet, but were captured in the early Solar System \citep{Sheppard2005}.
 
@@
 
-In order to obtain precise fundamental physical parameters like size and shape, thus constraining the albedo and in a broader sense also the composition for the irregular satellites and therefore to contribute to the study of their origin, we aim at observing stellar occultations, which provide  more accurate results than other ground-based techniques \citep{Sicardy2011, Ortiz2012, Braga-Ribas2014}.
-For that, reliable predictions of stellar occultations by these satellites are most needed.
-
-We present in this paper new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) for predict stellar occultations.
-This is a specific short-time ephemeris using only the positions obtained by \cite{GomesJunior2015}.
 \cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas.
@@
 They pointed out that the ephemeris of the irregular satellites have systematic errors that may reach 200 mas for some satellites.
-For an object at the distance of Jupiter, this represents an error langer than 700 km in the shadow path of an occultation.
 
-For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \cite{GomesJunior2015}, observations from Minor Planet Center and observations from Flagstaff.
-%For Nereid and Triton we used the most recent ephemeris available.
+We present in this paper new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) using only the positions obtained by \cite{GomesJunior2015}.
+We aim to eliminate the systematic errors in the JPL ephemeris (see Sec.
+\ref{Sec: integration}).
+%For an object at the distance of Jupiter, this represents an error langer than 700 km in the shadow path of an occultation.
+For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \cite{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Center and observations from Flagstaff.
 
-Also, with new and precise ephemeris, we present stellar occultation predictions for these 11 objects up to 2020.
 Phoebe, being the most studied object with a good measured size, can be used to calibrate and evaluate the technique for similar objects.
-Triton, although it is not an actual irregular satellite, was also include in the list because of its atmosphere which has a complicated and extreme seasonal cycle \citep{McKinnon2007, Elliot_2000}.
-
-%Triton is an uncommon satellite.
-Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
-Because Triton's orbit size is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded from the irregular satellites' class, but still studied together by many authors \citep{Sheppard2005, Jewitt2007}.
-Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have the same origin as the TNOs \citep{Agnor2006}.
-However, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
-The main motivation to study Triton by stellar occultations is to understand the evolution of its atmosphere due to Triton's complicated and extreme seasonal cycle \citep{McKinnon2007, Elliot_2000}.
-For all these reasons, Triton was also included as a target in this work.
-
-Excluding Triton \citep{Olkin1997, Elliot_2000}, no observation of a stellar occultation by an irregular satellite was published up to date.
+Up to date, no observation of a stellar occultation by an irregular satellite was published.
 Since their estimated sizes are very small (see Table \ref{Tab: satellite-diameter}), this may have discouraged earlier tries.
-But, in fact, given their relatively closer distances as compared to TNOs and Centaurs, and considering the current precision of their ephemeris and of star positions, we can now reliably predict the exact location and instant where the shadow of the occultation will cross the Earth.
-%For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds).
-Thus, in this case, if the accumulated error (2 sigma or 95\% confidence level) of ephemeris and star position be around 70 mas, we have a probability of about 30\% of observing a stellar occultation, which is quite satisfactory today (see discussion in section \ref{Sec: discussion}).
+But, in fact, given their relatively closer distances as compared to TNOs and Centaurs, and considering the precision of our ephemeris and of star positions, we can now reliably predict the exact location and instant where the shadow of the occultation will cross the Earth.
 
@@
 Phoebe & 32 & $212 \pm 1.4^{b}$ & 3 \tabularnewline
-\hdashline
-Nereid & 15 & $340 \pm 50^{c}$ & 4 \tabularnewline
-Triton & 124 & $2707 \pm 2.0^{c}$ & 5 \tabularnewline
+%\hdashline
+%Nereid & 15 & $340 \pm 50^{c}$ & 4 \tabularnewline
+%Triton & 124 & $2707 \pm 2.0^{c}$ & 5 \tabularnewline
 \hline
@@
 
-In section \ref{Sec: integration} we show the building of the new ephemeris.
-In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites and how they were made.
-Some tests made to check the accuracy of the predictions are presented in section \ref{Sec: testes} and the final discussion is presented in section \ref{Sec: discussion}.
+In section \ref{Sec: integration} the new determination of the orbits.
+In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites, including some tests made to check the accuracy of the predictions.
+The final discussion is presented in section \ref{Sec: discussion}.
 
-\section{Ephemeris} \label{Sec: integration}
+\section{Orbit computations} \label{Sec: integration}
 
-In order to improve ephemeris in short time spans in the near future for a more efficient prediction of stellar occultations by irregular satellites, we need predictions based on new and recent observations.
-In this context, \cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.
+\cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.
 
-Here we develop our own ephemeris based on the observations published in \cite{GomesJunior2015}.
-First, because the reduction was made with consistent and precise stellar catalogue and with a robust and reliable software \citep[PRAIA,][]{Assafin2011}.
+Here we compute new orbits based on the observations published in \cite{GomesJunior2015}.
+First, because the reduction was made with a consistent and precise stellar catalogue and with a robust astrometry \citep[PRAIA,][]{Assafin2011}.
 Second, besides recent observations, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets.
-For these reasons, only this set of positions was used for building the new ephemeris for the satellites of Jupiter.
+For these reasons, only this set of positions was used for the satellites of Jupiter.
+
+Due to the context of this work regarding to stellar occultations, the orbit fitting procedures used aimed primarily to derive precise ephemerides for the near future.
+The procedures allow for the improvement of orbits as new observations are added.
 
@@
 
-\subsection{Triton's and Nereid's ephemeris}
-
-Triton is an uncommon satellite.
+%\subsection{Triton's and Nereid's ephemeris}
+%
+%Triton is an uncommon satellite.
 Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
@@
 However, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
-
-For this satellite, we used the most recent ephemeris published by \cite{Emelyanov2015} which is based on more observations than previous ephemeris and covers 4 years more than the ephemeris of \cite{Jacobson2009}.
+%
+%For this satellite, we used the most recent ephemeris published by \cite{Emelyanov2015} which is based on more observations than previous ephemeris and covers 4 years more than the ephemeris of \cite{Jacobson2009}.
 In Fig.
@@
 This value is much smaller than the apparent size of Triton (124 mas, see Table \ref{Tab: satellite-diameter}), which indicates a good probability of observing an event by this object.
-
-\begin{figure*}
-\begin{centering}
-\subfigure[NEP081 JPL and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
-\subfigure[\protect\cite{Zhang2014} and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/Zhang-EME_Triton.eps} \label{Fig:eph-Triton-zhang}}
-\caption{Comparison between the \protect\cite{Emelyanov2015}, \protect\cite{Zhang2014} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Triton.}
-\label{Fig:eph-Triton}
-\end{centering}
-\end{figure*}
-
-For Nereid, we used the ephemeris published by \cite{Emelyanov2011}, which uses more recent observations than the JPL ephemeris published by \cite{Jacobson2009}.
+%
+%\begin{figure*}
+%\begin{centering}
+%\subfigure[NEP081 JPL and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
+%\subfigure[\protect\cite{Zhang2014} and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/Zhang-EME_Triton.eps} \label{Fig:eph-Triton-zhang}}
+%\caption{Comparison between the \protect\cite{Emelyanov2015}, \protect\cite{Zhang2014} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Triton.}
+%\label{Fig:eph-Triton}
+%\end{centering}
+%\end{figure*}
+%
+%For Nereid, we used the ephemeris published by \cite{Emelyanov2011}, which uses more recent observations than the JPL ephemeris published by \cite{Jacobson2009}.
 The comparison between the two ephemeris can be seen in Fig.
@@
 The declination ephemeris offsets are of the order of the apparent diameter (15 mas, see Table \ref{Tab: satellite-diameter}), resulting in fair chances (50\%) that the shadow crosses the expected Earth latitude predicted for an occultation.
-
-\begin{figure}
-\begin{centering}
-\includegraphics[width=8.8cm]{figures/JPL-EME_Nereid.eps}
-\caption{Comparison between the \protect\cite{Emelyanov2011} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Nereid.}
-\label{Fig:eph-Nereid}
-\end{centering}
-\end{figure}
-
-\cite{GomesJunior2015} published 902 new positions for Nereid from observations between 1992 and 2014.
+%
+%\begin{figure}
+%\begin{centering}
+%\includegraphics[width=8.8cm]{figures/JPL-EME_Nereid.eps}
+%\caption{Comparison between the \protect\cite{Emelyanov2011} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Nereid.}
+%\label{Fig:eph-Nereid}
+%\end{centering}
+%\end{figure}
+%
+%\cite{GomesJunior2015} published 902 new positions for Nereid from observations between 1992 and 2014.
 The development of a better ephemeris for Nereid using these new positions in the context discussed here of near future stellar occultations is desirable, but is out of the scope of this paper.
 
-\section{Prediction of occultations} \label{Sec: predictions}
+\section{Prediction of stellar occultations} \label{Sec: predictions}
+
+\subsection{Candidate events}
 
@@
 
-We predicted occultations for the 8 major irregular satellites of Jupiter,  Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope, and for Phoebe of Saturn and Triton and Nereid of Neptune.
-
-For Triton and Nereid, the candidates for stellar occultations in 2016 were searched using the WFI catalogue in the same way as the predictions for Centaurs and TNOs occultations by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.
-This catalogue contains the stars in the path of Neptune in the sky up to mid-2016.
-The catalogue was generated by observations made at the ESO 2p2 telescope (IAU code 809) using the Wide Field Imager (WFI) CCD mosaic detector.
-The filter used was the broad-band R filter ESO\#844 with $\lambda_c$ = 651.725 nm and $\Delta\lambda$ = 162.184 nm.
+We predicted occultations for the 8 major irregular satellites of Jupiter,  Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope, and for Phoebe of Saturn.
 
@@
 This is because at that periods these satellites will cross the apparent galactic plane.
-In contrast, for Nereid and Triton, almost no occultation was found because Neptune is in a very low dense region of stars and move more slowly than the other planets.
-We call attention that about 10\% (516) of the events will involve stars brighter than magR=14 (and almost 25\% brighter than magR=15), which helps the attempt of amateur observers.
-
-Table \ref{Tab: occ-list} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps.
-Since these objects are very small, the duration of each event is a few seconds.
-All the occultation tables and maps will be publicly available at the CDS.
-No star brighter than MagR*=18 will be occulted by Triton in 2016.
-For this satellite, we cut events with stars fainter than R=18, since for Triton the flux drop during such an occultation would be very small.
-In Fig.
-\ref{Fig: ocultacao} we show an example of an occultation map.
-This is one of the only two found occultation by Triton and it will happen in October 05, 2017.
-This event can be observed from Europe and the east coast of USA and will be a great opportunity to study in high resolution the atmosphere of the satellite.
+We call attention that about 10\% of the events will involve stars brighter than magR=14 (and almost 25\% brighter than magR=15), which helps the attempt of amateur observers.
 
@@
 Phoebe & 32 & 98 & 238 & 79 & 13 & 460 \tabularnewline
-\hdashline
-Nereid & 11$^{a}$ & 1 & 1 & 0 & 0 & 13 \tabularnewline
-Triton & -- & 1 & 1 & 0 & 0 & 2 \tabularnewline
+%\hdashline
+%Nereid & 11$^{a}$ & 1 & 1 & 0 & 0 & 13 \tabularnewline
+%Triton & -- & 1 & 1 & 0 & 0 & 2 \tabularnewline
 \hline
@@
 \end{table}
+
+Table \ref{Tab: sample-cds} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps.
+Since these objects are very small, the duration of each event is a few seconds.
+All the occultation tables and maps will be publicly available at the CDS.
+In Fig.
+\ref{Fig: ocultacao} we show an example of an occultation map.
+This is an occultation by Elara that will happen in February 21, 2017.
+This event can be observed from Australia and it is one of the best opportunities for this object due to the slow velocity of the event and it involves a bright star (MagR*=12.4).
 
@@
 Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - at the central instant of occultation (corrected by proper motions); C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees (zero at north of the star, increasing clockwise); $\nu$: velocity in the plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; {\it D}: planet range to Earth, in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
-\label{Tab: occ-list}
 \end{flushleft}
@@
 \begin{centering}
-\includegraphics[width = 8.8cm]{figures/Triton_2017-10-05T23:51:53.png}   
-\caption{Occultation map for Triton.% regarding to the first event sampled in Table \ref{Tab: occ-list}.
+\includegraphics[width = 8.8cm]{figures/Elara_2017-02-21T15:28:40.png}   
+\caption{Occultation map for Elara. % regarding to the first event sampled in Table \ref{Tab: occ-list}.
 The central red dot show the geocentric closest approach of the shadow.
@@
 Since we cannot foresee exactly when the GAIA catalogue will be released and when new and re-reduced satellite positions in the GAIA frame will become available, we decided to publish predictions up to the end of 2020.
-In the GAIA era, the occultation predictions will be updated.
+In the GAIA era, the occultations predicted will be updated.
 
-\subsection{Prediction test} \label{Sec: testes}
+\subsection{Robustness of predictions} \label{Sec: testes}
 
@@
 Moreover, the irregular satellites are much closer to Earth which implies in a much smaller shadow path error in kilometers.
-These advantages may be somewhat balanced by the smaller sizes estimated for the irregular satellites.Thus, in comparison, the chances for a successful observation of an stellar occultation by an irregular satellite should be considered at least also as good as those by TNOs.
+These advantages may be somewhat balanced by the smaller sizes estimated for the irregular satellites.
+Thus, in comparison, the chances for a successful observation of an stellar occultation by an irregular satellite should be considered at least also as good as those by TNOs.
 
@@
 And, in our case, the shadow covers a very restricted area on Earth because of the size of the irregular satellites.
-Since no stellar occultation by an irregular satellite was observed up to date, with the exception of Triton, and since we want to be sure that we can start observational campaigns with reasonable chances of success, we tested an occultation prediction for a large target, to assess the quality of the prediction.
+Since no stellar occultation by an irregular satellite was observed up to date, and since we want to be sure that we can start observational campaigns with reasonable chances of success, we tested the robustness of an occultation prediction for a large target.
 
@@
 Thus, the relative positions between the two objects had minimal influence of the errors of the reference catalogue of stars used and possible field distortions \citep[and references therein]{Peng2008}.
-The relative positions of the star and satellite were used to check the original predictions.
-Notice that in the test we do not attempt to observe any actual occultation.
+The relative positions of the star and satellite were used to check the original prediction.
+Notice that in the test we did not attempt to observe any actual occultation.
 The test could be performed at any site, regardless of the Earth location where the occultation would in fact be visible.
@@
 
-We predict stellar occultations for the period of 2016-2020 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid.
-The procedure used was the same as that for the prediction of stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.
+We performed new numerical integration for improving the orbits of some of the larger irregular satellites.
+Consequently, with more precise ephemeris, we, then, predicted stellar occultations aiming to access fundamental parameters like size, shape, albedo, ultimately aiming to track the formation origin of these bodies.
 
-For this, we generated new numerical integration of the orbits of the irregular satellites of Jupiter.
-A specific short-time ephemeris were obtained using only the positions of \cite{GomesJunior2015}.
-The systematic errors found in the JPL ephemeris \citep{Jacobson2012} were corrected.
-We also updated the ephemeris of \cite{Desmars2013} using the observations of \cite{GomesJunior2015}, from MPC and from Flagstaff.
-For Nereid and Triton we used the most recent ephemeris available \citep[, respectively]{Emelyanov2011, Emelyanov2015}.
+For the irregular satellites of Jupiter (Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope), we produced ephemeris using only the observations of \cite{GomesJunior2015}.
+The systematic errors found in the JPL ephemeris \citep{Jacobson2012} and \cite{Emelyanov2008} were corrected.
+This new ephemeris was denominated Special Tailored Ephemeris.
 
-The candidate stars were searched in the UCAC4 catalogue, except for the candidates in 2016 for Triton and Nereid.
-In this case, we used the WFI catalogue that was generated from observations made with ESO2p2/WFI CCD mosaic that covered the path of Neptune in the sky-plane up to 2016 (see Sec.
-\ref{Sec: predictions}).
-From this, a total of \noccs events are foreseen.
+We also updated the ephemeris of Phoebe \citep{Desmars2013} using the observations of \cite{GomesJunior2015}, \cite{Peng2015} observations from  MPC and from Flagstaff.
+A total of 5886 observations between 1989 and 2014 were used in the process.
+This represents and increase of about 75\% in the number used to generate the orbits of Phoebe by \cite{Desmars2013}.
+
+\textcolor{red}{Colocar mais alguma coisa sobre as efemérides}
+
+We predict stellar occultations for the period of 2016-2020 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; and one satellite of Saturn: Phoebe.
+The procedure used was the same as that for the prediction of stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.The candidate stars were searched in the UCAC4 catalogue.
+The occasional passage of Jupiter by the galactic plane in 2019-2020 and Saturn in 2018 creates the best opportunity of observing stellar occultations in the near future due to the great density of stars in the region.
+Indeed, a total of \noccs events are foreseen.
 
@@
 
+\cite{GomesJunior2015} also observed Sycorax (satellite of Uranus) and Nereid (satellites of Neptune).
+Sycorax had a few observations on 9 nights in two years which did not cover one orbital period.
+For Nereid, the observations covered many orbital periods, but due to Nereid's large orbital eccentricity there are no observations near the pericenter.
+
+Uranus and Neptune are crossing a very low dense region of stars.
+This results in almost no stellar occultation by these objects up to 2020.
+In fact, using JPL ephemeris, we identified only 2 events for each satellite in this period, but due to the bad conditions of the events (shadow far from observatories; faint stars) we chose not to publish any events here.
+For these reasons we did not attempt to generate new orbits for these satellites here.
+
 Continuous observations of the satellites are recommended and fitting of our dynamical model to those observations are expected to reduce the respective STE ephemeris errors.
 The first version of the GAIA catalogue is to be released up to the end of 2016 and will improve the position error of the stars to the 1-5 mas level.
-It will allow for the discovery of ocultations by fainter stars not present in the UCAC4 catalogue.
-The release of the GAIA catalogue should have a positive impact on both the astrometric precision of occulted stars and the reduction of astrometric positions of the satellites.
+Re-reduction of older positions, and reduction of new positions of irregular satellites with GAIA will improve new orbit determinations.
+It will also allow for the discovery of ocultations by more stars not present in the UCAC4 catalogue.
+The release of the GAIA catalogue should have a positive impact on both the astrometric precision of occulted stars and the reduction of new astrometric positions of the satellites.
 As a result, prediction of stellar occultations by irregular satellites shall increase in number as well as in success.

### 2016-04-15 relatorio/Occ-Irregular/Relatorio.tex
.

--- 2016-04-01 Irregular-occ/Irregular-occ.tex
@@
-% mn2esample.tex
-%
-% v2.1 released 22nd May 2002 (G. Hutton)
-%
-% The mnsample.tex file has been amended to highlight
-% the proper use of LaTeX2e code with the class file
-% and using natbib cross-referencing. These changes
-% do not reflect the original paper by A. V. Raveendran.
-%
-% Previous versions of this sample document were
-% compatible with the LaTeX 2.09 style file mn.sty
-% v1.2 released 5th September 1994 (M. Reed)
-% v1.1 released 18th July 1994
-% v1.0 released 28th January 1994
-
-\documentclass[useAMS,usenatbib]{mn2e}
-\usepackage[brazilian,english]{babel}
+\documentclass[12pt,a4paper]{report}
+\usepackage[brazilian, english]{babel}
 \usepackage[utf8]{inputenc}
 \usepackage[T1]{fontenc}
-\usepackage{prettyref}
-\usepackage{amsmath}
+\usepackage{amsmath,amsthm,amsfonts,amssymb,textcomp}
+%\usepackage{latexsym}
 \usepackage{graphicx}
+\graphicspath{{figuras}}
 \usepackage{subfigure}
+\usepackage{float}
+\usepackage{longtable}
+\usepackage{color}
 \usepackage{epstopdf}
-\usepackage{color}
+\usepackage{pdflscape}
 \usepackage[breaklinks=true]{hyperref}
-\usepackage{arydshln} % creates horizontal dashed lines in tables
+\usepackage[comma,authoryear]{natbib}
+\usepackage[nonumberlist]{glossaries}
+\usepackage{arydshln}
+\usepackage{footnote}
+\usepackage{longtable}
+\usepackage[small,bf,singlelinecheck=off]{caption}
+\usepackage[left=3cm,right=2cm,top=3cm,bottom=2cm]{geometry}
+\usepackage{arydshln}
+\usepackage{scalefnt}
+%\usepackage[alf]{abntcite}
 
-%%%%% AUTHORS - PLACE YOUR OWN MACROS HERE %%%%%
+%%% newcommand %%%%%%%%%%%%%%%%%
+\newcommand{\PE}{Perkin-Elmer }
+\newcommand{\BC}{Boller \& Chivens }
+\newcommand{\noccs}{5442 }
 
-\title[New orbits and predictions of stellar occultations by irregular satellites up to 2020]{New orbits for irregular satellites based on thousands of new observations and predictions of stellar occultations up to 2020}
-\author[A. R. Gomes-J\'unior, M. Assafin, L. Beauvalet et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
-M. Assafin$^{1,\dag}$,
-L. Beauvalet$^{2,3}$,
-J. Desmars$^{4}$,
-R. Vieira-Martins$^{1,2,5,\dag}$,\newauthor
-J. I. B. Camargo$^{2,5}$,
-B. E. Morgado$^{1,2}$
-F. Braga-Ribas$^{2,6}$
-\\
-$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
-CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
-$^{2}$Observat\'orio Nacional/MCTI, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
-$^{3}$Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France\\
-$^{4}$Institut de m\'ecanique c\'eleste et de calcul des \'eph\'em\'erides - Observatoire de Paris, UMR 8028 du CNRS, 77 Av. Denfert-Rochereau,\\ 75014 Paris, France\\
-$^{5}$Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil\\
-$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
-}
+\newcommand{\degr}{\ensuremath{^{\circ}}}%                    % degree symbol:  °
+\newcommand{\arcmin}{\ensuremath{^{\prime}}}%                    % degree symbol:  °
+\newcommand{\arcsec}{\ensuremath{^{\prime\prime}}}%                    % degree symbol:  °
+\newcommand{\fs}{\mbox{\ensuremath{.\!\!^s}}}
+\newcommand{\farcm}{\mbox{\ensuremath{.\mkern-4mu^\prime}}}%    % fractional arcminute symbol: 0.'0
+\newcommand{\farcs}{\mbox{\ensuremath{.\!\!^{\prime\prime}}}}%  % fractional arcsecond symbol: 0.''0
+\newcommand{\fdg}{\mbox{\ensuremath{.\!\!^\circ}}}%             % fractional degree symbol:     0.°0
+
+%\makeglossaries
+
+\makeatletter
+\renewcommand\chapter{\thispagestyle{plain}
+                \global\@topnum\z@
+                \@afterindentfalse
+                \secdef\@chapter\@schapter}
+\makeatother
+
+\makeatletter
+\renewcommand{\@makechapterhead}[1]{%
+\vspace*{50 pt}%
+{\setlength{\parindent}{0pt} \raggedright \normalfont
+\bfseries\Huge\thechapter.\ #1
+\par\nobreak\vspace{40 pt}}}
+\makeatother
+
+%\input{glossario}
+\author{Altair Ramos Gomes Júnior}
+\title{Occultation Predictions of Irregular Satellites}
 \begin{document}
-\newcommand{\noccs}{5442 } % total number of occultations
-
-\date{Accepted . Received ; in original form }
-
-\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2016}
 
@@
 
-\label{firstpage}
+\pagestyle{headings}
 
-\begin{abstract}
-%Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system.
-It is important to know their physical parameters, such as size, shape, albedo and composition, to trace back their true origin.
-The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects but this demands a very precise ephemeris based on new and good positions.
-%\cite{GomesJunior2015} published 6523 observations obtained between 1992 and 2014 for 18 irregular satellites with an estimated error in the positions of about 60 to 80 mas.
-%Based on this observations, we aimed to generate new numerical integration of the orbits of eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda.
-This is a short-time ephemeris that corrected systematic errors from previous ephemeris (JPL JUP300).
-For Phoebe we updated ephemeris from \cite{Desmars2013} using 75\% more positions than the previous one.
-%Also Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018.
-This creates a good opportunity to predict stellar occultations due to the large number of events that can be predicted in these periods.
-Thus, we aimed to predict stellar occultations for the same 9 satellites cited above.
-Using the UCAC4 catalogue we managed to identify \noccs candidates of stellar occultations between the period of January, 2016 and December, 2020.
-%We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
-\cite{GomesJunior2015} published 6523 new positions for 18 irregular satellites from observations made between 1992 and 2014 with an estimated error of about 60 to 80 mas.
-Based on this set of original positons, we performed numerical integrations for deriving new orbits for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda.
-This corrected systematic errors and resulted in an ephemeris more accurate/precise than JPL JUP300 for these bodies at least in the immediate future.
-For Phoebe we updated the ephemeris from \cite{Desmars2013} using 75\% more positions than the previous one.
-Due to their orbital configurations, it is common belief that the irregular satellites were captured by the giant planets in the early Solar System, but there is no consensus for a single model explaining where they were formed.
-Size, shape, albedo and composition would help a lot to trace back their true origin, but these physical parameters are yet poorly known for irregular satellites.
-The observation of stellar occultations would allow for the determination of such parameters.
-Indeed Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018, improving a lot the chances of observing such events in the near future.
-Using the derived ephemerides and the UCAC4 catalogue we managed to identify \noccs candidate stellar occultations between January 2016 and December 2020 for the 9 irregular satellites studied here.
-We discussed how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
-\end{abstract}
-
-\begin{keywords}
-occultations - planets and satellites: general - planets and satellites: individual: Jovian and Saturnian irregular satellites - planets and satellites: individual: Triton
-\end{keywords}
-
-\section{Introduction}\label{Sec: introducao}
+\section*{Introduction}\label{Sec: introducao}
 
@@
 
-\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas.
+Size, shape, albedo and composition would help a lot to trace back their true origin, but these physical parameters are yet poorly known for irregular satellites.
+The observation of stellar occultations would allow for the determination of such parameters.
+Indeed Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018, improving a lot the chances of observing such events in the near future.
+
+%\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas.
 For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}.
@@
 
-We present in this paper new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) using only the positions obtained by \cite{GomesJunior2015}.
-We aim to eliminate the systematic errors in the JPL ephemeris (see Sec.
-\ref{Sec: integration}).
+For this, we made new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) using only the positions obtained by \cite{GomesJunior2015}.
+We aim to eliminate the systematic errors in the JPL ephemeris pointed out by these authors.
 %For an object at the distance of Jupiter, this represents an error langer than 700 km in the shadow path of an occultation.
 For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \cite{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Center and observations from Flagstaff.
+With this, we predicted stellar occultation for the 9 objects.
+
+Using the derived ephemerides and the UCAC4 catalogue we managed to identify \noccs candidate stellar occultations between January 2016 and December 2020 for the 9 irregular satellites studied here.
 
@@
 
-In section \ref{Sec: integration} the new determination of the orbits.
-In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites, including some tests made to check the accuracy of the predictions.
-The final discussion is presented in section \ref{Sec: discussion}.
-
-\section{Orbit computations} \label{Sec: integration}
+\section*{Orbit computations} \label{Sec: integration}
 
@@
 
-\subsection{Special tailored Ephemerides (STE) for Jupiter irregular satellites}
+\subsection*{Special tailored Ephemerides (STE) for Jupiter irregular satellites}
 
@@
 
-Our numerical model describes the dynamical evolution of the irregular satellites of Jupiter in a jovicentric reference frame.
+%Our numerical model describes the dynamical evolution of the irregular satellites of Jupiter in a jovicentric reference frame.
 The satellites are submitted to the influence of the Sun and the rest of the solar system, as well as that of the Galilean satellites and the first harmonics of Jupiter's gravity field.
 The axes of the reference frame are those of the ICRS.
-
-We use the following notations:
-\begin{itemize}
-\item $i$ and $l$ one of the irregular satellites of Jupiter
-\item $J$ Jupiter 
-\item $j$ another body of the Solar System
-\item $M_j$ the mass of the $j$-th body, not an irregular satellite
-\item $m_i$ the mass of the irregular satellite $i$
-\item $\vec{r_i}$ the position of the $i$-th body with respect to the barycentre of Jupiter System
-\item $r_{ij}$ the distance between bodies $i$ and $j$  
-\item $R_J$ the radius of Jupiter
-\item $J_n$ the dynamic polar oblateness of the n-th order for Jupiter's gravity field
-\item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
-\item $\Phi_i$ is the inclination of the $i$-th satellite with respect to Jupiter's equator.
-\end{itemize}
-
-For an irregular satellite $i$, under the gravitational influence of Jupiter, the $\mathcal{N}-1$ other irregular satellites, the regular Jovian satellites and the rest of the Solar System ($\mathcal{N}$ bodies), the equation of motion is:
-\begin{equation}\begin{array}{ll}
-
-\ddot{\vec{r_i}}= & \displaystyle -GM_J\frac{\vec{r_J}-\vec{r_i}}{r_{iJ}^3}-\sum_{l=1,l\neq i}^\mathcal{N'}Gm_l\frac{\vec{r_l}-\vec{r_i}}{r_{il}^3}\\
-&\displaystyle -\sum_{j=1}^\mathcal{N}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3} \right)\\
- & \displaystyle +GM_J \nabla U_{\bar{l}\hat{J}} -\sum_{l=1}^\mathcal{N} Gm_l\nabla U_{\bar{l}\hat{J}}
-\end{array}
-\label{Eq:1}
-\end{equation}
-where the last term in the brackets and the last term in Eq. \ref{Eq:1} represent undirect perturbations. The oblateness potential seen by the body $i$ because of Jupiter is:
-\begin{equation}\begin{array}{ll}
-
-U_{\bar{l}\hat{J}}=&\displaystyle -\frac{R_J^2 J_2}{r_{iJ}^3}\left(\frac{3}{2}\sin^2 \Phi_i-\frac{1}{2}\right)\\ &\\ & 
-\displaystyle-\frac{R_J^4 J_4}{r_{iJ}^5}\left(\frac{35}{8}\sin^4 \Phi_i-\frac{15}{4}\sin^2 \Phi_i+\frac{3}{8}\right)\\
-& \\
-&\displaystyle-\frac{R_J^6 J_6}{r_{iJ}^7}\left(\frac{231}{16}\sin^6 \Phi_i-\frac{315}{16}\sin^4 \Phi_i+\frac{105}{16}\sin^2 \Phi_i-\frac{5}{16}\right)
-
-\end{array}
-\end{equation}
-
-The expressions of $\nabla_lU_{\bar{l}\, \hat{i}}$ and $\nabla_i U_{\bar{i}\, \hat{l}}.$ have been developed in \citet{Lainey2004}.
+%
+%We use the following notations:
+%\begin{itemize}
+%\item $i$ and $l$ one of the irregular satellites of Jupiter
+%\item $J$ Jupiter 
+%\item $j$ another body of the Solar System
+%\item $M_j$ the mass of the $j$-th body, not an irregular satellite
+%\item $m_i$ the mass of the irregular satellite $i$
+%\item $\vec{r_i}$ the position of the $i$-th body with respect to the barycentre of Jupiter System
+%\item $r_{ij}$ the distance between bodies $i$ and $j$  
+%\item $R_J$ the radius of Jupiter
+%\item $J_n$ the dynamic polar oblateness of the n-th order for Jupiter's gravity field
+%\item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
+%\item $\Phi_i$ is the inclination of the $i$-th satellite with respect to Jupiter's equator.
+%\end{itemize}
+%
+%For an irregular satellite $i$, under the gravitational influence of Jupiter, the $\mathcal{N}-1$ other irregular satellites, the regular Jovian satellites and the rest of the Solar System ($\mathcal{N}$ bodies), the equation of motion is:
+%\begin{equation}\begin{array}{ll}
+%
+%\ddot{\vec{r_i}}= & \displaystyle -GM_J\frac{\vec{r_J}-\vec{r_i}}{r_{iJ}^3}-\sum_{l=1,l\neq i}^\mathcal{N'}Gm_l\frac{\vec{r_l}-\vec{r_i}}{r_{il}^3}\\
+%&\displaystyle -\sum_{j=1}^\mathcal{N}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3} \right)\\
+% & \displaystyle +GM_J \nabla U_{\bar{l}\hat{J}} -\sum_{l=1}^\mathcal{N} Gm_l\nabla U_{\bar{l}\hat{J}}
+%\end{array}
+%\label{Eq:1}
+%\end{equation}
+%where the last term in the brackets and the last term in Eq. \ref{Eq:1} represent undirect perturbations. The oblateness potential seen by the body $i$ because of Jupiter is:
+%\begin{equation}\begin{array}{ll}
+%
+%U_{\bar{l}\hat{J}}=&\displaystyle -\frac{R_J^2 J_2}{r_{iJ}^3}\left(\frac{3}{2}\sin^2 \Phi_i-\frac{1}{2}\right)\\ &\\ & 
+%\displaystyle-\frac{R_J^4 J_4}{r_{iJ}^5}\left(\frac{35}{8}\sin^4 \Phi_i-\frac{15}{4}\sin^2 \Phi_i+\frac{3}{8}\right)\\
+%& \\
+%&\displaystyle-\frac{R_J^6 J_6}{r_{iJ}^7}\left(\frac{231}{16}\sin^6 \Phi_i-\frac{315}{16}\sin^4 \Phi_i+\frac{105}{16}\sin^2 \Phi_i-\frac{5}{16}\right)
+%
+%\end{array}
+%\end{equation}
+%
+%The expressions of $\nabla_lU_{\bar{l}\, \hat{i}}$ and $\nabla_i U_{\bar{i}\, \hat{l}}.$ have been developed in \citet{Lainey2004}.
 The equations of motion are integrated with the numerical integrator RADAU \citep{Everhart1985}.
-Our model was fitted to the observations through a least-squares procedure.
+%Our model was fitted to the observations through a least-squares procedure.
 The satellites were integrated one dynamical family at a time, to gain computing time, while losing minimum precision.
 Indeed, the interactions between satellites not belonging to the same dynamical family are negligible considering the short timespan of our integration.
-
-The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}.
-
-\begin{table*}
-\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0.
- }\label{Tab: sat_ell}
-\begin{center}
-\begin{tabular}{ccccccc}
-\hline\hline
-Satellite & a (km) & e & I$\degr$ & $\Omega\degr$ & $\omega\degr$ & $v\degr$ \\ 
-\hline
-Himalia &   11372100 $\pm$ 500    &    0.166 $\pm$ 0.002      &   45.14 $\pm$ 0.15      &   39.77 $\pm$ 0.19      &   351.48 $\pm$ 0.46      &   97.35 $\pm$ 0.48    \\
-Elara &   11741170 $\pm$ 690  &      0.222 $\pm$ 0.002      &   28.64 $\pm$ 0.18      &   68.42 $\pm$ 0.43      &   179.82 $\pm$ 0.56      &   339.08 $\pm$ 0.82  \\
-Lysithea &   11739900 $\pm$ 1300  &      0.136 $\pm$ 0.004      &    51.12 $\pm$ 0.27     &   5.53 $\pm$ 0.52      &   53.0 $\pm$ 1.5      &   318.9 $\pm$ 2.0   \\
-Leda &   11140300  $\pm$ 4300  &     0.173  $\pm$ 0.007     &   16.15  $\pm$ 0.75    &   272.6  $\pm$ 1.7    &   212.2  $\pm$ 3.6          &   218.8  $\pm$ 3.2  \\
-Pasiphae &  23425000  $\pm$ 5000    &     0.379  $\pm$ 0.001       &   152.44 $\pm$ 0.10      &   284.59 $\pm$ 0.21      &   135.96 $\pm$ 0.19      &   236.97 $\pm$ 0.16 \\
-Sinope &   22968800 $\pm$ 5200   &     0.316 $\pm$ 0.002      &   157.76 $\pm$ 0.12      &   256.62 $\pm$ 0.55      &   298.38 $\pm$ 0.55      &   167.57 $\pm$ 0.19    \\
-Carme &   24202924 $\pm$ 4800      &  0.242 $\pm$ 0.001      &   147.13 $\pm$ 0.10      &   154.01 $\pm$ 0.25      &   47.90 $\pm$ 0.29      &   234.41 $\pm$ 0.19  \\
-Ananke &  21683800  $\pm$ 7200  &     0.380 $\pm$ 0.002      &   172.29 $\pm$ 0.20      &   56.9 $\pm$ 1.2      &   123.3 $\pm$ 1.2      &   231.24 $\pm$ 0.21  \\
-%Themisto  & 7393800    &     0.198       &   25.77  &   220.0     &   216.3    &   262.1     \\
-\hline
-\end{tabular} 
-\end{center}
-\begin{flushleft}
-\textbf{Notes}: a: semimajor axis; e : excentricity; I: inclination relative to the equatorial reference plane J2000; $\Omega$: longitude of the ascending node; $\omega$: argument of periapsis; $v$: true anomaly.
-\end{flushleft}
-\end{table*}
+%
+%The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}.
+%
+%\begin{table*}
+%\scalefont{0.8}
+%\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0.
+% }\label{Tab: sat_ell}
+%\begin{center}
+%\begin{tabular}{ccccccc}
+%\hline\hline
+%Satellite & a (km) & e & I$\degr$ & $\Omega\degr$ & $\omega\degr$ & $v\degr$ \\ 
+%\hline
+%Himalia &   11372100 $\pm$ 500    &    0.166 $\pm$ 0.002      &   45.14 $\pm$ 0.15      &   39.77 $\pm$ 0.19      &   351.48 $\pm$ 0.46      &   97.35 $\pm$ 0.48    \\
+%Elara &   11741170 $\pm$ 690  &      0.222 $\pm$ 0.002      &   28.64 $\pm$ 0.18      &   68.42 $\pm$ 0.43      &   179.82 $\pm$ 0.56      &   339.08 $\pm$ 0.82  \\
+%Lysithea &   11739900 $\pm$ 1300  &      0.136 $\pm$ 0.004      &    51.12 $\pm$ 0.27     &   5.53 $\pm$ 0.52      &   53.0 $\pm$ 1.5      &   318.9 $\pm$ 2.0   \\
+%Leda &   11140300  $\pm$ 4300  &     0.173  $\pm$ 0.007     &   16.15  $\pm$ 0.75    &   272.6  $\pm$ 1.7    &   212.2  $\pm$ 3.6          &   218.8  $\pm$ 3.2  \\
+%Pasiphae &  23425000  $\pm$ 5000    &     0.379  $\pm$ 0.001       &   152.44 $\pm$ 0.10      &   284.59 $\pm$ 0.21      &   135.96 $\pm$ 0.19      &   236.97 $\pm$ 0.16 \\
+%Sinope &   22968800 $\pm$ 5200   &     0.316 $\pm$ 0.002      &   157.76 $\pm$ 0.12      &   256.62 $\pm$ 0.55      &   298.38 $\pm$ 0.55      &   167.57 $\pm$ 0.19    \\
+%Carme &   24202924 $\pm$ 4800      &  0.242 $\pm$ 0.001      &   147.13 $\pm$ 0.10      &   154.01 $\pm$ 0.25      &   47.90 $\pm$ 0.29      &   234.41 $\pm$ 0.19  \\
+%Ananke &  21683800  $\pm$ 7200  &     0.380 $\pm$ 0.002      &   172.29 $\pm$ 0.20      &   56.9 $\pm$ 1.2      &   123.3 $\pm$ 1.2      &   231.24 $\pm$ 0.21  \\
+%%Themisto  & 7393800    &     0.198       &   25.77  &   220.0     &   216.3    &   262.1     \\
+%\hline
+%\end{tabular} 
+%\end{center}
+%\begin{flushleft}
+%\textbf{Notes}: a: semimajor axis; e : excentricity; I: inclination relative to the equatorial reference plane J2000; $\Omega$: longitude of the ascending node; $\omega$: argument of periapsis; $v$: true anomaly.
+%\end{flushleft}
+%\end{table*}
 
@@
 \begin{figure}
-\includegraphics[width=8.8cm]{figures/Carme_ephemeris.eps}
+\includegraphics[width=16.0cm]{figures/Carme_ephemeris.eps}
 \caption{Offsets in declination of the positions published by \protect\cite{GomesJunior2015} for Carme.
@@
 
-\subsection{Phoebe's ephemeris}
+\subsection*{Phoebe's ephemeris}
 
@@
 \begin{centering}
-\includegraphics[width=8.8cm]{figures/Phoebe.eps} 
+\includegraphics[width=16.0cm]{figures/Phoebe.eps} 
 \caption{Comparison between the PH15, SAT375 JPL and \protect\cite{Emelyanov2007} ephemeris for the satellite Phoebe.}
@@
 
-\section{Prediction of stellar occultations} \label{Sec: predictions}
+\section*{Prediction of stellar occultations} \label{Sec: predictions}
 
-\subsection{Candidate events}
+\subsection*{Candidate events}
 
-The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the ephemeris presented in Sec.
-\ref{Sec: integration}.
+The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the ephemeris presented in the previous section.
 The search for stellar candidates follows the same procedure as presented by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.
@@
 
-\begin{table}
+\begin{table}[h]
 \caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from January, 2016 up to December, 2020.}
@@
 \par \end{centering}
-$^{a}$ {Using the WFI catalogue as explained in Sec. \ref{Sec: predictions}}.
 \end{table}
 
-Table \ref{Tab: sample-cds} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps.
+%Table \ref{Tab: sample-cds} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps.
 Since these objects are very small, the duration of each event is a few seconds.
@@
 
-\begin{table*}
-\caption{\label{Tab: sample-cds} A sample of stellar occultation predictions for Pasiphae}
-\begin{center}
-\begin{tabular}{cccrcccrcrr}
-\hline
-\hline
-d m Year~~~~h~~m~~~s & RA~~~(ICRS)~~~Dec & C/A & P/A & $\nu$ & {\it D} & $R^*$ & $\lambda$ & LST & $\mu_{\alpha*}$ & $\mu_{\delta}$ \tabularnewline
-\hline
-% & \multicolumn{2}{c}{Mean errors} & Nr & Nr &  UCAC4       \tabularnewline
-09 04 2016 03:58:19. & 11 14 36.7707 +07 39 20.7610 & 1.003 &  17.9 & -12.88 &  4.54 & 14.9 & 271. & 22:03 &  12. & -33. \tabularnewline 
-13 06 2016 00:16:12. & 11 12 48.5020 +07 06 43.3520 & 0.661 &  30.0 & +14.32 &  5.50 & 13.9 & 262. & 17:45 &  -1. &   1. \tabularnewline 
-27 06 2016 13:56:09. & 11 18 03.4160 +06 23 45.1940 & 1.707 &  28.0 & +20.29 &  5.74 & 11.7 &  44. & 16:53 &   4. & -10. \tabularnewline 
-18 07 2016 15:07:24. & 11 28 15.5076 +05 05 31.8060 & 0.942 &  26.7 & +27.80 &  6.05 & 14.0 &   8. & 15:40 &   4. &   4. \tabularnewline 
-22 07 2016 16:15:07. & 11 30 30.4310 +04 48 43.4340 & 0.644 & 206.5 & +29.04 &  6.11 & 14.6 & 348. & 15:27 &  23. & -24. \tabularnewline 
-24 07 2016 01:37:34. & 11 31 17.8471 +04 42 49.0540 & 0.029 & 206.6 & +29.46 &  6.12 & 15.1 & 206. & 15:22 &   2. &  -8. \tabularnewline 
-24 07 2016 17:37:18. & 11 31 40.7472 +04 39 57.5060 & 0.840 &  26.5 & +29.66 &  6.13 & 14.9 & 326. & 15:20 & -11. &  -1. \tabularnewline 
-\hline
-\end{tabular}
-\end{center}
-\begin{flushleft}
-\textbf{Notes}.
+%\begin{table*}
+%\caption{\label{Tab: sample-cds} A sample of stellar occultation predictions for Pasiphae}
+%\begin{center}
+%\begin{tabular}{cccrcccrcrr}
+%\hline
+%\hline
+%d m Year~~~~h~~m~~~s & RA~~~(ICRS)~~~Dec & C/A & P/A & $\nu$ & {\it D} & $R^*$ & $\lambda$ & LST & $\mu_{\alpha*}$ & $\mu_{\delta}$ \tabularnewline
+%\hline
+%% & \multicolumn{2}{c}{Mean errors} & Nr & Nr &  UCAC4       \tabularnewline
+%09 04 2016 03:58:19. & 11 14 36.7707 +07 39 20.7610 & 1.003 &  17.9 & -12.88 &  4.54 & 14.9 & 271. & 22:03 &  12. & -33. \tabularnewline 
+%13 06 2016 00:16:12. & 11 12 48.5020 +07 06 43.3520 & 0.661 &  30.0 & +14.32 &  5.50 & 13.9 & 262. & 17:45 &  -1. &   1. \tabularnewline 
+%27 06 2016 13:56:09. & 11 18 03.4160 +06 23 45.1940 & 1.707 &  28.0 & +20.29 &  5.74 & 11.7 &  44. & 16:53 &   4. & -10. \tabularnewline 
+%18 07 2016 15:07:24. & 11 28 15.5076 +05 05 31.8060 & 0.942 &  26.7 & +27.80 &  6.05 & 14.0 &   8. & 15:40 &   4. &   4. \tabularnewline 
+%22 07 2016 16:15:07. & 11 30 30.4310 +04 48 43.4340 & 0.644 & 206.5 & +29.04 &  6.11 & 14.6 & 348. & 15:27 &  23. & -24. \tabularnewline 
+%24 07 2016 01:37:34. & 11 31 17.8471 +04 42 49.0540 & 0.029 & 206.6 & +29.46 &  6.12 & 15.1 & 206. & 15:22 &   2. &  -8. \tabularnewline 
+%24 07 2016 17:37:18. & 11 31 40.7472 +04 39 57.5060 & 0.840 &  26.5 & +29.66 &  6.13 & 14.9 & 326. & 15:20 & -11. &  -1. \tabularnewline 
+%\hline
+%\end{tabular}
+%\end{center}
+%\begin{flushleft}
+%\textbf{Notes}.
 Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - at the central instant of occultation (corrected by proper motions); C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees (zero at north of the star, increasing clockwise); $\nu$: velocity in the plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; {\it D}: planet range to Earth, in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
-\end{flushleft}
-\end{table*}
+%\end{flushleft}
+%\end{table*}
 
@@
 \begin{centering}
-\includegraphics[width = 8.8cm]{figures/Elara_2017-02-21T15:28:40.png}   
+\includegraphics[width = 16.0cm]{figures/Elara_2017-02-21T15:28:40.png}   
 \caption{Occultation map for Elara. % regarding to the first event sampled in Table \ref{Tab: occ-list}.
@@
 
-The first preliminary catalogue version of the ESA astrometry satellite GAIA \citep{deBruijne2012} is expected to be released up to the end of 2016 (The catalogue with five-parameter astrometric solutions is up to the end of 2017).
+%The first preliminary catalogue version of the ESA astrometry satellite GAIA \citep{deBruijne2012} is expected to be released up to the end of 2016 (The catalogue with five-parameter astrometric solutions is up to the end of 2017).
 The precise star positions to be derived by GAIA will provide better predictions with the main source of error being the ephemeris.
@@
 
-\subsection{Robustness of predictions} \label{Sec: testes}
+\subsection*{Robustness of predictions} \label{Sec: testes}
 
-Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
+%Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
 In contrast to TNOs, the irregular satellites have much better ephemeris because the orbits of their host planets are better  known, their observational time span is much wider and covers many orbital periods.
@@
 
-Observing a stellar occultation demands a great effort.
+%Observing a stellar occultation demands a great effort.
 And, in our case, the shadow covers a very restricted area on Earth because of the size of the irregular satellites.
-Since no stellar occultation by an irregular satellite was observed up to date, and since we want to be sure that we can start observational campaigns with reasonable chances of success, we tested the robustness of an occultation prediction for a large target.
+Since no stellar occultation by an irregular satellite was observed up to date, and since we want to be sure that we can start observational campaigns with reasonable chances of success,
 
+To check the availability of observing an occultations we tested the robustness of a prediction for a large target.
 The test design consisted in observing the object and star to be occulted near the date of the event predicted when the two objects were present in the same field of view (FOV), close to each other.
-Thus, the relative positions between the two objects had minimal influence of the errors of the reference catalogue of stars used and possible field distortions \citep[and references therein]{Peng2008}.
+Thus, the relative positions between the two objects had minimal influence of the errors of the reference catalogue of stars used and possible field distortions %\citep[and references therein]{Peng2008}.
 The relative positions of the star and satellite were used to check the original prediction.
@@
 We tested the occultation by Himalia predicted to occur on March 3, 2015. The shadow would cross the northern part of South America. For the event, four situations were considered:
-\begin{enumerate}[I]
-\item Our nominal, published prediction with the STE ephemeris (see Sec. \ref{Sec: integration}), and the nominal UCAC4 position of the star;
+\begin{enumerate}
+\item Our nominal, published prediction with the STE ephemeris, and the nominal UCAC4 position of the star;
 \item Prediction with the JPL ephemeris and the nominal UCAC4 position of the star;
@@
 
-\section{Discussion} \label{Sec: discussion}
+\section*{Discussion} \label{Sec: discussion}
 
@@
 
-\textcolor{red}{Colocar mais alguma coisa sobre as efemérides}
+%\textcolor{red}{Colocar mais alguma coisa sobre as efemérides}
 
@@
 
-\cite{GomesJunior2015} also observed Sycorax (satellite of Uranus) and Nereid (satellites of Neptune).
+\cite{GomesJunior2015} also observed Sycorax (satellite of Uranus) and Nereid (satellite of Neptune).
 Sycorax had a few observations on 9 nights in two years which did not cover one orbital period.
@@
 This results in almost no stellar occultation by these objects up to 2020.
-In fact, using JPL ephemeris, we identified only 2 events for each satellite in this period, but due to the bad conditions of the events (shadow far from observatories; faint stars) we chose not to publish any events here.
-For these reasons we did not attempt to generate new orbits for these satellites here.
+In fact, using JPL ephemeris, we identified only 2 events for each satellite in this period, but these are events with bad conditions (shadow far from observatories; faint stars).
+%For these reasons we did not attempt to generate new orbits for these satellites here.
 
@@
 
-\section*{Acknowledgements}
+\bibliography{references}
+\bibliographystyle{apalike}
 
-ARG-J thanks the financial support of CAPES.
-MA thanks the CNPq (Grants 473002/2013-2 and 308721/2011-0) and FAPERJ (Grant E-26/111.488/2013).
-RV-M thanks grants: CNPq-306885/2013, Capes/Cofecub-2506/2015, Faperj/PAPDRJ-45/2013.
-JIBC acknowledges CNPq for a PQ2 fellowship (process number 308489/2013-6).
-BEM thanks the financial support of CAPES.
-FB-R acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014.
-The numerical model of the satellites of Jupiter was developped during a post-doctoral contract funded by the Chinese Academy of Sciences (CAS) and supported by the National Scientific Fund of China (NSFC)
-
-\bibliographystyle{mn2e}
-\bibliography{references}
-
-\label{lastpage}
+%\bibliography{Referencias}
+%\bibliographystyle{apalike}
 

### 2016-04-27 relatorio/Occ-Irregular/Relatorio.tex
.

@@
 \end{center}
-References: (1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}; (4) \cite{Thomas1991}; (5) \cite{Thomas2000}.\\
+References: (1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}.\\
 $^{a}${Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
 $^{b}${From Cassini observations.}
-$^{c}${From Voyager-2 observations.}
 \par

### 2016-04-28 Irregular-occ/Irregular-occ.tex
.

@@
 \end{center}
-References: (1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}; (4) \cite{Thomas1991}; (5) \cite{Thomas2000}.\\
+References: (1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}..\\
 $^{a}${Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
 $^{b}${From Cassini observations.}
-$^{c}${From Voyager-2 observations.}
 \par
@@
 \par \end{centering}
-$^{a}$ {Using the WFI catalogue as explained in Sec. \ref{Sec: predictions}}.
 \end{table}

### 2016-05-23 Irregular-occ/Irregular-occ.tex
.

@@
 
-The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}.
+The obtained ephemeris is hereafter refered as STE, for special-tailored ephemeris. The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}.
 
@@
 
-We compared our ephemeris to the JPL for all the Jupiter satellites we fitted, until 2018.
-For instance, the divergence between 2015 and 2018 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.
+We compared the STE ephemeris to the JPL for all the Jupiter satellites we fitted, until 2021.
+For instance, the divergence between 2015 and 2021 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.
 
 Fig.
-\ref{Fig: JPL-STE} displays the offsets of the positions published by \cite{GomesJunior2015} for the satellite Carme in declination relative to our ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.We see that the systematic JPL ephemeris offsets pointed out by \cite{GomesJunior2015} are reduced with our ephemeris, as expected.
+\ref{Fig: JPL-STE} displays the offsets of the positions published by \cite{GomesJunior2015} for the satellite Carme in declination relative to the STE ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.We see that the systematic JPL ephemeris offsets pointed out by \cite{GomesJunior2015} are reduced with our ephemeris, as expected.
 
@@
 
-The obtained ephemeris is hereafter refered as STE, for special-tailored ephemeris.
+Table \ref{Tab:STE-error}
+
+\begin{table}
+\caption{\label{Tab:STE-error} Mean offsets of the \protect\cite{GomesJunior2015} positions relative to the STE, \protect\cite{Jacobson2012} and \protect\cite{Emelyanov2008} ephemeris .}
+\begin{centering}
+\begin{tabular}{lcccccc}
+\hline  \hline
+%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
+ & \multicolumn{2}{c}{STE} & \multicolumn{2}{c}{JPL} & \multicolumn{2}{c}{Eme2008}\tabularnewline
+Satellite  & RA & DEC & RA & DEC & RA & DEC \tabularnewline
+\hline
+Ananke & 12 & 16 & 49 & 359 & 187 & 623 \tabularnewline
+Carme & 20 & 14 & 30 & 369 & 220 & 653 \tabularnewline
+Elara & 14 & 16 & 33 & 305 & 193 & 561 \tabularnewline
+Himalia & 15 & 12 & 54 & 257 & 230 & 568 \tabularnewline
+Leda & 8 & 24 & 38 & 362 & 208 & 640 \tabularnewline
+Lysithea & 16 & 11 & 35 & 330 & 212 & 604 \tabularnewline
+Pasiphae & 20 & 19 & 44 & 362 & 206 & 651 \tabularnewline
+Sinope & 15 & 21 & 34 & 356 & 256 & 682 \tabularnewline
+\hline
+\end{tabular}
+\par \end{centering}
+\end{table}
 
@@
 \includegraphics[width=8.8cm]{figures/Phoebe.eps} 
-\caption{Comparison between the PH15, SAT375 JPL and \protect\cite{Emelyanov2007} ephemeris for the satellite Phoebe.}
+\caption{Comparison between the PH15 and SAT375 JPL ephemeris for the satellite Phoebe.}
 \label{Fig:eph-Phoebe}

### 2016-05-23 Irregular-occ/Josselin/graph.py

@@ -1,0 +1,90 @@
+import numpy as np
+import matplotlib.pyplot as plt
+import astropy.units as u
+
+
+#####################################################################
+
+plt.rcParams['text.latex.preamble']=[r"\usepackage{txfonts}"]
+
+params = {'text.usetex' : True,
+          'font.size' : 22,
+          'font.family' : 'txfonts',
+          'text.latex.unicode': True,
+          }
+          
+plt.rcParams.update(params)
+
+sizel = 17
+
+#####################################################################
+
+a = np.loadtxt('precision_all1898-2012.txt', unpack=True)
+b = np.loadtxt('precision_all1898-2014.txt', unpack=True)
+c = np.loadtxt('precision_gomesjunior2015.txt', unpack=True)
+
+plt.plot(b[0], b[3]*1000, label='all')
+plt.plot(a[0], a[3]*1000, label='all-GJ15')
+plt.plot(c[0], c[3]*1000, label='GJ15')
+plt.xlim(2016,2021)
+plt.ylim(0,20)
+plt.xlabel('Time (years)', fontsize=sizel)
+plt.ylabel('Error RA (mas)', fontsize=sizel)
+plt.tick_params(axis='both', which='major', labelsize=sizel)
+xt = np.arange(2016,2022,1)
+plt.xticks(xt, xt)
+plt.legend(labelspacing=0.25, borderpad=0.5, handlelength=1.7, prop={'size':sizel})
+fig =plt.gcf()
+fig.set_size_inches((17.6*u.cm).to(u.imperial.inch).value,(9.9*u.cm).to(u.imperial.inch).value)
+fig.savefig('PH15_err_RA.eps',dpi=300, format='eps', bbox_inches='tight')
+
+plt.clf()
+
+plt.plot(b[0], b[4]*1000, label='all')
+plt.plot(a[0], a[4]*1000, label='all-GJ15')
+plt.plot(c[0], c[4]*1000, label='GJ15')
+plt.xlim(2016,2021)
+plt.ylim(0,20)
+plt.xlabel('Time (years)', fontsize=sizel)
+plt.ylabel('Error DEC (mas)', fontsize=sizel)
+plt.tick_params(axis='both', which='major', labelsize=sizel)
+xt = np.arange(2016,2022,1)
+plt.xticks(xt, xt)
[…]

### 2016-05-24 Irregular-occ/Irregular-occ.tex
.

@@
 \newcommand{\noccs}{5442 } % total number of occultations
+\defcitealias{GomesJunior2015}{GJ15}
 
@@
 
-\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas.
+\cite{GomesJunior2015} (which shall be referred as \citetalias{GomesJunior2015} hereafter) obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas.
 For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}.
@@
 
-We present in this paper new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) using only the positions obtained by \cite{GomesJunior2015}.
+We present in this paper new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) using only the positions obtained by \citetalias{GomesJunior2015}.
 We aim to eliminate the systematic errors in the JPL ephemeris (see Sec.
@@
 %For an object at the distance of Jupiter, this represents an error langer than 700 km in the shadow path of an occultation.
-For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \cite{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Center and observations from Flagstaff.
+For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Center and observations from Flagstaff.
 
@@
 
-\cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.
+\citetalias{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.
 
-Here we compute new orbits based on the observations published in \cite{GomesJunior2015}.
+Here we compute new orbits based on the observations published in \citetalias{GomesJunior2015}.
 First, because the reduction was made with a consistent and precise stellar catalogue and with a robust astrometry \citep[PRAIA,][]{Assafin2011}.
@@
 Fig.
-\ref{Fig: JPL-STE} displays the offsets of the positions published by \cite{GomesJunior2015} for the satellite Carme in declination relative to the STE ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.We see that the systematic JPL ephemeris offsets pointed out by \cite{GomesJunior2015} are reduced with our ephemeris, as expected.
+\ref{Fig: JPL-STE} displays the offsets of the positions published by \citetalias{GomesJunior2015} for the satellite Carme in declination relative to the STE ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.We see that the systematic JPL ephemeris offsets pointed out by \citetalias{GomesJunior2015} are reduced with our ephemeris, as expected.
 
@@
 \includegraphics[width=8.8cm]{figures/Carme_ephemeris.eps}
-\caption{Offsets in declination of the positions published by \protect\cite{GomesJunior2015} for Carme.
+\caption{Offsets in declination of the positions published by \protect\citetalias{GomesJunior2015} for Carme.
 The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the JUP300 JPL ephemeris and the green dot relative to \protect\cite{Emelyanov2008}.
-As expected, the ephemeris systematic errors pointed out by \protect\cite{GomesJunior2015} are reduced with the STE ephemeris.
+As expected, the ephemeris systematic errors pointed out by \protect\citetalias{GomesJunior2015} are reduced with the STE ephemeris.
 \label{Fig: JPL-STE}}
@@
 
-Table \ref{Tab:STE-error}
+In Table \ref{Tab:STE-error} it is presented the mean offsets of the \citetalias{GomesJunior2015} positions relative to the same 3 ephemeris as above.
+We can see that the mean offsets are highly reduced with the STE ephemeris.
 
 \begin{table}
-\caption{\label{Tab:STE-error} Mean offsets of the \protect\cite{GomesJunior2015} positions relative to the STE, \protect\cite{Jacobson2012} and \protect\cite{Emelyanov2008} ephemeris .}
+\caption{\label{Tab:STE-error} Mean offsets of the \protect\citetalias{GomesJunior2015} positions relative to the STE, \protect\cite{Jacobson2012} and \protect\cite{Emelyanov2008} ephemeris .}
 \begin{centering}
@@
 \hline
-Ananke & 12 & 16 & 49 & 359 & 187 & 623 \tabularnewline
-Carme & 20 & 14 & 30 & 369 & 220 & 653 \tabularnewline
-Elara & 14 & 16 & 33 & 305 & 193 & 561 \tabularnewline
-Himalia & 15 & 12 & 54 & 257 & 230 & 568 \tabularnewline
-Leda & 8 & 24 & 38 & 362 & 208 & 640 \tabularnewline
-Lysithea & 16 & 11 & 35 & 330 & 212 & 604 \tabularnewline
-Pasiphae & 20 & 19 & 44 & 362 & 206 & 651 \tabularnewline
-Sinope & 15 & 21 & 34 & 356 & 256 & 682 \tabularnewline
+Ananke & $-10$ & $3$ & $60$ & $-108$ & $101$ & $-107$ \tabularnewline
+Carme & $14$ & $-1$ & $-3$ & $-80$ & $-6$ & $-45$ \tabularnewline
+Elara & $3$ & $-12$ & $20$ & $-50$ & $23$ & $-83$ \tabularnewline
+Himalia & $-15$ & $-7$  & $-19$  & $-11$ & $-18$  & $-13$ \tabularnewline
+Leda & $-9$ & $-8$ & $60$ & $-13$ & $166$ & $92$ \tabularnewline
+Lysithea & $15$ & $-21$ & $40$ & $-43$ & $117$ & $-76$ \tabularnewline
+Pasiphae & $4$ & $-16$ & $-17$ & $-82$ & $-10$ & $-54$ \tabularnewline
+Sinope & $9$ & $-4$ & $10$ & $-35$ & $11$ & $-52$ \tabularnewline
 \hline
@@
 The new ephemeris (PH15) used the same dynamical model, including the perturbations of the Sun and the eight planets, the eight major satellites of Saturn and the $J_2$ parameter.
-The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \cite{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center\footnote{\url{http://lnfm1.sai.msu.ru/neb/nss/bsapoouf.htm}}), and observations from Flagstaff \citep{NOFS} between 2012 and 2014.
+The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center\footnote{\url{http://lnfm1.sai.msu.ru/neb/nss/bsapoouf.htm}}), and observations from Flagstaff \citep{NOFS} between 2012 and 2014.
 It represents a total number of 5886 observations from 1898 to 2014.
@@
 \label{Fig:eph-Phoebe}
+\end{centering}
+\end{figure}
+
+We computed the precision of the PH15 ephemeris considering three sets of observations: all the positions available; only the positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.
+In Fig.
+\ref{Fig:err-Phoebe} is shown the comparison between them for the time spam 2016-2021.
+It is possible to see that even considering only the positions of \citetalias{GomesJunior2015}, the estimated error of the ephemeris is smaller than 12 mas.
+The computed precision does not take into account the precision in the position of Saturn.
+
+\begin{figure}
+\begin{centering}
+\includegraphics[width=8.8cm]{figures/PH15_err_angsep.eps}  
+\caption{Comparison of the precision of the PH15 ephemeris where three different sets of positions were used to compute the ephemeris: all observations; only positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.}
+\label{Fig:err-Phoebe}
 \end{centering}
@@
 The precise star positions to be derived by GAIA will provide better predictions with the main source of error being the ephemeris.
-Astrometric reduction of observations published in \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions will be improved.
+Astrometric reduction of observations published in \citetalias{GomesJunior2015} will be revised with the GAIA catalogue and the predictions will be improved.
 Since we cannot foresee exactly when the GAIA catalogue will be released and when new and re-reduced satellite positions in the GAIA frame will become available, we decided to publish predictions up to the end of 2020.
@@
 
-For the irregular satellites of Jupiter (Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope), we produced ephemeris using only the observations of \cite{GomesJunior2015}.
+For the irregular satellites of Jupiter (Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope), we produced ephemeris using only the observations of \citetalias{GomesJunior2015}.
 The systematic errors found in the JPL ephemeris \citep{Jacobson2012} and \cite{Emelyanov2008} were corrected.
@@
 
-We also updated the ephemeris of Phoebe \citep{Desmars2013} using the observations of \cite{GomesJunior2015}, \cite{Peng2015} observations from  MPC and from Flagstaff.
+We also updated the ephemeris of Phoebe \citep{Desmars2013} using the observations of \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from  MPC and from Flagstaff.
 A total of 5886 observations between 1989 and 2014 were used in the process.
@@
 
-\textcolor{red}{Colocar mais alguma coisa sobre as efemérides}
+As it was shown for Phoebe, when we use only the positions of \citetalias{GomesJunior2015}, the ephemeris presents a precision in the order of those where all the positions were used.
+Moreover, the case of Phoebe is particular because we have many observations in a large time spam (1898 to 2014) including observations from Cassini.
+For the Jovian satellites with less observations, the precision we have by using only \citetalias{GomesJunior2015} observations would be quite equivalent or even better than the precision of other ephemeris for the short time explored in this work.
 
@@
 
-\cite{GomesJunior2015} also observed Sycorax (satellite of Uranus) and Nereid (satellite of Neptune).
+\citetalias{GomesJunior2015} also observed Sycorax (satellite of Uranus) and Nereid (satellite of Neptune).
 Sycorax had a few observations on 9 nights in two years which did not cover one orbital period.

### 2016-05-25 Irregular-occ/Irregular-occ.tex
.

@@
 L. Beauvalet$^{2,3}$,
-J. Desmars$^{4}$,
-R. Vieira-Martins$^{1,2,5,\dag}$,\newauthor
+J. Desmars$^{4}$,\newauthor
+R. Vieira-Martins$^{1,2,5,\dag}$,
 J. I. B. Camargo$^{2,5}$,
@@
 
-In Table \ref{Tab:STE-error} it is presented the mean offsets of the \citetalias{GomesJunior2015} positions relative to the same 3 ephemeris as above.
-We can see that the mean offsets are highly reduced with the STE ephemeris.
+In Table \ref{Tab:STE-error} it is presented the mean offsets and the respective standard deviation of the \citetalias{GomesJunior2015} positions relative to the same 3 ephemeris as above.
+We can see that the mean offsets are greatly reduced with the STE ephemeris.
 
-\begin{table}
-\caption{\label{Tab:STE-error} Mean offsets of the \protect\citetalias{GomesJunior2015} positions relative to the STE, \protect\cite{Jacobson2012} and \protect\cite{Emelyanov2008} ephemeris .}
+\begin{table*}
+\caption{\label{Tab:STE-error} Mean offsets and standard deviation of the \protect\citetalias{GomesJunior2015} positions relative to the STE, \protect\cite{Jacobson2012} and \protect\cite{Emelyanov2008} ephemeris .}
 \begin{centering}
-\begin{tabular}{lcccccc}
+\begin{tabular}{l|cc|cc|cc}
 \hline  \hline
 %\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
- & \multicolumn{2}{c}{STE} & \multicolumn{2}{c}{JPL} & \multicolumn{2}{c}{Eme2008}\tabularnewline
+ & \multicolumn{2}{|c|}{STE} & \multicolumn{2}{|c|}{JPL} & \multicolumn{2}{|c}{Eme2008}\tabularnewline
 Satellite  & RA & DEC & RA & DEC & RA & DEC \tabularnewline
 \hline
-Ananke & $-10$ & $3$ & $60$ & $-108$ & $101$ & $-107$ \tabularnewline
-Carme & $14$ & $-1$ & $-3$ & $-80$ & $-6$ & $-45$ \tabularnewline
-Elara & $3$ & $-12$ & $20$ & $-50$ & $23$ & $-83$ \tabularnewline
-Himalia & $-15$ & $-7$  & $-19$  & $-11$ & $-18$  & $-13$ \tabularnewline
-Leda & $-9$ & $-8$ & $60$ & $-13$ & $166$ & $92$ \tabularnewline
-Lysithea & $15$ & $-21$ & $40$ & $-43$ & $117$ & $-76$ \tabularnewline
-Pasiphae & $4$ & $-16$ & $-17$ & $-82$ & $-10$ & $-54$ \tabularnewline
-Sinope & $9$ & $-4$ & $10$ & $-35$ & $11$ & $-52$ \tabularnewline
+Ananke & $-10\pm90$ & $3\pm73$ & $60\pm127$ & $-108\pm99$ & $101\pm180$ & $-107\pm120$ \tabularnewline
+Carme & $14\pm73$ & $-1\pm51$ & $-3\pm114$ & $-80\pm102$ & $-6\pm108$ & $-45\pm80$ \tabularnewline
+Elara & $3\pm92$ & $-12\pm57$ & $20\pm92$ & $-50\pm69$ & $23\pm94$ & $-83\pm81$ \tabularnewline
+Himalia & $-15\pm66$ & $-7\pm54$  & $-19\pm80$  & $-11\pm52$ & $-18\pm72$  & $-13\pm53$ \tabularnewline
+Leda & $-9\pm67$ & $-8\pm77$ & $60\pm117$ & $-13\pm95$ & $166\pm162$ & $92\pm95$ \tabularnewline
+Lysithea & $15\pm79$ & $-21\pm68$ & $40\pm92$ & $-43\pm77$ & $117\pm193$ & $-76\pm185$ \tabularnewline
+Pasiphae & $4\pm89$ & $-16\pm57$ & $-17\pm130$ & $-82\pm85$ & $-10\pm102$ & $-54\pm74$ \tabularnewline
+Sinope & $9\pm79$ & $-4\pm47$ & $10\pm228$ & $-35\pm76$ & $11\pm227$ & $-52\pm63$ \tabularnewline
 \hline
@@
 \par \end{centering}
-\end{table}
+\end{table*}
 
@@
 In Fig.
-\ref{Fig:err-Phoebe} is shown the comparison between them for the time spam 2016-2021.
+\ref{Fig:err-Phoebe} we show the comparison between them for the time spam 2016-2021 in angular separation of Saturn.
 It is possible to see that even considering only the positions of \citetalias{GomesJunior2015}, the estimated error of the ephemeris is smaller than 12 mas.
@@
 \includegraphics[width=8.8cm]{figures/PH15_err_angsep.eps}  
-\caption{Comparison of the precision of the PH15 ephemeris where three different sets of positions were used to compute the ephemeris: all observations; only positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.}
+\caption{Comparison of the precision in angular separation of Saturn for the PH15 ephemeris where three different sets of positions were used to compute the ephemeris: all observations; only positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.}
 \label{Fig:err-Phoebe}
@@
 Moreover, the case of Phoebe is particular because we have many observations in a large time spam (1898 to 2014) including observations from Cassini.
-For the Jovian satellites with less observations, the precision we have by using only \citetalias{GomesJunior2015} observations would be quite equivalent or even better than the precision of other ephemeris for the short time explored in this work.
+For the Jovian satellites with less observations, the precision we have by using only \citetalias{GomesJunior2015} observations would be quite equivalent or even better than the precision of other ephemeris for the short time span explored in this work.
 

### 2016-05-31 Irregular-occ/Irregular-occ.tex
.

@@
 
-\title[New orbits and predictions of stellar occultations by irregular satellites up to 2020]{New orbits for irregular satellites based on thousands of new observations and predictions of stellar occultations up to 2020}
+\title[New orbits and predictions of stellar occultations by irregular satellites up to 2020]{New orbits of irregular satellites based on thousands of new observations and predictions for stellar occultations up to 2020}
 \author[A. R. Gomes-J\'unior, M. Assafin, L. Beauvalet et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
@@
 \cite{GomesJunior2015} published 6523 new positions for 18 irregular satellites from observations made between 1992 and 2014 with an estimated error of about 60 to 80 mas.
-Based on this set of original positons, we performed numerical integrations for deriving new orbits for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda.
-This corrected systematic errors and resulted in an ephemeris more accurate/precise than JPL JUP300 for these bodies at least in the immediate future.
+Based on this set of original positions, we performed numerical integrations for deriving new orbits for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda.
+This corrected systematic errors and resulted in an ephemeris more accurate/precise than JPL jup300 for these bodies at least in the immediate future.
 For Phoebe we updated the ephemeris from \cite{Desmars2013} using 75\% more positions than the previous one.
-Due to their orbital configurations, it is common belief that the irregular satellites were captured by the giant planets in the early Solar System, but there is no consensus for a single model explaining where they were formed.
+Due to their orbital characteristics, it is common belief that the irregular satellites were captured by the giant planets in the early Solar System, but there is no consensus for a single model explaining where they were formed.
 Size, shape, albedo and composition would help a lot to trace back their true origin, but these physical parameters are yet poorly known for irregular satellites.
@@
 Using the derived ephemerides and the UCAC4 catalogue we managed to identify \noccs candidate stellar occultations between January 2016 and December 2020 for the 9 irregular satellites studied here.
-We discussed how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
+We discussed how the successful observation of a stellar occultation by these objects is possible and present some of those potential occultations.
 \end{abstract}
@@
 \begin{keywords}
-occultations - planets and satellites: general - planets and satellites: individual: Jovian and Saturnian irregular satellites - planets and satellites: individual: Triton
+occultations - planets and satellites: general - planets and satellites: individual: Jovian and Saturnian irregular satellites
 \end{keywords}
@@
 \cite{Grav2003} and \cite{Grav2007} showed that the irregular satellites from the giant planets have their colors and spectral slopes similar to C-, D- and P-type asteroids, Centaurs and trans-neptunian objects (TNOs).
-This suggests that they may have come from different locations in the early solar system.
+This suggests that they may have come from different locations in the early Solar System.
 
@@
 
-We present in this paper new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) using only the positions obtained by \citetalias{GomesJunior2015}.
-We aim to eliminate the systematic errors in the JPL ephemeris (see Sec.
-\ref{Sec: integration}).
-%For an object at the distance of Jupiter, this represents an error langer than 700 km in the shadow path of an occultation.
+We present in this paper new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) using only the positions obtained by \citetalias{GomesJunior2015}%.
+We aim to eliminate the systematic errors in the JPL ephemeris
+ (see Sec. \ref{Sec: integration}). %For an object at the distance of Jupiter, this represents an error langer than 700 km in the shadow path of an occultation.
 For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Center and observations from Flagstaff.
@@
 \end{center}
-References: (1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}..\\
-$^{a}${Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
+References: (1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}.\\
+$^{a}${Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}\\
 $^{b}${From Cassini observations.}
@@
 
-In section \ref{Sec: integration} the new determination of the orbits.
+In section \ref{Sec: integration}, we present the new determination of the orbits.
 In section \ref{Sec: predictions}, we present the predictions of the stellar occultations by irregular satellites, including some tests made to check the accuracy of the predictions.
@@
 First, because the reduction was made with a consistent and precise stellar catalogue and with a robust astrometry \citep[PRAIA,][]{Assafin2011}.
-Second, besides recent observations, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets.
+Second, besides recent observations, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, allowing to fully constrain the orbit for the short time span explored in this work.
 For these reasons, only this set of positions was used for the satellites of Jupiter.
@@
 Due to the context of this work regarding to stellar occultations, the orbit fitting procedures used aimed primarily to derive precise ephemerides for the near future.
-The procedures allow for the improvement of orbits as new observations are added.
+The procedures allow for the improvement of orbits as recent observations are added.
 
-\subsection{Special tailored Ephemerides (STE) for Jupiter irregular satellites}
+\subsection{Special-Tailored Ephemerides (STE) for Jupiter irregular satellites}
 
-The last observations used to develop JPL current ephemeris of the irregular satellites of Jupiter were obtained in 2012 \citep{Jacobson2012}.
-As a result, the errors in the JPL ephemeris for the current epoch are large enough to prevent accurate predictions of stellar occultations without any corrections.
+The last observations used to develop current JPL ephemeris of the irregular satellites of Jupiter were obtained in 2012 \citep{Jacobson2012}.
+As a result, the errors in the JPL ephemeris for the current epoch may be probably too large to prevent accurate predictions of stellar occultations without any corrections.
 
@@
 
-The obtained ephemeris is hereafter refered as STE, for special-tailored ephemeris. The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}.
+The obtained ephemeris is hereafter refered as STE, for Special-Tailored Ephemeris. The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}.
 
 \begin{table*}
-\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0.
+\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0 with respect to the center of Jupiter.
  }\label{Tab: sat_ell}
@@
 The main risk of divergence over time comes from the possible absence of long-term effects when fitting to a short timespan of observations.
-If that were the case, our ephemeris would diverge too quickly to be of any use.
+If that was the case, our ephemeris would diverge too quickly to be of any use.
 JPL ephemerides are fitted over all the available observations.
 As a result, they will diverge less quickly than our own.
-Though they are no longer precise enough for our use, they remain a precious reference to identify whether our own model presents a quick divergence.
+Though they are not as much precise as required for the use, they remain a precious reference to identify whether our own model presents a quick divergence.
 
@@
 Fig.
-\ref{Fig: JPL-STE} displays the offsets of the positions published by \citetalias{GomesJunior2015} for the satellite Carme in declination relative to the STE ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.We see that the systematic JPL ephemeris offsets pointed out by \citetalias{GomesJunior2015} are reduced with our ephemeris, as expected.
+\ref{Fig: JPL-STE} displays the offsets of the positions published by \citetalias{GomesJunior2015} for the satellite Carme in declination relative to the STE ephemeris, to \cite{Jacobson2012} jup300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.We see that the systematic JPL ephemeris offsets pointed out by \citetalias{GomesJunior2015} are reduced with our ephemeris, as expected.
 
@@
 \caption{Offsets in declination of the positions published by \protect\citetalias{GomesJunior2015} for Carme.
-The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the JUP300 JPL ephemeris and the green dot relative to \protect\cite{Emelyanov2008}.
+The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the jup300 JPL ephemeris and the green dot relative to \protect\cite{Emelyanov2008}.
 As expected, the ephemeris systematic errors pointed out by \protect\citetalias{GomesJunior2015} are reduced with the STE ephemeris.
@@
 In Table \ref{Tab:STE-error} it is presented the mean offsets and the respective standard deviation of the \citetalias{GomesJunior2015} positions relative to the same 3 ephemeris as above.
-We can see that the mean offsets are greatly reduced with the STE ephemeris.
+We can see that the mean offsets as well as most of their standard deviations are greatly reduced with the STE ephemeris.
 
@@
  & \multicolumn{2}{|c|}{STE} & \multicolumn{2}{|c|}{JPL} & \multicolumn{2}{|c}{Eme2008}\tabularnewline
-Satellite  & RA & DEC & RA & DEC & RA & DEC \tabularnewline
+Satellite  & $\Delta\alpha\cos\delta$ (mas) & $\Delta\delta$ (mas) & $\Delta\alpha\cos\delta$ (mas) & $\Delta\delta$ (mas) & $\Delta\alpha\cos\delta$ (mas) & $\Delta\delta$ (mas) \tabularnewline
 \hline
@@
 In contrast, in \cite{Desmars2013} was used 3367 observations from 1898 to 2012.
-It represents an increase of almost 75\% in the number of observations.
+This represents an increase of almost 75\% in the number of observations, mainly with recent observations which is required for our purpose.
 
 In Fig.
-\ref{Fig:eph-Phoebe} we compare our ephemeris (PH15) with the SAT375 JPL\footnote{Jacobson, R.A.
+\ref{Fig:eph-Phoebe} we compare our ephemeris (PH15) with the sat375 JPL\footnote{Jacobson, R.A.
 2015-Feb-27.
-"Satellite Ephemeris: SAT375", JPL Satellite Ephemeris File Release, \url{ftp://ssd.jpl.nasa.gov/pub/eph/satellites/nio/LINUX_PC/sat375l.txt}} ephemeris.
+"Satellite Ephemeris: sat375", JPL Satellite Ephemeris File Release, \url{ftp://ssd.jpl.nasa.gov/pub/eph/satellites/nio/LINUX_PC/sat375l.txt}} ephemeris.
 The difference between them is smaller than 30 mas (< 10 mas in Declination).
@@
 \includegraphics[width=8.8cm]{figures/Phoebe.eps} 
-\caption{Comparison between the PH15 and SAT375 JPL ephemeris for the satellite Phoebe.}
+\caption{Comparison between the PH15 and sat375 JPL ephemeris for the satellite Phoebe.}
 \label{Fig:eph-Phoebe}
@@
 In Fig.
-\ref{Fig:err-Phoebe} we show the comparison between them for the time spam 2016-2021 in angular separation of Saturn.
+\ref{Fig:err-Phoebe} we show the comparison between them for the time span 2016-2021 in angular separation\footnote{\label{ft-angsep}The precision in angular separation is computed  by $\sigma_s = \sqrt{\sigma^2_{\alpha} \cos^2 \delta + \sigma^2_{\delta}}$ where $\sigma_{\alpha}$ and $\sigma_{\delta}$ are the precision in right ascension and declination.} of Saturn.
 It is possible to see that even considering only the positions of \citetalias{GomesJunior2015}, the estimated error of the ephemeris is smaller than 12 mas.
@@
 \includegraphics[width=8.8cm]{figures/PH15_err_angsep.eps}  
-\caption{Comparison of the precision in angular separation of Saturn for the PH15 ephemeris where three different sets of positions were used to compute the ephemeris: all observations; only positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.}
+\caption{Comparison of the precision in angular separation\textsuperscript{\ref{ft-angsep}} of Saturn for the PH15 ephemeris where three different sets of positions were used to compute the ephemeris: all observations; only positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.}
 \label{Fig:err-Phoebe}
@@
 
-Table \ref{Tab: sample-cds} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps.
+Table \ref{Tab: sample-cds} shows a sample of the catalogue of predicted occultations and their parameters, which are necessary to produce occultation maps.
 Since these objects are very small, the duration of each event is a few seconds.
@@
 Astrometric reduction of observations published in \citetalias{GomesJunior2015} will be revised with the GAIA catalogue and the predictions will be improved.
-Since we cannot foresee exactly when the GAIA catalogue will be released and when new and re-reduced satellite positions in the GAIA frame will become available, we decided to publish predictions up to the end of 2020.
-In the GAIA era, the occultations predicted will be updated.
+In that context, in the GAIA era, the occultations predicted will be updated.
 
@@
 \begin{enumerate}[I]
-\item Our nominal, published prediction with the STE ephemeris (see Sec. \ref{Sec: integration}), and the nominal UCAC4 position of the star;
-\item Prediction with the JPL ephemeris and the nominal UCAC4 position of the star;
-\item From star and satellite offsets calculated from observations made a few days before the occultation when the objects were very separated (different FOVs);
-\item Same as 3 but with the star and the satellite close in the same FOV.
+\item\label{item1} Our nominal, published prediction with the STE ephemeris (see Sec. \ref{Sec: integration}), and the nominal UCAC4 position of the star;
+\item\label{item2} Prediction with the JPL ephemeris and the nominal UCAC4 position of the star;
+\item\label{item3} Prediction from star and satellite offsets calculated from observations made a few days before the occultation when the objects were very separated (different FOVs);
+\item\label{item4} Same as \ref{item3} but with the star and the satellite close in the same FOV.
 \end{enumerate}
@@
 Table \ref{Tab: comparison-Himalia} shows the differences between the predictions in the four situations.
-For situation 3 we observed the objects on February 22 with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / pixel$) at the Observatório do Pico dos Dias, Brazil (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
+For situation \ref{item3} we observed the objects on February 22 with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / pixel$) at the Observatório do Pico dos Dias, Brazil (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
 On that day, Himalia and the star were observed in separate FOVs as they were still far apart.
 On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope (diameter = 1.6m; FOV = $5\farcm8$; pixel scale = $0\farcs17 / pixel$) at OPD just over an hour after the time scheduled for the event.
-Satellite and star were separated by about 16 arcsec, so very close to each other (situation 4).
+Satellite and star were separated by about 16 arcsec, so very close to each other (situation \ref{item4}).
 From the calculated offsets, the center of the shadow was obtained.
@@
 In the context of the test, for a 0 mas offset in C/A we would have 100\% probability of observing the occultation, and 0\% in the case of a C/A offset equal to or larger than 20 mas, the radius of Himalia.
-From Table 5, we have nearly 0\% probability of success in situation 3, for which the offset in C/A was -20 mas, but when the relative astrometry was poor, 10 days prior to the event.
-Once at the day of the event in situation 4, the C/A offset dropped to -9 mas only, corresponding to a 55\% probability of success.
-Comparison with the prediction using the JPL ephemeris (situation 2) gives a +11 mas C/A offset, or a compatibility of 45\% between the ephemerides.
+From Table 5, we have nearly 0\% probability of success in situation \ref{item3}, for which the offset in C/A was -20 mas, but when the relative astrometry was poor, 10 days prior to the event.
+Once at the day of the event in situation \ref{item4}, the C/A offset dropped to -9 mas only, corresponding to a 55\% probability of success.
+Comparison with the prediction using the JPL ephemeris (situation \ref{item2}) gives a +11 mas C/A offset, or a compatibility of 45\% between the ephemerides.
 All this suggests that there was a good probability of observing the event.
@@
 \hline
-STE & 00:39:51 UTC & $0\farcs703$ & i \tabularnewline
-JPL & -26s & +11mas (36km) & ii \tabularnewline % (284km)
-Feb. 22 Obs. & -14s & -20mas (65km) & iii \tabularnewline % (153km)
-Mar. 03 Obs. & -36s & -09mas (29km) & iv \tabularnewline % (393km)
+STE & 00:39:51 UTC & $0\farcs703$ & \ref{item1} \tabularnewline
+JPL & -26s & +11mas (36km) & \ref{item2} \tabularnewline % (284km)
+Feb. 22 Obs. & -14s & -20mas (65km) & \ref{item3} \tabularnewline % (153km)
+Mar. 03 Obs. & -36s & -09mas (29km) & \ref{item4} \tabularnewline % (393km)
 \hline
@@
 We performed new numerical integration for improving the orbits of some of the larger irregular satellites.
-Consequently, with more precise ephemeris, we, then, predicted stellar occultations aiming to access fundamental parameters like size, shape, albedo, ultimately aiming to track the formation origin of these bodies.
+Consequently, with more precise ephemeris, we predicted stellar occultations aiming to access fundamental parameters like size, shape, albedo, ultimately aiming to track the formation origin of these bodies.
 
 For the irregular satellites of Jupiter (Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope), we produced ephemeris using only the observations of \citetalias{GomesJunior2015}.
-The systematic errors found in the JPL ephemeris \citep{Jacobson2012} and \cite{Emelyanov2008} were corrected.
-This new ephemeris was denominated Special Tailored Ephemeris.
+The systematic errors found in the JPL ephemeris \citep{Jacobson2012} and \cite{Emelyanov2008} were improved.
+These new ephemerides are denominated Special-Tailored Ephemeris.
 
@@
 A total of 5886 observations between 1989 and 2014 were used in the process.
-This represents and increase of about 75\% in the number used to generate the orbits of Phoebe by \cite{Desmars2013}.
+This represents and increase of about 75\% in the number used to generate the ephemeris of Phoebe in \cite{Desmars2013}.
 
@@
 Indeed, a total of \noccs events are foreseen.
+The next time that Jupiter will cross the central side of the galactic plane will be in 2031 and Saturn in 2046-2047.
 
@@
 Observations a few days before the date of occultation predicted may improve the combined errors to 40-80 mas, depending on the magnitude of the objects.
+This probability is estimated for a single observing site, and we expect to reach higher probability with multi-sites.
 
@@
 Re-reduction of older positions, and reduction of new positions of irregular satellites with GAIA will improve new orbit determinations.
-It will also allow for the discovery of ocultations by more stars not present in the UCAC4 catalogue.
+It will also allow for the discovery of occultations by more stars not present in the UCAC4 catalogue.
 The release of the GAIA catalogue should have a positive impact on both the astrometric precision of occulted stars and the reduction of new astrometric positions of the satellites.
-As a result, prediction of stellar occultations by irregular satellites shall increase in number as well as in success.
+As a result, prediction of stellar occultations by irregular satellites shall increase in number as well as in precision and success.
 

### 2016-06-01 Irregular-occ/mnras-source/Irregular-occ.tex
.

--- 2016-05-31 Irregular-occ/Irregular-occ.tex
@@
 \usepackage[T1]{fontenc}
+\usepackage{times}  %%% neede to embed font 1 text
 \usepackage{prettyref}
@@
 
-\title[New orbits and predictions of stellar occultations by irregular satellites up to 2020]{New orbits of irregular satellites based on thousands of new observations and predictions for stellar occultations up to 2020}
+\title[Orbits and predictions for occultations up to 2020]{New orbits of irregular satellites based on thousands of new observations and predictions for stellar occultations up to 2020}
 \author[A. R. Gomes-J\'unior, M. Assafin, L. Beauvalet et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
@@
 %We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
-\cite{GomesJunior2015} published 6523 new positions for 18 irregular satellites from observations made between 1992 and 2014 with an estimated error of about 60 to 80 mas.
-Based on this set of original positions, we performed numerical integrations for deriving new orbits for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda.
+\cite{GomesJunior2015} published 6523 positions for 18 irregular satellites from observations made between 1992 and 2014 with an estimated error of about 60 to 80 mas.
+Based on this set of positions, we performed numerical integrations for deriving new orbits for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda.
 This corrected systematic errors and resulted in an ephemeris more accurate/precise than JPL jup300 for these bodies at least in the immediate future.
@@
 Due to their orbital characteristics, it is common belief that the irregular satellites were captured by the giant planets in the early Solar System, but there is no consensus for a single model explaining where they were formed.
-Size, shape, albedo and composition would help a lot to trace back their true origin, but these physical parameters are yet poorly known for irregular satellites.
+Size, shape, albedo and composition would help to trace back their true origin, but these physical parameters are yet poorly known for irregular satellites.
 The observation of stellar occultations would allow for the determination of such parameters.
 Indeed Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018, improving a lot the chances of observing such events in the near future.
-Using the derived ephemerides and the UCAC4 catalogue we managed to identify \noccs candidate stellar occultations between January 2016 and December 2020 for the 9 irregular satellites studied here.
-We discussed how the successful observation of a stellar occultation by these objects is possible and present some of those potential occultations.
+Using the derived ephemerides and the UCAC4 catalogue we managed to identify \noccs candidate stellar occultations between January 2016 and December 2020 for the 9 satellites studied here.
+We discussed how the successful observation of a stellar occultation by these objects is possible and present some potential occultations.
 \end{abstract}
@@
 \begin{keywords}
-occultations - planets and satellites: general - planets and satellites: individual: Jovian and Saturnian irregular satellites
+occultations - ephemerides - planets and satellites: general - planets and satellites: individual: Jovian and Saturnian irregular satellites
 \end{keywords}

### 2016-06-11 Irregulares/dados/Irregulares_results/final_CDS/teste.py
.

@@ -1,0 +1,12 @@
+for i in a:
+    n, ra, dec = np.loadtxt(i, usecols=[31,28,29], unpack=True)
+    print '{:45s}'.format(i), '{:3.0f} & {:3.0f}$\pm${:3.0f} & {:3.0f}$\pm${:3.0f}'.format(n.mean(), ra.mean()*1000, ra.std()*1000, dec.mean()*1000, dec.std()*1000)
+    
+def ret(s):
+    ra, dec = np.array([]), np.array([])
+    for i in a:
+        if s in i:
+            x, y = np.loadtxt(i, usecols=[0,1], unpack=True)
+            ra = np.hstack((ra,x))
+            dec = np.hstack((dec,y))
+    print '{:10s}'.format(s), '{} & {:3.0f}$\pm${:3.0f} & {:3.0f}$\pm${:3.0f}'.format(len(ra), ra.mean()*1000, ra.std()*1000, dec.mean()*1000, dec.std()*1000)

### 2016-06-11 Irregulares/dados/Irregulares_results/grafico_anom.py

@@ -20,10 +20,11 @@ nobj= [512, 511, 507, 506, 513, 510, 802, 508, 609, 509]
 pee = ['160-1', '160-total', '160-total', '160-total', '160-total', '160-1', '160-total', '160-total', '160-total', '160-total']
 iage = ['IAG_total', 'IAG1', 'IAG_total', 'IAG_total', 'IAG_total', 'IAG1', 'IAG_total', 'IAG1', 'IAG_total', 'IAG1']
 zeie = ['ZEI1', 'ZEI1', 'ZEI1', 'ZEI_total', 'ZEI1', 'ZEI1', 'ZEI1', 'ZEI1', 'ZEI1', 'ZEI1']
-locald = [2, 1, 1, 1, 1, 1, 1, 1, 1, 1]
+locald = [2, 3, 1, 1, 1, 1, 1, 1, 1, 1]
 localr = [2, 1, 1, 1, 1, 1, 1, 1, 1, 1]
 
 sizel = 18
+sizek = 14
 
 def julian(idx):
     f = open('julian_date_{}'.format(obj[idx]), 'w')
@@ -103,12 +104,12 @@ def plot(idx):
     plt.xlabel('True Anomaly', fontsize=sizel)
     plt.xticks(np.arange(0,361,45))
     plt.ylabel('Offset (mas)', fontsize=sizel)
-#    plt.legend(loc=locald[idx], numpoints=1, prop={'size':sizel})
+    plt.legend(loc=locald[idx], numpoints=1, prop={'size':sizek})
     plt.axhline(0, color='black')
     fig =plt.gcf()
     fig.set_size_inches(7.2,4.05)
     plt.tick_params(axis='both', which='major', labelsize=sizel)
-    fig.savefig('{}_DEC.eps'.format(obj[idx]),dpi=300, format='eps', bbox_inches='tight')
+    fig.savefig('{}_DEC.png'.format(obj[idx]),dpi=300, format='png', bbox_inches='tight')
     plt.clf()
 
     if os.path.isfile(arq160) == True:
@@ -127,12 +128,12 @@ def plot(idx):
     plt.xlabel('True Anomaly', fontsize=sizel)
     plt.xticks(np.arange(0,361,45))
     plt.ylabel('Offset (mas)', fontsize=sizel)
-#    plt.legend(loc=localr[idx], numpoints=1, prop={'size':sizel})
+    plt.legend(loc=localr[idx], numpoints=1, prop={'size':sizek})
[…]

### 2016-06-13 Irregulares/dados/Irregulares_results/grafico_anom.py
.

@@ -20,7 +20,7 @@ nobj= [512, 511, 507, 506, 513, 510, 802, 508, 609, 509]
 pee = ['160-1', '160-total', '160-total', '160-total', '160-total', '160-1', '160-total', '160-total', '160-total', '160-total']
 iage = ['IAG_total', 'IAG1', 'IAG_total', 'IAG_total', 'IAG_total', 'IAG1', 'IAG_total', 'IAG1', 'IAG_total', 'IAG1']
 zeie = ['ZEI1', 'ZEI1', 'ZEI1', 'ZEI_total', 'ZEI1', 'ZEI1', 'ZEI1', 'ZEI1', 'ZEI1', 'ZEI1']
-locald = [2, 3, 1, 1, 1, 1, 1, 1, 1, 1]
+locald = [2, 3, 1, 1, 1, 1, 1, 2, 1, 1]
 localr = [2, 1, 1, 1, 1, 1, 1, 1, 1, 1]
 
 sizel = 18

### 2016-06-22 Irregular-occ/Irregular-occ.tex
.

@@
 Our numerical model describes the dynamical evolution of the irregular satellites of Jupiter in a jovicentric reference frame.
-The satellites are submitted to the influence of the Sun and the rest of the solar system, as well as that of the Galilean satellites and the first harmonics of Jupiter's gravity field.
-The axes of the reference frame are those of the ICRS.
+The satellites are submitted to the influence of the Sun and the \textbf{main bodies of the solar system (from Mercury to Pluto, plus the Moon)}, as well as those of the Galilean satellites and the first harmonics of Jupiter's gravity field.
+The axis of the reference frame are \textbf{expected to be those of the equatorial reference frame J2000}.
 
-We use the following notations:
-\begin{itemize}
-\item $i$ and $l$ one of the irregular satellites of Jupiter
+We use the following notations: \begin{itemize}
+\item \textbf{in one dynamical family consisting of $\mathcal{N}$ irregular satellites, $i$ will stand for the one whose equation of motion we are considering, $l$ will stand for another irregular satellite in gravitationnal interaction belonging to the same family}
 \item $J$ Jupiter
-\item $j$ another body of the Solar System
-\item $M_j$ the mass of the $j$-th body, not an irregular satellite
+\item $j$ another body of the Solar System, \textbf{among the 10 included (Sun+planets+Pluto+Moon)}
+\item $M_j$ the mass of the $j$ body, not an irregular satellite
 \item $m_i$ the mass of the irregular satellite $i$
-\item $\vec{r_i}$ the position of the $i$-th body with respect to the barycentre of Jupiter System
+\item $\vec{r_i}$ the position of the $i$ body with respect \textbf{to the center of Jupiter}
 \item $r_{ij}$ the distance between bodies $i$ and $j$
 \item $R_J$ the radius of Jupiter
-\item $J_n$ the dynamic polar oblateness of the n-th order for Jupiter's gravity field
+\item $J_n$ the dynamic polar oblateness of the nth order for Jupiter's gravity field
 \item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
-\item $\Phi_i$ is the inclination of the $i$-th satellite with respect to Jupiter's equator.
+\item $\Phi_i$ is the latitude of the $i$ satellite with respect to Jupiter's equator.
 \end{itemize}
 
-For an irregular satellite $i$, under the gravitational influence of Jupiter, the $\mathcal{N}-1$ other irregular satellites, the regular Jovian satellites and the rest of the Solar System ($\mathcal{N}$ bodies), the equation of motion is:
+For an irregular satellite $i$, under the gravitational influence of Jupiter, the \textbf{seven other irregular satellites}, the regular Jovian satellites and\textbf{ the main bodies of the Solar System}, the equation of motion is:
+\begin{equation}
+\label{Eq:eq1}
+\begin{array}{ll}
+
+\ddot{\vec{r_i}}= & \displaystyle -GM_J\frac{\vec{r_J}-\vec{r_i}}{r_{iJ}^3}-\sum_{l=1,l\neq i}^\mathcal{N} Gm_l\frac{\vec{r_l}-\vec{r_i}}{r_{il}^3}\\
+&\displaystyle -\sum_{j=1}^{10}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3} \right)\\
+ & \displaystyle +GM_J \nabla U_{\bar{i}\hat{J}} -\sum_{l=1}^\mathcal{N} Gm_l\nabla U_{\bar{l}\hat{J}}
+\end{array}
+\end{equation}
+
+where the last term in brackets and the last term in Eq.
+\ref{Eq:eq1} represent undirect perturbations.
+The oblateness potential seen by \textbf{a satellite $i$ because of Jupiter is (with a similar expression for the oblateness seen by a satellite $l$)}:
 \begin{equation}\begin{array}{ll}
 
-\ddot{\vec{r_i}}= & \displaystyle -GM_J\frac{\vec{r_J}-\vec{r_i}}{r_{iJ}^3}-\sum_{l=1,l\neq i}^\mathcal{N'}Gm_l\frac{\vec{r_l}-\vec{r_i}}{r_{il}^3}\\
-&\displaystyle -\sum_{j=1}^\mathcal{N}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3} \right)\\
- & \displaystyle +GM_J \nabla U_{\bar{l}\hat{J}} -\sum_{l=1}^\mathcal{N} Gm_l\nabla U_{\bar{l}\hat{J}}
-\end{array}
-\label{Eq:1}
-\end{equation}
-where the last term in the brackets and the last term in Eq. \ref{Eq:1} represent undirect perturbations. The oblateness potential seen by the body $i$ because of Jupiter is:
-\begin{equation}\begin{array}{ll}
-
-U_{\bar{l}\hat{J}}=&\displaystyle -\frac{R_J^2 J_2}{r_{iJ}^3}\left(\frac{3}{2}\sin^2 \Phi_i-\frac{1}{2}\right)\\ &\\ & 
+U_{\bar{i}\hat{J}}=&\displaystyle -\frac{R_J^2 J_2}{r_{iJ}^3}\left(\frac{3}{2}\sin^2 \Phi_i-\frac{1}{2}\right)\\ &\\ &
 \displaystyle-\frac{R_J^4 J_4}{r_{iJ}^5}\left(\frac{35}{8}\sin^4 \Phi_i-\frac{15}{4}\sin^2 \Phi_i+\frac{3}{8}\right)\\
@@
 
-The expressions of $\nabla_lU_{\bar{l}\, \hat{i}}$ and $\nabla_i U_{\bar{i}\, \hat{l}}.$ have been developed in \citet{Lainey2004}.
-The equations of motion are integrated with the numerical integrator RADAU \citep{Everhart1985}.
+\textbf{The expressions of $\nabla U$ have been developed in \cite{Lainey2004}}.
+\textbf{The equations of motion are integrated with the 15th order numerical integrator RADAU \citep{Everhart1985} using a constant step of one day}.
+\textbf{The positions of the objects of the solar system are provided by the DE423 ephemeris \citep{Folkner2010}, while those of the galilean satellites are provided by NOE2010 \citep{Lainey2004}}.
 Our model was fitted to the observations through a least-squares procedure.

### 2016-06-27 Irregular-occ/Irregular-occ.tex
.

@@
 
-\title[Orbits and predictions for occultations up to 2020]{New orbits of irregular satellites based on thousands of new observations and predictions for stellar occultations up to 2020}
+\title[Orbits and predictions for occultations up to 2020]{New orbits of irregular satellites for the predictions of stellar occultations up to 2020}
+%\title[Orbits and predictions for occultations up to 2020]{New orbits of irregular satellites based on thousands of new observations and predictions for stellar occultations up to 2020}
 \author[A. R. Gomes-J\'unior, M. Assafin, L. Beauvalet et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail: altair08@astro.ufrj.br},
@@
 %We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
-\cite{GomesJunior2015} published 6523 positions for 18 irregular satellites from observations made between 1992 and 2014 with an estimated error of about 60 to 80 mas.
+\cite{GomesJunior2015} published \textbf{3613 positions for the 8 largest irregular satellites of Jupiter and 1787 positions for the largest irregular satellite of Saturn, Phoebe}.
+These observations were made between \textbf{1995} and 2014 and have an estimated error of about 60 to 80 mas.
 Based on this set of positions, we performed numerical integrations for deriving new orbits for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda.
-This corrected systematic errors and resulted in an ephemeris more accurate/precise than JPL jup300 for these bodies at least in the immediate future.
+%This corrected systematic errors and resulted in an ephemeris more accurate/precise than JPL jup300 for these bodies at least in the immediate future.
 For Phoebe we updated the ephemeris from \cite{Desmars2013} using 75\% more positions than the previous one.
@@
 
-\citetalias{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.
+\citetalias{GomesJunior2015} published \textbf{3613} precise positions for \textbf{the 8 largest irregular satellites of Jupiter} from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between \textbf{1995} and 2014.
 
@@
 
-The obtained ephemeris is hereafter refered as STE, for Special-Tailored Ephemeris. The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}.
+The obtained ephemeris is hereafter refered as STE, for Special-Tailored Ephemeris.
+The initial osculating elements at the origin of integration, \textbf{the number of observations used and the time-span where these observations are distributed} are presented in Table \ref{Tab: sat_ell}.
 
@@
 \begin{center}
-\begin{tabular}{ccccccc}
+\begin{tabular}{lcccccccc}
 \hline\hline
-Satellite & a (km) & e & I$\degr$ & $\Omega\degr$ & $\omega\degr$ & $v\degr$ \\ 
+Satellite & N & Time-span & a (km) & e & I$\degr$ & $\Omega\degr$ & $\omega\degr$ & $v\degr$ \\ 
 \hline
-Himalia &   11372100 $\pm$ 500    &    0.166 $\pm$ 0.002      &   45.14 $\pm$ 0.15      &   39.77 $\pm$ 0.19      &   351.48 $\pm$ 0.46      &   97.35 $\pm$ 0.48    \\
-Elara &   11741170 $\pm$ 690  &      0.222 $\pm$ 0.002      &   28.64 $\pm$ 0.18      &   68.42 $\pm$ 0.43      &   179.82 $\pm$ 0.56      &   339.08 $\pm$ 0.82  \\
-Lysithea &   11739900 $\pm$ 1300  &      0.136 $\pm$ 0.004      &    51.12 $\pm$ 0.27     &   5.53 $\pm$ 0.52      &   53.0 $\pm$ 1.5      &   318.9 $\pm$ 2.0   \\
-Leda &   11140300  $\pm$ 4300  &     0.173  $\pm$ 0.007     &   16.15  $\pm$ 0.75    &   272.6  $\pm$ 1.7    &   212.2  $\pm$ 3.6          &   218.8  $\pm$ 3.2  \\
-Pasiphae &  23425000  $\pm$ 5000    &     0.379  $\pm$ 0.001       &   152.44 $\pm$ 0.10      &   284.59 $\pm$ 0.21      &   135.96 $\pm$ 0.19      &   236.97 $\pm$ 0.16 \\
-Sinope &   22968800 $\pm$ 5200   &     0.316 $\pm$ 0.002      &   157.76 $\pm$ 0.12      &   256.62 $\pm$ 0.55      &   298.38 $\pm$ 0.55      &   167.57 $\pm$ 0.19    \\
-Carme &   24202924 $\pm$ 4800      &  0.242 $\pm$ 0.001      &   147.13 $\pm$ 0.10      &   154.01 $\pm$ 0.25      &   47.90 $\pm$ 0.29      &   234.41 $\pm$ 0.19  \\
-Ananke &  21683800  $\pm$ 7200  &     0.380 $\pm$ 0.002      &   172.29 $\pm$ 0.20      &   56.9 $\pm$ 1.2      &   123.3 $\pm$ 1.2      &   231.24 $\pm$ 0.21  \\
+Himalia & 1234 & 1995-2014 &   11372100 $\pm$ 500    &    0.166 $\pm$ 0.002      &   45.14 $\pm$ 0.15      &   39.77 $\pm$ 0.19      &   351.48 $\pm$ 0.46      &   97.35 $\pm$ 0.48    \\
+Elara & 636 & 1996-2014 &   11741170 $\pm$ 690  &      0.222 $\pm$ 0.002      &   28.64 $\pm$ 0.18      &   68.42 $\pm$ 0.43      &   179.82 $\pm$ 0.56      &   339.08 $\pm$ 0.82  \\
+Lysithea & 234 & 1996-2010 &   11739900 $\pm$ 1300  &      0.136 $\pm$ 0.004      &    51.12 $\pm$ 0.27     &   5.53 $\pm$ 0.52      &   53.0 $\pm$ 1.5      &   318.9 $\pm$ 2.0   \\
+Leda & 98 & 1996-2009 &   11140300  $\pm$ 4300  &     0.173  $\pm$ 0.007     &   16.15  $\pm$ 0.75    &   272.6  $\pm$ 1.7    &   212.2  $\pm$ 3.6          &   218.8  $\pm$ 3.2  \\
+Pasiphae & 609 & 1996-2013 &  23425000  $\pm$ 5000    &     0.379  $\pm$ 0.001       &   152.44 $\pm$ 0.10      &   284.59 $\pm$ 0.21      &   135.96 $\pm$ 0.19      &   236.97 $\pm$ 0.16 \\
+Sinope & 221 & 1996-2009 &   22968800 $\pm$ 5200   &     0.316 $\pm$ 0.002      &   157.76 $\pm$ 0.12      &   256.62 $\pm$ 0.55      &   298.38 $\pm$ 0.55      &   167.57 $\pm$ 0.19    \\
+Carme & 331 & 1996-2013 &   24202924 $\pm$ 4800      &  0.242 $\pm$ 0.001      &   147.13 $\pm$ 0.10      &   154.01 $\pm$ 0.25      &   47.90 $\pm$ 0.29      &   234.41 $\pm$ 0.19  \\
+Ananke & 250 & 1996-2010 &  21683800  $\pm$ 7200  &     0.380 $\pm$ 0.002      &   172.29 $\pm$ 0.20      &   56.9 $\pm$ 1.2      &   123.3 $\pm$ 1.2      &   231.24 $\pm$ 0.21  \\
 %Themisto  & 7393800    &     0.198       &   25.77  &   220.0     &   216.3    &   262.1     \\
@@
 \begin{flushleft}
-\textbf{Notes}: a: semimajor axis; e : excentricity; I: inclination relative to the equatorial reference plane J2000; $\Omega$: longitude of the ascending node; $\omega$: argument of periapsis; $v$: true anomaly.
+\textbf{Notes}: N: number of observations used; a: semimajor axis; e : excentricity; I: inclination relative to the equatorial reference plane J2000; $\Omega$: longitude of the ascending node; $\omega$: argument of periapsis; $v$: true anomaly.
 \end{flushleft}
@@
 Fig.
-\ref{Fig: JPL-STE} displays the offsets of the positions published by \citetalias{GomesJunior2015} for the satellite Carme in declination relative to the STE ephemeris, to \cite{Jacobson2012} jup300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.We see that the systematic JPL ephemeris offsets pointed out by \citetalias{GomesJunior2015} are reduced with our ephemeris, as expected.
+\ref{Fig: JPL-STE} displays the offsets of the positions published by \citetalias{GomesJunior2015} for the satellite Carme in declination relative to the STE ephemeris, to \cite{Jacobson2012} jup300 JPL ephemeris and \cite{Emelyanov2005}\footnote{\textbf{Last update: February 19, 2012}} ephemeris.
+\textbf{The DE431 planetary ephemeris \citep{Folkner2014} was used to calculate the positions of Jupiter for the 3 models}.
+We see that the systematic JPL ephemeris offsets pointed out by \citetalias{GomesJunior2015} are reduced with our ephemeris, as expected.
 
@@
 \caption{Offsets in declination of the positions published by \protect\citetalias{GomesJunior2015} for Carme.
-The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the jup300 JPL ephemeris and the green dot relative to \protect\cite{Emelyanov2008}.
+The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the jup300 JPL ephemeris and the green dot relative to \protect\cite{Emelyanov2005}.
 As expected, the ephemeris systematic errors pointed out by \protect\citetalias{GomesJunior2015} are reduced with the STE ephemeris.
@@
 \begin{table*}
-\caption{\label{Tab:STE-error} Mean offsets and standard deviation of the \protect\citetalias{GomesJunior2015} positions relative to the STE, \protect\cite{Jacobson2012} and \protect\cite{Emelyanov2008} ephemeris .}
+\caption{\label{Tab:STE-error} Mean offsets and standard deviation of the \protect\citetalias{GomesJunior2015} positions relative to the STE, \protect\cite{Jacobson2012} and \protect\cite{Emelyanov2005} ephemeris .}
 \begin{centering}
@@
 \hline
-Ananke & $-10\pm90$ & $3\pm73$ & $60\pm127$ & $-108\pm99$ & $101\pm180$ & $-107\pm120$ \tabularnewline
-Carme & $14\pm73$ & $-1\pm51$ & $-3\pm114$ & $-80\pm102$ & $-6\pm108$ & $-45\pm80$ \tabularnewline
+Himalia & $-15\pm66$ & $-7\pm54$  & $-19\pm80$  & $-11\pm52$ & $-18\pm72$  & $-13\pm53$ \tabularnewline
 Elara & $3\pm92$ & $-12\pm57$ & $20\pm92$ & $-50\pm69$ & $23\pm94$ & $-83\pm81$ \tabularnewline
-Himalia & $-15\pm66$ & $-7\pm54$  & $-19\pm80$  & $-11\pm52$ & $-18\pm72$  & $-13\pm53$ \tabularnewline
+Lysithea & $15\pm79$ & $-21\pm68$ & $40\pm92$ & $-43\pm77$ & $117\pm193$ & $-76\pm185$ \tabularnewline
 Leda & $-9\pm67$ & $-8\pm77$ & $60\pm117$ & $-13\pm95$ & $166\pm162$ & $92\pm95$ \tabularnewline
-Lysithea & $15\pm79$ & $-21\pm68$ & $40\pm92$ & $-43\pm77$ & $117\pm193$ & $-76\pm185$ \tabularnewline
 Pasiphae & $4\pm89$ & $-16\pm57$ & $-17\pm130$ & $-82\pm85$ & $-10\pm102$ & $-54\pm74$ \tabularnewline
 Sinope & $9\pm79$ & $-4\pm47$ & $10\pm228$ & $-35\pm76$ & $11\pm227$ & $-52\pm63$ \tabularnewline
+Carme & $14\pm73$ & $-1\pm51$ & $-3\pm114$ & $-80\pm102$ & $-6\pm108$ & $-45\pm80$ \tabularnewline
+Ananke & $-10\pm90$ & $3\pm73$ & $60\pm127$ & $-108\pm99$ & $101\pm180$ & $-107\pm120$ \tabularnewline
 \hline
@@
 For the irregular satellites of Jupiter (Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope), we produced ephemeris using only the observations of \citetalias{GomesJunior2015}.
-The systematic errors found in the JPL ephemeris \citep{Jacobson2012} and \cite{Emelyanov2008} were improved.
+The systematic errors found in the JPL ephemeris \citep{Jacobson2012} and \cite{Emelyanov2005} were improved.
 These new ephemerides are denominated Special-Tailored Ephemeris.
@@
 FB-R acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014.
-The numerical model of the satellites of Jupiter was developped during a post-doctoral contract funded by the Chinese Academy of Sciences (CAS) and supported by the National Scientific Fund of China (NSFC)
+The numerical model of the satellites of Jupiter was developped during a post-doctoral contract funded by the Chinese Academy of Sciences (CAS) and supported by the National Scientific Fund of China (NSFC).
 

### 2016-06-28 Irregular-occ/Irregular-occ.tex
.

@@
 
+\usepackage[normalem]{ulem} %%% para mostrar o texto riscado
+
 %%%%% AUTHORS - PLACE YOUR OWN MACROS HERE %%%%%
 
-\title[Orbits and predictions for occultations up to 2020]{New orbits of irregular satellites for the predictions of stellar occultations up to 2020}
+\title[Orbits and predictions for occultations up to 2020]{New short-term orbits of irregular satellites in order to predict stellar occultations up to 2020}
 %\title[Orbits and predictions for occultations up to 2020]{New orbits of irregular satellites based on thousands of new observations and predictions for stellar occultations up to 2020}
@@
 \item $j$ another body of the Solar System, \textbf{among the 10 included (Sun+planets+Pluto+Moon)}
-\item $M_j$ the mass of the $j$ body, not an irregular satellite
+\item $M_j$ the mass of the $j$-th body, not an irregular satellite
 \item $m_i$ the mass of the irregular satellite $i$
-\item $\vec{r_i}$ the position of the $i$ body with respect \textbf{to the center of Jupiter}
+\item $\vec{r_i}$ the position of the $i$-th body with respect \textbf{to the center of Jupiter}
 \item $r_{ij}$ the distance between bodies $i$ and $j$
@@
 \item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
-\item $\Phi_i$ is the latitude of the $i$ satellite with respect to Jupiter's equator.
+\item $\Phi_i$ is the \textbf{latitude} of the $i$-th satellite with respect to Jupiter's equator.
 \end{itemize}
@@
 \caption{Offsets in declination of the positions published by \protect\citetalias{GomesJunior2015} for Carme.
-The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the jup300 JPL ephemeris and the green dot relative to \protect\cite{Emelyanov2005}.
+The red "x" relative to the special-tailored ephemeris \textbf{(rms = 51)}, the blue "+" relative to the jup300 JPL ephemeris \textbf{(rms = 130)} and the green dot relative to \protect\cite{Emelyanov2005} \textbf{(rms = 92)}.
 As expected, the ephemeris systematic errors pointed out by \protect\citetalias{GomesJunior2015} are reduced with the STE ephemeris.
@@
 We can see that the mean offsets as well as most of their standard deviations are greatly reduced with the STE ephemeris.
+\textbf{It is obvious because the STE were based on the same observations with which they are compared.
+In contrast, two other ephemeris
+were based on different sets of observations.}
 
@@
 The new ephemeris (PH15) used the same dynamical model, including the perturbations of the Sun and the eight planets, the eight major satellites of Saturn and the $J_2$ parameter.
-The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center\footnote{\url{http://lnfm1.sai.msu.ru/neb/nss/bsapoouf.htm}}), and observations from Flagstaff \citep{NOFS} between 2012 and 2014.
+The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center \textbf{\citep{Arlot2009}}, and observations from Flagstaff \citep{NOFS} between 2012 and 2014.
 It represents a total number of 5886 observations from 1898 to 2014.
@@
 In Fig.
-\ref{Fig:err-Phoebe} we show the comparison between them for the time span 2016-2021 in angular separation\footnote{\label{ft-angsep}The precision in angular separation is computed  by $\sigma_s = \sqrt{\sigma^2_{\alpha} \cos^2 \delta + \sigma^2_{\delta}}$ where $\sigma_{\alpha}$ and $\sigma_{\delta}$ are the precision in right ascension and declination.} of Saturn.
+\ref{Fig:err-Phoebe} we show the comparison between them for the time span 2016-2021 in on-sky Phoebe-Saturn angular separation\footnote{\label{ft-angsep}The precision in angular separation is computed  by $\sigma_s = \sqrt{\sigma^2_{\alpha} \cos^2 \delta + \sigma^2_{\delta}}$ where $\sigma_{\alpha}$ and $\sigma_{\delta}$ are the precision in right ascension and declination.}.
 It is possible to see that even considering only the positions of \citetalias{GomesJunior2015}, the estimated error of the ephemeris is smaller than 12 mas.
@@
 \includegraphics[width=8.8cm]{figures/PH15_err_angsep.eps}  
-\caption{Comparison of the precision in angular separation\textsuperscript{\ref{ft-angsep}} of Saturn for the PH15 ephemeris where three different sets of positions were used to compute the ephemeris: all observations; only positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.}
+\caption{Comparison of the precision in on-sky Phoebe-Saturn angular separation\textsuperscript{\ref{ft-angsep}} for the PH15 ephemeris where three different sets of positions were used to compute the ephemeris: all observations; only positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.}
 \label{Fig:err-Phoebe}
@@
 This is because at that periods these satellites will cross the apparent galactic plane.
-We call attention that about 10\% of the events will involve stars brighter than magR=14 (and almost 25\% brighter than magR=15), which helps the attempt of amateur observers.
+We call attention that about 10\% of the events will involve stars brighter than magR=14  (and almost 25\% brighter than magR=15), \textbf{where magR stands for the UCAC4 magnitude in the R-band}, which helps the attempt of amateur observers.
 
@@
 Since these objects are very small, the duration of each event is a few seconds.
-All the occultation tables and maps will be publicly available at the CDS.
+All the occultation tables and maps will be publicly available at the CDS \textbf{(Centre de Données astronomiques de Strasbourg)}.
 In Fig.
@@
 \caption{Occultation map for Elara. % regarding to the first event sampled in Table \ref{Tab: occ-list}.
-The central red dot show the geocentric closest approach of the shadow.
-The small ones shows the center of the shadow separated by 60s.
+The central red dot shows the geocentric closest approach of the shadow.
+The small ones show the center of the shadow separated by 60s.
 The lines show the path of the shadow over the Earth.
 The shadow moves from right to left.
-\textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for a central observation; C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; vel: velocity of event in km/s; Delta: Geocentric distance to the occulting object in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$; long: east longitude of subplanet point in degrees, positive towards east.}
+\textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for\textbf{ a observation in the center of the shadow}; C/A: \textbf{apparent geocentric distance between the satellite and the star (a.k.a.
+the distance between the shadow and the center of the Earth) at the moment of the geocentric closest approach}, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; vel: \textbf{relative} velocity of event in km/s; Delta: Geocentric distance to the occulting object in AU; $R^*$: normalized \textbf{UCAC4 magnitude in the R-band} to a common shadow velocity of 20 km s$^{-1}$; long: east longitude of subplanet point in degrees, positive towards east, \textbf{at the instant of the geocentric closest approach}.}
 \label{Fig: ocultacao}
@@
 We performed new numerical integration for improving the orbits of some of the larger irregular satellites.
-Consequently, with more precise ephemeris, we predicted stellar occultations aiming to access fundamental parameters like size, shape, albedo, ultimately aiming to track the formation origin of these bodies.
+Consequently, with \textbf{our} ephemeris, we predicted stellar occultations aiming to access fundamental parameters like size, shape, albedo, ultimately aiming to track the formation origin of these bodies.
 
 For the irregular satellites of Jupiter (Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope), we produced ephemeris using only the observations of \citetalias{GomesJunior2015}.
-The systematic errors found in the JPL ephemeris \citep{Jacobson2012} and \cite{Emelyanov2005} were improved.
-These new ephemerides are denominated Special-Tailored Ephemeris.
+\textcolor{red}{The systematic errors found in the JPL ephemeris \citep{Jacobson2012} and \cite{Emelyanov2005} were improved.} These new ephemerides are denominated Special-Tailored Ephemeris.
 
@@
 Moreover, the case of Phoebe is particular because we have many observations in a large time spam (1898 to 2014) including observations from Cassini.
-For the Jovian satellites with less observations, the precision we have by using only \citetalias{GomesJunior2015} observations would be quite equivalent or even better than the precision of other ephemeris for the short time span explored in this work.
+For the Jovian satellites with less observations, the precision we have by using only \citetalias{GomesJunior2015} observations \textbf{may} be quite equivalent or even better than the precision of other ephemeris for the short time span explored in this work.
 
@@
 \citetalias{GomesJunior2015} also observed Sycorax (satellite of Uranus) and Nereid (satellite of Neptune).
-Sycorax had a few observations on 9 nights in two years which did not cover one orbital period.
+\textbf{There were few observations of Sycorax distributed in 9 nights over two years} which did not cover one orbital period.
 For Nereid, the observations covered many orbital periods, but due to Nereid's large orbital eccentricity there are no observations near the pericenter.

### 2016-06-29 Irregular-occ/referee_response.txt
.

Comments to the Author

REVIEWER:    This paper describes a very necessary work.
The authors recently published a huge portion of new astrometric observations of the eight largest irregular satellites of Jupiter and Phoebe, the ninth satellite of Saturn.
Thus there was an opportunity to create a new, more accurate ephemeris of these satellites.
The paper describes an attempt to do so.
    It is known that the ephemeris accuracy improves with an increase in observation time interval.
    Nevertheless, the authors limited the observation time interval for the eight irregular satellites of Jupiter using only their own observations on the time interval of 1996-2013 that does 1/7 part
of the full time interval of published observations.
     The authors apparently feared of the effort needed to make in order to use all the available observations of these satellites.
     Thus, the powerful potential of new observations was not realized.
It's disappointing the importance of the proposed article. This reduce the applicability of the new ephemeris.
Believing that they have the most accurate ephemeris the authors used them to predict stellar occultations by irregular satellites in the immediate future.
    Despite the above disadvantage I recommend publication of the article in MNRAS.
    Moderate revision of the manuscript is requested before it is considered for publication (see below).

AUTHORS: The main goal of the paper is to publish the occultation predictions of the irregular satellites.
This predictions demands a very good ephemeris in order to observe such an event.
But recent observations show differences from current ephemeris that are large enough to not allow a very accurate prediction.
With this in mind, we decided to create new ephemeris based only in recent observations for the near future.
The use of only the GJ15 positions was due to the trust we have on this observations for a fast numerical integration.
The creation of a complete orbit using all the observations available was not the scope of this work.
But we plan to do these work and then update our predictions.
We see that the title of the paper may confuse the reader.
So we decided to change the title to show that the ephemeris has a specific purpose.

REVIEWER: First of all I point to the incorrect concept of the authors on the ephemeris precision. In fact, no assessment the accuracy of the ephemeris is made by the authors.
The authors state "The residuals are smaller than those obtained with JPL ephemeris"
(see line 37 on the page 3 left column).
In fact, the smaller is the observation time interval the smaller residuals we get, however, the worst ephemeris accuracy is obtained.
    It is true that "the accuracy of an ephemeris decreases when we get further from the time of observations" (see lines 38, 39 on the page 3 left column).
But a longer observation time interval may be more important for the accuracy of the ephemeris than proximity to the observation time.
In general, ephemerides based on the limited set of observations carry the specific errors of this set of observations.
The ephemerides based on an extended set of observations are more precise as observational errors have more possibility of being averaged.
    For these reasons the conclusion (page 2, right column, lines 20-21)
"The procedures allow for the improvement of orbits as recent observations are added"
is not correct and must be excluded.

AUTHORS: I don't understand why it is not correct.
Even if we have worse ephemeris, once new observations are added to the database the number of observations and time-span would increase.
This, accordingly to your own statements, should improved the orbits.

REVIEWER:     The authors write in the manuscript that
"GJ15 published 6523 precise positions for 18 irregular satellites
from observations made between 1992 and 2014.".
But not all these observations were used for STE.
Please communicate in your paper that you used 3613 observations of the irregular satellites of Jupiter made on the time interaval 1996-2013 (for the most of 8 satellites) to construct your STE.
Please provide a Table disclosing the number and time interval of observations you used for each of 8 irregular satellites of Jupiter.
It is quite necessary. Otherwise, you will mislead the reader.

It may be a table as following

Satellite  N    beg    end
Himalia   1234  1995   2014
Elara      636  1996   2014
Pasiphae   609  1996   2013
Sinope     221  1996   2009
Lysithea   234  1996   2010
Carme      331  1996   2013
Ananke     250  1996   2010
Leda        98  1996   2009
 --------------
In sum    3613

AUTHORS: The correct number of observations used was modified in the text.
The number of observations by satellite and the time-span of the observations were added in the table 2.

REVIEWER: Please precise in the Abstract how many observations of 8 irregular satellites of Jupiter you used on what time intervals.

AUTHORS: This information was added to the abstract.

REVIEWER:    In Arstract the phrase
"Based on this set of positions, we performed numerical integrations ..."
is not correct. Not the numerical integrations is based on this set of positions but the fit of the initial conditions (obital parameters). Please fix it.

AUTHORS: A modification to this statement was made.

REVIEWER:    In Abstract the phrase
"This corrected systematic errors and resulted in an ephemeris more accurate/precise than JPL jup300 for these bodies at least in the immediate future"
mus be excluded. It is not justified.

AUTHORS: The sentence was removed.

REVIEWER:    It is not clear what means 'long-term effects' in the line 41 (page 3 left column).
Long-term perturbations are usually neglected when an analytical theory is applied. The authors do not
use any analytical theory but the numerical integration of the equations of motion where all necessary terms are included. Only a tidal dissipation can generate long-term effects. And being mentionned these effects must be evaluated.
Please explain what do you mean.

AUTHORS: The text was modified and the sentence was removed.

REVIEWER: It is not clear what kind of divergence is in question in the lines 40-46 (page 3 left column).
Please explain.

AUTHORS: The text was modified and the divergence was explained.
This is a divergence between the ephemeris and the real position of the satellite, which would mean an increase in the residuals, especially an increase in their mean value.

REVIEWER: The statement
"they (JPL ephemeris) are not as much precise as required for the use."
in the lines 44-45 (page 3 left column) is unjustified and must be excluded.
Furthermore, JPL ephemeris being based on the more large time interval may provide us by more precise ephemerides.

AUTHORS: Although JPL ephemeris are based on more large time interval observations, this does not mean a more accurate ephemeris.
They may have a smaller internal error due to these large number of observations within a large time-span
However, as can be seen in GJ15, recent observations observed in different sites show large offsets from the JPL ephemeris for some satellites, in some cases bigger than 200 mas
This is not enough precise for our use since we need precisions in the order of the satellites' sizes.

REVIEWER:    In any case the residuals are not an indicator of the ephemeris precision.
The O-C values include the errors of the planetary theory and systematic errors.
Small residuals may lead to incorrect positions and to incorrect ephemeris.
    As a result, I recommend that authors excluded the paragraphe in the lines 37-46 (page 3 left column).
Instead of this it must be noted in the article that a reliable method to evaluate the real precision of the irregular satellites ephemeris created from real observation is described in the paper
Emelyanov N. (2010)
Precision of the ephemerides of outer planetary satellites.
Planetary and Space Science. V. 58. P. 411-420.
Whith this method the real precision of ephemeris can by obtained for any irregular satellite at any given time moment using the natiral satellite ephemeride service MULTI-SAT
(Emel'yanov, Arlot, 2008).
These ephemerides are constructed from all available to date observations.
(Emelyanov N.V., 2005,
Ephemerides of the outer Jovian satellites.
Astronomy and Astrophysics. V. 435, p. 1173-1179.)
Last update is on 19.02.2012.
    The authors should mention these circonstances in the article and note that they set aside this method and limited to see the observation residual deviations and comparisons of their ephemeris with the results of other authors.
No conclusion that the STE are more precise can be made from these deviations and comparisons with other ephemeris.

AUTHORS: We did not exclude the paragraph cited but we update the Section 2.1 based on the comments above.

REVIEWER: Please exclude anywhere in the article conclusions about the precision of your ephemeris of the irregular satellites of Jupiter. You do not evaluate the precision correctly.

AUTHORS: The texts about precision were removed.

REVIEWER: Citation Emel’yanov & Arlot (2008) for the ephemerides in question is not correct. In reality these ephemerides were described in the paper
Emelyanov N.V. (2005) Ephemerides of the outer Jovian satellites.
Astronomy and Astrophysics. V. 435, p. 1173-1179.
The ephemerides were included in the satellite ephemerides server MULTI-SAT described in (Emel’yanov & Arlot, 2008).
    Furthermore, updates of these ephemerides were made since the publication by Emelyanov N.V. (2005). The date of last update can be read in the pages 'Sources' of the  server MULTI-SAT.
Really last update was made on 19.02.2012.

AUTHORS: The reference was changed to the correct one.

REVIEWER:   Concerning the Fig. 1 I recommend the authors to calculate and put in the Fig. 1 caption the rms of the offsets in declination of the positions  published by GJ15 for Carme for all three models considered.

AUTHORS: As recommended, the values of the root mean square of the offets of the positions relative to the 3 models was added to the caption of the figure.

REVIEWER:   Concerning the Table 3 it is clear (and it must be noted in the paper)  that the deviation of the GJ15 positions relative to the ephemeris includes the error of the calculated planet position. Therefore the planetary motion model used for the deviation calculation must be specified for each
ephemeride (STE, Jacobson et al. (2012) and Emelyanov (2005).
For correct comparison it must be the same planetary motion model.

AUTHORS: The information about the planetary ephemeris was added to the text. (See comment at the bottom of this report)

REVIEWER:    After the conclusion in the lines 58-60 (page 3 left column)
"We can see that ..." the following explanation must be added
"It is obvious because the STE were based on the same observations
with which they are compared. In contrast, two other ephemeris
were based on different sets of observations."
Because it is not surprising that the deviations of other ephemeris with GJ15 are greater.
Note for the authors: this fact does not approve better accuracy of STE.

AUTHORS: The explanation suggested was added to the text.

REVIEWER:    Note that all these comments relate only to the STE and not to the ephemeris of Phoebe being constructed from extended set of observatiobns. Really the ephemeris of Phoebe can be expected more precise.

REVIEWER:   See line 36 on the page 2 right column. You point to the jovicentric reference frame. In contrast, in the following notations (line 48 on the page 2 right column) the positions are with respect to the barycentre of Jupiter System. Please exclude this contradiction.

AUTHORS: The contradiction was corrected. (See comment at the bottom of this report)

REVIEWER:See line 40 on the page 2 right column. The statement
"The axes of the reference frame are those of the ICRS"
must be corrected as
"The axes of the reference frame are expected those of the ICRS"
because only the observations related to the ICRS may provide a link
to the ICRS.

AUTHORS: The statement was corrected.

REVIEWER:    See line 42 on the page 2 right column and following lines.
THe indices i, l and j denote the numbers of bodies and not the bodies themselves. Please fix it and explane correctly.

AUTHORS: The explanation was added to the text. (See comment at the bottom of this report)

REVIEWER: See line 55 on the page 2 right column. Phi_i is not the inclination of satellite (satellite can not have an inclination, it may be inclination of an orbit). In this case Phi_i is the latitude of the
satellte with respect to the Jupiter equator. Please fix it.

AUTHORS: The sentence was corrected.

REVIEWER: See line 58-60 on the page 2 right column. What means the number N in each case? There are confusions with the equations (1).
What is the number of irregular satellites?
What is the number of regular satellites satellites?
What is the number of other bodies?
What is the number of the Sun?
Please fix it.

AUTHORS: The  number N was explained. (See comment at the bottom of this report)

REVIEWER:The expression "the rest of the Solar System" is not convenient in this case. The rest of the Solar System includes many asteroids and satellites of other planets.
Also above in the lines 37-38 "the rest of the Solar System" must be corrected.

AUTHORS: The correct objects were described.

REVIEWER:   Please specify what planets were taken into account. Do Mercure and Pluto be included?

AUTHORS: The planets involved were specified. (See comment at the bottom of this report)

REVIEWER: The equations (1) are not correctly written. In the third line what means the index "l" (also in the expression (2))?
Where is the index "i" in the third line of the equation (1)?
In the expression (2) what is a link between the indexes "l" and "i"?
Please revise the equations (1) and (2) and exclude all confusions.

AUTHORS: The equations were revised and corrected. (See comment at the bottom of this report)

REVIEWER: Please verify the caption in the Table 2
"... with respect to the center of Jupiter".
Is it not the barycenter of the jovian system?

AUTHORS: with respect to the center of Jupiter (See comment at the bottom of this report)

REVIEWER:    Please specify planetary theory was used to calculete the coordinates of the Sun and planets (DE431, INPOP10, ???).
Reference?

AUTHORS: The planetary theory was specified. (See comment at the bottom of this report)

REVIEWER: Please specify how the coordinates of the Galilean satellites were calculated. With what model? theory? Reference?

AUTHORS: The ephemeris used for the Galilean satellies and the respective reference was added to the text. (See comment at the bottom of this report)

REVIEWER: See line 28 on the page 3 left column. Please specify the order in the RADAU integrator used and specify the constant or automatic step.

AUTHORS: The information required was added to the text.

REVIEWER: See line 35 on the page 3 right column. Please replace the footnote 1 by the citation of the paper
Arlot J.-E., Emelyanov N. V. (2009)
The NSDB natural satellites astrometric database.
Astronomy and Astrophysics. V. 503. P. 631–638.

AUTHORS: The footnote was replaced by the citation

REVIEWER: See line 47 on the page 3 right column. Please explain what means "in angular separation of Saturn"
or use other term. From what Saturn is separated?

AUTHORS: The term "in angular separation of Saturn" was replaced by "on-sky Phoebe-Saturn angular separation "

REVIEWER: See footnote 3 on the page 3 right column, also caption in Figure 3.
The expressions "precision in angular separation",
"the precision in right ascension and declination" are
not clear. How defined this "precision" is?
Really it may not be precision but deviation. Please explain or replace this term.

AUTHORS: JOSSELIN

REVIEWER: See line 54 on the page 4 right column. Please explain what is "magR" or express the idea otherwise.

AUTHORS: An explanation on what "magR" means was added to the text.

REVIEWER: See line 59 on the page 4 right column. Please explain: "CDS".

AUTHORS: The term "CDS" was explained

REVIEWER: Table 5. Notes. These notes are not clear. What is "the central instant of occultation". Please explain.
Expression "C/A:  the geocentric closest approach" is not correct.
This is a distance between ... and ... at the moment of the geocentric closest approach.
May be better to replace C/A by CAD (closest approach distance) with corresponding explanation.
Expression  "velocity in the plane of sky" is not correct.
In the plane of sky one can measure angles and not linear objects. The expression
"R*: normalized magnitude to a common shadow velocity of 20 km s^-1 ..."
is not clear at all. Please explain in the text what it is.
In the expression "east longitude of subplanet point in degrees"
please explain to what instant it relates.

AUTHORS: These notations are standard and used in previous works such as Assafin et al. (2010,2012) and Camargo et al. (2013).
So they'll be kepp as they are. But to avoid confusion, more explanations were added to the label of Table 5 and Fig. 4.

REVIEWER: Section 4. Discussion. First paragraphe. Plase replace
"Consequently, with more precise ephemeris, we predicted ..."
by
"Consequently, with our ephemeris, we predicted ..."
because the fact that the ephemeris are more precise is not justified.

AUTHORS: The phrase was replaced.

REVIEWER: Second paragraphe.
Phrase "The systematic errors found in the JPL ephemeris (Jacobson et al. 2012)
and Emel’yanov & Arlot (2008) were improved."
must be excluded as no systematic errors were shown in these ephemerides.

AUTHORS: The text was removed

REVIEWER: Section 4. Discussion. Fourth paragraphe, last phrase.
The fragment
"... the precision we have by using only GJ15 observations would be quite
equivalent or even better ..."
replace by
"... the precision we have by using only GJ15 observations may be quite
equivalent or even better ..."

AUTHORS: The sentence was replaced.

REVIEWER:See line 12 on the page 7 left column.
Please correct the expression
"Sycorax had a few observations ..."
Sycorax is not an observer nor researcher.

AUTHORS: In the previous sentence, it was said that Sycorax is a satellite.
But to avoid confusion, the sentenced was modified.

REVIEWER: In conclusion, I suggest the authors to address to the managers of the
ephemeris service MULTI-SAT (Emel’yanov & Arlot, 2008)
to include their ephemeris to the server. This will really evaluate the ephemeris and enable colleagues to profit advantage of new achievement. To do this, of course, the satellite positions must be developed in
expansions in Chebyshev polynomials.

AUTHORS: Concerning the possibility to make STE available to the public, the authors of this article are a bit reserved on the matter, and I (LB) do not think it would be a good idea.
Our ephemeris has been built to predict occultations on a rather short-term basis, with observations whose quality and astrometric reduction are best known by the people from Rio.
It is more a tool to predict observations than a solidly built ephemeris.
As you said, not using all the available observations lessens considerably the scientific interest of this product, and it is certainly useful for the rest of the community.
A complete integration and fitting of the irregular satellites is still in progress with my model, but it is a rather long process, and we needed a quick, easily adaptable ephemeris for stellar occultation.
As you probably know, occultation prediction uses many observations just before said occultations, and so we need to be able to update our ephemeris rather quickly. With only our observations, it is a matter of a few hours.
With all the available observations, it is rather a matter of days, and that is true only when we know exactly what really needs to be in the model or not (like, including all the satellites at the same time, integrating by dynamical family...).
These are the motivations for creating STE, which, as its name suggest, is specialy tailored to our use of it.
The extensive presentation of this ephemeris is more to convince the reader that our occultations predictions are sound rather than to insist on it's value for general scientific purpose.

Concerning the PH15 ephemeris, the managers of the MULTI-SAT system are being contacted to include this ephemeris to the server.

AUTHORS: I (Laurène Beauvalet) thank the referee for his attention in the reviewing and for pointing the errors that appeared while writing the tex equations.
I did most of the corrections he suggested since they were indeed errors on my part, or may have lead to confusion for the reader.
Contradictions between jovicentric or barycentric have been removed, everything is expressed with respect to the center of Jupiter
(which is why undirect interactions appear in the equation of motion since we are not in a galilean reference frame).
I also added the references to the theories used for the planets and the galileans satellites.

Concerning the use of indices, I used a rather common way of writing equations of motions, and "i" or "l" are indices for satellites of Jupiter.
This is the way I have seen most descriptions of dynamical models describe the equations of motion they use, see Lainey 2004 or Peters 1981 for example.
To try to make it clearer for the reader, I have put more details in the description of the model and suppressed the N number, which was just the number of bodies of big objects in the Solar System+galilean satellites taken into account.
Now N designates the number of irregular satellites in a dynamical family, and "l" designates one of those satellite, different form i. 
I hope that the corrections make the model clearer. If not, please do not hesitate to contact me, the author responsible for this part, Laurene Beauvalet (laurene.beauvalet@obspm.fr),
I would gladly discuss how to improve this part of the article.

### 2016-06-29 Irregular-occ/Irregular-occ.tex
.

@@
 These observations were made between \textbf{1995} and 2014 and have an estimated error of about 60 to 80 mas.
-Based on this set of positions, we performed numerical integrations for deriving new orbits for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda.
+Based on this set of positions, \textbf{we derived} new orbits for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda.
 %This corrected systematic errors and resulted in an ephemeris more accurate/precise than JPL jup300 for these bodies at least in the immediate future.
@@
 \item $J$ Jupiter
-\item $j$ another body of the Solar System, \textbf{among the 10 included (Sun+planets+Pluto+Moon)}
+\item \textbf{$j$ another body of the Solar System, among the galilean satellites, the Sun, the planets, Pluto and the Moon (14 bodies)}
 \item $M_j$ the mass of the $j$-th body, not an irregular satellite
@@
 \ddot{\vec{r_i}}= & \displaystyle -GM_J\frac{\vec{r_J}-\vec{r_i}}{r_{iJ}^3}-\sum_{l=1,l\neq i}^\mathcal{N} Gm_l\frac{\vec{r_l}-\vec{r_i}}{r_{il}^3}\\
-&\displaystyle -\sum_{j=1}^{10}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3} \right)\\
+&\displaystyle -\sum_{j=1}^{14}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3} \right)\\
  & \displaystyle +GM_J \nabla U_{\bar{i}\hat{J}} -\sum_{l=1}^\mathcal{N} Gm_l\nabla U_{\bar{l}\hat{J}}
@@
 \end{equation}
-
 where the last term in brackets and the last term in Eq.
@@
 
-All the orbits determined for the satellites show satisfying residuals.
-The residuals are smaller than those obtained with JPL ephemeris, which was expected because the accuracy of an ephemeris decreases when we get further from the time of observations.
-The main risk of divergence over time comes from the possible absence of long-term effects when fitting to a short timespan of observations.
-If that was the case, our ephemeris would diverge too quickly to be of any use.
-JPL ephemerides are fitted over all the available observations.
-As a result, they will diverge less quickly than our own.
-Though they are not as much precise as required for the use, they remain a precious reference to identify whether our own model presents a quick divergence.
+\textbf{Some methods to derive the errors of the ephemeris of irregular satellites can be found in \cite{Emelyanov2010}.
+In the Natural Satellites Ephemeride Server MULTI-SAT \citep{Emelyanov2008} the precisions can be obtained for the satellites at any given time from the \cite{Emelyanov2005} ephemeris updated to February 19, 2012.
+For our case, instead of calculate the error, we chose to compare the residuals of the \citetalias{GomesJunior2015} positions with 3 different ephemeris.}
 
@@
 \end{table*}
+
+\textbf{The residuals obtained with STE are smaller than those obtained with JPL ephemeris, which was expected because the accuracy of an ephemeris decreases when we get further from the time of observations.
+If the ephemeris is fitted over too-short a timespan of observations, the initial conditions (positions, velocities, masses) used in the model will not be good enough and the model will differ greatly from the observed position of the satellite when we will accumulate new observations.
+The ephemeris is then said to be diverging.
+%The main risk of divergence over time comes from the possible absence of long-term effects when fitting to a short timespan of observations.
+If that was the case, our ephemeris would diverge too quickly to be of any use.
+JPL ephemerides are fitted over all the available observations.
+As a result, they will diverge less quickly than our own.
+Though they are not as much precise as required for the use, they remain a precious reference to identify whether our own model presents a quick divergence.}
 
@@
 \hline
-d m Year~~~~h~~m~~~s & RA~~~(ICRS)~~~Dec & C/A & P/A & $\nu$ & {\it D} & $R^*$ & $\lambda$ & LST & $\mu_{\alpha*}$ & $\mu_{\delta}$ \tabularnewline
+d m Year~~~~h~~m~~~s & RA~~~(ICRS)~~~Dec & C/A & P/A & $\nu$ &{\it D} & $R^*$ & $\lambda$ & LST & $\mu_{\alpha*}$ & $\mu_{\delta}$ \tabularnewline
 \hline
@@
 \textbf{Notes}.
-Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - at the central instant of occultation (corrected by proper motions); C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees (zero at north of the star, increasing clockwise); $\nu$: velocity in the plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; {\it D}: planet range to Earth, in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
+Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - \textbf{at the instant of the occultation} (corrected by proper motions); C/A: textbf{apparent geocentric distance between the satellite and the star (a.k.a.
+the distance between the shadow and the center of the Earth) at the moment of the geocentric closest approach}, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees (zero at north of the star, increasing clockwise); $\nu$: \textbf{relative velocity of event in km s$^{-1}$}: positive = prograde, negative = retrograde; {\it D}: \textbf{Geocentric distance to the occulting object in AU}; $R^*$: normalized \textbf{UCAC4 magnitude in the R-band} to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$, \textbf{the value 20 km s$^{-1}$ is typical of events around the opposition}; $\lambda$: east longitude of subplanet point in degrees, positive towards east, \textbf{at the instant of the geocentric closest approach}; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
 \end{flushleft}
@@
 \textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for\textbf{ a observation in the center of the shadow}; C/A: \textbf{apparent geocentric distance between the satellite and the star (a.k.a.
-the distance between the shadow and the center of the Earth) at the moment of the geocentric closest approach}, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; vel: \textbf{relative} velocity of event in km/s; Delta: Geocentric distance to the occulting object in AU; $R^*$: normalized \textbf{UCAC4 magnitude in the R-band} to a common shadow velocity of 20 km s$^{-1}$; long: east longitude of subplanet point in degrees, positive towards east, \textbf{at the instant of the geocentric closest approach}.}
+the distance between the shadow and the center of the Earth) at the moment of the geocentric closest approach}, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; $\nu$: \textbf{relative} velocity of event in km/s; {\it D}: Geocentric distance to the occulting object in AU; $R^*$: normalized \textbf{UCAC4 magnitude in the R-band} to a common shadow velocity of 20 km s$^{-1}$; $\lambda$: east longitude of subplanet point in degrees, positive towards east, \textbf{at the instant of the geocentric closest approach}.}
 \label{Fig: ocultacao}
@@
 In the context of the test, for a 0 mas offset in C/A we would have 100\% probability of observing the occultation, and 0\% in the case of a C/A offset equal to or larger than 20 mas, the radius of Himalia.
-From Table 5, we have nearly 0\% probability of success in situation \ref{item3}, for which the offset in C/A was -20 mas, but when the relative astrometry was poor, 10 days prior to the event.
+From Table \ref{Tab: comparison-Himalia}, we have nearly 0\% probability of success in situation \ref{item3}, for which the offset in C/A was -20 mas, but when the relative astrometry was poor, 10 days prior to the event.
 Once at the day of the event in situation \ref{item4}, the C/A offset dropped to -9 mas only, corresponding to a 55\% probability of success.
@@
 For the irregular satellites of Jupiter (Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope), we produced ephemeris using only the observations of \citetalias{GomesJunior2015}.
-\textcolor{red}{The systematic errors found in the JPL ephemeris \citep{Jacobson2012} and \cite{Emelyanov2005} were improved.} These new ephemerides are denominated Special-Tailored Ephemeris.
+These new ephemerides are denominated Special-Tailored Ephemeris.
 
@@
 The numerical model of the satellites of Jupiter was developped during a post-doctoral contract funded by the Chinese Academy of Sciences (CAS) and supported by the National Scientific Fund of China (NSFC).
+\textbf{The authors thank Dr.
+Nikolay Emelyanov for his comments to improve the paper.}
 

### 2016-07-01 Irregular-occ/Irregular-occ.tex
.

@@
 We computed the precision of the PH15 ephemeris considering three sets of observations: all the positions available; only the positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.
+\textbf{The precision is computed by propagation of the covariance matrix of the orbit determination process and by linear transformations giving the covariance matrix in spherical coordinates (right ascension and declination) at a specific date \citep[for more details, see][]{Desmars2013a}.
+This last matrix then provides the standard deviation in right ascension $\sigma_\alpha$ and declination  $\sigma_\delta$ at the required date and finally, $\sigma_s=\sqrt{\sigma^2_\alpha\cos^2\delta+\sigma^2_\delta}$ is considered as an estimation of the precision in angular separation on the celestial sphere (or the sky plane uncertainty).}
+
 In Fig.
-\ref{Fig:err-Phoebe} we show the comparison between them for the time span 2016-2021 in on-sky Phoebe-Saturn angular separation\footnote{\label{ft-angsep}The precision in angular separation is computed  by $\sigma_s = \sqrt{\sigma^2_{\alpha} \cos^2 \delta + \sigma^2_{\delta}}$ where $\sigma_{\alpha}$ and $\sigma_{\delta}$ are the precision in right ascension and declination.}.
+\ref{Fig:err-Phoebe} we show the comparison between them for the time span 2016-2021 in \textbf{on-sky Phoebe-Saturn angular separation}.
 It is possible to see that even considering only the positions of \citetalias{GomesJunior2015}, the estimated error of the ephemeris is smaller than 12 mas.

### 2016-07-04 Irregular-occ/referee_response_2016-07-04.txt
.

--- 2016-06-29 Irregular-occ/referee_response.txt
@@
 
-AUTHORS: The main goal of the paper is to publish the occultation predictions of the irregular satellites.
-This predictions demands a very good ephemeris in order to observe such an event.
-But recent observations show differences from current ephemeris that are large enough to not allow a very accurate prediction.
-With this in mind, we decided to create new ephemeris based only in recent observations for the near future.
-The use of only the GJ15 positions was due to the trust we have on this observations for a fast numerical integration.
-The creation of a complete orbit using all the observations available was not the scope of this work.
-But we plan to do these work and then update our predictions.
-We see that the title of the paper may confuse the reader.
-So we decided to change the title to show that the ephemeris has a specific purpose.
+AUTHORS: Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018. This improves a lot the chances of observing
+stellar occultations by their irregular satellites in the near future, allowing for precise measurements of their poorly known 
+physical parameters (sizes, albedo), values mostly needed for better understanding their physical and dynamical origin and 
+evolution. In the GJ15 paper, thousands of certified, precise positions were obtained from almost 20 years of systematic 
+observations of these satellites. They were made with very distinct instruments at distant sites located at both Earth hemispheres 
+(good parallax angle coverage). This makes this set of positions at the same time versatile and trustfull. Because the positions 
+were obtained by the same authors of this work, we know them better, a clear advantage in the computation of orbits. The 
+investigation in the GJ15 paper showed that the ephemeris published up to date present offsets with respect to the GJ15 positions 
+much larger than explainable by errors in the astrometry of the GJ15 positions. Above all, these ephemeris offsets are not 
+acceptable for suitable stellar occultation predictions. Having this in mind, we generated a more trustfull ephemeris for the 
+immediate future, optimally designed for improving the stellar occultation predictions for the next few years. For that, we 
+profited from the use of this large set of GJ15 positions alone, which as said is well known by the authors and contains recent 
+but also past observations well distributed in time and in Earth location. It is our point of view that this objective was 
+successfully accomplished, as can be seen for instance in the ephemeris comparisons in Table 3 and Figs. 2 and 3. Efforts for the 
+determination of the orbits by using all the available observations are under way, but out of the scope of this work (see more 
+comments on this subject at the end of this report). In order to enphasize for the reader, from the begining, the pragmatic aspect 
+of this work, we changed the title a bit, which now reads: "New orbits of irregular satellites designed for the predictions of 
+stellar occultations up to 2020, based on thousands of new observations".
 
@@
 
-AUTHORS: I don't understand why it is not correct.
-Even if we have worse ephemeris, once new observations are added to the database the number of observations and time-span would increase.
-This, accordingly to your own statements, should improved the orbits.
+AUTHORS: We basically agree with the comments made by the referee here, but it seems that he misinterpreted the sentence quoted 
+above. By no means the sentence aims to give a sort of conclusion to the issues exposed above by the referee. It rather and simply 
+aimed at enphasizing one usefull technical aspect of the orbit fitting procedure, which is easily allowing for the continuous addition of more observations (old, new), aiming at refining the orbit fit.
+In order to make the sentence more clear in this respect, we rewrote it.
 
@@
 
-AUTHORS: The correct number of observations used was modified in the text.
-The number of observations by satellite and the time-span of the observations were added in the table 2.
+AUTHORS: The number and time span of the observations for each individual satellite are now given in Table 2.
 
@@
 
-AUTHORS: The text was modified and the sentence was removed.
+AUTHORS: The sentence is indeed meaningless and was removed.
 
@@
 
-AUTHORS: The text was modified and the divergence was explained.
-This is a divergence between the ephemeris and the real position of the satellite, which would mean an increase in the residuals, especially an increase in their mean value.
+AUTHORS: The text was modified for a better explanation.
+We refer to the maximum difference between the compared ephemerides of the satellites, which could or not indicate a significant discrepancy, having in mind the observed position errors of 60 mas - 80 mas stated in GJ15.
 
-REVIEWER: The statement
-"they (JPL ephemeris) are not as much precise as required for the use."
+REVIEWER: The statement "they (JPL ephemeris) are not as much precise as required for the use."
 in the lines 44-45 (page 3 left column) is unjustified and must be excluded.
@@
 
-AUTHORS: Although JPL ephemeris are based on more large time interval observations, this does not mean a more accurate ephemeris.
-They may have a smaller internal error due to these large number of observations within a large time-span
-However, as can be seen in GJ15, recent observations observed in different sites show large offsets from the JPL ephemeris for some satellites, in some cases bigger than 200 mas
-This is not enough precise for our use since we need precisions in the order of the satellites' sizes.
+AUTHORS: This part of the text was modified for better clarification of the error issues regarding to our ephemeris.
 
@@
 
-AUTHORS: We did not exclude the paragraph cited but we update the Section 2.1 based on the comments above.
+AUTHORS: This part of the text was modified for better clarification of the error issues regarding to our ephemeris, taking
+into consideration the comments above by the referee.
 
@@
 
-AUTHORS: The texts about precision were removed.
+AUTHORS: The conclusions about precision throughout the text were modified for better clarification of the error issues regarding to our ephemeris, following the guidelines pointed out by the referee.
 
@@
 
-AUTHORS: The explanation suggested was added to the text.
+AUTHORS: We agree in good part with the referee. Of course, by construction, we do expect smaller offsets in the comparison of
+GJ15 positions with STE. However, it is not obvious for us that these offsets should be that smaller in comparison with the other
+ephemeris offsets (see Table 3). Notice that the GJ15 positions come from observations made with very distinct instruments at
+distant sites located at both Earth hemispheres (good parallax angle coverage), making this set of positions not remarkably 
+distinct than any set of positions that were used in the construction of the JPL and Emelyanov & Arlot ephemeris. Thus, from our 
+point of view, these ephemeris offsets comparisons do suggest that the accuracy of the STE ephemeris is at least slightly better
+than that of the other ephemeris, at least for the time span of our satellites´ observations. As a consequence, this supports the
+STE as the best ephemeris for the next few years, making it the best choice to use in stellar occultation predictions in the short
+future. In the text, we added this comment, assimilating the sentence suggested by the referee.
 
@@
 
-AUTHORS: The contradiction was corrected. (See comment at the bottom of this report)
+AUTHORS: The contradiction was corrected. (See also the comment at the bottom of this report)
 
@@
 
-AUTHORS: The  number N was explained. (See comment at the bottom of this report)
+AUTHORS: The number N is now explained. (See comment at the bottom of this report)
 
@@
 
-AUTHORS: The correct objects were described.
+AUTHORS: The used objects are now clearly cited.
 
@@
 
-AUTHORS: with respect to the center of Jupiter (See comment at the bottom of this report)
+AUTHORS: It is with respect to the center of Jupiter (See comment at the bottom of this report)
 
@@
 
-AUTHORS: The planetary theory was specified. (See comment at the bottom of this report)
+AUTHORS: The planetary theory is now specified. (See comment at the bottom of this report)
 
@@
 
-AUTHORS: The term "in angular separation of Saturn" was replaced by "on-sky Phoebe-Saturn angular separation "
+AUTHORS: An explanation was added to the text.
 
@@
 
-AUTHORS: JOSSELIN
+AUTHORS: An explanation was added to the text
 
@@
 
-AUTHORS: An explanation on what "magR" means was added to the text.
+AUTHORS: It means R (red band) magnitude; it is now explicitly written in the text.
 
@@
 
-AUTHORS: These notations are standard and used in previous works such as Assafin et al. (2010,2012) and Camargo et al. (2013).
-So they'll be kepp as they are. But to avoid confusion, more explanations were added to the label of Table 5 and Fig. 4.
+AUTHORS: These notations are standard in stellar occultation work and have already been extensively used in many previous works
+such as Assafin et al. (2010, 2012) and  Camargo et al. (2013), and references therein, so they'll be kept as they are. But to let 
+it easier for understanding by a broader community of readers, some more explanations were added to the label of Table 5 and
+Fig.4, and a reference for these definitions is given (Assafin et al., 2010).
 
@@
 AUTHORS: Concerning the possibility to make STE available to the public, the authors of this article are a bit reserved on the matter, and I (LB) do not think it would be a good idea.
-Our ephemeris has been built to predict occultations on a rather short-term basis, with observations whose quality and astrometric reduction are best known by the people from Rio.
+Our ephemeris has been built to predict occultations on a rather short time span basis, with observations whose quality and astrometric reduction are best known by the people from Rio.
 It is more a tool to predict observations than a solidly built ephemeris.
-As you said, not using all the available observations lessens considerably the scientific interest of this product, and it is certainly useful for the rest of the community.
-A complete integration and fitting of the irregular satellites is still in progress with my model, but it is a rather long process, and we needed a quick, easily adaptable ephemeris for stellar occultation.
+As you know, we did not use all the available observations. A complete integration and fitting of the irregular satellites is
+still in progress with my model, but it is a rather long process, and we needed a quick, easily adaptable ephemeris for stellar occultation predictions.
 As you probably know, occultation prediction uses many observations just before said occultations, and so we need to be able to update our ephemeris rather quickly.
 With only our observations, it is a matter of a few hours.
-With all the available observations, it is rather a matter of days, and that is true only when we know exactly what really needs to be in the model or not (like, including all the satellites at the same time, integrating by dynamical family...).
+With all the available observations, it is rather a matter of days, and that is true only when we know exactly what really needs to be in the model or not (like including all the satellites at the same time, integrating by dynamical family...), not to
+mention weighting schemes, etc.
 These are the motivations for creating STE, which, as its name suggest, is specialy tailored to our use of it.
@@
 
-AUTHORS: I (Laurène Beauvalet) thank the referee for his attention in the reviewing and for pointing the errors that appeared while writing the tex equations.
-I did most of the corrections he suggested since they were indeed errors on my part, or may have lead to confusion for the reader.
+AUTHORS: I (LB) thank the referee for his attention in the reviewing and for pointing the errors that appeared while writing the tex equations.
+I did most of the corrections he suggested since they were indeed errors on my part, or might have lead to confusion for the reader.
 Contradictions between jovicentric or barycentric have been removed, everything is expressed with respect to the center of Jupiter
@@
 This is the way I have seen most descriptions of dynamical models describe the equations of motion they use, see Lainey 2004 or Peters 1981 for example.
-To try to make it clearer for the reader, I have put more details in the description of the model and suppressed the N number, which was just the number of bodies of big objects in the Solar System+galilean satellites taken into account.
-Now N designates the number of irregular satellites in a dynamical family, and "l" designates one of those satellite, different form i. 
+In order to make it clearer for the reader, I have put more details in the description of the model and suppressed the N number, which was just the number of bodies of big objects in the Solar System+galilean satellites taken into account.
+Now N designates the number of irregular satellites in a dynamical family, and "l" designates one of those satellite, different from i. 
 I hope that the corrections make the model clearer. If not, please do not hesitate to contact me, the author responsible for this part, Laurene Beauvalet (laurene.beauvalet@obspm.fr),

### 2016-07-04 Irregular-occ/Irregular-occ.tex
.

@@
 \usepackage[T1]{fontenc}
-\usepackage{times}  %%% neede to embed font 1 text
+\usepackage{times}  %%% needed to embed font 1 text
 \usepackage{prettyref}
@@
 \usepackage{color}
-\usepackage[breaklinks=true]{hyperref}  
+\usepackage[breaklinks=true, draft]{hyperref}  % draft because there is an error = 'pdfTeX error (ext4): \pdfendlink ended up in different nesting level than \pdfstartlink'.
 \usepackage{arydshln} % creates horizontal dashed lines in tables
@@
 
-\title[Orbits and predictions for occultations up to 2020]{New short-term orbits of irregular satellites in order to predict stellar occultations up to 2020}
+\title[Orbits and predictions for occultations up to 2020]{New orbits of irregular satellites designed for the predictions of stellar occultations up to 2020, based on thousands of new observations}
 %\title[Orbits and predictions for occultations up to 2020]{New orbits of irregular satellites based on thousands of new observations and predictions for stellar occultations up to 2020}
@@
 $^{5}$Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil\\
-$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
+$^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil\\
+$^\dag$Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France
 }
@@
 Due to the context of this work regarding to stellar occultations, the orbit fitting procedures used aimed primarily to derive precise ephemerides for the near future.
-The procedures allow for the improvement of orbits as recent observations are added.
+Technically, the procedures easily allow for the continuous addition of more observations (old, new) aiming at refining the orbit fits.
 
@@
 The satellites are submitted to the influence of the Sun and the \textbf{main bodies of the solar system (from Mercury to Pluto, plus the Moon)}, as well as those of the Galilean satellites and the first harmonics of Jupiter's gravity field.
-The axis of the reference frame are \textbf{expected to be those of the equatorial reference frame J2000}.
+The axes of the reference frame are \textbf{expected to be those of the ICRS}.
 
 We use the following notations: \begin{itemize}
-\item \textbf{in one dynamical family consisting of $\mathcal{N}$ irregular satellites, $i$ will stand for the one whose equation of motion we are considering, $l$ will stand for another irregular satellite in gravitationnal interaction belonging to the same family}
+\item \textbf{in one dynamical family consisting of $\mathcal{N}$ irregular satellites, $i$ will stand for the one whose equation of motion we are considering, $l$ will stand for another irregular satellite in gravitational interaction belonging to the same family}
 \item $J$ Jupiter
-\item \textbf{$j$ another body of the Solar System, among the galilean satellites, the Sun, the planets, Pluto and the Moon (14 bodies)}
+\item \textbf{$j$ another body of the Solar System, among the Galilean satellites, the Sun, the planets, Pluto and the Moon (14 bodies)}
 \item $M_j$ the mass of the $j$-th body, not an irregular satellite
@@
 
-The obtained ephemeris is hereafter refered as STE, for Special-Tailored Ephemeris.
-The initial osculating elements at the origin of integration, \textbf{the number of observations used and the time-span where these observations are distributed} are presented in Table \ref{Tab: sat_ell}.
+The obtained ephemeris is hereafter referred to as STE, for Special-Tailored Ephemeris.
+The initial osculating elements at the origin of integration, \textbf{the number and time span of the observations of each satellite} are presented in Table \ref{Tab: sat_ell}.
 
@@
 In the Natural Satellites Ephemeride Server MULTI-SAT \citep{Emelyanov2008} the precisions can be obtained for the satellites at any given time from the \cite{Emelyanov2005} ephemeris updated to February 19, 2012.
-For our case, instead of calculate the error, we chose to compare the residuals of the \citetalias{GomesJunior2015} positions with 3 different ephemeris.}
+However, since the practical realization of the STE ephemeris is for help improving the prediction of stellar occultations by the irregular satellites in the immediate future, it is interesting to compare the STE with the other relevant ephemerides for the next few years.}
 
 We compared the STE ephemeris to the JPL for all the Jupiter satellites we fitted, until 2021.
-For instance, the divergence between 2015 and 2021 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.
+For instance, the \textbf{maximum difference} between 2015 and 2021 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.
 
@@
 \caption{Offsets in declination of the positions published by \protect\citetalias{GomesJunior2015} for Carme.
-The red "x" relative to the special-tailored ephemeris \textbf{(rms = 51)}, the blue "+" relative to the jup300 JPL ephemeris \textbf{(rms = 130)} and the green dot relative to \protect\cite{Emelyanov2005} \textbf{(rms = 92)}.
-As expected, the ephemeris systematic errors pointed out by \protect\citetalias{GomesJunior2015} are reduced with the STE ephemeris.
+The red "x" relate to the special-tailored ephemeris \textbf{(rms = 51)}, the blue "+" to the jup300 JPL ephemeris \textbf{(rms = 130)} and the green dot to \protect\cite{Emelyanov2005} \textbf{(rms = 92)}.
+As expected, the ephemeris \textbf{offsets} pointed out by \protect\citetalias{GomesJunior2015} are reduced with the STE ephemeris.
 \label{Fig: JPL-STE}}
@@
 
-In Table \ref{Tab:STE-error} it is presented the mean offsets and the respective standard deviation of the \citetalias{GomesJunior2015} positions relative to the same 3 ephemeris as above.
+In Table \ref{Tab:STE-error} we present the mean offsets and the respective standard deviation of the \citetalias{GomesJunior2015} positions relative to the same 3 ephemeris as above.
 We can see that the mean offsets as well as most of their standard deviations are greatly reduced with the STE ephemeris.
-\textbf{It is obvious because the STE were based on the same observations with which they are compared.
-In contrast, two other ephemeris
-were based on different sets of observations.}
+\textbf{Of course, by construction, we should expect smaller offsets in the comparison of \citetalias{GomesJunior2015} positions with STE.
+However, it is not obvious that these offsets should be that smaller in comparison with the other ephemeris offsets.
+Notice that the \citetalias{GomesJunior2015} positions come from observations made with very distinct instruments at distant sites located at both Earth hemispheres (good parallax angle coverage), making this set of positions not remarkably distinct than any set of positions that were used in the construction of the other two ephemeris.
+Thus, these ephemeris offsets comparisons suggest that the accuracy of the STE ephemeris is at least slightly better than that of the other ephemeris, at least for the time span of our satellites' observations.
+This supports the utility of the STE ephemeris for the next few years, making it one of the best choices to use in stellar occultation predictions in the short future for these satellites.}
 
@@
 
-\textbf{The residuals obtained with STE are smaller than those obtained with JPL ephemeris, which was expected because the accuracy of an ephemeris decreases when we get further from the time of observations.
+%\textbf{The residuals obtained with STE are smaller than those obtained with JPL ephemeris, which was expected because the accuracy of an ephemeris decreases when we get further from the time of observations.
 If the ephemeris is fitted over too-short a timespan of observations, the initial conditions (positions, velocities, masses) used in the model will not be good enough and the model will differ greatly from the observed position of the satellite when we will accumulate new observations.
@@
 If that was the case, our ephemeris would diverge too quickly to be of any use.
-JPL ephemerides are fitted over all the available observations.
+%JPL ephemerides are fitted over all the available observations.
 As a result, they will diverge less quickly than our own.
@@
 \textbf{The precision is computed by propagation of the covariance matrix of the orbit determination process and by linear transformations giving the covariance matrix in spherical coordinates (right ascension and declination) at a specific date \citep[for more details, see][]{Desmars2013a}.
-This last matrix then provides the standard deviation in right ascension $\sigma_\alpha$ and declination  $\sigma_\delta$ at the required date and finally, $\sigma_s=\sqrt{\sigma^2_\alpha\cos^2\delta+\sigma^2_\delta}$ is considered as an estimation of the precision in angular separation on the celestial sphere (or the sky plane uncertainty).}
+This last matrix then provides the standard deviation in right ascension $\sigma_\alpha$ and declination  $\sigma_\delta$ at the required date, with $\sigma_s=\sqrt{\sigma^2_\alpha\cos^2\delta+\sigma^2_\delta}$ being the total error in the celestial sphere.}
 
@@
 \includegraphics[width=8.8cm]{figures/PH15_err_angsep.eps}  
-\caption{Comparison of the precision in on-sky Phoebe-Saturn angular separation\textsuperscript{\ref{ft-angsep}} for the PH15 ephemeris where three different sets of positions were used to compute the ephemeris: all observations; only positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.}
+\caption{Comparison of the precision in on-sky Phoebe-Saturn angular separation for the PH15 ephemeris where three different sets of positions were used to compute the ephemeris: all observations; only positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.}
 \label{Fig:err-Phoebe}
@@
 This is because at that periods these satellites will cross the apparent galactic plane.
-We call attention that about 10\% of the events will involve stars brighter than magR=14  (and almost 25\% brighter than magR=15), \textbf{where magR stands for the UCAC4 magnitude in the R-band}, which helps the attempt of amateur observers.
+We call attention that about 10\% of the events will involve stars brighter than \textbf{magnitude R=14}  (and almost 25\% brighter than R=15), which helps the attempt of amateur observers.
 
@@
 This is an occultation by Elara that will happen in February 21, 2017.
-This event can be observed from Australia and it is one of the best opportunities for this object due to the slow velocity of the event and it involves a bright star (MagR*=12.4).
+This event can be observed from Australia and it is one of the best opportunities for this object due to the slow velocity of the event and it involves a bright star (R*=12.4).
 
@@
 \textbf{Notes}.
-Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - \textbf{at the instant of the occultation} (corrected by proper motions); C/A: textbf{apparent geocentric distance between the satellite and the star (a.k.a.
-the distance between the shadow and the center of the Earth) at the moment of the geocentric closest approach}, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees (zero at north of the star, increasing clockwise); $\nu$: \textbf{relative velocity of event in km s$^{-1}$}: positive = prograde, negative = retrograde; {\it D}: \textbf{Geocentric distance to the occulting object in AU}; $R^*$: normalized \textbf{UCAC4 magnitude in the R-band} to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$, \textbf{the value 20 km s$^{-1}$ is typical of events around the opposition}; $\lambda$: east longitude of subplanet point in degrees, positive towards east, \textbf{at the instant of the geocentric closest approach}; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
+Entries included: day of the year and UTC \textbf{central instant} of the prediction; right ascension and declination of the occulted star - \textbf{at the central instant of the occultation} (corrected by proper motions); C/A: \textbf{apparent geocentric distance between the satellite and the star (a.k.a.
+the distance between the shadow and the center of the Earth) at the moment of the geocentric closest approach}, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees (zero at north of the star, increasing clockwise); $\nu$: \textbf{relative velocity of event in km s$^{-1}$}: positive = prograde, negative = retrograde; {\it D}: \textbf{Geocentric distance to the occulting object in AU}; $R^*$: normalized \textbf{UCAC4 magnitude in the R-band} to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{UCAC4}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$, \textbf{the value 20 km s$^{-1}$ is typical of events around the opposition}; $\lambda$: east longitude of subplanet point in degrees, positive towards east, \textbf{at the instant of the geocentric closest approach}; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
+\textbf{For more detailed information about the definition and use of these stellar occultation geometric elements see \cite{Assafin2010}}.
 \end{flushleft}
@@
 The shadow moves from right to left.
-\textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for\textbf{ a observation in the center of the shadow}; C/A: \textbf{apparent geocentric distance between the satellite and the star (a.k.a.
-the distance between the shadow and the center of the Earth) at the moment of the geocentric closest approach}, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; $\nu$: \textbf{relative} velocity of event in km/s; {\it D}: Geocentric distance to the occulting object in AU; $R^*$: normalized \textbf{UCAC4 magnitude in the R-band} to a common shadow velocity of 20 km s$^{-1}$; $\lambda$: east longitude of subplanet point in degrees, positive towards east, \textbf{at the instant of the geocentric closest approach}.}
+\textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for\textbf{ an observation in the center of the shadow}; C/A: \textbf{apparent geocentric distance between the satellite and the star (a.k.a.
+the apparent distance in the plane of the sky between the shadow and the center of the Earth) at the moment of the geocentric closest approach}, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; $\nu$: \textbf{relative} velocity of event in km/s; {\it D}: Geocentric distance to the occulting object in AU; $R^*$: normalized \textbf{UCAC4 magnitude in the R-band} to a common shadow velocity of 20 km s$^{-1}$; $\lambda$: east longitude of subplanet point in degrees, positive towards east, \textbf{at the central instant of the geocentric closest approach (see the notes of Table 5)}.}
 \label{Fig: ocultacao}
@@
 
-We performed new numerical integration for improving the orbits of some of the larger irregular satellites.
+We performed new numerical integrations for improving the orbits of some of the larger irregular satellites.
 Consequently, with \textbf{our} ephemeris, we predicted stellar occultations aiming to access fundamental parameters like size, shape, albedo, ultimately aiming to track the formation origin of these bodies.

### 2016-07-04 Irregular-occ/Irregular-occ.tex
.

@@
 \usepackage{color}
-\usepackage[breaklinks=true, draft]{hyperref}  % draft because there is an error = 'pdfTeX error (ext4): \pdfendlink ended up in different nesting level than \pdfstartlink'.
+\usepackage[breaklinks=true]{hyperref}  % draft because there is an error = 'pdfTeX error (ext4): \pdfendlink ended up in different nesting level than \pdfstartlink'.
 \usepackage{arydshln} % creates horizontal dashed lines in tables
@@
 $^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
-CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
+CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\,
 $^{2}$Observat\'orio Nacional/MCTI, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
 $^{3}$Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France\\
-$^{4}$Institut de m\'ecanique c\'eleste et de calcul des \'eph\'em\'erides - Observatoire de Paris, UMR 8028 du CNRS, 77 Av. Denfert-Rochereau,\\ 75014 Paris, France\\
+$^{4}$IMCCE, Observatoire de Paris, PSL Research University, CNRS, Sorbonne Universités, UPMC Univ Paris 06, Univ. Lille, 77 Av. Denfert-Rochereau, 75014 Paris, France\\
 $^{5}$Laborat\'orio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil\\
@@
 MA thanks the CNPq (Grants 473002/2013-2 and 308721/2011-0) and FAPERJ (Grant E-26/111.488/2013).
-RV-M thanks grants: CNPq-306885/2013, Capes/Cofecub-2506/2015, Faperj/PAPDRJ-45/2013.
+RV-M thanks grants: CNPq-306885/2013, Capes/Cofecub-2506/2015, Faperj: PAPDRJ-45/2013 and Cient.Est-05/2015.
 JIBC acknowledges CNPq for a PQ2 fellowship (process number 308489/2013-6).
@@
 The numerical model of the satellites of Jupiter was developped during a post-doctoral contract funded by the Chinese Academy of Sciences (CAS) and supported by the National Scientific Fund of China (NSFC).
-\textbf{The authors thank Dr.
-Nikolay Emelyanov for his comments to improve the paper.}
+\textbf{The authors thank Dr. Nikolay Emelyanov for his comments to improve the paper.}
 

### 2016-07-08 Irregular-occ/mnras-source-revised/Irregular-occ.tex
.

--- 2016-07-04 Irregular-occ/Irregular-occ.tex
@@
 \begin{abstract}
-%Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system.
-It is important to know their physical parameters, such as size, shape, albedo and composition, to trace back their true origin.
-The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects but this demands a very precise ephemeris based on new and good positions.
-%\cite{GomesJunior2015} published 6523 observations obtained between 1992 and 2014 for 18 irregular satellites with an estimated error in the positions of about 60 to 80 mas.
-%Based on this observations, we aimed to generate new numerical integration of the orbits of eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda.
-This is a short-time ephemeris that corrected systematic errors from previous ephemeris (JPL JUP300).
-For Phoebe we updated ephemeris from \cite{Desmars2013} using 75\% more positions than the previous one.
-%Also Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018.
-This creates a good opportunity to predict stellar occultations due to the large number of events that can be predicted in these periods.
-Thus, we aimed to predict stellar occultations for the same 9 satellites cited above.
-Using the UCAC4 catalogue we managed to identify \noccs candidates of stellar occultations between the period of January, 2016 and December, 2020.
-%We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
-\cite{GomesJunior2015} published \textbf{3613 positions for the 8 largest irregular satellites of Jupiter and 1787 positions for the largest irregular satellite of Saturn, Phoebe}.
-These observations were made between \textbf{1995} and 2014 and have an estimated error of about 60 to 80 mas.
-Based on this set of positions, \textbf{we derived} new orbits for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda.
+\cite{GomesJunior2015} published 3613 positions for the 8 largest irregular satellites of Jupiter and 1787 positions for the largest irregular satellite of Saturn, Phoebe.
+These observations were made between 1995 and 2014 and have an estimated error of about 60 to 80 mas.
+Based on this set of positions, we derived new orbits for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda.
 %This corrected systematic errors and resulted in an ephemeris more accurate/precise than JPL jup300 for these bodies at least in the immediate future.
@@
 
-We present in this paper new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) using only the positions obtained by \citetalias{GomesJunior2015}%.
-We aim to eliminate the systematic errors in the JPL ephemeris
- (see Sec. \ref{Sec: integration}). %For an object at the distance of Jupiter, this represents an error langer than 700 km in the shadow path of an occultation.
+We present in this paper new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) using only the positions obtained by \citetalias{GomesJunior2015} (see Sec.
+\ref{Sec: integration}).
 For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Center and observations from Flagstaff.
@@
 Phoebe & 32 & $212 \pm 1.4^{b}$ & 3 \tabularnewline
-%\hdashline
-%Nereid & 15 & $340 \pm 50^{c}$ & 4 \tabularnewline
-%Triton & 124 & $2707 \pm 2.0^{c}$ & 5 \tabularnewline
 \hline
@@
 
-%Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
-In contrast to TNOs, the irregular satellites have much better ephemeris because the orbits of their host planets are better  known, their observational time span is much wider and covers many orbital periods.
-Moreover, the irregular satellites are much closer to Earth which implies in a much smaller shadow path error in kilometers.
-These advantages may be somewhat balanced by the smaller sizes estimated for the irregular satellites.Thus, in comparison, the chances for a successful observation of an stellar occultation by an irregular satellite should be considered at least also as good as those by TNOs.
-
 In section \ref{Sec: integration}, we present the new determination of the orbits.
@@
 
-\citetalias{GomesJunior2015} published \textbf{3613} precise positions for \textbf{the 8 largest irregular satellites of Jupiter} from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between \textbf{1995} and 2014.
+\citetalias{GomesJunior2015} published 3613 precise positions for the 8 largest irregular satellites of Jupiter from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1995 and 2014.
 
@@
 Our numerical model describes the dynamical evolution of the irregular satellites of Jupiter in a jovicentric reference frame.
-The satellites are submitted to the influence of the Sun and the \textbf{main bodies of the solar system (from Mercury to Pluto, plus the Moon)}, as well as those of the Galilean satellites and the first harmonics of Jupiter's gravity field.
-The axes of the reference frame are \textbf{expected to be those of the ICRS}.
+The satellites are submitted to the influence of the Sun and the main bodies of the solar system (from Mercury to Pluto, plus the Moon), as well as those of the Galilean satellites and the first harmonics of Jupiter's gravity field.
+The axes of the reference frame are expected to be those of the ICRS.
 
 We use the following notations: \begin{itemize}
-\item \textbf{in one dynamical family consisting of $\mathcal{N}$ irregular satellites, $i$ will stand for the one whose equation of motion we are considering, $l$ will stand for another irregular satellite in gravitational interaction belonging to the same family}
+\item in one dynamical family consisting of $\mathcal{N}$ irregular satellites, $i$ will stand for the one whose equation of motion we are considering, $l$ will stand for another irregular satellite in gravitational interaction belonging to the same family
 \item $J$ Jupiter
-\item \textbf{$j$ another body of the Solar System, among the Galilean satellites, the Sun, the planets, Pluto and the Moon (14 bodies)}
+\item $j$ another body of the Solar System, among the Galilean satellites, the Sun, the planets, Pluto and the Moon (14 bodies)
 \item $M_j$ the mass of the $j$-th body, not an irregular satellite
 \item $m_i$ the mass of the irregular satellite $i$
-\item $\vec{r_i}$ the position of the $i$-th body with respect \textbf{to the center of Jupiter}
+\item $\vec{r_i}$ the position of the $i$-th body with respect to the center of Jupiter
 \item $r_{ij}$ the distance between bodies $i$ and $j$
@@
 \item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
-\item $\Phi_i$ is the \textbf{latitude} of the $i$-th satellite with respect to Jupiter's equator.
+\item $\Phi_i$ is the latitude of the $i$-th satellite with respect to Jupiter's equator.
 \end{itemize}
 
-For an irregular satellite $i$, under the gravitational influence of Jupiter, the \textbf{seven other irregular satellites}, the regular Jovian satellites and\textbf{ the main bodies of the Solar System}, the equation of motion is:
+For an irregular satellite $i$, under the gravitational influence of Jupiter, the seven other irregular satellites, the regular Jovian satellites and the main bodies of the Solar System, the equation of motion is:
 \begin{equation}
@@
 \ref{Eq:eq1} represent undirect perturbations.
-The oblateness potential seen by \textbf{a satellite $i$ because of Jupiter is (with a similar expression for the oblateness seen by a satellite $l$)}:
+The oblateness potential seen by a satellite $i$ because of Jupiter is (with a similar expression for the oblateness seen by a satellite $l$):
 \begin{equation}\begin{array}{ll}
@@
 
-\textbf{The expressions of $\nabla U$ have been developed in \cite{Lainey2004}}.
-\textbf{The equations of motion are integrated with the 15th order numerical integrator RADAU \citep{Everhart1985} using a constant step of one day}.
-\textbf{The positions of the objects of the solar system are provided by the DE423 ephemeris \citep{Folkner2010}, while those of the galilean satellites are provided by NOE2010 \citep{Lainey2004}}.
+The expressions of $\nabla U$ have been developed in \cite{Lainey2004}.
+The equations of motion are integrated with the 15th order numerical integrator RADAU \citep{Everhart1985} using a constant step of one day.
+The positions of the objects of the solar system are provided by the DE423 ephemeris \citep{Folkner2010}, while those of the galilean satellites are provided by NOE2010 \citep{Lainey2004}.
 Our model was fitted to the observations through a least-squares procedure.
@@
 The obtained ephemeris is hereafter referred to as STE, for Special-Tailored Ephemeris.
-The initial osculating elements at the origin of integration, \textbf{the number and time span of the observations of each satellite} are presented in Table \ref{Tab: sat_ell}.
+The initial osculating elements at the origin of integration, the number and time span of the observations of each satellite are presented in Table \ref{Tab: sat_ell}.
 
@@
 
-\textbf{Some methods to derive the errors of the ephemeris of irregular satellites can be found in \cite{Emelyanov2010}.
+Some methods to derive the errors of the ephemeris of irregular satellites can be found in \cite{Emelyanov2010}.
 In the Natural Satellites Ephemeride Server MULTI-SAT \citep{Emelyanov2008} the precisions can be obtained for the satellites at any given time from the \cite{Emelyanov2005} ephemeris updated to February 19, 2012.
-However, since the practical realization of the STE ephemeris is for help improving the prediction of stellar occultations by the irregular satellites in the immediate future, it is interesting to compare the STE with the other relevant ephemerides for the next few years.}
+However, since the practical realization of the STE ephemeris is for help improving the prediction of stellar occultations by the irregular satellites in the immediate future, it is interesting to compare the STE with the other relevant ephemerides for the next few years.
 
 We compared the STE ephemeris to the JPL for all the Jupiter satellites we fitted, until 2021.
-For instance, the \textbf{maximum difference} between 2015 and 2021 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.
+For instance, the maximum difference between 2015 and 2021 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.
 
 Fig.
-\ref{Fig: JPL-STE} displays the offsets of the positions published by \citetalias{GomesJunior2015} for the satellite Carme in declination relative to the STE ephemeris, to \cite{Jacobson2012} jup300 JPL ephemeris and \cite{Emelyanov2005}\footnote{\textbf{Last update: February 19, 2012}} ephemeris.
-\textbf{The DE431 planetary ephemeris \citep{Folkner2014} was used to calculate the positions of Jupiter for the 3 models}.
+\ref{Fig: JPL-STE} displays the offsets of the positions published by \citetalias{GomesJunior2015} for the satellite Carme in declination relative to the STE ephemeris, to \cite{Jacobson2012} jup300 JPL ephemeris and \cite{Emelyanov2005}\footnote{Last update: February 19, 2012} ephemeris.
+The DE431 planetary ephemeris \citep{Folkner2014} was used to calculate the positions of Jupiter for the 3 models.
 We see that the systematic JPL ephemeris offsets pointed out by \citetalias{GomesJunior2015} are reduced with our ephemeris, as expected.
@@
 \caption{Offsets in declination of the positions published by \protect\citetalias{GomesJunior2015} for Carme.
-The red "x" relate to the special-tailored ephemeris \textbf{(rms = 51)}, the blue "+" to the jup300 JPL ephemeris \textbf{(rms = 130)} and the green dot to \protect\cite{Emelyanov2005} \textbf{(rms = 92)}.
-As expected, the ephemeris \textbf{offsets} pointed out by \protect\citetalias{GomesJunior2015} are reduced with the STE ephemeris.
+The red "x" relate to the special-tailored ephemeris (rms = 51), the blue "+" to the jup300 JPL ephemeris (rms = 130) and the green dot to \protect\cite{Emelyanov2005} (rms = 92).
+As expected, the ephemeris offsets pointed out by \protect\citetalias{GomesJunior2015} are reduced with the STE ephemeris.
 \label{Fig: JPL-STE}}
@@
 We can see that the mean offsets as well as most of their standard deviations are greatly reduced with the STE ephemeris.
-\textbf{Of course, by construction, we should expect smaller offsets in the comparison of \citetalias{GomesJunior2015} positions with STE.
+Of course, by construction, we should expect smaller offsets in the comparison of \citetalias{GomesJunior2015} positions with STE.
 However, it is not obvious that these offsets should be that smaller in comparison with the other ephemeris offsets.
@@
 Thus, these ephemeris offsets comparisons suggest that the accuracy of the STE ephemeris is at least slightly better than that of the other ephemeris, at least for the time span of our satellites' observations.
-This supports the utility of the STE ephemeris for the next few years, making it one of the best choices to use in stellar occultation predictions in the short future for these satellites.}
+This supports the utility of the STE ephemeris for the next few years, making it one of the best choices to use in stellar occultation predictions in the short future for these satellites.
 
@@
 
-%\textbf{The residuals obtained with STE are smaller than those obtained with JPL ephemeris, which was expected because the accuracy of an ephemeris decreases when we get further from the time of observations.
-If the ephemeris is fitted over too-short a timespan of observations, the initial conditions (positions, velocities, masses) used in the model will not be good enough and the model will differ greatly from the observed position of the satellite when we will accumulate new observations.
-The ephemeris is then said to be diverging.
-%The main risk of divergence over time comes from the possible absence of long-term effects when fitting to a short timespan of observations.
-If that was the case, our ephemeris would diverge too quickly to be of any use.
-%JPL ephemerides are fitted over all the available observations.
-As a result, they will diverge less quickly than our own.
-Though they are not as much precise as required for the use, they remain a precious reference to identify whether our own model presents a quick divergence.}
-
 \subsection{Phoebe's ephemeris}
@@
 The new ephemeris (PH15) used the same dynamical model, including the perturbations of the Sun and the eight planets, the eight major satellites of Saturn and the $J_2$ parameter.
-The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center \textbf{\citep{Arlot2009}}, and observations from Flagstaff \citep{NOFS} between 2012 and 2014.
+The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center \citep{Arlot2009}, and observations from Flagstaff \citep{NOFS} between 2012 and 2014.
 It represents a total number of 5886 observations from 1898 to 2014.
@@
 We computed the precision of the PH15 ephemeris considering three sets of observations: all the positions available; only the positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.
-\textbf{The precision is computed by propagation of the covariance matrix of the orbit determination process and by linear transformations giving the covariance matrix in spherical coordinates (right ascension and declination) at a specific date \citep[for more details, see][]{Desmars2013a}.
-This last matrix then provides the standard deviation in right ascension $\sigma_\alpha$ and declination  $\sigma_\delta$ at the required date, with $\sigma_s=\sqrt{\sigma^2_\alpha\cos^2\delta+\sigma^2_\delta}$ being the total error in the celestial sphere.}
+The precision is computed by propagation of the covariance matrix of the orbit determination process and by linear transformations giving the covariance matrix in spherical coordinates (right ascension and declination) at a specific date \citep[for more details, see][]{Desmars2013a}.
+This last matrix then provides the standard deviation in right ascension $\sigma_\alpha$ and declination  $\sigma_\delta$ at the required date, with $\sigma_s=\sqrt{\sigma^2_\alpha\cos^2\delta+\sigma^2_\delta}$ being the total error in the celestial sphere.
 
 In Fig.
-\ref{Fig:err-Phoebe} we show the comparison between them for the time span 2016-2021 in \textbf{on-sky Phoebe-Saturn angular separation}.
+\ref{Fig:err-Phoebe} we show the comparison between them for the time span 2016-2021 in on-sky Phoebe-Saturn angular separation.
 It is possible to see that even considering only the positions of \citetalias{GomesJunior2015}, the estimated error of the ephemeris is smaller than 12 mas.
@@
 \end{figure}
-
-%\subsection{Triton's and Nereid's ephemeris}
-%
-%Triton is an uncommon satellite.
-Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
-Because Triton's orbit size is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded from the irregular satellites' class, but still studied together by many authors \citep{Sheppard2005, Jewitt2007}.
-Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have the same origin as the TNOs \citep{Agnor2006}.
-However, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
-%
-%For this satellite, we used the most recent ephemeris published by \cite{Emelyanov2015} which is based on more observations than previous ephemeris and covers 4 years more than the ephemeris of \cite{Jacobson2009}.
-In Fig.
-\ref{Fig:eph-Triton} we compare it with the ephemeris published in \cite{Zhang2014} and the NEP081 JPL ephemeris \citep{Jacobson2009}.
-The offsets between the three ephemeris are smaller than 20 mas for the period 2015-2021.
-This value is much smaller than the apparent size of Triton (124 mas, see Table \ref{Tab: satellite-diameter}), which indicates a good probability of observing an event by this object.
-%
-%\begin{figure*}
-%\begin{centering}
-%\subfigure[NEP081 JPL and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
-%\subfigure[\protect\cite{Zhang2014} and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/Zhang-EME_Triton.eps} \label{Fig:eph-Triton-zhang}}
-%\caption{Comparison between the \protect\cite{Emelyanov2015}, \protect\cite{Zhang2014} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Triton.}
-%\label{Fig:eph-Triton}
-%\end{centering}
-%\end{figure*}
-%
-%For Nereid, we used the ephemeris published by \cite{Emelyanov2011}, which uses more recent observations than the JPL ephemeris published by \cite{Jacobson2009}.
-The comparison between the two ephemeris can be seen in Fig.
-\ref{Fig:eph-Nereid}.
-The right ascension offsets between them are smaller than 60 mas and the declination ones are smaller than 15 mas.
-For Nereid, ephemeris errors in right ascension translate to uncertainties in the central instant of a stellar occultation, so offsets of 60 mas are not critical.
-The declination ephemeris offsets are of the order of the apparent diameter (15 mas, see Table \ref{Tab: satellite-diameter}), resulting in fair chances (50\%) that the shadow crosses the expected Earth latitude predicted for an occultation.
-%
-%\begin{figure}
-%\begin{centering}
-%\includegraphics[width=8.8cm]{figures/JPL-EME_Nereid.eps}
-%\caption{Comparison between the \protect\cite{Emelyanov2011} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Nereid.}
-%\label{Fig:eph-Nereid}
-%\end{centering}
-%\end{figure}
-%
-%\cite{GomesJunior2015} published 902 new positions for Nereid from observations between 1992 and 2014.
-The development of a better ephemeris for Nereid using these new positions in the context discussed here of near future stellar occultations is desirable, but is out of the scope of this paper.
 
@@
 This is because at that periods these satellites will cross the apparent galactic plane.
-We call attention that about 10\% of the events will involve stars brighter than \textbf{magnitude R=14}  (and almost 25\% brighter than R=15), which helps the attempt of amateur observers.
+We call attention that about 10\% of the events will involve stars brighter than magnitude R=14  (and almost 25\% brighter than R=15), which helps the attempt of amateur observers.
 
@@
 Phoebe & 32 & 98 & 238 & 79 & 13 & 460 \tabularnewline
-%\hdashline
-%Nereid & 11$^{a}$ & 1 & 1 & 0 & 0 & 13 \tabularnewline
-%Triton & -- & 1 & 1 & 0 & 0 & 2 \tabularnewline
 \hline
@@
 Since these objects are very small, the duration of each event is a few seconds.
-All the occultation tables and maps will be publicly available at the CDS \textbf{(Centre de Données astronomiques de Strasbourg)}.
+All the occultation tables and maps will be publicly available at the CDS (Centre de Données astronomiques de Strasbourg).
 In Fig.
@@
 \hline
-% & \multicolumn{2}{c}{Mean errors} & Nr & Nr &  UCAC4       \tabularnewline
 09 04 2016 03:58:19. & 11 14 36.7707 +07 39 20.7610 & 1.003 &  17.9 & -12.88 &  4.54 & 14.9 & 271. & 22:03 &  12. & -33. \tabularnewline 
@@
 \textbf{Notes}.
-Entries included: day of the year and UTC \textbf{central instant} of the prediction; right ascension and declination of the occulted star - \textbf{at the central instant of the occultation} (corrected by proper motions); C/A: \textbf{apparent geocentric distance between the satellite and the star (a.k.a.
-the distance between the shadow and the center of the Earth) at the moment of the geocentric closest approach}, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees (zero at north of the star, increasing clockwise); $\nu$: \textbf{relative velocity of event in km s$^{-1}$}: positive = prograde, negative = retrograde; {\it D}: \textbf{Geocentric distance to the occulting object in AU}; $R^*$: normalized \textbf{UCAC4 magnitude in the R-band} to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{UCAC4}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$, \textbf{the value 20 km s$^{-1}$ is typical of events around the opposition}; $\lambda$: east longitude of subplanet point in degrees, positive towards east, \textbf{at the instant of the geocentric closest approach}; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
-\textbf{For more detailed information about the definition and use of these stellar occultation geometric elements see \cite{Assafin2010}}.
+Entries included: day of the year and UTC central instant of the prediction; right ascension and declination of the occulted star - at the central instant of the occultation (corrected by proper motions); C/A: apparent geocentric distance between the satellite and the star (a.k.a.
+the distance between the shadow and the center of the Earth) at the moment of the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees (zero at north of the star, increasing clockwise); $\nu$: relative velocity of event in km s$^{-1}$: positive = prograde, negative = retrograde; {\it D}: Geocentric distance to the occulting object in AU; $R^*$: normalized UCAC4 magnitude in the R-band to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{UCAC4}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$, the value 20 km s$^{-1}$ is typical of events around the opposition; $\lambda$: east longitude of subplanet point in degrees, positive towards east, at the instant of the geocentric closest approach; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
+For more detailed information about the definition and use of these stellar occultation geometric elements see \cite{Assafin2010}.
 \end{flushleft}
@@
 \includegraphics[width = 8.8cm]{figures/Elara_2017-02-21T15:28:40.png}   
-\caption{Occultation map for Elara. % regarding to the first event sampled in Table \ref{Tab: occ-list}.
+\caption{Occultation map for Elara.
 The central red dot shows the geocentric closest approach of the shadow.
@@
 The shadow moves from right to left.
-\textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for\textbf{ an observation in the center of the shadow}; C/A: \textbf{apparent geocentric distance between the satellite and the star (a.k.a.
-the apparent distance in the plane of the sky between the shadow and the center of the Earth) at the moment of the geocentric closest approach}, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; $\nu$: \textbf{relative} velocity of event in km/s; {\it D}: Geocentric distance to the occulting object in AU; $R^*$: normalized \textbf{UCAC4 magnitude in the R-band} to a common shadow velocity of 20 km s$^{-1}$; $\lambda$: east longitude of subplanet point in degrees, positive towards east, \textbf{at the central instant of the geocentric closest approach (see the notes of Table 5)}.}
+\textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for an observation in the center of the shadow; C/A: apparent geocentric distance between the satellite and the star (a.k.a.
+the apparent distance in the plane of the sky between the shadow and the center of the Earth) at the moment of the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; $\nu$: relative velocity of event in km/s; {\it D}: Geocentric distance to the occulting object in AU; $R^*$: normalized UCAC4 magnitude in the R-band to a common shadow velocity of 20 km s$^{-1}$; $\lambda$: east longitude of subplanet point in degrees, positive towards east, at the central instant of the geocentric closest approach (see the notes of Table 5).}
 \label{Fig: ocultacao}
@@
 We performed new numerical integrations for improving the orbits of some of the larger irregular satellites.
-Consequently, with \textbf{our} ephemeris, we predicted stellar occultations aiming to access fundamental parameters like size, shape, albedo, ultimately aiming to track the formation origin of these bodies.
+Consequently, with our ephemeris, we predicted stellar occultations aiming to access fundamental parameters like size, shape, albedo, ultimately aiming to track the formation origin of these bodies.
 
@@
 Moreover, the case of Phoebe is particular because we have many observations in a large time spam (1898 to 2014) including observations from Cassini.
-For the Jovian satellites with less observations, the precision we have by using only \citetalias{GomesJunior2015} observations \textbf{may} be quite equivalent or even better than the precision of other ephemeris for the short time span explored in this work.
+For the Jovian satellites with less observations, the precision we have by using only \citetalias{GomesJunior2015} observations may be quite equivalent or even better than the precision of other ephemeris for the short time span explored in this work.
 
@@
 \citetalias{GomesJunior2015} also observed Sycorax (satellite of Uranus) and Nereid (satellite of Neptune).
-\textbf{There were few observations of Sycorax distributed in 9 nights over two years} which did not cover one orbital period.
+There were few observations of Sycorax distributed in 9 nights over two years which did not cover one orbital period.
 For Nereid, the observations covered many orbital periods, but due to Nereid's large orbital eccentricity there are no observations near the pericenter.
@@
 The numerical model of the satellites of Jupiter was developped during a post-doctoral contract funded by the Chinese Academy of Sciences (CAS) and supported by the National Scientific Fund of China (NSFC).
-\textbf{The authors thank Dr. Nikolay Emelyanov for his comments to improve the paper.}
+The authors thank Dr. Nikolay Emelyanov for his comments to improve the paper.
 

### 2016-10-25 relatorio/OV-2016/Relatorio.tex
.

--- 2015-11-11 relatorio_2015/Relatorio.tex
@@
 
+\indent \indent O Sistema Solar, apesar de próximo, em um âmbito astronômico, ainda não é completamente conhecido.
+Uma de suas surpresas mais recentes, por exemplo, foi a descoberta de um sistema de anéis ao redor do Centauro Chariklo \citep{BragaRibas2014}, característica nunca antes observada em um objeto que não fosse um planeta gigante.
+
+Objetos como Centauros, Objetos Trans-Netunianos (TNOs) e satélites irregulares estão localizados a grandes distâncias do Sol, em regiões de baixa temperatura.
+Portanto, comparados a objetos formados mais próximos do Sol, eles provavelmente sofreram pouca diferenciação, seja por mecanismos internos ou por choques com outros objetos.
+Dessa forma, acredita-se que os TNOs guardam estruturas e composições relativamente inalteradas de sua época de formação.
+Por isso TNOs são muito importantes para estudar a formação do Sistema Solar.
+
 \indent \indent O estudo de objetos como objetos trans-Neptunianos (TNOs), Centauros e Satélites Irregulares nos ajudam a compreender a formação e evolução do Sistema Solar Exterior.
@@
 
-Centauros são uma população transiente entre TNOs e cometas da família de Júpiter, orbitando em uma região entre Júpiter e Netuno.
-Atualmente, cerca de 400 deles são conhecidos e é geralmente aceito que eles compartilham uma origem comum com os objetos do cinturão de Kuiper.
+Centauros são uma população transiente entre TNOs e cometas da família de Júpiter, orbitando em uma região entre Júpiter e Netuno.% Atualmente, cerca de 400 deles são conhecidos e é geralmente aceito que eles compartilham uma origem comum com os objetos do cinturão de Kuiper.
 Uma vez que Centauros são tipicamente mais brilhantes que TNOs por estarem mais próximos, eles servem como representantes a partir dos quais é possível inferir propriedades mais gerais sobre os objetos mais distantes \citep{Fernandez2002}.
-Dessa forma, a partir daqui, TNOs e Centauros serão tratados indistintamente como uma única população e referida unicamente como "TNOs".
+%Dessa forma, a partir daqui, TNOs e Centauros serão tratados indistintamente como uma única população e referida unicamente como "TNOs".
 
-Atualmente, é aceito que TNOs tenham sido formados nas partes mais internas do Sistema Solar.
+%Atualmente, é aceito que TNOs tenham sido formados nas partes mais internas do Sistema Solar.
 Eles teriam então sido colocados em suas posições atuais devido a troca de momento angular entre os planetas e planetésimos quando da migração dos planetas gigantes.
@@
 
-Os sistemas de Júpiter e Saturno já foram visitados pelas Voyager I e II, Galileu (Júpiter) e Cassini (Saturno), porém apenas Saturno continua sendo investigado por uma sonda.
+%Os sistemas de Júpiter e Saturno já foram visitados pelas Voyager I e II, Galileu (Júpiter) e Cassini (Saturno), porém apenas Saturno continua sendo investigado por uma sonda.
 Todavia, as sondas observaram apenas os planetas, os anéis e satélites mais internos.
@@
 
-Já no caso de Urano e Netuno, nenhuma sonda exclusiva foi enviada, apenas as \textit{Voyagers I} e \textit{II} os visitaram, mas não permaneceram nos sistemas.
+%Já no caso de Urano e Netuno, nenhuma sonda exclusiva foi enviada, apenas as \textit{Voyagers I} e \textit{II} os visitaram, mas não permaneceram nos sistemas.
 Recentemente, a sonda \textit{New Horizons} visitou Plutão \citep{Stern2015} e obteu parâmetros físicos para o planeta anão e seus satélites (primeira visita por sonda a um objeto do cinturão de Kuiper), porém foi uma passagem rápida e o acompanhamento da evolução do sistema, incluindo a evolução da atmosfera de Plutão se dará por observações de solo.
 
-A quantidade de objetos descobertos além da órbita de Saturno tem aumentado muito desde o fim do século passado.
-Como são raras as oportunidades em que uma sonda se aproxima desses objetos, a obtenção de suas características físicas ficam a cargo de observações de solo ou de
-telescópios espaciais.
+%A quantidade de objetos descobertos além da órbita de Saturno tem aumentado muito desde o fim do século passado.
+Como são raras as oportunidades em que uma sonda se aproxima desses objetos, a obtenção de suas características físicas ficam a cargo de observações de solo ou de telescópios espaciais.
 
 Um método que tem se mostrado eficiente para a obtenção desses parâmetros é o método de ocultações estelares, que proporciona medidas tão precisas que são apenas superadas por medidas oriundas de sondas.
-De fato, a sonda \textit{Dawn} já está orbitando Ceres, um dos objetos deste estudo, e a sonda \textit{New Horizons} visitou Plutão.
+De fato, a sonda \textit{Dawn} já está orbitando Ceres, um dos objetos deste estudo, e a sonda \textit{New Horizons} visitou Plutão \citep{Stern2015}.
 As observações feitas por essas sondas trazem a oportunidade perfeita para comparar os resultados com os obtidos pelas nossas ocultações e calibrar a técnica, de forma que ocultações feitas para outros objetos possam ser mais precisas.
@@
 
-\section{Introdução}
-\label{Sec: TNOs-intro}
+%\section{Introdução}
+%\label{Sec: TNOs-intro}
 
@@
 
-Estimar parâmetros físicos para esses objetos é um desafio, porém essencial para acessar a massa atual e composição (material) da população e recuperar sua história evolutiva.
-É possível obter seus tamanhos através de modelos, por exemplo combinando o brilho no visível e emissão térmica obtida em infra-vermelho.
+Estimar parâmetros físicos como tamanho, albedo, densidade, etc para esses objetos é um desafio, porém essencial para acessar a massa atual e composição (material) da população e recuperar sua história evolutiva.
+%É possível obter seus tamanhos através de modelos, por exemplo combinando o brilho no visível e emissão térmica obtida em infra-vermelho.
 Todavia os erros obtidos são piores que 10 \% \citep{Camargo2013}.
+Para se obter esses parâmetros, de forma muito precisa e acurada, sem a necessidade de se adotar modelos, utilizamos a técnica de ocultações estelares.
+Até 2009, quando foi observada uma ocultação de 2002TX300 \citep{Elliot2010}, Plutão e Caronte eram os únicos TNOs com ocultações observadas.
+Até hoje, cerca de uma dezena de TNOs e Centauros tiveram ocultações observadas.%, alguns com apenas uma corda, como Varuna, o que permite apenas determinar um tamanho mínimo para o objeto.
 
-Para obter parâmetros como tamanho, albedo, densidade, etc, de forma muito mais precisa e acurada, sem a necessidade de se adotar modelos, utilizamos a técnica de ocultações estelares.
-Até 2009, quando foi observada uma ocultação de 2002TX300 \citep{Elliot2010}, Plutão e Caronte eram os únicos TNOs com ocultações observadas.
-Até hoje, cerca de uma dezena de TNOs e Centauros tiveram ocultações observadas, alguns com apenas uma corda, como Varuna, o que permite apenas determinar um tamanho mínimo para o objeto.
-
-Com o objetivo de prever e observar ocultações estelares por TNOs, o nosso grupo conta com colaboração de pesquisadores nacionais e internacionais, profissionais e amadores.
+%Com o objetivo de prever e observar ocultações estelares por TNOs, o nosso grupo conta com colaboração de pesquisadores nacionais e internacionais, profissionais e amadores.
 Utilizamos telescópios de pequeno e grande porte e câmera rápida para observar eventos que proporcionaram grandes descobertas, como por exemplo a descoberta do sistema de anéis ao redor do Centauro Chariklo \citep{BragaRibas2014}.
 
-\section{Astrometria}
-\label{Sec: TNOs-astrometria}
+\section{Ocultações Estelares}
+\label{Sec: TNO-occ}
 
@@
 
-Da mesma forma, estando muito distantes, eles se movem relativamente devagar no céu ocultando relativamente poucas estrelas por ano.
+%Da mesma forma, estando muito distantes, eles se movem relativamente devagar no céu ocultando relativamente poucas estrelas por ano.
 Porém, sendo menos brilhantes (Mag $V > 18$) é possível observar ocultações de estrelas mais fracas (estrelas até Mag $R = 17$), onde a queda de brilho da estrela durante a ocultação poderá ser detectada e, mesmo com baixo sinal-ruído (ex: 5 a 10), o tamanho do corpo pode ser calculado com precisão de pelo menos 10 a 20 km, melhorando à medida em que a velocidade da sombra seja menor que 20 km/s (velocidade típica para um corpo em oposição a 30 UA).
@@
 
-\section{Ocultações Estelares}
-\label{Sec: TNO-occ}
-
-\indent \indent Com o objetivo de obter os parâmetros físicos desses objetos, para melhor caracterizar as propriedades básicas dessas populações, uma grande campanha observacional é realizada para se observar ocultações estelares por TNOs e Centauros.
-Essa campanha é liderada por pesquisadores brasileiros, franceses e espanhóis e tem colaboração com observadores espalhados por todo o mundo.
-
-Os objetos foram basicamente selecionados de acordo com o grau de interesse científico (representar uma classe de TNO, ser um  objeto peculiar, etc), tamanho (quanto maior, mais chances de observar a ocultação) e região do céu (quanto mais estrelas, maior  a probabilidade de ocultar estrelas brilhantes, permitindo curvas de luz de melhor resolução espacial).
-
-Vários objetos já foram observados pelo grupo, sendo que o Sistema Plutão-Caronte conta com várias ocultações observadas além de um trabalho astrométrico minucioso baseado em 19 anos de observações no OPD, mais observações no ESO \citep{Rossi2014}.
+%Com o objetivo de obter os parâmetros físicos desses objetos, para melhor caracterizar as propriedades básicas dessas populações,
+Uma grande campanha observacional é realizada para se observar ocultações estelares por TNOs e Centauros.
+Essa campanha é liderada por pesquisadores brasileiros, franceses e espanhóis e tem colaboração com observadores espalhados por todo o mundo
+%Os objetos foram basicamente selecionados de acordo com o grau de interesse científico (representar uma classe de TNO, ser um  objeto peculiar, etc), tamanho (quanto maior, mais chances de observar a ocultação) e região do céu (quanto mais estrelas, maior  a probabilidade de ocultar estrelas brilhantes, permitindo curvas de luz de melhor resolução espacial).
+resultando em vários objetos já observados.
+% pelo grupo, sendo que o Sistema Plutão-Caronte conta com várias ocultações observadas além de um trabalho astrométrico minucioso baseado em 19 anos de observações no OPD, mais observações no ESO \citep{Rossi2014}.
 As observações bem sucedidas de ocultações estelares resultantes desse trabalho de predição já geraram diversas publicações, inclusive na revista Nature, por exemplo \cite{Sicardy2011} e \cite{BragaRibas2014} nas quais contribuí em caráter observacional.
@@
 
-No dia 29 de Junho de 2015 houve uma ocultação de Plutão de máxima importância devido a esta ser a última antes da chegada da sonda \textit{New Horizons} no sistema e foi observada na Austrália e Nova Zelândia.
+Além desses contribuí com outras ocultações como a de Plutão observada na Austrália e Nova Zelândia em 29 de Junho de 2015 e a de Chariklo em 29 de Abril de 2014 reduzindo curvas de luz e observei uma ocultação do TNO 2003VS2 dia 07 de Novembro de 2014 no OPD.
+Esses trabalhos ainda não foram publicados.
+
+%No dia 29 de Junho de 2015 houve uma ocultação de Plutão de máxima importância devido a esta ser a última antes da chegada da sonda \textit{New Horizons} no sistema e foi observada na Austrália e Nova Zelândia.
 Os modelos de perfil de atmosfera para Plutão poderão ser calibrados com os dados da \textit{New Horizons} para serem utilizados em futuras ocultações e estudar a evolução da atmosfera do planeta anão.
 
-Em 29 de Abril de 2014, uma ocultação de Chariklo foi observada na África.
+%Em 29 de Abril de 2014, uma ocultação de Chariklo foi observada na África.
 Eu trabalhei com a redução de três curvas de luz (Gifberb, SAAO e Springbok).
@@
 
-Outro evento que observei foi uma ocultação do TNO 2003VS2 dia 07 de Novembro de 2014 no OPD.
-As variações do céu não permitiram uma boa observação de forma que a curva de luz estava muito ruidosa, mas boa o suficiente próximo ao instante do evento.
+%Outro evento que observei foi  As variações do céu não permitiram uma boa observação de forma que a curva de luz estava muito ruidosa, mas boa o suficiente próximo ao instante do evento.
 Um trabalho foi feito para identificar alguma ocultação mas nada pôde ser extraído.
@@
 
-\section{Introdução}
-\label{Sec: Ceres-intro}
+%\section{Introdução}
+%\label{Sec: Ceres-intro}
 
@@
 
-Foram preditos dois eventos por Steve Preston\footnote{Predições publicadas em \url{http://asteroidoccultation.com}.} para a IOTA (International Occultation Timing Association), durante predições de rotina de ocultações de estrelas brilhantes por asteroides.
-Os caminhos das sombras podem ser visualizados na Fig.
+Foram preditos dois eventos por Steve Preston\footnote{Predições publicadas em \url{http://asteroidoccultation.com}.} para a IOTA (International Occultation Timing Association), durante predições de rotina de ocultações de estrelas brilhantes por asteroides.% Os caminhos das sombras podem ser visualizados na Fig.
 \ref{Fig: Ceres-map}.
 
-\begin{figure}[h]
-\begin{centering}
-\subfigure[Ocultação de 2010]{\includegraphics[scale=0.39]{figuras/Ceres_2010.eps}\label{Fig: Ceres-map-2010}}
-\subfigure[Ocultação de 2013]{\includegraphics[scale=0.39]{figuras/Ceres_2013-zoom.eps}\label{Fig: Ceres-map-2013}}
-\caption{Reconstrução pós-ocultação do caminho da sombra de Ceres na Terra para os eventos de 17 de Agosto de 2010 (a) e 25 de Outubro de 2013 (b).
+%\begin{figure}[h]
+%\begin{centering}
+%\subfigure[Ocultação de 2010]{\includegraphics[scale=0.39]{figuras/Ceres_2010.eps}\label{Fig: Ceres-map-2010}}
+%\subfigure[Ocultação de 2013]{\includegraphics[scale=0.39]{figuras/Ceres_2013-zoom.eps}\label{Fig: Ceres-map-2013}}
+%\caption{Reconstrução pós-ocultação do caminho da sombra de Ceres na Terra para os eventos de 17 de Agosto de 2010 (a) e 25 de Outubro de 2013 (b).
 Os pontos em azul são os sítios que observaram os eventos.
@@
 Os pontos vermelhos são os centros da sombra separados por 15 s.
-\label{Fig: Ceres-map}}
-\end{centering}
-\end{figure}
+%\label{Fig: Ceres-map}}
+%\end{centering}
+%\end{figure}
 
@@
 
-\indent O primeiro evento foi observado no Brasil a partir de cinco diferentes sítios (ver Fig.
-\ref{Fig: Ceres-map-2010}).
+O primeiro evento foi observado no Brasil a partir de cinco diferentes sítios.
 Destes, 4 obtiveram cordas positivas enquanto UFSC teve uma corda negativa.
 Das positivas, a observação proveniente do INPE iniciou-se após o início do evento devido a dificuldades técnicas e, portanto, apenas a emersão da curva de luz foi detectada.
-O segundo foi observado na costa Leste dos Estados Unidos (\ref{Fig: Ceres-map-2013}).
+O segundo foi observado na costa Leste dos Estados Unidos.
 
-Uma das características mais importantes desse evento foi a velocidade com que ocorreu (apenas 3.9 km s$^{-1}$) acarretando que mesmo exposições de poucos segundos representariam resoluções espaciais significantes.
+%Uma das características mais importantes desse evento foi a velocidade com que ocorreu (apenas 3.9 km s$^{-1}$) acarretando que mesmo exposições de poucos segundos representariam resoluções espaciais significantes.
 
-Todas as observações foram feitas com a utilização de CCDs.
+%Todas as observações foram feitas com a utilização de CCDs.
 As curvas de luz de cada observação foram obtidas das imagens FITS com a utilização do pacote PRAIA \citep[Plataforma de Redução Astrométrica de Imagens Astronômicas,][]{Assafin2011}.
-As curvas foram normalizadas para o fluxo da estrela mais Ceres, uma vez que eles estavam indistinguíveis logo antes e depois da ocultação.
+%As curvas foram normalizadas para o fluxo da estrela mais Ceres, uma vez que eles estavam indistinguíveis logo antes e depois da ocultação.
 Por fim, elas foram normalizadas pelo ajuste de uma curva polinomial (de primeira ou segunda ordem) fora da queda de fluxo assim fixando em 1 a razão de fluxo fora da ocultação.
-
 Os instantes de ingresso e egresso foram obtidas de cada curva de luz ajustando-se um modelo de poço quadrado levando em consideração a difração de Fresnel, a banda do CCD, o diâmetro aparente da estrela e o tempo de exposição utilizado \citep[ver][]{Widemann2009, Braga-Ribas2013}.
@@
 
-O ajuste dos dados da ocultação consiste em minimizar uma função de $\chi^{2}$ clássica para cada curva de luz, como descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}.
+%O ajuste dos dados da ocultação consiste em minimizar uma função de $\chi^{2}$ clássica para cada curva de luz, como descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}.
 Os parâmetros livres para ajustar são os instantes de ingresso e egresso que fornece o valor mínimo de $\chi^{2}$ ($\chi^{2}_{min}$).
@@
 
-\begin{figure}
-\begin{centering}
-\subfigure[Ocultação de 2010]{\includegraphics[scale=0.35]{figuras/Ceres_2010_sphere.eps}\label{Fig: Ceres-limb-2010} }
-\subfigure[Ocultação de 2013]{\includegraphics[scale=0.35]{figuras/Ceres_2013_sphere.pdf}\label{Fig: Ceres-limb-2013} }
-\caption{Melhores ajustes elípticos para as cordas das ocultações de 2010 e 2013.
+%\begin{figure}[h]
+%\begin{centering}
+%\subfigure[Ocultação de 2010]{\includegraphics[scale=0.35]{figuras/Ceres_2010_sphere.eps}\label{Fig: Ceres-limb-2010} }
+%\subfigure[Ocultação de 2013]{\includegraphics[scale=0.35]{figuras/Ceres_2013_sphere.pdf}\label{Fig: Ceres-limb-2013} }
+%\caption{Melhores ajustes elípticos para as cordas das ocultações de 2010 e 2013.
 As setas indicam a direção de movimento, as linhas azuis são as cordas observadas, os segmentos vermelhos são as barras de erro dos ingressos, egressos e centro da ocultação em $1\sigma$.
@@
 Os instantes marcados em verde em (b) não foram utilizados para o ajuste, como descrito no texto.
-\label{Fig: Ceres-limb}}
-\end{centering}
-\end{figure}
+%\label{Fig: Ceres-limb}}
+%\end{centering}
+%\end{figure}
 
@@
 Como foi mostrado, usar as coordenadas do polo de Ceres determinadas por \cite{Drummond2014} para limitar o ângulo de posição não foi um procedimento eficiente para o evento de 2013.
-Por outro lado, fixar o ângulo de posição para a ocultação de 2010 reduziu as barras de erro dos outros parâmetros (com exceção do achatamento).
-Por fim, esse procedimento resultou em um excelente acordo entre os raios equatoriais obtidos para ambos os eventos.
+Por outro lado, reduziu as barras de erro dos outros parâmetros (com exceção do achatamento) para a ocultação de 2010.% Por fim, esse procedimento resultou em um excelente acordo entre os raios equatoriais obtidos para ambos os eventos.
 
@@
 
-A sonda da NASA \textit{Dawn} poderá elucidar muitas perguntas sobre Ceres e seus resultados serão para calibrar todas as técnicas usadas até agora para o estudo das propriedades físicas dos pequenos objetos do Sistema Solar, como as ocultações estelares.
+A sonda da NASA \textit{Dawn} poderá elucidar muitas perguntas sobre Ceres e seus resultados serão utilizados para calibrar todas as técnicas usadas até agora para o estudo das propriedades físicas dos pequenos objetos do Sistema Solar, como as ocultações estelares.
 
@@
 Apenas Himalia, Phoebe e Nereida foram observados por sondas, apesar de não serem medidas ideais, já que foram alvos de oportunidade.
-A sonda \textit{Cassini} observou Himalia em 2000 ao passar próximo a Júpiter e obteve o tamanho de Himalia com um erro médio de 10 km \citep{Porco2003}.
+Destes apenas Phoebe teve seu tamanho estimado a partir de imagens de alta resolução da Cassini com um erro médio de 0.7 km.%A sonda \textit{Cassini} observou Himalia em 2000 ao passar próximo a Júpiter e obteve o tamanho de Himalia com um erro médio de 10 km \citep{Porco2003}.
 Em 2004, a \textit{Cassini}, aproximando-se de Saturno, observou Phoebe em alta resolução obtendo um erro médio na medida de seu tamanho de 0.7 km \citep{Thomas2010}.
@@
 Se esses objetos foram capturados, permanece a pergunta de onde eles vieram.
-Estudos como os de \citealp{Grav2003, Clark2005, Grav2007} mostraram através de suas cores e inclinações espectrais que esses satélites contêm uma fração mais ou menos igual de objetos semelhantes a asteroides tipo C, P e D ou a Centauros e TNOs.
+Estudos como os de \citealp{Grav2003, Clark2005} e \citealp{Grav2007} mostraram através de suas cores e inclinações espectrais que esses satélites contêm uma fração mais ou menos igual de objetos semelhantes a asteroides tipo C, P e D ou a Centauros e TNOs.
 Esses trabalhos sugerem origens diferentes para os satélites irregulares.
@@
 
-Um total de 6523 posições de 18 satélites foram identificadas. Esse trabalho está publicado \citep{GomesJunior2015-Irregular} e o catálogo com as posições está disponível no CDS.
-
+Um total de 6523 posições de 18 satélites foram identificadas.
+Esse trabalho está publicado \citep{GomesJunior2015-Irregular} e o catálogo com as posições está disponível no CDS.
 Um dos principais resultados obtidos foi a grande quantidade de posições identificadas para os satélites irregulares em comparação com a quantidade utilizada nas integrações numéricas atuais publicadas por \cite{Jacobson2012}.
@@
 Observando os offsets das posições com respeito às efemérides ao longo do tempo e anomalia verdadeira foram identificados erros sistemáticos nas efemérides.
-O erro mais claro é na declinação do satélite Carme o qual foi associado a um erro em sua inclinação orbital.
-
+O erro mais evidente é na declinação do satélite Carme o qual foi associado a um erro em sua inclinação orbital.
 %\begin{figure}[h]
@@
 %\end{figure}
-
 Esse padrão em declinação também foi identificado para outros satélites como Pasiphae e Ananke.
@@
 
-\indent \indent Com o objetivo de obter parâmetros físicos para os satélites irregulares com maior precisão foram feitas predições de ocultações estelares para os 7 maiores satélites de Júpiter (Himalia, Elara, Pasiphae, Sinope, Lysithea, Carme, Ananke), o satélite Phoebe de Saturno e Nereida de Netuno.
+\indent \indent Com o objetivo de obter parâmetros físicos para os satélites irregulares com maior precisão foram feitas predições de ocultações estelares para os 7 maiores satélites de Júpiter (Himalia, Elara, Pasiphae, Sinope, Lysithea, Carme, Ananke), o satélite Phoebe de Saturno e Tritão e Nereida de Netuno.
 Esses objetos são pequenos, se comparados aos TNOs estudados no capítulo \ref{Cap: TNOs}, sendo que o menor dos satélites da amostra, Ananke, possui um diâmetro estimado de 28 km.
@@
 Essa integração foi realizada pela pós-doc Laurène Beauvalet.
-Para Phoebe e Nereida foram utilizadas efemérides do JPL.
+Para Phoebe, Tritão e Nereida foram utilizadas efemérides do JPL.
 
-Um total de 588 eventos foram obtidos entre 01 de Janeiro de 2015 e 31 de Dezembro de 2017 para os 9 objetos sendo que a maioria será descartada por passar em regiões de oceano ou onde não há observadores disponíveis.
+Aproximadamente 600 eventos foram obtidos entre 01 de Janeiro de 2015 e 31 de Dezembro de 2017 para os 10 objetos sendo que a maioria será descartada por passar em regiões de oceano ou onde não há observadores disponíveis.
 Estima-se que restem ao final cerca de 10\% desse número.
-Para Nereida, apenas uma estrela UCAC4 foi identificada para ser ocultada.
-Nesse caso, utilizamos um catálogo de estrelas feito a partir do caminho aparente do satélite de Netuno, Tritão, no céu para predições de ocultações estelares pelo mesmo.
+Para Tritão e Nereida% , apenas uma estrela UCAC4 foi identificada para ser ocultada.
+Nesse caso,
+utilizamos um catálogo de estrelas feito a partir do caminho aparente do satélite de Netuno, Tritão, no céu para predições de ocultações estelares pelo mesmo.
 
@@
 
-Observar uma ocultação estelar de um satélite irregular exige um esforço muito grande por parte dos observadores.
+%Observar uma ocultação estelar de um satélite irregular exige um esforço muito grande por parte dos observadores.
 A sombra cobrirá uma região muito restrita na Terra devido aos satélites irregulares serem pequenos.
@@
 
-Um pedido de tempo foi feito para o telescópio Zeiss para se tentar observar uma ocultação de Himalia  dia 23 de Outubro de 2015.
+%Um pedido de tempo foi feito para o telescópio Zeiss para se tentar observar uma ocultação de Himalia  dia 23 de Outubro de 2015.
 A sombra dessa ocultação estava predita para passar pela região sudeste do Brasil.
@@
 
-\section{Introdução}
-\label{Sec: Netuno-intro}
+%\section{Introdução}
+%\label{Sec: Netuno-intro}
 
@@
 
-\indent \indent Nos mesmos moldes em que foram feitas observações dos caminhos de TNOs para predição de ocultações estelares (ver seção \ref{Sec: TNOs-astrometria}), o caminho de Tritão também foi observado gerando um catálogo de estrelas nos moldes do catálogo WFI.
+%\indent \indent Nos mesmos moldes em que foram feitas observações dos caminhos de TNOs para predição de ocultações estelares (ver seção \ref{Sec: TNO-occ}), o caminho de Tritão também foi observado gerando um catálogo de estrelas nos moldes do catálogo WFI.
 Infelizmente, devido a Netuno estar passando por regiões pouco populadas por estrelas brilhantes, nenhum evento pôde ser observado oriundo deste trabalho.
@@
 
-Em 21 de Maio de 2008 houve uma ocultação por Tritão cuja sombra cruzou o sul da África.
+\indent \indent Em 21 de Maio de 2008 houve uma ocultação por Tritão cuja sombra cruzou o sul da África.
 Dos sítios que observaram, apenas duas cordas foram positivas.
@@
 \indent \indent Com o objetivo de estudar os efeitos de maré sobre o sistema Netuno-Tritão, serão obtidas as posições desses objetos em mais de 7000 imagens observadas no OPD desde 1992 até o presente.
-Uma vez que eles estão muito próximos no céu e Netuno é muito mais brilhante do que Tritão, a posição do satélite pode estar afetada pela luz do planeta.
+%Uma vez que eles estão muito próximos no céu e Netuno é muito mais brilhante do que Tritão, a posição do satélite pode estar afetada pela luz do planeta.
 Para eliminar essa influência será utilizada uma técnica chamada de "coronografia digital" \citep{Assafin2009}.
-
-A vantagem desse banco de dados é sua cobertura temporal (1992 - 2015), o grande número de observações e um tratamento astrométrico diferenciado.
-Faremos uso de coronografia digital para eliminar a influência da luz difusa de Netuno sobre Tritão, e faremos uma determinação direta de posições de Netuno (com PSFs numéricas especialmente deduzidas para o caso de objetos extensos), com Tritão no mesmo frame.
+A vantagem desse banco de dados é sua cobertura temporal, o grande número de observações e um tratamento astrométrico diferenciado.
+Faremos uso de coronografia digital \citep{Assafin2009} para eliminar a influência da luz difusa de Netuno sobre Tritão, e faremos uma determinação direta de posições de Netuno (com PSFs numéricas especialmente deduzidas para o caso de objetos extensos), com Tritão no mesmo frame.
 Esses dados tem o mesmo valor que o de posições oriundas de sondas, em que pesem efeitos dinâmicos de longo período \citep{Lainey2012}.
 
-Além das posições de solo, irei trabalhar com posições telemétricas de sondas espaciais dentro do projeto ESPaCE (European Satellite PArtnership for Computing Ephemerides).
+Além das posições de solo, irei trabalhar com posições de telemetria de sondas espaciais dentro do projeto ESPaCE (European Satellite PArtnership for Computing Ephemerides).
 Esse projeto tem por objetivo extrair e analisar dados astrométricos de medidas feitas por sondas espaciais e utilizá-las para cálculo de órbitas e geração de efemérides dos objetos dos Sistema Solar.
@@
 A combinação desses dois bancos de dados, junto com as posições publicadas na literatura permitirá estudar efeitos dissipativos de energia no interior de Tritão.
-Como dito na seção \ref{Sec: Netuno-intro}, a circularização da órbita de Tritão, a partir de uma órbita altamente excêntrica, após a captura aconteceu por dissipação oriunda de efeitos de maré \citep{Nimmo2015}.
+%Como dito na seção \ref{Sec: Netuno-intro}, a circularização da órbita de Tritão, a partir de uma órbita altamente excêntrica, após a captura aconteceu por dissipação oriunda de efeitos de maré \citep{Nimmo2015}.
 
@@
 \indent \indent Esse trabalho tem por meta estudar os objetos do Sistema Solar exterior, de forma a obter características físicas e dinâmicas relevantes para entender a formação e evolução do Sistema Solar.
-As técnicas utilizadas consistem em obter posições astrométricas precisas desses corpos para corrigir erros sistemáticos em suas efemérides, ou mesmo permitir a determinação de novas órbitas com novas integrações numéricas, a fim de melhorar a predição de futuras ocultações estelares.% A técnica de fotometria diferencial de abertura é empregada para gerar as curvas de luz das ocultações estelares.
+As técnicas utilizadas consistem em obter posições astrométricas precisas desses corpos para corrigir erros sistemáticos em suas efemérides, ou mesmo permitir a determinação de novas órbitas com novas integrações numéricas, a fim de melhorar a predição de futuras ocultações estelares.
+% A técnica de fotometria diferencial de abertura é empregada para gerar as curvas de luz das ocultações estelares.
 Prever, observar e reduzir ocultações estelares por esses objetos e obter seus parâmetros físicos (tamanho, forma, albedo, densidade, presença de atmosfera, satélites ou anéis) são os objetivos gerais desta Tese.
@@
 Ocultações estelares por esse objeto permitem estudar a evolução de sua atmosfera, a qual será calibrada a partir de observações oriundas da sonda \textit{New Horizons}.
-Para isso uma ocultação que ocorrei próxima à data da chegada da sonda foi observada.
+%Para isso uma ocultação que ocorreu próxima à data da chegada da sonda foi observada.
 Das observações às quais participei, dois estudos sobre a atmosfera de Plutão puderam ser feitos: \cite{Olkin2015} e \cite{DiasOliveira2015}.
@@
 Diversas ocultações já estão preditas e a publicação destas predições está em preparação (Gomes-Júnior et al.
-2016, A\&A, in prep.).
-Observações de ocultações por Himalia, satélite de Júpiter, e Tritão estão sendo preparadas.
+2016, A\&A, in prep.).% Observações de ocultações por Himalia, satélite de Júpiter, e Tritão estão sendo preparadas.
 

### 2016-11-07 relatorio/OV-2016/Relatorio.tex
.

@@
 \usepackage{arydshln}
+%\usepackage{floatrow}
 \usepackage{footnote}
 \usepackage[left=3cm,right=2cm,top=3cm,bottom=2cm]{geometry}
+%\usepackage{blindtext}
+
 %\usepackage[alf]{abntcite}
@@
 \newcommand{\fdg}{\mbox{\ensuremath{.\!\!^\circ}}}%             % fractional degree symbol:     0.°0
+%\newfloatcommand{capbtabbox}{table}[][\FBwidth]
 
@@
 \renewcommand{\@makechapterhead}[1]{%
-\vspace*{50 pt}%
+\vspace*{20 pt}%
 {\setlength{\parindent}{0pt} \raggedright \normalfont
-\bfseries\Huge\thechapter.\ #1
-\par\nobreak\vspace{40 pt}}}
+\bfseries\Large\thechapter.\ #1
+\par\nobreak\vspace{8 pt}}}
 \makeatother
@@
 %\input{glossario}
+\date{}
 \author{Altair Ramos Gomes Júnior}
-\title{Um Estudo Astrométrico e Fotométrico do Sistema Solar Exterior}
+\title{Relatório das Atividades de PG 2015-2016}
 \begin{document}
@@
 Objetos como Centauros, Objetos Trans-Netunianos (TNOs) e satélites irregulares estão localizados a grandes distâncias do Sol, em regiões de baixa temperatura.
-Portanto, comparados a objetos formados mais próximos do Sol, eles provavelmente sofreram pouca diferenciação, seja por mecanismos internos ou por choques com outros objetos.
-Dessa forma, acredita-se que os TNOs guardam estruturas e composições relativamente inalteradas de sua época de formação.
-Por isso TNOs são muito importantes para estudar a formação do Sistema Solar.
+Portanto, comparados a objetos formados mais próximos do Sol, eles provavelmente sofreram pouca diferenciação, seja por mecanismos internos ou por colisões frequentes (choques reais ou passagens gravitacionalmente próximas) com outros objetos.
 
-\indent \indent O estudo de objetos como objetos trans-Neptunianos (TNOs), Centauros e Satélites Irregulares nos ajudam a compreender a formação e evolução do Sistema Solar Exterior.
-Nesta região distante do Sol, de baixas temperaturas, objetos principalmente de tamanho relativamente pequeno, e mais dispersos no espaço, provavelmente sofreram muito pouca diferenciação, seja por mecanismos internos, seja por choques com outros corpos, comparados a objetos formados mais próximos do Sol.
-
-Por serem corpos asteroidais localizados de forma dispersa além da órbita de Netuno, considera-se que os TNOs guardam estruturas e composições relativamente inalteradas em relação a sua época de formação, constituindo-se assim em corpos de prova de grande valor para o estudo da origem do Sistema Solar, ao menos para essa região exterior.
-
+Acredita-se, por isso, que os TNOs guardam estruturas e composições relativamente inalteradas de sua época de formação.
+Portanto eles são muito importantes para estudar a formação do Sistema Solar.
 Tirando Plutão, o primeiro TNO foi descoberto há pouco mais de 20 anos \citep{Jewitt1993}.
@@
 
-Centauros são uma população transiente entre TNOs e cometas da família de Júpiter, orbitando em uma região entre Júpiter e Netuno.% Atualmente, cerca de 400 deles são conhecidos e é geralmente aceito que eles compartilham uma origem comum com os objetos do cinturão de Kuiper.
+Centauros são uma população transiente entre TNOs e cometas da família de Júpiter, orbitando em uma região entre Júpiter e Netuno.
 Uma vez que Centauros são tipicamente mais brilhantes que TNOs por estarem mais próximos, eles servem como representantes a partir dos quais é possível inferir propriedades mais gerais sobre os objetos mais distantes \citep{Fernandez2002}.
-%Dessa forma, a partir daqui, TNOs e Centauros serão tratados indistintamente como uma única população e referida unicamente como "TNOs".
 
-%Atualmente, é aceito que TNOs tenham sido formados nas partes mais internas do Sistema Solar.
-Eles teriam então sido colocados em suas posições atuais devido a troca de momento angular entre os planetas e planetésimos quando da migração dos planetas gigantes.
-A evolução se deu de tal forma que a passagem dos planetesimais e planetas por zonas de ressonância de movimento médio redefiniu as órbitas desses corpos \citep{Tsiganis2005}.
-
-Muitos dos objetos que pertenciam ao cinturão de Kuiper primordial acabaram sendo enviados pela interação com os planetas gigantes para as partes mais internas do Sistema Solar.
-Alguns podem ter sido capturados pelos planetas criando a população de satélites irregulares ou satélites externos \citep{Nesvorny2007}, troianos \citep{Morbidelli2005} ou até mesmo para o cinturão principal de asteroides como proposto para Ceres por \cite{McKinnon2012}.
+Segundo o modelo de Nice \citep{Morbidelli2005,Gomes2005,Tsiganis2005}, muitos dos objetos que pertenciam ao cinturão de Kuiper primordial acabaram sendo enviados pela interação com os planetas gigantes para as partes mais internas do Sistema Solar.
+Alguns podem ter sido capturados pelos planetas criando a população de satélites irregulares ou satélites externos \citep{Nesvorny2007}, troianos \citep{Morbidelli2005} ou até mesmo para o cinturão principal de asteroides como proposto para Ceres por \cite{McKinnon2012}, hipótese reforçada por resultados recentes \citep{DeSanctis2015} a partir de observações da sonda \textit{Dawn}.
 Estudar esses objetos é de grande importância para entender a evolução do Sistema Solar.
@@
 
-Sabe-se que poucas sondas espaciais foram enviadas para estudar o Sistema Solar externo e que a quantidade de objetos estudados é muito pequena.
+Com exceção dos planetas gigantes e seus satélites internos, apenas Plutão e Phoebe, satélite irregular de Saturno, foram observados, com grande qualidade de dados, por sondas espaciais além da órbita de Júpiter.
 Por isso, ainda hoje, as observações de solo tem se mostrado de grande importância.
 
-%Os sistemas de Júpiter e Saturno já foram visitados pelas Voyager I e II, Galileu (Júpiter) e Cassini (Saturno), porém apenas Saturno continua sendo investigado por uma sonda.
-Todavia, as sondas observaram apenas os planetas, os anéis e satélites mais internos.
-Os satélites externos ou foram pouco observados (como Phoebe) ou simplesmente não foram observados.
-
-%Já no caso de Urano e Netuno, nenhuma sonda exclusiva foi enviada, apenas as \textit{Voyagers I} e \textit{II} os visitaram, mas não permaneceram nos sistemas.
-Recentemente, a sonda \textit{New Horizons} visitou Plutão \citep{Stern2015} e obteu parâmetros físicos para o planeta anão e seus satélites (primeira visita por sonda a um objeto do cinturão de Kuiper), porém foi uma passagem rápida e o acompanhamento da evolução do sistema, incluindo a evolução da atmosfera de Plutão se dará por observações de solo.
-
-%A quantidade de objetos descobertos além da órbita de Saturno tem aumentado muito desde o fim do século passado.
-Como são raras as oportunidades em que uma sonda se aproxima desses objetos, a obtenção de suas características físicas ficam a cargo de observações de solo ou de telescópios espaciais.
-
 Um método que tem se mostrado eficiente para a obtenção desses parâmetros é o método de ocultações estelares, que proporciona medidas tão precisas que são apenas superadas por medidas oriundas de sondas.
-De fato, a sonda \textit{Dawn} já está orbitando Ceres, um dos objetos deste estudo, e a sonda \textit{New Horizons} visitou Plutão \citep{Stern2015}.
-As observações feitas por essas sondas trazem a oportunidade perfeita para comparar os resultados com os obtidos pelas nossas ocultações e calibrar a técnica, de forma que ocultações feitas para outros objetos possam ser mais precisas.
+De fato, a sonda \textit{Dawn} já está orbitando Ceres e a sonda \textit{New Horizons} visitou Plutão \citep{Stern2015}.
+As observações feitas por essas sondas trazem a oportunidade perfeita para comparar os resultados com os obtidos pelas nossas ocultações e calibrar a técnica, de forma que ocultações feitas para outros objetos possam ser ainda mais precisas.
 
 O objetivo desse Doutorado é observar objetos do Sistema Solar exterior, fazer astrometria para melhorar suas efemérides de forma que ocultações estelares por esses objetos possam ser preditas com precisão, observar as ocultações e obter os parâmetros físicos.
-Além disso, será feito um estudo astrométrico do sistema Netuno-Tritão para obter parâmetros relacionados a efeitos de maré, que permitem em princípio inferir a estrutura interna desses corpos.
 
@@
 
-%Nos capítulos \ref{Cap: TNOs}, \ref{Cap: Ceres}, \ref{Cap: Irregulares} e \ref{Cap: Netuno-Tritao} relatamos os trabalhos desenvolvidos envolvendo predições e observações de ocultações estelares com Centauros e TNOs, Ceres, satélites irregulares dos planetas gigantes e com o sistema Netuno-Tritão, respectivamente.
-Por fim, no capítulo \ref{Cap: perspectivas} resumimos o que já foi desenvolvido e descrevemos o que ainda está para ser desenvolvido.
-
-%\chapter{Astrometria}
-%\label{Cap: Astrometria}
-%
-%
-%\indent \indent O tratamento astrométrico das imagens é feito utilizando a Plataforma para Redução Automática de Imagens Astronômicas \citep[PRAIA][]{Assafin2011}.
-As medidas (x, y) são obtidas através de uma Gaussiana bidimensional simétrica ajustada até uma largura à altura (FWHM = seeing).
-Em uma FWHM, o perfil da imagem é bem descrito por um perfil gaussiano, livre das distorções de borda que prejudicam a determinação do centro.
-O PRAIA identifica automaticamente as estrelas de catálogo e determina ($\alpha$, $\delta$) com um modelo definido pelo usuário que relaciona as medidas (x, y) da imagem e as coordenadas projetadas no plano do céu (X, Y).
-Os alvos são automaticamente identificados nas reduções, a partir das suas efemérides (JPL, NAIF, etc) e suas posições, offsets de efeméride e estimativa de erros de medida e de posição guardados para análise posterior.
-%
-%Majoritariamente, o catálogo UCAC4 \citep{Zacharias2013} é utilizado como representante prático do ICRS.
-Como será exposto na seção \ref{Sec: TNOs-astrometria}, o catálogo WFI é utilizado para determinados TNOs até 2015.
-Esses catálogos serão substituídos pelo catálogo GAIA uma vez que seja lançado em meados de 2016.
-%
-%Para cada frame, usamos um modelo polinomial que pode ser de grau 1, 2 ou 3 para relacionar as medidas (x, y) com as coordenadas (X, Y) do plano tangente.
-Para observações feitas no ESO, o procedimento astrométrico seguido é o mesmo descrito em \citep{Assafin2012}; as medidas (x, y) dos CCDs individuais são pré-corrigidos por um padrão de distorção de campo e todas as posições vindas de diferentes CCDs e mosaicos são combinadas usando o polinômio de grau três para produzir uma solução global para cada noite e campo observado, a as posições ($\alpha$, $\delta$) são então obtidas no sistema do catálogo utilizado.
-Estrelas de referência com resíduos de posição (O-C) que desviam mais de 120 mas\footnote{mili-segundos de arco (mili-arcsecond, \textit{mas}, em inglês)} (ou valor similar adotado pelo usuário, usualmente 3 sigma do erro do catálogo) são automaticamente eliminadas uma a uma.
-%
-%Um fator importante do trabalho está na utilidade estatística de se usar uma base de dados com um número extremamente elevado de observações (imagens).
-Esta é a maneira pela qual os efeitos atmosféricos de turbulência e extinção são efetivamente alisados e as informações de posição e de brilho são corretamente extraídas das imagens CCD.
-%
-%O tratamento astrométrico e fotométrico do PRAIA é capaz de lidar com uma imensa quantidade de dados de forma rápida, sendo capaz de obter a qualidade necessária para atingir os objetivos científicos dos programas observacionais em tempo recorde de processamento.
-As implementações incorporadas no software refletem experiência de mais de 20 anos de atuação na área da redução astrométrica e fotométrica de imagens astronômicas digitalizadas.
-%
-%Há ainda o cálculo de magnitudes instrumentais via fotometria de abertura e PSF, calculando-se o ponto zero em relação ao sistema fotométrico do catálogo de referência, como o 2MASS, UCAC2 ou o UCAC4 (o ponto zero pode ser re-definido).
-A magnitude do fundo de céu também é calculada.
-A FWHM das estrelas (seeing) é calculada automaticamente, sendo dados valores individuais por estrela, e valores médios, com as respectivas incertezas.
-%
-%Todas as informações fotométricas e astrométricas da redução, para todos os catálogos, e para cada campo, são guardadas em arquivos de dados.
-Também são gerados arquivos listando as estatísticas gerais de todos os campos.
-Os arquivos são de fácil manuseio nas tarefas de análise numérica e gráfica dos resultados.
-Além das posições obtidas, listam-se os parâmetros Gaussianos dos ajustes de cada estrela, os erros de medida em (x,y) e em posição, número de estrelas de referência, movimentos próprios, época da observação, magnitudes UCAC2, UCAC4, 2MASS, magnitudes instrumentais, informações da observação (filtros usados, tempo de exposição, etc).
-Em particular, são listados em arquivos separados os dados pertinentes aos alvos identificados nos campos, incluindo-se as diferenças de posição entre os ($ \alpha $,$ \delta $) observados reduzidos com UCAC2, UCAC4 e 2MASS e as coordenadas de referência fornecidas previamente pelo usuário em arquivo (efeméride, catálogo, etc.), chamados de offsets.
-Um esquema ilustrativo do PRAIA é mostrado na Fig \ref{Fig:PRAIA}.
-%
-%\begin{figure}[h]
-%\centering
-%\includegraphics[scale=0.4]{figuras/PRAIA.png}
-%\caption[Esquema ilustrativo do pacote PRAIA]{Esquema ilustrativo do pacote PRAIA: a sequência a ser seguida pelo PRAIA está no sentido horário, desde a extração dos dados das imagens, até a identificação dos alvos nos campos CCD e, consequentemente, a obtenção de resultados.}
-%\label{Fig:PRAIA}
-%\end{figure}
-%
-%\chapter{Ocultações Estelares}
-%\label{Cap: ocultacoes}
-%
-%\indent \indent Ocultação estelar é um evento que ocorre quando um objeto do Sistema Solar passa na frente de uma estrela de forma que o brilho da estrela é completamente ou parcialmente bloqueado durante um intervalo de tempo.
-A variação de brilho da estrela pode ser medida fazendo-se observações com exposições curtas.
-O produto gerado é uma curva de fluxo de luz x tempo.
-Essa técnica pode ser utilizada para se obter parâmetros físicos dos objetos ocultantes como tamanho, forma, albedo, densidade (se a massa é conhecida), detecção da presença de atmosfera, satélites, anéis, etc \citep{Elliot1979}.
-%
-%A Fig.
-\ref{Fig: occ-scheme} mostra uma curva de luz esquemática de uma ocultação estelar central por um planeta com atmosfera e anéis.
-Ao ocultar a estrela, os anéis passarão primeiro, a densidade do anel, tamanho e tempo de exposição determinará a queda de brilho da estrela.
-Ao chegar no corpo principal, caso não possua atmosfera, a queda de brilho será abrupta, caso possua, a queda de brilho será mais suave devido à refração da luz da estrela na atmosfera dificultando a determinação da superfície do objeto.
-%
-%\begin{figure}[h]
-%\begin{centering}
-%\includegraphics[scale=0.75]{figuras/stellarocc.png} 
-%\caption{Curva de luz esquemática de uma ocultação estelar por um planeta com atmosfera e anéis. Fonte: \cite{Elliot1979}}
-%\label{Fig: occ-scheme}
-%\end{centering} 
-%\end{figure}
-%
-%A suavidade da curva portanto será determinada pelo gradiente de temperatura, pressão e densidade da atmosfera, presença de nuvens, etc.
-A luz da estrela nunca será completamente apagada quando ocultada por um objeto com atmosfera devido à refração da luz na atmosfera onde parte dela convergirá na direção do observador.
-Caso a ocultação seja central para o observador haverá um aumento de brilho no meio do evento ("flash") pois a convergência da luz será máxima.
-A atmosfera, no caso, agiria como uma lente como mostrado na Fig.
-\ref{Fig: occ-scheme}.
-%
-%Quando o evento é observado por diversos telescópios espalhados, cada observador verá a ocultação em latitudes diferentes.
-Essas diversas curvas de luz permitirão reconstruir a forma do objeto, possivelmente identificando estruturas superficiais como montanhas e crateras.
-Como a estrela está muito distante, podendo ser considerada quase pontual, a sombra do objeto durante a ocultação é muito próxima ao seu tamanho real.
-%
-%Para se observar esses eventos, portanto, é necessário uma campanha grande com observadores espalhados por várias regiões para cobrir a sombra do objeto.
-A precisão nas posições da estrela e do objeto são cruciais pois o evento ocorre em instantes específicos e passará por regiões específicas da Terra alheias à escolha do observador.
-%
-%As curvas de luz são geradas a partir das observações seguindo a técnica da fotometria diferencial.
-Em nosso trabalho usamos as tasks de fotometria diferencial do pacote PRAIA.
-As imagens são previamente calibradas de bias e "flatfield".
-Os alvos e objetos de calibração são marcados em uma imagem de referência (normalmente a primeira da série) e depois automaticamente identificados no resto da sequência.
-Técnicas de cálculo do fundo de céu são empregadas, envolvendo a análise da moda de histogramas, e o fluxo dos objetos e do alvo são calculados.
-Diversas aberturas são testadas até que a menor dispersão seja obtida para a curva (melhor relação sinal-ruído).
-Ao final, obtêm-se as razões de fluxo alvo/calibrador(es), sinal-ruído, erros da razão de fluxo, entre outras informações.
-O instante de tempo a cada ponto da curva também é registrado.
-%
-%Nas seções \ref{Sec: Ceres-2010-occ} e \ref{Sec: Ceres-2013-occ} detalhamos os procedimentos da análise da curva para obtenção dos instantes de imersão e emersão, para o cálculo da dimensão e forma do alvo, no caso de ausência de atmosfera.
-Na presença de atmosfera, técnicas de "ray-tracing" são utilizadas para ajustar a altura e perfis de densidade/temperatura a partir da curva de luz obtida.
-Maiores detalhes desta técnica são dadas em \cite{DiasOliveira2015} e referências.
-
-\chapter{Centauros e TNOs}
-\label{Cap: TNOs}
+%\chapter{Centauros e TNOs}
+%\label{Cap: TNOs}
 
@@
 
-\indent \indent Dentro do contexto explicitado no capítulo \ref{Cap: intro}, os TNOs são uns dos objetos mais significantes por possivelmente serem a fonte de outras populações do Sistema Solar, como por exemplo Centauros \citep{Fernandez2002}, Satélites Irregulares \citep{Nesvorny2007} e Troianos \citep{Morbidelli2005}.
+%\indent \indent Dentro do contexto explicitado no capítulo \ref{Cap: intro}, os TNOs são uns dos objetos mais significantes por possivelmente serem a fonte de outras populações do Sistema Solar, como por exemplo Centauros \citep{Fernandez2002}, Satélites Irregulares \citep{Nesvorny2007} e Troianos \citep{Morbidelli2005}.
 Também existe a possibilidade de que Ceres tenha uma origem como TNO \citep{McKinnon2012}, assim como Tritão \citep{Agnor2006}.
 
-Pouco se conhece sobre Centauros e TNOs, pois nunca foram visitados por sondas, com a exceção de Plutão, e a grande maioria das observações são de origem fotométrica usual (magnitudes, cores) ou, em menor parte, espectroscópica, ou em menor número ainda por telescópios espaciais.
+%Pouco se conhece sobre Centauros e TNOs, pois nunca foram visitados por sondas, com a exceção de Plutão, e a grande maioria das observações são de origem fotométrica usual (magnitudes, cores) ou, em menor parte, espectroscópica, ou em menor número ainda por telescópios espaciais.
 
-Estimar parâmetros físicos como tamanho, albedo, densidade, etc para esses objetos é um desafio, porém essencial para acessar a massa atual e composição (material) da população e recuperar sua história evolutiva.
+%Estimar parâmetros físicos como tamanho, albedo, densidade, etc para esses objetos é um desafio, porém essencial para acessar a massa atual e composição (material) da população e recuperar sua história evolutiva.
 %É possível obter seus tamanhos através de modelos, por exemplo combinando o brilho no visível e emissão térmica obtida em infra-vermelho.
 Todavia os erros obtidos são piores que 10 \% \citep{Camargo2013}.
-Para se obter esses parâmetros, de forma muito precisa e acurada, sem a necessidade de se adotar modelos, utilizamos a técnica de ocultações estelares.
+%Para se obter esses parâmetros, de forma muito precisa e acurada, sem a necessidade de se adotar modelos, utilizamos a técnica de ocultações estelares.
 Até 2009, quando foi observada uma ocultação de 2002TX300 \citep{Elliot2010}, Plutão e Caronte eram os únicos TNOs com ocultações observadas.
@@
 
-\section{Ocultações Estelares}
+\chapter{Ocultações Estelares}
+\label{Cap: Occ}
+
+\indent \indent Uma ocultação estelar ocorre quando um objeto do Sistema Solar passa em frente a uma estrela.
+A sombra causada por esse fenômeno passaria sobre a Terra e um observador que estivesse sob essa sombra veria uma variação no brilho do par objeto-estrela ao longo do tempo.
+A curva de luz oriunda da ocultação pode ser usada para inferir parâmetros físicos como tamanho, albedo, densidade, etc, para o objeto ocultante.
+
+Com várias curvas de luz observadas no mesmo evento e obtidas em diferentes regiões da sombra se pode obter a forma e tamanho do objeto.
+Conhecendo-se a magnitude absoluta, calcula-se o albedo.
+Se a massa é conhecida, calcula-se sua densidade.
+Atmosferas e anéis também podem ser descobertos por esse método.
+
+\section{TNOs}
 \label{Sec: TNO-occ}
 
-\indent \indent Observar uma ocultação exige um grande trabalho de predição e de sua melhoria.
+Até 2009, quando foi observada uma ocultação de 2002TX300 \citep{Elliot2010}, Plutão e Caronte eram os únicos TNOs com ocultações observadas.
+Até hoje, cerca de uma dezena de TNOs e Centauros tiveram ocultações observadas.
+
+Observar uma ocultação estelar exige um grande trabalho de predição e de sua melhoria.
 Os TNOs estão muito distantes significando que mesmo sem grandes erros nas suas efemérides ou na posição da estrela o erro do local por onde a sombra irá passar pode ser de milhares de quilômetros.
@@
 
-%Da mesma forma, estando muito distantes, eles se movem relativamente devagar no céu ocultando relativamente poucas estrelas por ano.
-Porém, sendo menos brilhantes (Mag $V > 18$) é possível observar ocultações de estrelas mais fracas (estrelas até Mag $R = 17$), onde a queda de brilho da estrela durante a ocultação poderá ser detectada e, mesmo com baixo sinal-ruído (ex: 5 a 10), o tamanho do corpo pode ser calculado com precisão de pelo menos 10 a 20 km, melhorando à medida em que a velocidade da sombra seja menor que 20 km/s (velocidade típica para um corpo em oposição a 30 UA).
+Devido ao grande esforço necessário para se prever e observar ocultações estelares (a sombra do objeto pode passar em qualquer lugar na Terra) e o interesse científico que o método proporciona, o nosso grupo conta com colaboração de pesquisadores nacionais e internacionais, profissionais e amadores.
+Utilizamos telescópios de pequeno e grande porte e câmera rápida para observar eventos que proporcionaram grandes descobertas, como por exemplo a descoberta do sistema de anéis ao redor do Centauro Chariklo \citep{BragaRibas2014}, do qual colaborei.
 
-Para identificar candidatas a ocultações estelares foi realizada uma campanha observacional no ESO para observar os caminhos aparentes dos TNOs entre 2009 e 2015.
-Assim, estrelas que não estão no UCAC4, catálogo comumente utilizado como referência astrométrica, foram identificadas bem como estrelas que estão mas não tem posições ou movimentos próprios com precisão suficiente.
-Tanto os catálogos de estrelas quanto as estrelas candidatas a ocultação foram publicados \citep{Assafin2010, Assafin2012, Camargo2013}.
-Esses catálogos são utilizados como representante prático do ICRS no tratamento astrométrico e o conjunto é denominado catálogo WFI.
+Dentro do meu programa de doutorado, participei e participo ativamente na observação e redução astrométrica de TNOs e estrelas a serem ocultadas.
+Essas observações são utilizadas na melhoria das predições dos eventos diminuindo erros na localização na Terra onde a sombra passaria.
+
+Até há pouco tempo um grande esforço era feito para melhorar as posições das estrelas uma vez que o melhor catálogo de estrelas de referência disponível, o UCAC4 \citep{Zacharias2013}, tinha erros maiores do que o necessário para uma boa predição.
+Porém, recentemente tivemos o primeiro release do catálogo GAIA \citep{Brown2016}, que proporcionará melhores posições para as estrelas, na casa dos micro-segundos de arco.
 
 Não somente as posições das estrelas precisam ser atualizadas, mas também as posições dos TNOs.
-Portanto os TNOs também foram observados tanto no ESO quanto no Observatório do Pico dos Dias (OPD, LNA/MCTI, Itajuba, MG., código IAU 874, $45^{\circ} ~34\arcmin ~57\arcsec$ W, $22^{\circ} ~32\arcmin ~04\arcsec$ S, 1864 m).
+Portanto os TNOs também foram observados tanto no ESO quanto no Observatório do Pico dos Dias.
+%(OPD, LNA/MCTI, Itajuba, MG., código IAU 874, $45^{\circ} ~34\arcmin ~57\arcsec$ W, $22^{\circ} ~32\arcmin ~04\arcsec$ S, 1864 m).
 A princípio as predições das ocultações eram atualizadas utilizando o método de offsets, porém, em colaboração com o Dr.
-Josselin Desmars, integrações numéricas das órbitas dos TNOs foram realizadas desenvolvendo-se uma ferramenta denominada NIMA\footnote{Numerical Integration of the Motion of an Asteroid} \citep[aceito]{Desmars2015}.
+Josselin Desmars, integrações numéricas das órbitas dos TNOs foram realizadas desenvolvendo-se uma ferramenta denominada NIMA\footnote{Numerical Integration of the Motion of an Asteroid}, cuja publicação \citep{Desmars2015} sou co-autor.
 
-%Nesse trabalho, posições obtidas de ocultações já realizadas tem um peso muito maior que as restantes, diferentemente do que é feito pelo JPL.
-Da mesma forma, noites em que houveram um número maior de observações que as restantes terão um peso moderado já que se fosse utilizado um peso individual, o peso final de uma noite poderia estar superestimado levando a um erro na integração.
-Neste trabalho contribuí com observações e reduções astrométricas de imagens de TNOs feitas no OPD.
-
-%Com o objetivo de obter os parâmetros físicos desses objetos, para melhor caracterizar as propriedades básicas dessas populações,
 Uma grande campanha observacional é realizada para se observar ocultações estelares por TNOs e Centauros.
-Essa campanha é liderada por pesquisadores brasileiros, franceses e espanhóis e tem colaboração com observadores espalhados por todo o mundo
-%Os objetos foram basicamente selecionados de acordo com o grau de interesse científico (representar uma classe de TNO, ser um  objeto peculiar, etc), tamanho (quanto maior, mais chances de observar a ocultação) e região do céu (quanto mais estrelas, maior  a probabilidade de ocultar estrelas brilhantes, permitindo curvas de luz de melhor resolução espacial).
-resultando em vários objetos já observados.
-% pelo grupo, sendo que o Sistema Plutão-Caronte conta com várias ocultações observadas além de um trabalho astrométrico minucioso baseado em 19 anos de observações no OPD, mais observações no ESO \citep{Rossi2014}.
+Essa campanha é liderada por pesquisadores brasileiros, franceses e espanhóis e tem colaboração com observadores espalhados por todo o mundo resultando em vários objetos já observados.
 As observações bem sucedidas de ocultações estelares resultantes desse trabalho de predição já geraram diversas publicações, inclusive na revista Nature, por exemplo \cite{Sicardy2011} e \cite{BragaRibas2014} nas quais contribuí em caráter observacional.
@@
 
-Além desses contribuí com outras ocultações como a de Plutão observada na Austrália e Nova Zelândia em 29 de Junho de 2015 e a de Chariklo em 29 de Abril de 2014 reduzindo curvas de luz e observei uma ocultação do TNO 2003VS2 dia 07 de Novembro de 2014 no OPD.
-Esses trabalhos ainda não foram publicados.
+Além desses contribuí com outras ocultações como a de Plutão observada na Austrália e Nova Zelândia em 29 de Junho de 2015 e a de Chariklo em 29 de Abril de 2014 reduzindo curvas de luz, observei uma ocultação do TNO 2003VS2 dia 07 de Novembro de 2014 no OPD e trabalhei na predição de eventos como os dos TNOs 2003AZ84 e 2006UK126.
+Alguns desses trabalhos ainda não foram publicados.
 
-%No dia 29 de Junho de 2015 houve uma ocultação de Plutão de máxima importância devido a esta ser a última antes da chegada da sonda \textit{New Horizons} no sistema e foi observada na Austrália e Nova Zelândia.
+\cite{Sicardy2016} obteve os perfis de atmosfera para Plutão a partir do evento do dia 29 de Junho de 2015 Esse evento foi de máxima importância devido a esta ser a última antes da chegada da sonda \textit{New Horizons} no sistema e foi observada na Austrália e Nova Zelândia.
 Os modelos de perfil de atmosfera para Plutão poderão ser calibrados com os dados da \textit{New Horizons} para serem utilizados em futuras ocultações e estudar a evolução da atmosfera do planeta anão.
 
-%Em 29 de Abril de 2014, uma ocultação de Chariklo foi observada na África.
-Eu trabalhei com a redução de três curvas de luz (Gifberb, SAAO e Springbok).
-Um fato interessante dessa ocultação foi que a estrela era uma binária não catalogada.
-Essa ocultação foi importante para obter com total certeza a direção do polo de Chariklo.
-Além disso, as câmeras utilizadas eram de alta qualidade possibilitando obter que pelo menos um dos anéis é opaco.
-Uma das cordas, fortuitamente, passou contígua ao anel observando apenas o anel externo.
-Os resultados ainda não foram publicados.
+\cite{Alex2016}, submetido, identificou no TNO 2003AZ84 uma possível fenda através de uma ocultação que ocorreu em 3 de Fevereiro de 2014.
+Ocasionalmente, uma das cordas do evento foi observada na borda da sombra.
+Esse não é o primeiro objeto com fenda a ser descoberto nessa região.
+Na verdade, a sonda \textit{New Horizons} observou uma fenda de grandes proporções no maior satélite de Plutão, Caronte \citep{Stern2015}.
 
-%Outro evento que observei foi  As variações do céu não permitiram uma boa observação de forma que a curva de luz estava muito ruidosa, mas boa o suficiente próximo ao instante do evento.
-Um trabalho foi feito para identificar alguma ocultação mas nada pôde ser extraído.
-Nesse evento foram obtidas 3 cordas positivas e 7 negativas.
-A redução preliminar revelou que 2003VS2 possui um corpo levemente alongado com aproximadamente 600 km de diâmetro.
-O trabalho ainda está em análise.
+\cite{Rossi2016}, aceito, publicou resultados da ocultação do TNO 2007UK126.
+Essa foi a primeira ocultação observada pelo projeto RECON.
+Esse projeto possui diversos telescópios que foram espalhados pela costa oeste dos Estados Unidos.
+A importância desse grupo se deve à grande área que ele cobre.
+Isso possibilitará que diversos outros eventos possam ser observados com grande qualidade espacial.
 
-\chapter{Ceres}
-\label{Cap: Ceres}
-
-%\section{Introdução}
-%\label{Sec: Ceres-intro}
+\section{Ceres}
+\label{Sub: Ceres}
 
@@
 Na verdade, foi proposto que a origem de Ceres pode ser comum aos objetos trans-Netunianos \citep{McKinnon2012}, espalhado posteriormente para o cinturão principal de asteroides devido à migração dos planetas gigantes predito pelo Modelo de Nice \citep{Gomes2005}.
+
+\cite{DeSanctis2015} reforça essa teoria.
+A partir do observações do espectrômetro VIMS da sonda \textit{Dawn} ele identificou filossilicatos amoniados presentes na superfície de Ceres.
+Como amônia é mais abundante no Sistema Solar Exterior, o trabalho sugere que Ceres ou tem uma origem nessa região ou ele acretou muito material que foi transportado para o Cinturão Principal de Asteroides.
 
@@
 
-%A primeira ocultação estelar por Ceres foi observada em 1984 \citep{Millis1987} e determinou seu tamanho com precisão de alguns quilômetros em uma época que as incertezas eram, normalmente, dez vezes maiores.
-Devido ao brilho aparente de Ceres ser alto, comparado à maioria dos asteroides, estrelas capazes de causar uma queda de magnitude detectável quando ocultadas são limitadas às mais brilhantes.
-Por exemplo, depois do evento de 1984, apenas 4 ocultações estelares por Ceres foram observadas \citep{Dunham2014}.
-Duas delas tiveram apenas duas cordas cada que não foram suficientes para prover resultados acurados\footnote{Esses eventos ocorreram em 22 de Agosto de 1994 e 30 de Outubro de 2010.}.
-Os dois restantes, que ocorreram em 17 de Agosto de 2010 e 25 de Outubro de 2013, foram trabalhos por mim em colaboração com o grupo do Rio \citep{GomesJunior2015-Ceres}.
+Com este fim, foram preditos dois eventos por Steve Preston\footnote{Predições publicadas em \url{http://asteroidoccultation.com}.} para a IOTA (International Occultation Timing Association), durante predições de rotina de ocultações de estrelas brilhantes por asteroides.
+As ocultações desse objeto me renderam uma publicação como primeiro autor, \citep{GomesJunior2015-Ceres}.
 
-Foram preditos dois eventos por Steve Preston\footnote{Predições publicadas em \url{http://asteroidoccultation.com}.} para a IOTA (International Occultation Timing Association), durante predições de rotina de ocultações de estrelas brilhantes por asteroides.% Os caminhos das sombras podem ser visualizados na Fig.
-\ref{Fig: Ceres-map}.
+%Foram preditos dois eventos por Steve Preston\footnote{Predições publicadas em \url{http://asteroidoccultation.com}.} para a IOTA (International Occultation Timing Association), durante predições de rotina de ocultações de estrelas brilhantes por asteroides.
 
-%\begin{figure}[h]
-%\begin{centering}
-%\subfigure[Ocultação de 2010]{\includegraphics[scale=0.39]{figuras/Ceres_2010.eps}\label{Fig: Ceres-map-2010}}
-%\subfigure[Ocultação de 2013]{\includegraphics[scale=0.39]{figuras/Ceres_2013-zoom.eps}\label{Fig: Ceres-map-2013}}
-%\caption{Reconstrução pós-ocultação do caminho da sombra de Ceres na Terra para os eventos de 17 de Agosto de 2010 (a) e 25 de Outubro de 2013 (b).
-Os pontos em azul são os sítios que observaram os eventos.
-a) O ponto grande vermelho é a máxima aproximação geocêntrica às 22:40:25 UT.
-Os pequenos representam o centro da sombra separados por um minuto.
-b) Visão superior da ocultação sobre os sítios que observaram o evento de 25 de Outubro de 2013.
-Os pontos vermelhos são os centros da sombra separados por 15 s.
-%\label{Fig: Ceres-map}}
-%\end{centering}
-%\end{figure}
+O primeiro evento foi observado no Brasil em 17 de Agosto de 2010 em cinco diferentes sítios.
+Destes, 4 obtiveram cordas positivas enquanto UFSC teve corda negativa.
+Das positivas, a observação proveniente do INPE iniciou-se após o início do evento devido a dificuldades técnicas e, portanto, apenas a emersão da curva de luz foi detectada.
+O segundo foi observado no do dia 25 de Outubro de 2013 na costa Leste dos Estados Unidos.
 
-\section{Ocultaç\~oes de 2010 e 2013}
-\label{Sec: Ceres-2010-occ}
-
-\indent \indent Em 17 de Agosto de 2010 Ceres ocultou a estrela TYC 6833-163-1 (UCAC4 313-111823), cuja magnitude é $V = 11.55$ e tem posição no ICRS para a data do evento baseada no catálogo UCAC4 \citep{Zacharias2013}: $\alpha = 17^{h}18^{m}29\fs 0085$  e $\delta = -27\degr 26\arcmin 38\farcs 867$.
-E em 25 de Outubro de 2013, Ceres ocultou a estrela TYC 865-911-1 (UCAC4 496-058191), de magnitude $V = 10.05$.
-Baseada no UCAC4 \citep{Zacharias2013}, sua posição ICRS para a data da ocultação é: $\alpha = 11^{h}57^{m}52\fs7641$ e $\delta = +09\degr 07\arcmin 49\farcs835$.
-
-%\begin{equation}
-%\left\{ 
-%  \begin{array}{l l}
-%    \alpha = 17^{h}18^{m}29\fs 0085\\
-%    \delta = -27\degr 26\arcmin 38\farcs 867
-%  \end{array}
-%\right.
-%\label{Eq: 2010-coord}
-%\end{equation}
-
-O primeiro evento foi observado no Brasil a partir de cinco diferentes sítios.
-Destes, 4 obtiveram cordas positivas enquanto UFSC teve uma corda negativa.
-Das positivas, a observação proveniente do INPE iniciou-se após o início do evento devido a dificuldades técnicas e, portanto, apenas a emersão da curva de luz foi detectada.
-O segundo foi observado na costa Leste dos Estados Unidos.
-
-%Uma das características mais importantes desse evento foi a velocidade com que ocorreu (apenas 3.9 km s$^{-1}$) acarretando que mesmo exposições de poucos segundos representariam resoluções espaciais significantes.
-
-%Todas as observações foram feitas com a utilização de CCDs.
 As curvas de luz de cada observação foram obtidas das imagens FITS com a utilização do pacote PRAIA \citep[Plataforma de Redução Astrométrica de Imagens Astronômicas,][]{Assafin2011}.
-%As curvas foram normalizadas para o fluxo da estrela mais Ceres, uma vez que eles estavam indistinguíveis logo antes e depois da ocultação.
-Por fim, elas foram normalizadas pelo ajuste de uma curva polinomial (de primeira ou segunda ordem) fora da queda de fluxo assim fixando em 1 a razão de fluxo fora da ocultação.
+O evento de 2010 foi reduzido pelo Felipe Braga Ribas enquanto eu reduzi o evento de 2013.
 Os instantes de ingresso e egresso foram obtidas de cada curva de luz ajustando-se um modelo de poço quadrado levando em consideração a difração de Fresnel, a banda do CCD, o diâmetro aparente da estrela e o tempo de exposição utilizado \citep[ver][]{Widemann2009, Braga-Ribas2013}.
 Algumas cordas do evento de 2013 não foram utilizadas por haver erros nos tempos fornecidos identificados durante o tratamento das imagens.
-
-%O ajuste dos dados da ocultação consiste em minimizar uma função de $\chi^{2}$ clássica para cada curva de luz, como descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}.
-Os parâmetros livres para ajustar são os instantes de ingresso e egresso que fornece o valor mínimo de $\chi^{2}$ ($\chi^{2}_{min}$).
-%O melhor ajuste das curvas de luz para as ocultações está mostrado na Fig.
-\ref{Fig: Ceres-curves}.
-
-%\begin{figure}
-%\begin{centering}
-%\subfigure[Ocultação de 2010]{\includegraphics[scale=0.51]{figuras/Ceres_2010_fluxratio.eps}\label{Fig: Ceres-2010-curves} }
-%\subfigure[Ocultação de 2013]{\includegraphics[scale=0.51]{figuras/Ceres_2013_fluxratio.eps}\label{Fig: Ceres-2013-curves} }
-%\caption{Curvas de luz normalizadas das cordas positivas dos eventos.
-As curvas estão desviadas por um fator de 0.2 (a) e 1.0 (b) para melhor visualização.
-As linhas pretas são os melhores ajustes com o modelo de poço quadrado.
-As linhas vermelhas são os melhores ajustes com o modelo de poço quadrado, porém levando em conta a difração de Fresnel, o diâmetro da estrela e o tempo de exposição.
-Os instantes médios de cada curva não coincidem devido às diferentes longitudes dos sítios.
-A curva de luz de Brookline (b) está desviada por um fator de -64 s como explicado no texto.
-%\label{Fig: Ceres-curves}}
-%\end{centering}
-%\end{figure}
-
-A metodologia usada para analisar o perfil de Ceres a partir das observações é o mesmo descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}.
-%Cada combinação de posição do sítio, instantes de ingresso e egresso, junto com as coordenadas da estrela e as efemérides de Ceres, correspondem a um ponto no plano do céu.
-A coleção de todos esses pontos determina o limbo aparente de Ceres.
-%Adotamos um modelo elíptico para o perfil do limbo, resultante da projeção de um esferoide com achatamento nos polos no plano do céu.
-Essa escolha é suportada pelo trabalho de \cite{Drummond2014}, por meio de imagem direta de Ceres.
-Dessa forma, nós temos $N=7$ extremidades das cordas para ajustar $M=5$ parâmetros que definem uma elipse: semi-eixo maior e semi-eixo menor aparentes ($a^\prime$ and $b^\prime$, respectivamente), ângulo de posição $P$ do seu semi-eixo maior e as posição $(f_c,g_c)$ do seu centro com respeito à estrela ocultada.
-O semi-eixo maior {$a^\prime$} é equivalente ao raio equatorial $R_{equa}$ do elipsoide.
-%As coordenadas $f_{c}$ e $g_{c}$, em quilômetros, foram calculadas usando a efeméride de Ceres JPL\#33 \citep{Giorgini1996} e a posição da estrela ocultada.
-Elas são positivas na direção Leste e Norte celestes, respectivamente.
-O ângulo de posição $P$ é contado positivamente a partir do norte celeste local em direção ao leste celeste.
-O achatamento aparente pode ser definido por $\epsilon^\prime = 1 - (b^\prime/a^\prime)$.
-O melhor ajuste é obtido minimizando uma função de $\chi^{2}_{r}$ reduzido, onde definimos o número de graus de liberdade do problema como $\mathcal{N} \equiv N - M$.
-Todos os procedimentos que permitem a determinação das barras de erro dos parâmetros físicos podem ser encontradas em \cite{Braga-Ribas2013}.
 
@@
 
-%Para o evento de 2010, a solução nominal teve como melhor ajuste $\chi^2_{r,min} = 0.24$, que podem ser interpretadas como as barras de erro estarem superestimadas com respeito à boa qualidade do ajuste.
-Porém, como o problema tem somente dois graus de liberdade, $\chi^2_{r,min}$ relativamente pequenos são aceitáveis.
-Os resultados obtidos para o diâmetro equatorial, achatamento, ângulo de posição e coordenadas do centro são apresentadas na segunda coluna da tabela \ref{Tab: resultados}.
-As soluções para cada evento estão esquematizadas na Fig.
-\ref{Fig: Ceres-limb}
-
-\begin{table}[h]
- \begin{centering}
-  \caption{Resultados do ajuste de limbo de Ceres com os dados dos eventos de 2010 e 2013.\label{Tab: resultados}}
-  \begin{tabular}{@{}lcccc}
-  \hline
-     Solução & 2010/Nominal & \textbf{2010/Polo fixo} & 2013/Nominal & 2013/Polo fixo \\
-\hline
-Diam. equat.  (km) & 982 $\pm$ 14 & \textbf{972 $\pm$ 6}  & 971 $\pm$ 7  & 971 $\pm$ 7\\
-%Equatorial radius (km) & 491 $\pm$ 7  & \textbf{486 $\pm$ 3}  & 485.5 $\pm$ 3.5 & 485.5 $\pm$ 3.5\\
-Achatamento        & 0.08 $\pm$ 0.03 & \textbf{0.08 $\pm$ 0.03} & 0.08 $\pm$ 0.04 & 0.08 $\pm$ 0.04\\
-Âng. de pos. (deg)   & 5 $\pm$ 10    & \textbf{12 $\pm$ 3} (*)& 22 $\pm$ 5    & 25 $\pm$ 3 (*)\\
-$f_c$ (km)             & 97 $\pm$ 9   & \textbf{102 $\pm$ 5}   & 77 $\pm$ 6    & 78 $\pm$ 6\\
-$g_c$ (km)             & 16 $\pm$ 15  & \textbf{21 $\pm$ 11}  & 13 $\pm$ 16   & 13 $\pm$ 16\\
-$\chi^2_{r,min}$       & 0.24          &  \textbf{0.42}         & 1.27          & 1.27\\
-\hline
-\end{tabular}
-\textbf{Notas:} Em negrito destaca-se a melhor solução obtida.
-As barras de erro estão no nível de 1$\sigma$.
-O diâmetro polar ($D_{pol}$) pode ser facilmente calculado a partir de $D_{pol}=D_{equa}(1 - \epsilon)$.
-(*) Ângulos de posição derivados a partir das coordenadas do polo de Ceres determinadas por \cite{Drummond2014}.
-\end{centering}
-\end{table}
-
-%Como pode ser visto, o parâmetro com a maior incerteza é o ângulo de posição cobrindo um intervalo de 20$\degr$.
-Claramente, fixar as coordenadas do pólo pode melhorar a solução.
-Por fim, a correção do achatamento devido ao ângulo do aspecto polar está dentro da barra de erro 1$\sigma$ e não tem relevância estatística, dessa forma $\epsilon = 0.08 \pm 0.03$.
-
-%No momento da ocultação, as coordenadas do polo de Ceres correspondiam a um ângulo de posição $P = (12 \pm 3)\degr$.
-Explorar o espaço de parâmetros restringindo a elipses com ângulo de posição dentro deste intervalo resulta na solução de pólo fixo.
-O parâmetros físicos do melhor ajuste estão mostrados na Tab.
-\ref{Tab: resultados} enquanto a solução está esquematizada na Fig.
-\ref{Fig: Ceres-limb-2010}.
-
-%Essa solução corresponde ao limite superior da barra de erro 1$\sigma$ da solução nominal para $P$.
-Por outro lado, ela obtém os menores valores para o diâmetro equatorial, melhorando sua determinação por um fator de 2.
-
-%\begin{figure}[h]
-%\begin{centering}
-%\subfigure[Ocultação de 2010]{\includegraphics[scale=0.35]{figuras/Ceres_2010_sphere.eps}\label{Fig: Ceres-limb-2010} }
-%\subfigure[Ocultação de 2013]{\includegraphics[scale=0.35]{figuras/Ceres_2013_sphere.pdf}\label{Fig: Ceres-limb-2013} }
-%\caption{Melhores ajustes elípticos para as cordas das ocultações de 2010 e 2013.
-As setas indicam a direção de movimento, as linhas azuis são as cordas observadas, os segmentos vermelhos são as barras de erro dos ingressos, egressos e centro da ocultação em $1\sigma$.
-A linha verde em (a) é uma corda negativa.
-Os instantes marcados em verde em (b) não foram utilizados para o ajuste, como descrito no texto.
-%\label{Fig: Ceres-limb}}
-%\end{centering}
-%\end{figure}
-
-%\section{Ocultação de 2013}
-%\label{Sec: Ceres-2013-occ}
-
-%\indent \indent Em 25 de Outubro de 2013, Ceres ocultou a estrela TYC 865-911-1 (UCAC4 496-058191), de magnitude $V = 10.05$.
-Baseada no UCAC4 \citep{Zacharias2013}, sua posição ICRS para a data da ocultação é:
-%
-%%
-%\begin{equation}
-%\left\{ 
-%  \begin{array}{l l}
-%    \alpha = 11^{h}57^{m}52\fs7641\\
-%    \delta = +09\degr 07\arcmin 49\farcs835
-%  \end{array}
-%\right.
-%\end{equation}
-%%
-%Esse evento foi observado na costa Leste dos Estados Unidos logo antes do amanhecer, como mostrado na Fig. \ref{Fig: Ceres-map-2013}.
-%
-%Nove cordas positivas foram obtidas em vários sítios (ver Fig.
-\ref{Fig: Ceres-map-2013}).
-Cada estação foi equipada com uma câmera de vídeo com tempo de leitura desprezível.
-Isso é particularmente importante, já que a velocidade da sombra de Ceres para esse evento foi de 42.6 km s$^{-1}$, muito mais rápido que o evento de 2010.
-%
-%Durante o evento Ceres estava muito baixo no céu com alturas entre $15\degr$ (Winchester) e $20\degr$ (Hampton).
-Forte cintilação era esperada e, combinada com o curto tempo de integração e baixa diminuição de brilho, resultou em curvas de luz ruidosas e assim grandes incertezas nos instantes de imersão e emersão.
-%
-%Todos os vídeos foram convertidos para imagens FITS e a fotometria foi obtida via \textsc{praia} \cite{Assafin2011}.
-As curvas de luz foram normalizadas por uma estrela de referência quando havia uma estrela no campo.
-%
-%Para reduzir o ruído, os dados foram binados por grupos de cinco imagens -- com exceção de Greenbelt, onde grupos de dez imagens foram utilizadas.
-Esse procedimento aumenta o tempo de integração efetivo reportado por um fator de 5 (ou 10).
-Da mesma forma que para o evento de 2010, uma normalização adicional por um polinômio foi aplicada.
-%
-%Os instantes de ingresso e egresso da ocultação foram obtidos pelo mesmo procedimento descrito na seção \ref{Sec: Ceres-2010-occ}.
-Uma vez que o tempo de integração efetivo usado (0.17 s) representa aproximadamente 7 km no plano do céu, e a escala de Fresnel e  o diâmetro da estrela estão novamente na ordem de centenas de metros, o erro da determinação dos instantes de imersão e emersão são dominados principalmente pelo tempo de integração, da mesma forma que para o evento de 2010.
-Os melhores ajustes para as curvas de luz da ocultação estão mostrados na Fig.
-\ref{Fig: Ceres-2013-curves}.
-%
-%Uma comparação entre os tempos obtidos pela estação de Brookline e o restante mostrou que Brookline têm um atraso de aproximadamente 64 s.
-Dessa forma, não utilizamos os tempos de Brookline na análise.
-%
-%Perfis elípticos foram ajustados para todas as cordas restantes.
-pelo mesmo procedimento descrito na seção \ref{Sec: Ceres-2010-occ}.
-O resultado foi $\chi^2_{r,min} = 13$ , sugerindo que um modelo elíptico não é satisfatório para os dados.
-De fato, ao olharmos para a Fig.
-\ref{Fig: Ceres-limb-2013} vemos que a corda de Varina adiantada com respeito às outras.
-Como nessa estação o tempo não foi inserido diretamente nos frames do vídeo é possível que essa diferença seja oriunda de um eventual problema da correspondência entre os tempos do camcorder e do GPS.
-%
-%A imersão gravada em Owings também parece atrasada com respeito às cordas próximas (ver Fig.
-\ref{Fig: Ceres-limb-2013}).
-Essa corda tem aproximadamente o mesmo tamanho da corda de Mechanicsville, apesar de estarem separadas por cerca de 100 km.
-Diferentemente de Varina, essa estação teve os tempos inseridos em cada frame do vídeo o que torna mais difícil justificar um problema de tempo.
-Outras possibilidades seriam uma má determinação dos instantes de ingresso e egresso dessa curva ou uma característica do relevo de Ceres.
-%
-%Em um segundo ajuste, não consideramos as cordas de Brookline, Varina e Owings.
-O ajuste dos cinco parâmetros que definem uma elipse para os doze contatos resultou em $\chi^2_{r,min} = 1.27$, indicando  que está em bom acordo com os dados observados dentro das barras de erro.
-Essa é a solução mostrada na Fig.
-\ref{Fig: Ceres-limb-2013}, onde podemos ver que o tamanho da corda de Brookline é compatível com o modelo.
-%
-%Os resultados das soluções nominais e de polo fixo não são significantemente diferentes.
-Isso se deve ao fato da barra de erro do ângulo de posição da solução nominal (que é muito menor que o da solução nominal de 2010) ser muito similar à barra de erro do ângulo de posição da solução de polo fixo ($P = (25 \pm 3)\degr$).
-Os dois resultados estão apresentados nas colunas 4 e 5 da tabela \ref{Tab: resultados}.
-%
-%Sobre a hipótese do atraso observado na imersão da curva de luz de Owings ser associado a uma característica topográfica, o contato gravado corresponderia a uma elevação negativa de 31 $\pm$ 4 km com respeito à elipse de melhor ajuste.
-Porém modelos teóricos preveem que relevos em Ceres não devem ser maiores que 10--20 km \citep{Johnson1973}, enquanto dados observacionais limitam o contorno em 18 km \citep{Carry2008}.
-Imagens mais recentes da sonda \textit{Dawn} também revelam uma superfície mais suave.
-Portanto, a associação da imersão em Owings com um relevo é improvável.
-
-\section{Discussão}
-\label{Sec: Ceres-discussao}
-
-\indent \indent Os resultados apresentados na tabela \ref{Tab: resultados} mostram um acordo entre os parâmetros físicos obtidos nas duas ocultações, especialmente no diâmetro equatorial.
-As diferenças ocorrem basicamente nos tamanhos das barras de erro e podem ser justificadas pelas particularidades de cada conjunto de dados.
-
-%O evento de 2010, por exemplo, teve somente sete contatos, porém bem distribuídos sobre o disco de Ceres.
+O evento de 2010, por exemplo, teve somente sete contatos, porém bem distribuídos sobre o disco de Ceres.
 Por outro lado, o evento de 2013 teve cinco contatos a mais, todavia concentrados em certas regiões do corpo.
 Em particular, a ausência de cordas próximas ao polo sul fez seu achatamento ser pior determinado para o evento de 2013 que para o evento de 2010.
-Mesmo nossa melhor medida de achatamento, $\epsilon=0.08 \pm 0.03$, tem alta incerteza se comparado com outros valores publicados na literatura, como mostra a tabela \ref{Tab: Ceres-final}.
+Mesmo nossa melhor medida de achatamento, $\epsilon=0.08 \pm 0.03$, tem alta incerteza se comparado com outros valores publicados na literatura.
 
-Como foi mostrado, usar as coordenadas do polo de Ceres determinadas por \cite{Drummond2014} para limitar o ângulo de posição não foi um procedimento eficiente para o evento de 2013.
-Por outro lado, reduziu as barras de erro dos outros parâmetros (com exceção do achatamento) para a ocultação de 2010.% Por fim, esse procedimento resultou em um excelente acordo entre os raios equatoriais obtidos para ambos os eventos.
+Usar as coordenadas do polo de Ceres determinadas por \cite{Drummond2014} para limitar o ângulo de posição não foi um procedimento eficiente para o evento de 2013.
+Por outro lado, reduziu as barras de erro dos outros parâmetros (com exceção do achatamento) para a ocultação de 2010.
+Por fim, esse procedimento resultou em um excelente acordo entre os raios equatoriais obtidos para ambos os eventos.
 
 Uma comparação do diâmetro equatorial de Ceres medido por diferentes técnicas mostra um acordo entre os nossos resultados e aqueles obtidos por imageamente direto do Hubble Space Telescope (HST) \citep{Thomas2005}, do Keck Observatory e do ESO VLT \citep{Drummond2014}.
-O valor diferente e menor reportado por \cite{Carry2008} pode ser justificado pelo fato desse estudo não levar em conta o efeito de escurecimento de bordo.
 
-%O evento de 1984 \citep{Millis1987} é a única outra ocultação com a qual podemos comparar nossos resultados.
-As medidas do diâmetro não estão de acordo dentro de um fator de 2$\sigma$.
-É difícil dizer com certeza as razões dessa divergência.
-Uma forma de clarificar o problema seria redeterminar os instantes de imersão e emersão das curvas de luz originais usando a mesma metodologia aplicada nesse trabalho.
-Infelizmente, não temos acesso aos dados da curva de luz original do evento de 1984.
+A sonda da NASA \textit{Dawn} chegou e está obtendo dados de Ceres desde o começo de 2015 \citep{Russell2016}.
+Seus resultados poderão elucidar muitas perguntas sobre Ceres e serão utilizados para calibrar todas as técnicas usadas até agora para o estudo das propriedades físicas dos pequenos objetos do Sistema Solar, como as ocultações estelares.
 
-A sonda da NASA \textit{Dawn} poderá elucidar muitas perguntas sobre Ceres e seus resultados serão utilizados para calibrar todas as técnicas usadas até agora para o estudo das propriedades físicas dos pequenos objetos do Sistema Solar, como as ocultações estelares.
+O resultado deste trabalho foi publicado em \cite{GomesJunior2015-Ceres}.
 
-O resultado deste trabalho está foi publicado em \cite{GomesJunior2015-Ceres}
-
-%\begin{table}
-%  \caption{Diâmetro equatorial e achatamento de Ceres \label{Tab: Ceres-final}}
-%  \begin{centering}
-%  \begin{tabular}{@{}lccc}
-%  \hline
-%     Diâmetro Equatorial (km) & Achatamento & Método & Ref. \\
-%\hline
-%$972 \pm 6$  & $0.08  \pm 0.03$  & Ocultação & 1\\
-%$967 \pm 10$ & $0.078 \pm 0.015$ & Keck+VTL    & 2\\
-%$959 \pm 5$  & $0.074 \pm 0.007$ & Keck        & 3\\
-%$975 \pm 4$  & $0.067 \pm 0.005$ & HST         & 4\\
-%$959 \pm 5$  & $0.05  \pm 0.01$  & Ocultação & 5\\
-%\hline
-%\end{tabular}
-%\par\end{centering}
-%\textbf{Referências.} 1: \cite{GomesJunior2015-Ceres}. 2: \cite{Drummond2014}. 3: \cite{Carry2008}. 4: \cite{Thomas2005}. 5: \cite{Millis1987}.
-%\end{table}
-
+%\newpage
 \chapter{Satélites Irregulares dos planetas gigantes}
@@
 
-\indent \indent Os satélites irregulares dos planetas gigantes são menores que os regulares, possuindo órbitas mais excêntricas, inclinadas e distantes.
+\indent \indent Os satélites irregulares, também denominados satélites externos, dos planetas gigantes são menores que os regulares, possuindo órbitas mais excêntricas, inclinadas e distantes.
 Na maioria dos casos, essas órbitas são retrógradas.
@@
 
-Por serem pequenos, eles são pouco brilhantes e só foram descobertos no último século.
+Por serem relativamente pequenos, eles são pouco brilhantes e só foram descobertos no último século.
 Dentre os satélite irregulares dos planetas gigantes, poucos são aqueles que possuem medidas de seus parâmetros físicos.
 Apenas Himalia, Phoebe e Nereida foram observados por sondas, apesar de não serem medidas ideais, já que foram alvos de oportunidade.
-Destes apenas Phoebe teve seu tamanho estimado a partir de imagens de alta resolução da Cassini com um erro médio de 0.7 km.%A sonda \textit{Cassini} observou Himalia em 2000 ao passar próximo a Júpiter e obteve o tamanho de Himalia com um erro médio de 10 km \citep{Porco2003}.
-Em 2004, a \textit{Cassini}, aproximando-se de Saturno, observou Phoebe em alta resolução obtendo um erro médio na medida de seu tamanho de 0.7 km \citep{Thomas2010}.
-Por fim, Nereida foi observado em 1989 pela sonda \textit{Voyager II} e seu tamanho foi obtido com um erro de 25 km \citep{Smith1989}.
-Para outros satélites irregulares de Júpiter, seus tamanhos foram estimados por \cite{Rettig2001} impondo um albedo de 4\% com a justificativa de que esse valor é um albedo nominalmente utilizado para objetos do Sistema Solar Externo.
-
-%Existem alguns mecanismos de captura de objetos por planetas gigantes propostos na literatura.
-Há o Arrasto Gasoso na nebulosa circumplanetária primordial \citep{Sheppard2005} onde o movimento do objeto seria afetado pelo gás e sua velocidade diminuiria até que fosse capturado pelo planeta.
-Outro mecanismo se chama captura pull-down \citep{Sheppard2005}, onde a massa do planeta aumentaria enquanto o objeto estivesse temporariamente capturado.
-
-%Um mecanismo baseado no modelo de Nice \citep{Morbidelli2005, Tsiganis2005, Gomes2005} foi proposto por \cite{Nesvorny2007} e, especificamente para Júpiter, por \cite{Nesvorny2014}.
-Durante a instabilidade do Sistema Solar primordial, ocorreram encontros entre os planetas externos.
-Nesses encontros planetários poderia haver troca de energia e momento angular entre os planetas e os objetos próximos tornando possível a captura de um corpo irregular por um planeta gigante.
-Nesse cenário, a taxa de sobrevivência de satélites anteriores ao LHB (Bombardeamento Pesado Tardio\footnote{"Late Heavy Bombardment", em inglês}) é muito pequena.
-
-%Um outro mecanismo importante é a captura através de interações colisionais \citep{Sheppard2005}.
-Uma colisão entre dois corpos pequenos dentro da esfera de Hill do planeta poderia gerar objetos fragmentados e a energia dissipativa poderia ser tal que alguns desses objetos seriam capturados.
-
-%Alguns desses objetos formam grupos dinâmicos com elementos orbitais semelhantes, chamados famílias, similar às famílias encontradas no Cinturão Principal de Asteroides.
-Essas famílias podem ter sido criadas por um corpo pai que se fragmentou por colisões com cometas ou outros asteroides \citep{Nesvorny2004}.
-Colisões com cometas tem uma probabilidade maior de ocorrer durante o Bombardeamento Pesado Tardio (LHB) \citep{Gomes2005}.
-
-%\cite{Nesvorny2003} estudou as taxas de colisões entre satélites irregulares e concluiu que alguns satélites podem ter sido removidos por colisão com um satélite maior.
-A taxa de colisão entre os satélites do grupo de Himalia (Himalia, Elara, Lysithea and Leda, principalmente), por exemplo, foi encontrado como sendo maior que 1 durante a idade do Sistema Solar sugerindo que sua estrutura atual foi originada por colisão satélite-satélite.
-
-%Para Phoebe, materiais ejetados de sua superfície causadas por impactos poderiam evoluir devido ao arrasto de Poynting-Robertson e colidir com Iapetus causando a grande variação de albedo observada no satélite \citep{Nesvorny2003}.
-De fato, a sonda \textit{Cassini} detectou em Phoebe uma característica de absorção em 2.42 $\mu m$, provavelmente combinações de CN, que foi também detectada no lado escuro de Iapetus \citep{Clark2005}
+Destes apenas Phoebe teve seu tamanho estimado a partir de imagens de alta resolução da Cassini com um erro médio de 0.7 km.
 
@@
 
-Nesse contexto, durante o Mestrado, foi realizado junto ao grupo um trabalho de caráter astrométrico para a melhoria das efemérides dos satélites irregulares de Júpiter e Saturno.
+Nesse contexto, durante o Mestrado, foi realizado junto ao grupo um trabalho astrométrico para a melhoria das efemérides dos satélites irregulares de Júpiter e Saturno.
 Em colaboração com o Dr.
@@
 
-%\begin{table}[h]
-%\caption{Comparação entre o número de posições obtidas em \cite{GomesJunior2015-Irregular} e o número utilizado nas integrações numéricas das órbitas pelo JPL como publicado por  \cite{Jacobson2012}.
-Os satélites estão separados por planeta (linha reta) e por família orbital (linha tracejada).}
-%\label{Tab: comparison-horizons}
-%\begin{centering}
-%\begin{tabular}{lccccccc}
-%\hline  \hline
-%Satélite & Diâm. (km)\footnote{Planetary Satellite Physical Parameters - JPL: \url{http://ssd.jpl.nasa.gov/?sat_phys_par}} & Mag V  & OPD  & OHP & ESO & Total  & Jacobson \tabularnewline
-%\hline
-%Himalia & 170 & 14 & 854 & 357 & 23 & 1234 & 1757 \tabularnewline
-%Elara & 86 & 16 & 403 & 187 & 46 & 636 & 1115 \tabularnewline
-%Lysithea & 36 & 18 & 60 & 84 & 90 & 234 & 431 \tabularnewline
-%Leda & 20 & 19 & 6 & 48 & 44 & 98 & 178 \tabularnewline
-%\hdashline
-%Pasiphae & 60 & 17 & 295 & 248 & 66 & 609 & 1629 \tabularnewline
-%Callirrhoe & 9 & 21 & 9 & -  &  16 & 25 & 95 \tabularnewline
-%Megaclite & 5 & 22 & - & -  &  10 & 10 & 50  \tabularnewline
-%\hdashline
-%Ananke & 28 & 18 & 52 & 141 & 57 & 250 & 600 \tabularnewline
-%Praxidike & 7 & 21 & - & -  &   2 & 2 & 59 \tabularnewline
-%\hdashline
-%Carme & 46 & 18 & 90 & 204 & 37 & 331 & 973 \tabularnewline
-%Sinope & 38 & 18 & 41 & 169 & 11 & 221 & 854 \tabularnewline
-%Themisto & 8 & 21 & - & - & 16 & 16 & 55 \tabularnewline
-%\hline
-%Phoebe & 213 & 16 & 1239 & 516 & 32 & 1787 & 3479 \tabularnewline
-%\hdashline
-%Siarnaq & 40 & 20 & - & 20 & 56 & 76 & 239 \tabularnewline
-%Paaliaq & 22 & 21 & - & - & 11 & 11 & 82 \tabularnewline
-%\hdashline
-%Albiorix & 32 & 20 & - & - & 46 & 46 & 137 \tabularnewline
-%\hline
-%Sycorax & 150 & 21 & - & - & 35 & 35 & 237 \tabularnewline
-%\hline
-%Nereid & 340 & 19 & 803 & - & 99 & 902 & 716 \tabularnewline
-%\hline
-%\end{tabular}
-%\par\end{centering}
-%\end{table}
-
 Observando os offsets das posições com respeito às efemérides ao longo do tempo e anomalia verdadeira foram identificados erros sistemáticos nas efemérides.
 O erro mais evidente é na declinação do satélite Carme o qual foi associado a um erro em sua inclinação orbital.
-%\begin{figure}[h]
-%\begin{centering}
-%\subfigure[Right Ascension]{\includegraphics[scale=0.45]{figuras/Carme_RA}\label{Fig: carme_alfa}}
-%\subfigure[Declination]{\includegraphics[scale=0.45]{figuras/Carme_DEC}\label{Fig: carme_delta}}
-%\caption{Offsets médios das efemérides e dispersões (barra de erro 1$\sigma$) nas coordenadas de Carme tomadas noite a noite por anomalia verdadeira para cada telescópio.
-O quadrado vermelho são as observações com o telescópio do OPD Perkin-Elmer, o círculo azul para o Boller \& Chivens também do OPD, o triângulo pra cima preto para o OHP e a estrela verde para o ESO.
-%\label{Fig: carme_anom}}
-%\end{centering}
-%\end{figure}
 Esse padrão em declinação também foi identificado para outros satélites como Pasiphae e Ananke.
@@
 
-%Um método que tem se mostrado eficiente para a obtenção dos parâmetros físicos (tamanho, forma, albedo, etc) desses objetos é o método de ocultações estelares, que proporciona medidas tão precisas que são apenas superadas por medidas oriundas de sondas.
-Uma vez que suas efemérides não são muito precisas dado os seus tamanhos estimados, predizer e observar ocultações estelares é muito difícil e nenhuma observação de um evento desses para um satélite irregular ainda foi relatado na literatura.
-
 \section{Predição das ocultações}
@@
 
-\indent \indent Com o objetivo de obter parâmetros físicos para os satélites irregulares com maior precisão foram feitas predições de ocultações estelares para os 7 maiores satélites de Júpiter (Himalia, Elara, Pasiphae, Sinope, Lysithea, Carme, Ananke), o satélite Phoebe de Saturno e Tritão e Nereida de Netuno.
-Esses objetos são pequenos, se comparados aos TNOs estudados no capítulo \ref{Cap: TNOs}, sendo que o menor dos satélites da amostra, Ananke, possui um diâmetro estimado de 28 km.
+\indent \indent Com o objetivo de obter parâmetros físicos para os satélites irregulares com maior precisão foram feitas predições de ocultações estelares para os 8 maiores satélites de Júpiter (Himalia, Elara, Pasiphae, Sinope, Lysithea, Carme, Ananke e Leda) e o satélite Phoebe de Saturno.
+Esses objetos são pequenos, se comparados aos TNOs estudados na seção \ref{Sec: TNO-occ}, sendo que o menor dos satélites da amostra, Leda, possui um diâmetro estimado de 20 km.
 
 As predições foram feitas utilizando as posições de estrelas dadas no catálogo UCAC4.
-Após o lançamento do catálogo GAIA novas predições serão feitas.
 Para as posições dos satélites de Júpiter foi feita uma nova integração numérica utilizando as posições publicadas em \cite{GomesJunior2015-Irregular}.
-Essa integração foi realizada pela pós-doc Laurène Beauvalet.
-Para Phoebe, Tritão e Nereida foram utilizadas efemérides do JPL.
+Essa integração foi realizada pela pós-doc do grupo no ON Laurène Beauvalet.
+Outras posições publicadas na literatura não foram utilizadas.
+Isso se deve por não estarmos interessados em efemérides a longo prazo uma vez que só faremos predições até o fim de 2020.
+Além disso, a confiança que temos nos nossos dados é muito maior, e analisar as posições de outros observadores exigiria muito tempo e esforço fora do escopo do trabalho.
 
-Aproximadamente 600 eventos foram obtidos entre 01 de Janeiro de 2015 e 31 de Dezembro de 2017 para os 10 objetos sendo que a maioria será descartada por passar em regiões de oceano ou onde não há observadores disponíveis.
+Para Phoebe atualizamos as efemérides publicadas por \cite{Desmars2013} adicionando as posições de \cite{GomesJunior2015-Irregular}, \cite{Peng2015}, MPC e Flagstaff.
+Um total de 5886 posições foram utilizadas.
+Isso representou um acréscimo de cerca de 75\% em comparação à \cite{Desmars2013}, a maioria com observações recentes, o que é requerido ao nosso propósito.
+
+Aproximadamente 5442 eventos foram obtidos entre 01 de Janeiro de 2016 e 31 de Dezembro de 2020 para os 9 objetos.
+A maioria será descartada por passar em regiões de oceano ou onde não há observadores disponíveis.
 Estima-se que restem ao final cerca de 10\% desse número.
-Para Tritão e Nereida% , apenas uma estrela UCAC4 foi identificada para ser ocultada.
-Nesse caso,
-utilizamos um catálogo de estrelas feito a partir do caminho aparente do satélite de Netuno, Tritão, no céu para predições de ocultações estelares pelo mesmo.
 
-%Como esses objetos são muito pequenos, em sua maioria as ocultações durarão poucos segundos, portanto apenas eventos com estrelas brilhantes serão selecionados e se houver câmeras de integração rápida disponíveis.
+A razão desse grande número de eventos se deve ao fato que Júpiter, em seu caminho aparente no céu, irá passar pelo plano da galáxia em 2019-2020 e Saturno em 2018.
+Devido à grande densidade de estrelas nessa região, o número de eventos por ano é cerca de 15 vezes maior do que quando fora dessa região.
+Por exemplo Carme, com 20 ocultações previstas em 2016, ocultará 369 estrelas em 2019.
+
+Como esses objetos são muito pequenos, em sua maioria as ocultações durarão poucos segundos, portanto apenas eventos com estrelas brilhantes serão selecionados e se houver câmeras de integração rápida disponíveis.
 Por outro lado, os satélites de Júpiter estão muito mais perto que os TNOs e como o erro astrométrico é uma medida angular, consequentemente, o erro da sombra da ocultação projetada na Terra será muito menor que para TNOs.
 Assim, as latitudes a serem cobertas para que uma ocultação por satélite irregular de Júpiter seja detectada corresponde a poucas centenas de quilômetros.
-
-%\section{Testes}
-%\label{Sec: Irr-testes}
-
-%\indent \indent Até que o primeiro catálogo GAIA seja lançado, as posições das estrelas usadas serão as do catálogo UCAC4.
-Porém esse catálogo possui imprecisões relevantes principalmente para as estrelas mais fracas.
-Uma das estrelas testadas apresentou um offset da posição do catálogo para a data da observação de mais de 600 mas.
-Para a estrela em questão, esse valor representa uma variação de aproximadamente 1900 km da sombra projetada na Terra.
-
-%Para corrigir esses erros, será necessário observar as estrelas a serem ocultadas dias antes do evento obtendo os offsets correspondentes e calibrando a predição.
-Erros em ascensão reta interferem majoritariamente no instante em que a ocultação ocorrerá.
-Já erros em declinação principalmente na latitude geográfica em que a sombra passará.
-
-%Outra fonte de erro na predição é a posição do satélite.
-Como mostradas na seção \ref{Sec: Irr-astrometria}, as órbitas dos satélites irregulares não são muito precisas havendo erros sistemáticos em seus elementos orbitais.
-Infelizmente o JPL não provê suas efemérides com os erros estimados das posições.
-Dessa forma, há três alternativas para a correção das posições dos satélites.
-
-%A primeira delas é observar o satélite dias antes, quanto mais próximo do evento melhor, e determinar um offset.
-O offset, portanto, seria considerado estável dentro desses poucos dias uma vez que ele não se move muito em sua órbita.
-Porém, nem sempre é possível observar o alvo com antecedência.
-
-%A segunda alternativa é recalcular as órbitas dos satélites, gerar novas efemérides e estimar seus erros.
-Espera-se que com o grande número de posições publicadas em \cite{GomesJunior2015-Irregular} as órbitas sejam calculadas com maior precisão.
-Essa integração numérica será trabalhada durante o doutorado sanduíche em Paris em colaboração com o Dr.
-Jean-Eudes Arlot e o Dr.
-Josselin Desmars como foi feito para o trabalho de ocultações de TNOs na seção \ref{Sec: TNO-occ}.
-
-%Por fim, a terceira alternativa consiste em se fazer um ajuste dos offsets de efeméride obtidos em \cite{GomesJunior2015-Irregular} em função da anomalia verdadeira, tempo e/ou anomalia média.
-Essa escolha é suportada pelas dispersões dos offsets em função da anomalia verdadeira (por exemplo, Fig.
-\ref{Fig: carme_anom}, para Carme) que apresentaram características notáveis.
-Os parâmetros a serem utilizados para cada satélite serão estudados e aplicados.
-Esse é o método que será utilizado para o artigo de predições de satélites irregulares que será submetido logo.
-
-%Os primeiros ajustes foram feitos apenas considerando os offsets em declinação, pois são os que apresentaram características mais evidentes e, por seus erros representarem principalmente erros em latitudes, são os principais responsáveis pela não detecção da ocultação.
-Nesse caso, utilizamos um ajuste em função de três parâmetros: sen${f}, \cos{f}$ e uma constante, onde $f$ é a anomalia verdadeira dada pela efeméride.
-Esse ajuste é mostrado na Fig.
-\ref{Fig: Carme-dec-fit}.
-
-%\begin{figure}
-%\begin{centering}
-%\includegraphics[scale=0.45]{figuras/Carme-DEC_fit.png} 
-%\caption{Ajuste dos offsets em declinação de Carme publicados em \cite{GomesJunior2015-Irregular} em função da anomalia verdadeira}
-%\label{Fig: Carme-dec-fit}
-%\end{centering}
-%\end{figure}
-
-%É possível ver para o caso de Carme que o ajuste representa razoavelmente bem a dispersão dos offsets em função da anomalia verdadeira.
-Para outros satélites, ajuste em primeira ordem do seno e do cosseno da anomalia verdadeira também representam razoavelmente bem os offsets em declinação.
-Para os offsets em ascensão reta, o ajuste com três parâmetros não é suficiente.
-
-%Observar uma ocultação estelar de um satélite irregular exige um esforço muito grande por parte dos observadores.
-A sombra cobrirá uma região muito restrita na Terra devido aos satélites irregulares serem pequenos.
-Portanto, antes de iniciar uma grande campanha observacional, testamos algumas predições de ocultações para objetos maiores, para avaliar a qualidade da predição.
-
-Esse teste consiste em se observar o objeto ocultante e a estrela a ser ocultada próximo ao evento, quando os dois objetos estiverem presentes no mesmo campo, preferencialmente próximo ao instante do evento quando os objetos estarão próximos um do outro.
-Assim as posições relativas entre os dois objetos terão influências mínimas dos erros das estrelas de catálogo utilizadas na redução e das possíveis distorções de campo.
+Destes, cerca de 10\% dos eventos envolvem estrelas mais brilhantes de magnitude R=14 e 25\% mais brilhantes que R=15.
+Isso facilitará a observação por observadores amadores.
 
@@
 Foram comparadas predições oriundas da nossa integração numérica, das efemérides do JPL, e de offsets de observações tomadas próximas à data do evento.
-%Para cada evento quatro mapas foram gerados: o primeiro com as posições nominais da estrela e do satélite para o instante previsto; o segundo com o offset calculado como descrito anteriormente; o terceiro com os offsets da estrela e do satélite a partir de observações feitas alguns dias antes da ocultação quando os dois objetos estavam separados (campos diferentes); e o quarto a partir de observações feitas com menos de 24 horas de diferença da data prevista com a estrela e o satélite próximos no mesmo campo.
-
-%A Fig.
-\ref{Fig: occ-Himalia} mostra os quatro mapas para o teste da ocultação de Himalia de 03 de Março de 2015.
-O mapa \ref{Fig: occ-Himalia-off22fev} foi feito a partir de posições obtidas em 22 de Fevereiro observadas no telescópio Zeiss do Observatório do Pico dos Dias (OPD).
-Nesse dia, Himalia e a estrela a ser ocultada foram observados em campos separados por ainda estarem distantes.
-Na noite do evento, 03 de Março, os objetos foram observados com o telescópio Perkin-Elmer do OPD pouco mais de uma hora após os instante previsto para o evento.
-Satélite e estrela estavam separados por aproximadamente 16 arcsec, portanto muito próximos no campo.
-A partir dos offsets calculados, o mapa \ref{Fig: occ-Himalia-off03mar} foi gerado.
-
-%\begin{figure}[h!]
-%\begin{centering}
-%\subfigure[Sem offsets]{\includegraphics[scale=0.16]{figuras/Himalia1_2015-03-03T00:39:25.png}  \label{Fig: occ-Himalia-nooff}}
-%\subfigure[Offset de Himalia a partir do ajuste]{\includegraphics[scale=0.16]{figuras/Himalia2_2015-03-03T00:39:18.png}  \label{Fig: occ-Himalia-ajuste}}
-%\subfigure[Obs. dia 22 de Fevereiro (campos separados)]{\includegraphics[scale=0.16]{figuras/Himalia3_2015-03-03T00:39:40.png}   \label{Fig: occ-Himalia-off22fev}}
-%\subfigure[Obs. dia 03 de Março (mesmo campo)]{\includegraphics[scale=0.16]{figuras/Himalia4_2015-03-03T00:39:15.png}   \label{Fig: occ-Himalia-off03mar}}
-%\caption{Predições para Himalia: O ponto grande em vermelho mostra o ponto de máxima aproximação geocêntrica da sombra na Terra, os pontos vermelhos menores são os centros da ocultação separados por um minuto, as linhas retas são os limites das sombras dado o tamanho estimado do objeto e as linhas tracejadas equivalem a uma diferença de 40 mas do centro da sombra representando o erro estimado da predição.
-(a) é o mapa utilizando apenas as posições nominais da estrela e do satélite.
-(b) mostra a sombra dado um offset estimado para a posição de Himalia dada sua anomalia verdadeira segundo um ajuste feito em cima dos offsets de efeméride encontrados em \cite{GomesJunior2015-Irregular}.
-Em (c) são aplicados offsets às posições da estrela e do satélite a partir de observações feitas em 22 de Fevereiro no telescópio Zeiss.
-Em (d), temos o mesmo que para (c) porém com offsets obtidos de observações feitas em 03 de Março no telescópio Perkin-Elmer quando os objetos estavam próximos no campo.}
-%\label{Fig: occ-Himalia}
-%\end{centering}
-%\end{figure}
 
@@
 Para o evento de Elara as observações tomadas no Zeiss prejudicaram a qualidade das posições uma vez que Elara é um objeto pouco brilhante.
-%Na Fig.
-\ref{Fig: occ-Elara} temos os mapas para a ocultação de Elara do dia 30 de Março de 2015.
-As observações de teste foram tomadas nos dias 25 de Março e 02 de Abril de 2015 no Zeiss.
-Na noite do dia 02 eles ainda puderam ser observadas no mesmo campo.
-Devido à Elara ser bem mais fraco, as dispersões das posições do satélite nas duas noites acabaram sendo maiores que para Himalia.
-Ainda assim, as diferenças entre os mapas obtidos foram relativamente pequenas. A maior diferença entre elas é de $48 s$ e $302 km$.
+Ainda assim, as diferenças entre os mapas obtidos foram relativamente pequenas.
+A maior diferença entre elas é de $48 s$ e $302 km$.
 
-%\begin{figure}[h]
-%\begin{centering}
-%\subfigure[Sem offsets]{\includegraphics[scale=0.16]{figuras/Elara1_2015-03-30T01:45:15.png} \label{Fig: occ-Elara-nooff}}
-%\subfigure[Offset de Elara a partir do ajuste]{\includegraphics[scale=0.16]{figuras/Elara2_2015-03-30T01:44:59.png} \label{Fig: occ-Elara-ajuste}}
-%\subfigure[Obs. dia 25 de Março (campos separados)]{\includegraphics[scale=0.16]{figuras/Elara3_2015-03-30T01:44:17.png}  \label{Fig: occ-Elara-off25mar}}
-%\subfigure[Obs. dia 02 de Abril (mesmo campo)]{\includegraphics[scale=0.16]{figuras/Elara4_2015-03-30T01:45:30.png}  \label{Fig: occ-Elara-off02abr}}
-%\caption{Predições para Elara: Mesmo que a Fig.
-\ref{Fig: occ-Himalia}, mas para Elara.
-O mapa (c) feito a partir de observações tomadas no dia 25 de Março e (d) a partir de observações de 02 de Abril de 2015.}
-%\label{Fig: occ-Elara}
-%\end{centering}
-%\end{figure}
+Uma ocultação de Himalia, maior satélite irregular de Júpiter, estava prevista para 23 de Outubro de 2015 quando a sombra passaria no Brasil.
+As condições não eram muito propícias.
+A estrela possuía magnitude R=15.5, os objetos estavam baixo no céu, cerca de 30 graus de altura, e próximo ao amanhecer.
 
-%Um pedido de tempo foi feito para o telescópio Zeiss para se tentar observar uma ocultação de Himalia  dia 23 de Outubro de 2015.
-A sombra dessa ocultação estava predita para passar pela região sudeste do Brasil.
-Infelizmente o tempo estava fechado e nenhum dos sítios disponíveis pôde observar o evento.
+De toda forma, organizei junto com o Dr.
+Felipe Braga Ribas, do Instituto Tecnológico Federal do Paraná, uma campanha para a observação do evento.
+Reunimos observadores em Guaratinguetá-SP, São Paulo-SP, Valinhos-SP, Vitória-ES, Brasília-DF, Curitiba-PR, Ponta-Grossa-PR, Oliveira-MG e no Observatório do Pico dos Dias.
+No fim das contas, todos os observadores não puderam observar o evento devido ao mau tempo.
 
-Uma publicação das predições das ocultações está sendo preparada. A observação de novos eventos serão tentados.
+Também foram feitas predições para Sycorax, satélite irregular de Urano, e Nereida, satélite irregular de Netuno.
+Porém, dentro do intervalo de 5 anos pesquisado, apenas 2 eventos por satélite foram identificados.
+Isso se deve ao fato de Urano e Netuno estarem passando por uma região pouco densa de estrelas no céu.
+Como esses eventos apresentavam más condições de observação, estrela fraca, sombra passando pelo oceano, etc., eles não foram incluídos no trabalho.
 
-%\begin{figure}[h]
-%\includegraphics[scale=0.35]{figuras/Himalia_2015-10-23T07:54:47.png}
-%\caption{Mapa da ocultação de Himalia que ocorrerá em 23 de Outubro de 2015}
-%\label{Fig: occ-Himalia-itajuba}
-%\end{figure}
+A próxima etapa é atualizar as predições feitas com o UCAC4 utilizando as posições GAIA, cujo primeiro release se deu em setembro de 2016 \citep{Brown2016}.
+A falta de movimentos próprios no release I não afetará as predições uma vez que estas tem datas próximas à data de observação do GAIA (época média do catálogo).
+
+O trabalho de predições usando o catálogo UCAC4, junto com tabelas de informações sobre os eventos, foi publicado por \cite{GomesJunior2016}.
+Essas predições também são disponibilizadas na rede \textit{OccultWatcher}\footnote{http://www.occultwatcher.net/} onde podem ser visualizadas por observadores, em sua maioria amadores, em todo o mundo.
 
@@
 
-%Dentre os diversos mecanismos propostos para a captura do satélite, o mais provável foi proposto por \cite{Agnor2006} onde Tritão seria capturado através da dissociação de um sistema binário de planetesimais ao passar próximo a Netuno.
-Após a captura, a órbita de Tritão seria encolhida e circularizada para sua órbita atual.
-Um dos possíveis mecanismos seria a interação gravitacional com Netuno, i.e., efeito de maré \citep{Nogueira2011}.
-
-%A combinação entre a obliquidade do eixo de rotação de Netuno em relação ao seu eixo orbital e a alta inclinação da órbita de Tritão dá a Tritão um complicado e extremo ciclo sazonal \citep{McKinnon2007}.
-As marés associadas à sua captura podem ter fortemente aquecido e transformado seu interior.
-
 Tritão é um dos poucos objetos do sistema solar que possuem atmosfera e é constituída principalmente de $N_{2}$.
@@
 
-%\section{Astrometria}
-%\label{Sec: Netuno-astrometria}
+Tritão e Netuno foram observados pela sonda \textit{Voyager II} em 1989 \citep{Smith1989}, porém foi uma passagem rápida.
+O acompanhamento orbital de Tritão e Netuno assim como o estudo de sua atmosfera deve ser feita por observações de solo.
+Este último através de ocultação estelar.
 
-%\indent \indent Com o objetivo de estudar os efeitos de maré sobre o sistema Netuno-Tritão, serão obtidas as posições desses objetos em mais de 7000 imagens observadas no OPD desde 1992 até o presente.
-Uma vez que eles estão muito próximos no céu e Netuno é muito mais brilhante do que Tritão, a posição do satélite pode estar afetada pela luz do planeta.
-Para eliminar essa influência será utilizada uma técnica chamada de "coronografia digital" \citep{Assafin2009}.
+Regularmente, o nosso grupo tem observado o sistema Netuno-Tritão ao longo do tempo.
+Isso produziu um grande banco de dados com mais de 5000 observações CCD desde 1992.
+Como comparação, \cite{Emelyanov2015} produziram novas efemérides de Tritão a partir de 10254 observações feitas entre 1847 e 2012.
 
-%O procedimento de coronografia digital faz parte do pacote PRAIA e foi desenvolvido para se observar os satélites principais de Urano durante os fenômenos mútuos de 2007 \citep{Assafin2009}.
-A técnica consiste em computar o perfil de luz do objeto mais brilhante dado uma área.
-A imagem original, então, é subtraída do perfil do objeto gerando uma imagem coronografada.
+Uma das principais vantagens das nossas observações é a que na maioria dos casos (cerca de 3400) temos posições de Netuno e Tritão oriundas da mesma imagem.
+Isso ajudará a diminuir possíveis erros sistemáticos presentes na efeméride Tritão causadas pela efeméride planetária.
 
-%Para se obter o perfil do objeto, é calculada a distância $d$ de cada pixel ($j$,$i$) às coordenadas centrais do baricentro ($x_{c}$,$y_{c}$) da fonte brilhante.
-Para cada anel de 1 pixel de largura a uma distância $d$ do baricentro é feito um histograma de contagens.
-Por fim, o perfil é obtido sendo que, para cada pixel, a contagem é estimada como a moda do histograma de contagens de seu respectivo anel.
+Apesar de sempre termos os dois objetos na mesma imagem, a diferença de brilho entre eles dificulta a observação em boa qualidade dos dois ao mesmo tempo.
+Netuno é muito brilhante e exige exposições mais curtas, porém Tritão acaba ficando pouco exposto e o número de estrelas do catálogo de referência fica reduzido, prejudicando a redução astrométrica.
+Ao ter Tritão bem exposto e com bom número de estrelas de referência, Netuno fica saturado na imagem, inviabilizando seu uso.
 
-%No caso, a luz de Netuno será removida diminuindo a sua influência sobre a posição de Tritão. Como subproduto, as coordenadas do baricentro calculadas são uma boa estimativa da posição de Netuno.
+A redução da posição de Netuno ainda encontra duas dificuldades.
+São elas a refração cromática diferencial e o seu diâmetro angular aparente de 2,3 segundos de arco, da ordem do seeing, de forma que uma PSF (point spread function) gaussiana usual pode não ser adequada.
 
-%Como o tamanho aparente de Netuno ($\approx 2\arcsec$) é da ordem do seeing, uma técnica está sendo desenvolvida para determinar o centroide de objetos extensos.
-No caso, supõe-se que, para cada ponto da imagem, a luz é espalhada sobre o CCD como uma gaussiana e a imagem correspondente é a soma de todas estas gaussianas.
-O perfil resultante não tem forma analítica fechada, devendo ser calculado numericamente.
-Ainda estamos estudando a possibilidade de se obter funções analíticas para a PSF.
+A refração cromática diferencial é causada principalmente pelo fato de Netuno possuir uma cor mais azulada em relação às estrelas de referência, cujas cores têm uma tendência mais avermelhada devido ao espalhamento causado por poeira.
+Esse efeito se intensifica quanto mais próximo do plano galático.
 
-\section{Ocultaç\~oes}
-\label{Sec: Netuno-predic}
+Para corrigir esse efeito utilizamos a técnica adotada por \cite{Rossi2014} onde a refração cromática é modelada a partir dos offsets de efeméride da posição de Netuno.
+O modelo é aplicado em ascensão reta uma vez que, devido à latitude do sítio observacional é próxima à declinação de Netuno na data da observação, a refração é mais significativa.
+Os parâmetros obtidos são, então, aplicados aos offsets em declinação.
 
-%\indent \indent Nos mesmos moldes em que foram feitas observações dos caminhos de TNOs para predição de ocultações estelares (ver seção \ref{Sec: TNO-occ}), o caminho de Tritão também foi observado gerando um catálogo de estrelas nos moldes do catálogo WFI.
-Infelizmente, devido a Netuno estar passando por regiões pouco populadas por estrelas brilhantes, nenhum evento pôde ser observado oriundo deste trabalho.
+Essa técnica se mostrou efetiva reduzindo pela metade a dispersão dos offsets em ascensão reta na maioria dos casos.
+Para Tritão, como esperado devido à sua cor mais avermelhada, a técnica não se mostrou efetiva.
 
-%\section{Ocultação de Maio/2008}
-%\label{Sec: Netuno-maio2008}
+A Fig.
+\ref{Fig:refracao} mostra a distribuição dos offsets de Netuno observados no telescópio Perkin-Elmer no OPD antes e depois da correção de refração cromática.
+Pode-se ver que o método diminuiu consideravelmente a dispersão em ascensão reta.
+Isso mostra a significância desse procedimento.
 
-\indent \indent Em 21 de Maio de 2008 houve uma ocultação por Tritão cuja sombra cruzou o sul da África.
-Dos sítios que observaram, apenas duas cordas foram positivas.
-As observações foram feitas de La Réunion e da Namíbia com câmeras rápidas.
-Dentre elas, uma observou somente a atmosfera do objeto.
-Foi a primeira ocultação de Tritão desde 1997 \citep{Elliot2000}.
+\begin{figure}[h]
+\includegraphics[width=16cm]{figuras/dist_Netuno_160.png} 
+\caption{Distribuição dos offsets de Netuno observados no telescópio Perkin-Elmer no OPD antes e depois da correção de refração cromática. \label{Fig:refracao}}
+\end{figure}
 
-Os perfis de pressão da atmosfera de Tritão foram calculados (Sicardy et al., comunicação privada), porém, devido a uma das cordas ter pego somente atmosfera, não foi possível determinar a posição das cordas em relação ao corpo com precisão.
-Isso resultou em valores de pressão próximo à superfície com pequenas barras de erro interno, mas que podem estar desviados em mais de 3 sigmas do valor real.
+O segundo problema encontrado na posição de Netuno foi o fato dele ser aparentemente muito grande, maior que o seeing médio das estrelas.
+O ajuste com a gaussiana não se adequa muito bem ao topo da PSF de Netuno.
+Como Netuno está sempre bem amostrado, é possível que o ajuste da Gaussiana usando apenas as laterais da imagem de Netuno seja bom o suficiente.
+Porém para um ajuste completo, o Dr.
+Roberto Viera Martins desenvolveu uma PSF numérica para ajustar Netuno com melhor precisão.
 
-Novas ocultações são necessárias para estudar o perfil da atmosfera de Tritão e sua evolução ao longo do tempo.
-Para 2015, antes que o catálogo WFI não cubra mais o caminho orbital de Tritão, ainda haverá uma oportunidade de observar uma ocultação estelar pelo objeto.
-O evento está predito para ocorrer dia 10 de Novembro de 2015 com a sombra passando pela China.
-Dependendo do offset da estrela e de Tritão, há a possibilidade da sombra ir mais ao sul passando pelo Japão ou Austrália, onde temos colaboradores que poderão observar o evento.
+A PSF numérica consiste em uma integral sobre um disco de raio R onde cada ponto no disco gera uma dispersão gaussiana de mesma largura à meia altura das estrelas da imagem.
+Outras características como escurecimento de bordo e ângulo de fase podem ser consideradas no cálculo.
 
-%\begin{figure}[h]
-%\begin{centering}
-%\includegraphics[scale=0.58]{figuras/tritao-2008.png} 
-%\caption{Reconstrução pós-ocultação da forma de Tritão. As linhas são as cordas observadas. Como é possível ver, a corda da Namíbia observou apenas a atmosfera de Tritão.}
-%\label{Fig: Tritao-body}
-%\end{centering}
-%\end{figure}
-%
-%\begin{figure}[h]
-%\begin{centering}
-%\includegraphics[scale=0.58]{figuras/pressao-2008.png} 
-%\caption{Comparação entre a pressão atmosférica na superfície de Tritão obtida por Sicardy et al.
-(comunicação privada) e pressões publicadas na literatura: \cite{Gurola1995, Olkin1997, Sicardy1998} e \cite{Elliot2000}}.
-%\label{Fig: Tritao-pressao}
-%\end{centering}
-%\end{figure}
+Essa técnica está sendo testada em algumas posições de Netuno.
+O objetivo é verificar se o fotocentro calculado com a PSF numérica varia em relação à Gaussiana de forma significativa.
+Esse teste levará em conta o ângulo de fase da observação.
 
-%\cite{Hansen1992} geraram modelos que preveem que a pressão atmosférica de Tritão em sua superfície estaria em ascensão.
-Neste trabalho, a pressão atmosférica foi obtida com um valor muito diferente do valor esperado mesmo ao nível de $3\sigma$.
-O resultado obtido também é inconsistente com outros valores da literatura que confirmam o esperado por \cite{Hansen1992}.
+As posições finais a serem obtidas para o sistema Netuno-Tritão serão utilizadas em nova integração numérica que será realizada durante o meu doutorado sanduíche, o qual comecei em Setembro de 2016.
+Esse estágio tem duração de um ano e está sendo realizado no \textit{Institut de Méchanique Céleste et de Calcul des Éphémérides} (IMCCE) e é financiado pelo programa CAPES-COFECUB.
 
-%Para recuperar essa ocultação, um trabalho astrométrico deve ser feito de forma a medir com precisão maior a posição do centro do corpo no instante da ocultação.
-Dessa forma, será possível localizar as cordas observadas às suas respectivas latitudes, diminuindo as barras de erro.
+Essa nova integração numérica será usada para melhorar as efemérides de Tritão em relação a Netuno.
+Além disso, será possível analisar a efeméride de Netuno em relação ao Sol devido às diferenças das posições obtidas e das efemérides.
+Uma publicação é esperada desse trabalho.
 
-\section{Astrometria}
-\label{Sec: Netuno-mares}
-
-\indent \indent Com o objetivo de estudar os efeitos de maré sobre o sistema Netuno-Tritão, serão obtidas as posições desses objetos em mais de 7000 imagens observadas no OPD desde 1992 até o presente.
-%Uma vez que eles estão muito próximos no céu e Netuno é muito mais brilhante do que Tritão, a posição do satélite pode estar afetada pela luz do planeta.
-Para eliminar essa influência será utilizada uma técnica chamada de "coronografia digital" \citep{Assafin2009}.
-A vantagem desse banco de dados é sua cobertura temporal, o grande número de observações e um tratamento astrométrico diferenciado.
-Faremos uso de coronografia digital \citep{Assafin2009} para eliminar a influência da luz difusa de Netuno sobre Tritão, e faremos uma determinação direta de posições de Netuno (com PSFs numéricas especialmente deduzidas para o caso de objetos extensos), com Tritão no mesmo frame.
-Esses dados tem o mesmo valor que o de posições oriundas de sondas, em que pesem efeitos dinâmicos de longo período \citep{Lainey2012}.
-
-Além das posições de solo, irei trabalhar com posições de telemetria de sondas espaciais dentro do projeto ESPaCE (European Satellite PArtnership for Computing Ephemerides).
-Esse projeto tem por objetivo extrair e analisar dados astrométricos de medidas feitas por sondas espaciais e utilizá-las para cálculo de órbitas e geração de efemérides dos objetos dos Sistema Solar.
-Essas medidas são ultra precisas e poderão prover novos modelos dinâmicos para muitos satélites naturais\footnote{\url{http://espace.oma.be/}}.
-
-A combinação desses dois bancos de dados, junto com as posições publicadas na literatura permitirá estudar efeitos dissipativos de energia no interior de Tritão.
-%Como dito na seção \ref{Sec: Netuno-intro}, a circularização da órbita de Tritão, a partir de uma órbita altamente excêntrica, após a captura aconteceu por dissipação oriunda de efeitos de maré \citep{Nimmo2015}.
-
-%Durante a circularização da órbita, o aquecimento causado por efeito de maré teria aumentado chegando a um fluxo de calor de $\approx 2 - 4 W m^{-2}$, comparáveis às medidas atualmente para Io.
-Após o aquecimento, o interior de Tritão começou a esfriar, mas mesmo hoje esse esfriamento não deve ter sido completado \citep{McKinnon2007}.
-
-%Pela órbita de Tritão ser retrógrada, espera-se que o semi-eixo de sua órbita diminua devido ao efeito de maré que provoca em Netuno e que seja destruído ao se aproximar do planeta devido à essa mesma força \citep{McKinnon2007}.
-Estima-se que a órbita de Tritão encolha em aproximadamente 15\% nos próximos 5 bilhões de anos.
+Uma grande importância da melhoria na posição de Tritão é que está previsto para o dia 5 de Outubro de 2017 uma ocultação de uma estrela de magnitude R=12 por Tritão.
+A sombra prevista passará pela Europa e pela costa leste dos Estados Unidos e será uma grande oportunidade para estudar a atmosfera do satélite.
+Isso se torna mais significativo ao lembrar que Netuno está em uma região pouco densa de estrelas e ocultações são raras.
 
@@
 As técnicas utilizadas consistem em obter posições astrométricas precisas desses corpos para corrigir erros sistemáticos em suas efemérides, ou mesmo permitir a determinação de novas órbitas com novas integrações numéricas, a fim de melhorar a predição de futuras ocultações estelares.
-% A técnica de fotometria diferencial de abertura é empregada para gerar as curvas de luz das ocultações estelares.
-Prever, observar e reduzir ocultações estelares por esses objetos e obter seus parâmetros físicos (tamanho, forma, albedo, densidade, presença de atmosfera, satélites ou anéis) são os objetivos gerais desta Tese.
 
-Participo ativamente em observações de TNOs, satélites, e estrelas com o objetivo de melhorar as efemérides e/ou predição de ocultações.
-Nesse âmbito já escrevi 1 trabalho: \cite{GomesJunior2015-Irregular}; e contribuo em 2 trabalhos: \cite{Camargo2015} e \cite{Desmars2015}.
+Como participo ativamente em observações de TNOs e satélites com o objetivo de melhorar as efemérides e/ou predição de ocultações contribuí em mais 2 trabalhos.
+O primeiro sobra astrometria dos satélites internos de Urano publicado por \cite{Camargo2015}.
+O segundo sobre uma nova técnica de medidas relativas de grande precisão entre os satélites Galileanos \citep{Morgado2016}.
+Devido à grande quantidade de atividade desenvolvida em grupo, outras publicações em colaboração são esperadas.
+Também sou autor e co-autor de pôsteres e apresentações em congressos.
 
-%A busca pela predição de uma ocultação e sua melhoria na era pré-GAIA vai demandar ainda diversas missões observacionais.
-Quando o catálogo do GAIA for lançado, as posições das estrelas serão boas o suficientes para não precisar mais de acompanhamento, porém os objetos do Sistema Solar ainda precisaram ser observados e suas efemérides melhoradas pois serão, portanto, a única fonte de erro na predição de uma ocultação.
+Além disso, no doutorado sanduíche, estou desenvolvendo, junto com o Dr.
+Valéry Lainey, pesquisador do IMCCE do Observatório de Paris, França, um código para integração numérica das órbitas de objetos do Sistema Solar.
+Esse trabalho é muito importante pois não existe no Brasil pessoas especializadas em integração numérica para geração de efemérides.
 
-Observar uma ocultação pode revelar características inesperadas e até dar origem a novas descobertas no Sistema Solar.
-Exemplos disso são a descoberta de um sistema de anéis, estrutura antes conhecida apenas nos planetas gigantes, ao redor de um Centauro e a identificação de um alto albedo geométrico em Eris, objeto que está muito distante do Sol e possui um tamanho bem menor do que se supunha, similar ao de Plutão e Tritão.
-Esses dois trabalhos foram publicados na Nature em \cite{BragaRibas2014} e \cite{Sicardy2011}, respectivamente, nos quais contribuí em caráter observacional.
+Um código próprio irá enriquecer o conteúdo de nossos trabalho dando um passo a mais além da publicação de posições astrométricas apenas.
+Isso também nos permitirá ter uma maior independência onde até agora precisávamos procurar uma colaboração.
 
-Particularmente, Plutão possui um interesse especial.
-Recém visitado por uma sonda espacial, ele é um planeta anão que possui atmosfera, cinco satélites, sendo que o maior deles possui um diâmetro pouco maior que a metade do de Plutão.
-Ocultações estelares por esse objeto permitem estudar a evolução de sua atmosfera, a qual será calibrada a partir de observações oriundas da sonda \textit{New Horizons}.
-%Para isso uma ocultação que ocorreu próxima à data da chegada da sonda foi observada.
-Das observações às quais participei, dois estudos sobre a atmosfera de Plutão puderam ser feitos: \cite{Olkin2015} e \cite{DiasOliveira2015}.
+Esse ano será suficiente para obter um código capaz de calcular órbitas de satélites irregulares, TNOs entre outros objetos, inclusive ajustando a órbita às observações.
+As observações de Tritão e Netuno serão utilizadas como teste.
 
-Da mesma forma que para Plutão, Ceres é um planeta anão e está sendo visitado por uma sonda.
-A partir de duas ocultações, calculamos o tamanho de Ceres e obtivemos posições muito precisas.
-Os resultados foram publicados em \cite{GomesJunior2015-Ceres} e serviram como aprendizado pessoal quanto à utilização da técnica.
-Quando os resultados da sonda \textit{Dawn} forem publicados, será possível calibrar a técnica de ocultações estelares utilizada nesse trabalho.
+Por fim, pretendemos observar ocultações estelares de satélites irregulares.
+Com o release do GAIA temos posições para as estrelas com precisão muito melhor às utilizadas até o momento.
+Além disso, trabalharemos na integração numérica das órbitas dos satélites, inclusive usando observações publicadas na literatura melhorando, portanto, suas predições.
 
-%Muitos outros objetos estão na lista para se observar ocultações estelares, todos com grandes potenciais científicos.
-Por exemplo, há a possibilidade de que o Centauro (2060) Chiron também possua anéis \citep{Ruprecht2015}.
-Como o número de ocultações detectadas pelo nosso grupo vem aumentando ao longo do tempo e aumentará muito mais com o GAIA, diversas publicações ainda estão por vir.
-
-Dentre os objetos a serem ocultados, inserimos na lista os satélites irregulares dos planetas gigantes, nunca antes observados por ocultações.
-Diversas ocultações já estão preditas e a publicação destas predições está em preparação (Gomes-Júnior et al.
-2016, A\&A, in prep.).% Observações de ocultações por Himalia, satélite de Júpiter, e Tritão estão sendo preparadas.
-
-Por fim, estudos sobre efeitos de maré no sistema Netuno-Tritão para obter parâmetros internos dos satélites serão feitos durante o Doutorado Sanduíche que ocorrerá na França em colaboração com o Doutor Valéry Lainey.
-Nesse trabalho, focaremos na obtenção de posições astrométricas do sistema a partir de imagens de solo observadas no OPD e de sondas espaciais dentro do projeto ESPaCE.
-
+\newpage
 \bibliography{Referencias}

### 2017-08-16 Ceres/Old_versions/Altair 23.03.2015/CeresV3.tex
mod 2017 ago16

--- 2015-05-16 Ceres/Old_versions/CeresV2.tex
@@
 Far from being a mere taxonomic information, this suggests the great impact its study can have on the understanding of planetary formation and evolution of the Solar System.
-Indeed, it was proposed that Ceres could have its origin as a transneptunian object, latter scattered to the Main Belt due to the giant planets' migration predicted by the ``Nice Model''.
+Indeed, it was proposed that Ceres could have its origin as a transneptunian object, later scattered to the Main Belt due to the giant planets' migration predicted by the ``Nice Model''.
 It may well had been formed close to its current location, however, even on this scenario, the dynamical history of the Solar System must have let its signatures on Ceres.
@@
 The two remaining events, which occurred on 17 August 2010 and 25 October 2013, are reported on the present work.
-The former was observed in Brazil from 5 stations, while the later was from the United States and led to 9 chords.
+The former was observed in Brazil from 5 stations, while the latter was from the United States and led to 9 chords.
 Throughout the paper we shall refer to these events as the ``2010'' and the ``2013 occultation''.
@@
 
-Of course, the apparent semimajor axis $a^\prime$ equals the equatorial radius $a$ of the ellipsoid. %The polar radius $b$ can then be determined from $\epsilon = 1 - (b/a)$.
+Of course, the apparent semimajor axis $a^\prime$ equals the equatorial radius $R_{equa}$ of the ellipsoid. %The polar radius $b$ can then be determined from $\epsilon = 1 - (b/a)$.
 
@@
 
-Evaluation of the reduced $\chi^2$ function over the $\lbrace a, \epsilon^\prime, P, f_c, g_c \rbrace$--space produces a map related to the likelihood of a solution.
+Evaluation of the reduced $\chi^2$ function over the $\lbrace R_{equa}, \epsilon^\prime, P, f_c, g_c \rbrace$--space produces a map related to the likelihood of a solution.
 For instance, the 1--sigma reported time uncertainties correspond to the subset of the parameter space for which $\chi^2_r$ lies between ${\chi^2_r}_{min}$ and ${\chi^2_r}_{min} + 1$.
@@
 
-The resulting values of semimajor axis, oblateness, position angle and centre coordinates are presented in the second column of Table~2.%, and the corresponding ellipse is plotted with the observed chords in Fig.~2.
+The resulting values of equatorial radius, oblateness, position angle and centre coordinates are presented in the second column of Table~2.%, and the corresponding ellipse is plotted with the observed chords in Fig.~2.
 % The equivalent radius of the best fit solution is $R_{equiv} = 469 \pm 10$ km.
@@
 We notice that the constraint corresponds to the upper limit of nominal solution's 1-sigma error bar.
-On the other hand, it selects the smallest values of semimajor axis, improving its determination by a factor of about 2.
+On the other hand, it selects the smallest values of equatorial radius, improving its determination by a factor of about 2.
 Notwithstanding, oblateness' figures remain the same.
@@
 \hline
-Equatorial Radius (km) & 491 $\pm$ 7   & 486 $\pm$ 3   & 486 $\pm$ 2   & \\
+Equatorial radius (km) & 491 $\pm$ 7   & 486 $\pm$ 3   & 486 $\pm$ 2   & \\
 Oblateness             & 0.08$\pm$0.03 & 0.08$\pm$0.03 & 0.09$\pm$0.05 & \\
-Position angle (deg)& 5 $\pm$ 10    & 12 $\pm$ 3 (*)& 22 $\pm$ 5    &  \\
+Position angle (deg)   & 5 $\pm$ 10    & 12 $\pm$ 3 (*)& 22 $\pm$ 5    &  (*)\\
 $f_c$ (km)             & 133 $\pm$ 9   & 138 $\pm$ 5   & 78 $\pm$ 7    & \\
@@
 \end{tabular}
-\textbf{Notes.} Error bars are at 1-sigma level. The reported oblateness derived from the 2010 event are the true ones, see text for details. \\
+\textbf{Notes.} Error bars are at 1-sigma level. The reported oblateness derived from the 2010 event are the physical ones (not apparent), see text for details. \\
 (*) Position angle derived from Ceres' rotational pole coordinates determined by \cite{Drummond2014}.
@@
 
-Nine positive chords were obtained for the 2013 occutation by a variety of equipment as listed in Table \ref{Tab: obs-2013}.
+Nine positive chords were obtained for the 2013 occultation by a variety of equipment as listed in Table \ref{Tab: obs-2013}.
 Each station was equipped with a video camera (whose readout time may be considered negligible).
@@
 
-From the occultation times obtained, the shape for Ceres' limb was adjusted by an ellipse characterized by five parameters: the coordinates of the body center, relative to the star in the plane of the sky ($f_{c}, g_{c}$); the apparent semimajor axis a, equivalent to the equatorial radius $R_{Equa}$ ; the apparent oblateness = $(a'-b')/a'$ (where $b'$ is the apparent semiminor axis); and the position angle P of the semiminor axis $b'$ .
+From the occultation times obtained, the shape for Ceres' limb was adjusted by an ellipse characterized by five parameters: the coordinates of the body center, relative to the star in the plane of the sky ($f_{c}, g_{c}$); the apparent semimajor axis a, equivalent to the equatorial radius $R_{Equa}$ ; the apparent oblateness = $(a'-b')/a'$ (where $b'$ is the apparent semiminor axis or Polar Radius); and the position angle P of the semiminor axis $b'$ .
 The coordinates $f_{c}$ and $g_{c}$, in kilometers, are calculated using the JPL\#33  Ceres' ephemeris and the star position given in Section \ref{Sec: observation-2013}.

### 2017-08-16 Ceres/Old_versions/Altair_25_03_2015/CeresV3.tex
mod 2017 ago16

--- 2017-08-16 Ceres/Old_versions/Altair 23.03.2015/CeresV3.tex
@@
 \caption{Post-occultation reconstruction of Ceres' shadow path on Earth for the 2010 August 17 event.
-The big red dot is the geocentric closest approach at 22:41:28 UT.
+The big red dot is the geocentric closest approach at 22:40:25 UT.
 The small red ones represents the center of the shadow separated by one minute where the shadow moves from the left to the right.
@@
 \hline
-Equatorial radius (km) & 491 $\pm$ 7   & 486 $\pm$ 3   & 486 $\pm$ 2   & \\
-Oblateness             & 0.08$\pm$0.03 & 0.08$\pm$0.03 & 0.09$\pm$0.05 & \\
-Position angle (deg)   & 5 $\pm$ 10    & 12 $\pm$ 3 (*)& 22 $\pm$ 5    &  (*)\\
-$f_c$ (km)             & 133 $\pm$ 9   & 138 $\pm$ 5   & 78 $\pm$ 7    & \\
-$g_c$ (km)             & -17 $\pm$ 15  & -11 $\pm$ 11  & 14 $\pm$ 16   & \\
-$\chi^2_{r,min}$       & 0.24          &  0.42         & 1.27          & \\
+Equatorial radius (km) & 491 $\pm$ 7   & 486 $\pm$ 3   & 486 $\pm$ 2   & 485 $\pm$ 4\\
+Oblateness             & 0.08 $\pm$ 0.03 & 0.08 $\pm$ 0.03 & 0.09 $\pm$ 0.05 & 0.08 $\pm$ 0.04\\
+Position angle (deg)   & 5 $\pm$ 10    & 12 $\pm$ 3 (*)& 22 $\pm$ 5    &  25$\pm$3(*)\\
+$f_c$ (km)             & 133 $\pm$ 9   & 138 $\pm$ 5   & 78 $\pm$ 7    & 78 $\pm$ 6\\
+$g_c$ (km)             & -17 $\pm$ 15  & -11 $\pm$ 11  & 14 $\pm$ 16   & 13 $\pm$ 16\\
+$\chi^2_{r,min}$       & 0.24          &  0.42         & 1.27          & 1.27\\
 \hline
@@
 The red dots represents the center of the shadow separated by one minute where the shadow moves from the left to the right.
-The geocentric closest approach at 09:43:11 UT.
+The geocentric closest approach at 09:43:03 UT.
 The blue dots are the sites that have observed the event.\label{Fig: Ceres-2013-map}}

### 2017-08-16 Ceres/Old_versions/Altair_27_03_2015/CeresV4.tex
mod 2017 ago16

--- 2017-08-16 Ceres/Old_versions/Altair_25_03_2015/CeresV3.tex
@@
 email@address (AVR); otheremail@otheraddress (ANO)} and A. N.
-Other$^{2}$\footnotemark[1]\thanks{This file has been amended to
-highlight the proper use of \LaTeXe\ code with the class file.
-These changes are for illustrative purposes and do not reflect the
-original paper by A. V. Raveendran.}\\
-$^{1}$Indian Institute of Astrophysics, Bangalore 560034, India\\
-$^{2}$Building, Institute, Street Address, City, Code, Country}
+Other$^{2}$\footnotemark[1]\thanks{}\\
+$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
+CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
+$^{2}$Observat\'orio Nacional/MCT, R. General Jos\'e Cristino 77,
+CEP 20921-400 Rio de Janeiro - RJ, Brazil}
 \begin{document}
 
-\date{Accepted 1988 December 15. Received 1988 December 14; in original form 1988 October 11}
+\date{Accepted . Received ; in original form }
 
-\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2002}
+\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2015}
 
@@
 \begin{keywords}
-keywords.
+minor planets, asteroids: individual (1, Ceres) - occultations - planets and satellites: fundamental parameters
 \end{keywords}
@@
 
+TYC 6833-163-1
+
 \begin{equation}
@@
 The occultation was detected on four of them.
-Among these, one started the shots with the event already in progress and only provided star's reappearance time; the other three recorded the whole phenomenon.
+Among these, INPE started the shots with the event already in progress and only provided star's reappearance time; the other three recorded the whole phenomenon.
 Since this event had a very low velocity -- only 3.9 km/s -- the exposure times (which for faster occultations would result in a big time uncertainty) led to quite small error bars, with the exception of the UEPG chord.
@@
  \begin{minipage}{140mm}
-  \caption{Circumstances of observation for all observing stations of the 2010 event.}
+  \caption{Circumstances of observation for all observing stations of the 2010 event. \label{Tab: obs-2010}}
   \begin{tabular}{@{}lccccc}
@@
 
-\subsection{Light curves and timing}
+\subsection{Light curves and timing \label{Sec: light-curve-2010}}
 
-[Curvas de luz, modelagem, difraç\~ao, di\^ametro estelar.]
+The flux of the star in the five occultation chords was obtained from the FITS images with the Platform for Reduction of Astronomical Images Automatically (PRAIA) \citep{2011gfun.conf...85A}.
+The light curves were normalized to an unocculted star flux.
+A linear or second-degree polynomial fit was applied to the observed flux before and after the occultation (with exception of INPE).
+
+The start and end times of the occultation were obtained for each light curve by fitting a sharp edge occultation model.
+This model is convolved by Fresnel diffraction, the CCD bandwidth, the stellar diameter in kilometers, and the finite integration time; see \cite{Widemann2009}.
+
+The Fresnel scale ($F = \sqrt{\lambda D/2}$) for the present Ceres’ geocentric distance $D = 2.29 AU = 3.42 \times 10^{8} km$ is 0.3 km for a typical wavelength of $\lambda = 0.65 \micron m$.The star diameter is estimated using the formulae of \cite{vanBelle1999}.
+Its B, V , and K apparent magnitudes are 12.00, 11.78 and 9.18, respectively, in the Tycho catalog \citep{Hog2000}.
+This yields a diameter of about $0.11 km$ projected at Ceres’s distance.
+The smallest integration time used in the positive observations was $1.0 s$, which translates to almost $3.9 km$ in the celestial plane.
+Therefore, the occultation light curves are largely dominated by the integration times, not by Fresnel diffraction or star diameter.
+
+The occultation fits consist in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011}.
+The free parameter to adjust is the ingress (disappearance) or egress (re-appearance) time $t_{occ}$, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$.
+
+The best fits to the occultation light curves are shown in Figure \ref{Fig: Ceres-2010-curves}, and the derived occultation times are listed in Table \ref{Tab: obs-2010}.
 
@@
 \hline
-Equatorial radius (km) & 491 $\pm$ 7   & 486 $\pm$ 3   & 486 $\pm$ 2   & 485 $\pm$ 4\\
-Oblateness             & 0.08 $\pm$ 0.03 & 0.08 $\pm$ 0.03 & 0.09 $\pm$ 0.05 & 0.08 $\pm$ 0.04\\
+Equatorial radius (km) & 491 $\pm$ 7   & 486 $\pm$ 3   & 486 $\pm$ 4   & 485 $\pm$ 4\\
+Oblateness             & 0.08 $\pm$ 0.03 & 0.08 $\pm$ 0.03 & 0.08 $\pm$ 0.04 & 0.08 $\pm$ 0.04\\
 Position angle (deg)   & 5 $\pm$ 10    & 12 $\pm$ 3 (*)& 22 $\pm$ 5    &  25$\pm$3(*)\\
-$f_c$ (km)             & 133 $\pm$ 9   & 138 $\pm$ 5   & 78 $\pm$ 7    & 78 $\pm$ 6\\
-$g_c$ (km)             & -17 $\pm$ 15  & -11 $\pm$ 11  & 14 $\pm$ 16   & 13 $\pm$ 16\\
+$f_c$ (km)             & 133 $\pm$ 9   & 138 $\pm$ 5   & 77 $\pm$ 6    & 78 $\pm$ 6\\
+$g_c$ (km)             & -17 $\pm$ 15  & -11 $\pm$ 11  & 13 $\pm$ 16   & 13 $\pm$ 16\\
 $\chi^2_{r,min}$       & 0.24          &  0.42         & 1.27          & 1.27\\
@@
 \includegraphics[scale=0.36]{figures/Ceres_2010_body.pdf} 
-\caption{\label{Fig:Ceres-2010-body}}
+\caption{The best elliptical fit for the occultation chords for the event of 2010 using timing from Table \ref{Tab: obs-2010} and the pole-constrained solution. \label{Fig:Ceres-2010-body}}
 \end{figure}
@@
 \section[]{The 2013 occultation}
+
+TYC 865-911-1
 
@@
 \caption{Post-occultation reconstruction of Ceres' shadow path on Earth for the 2013 October 25 event at the east coast of USA.
-The red dots represents the center of the shadow separated by one minute where the shadow moves from the left to the right.
-The geocentric closest approach at 09:43:03 UT.
+The big dot is the geocentric closest approach at 09:43:03 UT and the small ones represents the center of the shadow separated by 30 seconds where the shadow moves from the left to the right.
 The blue dots are the sites that have observed the event.\label{Fig: Ceres-2013-map}}
@@
 
-All videos were converted to FITS images and the flux of the star was obtained with the Platform for Reduction of Astronomical Images Automatically (PRAIA) \citep{2011gfun.conf...85A}.
+All videos were converted to FITS images and the flux of the star was also obtained with the PRAIA \citep{2011gfun.conf...85A}.
 The light curves were normalized to an unocculted star flux when a star was available.
-
-The data were binned by 5 or 10 points to reduce the noise. A linear or second-degree polynomial fit was applied to the observed flux before and after the occultation.
 
@@
 
-The start and end times of the occultation were obtained for each light curve by fitting a sharp edge occultation model.
-This model is convolved by Fresnel diffraction, the CCD bandwidth, the stellar diameter in kilometers, and the finite integration time; see \cite{Widemann2009}.
+Due to the small integration time, the flux star was very noisy.
+To reduce the noise, the data were binned by 5 points for all curves, except for Greenbelt where a 10 points binning was needed.
+This made an effective integration time 5 or 10 times bigger.
 
-The Fresnel scale ($F = \sqrt{\lambda D/2}$) for the present Ceres’ geocentric distance $D = 3.28 AU = 4.91 \times 10^{8} km$ is 0.4 km for a typical wavelength of $\lambda = 0.65 \micron m$.The star diameter is estimated using the formulae of \cite{vanBelle1999}.
-Its B, V , and K apparent magnitudes are 10.47, 9.97 and 8.76, respectively, in the Tycho catalog \citep{Hog2000}.
-This yields a diameter of about 0.16 km projected at Ceres’s distance.
-The smallest integration time used in the positive observations was 0.033 s, which translates to almost 1.4 km in the celestial plane.
-Therefore, the occultation light curves are largely dominated by the integration times, not by Fresnel diffraction or star diameter.
-
-The occultation fits consist in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011}.
-The free parameter to adjust is the ingress (disappearance) or egress (re-appearance) time $t_{occ}$, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$.
+The star and end times were obtained by the same procedure explained in Section \ref{Sec: light-curve-2010}.
+The Fresnel scale of $0.4 km$ for this event was calculated using the Ceres’ geocentric distance $D = 3.28 AU = 4.91 \times 10^{8} km$and a typical wavelength of $\lambda = 0.65 \micron m$.The star diameter projected at Ceres' distance of $0.16 km$ is estimated using the B = 10.47, V=9.97, and K=8.76 apparent magnitudes in the Tycho catalog \citep{Hog2000}.
+The smallest effective integration time used in the positive observations was $0.165 s$, which translates to almost $7.0 km$ in the celestial plane.
+As for the 2010 event, the occultation light curves are largely dominated by the integration times, not by Fresnel diffraction or star diameter.
 
@@
 \includegraphics[scale=0.58]{figures/Ceres_2013_fluxratio} 
-\caption{Same as in Fig. \ref{Fig: Ceres-2010-curves} for the nini light curves of the 2013 event. \label{Fig: Ceres-2013-curves}}
+\caption{The nine occultation light curves normalized and vertically shifted by a factor of 1.0 for better viewing.
+The solid black lines are the best fit of the square-well model to the data.
+The red lines are the square-well model convoluted with the Fresnel diffraction, the star diameter, and the applied exposure time.
+The mid-times of the occultations do not coincide due to the propagation delays of the shadow due to the distinct longitude of the sites.
+The light curve of Brookline is shifted by -64s as explained in the text.
+\label{Fig: Ceres-2013-curves}}
 \end{figure}
@@
 
-From the occultation times obtained, the shape for Ceres' limb was adjusted by an ellipse characterized by five parameters: the coordinates of the body center, relative to the star in the plane of the sky ($f_{c}, g_{c}$); the apparent semimajor axis a, equivalent to the equatorial radius $R_{Equa}$ ; the apparent oblateness = $(a'-b')/a'$ (where $b'$ is the apparent semiminor axis or Polar Radius); and the position angle P of the semiminor axis $b'$ .
+From the occultation times obtained, the shape for Ceres' limb was adjusted by an ellipse characterized by five parameters: the coordinates of the body center, relative to the star in the plane of the sky ($f_{c}, g_{c}$); the apparent semimajor axis $a'$, equivalent to the equatorial radius $R_{Equa}$ ; the apparent oblateness = $(a'-b')/a'$ (where $b'$ is the apparent semiminor axis); and the position angle P of the semiminor axis $b'$ .
 The coordinates $f_{c}$ and $g_{c}$, in kilometers, are calculated using the JPL\#33  Ceres' ephemeris and the star position given in Section \ref{Sec: observation-2013}.
@@
 \includegraphics[scale=0.36]{figures/Ceres_2013_body.pdf}
-\caption{\label{Fig: Ceres-2013-body}}
+\caption{The best elliptical fit for the occultation chords for the event of 2013 using timing from Table \ref{Tab: obs-2013} and the pole-constrained solution.
+The chord of Brookline if shifted by -64s as explained in the text\label{Fig: Ceres-2013-body}}
 \end{figure}

### 2017-08-16 Ceres/Old_versions/ArtigoV6/CeresV6.tex
mod 2017 ago16

--- 2015-05-16 Ceres/Old_versions/ArtigoV7/CeresV7.tex
@@
 
-\title[Results of two stellar occultations by Ceres]{Results of two multi-chords stellar occultations by dwarf planet (1) Ceres}
-\author[A. R. Gomes-J\'unior, B. L. Giacchini, F. Braga-Ribas et al.]{A. R. Gomes-J\'unior$^{1}$\thanks{E-mail:
-altair08@astro.ufrj.br},
-B. L. Giacchini$^{2,3,4}$\thanks{E-mail: breno@cbpf.br},
-F. Braga-Ribas$^{5, 6}$,
-M. Assafin$^{1}$,\newauthor
-R. Vieira-Martins$^{1,5}$
-T. Blank$^{4}$, 
-J. Brooks$^{4}$, 
-J. Broughton$^{7}$, 
-R. F. Dantowitz$^{8}$, \newauthor
-D. W. Dunham$^{4}$, 
-J. B. Dunham$^{4}$, 
-C. K. Ellington$^{4}$, 
-T. George$^{4}$, 
-P. D. Maley$^{4,9}$, \newauthor
-B. Timerson$^{4}$, 
-N. S. Weber$^{8}$\\
+\title[Results of two stellar occultations by Ceres]{Results of two stellar occultations by dwarf planet (1) Ceres}
+\author[A. V. Raveendran and A. N. Other]{A. V. Raveendran$^{1}$\thanks{E-mail:
+email@address (AVR); otheremail@otheraddress (ANO)} and A. N.
+Other$^{2}$\footnotemark[1]\thanks{}\\
 $^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
 CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
-$^{2}$Centro Brasileiro de Pesquisas F\'isicas, Rua Dr. Xavier Sigaud 150, Rio de Janeiro  22290-180, Brazil\\
-$^{3}$Se\c{c}\~ao de Oculta\c{c}\~oes/REA-Brasil, Av. Prof. Magalh\~aes Penido 403/302, Belo Horizonte 31270-700, Brazil\\
-$^{4}$International Occultation Timing Association, P.O. Box 131034, Houston, TX 77219-1034, USA\\
-$^{5}$Observat\'orio Nacional/MCT, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
-$^{6}$Federal University of Technology - Paraná (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil\\
-$^{7}$RASNZ Occultation Section, 18 Branch Cr. Reedy Creek, QLD 4227, Australia\\
-$^{8}$Clay Center Observatory at Dexter Southfield, 20 Newton Street, Brookline, MA 02445, USA\\
-$^{9}$NASA Johnson Space Center Astronomical Society, 4535 Cedar Ridge Trail, Houston, TX 77059 USA}
-
+$^{2}$Observat\'orio Nacional/MCT, R. General Jos\'e Cristino 77,
+CEP 20921-400 Rio de Janeiro - RJ, Brazil}
 \begin{document}
@@
 The low apparent magnitude of Ceres, if compared to most asteroids, imposes a somewhat strong constraint on the stars capable of causing a detectable magnitude drop when occulted.
-For instance, after the 1984 event, to our knowledge, only four stellar occultations have been observed.
+For instance, after the 1984 event, to our knowledge, only four like this have been observed.
 Two of them had only two chords each, not sufficient, thus, for providing accurate results\footnote{These events took place on 22 August 1994 and 30 October 2010.}.
@@
 
-This work is organized as follows. In Sections 2 and 3 we analyse, the 2010 and the 2013 events, respectively. The comparison of both results to those on the literature is carried out in Section 4.
+This work is organized as follows. In Sections 2 and 3 we analyse, on this order, the 2010 and the 2013 events. The comparison of both results to those on the literature is carried out in Section 4.
 
-\section{The 2010 occultation}
+\section[]{The 2010 occultation}
 
@@
 %
-On 17 August 2010 Ceres was predicted to occult the star TYC 6833-163-1 (UCAC2 20678706), which has magnitude $K=9.18$ and whose ICRS position is \citep{Zacharias2004}:
-% UCAC4 313-111823
+On 17 August 2010 Ceres was predicted to occult the star TYC 6833-163-1 (UCAC2 20678706), which has magnitude $K=9.18$ and whose ICRF/J200 position is \citep{Zacharias2004}:
+%
 \begin{equation}
@@
 \right.
-\label{Eq: 2010-coord}
 \end{equation}
-%
-% Abaixo a coordenada UCAC4 aplicada de Movimento Próprio
-%
-%\begin{equation}
-%\left\{ 
-%  \begin{array}{l l}
-%    \alpha = 17^{\text{h}}18^{\text{m}}29^{\text{s}}.009\\
-%    \delta = -27\degr 26\arcmin 38\arcsec.866.
-%  \end{array}
-%\right.
-%\label{Eq: 2010-coord}
-%\end{equation}
 %
@@
 
-Observations were carried out in Brazil at five stations as displayed in Table \ref{Tab: obs-2010} and Fig.
-\ref{Fig: Ceres-2010-map}.
+Observations were carried out in Brazil at five stations as displayed in Table~1 and Fig.~\ref{Fig: Ceres-2010-map}.
 The occultation was detected on four of them.
-UFSC had a negative chord.
-From the positive ones, INPE started the shots with the event already in progress and only provided star's reappearance time; the other three recorded the whole phenomenon.
+Among these, one started the shots with the event already in progress and only provided star's reappearance time; the other three recorded the whole phenomenon.
 
-A remarkable circumstance of this event was the low velocity of Ceres: only $3.9~km.s^{-1}$ in the plane of the sky.
+A remarkable circumstance of this event was the low velocity of Ceres: only 3.9 km s$^{-1}$ in the plane of the sky.
 Therefore, even exposures of a few seconds would translate in a relevant spatial resolution.
@@
   \hline
-          & Longitude & Telescope & Exposure & Result &   \\
-     Site & Latitude  & Aperture  & Cycle time & Ingress & Observer\\          
-          & Height    & Camera    &        & Egress    & \\          
+     Site & Longitude & Telescope & Exposure & Result & Observer  \\
+          & Latitude  & Aperture  & Cycle time & Ingress & \\          
+          & Height    & Camera    &    S/N    & Egress    & \\          
 \hline
@@
 The light curves were normalized to an unocculted star flux.
-Additionally, with exception of INPE, they were normalized by fitting a polynomial curve (of first or second order) outside the flux drop, so that the flux ratio was set to 1 outside the occultation.
+Additionally --with exception of INPE-- they were were normalized by fitting a polynomial curve (of first or second order) outside the flux drop, so that the flux ratio was set to 1 outside the occultation.
 
@@
 The free parameter to adjust is the ingress (disappearance) or egress (reappearance) time $t_{occ}$, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$.
-The best fittings to the 2010 occultation light curves are shown in Fig.
-\ref{Fig: Ceres-2010-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2010}.
+The best fittings to the occultation light curves are shown in Fig.~\ref{Fig: Ceres-2010-curves}, and the derived occultation times are listed in Table~\ref{Tab: obs-2010}.
 
@@
 This model is supported by the work of \cite{Drummond2014}, by means of direct imaging of Ceres.
-Hence, we have $N=7$ chord extremities to adjust the $M=5$ parameters which define an ellipse: apparent semimajor and semiminor axis ($a^\prime$ and $b^\prime$, respectively), position angle $P$ of its semiminor axis and the position $(f_c,g_c)$ of its centre with respect to the occulted star.
-The coordinates $f_{c}$ and $g_{c}$, in kilometres, are calculated using the JPL\#26 Ceres' ephemeris and the occulted star's position.
+Hence, we have $N=7$ chord extremities to adjust the $M=5$ parameters which define an ellipse: apparent semimajor and semiminor axis --$a^\prime$ and $b^\prime$, respectively--, position angle $P$ of its semiminor axis and the position $(f_c,g_c)$ of its centre with respect to the occulted star\footnote{The coordinates $f_{c}$ and $g_{c}$, in kilometres, are calculated using the JPL\#33 Ceres' ephemeris and the occulted star's position.
 They are positive toward the local celestial east and north, respectively.
-The position angle $P$ is counted positively from the direction of local celestial north to celestial east.
+The position angle $P$ is counted positively from the direction of local celestial north to celestial east.}.
 The apparent oblateness can be defined by $\epsilon^\prime = 1 - (b^\prime/a^\prime)$.
@@
 \epsilon^\prime = 1 - \sqrt{\cos^2(\zeta) + (1-\epsilon)^2\sin^2(\zeta)},
-\label{Eq: TrueObla}
+\label{TrueObla}
 \end{equation}
@@
 As we shortly show, it led to a rather large error bar on the position angle.
-Furthermore, the nominal solution solely is not capable of returning the true oblateness, which can be evaluated through equation \ref{Eq: TrueObla}, provided that the angle $\zeta$ is known.% or constrained by some physical hypothesis.
+Furthermore, the nominal solution solely is not capable of returning the true oblateness, which can be evaluated through equation \eqref{TrueObla}, provided that the angle $\zeta$ is known.% or constrained by some physical hypothesis.
 
@@
 %\end{array} \right .
-\label{Eq: Pole}
+\label{Pole}
 \end{equation}
@@
 The knowledge of Ceres' pole not only allows the determination of its polar aspect, it suffices to set its position angle.
-Therefore Eq.
-\ref{Eq: Pole} may act as a constraint for $P$, and a second solution can be obtained by probing the parameter space with the restriction that the position angle is confined to the range that follows from Eq.
-\ref{Eq: Pole}.
+Therefore \eqref{Pole} may act as a constraint for $P$, and a second solution can be obtained by probing the parameter space with the restriction that the position angle is confined to the range that follows from \eqref{Pole}.
 We call this the ``pole-constrained solution''.
@@
 Already mentioned, the parameter with the largest uncertainty is the position angle: spanning on a 20$\degr$ interval, its determination has a relative precision worse than 10$\%$.
-Clearly, the coordinates of Ceres' pole (Eq.
-\ref{Eq: Pole}) can impose a strong constraint on the position angle, as the next solution shows.
+Clearly, the coordinates of Ceres' pole (\ref{Pole}) can impose a strong constraint on the position angle, as the next solution shows.
 
@@
 
-At the moment of the occultation, the coordinates (Eq.
-\ref{Eq: Pole}) of Ceres' rotational pole correspond to the position angle $P = (12 \pm 3)\degr$.
-Exploration of the parameter space, restricted to ellipses whose position angle lies on this range, results in the pole-constrained solution.
-The related physical parameters are displayed on the third column of Table \ref{Tab: results}, while the best-fitting solution is depicted in Fig.\ref{Fig:Ceres-2010-body}.
+At the moment of the occultation, the coordinates (\ref{Pole}) of Ceres' rotational pole correspond to the position angle $P = (12 \pm 3)\degr$.
+Exploration of the parameter space, restricted to ellipses whose position angle lie on this range, results in the pole-constrained solution.
+The related physical parameters are displayed on the third column of Table~2, while the best-fitting solution is depicted in Fig.\ref{Fig:Ceres-2010-body}.
 
@@
 
-\section{The 2013 occultation}
+\section[]{The 2013 occultation}
 
@@
 
-The event which took place on 25 October 2013 involved the star TYC 865-911-1 (UCAC4 496-058191), of magnitude $K = 8.76$.
-Based on UCAC4 \citep{Zacharias2013}, the ICRS position to the date of occultation corrected of proper motion is :
+The event which took place on 25 October 2013 involved the star TYC 865-911-1 (UCAC2 35031360), of magnitude $K = 8.76$. Its ICRF/J2000 position is \citep{Zacharias2004}:
 %
@@
     \alpha = 11^{h}57^{m}52^{s}.7641\\
-    \delta = +09\degr 07\arcmin 49\arcsec.835
+    \delta = +09\degr 07\arcmin 49\arcsec.835.
   \end{array}
@@
 
-\begin{figure*}
-\begin{centering}
-\subfigure{\includegraphics[scale=0.42]{figures/Ceres_2013.png}}
-\hspace{5mm}
-\subfigure{\includegraphics[scale=0.42]{figures/Ceres_2013-zoom.png}}
+\begin{figure}
+\includegraphics[scale=0.42]{figures/Ceres_2013.png} 
 \caption{Post-occultation reconstruction of Ceres' shadow path on Earth for the 2013 October 25 event at the east coast of USA.
-Left panel: Overview of the occultation.
-The big red dot is the geocentric closest approach at 09:43:03 UT and the small ones represent the centre of the shadow separated by 30 seconds where the shadow moves from the left to the right.
-The blue dots are the sites that have observed the event.
-Right Panel: Upper view of the occultation over the sites that observed the event.
-Red points are the centre of the shadow separated by 5 seconds.
-\label{Fig: Ceres-2013-map}}
-\end{centering}
-\end{figure*}
+The big dot is the geocentric closest approach at 09:43:03 UT and the small ones represent the centre of the shadow separated by 30 seconds where the shadow moves from the left to the right.
+The blue dots are the sites that have observed the event.\label{Fig: Ceres-2013-map}}
+\end{figure}
 
@@
 
-All videos were converted to FITS images and the photometry of the target was obtained via PRAIA \citep{2011gfun.conf...85A}.
+All videos were converted to FITS images and the photometry of the target was obtained via the PRAIA \citep{2011gfun.conf...85A}.
 The light curves were normalized to the flux of a reference star when such was available on the field.
@@
 
-Elliptic limb profiles were adjusted to the eight\footnote{Brookline's chord was not used for limb fitting since it actually measured duration instead of absolute times.} chords by the same procedure described in Section \ref{Sec: limbfittingmethod}\footnote{For this event, the JPL\#33 Ceres' ephemeris was used to obtain the $f_{c}$ and $g_{c}$ coordinates.}.
+Elliptic limb profiles were adjusted to the eight\footnote{Brookline's chord was not used for limb fitting since it actually measured duration instead of absolute times.} chords by the same procedure described in Section~\ref{Sec: limbfittingmethod}.
 This yielded $\chi^2_{r,min} = 13$, suggesting that an elliptic model is not satisfactory to the data.
-Indeed, a quick glance at a plot of the observed chords, such as in Fig.
-\ref{Fig: Ceres-2013-body}, shows that the Varina chord seems to be somewhat advanced with respect to the others.
+Indeed, a quick glance at a plot of the observed chords --such as in Fig.~\ref{Fig: Ceres-2013-body}-- shows that the Varina chord seems to be somewhat advanced with respect to the others.
 Taking into account that in this station time stamps were not inserted on the video frames, it is always possible to attribute this advance to an eventual problem on the correspondence between camcorder's and GPS' times.
@@
 
-The immersion recorded at Owings also seems to be shifted (delayed) with respect to the nearby chords (see Fig.
-\ref{Fig: Ceres-2013-body}).
-This chord, actually, has roughly the same length of Mechanicsville's, despite the fact that they are separated by circa $100~km$.
+The immersion recorded at Owings also seems to be shifted (delayed) with respect to the nearby chords --see Fig.~\ref{Fig: Ceres-2013-body}.
+This chord, actually, has roughly the same length of Mechanicsville's, despite the fact that they are separated by circa 100~km.
 Differently from Varina, though, this station had time stamps inserted in each frame of the video, which makes it more unlike to justify an eventual timing issue.
@@
 Actually, the uncertainty of the former is of $5\degr$ in contrast to $10\degr$ of the latter.
-This suggests that the pole-constrained solution obtained via the position of Ceres' pole (Eq.
-\ref{Eq: Pole}) would not be significantly different of the nominal one.
+This suggests that the pole-constrained solution obtained via the position of Ceres' pole (\ref{Pole}) would not be significantly different of the nominal one.
 
-This assumption is confirmed when we carry out the limb fitting with the constraint $P = (25 \pm 3)\degr$, the position angle at the moment of the occultation which follows from (Eq.
-\ref{Eq: Pole}).
+This assumption is confirmed when we carry out the limb fitting with the constraint $P = (25 \pm 3)\degr$ --the position angle at the moment of the occultation which follows from (\ref{Pole}).
 The physical parameters related to this pole-constrained solution are presented in the last column of Table~2, and are essentially the same of the nominal solution.
@@
 
-\section{Discussion}
+\section[]{Discussion}
 
@@
 
-The 2010 event, for example, had only seven contacts; none the less, they were well distributed over Ceres' disc (see Fig.
-\ref{Fig:Ceres-2010-body}) acting as a constraint to is shape.
+The 2010 event, for example, had only seven contacts; none the less, they were well distributed over Ceres' disc (see Fig.~\ref{Fig:Ceres-2010-body}) acting as a constraint to is shape.
 On the other hand, the 2013 event had five more contacts, but they were concentrated in certain regions of the body.
@@
 
-Comparison of Ceres' equatorial diameter as measured by different techniques is carried out in Table \ref{Tab: Ceres-final}.
+Comparison of Ceres' equatorial diameter as measured by different techniques is carried out in Table~4.
 We note an overall agreement between our result to those obtained via direct imaging by the Hubble Space Telescope (HST) \citep{Thomas2005}, the Keck Observatory and the ESO VLT \citep{Drummond2014}.
@@
 
-The geocentric positions for Ceres at the time of the occultations could also be obtained with great accuracy. These are given in Eqs. \ref{Eq: Ceres-2010_pos} and \ref{Eq: Ceres-2013_pos}.
-
-\begin{equation}
-2010 ~ Aug ~ 17
-\left\{ 
-  \begin{array}{l l}
-    Time = 22:40:25\\
-    \alpha = 17^{\text{h}}18^{\text{m}}29^{\text{s}}.008\\
-    \delta = -27\degr 26\arcmin 38\arcsec.890.
-  \end{array}
-\right.
-\label{Eq: Ceres-2010_pos}
-\end{equation}
-
-\begin{equation}
-2013 ~ Oct ~ 25
-\left\{ 
-  \begin{array}{l l}
-    Time = 09:43:03\\
-    \alpha = 11^{h}57^{m}52^{s}.7641\\
-    \delta = +09\degr 07\arcmin 49\arcsec.835
-  \end{array}
-\right.
-\label{Eq: Ceres-2013_pos}
-\end{equation}
-
-\begin{table}
-  \caption{Ceres' equatorial diameter and oblateness \label{Tab: Ceres-final}}
-  \begin{centering}
+\begin{table*}
+ \centering
+ \begin{minipage}{140mm}
+  \caption{Ceres' equatorial diameter and oblateness}
   \begin{tabular}{@{}llcc}
@@
 \end{tabular}
-\par\end{centering}
 \textbf{References.} 1: Present work. 2: \cite{Drummond2014}. 3: \cite{Carry2008}. 4: \cite{Thomas2005}. 5: \cite{Millis1987}.
-\end{table}
+\end{minipage}
+\end{table*}
 
@@
 
-ARGJ thanks the financial support of CAPES.
-BLG thanks CNPq.
-F.B.R.
-acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014.
-We also acknowledge Steve Preston for the predictions of the occultations.
+Agradecimentos
 
@@
 
+\bibitem[\protect\citeauthoryear{Widemann et al.}{2009}] {Widemann2009} Widemann T. et al., 2009,
+Icarus, 199, 458
+
 \bibitem[\protect\citeauthoryear{van Belle}{1999}] {vanBelle1999} van Belle G.T., 1999,
@@
 
-\bibitem[\protect\citeauthoryear{Widemann et al.}{2009}] {Widemann2009} Widemann T. et al., 2009,
-Icarus, 199, 458
-
 \bibitem[\protect\citeauthoryear{Zacharias et al.}{2004}]{Zacharias2004} Zacharias N. et al., 2004,
 AJ, 127, 3043
-
-\bibitem[\protect\citeauthoryear{Zacharias et al.}{2013}]{Zacharias2013} Zacharias N. et al., 2013,
-AJ, 145, 44
 

### 2017-08-16 Ceres/Old_versions/ArtivoV9/CeresV9.tex
mod 2017 ago16

--- 2015-05-16 Ceres/paper-Ceres-source/paper-Ceres.tex
@@
 \usepackage{subfigure}
-\usepackage{epstopdf}
 \usepackage{color}
@@
 B. L. Giacchini$^{2,3,4}$,%\thanks{E-mail: breno@cbpf.br},
-F. Braga-Ribas$^{5, 6, \dag}$,
+F. Braga-Ribas$^{5, 6, \dag}$
 M. Assafin$^{1, \dag, \ddag}$,\newauthor
@@
 F.R. Herpich$^{11}$,  \newauthor
-C. Jacques$^{3, 12}$,
+C. Jacques$^{12}$,
 P. D. Maley$^{4,13}$,
 L. Mehret$^{10}$,
-A.J.T. Mello$^{14}$,
-A.C. Milone$^{15}$,\newauthor
-E. Pimentel$^{3, 12}$,  
+A.J.T. Mello$^{11}$,
+A.C. Milone$^{14}$,\newauthor
+E. Pimentel$^{12}$,  
 W. Schoenell$^{11}$,
@@
 $^{2}$Centro Brasileiro de Pesquisas F\'isicas, Rua Dr. Xavier Sigaud 150, Rio de Janeiro  22290-180, Brazil\\
-$^{3}$Se\c{c}\~ao de Oculta\c{c}\~oes/REA-Brasil, Belo Horizonte, Brazil\\
+$^{3}$Se\c{c}\~ao de Oculta\c{c}\~oes/REA-Brasil, Av. Prof. Magalh\~aes Penido 403/302, Belo Horizonte 31270-700, Brazil\\
 $^{4}$International Occultation Timing Association, P.O. Box 131034, Houston, TX 77219-1034, USA\\
-$^{5}$Observat\'orio Nacional/MCTI, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
+$^{5}$Observat\'orio Nacional/MCT, R. General Jos\'e Cristino 77, CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
 $^{6}$Federal University of Technology - Paran\'a (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil\\
@@
 $^{13}$NASA Johnson Space Center Astronomical Society, Houston, Texas, USA\\
-$^{14}$Federal University of Technology - Paran\'a (UTFPR / DAELT), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil\\
-$^{15}$Instituto Nacional de Pesquisas Espaciais, S\~ao Jos\'e dos Campos, Brazil\\
+$^{14}$Instituto Nacional de Pesquisas Espaciais, S\~ao Jos\'e dos Campos, Brazil\\
 $^\dag$ Associated to Laborat\'{o}rio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, CEP 20921-400,\\ Rio de Janeiro, Brazil\\
@@
 The first one is a nominal solution with an indeterminate polar aspect angle.
-The second one was constrained by the pole coordinates as given by Drummond et al.
-Assuming a Maclaurin spheroid, we determine an equatorial diameter of 972 $\pm$ 6 km and an apparent oblateness of $0.08 \pm 0.03$ as our best solution.
+The second one was constrained by the pole coordinates as given by \cite{Drummond2014}.
+Assuming a Maclaurin spheroid, we determine an equatorial diameter of $972 \pm 6~km$ and an apparent oblateness of $0.08 \pm 0.03$ as our best solution.
 These results are compared to all available size and shape determinations for Ceres made so far, and shall be confirmed by the NASA's \textit{Dawn} space mission.
@@
 In Section 4 we give the geocentric positions of Ceres derived from the occultations.
-The comparison of our results to those in the literature is carried out in Section 5.
+The comparison of both results to those in the literature is carried out in Section 5.
 
@@
 
-As predicted, on 2010 August 17 Ceres occulted the star TYC 6833-163-1 (UCAC4 313-111823), which has magnitude $V = 11.55$ and has the ICRS position, based on UCAC4 catalogue \citep{Zacharias2013}, to the date of occultation:
+As predicted, on 2010 August 17 Ceres occulted the star TYC 6833-163-1 (UCAC4 313-111823), which has magnitude $V = 11.55$ and has the ICRS position, based on UCAC4 catalogue \citep{Zacharias2013}, to the date of occultation, corrected of proper motion:
 
@@
     \alpha = 17^{\rmn{h}}18^{\rmn{m}}29\fs0085\\
-    \delta = -27\degr 26\arcmin 38\farcs 867
+    \delta = -27\degr 26\arcmin 38\farcs 867.
   \end{array}
@@
 The southern-most one (UFSC) had a negative chord.
-From the positive sites, the one named INPE started observing after the event was already in progress, due to technical difficulties, thus providing only the star's reappearance time; the other three recorded the whole phenomenon.
+From the positive ones, the one named INPE started observing after the event was already in progress, thus providing only the star's reappearance time; the other three recorded the whole phenomenon.
 
-A remarkable circumstance of this event was the low velocity of Ceres: only 3.9 km s$^{-1}$ in the plane of the sky.
+A remarkable circumstance of this event was the low velocity of Ceres: only $3.9~km~s^{-1}$ in the plane of the sky.
 Therefore, even exposures of a few seconds would translate in a relevant spatial resolution.
@@
 \hline
- Belo Horizonte & 43$\degr$59\arcmin 51\farcs1~W & LX200 & 5 s & Positive & C. Jacques  \\
- CEAMIG-REA &19$\degr$49\arcmin 49\farcs0~S & 31 cm & 12 s & 22:39:03.9 $\pm$ 0.6 s &  E. Pimentel \\
+ Belo Horizonte & 43$\degr$59\arcmin 51.1\arcsec~W & LX200 & 5 s & Positive & C. Jacques  \\
+ CEAMIG-REA &19$\degr$49\arcmin 49.0\arcsec~S & 31 cm & 12 s & 22:39:03.9 $\pm$ 0.6 s &  E. Pimentel \\
             & 825 m       &  f/10         &   SBIG ST10        & 22:40:20 $\pm$ 5 s &   \\
  & & & & & \\
- Pico dos Dias    & 45$\degr$34\arcmin45\farcs1~W &  Zeiss     & 1 s & Positive & J. I. B. Camargo \\
- LNA    &22$\degr$32\arcmin03\farcs7~S & 60 cm & 2 s & 22:37:30.3 $\pm$ 0.6 s &  G. B. Rossi \\
+ Pico dos Dias    & 45$\degr$34\arcmin45.1\arcsec~W &  Zeiss     & 1 s & Positive & J. I. B. Camargo \\
+ LNA    &22$\degr$32\arcmin03.7\arcsec~S & 60 cm & 2 s & 22:37:30.3 $\pm$ 0.6 s &  G. B. Rossi \\
             & 1864 m     & f/12.5          &    Andor Ikon       & 22:41:55.3 $\pm$ 0.7 s &              \\
  & & & & & \\
- S\~ao Jos\'e dos       & 45$\degr$51\arcmin44\farcs0~W &  C11  & 2 s & Egress only & A. C. Milone\\
- Campos       &23$\degr$12\arcmin33\farcs0~S & 28 cm & 5 s & Start obs.: 22:39:44 & T. Maldonado\\
+ S\~ao Jos\'e dos       & 45$\degr$51\arcmin43.0\arcsec~W &  C11  & 2 s & Egress only & A. C. Milone\\
+ Campos       &23$\degr$12\arcmin32.0\arcsec~S & 28 cm & 5 s & Start obs.: 22:39:44 & T. Maldonado\\
  INPE      & 620 m      & f/6.3          & SBIG ST7     & 22:42:03.0 $\pm$ 0.2 s & M. Okada    \\
  & & & & & \\
- Ponta Grossa       & 50$\degr$05\arcmin56\farcs0~W & RCX 400 &30 s & Positive & M. Emilio   \\
- UEPG       &25$\degr$05\arcmin22\farcs2~S & 40 cm & 52 s & 22:37:17 $\pm$ 13 s & L. Mehret   \\
+ Ponta Grossa       & 50$\degr$05\arcmin56.0\arcsec~W & RCX 400 &30 s & Positive & M. Emilio   \\
+ UEPG       &25$\degr$05\arcmin22.2\arcsec~S & 40 cm & 52 s & 22:37:17 $\pm$ 13 s & L. Mehret   \\
             & 910 m      & f/8          & SBIG STL6E     & 22:39:56 $\pm$ 13 s & \\
  & & & & & \\
- Florian\'opolis       & 48$\degr$31\arcmin20\farcs5~W &   C11      & 3 s & No occultation  & W. Schoenell\\
- UFSC       &27$\degr$36\arcmin12\farcs3~S & 28 cm & 6 s&  Start obs.: 21:49:27 & A. J. T. Mello\\
+ Florian\'opolis       & 48$\degr$31\arcmin20.5\arcsec~W &   C11      & 3 s & No occultation  & W. Schoenell\\
+ UFSC       &27$\degr$36\arcmin12.3\arcsec~S & 28 cm & 6 s&  Start obs.: 21:49:27 & A. J. T. Mello\\
             & 20 m       & f/10          &  SBIG ST7      &   End obs.: 22:51:21      & F. R. Herpich \\
@@
 
-\subsection{Light curves \label{Sec: light-curve-2010}}
+\subsection{Light curves and timing \label{Sec: light-curve-2010}}
 
@@
 This yields a diameter of about 0.11~km projected at Ceres' distance.
-The smallest integration time used in the positive observations was 1.0~s, which translates to almost 3.9 km in the celestial plane.
-Therefore, the error on the time determination of the ingress and egress is largely dominated by the integration times, not by Fresnel diffraction or star diameter, which are both on the order of a few hundred meters for this event.
+The smallest integration time used in the positive observations was 1.0~s, which translates to almost 3.9~km in the celestial plane.
+Therefore, the time error of ingress and egress are largely dominated by the integration times, not by Fresnel diffraction or star diameter, which are on the order a few hundreds meters for this event.
 The occultation data fit consists in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011} and BR13.\ The free parameter to adjust is the ingress or egress instant, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$.
@@
 The apparent oblateness can be defined by $\epsilon^\prime = 1 - (b^\prime/a^\prime)$.
-The best-fitting solution is obtained minimizing a reduced $\chi^{2}_{r}$ function, where we define the number of degrees of freedom of the problem as $\mathcal{N} \equiv N - M$.
+The best-fitting solution is obtained minimizing a reduced $\chi^{2}$ function, where we define the number of degrees of freedom of the problem as $\mathcal{N} \equiv N - M$.
 All the procedures that allow the determination of the error bars of the physical parameters can be found on BR13.
@@
 Two possible solutions were considered for the limb fitting.
-The first, which we call nominal solution, consists of determining the five parameters that characterize an ellipse from the seven observed contacts.
+The first is the nominal one, which consists of determining the five parameters that characterize an ellipse from the seven observed contacts.
 As we shortly show, it led to a rather large uncertainty on the position angle.
@@
 Position angle (deg)   & 5 $\pm$ 10    & \textbf{12 $\pm$ 3} (*)& 22 $\pm$ 5    & 25 $\pm$ 3 (*)\\
-$f_c$ (km)             & 97 $\pm$ 9   & \textbf{102 $\pm$ 5}   & 77 $\pm$ 6    & 78 $\pm$ 6\\
-$g_c$ (km)             & 16 $\pm$ 15  & \textbf{21 $\pm$ 11}  & 13 $\pm$ 16   & 13 $\pm$ 16\\
+$f_c$ (km)             & 133 $\pm$ 9   & \textbf{138 $\pm$ 5}   & 77 $\pm$ 6    & 78 $\pm$ 6\\
+$g_c$ (km)             & -17 $\pm$ 15  & \textbf{-11 $\pm$ 11}  & 13 $\pm$ 16   & 13 $\pm$ 16\\
 $\chi^2_{r,min}$       & 0.24          &  \textbf{0.42}         & 1.27          & 1.27\\
@@
 \begin{figure}
-\includegraphics[scale=0.37, angle=-90]{figures/Ceres_2010_sphere.eps} 
+\includegraphics[scale=0.36]{figures/Ceres_2010_sphere.pdf} 
 \caption{The best elliptical fit for the occultation chords for the event of 2010 using the times from Table~\ref{Tab: obs-2010} and the pole-constrained solution.
@@
 The event which took place on 2013 October 25 involved the star TYC 865-911-1 (UCAC4 496-058191), of magnitude $V = 10.05$.
-Based on UCAC4 \citep{Zacharias2013}, the ICRS position to the date of occultation is:
+Based on UCAC4 \citep{Zacharias2013}, the ICRS position to the date of occultation corrected of proper motion is:
 %
@@
 %
-The occultation could only be visible in the United States, before dawn, as depicted in Fig. \ref{Fig: Ceres-2013-map}.% The shadow would then cross the Atlantic and Africa, but during daylight time.
+The occultation could only be visible in the United States, before dawn.% The shadow would then cross the Atlantic and Africa, but during daylight time.
 
@@
 Each station was equipped with a video camera with negligible readout times.
-This was of particular importance, since in this event Ceres' shadow speed was 42.6 km s$^{-1}$, much faster than in the 2010 event.
+This was of particular importance, since in this event Ceres' shadow speed was $42.6~ km~s^{-1}$, much faster than in the 2010 event.
+All the observing sites are in the United States, as depicted in Fig.~\ref{Fig: Ceres-2013-map}.
 
@@
 \hline
- Hampton & 70$\degr$48\arcmin59\farcs7~W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
-  &42$\degr$53\arcmin52\farcs8~N & f/5 & 0.033 s     & 09:40:46.9 $\pm$ 0.1 s &   \\
+ Hampton & 70$\degr$48\arcmin59.7\arcsec~W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
+  &42$\degr$53\arcmin52.8\arcsec~N & f/5 & 0.033 s     & 09:40:46.9 $\pm$ 0.1 s &   \\
             & 7 m       &  &     & 09:40:57.26  $\pm$ 0.08 s &   \\
  & & & & & \\
- Topsfield & 70$\degr$55\arcmin16\farcs6~W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
-  &42$\degr$37\arcmin55\farcs9~N & f/5      & 0.033 s   & 09:40:45.4 $\pm$ 0.1 s &   \\
+ Topsfield & 70$\degr$55\arcmin16.6\arcsec~W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
+  &42$\degr$37\arcmin55.9\arcsec~N & f/5      & 0.033 s   & 09:40:45.4 $\pm$ 0.1 s &   \\
             & 45 m      &  &     & 09:40:58.0  $\pm$ 0.1 s &   \\
  & & & & & \\
- Brookline & 71$\degr$08\arcmin14\farcs5~W & 64 cm & Infinity2-1R & Positive & N. Weber \\
-  &42$\degr$18\arcmin27\farcs4~N & f/9.6 & 0.015 s     & 09:41:48.00 $\pm$ 0.01 s &  R. Dantowitz  \\
+ Brookline & 71$\degr$08\arcmin14.5\arcsec~W & 64 cm & Infinity2-1R & Positive & N. Weber \\
+  &42$\degr$18\arcmin27.4\arcsec~N & f/9.6 & 0.015 s     & 09:41:48.00 $\pm$ 0.01 s &  R. Dantowitz  \\
             & 109 m     &       &     & 09:42:02.93 $\pm$ 0.01 s   \\
  & & & & & \\
- Winchester & 78$\degr$14\arcmin39\farcs6~W & 36 cm & PC164C & Positive & J. Brooks \\
-  &39$\degr$16\arcmin21\farcs5~N & f/5 & 0.033 s    & 09:40:33.26 $\pm$ 0.08 s &   \\
+ Winchester & 78$\degr$14\arcmin39.6\arcsec~W & 36 cm & PC164C & Positive & J. Brooks \\
+  &39$\degr$16\arcmin21.5\arcsec~N & f/5 & 0.033 s    & 09:40:33.26 $\pm$ 0.08 s &   \\
             & 211 m     &  &     & 09:40:55.86  $\pm$ 0.09 s &   \\
  & & & & & \\
- Greenbelt & 76$\degr$52\arcmin09\farcs4~W & 12 cm & PC164C-EX2 & Positive & J. Dunham \\
-  &38$\degr$59\arcmin12\farcs1~N & f/2.5 & 0.033 s    & 09:40:33.6 $\pm$ 0.1 s & D. Dunham  \\
+ Greenbelt & 76$\degr$52\arcmin09.4\arcsec~W & 12 cm & PC164C-EX2 & Positive & J. Dunham \\
+  &38$\degr$59\arcmin12.1\arcsec~N & f/2.5 & 0.033 s    & 09:40:33.6 $\pm$ 0.1 s & D. Dunham  \\
             & 52 m      &   &     & 09:40:56.4  $\pm$ 0.1 s &   \\
  & & & & & \\
- Alexandria & 77$\degr$02\arcmin28\farcs3~W & 7 cm & Watec120N & Positive & P. Maley \\
-  &38$\degr$49\arcmin19\farcs1~N & f/10 &  0.067 s   & 09:40:33.3 $\pm$ 0.1 s &   \\
+ Alexandria & 77$\degr$02\arcmin28.3\arcsec~W & 7 cm & Watec120N & Positive & P. Maley \\
+  &38$\degr$49\arcmin19.1\arcsec~N & f/10 &  0.067 s   & 09:40:33.3 $\pm$ 0.1 s &   \\
             & 8 m       &   &     & 09:40:56.1  $\pm$ 0.1 s &   \\
  & & & & & \\
- Owings & 76$\degr$38\arcmin06\farcs3~W & 25 cm & PC164C  & Positive & C. Ellington \\
-  &38$\degr$41\arcmin26\farcs5~N & f/3.3 &  0.033 s & 09:40:34.27 $\pm$ 0.05 s &   \\
+ Owings & 76$\degr$38\arcmin06.3\arcsec~W & 25 cm & PC164C  & Positive & C. Ellington \\
+  &38$\degr$41\arcmin26.5\arcsec~N & f/3.3 &  0.033 s & 09:40:34.27 $\pm$ 0.05 s &   \\
             & 38 m                &       &     & 09:40:56.0  $\pm$ 0.2 s &   \\
  & & & & & \\
- Mechanicsville & 77$\degr$23\arcmin06\farcs7~W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
-  &37$\degr$41\arcmin26\farcs1~N & f/2.5 & 0.033 s    & 09:40:33.0 $\pm$ 0.1 s &   \\
+ Mechanicsville & 77$\degr$23\arcmin06.7\arcsec~W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
+  &37$\degr$41\arcmin26.1\arcsec~N & f/2.5 & 0.033 s    & 09:40:33.0 $\pm$ 0.1 s &   \\
             & 60 m      &   &     & 09:40:54.8  $\pm$ 0.1 s &   \\
  & & & & & \\
- Varina & 77$\degr$19\arcmin49\farcs3~W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
-  &37$\degr$25\arcmin58\farcs6~N & f/2.5 &  0.033 s   & 09:40:32.4 $\pm$ 0.3 s &   \\
+ Varina & 77$\degr$19\arcmin49.3\arcsec~W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
+  &37$\degr$25\arcmin58.6\arcsec~N & f/2.5 &  0.033 s   & 09:40:32.4 $\pm$ 0.3 s &   \\
             & 19 m      &   &     & 09:40:53.1  $\pm$ 0.2 s &   \\
@@
 
-\subsection{Light curves}
+\subsection{Light curves and timing}
 
@@
 \ref{Fig: Ceres-2013-body}).
-This chord, actually, has roughly the same length of Mechanicsville's, despite the fact that they are separated by about 100 km.
+This chord, actually, has roughly the same length of Mechanicsville's, despite the fact that they are separated by about $100~km$.
 Differently from Varina, though, this station had time stamps inserted in each frame of the video, which makes it more unlikely to justify an eventual timing issue.
@@
 
-In the hypothesis that the delay observed in the immersion at Owings could be associated to a limb topography feature, the recorded contact would then correspond to a negative elevation of 31 $\pm$ 4 km with respect to the best-fitting ellipse.
+In the hypothesis that the delay observed in the immersion at Owings could be associated to a limb topography feature, the recorded contact would then correspond to a negative elevation of $31 \pm 4$ km with respect to the best-fitting ellipse.
 Theoretical models, however, predict that reliefs in Ceres should not be higher than about 10--20 km \citep{Johnson1973}, while published observational data sets the bound of 18 km \citep{Carry2008}.
@@
 The objects' geocentric positions derived from stellar occultations are most valuable for improving their orbits \citep{Desmars2015}.
-Usually the precision of the astrometric positions are limited by the accuracy of the occulted star position, not by the limb fit.
-Ceres' geocentric J2000 positions at the time of each occultation are displayed in Eqs.~\ref{Eq: Ceres-2010_pos} and \ref{Eq: Ceres-2013_pos}.
-The errors of the positions come from the errors of the star positions, taken from the catalogues and from the errors of the relative apparent distances between star and Ceres, derived from the limb fit (which are displayed as $\Delta\alpha \cos\delta$, $\Delta\delta$).
+The precision of these astrometric positions are only limited by the accuracy of the occulted star position, not by the limb fit.
+The geocentric J2000 positions of Ceres at the time of each occultation are displayed in Eqs.~\ref{Eq: Ceres-2010_pos} and \ref{Eq: Ceres-2013_pos}.
+The errors of Ceres positions come from the errors of the star positions, taken from the respective catalogues for the occultation epochs, and the errors of the relative apparent distances between star and Ceres, derived from the limb fits.
+For the record, we also display the small errors of the relative apparent distances ($\Delta\alpha$, $\Delta\delta$).
 
@@
 \begin{equation}
-2010 ~ \text{Aug} ~ 17
+2010 ~ Aug ~ 17
 \left\{
  \begin{array}{l l}
-    \text{Time} = 22:40:00\\
+    Time = 22:40:00\\
     \alpha = 17^{\rmn{h}}18^{\rmn{m}}29\fs0122\pm0\farcs027\\
     \delta = -27\degr 26\arcmin 38\farcs617\pm0\farcs028\\
-    \Delta\alpha\cos\delta = 0\farcs003; ~~\Delta\delta = 0\farcs007
+    \Delta\alpha = 0\farcs003; ~~\Delta\delta = 0\farcs007
   \end{array}
@@
 \begin{equation}
-2013 ~ \text{Oct} ~ 25
+2013 ~ Oct ~ 25
 \left\{
   \begin{array}{l l}
-    \text{Time} = 09:45:00\\
+    Time = 09:45:00\\
     \alpha = 11^{\rmn{h}}57^{\rmn{m}}52\fs9154\pm0\farcs019\\
     \delta = +09\degr 07\arcmin 49\farcs865\pm0\farcs021\\
-    \Delta\alpha\cos\delta = 0\farcs002; ~~\Delta\delta = 0\farcs007
+    \Delta\alpha = 0\farcs002; ~~\Delta\delta = 0\farcs007
   \end{array}
@@
 The differences of the solutions occur basically on the size of their error bars, and can be justified by the particularities of each set of data, as discussed below.
-
-The 2010 event, for example, had only seven contacts; none the less, they were well distributed over Ceres' disc (see Fig.
-\ref{Fig:Ceres-2010-body}) acting as a constraint to its shape.
-On the other hand, the 2013 event had five more exploitable contacts, but they were concentrated in certain regions of the body.
-In particular, the absence of chords close to Ceres' south pole made its oblateness worse determined here than in the 2010 event.
-
-However, even our best measurement for the oblateness, $\epsilon=0.08 \pm 0.03$, has still a higher uncertainty with regard to other figures published in the literature, as Table~\ref{Tab: Ceres-final} shows.
-A larger number of uniformly spaced chords would be necessary to offer a best constraint to the oblateness.
-
-The few chords of the 2010 occultation could themselves only constraint the position angle of the object to a uncertainty of $10\degr$.
-This uncertainty was reduced by a factor of two in the 2013 event, approaching --and verifying-- the result predicted by the work of \cite{Drummond2014}.
-As was shown, using the coordinates of Ceres' polar axis to limit the position angle was not an efficient procedure in the 2013 event, in the sense that it did not result in significant changes in the parameters obtained in the nominal solution (see Table \ref{Tab: results}).
-
-On the other hand, constrain the position angle on the 2010 occultation was proved to reduce the error bars of the other parameters (disregarding oblateness).
-Moreover, this procedure resulted in an excellent agreement between the equatorial radius figures of both events.
-
-The 2013 occultation, therefore, offers not only an independent verification of the figures resulted from the 2010 event, but also validates the procedure carried out there which led to the best-constrained parameters in this work.
-
-Comparison of Ceres' equatorial diameter as measured by different techniques is carried out in Table \ref{Tab: Ceres-final}.
-We note an overall agreement between our result to those obtained via direct imaging by the Hubble Space Telescope (HST) \citep{Thomas2005}, the Keck Observatory and the ESO VLT \citep{Drummond2014}.
-The smaller figure reported by \cite{Carry2008} may be justified by the fact that in this study the effects of limb-darking were not taken into account, as pointed out by \cite{Drummond2014}.
-
-As mentioned in the Introduction, the 1984 event \citep{Millis1987} is the only occultation data to which we can compare our result.
-The measured diameters do not agree, ours being larger by 2-sigma.
-It is difficult to state for sure the reasons for this divergence since we had no access to the original 1984 data or reduction methodology.
-We note, however, that in this occultation a variation of 0.5 s in the contact times correspond to almost 7 km on Ceres' limb, which is on the same order of the residuals of their elliptical fit.
-Therefore, in order to properly compare their result to ours it would be necessary to reduce their light curves with the same methodology used in the present work.
-
-NASA's \textit{Dawn} mission shall bring light to these questions, which will be important not only for the knowledge of Ceres itself, but also for all the techniques used so far to study the physical properties of small Solar System objects, such as the stellar occultations.
 
@@
 $975 \pm 4$  & $0.067 \pm 0.005$ & HST         & 4\\
-$959 \pm 5$  & $0.05  \pm 0.01$  & Occultation & 5\\
+$960 \pm 5$  & $0.05  \pm 0.01$  & Occultation & 5\\
 \hline
@@
 \end{table}
+
+The 2010 event, for example, had only seven contacts; none the less, they were well distributed over Ceres' disc (see Fig.
+\ref{Fig:Ceres-2010-body}) acting as a constraint to its shape.
+On the other hand, the 2013 event had five more exploitable contacts, but they were concentrated in certain regions of the body.
+In particular, the lack of chords in Ceres' southern hemisphere made its oblateness worse determined here than in the 2010 event.
+
+However, even our best measurement for the oblateness, $\epsilon=0.08 \pm 0.03$, has still a higher uncertainty with regard to other figures published in the literature, as Table~\ref{Tab: Ceres-final} shows.
+A larger number of uniformly spaced chords would be necessary to offer a best constraint to the oblateness.
+
+The few chords of the 2010 occultation could themselves only constraint the position angle of the object to a uncertainty of $10\degr$.
+This uncertainty was reduced by a factor of two in the 2013 event, approaching --and verifying-- the result predicted by the work of \cite{Drummond2014}.
+As was shown, using the coordinates of Ceres' polar axis to limit the position angle was not an efficient procedure in the 2013 event, in the sense that it did not result in significant changes in the parameters obtained in the nominal solution (see Table \ref{Tab: results}).
+
+On the other hand, constrain the position angle on the 2010 occultation was proved to reduce the error bars of the other parameters (disregarding oblateness).
+Moreover, this procedure resulted in an excellent agreement between the equatorial radius figures of both events.
+
+The 2013 occultation, therefore, offers not only an independent verification of the figures resulted from the 2010 event, but also validates the procedure carried out there --and which led to the best-constrained parameters in this work.
+
+Comparison of Ceres' equatorial diameter as measured by different techniques is carried out in Table \ref{Tab: Ceres-final}.
+We note an overall agreement between our result to those obtained via direct imaging by the Hubble Space Telescope (HST) \citep{Thomas2005}, the Keck Observatory and the ESO VLT \citep{Drummond2014}.
+The smaller figure reported by \cite{Carry2008} may be justified by the fact that in this study the effects of limb-darking were not taken into account, as pointed out by \cite{Drummond2014}.
+
+As mentioned in the Introduction, the 1984 event \citep{Millis1987} is the only occultation data to which we can compare our result.
+The measured diameters do not agree, ours being larger by 2-sigma.
+It is difficult to state for sure the reasons for this divergence since we had no access to the original 1984 data or reduction methodology.
+We note, however, that in this occultation a variation of 0.5 s in the contact times correspond to almost $7~km$ on Ceres' limb, which is on the same order of the residuals of their elliptical fit.
+Therefore, in order to properly compare their result to ours it would be necessary to reduce their light curves with the same methodology used in the present work.
+
+NASA's \textit{Dawn} mission shall bring light to these questions, which will be important not only for the knowledge of Ceres itself, but also for all the techniques used so far to study the physical properties of small Solar System objects, such as the stellar occultations.
 
@@
 
-\bibitem[\protect\citeauthoryear{Assafin et al.}{2011}] {2011gfun.conf...85A} Assafin M., et al., 2011, in Tanga P., Thuillot W., eds, Gaia follow-up network for the solar system objects: Gaia FUN-SSO workshop proceedings.
-IMCCE-Paris Observatory, Paris, p.
-85
+\bibitem[\protect\citeauthoryear{Assafin et al.}{2011}] {2011gfun.conf...85A} Assafin M.
+et al., 2011, Gaia follow-up network for the solar system objects : Gaia FUN-SSO workshop proceedings, held at IMCCE -Paris Observatory, France, November 29 - December 1, 2010.
+ISBN 2-910015-63-7
 
@@
 
-\bibitem[\protect\citeauthoryear{Castillo-Rogez}{2011}]{CastilloR2011} Castillo-Rogez J.~C.,
+\bibitem[\protect\citeauthoryear{Castillo-Rogez}{2011}]{CastilloR2011} Castillo-Rogez, J.~C. 
 2011, Icarus, 215, 599
 
-\bibitem[\protect\citeauthoryear{Desmars et al.}{2015}]{Desmars2015} Desmars J. et al., 2015, A\&A, submitted
+\bibitem[\protect\citeauthoryear{Desmars et al.}{2015}]{Desmars2015} Desmars, J. et al., 2015, A\&A, submitted
 
@@
 \bibitem[\protect\citeauthoryear{Dunham et al.}{2014}]{Dunham2014}
-Dunham D.W. et al., 2014, Asteroid Occultations V12.0. EAR-A-3-RDR-OCCULTATIONS-V12.0. NASA Planetary Data System
+Dunham, D.W. et al., 2014, Asteroid Occultations V12.0. EAR-A-3-RDR-OCCULTATIONS-V12.0. NASA Planetary Data System
 
-\bibitem[\protect\citeauthoryear{Gomes et al.}{2005}]{Gomes05} Gomes R., Levison H.~F., Tsiganis K., Morbidelli A.,\ 2005, Nature, 435, 466
+\bibitem[\protect\citeauthoryear{Gomes et al.}{2005}]{Gomes05} Gomes, R., Levison, H.~F., Tsiganis, K., \& Morbidelli, A.\ 2005, Nature, 435, 466
 
-\bibitem[\protect\citeauthoryear{Giorgini et al.}{1996}]{Giorgini1996} Giorgini, J. D. et al., 1996, BAAS, 28, 1158
+\bibitem[\protect\citeauthoryear{Giorgini et al.}{1996}]{Giorgini1996} Giorgini, J. D. et al. 1996, BAAS, 28, 1158
 
@@
 
-\bibitem[\protect\citeauthoryear{McKinnon}{2012}]{McKinnon12} McKinnon W.~B.,\ 2012, LPI 
+\bibitem[\protect\citeauthoryear{McKinnon}{2012}]{McKinnon12} McKinnon, W.~B.\ 2012, LPI 
 Contributions, 1667, 6475
@@
 
-\bibitem[\protect\citeauthoryear{Russell et al.}{2004}]{Russell2004} Russell C.~T. et al.,\ 2004, P\&SS, 52, 465
+\bibitem[\protect\citeauthoryear{Russell et al.}{2004}]{Russell2004} Russell, C.~T., Coradini, A., Christensen, U., et al.\ 2004, P\&SS, 52, 465
 

### 2017-08-16 Ceres/Old_versions/CeresV3.tex
mod 2017 ago16

--- 2017-08-16 Ceres/Old_versions/Altair_27_03_2015/CeresV4.tex
@@
 email@address (AVR); otheremail@otheraddress (ANO)} and A. N.
-Other$^{2}$\footnotemark[1]\thanks{}\\
-$^{1}$Observat\'orio do Valongo/UFRJ, Ladeira Pedro Ant\^onio 43,
-CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
-$^{2}$Observat\'orio Nacional/MCT, R. General Jos\'e Cristino 77,
-CEP 20921-400 Rio de Janeiro - RJ, Brazil}
+Other$^{2}$\footnotemark[1]\thanks{This file has been amended to
+highlight the proper use of \LaTeXe\ code with the class file.
+These changes are for illustrative purposes and do not reflect the
+original paper by A. V. Raveendran.}\\
+$^{1}$Indian Institute of Astrophysics, Bangalore 560034, India\\
+$^{2}$Building, Institute, Street Address, City, Code, Country}
 \begin{document}
 
-\date{Accepted . Received ; in original form }
+\date{Accepted 1988 December 15. Received 1988 December 14; in original form 1988 October 11}
 
-\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2015}
+\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2002}
 
@@
 \section[]{The 2010 occultation}
-
-TYC 6833-163-1
 
@@
 
-The flux of the star in the five occultation chords was obtained from the FITS images with the Platform for Reduction of Astronomical Images Automatically (PRAIA) \citep{2011gfun.conf...85A}.
+The flux of the star in the five occultation chords was obtained from the FITS images with the Platform for Reduction of Astronomical Images Automatically (PRAIA) \citep{Assafin2011}.
 The light curves were normalized to an unocculted star flux.
@@
 \section[]{The 2013 occultation}
-
-TYC 865-911-1
 

### 2017-08-16 Ceres/Old_versions/Felipe_04abr2015/Felipe_04abr2015/CeresV8.tex
mod 2017 ago16

--- 2017-08-16 Ceres/Old_versions/ArtivoV9/CeresV9.tex
@@
 \usepackage{color}
-\usepackage[breaklinks=true]{hyperref}
+\usepackage[breaklinks=true]{hyperref} %% to avoid \citeads line
 
@@
 
-\title[Results of two stellar occultations by Ceres]{Results of two multi-chord stellar occultations by dwarf planet (1) Ceres}
+\title[Results of two stellar occultations by Ceres]{Results of two multi-chords stellar occultations by dwarf planet (1) Ceres}
 \author[A. R. Gomes-J\'unior, B. L. Giacchini, F. Braga-Ribas et al.]{A. R. Gomes-J\'unior$^{1,}$\thanks{E-mail: altair08@astro.ufrj.br},
@@
 F. Braga-Ribas$^{5, 6, \dag}$
-M. Assafin$^{1, \dag, \ddag}$,\newauthor
-R. Vieira-Martins$^{1,5, \dag, \ddag}$,
+M. Assafin$^{1, \dag}$,\newauthor
+R. Vieira-Martins$^{1,5, \dag}$,
 J.I.B. Camargo$^{5, \dag}$,
@@
 $^{7}$LESIA, Observatoire de Paris, CNRS UMR 8109, Universit\'{e} Pierre et Marie Curie, Universit\'{e} Paris-Diderot, 5 place Jules Janssen,\\ F-92195 Meudon Cedex, France\\
-$^{8}$RASNZ Occultation Section, P.O. Box 3181, Wellington, New Zealand\\
+$^{8}$RASNZ Occultation Section, 18 Branch Cr. Reedy Creek, QLD 4227, Australia\\
 $^{9}$Clay Center Observatory at Dexter Southfield, 20 Newton Street, Brookline, MA 02445, USA\\
@@
 $^{12}$Centro de Estudos Astron\^omicos de Minas Gerais, Belo Horizonte, Brazil\\
-$^{13}$NASA Johnson Space Center Astronomical Society, Houston, Texas, USA\\
-$^{14}$Instituto Nacional de Pesquisas Espaciais, S\~ao Jos\'e dos Campos, Brazil\\
-$^\dag$ Associated to Laborat\'{o}rio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, CEP 20921-400,\\ Rio de Janeiro, Brazil\\
-$^\ddag$Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France
+$^{13}$NASA Johnson Space Center Astronomical Society, 4535 Cedar Ridge Trail, Houston, TX 77059 USA\\
+$^{14}$Instituto Nacional de Pesquisas Espaciais, São José dos Campos, Brazil\\
+$^\dag$ Associated to Laborat\'{o}rio Interinstitucional de e-Astronomia - LIneA, Rua Gal. Jos\'e Cristino 77, CEP 20921-400,\\ Rio de Janeiro, Brazil
 }
@@
 \begin{abstract}
-We report the results of two multi-chord stellar occultations by the dwarf planet (1) Ceres that were observed  from Brazil on 2010 August 17, and from the USA on 2013 October 25.
+We report the results of two multi-chord stellar occultations by the dwarf planet (1) Ceres that were observed  from Brazil on August 17 2010, and from USA on October 25 2013.
 Four positive detections were obtained for the 2010 occultation, and nine for the 2013 occultation.
-Elliptical models were adjusted to the observed chords to obtain Ceres' size and shape.
-Two limb fitting solutions were studied for each event.
+Elliptical models were adjusted to the observed chords to obtain the Ceres' size and shape.
+Two limb fitting solution were studied for each event.
 The first one is a nominal solution with an indeterminate polar aspect angle.
@@
 Assuming a Maclaurin spheroid, we determine an equatorial diameter of $972 \pm 6~km$ and an apparent oblateness of $0.08 \pm 0.03$ as our best solution.
-These results are compared to all available size and shape determinations for Ceres made so far, and shall be confirmed by the NASA's \textit{Dawn} space mission.
+These results are compared to all available size and shape determinations for Ceres made so far, and shall be confirmed by the Nasa's \textit{Dawn} space mission.
 \end{abstract}
@@
 
-Ceres is the sole example of a dwarf planet in the inner Solar System.
-Far from being mere taxonomic information, this suggests the great impact its study can have on the understanding of planetary formation and evolution of the Solar System.
-Indeed, it was proposed that Ceres' origin could have been as a transneptunian object \citep{McKinnon12}, later scattered to the Main Belt due to the giant planets' migration predicted by the `Nice Model' \citep{Gomes05}.
-Even if it was formed close to its current location, the dynamical history of the Solar System must have let its signatures on Ceres.
-These could include not only the late heavy bombardment features that might exist on its surface, but also the makeup of its volatiles, which could have been transported from the outer region.
+Ceres is the sole exemplar of a dwarf planet in the inner Solar System.
+Far from being a mere taxonomic information, this suggests the great impact its study can have on the understanding of planetary formation and evolution of the Solar System.
+Indeed, it was proposed that Ceres could have its origin as a transneptunian object \citep{McKinnon12}, later scattered to the Main Belt due to the giant planets' migration predicted by the `Nice Model' \citep{Gomes05}.
+It may well had been formed close to its current location, however, even on this scenario, the dynamical history of the Solar System must have let its signatures on Ceres.
+Not only on the late heavy bombardment features that might exist on its surface, but also on its volatiles, which could have been transported from the outer region.
 
-Since the 1970's it has been speculated that Ceres could contain water ice, which was recently verified \citep{Kuppers2014}.
+Since the 1970's it has been speculated that Ceres could contain water ice, what was recently verified \citep{Kuppers2014}.
 Although the water regime on this object is still unknown, some internal structure models suggest the existence of a water ice -- or even a liquid water -- layer \citep{CastilloR2011}.
-Yet the very question of whether Ceres underwent differentiation is open and, on the assumption of an affirmative answer, it is natural to ask if it ever had tectonics, what its geological evolution was, and if it is still active.
-Inarguably NASA's \textit{Dawn} mission \citep{Russell2004}, which is currently orbiting the dwarf planet, will shed light on several open issues concerning Ceres.
+Yet the very question of whether Ceres suffered differentiation is open and, on the assumption of an affirmative answer, it is natural to ask if it ever had tectonics, how its geological evolution was, and if it is still active.
+Inarguably NASA's \textit{Dawn} mission \citep{Russell2004}, that is currently orbiting the dwarf planet, will shed light on several open issues concerning Ceres.
 
-Containing approximately one fifth of the whole Main Belt's mass, Ceres is expected to have an equilibrium figure, i.e.
+Owning approximately one fifth of the whole Main Belt's mass, Ceres is expected to have an equilibrium figure, i.e.
 a Maclaurin or a Jacobi ellipsoid.
@@
 
-The first stellar occultation by Ceres was observed in 1984 \citep{Millis1987} and led to the determination of its size to the precision of some kilometres, at a time when the uncertainties were often ten times larger.
-The high apparent brightness of Ceres, as compared to most asteroids, imposes a somewhat strong constraint on the stars capable of causing a detectable magnitude drop when occulted.
-For instance, after the 1984 event, to our knowledge, only four stellar occultations have been observed \citep{Dunham2014}.
-Two of them had only two chords each, thus not sufficient for providing accurate results\footnote{These events took place on 1994 August 22 and 2010 October 30.}.
-The two remaining events, which occurred on 2010 August 17 and 2013 October 25, are reported in the present work and are the first ones that used charged-couple devices (CCD) as recording systems.
+The first stellar occultation by Ceres was observed in 1984 \citep{Millis1987} and led to the determination of its size to the precision of some kilometres, in a time when the uncertainties were often ten times larger.
+The low apparent magnitude of Ceres, if compared to most asteroids, imposes a somewhat strong constraint on the stars capable of causing a detectable magnitude drop when occulted.
+For instance, after the 1984 event, to our knowledge, only four stellar occultations have been observed.
+Two of them had only two chords each, not sufficient, thus, for providing accurate results\footnote{These events took place on 22 August 1994 and 30 October 2010.}.
+The two remaining events, which occurred on 17 August 2010 and 25 October 2013, are reported on the present work and are the first ones that used charged-couple devices (CCD) as recording systems.
 % The former was observed in Brazil from 4 stations, while the latter was from the United States and led to 9 chords.
@@
 This work is organized as follows.
-In Sections 2 and 3 we analyse the 2010 and the 2013 occultations, respectively.
+In Sections 2 and 3 we analyse, the 2010 and the 2013 occultations, respectively.
 In Section 4 we give the geocentric positions of Ceres derived from the occultations.
-The comparison of both results to those in the literature is carried out in Section 5.
+The comparison of both results to those on the literature is carried out in Section 5.
 
@@
 
-As predicted, on 2010 August 17 Ceres occulted the star TYC 6833-163-1 (UCAC4 313-111823), which has magnitude $V = 11.55$ and has the ICRS position, based on UCAC4 catalogue \citep{Zacharias2013}, to the date of occultation, corrected of proper motion:
+On 17 August 2010 Ceres was predicted to occult the star TYC 6833-163-1, %(UCAC2 20678706),
+ which has magnitude $V = 11.55$ and has the ICRS position:
+% based on UCAC4 catalogue \citep{Zacharias2013} to the date of occultation, corrected of proper motion:
 
@@
   \begin{array}{l l}
-    \alpha = 17^{\rmn{h}}18^{\rmn{m}}29\fs0085\\
-    \delta = -27\degr 26\arcmin 38\farcs 867.
+    \alpha = 17^{\text{h}}18^{\text{m}}29^{\text{s}}.008\\
+    \delta = -27\degr 26\arcmin 38\arcsec.890.
   \end{array}
@@
 \begin{figure}
-\includegraphics[scale=0.42]{figures/Ceres_2010.eps} 
+\includegraphics[scale=0.42]{figures/Ceres_2010.png} 
 \caption{Post-occultation reconstruction of Ceres' shadow path on Earth for the 2010 August 17 event.
@@
 \ref{Fig: Ceres-2010-map}.
-The occultation was detected from four of them.
-The southern-most one (UFSC) had a negative chord.
-From the positive ones, the one named INPE started observing after the event was already in progress, thus providing only the star's reappearance time; the other three recorded the whole phenomenon.
+The occultation was detected on four of them.
+The south most one (UFSC) had a negative chord.
+From the positive ones, the one named INPE started the observation when the event was already in progress, thus providing only the star's reappearance time; the other three recorded the whole phenomenon.
 
-A remarkable circumstance of this event was the low velocity of Ceres: only $3.9~km~s^{-1}$ in the plane of the sky.
+A remarkable circumstance of this event was the low velocity of Ceres: only $3.9~km.s^{-1}$ in the plane of the sky.
 Therefore, even exposures of a few seconds would translate in a relevant spatial resolution.
@@
 The fraction of a second could be retrieved as described on the Section~3.1 of \cite{BragaRibas2013}  (which shall be referred as BR13 hereafter).
-Cycle times (exposure plus read-out) varied from 2 to 52 seconds, as can be verified on Table~\ref{Tab: obs-2010}, making it a heterogeneous set of observations and imposing an error of a few seconds for the ingress / egress times of some sites.
+Cycle times (exposure plus read-out) varied from 2 to 52 seconds, as can be verified on Table~\ref{Tab: obs-2010}, making of it an heterogeneous set of observations and imposing an error of a few seconds for the ingress / egress times of some sites.
 
@@
 \hline
- Belo Horizonte & 43$\degr$59\arcmin 51.1\arcsec~W & LX200 & 5 s & Positive & C. Jacques  \\
- CEAMIG-REA &19$\degr$49\arcmin 49.0\arcsec~S & 31 cm & 12 s & 22:39:03.9 $\pm$ 0.6 s &  E. Pimentel \\
-            & 825 m       &  f/10         &   SBIG ST10        & 22:40:20 $\pm$ 5 s &   \\
+ Belo Horizonte & 43$\degr$59'51.1'' W & LX200 & 5 s & Positive & C. Jacques  \\
+ CEAMIG-REA &19$\degr$49'49.0'' S & 31 cm & 12 s & 22:39:03.9 $\pm$ 0.6 &  E. Pimentel \\
+            & 825 m       &  f/10         &   SBIG ST10        & 22:40:20 $\pm$ 5 &   \\
  & & & & & \\
- Pico dos Dias    & 45$\degr$34\arcmin45.1\arcsec~W &  Zeiss     & 1 s & Positive & J. I. B. Camargo \\
- LNA    &22$\degr$32\arcmin03.7\arcsec~S & 60 cm & 2 s & 22:37:30.3 $\pm$ 0.6 s &  G. B. Rossi \\
-            & 1864 m     & f/12.5          &    Andor Ikon       & 22:41:55.3 $\pm$ 0.7 s &              \\
+ Pico dos Dias    & 45$\degr$34'45.1'' W &  Zeiss     & 1 s & Positive & J. I. B. Camargo \\
+ OPD/LNA    &22$\degr$32'03.7'' S & 60 cm & 2 s & 22:37:30.3 $\pm$ 0.6 &  G. B. Rossi \\
+            & 1864 m     & f/12.5          &    Andor Ikon       & 22:41:55.3 $\pm$ 0.7 &              \\
  & & & & & \\
- S\~ao Jos\'e dos       & 45$\degr$51\arcmin43.0\arcsec~W &  C11  & 2 s & Egress only & A. C. Milone\\
- Campos       &23$\degr$12\arcmin32.0\arcsec~S & 28 cm & 5 s & Start obs.: 22:39:44 & T. Maldonado\\
- INPE      & 620 m      & f/6.3          & SBIG ST7     & 22:42:03.0 $\pm$ 0.2 s & M. Okada    \\
+ S\~ao Jos\'e dos       & 45$\degr$51'44.0'' W &  C11  & 2 s & Egress only & A. C. Milone\\
+ Campos       &23$\degr$12'33.0'' S & 28 cm & 5 s & Start obs.: 22:39:44 & T. Maldonado\\
+ INPE      & 975 m      & f/6.3          & SBIG ST7     & 22:42:03.0 $\pm$ 0.2 & M. Okada    \\
  & & & & & \\
- Ponta Grossa       & 50$\degr$05\arcmin56.0\arcsec~W & RCX 400 &30 s & Positive & M. Emilio   \\
- UEPG       &25$\degr$05\arcmin22.2\arcsec~S & 40 cm & 52 s & 22:37:17 $\pm$ 13 s & L. Mehret   \\
-            & 910 m      & f/8          & SBIG STL6E     & 22:39:56 $\pm$ 13 s & \\
+ Ponta Grossa       & 50$\degr$05'56.0'' W & RCX 400 &30 s & Positive & M. Emilio   \\
+ UEPG       &25$\degr$05'22.2'' S & 40 cm & 52 s & 22:37:17 $\pm$ 13 & L. Mehret   \\
+            & 910 m      & f/8          & SBIG STL6E     & 22:39:56 $\pm$ 13 & \\
  & & & & & \\
- Florian\'opolis       & 48$\degr$31\arcmin20.5\arcsec~W &   C11      & 3 s & No occultation  & W. Schoenell\\
- UFSC       &27$\degr$36\arcmin12.3\arcsec~S & 28 cm & 6 s&  Start obs.: 21:49:27 & A. J. T. Mello\\
+ Florian\'opolis       & 48$\degr$31'20.5'' W &   C11      & 3 s & No occultation  & W. Schoenell\\
+ UFSC       &27$\degr$36'12.3'' S & 28 cm & 6 s&  Start obs.: 21:49:27 & A. J. T. Mello\\
             & 20 m       & f/10          &  SBIG ST7      &   End obs.: 22:51:21      & F. R. Herpich \\
@@
 
-The flux of the star in the five occultation chords was obtained from the FITS images with the \textsc{platform for reduction of astronomical images automatically (praia)} \citep{2011gfun.conf...85A}.
+The flux of the star in the five occultation chords was obtained from the FITS images with the Platform for Reduction of Astronomical Images Automatically (PRAIA) \citep{2011gfun.conf...85A}.
 The light curves were normalized to the flux of the star plus Ceres, as they were merged right before and after the occultation.
@@
 
-The ingress (disappearance) and egress (reappearance) instants of the occultation were obtained for each light curve by fitting a square-well model convoluted with the Fresnel diffraction, the CCD bandwidth, the stellar apparent diameter, and the applied finite exposure time; see \cite{Widemann2009} and BR13.
+The start and end times of the occultation were obtained for each light curve by fitting a square-well model convoluted with the Fresnel diffraction, the CCD bandwidth, the stellar apparent diameter, and the applied finite exposure time; see \cite{Widemann2009} and BR13.
 %\cite{BragaRibas2013}.
@@
 The smallest integration time used in the positive observations was 1.0~s, which translates to almost 3.9~km in the celestial plane.
-Therefore, the time error of ingress and egress are largely dominated by the integration times, not by Fresnel diffraction or star diameter, which are on the order a few hundreds meters for this event.
-The occultation data fit consists in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011} and BR13.\ The free parameter to adjust is the ingress or egress instant, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$.
+Therefore, the occultation light curves are largely dominated by the integration times, not by Fresnel diffraction or star diameter, which are on the order a few hundreds meters for this event.
+The occultation fits consist in minimizing a classical $\chi^{2}$ function for each light curve, as described in \cite{Sicardy2011} and BR13.\ The free parameter to adjust is the ingress (disappearance) or egress (reappearance) time, which provides the minimum value of $\chi^{2}$, denoted as $\chi^{2}_{min}$.
 The best fittings to the 2010 occultation light curves are shown in Fig.
@@
 \begin{figure}
-\includegraphics[scale=0.58]{figures/Ceres_2010_fluxratio.eps} 
+\includegraphics[scale=0.58]{figures/Ceres_2010_fluxratio.png} 
 \caption{The four occultation light curves normalized and vertically shifted by a factor of 0.2 for better visualisation.
@@
 The methodology used to analyse Ceres' profile from the observations is the same described by \cite{Sicardy2011} and BR13.%\cite{BragaRibas2013}. 
- ~Each combination of site position and recorded ingress/egress instant, together with star coordinates and Ceres' ephemeris, corresponds to a point in the plane of the sky.
-The collection of all these points ideally determines the apparent limb of Ceres.
+ ~To each combination of site position and recorded ingress/egress time, together with star coordinates and Ceres' ephemeris, corresponds a point on the plane of the sky.
+The collection of those points ideally determines the apparent limb of Ceres.
 %~To each combination of site position and recorded ingress/egress time, together with star coordinates and Ceres' ephemeris, corresponds a point $(f_{i,obs},g_{i,obs})$ on the plane of the sky.
@@
 We adopt an elliptic model for the limb profile, resulting from the projection of an oblate spheroid onto the sky plane.
-This choice is supported by the work of \cite{Drummond2014}, by means of direct imaging of Ceres.
+This model is supported by the work of \cite{Drummond2014}, by means of direct imaging of Ceres.
 Hence, we have $N=7$ chord extremities to adjust the $M=5$ parameters which define an ellipse: apparent semimajor and semiminor axis ($a^\prime$ and $b^\prime$, respectively), position angle $P$ of its semiminor axis and the position $(f_c,g_c)$ of its centre with respect to the occulted star.
 Of course, the apparent semimajor axis {$a^\prime$} is equivalent to the equatorial radius $R_{equa}$ of the ellipsoid.
-The coordinates $f_{c}$ and $g_{c}$, in kilometres, are calculated using the JPL\#33 Ceres' ephemeris \citep{Giorgini1996} and the occulted star's position.
+The coordinates $f_{c}$ and $g_{c}$, in kilometres, are calculated using the JPL\#26 Ceres' ephemeris and the occulted star's position.
 They are positive toward the local celestial east and north, respectively.
@@
 The apparent oblateness can be defined by $\epsilon^\prime = 1 - (b^\prime/a^\prime)$.
-The best-fitting solution is obtained minimizing a reduced $\chi^{2}$ function, where we define the number of degrees of freedom of the problem as $\mathcal{N} \equiv N - M$.
-All the procedures that allow the determination of the error bars of the physical parameters can be found on BR13.
+We define the number of degrees of freedom of the problem as $\mathcal{N} \equiv N - M$.
+All the procedure that allows the determination of the error bars of the physical parameters can be found on BR13.
 
@@
 As we shortly show, it led to a rather large uncertainty on the position angle.
-Furthermore, the nominal solution alone is not capable of returning the true oblateness, which can be evaluated through equation 2 of BR13  provided that the aspect angle $\zeta$ is known.% or constrained by some physical hypothesis.
+Furthermore, the nominal solution solely is not capable of returning the true oblateness, which can be evaluated through equation 2 of BR13.% \ref{Eq: TrueObla}, provided that the angle $\zeta$ is known.% or constrained by some physical hypothesis.
 
-From observations with adaptive optics spanning a 10-year period, \cite{Drummond2014} determined the position of Ceres' polar axis within an error of only 3$\degr$:
+From observations with adaptive optics spanning a 10-year period, \cite{Drummond2014} determined the position of Ceres' polar axis within a range of only 3$\degr$:
 \begin{equation}
@@
 Therefore Eq.~\ref{Eq: Pole} may act as a constraint for $P$, and a second solution can be obtained by probing the parameter space with the restriction that the position angle is confined to the range that follows from Eq.~\ref{Eq: Pole}.
-We call this the `pole-constrained solution'.
+We call this the ``pole-constrained solution''.
 
@@
 For the best-fitting solution we find $\chi^2_{r,min} = 0.24$, which could be interpreted as a slightly overestimation of the error bars with regard to the good quality of the fit.
-However, inasmuch as the problem has only two degrees of freedom, it is far from the statistical realm and relatively small $\chi^2_{r,min}$ are acceptable.
+However, inasmuch as the problem has only two degrees of freedom, it is far from the statistical realm and low $\chi^2_{r,min}$ may eventually happen.
 
@@
 
-Already mentioned, the parameter with the largest uncertainty is the position angle: spanning on a 20$\degr$ interval, its determination has a relative precision worse than 10 per cent.
+Already mentioned, the parameter with the largest uncertainty is the position angle: spanning on a 20$\degr$ interval, its determination has a relative precision worse than 10$\%$.
 Clearly, the coordinates of Ceres' pole (Eq.
@@
 
-At the occultation, the coordinates (Eq.
+At the moment of the occultation, the coordinates (Eq.
 \ref{Eq: Pole}) of Ceres' rotational pole correspond to the position angle $P = (12 \pm 3)\degr$.
-Exploration of the parameter space, restricted to ellipses with position angles laying within this range, results in the pole-constrained solution.
+Exploration of the parameter space, restricted to ellipses whose position angle lies on this range, results in the pole-constrained solution.
 The related physical parameters are displayed on the third column of Table \ref{Tab: results} in boldface, while the best-fitting solution is depicted in Fig.\ref{Fig:Ceres-2010-body}.
@@
 
-The event which took place on 2013 October 25 involved the star TYC 865-911-1 (UCAC4 496-058191), of magnitude $V = 10.05$.
+The event which took place on 25 October 2013 involved the star TYC 865-911-1 (UCAC4 496-058191), of magnitude $V = 10.05$.
 Based on UCAC4 \citep{Zacharias2013}, the ICRS position to the date of occultation corrected of proper motion is:
@@
   \begin{array}{l l}
-    \alpha = 11^{\rmn{h}}57^{\rmn{m}}52\fs7641\\
-    \delta = +09\degr 07\arcmin 49\farcs835
+    \alpha = 11^{h}57^{m}52^{s}.7641\\
+    \delta = +09\degr 07\arcmin 49\arcsec.835
   \end{array}
@@
 %\subfigure
-{\includegraphics[scale=0.42]{figures/Ceres_2013-zoom.eps}}
+{\includegraphics[scale=0.42]{figures/Ceres_2013-zoom.png}}
 \caption{Post-occultation reconstruction of Ceres' shadow path on Earth for the 2013 October 25 event at the east coast of USA. 
@@
 The blue dots are the sites that have observed the event.
-Upper view of the occultation over the sites that observed the event (blue dots). Red points are the centre of the shadow separated by 15 seconds.
+Right Panel: Upper view of the occultation over the sites that observed the event (blue dots). Red points are the centre of the shadow separated by 5 seconds.
 \label{Fig: Ceres-2013-map}}
@@
 
-Nine positive chords were obtained by the variety of instruments listed in Table \ref{Tab: obs-2013}.
-Each station was equipped with a video camera with negligible readout times.
-This was of particular importance, since in this event Ceres' shadow speed was $42.6~ km~s^{-1}$, much faster than in the 2010 event.
+Nine positive chords were obtained by the variety of equipment listed in Table \ref{Tab: obs-2013}.
+Each station was equipped with a video camera (whose readout time can be considered negligible).
+This is of particular importance in this event, since Ceres' shadow had the much faster speed of 42.6~km~s$^{-1}$, if compared to the 2010 event.
 All the observing sites are in the United States, as depicted in Fig.~\ref{Fig: Ceres-2013-map}.
@@
 The timing method consisted in the comparison of the camcorder internal clock to a 1PPS GPS signal, before and after the recording of the occultation.
-Absolute timing errors of this procedure are expected to be less than 0.1 s.
+Absolute timing errors of this procedure are expected to be inferior to 0.1 s.
 
-During the event Ceres was low in the horizon, with altitudes between $15\degr$ (Winchester) and $20\degr$ (Hampton).
-Strong scintillation is expected in such a scenario which, combined to short integration times and the low magnitude drop of the event, resulted in rather noisy light curves and thus larger uncertainties in the time of the contacts, as is shown in the next section.
+A remarkable observing circumstance of this event is the low altitude of Ceres with respect to the horizon.
+Precisely, its altitude at moment of the occultation ranged between $20\degr$ (Hampton) and $15\degr$ (Winchester).
+Strong scintillation is expected in such a scenario which, combined to short integration times and the low magnitude drop of the event, resulted in rather noisy light curves and thus larger uncertainty in the time of the contacts, as it is shown in the next section.
 
@@
 \hline
- Hampton & 70$\degr$48\arcmin59.7\arcsec~W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
-  &42$\degr$53\arcmin52.8\arcsec~N & f/5 & 0.033 s     & 09:40:46.9 $\pm$ 0.1 s &   \\
-            & 7 m       &  &     & 09:40:57.26  $\pm$ 0.08 s &   \\
+ Hampton & 70$\degr$48'59.7'' W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
+  &42$\degr$53'52.8'' N & f/5 & 0.033 s     & 09:40:46.9 $\pm$ 0.1 &   \\
+            & 7 m       &  &     & 09:40:57.26  $\pm$ 0.08 &   \\
  & & & & & \\
- Topsfield & 70$\degr$55\arcmin16.6\arcsec~W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
-  &42$\degr$37\arcmin55.9\arcsec~N & f/5      & 0.033 s   & 09:40:45.4 $\pm$ 0.1 s &   \\
-            & 45 m      &  &     & 09:40:58.0  $\pm$ 0.1 s &   \\
+ Topsfield & 70$\degr$55'16.6'' W & 12 cm & PC164C-EX2 & Positive & T. Blank \\
+  &42$\degr$37'55.9'' N & f/5      & 0.033 s   & 09:40:45.4 $\pm$ 0.1 &   \\
+            & 45 m      &  &     & 09:40:58.0  $\pm$ 0.1 &   \\
  & & & & & \\
- Brookline & 71$\degr$08\arcmin14.5\arcsec~W & 64 cm & Infinity2-1R & Positive & N. Weber \\
-  &42$\degr$18\arcmin27.4\arcsec~N & f/9.6 & 0.015 s     & 09:41:48.00 $\pm$ 0.01 s &  R. Dantowitz  \\
-            & 109 m     &       &     & 09:42:02.93 $\pm$ 0.01 s   \\
+ Brookline & 71$\degr$08'14.5'' W & 64 cm & Infinity2-1R & Positive & N. Weber \\
+  &42$\degr$18'27.4'' N & f/9.6 & 0.015 s     & 09:41:48.00 $\pm$ 0.01&  R. Dantowitz  \\
+            & 109 m     &       &     & 09:42:02.93 $\pm$ 0.01   \\
  & & & & & \\
- Winchester & 78$\degr$14\arcmin39.6\arcsec~W & 36 cm & PC164C & Positive & J. Brooks \\
-  &39$\degr$16\arcmin21.5\arcsec~N & f/5 & 0.033 s    & 09:40:33.26 $\pm$ 0.08 s &   \\
-            & 211 m     &  &     & 09:40:55.86  $\pm$ 0.09 s &   \\
+ Winchester & 78$\degr$14'39.6'' W & 36 cm & PC164C & Positive & J. Brooks \\
+  &39$\degr$16'21.5'' N & f/5 & 0.033 s    & 09:40:33.26 $\pm$ 0.08 &   \\
+            & 211 m     &  &     & 09:40:55.86  $\pm$ 0.09 &   \\
  & & & & & \\
- Greenbelt & 76$\degr$52\arcmin09.4\arcsec~W & 12 cm & PC164C-EX2 & Positive & J. Dunham \\
-  &38$\degr$59\arcmin12.1\arcsec~N & f/2.5 & 0.033 s    & 09:40:33.6 $\pm$ 0.1 s & D. Dunham  \\
-            & 52 m      &   &     & 09:40:56.4  $\pm$ 0.1 s &   \\
+ Greenbelt & 76$\degr$52'09.4'' W & 12 cm & PC164C-EX2 & Positive & J. Dunham \\
+  &38$\degr$59'12.1'' N & f/2.5 & 0.033 s    & 09:40:33.6 $\pm$ 0.1 & D. Dunham  \\
+            & 52 m      &   &     & 09:40:56.4  $\pm$ 0.1 &   \\
  & & & & & \\
- Alexandria & 77$\degr$02\arcmin28.3\arcsec~W & 7 cm & Watec120N & Positive & P. Maley \\
-  &38$\degr$49\arcmin19.1\arcsec~N & f/10 &  0.067 s   & 09:40:33.3 $\pm$ 0.1 s &   \\
-            & 8 m       &   &     & 09:40:56.1  $\pm$ 0.1 s &   \\
+ Alexandria & 77$\degr$02'28.3'' W & 7 cm & Watec120N & Positive & P. Maley \\
+  &38$\degr$49'19.1'' N & f/10 &  0.067 s   & 09:40:33.3 $\pm$ 0.1 &   \\
+            & 8 m       &   &     & 09:40:56.1  $\pm$ 0.1 &   \\
  & & & & & \\
- Owings & 76$\degr$38\arcmin06.3\arcsec~W & 25 cm & PC164C  & Positive & C. Ellington \\
-  &38$\degr$41\arcmin26.5\arcsec~N & f/3.3 &  0.033 s & 09:40:34.27 $\pm$ 0.05 s &   \\
-            & 38 m                &       &     & 09:40:56.0  $\pm$ 0.2 s &   \\
+ Owings & 76$\degr$38'06.3'' W & 25 cm & PC164C  & Positive & C. Ellington \\
+  &38$\degr$41'26.5'' N & f/3.3 &  0.033 s & 09:40:34.27 $\pm$ 0.05 &   \\
+            & 38 m                &       &     & 09:40:56.0  $\pm$ 0.2 &   \\
  & & & & & \\
- Mechanicsville & 77$\degr$23\arcmin06.7\arcsec~W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
-  &37$\degr$41\arcmin26.1\arcsec~N & f/2.5 & 0.033 s    & 09:40:33.0 $\pm$ 0.1 s &   \\
-            & 60 m      &   &     & 09:40:54.8  $\pm$ 0.1 s &   \\
+ Mechanicsville & 77$\degr$23'06.7'' W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
+  &37$\degr$41'26.1'' N & f/2.5 & 0.033 s    & 09:40:33.0 $\pm$ 0.1 &   \\
+            & 60 m      &   &     & 09:40:54.8  $\pm$ 0.1 &   \\
  & & & & & \\
- Varina & 77$\degr$19\arcmin49.3\arcsec~W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
-  &37$\degr$25\arcmin58.6\arcsec~N & f/2.5 &  0.033 s   & 09:40:32.4 $\pm$ 0.3 s &   \\
-            & 19 m      &   &     & 09:40:53.1  $\pm$ 0.2 s &   \\
+ Varina & 77$\degr$19'49.3'' W & 12 cm & PC164C-EX2 & Positive & D. Dunham \\
+  &37$\degr$25'58.6'' N & f/2.5 &  0.033 s   & 09:40:32.4 $\pm$ 0.3 &   \\
+            & 19 m      &   &     & 09:40:53.1  $\pm$ 0.2 &   \\
 
@@
 
-All videos were converted to FITS images and the photometry of the target was obtained via \textsc{praia} \citep{2011gfun.conf...85A}.
-The light curves were normalized to the flux of a reference star when available on the field.
+All videos were converted to FITS images and the photometry of the target was obtained via PRAIA \citep{2011gfun.conf...85A}.
+The light curves were normalized to the flux of a reference star when such was available on the field.
 
@@
 
-The ingress and egress instants of the occultation were obtained by the same procedure explained in Section \ref{Sec: light-curve-2010}.
-%The Fresnel scale of 0.4~km for this event was calculated using Ceres' geocentric distance $D = 3.28 \text{ au} = 4.91 \times 10^{8}$~km, for a typical wavelength of $\lambda = 0.65$~$\micron$.
+The start and end times of the occultation were obtained by the same procedure explained in Section~2.2, %The Fresnel scale of 0.4~km for this event was calculated using Ceres' geocentric distance $D = 3.28 \text{ au} = 4.91 \times 10^{8}$~km, for a typical wavelength of $\lambda = 0.65$~$\micron$.
 Considering the apparent magnitudes of the star, $B = 10.47$, $V=9.97$, and $K=8.76$ \citep{Hog2000}, we estimate its diameter projected at Ceres' distance as 0.16~km.
-Since the typical effective integration time used (0.17~s) translates to about 7 km in the celestial plane, and the Fresnel scale and star diameter are again of the order of a few hundreds of meters, the theoretical occultation light curves are largely dominated by the integration times, as for the 2010 event.
+Since the typical effective integration time used (0.17~s) translates to about 7 km in the celestial plane, and the Fresnel scale and star diameter are again on the order of few hundreds of meters, the theoretical occultation light curves are largely dominated by the integration times, as for the 2010 event.
 
@@
 \begin{figure}
-\includegraphics[scale=0.58]{figures/Ceres_2013_fluxratio.eps} 
+\includegraphics[scale=0.58]{figures/Ceres_2013_fluxratio} 
 \caption{The nine occultation light curves normalized and vertically shifted by a factor of 1.0 for better viewing, see Fig.~\ref{Fig: Ceres-2010-curves} for the explanation of the graph.% The solid black lines are the best fit of the square-well model to the data.
@@
 
-Elliptic limb profiles were adjusted to all the available\footnote{Brookline's chord was not used for limb fitting since it had an inaccurate absolute time.} chords by the same procedure described in Section \ref{Sec: limbfittingmethod}.%\footnote{For this event, the JPL\#33 Ceres' ephemeris was used to obtain the $f_{c}$ and $g_{c}$ coordinates.}
+Elliptic limb profiles were adjusted to all the available\footnote{Brookline's chord was not used for limb fitting since it had an inaccurate absolute time.} chords by the same procedure described in Section \ref{Sec: limbfittingmethod}\footnote{For this event, the JPL\#33 Ceres' ephemeris was used to obtain the $f_{c}$ and $g_{c}$ coordinates.}.
 This yielded $\chi^2_{r,min} = 13$, suggesting that an elliptic model is not satisfactory to the data.
@@
 \ref{Fig: Ceres-2013-body}, shows that the Varina chord seems to be somewhat advanced with respect to the others.
-Taking into account that in this station time stamps were not inserted on the video frames, it is fairly possible to attribute this advance to an eventual problem on the correspondence between camcorder's and GPS' times.
+Taking into account that in this station time stamps were not inserted on the video frames, it is always possible to attribute this advance to an eventual problem on the correspondence between camcorder's and GPS' times.
 This could be caused, for example, if the camcorder delayed to start the recording and, since this was an unattended pre-pointed station, this fact would not be noticed by the analysis of the video itself.
@@
 This chord, actually, has roughly the same length of Mechanicsville's, despite the fact that they are separated by about $100~km$.
-Differently from Varina, though, this station had time stamps inserted in each frame of the video, which makes it more unlikely to justify an eventual timing issue.
-Another possible explanation for the apparent problem of the times of this chord is the determination of the ingress and egress instants in the light curve analysis, which could have been affected by noise.
+Differently from Varina, though, this station had time stamps inserted in each frame of the video, which makes it more unlike to justify an eventual timing issue.
+Another possible explanation for the apparent problem of the times of this chord is the determination of the events in the light curve analysis, which could have been affected by noise.
 Finally, the delay during the ingress could be caused by a relief feature in Ceres; we shall soon return to this hypothesis.
@@
 
-For this event the polar aspect angle is $\zeta = 90.7\degr$, which makes the true oblateness equal to the apparent one, within the error bars.
+For this event the polar aspect angle is $\zeta = 90.7\degr$, which makes true oblateness be equal to the apparent one, within the error bars.
 
-The position angle of this nominal solution is better constrained than the 2010 occultation.
-Actually, the uncertainty in the former is of $5\degr$ in contrast to $10\degr$ of the latter.
+It is worth noting that the position angle of this nominal solution is better constrained than the respective of the 2010 occultation.
+Actually, the uncertainty of the former is of $5\degr$ in contrast to $10\degr$ of the latter.
 This suggests that the pole-constrained solution obtained via the position of Ceres' pole (Eq.
-\ref{Eq: Pole}) would not be significantly different from the nominal one.
+\ref{Eq: Pole}) would not be significantly different of the nominal one.
 
-This assumption was confirmed when we carried out the limb fitting with the constraint $P = (25 \pm 3)\degr$, the position angle at the occultation which follows from Eq.
-\ref{Eq: Pole}.
+This assumption is confirmed when we carry out the limb fitting with the constraint $P = (25 \pm 3)\degr$, the position angle at the moment of the occultation which follows from (Eq.
+\ref{Eq: Pole}).
 The physical parameters related to this pole-constrained solution are presented in the last column of Table~\ref{Tab: results}, and are essentially the same of the nominal solution.
 
-In the hypothesis that the delay observed in the immersion at Owings could be associated to a limb topography feature, the recorded contact would then correspond to a negative elevation of $31 \pm 4$ km with respect to the best-fitting ellipse.
-Theoretical models, however, predict that reliefs in Ceres should not be higher than about 10--20 km \citep{Johnson1973}, while published observational data sets the bound of 18 km \citep{Carry2008}.
+We close this section returning to the hypothesis that the delay observed in the immersion at Owings could be associated to a limb topography feature.
+The recorded contact would then correspond to a negative elevation of $31 \pm 4$ km with respect to the best-fitting ellipse.
+Theoretical models, however, predict that relief in Ceres should be no higher than about 10--20 km \citep{Johnson1973}, while published observational data sets the bound of 18 km \citep{Carry2008}.
 More recently, images by the probe \textit{Dawn} also reveal an even smoother surface.
-Therefore, the association of Owings first contact to a relief is improbable.
+The association of Owings first contact to a relief must, therefore, be ruled out.
 
@@
 \includegraphics[scale=0.36]{figures/Ceres_2013_sphere.pdf}
-\caption{The best elliptical fit for the occultation chords for the event of 2013 using timing from Table \ref{Tab: obs-2013} and the pole-constrained solution.
+\caption{The best elliptical fit for the occultation chords for the event of 2013 using timing from Table \ref{Tab: obs-2013} and the nominal solution.
 The arrow indicates the direction of motion, blue lines are the observed chords, the red and green segments are the ingress, egress and mid-occultation error bars at 1-$\sigma$ level.
@@
 
-The objects' geocentric positions derived from stellar occultations are most valuable for improving their orbits \citep{Desmars2015}.
-The precision of these astrometric positions are only limited by the accuracy of the occulted star position, not by the limb fit.
-The geocentric J2000 positions of Ceres at the time of each occultation are displayed in Eqs.~\ref{Eq: Ceres-2010_pos} and \ref{Eq: Ceres-2013_pos}.
-The errors of Ceres positions come from the errors of the star positions, taken from the respective catalogues for the occultation epochs, and the errors of the relative apparent distances between star and Ceres, derived from the limb fits.
-For the record, we also display the small errors of the relative apparent distances ($\Delta\alpha$, $\Delta\delta$).
+Derive geocentric position of an object from a stellar occultation was proved to be a very precious source for orbit fit proposes \citep{Desmars2015}.
+Usually the astrometric position is limited to the accuracy of the occulted star position, not by the limb fit.
+The geocentric J2000 positions of Ceres at the time of each occultation are displayed in Eqs.~\ref{Eq: Ceres-2010_pos} and \ref{Eq: Ceres-2013_pos}, the error on the star position were not taken into account.
 
@@
     Time = 22:40:00\\
-    \alpha = 17^{\rmn{h}}18^{\rmn{m}}29\fs0122\pm0\farcs027\\
-    \delta = -27\degr 26\arcmin 38\farcs617\pm0\farcs028\\
-    \Delta\alpha = 0\farcs003; ~~\Delta\delta = 0\farcs007
+    \alpha = 17^{\text{h}}18^{\text{m}}29^{\text{s}}.0122\pm0\arcsec.003\\
+    \delta = -27\degr 26\arcmin 38\arcsec.617\pm0\arcsec.007
   \end{array}
@@
     Time = 09:45:00\\
-    \alpha = 11^{\rmn{h}}57^{\rmn{m}}52\fs9154\pm0\farcs019\\
-    \delta = +09\degr 07\arcmin 49\farcs865\pm0\farcs021\\
-    \Delta\alpha = 0\farcs002; ~~\Delta\delta = 0\farcs007
+    \alpha = 11^{h}57^{m}52^{s}.9154\pm0\arcsec.002\\
+    \delta = +09\degr 07\arcmin 49\arcsec.865\pm0\arcsec.007
   \end{array}
@@
 
-A quick glance at Table~\ref{Tab: results} shows an overall agreement between the physical parameters derived from both occultations, specially in the equatorial diameter.
+A quick glance at Table~\ref{Tab: results} shows there is an overall agreement between the physical parameters derived from both occultations, specially in the equatorial diameter.
 The differences of the solutions occur basically on the size of their error bars, and can be justified by the particularities of each set of data, as discussed below.
@@
   \begin{centering}
-  \begin{tabular}{@{}lccc}
+  \begin{tabular}{@{}llcc}
   \hline
@@
 $959 \pm 5$  & $0.074 \pm 0.007$ & Keck        & 3\\
-$975 \pm 4$  & $0.067 \pm 0.005$ & HST         & 4\\
-$960 \pm 5$  & $0.05  \pm 0.01$  & Occultation & 5\\
+$975 \pm 4$  & $0.068 \pm 0.005$ & HST         & 4\\
+$959 \pm 7$  & $0.05  \pm 0.01$  & Occultation & 5\\
 \hline
@@
 The 2010 event, for example, had only seven contacts; none the less, they were well distributed over Ceres' disc (see Fig.
-\ref{Fig:Ceres-2010-body}) acting as a constraint to its shape.
+\ref{Fig:Ceres-2010-body}) acting as a constraint to is shape.
 On the other hand, the 2013 event had five more exploitable contacts, but they were concentrated in certain regions of the body.
-In particular, the lack of chords in Ceres' southern hemisphere made its oblateness worse determined here than in the 2010 event.
+In particular, the absence of chords closer to Ceres' southern limb made its oblateness be worse determined here than in the 2010 event.
 
@@
 
-The few chords of the 2010 occultation could themselves only constraint the position angle of the object to a uncertainty of $10\degr$.
-This uncertainty was reduced by a factor of two in the 2013 event, approaching --and verifying-- the result predicted by the work of \cite{Drummond2014}.
-As was shown, using the coordinates of Ceres' polar axis to limit the position angle was not an efficient procedure in the 2013 event, in the sense that it did not result in significant changes in the parameters obtained in the nominal solution (see Table \ref{Tab: results}).
+The few chords of the 2010 occultation could themselves only constraint the position angle of the object to a $20\degr$ interval.
+This range was reduced by a factor of two in the 2013 event, approaching --and verifying-- the result predicted by the work of \cite{Drummond2014}.
+As it was shown, using the coordinates of Ceres' polar axis to limit the position angle, was not an efficient procedure in the case of this occultation, in the sense that the aforementioned limitation had not correspond to a stronger constraint on the other parameters.
 
@@
 
-As mentioned in the Introduction, the 1984 event \citep{Millis1987} is the only occultation data to which we can compare our result.
-The measured diameters do not agree, ours being larger by 2-sigma.
-It is difficult to state for sure the reasons for this divergence since we had no access to the original 1984 data or reduction methodology.
-We note, however, that in this occultation a variation of 0.5 s in the contact times correspond to almost $7~km$ on Ceres' limb, which is on the same order of the residuals of their elliptical fit.
+As mentioned in the Introduction, the 1984 event \citep{Millis1987} is the only occultation data to which we can compare our result; and it is interesting to notice that the diameters measured in each case do not agree, being our result larger than the other.
+It is difficult to state with sure the reasons for this divergence since we had no access to the original 1987 data or reduction methodology.
+We note, however, that in this occultation a variation of 0.5 s in the contact times correspond to almost 7~km on Ceres' limb, which is on the same order of the residuals of their elliptical fit.
 Therefore, in order to properly compare their result to ours it would be necessary to reduce their light curves with the same methodology used in the present work.
 
-NASA's \textit{Dawn} mission shall bring light to these questions, which will be important not only for the knowledge of Ceres itself, but also for all the techniques used so far to study the physical properties of small Solar System objects, such as the stellar occultations.
+As mentioned, NASA's \textit{Dawn} mission shall bring some light to these questions, what will be important not only for the knowledge of Ceres it self, but also for all the techniques used so far to study the physical properties of small Solar System objects, such as the stellar occultations.
 
@@
 acknowledges PAPDRJ-FAPERJ/CAPES E-43/2013 number 144997, E-26/101.375/2014 and CDFB-CAPES/Brazil.
-M.A.
-acknowledges CNPq grants 473002/2013-2, 482080/2009-4 and 312394/2014-4, and FAPERJ grant 111.488/2013.
-RVM thanks grants: CNPq-306885/2013, Capes/Cofecub-2506/2015, Faperj/PAPDRJ-45/2013.
-J.I.B.
-Camargo acknowledges CNPq for a PQ2 fellowship (process number 308489/2013-6).
 We also acknowledge Steve Preston for the predictions of the occultations.
@@
 
-\bibitem[\protect\citeauthoryear{Carry et al.}{2008}] {Carry2008} Carry B. et al., 2008,
-A\&A, 478, 235
-
 \bibitem[\protect\citeauthoryear{Castillo-Rogez}{2011}]{CastilloR2011} Castillo-Rogez, J.~C. 
@@
 
-\bibitem[\protect\citeauthoryear{Dunham et al.}{2014}]{Dunham2014}
-Dunham, D.W. et al., 2014, Asteroid Occultations V12.0. EAR-A-3-RDR-OCCULTATIONS-V12.0. NASA Planetary Data System
+\bibitem[\protect\citeauthoryear{Carry et al.}{2008}] {Carry2008} Carry B. et al., 2008,
+A\&A, 478, 235
 
@@
 
-\bibitem[\protect\citeauthoryear{Giorgini et al.}{1996}]{Giorgini1996} Giorgini, J. D. et al. 1996, BAAS, 28, 1158
-
 %\bibitem[\protect\citeauthoryear{Hog et al.}{2000}] {Hog2000} Hog E. et al., 2000,
-%A\&A, 355, L27
+A\&A, 355, L27
 
@@
 %\bibitem[\protect\citeauthoryear{van Belle}{1999}] {vanBelle1999} van Belle G.T., 1999,
-%PASP, 111, 1515
+PASP, 111, 1515
 
@@
 %\bibitem[\protect\citeauthoryear{Zacharias et al.}{2004}]{Zacharias2004} Zacharias N. et al., 2004,
-%AJ, 127, 3043
+AJ, 127, 3043
 

### 2017-08-16 Ceres/paper-Ceres-revised/paper-Ceres.tex
mod 2017 ago16

--- 2015-05-16 Ceres/paper-Ceres-source/paper-Ceres.tex
@@
 Indeed, it was proposed that Ceres' origin could have been as a transneptunian object \citep{McKinnon12}, later scattered to the Main Belt due to the giant planets' migration predicted by the `Nice Model' \citep{Gomes05}.
-Even if it was formed close to its current location, the dynamical history of the Solar System must have let its signatures on Ceres.
+Even if it was formed close to its current location, the dynamical history of the Solar System must have left its signatures on Ceres.
 These could include not only the late heavy bombardment features that might exist on its surface, but also the makeup of its volatiles, which could have been transported from the outer region.
@@
 Since the 1970's it has been speculated that Ceres could contain water ice, which was recently verified \citep{Kuppers2014}.
-Although the water regime on this object is still unknown, some internal structure models suggest the existence of a water ice -- or even a liquid water -- layer \citep{CastilloR2011}.
+Although the water regime on this object is still unknown, some internal structure models suggest the existence of a water vapour -- or even a liquid water -- layer \citep{CastilloR2011}.
 Yet the very question of whether Ceres underwent differentiation is open and, on the assumption of an affirmative answer, it is natural to ask if it ever had tectonics, what its geological evolution was, and if it is still active.
@@
 
+\textbf{It is possible to derive the pole position $(\zeta, P)$ from the coordinates of Ceres' pole $(\alpha_{p}, \delta_{p})$ and its ephemeris $(\alpha, \delta)$ via}
+%
+\begin{equation}
+\left\{
+\begin{array}{l l}
+\cos(\zeta)= -\sin(\delta_{p})\sin(\delta) - \cos(\delta_{p})\cos(\delta)\cos(\alpha_{p}-\alpha)\\
+\tan(P) = \dfrac{\sin(\alpha_{p}-\alpha)} {\tan(\delta_{p})\cos(\delta) - \sin(\delta)\cos(\alpha_{p}-\alpha)},\\
+%\cos(B).\cos(P)= \sin(\delta_{p}).\cos(\delta) - \cos(\delta_{p}).\sin(\delta).\cos(\alpha_{p}-\alpha)\\
+%\cos(B).\sin(P)=\cos(\delta_{p}).\sin(\alpha_{p}-\alpha)\\
+%\zeta = 90\degr - B
+\end{array}
+\right.
+\label{Eq: pole-position}
+\end{equation}
+%
+\textbf{where $\zeta = 0\degr$ and $\zeta=90\degr$ respectively correspond to a pole-on and an equator-on geometry.}
+
 From observations with adaptive optics spanning a 10-year period, \cite{Drummond2014} determined the position of Ceres' polar axis within an error of only 3$\degr$:
@@
 %\begin{array}{l}
-\alpha = (287 \pm 3) \degr, \quad %\\
-\delta = (+64 \pm 3) \degr,
+\alpha_{p} = (287 \pm 3) \degr, \quad %\\
+\delta_{p} = (+64 \pm 3) \degr,
 %\end{array} \right .
@@
 The knowledge of Ceres' pole not only allows the determination of its polar aspect, it suffices to set its position angle.
-Therefore Eq.~\ref{Eq: Pole} may act as a constraint for $P$, and a second solution can be obtained by probing the parameter space with the restriction that the position angle is confined to the range that follows from Eq.~\ref{Eq: Pole}.
+Therefore \textbf{Eqs.
+\ref{Eq: pole-position} and \ref{Eq: Pole}} may act as a constraint for $P$, and a second solution can be obtained by probing the parameter space with the restriction that the position angle is confined to the range that follows from Eq.
+\ref{Eq: Pole}.
 We call this the `pole-constrained solution'.
@@
 On the other hand, the 2013 event had five more exploitable contacts, but they were concentrated in certain regions of the body.
-In particular, the absence of chords close to Ceres' south pole made its oblateness worse determined here than in the 2010 event.
+In particular, the absence of chords close to Ceres' south pole made its oblateness less well determined here than in the 2010 event.
 
@@
 
-The few chords of the 2010 occultation could themselves only constraint the position angle of the object to a uncertainty of $10\degr$.
+The few chords of the 2010 occultation could themselves only \textbf{constrain} the position angle of the object to a uncertainty of $10\degr$.
 This uncertainty was reduced by a factor of two in the 2013 event, approaching --and verifying-- the result predicted by the work of \cite{Drummond2014}.
@@
 
-On the other hand, constrain the position angle on the 2010 occultation was proved to reduce the error bars of the other parameters (disregarding oblateness).
+On the other hand, \textbf{constraining} the position angle on the 2010 occultation was proved to reduce the error bars of the other parameters (disregarding oblateness).
 Moreover, this procedure resulted in an excellent agreement between the equatorial radius figures of both events.
@@
 The measured diameters do not agree, ours being larger by 2-sigma.
-It is difficult to state for sure the reasons for this divergence since we had no access to the original 1984 data or reduction methodology.
-We note, however, that in this occultation a variation of 0.5 s in the contact times correspond to almost 7 km on Ceres' limb, which is on the same order of the residuals of their elliptical fit.
-Therefore, in order to properly compare their result to ours it would be necessary to reduce their light curves with the same methodology used in the present work.
+\textbf{It is difficult to state for sure the reasons for this divergence.
+One way to clarify the issue is to redetermine the immersion and emersion instants from the original light curves using the same methodology presented in our work.
+Note that in this occultation a variation of 0.5 s in the contact times correspond to almost 7 km on Ceres' limb, which is on the same order of the residuals of their elliptical fit.
+However, we could not do that, as we had no access to the original 1984 light curve data.
+Moreover, we have no description of how the immersion and emersion instants of the chords were actually determined from these light curves in \cite{Millis1987}.}
 
@@
 $972 \pm 6$  & $0.08  \pm 0.03$  & Occultation & 1\\
-$967 \pm 10$ & $0.078 \pm 0.015$ & Keck+VTL    & 2 \\
+$967 \pm 10$ & $0.078 \pm 0.015$ & Keck+VTL    & 2\\
 $959 \pm 5$  & $0.074 \pm 0.007$ & Keck        & 3\\
@@
 Camargo acknowledges CNPq for a PQ2 fellowship (process number 308489/2013-6).
-We also acknowledge Steve Preston for the predictions of the occultations.
+We also acknowledge Steve Preston for the predictions of the occultations \textbf{and the referee Lawrence H.
+Wasserman (Lowell Observatory) for his contributions to improve the text.}
 

### 2017-08-16 Irregular-occ/mnras-source_arxiv/Irregular-occ.tex
mod 2017 ago16

\documentclass[a4paper]{article}
\pdfoutput=1
\usepackage{hyperref}
\hypersetup{
  pdfinfo={
    Title={New orbits of irregular satellites designed for the predictions of stellar occultations up to 2020, based on thousands of new observations},
    Author={A. R. Gomes-Júnior, M. Assafin, L. Beauvalet, J. Desmars, R. Vieira-Martins, J. I. B. Camargo, B. E. Morgado, F. Braga-Ribas},
    Keywords={occultations - ephemerides - planets and satellites: general - planets and satellites: individual: Jovian and Saturnian irregular satellites}
  }
}
\usepackage{pdfpages}

\begin{document}
\includepdf[pages=1-last]{Irregular.pdf}
\end{document}

### 2017-08-16 projeto/Projeto.tex
mod 2017 ago16

\documentclass[12pt,a4paper]{report}
\usepackage[brazilian]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amsthm,amsfonts,amssymb,textcomp}
%\usepackage{latexsym}
\usepackage{graphicx}
\graphicspath{{figuras}}
\usepackage{subfigure}
\usepackage{float}
\usepackage{longtable}
\usepackage{color}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage[breaklinks=true]{hyperref}
\usepackage[comma,authoryear]{natbib}
%\usepackage[nonumberlist]{glossaries}
\usepackage{arydshln}
\usepackage{footnote}
\usepackage{longtable}
\usepackage[small,bf,singlelinecheck=off]{caption}
\usepackage[left=3cm,right=2cm,top=3cm,bottom=2cm]{geometry}
%\usepackage[alf]{abntcite}

%%% newcommand %%%%%%%%%%%%%%%%%
\newcommand{\PE}{Perkin-Elmer }
\newcommand{\BC}{Boller \& Chivens }

\newcommand{\degr}{\ensuremath{^{\circ}}}%                    % degree symbol:  °
\newcommand{\arcmin}{\ensuremath{^{\prime}}}%                    % degree symbol:  °
\newcommand{\arcsec}{\ensuremath{^{\prime\prime}}}%                    % degree symbol:  °
\newcommand{\fs}{\mbox{\ensuremath{.\!\!^s}}}
\newcommand{\farcm}{\mbox{\ensuremath{.\mkern-4mu^\prime}}}%    % fractional arcminute symbol: 0.'0
\newcommand{\farcs}{\mbox{\ensuremath{.\!\!^{\prime\prime}}}}%  % fractional arcsecond symbol: 0.''0
\newcommand{\fdg}{\mbox{\ensuremath{.\!\!^\circ}}}%             % fractional degree symbol:     0.°0

%\makeglossaries

\makeatletter
\renewcommand\chapter{\thispagestyle{plain}
                \global\@topnum\z@
                \@afterindentfalse
                \secdef\@chapter\@schapter}
\makeatother

\makeatletter
\renewcommand{\@makechapterhead}[1]{%
\vspace*{50 pt}%
{\setlength{\parindent}{0pt} \raggedright \normalfont
\bfseries\Huge\thechapter.\ #1
\par\nobreak\vspace{40 pt}}}
\makeatother

\renewcommand{\thesection}{\arabic{section}}

%\input{glossario}
\author{Altair Ramos Gomes Júnior}
\title{Ocultações Estelares por Satélites Irregulares}
\begin{document}

\maketitle

\pagestyle{headings}

\section{Introdução} \label{Chap:Intro}

\indent \indent O estudo de objetos como objetos trans-Neptunianos (TNOs), Centauros e Satélites Irregulares nos ajudam a compreender a formação e evolução do Sistema Solar Exterior. Nesta região distante do Sol, de baixas temperaturas, objetos principalmente de tamanho relativamente pequeno, e mais dispersos no espaço, provavelmente sofreram muito pouca diferenciação, seja por mecanismos internos, seja por choques com outros corpos, comparados a objetos formados mais próximos do Sol.

Por serem corpos asteroidais localizados de forma dispersa além da órbita de Netuno, considera-se que os TNOs guardam estruturas e composições relativamente inalteradas em relação a sua época de formação, constituindo-se assim em corpos de prova de grande valor para o estudo da origem do Sistema Solar, ao menos para essa região exterior.

Tirando Plutão, o primeiro TNO foi descoberto há pouco mais de 20 anos \citep{Jewitt1993}. Por isso, as propriedades básicas desta população ainda não estão inteiramente estabelecidas, como a distribuição de tamanhos, composição, estruturas internas.

Atualmente, é aceito que TNOs tenham sido formados nas partes mais internas do Sistema Solar. Eles teriam então sido colocados em suas posições atuais devido a troca de momento angular entre os planetas e planetésimos quando da migração dos planetas gigantes. A evolução se deu de tal forma que a passagem dos planetesimais e planetas por zonas de ressonância de movimento médio redefiniu as órbitas desses corpos \citep{Tsiganis2005}.

Muitos dos objetos que pertenciam ao cinturão de Kuiper primordial acabaram sendo enviados pela interação com os planetas gigantes para as partes mais internas do Sistema Solar. Alguns podem ter sido capturados pelos planetas criando a população de satélites irregulares ou satélites externos \citep{Nesvorny2007}, troianos \citep{Morbidelli2005} ou até mesmo para o cinturão principal de asteroides como proposto para Ceres por \cite{McKinnon2012}. Estudar esses objetos é de grande importância para entender a evolução do Sistema Solar.

Os satélites irregulares dos planetas gigantes são menores que os regulares, possuindo órbitas mais excêntricas, inclinadas e distantes. Na maioria dos casos, essas órbitas são retrógradas. Devido a suas configurações orbitais, é amplamente aceito que estes objetos foram capturados nos estágios iniciais da formação do Sistema Solar \citep{Sheppard2003}. Figura \ref{Fig:sat-conf} resume as características orbitais dos satélites irregulares de Júpiter e Saturno.

\begin{figure}[h]
\begin{centering}
\subfigure[Satélites de Júpiter]{\includegraphics[scale=0.13]{figuras/Jupiter2.jpg}\label{Fig:sat-jup} }
\subfigure[Satélites de Saturno]{\includegraphics[scale=0.13]{figuras/Saturno2.jpg}\label{Fig:sat-sat} }
\caption{Esquema ilustrativo das órbitas dos satélites irregulares: a escala vertical se refere à inclinação em graus da órbita dos satélites em relação ao planeta; a escala horizontal é a distância em milhões de Km ao planeta; as linhas representam a variação de distância do satélite ao planeta durante o período orbital; os círculos são os tamanhos relativos dos satélites.(baseado em dados obtidos do JPL, via site Horizons).
\label{Fig:sat-conf}}
\end{centering}
\end{figure}

Por serem pequenos, eles são pouco brilhantes e só foram descobertos no último século\footnote{Website: \url{http://ssd.jpl.nasa.gov/?sat\_discovery}}. Até o momento, apenas os planetas Júpiter, Saturno, Urano e Netuno possuem satélites irregulares conhecidos, embora a possibilidade de que outros corpos, como Plutão, possuam tais satélites não são nulas \citep{Michaely2017}. Devido a isso, as menções a "satélites irregulares" presentes neste texto referem-se aos satélites irregulares dos planetas gigantes.

%Existem alguns mecanismos de captura de objetos por planetas gigantes propostos na literatura. Há o Arrasto Gasoso na nebulosa circumplanetária primordial \citep{Sheppard2005} onde o movimento do objeto seria afetado pelo gás e sua velocidade diminuiria até que fosse capturado pelo planeta. Outro mecanismo se chama captura pull-down \citep{Sheppard2005}, onde a massa do planeta aumentaria enquanto o objeto estivesse temporariamente capturado.

%Um mecanismo baseado no modelo de Nice \citep{Morbidelli2005, Tsiganis2005, Gomes2005} foi proposto por \cite{Nesvorny2007} e, especificamente para Júpiter, por \cite{Nesvorny2014}. Durante a instabilidade do Sistema Solar primordial, ocorreram encontros entre os planetas externos. Nesses encontros planetários poderia haver troca de energia e momento angular entre os planetas e os objetos próximos tornando possível a captura de um corpo irregular por um planeta gigante. Nesse cenário, a taxa de sobrevivência de satélites anteriores ao LHB (Bombardeamento Pesado Tardio\footnote{"Late Heavy Bombardment", em inglês}) é muito pequena.

%Um terceiro mecanismo importante é a captura através de interações colisionais \citep{Sheppard2005}. Uma colisão entre dois corpos pequenos dentro da esfera de Hill do planeta poderia gerar objetos fragmentados e a energia dissipativa poderia ser tal que alguns desses objetos seriam capturados.

%Alguns desses objetos formam grupos dinâmicos com elementos orbitais semelhantes, chamados famílias, similar às famílias encontradas no Cinturão Principal de Asteroides. Essas famílias podem ter sido criadas a partir de um corpo pai que se fragmentou por colisões com cometas ou outros asteroides \citep{Nesvorny2004}. Colisões com cometas tem uma probabilidade maior de ocorrer durante o Bombardeamento Pesado Tardio (LHB) \citep{Gomes2005}.

%\cite{Nesvorny2003} estudou as taxas de colisões entre satélites irregulares e concluiu que alguns satélites podem ter sido removidos por colisão com um satélite maior. A taxa de colisão entre os satélites do grupo de Himalia (Himalia, Elara, Lysithea and Leda, principalmente), por exemplo, foi encontrado como sendo maior que 1 durante a idade do Sistema Solar sugerindo que sua estrutura atual foi originada por colisão satélite-satélite.

%Para Phoebe, materiais ejetados de sua superfície causadas por impactos poderiam evoluir devido ao arrasto de Poynting-Robertson e colidir com Iapetus causando a grande variação de albedo observada no satélite \citep{Nesvorny2003}. De fato, a sonda \textit{Cassini} detectou em Phoebe uma característica de absorção em 2.42 $\mu m$, provavelmente combinações de CN, que foi também detectada no lado escuro de Iapetus \citep{Clark2005}.

Se esses objetos foram capturados, permanece a pergunta de onde eles vieram. \cite{Clark2005} mostraram a partir de espectroscopia da Cassini que Phoebe tem uma superfície provavelmente coberta por material do Sistema Solar externo e \cite{Grav2003} mostraram que os satélites de grupo prógrado de Júpiter Himalia tem cores cinzas implicando que suas superfícies são similares a asteróides tipo C. Nesse mesmo trabalho, o grupo retrógrado de Júpiter Carme foi encontrado como tendo cores superficiais semelhantes à asteroides tipo D como a família de Hilda ou famílias de troianos, enquanto J13 Kalyke tem uma cor mais avermelhada como centauros ou objetos trans-Netunianos (TNOs).

Para os satélites de saturno, \cite{Grav2007} mostraram através de suas cores e inclinações espectrais que esses satélites contêm uma fração mais ou menos igual de objetos semelhantes a asteroides tipo C, P e D, mas S22 Ijiraq é um pouco mais avermelhado que objetos tipo D. Esses trabalhos sugerem origens diferentes para os satélites irregulares.

Dentre os satélite irregulares, poucos são aqueles que possuem medidas de seus parâmetros físicos. Apenas Himalia, Phoebe e Nereida foram observados por sondas, apesar de não serem medidas ideais, já que foram alvos de oportunidade. A sonda \textit{Cassini} observou Himalia em 2000 ao passar próximo a Júpiter e obteve o tamanho de Himalia com um erro médio de 10 km \citep{Porco2003}.

Em 2004, a \textit{Cassini}, aproximando-se de Saturno, observou Phoebe em alta resolução obtendo um erro médio na medida de seu tamanho de 0.7 km \citep{Thomas2010}. Por fim, Nereida foi observado em 1989 pela sonda \textit{Voyager II} e seu tamanho foi obtido com um erro de 25 km \citep{Smith1989}. Para outros satélites irregulares de Júpiter, seus tamanhos foram estimados por \cite{Rettig2001} impondo um albedo de 4\% com a justificativa de que esse valor é um albedo nominalmente utilizado para objetos do Sistema Solar Externo.

\cite{Grav2015}, utilizando observações em infra-vermelho obtidos dentro do projeto WISE, da NASA, inferiu os tamanhos de 11 satélites irregulares de Júpiter e 3 de Saturno. Os diâmetros médios obtidos têm erros da ordem de 1 km para os satélites de Júpiter e 5 km para os de Saturno. Porém, como se espera para objetos desses tamanhos, eles não devem possuir formatos esféricos e serem craterizados.

Um método que tem se mostrado eficiente para a obtenção de características físicas de objetos distantes a partir de observações de solo é o método de ocultações estelares. Essa técnica proporciona medidas tão precisas que são apenas superadas por medidas oriundas de sondas, in loco. Porém, devido às escalas envolvidas nessa técnica, um grande trabalho de predição e atualização desses eventos deve ser feito para gerar resultados satisfatórios.

Como previsto por \cite{GomesJunior2016}, em 2018 Saturno irá atravessar o plano da Galáxia, aumentando em mais de 20 vezes a probabilidade de uma ocultação estelar. O mesmo para Júpiter, em 2019-2020. Com a publicação da primeira versão do catálogo Gaia em Setembro de 2016 e a segunda versão prevista para Abril de 2018, a posição das estrelas no céu não será mais um problema. Restará, contudo, a melhoria no conhecimentos das órbitas dos satélites irregulares.

O objetivo do pós-doc proposto é, portanto, composta de três etapas: a observação contínua dos satélites irregulares e obtenção de posições de alta precisão através de astrometria de campo (Seção \ref{Sec:astrometria}); a integração numérica das órbitas desses satélites utilizando posições antigas e novas, bem como pesando essas posições de forma adequada (Seção \ref{Sec:Integ}); e, por fim, gerenciar campanhas de observações de ocultações por satélites irregulares, nos mesmos moldes como tem sido feito para ocultações por TNOs e que deram ótimos resultados, e, fortuitamente, observar tais ocultações e publicar seus resultados (Seção \ref{Sec:Occ}).

\section{Astrometria} \label{Sec:astrometria}

\indent \indent O primeiro satélite irregular a ser descoberto foi Phoebe, satélite de Saturno, em 1898. Porém, até 1997, apenas uma dezena de satélites haviam sido descobertos, sendo 8 de Júpiter, 1 de Saturno e 1 de Netuno \citep{Sheppard2005}. O número de satélites irregulares descobertos, porém, aumentou consideravelmente com o início da era dos grandes surveys, nos anos 2000 \citep{Sheppard2003}. Atualmente, se conhece quase 100 satélites irregulares, sendo metade deles orbitando Júpiter.

Devido a serem objetos tão recentes e pouco brilhantes, poucas observações foram feitas até o momento. Por exemplo, os modelos orbitais mais recentes para os satélites irregulares de Júpiter, publicados por \cite{Brozovic2017}, fazem uso de menos cem observações distribuídas em dez anos para a maioria dos satélites. Apenas 6 possuem mais que mil observações.

O Grupo de Astrometria do Rio de Janeiro tem feito contínuas observações de diversos objetos do Sistema Solar no Observatório do Pico dos Dias (OPD). Dentre eles, os satélites irregulares tem sido observados regularmente desde 1995, e algumas poucas observações de Nereida, satélite de Netuno, desde 1992.

Nesse contexto, durante o Mestrado e Doutorado, foi realizado junto ao grupo um trabalho de caráter astrométrico para a melhoria das efemérides dos satélites irregulares. Reduzi o banco de dados com mais de 4 mil imagens obtidas no OPD entre 1992 e 2014. Em colaboração com o Dr. Jean-Eudes Arlot do \textit{Institut de Mécanique Céleste et de Calcul des Éphémérides} (IMCCE, em Paris), reduzi um banco de dados observado no \textit{Observatoire Haute-Provence} (OHP, França) entre 1998 e 2008, contendo mais de mil posições. Além disso reduzi 810 observações feitas no \textit{European Southern Observatory} (ESO, Chile) em 24 noites utilizando o detector mosaico CCD \textit{Wide Field Imager} (WFI).

Mais de 8000 posições foram identificadas como pertencentes a satélites irregulares. Porém, devido à grande gama de configurações (3 sítios, 5 telescópios, mais de 10 câmeras e mais de 10 filtros) e às condições observacionais de algumas noites, 6523 posições foram selecionadas como boas posições, ou seja, possuem cinco ou mais estrelas do catálogo de referência \citep[UCAC4, ][]{Zacharias2013}, estão dentro de 2$\sigma$ da dispersão das posições da noite e a dispersão da noite à qual pertence não é maior que 2$\sigma$ da média das dispersões do conjunto total de noites. Essa estatística é feita satélite por satélite. O trabalho foi publicado em \cite{GomesJunior2015-Irregular} e o catálogo de posições está disponível no CDS.

Um dos principais resultados desse trabalho foi a grande quantidade de posições obtidas para os satélites irregulares. É importante notar que \cite{Brozovic2017} já utiliza nossas posições em seu trabalho. Em comparação com a quantidade de posições utilizadas anteriormente nas integrações numéricas, por exemplo em \cite{Jacobson2012}, a nossa publicação aumentou em mais 50\% o número de observações para satélites como Himalia, Elara, Lysithea e Leda, satélites de Júpiter e em mais de 100\% para Nereida, satélite de Netuno.

Desde a publicação dessas posições, novos catálogos com estrelas de referência foram publicados. O principal deles é o catálogo Gaia, cuja primeira versão foi publicada em Setembro de 2016. Em comparação com o catálogo UCAC4, o Gaia aumenta a precisão das posições das estrelas de 30 mas (milisegundo de arco) para menos de 1 mas, e limitando a quantidade de estrelas em magnitude G=21 (UCAC4 só possui estrelas até magnitude R=16).

Porém o Gaia-DR1 não possui movimentos próprios para as estrelas, o que aumenta os erros provenientes de reduações que o utilizam para datas distantes da época média do catálogo. Felizmente, o GAIA-DR2, cuja publicação está prevista para Abril de 2018 terá movimentos próprios em alta qualidade.

A re-redução das posições dos satélites irregulares publicadas em \cite{GomesJunior2015-Irregular} utilizando o Gaia-DR2 como representação do ICRS é essencial para a melhoria da integração numérica de suas órbitas e, consequentemente, uma melhor predição de ocultações estelares. Além disso, desde a publicação, novas imagens foram (quase mil observações no OPD) e continuam sendo obtidas desses satélites. %A minha experiência com observação (OPD e OHP, durante meu doutorado sanduíche na França) e redução será importante nesse banco de dados que continua a crescer.

\section{Integração Numérica das Órbitas} \label{Sec:Integ}

\indent \indent Com o objetivo de prever ocultações estelares, é importante que se tenha uma ferramenta de integração numérica das órbitas dos objetos alvos que possa atualizá-las cotidiana, com novas observações. Porém, não são muitos os grupos que trabalham com integração numérica voltada para utilização de observações e cáclculo de efemérides.

Os principais grupos envolvidos na geração de efemérides são: o JPL\footnote{https://ssd.jpl.nasa.gov}, vinculado à NASA; e o IMCCE, do Observatório de Paris. Este último, em conjunto com pesquisadores da Universidade de Moscou, sustentam o serviço de efemérides para satélites naturais, o MULTI-SAT\footnote{http://lnfm1.sai.msu.ru/neb/nss/nssephme.htm}.

No Brasil, embora existam pesquisadores na área de dinâmica celeste, não existem pesquisadores trabalhando na geração de efemérides. Para as ocultações por TNOs, uma ferramenta foi criada nesse sentido, o NIMA \citep{Desmars2015}. Porém, ela é mantida por um pesquisador francês que não está mais no Brasil.

No caso dos satélites irregulares, em colaboração com uma pesquisadora francesa chamada Lauréne Beauvalet, nós realizamos a integração numérica para os 8 satélites irregulares principais de Júpiter. As posições utilizadas foram apenas aquelas publicadas em \cite{GomesJunior2015-Irregular}. Uma vez que não pretendíamos obter uma órbita para mais do 5 anos a frente, o trabalho de integração numérica usando todas as posições disponíveis na literatura foi deixada para uma outra oportunidade. O resultado foi publicado em \cite{GomesJunior2016}. Porém, a pesquisadora acima nunca nos disponibilizou a ferramenta de integração e, pouco após esse trabalho, ela deixou a pesquisa.

Para contornar esse problema, eu realizei, de Setembro de 2016 a Agosto de 2017, um Doutorado Sanduíche no IMCCE. Nesse período, eu desenvolvi, em colaboração com o doutorando Bruno Morgado, e supervisionado pelo Dr. Valéry Lainey, especialista na área de dinâmica orbital de satélites naturais, uma ferramenta de integração numérica, ajuste às observações e geração de efemérides.

A experiência do Dr. Valéry Lainey fez com que, dentro de um ano, fosse possível desenvolver e concluir um código de integração numérica que pode ser aplicado a satélites naturais e objetos com órbitas heliocêntricas. Ele leva em conta perturbações causadas por outros objetos do Sistema Solar e a perturbação causada pelo achatamento do planeta central.

Ou seja, o código pode ser utilizado para integrar a órbita de outros objetos, além daqueles que serão tratatos nesse trabalho. Isso abre uma possibilidade de colaboração com outros pesquisadores. 
%Para comparar com os resultados de \cite{GomesJunior2016}, eu fiz a integração numérica de alguns satélites irregulares utilizando as mesmas posições da órbita anterior e os mesmos perturbadores. Os resultados foram

Para os satélites irregulares, será importante recuperar todas as posições publicadas na literatura. Além disso, será necessário estudar tais posições para identificar os pesos adequados a serem aplicados a elas durante a integração numérica. Por fim, todos os satélites terão novas órbitas determinadas, de modo a melhorar a predição das ocultações.

\section{Ocultações Estelares} \label{Sec:Occ}

\indent \indent Com o objetivo de obter parâmetros físicos para os satélites irregulares com maior precisão foram feitas predições de ocultações estelares para os 8 maiores satélites de Júpiter (Himalia, Elara, Pasiphae, Sinope, Lysithea, Carme, Ananke e Leda) e o satélite Phoebe de Saturno. Esses objetos são pequenos, se comparados aos TNOs, sendo que o menor dos satélites da amostra, Leda, possui um diâmetro estimado de 20 km.

As predições foram feitas utilizando as posições de estrelas dadas no catálogo UCAC4. Tomou-se como referência as efemérides do STE. Cerca de 3800 eventos foram identificados entre 01 de Janeiro de 2017 e 31 de Dezembro de 2020 para os 9 objetos envolvendo estrelas até magnitude 16. A distribuição da quantidade de ocultações por ano pode ser encontrada na Tabela \ref{Tab: satellite-occultation}.

\begin{table}[h]
\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from January, 2016 up to December, 2020.}
\begin{centering}
\begin{tabular}{lcccccc}
\hline  \hline
%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
Satellite & Diâm. (km) & 2017 & 2018 & 2019 & 2020 & Total \tabularnewline
\hline
Ananke & 28 & 12 & 33 & 182 & 62 & 289 \tabularnewline
Carme & 46 & 11 & 26 & 200 & 167 & 404 \tabularnewline
Elara & 86 & 13 & 31 & 187 & 44 & 275 \tabularnewline
Himalia & 170 & 9 & 44 & 243 & 156 & 452 \tabularnewline
Leda & 20 & 19 & 36 & 321 & 159 & 535 \tabularnewline
Lysithea & 36 & 5 & 27 & 269 & 158 & 459 \tabularnewline
Pasiphae & 60 & 12 & 29 & 289 & 158 & 488 \tabularnewline
Sinope & 38 & 18 & 32 & 327 & 206 & 583 \tabularnewline
\hdashline
Phoebe & 212 & 130 & 180 & 55 & 12 & 377 \tabularnewline
\hline
\end{tabular}
\par \end{centering}
\end{table}

A grande quantidade de ocultações encontradas se devem à passagem de Júpiter em frente ao Plano da Galáxia em 2019-2020 e de Saturno em 2018. Essa condição aumentou em mais de 20 vezes a probabilidade de uma ocultação entre 2017 e 2019. Para os satélites de Urano e Netuno não foram identificadas ocultações em boas condições de observação, uma vez que eles estão em regiões no céu com baixa densidade de estrelas. As predições foram publicadas em \cite{GomesJunior2016}.

Como esses objetos são muito pequenos (ver tamanho estimado na tabela \ref{Tab: satellite-occultation}), em sua maioria as ocultações durarão poucos segundos, portanto apenas eventos com estrelas brilhantes foram selecionados, mas é necessário que hajam câmeras de integração rápida disponíveis. Por outro lado, os satélites de Júpiter estão muito mais perto que os TNOs e como o erro astrométrico é uma medida angular, consequentemente, o erro da sombra da ocultação projetada na Terra será muito menor que para TNOs. Assim, as latitudes a serem cobertas para que uma ocultação por satélite irregular de Júpiter seja detectada corresponde a poucas centenas de quilômetros.

As ocultações preditas estão sendo divulgadas em ferramentas acessadas por observadores amadores de todo o mundo, como o Occult Watcher. Elas serão pouco a pouco atualizadas conforme novos catálogos form publicados e novas órbitas forem obtidas.

\begin{figure}[h]
\begin{centering}
\includegraphics[width = 16.0cm]{figuras/Phoebe_2017-07-06.png}   
\caption{Mapa de Ocultação por Phoebe em 06 de Julho de 2017 observada no Japão.
%The central red dot show the geocentric closest approach of the shadow. The small ones shows the center of the shadow separated by 60s. The lines show the path of the shadow over the Earth. The shadow moves from right to left.
%\textbf{Labels:} Diam: Diâmetro do Objeto; Tmax: Duração máxima do evento para uma observação central; C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; vel: velocity of event in km/s; Delta: Geocentric distance to the occulting object in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$; long: east longitude of subplanet point in degrees, positive towards east.
}
\label{Fig: ocultacao}
\end{centering}
\end{figure}

Até o momento, algumas tentativas foram feitas para observar ocultação por satélites irregulares. Porém, só uma pôde ser observada. Essa foi uma ocultação pelo satélite de Saturno, Phoebe, observada por dois observadores no Japão em 6 de Julho de 2017. É a primeira ocultação por satélite irregular já detectada. A figura \ref{Fig: ocultacao} mostra no mapa a região em que a sombra passou sobre a Terra. As observações já foram recuperadas e estão sendo tratadas.

Essa ocultação e outras que poderão ser observadas serão tratadas por mim. Eu tenho experiência na redução de curvas de luz de ocultação, onde já reduzi curvas de diversas ocultações de Chariklo, Plutão e outros TNOs. Além disso, trabalhei integralmente nas reduções das ocultações de Ceres de 17 de Agosto de 2010 e 25 de Outubro de 2013, inclusive ajustando a forma do corpo aos instantes de imersão e emersão das curvas de luz. Esse trabalho foi publicado em \cite{GomesJunior2015-Ceres}.

\section*{Cronograma para o pós-doc}

\indent \indent Como explicado nos capítulos anteriores, o projeto se divide em três etapas: observação e astrometria dos satélites irregulares; integração numérica de suas órbitas; e observação de ocultações estelares. Para o período de pós-doc de Março de 2018 a Fevereiro de 2019, pretendo realizar as seguintes tarefas:

Observação dos satélites irregulares para astrometria: dentro do programa de observações que o grupo do Rio faz a partir do OPD, os satélites irregulares serão observados. Para os satélites de Júpiter, entre Fevereiro e Setembro de 2018 eles estarão observáveis. Para os satélites de Saturno, entre Março e Outubro. E para os satélites de Netuno, de Maio a Dezembro.

Astrometria: com a publicação da segunda versão do catálogo GAIA em Abril de 2018, todas as observações antigas serão re-reduzidas com esse nova versão. Todo o banco de dados, incluindo as novas imagens, será preparado para uma rápida redução com o novo catálogo. Provavelmente, menos que três meses a partir da publicação serão precisos para o processo de re-redução, considerando as adaptações necessárias para a implementação do GAIA-DR2.

Integração Numérica: como primeira medida, será necessário recuperar na literatura todas as posições publicadas para os objetos estudados. Algumas dessas posições deverão ser convertidas para o Sistema de Referência mais moderno. E todas serão estudadas de forma a se identificar o peso adequado a ser usado na integração numérica. Dentro desse período de um ano, pretendo gerar novas órbitas para os oito satélites irregulares principais de Júpiter e o satélite irregular principal de Netuno, Nereida. Após a re-redução das imagens do OPD ser refeita com o GAIA-DR2, essas órbitas serão atualizadas.  O trabalho de atualização será constante ao longo de todo o período do pós-doc.

Observação de Ocultações Estelares: como mostrado na tabela \ref{Tab: satellite-occultation} existem algumas ocultações previstas para 2018 para os satélites de Júpiter. Esses eventos serão selecionados de acordo com a probabilidade de observação. Porém, devido à recente observação da ocultação por Phoebe, e uma vez que depois de 2018 Phoebe não estará mais cruzando a Via Láctea, uma maior atenção será dada a esse satélite. Uma maior quantidade de eventos detectados permitirá fazer uma composição da forma do satélite em latitudes que não foram observadas pela sonda Cassini. A órbita desse satélite é a melhor conhecida dentre todas os satélites irregulares. O período em que esses eventos poderão ser observados é semelhante ao descrito acima para a observação dos satélites de carátes astrométrico.

Por fim, nesse período, publicarei em revistas científicas de prestígio internacional, a ocultação de Phoebe observada, dentre outras que poderão ser observadas. Publicarei também um novo catálogo de posições astrométricas dos satélites, reduzidas com o GAIA-DR2. Além disso, as órbitas determinadas para esses satélites também serão publicadas.

\bibliography{Referencias}
\bibliographystyle{apalike}

\end{document}

### 2017-08-16 qualify/Qualify-altair.tex
mod 2017 ago16

@@
 \resumo{Resumo}
-\indent \indent O estudo de objetos como TNOs, Centauros e Satélites Irregulares (remanescentes relativamente inalterados da formação do Sistema Solar) nos ajudam a compreender a formação e evolução do Sistema Solar.
+\indent \indent O estudo de objetos como objetos trans-Netunianos (TNOs), Centauros e Satélites Irregulares (remanescentes relativamente inalterados da formação do Sistema Solar) nos ajudam a compreender a formação e evolução do Sistema Solar.
 Poucas foram as sondas enviadas para explorar essas regiões e muito do conhecimento sobre seus constituintes foram adquiridos a partir de observações de solo.
@@
 
-Sabe-se que poucas sondas espaciais foram enviados para estudar o Sistema Solar externo e que a quantidade de objetos estudados é muito pequena.
+Sabe-se que poucas sondas espaciais foram enviadas para estudar o Sistema Solar externo e que a quantidade de objetos estudados é muito pequena.
 Por isso, ainda hoje, as observações de solo tem se mostrado de grande importância.
@@
 Para observações feitas no ESO, o procedimento astrométrico seguido é o mesmo descrito em \citep{Assafin2012}; as medidas (x, y) dos CCDs individuais são pré-corrigidos por um padrão de distorção de campo e todas as posições vindas de diferentes CCDs e mosaicos são combinadas usando o polinômio de grau três para produzir uma solução global para cada noite e campo observado, a as posições ($\alpha$, $\delta$) são então obtidas no sistema do catálogo utilizado.
-Estrelas de referência com resíduos de posição (O-C) que desviam mais de 120 mas (ou valor similar adotado pelo usuario, usualmente 3 sigma do erro do catálogo) são automaticamente eliminadas uma a uma.
+Estrelas de referência com resíduos de posição (O-C) que desviam mais de 120 mas\footnote{mili-segundos de arco (mili-arcsecond, \textit{mas}, em inglês)} (ou valor similar adotado pelo usuário, usualmente 3 sigma do erro do catálogo) são automaticamente eliminadas uma a uma.
 
@@
 Da mesma forma, estando muito distantes, eles se movem relativamente devagar no céu ocultando relativamente poucas estrelas por ano.
-Porém, sendo menos brilhantes (Mag $V > 18$) é possível observar ocultações de estrelas mais fracas (estrelas até $R = 17$), onde a queda de brilho da estrela durante a ocultação poderá ser detectada e, mesmo com baixo sinal-ruído (ex: 5 a 10), o tamanho do corpo pode ser calculado com precisão de pelo menos 10 a 20 km, melhorando à medida em que a velocidade da sombra seja menor que 20 km/s (velocidade típica para um corpo em oposição a 30 UA).
+Porém, sendo menos brilhantes (Mag $V > 18$) é possível observar ocultações de estrelas mais fracas (estrelas até Mag $R = 17$), onde a queda de brilho da estrela durante a ocultação poderá ser detectada e, mesmo com baixo sinal-ruído (ex: 5 a 10), o tamanho do corpo pode ser calculado com precisão de pelo menos 10 a 20 km, melhorando à medida em que a velocidade da sombra seja menor que 20 km/s (velocidade típica para um corpo em oposição a 30 UA).
 
@@
 Assim, estrelas que não estão no UCAC4, catálogo comumente utilizado como referência astrométrica, foram identificadas bem como estrelas que estão mas não tem posições ou movimentos próprios com precisão suficiente.
-Tantos os catálogos de estrelas quanto as cadidatas a ocultação foram publicados \citep{Assafin2010, Assafin2012, Camargo2013}.
+Tanto os catálogos de estrelas quanto as estrelas candidatas a ocultação foram publicados \citep{Assafin2010, Assafin2012, Camargo2013}.
 Esses catálogos são utilizados como representante prático do ICRS no tratamento astrométrico e o conjunto é denominado catálogo WFI.
@@
 Portanto os TNOs também foram observados tanto no ESO quanto no Observatório do Pico dos Dias (OPD, LNA/MCTI, Itajuba, MG., código IAU 874, $45^{\circ} ~34\arcmin ~57\arcsec$ W, $22^{\circ} ~32\arcmin ~04\arcsec$ S, 1864 m).
-À princípio as predições das ocultações eram atualizadas utilizando o método de offsets, porém, em colaboração com o Dr.
+A princípio as predições das ocultações eram atualizadas utilizando o método de offsets, porém, em colaboração com o Dr.
 Josselin Desmars, integrações numéricas das órbitas dos TNOs foram realizadas desenvolvendo-se uma ferramenta denominada NIMA\footnote{Numerical Integration of the Motion of an Asteroid} \citep[submetido]{Desmars2015}.
@@
 Nesse trabalho, posições obtidas de ocultações já realizadas tem um peso muito maior que as restantes, diferentemente do que é feito pelo JPL.
-Da mesma forma, noites em que houveram um número de observações maior que as restantes terá um peso moderado já que se fosse utilizado um peso individual, o peso final de uma noite poderia estar superestimado levando a um erro na integração.
+Da mesma forma, noites em que houveram um número maior de observações que as restantes terão um peso moderado já que se fosse utilizado um peso individual, o peso final de uma noite poderia estar superestimado levando a um erro na integração.
 Neste trabalho contribuí com observações e reduções astrométricas de imagens de TNOs feitas no OPD.
@@
 
-\indent \indent Com o objetivo de obter os parâmetro físicos desses objetos, para melhor caracterizar as propriedades básicas dessas populações, uma grande campanha observacional é realizada para se observar ocultações estelares por TNOs e Centauros.
+\indent \indent Com o objetivo de obter os parâmetros físicos desses objetos, para melhor caracterizar as propriedades básicas dessas populações, uma grande campanha observacional é realizada para se observar ocultações estelares por TNOs e Centauros.
 Essa campanha é liderada por pesquisadores brasileiros, franceses e espanhóis e tem colaboração com observadores espalhados por todo o mundo.
 
-Os objetos foram basicamente selecionados de acordo o grau de interesse científico (representar uma classe de TNO, ser um  objeto peculiar, etc), tamanho (quanto maior, mais chances de observar a ocultação) e região do céu (quanto mais estrelas, maior  a probabilidade de ocultar estrelas brilhantes, permitindo curvas de luz de melhor resolução espacial).
+Os objetos foram basicamente selecionados de acordo com o grau de interesse científico (representar uma classe de TNO, ser um  objeto peculiar, etc), tamanho (quanto maior, mais chances de observar a ocultação) e região do céu (quanto mais estrelas, maior  a probabilidade de ocultar estrelas brilhantes, permitindo curvas de luz de melhor resolução espacial).
 
@@
 
-Outro evento que observei foi uma ocultação de 2003VS2 dia 07 de Novembro de 2014 no OPD.
+Outro evento que observei foi uma ocultação do TNO 2003VS2 dia 07 de Novembro de 2014 no OPD.
 As variações do céu não permitiram uma boa observação de forma que a curva de luz estava muito ruidosa, mas boa o suficiente próximo ao instante do evento.
@@
 Nesse evento foram obtidas 3 cordas positivas e 7 negativas.
-A redução preliminar revelou que 2003VS2 é possui um corpo levemente alongado com aproximadamente 600 km de diâmetro.
+A redução preliminar revelou que 2003VS2 possui um corpo levemente alongado com aproximadamente 600 km de diâmetro.
 O trabalho ainda está em análise.
@@
 De fato, observações diretas de Ceres com a utilização de ótica adaptativa indica que ele é um esferoide achatado nos polos \citep{Drummond2014}.
-O conhecimento preciso de seu tamanho e forma é extrema importância para modelos de densidade, estrutura interna e diferenciação.
+O conhecimento preciso de seu tamanho e forma é de extrema importância para modelos de densidade, estrutura interna e diferenciação.
 
@@
 
-Os dois eventos foram preditos por Steve Preston\footnote{Predições publicadas em \url{http://asteroidoccultation.com}.} para a IOTA (International Occultation Timing Association), durante predições de rotina de ocultações de asteroides de estrelas brilhantes.
+Os dois eventos foram preditos por Steve Preston\footnote{Predições publicadas em \url{http://asteroidoccultation.com}.} para a IOTA (International Occultation Timing Association), durante predições de rotina de ocultações de estrelas brilhantes por asteroides.
 Os caminhos das sombras podem ser visualizados na Fig.
@@
 
-O ajuste dos dados da ocultação consiste em minimizar uma função dep $\chi^{2}$ clássica para cada curva de luz, como descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}.
+O ajuste dos dados da ocultação consiste em minimizar uma função de $\chi^{2}$ clássica para cada curva de luz, como descrito em \cite{Sicardy2011} e \cite{Braga-Ribas2013}.
 Os parâmetros livres para ajustar são os instantes de ingresso e egresso que fornece o valor mínimo de $\chi^{2}$ ($\chi^{2}_{min}$).
@@
 Para reduzir o ruído, os dados foram binados por grupos de cinco imagens -- com exceção de Greenbelt, onde grupos de dez imagens foram utilizadas.
-Esse procedimento o tempo de integração efetivo por um fator de 5 (ou 10).
+Esse procedimento aumenta o tempo de integração efetivo reportado por um fator de 5 (ou 10).
 Da mesma forma que para o evento de 2010, uma normalização adicional por um polinômio foi aplicada.
@@
 
-Uma comparação entre os tempos obtidos pela estação de Brookline e o restante mostrou que esses têm um atraso de aproximadamente 64 s. Dessa forma, não utilizamos os tempos de Brookline na análise.
+Uma comparação entre os tempos obtidos pela estação de Brookline e o restante mostrou que Brookline têm um atraso de aproximadamente 64 s.
+Dessa forma, não utilizamos os tempos de Brookline na análise.
 
@@
 
-O evento de 1984 \citep{Millis1987} é a única outra ocultação que podemos comparar nosssos resultados.
-As medidas do diâmetro não se acordam dentro de 2$\sigma$.
+O evento de 1984 \citep{Millis1987} é a única outra ocultação com a qual podemos comparar nossos resultados.
+As medidas do diâmetro não estão de acordo dentro de um fator de 2$\sigma$.
 É difícil dizer com certeza as razões dessa divergência.
@@
 \hline
-$972 \pm 6$  & $0.08  \pm 0.03$  & Occultação & 1\\
+$972 \pm 6$  & $0.08  \pm 0.03$  & Ocultação & 1\\
 $967 \pm 10$ & $0.078 \pm 0.015$ & Keck+VTL    & 2\\
@@
 $975 \pm 4$  & $0.067 \pm 0.005$ & HST         & 4\\
-$959 \pm 5$  & $0.05  \pm 0.01$  & Occultação & 5\\
+$959 \pm 5$  & $0.05  \pm 0.01$  & Ocultação & 5\\
 \hline
@@
 Por fim, Nereida foi observado em 1989 pela sonda \textit{Voyager II} e seu tamanho foi obtido com um erro de 25 km \citep{Smith1989}.
-Para outros satélites irregulares de Júpiter, seus tamanhos foram estimados por \cite{Rettig2001} impondo um albedo de 4\% com a justifica de que esse valor é um albedo nominalmente utilizado para objetos do Sistema Solar Externo.
+Para outros satélites irregulares de Júpiter, seus tamanhos foram estimados por \cite{Rettig2001} impondo um albedo de 4\% com a justificativa de que esse valor é um albedo nominalmente utilizado para objetos do Sistema Solar Externo.
 
@@
 Durante a instabilidade do Sistema Solar primordial, ocorreram encontros entre os planetas externos.
-Esses encontros planetários poderia haver troca de energia e momento angular entre os planetas e os objetos próximos tornando possível a captura de um corpo irregular por um planeta gigante.
+Nesses encontros planetários poderia haver troca de energia e momento angular entre os planetas e os objetos próximos tornando possível a captura de um corpo irregular por um planeta gigante.
 Nesse cenário, a taxa de sobrevivência de satélites anteriores ao LHB (Bombardeamento Pesado Tardio\footnote{"Late Heavy Bombardment", em inglês}) é muito pequena.
@@
 \cite{Nesvorny2003} estudou as taxas de colisões entre satélites irregulares e concluiu que alguns satélites podem ter sido removidos por colisão com um satélite maior.
-A taxa de colisão entre os satélites do grupo de Himalia (Himalia, Elara, Lysithea and Leda, principalmente), por exemplo, foi encontrado como sendo maior que "um" durante a idade do Sistema Solar sugerindo que sua estrutura atual foi originada por colisão satélite-satélite.
-
-Para Phoebe, materiais ejetados de sua superfície causadas por impactos poderiam evoluir devido ao arrasto de Poynting-Robertson e colidir com Iapetus causando causando a grande variação de albedo observada no satélite \citep{Nesvorny2003}.
+A taxa de colisão entre os satélites do grupo de Himalia (Himalia, Elara, Lysithea and Leda, principalmente), por exemplo, foi encontrado como sendo maior que 1 durante a idade do Sistema Solar sugerindo que sua estrutura atual foi originada por colisão satélite-satélite.
+
+Para Phoebe, materiais ejetados de sua superfície causadas por impactos poderiam evoluir devido ao arrasto de Poynting-Robertson e colidir com Iapetus causando a grande variação de albedo observada no satélite \citep{Nesvorny2003}.
 De fato, a sonda \textit{Cassini} detectou em Phoebe uma característica de absorção em 2.42 $\mu m$, provavelmente combinações de CN, que foi também detectada no lado escuro de Iapetus \citep{Clark2005}
@@
 
-Nesse contexto, durante o Mestrado, foi realizado junto ao grupo um trabalho de caráter astrométrico para a melhoria das efemérides dos satélites irregulares de Júpiter Saturno.
+Nesse contexto, durante o Mestrado, foi realizado junto ao grupo um trabalho de caráter astrométrico para a melhoria das efemérides dos satélites irregulares de Júpiter e Saturno.
 Em colaboração com o Dr.
@@
 
-Um dos principais resultados desse trabalho foi a grande quantidade de posições obtidas para os satélites irregulares em comparação com a quantidade utilizada nas integrações numéricas atuais (ver tabela \ref{Tab: comparison-horizons}).
+Um dos principais resultados desse trabalho foi a grande quantidade de posições obtidas para os satélites irregulares em comparação com a quantidade utilizada nas integrações numéricas atuais publicadas por \cite{Jacobson2012} (ver tabela \ref{Tab: comparison-horizons}).
+Dessa forma, é de se esperar que nossas posições contribuam significativamente para a melhoria da órbita desses objetos.
 
@@
 
-Com o objetivo ver o potencial das nossas posições em melhorar as órbitas dos satélites irregulares, analisamos os offsets das nossas posições com respeito às efemérides do JPL.
+Com o objetivo de ver o potencial das nossas posições em melhorar as órbitas dos satélites irregulares, analisamos os offsets das nossas posições com respeito às efemérides do JPL.
 Tomando Carme como exemplo, plotamos os offsets médios das efemérides para cada noite na Fig.
@@
 Um método que tem se mostrado eficiente para a obtenção dos parâmetros físicos (tamanho, forma, albedo, etc) desses objetos é o método de ocultações estelares, que proporciona medidas tão precisas que são apenas superadas por medidas oriundas de sondas.
-Uma vez que suas efemérides não são muito precisas, predizer e observar ocultações estelares é muito difícil e nenhuma observação de um evento desses para um satélite irregular ainda foi relatado na literatura.
+Uma vez que suas efemérides não são muito precisas dado os seus tamanhos estimados, predizer e observar ocultações estelares é muito difícil e nenhuma observação de um evento desses para um satélite irregular ainda foi relatado na literatura.
 
@@
 Um total de 588 eventos foram obtidos entre 01 de Janeiro de 2015 e 31 de Dezembro de 2017 para os 9 objetos sendo que a maioria será descartada por passar em regiões de oceano ou onde não há observadores disponíveis.
+Estima-se que restem ao final cerca de 10\% desse número.
 Para Nereida, apenas uma estrela UCAC4 foi identificada para ser ocultada.
@@
 Erros em ascensão reta interferem majoritariamente no instante em que a ocultação ocorrerá.
-Já erros em declinação, principalmente na latitude geográfica em que a sombra passará.
+Já erros em declinação principalmente na latitude geográfica em que a sombra passará.
 
@@
 Espera-se que com o grande número de posições publicadas em \cite{GomesJunior2015-Irregular} as órbitas sejam calculadas com maior precisão.
-Esse integração numérica será trabalhada durante o doutorado sanduíche em Paris em colaboração com o Dr.
+Essa integração numérica será trabalhada durante o doutorado sanduíche em Paris em colaboração com o Dr.
 Jean-Eudes Arlot e o Dr.
@@
 
-Esse teste consiste em se observar o objeto ocultante e a estrela a ser ocultada próximo ao evento, quando os dois objetos estiverem presentes no mesmo campo, preferencialmente próximo ao instante do evento quando os objetos estarão próximos pum do outro.
-Assim as posições relativas entre os dois objetos terá mínima influência dos erros das estrelas de catálogo utilizadas na redução e das possíveis distorções de campo.
+Esse teste consiste em se observar o objeto ocultante e a estrela a ser ocultada próximo ao evento, quando os dois objetos estiverem presentes no mesmo campo, preferencialmente próximo ao instante do evento quando os objetos estarão próximos um do outro.
+Assim as posições relativas entre os dois objetos terão influências mínimas dos erros das estrelas de catálogo utilizadas na redução e das possíveis distorções de campo.
 
@@
 
-\indent \indent Com o objetivo de estudar os efeitos de maré sobre o sistema Netuno-Tritão, serão obtidos as posições desses objetos em mais de 7000 imagens observadas no OPD desde 1992 até o presente.
+\indent \indent Com o objetivo de estudar os efeitos de maré sobre o sistema Netuno-Tritão, serão obtidas as posições desses objetos em mais de 7000 imagens observadas no OPD desde 1992 até o presente.
 Uma vez que eles estão muito próximos no céu e Netuno é muito mais brilhante do que Tritão, a posição do satélite pode estar afetada pela luz do planeta.
@@
 O perfil resultante não tem forma analítica fechada, devendo ser calculado numericamente.
-Este estudo ainda está em andamento.
+Ainda estamos estudando a possibilidade de se obter funções analíticas para a PSF.
 
@@
 
-\indent \indent Nos mesmos moldes em que foram feitas observações dos caminhos de TNOs para predição de ocultações estelares (ver seção \ref{Sec: TNOs-astrometria}), o caminho de Tritão também foi observado gerando um catálogo de estrelas inclusos no catálogo WFI.
+\indent \indent Nos mesmos moldes em que foram feitas observações dos caminhos de TNOs para predição de ocultações estelares (ver seção \ref{Sec: TNOs-astrometria}), o caminho de Tritão também foi observado gerando um catálogo de estrelas nos moldes do catálogo WFI.
 Infelizmente, devido a Netuno estar passando por regiões pouco populadas por estrelas brilhantes, nenhum evento pôde ser observado oriundo deste trabalho.

### 2017-08-16 relatorio/Netuno/Relatorio_2.tex
mod 2017 ago16

--- 2016-04-27 relatorio/Occ-Irregular/Relatorio.tex
@@
 \usepackage[left=3cm,right=2cm,top=3cm,bottom=2cm]{geometry}
-\usepackage{arydshln}
-\usepackage{scalefnt}
 %\usepackage[alf]{abntcite}
@@
 \newcommand{\BC}{Boller \& Chivens }
-\newcommand{\noccs}{5442 }
 
@@
 \author{Altair Ramos Gomes Júnior}
-\title{Occultation Predictions of Irregular Satellites}
+\title{Astrometry of the Neptune-Triton System}
 \begin{document}
@@
 
-\section*{Introduction}\label{Sec: introducao}
+\section*{Chromatic Refraction}
 
-Irregular satellites revolve around giant planets at large distances in eccentric, highly inclined and frequently retrograde orbits.
-Because of these peculiar orbits, it is largely accepted that these objects did not form by accretion around their planet, but were captured in the early Solar System \citep{Sheppard2005}.
+Since the colors of Neptune and Triton are different, it is expected that their positions have influence of chromatic refraction with different intensities.
+The apparent position of Neptune, which is more blue than Triton, would be more shifted towards the zenith than the Triton's position.
 
-There is no consensus for a single model explaining where the irregular satellites were formed.
-\cite{Cuk2004} showed that the progenitor of the Himalia group may have originated in heliocentric orbits similar to the Hilda asteroid group.
-\cite{Sheppard2005} stated that the irregular satellites may be some of the objects that were formed within the giant planets region.
+To test the effects of chromatic refraction I looked for nights with many observations distributed over the night.
+I selected two of them (August 22, 2004 and September 25, 2011) with more than 4 hours between the first and last observations.
+In the night of 2004 it was used a filter called "Dark" which we suppose it means no filter, however we are no sure so I classified it as unknown.
+In the night of 2011 it was used a I filter.
 
-\cite{Grav2003} and \cite{Grav2007} showed that the irregular satellites from the giant planets have their colors and spectral slopes similar to C-, D- and P-type asteroids, Centaurs and trans-neptunian objects (TNOs).
-This suggests that they may have come from different locations in the early solar system.
+In Fig \ref{Fig:refraction} it is plotted the difference in the positions of Triton and Neptune compared to the difference of their ephemeris in RA and DEC for both nights.
 
-\cite{Sheppard2005} and \cite{Jewitt2007} also explored the possibility that the irregular satellites originated as comets or TNOs.
-TNOs are highly interesting objects that, due to their large heliocentric distances, may be highly preserved with physical properties similar to those they had when they were formed \citep{Barucci2008}.
-This is even more true for the smaller objects, since in principle larger sizes favour physical differentiation processes in the body and vice-versa.
-However, due to the distance, the smaller TNOs from this region are more difficult to observe.
-Thus, if irregular satellites - or at least a few of them - do share a common origin with small TNOs, and since these objects are situated at much closer heliocentric distances now, this gives a unique chance of observing and studying representatives of this specific TNO population in much greater detail than could ever be possible by direct observation of this population in the Kuiper Belt.
+\begin{figure}[h]
+\includegraphics[width=16.0cm]{plot_hour_gr1.png} 
+\caption{Distribution of the difference of offsets in the sense Triton - Neptune for the nights August 22, 2004 and September 25, 2011.}
+\label{Fig:refraction}
+\end{figure}
 
-Phoebe is the most studied irregular satellite.
-\cite{Clark2005} suggest that its surface is probably covered by material of cometary origin.
-It was also stated by \cite{Johnson2005} that if the porosity of Phoebe is 15\%, Phoebe would have an uncompressed density similar to those of Pluto and Triton.
+Teoretically, the difference in the positions of the objects in the sense Triton - Neptune compared to the difference in the ephemeris over a night would cause the following effects.
 
-Size, shape, albedo and composition would help a lot to trace back their true origin, but these physical parameters are yet poorly known for irregular satellites.
-The observation of stellar occultations would allow for the determination of such parameters.
-Indeed Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018, improving a lot the chances of observing such events in the near future.
+\begin{itemize}
+\item RA: the difference in the offsets would be positives in the East side of the sky, negative in the West side and zero in the culmination, with the assumption that only chromatic refraction affects the offsets.
+\item DEC: the difference in the offsets in the culmination would be positive (Neptune is in the North side in both nights for the site).
+Farther from the culmination the difference in the offsets would be more positive.
+\end{itemize}
 
-%\cite{GomesJunior2015} obtained 6523 suitable positions for 18 irregular satellites between 1992 and 2014 with an estimated error in the positions of about 60 to 80 mas.
-For some satellites the number of positions obtained is comparable to the number used in the numerical integration of orbits by the JPL \citep{Jacobson2012}.
-They pointed out that the ephemeris of the irregular satellites have systematic errors that may reach 200 mas for some satellites.
+Fig \ref{Fig:refraction} clearly shows that that the chromatic refraction is affecting the offsets confirming our expectation.
 
-For this, we made new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) using only the positions obtained by \cite{GomesJunior2015}.
-We aim to eliminate the systematic errors in the JPL ephemeris pointed out by these authors.
-%For an object at the distance of Jupiter, this represents an error langer than 700 km in the shadow path of an occultation.
-For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \cite{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Center and observations from Flagstaff.
-With this, we predicted stellar occultation for the 9 objects.
-
-Using the derived ephemerides and the UCAC4 catalogue we managed to identify \noccs candidate stellar occultations between January 2016 and December 2020 for the 9 irregular satellites studied here.
-
-Phoebe, being the most studied object with a good measured size, can be used to calibrate and evaluate the technique for similar objects.
-Up to date, no observation of a stellar occultation by an irregular satellite was published.
-Since their estimated sizes are very small (see Table \ref{Tab: satellite-diameter}), this may have discouraged earlier tries.
-But, in fact, given their relatively closer distances as compared to TNOs and Centaurs, and considering the precision of our ephemeris and of star positions, we can now reliably predict the exact location and instant where the shadow of the occultation will cross the Earth.
-
-\begin{table}
-\caption{\label{Tab: satellite-diameter} Estimated diameter of the satellites and correspondent apparent diameter}
-\begin{center}
-\begin{tabular}{lccc}
-\hline  \hline
-\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
-Satellite  & mas$^ {a}$  & km & Ref. \tabularnewline
+\begin{table*}[h]
+\centering
+\caption{Difference in the offsets of Triton-Neptune of the nights used as tests of chromatic refraction with sigma-clipping procedure}
+\label{Tab:dados}
+\begin{tabular}{|c|c|c|c|c|c|c|}
 \hline 
-Ananke & 8 & 29 & 1 \tabularnewline
-Carme & 13 & 46 & 1 \tabularnewline
-Elara & 24 & 86 & 1 \tabularnewline
-Himalia & 41 & $(150\times120) \pm 20^{b}$ & 2 \tabularnewline
-Leda & 5 & 20 & 1 \tabularnewline
-Lysithea & 10 & 36 & 1 \tabularnewline
-Pasiphae & 17 & 62 & 1 \tabularnewline
-Sinope & 10 & 37 & 1 \tabularnewline
-\hdashline
-Phoebe & 32 & $212 \pm 1.4^{b}$ & 3 \tabularnewline
-%\hdashline
-%Nereid & 15 & $340 \pm 50^{c}$ & 4 \tabularnewline
-%Triton & 124 & $2707 \pm 2.0^{c}$ & 5 \tabularnewline
+Night & N & $\mu$ RA & $\mu$ DEC & $\sigma$ RA & $\sigma$ DEC & Ncat\\
+\hline
+2004-08-22 & 29 & 2 & 45 & 34 & 28 & 13 \\ 
+2011-09-25 & 106 & -20 & 67 & 11 & 11 & 17\\
 \hline 
 \end{tabular}
-\end{center}
-References: (1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}.\\
-$^{a}${Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
-$^{b}${From Cassini observations.}
-\par
-\end{table}
-
-%Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
-In contrast to TNOs, the irregular satellites have much better ephemeris because the orbits of their host planets are better  known, their observational time span is much wider and covers many orbital periods.
-Moreover, the irregular satellites are much closer to Earth which implies in a much smaller shadow path error in kilometers.
-These advantages may be somewhat balanced by the smaller sizes estimated for the irregular satellites.Thus, in comparison, the chances for a successful observation of an stellar occultation by an irregular satellite should be considered at least also as good as those by TNOs.
-
-\section*{Orbit computations} \label{Sec: integration}
-
-\cite{GomesJunior2015} published 6523 precise positions for 18 irregular satellites from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1992 and 2014.
-
-Here we compute new orbits based on the observations published in \cite{GomesJunior2015}.
-First, because the reduction was made with a consistent and precise stellar catalogue and with a robust astrometry \citep[PRAIA,][]{Assafin2011}.
-Second, besides recent observations, this consistent set of numerous and precise positions covers many orbital periods at many distinct orbital plane sights, so that the orbital inclinations along with all other orbital elements could be satisfactorily derived without the need of further position sets.
-For these reasons, only this set of positions was used for the satellites of Jupiter.
-
-Due to the context of this work regarding to stellar occultations, the orbit fitting procedures used aimed primarily to derive precise ephemerides for the near future.
-The procedures allow for the improvement of orbits as new observations are added.
-
-\subsection*{Special tailored Ephemerides (STE) for Jupiter irregular satellites}
-
-The last observations used to develop JPL current ephemeris of the irregular satellites of Jupiter were obtained in 2012 \citep{Jacobson2012}.
-As a result, the errors in the JPL ephemeris for the current epoch are large enough to prevent accurate predictions of stellar occultations without any corrections.
-
-%Our numerical model describes the dynamical evolution of the irregular satellites of Jupiter in a jovicentric reference frame.
-The satellites are submitted to the influence of the Sun and the rest of the solar system, as well as that of the Galilean satellites and the first harmonics of Jupiter's gravity field.
-The axes of the reference frame are those of the ICRS.
-%
-%We use the following notations:
-%\begin{itemize}
-%\item $i$ and $l$ one of the irregular satellites of Jupiter
-%\item $J$ Jupiter 
-%\item $j$ another body of the Solar System
-%\item $M_j$ the mass of the $j$-th body, not an irregular satellite
-%\item $m_i$ the mass of the irregular satellite $i$
-%\item $\vec{r_i}$ the position of the $i$-th body with respect to the barycentre of Jupiter System
-%\item $r_{ij}$ the distance between bodies $i$ and $j$  
-%\item $R_J$ the radius of Jupiter
-%\item $J_n$ the dynamic polar oblateness of the n-th order for Jupiter's gravity field
-%\item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
-%\item $\Phi_i$ is the inclination of the $i$-th satellite with respect to Jupiter's equator.
-%\end{itemize}
-%
-%For an irregular satellite $i$, under the gravitational influence of Jupiter, the $\mathcal{N}-1$ other irregular satellites, the regular Jovian satellites and the rest of the Solar System ($\mathcal{N}$ bodies), the equation of motion is:
-%\begin{equation}\begin{array}{ll}
-%
-%\ddot{\vec{r_i}}= & \displaystyle -GM_J\frac{\vec{r_J}-\vec{r_i}}{r_{iJ}^3}-\sum_{l=1,l\neq i}^\mathcal{N'}Gm_l\frac{\vec{r_l}-\vec{r_i}}{r_{il}^3}\\
-%&\displaystyle -\sum_{j=1}^\mathcal{N}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3} \right)\\
-% & \displaystyle +GM_J \nabla U_{\bar{l}\hat{J}} -\sum_{l=1}^\mathcal{N} Gm_l\nabla U_{\bar{l}\hat{J}}
-%\end{array}
-%\label{Eq:1}
-%\end{equation}
-%where the last term in the brackets and the last term in Eq. \ref{Eq:1} represent undirect perturbations. The oblateness potential seen by the body $i$ because of Jupiter is:
-%\begin{equation}\begin{array}{ll}
-%
-%U_{\bar{l}\hat{J}}=&\displaystyle -\frac{R_J^2 J_2}{r_{iJ}^3}\left(\frac{3}{2}\sin^2 \Phi_i-\frac{1}{2}\right)\\ &\\ & 
-%\displaystyle-\frac{R_J^4 J_4}{r_{iJ}^5}\left(\frac{35}{8}\sin^4 \Phi_i-\frac{15}{4}\sin^2 \Phi_i+\frac{3}{8}\right)\\
-%& \\
-%&\displaystyle-\frac{R_J^6 J_6}{r_{iJ}^7}\left(\frac{231}{16}\sin^6 \Phi_i-\frac{315}{16}\sin^4 \Phi_i+\frac{105}{16}\sin^2 \Phi_i-\frac{5}{16}\right)
-%
-%\end{array}
-%\end{equation}
-%
-%The expressions of $\nabla_lU_{\bar{l}\, \hat{i}}$ and $\nabla_i U_{\bar{i}\, \hat{l}}.$ have been developed in \citet{Lainey2004}.
-The equations of motion are integrated with the numerical integrator RADAU \citep{Everhart1985}.
-%Our model was fitted to the observations through a least-squares procedure.
-The satellites were integrated one dynamical family at a time, to gain computing time, while losing minimum precision.
-Indeed, the interactions between satellites not belonging to the same dynamical family are negligible considering the short timespan of our integration.
-%
-%The initial osculating elements at the origin of integration are presented in Table \ref{Tab: sat_ell}.
-%
-%\begin{table*}
-%\scalefont{0.8}
-%\caption{Initial osculating elements for Jupiter irregular satellites at JD 2451545.0.
-% }\label{Tab: sat_ell}
-%\begin{center}
-%\begin{tabular}{ccccccc}
-%\hline\hline
-%Satellite & a (km) & e & I$\degr$ & $\Omega\degr$ & $\omega\degr$ & $v\degr$ \\ 
-%\hline
-%Himalia &   11372100 $\pm$ 500    &    0.166 $\pm$ 0.002      &   45.14 $\pm$ 0.15      &   39.77 $\pm$ 0.19      &   351.48 $\pm$ 0.46      &   97.35 $\pm$ 0.48    \\
-%Elara &   11741170 $\pm$ 690  &      0.222 $\pm$ 0.002      &   28.64 $\pm$ 0.18      &   68.42 $\pm$ 0.43      &   179.82 $\pm$ 0.56      &   339.08 $\pm$ 0.82  \\
-%Lysithea &   11739900 $\pm$ 1300  &      0.136 $\pm$ 0.004      &    51.12 $\pm$ 0.27     &   5.53 $\pm$ 0.52      &   53.0 $\pm$ 1.5      &   318.9 $\pm$ 2.0   \\
-%Leda &   11140300  $\pm$ 4300  &     0.173  $\pm$ 0.007     &   16.15  $\pm$ 0.75    &   272.6  $\pm$ 1.7    &   212.2  $\pm$ 3.6          &   218.8  $\pm$ 3.2  \\
-%Pasiphae &  23425000  $\pm$ 5000    &     0.379  $\pm$ 0.001       &   152.44 $\pm$ 0.10      &   284.59 $\pm$ 0.21      &   135.96 $\pm$ 0.19      &   236.97 $\pm$ 0.16 \\
-%Sinope &   22968800 $\pm$ 5200   &     0.316 $\pm$ 0.002      &   157.76 $\pm$ 0.12      &   256.62 $\pm$ 0.55      &   298.38 $\pm$ 0.55      &   167.57 $\pm$ 0.19    \\
-%Carme &   24202924 $\pm$ 4800      &  0.242 $\pm$ 0.001      &   147.13 $\pm$ 0.10      &   154.01 $\pm$ 0.25      &   47.90 $\pm$ 0.29      &   234.41 $\pm$ 0.19  \\
-%Ananke &  21683800  $\pm$ 7200  &     0.380 $\pm$ 0.002      &   172.29 $\pm$ 0.20      &   56.9 $\pm$ 1.2      &   123.3 $\pm$ 1.2      &   231.24 $\pm$ 0.21  \\
-%%Themisto  & 7393800    &     0.198       &   25.77  &   220.0     &   216.3    &   262.1     \\
-%\hline
-%\end{tabular} 
-%\end{center}
-%\begin{flushleft}
-%\textbf{Notes}: a: semimajor axis; e : excentricity; I: inclination relative to the equatorial reference plane J2000; $\Omega$: longitude of the ascending node; $\omega$: argument of periapsis; $v$: true anomaly.
-%\end{flushleft}
-%\end{table*}
-
-All the orbits determined for the satellites show satisfying residuals.
-The residuals are smaller than those obtained with JPL ephemeris, which was expected because the accuracy of an ephemeris decreases when we get further from the time of observations.
-The main risk of divergence over time comes from the possible absence of long-term effects when fitting to a short timespan of observations.
-If that were the case, our ephemeris would diverge too quickly to be of any use.
-JPL ephemerides are fitted over all the available observations.
-As a result, they will diverge less quickly than our own.
-Though they are no longer precise enough for our use, they remain a precious reference to identify whether our own model presents a quick divergence.
-
-We compared our ephemeris to the JPL for all the Jupiter satellites we fitted, until 2018.
-For instance, the divergence between 2015 and 2018 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.
-
-Fig.
-\ref{Fig: JPL-STE} displays the offsets of the positions published by \cite{GomesJunior2015} for the satellite Carme in declination relative to our ephemeris, to \cite{Jacobson2012} JUP300 JPL ephemeris and \cite{Emelyanov2008} ephemeris.We see that the systematic JPL ephemeris offsets pointed out by \cite{GomesJunior2015} are reduced with our ephemeris, as expected.
-
-\begin{figure}
-\includegraphics[width=16.0cm]{figures/Carme_ephemeris.eps}
-\caption{Offsets in declination of the positions published by \protect\cite{GomesJunior2015} for Carme.
-The red "x" relative to the special-tailored ephemeris, the blue "+" relative to the JUP300 JPL ephemeris and the green dot relative to \protect\cite{Emelyanov2008}.
-As expected, the ephemeris systematic errors pointed out by \protect\cite{GomesJunior2015} are reduced with the STE ephemeris.
-\label{Fig: JPL-STE}}
-\end{figure}
-
-The obtained ephemeris is hereafter refered as STE, for special-tailored ephemeris.
-
-\subsection*{Phoebe's ephemeris}
-
-For the specific case of Phoebe, the ninth satellite of Saturn, we have updated the ephemeris published in \cite{Desmars2013}.
-The new ephemeris (PH15) used the same dynamical model, including the perturbations of the Sun and the eight planets, the eight major satellites of Saturn and the $J_2$ parameter.
-The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \cite{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center\footnote{\url{http://lnfm1.sai.msu.ru/neb/nss/bsapoouf.htm}}), and observations from Flagstaff \citep{NOFS} between 2012 and 2014.
-It represents a total number of 5886 observations from 1898 to 2014.
-In contrast, in \cite{Desmars2013} was used 3367 observations from 1898 to 2012.
-It represents an increase of almost 75\% in the number of observations.
-
-In Fig.
-\ref{Fig:eph-Phoebe} we compare our ephemeris (PH15) with the SAT375 JPL\footnote{Jacobson, R.A.
-2015-Feb-27.
-"Satellite Ephemeris: SAT375", JPL Satellite Ephemeris File Release, \url{ftp://ssd.jpl.nasa.gov/pub/eph/satellites/nio/LINUX_PC/sat375l.txt}} ephemeris.
-The difference between them is smaller than 30 mas (< 10 mas in Declination).
-This difference is smaller than the apparent diameter of Phoebe (see Table \ref{Tab: satellite-diameter})
-
-\begin{figure}
-\begin{centering}
-\includegraphics[width=16.0cm]{figures/Phoebe.eps} 
-\caption{Comparison between the PH15, SAT375 JPL and \protect\cite{Emelyanov2007} ephemeris for the satellite Phoebe.}
-\label{Fig:eph-Phoebe}
-\end{centering}
-\end{figure}
-
-%\subsection{Triton's and Nereid's ephemeris}
-%
-%Triton is an uncommon satellite.
-Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones.
-Because Triton's orbit size is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded from the irregular satellites' class, but still studied together by many authors \citep{Sheppard2005, Jewitt2007}.
-Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have the same origin as the TNOs \citep{Agnor2006}.
-However, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
-%
-%For this satellite, we used the most recent ephemeris published by \cite{Emelyanov2015} which is based on more observations than previous ephemeris and covers 4 years more than the ephemeris of \cite{Jacobson2009}.
-In Fig.
-\ref{Fig:eph-Triton} we compare it with the ephemeris published in \cite{Zhang2014} and the NEP081 JPL ephemeris \citep{Jacobson2009}.
-The offsets between the three ephemeris are smaller than 20 mas for the period 2015-2021.
-This value is much smaller than the apparent size of Triton (124 mas, see Table \ref{Tab: satellite-diameter}), which indicates a good probability of observing an event by this object.
-%
-%\begin{figure*}
-%\begin{centering}
-%\subfigure[NEP081 JPL and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
-%\subfigure[\protect\cite{Zhang2014} and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/Zhang-EME_Triton.eps} \label{Fig:eph-Triton-zhang}}
-%\caption{Comparison between the \protect\cite{Emelyanov2015}, \protect\cite{Zhang2014} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Triton.}
-%\label{Fig:eph-Triton}
-%\end{centering}
-%\end{figure*}
-%
-%For Nereid, we used the ephemeris published by \cite{Emelyanov2011}, which uses more recent observations than the JPL ephemeris published by \cite{Jacobson2009}.
-The comparison between the two ephemeris can be seen in Fig.
-\ref{Fig:eph-Nereid}.
-The right ascension offsets between them are smaller than 60 mas and the declination ones are smaller than 15 mas.
-For Nereid, ephemeris errors in right ascension translate to uncertainties in the central instant of a stellar occultation, so offsets of 60 mas are not critical.
-The declination ephemeris offsets are of the order of the apparent diameter (15 mas, see Table \ref{Tab: satellite-diameter}), resulting in fair chances (50\%) that the shadow crosses the expected Earth latitude predicted for an occultation.
-%
-%\begin{figure}
-%\begin{centering}
-%\includegraphics[width=8.8cm]{figures/JPL-EME_Nereid.eps}
-%\caption{Comparison between the \protect\cite{Emelyanov2011} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Nereid.}
-%\label{Fig:eph-Nereid}
-%\end{centering}
-%\end{figure}
-%
-%\cite{GomesJunior2015} published 902 new positions for Nereid from observations between 1992 and 2014.
-The development of a better ephemeris for Nereid using these new positions in the context discussed here of near future stellar occultations is desirable, but is out of the scope of this paper.
-
-\section*{Prediction of stellar occultations} \label{Sec: predictions}
-
-\subsection*{Candidate events}
-
-The prediction of the occultations was made by crossing the stellar coordinates and proper motions of the UCAC4 catalogue \citep{Zacharias2013} with the ephemeris presented in the previous section.
-The search for stellar candidates follows the same procedure as presented by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}.
-
-We predicted occultations for the 8 major irregular satellites of Jupiter,  Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope, and for Phoebe of Saturn.
-
-A total of \noccs events were identified between January 2016 and December 2020.
-In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite.
-It is possible to see an increase in the number of events found for Phoebe in 2018 and for the satellites of Jupiter in 2019-2020.
-This is because at that periods these satellites will cross the apparent galactic plane.
-We call attention that about 10\% of the events will involve stars brighter than magR=14 (and almost 25\% brighter than magR=15), which helps the attempt of amateur observers.
-
-\begin{table}[h]
-\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from January, 2016 up to December, 2020.}
-\begin{centering}
-\begin{tabular}{lcccccc}
-\hline  \hline
-%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
-Satellite  & 2016 & 2017 & 2018 & 2019 & 2020 & Total \tabularnewline
-\hline
-Ananke & 12 & 16 & 49 & 359 & 187 & 623 \tabularnewline
-Carme & 20 & 14 & 30 & 369 & 220 & 653 \tabularnewline
-Elara & 14 & 16 & 33 & 305 & 193 & 561 \tabularnewline
-Himalia & 15 & 12 & 54 & 257 & 230 & 568 \tabularnewline
-Leda & 8 & 24 & 38 & 362 & 208 & 640 \tabularnewline
-Lysithea & 16 & 11 & 35 & 330 & 212 & 604 \tabularnewline
-Pasiphae & 20 & 19 & 44 & 362 & 206 & 651 \tabularnewline
-Sinope & 15 & 21 & 34 & 356 & 256 & 682 \tabularnewline
-\hdashline
-Phoebe & 32 & 98 & 238 & 79 & 13 & 460 \tabularnewline
-%\hdashline
-%Nereid & 11$^{a}$ & 1 & 1 & 0 & 0 & 13 \tabularnewline
-%Triton & -- & 1 & 1 & 0 & 0 & 2 \tabularnewline
-\hline
-\end{tabular}
-\par \end{centering}
-\end{table}
-
-%Table \ref{Tab: sample-cds} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps.
-Since these objects are very small, the duration of each event is a few seconds.
-All the occultation tables and maps will be publicly available at the CDS.
-In Fig.
-\ref{Fig: ocultacao} we show an example of an occultation map.
-This is an occultation by Elara that will happen in February 21, 2017.
-This event can be observed from Australia and it is one of the best opportunities for this object due to the slow velocity of the event and it involves a bright star (MagR*=12.4).
-
-%\begin{table*}
-%\caption{\label{Tab: sample-cds} A sample of stellar occultation predictions for Pasiphae}
-%\begin{center}
-%\begin{tabular}{cccrcccrcrr}
-%\hline
-%\hline
-%d m Year~~~~h~~m~~~s & RA~~~(ICRS)~~~Dec & C/A & P/A & $\nu$ & {\it D} & $R^*$ & $\lambda$ & LST & $\mu_{\alpha*}$ & $\mu_{\delta}$ \tabularnewline
-%\hline
-%% & \multicolumn{2}{c}{Mean errors} & Nr & Nr &  UCAC4       \tabularnewline
-%09 04 2016 03:58:19. & 11 14 36.7707 +07 39 20.7610 & 1.003 &  17.9 & -12.88 &  4.54 & 14.9 & 271. & 22:03 &  12. & -33. \tabularnewline 
-%13 06 2016 00:16:12. & 11 12 48.5020 +07 06 43.3520 & 0.661 &  30.0 & +14.32 &  5.50 & 13.9 & 262. & 17:45 &  -1. &   1. \tabularnewline 
-%27 06 2016 13:56:09. & 11 18 03.4160 +06 23 45.1940 & 1.707 &  28.0 & +20.29 &  5.74 & 11.7 &  44. & 16:53 &   4. & -10. \tabularnewline 
-%18 07 2016 15:07:24. & 11 28 15.5076 +05 05 31.8060 & 0.942 &  26.7 & +27.80 &  6.05 & 14.0 &   8. & 15:40 &   4. &   4. \tabularnewline 
-%22 07 2016 16:15:07. & 11 30 30.4310 +04 48 43.4340 & 0.644 & 206.5 & +29.04 &  6.11 & 14.6 & 348. & 15:27 &  23. & -24. \tabularnewline 
-%24 07 2016 01:37:34. & 11 31 17.8471 +04 42 49.0540 & 0.029 & 206.6 & +29.46 &  6.12 & 15.1 & 206. & 15:22 &   2. &  -8. \tabularnewline 
-%24 07 2016 17:37:18. & 11 31 40.7472 +04 39 57.5060 & 0.840 &  26.5 & +29.66 &  6.13 & 14.9 & 326. & 15:20 & -11. &  -1. \tabularnewline 
-%\hline
-%\end{tabular}
-%\end{center}
-%\begin{flushleft}
-%\textbf{Notes}.
-Entries included: day of the year and UTC time of the prediction; right ascension and declination of the occulted star - at the central instant of occultation (corrected by proper motions); C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees (zero at north of the star, increasing clockwise); $\nu$: velocity in the plane of sky, in km s$^{-1}$: positive = prograde, negative = retrograde; {\it D}: planet range to Earth, in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$ by the relationship $\textrm{R}^* = \textrm{R}_{\textrm{actual}} + 2.5 \times \log 10 \left(\frac{\textrm{velocity}} {20 \textrm{km}\, \textrm{s}^{-1}} \right)$; $\lambda$: east longitude of subplanet point in degrees, positive towards east; LST: UT + $\lambda$: local solar time at subplanet point, hh:mm; $\mu_{\alpha *}$ and $\mu_{\delta}$: proper motions in right ascension and declination, respectively (mas/year).
-%\end{flushleft}
-%\end{table*}
-
-\begin{figure}
-\begin{centering}
-\includegraphics[width = 16.0cm]{figures/Elara_2017-02-21T15:28:40.png}   
-\caption{Occultation map for Elara. % regarding to the first event sampled in Table \ref{Tab: occ-list}.
-The central red dot show the geocentric closest approach of the shadow.
-The small ones shows the center of the shadow separated by 60s.
-The lines show the path of the shadow over the Earth.
-The shadow moves from right to left.
-\textbf{Labels:} Diam: Diameter of the object; Tmax: Maximum duration of the event for a central observation; C/A: the geocentric closest approach, in arcseconds; P/A: the satellite position angle with respect to the occulted star at C/A, in degrees; vel: velocity of event in km/s; Delta: Geocentric distance to the occulting object in AU; $R^*$: normalized magnitude to a common shadow velocity of 20 km s$^{-1}$; long: east longitude of subplanet point in degrees, positive towards east.}
-\label{Fig: ocultacao}
-\end{centering}
-\end{figure}
-
-%The first preliminary catalogue version of the ESA astrometry satellite GAIA \citep{deBruijne2012} is expected to be released up to the end of 2016 (The catalogue with five-parameter astrometric solutions is up to the end of 2017).
-The precise star positions to be derived by GAIA will provide better predictions with the main source of error being the ephemeris.
-Astrometric reduction of observations published in \cite{GomesJunior2015} will be revised with the GAIA catalogue and the predictions will be improved.
-Since we cannot foresee exactly when the GAIA catalogue will be released and when new and re-reduced satellite positions in the GAIA frame will become available, we decided to publish predictions up to the end of 2020.
-In the GAIA era, the occultations predicted will be updated.
-
-\subsection*{Robustness of predictions} \label{Sec: testes}
-
-%Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved.
-In contrast to TNOs, the irregular satellites have much better ephemeris because the orbits of their host planets are better  known, their observational time span is much wider and covers many orbital periods.
-Moreover, the irregular satellites are much closer to Earth which implies in a much smaller shadow path error in kilometers.
-These advantages may be somewhat balanced by the smaller sizes estimated for the irregular satellites.
-Thus, in comparison, the chances for a successful observation of an stellar occultation by an irregular satellite should be considered at least also as good as those by TNOs.
-
-%Observing a stellar occultation demands a great effort.
-And, in our case, the shadow covers a very restricted area on Earth because of the size of the irregular satellites.
-Since no stellar occultation by an irregular satellite was observed up to date, and since we want to be sure that we can start observational campaigns with reasonable chances of success,
-
-To check the availability of observing an occultations we tested the robustness of a prediction for a large target.
-The test design consisted in observing the object and star to be occulted near the date of the event predicted when the two objects were present in the same field of view (FOV), close to each other.
-Thus, the relative positions between the two objects had minimal influence of the errors of the reference catalogue of stars used and possible field distortions %\citep[and references therein]{Peng2008}.
-The relative positions of the star and satellite were used to check the original prediction.
-Notice that in the test we did not attempt to observe any actual occultation.
-The test could be performed at any site, regardless of the Earth location where the occultation would in fact be visible.
-
-We tested the occultation by Himalia predicted to occur on March 3, 2015. The shadow would cross the northern part of South America. For the event, four situations were considered:
-\begin{enumerate}
-\item Our nominal, published prediction with the STE ephemeris, and the nominal UCAC4 position of the star;
-\item Prediction with the JPL ephemeris and the nominal UCAC4 position of the star;
-\item From star and satellite offsets calculated from observations made a few days before the occultation when the objects were very separated (different FOVs);
-\item Same as 3 but with the star and the satellite close in the same FOV.
-\end{enumerate}
-
-Table \ref{Tab: comparison-Himalia} shows the differences between the predictions in the four situations.
-For situation 3 we observed the objects on February 22 with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / pixel$) at the Observatório do Pico dos Dias, Brazil (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m).
-On that day, Himalia and the star were observed in separate FOVs as they were still far apart.
-On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope (diameter = 1.6m; FOV = $5\farcm8$; pixel scale = $0\farcs17 / pixel$) at OPD just over an hour after the time scheduled for the event.
-Satellite and star were separated by about 16 arcsec, so very close to each other (situation 4).
-From the calculated offsets, the center of the shadow was obtained.
-Notice that the shadow path was not predicted to cross the OPD (which was located at almost 2000 km south from the shadow path).
-This was not necessary for testing the prediction.
-
-The critical parameter in the comparisons is the C/A, which here is related to latitudes.
-The apparent radius of Himalia is about 20 mas (see Table 1).
-In the context of the test, for a 0 mas offset in C/A we would have 100\% probability of observing the occultation, and 0\% in the case of a C/A offset equal to or larger than 20 mas, the radius of Himalia.
-From Table 5, we have nearly 0\% probability of success in situation 3, for which the offset in C/A was -20 mas, but when the relative astrometry was poor, 10 days prior to the event.
-Once at the day of the event in situation 4, the C/A offset dropped to -9 mas only, corresponding to a 55\% probability of success.
-Comparison with the prediction using the JPL ephemeris (situation 2) gives a +11 mas C/A offset, or a compatibility of 45\% between the ephemerides.
-All this suggests that there was a good probability of observing the event.
-The largest differences between the shadows of the four situations were 36s in time along the shadow path and 101km (31 mas) in the direction perpendicular to the shadows, suggesting that observers should be spread in narrow latitude ranges 100 km wide.
-
-\begin{table}
-\caption{\label{Tab: comparison-Himalia} Comparison between the predictions of the Himalia occultation at March 03, 2015.}
-\begin{centering}
-\begin{tabular}{lccc}
-\hline  \hline
-\multicolumn{4}{c}{Differences with respect to the STE prediction} \tabularnewline
-Method  & Instant of C/A  & C/A & Sit.   \tabularnewline
-\hline
-STE & 00:39:51 UTC & $0\farcs703$ & i \tabularnewline
-JPL & -26s & +11mas (36km) & ii \tabularnewline % (284km)
-Feb. 22 Obs. & -14s & -20mas (65km) & iii \tabularnewline % (153km)
-Mar. 03 Obs. & -36s & -09mas (29km) & iv \tabularnewline % (393km)
-\hline
-\end{tabular}
-\par\end{centering}
-C/A: geocentric closest approach; Sit: Situation test considered.
-\end{table}
-
-\section*{Discussion} \label{Sec: discussion}
-
-We performed new numerical integration for improving the orbits of some of the larger irregular satellites.
-Consequently, with more precise ephemeris, we, then, predicted stellar occultations aiming to access fundamental parameters like size, shape, albedo, ultimately aiming to track the formation origin of these bodies.
-
-For the irregular satellites of Jupiter (Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae and Sinope), we produced ephemeris using only the observations of \cite{GomesJunior2015}.
-The systematic errors found in the JPL ephemeris \citep{Jacobson2012} and \cite{Emelyanov2008} were corrected.
-This new ephemeris was denominated Special Tailored Ephemeris.
-
-We also updated the ephemeris of Phoebe \citep{Desmars2013} using the observations of \cite{GomesJunior2015}, \cite{Peng2015} observations from  MPC and from Flagstaff.
-A total of 5886 observations between 1989 and 2014 were used in the process.
-This represents and increase of about 75\% in the number used to generate the orbits of Phoebe by \cite{Desmars2013}.
-
-%\textcolor{red}{Colocar mais alguma coisa sobre as efemérides}
-
-We predict stellar occultations for the period of 2016-2020 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; and one satellite of Saturn: Phoebe.
-The procedure used was the same as that for the prediction of stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.The candidate stars were searched in the UCAC4 catalogue.
-The occasional passage of Jupiter by the galactic plane in 2019-2020 and Saturn in 2018 creates the best opportunity of observing stellar occultations in the near future due to the great density of stars in the region.
-Indeed, a total of \noccs events are foreseen.
-
-In a broader, general sense, the probability of successfully observing an occultation is roughly the ratio of the satellite's radius by the budget error (2 sigma for a 95\% confidence level) of ephemeris and star position.
-Thus, UCAC4 errors ranging between 20 mas - 50 mas (1 sigma) combined with a mean error (1 sigma) in the JPL ephemeris of 30 mas for Himalia and 150 mas for Leda published in Table 2 of \cite{Jacobson2012} would give 28\%-17\% probability of observing such an event by Himalia and $\approx2$\% for Leda, the smallest irregular satellite in the sample.
-Observations a few days before the date of occultation predicted may improve the combined errors to 40-80 mas, depending on the magnitude of the objects.
-
-The test made with an occultation expected to happen in March 03, 2015 for Himalia showed that this event would probably have been observed successfully in case there were observers available in the shadow area.
-The results show satisfying small offsets with respect to the local of the prediction.
-
-\cite{GomesJunior2015} also observed Sycorax (satellite of Uranus) and Nereid (satellite of Neptune).
-Sycorax had a few observations on 9 nights in two years which did not cover one orbital period.
-For Nereid, the observations covered many orbital periods, but due to Nereid's large orbital eccentricity there are no observations near the pericenter.
-
-Uranus and Neptune are crossing a very low dense region of stars.
-This results in almost no stellar occultation by these objects up to 2020.
-In fact, using JPL ephemeris, we identified only 2 events for each satellite in this period, but these are events with bad conditions (shadow far from observatories; faint stars).
-%For these reasons we did not attempt to generate new orbits for these satellites here.
-
-Continuous observations of the satellites are recommended and fitting of our dynamical model to those observations are expected to reduce the respective STE ephemeris errors.
-The first version of the GAIA catalogue is to be released up to the end of 2016 and will improve the position error of the stars to the 1-5 mas level.
-Re-reduction of older positions, and reduction of new positions of irregular satellites with GAIA will improve new orbit determinations.
-It will also allow for the discovery of ocultations by more stars not present in the UCAC4 catalogue.
-The release of the GAIA catalogue should have a positive impact on both the astrometric precision of occulted stars and the reduction of new astrometric positions of the satellites.
-As a result, prediction of stellar occultations by irregular satellites shall increase in number as well as in success.
-
-\bibliography{references}
-\bibliographystyle{apalike}
-
-%\bibliography{Referencias}
-%\bibliographystyle{apalike}
+\\N: Number of positions after sigma-clipping procedure. Ncat is the number UCAC4 stars used in the reductions.
+\end{table*}
 

### 2017-08-16 relatorio/OV-2015/Relatorio.tex
mod 2017 ago16

--- 2015-11-11 relatorio_2015/Relatorio.tex
@@
 
-\begin{figure}
+\begin{figure}[h]
 \begin{centering}

### 2017-08-16 Irregular-occ/dados/gaia/tabela.py

@@ -1,0 +1,72 @@
+import numpy as np
+from astropy.coordinates import SkyCoord
+from astropy.time import Time
+import astropy.units as u
+from astropy.table import Table, Column, vstack
+import matplotlib.pyplot as plt
+import matplotlib.image as mpimg
+import csv
+
+f = open('lista_tables', 'r')
+lista = f.readlines()
+f.close()
+
+tabela = Table(names=('object', 'instantes', 'RA (ICRS) DEC', 'C/A', 'P/A', 'vel', 'distance', 'magR', 'magK', 'long', 'caminho'),\
+    dtype=(np.str, np.str, np.str, np.float, np.float, np.float, np.float, np.float, np.float, np.float, np.str))
+
+c = csv.writer(open("agenda.csv", "wb"))
+c.writerow(["Subject","Start Date","Start Time","End Date","End Time","All Day Event","Description","Private"])
+    
+def mostrar(i):
+    imagem = tabela[np.where(tabela['numero'] == i)]
+    imagem.pprint(max_width=-1)
+    img = mpimg.imread(imagem['caminho'][0])
+    plt.imshow(img)
+    plt.show()
+
+for i in lista:
+    arquivo = i.strip()
+    print i.strip()
+    dados = np.loadtxt(arquivo, skiprows=41, usecols=(0, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 18, 19, 20, 21, 22, 25, 26, 28, 29), \
+        dtype={'names': ('dia', 'mes', 'ano', 'hor', 'min', 'sec', 'afh', 'afm', 'afs', 'ded', 'dem', 'des', 'ca', 'pa', 'vel', 'delta', 'mR', 'mK', 'long', 'ora', 'ode'), \
+       'formats': ('S30', 'S30', 'S30','S30', 'S30', 'S30','S20', 'S20', 'S20','S20', 'S20', 'S20', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8', 'f8')}, ndmin=1)
+################## lendo coordenadas #################
+    coor = dados['afh']
+    for i in ['afm', 'afs', 'ded', 'dem', 'des']:
+        coor = np.core.defchararray.add(coor, ' ')
+        coor = np.core.defchararray.add(coor, dados[i])
+    stars = SkyCoord(coor, frame='fk5', unit=(u.hourangle, u.degree))
[…]
